### 2012-05-08 fibrational/fibration-model.tex
Import all the latex files

\documentclass{article}

\usepackage[usenames]{color}
\definecolor{citationcolour}{rgb}{0,0.4,0.2}
\definecolor{linkcolour}{rgb}{0,0,0.8}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
            urlcolor=linkcolour,
            linkcolor=linkcolour,
            citecolor=citationcolour,
%            pdftitle=Relational Parametricity for Algebraically-Indexed Types,
%            pdfauthor={Robert Atkey, Patricia Johann, Neil Ghani},
            pdfkeywords={}}  
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}

\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage[all]{xy}

\newcommand{\cat}[1]{\mathbb{#1}}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\id}{\mathit{id}}
\newcommand{\Id}{\mathit{Id}}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Sub}{\mathrm{Sub}}

\newenvironment{example}{\noindent\textbf{Example}}{\medskip}

% copied from Andrej Bauer
% http://www.tug.org/pipermail/xy-pic/2001-July/000015.html
% (but they were the wrong way round?)
\newcommand{\pushoutcorner}[1][dr]{\save*!/#1+1.2pc/#1:(1,-1)@^{|-}\restore}
\newcommand{\pullbackcorner}[1][dr]{\save*!/#1-1.2pc/#1:(-1,1)@^{|-}\restore}

\title{Fibrational Models of Units of Measure \\ \textbf{Draft}}
\author{Robert Atkey}

\begin{document}

\maketitle

\section{Fibrational Models of Units of Measure Types}
\label{sec:fibrational-models}

A fibrational model of Andrew's units of measure type system consists
of a fibration $p : \cat{E} \to \cat{B}$, with the following
conditions:
\begin{enumerate}
\item The category $\cat{B}$ has finite products. We write $1$ for the
  terminal object, $! : A \to 1$ for the unique map from any object
  $A$ to $1$, $\mathsf{fst}_{AB} : A \times B \to A$ for the left
  projection and $\mathsf{snd}_{AB} : A \times B \to B$ for the right
  projection. Given any two arrows $f : C \to A$ and $g : C \to B$ we
  write their pairing as $\pair{f}{g} : C \to A \times B$. For arrows
  $f : A \to B$ and $g : C \to D$, their product $f \times g : A
  \times C \to B \times D$ is defined as $f \times g = \pair{f \circ
    \mathsf{fst}_{AC}}{g \circ \mathsf{snd}_{AC}}$. We write
  $\alpha_{ABC} : A \times (B \times C) \to (A \times B) \times C$ for
  the canonical associativity maps defined using the pairing and
  projections. We use $\sigma_{AB} : A \times B \to B \times A$ for
  the swapping morphism defined as $\sigma_{AB} =
  \pair{\mathsf{snd}_{AB}}{\mathsf{fst}_{AB}}$.
\item The category $\cat{B}$ has an abelian group object: this
  consists of an object $G$ of $\cat{B}$, along with morphisms $1
  \stackrel{e}\longrightarrow G$, $G \times G
  \stackrel{m}\longrightarrow G$ and $G
  \stackrel{\cdot^{-1}}\longrightarrow G$ making the following
  diagrams commute:

  \begin{tabular*}{1.0\linewidth}{@{\extracolsep{\fill}}rr}
    $\xymatrix{
      {G} \ar[r]^(.4){\pair{e \circ !}{\id}} \ar[dr]_\id & {G \times G} \ar[d]^m & {G} \ar[l]_(.4){\pair{\id}{e \circ !}} \ar[dl]^\id \\
      & {G}
    }$
    &(left and right unit)
  \end{tabular*}

  \begin{tabular*}{1.0\linewidth}{@{\extracolsep{\fill}}rr}
    $\xymatrix{
      {G \times (G \times G)} \ar[r]^\alpha \ar[d]_{\id_G \times m} & {(G \times G) \times G} \ar[d]^{m \times \id_G} \\
      {G \times G} \ar[dr]_m & {G \times G} \ar[d]^m \\
      & {G}
    }$
    &(associativity)
  \end{tabular*}

  \begin{tabular*}{1.0\linewidth}{@{\extracolsep{\fill}}rr}
    $\xymatrix{
      {G \times G} \ar[r]^\sigma \ar[dr]_m & {G \times G} \ar[d]^m \\
      & {G}
    }$
    &(commutativity)
  \end{tabular*}

  \begin{tabular*}{1.0\linewidth}{@{\extracolsep{\fill}}rr}
    $\xymatrix{
      {G} \ar[r]^(.4){\pair{\id}{\cdot^{-1}}} \ar[d]_{!} & {G \times G} \ar[d]^m \\
      {1} \ar[r]^e & {G}
    }$
    &(inverses)
  \end{tabular*}
\item Each fibre of the fibration $p : \cat{E} \to \cat{B}$ is
  cartesian closed and re-indexing preserves the cartesian closed
  structure.
\item For every object $A$ in $\cat{B}$, the re-indexing functor along
  the projection morphism $\mathsf{fst}_{AG} : A \times G \to A$
  (where $G$ is the abelian group object in $\cat{B}$),
  $\mathsf{fst}_{AG}^* : \cat{E}_A \to \cat{E}_{A \times G}$, has a
  right adjoint $\Pi_A : \cat{E}_{A \times G} \to \cat{E}_A$. We write
  $\eta : \Id \Rightarrow \Pi_A \mathsf{fst}^*_{AG}$ for the unit of
  this adjunction and $\epsilon : \mathsf{fst}^*_{AG} \Pi_A
  \Rightarrow \Id$ for the counit.

  This adjunction must satisfy the Beck-Chevalley condition: for any
  morphism $u : A \to A'$ in $\cat{B}$, the natural transformation
  \begin{displaymath}
    \xymatrix{
      {u^* \Pi_{A'}} \ar[r]^(.37){\eta_{u^*\Pi_{A'}}}
      &
      {\Pi_A \mathsf{fst}^*_{AG} u^* \Pi_{A'}} \ar[r]^(.4){\cong}
      &
      {\Pi_A (u \times \id_G)^* \mathsf{fst}^*_{A'G} \Pi_{A'}} \ar[r]^(.6){\Pi_A(u \times \id_G)^*\epsilon}
      &
      {\Pi_A (u \times \id_G)^*}
    }
  \end{displaymath}
  is an isomorphism.

\item The fibre over $G$ contains a distinguished object
  $\mathsf{num}$.
\end{enumerate}

\section{Sketch of the Interpretation of the Syntax}

I'll use Andrew's notation for this section: unit variable contexts
are ranged over by $\mathcal{V}$ and consist of sequences of distinct
unit variable names: $u_1, ..., u_n$, unit expressions are ranged over
by $\mu, \mu_1, \mu_2$ and are built from unit variables (from some
unit variable context), group multiplication, inverses and units.

\paragraph{Caveat} As is normal with fibrational models, the
interpretation of substitution here is only up-to isomorphism, not
equality. The usual way to deal with this is to switch to using a
split fibration.

\medskip

Unit contexts $\mathcal{V} = u_1,...,u_n$ are interpreted as $n$-ary
products of the abelian group object in $\cat{B}$: $\sem{\mathcal{V}}
= G^{|\mathcal{V}|}$, where $|\mathcal{V}|$ is the number of variables
in $\mathcal{V}$.

Unit expressions $\mathcal{V} \vdash \mu$ are interpreted as morphisms
$\sem{\mu} : \sem{\mathcal{V}} \to G$ in $\cat{B}$. Multiplication of
unit expressions is interpreted using the morphism $m : G \times G \to
G$, the unit $1$ is interpreted using the morphism $e : 1 \to G$ and
unit inverses are interpreted using $\cdot^{-1} : G \to G$. The
commuting diagrams required above ensure that if $\mu_1 = \mu_2$ under
the abelian group axioms, then $\sem{\mu_1} = \sem{\mu_2}$.

Well-formed types $\mathcal{V} \vdash A$ are interpreted as objects in
the fibre over $\sem{\mathcal{V}}$: objects $\sem{A}$ in the category
$\cat{E}_{\sem{\mathcal{V}}}$. The types $\texttt{num}\ \mu$, where
$\mathcal{V} \vdash \mu$, are interpreted as $\sem{\mu}^*\mathsf{num}
\in \cat{E}_{\sem{\mathcal{V}}}$. To interpret the $\forall$-types, we
use the $\Pi_A$ functors: given a type $\mathcal{V}, u \vdash A$,
which is interpreted as an object $\sem{A} \in
\cat{E}_{\sem{\mathcal{V},u}}$ then the type $\mathcal{V} \vdash
\forall u. A$ is interpreted as $\Pi_{\sem{\mathcal{V}}}\sem{A} \in
\cat{E}_{\sem{\mathcal{V}}}$. The rest of the type structure is
interpreted using the cartesian closed structure of the fibres.

Simultaneous substitutions $u$ of unit expressions for unit variables
are modelled as morphisms $\sem{u} : \sem{\mathcal{V}} \to
\sem{\mathcal{V'}}$ in $\cat{B}$. Given a type $\mathcal{V'} \vdash A$
and a simultaneous substitution $u$ of unit expressions well-formed in
$\mathcal{V}$ for all unit variables in $\mathcal{V'}$, then it is the
case that $\sem{u}^*\sem{A} \cong \sem{u^*A}$, by the preservation of
cartesian closed structure by re-indexing, and the Beck-Chevalley
condition.

For a unit variable context $\mathcal{V}$, well-formed typing contexts
$\mathcal{V} \vdash \Gamma$ consist of lists of variable name/type
pairs $x_i : A_i$, where $\mathcal{V} \vdash A_i$, for all $i$. A
typing context $\mathcal{V} \vdash \Gamma$ is interpreted as the
object $\sem{A_1} \times ... \times \sem{A_n}$ in the fibre
$\cat{E}_{\sem{\mathcal{V}}}$, using the finite products of the fibre.

A term $\mathcal{V}; \Gamma \vdash M : A$ is interpreted as an arrow
$\sem{\Gamma} \to \sem{A}$ in the category
$\cat{E}_{\sem{\mathcal{V}}}$. For example, $\forall$-introduction is
interpreted as follows: given $\mathcal{V}, u; \Gamma \vdash e : A$,
where $u \not\in \mathit{fv}(\Gamma)$, then $e$ has an interpetation
$\sem{e} : \mathsf{fst}^*_{\sem{\mathcal{V}}G}\sem{\Gamma} \to
\sem{A}$ in $\cat{E}_{\sem{\mathcal{V},u}}$. The interpretation of
$\mathcal{V}; \Gamma \vdash \Lambda u. e : \forall u. A$ is then given
by $\phi_{\sem{\Gamma}\sem{A}}(\sem{e}) : \sem{\Gamma} \to \Pi_{\sem{\mathcal{V}}}\sem{A}$
in $\cat{E}_{\sem{\mathcal{V}}}$, where $\phi_{XY} :
\cat{E}_{\sem{\mathcal{V},u}}(\mathsf{fst}^*_{\sem{\mathcal{V}}G}X,Y)
\to \cat{E}_{\sem{\mathcal{V}}}(X,\Pi_{\sem{\mathcal{V}}}Y)$ is one
half of the isomorphism of homsets derived from the adjunction
$\mathsf{fst}^*_{\sem{\mathcal{V}}G} \dashv \Pi_{\sem{\mathcal{V}}}$.

\section{A Class of Logical Relation Models}

I'll now show that there exists a large class of examples of
fibrations satisfying the conditions in
\autoref{sec:fibrational-models}, built from fibrations describing a
logic, and some additional data which I describe below. All of these
examples will share the same base category $\cat{B}$ which is defined
as follows:
\begin{itemize}
\item Objects of $\cat{B}$ are natural numbers;
\item The set of morphisms $n \to 1$ in $\cat{B}$ is the set of terms
  in the theory of abelian groups over $n$ free variables, quotiented
  by the equational theory of abelian groups. Morphisms $n \to n'$
  consist of $n'$-many morphisms $n \to 1$.
\end{itemize}
Note that $\cat{B}$ has finite products: the terminal object is $0$,
and the product of $n$ and $n'$ is $n + n'$. The object $1$ is an
abelian group object in this category. More abstractly, $\cat{B}$ is
the Lawvere theory describing the theory of abelian groups.

To build a model of the units of measure system, we require:
\begin{enumerate}
\item A fibration $q : \cat{P} \to \cat{C}$, with right adjoints to
  all re-indexing functors (satisfying Beck-Chevalley), and such that
  all fibres are cartesian closed and re-indexing preserves the
  cartesian closed structure. The category $\cat{C}$ must have finite
  products.
\item A functor $R : \cat{B} \to \cat{C}$ satisfying the following
  property: for any arrow $u : m \to n$ in $\cat{B}$, the following
  diagram is a pullback:
  \begin{displaymath}
    \xymatrix{
      {R(m + 1)} \ar[r]^{\mathsf{fst}} \ar[d]_{R(u + 1)} \pullbackcorner & {R(m)} \ar[d]^{R(u)} \\
      {R(n + 1)} \ar[r]^{\mathsf{fst}} & {R(n)}
    }
  \end{displaymath}
  The functor $R$ is used to interpret unit expressions as objects
  that can be quantified over in the internal logic of $q$. The
  pullback condition is used to ensure that the Beck-Chevalley
  condition required in \autoref{sec:fibrational-models} holds.
\item A chosen object $N$ of $\cat{C}$ and a chosen object $P_N$ of
  $\cat{P}_{R(1) \times N \times N}$. This will be used for the
  $\mathsf{num}$ object.
\end{enumerate}

\begin{example}
  Let the fibration $q$ be the sub-object fibration over sets, $q :
  \Sub(\Set) \to \Set$, where objects of $\Sub(\Set)$ are pairs of
  sets $A$ and a predicate $P$ on $A$. This has right-adjoints to
  re-indexing and fibred cartesian closed structure.

  Define the functor $R : \cat{B} \to \Set$ on objects as $R(n) =
  (\mathbb{Q}^{>0})^n$, where $\mathbb{Q}^{>0}$ is the set of strictly
  positive rational numbers. Morphisms $u : n \to 1$ are exactly terms
  in the theory of abelian groups, so we interpret them as functions
  $R(u) : (\mathbb{Q}^{>0})^n \to \mathbb{Q}^{>0}$ using the
  multiplicative abelian group structure of $\mathbb{Q}^{>0}$. It can
  be checked that the pullback property required of $R$ holds.

  Choose $N$ to be the set $\mathbb{Q}$, and $P(k \in \mathbb{Q}^{>0},
  q, q') \Leftrightarrow q' = kq$.
\end{example}

Given the data above, we can now generate an instance of the structure
in \autoref{sec:fibrational-models}. We first generate a new fibration
via a pullback:
\begin{displaymath}
  \xymatrix{
    {\cat{E}} \ar[d]_r \ar[rr] \pullbackcorner
    &
    &
    {\cat{P}} \ar[d]^q
    \\
    {\cat{B} \times \cat{C}} \ar[r]^(.4){R \times \Delta}
    &
    {\cat{C} \times (\cat{C} \times \cat{C})} \ar[r]^(.7)\times
    &
    {\cat{C}}
  }
\end{displaymath}
In this diagram, the functor $\Delta : \cat{C} \to \cat{C} \times
\cat{C}$ duplicates, and the functor $\times : \cat{C} \times (\cat{C}
\times \cat{C}) \to \cat{C}$ uses the finite product structure of
$\cat{C}$ to turn triples $(A,(B,C))$ of objects of $\cat{C}$ into
objects $A \times B \times C$ of $\cat{C}$.

Concretely, the objects of $\cat{E}$ are triple $(n, A, P)$, where $n$
is an object of $\cat{B}$ (i.e. a natural number), $A$ is an object of
$\cat{C}$ and $P$ is an object of the fibre $\cat{P}_{R(n) \times A
  \times A}$.

Set $p = \mathsf{fst} \circ r : \cat{E} \to \cat{B}$. We know $p$ to
be a fibration because $r$ is a fibration by construction as a
pullback, $\mathsf{fst}$ is trivially a fibration, and the composition
of fibrations always gives fibrations.

We now show that $p : \cat{E} \to \cat{B}$ has all the structure we
required in \autoref{sec:fibrational-models}. We already know that
$\cat{B}$ has finite products and an abelian group object, so it
remains to show that $p$ is fibred cartesian closed, and that the
required right adjoints to re-indexing along projections $\mathsf{fst}
: n + 1 \to n$ exist.

{\huge Not finished yet...}

\end{document}

### 2012-05-08 geometry-notes/invariance.tex
Import all the latex files

\documentclass{article}

\usepackage[usenames]{color}
\definecolor{citationcolour}{rgb}{0,0.4,0.2}
\definecolor{linkcolour}{rgb}{0,0,0.8}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
            urlcolor=linkcolour,
            linkcolor=linkcolour,
            citecolor=citationcolour,
            pdftitle=Geometric Invariance and Relational Parametricity,
            pdfauthor={Robert Atkey},
            pdfkeywords={}}
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}

\usepackage{amssymb}
\usepackage{stmaryrd}

\newcommand{\sepbar}{\mathrel|}

\newcommand{\GA}{\mathrm{GA}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Is}{\mathrm{Is}}
\newcommand{\Orth}{\mathrm{O}}
\newcommand{\SOrth}{\mathrm{SO}}
\newcommand{\Transl}{\mathrm{T}}
\newcommand{\Scale}{\mathrm{Scale}}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\isDefinedAs}{\stackrel{\mathit{def}}=}

\newcommand{\statementref}[1]{\hyperref[#1]{Statement \ref*{#1}}}

\newenvironment{example}{\vspace{0.1in}\noindent\textbf{Example}\quad}{}

\newcommand{\ignore}[1]{}

\title{Geometric Invariance and Relational Parametricity}
\author{Robert Atkey}

\begin{document}

\maketitle

\section{Coordinate-free Geometry and Invariance}

When writing programs that manipulate $n$-dimensional geometric data,
the basic data structure is the $n$-tuple of numbers. In the
two-dimensional case, tuples $\vec{v} = (x,y)$ can be called upon to
represent both points -- as offsets from some origin -- and vectors --
as offsets in their own right. Geometrically, the two notions of
\emph{point} and \emph{vector} are very different despite their common
representation. By representing a point concretely as an offset from
some privileged origin, we inadvertently fix a particular origin,
$(0,0)$. However, as made clear by Gallieo, many of the laws of
physics are invariant under change of the frame of reference: if we
change where we think the origin is, then we do not change the
fundamental laws. Putting this in concrete terms, if we have two
frames of reference, one with $(0,0)$ as the origin, and one with
$(10,20)$ as the origin, then the tuple $(1,1)$ with respect to the
first frame and $(11,21)$ with respect to the second frame represent
the \emph{same} point. It is merely an artifact of our representation
of points as pairs of numbers that these appear to be
different. Therefore, when we write programs that manipulate data that
represents points, then we ought to follow the general principle that
our programs are invariant with respect to changes in the choice of
origin.

Vectors are not intended to be regarded as invariant under change of
choice of origin. Vectors represent offsets, or movements, and the
vector $(0,0)$ always represents the zero offset. However, they are
intended to be invariant under change of basis: the vector $(2,3)$,
with respect to the basis $(1,0),(0,1)$, and the vector $(1,1)$, with
respect to the basis $(2,0),(0,3)$, should be regarded as the same
vector. Invariance under change of basis also applies to the
representation of points. We therefore refine the principle of
invariance under change of origin for points to also include invariant
under change of basis for points and vectors. A particular choice of
origin and basis is called a \emph{frame} (\cite{gallier11geometric},
Chapter 2). In general, the technique of maintaining geometric
invariants such as frame invariance has been referred to as
\emph{coordinate-free geometry}: we should aim to think in terms of
abstract concepts such as points and vectors, rather than their actual
representation as tuples of numbers.

Invariance under change of representation immediately recalls
Reynolds' famous story of two professors teaching the theory of
complex numbers to students \cite{reynolds83types}. One professor
chooses rectangular coordinates ($x + iy$), while the other chooses
polar coordinates ($\alpha\cos\theta + i\alpha\sin\theta$). Happily,
the two classes can interact because the theory of complex numbers is
invariant with respect to the choice of representation. Reynolds
formalises the idea of invariance with respect to changes of
representation by using the notion of preservation of relations. If a
binary relation $R$ relates two representations of a concept---for
example the rectangular and polar representations of the same complex
number are related---then a program that manipulates complex numbers
at a level of abstraction above their specific representation should
preserve the relation $R$. Concretely, let $f$ be a program taking
complex numbers to complex numbers that is intended to respect
invariance under change of representation. Then if $c$ is a complex
number in rectangular form, and $c'$ is a complex number in polar form
such that $R$ relates them, which we write as $(c,c') \in R$, then it
should be the case that $(f(c), f(c')) \in R$.

We now apply Reynolds' relational approach to defining invariance
under changes of representation to the geometric setting. Following
Reynolds, we express the fact that two tuples in $\mathbb{R}^2$ are
representatives of the same point or vector by defining binary
relations over $\mathbb{R}^2$. We wish to ensure that invariance with
respect to all choices of frame is respected, so we parameterise our
relations by the particular choice made. A neat way to represent a
choice of frame is as an invertible affine map from $\mathbb{R}^2$ to
$\mathbb{R}^2$. Affine maps are of the form $F(\vec{v})^\top =
A\vec{v}^\top + b$, where $A$ is an invertible real $2 \mathord\times
2$-matrix and $b$ is a vector in $\mathbb{R}^2$. The matrix $A$
represents the change of basis, and $b$ represents the change of
origin. The map $F$ tells us how to translate the representative of a
point in our chosen frame to the representative of the same point in
the ``standard frame'' with origin represented by $(0,0)$ and basis
$(1,0),(0,1)$. The set of all such maps forms the general affine group
$\GA(2)$.

We define a $\GA(2)$-parameterised family of binary relations $\{
R_{\texttt{pt}}(F) \subseteq \mathbb{R}^2 \times \mathbb{R}^2 \}_{F
  \in \GA(2)}$. We intend that two tuples $\vec{v_1} = (x_1,y_1)$ and
$\vec{v_2} = (x_2,y_2)$ are related for $F \in \GA(2)$, if, when the
former is interpreted with respect to the standard frame and the
latter is interpreted with respect to the frame described by $F$, the
two tuples represent the same point.
\begin{displaymath}
  R_{\texttt{pt}}(F) = \{ (\vec{v_1}, \vec{v_2}) \sepbar F\vec{v_2} = \vec{v_1} \}
\end{displaymath}

Now consider a function $f$ that, for example, takes two tuples in
$\mathbb{R}^2$ and returns a single tuple in $\mathbb{R}^2$. We intend
that all these tuples represent points in the same frame, but that the
function itself should be invariant with respect to the choice of
frame. We use Reynolds' idea of preservation of relations to formalise
this: for any $F \in \GA(2)$ (i.e.~any choice of frame), the function
$f$ should satisfy the following statement:
\begin{equation}\label{eq:f-preserve-rel-frame}
  \forall (\vec{v_1},\vec{v'_1}) \in R_{\texttt{pt}}(F), (\vec{v_2},\vec{v'_2}) \in R_{\texttt{pt}}(F).\ (f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\texttt{pt}}(F).
\end{equation}
Unfolding the definition of $R_{\texttt{pt}}$, we can see that this
statement is equivalent to the following statement expressing
invariance of $f$ with respect to changes of frame: for all $F \in
\GA(2)$,
\begin{equation}\label{eq:f-invariant-frame}
  \forall \vec{v_1}, \vec{v_2}.\ f(F\vec{v_1},F\vec{v_2}) = F(f(\vec{v_1},\vec{v_2})).
\end{equation}
So Reynolds' idea of preservation of relations, when instantiated with
the family of relations $R_{\texttt{pt}}$, yields exactly the
geometric property of invariance under change of frame.

An example function $f$ that satisfies
\statementref{eq:f-invariant-frame} is the following function that
computes a particular affine combination of two points by working
directly on their coordinate representation:
\begin{displaymath}
  f(\vec{v_1}, \vec{v_2}) = \frac{1}{2}\vec{v_1} + \frac{1}{2}\vec{v_2}.
\end{displaymath}
In general, affine combinations of points $\lambda_1\vec{v_1} +
\lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$, satisfy
\statementref{eq:f-invariant-frame}. Affine combination is one of the
fundamental building blocks of affine geometry -- the properties of
points invariant under invertible affine maps
(\cite{gallier11geometric}, Chapter 2). If we drop the condition that
$\lambda_1 + \lambda_2 = 1$, then we are dealing with linear
combinations of vectors, and we are no longer invariant with respect
to changes of frame. However, linear combinations are invariant with
respect to change of basis. We can represent changes of basis as
linear invertible maps $B : \mathbb{R}^2 \to \mathbb{R}^2$. The set of
all such maps forms the general linear group $\GL(2)$. We now define
another family of relations $\{R_{\texttt{vec}}(B) \subseteq
\mathbb{R}^2 \times \mathbb{R}^2 \}_{B \in \GL(2)}$ that relates two
points up to change of basis:
\begin{displaymath}
  R_{\texttt{vec}}(B) = \{ (\vec{v_1},\vec{v_2}) \sepbar B\vec{v_2} = \vec{v_1} \}.
\end{displaymath}
Now the functions $f_{\lambda_1\lambda_2}(\vec{v_1},\vec{v_2}) =
\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$, for arbitrary $\lambda_1$
and $\lambda_2$, do preserve the relations $R_{\texttt{vec}}(B)$, for
all $B \in \GL(2)$. Unfolding the definition of $R_{\texttt{vec}}(B)$
in the analogous statement to \statementref{eq:f-preserve-rel-frame}
for $R_{\texttt{vec}}$ instead of $R_{\texttt{pt}}$, we can see that
preservation of the relations $R_{\texttt{vec}}$ characterises the
functions that are invariant under change of basis. For all $B \in \GL(2)$, we have
\begin{displaymath}
  \forall \vec{v_1}, \vec{v_2}.\ f(B\vec{v_1},B\vec{v_2}) = B(f(\vec{v_1},\vec{v_2})),
\end{displaymath}
and this is exactly the property of invariance under change of basis
we required above of programs manipulating vectors.

Note that the family of relations $R_{\texttt{vec}}$ is just
$R_{\texttt{pt}}$ when restricted to elements of the group
$\GL(2)$. In the type system we introduce in the next section, we
combine points and vectors into the same data type. Whether it
represents a point or a vector depends on the group of geometric
transformations that we expect it to be invariant under.

\section{A Type System for Geometric Invariance}
\label{sec:type-system-geom-intro}

In the last section, we saw how Reynolds' notion of preservation of
relations can be used to define when it is the case that programs
manipulating representatives of points and vectors make geometric
sense. Reynolds further insight was to see how preservation of
relations can automatically be established by following a typing
discipline. In Reynolds' case, the typing discipline was the
second-order polymorphic $\lambda$-calculus, also known as System F,
where new types can be constructed by universal quantification over
all types. Universal type quantification is commonly written $\forall
\alpha. A$, which denotes a type that may be instantiated with any
other type $B$ by replacing the type variable $\alpha$ with $B$ in
$A$.

In terms of relations, Reynolds interprets universal quantification
over types as quantification over binary relations between denotations
of types. In the statements of geometric invariance that we stated in
the previous section (e.g., \statementref{eq:f-invariant-frame}), we
did not quantify over all relations. Instead, we quantified over all
elements of a geometric group such as $\GA(2)$ or $\GL(2)$, and used
those elements to select relations from the families $R_{\texttt{pt}}$
and $R_{\texttt{vec}}$. Driven by this observation, we introduce
quantification over elements of geometric groups into the language of
types. We use the notation $\forall X \mathord: G. A$ for
quantification over elements of geometric groups, where $G$ is the
name of some geometric group, and $X$ is a variable name standing for
an element of $G$. We refer to $G$ as the \emph{sort} of $X$. As we
proceed, we will introduce the sorts representing geometric groups
that we will be using in addition to $\GA(2)$ and $\GL(2)$.

The sorts $\GA(2)$ and $\GL(2)$ represent groups, so we can combine
their elements using the usual group operations. We write operations
multiplicatively, using juxtaposition ($e_1e_2$) for multiplication,
$e^{-1}$ for inverse and $1$ for the unit. We overload the same
notation to stand for the group operations in both $\GA(2)$ and
$\GL(2)$. We also regard expressions built from variables and the
group operations up to the group axioms. So, for example,
$F_1(F_2F_3)$ and $(F_1F_2)F_3$ are to be regarded as equivalent.

Our language of types also includes the common type constructors for
function types, $A \to B$, sum types $A + B$, unit type
$\mathsf{Unit}$ and tuple types $A \times B$. We also assume a
primitive type $\mathsf{real}$, used to represent scalars. As we noted
at the start of the previous section, the central data structure in
geometric applications is the tuple of real numbers for representing
points and vectors. However, we cannot simply express this as the type
$\mathsf{real} \times \mathsf{real}$ because it will not have the
correct relational interpretation. (Two elements of type
$\mathsf{real}$ will be related if and only if they are equal, and
hence two elements of $\mathsf{real} \times \mathsf{real}$ will be
related if and only if they are equal, by Reynolds' definition of the
relational interpretation of tuple types.) Therefore, we introduce a
new type parameterised by expressions $e$ of sort $\GA(2)$ to
represent points (and later, vectors): $\mathsf{vec}\langle e
\rangle$. We have used the name $\mathsf{vec}$, even though we have
taken pains to separate geometric points and vectors, to recall the
computer science notion of vector as a sequence of values of
homogeneous type which has a known length (in this case two).

% FIXME: Denotational semantics of $\mathsf{vec}$? and transformation
% quantification?

Equipped with quantification over the group $\GA(2)$ of invertible
affine maps, and the parameterised types $\mathsf{vec}\langle \cdot
\rangle$, we can now express geometric invariance properties as
types. In the previous section, we gave examples of functions $f :
\mathbb{R}^2 \times \mathbb{R}^2 \to \mathbb{R}^2$ that have the
property that they preserve all invertible affine maps. We can now
express this property as a type:
\begin{displaymath}
  \mathrm{f} : \forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \times \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle
\end{displaymath}
We use the variable name $F$ for elements of $\GA(2)$ to suggest the
intuition that this typing indicates that $\mathrm{f}$ takes a pair of
points in the $F$rame $F$ and returns a point in the same frame.

\subsection{Affine Combination of Points}
\label{sec:affine-combination}

As we stated in the previous section, the primitive operation of
affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
\lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
affine combination can be interpreted as interpolation between the
points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
combination of points to our calculus, with the following typing and
intended denotation.
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affineCombination} : \forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \\
    \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ t\ (x_2,y_2) = ((1-t)x_1 + tx_2, (1-t)y_1 + ty_2)
  \end{array}
\end{displaymath}
By defining the primitive function $\mathrm{affineCombination}$ to
take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
we are taking the affine combination of two representatives of points,
and not just an arbitrary linear combination. Note that the intended
denotation in the second line does not take any argument corresponding
to the universally quantified $F$ in the type. In our calculus,
geometric group expressions are erased at runtime.

% In Linear Algebra notation, we could have written the right-hand side
% as $(1-t)\vec{v_1} + t\vec{v_2}$, where $\vec{v_1} = (x_1,y_1)$ and
% $\vec{v_2} = (x_2,y_2)$.

\begin{example}
  With just the single primitive operation of affine combination, we
  can already express some useful programs. For example, the
  evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves with
  two endpoints and a single control point) can be expressed using the
  affine combination primitive and three steps of De Casteljau's
  algorithm:
  \begin{displaymath}
    \begin{array}{l}
      \mathrm{quadraticB\acute{e}zier} : \forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \\
      \mathrm{quadraticB\acute{e}zier}\ [F]\ p_0\ p_1\ p_2\ t = \\
      \quad
      \begin{array}{l@{\hspace{-0.002cm}}l}
        \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ t\ p_1) \\
        & t \\
        & (\mathrm{affineCombination}\ [F]\ p_1\ t\ p_2)
      \end{array}
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$ and a control point $p_1$,
  $\mathrm{quadraticB\acute{e}zier}\ p_0\ p_1\ p_2\ t$ for $t \in
  [0,1]$ gives us the point on the curve at ``time'' $t$. Note that
  the type of $\mathrm{quadraticB\acute{e}zier}$ tells us immediately
  that it preserves all invertible affine maps.
  %FIXME: forward ref to ``free theorem''.
\end{example}

\subsection{Vector Space Operations}
\label{sec:vector-space-ops}

Affine combination only provides us with an operation on points. For
the purposes of computational geometry, we are also interested in
operations on the vectors representing offsets between points. We now
examine the correct types to assign to the vector space operations of
addition of vectors, negation of vectors, multiplication by a scalar
and the zero vector.

In our discussion in the previous section, we observed that vector
operations are those that are preserved by change-of-basis operations
-- i.e.~under the action of elements of the group $\GL(2)$ of
invertible linear maps. Since $\GL(2)$ is a subgroup of $\GA(2)$ we
may form the type $\mathsf{vec}\langle e \rangle$ when $e$ is of sort
$\GL(2)$ as well as $\GA(2)$. The obvious type assignment for vector
addition is now the following:
\begin{displaymath}
  (+) : \forall B\mathord:\GL(2).\ \mathsf{vec}\langle B \rangle \to \mathsf{vec}\langle B \rangle \to \mathsf{vec}\langle B \rangle
\end{displaymath}
(we write binary operators intended to be used infix in parentheses
when not appearing in infix position, following the Haskell
syntax.) The use of $B$ for names of elements of $\GL(2)$ is meant to
suggest that the type $\mathsf{vec}\langle B \rangle$ denotes vectors
with respect to the $B$asis $B$.

However, it is possible to give vector addition a more precise type
that describes its effect on the chosen origin as well as its
invariance under change of basis. We now introduce a new sort
$\Transl(2)$ of translations. Translations form an abelian group,
which we write additively, using $+$,$-$ and $0$, and we use lower
case $t$, $t_1$, etc.~for $t$ranslation variables. Invertible affine
maps are constructed from invertible linear maps and translations, so
given a $B : \GL(2)$ and a $t : \Transl(2)$, we can form an invertible
affine map $B \ltimes t : \GA(2)$. Conversely, given $F : \GA(2)$, we
have an invertible linear map $\pi_L(F) : \GL(2)$, and a translation
$\pi_T(F) : \Transl(2)$. Thinking of elements of $\GA(2)$ as frames,
$\pi_L$ returns the basis of the frame, and $\pi_T$ returns the
origin.

We can now give vector addition a more general type, which
demonstrates how adding vectors has an effect on translations:
\begin{displaymath}
  (+) : \forall B \mathord: \GL(2), t_1, t_2 \mathord: T(2).\ \mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2 \rangle \to \mathsf{vec}\langle B \ltimes (t_1 + t_2)\rangle
\end{displaymath}
We can also negate vectors, yielding a vector which points in the
opposite direction. Negation is invariant under invertible linear maps
but negates translations:
\begin{displaymath}
  \mathrm{negate} : \forall B \mathord: \GL(2), t \mathord: T(2).\ \mathsf{vec}\langle B \ltimes t\rangle \to \mathsf{vec}\langle B\ltimes(-t)\rangle
\end{displaymath}
With addition and negation of vectors, we can define the derived
operation of vector subtraction with the following type:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall B \mathord:\GL(2), t_1,t_2 \mathord:T(2).\ \mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2\rangle \to \mathsf{vec}\langle B \ltimes (t_1 - t_2)\rangle \\
    (-)\ [B]\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
\end{array}
\end{displaymath}

Given two points in the same frame---i.e.~two values of type
$\mathsf{vec}\langle F \rangle$---we can compute their offset, which
is a vector expressed with respect to the basis part of $F$, as a
special case of vector subtraction:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle \pi_L(F) \ltimes 0 \rangle \\
    \mathrm{offset}\ [F]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
The subtraction operation in the body of $\mathrm{offset}$ is
instantiated with $B = \pi_L(F)$, $t_1 = \pi_T(F)$ and $t_2 =
\pi_T(F)$. The translation $\pi_T(F)$ cancels itself out, and we are
left with a vector in the basis $\pi_L(F)$ of the frame $F$.

Similarly, the type of the vector addition operation can be
specialised to the case of moving a point by an offset represented by
a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle \pi_L(F) \ltimes 0 \rangle \to \mathsf{vec}\langle F \rangle \\
    \mathrm{moveBy}\ [F]\ p\ v = p + v
  \end{array}
\end{displaymath}
The addition operation in the body of $\mathrm{moveBy}$ is
instantiated with $B = \pi_L(F)$, $t_1 = \pi_T(F)$ and $t_2 =
0$. Since $t_1 + t_2 = \pi_T(F) + 0 = \pi_T(F)$ in this case, we are
left with a value of type $\mathsf{vec}\langle \pi_L(F) \ltimes
\pi_T(F) \rangle$, which is equivalent to the type
$\mathsf{vec}\langle F \rangle$.

The types we assign to the remaining vector space primitives, the zero
vector and multiplication by a scalar, do not describe any interesting
effect on translations%\footnote{but they could...}
. The zero vector has the following polymorphic type, indicating that
for any basis $B$, there is a zero vector.
\begin{displaymath}
  \mathrm{0} : \forall B \mathord: \GL(2).\ \mathsf{vec}\langle B \ltimes 0 \rangle
\end{displaymath}
Intuitively, there cannot be a zero vector for any frame $F$, because
this would entail picking a privileged origin. For multiplication of a
vector by a scalar, we assign the following type:
\begin{displaymath}
  (*) : \forall B \mathord: \GL(2).\ \mathsf{real} \to \mathsf{vec}\langle B \ltimes 0 \rangle \to \mathsf{vec}\langle B \ltimes 0 \rangle
\end{displaymath}
This type indicates that scaling a vector is preserved by change of
basis.

\begin{example}
  The vector space primitives we have defined in this section, along
  with the geometric invariance properties that follow from their
  types, allow us to establish a useful type isomorphism. Consider
  functions with types conforming to the following schema:
  \begin{displaymath}
    \tau_{n,A} \isDefinedAs \forall F\mathord:\GA(2).\ \mathsf{vec}\langle F \rangle^{n+1} \to A[F]
  \end{displaymath}
  We take $n$ to be a natural number, and $A^n$ for some type $A$
  denotes the $n$-ary tupling of $A$ with itself: i.e.~$A \times
  ... \times A$, with $A$ repeated $n$ times. The notation $A[-]$
  denotes a type with a ``hole'' that can be filled with an
  expression. In this case, we require that $A[e]$ be a well-formed
  type when $e$ has sort $\GA(2)$.

  Just by looking at the types $\tau_{n,A}$, we know that their
  inhabitants will be invariant under change of frame, due to the
  quantification over all $F$ in $\GA(2)$. Therefore, we may as well
  pick one of the input points as the origin and assume that all the
  other points are defined with respect to this chosen origin. This is
  similar to the common mathematical step of stating that, ``without
  loss of generality'', we may pick some point in a description of a
  problem to be the origin, as long as the problem statement is
  invariant under translation.

  The types $\tau_{n,A}$ are isomorphic to the corresponding types
  $\sigma_{n,A}$, defined below. Inhabitants of the types
  $\sigma_{n,A}$ take one fewer points (now seen as vector offsets
  from some origin), but are invariant only under change of basis.
  \begin{displaymath}
    \sigma_{n,A} \isDefinedAs \forall B\mathord:\GL(2).\ \mathsf{vec}\langle B \rangle^n \to A[B \ltimes 0]
  \end{displaymath}
  The isomorphism between these two families of types can be witnessed
  by the following two functions.
  \begin{displaymath}
    \begin{array}{l}
      i_{n,A} : \tau_{n,A} \to \sigma_{n,A} \\
      i_{n,A}\ f\ [B]\ (p_1, p_2, ..., p_n) = f\ [B \ltimes 0]\ (0, p_1, ..., p_n) \\
      \\
      i^{-1}_{n,A} : \sigma_{n,A} \to \tau_{n,A} \\
      i^{-1}_{n,A}\ g\ [F]\ (p_0, ..., p_n) = g\ [\pi_L(F)]\ (p_1-p_0, ..., p_n-p_0)
    \end{array}
  \end{displaymath}
  The direction defined by the function $i_{n,A}$ treats the inputs
  $p_1,...,p_n$ as vector offsets from the origin $0$. The direction
  witnessed by $i^{-1}_{n,A}$ picks the first point to act as the
  origin, and uses the operation of vector subtraction we defined
  above to turn each of the other points into offsets from this
  point. Note that this isomorphism is not unique: for each $n$ we can
  pick any of the $n+1$ inputs to act as the distinguished
  origin. Thus the families of types $\tau_{n,A}$ and $\sigma_{n,A}$
  are isomorphic, but not uniquely isomorphic.
\end{example}

\subsection{Euclidean Geometry}
\label{sec:euclidean-geom}

\subsubsection{Dot Product}

Euclidean geometry extends affine geometry with the operation of
\emph{dot product} (also known as the \emph{inner product}) of two
vectors, which produces a real number. The standard Euclidean dot
product is defined on the coordinate representation in the following
way:
\begin{displaymath}
  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2
\end{displaymath}
The dot product is used to define properties of vectors such as their
\emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
of the length of the vector $\vec{v}$, and the notion of
orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
if $\vec{v_1}\cdot\vec{v_2} = 0$.

To incorporate the operation of dot product into our calculus, we must
give it a type. It is not the case that the dot product is invariant
under the action of any of the groups $\GA(2)$, $\GL(2)$ or
$\Transl(2)$ that we have considered so far. In fact, the dot product
is invariant under the subgroup $\Orth(2)$ of $\GL(2)$ of
\emph{orthogonal} linear transformations. The group $\Orth(2)$
consists of invertible linear maps whose determinant is either $1$ or
$-1$. We introduce orthogonal transformations as a new sort
$\Orth(2)$, and overload the multiplicative group operations syntax
for inhabitants of $\Orth(2)$. We also assume a function $\iota_O$
that takes $e : \Orth(2)$ to $\iota_O(e) : \GL(2)$.

We may now assign the following type to the dot product primitive:
\begin{displaymath}
  (\cdot) : \forall O \mathord: \Orth(2).\ \mathsf{vec}\langle \iota_O(O) \ltimes 0 \rangle \to \mathsf{vec}\langle \iota_O(O) \ltimes 0 \rangle \to \mathsf{real}
\end{displaymath}
With dot product as a primitive operation, we can define several other
useful derived operations. First, the norm of a vector, which in the
case of standard Euclidean geometry is the square of the length of the
vector. The norm is invariant under orthogonal changes of basis, but
not arbitrary changes of basis. Hence, we quantify over all orthogonal
transformations in the type of $\mathrm{norm}$:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{norm} : \forall O \mathord: \Orth(2).\ \mathsf{vec}\langle O \rangle \to \mathsf{real} \\
    \mathrm{norm}\ [O]\ v = v \cdot v
  \end{array}
\end{displaymath}
Given the norm function, we can compute the distance between a pair of
points in a frame, provided that the frame has an orthogonal basis
described by some orthogonal transformation $O$:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{distance} : \forall O \mathord: \Orth(2), t \mathord: \Transl(2).\ \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{real} \\
    \mathrm{distance}\ [O]\ [t]\ p_1\ p_2 = \mathrm{sqrt}\ (\mathrm{norm}\ (\mathrm{offset}\ p_1\ p_2))
  \end{array}
\end{displaymath}
In the definition of $\mathrm{distance}$, we have made use of a square
root function on the reals $\mathrm{sqrt} : \mathsf{real} \to
\mathsf{real}$.

The dot product is also used to determine whether two vectors are
orthogonal to one another. In the case of the standard dot product in
$2$-dimensional space, two vectors are orthogonal if they are at right
angles to one another.
\begin{displaymath}
  \begin{array}{l}
    \mathrm{orthogonal} : \forall O \mathord: \Orth(2).\ \mathsf{vec}\langle O \rangle \to \mathsf{vec}\langle O \rangle \to \mathsf{bool} \\
    \mathrm{orthogonal}\ [O]\ v_1\ v_2 = v_1 \cdot v_2 \equiv 0
  \end{array}
\end{displaymath}

\subsubsection{Cross Product}

\begin{figure}[t]
  \centering
  \begin{eqnarray*}
    0   &:& \forall S \mathord:\Scale.\ \mathsf{real}\langle S \rangle \\
    1   &:& \mathsf{real}\langle 1 \rangle \\
    (+) &:& \forall S \mathord:\Scale.\ \mathsf{real}\langle S \rangle \to \mathsf{real}\langle S \rangle \to \mathsf{real}\langle S \rangle \\
    (-) &:& \forall S \mathord:\Scale.\ \mathsf{real}\langle S \rangle \to \mathsf{real}\langle S \rangle \to \mathsf{real}\langle S \rangle \\
    (*) &:& \forall S_1,S_2 \mathord:\Scale.\ \mathsf{real}\langle S_1 \rangle \to \mathsf{real}\langle S_2 \rangle \to \mathsf{real}\langle S_1S_2 \rangle \\
    (/) &:& \forall S_1,S_2 \mathord:\Scale.\ \mathsf{real}\langle S_1 \rangle \to \mathsf{real}\langle S_2 \rangle \to \mathsf{real}\langle S_1S_2^{-1} \rangle + 1 \\
    \mathrm{abs} &:& \forall S \mathord:\Scale.\ \mathsf{real}\langle S \rangle \to \mathsf{real}\langle |S| \rangle \\
    \mathrm{sqrt} &:& \mathsf{real}\langle 1 \rangle \to \mathsf{real}\langle 1 \rangle
  \end{eqnarray*}
  \caption{Operations on scaled real numbers}
  \label{fig:real-ops}
\end{figure}

The cross product is another operation that takes pairs of vectors and
produces a real number. Algebraically, the cross product is defined on
coordinate representations as follows:
\begin{displaymath}
  (x_1,y_1) \times (x_2,y_2) = x_1y_2 - x_2y_1
\end{displaymath}
Geometrically, the cross product of two vectors is the signed area of
the parallelogram described by the pair of input vectors. It turns out
that the value of the cross product of two vectors varies with the
determinant of the matrix describing the basis of the vector space. We
therefore augment our calculus with a new sort $\Scale$ of scale
factors. Semantically, $\Scale$ ranges over the non-zero real numbers
and forms an abelian group, which we write multiplicatively. We also
add two new operations: determinant, $\det B$, which takes an
inhabitant of $\GL(2)$ to its determinant in $\Scale$, and the
operation of absolute value that takes scaling factors to scaling
factors which we write as $|e|$. We also refine the type
$\mathsf{real}$ of real numbers to take a parameter of sort $\Scale$:
$\mathsf{real}\langle e \rangle$, where the old type $\mathsf{real}$
is just $\mathsf{real}\langle 1 \rangle$. The collection of operations
on real numbers with scaling factors is shown in
\autoref{fig:real-ops}.

With the additional sort $\Scale$, and the refined type of real
numbers, we can state the type of the cross product.
\begin{displaymath}
  (\times) : \forall B \mathord: \GL(2).\ \mathsf{vec}\langle B \rangle \to \mathsf{vec}\langle B \rangle \to \mathsf{real}\langle \det B \rangle
\end{displaymath}

For $O : \Orth(2)$, the absolute value of the determinant is always
$1$, so we assume the equation $|\det O| = 1$ to hold.

\begin{example}
  Using the structure we have defined in this subsection, we can
  define the following function that computes the area of a
  triangle. By the type of this function, we can see that the area of
  a triangle is invariant under orthognal transformations and
  arbitrary translations.
  \begin{displaymath}
    \begin{array}{l}
      \mathrm{area} : \forall O\mathord:\Orth(2), t\mathord:\Transl(2).\ \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{real}\langle 1 \rangle \\
      \mathrm{area}\ [O]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
    \end{array}
  \end{displaymath}
  In the body of this function, we have performed the calculation in
  several steps, each of which removes some of the symmetry described
  by the types. First we have computed the two offset vectors $p_2 -
  p_1$ and $p_3 - p_1$. These operations remove the effect of
  translations on the result, in exactly the same way as the type
  isomorphism we demonstrated above. We then compute the cross product
  of the two vectors to determine the area of the parallelogram
  described by the sides of the triangle. This value has type
  $\mathsf{real}\langle \det O \rangle$. We know that the determinant
  of an orthogonal transformation is always either $-1$ or $1$, so we
  have a result which varies according to possible reflections of the
  space. To remove this variation, we apply the $\mathrm{abs}$
  function to leave us with a value of type $\mathsf{real}\langle
  |\det O| \rangle$. This type is equal to the type
  $\mathsf{real}\langle 1 \rangle$. Finally, we multiply by
  $\frac{1}{2}$ to recover the area of the triangle, rather than the
  whole parallelogram.
\end{example}

% \subsection{Transformations Within and Between Frames}
% \label{sec:runtime-rep-frames}

% \begin{enumerate}
% \item A translation with respect to some frame is simply a vector with
%   respect to that frame's basis. We can apply it to points just by 
% \item 
% \end{enumerate}

% Singleton types

% Representing frames, a type representing translations:
% \begin{displaymath}
%   \mathsf{translation}\langle T \rangle
% \end{displaymath}
% And representing bases, as linear maps
% \begin{displaymath}
%   \mathsf{invLinearMap}\langle B \rangle
% \end{displaymath}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FIXME: write down all the groups, and the functions between them

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plain}
\bibliography{paper}

\end{document}

### 2012-05-08 paper/paper.tex
Import all the latex files

\documentclass[natbib,preprint]{sigplanconf}

\usepackage[usenames]{color}
\definecolor{citationcolour}{rgb}{0,0.4,0.2}
\definecolor{linkcolour}{rgb}{0,0,0.8}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
            urlcolor=linkcolour,
            linkcolor=linkcolour,
            citecolor=citationcolour,
            pdftitle=Relational Parametricity for Algebraically-Indexed Types,
            pdfauthor={Robert Atkey, Patricia Johann, Neil Ghani},
            pdfkeywords={}}  
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}

\title{Relational Parametricity for Algebraically Indexed Types}

\authorinfo{Robert Atkey\and Patricia Johann\and Neil Ghani}
           {University of Strathclyde}
           {\{Robert.Atkey,Patricia.Johann,Neil.Ghani\}@cis.strath.ac.uk}

\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage[all]{xy}

\newcommand{\sepbar}{\mathrel|}

\newcommand{\Rel}{\mathrm{Rel}}
\newcommand{\relArrow}{\mathrel{\widehat\to}}
\newcommand{\relTimes}{\mathrel{\widehat\times}}
\newcommand{\relSum}{\mathrel{\widehat+}}

\newcommand{\setOfIntegers}{\mathbb{Z}}
\newcommand{\setOfBooleans}{\{\tmTT, \tmFF\}}

\newcommand{\id}{\mathrm{id}}

\newcommand{\indexOp}[1]{\texttt{#1}}
\newcommand{\idxTms}[2]{\mathrm{IdxTm}(#1 \vdash #2)}
\newcommand{\idxTmAS}[1]{\mathrm{IdxTm}(#1)}

\newcommand{\sortType}{\texttt{type}}
\newcommand{\tyInt}{\texttt{int}}
\newcommand{\tyBool}{\texttt{bool}}
\newcommand{\tyUnit}{\texttt{unit}}
\newcommand{\tyPrim}[2]{\texttt{#1}\langle #2 \rangle}
\newcommand{\tyPrimNm}[1]{\texttt{#1}}

\newcommand{\tyArr}{\to}
\newcommand{\tyProduct}{\times}
\newcommand{\tyX}[1]{\texttt{X}\langle #1 \rangle}
\newcommand{\isType}{\textsf{ type}}
\newcommand{\isCtxt}{\textsf{ ctxt}}
\newcommand{\Ty}{\textsf{Ty}}
\newcommand{\ty}{\textsf{ty}}
\newcommand{\subst}[2]{\mathrm{Subst}(#1,#2)}

\newcommand{\tmTT}{\texttt{tt}}
\newcommand{\tmFF}{\texttt{ff}}

\newcommand{\relEnv}[1]{\mathcal{#1}}
\newcommand{\tySem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}}}
\newcommand{\ctxtSem}[1]{\llbracket #1 \rrbracket^{\mathcal{C}}}
\newcommand{\tmSem}[1]{\llbracket #1 \rrbracket^{\mathit{tm}}}
\newcommand{\tyPrimSem}[1]{\llbracket \tyPrimNm{#1} \rrbracket^{\mathcal{T}_0}}
\newcommand{\rsem}[3]{\llbracket #1 \rrbracket^{\mathcal{R}}_{#2}{#3}}
\newcommand{\extends}[2]{\mathsf{ext}(#1,#2)}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
\newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
\newcommand{\corref}[1]{\hyperref[#1]{Corollary \ref*{#1}}}
\newcommand{\conref}[1]{\hyperref[#1]{Conjecture \ref*{#1}}}
\newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}

\begin{document}

\maketitle

\begin{abstract}

\end{abstract}

\category{D.1.1}{Programming techniques}{Applicative (functional)
  programming} \category{D.2.4}{Software Engineering}{Software/Program
  Verification} \category{D.3.3}{Programming Languages}{Language
  Constructs and Features---Data types and structures}

\terms
  Languages, Theory, Types

\keywords
  parametricity, dimension types, computational geometry

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

\begin{enumerate}
\item Introduction using the area-of-a-triangle example
\item Narrowing of the semantic gap by use of advanced type systems
\item Importance of the use of symmetry in computer science,
  mathematics and the physical sciences
\item Development of the theory of parametricity
\end{enumerate}

\subsection{Contributions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background and Related Work}

Reynolds (Parametricity), Wadler (Free Theorems).

Kennedy (Units of Measure)

Something about symmetry?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Type System}

We now introduce our type system. We parameterise our development by
an arbitrary multi-sorted algebraic theory, in the sense of universal
algebra. We present the precise form of the theories we allow in
\autoref{sec:index-exp}, and define some auxiliary notions of
substitution and terms-up-to-equality. Several examples of useful
algebraic theories will be given in
\autoref{sec:example-systems}. Once an algebraic theory has been
fixed, we define the well-formed types and terms with respect to this
theory in \autoref{sec:types-and-terms}.

\subsection{Algebraic Theories for Indexing}\label{sec:index-exp}

\begin{figure*}[t]
  \centering
  Well-sorted terms
  \begin{mathpar}
    \inferrule*
    {i : s \in \Delta}
    {\Delta \vdash x : s}

    \inferrule*
    {\indexOp{f} : s_1,...,s_n \to s \\
      \Delta \vdash e_1 : s_1 \\
      \dots \\
      \Delta \vdash e_n : s_n}
    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) : s}
  \end{mathpar}

  \medskip

  Equational rules
  \begin{mathpar}
    \inferrule*
    {e_1 \stackrel{ax}= e_2 : s \in \mathit{Ax}(\Delta)}
    {\Delta \vdash e_1 \equiv e_2 : s}

    \inferrule*
    {\indexOp{f} : s_1,...,s_n \to s \\
      \Delta \vdash e_1 \equiv e'_1 : s_1 \\
      \dots \\
      \Delta \vdash e_n \equiv e'_n : s_n}
    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) \equiv \indexOp{f}(e'_1, ..., e'_n) : s}

    \inferrule*
    {\Delta \vdash e : s}
    {\Delta \vdash e \equiv e : s}

    \inferrule*
    {\Delta \vdash e_1 \equiv e_2 : s}
    {\Delta \vdash e_2 \equiv e_1 : s}

    \inferrule*
    {\Delta \vdash e_1 \equiv e_2 : s \\ \Delta \vdash e_2 \equiv e_3 : s}
    {\Delta \vdash e_1 \equiv e_3 : s}
  \end{mathpar}
  \caption{Algebraic Theories for Indexing}
  \label{fig:alg-theories}
\end{figure*}

We assume a set $\mathit{Sort}$ of \emph{index sorts}, and we will use
$s, s_1, s_2, ...$ to range over elements of $\mathit{Sort}$. An index
context $\Delta$ is list of index variable and sort pairs, $i : s$,
and we use the symbol $\epsilon$ to denote the empty index context (we
will also use this symbol below to denote the empty typing
context). We use letters like $i$ and $j$, sometimes with subscripts,
to range over index variables. We also assume a collection
$\mathit{IndexOp}$ of sorted index operations $\indexOp{f} :
s_1,...,s_n \to s$. From the index operations we generate the
well-sorted index expressions by the rules in
\autoref{fig:alg-theories}.

For each index context $\Delta$, we assume a set $\mathit{Ax}(\Delta)$
of axioms $e_1 \stackrel{ax}= e_2 : s$ where both $\Delta \vdash e_1 :
s$ and $\Delta \vdash e_2 : s$. From this collection of axioms, we
generate the equational theory of index expressions by the rules in
\autoref{fig:alg-theories} by closing under congruence, reflexivity,
symmetry and transitivity.

% For a pair of index contexts $\Delta$ and $\Delta'$, a substitution
% $\sigma \in \subst{\Delta}{\Delta'}$ is a function $\sigma$ mapping
% variables in $\Delta$ to well-sorted index expressions in $\Delta'$.

Index context morphisms $\sigma : \Delta \Rightarrow \Delta'$ consist
of a sequence of terms, one for each element of $\Delta'$, all
well-sorted in the index context $\Delta$.
\begin{mathpar}
  \inferrule*
  { }
  {\cdot : \Delta \Rightarrow \epsilon}

  \inferrule*
  {\sigma : \Delta \Rightarrow \Delta' \\ \Delta \vdash e : s}
  {\sigma, e : \Delta \Rightarrow \Delta', i:s}
\end{mathpar}
Directly from the definition of index context morphisms, we can see
that given an index context morphism $\sigma : \Delta \Rightarrow
\Delta'$, then for any index variable $i : s \in \Delta'$ there is an
index expression $\sigma(i)$ such that $\Delta \vdash \sigma(i) : s$.

We write $\id_\Delta : \Delta \Rightarrow \Delta$ for the index
context morphism comprised of the terms built from the variables of
$\Delta$, in order. We write $\pi_{i\mathord:s} : \Delta, i : s
\Rightarrow \Delta$ for the context morphism again comprised of the
terms built from the variables of $\Delta$, but this time typed in a
larger context. For an index context morphism $\sigma : \Delta \to
\Delta'$ we write $\sigma_s : \Delta, i\mathord:s \Rightarrow \Delta',
i\mathord:s$ for the index context morphism generated by lifting each
term to the larger context $\Delta, i\mathord:s$, and mapping the
variable $i\mathord:s$ to the term $i$.

Given an index expression $\Delta' \vdash e : s$, we define
re-indexing by an index context morphism $\sigma : \Delta \Rightarrow
\Delta'$ by the following clauses:
\begin{eqnarray*}
  \sigma^*i & = & \sigma(i) \\
  \sigma^*\indexOp{f}(e_1, ..., e_n) & = & \indexOp{f}(\sigma^*e_1,...,\sigma^*e_n)
\end{eqnarray*}
It is easy to check that these clauses preserve well-formedness of
index expressions, so that we obtain $\Delta \vdash \sigma^*e :
s$. Moreover re-indexing by index context morphisms preserves
equality:
\begin{lemma}\label{lem:indexeq-subst}
  If $\Delta \vdash e_1 \equiv e_2 : s$, then for any $\sigma :
  \Delta' \Rightarrow \Delta$, we have $\Delta' \vdash \sigma^*e_1
  \equiv \sigma^*e_2 : s$.
\end{lemma}

Also, projection commutes with lifting of substitutions. This is used
in the proof of \lemref{lem:tysubst-rel}.
\begin{lemma}
  $\sigma^* \circ \pi^*_{i\mathord:s} = \pi^*_{i\mathord:s} \circ \sigma^*_s$
\end{lemma}

For an index context $\Delta$ and sort $s$, we write
$\idxTms{\Delta}{s}$ for the set of index expressions $e$ such that
$\Delta \vdash e : s$, quotiented by the equivalence relation $\Delta
\vdash - \equiv - : s$. In light of \lemref{lem:indexeq-subst}, we can
see that re-indexing by an index context morphism $\sigma : \Delta
\Rightarrow \Delta'$ can be used as a function $\sigma^* :
\idxTms{\Delta'}{s} \to \idxTms{\Delta}{s}$, for any sort $s$.

\subsection{Types and Terms}\label{sec:types-and-terms}

\begin{figure*}[t]
  \centering
  Well-indexed Types
  \begin{mathpar}
    \inferrule* [right=TyPrimIdx]
    {\tyPrimNm{X} \in \mathit{PrimType}(s) \\ \Delta \vdash e : s}
    {\Delta \vdash \tyPrim{X}{e} : \sortType}

    \inferrule* [right=TyPrimNoIdx]
    {\tyPrimNm{X} \in \mathit{PrimType}(\cdot)}
    {\Delta \vdash \tyPrimNm{X} : \sortType}

    \inferrule* [right=TyUnit]
    { }
    {\Delta \vdash \tyUnit : \sortType}

    \inferrule* [right=TyArr]
    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
    {\Delta \vdash A \tyArr B : \sortType}

    \inferrule* [right=TyProd]
    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
    {\Delta \vdash A \tyProduct B : \sortType}

    \inferrule* [right=TySum]
    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
    {\Delta \vdash A + B : \sortType}

    \inferrule* [right=TyForall] %FIXME: macroize forall
    {\Delta, i : s \vdash A : \sortType}
    {\Delta \vdash \forall i \mathord: s. A : \sortType}
  \end{mathpar}

  \medskip
  Equality of well-indexed types
  \begin{mathpar}
    \inferrule* [right=TyEqPrimIdx]
    {\Delta \vdash e_1 \equiv e_2 : s}
    {\Delta \vdash \tyPrim{X}{e_1} \equiv \tyPrim{X}{e_2} : \sortType}

    \inferrule* [right=TyEqPrimNoIdx]
    { }
    {\Delta \vdash \tyPrimNm{X} \equiv \tyPrimNm{X} : \sortType}

    \inferrule* [right=TyEqUnit]
    { }
    {\Delta \vdash \tyUnit \equiv \tyUnit : \sortType}

    \inferrule* [right=TyEqArr]
    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
    {\Delta \vdash A \tyArr B \equiv A' \tyArr B' : \sortType}

    \inferrule* [right=TyEqProd]
    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
    {\Delta \vdash A \tyProduct B \equiv A' \tyProduct B' : \sortType}

    \inferrule* [right=TyEqSum]
    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
    {\Delta \vdash A + B \equiv A' + B' : \sortType}

    \inferrule* [right=TyEqForall]
    {\Delta, i : s \vdash A \equiv A' : \sortType}
    {\Delta \vdash \forall i \mathord: s. A \equiv \forall i \mathord: s. A' : \sortType}
  \end{mathpar}

  \medskip
  Well-indexed contexts
  \begin{mathpar}
    \inferrule*
    { }
    {\Delta \vdash \epsilon \isCtxt}

    \inferrule*
    {\Delta \vdash \Gamma \isCtxt \\ \Delta \vdash A : \sortType \\ x \not\in \Gamma}
    {\Delta \vdash \Gamma, x : A \isCtxt}
  \end{mathpar}
  \caption{Well-indexed Types and Contexts}
  \label{fig:types}
\end{figure*}

For each sort $s \in \mathit{Sort}$, we assume we are given a set of
names of primitive types $\mathit{PrimType}(s)$. We assume that the
sets $\mathit{PrimType}(s)$ are disjoint for non-equal sorts $s_1,
s_2$. For example, in the units of measure example in
\autoref{sec:units-example}, there is a single sort $\mathsf{unit}$,
and $\mathit{PrimType}(\mathsf{unit}) = \{\tyPrimNm{num}\}$. For each
primitive type $\tyPrimNm{X} \in \mathit{PrimType}(s)$, and index
expression $\Delta \vdash e : s$, we form the type $\tyPrim{X}{e}$, by
the \TirName{TyPrimIdx} rule in \autoref{fig:types}. Thus, in the
units-of-measure example, the type $\tyPrim{num}{u}$ represents
numbers measured by the units $u$.

We also assume that there is an additional set of primitive types
$\mathit{PrimType}(\cdot)$ that is disjoint from all the other sets of
primitive types. These primitive types do not need to be paired with
an index expression to form well-indexed types. Such types are created
by the \TirName{TyPrimNoIdx} rule. An example of such a primitive type
is the type $\tyPrimNm{real}$ of real-valued scalars in the Euclidean
plane geometry example in \autoref{sec:2d-example}.

The rules for generating the rest of the types of the system,
including unit, product, sum, function and universal types, are given
in \autoref{fig:types}. We use capital roman letters like $A$ and $B$
to range over well-indexed types. Well-indexed types come with an
equational theory, derived from the equational theory on index
expressions. This equational theory is also presented in
\autoref{fig:types}.

We define substitution by an index context morphism $\sigma : \Delta
\Rightarrow \Delta'$ on types by structural induction on the structure
of the type: given a type $\Delta' \vdash A : \sortType$, we obtain a
type $\Delta \vdash \sigma^*A : \sortType$. The only interesting case
is for \TirName{TyPrimIdx}:
\begin{displaymath}
  \sigma^*\tyPrim{X}{e} = \tyPrim{X}{\sigma^*e}
\end{displaymath}

Well-formed contexts are defined to be lists of variables paired with
well-formed types with respect to some index context $\Delta$, with
the condition that no variable names are repeated. Formally, contexts
are defined by the two rules shown in
\autoref{fig:types}. Substitution by an index context morphism extends
to contexts just by doing substitution on each individual type.

The typing rules for our system define the well-typed terms with
respect to an index context $\Delta$ and a type context $\Delta \vdash
\Gamma \isCtxt$. The judgement $\Delta; \Gamma \vdash M : A$ is
defined in \autoref{fig:terms}. The equational theory on types is
incorporated into the type system via the rule \TirName{TyEq}, which
allows for a term that is judged to have type $A$ to also have any
other equal type $B$.
\begin{figure*}[t]
  \centering
  \begin{mathpar}
    \inferrule* [right=Var]
    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
    {\Delta; \Gamma \vdash x : A}

    \inferrule* [right=TyEq]
    {\Delta; \Gamma \vdash M : A \\ \Delta \vdash A \equiv B : \sortType}
    {\Delta; \Gamma \vdash M : B}

    \inferrule* [right=Unit]
    { }
    {\Delta; \Gamma \vdash * : 1}

    \inferrule* [right=Pair]
    {\Delta; \Gamma \vdash M : A \\
      \Delta; \Gamma \vdash N : B}
    {\Delta; \Gamma \vdash (M, N) : A \tyProduct B}

    \inferrule* [right=Proj1]
    {\Delta; \Gamma \vdash M : A \tyProduct B}
    {\Delta; \Gamma \vdash \pi_1 M : A}

    \inferrule* [right=Proj2]
    {\Delta; \Gamma \vdash M : A \tyProduct B}
    {\Delta; \Gamma \vdash \pi_2 M : B}

    \inferrule* [right=Inl]
    {\Delta; \Gamma \vdash M : A}
    {\Delta; \Gamma \vdash \mathrm{inl}\ M : A + B}

    \inferrule* [right=Inr]
    {\Delta; \Gamma \vdash M : B}
    {\Delta; \Gamma \vdash \mathrm{inr}\ M : A + B}

    \inferrule* [right=Cond]
    {\Delta; \Gamma \vdash M : A + B \\
      \Delta; \Gamma, x : A \vdash N_1 : C \\
      \Delta; \Gamma, y : B \vdash N_2 : C}
    {\Delta; \Gamma \vdash \textrm{case}\ M\ \textrm{of}\ \textrm{inl}\ x.N_1; \textrm{inr}\ y.N_2 : C}

    \inferrule* [right=Abs]
    {\Delta; \Gamma, x : A \vdash M : B}
    {\Delta; \Gamma \vdash \lambda x.M : A \tyArr B}

    \inferrule* [right=App]
    {\Delta; \Gamma \vdash M : A \tyArr B \\
      \Delta; \Gamma \vdash N : A}
    {\Delta; \Gamma \vdash M N : B}

    \inferrule* [right=UnivAbs]
    {\Delta, i : s; \pi_{i\mathord:s}^*\Gamma \vdash M : A}
    {\Delta; \Gamma \vdash \Lambda i. M : \forall i\mathord:s. A}

    \inferrule* [right=UnivApp]
    {\Delta; \Gamma \vdash M : \forall i\mathord:s. A \\ \Delta \vdash e : s}
    {\Delta; \Gamma \vdash M [e] : (\id_\Delta, e)^*A}
  \end{mathpar}

  \caption{Well-typed terms}
  \label{fig:terms}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example Systems}\label{sec:example-systems}

\subsection{Units of Measure}\label{sec:units-example}

We now show that our general framework encompasses Kennedy's
units-of-measure system \cite{kennedy97relational}. For our indexing
algebraic theory, we have a single sort $\mathsf{unit}$ intended to
represent composite units such as $\mathit{m}\ \mathit{s}^{-1}$ or
$\mathit{kg}\ \mathit{m}^{-2}$. Units of measure form an abelian
group, so we have three operations:
\begin{mathpar}
  1 : \mathsf{unit},

  \cdot : \mathsf{unit}, \mathsf{unit} \to \mathsf{unit},

  -^{-1} : \mathsf{unit} \to \mathsf{unit},
\end{mathpar}
subject to the normal abelian group axioms. We will elide the group
multiplication operation and just write multiplication as
juxtaposition: $u_1u_2$.

We let $\mathit{PrimType}(\mathsf{unit}) = \{\tyPrimNm{num}$, and
$\mathit{PrimType}(\cdot) = \{\tyPrimNm{rational}\}$. The intended
reading is that the type $\tyPrim{num}{u_1u^2_2}$ represents
quantities in units of $u_1u^2_2$, while values of type
$\tyPrimNm{rational}$ represent dimensionless constants. We make a
distinction between $\tyPrim{num}{1}$ and $\tyPrimNm{rational}$ for
technical reasons arising from the relational interpretation below: we
want the relation on $\tyPrim{num}{1}$ to be the identity relation,
and the existence of $\mathit{toRat}$ and $\mathit{fromRat}$ forces
this\footnote{Kennedy assumes that all relational environments set the
  relation on $\tyPrim{num}{1}$ to be the identity relation, but we
  cannot do this in general, because we do not necessarily have a
  group structure in the indexing theory.}. To write programs using
the $\tyPrimNm{num}$ type, we assume an typing context
$\Gamma_{\textrm{units}}$ consisting of the following typed
operations:
\begin{eqnarray*}
  \mathit{toRat} & : & \tyPrim{num}{1} \to \tyPrimNm{rational} \\
  \mathit{fromRat} & : & \tyPrim{num}{1} \to \tyPrimNm{rational} \\
  0 & : & \forall u. \tyPrim{num}{u} \\
  1 & : & \tyPrim{num}{1} \\
  + & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyPrim{num}{u} \\
  - & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyPrim{num}{u} \\
  * & : & \forall u_1 u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to \tyPrim{num}{u_1u_2} \\
  / & : & \forall u_1 u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to \tyPrim{num}{u_1u_2^{-1}} + 1 \\
  < & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyBool
\end{eqnarray*}
There are constants for zero and one, but only the constant for zero
is polymorphic in the units. In \autoref{sec:units-semantics} below,
we will following Kennedy by interpreting units of measure in terms of
\emph{scalings} by rational numbers, where it will be apparent that
the constant one cannot be unit-polymorphic: we can scale $0$ and
still get $0$, but we cannot scale $1$ by an arbitrary rational number
and expect to get $1$.

\subsection{Euclidean Plane Geometry}\label{sec:2d-example}

We now consider a type system for writing programs that are invariant
under the translational symmetries of the Euclidean plane. We assume a
single sort $\mathsf{translation}$. The sort $\mathsf{translation}$ is
intended to represent translations in the plane, and comes equipped
with abelian group structure that we write additively:
\begin{mathpar}
  0 : \mathsf{translation}

  + : \mathsf{translation},\mathsf{translation} \to \mathsf{translation}

  - : \mathsf{translation} \to \mathsf{translation}
\end{mathpar}
In indexing expressions involving variables of sort
$\mathsf{translation}$, we use the shorthand $t_1 - t_2$ to stand for
$t_1 + (-t_2)$.

There are two primitive types:
$\mathit{PrimType}(\mathsf{translation}) = \{\tyPrimNm{vec}\}$, and
$\mathit{PrimType}(\cdot) = \{\tyPrimNm{real}\}$. The types
$\tyPrim{vec}{t}$ represent two-dimensional vectors, indistinguishable
up to some translation $t$. Values of type $\tyPrimNm{real}$ are just
real numbers. To write programs using these types we assume a typing
context $\Gamma_{\mathit{2D}}$ consisting of the following operations:
\begin{eqnarray*}
  + & : & \forall t_1 t_2\mathord:\mathsf{translation}.\tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2} \\
  - & : & \forall t\mathord:\mathsf{translation}. \tyPrim{vec}{t} \to \tyPrim{vec}{-t} \\
  0 & : & \tyPrim{vec}{0} \\
  * & : & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0} \\
  \mathrm{getX} & : & \tyPrim{vec}{0} \to \tyPrimNm{real} \\
  \mathrm{getY} & : & \tyPrim{vec}{0} \to \tyPrimNm{real} \\
  \mathrm{vec}  & : & \tyPrimNm{real} \to \tyPrimNm{real} \to \tyPrim{vec}{0}
\end{eqnarray*}
The first three operations express the abelian group structure of
2-dimensional real vectors. Note how the indexing tracks the
operations on vectors: e.g.~addition on vectors is reflected by
addition on the indicies. The final four operations are operations
that are not invariant under arbitrary translation: multiplication by
a scalar, and projection from and constructon of vectors. For the
purposes of examples, we also assume a standard set of arithmetic and
other operations on real numbers.

As an example of the use of the above operations, consider the
following type of a function for computing the area of a triangle
described by three points. Since the area of a triangle is invariant
under translations that are applied to all three points, we can
quantify over all translations $t$:
\begin{displaymath}
  \mathit{area} : \forall t\mathord:\mathsf{translation}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real}
\end{displaymath}
Since the result is invariant under translation, it safe to assume
that one of the vertexes of the triangle is the origin, and only
specify two points to describe the triangle. So we might expect that
the type of $\mathit{area}$ is isomorphic to this type:
\begin{displaymath}
  \mathit{area'} : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrimNm{real}
\end{displaymath}
Given $\mathit{area'}$, we can define $\mathit{area}$ using the
operations in $\Gamma_{\mathit{2D}}$:
\begin{displaymath}
  \mathit{area} = \Lambda t. \lambda v_1 v_2 v_3.\ \mathit{area'}\ (v_2 - v_1)\ (v_3 - v_1)
\end{displaymath}
and vice versa, we can define $\mathit{area'}$ in terms of
$\mathit{area}$:
\begin{displaymath}
  \mathit{area'} = \lambda v_1 v_2. \mathit{area}\ [0]\ 0\ v_1\ v_2
\end{displaymath}
In \autoref{sec:2d-semantics} below, we will show that these two
translations form a type isomorphism between the types of
$\mathit{area}$ and $\mathit{area'}$.

\paragraph{Alternative Transformation Groups?}
We should also consider other groups of transformations for indexing:
isometries, orthogonal transformations, rotations and reflections.

\subsection{Security Typing and Non-interference}

This example shows how our framework can accommodate non-interference
properties common in security type systems. This example has a
different flavour to the two previous examples because it does not
concern invariance under the action of some group. We assume a single
sort $\mathsf{principal}$, intended to represent (possibly composite)
principals in a system. Principals are combined using the connectives
of boolean logic:
\begin{mathpar}
  \top : \mathsf{principal}

  \bot : \mathsf{principal}

  \land : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}

  \lor : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}

  \Rightarrow : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}

  \lnot : \mathsf{principal} \to \mathsf{principal}
\end{mathpar}
For the equational theory of principals, we assume all the axioms of
boolean algebra.

Let $\mathit{PrimType}(\mathsf{principal}) = \{\tyPrimNm{T}\}$. The
intended reading of a type $\tyPrim{T}{p}$ for some principal $p$ is
that the existence of a value of this type indicates that the
principal $p$ is ``happy''. To write programs, we assume the following
typing context $\Gamma_{\mathit{sec}}$ of operations:
\begin{eqnarray*}
  \mathrm{proj} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1 \land p_2} \to \tyPrim{T}{p_1} \\
  \mathrm{combine} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1} \to \tyPrim{T}{p_2} \to \tyPrim{T}{p_1 \land p_2}
\end{eqnarray*}

To illustrate the use of this application of our framework, we define
a type synonym, for each type $A$ and principal $p$:
\begin{displaymath}
  T_pA = \tyPrim{T}{p} \to A
\end{displaymath}
For every principal $p$, we can endow the types $T_p-$ with the
structure of a monad. This is due to the fact that it is an instance
of the ``environment'' (or ``reader'') monad. Explicitly, the
$\mathit{return}$ and $>>=$ of these monads are given by:
\begin{eqnarray*}
  \mathit{return} & : & A \to T_pA \\
  \mathit{return} & = & \lambda a\ t.\ a \\
  \\
  (>>=) & : & T_pA \to (A \to T_pB) \to T_pB \\
  (>>=) & = & \lambda a\ f\ t.\ f\ (a\ t)\ t
\end{eqnarray*}
The intended reading of a value of type $T_pA$ is as a piece of data
of type $A$ that can only be accessed if the principal $p$ is
happy. The existence of such a monad should allow us to draw parallels
between our system and the Dependency Core Calculus (DCC) system of
Abadi \emph{et al.} \cite{abadi99core}. The definition of a DCC monad
as an instance of an environment monad follows Tse and Zdancewic's
interpretation of DCC in System F, where they use System F
parametricity to ensure non-interference properties.

There is also a coercion operator between instances of $T_pA$ for
different principals. Reading the type in terms of happiness, this
says that if we have a piece of data that is protected by $p$, and
happiness of $q$ implies happiness of $p$, then this data can also be
accessed by $q$.
\begin{eqnarray*}
  \mathit{coerce} & : & \tyPrim{T}{q \Rightarrow p} \to T_pA \to T_qA \\
  \mathit{coerce} & = & \lambda t_{qp}\ a\ t_q.\ a\ (\mathrm{combine}\ t_{qp}\ t_q)
\end{eqnarray*}
Here we have used the fact that the result of the application of
$\mathrm{combine}$ has type $\tyPrim{T}{(q \Rightarrow p) \land
  q}$. By the laws of boolean algebras---which we have assumed for our
indexing theory---this is equal to the type $\tyPrim{T}{p}$. Thus we
get access to the data stored in $a$, which is protected by $p$.

Now suppose we have a term $M$ with the following typing:
\begin{displaymath}
  p,q:\mathsf{principal}; t_{qp} : \tyPrim{T}{q \Rightarrow p}, x : T_q (1 + 1) \vdash M : T_p (1 + 1)
\end{displaymath}
Intuitively, given the operations $\Gamma_{\mathit{sec}}$ we have
defined above, it should be the case that $M$ can only be equivalent
to the program which is constantly $\lambda t. \mathrm{inl}(*)$ or
$\lambda t. \mathrm{inr}(*)$, because there is no useful data to be
gained from the $t_{qp}$ argument, and a value of $\tyPrim{T}{q}$
cannot be generated in order to get access to the data within the $x$
argument. The application of our framework that we have presented in
this section has a ``non-interference'' property: the value of $x$
cannot interfere with the output. In \autoref{sec:sec-semantics} we
show that this is indeed the case.

\paragraph{Alternative Logics?}
Above, we assumed that the indexing theory of principals was the
theory of boolean algebras. As a consequence, a value of type $T_{p
  \lor \lnot p}A$ can always be accessed, due to the exlcuded middle
property of boolean logic. An interesting question is: what happens we
we consider alternative logics. If we take the theory of Heyting
algebras for our theory of principals, then data typed as $T_{p \lor
  \lnot p}A$ can only be accessed if either $p$ or $\lnot p$ can be
demonstrated. Considering an alternative logical theory alters the
non-interference property of the type system.

\paragraph{Alternative Presentations?}
In the above we presented the theory of boolean logic by means of the
equational theory of boolean algebras. This isn't necessarily the most
convenient way of presenting a logic. An alternative approach would be
to consider a trivial equational theory with no axioms, and
incorporate the rules of logic consequence as operations in
$\Gamma_{\mathit{sec}}$. For example, as a Hilbert system:
\begin{eqnarray*}
  \mathit{MP} & : & \forall p,q.\ \tyPrim{T}{p \Rightarrow q} \to \tyPrim{T}{p} \to \tyPrim{T}{q} \\
  \mathit{k} & : & \forall p,q.\ \tyPrim{T}{p \Rightarrow q \Rightarrow p} \\
  \mathit{s} & : & \forall p,q,r.\ \tyPrim{T}{(p \Rightarrow q \Rightarrow r) \Rightarrow (p \Rightarrow q) \Rightarrow p \Rightarrow r}
\end{eqnarray*}
Plus axioms for the other connectives.  Or possibly some kind of
encoding of a natural deduction system?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semantics and Parametricity Theorem}

There is an index-erasure semantics, and a relational semantics. The
relational semantics allows us to derive free theorems about the
index-erasure semantics.

\subsection{Index Erasure Semantics}

We now define a semantics of types and terms that ignores all indexing
information. By defining this semantics we can easily see that the
indexing information can be discarded before execution.

\subsubsection{Index-Erasure Semantics of Types}

For each primitive type $\tyPrimNm{X} \in \mathit{PrimType}(s)$, for
some sort $s$, and for the primitive types $\tyPrimNm{X} \in
\mathit{PrimType}(\cdot)$, we assume that $\tyPrimNm{X}$ is assigned a
set $\tyPrimSem{X}$. We extend this to assign a set to every
well-formed type by induction on the type structure. To interpret the
universal quantifier, we simply ignore the indexing:
\begin{eqnarray*}
  \tySem{\tyUnit} & = & \{*\} \\
  \tySem{\tyPrim{X}{e}} & = & \tyPrimSem{X} \\
  \tySem{\tyPrim{X}} & = & \tyPrimSem{X} \\
  \tySem{A \tyProduct B} & = & \tySem{A} \times \tySem{B} \\
  \tySem{A \tyArr B} & = & \tySem{A} \to \tySem{B} \\
  \tySem{A + B} & = & \tySem{A} + \tySem{B} \\
  \tySem{\forall i\mathord:s. A} & = & \tySem{A}
\end{eqnarray*}

Since our definition has completely ignored the indexing, and type
equality is defined as an extension of index equality, we can easily
see that equal types have equal denotations in our index-erasure
semantics:
\begin{lemma}\label{lem:tyeq-erasure}
  If $\Delta \vdash A \equiv B : \sortType$ then $\tySem{A} =
  \tySem{B}$.
\end{lemma}

\begin{proof}
  By induction on the derivation of $\Delta \vdash A \equiv B :
  \sortType$:
  \begin{description}
  \item[Case \TirName{TyEqUnit}] In this case, $A = B = \tyUnit$, so
    $\tySem{A} = \tySem{B}$.
  \item[Case \TirName{TyEqPrimIdx}] We have $A = \tyPrim{X}{e_1}$ and
    $B = \tyPrim{X}{e_2}$, for some $e_1$ and $e_2$ of the same
    sort. The erasure semantics ignores the index arguments, so
    $\tySem{A} = \tySem{\tyPrim{X}{e_1}} = \tyPrimSem{X} =
    \tySem{\tyPrim{X}{e_2}} = \tySem{B}$, as required.
  \item[Case \TirName{TyEqPrimNoIdx}] We have $A = \tyPrimNm{X}$ and
    $B = \tyPrimNm{X}$. So $\tySem{A} = \tyPrimSem{X} = \tySem{B}$, as
    required.
  \item[Case \TirName{TyEqArr}] Here, $A = A_1 \tyArr A_2$ and $B =
    B_1 \tyArr B_2$, and by the induction hypothesis we know that
    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
    $\tySem{A} = \tySem{A_1 \tyArr A_2} = \tySem{A_1} \to \tySem{A_2}
    = \tySem{B_1} \to \tySem{B_2} = \tySem{B_1 \tyArr B_2} =
    \tySem{B}$.
  \item[Case \TirName{TyEqProd}] In this case, $A = A_1 \tyProduct
    A_2$ and $B = B_1 \tyProduct B_2$, and by the induction hypothesis
    we know that $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} =
    \tySem{B_2}$. Hence $\tySem{A} = \tySem{A_1 \tyProduct A_2} =
    \tySem{A_1} \times \tySem{A_2} = \tySem{B_1} \times \tySem{B_2} =
    \tySem{B_1 \tyProduct B_2} = \tySem{B}$.
  \item[Case \TirName{TyEqSum}] In this case, $A = A_1 + A_2$ and $B =
    B_1 + B_2$, and by the induction hypothesis we know that
    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
    $\tySem{A} = \tySem{A_1 + A_2} = \tySem{A_1} + \tySem{A_2} =
    \tySem{B_1} + \tySem{B_2} = \tySem{B_1 + B_2} = \tySem{B}$.
  \item[Case \TirName{TyEqForall}] We have $A = \forall
    i\mathord:s.A'$ and $B = \forall i\mathord:s.B'$. By the induction
    hypothesis we know that $\tySem{A'} = \tySem{B'}$, and hence
    $\tySem{A} = \tySem{\forall i\mathord:s. A'} = \tySem{A'} =
    \tySem{B'} = \tySem{\forall i\mathord:s. B'} = \tySem{B}$.
  \end{description}
\end{proof}

Moreover, the index-erasure semantics is insensitive to index
substitutions.
\begin{lemma}\label{lem:tysubst-erasure}
  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
  \Delta$, then $\tySem{\sigma^*A} = \tySem{A}$.
\end{lemma}

\begin{proof}
  By induction on the derivation of $\Delta \vdash A : \sortType$.
  \begin{description}
  \item[Case \TirName{TyUnit}] Directly from the definition of
    substitution on types of the form $\tyUnit$, we have
    $\tySem{\sigma^*\tyUnit} = \tySem{\tyUnit}$, as required.
  \item[Case \TirName{TyPrimIdx}] Since the index-erasure semantics
    ignores all index expessions, we have
    $\tySem{\sigma^*\tyPrim{X}{e}} = \tySem{\tyPrim{X}{\sigma^*e}} =
    \tyPrimSem{X} = \tySem{\tyPrim{X}{e}}$.
  \item[Case \TirName{TyPrimNoIdx}] Similar to the previous case.
  \item[Case \TirName{TyArr}] We have $A = A_1 \tyArr A_2$, and by the
    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
    \tySem{A_1}$ and $\tySem{\sigma^*B_1} = \tySem{B_1}$. Thus, by the
    definition of substitution on arrow types, we have
    $\tySem{\sigma^*(A_1 \tyArr A_2)} = \tySem{\sigma^*A_1 \tyArr
      \sigma^*A_2} = \tySem{\sigma^*A_1} \to \tySem{\sigma^*A_2} =
    \tySem{A_1} \to \tySem{A_2} = \tySem{A_1 \tyArr A_2}$, as
    required.
  \item[Case \TirName{TyProd}] We have $A = A_1 \tyProduct A_2$, and
    by the induction hypothesis, we know that $\tySem{\sigma^*A_1} =
    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
    definition of substitution on product types, we have
    $\tySem{\sigma^*(A_1 \tyProduct A_2)} = \tySem{\sigma^*A_1
      \tyProduct \sigma^*A_2} = \tySem{\sigma^*A_1} \times
    \tySem{\sigma^*A_2} = \tySem{A_1} \times \tySem{A_2} = \tySem{A_1
      \tyProduct A_2}$, as required.
  \item[Case \TirName{TySum}] We have $A = A_1 + A_2$, and by the
    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
    definition of substitution on sum types, we have
    $\tySem{\sigma^*(A_1 + A_2)} = \tySem{\sigma^*A_1 + \sigma^*A_2} =
    \tySem{\sigma^*A_1} + \tySem{\sigma^*A_2} = \tySem{A_1} +
    \tySem{A_2} = \tySem{A_1 + A_2}$, as required.
  \item[Case \TirName{TyForall}] We have $A = \forall i\mathord:s.A'$,
    and by the induction hypothesis we know that, for all $\sigma'$,
    $\tySem{\sigma'^*A'} = \tySem{A'}$. Therefore, from the definition
    of substitution for $\forall$-types, we have
    $\tySem{\sigma^*\forall i\mathord:s.A'} = \tySem{\forall
      i\mathord:s.\sigma_s^{*}A'} = \tySem{\sigma_s^{*}A'} = \tySem{A'}
    = \tySem{\forall i\mathord:s.A'}$, where we have relied on the
    fact that the index-erasure semantics of types ignores the index
    quantification.
  \end{description}
\end{proof}

For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
\ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
one-element set.

\subsubsection{Index-Erasure Semantics of Terms}

For a well-formed term $\Delta; \Gamma \vdash e : A$, we define a
function $\tmSem{e} : \ctxtSem{\Gamma} \to \tySem{A}$, completely
ignoring the indexing information. In light of
\lemref{lem:tyeq-erasure} and \lemref{lem:tysubst-erasure}, we may
define this function directly on the syntax of well-typed terms,
rather than on the typing derivations. We define $\tmSem{e}$ by the
following clauses:
\begin{eqnarray*}
  \tmSem{x}\eta & = & \eta_x \\
  \tmSem{(M,N)}\eta & = & (\tmSem{M}\eta, \tmSem{N}\eta) \\
  \tmSem{\pi_1M}\eta & = & \pi_1(\tmSem{M}\eta) \\
  \tmSem{\pi_2M}\eta & = & \pi_2(\tmSem{M}\eta) \\
  \tmSem{\mathrm{inl}\ M}\eta & = & \mathrm{inl}\ (\tmSem{M}\eta) \\
  \tmSem{\mathrm{inr}\ M}\eta & = & \mathrm{inr}\ (\tmSem{M}\eta) \\
  \left\llbracket
    \begin{array}{l}
      \textrm{case}\ M\ \textrm{of}\\
      \textrm{inl}\ x.N_1;\\
      \textrm{inr}\ y.N_2
    \end{array}\right\rrbracket^{\mathit{tm}}\eta & = &
  \left\{
    \begin{array}{ll}
      \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
      \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
    \end{array}
  \right. \\
  \tmSem{\lambda x. M}\eta & = & \lambda v. \tmSem{M}(\eta, v) \\
  \tmSem{M N}\eta & = & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
  \tmSem{\Lambda i.\ M}\eta & = & \tmSem{M}\eta \\
  \tmSem{M[e]}\eta & = & \tmSem{M}\eta
\end{eqnarray*}
In the above, we have written $\eta_x$ to denote the appropriate
projection from $\eta$ to get the position corresponding to the
variable $x$ in the context.

\subsection{Relational Semantics}

\subsubsection{Relation Environments}

For a set $X$, let $\Rel(X)$ be the set of binary relations on
$X$. For an index context $\Delta$, a \emph{relation environment} is a
family of functions $\rho = \{ \rho_{\tyPrimNm{X}} :
\idxTms{\Delta}{s} \to \Rel(\tyPrimSem{X}) \}_{s \in \mathit{Sort},
  \tyPrimNm{X} \in \mathit{PrimType}(s)}$.

We define our relational semantics of types with respect to a given
collection of relation environments: we assume that, for each index
context $\Delta$, we are given a set of relation environments
$\relEnv{E}(\Delta)$. This family must satisfy the following two
properties:
\begin{enumerate}
\item Closure under re-indexing by index context morphisms: if $\rho
  \in \relEnv{E}(\Delta)$ and $\sigma : \Delta' \Rightarrow \Delta$,
  then $\rho \circ \sigma^* \in \relEnv{E}(\Delta')$.
\item If we have a pair of relation environments $\rho_1 \in
  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
  \relEnv{E}(\Delta)$, along with an index context morphism $\sigma :
  \Delta' \Rightarrow \Delta$, such that the outer edges of the
  following diagram commute, for all primitive types $\tyPrimNm{X} \in
  \mathit{PrimType}(s)$:
  \begin{displaymath}
    \xymatrix{
      {\idxTms{\Delta}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar[d]_{\sigma^*}
      &
      {\idxTms{\Delta,i\mathord:s'}{s}} \ar[d]^{\sigma_{s'}^*} \ar@/^/[rdd]^{\rho_1}
      \\
      {\idxTms{\Delta'}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2}
      &
      {\idxTms{\Delta',i\mathord:s'}{s}} \ar@{.>}[dr]_\rho
      \\
      &
      &
      {\Rel(X)}
    }
  \end{displaymath}
  Then there exists a relation environment $\rho \in
  \relEnv{E}(\Delta,i\mathord:s')$ such that the two triangles in the
  bottom right of the diagram commute.
\end{enumerate}

Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
additional index variable $i\mathord:s$ to be the following:
\begin{displaymath}
  \extends{\rho}{i\mathord:s} = \{ \rho' \in \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s} = \rho \}
\end{displaymath}

\subsubsection{Relational Semantics of Types}

%FIXME: re-word this
A given relation context $\rho \in \relEnv{E}(\Delta)$ is extended to
a relation interpretation of any type $\Delta \vdash A : \sortType$ by
induction its derivation. So we define a relation
$\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-formed
type $A$.
\begin{eqnarray*}
  \rsem{\tyUnit}{\relEnv{E}}\rho & = & \{(*,*)\} \\
  \rsem{\tyPrim{X}{e}}{\relEnv{E}}\rho & = & \rho_{\tyPrimNm{X}}(e) \\
  \rsem{A \tyArr B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A \tyProduct B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A + B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & = & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i:s} \}
\end{eqnarray*}
In this definition we have made use of the following two constructions
on binary relations. If $R \in \Rel(X)$ and $S \in \Rel(Y)$, then $R
\relArrow S \in \Rel(X \to Y)$ is defined as $\{ (f_1,f_2) \sepbar
\forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S \}$. With the same
assumptions on $R$ and $S$, the relation $R \relTimes S \in \Rel(X
\times Y)$ is defined as $\{ ((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2)
\in R \land (b_1,b_2) \in S \}$. FIXME: do $\relSum$.

We extend the relational semantics to contexts of well-formed types by
the following clauses:
\begin{eqnarray*}
  \rsem{\epsilon}{\relEnv{E}}{\rho} & = & \id_{\ctxtSem{\epsilon}} \\
  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & = & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
\end{eqnarray*}

\subsubsection{Parametricity}

This relies on \lemref{lem:tyeq-erasure} to make sure that it is
well-typed.

\begin{lemma}\label{lem:tyeq-rel}
  If $\Delta \vdash A \equiv B : \sortType$, then for all $\rho \in
  \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}\rho = \rsem{B}{\relEnv{E}}\rho$.
\end{lemma}

\begin{proof}
  By induction on the derivation of $\Delta \vdash A \equiv B :
  \sortType$.
  \begin{description}
  \item[Case \TirName{TyEqUnit}] In this case $A = B = \tyUnit$, so
    the statement is trivially statisfied.
  \item[Case \TirName{TyEqPrimIdx}] In this case we have $A =
    \tyPrim{X}{e_1}$ and $B = \tyPrim{X}{e_2}$, with $\Delta \vdash
    e_1 \equiv e_2 : s$. The index expressions $e_1$ and $e_2$ are in
    the same equivalence class, so by the definition of relation
    environments we have that $\rho_{\tyPrimNm{X}}(e_1) =
    \rho_{\tyPrimNm{X}}(e_2)$ and hence
    $\rsem{\tyPrim{X}{e_1}}{\relEnv{E}}\rho =
    \rsem{\tyPrim{X}{e_2}}{\relEnv{E}}\rho$.
  \item[Case \TirName{TyEqPrimNoIdx}] This case is similar to the case
    for \TirName{TyEqUnit}.
  \item[Case \TirName{TyEqArr}] We have $A = A_1 \tyArr A_2$ and $B =
    B_1 \tyArr B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$
    and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the induction
    hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_2}{\relEnv{E}}\rho$. Thus we may reason as follows:
    $\rsem{A_1 \tyArr A_2}{\relEnv{E}}\rho =
    \rsem{A_1}{\relEnv{E}}\rho \relArrow \rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho \relArrow \rsem{B_2}{\relEnv{E}}\rho =
    \rsem{B_1 \tyArr B_2}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{TyEqProd}] We have $A = A_1 \tyProduct A_2$ and
    $B = B_1 \tyProduct B_2$, with $\Delta \vdash A_1 \equiv B_1 :
    \sortType$ and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the
    induction hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
    $\rsem{A_1 \tyProduct A_2}{\relEnv{E}}\rho =
    \rsem{A_1}{\relEnv{E}}\rho \relTimes \rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho \relTimes \rsem{B_2}{\relEnv{E}}\rho =
    \rsem{B_1 \tyProduct B_2}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{TyEqSum}] We have $A = A_1 + A_2$ and $B = B_1 +
    B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$ and $\Delta
    \vdash A_2 \equiv B_2 : \sortType$. By the induction hypothesis,
    we know that $\rsem{A_1}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
    $\rsem{A_1 + A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}\rho
    \relSum \rsem{A_2}{\relEnv{E}}\rho = \rsem{B_1}{\relEnv{E}}\rho
    \relSum \rsem{B_2}{\relEnv{E}}\rho = \rsem{B_1 +
      B_2}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{TyEqForall}] We have $A = \forall
    i\mathord:s. A'$ and $B = \forall i\mathord:s.B'$, with $\Delta, i
    : s \vdash A' \equiv B' : \sortType$. By the induction hypothesis,
    we know that for all $\rho'$, $\rsem{A'}{\relEnv{E}}\rho' =
    \rsem{B'}\rho'$. We reason as follows:
    \begin{eqnarray*}
      &    & (x_1,x_2) \in \rsem{\forall i\mathord:s. A'}{\relEnv{E}}\rho \\
      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{A'}{\relEnv{E}}\rho' \\
      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{B'}{\relEnv{E}}\rho' \\
      &\iff& (x_1,x_2) \in \rsem{\forall i\mathord:s. B'}{\relEnv{E}}\rho
    \end{eqnarray*}
    Thus, by extensionality, we have $\rsem{\forall
      i\mathord:s.A'}\rho \equiv \rsem{\forall
      i\mathord:s.B'}{\relEnv{E}}\rho$, as required.
  \end{description}
\end{proof}

This next lemma relies on \lemref{lem:tysubst-erasure} to make sure
that it is well-typed.

\begin{lemma}\label{lem:tysubst-rel}
  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
  $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Likewise,
  if $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho = \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
\end{lemma}

\begin{proof}
  The first part of the lemma statement is proved by induction on the
  derivation of $\Delta \vdash A : \sortType$. We analyse each case in
  turn:
  \begin{description}
  \item[Case \TirName{TyUnit}] Directly from the definition of
    substitution on types of the form $\tyUnit$, and the fact that
    $\rsem{\tyUnit}{\relEnv{E}}\rho_1 =
    \rsem{\tyUnit}{\relEnv{E}}\rho_2$ for any pair $\rho_1,\rho_2$ of
    relation environments, we have
    $\rsem{\sigma^*\tyUnit}{\relEnv{E}}\rho =
    \rsem{\tyUnit}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as required.
  \item[Case \TirName{TyPrimIdx}] We have $A = \tyPrim{X}{e}$, therefore
    $\rsem{\sigma^*A}{\relEnv{E}}\rho =
    \rsem{\sigma^*\tyPrim{X}{e}}{\relEnv{E}}\rho =
    \rsem{\tyPrim{X}{\sigma^*e}}{\relEnv{E}}\rho =
    \rho_{\tyPrimNm{X}}(\sigma^*e) =
    \rsem{\tyPrim{X}{e}}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as
    required.
  \item[Case \TirName{TyPrimNoIdx}] Similar to the case for
    \TirName{TyUnit}.
  \item[Case \TirName{TyArr}] In this case, $A = A_1 \tyArr A_2$ and
    by the induction hypothesis, we know that $\rsem{\sigma^*A_1}{\relEnv{E}}\rho
    = \rsem{A_1}{\relEnv{E}}{(\rho \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we reason as
    follows: $\rsem{\sigma^*(A_1 \tyArr A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1
      \tyArr \sigma^*A_2}{\relEnv{E}}\rho = \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relArrow
    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho \circ \sigma^*)}
    \relArrow \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)} = \rsem{A_1 \tyArr
      A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as required.
  \item[Case \TirName{TyProd}] In this case, $A = A_1 \tyProduct A_2$
    and by the induction hypothesis, we know that
    $\rsem{\sigma^*A_1}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
      \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we
    reason as follows: $\rsem{\sigma^*(A_1 \tyProduct
      A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1 \tyProduct
      \sigma^*A_2}{\relEnv{E}}\rho =
    \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relTimes
    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
      \circ \sigma^*)} \relTimes \rsem{A_2}{\relEnv{E}}{(\rho \circ
      \sigma^*)} = \rsem{A_1 \tyProduct A_2}{\relEnv{E}}{(\rho \circ
      \sigma^*)}$, as required.
  \item[Case \TirName{TySum}] In this case, $A = A_1 + A_2$
    and by the induction hypothesis, we know that
    $\rsem{\sigma^*A_1}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
      \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we
    reason as follows: $\rsem{\sigma^*(A_1 +
      A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1 +
      \sigma^*A_2}{\relEnv{E}}\rho =
    \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relSum
    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
      \circ \sigma^*)} \relSum \rsem{A_2}{\relEnv{E}}{(\rho \circ
      \sigma^*)} = \rsem{A_1 + A_2}{\relEnv{E}}{(\rho \circ
      \sigma^*)}$, as required.
  \item[Case \TirName{TyForall}] We have $A = \forall
    i\mathord:s.A'$. We prove the required equation by demonstrating
    two inclusions. From left-to-right, we know that
    \begin{equation}
      \label{eq:tysubst-rel-forall1}
      (x_1,x_2) \in \rsem{\sigma^*\forall i\mathord:s. A'}{\relEnv{E}}\rho
    \end{equation}
    and we must show that $(x_1, x_2) \in \rsem{\forall
      i\mathord:s. A'}{\relEnv{E}}{(\rho \circ \sigma^*)}$. From
    (\ref{eq:tysubst-rel-forall1}) and the definition of substitution,
    we know that
    \begin{displaymath}
      \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{\sigma_s^*A'}{\relEnv{E}}\rho'
    \end{displaymath}
    By the induction hypothesis, we can deduce that
    \begin{equation}
      \label{eq:tysubst-rel-forall2}
      \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{A'}{\relEnv{E}}{(\rho' \circ \sigma_s^{*})}
    \end{equation}
    To show the required inclusion, we must show that for all $\rho_1
    \in \extends{\rho \circ \sigma^*}{i:s}$, it is the case that
    $(x_1,x_2) \in \rsem{A'}{\relEnv{E}}{\rho_1}$. We now make use of
    the pushout property of our family $\relEnv{E}$ of relation
    environments: we have a $\rho \in \relEnv{E}(\Delta)$ and $\rho_1
    \in \relEnv{E}(\Delta, i:s)$ such that $\rho_1 \circ \pi_{i:s} =
    \rho \circ \sigma^*$ (by the assumption that $\rho_1 \in
    \extends{\rho \circ \sigma^*}{i:s}$), so we can obtain a $\rho'
    \in \relEnv{E}(\Delta',i:s)$ such that $\rho' \circ \pi_{i:s} =
    \rho$ (so $\rho' \in \extends{\rho}{i:s}$), and $\rho' \circ
    \sigma^* = \rho_1$. Therefore we use
    (\ref{eq:tysubst-rel-forall2}) to deduce that $(x_1,x_2) \in
    \rsem{A'}{\relEnv{E}}{\rho_1}$, and the inclusion from
    left-to-right is shown.

    From right-to-left, we must now show
    (\ref{eq:tysubst-rel-forall2}) under the assumption that
    \begin{equation}
      \label{eq:tysubst-rel-forall3}
      \forall \rho_1 \in \extends{\rho \circ \sigma^*}{i:s}. (x_1,x_2) \in \rsem{A'}{\relEnv{E}}{\rho_1}.
    \end{equation}
    Given $\rho' \in \extends{\rho}{i:s}$, we let $\rho_1 = \rho'
    \circ \sigma^*$. Now, $\rho_1 \circ \pi_{i:s} = \rho' \circ
    \sigma^* \circ \pi_{i:s} = \rho' \circ \pi_{i:s} \circ \sigma_s^{*}
    = \rho \circ \sigma_s^{*}$. Thus $\rho_1 \in \extends{\rho \circ
      \sigma^*}{i:s}$ and we may use (\ref{eq:tysubst-rel-forall3}) to
    deduce that $(x_1,x_2) \in \rsem{A'}{\relEnv{E}}{(\rho' \circ \sigma_s^{*})}$,
    as required.
  \end{description}
  The second part of the lemma statement follows directly by induction
  on the derivation of $\Delta \vdash \Gamma \isCtxt$ and the first
  part. The step case of the induction is very similar to the
  \TirName{TyProd} case above.
\end{proof}

\begin{theorem}[Parametricity]
  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
\end{theorem}

\begin{proof}
  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
  three interesting cases follow:
  \begin{description}
  \item[Case \TirName{TyEq}] This case follows directly from
    \lemref{lem:tyeq-rel}.
  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
    $A = \forall i\mathord:s. A'$. We know from the induction
    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}\rho'$,
    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho'$. We are given $\rho \in
    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
    \rsem{\pi_{i:s}^*\Gamma}\rho''$ by \lemref{lem:tysubst-rel}. So we
    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}\rho''$, and thus
    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
      i.M'}\eta_2) \in \rsem{\forall
      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
    know that for all $\rho' \in \extends{\rho}{i:s}$,
    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho'$. If we let $\rho' = \rho \circ
    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
    by \lemref{lem:tysubst-rel}, conclude that $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
    as required.
  \end{description}
  The remaining cases are standard for proofs of the fundamental lemma
  of logical relations for simply-typed systems.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Consequences of Parametricity}

We now give instances of our semantic framework for each of the three
examples from \autoref{sec:example-systems}. For each of the three
examples, we defined an indexing theory, a collection of primitive
types, and a collection of primitive operations $\Gamma_X$. Below, we
assign the ``obvious'' index-erasure semantics to each of the
primitive types and operations: so we have $\tyPrimSem{X}$ for each
primitive type $\tyPrimNm{X}$, and a $\eta_X \in \ctxtSem{\Gamma_X}$.

The key question we wish to answer below is whether there is a family
of sets of relation environments $\{\relEnv{E}_X(\Delta)\}_{\Delta}$
that satisfies the following completeness property identified by
Kennedy.
\begin{equation}\label{eq:kennedy-completeness}
  (\forall \Delta. \forall \rho \in \relEnv{E}'(\Delta). (\eta_X, \eta_X) \in \rsem{\Gamma_X}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_X
\end{equation}
where the subset relation on families of sets of relation environments
is defined as $\relEnv{E}_1 \subseteq \relEnv{E}_2$ if $\forall
\Delta. \relEnv{E}_1(\Delta) \subseteq \relEnv{E}_2(\Delta)$.

Such a completeness result allows us to derive inexpressibility
results for a given instance of our general framework. Kennedy
demonstrates this for the units of measure system by showing that a
putative square root function of type $\forall u.\ \tyPrim{num}{u^2}
\to \tyPrim{num}{u}$ can only ever be the constantly zero function (or
non-terminating).

\subsection{Units of Measure}\label{sec:units-semantics}

For the units of measure example, we set $\tyPrimSem{num} =
\tyPrimSem{rational} = \mathbb{Q}$, the set of rational numbers. We
define $\eta_{\mathit{units}} \in \Gamma_{\mathit{units}}$ are defined
to be the standard arithmetical and comparison operations on rational
numbers, and $\mathit{toRat}$ and $\mathit{fromRat}$ are defined to be
the identity functions. The division function (of type $/ : \forall
u_1,u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to
\tyPrim{num}{u_1u^{-1}_2} + 1$) returns $\mathrm{inr}(*)$ only when
its second argument is zero.

Kennedy defines the following family of sets of relational
environments. Note that, since we have assumed the abelian group
axioms, the set $\idxTms{\Delta}{\mathsf{unit}}$ of indexing
expressions quotiented by the equational theory, is an abelian group.
\begin{displaymath}
  \relEnv{E}_{\mathit{units}}(\Delta) = \left\{ \rho^{G,h} \sepbar
    \begin{array}{l}
      G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{unit}} \\
      h : G \to \mathbb{Q}^{>0} \\
      h\textrm{ is a group homomorphism}
    \end{array}
  \right\}
\end{displaymath}
where $\mathbb{Q}^{>0} = \{ q \in \mathbb{Q} \sepbar q > 0 \}$ and
\begin{displaymath}
  \rho^{G,h}_{\tyPrimNm{num}}(e) = \left\{
    \begin{array}{ll}
      \{(0,0)\} & \textrm{if }e \not\in G \\
      \{(r,h(e)r) \sepbar r \in \mathbb{Q} \} & \textrm{if }u \in G
    \end{array}
  \right.
\end{displaymath}

\begin{theorem}
  The family of sets of relational environments
  $\relEnv{E}_{\mathit{units}}$ satisfies the completeness property
  \hyperref[eq:kennedy-completeness]{(\ref*{eq:kennedy-completeness})}
  above. That is, for all $\relEnv{E}'$:
  \begin{displaymath}
    (\forall \Delta.\forall \rho \in \relEnv{E}'(\Delta). (\eta_{\mathit{units}}, \eta_{\mathit{units}}) \in \rsem{\Gamma_{\mathit{units}}}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_{\mathit{units}}
  \end{displaymath}
\end{theorem}

\begin{proof}
  (Following Kennedy).  From right to left, we have to check that for
  each $\rho \in \relEnv{E}'(\Delta)$, where $\relEnv{E}' \subseteq
  \relEnv{E}_{\mathit{units}}$, that each of the semantic
  interpretations of the operations satisfies
  $(\eta_{\mathit{units}}(\mathit{op}),
  \eta_{\mathit{units}}(\mathit{op})) \in
  \rsem{\Gamma(\mathit{op})}{\relEnv{E}'}\rho$. For instance,
  multiplication has to satisfy: for all subgroups $G \subseteq
  \idxTms{\Delta,u_1,u_2}{\mathsf{unit}}$ and all group homomorphisms
  $h : G \to \mathbb{Q}^{>0}$, then $(q_1,q'_1) \in \rho^{G,h}(u_1)$
  and $(q_2,q'_2) \in \rho^{G,h}(u_2)$ implies that $(q_1q_2,q'_1q'_2)
  \in \rho^{G,h}(u_1u_2)$. There are four cases, depending on whether
  $u_1 \in G$ and whether $u_2 \in G$. If $u_1 \not\in G$, then $q_1 =
  q'_1 = 0$, so $q_1q_2 = 0$ and $q'_1q'_2 = 0$. Since $(0,0) \in
  \rho^{G,h}(u')$ for any $u'$, $G$ and $h$, we are done. If $u_1 \in
  G$ and $u_2 \in G$, then we need to show that $(q_1q_2,
  h(u_1)q_1h(u_2)q_2) \in \rho^{G,h}(u_1u_2)$. By $h$ being a
  homomorphism, we have $(q_1q_2, h(u_1)q_1h(u_2)q_2) = (q_1q_1,
  h(u_1u_2)q_1q_2)$, which is in $\rho^{G,h}(u_1u_2)$. The cases for
  the other operations are similar.

  From left to right, we have the following properties of any
  relational environment $\rho_{\tyPrimNm{num}} \in
  \relEnv{E}'(\Delta)$, derived from the hypothesis that all the
  operations in $\eta_{\mathit{units}}$ satisfy their types'
  relational semantics.
  \begin{enumerate}
  \item\label{item:obsv-poly} By the existence of the function
    $\mathit{toRat}$, the relation $\rho_{\tyPrimNm{num}}(1)$ is the
    identity relation.
  \item\label{item:zero-poly} For any unit expression $e$, $(0,0) \in
    \rho_{\tyPrimNm{num}}(e)$.
  \item \label{item:one-poly} $(1,1) \in \rho_{\tyPrimNm{num}}(1)$.
  \item \label{item:mult-poly} For any unit expressions $e_1,e_2$,
    such that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e_1)$ and
    $(q_2,q'_2) \in \rho_{\tyPrimNm{num}}(e_2)$, we have $(q_1q_2,
    q'_1q'_2) \in \rho_{\tyPrimNm{num}}(e_1e_2)$.
  \item\label{item:div-poly} For any unit expressions $e_1,e_2$, such
    that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e_1)$ and $(q_2,q'_2)
    \in \rho_{\tyPrimNm{num}}(e_2)$, we have that either $q_2 = q'_2 =
    0$, or we have that $(q_1/q_2,q'_1,q'_2) \in
    \rho_{\tyPrimNm{num}}(e_1e_2^{-1})$.
  \item\label{item:compare-poly} For any unit expression $e$, such
    that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e)$ and $(q_2,q'_2) \in
    \rho_{\tyPrimNm{num}}(e)$, we have $q_1 < q_2$ iff $q'_1 < q'_2$.
  \end{enumerate}
  We now show that if $(q,q') \in \rho_{\tyPrimNm{num}}(e)$ for some
  unit expression $e$, then either $q = q' = 0$, or $q' = rq$ for some
  $r \in \mathbb{Q}^{>0}$, determined by $e$. We know from
  \ref{item:zero-poly} above that $q = q' = 0$ is a possibility. Now
  consider what happens if $q \not= 0$: in this case, by
  \ref{item:zero-poly} and \ref{item:compare-poly}, we know that $0 <
  q$ iff $0 < q'$, so $q' \not= 0$, and $q'$ has the same sign as
  $q$. So $q/q'$ exists and is in $\mathbb{Q}^{>0}$. Hence $q = rq'$
  for $r = q/q'$. So now we know that if $(q,q') \in
  \rho_{\tyPrimNm{num}}(e)$, and $q \not= 0$ or $q' \not= 0$ then
  there exists an $r$ such that $q = rq'$.

  We now show that the unit expression $e$ determines the value of
  $r$. Assume that we have $(q_1,r_1q_1) \in \rho_{\tyPrimNm{num}}(e)$
  and $(q_2,r_2q_2) \in \rho_{\tyPrimNm{num}}(e)$. By
  \ref{item:div-poly}, we know that $(q_1/q_2, (r_1q_1)/(r_2q_2)) \in
  \rho_{\tyPrimNm{num}}(1)$. By \ref{item:obsv-poly},
  $\rho_{\tyPrimNm{num}}(1)$ is the equality relation, so we can
  deduce that $r_1 = r_2$.

  Now let $G$ be the subgroup of $\idxTms{\Delta}{\mathsf{unit}}$ such
  that $e \in G$ implies that $\rho_{\tyPrimNm{num}}(e) \not=
  \{(0,0)\}$. By the argument above, for each $e \in G$, there is a
  unique $r \in \mathbb{Q}^{>0}$ such that $\rho_{\tyPrimNm{num}}(e) =
  \{ (q,rq) \sepbar q \in \mathbb{Q} \}$. Thus there is a function $h
  : G \to \mathbb{Q}^{>0}$. By \ref{item:one-poly} and
  \ref{item:mult-poly}, $h$ is a group homomorphism. Thus we have
  shown that $\rho_{\tyPrimNm{num}}$ has the form of
  $\rho^{G,h}_{\tyPrimNm{num}}$ for some $G$ and $h$. Since
  $\rho_{\tyPrimNm{num}}$ was an arbitrary element of
  $\relEnv{E}'(\Delta)$, and $\Delta$ was arbitrary, we have shown
  that $\relEnv{E}' \subseteq \relEnv{E}_{\mathit{units}}$.
\end{proof}

Some notes towards a generalisation:
\begin{itemize}
\item By inspecting Kennedy's proof, we can see that the operations of
  addition and subtraction do not play any role, save for an
  obligation to prove that they satisfy the relational semantics of
  their types. This follows just by distributivity of addition over
  multiplication.
\item Kennedy's proof does not rely on any particular features of
  $\mathbb{Q}$ other than the fact that it is an ordered field. By
  inspecting the proof, it is possible to see that if we remove the
  comparison operation $< : \forall u. \tyPrim{num}{u} \to
  \tyPrim{num}{u} \to \tyPrimNm{bool}$ from $\Gamma_{\mathit{units}}$,
  then the appropriate family of sets of relational environments uses
  homomorphisms into the set of non-zero rationals:
  \begin{displaymath}
    \begin{array}{l}
      \relEnv{E}_{\mathit{units}\backslash\{<\}}(\Delta) = \\
      \quad\left\{ \rho^{G,h} \sepbar
        \begin{array}{l}
          G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{unit}} \\
          h : G \to \mathbb{Q}^{\not=0} \\
          h\textrm{ is a group homomorphism}
        \end{array}
      \right\}
    \end{array}
  \end{displaymath}
  To complete the step in the proof where the comparison operation was
  used, we use the division operation to show that if $(q,q') \in
  \rho_{\tyPrimNm{num}}(e)$ for some $e$, and $q \not= 0$, then $q'
  \not= 0$ (but they need not necessarily have the same sign).

  Just as Kennedy's proof generalises to an arbitrary ordered field,
  if we remove the comparison operation we can generalise to an
  arbitrary field as the underlying semantics of $\tyPrimNm{num}$ and
  $\tyPrimNm{rational}$.
\end{itemize}

\subsection{Euclidean Plane Geometry}\label{sec:2d-semantics}

We set $\tyPrimSem{vec} = \mathbb{R}^2$, and $\tyPrimSem{real} =
\mathbb{R}$. The semantic interpretation $\eta_{\mathit{2D}}$ of the
operations in $\Gamma_{\mathit{2D}}$ consists of the standard vector
zero, addition, negation and scaling in $\mathbb{R}^2$, along with
projections to interpret $\mathit{getX}$ and $\mathit{getY}$, and
pairing to interpret $\mathit{vec}$.

Let
\begin{displaymath}
  \begin{array}{l}
    \relEnv{E}_{\mathit{2D}}(\Delta) = \\
    \quad\left\{ \rho^{G,h} \sepbar
      \begin{array}{l}
        G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{translation}} \\
        h : G \to \mathbb{R}^2 \\
        h\textrm{ is a group homomorphism}
      \end{array}
    \right\}
  \end{array}
\end{displaymath}
where
\begin{displaymath}
  \rho^{G,h}_{\tyPrimNm{vec}}(t) = \left\{
    \begin{array}{ll}
      \{\} & \textrm{if }t \not\in G \\
      \{(v,h(t) + v) \sepbar v \in \mathbb{R}^2 \} & \textrm{if }t \in G
    \end{array}
  \right.
\end{displaymath}
Note that, unlike the units of measure example above, the pair $(0,0)$
need not appear in every relation. This is because we do not have a
translation polymorphic constant like $0 : \forall u. \tyPrim{num}{u}$
in the units of measure example.

\begin{theorem}
  The family of sets of relational environments
  $\relEnv{E}_{\mathit{units}}$ satisfies the completeness property
  \hyperref[eq:kennedy-completeness]{(\ref*{eq:kennedy-completeness})}
  above. That is, for all $\relEnv{E}'$:
  \begin{displaymath}
    (\forall \Delta.\forall \rho \in \relEnv{E}'(\Delta). (\eta_{\mathit{2D}}, \eta_{\mathit{2D}}) \in \rsem{\Gamma_{\mathit{2D}}}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_{\mathit{2D}}
  \end{displaymath}
\end{theorem}

\begin{proof}
  From right to left, we must check that each operation satisfies the
  relational interpretation of its type. This is very similar to the
  case for the units of measure example above.

  From left to right, we have the following properties of any
  relational environment $\rho_{\tyPrimNm{vec}} \in
  \relEnv{E}'(\Delta)$, derived from the hypothesis that all the
  operations in $\eta_{\mathit{2D}}$ satisfy their types'
  relational semantics.
  \begin{enumerate}
  \item\label{item:obsv-vec-poly} By the existence of the functions
    $\mathit{getX}$ and $\mathit{getY}$, the relation
    $\rho_{\tyPrimNm{vec}}(0)$ is the identity relation.
  \item \label{item:zero-vec-poly} $(0,0) \in \rho_{\tyPrimNm{vec}}(0)$.
  \item \label{item:add-vec-poly} For any translation expressions $t_1,t_2$,
    such that $(v_1,v'_1) \in \rho_{\tyPrimNm{vec}}(t_1)$ and
    $(v_2,v'_2) \in \rho_{\tyPrimNm{vec}}(t_2)$, we have $(v_1 + v_2,
    v'_1 + v'_2) \in \rho_{\tyPrimNm{vec}}(t_1 + t_2)$.
  \item\label{item:neg-vec-poly} For any translation expression $t$,
    such that $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$, we have that
    $(-v,-v') \in \rho_{\tyPrimNm{vec}}(-t)$.
  \end{enumerate}
  We now show that if $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$ for some
  translation expression $t$, then $v' = w + v$ for some $w \in
  \mathbb{R}^2$, where $w$ is determined by $t$. If we are given
  $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$, then we can set $w = v' -
  v$. This is uniquely determined by $t$: if we have $(v, w + v) \in
  \rho_{\tyPrimNm{vec}}(t)$ and $(v, w' + v) \in
  \rho_{\tyPrimNm{vec}}(t)$, then by \ref{item:add-vec-poly} and
  \ref{item:neg-vec-poly}, we have $(v - v, (w + v) - (w' + v)) \in
  \rho_{\tyPrimNm{vec}}(0)$. By \ref{item:obsv-vec-poly}, we know that
  $\rho_{\tyPrimNm{vec}}(0)$ is the identity relation, so we can
  deduce that $w = w'$.

  Now define $G \subseteq \idxTms{\Delta}{\mathsf{translation}}$ to
  have $t \in G$ if $\rho_{\tyPrimNm{vec}}(t) \not= \{\}$. Since the
  translation factor $w$ is determined by $t$, we can define a
  function $h : G \to \mathbb{R}^2$. By \ref{item:zero-vec-poly} and
  \ref{item:add-vec-poly}, $h$ is a group homomorphism, and $(v,v')
  \in \rho_{\tyPrimNm{vec}}$ iff $v = h(t) + v'$. Thus
  $\rho_{\tyPrimNm{vec}}$ is of the form
  $\rho^{G,h}_{\tyPrimNm{vec}}$, as required.
\end{proof}

Some more notes:
\begin{enumerate}
\item The proof is slightly simpler than the case for the units of
  measure example above because the underlying semantics of
  $\tyPrimNm{vec}$ forms an abelian group, so we can just use inverses
  to determine the difference factor $w$, and we don't have to deal
  with the possibility of division by zero.
\item The proof above should generalise to any abelian group.
\item Generalising to non-abelian groups does not look
  straightforward: the step that shows that the translation factor $w$
  is uniquely determined by the translation expression $t$ fails. We
  end up with
  \begin{displaymath}
    0 = w + v + (-w') + (-v)
  \end{displaymath}
  from which it is not possible to conclude that $w = w'$. (Maybe
  something using normal subgroups would work?)
\item A slightly different example is obtained by considering
  rotations rather than translations (we consider rotations as an
  additive abelian group). In this case, we could have the following
  typed operations:
  \begin{eqnarray*}
    0 & : & \forall r.\ \tyPrim{vec}{r} \\
    + & : & \forall r.\ \tyPrim{vec}{r} \to \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
    - & : & \forall r.\ \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
    * & : & \forall r.\ \tyPrimNm{real} \to \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
    \mathit{norm} & : & \forall r. \tyPrim{vec}{r} \to \tyPrimNm{real} \\
    \mathit{angleOf} & : & \forall r. \tyPrim{vec}{r} \to \tyPrim{rot}{r} + 1 \\
    \mathit{rotateBy} & : & \forall r_1 r_2. \tyPrim{vec}{r_1} \to \tyPrim{rot}{r_2} \to \tyPrim{vec}{r_1 + r_2} \\
    0_{\mathit{rot}} & : & \tyPrim{rot}{0} \\
    +_{\mathit{rot}} & : & \forall r_1 r_2. \tyPrim{rot}{r_1} \to \tyPrim{rot}{r_2} \to \tyPrim{rot}{r_1 + r_2} \\
    -_{\mathit{rot}} & : & \forall r. \tyPrim{rot}{r} \to \tyPrim{rot}{-r}
  \end{eqnarray*}
  where $\tyPrimSem{vec} = \mathbb{R}^2$, as before, and
  $\tyPrimSem{rot} = \mathrm{SO}(2)$, the group of rotations of the
  plane. The operation $\mathit{angleOf}$ is intended to return
  $\mathrm{inr}(*)$ when the provided vector is zero.

  In this case, rotations form an abelian group, so we can re-use the
  proof above to give a family of sets of relational environments that
  is complete for the group operations on rotations. The fact that
  rotations are linear admits the typing of the first four
  operations. The existence of the $\mathit{angleOf}$ and
  $\mathit{rotateBy}$ functions allow us to determine ``rotation
  factors'' between arbitrary pairs of vectors.

  I conjecture that the following definition of
  $\relEnv{E}_{\mathit{2Drot}}$ satisfies the completeness criterion,
  with the standard interpretation of the operations above.
  \begin{displaymath}
    \begin{array}{l}
      \relEnv{E}_{\mathit{2Drot}}(\Delta) = \\
      \quad\left\{ \rho^{G,h} \sepbar
        \begin{array}{l}
          G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{rotation}} \\
          h : G \to \mathrm{SO}(2) \\
          h\textrm{ is a group homomorphism}
        \end{array}
      \right\}
    \end{array}
  \end{displaymath}
  where
  \begin{displaymath}
    \rho^{G,h}_{\tyPrimNm{vec}}(r) = \left\{
      \begin{array}{ll}
        \{(0,0)\} & \textrm{if }t \not\in G \\
        \{(v,h(r) v) \sepbar v \in \mathbb{R}^2 \} & \textrm{if }t \in G
      \end{array}
    \right.
  \end{displaymath}
  and
  \begin{displaymath}
    \rho^{G,h}_{\tyPrimNm{rot}}(r) = \left\{
      \begin{array}{ll}
        \{\} & \textrm{if }t \not\in G \\
        \{(s,h(r) s) \sepbar s \in \mathrm{SO}(2) \} & \textrm{if }t \in G
      \end{array}
    \right.
  \end{displaymath}
\end{enumerate}

\subsection{Security Typing}\label{sec:sec-semantics}

In this case, we let $\tyPrimSem{T} = \{*\}$, and the operations
$\mathit{proj}$ and $\mathit{combine}$ are given the only possible
interpretations.

I conjecture that the follow definition works:
\begin{displaymath}
  \begin{array}{l}
    \relEnv{E}_{\mathit{sec}}(\Delta) = \\
    \quad\left\{ \rho^{h} \sepbar
      \begin{array}{l}
        h : \idxTms{\Delta}{\mathsf{principal}} \to \{\top,\bot\} \\
        h\textrm{ is a boolean algebra homomorphism}
      \end{array}
    \right\}
  \end{array}
\end{displaymath}
where
\begin{displaymath}
  \rho^{h}_{\tyPrimNm{T}}(p) = \left\{
    \begin{array}{ll}
      \{\} & \textrm{if }h(p) = \bot \\
      \{(*,*) \} & \textrm{if }h(p) = \top
    \end{array}
  \right.
\end{displaymath}

One can compare this definition to the previous ones by considering
the boolean algebra homomorphism as defining a sub-boolean algebra of
$\idxTms{\Delta}{\mathsf{principal}}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Extensions}

\subsection{Inductive Types}

\subsection{Type Operators}

\subsection{Size indexing and Combinatorial Species}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plainnat}
\bibliography{paper}

\end{document}

### 2012-05-08 paper/paper.tex
Add Andrew to author list, and alphabetise the authors

@@
             pdftitle=Relational Parametricity for Algebraically-Indexed Types,
-            pdfauthor={Robert Atkey, Patricia Johann, Neil Ghani},
+            pdfauthor={Robert Atkey, Neil Ghani, Patricia Johann, Andrew Kennedy},
             pdfkeywords={}}  
@@
 
-\authorinfo{Robert Atkey\and Patricia Johann\and Neil Ghani}
+\authorinfo{Robert Atkey\and Neil Ghani\\\and Patricia Johann}
            {University of Strathclyde}
-           {\{Robert.Atkey,Patricia.Johann,Neil.Ghani\}@cis.strath.ac.uk}
+           {\{Robert.Atkey,Neil.Ghani,Patricia.Johann\}@cis.strath.ac.uk}
+
+\authorinfo{Andrew Kennedy}
+           {Microsoft Research}
+           {}
 

### 2012-05-17 paper/paper.tex
Move proofs of the lemmas to an appendix

@@
 
+Moreover, the index-erasure semantics is insensitive to index
+substitutions.
+\begin{lemma}\label{lem:tysubst-erasure}
+  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
+  \Delta$, then $\tySem{\sigma^*A} = \tySem{A}$.
+\end{lemma}
+
+For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
+index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
+structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
+\ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
+one-element set.
+
+\subsubsection{Index-Erasure Semantics of Terms}
+
+For a well-formed term $\Delta; \Gamma \vdash e : A$, we define a
+function $\tmSem{e} : \ctxtSem{\Gamma} \to \tySem{A}$, completely
+ignoring the indexing information. In light of
+\lemref{lem:tyeq-erasure} and \lemref{lem:tysubst-erasure}, we may
+define this function directly on the syntax of well-typed terms,
+rather than on the typing derivations. We define $\tmSem{e}$ by the
+following clauses:
+\begin{eqnarray*}
+  \tmSem{x}\eta & = & \eta_x \\
+  \tmSem{(M,N)}\eta & = & (\tmSem{M}\eta, \tmSem{N}\eta) \\
+  \tmSem{\pi_1M}\eta & = & \pi_1(\tmSem{M}\eta) \\
+  \tmSem{\pi_2M}\eta & = & \pi_2(\tmSem{M}\eta) \\
+  \tmSem{\mathrm{inl}\ M}\eta & = & \mathrm{inl}\ (\tmSem{M}\eta) \\
+  \tmSem{\mathrm{inr}\ M}\eta & = & \mathrm{inr}\ (\tmSem{M}\eta) \\
+  \left\llbracket
+    \begin{array}{l}
+      \textrm{case}\ M\ \textrm{of}\\
+      \textrm{inl}\ x.N_1;\\
+      \textrm{inr}\ y.N_2
+    \end{array}\right\rrbracket^{\mathit{tm}}\eta & = &
+  \left\{
+    \begin{array}{ll}
+      \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
+      \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
+    \end{array}
+  \right. \\
+  \tmSem{\lambda x. M}\eta & = & \lambda v. \tmSem{M}(\eta, v) \\
+  \tmSem{M N}\eta & = & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
+  \tmSem{\Lambda i.\ M}\eta & = & \tmSem{M}\eta \\
+  \tmSem{M[e]}\eta & = & \tmSem{M}\eta
+\end{eqnarray*}
+In the above, we have written $\eta_x$ to denote the appropriate
+projection from $\eta$ to get the position corresponding to the
+variable $x$ in the context.
+
+\subsection{Relational Semantics}
+
+\subsubsection{Relation Environments}
+
+For a set $X$, let $\Rel(X)$ be the set of binary relations on
+$X$. For an index context $\Delta$, a \emph{relation environment} is a
+family of functions $\rho = \{ \rho_{\tyPrimNm{X}} :
+\idxTms{\Delta}{s} \to \Rel(\tyPrimSem{X}) \}_{s \in \mathit{Sort},
+  \tyPrimNm{X} \in \mathit{PrimType}(s)}$.
+
+We define our relational semantics of types with respect to a given
+collection of relation environments: we assume that, for each index
+context $\Delta$, we are given a set of relation environments
+$\relEnv{E}(\Delta)$. This family must satisfy the following two
+properties:
+\begin{enumerate}
+\item Closure under re-indexing by index context morphisms: if $\rho
+  \in \relEnv{E}(\Delta)$ and $\sigma : \Delta' \Rightarrow \Delta$,
+  then $\rho \circ \sigma^* \in \relEnv{E}(\Delta')$.
+\item If we have a pair of relation environments $\rho_1 \in
+  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
+  \relEnv{E}(\Delta)$, along with an index context morphism $\sigma :
+  \Delta' \Rightarrow \Delta$, such that the outer edges of the
+  following diagram commute, for all primitive types $\tyPrimNm{X} \in
+  \mathit{PrimType}(s)$:
+  \begin{displaymath}
+    \xymatrix{
+      {\idxTms{\Delta}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar[d]_{\sigma^*}
+      &
+      {\idxTms{\Delta,i\mathord:s'}{s}} \ar[d]^{\sigma_{s'}^*} \ar@/^/[rdd]^{\rho_1}
+      \\
+      {\idxTms{\Delta'}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2}
+      &
+      {\idxTms{\Delta',i\mathord:s'}{s}} \ar@{.>}[dr]_\rho
+      \\
+      &
+      &
+      {\Rel(X)}
+    }
+  \end{displaymath}
+  Then there exists a relation environment $\rho \in
+  \relEnv{E}(\Delta,i\mathord:s')$ such that the two triangles in the
+  bottom right of the diagram commute.
+\end{enumerate}
+
+Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
+the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
+additional index variable $i\mathord:s$ to be the following:
+\begin{displaymath}
+  \extends{\rho}{i\mathord:s} = \{ \rho' \in \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s} = \rho \}
+\end{displaymath}
+
+\subsubsection{Relational Semantics of Types}
+
+%FIXME: re-word this
+A given relation context $\rho \in \relEnv{E}(\Delta)$ is extended to
+a relation interpretation of any type $\Delta \vdash A : \sortType$ by
+induction its derivation. So we define a relation
+$\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-formed
+type $A$.
+\begin{eqnarray*}
+  \rsem{\tyUnit}{\relEnv{E}}\rho & = & \{(*,*)\} \\
+  \rsem{\tyPrim{X}{e}}{\relEnv{E}}\rho & = & \rho_{\tyPrimNm{X}}(e) \\
+  \rsem{A \tyArr B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
+  \rsem{A \tyProduct B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
+  \rsem{A + B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
+  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & = & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i:s} \}
+\end{eqnarray*}
+In this definition we have made use of the following two constructions
+on binary relations. If $R \in \Rel(X)$ and $S \in \Rel(Y)$, then $R
+\relArrow S \in \Rel(X \to Y)$ is defined as $\{ (f_1,f_2) \sepbar
+\forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S \}$. With the same
+assumptions on $R$ and $S$, the relation $R \relTimes S \in \Rel(X
+\times Y)$ is defined as $\{ ((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2)
+\in R \land (b_1,b_2) \in S \}$. FIXME: do $\relSum$.
+
+We extend the relational semantics to contexts of well-formed types by
+the following clauses:
+\begin{eqnarray*}
+  \rsem{\epsilon}{\relEnv{E}}{\rho} & = & \id_{\ctxtSem{\epsilon}} \\
+  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & = & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
+\end{eqnarray*}
+
+\subsubsection{Parametricity}
+
+This relies on \lemref{lem:tyeq-erasure} to make sure that it is
+well-typed.
+
+\begin{lemma}\label{lem:tyeq-rel}
+  If $\Delta \vdash A \equiv B : \sortType$, then for all $\rho \in
+  \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}\rho = \rsem{B}{\relEnv{E}}\rho$.
+\end{lemma}
+
+This next lemma relies on \lemref{lem:tysubst-erasure} to make sure
+that it is well-typed.
+
+\begin{lemma}\label{lem:tysubst-rel}
+  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
+  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
+  $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Likewise,
+  if $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
+  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
+  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho = \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
+\end{lemma}
+
+\begin{theorem}[Parametricity]
+  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
+  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
+  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
+  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
+\end{theorem}
+
 \begin{proof}
+  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
+  three interesting cases follow:
+  \begin{description}
+  \item[Case \TirName{TyEq}] This case follows directly from
+    \lemref{lem:tyeq-rel}.
+  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
+    $A = \forall i\mathord:s. A'$. We know from the induction
+    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
+    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}\rho'$,
+    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
+    \rsem{A'}{\relEnv{E}}\rho'$. We are given $\rho \in
+    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
+    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
+    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
+    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
+    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
+      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
+    \rsem{\pi_{i:s}^*\Gamma}\rho''$ by \lemref{lem:tysubst-rel}. So we
+    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
+    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}\rho''$, and thus
+    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
+    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
+      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
+      i.M'}\eta_2) \in \rsem{\forall
+      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
+  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
+    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
+    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
+    know that for all $\rho' \in \extends{\rho}{i:s}$,
+    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
+    \rsem{A'}{\relEnv{E}}\rho'$. If we let $\rho' = \rho \circ
+    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
+    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
+    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
+    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
+    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
+    by \lemref{lem:tysubst-rel}, conclude that $(\tmSem{M'}\eta_1,
+    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
+    as required.
+  \end{description}
+  The remaining cases are standard for proofs of the fundamental lemma
+  of logical relations for simply-typed systems.
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Consequences of Parametricity}
+
+We now give instances of our semantic framework for each of the three
+examples from \autoref{sec:example-systems}. For each of the three
+examples, we defined an indexing theory, a collection of primitive
+types, and a collection of primitive operations $\Gamma_X$. Below, we
+assign the ``obvious'' index-erasure semantics to each of the
+primitive types and operations: so we have $\tyPrimSem{X}$ for each
+primitive type $\tyPrimNm{X}$, and a $\eta_X \in \ctxtSem{\Gamma_X}$.
+
+The key question we wish to answer below is whether there is a family
+of sets of relation environments $\{\relEnv{E}_X(\Delta)\}_{\Delta}$
+that satisfies the following completeness property identified by
+Kennedy.
+\begin{equation}\label{eq:kennedy-completeness}
+  (\forall \Delta. \forall \rho \in \relEnv{E}'(\Delta). (\eta_X, \eta_X) \in \rsem{\Gamma_X}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_X
+\end{equation}
+where the subset relation on families of sets of relation environments
+is defined as $\relEnv{E}_1 \subseteq \relEnv{E}_2$ if $\forall
+\Delta. \relEnv{E}_1(\Delta) \subseteq \relEnv{E}_2(\Delta)$.
+
+Such a completeness result allows us to derive inexpressibility
+results for a given instance of our general framework. Kennedy
+demonstrates this for the units of measure system by showing that a
+putative square root function of type $\forall u.\ \tyPrim{num}{u^2}
+\to \tyPrim{num}{u}$ can only ever be the constantly zero function (or
+non-terminating).
+
+\subsection{Units of Measure}\label{sec:units-semantics}
+
+For the units of measure example, we set $\tyPrimSem{num} =
+\tyPrimSem{rational} = \mathbb{Q}$, the set of rational numbers. We
+define $\eta_{\mathit{units}} \in \Gamma_{\mathit{units}}$ are defined
+to be the standard arithmetical and comparison operations on rational
+numbers, and $\mathit{toRat}$ and $\mathit{fromRat}$ are defined to be
+the identity functions. The division function (of type $/ : \forall
+u_1,u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to
+\tyPrim{num}{u_1u^{-1}_2} + 1$) returns $\mathrm{inr}(*)$ only when
+its second argument is zero.
+
+Kennedy defines the following family of sets of relational
+environments. Note that, since we have assumed the abelian group
+axioms, the set $\idxTms{\Delta}{\mathsf{unit}}$ of indexing
+expressions quotiented by the equational theory, is an abelian group.
+\begin{displaymath}
+  \relEnv{E}_{\mathit{units}}(\Delta) = \left\{ \rho^{G,h} \sepbar
+    \begin{array}{l}
+      G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{unit}} \\
+      h : G \to \mathbb{Q}^{>0} \\
+      h\textrm{ is a group homomorphism}
+    \end{array}
+  \right\}
+\end{displaymath}
+where $\mathbb{Q}^{>0} = \{ q \in \mathbb{Q} \sepbar q > 0 \}$ and
+\begin{displaymath}
+  \rho^{G,h}_{\tyPrimNm{num}}(e) = \left\{
+    \begin{array}{ll}
+      \{(0,0)\} & \textrm{if }e \not\in G \\
+      \{(r,h(e)r) \sepbar r \in \mathbb{Q} \} & \textrm{if }u \in G
+    \end{array}
+  \right.
+\end{displaymath}
+
+\begin{theorem}
+  The family of sets of relational environments
+  $\relEnv{E}_{\mathit{units}}$ satisfies the completeness property
+  \hyperref[eq:kennedy-completeness]{(\ref*{eq:kennedy-completeness})}
+  above. That is, for all $\relEnv{E}'$:
+  \begin{displaymath}
+    (\forall \Delta.\forall \rho \in \relEnv{E}'(\Delta).
+(\eta_{\mathit{units}}, \eta_{\mathit{units}}) \in \rsem{\Gamma_{\mathit{units}}}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_{\mathit{units}}
+  \end{displaymath}
+\end{theorem}
+
+\begin{proof}
+  (Following Kennedy).  From right to left, we have to check that for
+  each $\rho \in \relEnv{E}'(\Delta)$, where $\relEnv{E}' \subseteq
+  \relEnv{E}_{\mathit{units}}$, that each of the semantic
+  interpretations of the operations satisfies
+  $(\eta_{\mathit{units}}(\mathit{op}),
+  \eta_{\mathit{units}}(\mathit{op})) \in
+  \rsem{\Gamma(\mathit{op})}{\relEnv{E}'}\rho$. For instance,
+  multiplication has to satisfy: for all subgroups $G \subseteq
+  \idxTms{\Delta,u_1,u_2}{\mathsf{unit}}$ and all group homomorphisms
+  $h : G \to \mathbb{Q}^{>0}$, then $(q_1,q'_1) \in \rho^{G,h}(u_1)$
+  and $(q_2,q'_2) \in \rho^{G,h}(u_2)$ implies that $(q_1q_2,q'_1q'_2)
+  \in \rho^{G,h}(u_1u_2)$. There are four cases, depending on whether
+  $u_1 \in G$ and whether $u_2 \in G$. If $u_1 \not\in G$, then $q_1 =
+  q'_1 = 0$, so $q_1q_2 = 0$ and $q'_1q'_2 = 0$. Since $(0,0) \in
+  \rho^{G,h}(u')$ for any $u'$, $G$ and $h$, we are done. If $u_1 \in
+  G$ and $u_2 \in G$, then we need to show that $(q_1q_2,
+  h(u_1)q_1h(u_2)q_2) \in \rho^{G,h}(u_1u_2)$. By $h$ being a
+  homomorphism, we have $(q_1q_2, h(u_1)q_1h(u_2)q_2) = (q_1q_1,
+  h(u_1u_2)q_1q_2)$, which is in $\rho^{G,h}(u_1u_2)$. The cases for
+  the other operations are similar.
+
+  From left to right, we have the following properties of any
+  relational environment $\rho_{\tyPrimNm{num}} \in
+  \relEnv{E}'(\Delta)$, derived from the hypothesis that all the
+  operations in $\eta_{\mathit{units}}$ satisfy their types'
+  relational semantics.
+  \begin{enumerate}
+  \item\label{item:obsv-poly} By the existence of the function
+    $\mathit{toRat}$, the relation $\rho_{\tyPrimNm{num}}(1)$ is the
+    identity relation.
+  \item\label{item:zero-poly} For any unit expression $e$, $(0,0) \in
+    \rho_{\tyPrimNm{num}}(e)$.
+  \item \label{item:one-poly} $(1,1) \in \rho_{\tyPrimNm{num}}(1)$.
+  \item \label{item:mult-poly} For any unit expressions $e_1,e_2$,
+    such that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e_1)$ and
+    $(q_2,q'_2) \in \rho_{\tyPrimNm{num}}(e_2)$, we have $(q_1q_2,
+    q'_1q'_2) \in \rho_{\tyPrimNm{num}}(e_1e_2)$.
+  \item\label{item:div-poly} For any unit expressions $e_1,e_2$, such
+    that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e_1)$ and $(q_2,q'_2)
+    \in \rho_{\tyPrimNm{num}}(e_2)$, we have that either $q_2 = q'_2 =
+    0$, or we have that $(q_1/q_2,q'_1,q'_2) \in
+    \rho_{\tyPrimNm{num}}(e_1e_2^{-1})$.
+  \item\label{item:compare-poly} For any unit expression $e$, such
+    that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e)$ and $(q_2,q'_2) \in
+    \rho_{\tyPrimNm{num}}(e)$, we have $q_1 < q_2$ iff $q'_1 < q'_2$.
+  \end{enumerate}
+  We now show that if $(q,q') \in \rho_{\tyPrimNm{num}}(e)$ for some
+  unit expression $e$, then either $q = q' = 0$, or $q' = rq$ for some
+  $r \in \mathbb{Q}^{>0}$, determined by $e$. We know from
+  \ref{item:zero-poly} above that $q = q' = 0$ is a possibility. Now
+  consider what happens if $q \not= 0$: in this case, by
+  \ref{item:zero-poly} and \ref{item:compare-poly}, we know that $0 <
+  q$ iff $0 < q'$, so $q' \not= 0$, and $q'$ has the same sign as
+  $q$. So $q/q'$ exists and is in $\mathbb{Q}^{>0}$. Hence $q = rq'$
+  for $r = q/q'$. So now we know that if $(q,q') \in
+  \rho_{\tyPrimNm{num}}(e)$, and $q \not= 0$ or $q' \not= 0$ then
+  there exists an $r$ such that $q = rq'$.
+
+  We now show that the unit expression $e$ determines the value of
+  $r$. Assume that we have $(q_1,r_1q_1) \in \rho_{\tyPrimNm{num}}(e)$
+  and $(q_2,r_2q_2) \in \rho_{\tyPrimNm{num}}(e)$. By
+  \ref{item:div-poly}, we know that $(q_1/q_2, (r_1q_1)/(r_2q_2)) \in
+  \rho_{\tyPrimNm{num}}(1)$. By \ref{item:obsv-poly},
+  $\rho_{\tyPrimNm{num}}(1)$ is the equality relation, so we can
+  deduce that $r_1 = r_2$.
+
+  Now let $G$ be the subgroup of $\idxTms{\Delta}{\mathsf{unit}}$ such
+  that $e \in G$ implies that $\rho_{\tyPrimNm{num}}(e) \not=
+  \{(0,0)\}$. By the argument above, for each $e \in G$, there is a
+  unique $r \in \mathbb{Q}^{>0}$ such that $\rho_{\tyPrimNm{num}}(e) =
+  \{ (q,rq) \sepbar q \in \mathbb{Q} \}$. Thus there is a function $h
+  : G \to \mathbb{Q}^{>0}$. By \ref{item:one-poly} and
+  \ref{item:mult-poly}, $h$ is a group homomorphism. Thus we have
+  shown that $\rho_{\tyPrimNm{num}}$ has the form of
+  $\rho^{G,h}_{\tyPrimNm{num}}$ for some $G$ and $h$. Since
+  $\rho_{\tyPrimNm{num}}$ was an arbitrary element of
+  $\relEnv{E}'(\Delta)$, and $\Delta$ was arbitrary, we have shown
+  that $\relEnv{E}' \subseteq \relEnv{E}_{\mathit{units}}$.
+\end{proof}
+
+Some notes towards a generalisation:
+\begin{itemize}
+\item By inspecting Kennedy's proof, we can see that the operations of
+  addition and subtraction do not play any role, save for an
+  obligation to prove that they satisfy the relational semantics of
+  their types. This follows just by distributivity of addition over
+  multiplication.
+\item Kennedy's proof does not rely on any particular features of
+  $\mathbb{Q}$ other than the fact that it is an ordered field. By
+  inspecting the proof, it is possible to see that if we remove the
+  comparison operation $< : \forall u. \tyPrim{num}{u} \to
+  \tyPrim{num}{u} \to \tyPrimNm{bool}$ from $\Gamma_{\mathit{units}}$,
+  then the appropriate family of sets of relational environments uses
+  homomorphisms into the set of non-zero rationals:
+  \begin{displaymath}
+    \begin{array}{l}
+      \relEnv{E}_{\mathit{units}\backslash\{<\}}(\Delta) = \\
+      \quad\left\{ \rho^{G,h} \sepbar
+        \begin{array}{l}
+          G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{unit}} \\
+          h : G \to \mathbb{Q}^{\not=0} \\
+          h\textrm{ is a group homomorphism}
+        \end{array}
+      \right\}
+    \end{array}
+  \end{displaymath}
+  To complete the step in the proof where the comparison operation was
+  used, we use the division operation to show that if $(q,q') \in
+  \rho_{\tyPrimNm{num}}(e)$ for some $e$, and $q \not= 0$, then $q'
+  \not= 0$ (but they need not necessarily have the same sign).
+
+  Just as Kennedy's proof generalises to an arbitrary ordered field,
+  if we remove the comparison operation we can generalise to an
+  arbitrary field as the underlying semantics of $\tyPrimNm{num}$ and
+  $\tyPrimNm{rational}$.
+\end{itemize}
+
+\subsection{Euclidean Plane Geometry}\label{sec:2d-semantics}
+
+We set $\tyPrimSem{vec} = \mathbb{R}^2$, and $\tyPrimSem{real} =
+\mathbb{R}$. The semantic interpretation $\eta_{\mathit{2D}}$ of the
+operations in $\Gamma_{\mathit{2D}}$ consists of the standard vector
+zero, addition, negation and scaling in $\mathbb{R}^2$, along with
+projections to interpret $\mathit{getX}$ and $\mathit{getY}$, and
+pairing to interpret $\mathit{vec}$.
+
+Let
+\begin{displaymath}
+  \begin{array}{l}
+    \relEnv{E}_{\mathit{2D}}(\Delta) = \\
+    \quad\left\{ \rho^{G,h} \sepbar
+      \begin{array}{l}
+        G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{translation}} \\
+        h : G \to \mathbb{R}^2 \\
+        h\textrm{ is a group homomorphism}
+      \end{array}
+    \right\}
+  \end{array}
+\end{displaymath}
+where
+\begin{displaymath}
+  \rho^{G,h}_{\tyPrimNm{vec}}(t) = \left\{
+    \begin{array}{ll}
+      \{\} & \textrm{if }t \not\in G \\
+      \{(v,h(t) + v) \sepbar v \in \mathbb{R}^2 \} & \textrm{if }t \in G
+    \end{array}
+  \right.
+\end{displaymath}
+Note that, unlike the units of measure example above, the pair $(0,0)$
+need not appear in every relation. This is because we do not have a
+translation polymorphic constant like $0 : \forall u. \tyPrim{num}{u}$
+in the units of measure example.
+
+\begin{theorem}
+  The family of sets of relational environments
+  $\relEnv{E}_{\mathit{units}}$ satisfies the completeness property
+  \hyperref[eq:kennedy-completeness]{(\ref*{eq:kennedy-completeness})}
+  above. That is, for all $\relEnv{E}'$:
+  \begin{displaymath}
+    (\forall \Delta.\forall \rho \in \relEnv{E}'(\Delta).
+(\eta_{\mathit{2D}}, \eta_{\mathit{2D}}) \in \rsem{\Gamma_{\mathit{2D}}}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_{\mathit{2D}}
+  \end{displaymath}
+\end{theorem}
+
+\begin{proof}
+  From right to left, we must check that each operation satisfies the
+  relational interpretation of its type. This is very similar to the
+  case for the units of measure example above.
+
+  From left to right, we have the following properties of any
+  relational environment $\rho_{\tyPrimNm{vec}} \in
+  \relEnv{E}'(\Delta)$, derived from the hypothesis that all the
+  operations in $\eta_{\mathit{2D}}$ satisfy their types'
+  relational semantics.
+  \begin{enumerate}
+  \item\label{item:obsv-vec-poly} By the existence of the functions
+    $\mathit{getX}$ and $\mathit{getY}$, the relation
+    $\rho_{\tyPrimNm{vec}}(0)$ is the identity relation.
+  \item \label{item:zero-vec-poly} $(0,0) \in \rho_{\tyPrimNm{vec}}(0)$.
+  \item \label{item:add-vec-poly} For any translation expressions $t_1,t_2$,
+    such that $(v_1,v'_1) \in \rho_{\tyPrimNm{vec}}(t_1)$ and
+    $(v_2,v'_2) \in \rho_{\tyPrimNm{vec}}(t_2)$, we have $(v_1 + v_2,
+    v'_1 + v'_2) \in \rho_{\tyPrimNm{vec}}(t_1 + t_2)$.
+  \item\label{item:neg-vec-poly} For any translation expression $t$,
+    such that $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$, we have that
+    $(-v,-v') \in \rho_{\tyPrimNm{vec}}(-t)$.
+  \end{enumerate}
+  We now show that if $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$ for some
+  translation expression $t$, then $v' = w + v$ for some $w \in
+  \mathbb{R}^2$, where $w$ is determined by $t$. If we are given
+  $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$, then we can set $w = v' -
+  v$. This is uniquely determined by $t$: if we have $(v, w + v) \in
+  \rho_{\tyPrimNm{vec}}(t)$ and $(v, w' + v) \in
+  \rho_{\tyPrimNm{vec}}(t)$, then by \ref{item:add-vec-poly} and
+  \ref{item:neg-vec-poly}, we have $(v - v, (w + v) - (w' + v)) \in
+  \rho_{\tyPrimNm{vec}}(0)$. By \ref{item:obsv-vec-poly}, we know that
+  $\rho_{\tyPrimNm{vec}}(0)$ is the identity relation, so we can
+  deduce that $w = w'$.
+
+  Now define $G \subseteq \idxTms{\Delta}{\mathsf{translation}}$ to
+  have $t \in G$ if $\rho_{\tyPrimNm{vec}}(t) \not= \{\}$. Since the
+  translation factor $w$ is determined by $t$, we can define a
+  function $h : G \to \mathbb{R}^2$. By \ref{item:zero-vec-poly} and
+  \ref{item:add-vec-poly}, $h$ is a group homomorphism, and $(v,v')
+  \in \rho_{\tyPrimNm{vec}}$ iff $v = h(t) + v'$. Thus
+  $\rho_{\tyPrimNm{vec}}$ is of the form
+  $\rho^{G,h}_{\tyPrimNm{vec}}$, as required.
+\end{proof}
+
+Some more notes:
+\begin{enumerate}
+\item The proof is slightly simpler than the case for the units of
+  measure example above because the underlying semantics of
+  $\tyPrimNm{vec}$ forms an abelian group, so we can just use inverses
+  to determine the difference factor $w$, and we don't have to deal
+  with the possibility of division by zero.
+\item The proof above should generalise to any abelian group.
+\item Generalising to non-abelian groups does not look
+  straightforward: the step that shows that the translation factor $w$
+  is uniquely determined by the translation expression $t$ fails. We
+  end up with
+  \begin{displaymath}
+    0 = w + v + (-w') + (-v)
+  \end{displaymath}
+  from which it is not possible to conclude that $w = w'$. (Maybe
+  something using normal subgroups would work?)
+\item A slightly different example is obtained by considering
+  rotations rather than translations (we consider rotations as an
+  additive abelian group). In this case, we could have the following
+  typed operations:
+  \begin{eqnarray*}
+    0 & : & \forall r.\ \tyPrim{vec}{r} \\
+    + & : & \forall r.\ \tyPrim{vec}{r} \to \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
+    - & : & \forall r.\ \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
+    * & : & \forall r.\ \tyPrimNm{real} \to \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
+    \mathit{norm} & : & \forall r. \tyPrim{vec}{r} \to \tyPrimNm{real} \\
+    \mathit{angleOf} & : & \forall r. \tyPrim{vec}{r} \to \tyPrim{rot}{r} + 1 \\
+    \mathit{rotateBy} & : & \forall r_1 r_2. \tyPrim{vec}{r_1} \to \tyPrim{rot}{r_2} \to \tyPrim{vec}{r_1 + r_2} \\
+    0_{\mathit{rot}} & : & \tyPrim{rot}{0} \\
+    +_{\mathit{rot}} & : & \forall r_1 r_2. \tyPrim{rot}{r_1} \to \tyPrim{rot}{r_2} \to \tyPrim{rot}{r_1 + r_2} \\
+    -_{\mathit{rot}} & : & \forall r. \tyPrim{rot}{r} \to \tyPrim{rot}{-r}
+  \end{eqnarray*}
+  where $\tyPrimSem{vec} = \mathbb{R}^2$, as before, and
+  $\tyPrimSem{rot} = \mathrm{SO}(2)$, the group of rotations of the
+  plane. The operation $\mathit{angleOf}$ is intended to return
+  $\mathrm{inr}(*)$ when the provided vector is zero.
+
+  In this case, rotations form an abelian group, so we can re-use the
+  proof above to give a family of sets of relational environments that
+  is complete for the group operations on rotations. The fact that
+  rotations are linear admits the typing of the first four
+  operations. The existence of the $\mathit{angleOf}$ and
+  $\mathit{rotateBy}$ functions allow us to determine ``rotation
+  factors'' between arbitrary pairs of vectors.
+
+  I conjecture that the following definition of
+  $\relEnv{E}_{\mathit{2Drot}}$ satisfies the completeness criterion,
+  with the standard interpretation of the operations above.
+  \begin{displaymath}
+    \begin{array}{l}
+      \relEnv{E}_{\mathit{2Drot}}(\Delta) = \\
+      \quad\left\{ \rho^{G,h} \sepbar
+        \begin{array}{l}
+          G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{rotation}} \\
+          h : G \to \mathrm{SO}(2) \\
+          h\textrm{ is a group homomorphism}
+        \end{array}
+      \right\}
+    \end{array}
+  \end{displaymath}
+  where
+  \begin{displaymath}
+    \rho^{G,h}_{\tyPrimNm{vec}}(r) = \left\{
+      \begin{array}{ll}
+        \{(0,0)\} & \textrm{if }t \not\in G \\
+        \{(v,h(r) v) \sepbar v \in \mathbb{R}^2 \} & \textrm{if }t \in G
+      \end{array}
+    \right.
+  \end{displaymath}
+  and
+  \begin{displaymath}
+    \rho^{G,h}_{\tyPrimNm{rot}}(r) = \left\{
+      \begin{array}{ll}
+        \{\} & \textrm{if }t \not\in G \\
+        \{(s,h(r) s) \sepbar s \in \mathrm{SO}(2) \} & \textrm{if }t \in G
+      \end{array}
+    \right.
+  \end{displaymath}
+\end{enumerate}
+
+\subsection{Security Typing}\label{sec:sec-semantics}
+
+In this case, we let $\tyPrimSem{T} = \{*\}$, and the operations
+$\mathit{proj}$ and $\mathit{combine}$ are given the only possible
+interpretations.
+
+I conjecture that the follow definition works:
+\begin{displaymath}
+  \begin{array}{l}
+    \relEnv{E}_{\mathit{sec}}(\Delta) = \\
+    \quad\left\{ \rho^{h} \sepbar
+      \begin{array}{l}
+        h : \idxTms{\Delta}{\mathsf{principal}} \to \{\top,\bot\} \\
+        h\textrm{ is a boolean algebra homomorphism}
+      \end{array}
+    \right\}
+  \end{array}
+\end{displaymath}
+where
+\begin{displaymath}
+  \rho^{h}_{\tyPrimNm{T}}(p) = \left\{
+    \begin{array}{ll}
+      \{\} & \textrm{if }h(p) = \bot \\
+      \{(*,*) \} & \textrm{if }h(p) = \top
+    \end{array}
+  \right.
+\end{displaymath}
+
+One can compare this definition to the previous ones by considering
+the boolean algebra homomorphism as defining a sub-boolean algebra of
+$\idxTms{\Delta}{\mathsf{principal}}$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Extensions}
+
+\subsection{Inductive Types}
+
+\subsection{Type Operators}
+
+\subsection{Size indexing and Combinatorial Species}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusions}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bibliographystyle{plainnat}
+\bibliography{paper}
+
+\appendix
+
+\section{Details of proofs}
+
+\begin{proof}\textbf{[Of \lemref{lem:tyeq-erasure}]}
   By induction on the derivation of $\Delta \vdash A \equiv B :
@@
 
-Moreover, the index-erasure semantics is insensitive to index
-substitutions.
-\begin{lemma}\label{lem:tysubst-erasure}
-  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then $\tySem{\sigma^*A} = \tySem{A}$.
-\end{lemma}
-
-\begin{proof}
+\begin{proof}\textbf{[Of \lemref{lem:tysubst-erasure}]}
   By induction on the derivation of $\Delta \vdash A : \sortType$.
@@
 
-For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
-index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
-structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
-\ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
-one-element set.
-
-\subsubsection{Index-Erasure Semantics of Terms}
-
-For a well-formed term $\Delta; \Gamma \vdash e : A$, we define a
-function $\tmSem{e} : \ctxtSem{\Gamma} \to \tySem{A}$, completely
-ignoring the indexing information. In light of
-\lemref{lem:tyeq-erasure} and \lemref{lem:tysubst-erasure}, we may
-define this function directly on the syntax of well-typed terms,
-rather than on the typing derivations. We define $\tmSem{e}$ by the
-following clauses:
-\begin{eqnarray*}
-  \tmSem{x}\eta & = & \eta_x \\
-  \tmSem{(M,N)}\eta & = & (\tmSem{M}\eta, \tmSem{N}\eta) \\
-  \tmSem{\pi_1M}\eta & = & \pi_1(\tmSem{M}\eta) \\
-  \tmSem{\pi_2M}\eta & = & \pi_2(\tmSem{M}\eta) \\
-  \tmSem{\mathrm{inl}\ M}\eta & = & \mathrm{inl}\ (\tmSem{M}\eta) \\
-  \tmSem{\mathrm{inr}\ M}\eta & = & \mathrm{inr}\ (\tmSem{M}\eta) \\
-  \left\llbracket
-    \begin{array}{l}
-      \textrm{case}\ M\ \textrm{of}\\
-      \textrm{inl}\ x.N_1;\\
-      \textrm{inr}\ y.N_2
-    \end{array}\right\rrbracket^{\mathit{tm}}\eta & = &
-  \left\{
-    \begin{array}{ll}
-      \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
-      \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
-    \end{array}
-  \right. \\
-  \tmSem{\lambda x. M}\eta & = & \lambda v. \tmSem{M}(\eta, v) \\
-  \tmSem{M N}\eta & = & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
-  \tmSem{\Lambda i.\ M}\eta & = & \tmSem{M}\eta \\
-  \tmSem{M[e]}\eta & = & \tmSem{M}\eta
-\end{eqnarray*}
-In the above, we have written $\eta_x$ to denote the appropriate
-projection from $\eta$ to get the position corresponding to the
-variable $x$ in the context.
-
-\subsection{Relational Semantics}
-
-\subsubsection{Relation Environments}
-
-For a set $X$, let $\Rel(X)$ be the set of binary relations on
-$X$. For an index context $\Delta$, a \emph{relation environment} is a
-family of functions $\rho = \{ \rho_{\tyPrimNm{X}} :
-\idxTms{\Delta}{s} \to \Rel(\tyPrimSem{X}) \}_{s \in \mathit{Sort},
-  \tyPrimNm{X} \in \mathit{PrimType}(s)}$.
-
-We define our relational semantics of types with respect to a given
-collection of relation environments: we assume that, for each index
-context $\Delta$, we are given a set of relation environments
-$\relEnv{E}(\Delta)$. This family must satisfy the following two
-properties:
-\begin{enumerate}
-\item Closure under re-indexing by index context morphisms: if $\rho
-  \in \relEnv{E}(\Delta)$ and $\sigma : \Delta' \Rightarrow \Delta$,
-  then $\rho \circ \sigma^* \in \relEnv{E}(\Delta')$.
-\item If we have a pair of relation environments $\rho_1 \in
-  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
-  \relEnv{E}(\Delta)$, along with an index context morphism $\sigma :
-  \Delta' \Rightarrow \Delta$, such that the outer edges of the
-  following diagram commute, for all primitive types $\tyPrimNm{X} \in
-  \mathit{PrimType}(s)$:
-  \begin{displaymath}
-    \xymatrix{
-      {\idxTms{\Delta}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar[d]_{\sigma^*}
-      &
-      {\idxTms{\Delta,i\mathord:s'}{s}} \ar[d]^{\sigma_{s'}^*} \ar@/^/[rdd]^{\rho_1}
-      \\
-      {\idxTms{\Delta'}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2}
-      &
-      {\idxTms{\Delta',i\mathord:s'}{s}} \ar@{.>}[dr]_\rho
-      \\
-      &
-      &
-      {\Rel(X)}
-    }
-  \end{displaymath}
-  Then there exists a relation environment $\rho \in
-  \relEnv{E}(\Delta,i\mathord:s')$ such that the two triangles in the
-  bottom right of the diagram commute.
-\end{enumerate}
-
-Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
-the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
-additional index variable $i\mathord:s$ to be the following:
-\begin{displaymath}
-  \extends{\rho}{i\mathord:s} = \{ \rho' \in \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s} = \rho \}
-\end{displaymath}
-
-\subsubsection{Relational Semantics of Types}
-
-%FIXME: re-word this
-A given relation context $\rho \in \relEnv{E}(\Delta)$ is extended to
-a relation interpretation of any type $\Delta \vdash A : \sortType$ by
-induction its derivation. So we define a relation
-$\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-formed
-type $A$.
-\begin{eqnarray*}
-  \rsem{\tyUnit}{\relEnv{E}}\rho & = & \{(*,*)\} \\
-  \rsem{\tyPrim{X}{e}}{\relEnv{E}}\rho & = & \rho_{\tyPrimNm{X}}(e) \\
-  \rsem{A \tyArr B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{A \tyProduct B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{A + B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & = & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i:s} \}
-\end{eqnarray*}
-In this definition we have made use of the following two constructions
-on binary relations. If $R \in \Rel(X)$ and $S \in \Rel(Y)$, then $R
-\relArrow S \in \Rel(X \to Y)$ is defined as $\{ (f_1,f_2) \sepbar
-\forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S \}$. With the same
-assumptions on $R$ and $S$, the relation $R \relTimes S \in \Rel(X
-\times Y)$ is defined as $\{ ((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2)
-\in R \land (b_1,b_2) \in S \}$. FIXME: do $\relSum$.
-
-We extend the relational semantics to contexts of well-formed types by
-the following clauses:
-\begin{eqnarray*}
-  \rsem{\epsilon}{\relEnv{E}}{\rho} & = & \id_{\ctxtSem{\epsilon}} \\
-  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & = & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
-\end{eqnarray*}
-
-\subsubsection{Parametricity}
-
-This relies on \lemref{lem:tyeq-erasure} to make sure that it is
-well-typed.
-
-\begin{lemma}\label{lem:tyeq-rel}
-  If $\Delta \vdash A \equiv B : \sortType$, then for all $\rho \in
-  \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}\rho = \rsem{B}{\relEnv{E}}\rho$.
-\end{lemma}
-
-\begin{proof}
+\begin{proof}\textbf{[Of \lemref{lem:tyeq-rel}]}
   By induction on the derivation of $\Delta \vdash A \equiv B :
@@
 
-This next lemma relies on \lemref{lem:tysubst-erasure} to make sure
-that it is well-typed.
-
-\begin{lemma}\label{lem:tysubst-rel}
-  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
-  $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Likewise,
-  if $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
-  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho = \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
-\end{lemma}
-
-\begin{proof}
+\begin{proof}\textbf{[Of \lemref{lem:tysubst-rel}]}
   The first part of the lemma statement is proved by induction on the
@@
 
-\begin{theorem}[Parametricity]
-  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
-  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
-  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
-  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
-\end{theorem}
-
-\begin{proof}
-  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
-  three interesting cases follow:
-  \begin{description}
-  \item[Case \TirName{TyEq}] This case follows directly from
-    \lemref{lem:tyeq-rel}.
-  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
-    $A = \forall i\mathord:s. A'$. We know from the induction
-    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
-    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}\rho'$,
-    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho'$. We are given $\rho \in
-    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
-    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
-    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
-    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
-    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
-      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
-    \rsem{\pi_{i:s}^*\Gamma}\rho''$ by \lemref{lem:tysubst-rel}. So we
-    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}\rho''$, and thus
-    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
-      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
-      i.M'}\eta_2) \in \rsem{\forall
-      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
-    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
-    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
-    know that for all $\rho' \in \extends{\rho}{i:s}$,
-    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho'$. If we let $\rho' = \rho \circ
-    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
-    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
-    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
-    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
-    by \lemref{lem:tysubst-rel}, conclude that $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
-    as required.
-  \end{description}
-  The remaining cases are standard for proofs of the fundamental lemma
-  of logical relations for simply-typed systems.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Consequences of Parametricity}
-
-We now give instances of our semantic framework for each of the three
-examples from \autoref{sec:example-systems}. For each of the three
-examples, we defined an indexing theory, a collection of primitive
-types, and a collection of primitive operations $\Gamma_X$. Below, we
-assign the ``obvious'' index-erasure semantics to each of the
-primitive types and operations: so we have $\tyPrimSem{X}$ for each
-primitive type $\tyPrimNm{X}$, and a $\eta_X \in \ctxtSem{\Gamma_X}$.
-
-The key question we wish to answer below is whether there is a family
-of sets of relation environments $\{\relEnv{E}_X(\Delta)\}_{\Delta}$
-that satisfies the following completeness property identified by
-Kennedy.
-\begin{equation}\label{eq:kennedy-completeness}
-  (\forall \Delta. \forall \rho \in \relEnv{E}'(\Delta). (\eta_X, \eta_X) \in \rsem{\Gamma_X}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_X
-\end{equation}
-where the subset relation on families of sets of relation environments
-is defined as $\relEnv{E}_1 \subseteq \relEnv{E}_2$ if $\forall
-\Delta. \relEnv{E}_1(\Delta) \subseteq \relEnv{E}_2(\Delta)$.
-
-Such a completeness result allows us to derive inexpressibility
-results for a given instance of our general framework. Kennedy
-demonstrates this for the units of measure system by showing that a
-putative square root function of type $\forall u.\ \tyPrim{num}{u^2}
-\to \tyPrim{num}{u}$ can only ever be the constantly zero function (or
-non-terminating).
-
-\subsection{Units of Measure}\label{sec:units-semantics}
-
-For the units of measure example, we set $\tyPrimSem{num} =
-\tyPrimSem{rational} = \mathbb{Q}$, the set of rational numbers. We
-define $\eta_{\mathit{units}} \in \Gamma_{\mathit{units}}$ are defined
-to be the standard arithmetical and comparison operations on rational
-numbers, and $\mathit{toRat}$ and $\mathit{fromRat}$ are defined to be
-the identity functions. The division function (of type $/ : \forall
-u_1,u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to
-\tyPrim{num}{u_1u^{-1}_2} + 1$) returns $\mathrm{inr}(*)$ only when
-its second argument is zero.
-
-Kennedy defines the following family of sets of relational
-environments. Note that, since we have assumed the abelian group
-axioms, the set $\idxTms{\Delta}{\mathsf{unit}}$ of indexing
-expressions quotiented by the equational theory, is an abelian group.
-\begin{displaymath}
-  \relEnv{E}_{\mathit{units}}(\Delta) = \left\{ \rho^{G,h} \sepbar
-    \begin{array}{l}
-      G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{unit}} \\
-      h : G \to \mathbb{Q}^{>0} \\
-      h\textrm{ is a group homomorphism}
-    \end{array}
-  \right\}
-\end{displaymath}
-where $\mathbb{Q}^{>0} = \{ q \in \mathbb{Q} \sepbar q > 0 \}$ and
-\begin{displaymath}
-  \rho^{G,h}_{\tyPrimNm{num}}(e) = \left\{
-    \begin{array}{ll}
-      \{(0,0)\} & \textrm{if }e \not\in G \\
-      \{(r,h(e)r) \sepbar r \in \mathbb{Q} \} & \textrm{if }u \in G
-    \end{array}
-  \right.
-\end{displaymath}
-
-\begin{theorem}
-  The family of sets of relational environments
-  $\relEnv{E}_{\mathit{units}}$ satisfies the completeness property
-  \hyperref[eq:kennedy-completeness]{(\ref*{eq:kennedy-completeness})}
-  above. That is, for all $\relEnv{E}'$:
-  \begin{displaymath}
-    (\forall \Delta.\forall \rho \in \relEnv{E}'(\Delta).
-(\eta_{\mathit{units}}, \eta_{\mathit{units}}) \in \rsem{\Gamma_{\mathit{units}}}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_{\mathit{units}}
-  \end{displaymath}
-\end{theorem}
-
-\begin{proof}
-  (Following Kennedy).  From right to left, we have to check that for
-  each $\rho \in \relEnv{E}'(\Delta)$, where $\relEnv{E}' \subseteq
-  \relEnv{E}_{\mathit{units}}$, that each of the semantic
-  interpretations of the operations satisfies
-  $(\eta_{\mathit{units}}(\mathit{op}),
-  \eta_{\mathit{units}}(\mathit{op})) \in
-  \rsem{\Gamma(\mathit{op})}{\relEnv{E}'}\rho$. For instance,
-  multiplication has to satisfy: for all subgroups $G \subseteq
-  \idxTms{\Delta,u_1,u_2}{\mathsf{unit}}$ and all group homomorphisms
-  $h : G \to \mathbb{Q}^{>0}$, then $(q_1,q'_1) \in \rho^{G,h}(u_1)$
-  and $(q_2,q'_2) \in \rho^{G,h}(u_2)$ implies that $(q_1q_2,q'_1q'_2)
-  \in \rho^{G,h}(u_1u_2)$. There are four cases, depending on whether
-  $u_1 \in G$ and whether $u_2 \in G$. If $u_1 \not\in G$, then $q_1 =
-  q'_1 = 0$, so $q_1q_2 = 0$ and $q'_1q'_2 = 0$. Since $(0,0) \in
-  \rho^{G,h}(u')$ for any $u'$, $G$ and $h$, we are done. If $u_1 \in
-  G$ and $u_2 \in G$, then we need to show that $(q_1q_2,
-  h(u_1)q_1h(u_2)q_2) \in \rho^{G,h}(u_1u_2)$. By $h$ being a
-  homomorphism, we have $(q_1q_2, h(u_1)q_1h(u_2)q_2) = (q_1q_1,
-  h(u_1u_2)q_1q_2)$, which is in $\rho^{G,h}(u_1u_2)$. The cases for
-  the other operations are similar.
-
-  From left to right, we have the following properties of any
-  relational environment $\rho_{\tyPrimNm{num}} \in
-  \relEnv{E}'(\Delta)$, derived from the hypothesis that all the
-  operations in $\eta_{\mathit{units}}$ satisfy their types'
-  relational semantics.
-  \begin{enumerate}
-  \item\label{item:obsv-poly} By the existence of the function
-    $\mathit{toRat}$, the relation $\rho_{\tyPrimNm{num}}(1)$ is the
-    identity relation.
-  \item\label{item:zero-poly} For any unit expression $e$, $(0,0) \in
-    \rho_{\tyPrimNm{num}}(e)$.
-  \item \label{item:one-poly} $(1,1) \in \rho_{\tyPrimNm{num}}(1)$.
-  \item \label{item:mult-poly} For any unit expressions $e_1,e_2$,
-    such that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e_1)$ and
-    $(q_2,q'_2) \in \rho_{\tyPrimNm{num}}(e_2)$, we have $(q_1q_2,
-    q'_1q'_2) \in \rho_{\tyPrimNm{num}}(e_1e_2)$.
-  \item\label{item:div-poly} For any unit expressions $e_1,e_2$, such
-    that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e_1)$ and $(q_2,q'_2)
-    \in \rho_{\tyPrimNm{num}}(e_2)$, we have that either $q_2 = q'_2 =
-    0$, or we have that $(q_1/q_2,q'_1,q'_2) \in
-    \rho_{\tyPrimNm{num}}(e_1e_2^{-1})$.
-  \item\label{item:compare-poly} For any unit expression $e$, such
-    that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e)$ and $(q_2,q'_2) \in
-    \rho_{\tyPrimNm{num}}(e)$, we have $q_1 < q_2$ iff $q'_1 < q'_2$.
-  \end{enumerate}
-  We now show that if $(q,q') \in \rho_{\tyPrimNm{num}}(e)$ for some
-  unit expression $e$, then either $q = q' = 0$, or $q' = rq$ for some
-  $r \in \mathbb{Q}^{>0}$, determined by $e$. We know from
-  \ref{item:zero-poly} above that $q = q' = 0$ is a possibility. Now
-  consider what happens if $q \not= 0$: in this case, by
-  \ref{item:zero-poly} and \ref{item:compare-poly}, we know that $0 <
-  q$ iff $0 < q'$, so $q' \not= 0$, and $q'$ has the same sign as
-  $q$. So $q/q'$ exists and is in $\mathbb{Q}^{>0}$. Hence $q = rq'$
-  for $r = q/q'$. So now we know that if $(q,q') \in
-  \rho_{\tyPrimNm{num}}(e)$, and $q \not= 0$ or $q' \not= 0$ then
-  there exists an $r$ such that $q = rq'$.
-
-  We now show that the unit expression $e$ determines the value of
-  $r$. Assume that we have $(q_1,r_1q_1) \in \rho_{\tyPrimNm{num}}(e)$
-  and $(q_2,r_2q_2) \in \rho_{\tyPrimNm{num}}(e)$. By
-  \ref{item:div-poly}, we know that $(q_1/q_2, (r_1q_1)/(r_2q_2)) \in
-  \rho_{\tyPrimNm{num}}(1)$. By \ref{item:obsv-poly},
-  $\rho_{\tyPrimNm{num}}(1)$ is the equality relation, so we can
-  deduce that $r_1 = r_2$.
-
-  Now let $G$ be the subgroup of $\idxTms{\Delta}{\mathsf{unit}}$ such
-  that $e \in G$ implies that $\rho_{\tyPrimNm{num}}(e) \not=
-  \{(0,0)\}$. By the argument above, for each $e \in G$, there is a
-  unique $r \in \mathbb{Q}^{>0}$ such that $\rho_{\tyPrimNm{num}}(e) =
-  \{ (q,rq) \sepbar q \in \mathbb{Q} \}$. Thus there is a function $h
-  : G \to \mathbb{Q}^{>0}$. By \ref{item:one-poly} and
-  \ref{item:mult-poly}, $h$ is a group homomorphism. Thus we have
-  shown that $\rho_{\tyPrimNm{num}}$ has the form of
-  $\rho^{G,h}_{\tyPrimNm{num}}$ for some $G$ and $h$. Since
-  $\rho_{\tyPrimNm{num}}$ was an arbitrary element of
-  $\relEnv{E}'(\Delta)$, and $\Delta$ was arbitrary, we have shown
-  that $\relEnv{E}' \subseteq \relEnv{E}_{\mathit{units}}$.
-\end{proof}
-
-Some notes towards a generalisation:
-\begin{itemize}
-\item By inspecting Kennedy's proof, we can see that the operations of
-  addition and subtraction do not play any role, save for an
-  obligation to prove that they satisfy the relational semantics of
-  their types. This follows just by distributivity of addition over
-  multiplication.
-\item Kennedy's proof does not rely on any particular features of
-  $\mathbb{Q}$ other than the fact that it is an ordered field. By
-  inspecting the proof, it is possible to see that if we remove the
-  comparison operation $< : \forall u. \tyPrim{num}{u} \to
-  \tyPrim{num}{u} \to \tyPrimNm{bool}$ from $\Gamma_{\mathit{units}}$,
-  then the appropriate family of sets of relational environments uses
-  homomorphisms into the set of non-zero rationals:
-  \begin{displaymath}
-    \begin{array}{l}
-      \relEnv{E}_{\mathit{units}\backslash\{<\}}(\Delta) = \\
-      \quad\left\{ \rho^{G,h} \sepbar
-        \begin{array}{l}
-          G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{unit}} \\
-          h : G \to \mathbb{Q}^{\not=0} \\
-          h\textrm{ is a group homomorphism}
-        \end{array}
-      \right\}
-    \end{array}
-  \end{displaymath}
-  To complete the step in the proof where the comparison operation was
-  used, we use the division operation to show that if $(q,q') \in
-  \rho_{\tyPrimNm{num}}(e)$ for some $e$, and $q \not= 0$, then $q'
-  \not= 0$ (but they need not necessarily have the same sign).
-
-  Just as Kennedy's proof generalises to an arbitrary ordered field,
-  if we remove the comparison operation we can generalise to an
-  arbitrary field as the underlying semantics of $\tyPrimNm{num}$ and
-  $\tyPrimNm{rational}$.
-\end{itemize}
-
-\subsection{Euclidean Plane Geometry}\label{sec:2d-semantics}
-
-We set $\tyPrimSem{vec} = \mathbb{R}^2$, and $\tyPrimSem{real} =
-\mathbb{R}$. The semantic interpretation $\eta_{\mathit{2D}}$ of the
-operations in $\Gamma_{\mathit{2D}}$ consists of the standard vector
-zero, addition, negation and scaling in $\mathbb{R}^2$, along with
-projections to interpret $\mathit{getX}$ and $\mathit{getY}$, and
-pairing to interpret $\mathit{vec}$.
-
-Let
-\begin{displaymath}
-  \begin{array}{l}
-    \relEnv{E}_{\mathit{2D}}(\Delta) = \\
-    \quad\left\{ \rho^{G,h} \sepbar
-      \begin{array}{l}
-        G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{translation}} \\
-        h : G \to \mathbb{R}^2 \\
-        h\textrm{ is a group homomorphism}
-      \end{array}
-    \right\}
-  \end{array}
-\end{displaymath}
-where
-\begin{displaymath}
-  \rho^{G,h}_{\tyPrimNm{vec}}(t) = \left\{
-    \begin{array}{ll}
-      \{\} & \textrm{if }t \not\in G \\
-      \{(v,h(t) + v) \sepbar v \in \mathbb{R}^2 \} & \textrm{if }t \in G
-    \end{array}
-  \right.
-\end{displaymath}
-Note that, unlike the units of measure example above, the pair $(0,0)$
-need not appear in every relation. This is because we do not have a
-translation polymorphic constant like $0 : \forall u. \tyPrim{num}{u}$
-in the units of measure example.
-
-\begin{theorem}
-  The family of sets of relational environments
-  $\relEnv{E}_{\mathit{units}}$ satisfies the completeness property
-  \hyperref[eq:kennedy-completeness]{(\ref*{eq:kennedy-completeness})}
-  above. That is, for all $\relEnv{E}'$:
-  \begin{displaymath}
-    (\forall \Delta.\forall \rho \in \relEnv{E}'(\Delta).
-(\eta_{\mathit{2D}}, \eta_{\mathit{2D}}) \in \rsem{\Gamma_{\mathit{2D}}}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_{\mathit{2D}}
-  \end{displaymath}
-\end{theorem}
-
-\begin{proof}
-  From right to left, we must check that each operation satisfies the
-  relational interpretation of its type. This is very similar to the
-  case for the units of measure example above.
-
-  From left to right, we have the following properties of any
-  relational environment $\rho_{\tyPrimNm{vec}} \in
-  \relEnv{E}'(\Delta)$, derived from the hypothesis that all the
-  operations in $\eta_{\mathit{2D}}$ satisfy their types'
-  relational semantics.
-  \begin{enumerate}
-  \item\label{item:obsv-vec-poly} By the existence of the functions
-    $\mathit{getX}$ and $\mathit{getY}$, the relation
-    $\rho_{\tyPrimNm{vec}}(0)$ is the identity relation.
-  \item \label{item:zero-vec-poly} $(0,0) \in \rho_{\tyPrimNm{vec}}(0)$.
-  \item \label{item:add-vec-poly} For any translation expressions $t_1,t_2$,
-    such that $(v_1,v'_1) \in \rho_{\tyPrimNm{vec}}(t_1)$ and
-    $(v_2,v'_2) \in \rho_{\tyPrimNm{vec}}(t_2)$, we have $(v_1 + v_2,
-    v'_1 + v'_2) \in \rho_{\tyPrimNm{vec}}(t_1 + t_2)$.
-  \item\label{item:neg-vec-poly} For any translation expression $t$,
-    such that $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$, we have that
-    $(-v,-v') \in \rho_{\tyPrimNm{vec}}(-t)$.
-  \end{enumerate}
-  We now show that if $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$ for some
-  translation expression $t$, then $v' = w + v$ for some $w \in
-  \mathbb{R}^2$, where $w$ is determined by $t$. If we are given
-  $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$, then we can set $w = v' -
-  v$. This is uniquely determined by $t$: if we have $(v, w + v) \in
-  \rho_{\tyPrimNm{vec}}(t)$ and $(v, w' + v) \in
-  \rho_{\tyPrimNm{vec}}(t)$, then by \ref{item:add-vec-poly} and
-  \ref{item:neg-vec-poly}, we have $(v - v, (w + v) - (w' + v)) \in
-  \rho_{\tyPrimNm{vec}}(0)$. By \ref{item:obsv-vec-poly}, we know that
-  $\rho_{\tyPrimNm{vec}}(0)$ is the identity relation, so we can
-  deduce that $w = w'$.
-
-  Now define $G \subseteq \idxTms{\Delta}{\mathsf{translation}}$ to
-  have $t \in G$ if $\rho_{\tyPrimNm{vec}}(t) \not= \{\}$. Since the
-  translation factor $w$ is determined by $t$, we can define a
-  function $h : G \to \mathbb{R}^2$. By \ref{item:zero-vec-poly} and
-  \ref{item:add-vec-poly}, $h$ is a group homomorphism, and $(v,v')
-  \in \rho_{\tyPrimNm{vec}}$ iff $v = h(t) + v'$. Thus
-  $\rho_{\tyPrimNm{vec}}$ is of the form
-  $\rho^{G,h}_{\tyPrimNm{vec}}$, as required.
-\end{proof}
-
-Some more notes:
-\begin{enumerate}
-\item The proof is slightly simpler than the case for the units of
-  measure example above because the underlying semantics of
-  $\tyPrimNm{vec}$ forms an abelian group, so we can just use inverses
-  to determine the difference factor $w$, and we don't have to deal
-  with the possibility of division by zero.
-\item The proof above should generalise to any abelian group.
-\item Generalising to non-abelian groups does not look
-  straightforward: the step that shows that the translation factor $w$
-  is uniquely determined by the translation expression $t$ fails. We
-  end up with
-  \begin{displaymath}
-    0 = w + v + (-w') + (-v)
-  \end{displaymath}
-  from which it is not possible to conclude that $w = w'$. (Maybe
-  something using normal subgroups would work?)
-\item A slightly different example is obtained by considering
-  rotations rather than translations (we consider rotations as an
-  additive abelian group). In this case, we could have the following
-  typed operations:
-  \begin{eqnarray*}
-    0 & : & \forall r.\ \tyPrim{vec}{r} \\
-    + & : & \forall r.\ \tyPrim{vec}{r} \to \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
-    - & : & \forall r.\ \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
-    * & : & \forall r.\ \tyPrimNm{real} \to \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
-    \mathit{norm} & : & \forall r. \tyPrim{vec}{r} \to \tyPrimNm{real} \\
-    \mathit{angleOf} & : & \forall r. \tyPrim{vec}{r} \to \tyPrim{rot}{r} + 1 \\
-    \mathit{rotateBy} & : & \forall r_1 r_2. \tyPrim{vec}{r_1} \to \tyPrim{rot}{r_2} \to \tyPrim{vec}{r_1 + r_2} \\
-    0_{\mathit{rot}} & : & \tyPrim{rot}{0} \\
-    +_{\mathit{rot}} & : & \forall r_1 r_2. \tyPrim{rot}{r_1} \to \tyPrim{rot}{r_2} \to \tyPrim{rot}{r_1 + r_2} \\
-    -_{\mathit{rot}} & : & \forall r. \tyPrim{rot}{r} \to \tyPrim{rot}{-r}
-  \end{eqnarray*}
-  where $\tyPrimSem{vec} = \mathbb{R}^2$, as before, and
-  $\tyPrimSem{rot} = \mathrm{SO}(2)$, the group of rotations of the
-  plane. The operation $\mathit{angleOf}$ is intended to return
-  $\mathrm{inr}(*)$ when the provided vector is zero.
-
-  In this case, rotations form an abelian group, so we can re-use the
-  proof above to give a family of sets of relational environments that
-  is complete for the group operations on rotations. The fact that
-  rotations are linear admits the typing of the first four
-  operations. The existence of the $\mathit{angleOf}$ and
-  $\mathit{rotateBy}$ functions allow us to determine ``rotation
-  factors'' between arbitrary pairs of vectors.
-
-  I conjecture that the following definition of
-  $\relEnv{E}_{\mathit{2Drot}}$ satisfies the completeness criterion,
-  with the standard interpretation of the operations above.
-  \begin{displaymath}
-    \begin{array}{l}
-      \relEnv{E}_{\mathit{2Drot}}(\Delta) = \\
-      \quad\left\{ \rho^{G,h} \sepbar
-        \begin{array}{l}
-          G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{rotation}} \\
-          h : G \to \mathrm{SO}(2) \\
-          h\textrm{ is a group homomorphism}
-        \end{array}
-      \right\}
-    \end{array}
-  \end{displaymath}
-  where
-  \begin{displaymath}
-    \rho^{G,h}_{\tyPrimNm{vec}}(r) = \left\{
-      \begin{array}{ll}
-        \{(0,0)\} & \textrm{if }t \not\in G \\
-        \{(v,h(r) v) \sepbar v \in \mathbb{R}^2 \} & \textrm{if }t \in G
-      \end{array}
-    \right.
-  \end{displaymath}
-  and
-  \begin{displaymath}
-    \rho^{G,h}_{\tyPrimNm{rot}}(r) = \left\{
-      \begin{array}{ll}
-        \{\} & \textrm{if }t \not\in G \\
-        \{(s,h(r) s) \sepbar s \in \mathrm{SO}(2) \} & \textrm{if }t \in G
-      \end{array}
-    \right.
-  \end{displaymath}
-\end{enumerate}
-
-\subsection{Security Typing}\label{sec:sec-semantics}
-
-In this case, we let $\tyPrimSem{T} = \{*\}$, and the operations
-$\mathit{proj}$ and $\mathit{combine}$ are given the only possible
-interpretations.
-
-I conjecture that the follow definition works:
-\begin{displaymath}
-  \begin{array}{l}
-    \relEnv{E}_{\mathit{sec}}(\Delta) = \\
-    \quad\left\{ \rho^{h} \sepbar
-      \begin{array}{l}
-        h : \idxTms{\Delta}{\mathsf{principal}} \to \{\top,\bot\} \\
-        h\textrm{ is a boolean algebra homomorphism}
-      \end{array}
-    \right\}
-  \end{array}
-\end{displaymath}
-where
-\begin{displaymath}
-  \rho^{h}_{\tyPrimNm{T}}(p) = \left\{
-    \begin{array}{ll}
-      \{\} & \textrm{if }h(p) = \bot \\
-      \{(*,*) \} & \textrm{if }h(p) = \top
-    \end{array}
-  \right.
-\end{displaymath}
-
-One can compare this definition to the previous ones by considering
-the boolean algebra homomorphism as defining a sub-boolean algebra of
-$\idxTms{\Delta}{\mathsf{principal}}$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Extensions}
-
-\subsection{Inductive Types}
-
-\subsection{Type Operators}
-
-\subsection{Size indexing and Combinatorial Species}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{plainnat}
-\bibliography{paper}
-
 \end{document}

### 2012-05-17 paper/paper.tex
Fill in definition of \relSum

@@
 \end{eqnarray*}
-In this definition we have made use of the following two constructions
-on binary relations. If $R \in \Rel(X)$ and $S \in \Rel(Y)$, then $R
-\relArrow S \in \Rel(X \to Y)$ is defined as $\{ (f_1,f_2) \sepbar
-\forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S \}$. With the same
-assumptions on $R$ and $S$, the relation $R \relTimes S \in \Rel(X
-\times Y)$ is defined as $\{ ((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2)
-\in R \land (b_1,b_2) \in S \}$. FIXME: do $\relSum$.
+In this definition we have made use of the following three
+constructions on binary relations. If $R \in \Rel(X)$ and $S \in
+\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is defined as $\{
+(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
+\}$. With the same assumptions on $R$ and $S$, the relation $R
+\relTimes S \in \Rel(X \times Y)$ is defined as $\{
+((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
+\}$. Finally, the relation $R \relSum S \in \Rel(X + Y)$ is defined as
+$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
+\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.
 

### 2012-05-22 paper/geometry-twocolumn.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

--- 2012-05-08 geometry-notes/invariance.tex
@@
-\documentclass{article}
+\section{Motivating Example: 2D Geometry}
 
-\usepackage[usenames]{color}
-\definecolor{citationcolour}{rgb}{0,0.4,0.2}
-\definecolor{linkcolour}{rgb}{0,0,0.8}
-\usepackage{hyperref}
-\hypersetup{colorlinks=true,
-            urlcolor=linkcolour,
-            linkcolor=linkcolour,
-            citecolor=citationcolour,
-            pdftitle=Geometric Invariance and Relational Parametricity,
-            pdfauthor={Robert Atkey},
-            pdfkeywords={}}
-\def\sectionautorefname{Section}
-\def\subsectionautorefname{Section}
-
-\usepackage{amssymb}
-\usepackage{stmaryrd}
-
-\newcommand{\sepbar}{\mathrel|}
-
-\newcommand{\GA}{\mathrm{GA}}
-\newcommand{\GL}{\mathrm{GL}}
-\newcommand{\Is}{\mathrm{Is}}
-\newcommand{\Orth}{\mathrm{O}}
-\newcommand{\SOrth}{\mathrm{SO}}
-\newcommand{\Transl}{\mathrm{T}}
-\newcommand{\Scale}{\mathrm{Scale}}
-
-\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
-\newcommand{\isDefinedAs}{\stackrel{\mathit{def}}=}
-
-\newcommand{\statementref}[1]{\hyperref[#1]{Statement \ref*{#1}}}
-
-\newenvironment{example}{\vspace{0.1in}\noindent\textbf{Example}\quad}{}
-
-\newcommand{\ignore}[1]{}
-
-\title{Geometric Invariance and Relational Parametricity}
-\author{Robert Atkey}
-
-\begin{document}
-
-\maketitle
-
-\section{Coordinate-free Geometry and Invariance}
+\subsection{Coordinate-free Geometry and Invariance}
 
@@
 \begin{equation}\label{eq:f-preserve-rel-frame}
-  \forall (\vec{v_1},\vec{v'_1}) \in R_{\texttt{pt}}(F), (\vec{v_2},\vec{v'_2}) \in R_{\texttt{pt}}(F).\ (f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\texttt{pt}}(F).
+  \begin{array}{l}
+    \forall (\vec{v_1},\vec{v'_1}) \in R_{\texttt{pt}}(F), (\vec{v_2},\vec{v'_2}) \in R_{\texttt{pt}}(F).\\
+    \hspace{2cm}(f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\texttt{pt}}(F).
+  \end{array}
 \end{equation}
@@
 
-\section{A Type System for Geometric Invariance}
+\subsection{A Type System for Geometric Invariance}
 \label{sec:type-system-geom-intro}
@@
   \begin{array}{l}
-    \mathrm{affineCombination} : \forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \\
-    \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ t\ (x_2,y_2) = ((1-t)x_1 + tx_2, (1-t)y_1 + ty_2)
+    \mathrm{affineCombination} :\\
+    \hspace{1.5cm}\forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \\
+    \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ t\ (x_2,y_2) = \\
+    \hspace{3cm}((1-t)x_1 + tx_2, (1-t)y_1 + ty_2)
   \end{array}
@@
     \begin{array}{l}
-      \mathrm{quadraticB\acute{e}zier} : \forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \\
+      \mathrm{quadraticB\acute{e}zier} :\\
+      \hspace{0.5cm}\forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \\
       \mathrm{quadraticB\acute{e}zier}\ [F]\ p_0\ p_1\ p_2\ t = \\
@@
   \end{displaymath}
-  For two endpoints $p_0$ and $p_2$ and a control point $p_1$,
-  $\mathrm{quadraticB\acute{e}zier}\ p_0\ p_1\ p_2\ t$ for $t \in
-  [0,1]$ gives us the point on the curve at ``time'' $t$. Note that
-  the type of $\mathrm{quadraticB\acute{e}zier}$ tells us immediately
-  that it preserves all invertible affine maps.
+  For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
+  application $\mathrm{quadraticB\acute{e}zier}\ p_0\ p_1\ p_2\ t$ for
+  $t \in [0,1]$ gives us the point on the curve at ``time'' $t$. Note
+  that the type of $\mathrm{quadraticB\acute{e}zier}$ tells us
+  immediately that it preserves all invertible affine maps.
   %FIXME: forward ref to ``free theorem''.
@@
 \begin{displaymath}
-  (+) : \forall B \mathord: \GL(2), t_1, t_2 \mathord: T(2).\ \mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2 \rangle \to \mathsf{vec}\langle B \ltimes (t_1 + t_2)\rangle
+  \begin{array}{l}
+    (+) : \forall B \mathord: \GL(2), t_1, t_2 \mathord: T(2).\ \\
+    \quad\quad\quad\quad\mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2 \rangle \to \mathsf{vec}\langle B \ltimes (t_1 + t_2)\rangle
+\end{array}
 \end{displaymath}
@@
   \begin{array}{l}
-    (-) : \forall B \mathord:\GL(2), t_1,t_2 \mathord:T(2).\ \mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2\rangle \to \mathsf{vec}\langle B \ltimes (t_1 - t_2)\rangle \\
+    (-) : \forall B \mathord:\GL(2), t_1,t_2 \mathord:T(2).\ \\
+    \hspace{1cm}\mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2\rangle \to \mathsf{vec}\langle B \ltimes (t_1 - t_2)\rangle \\
     (-)\ [B]\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
@@
     \begin{array}{l}
-      \mathrm{area} : \forall O\mathord:\Orth(2), t\mathord:\Transl(2).\ \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{real}\langle 1 \rangle \\
+      \mathrm{area} : \forall O\mathord:\Orth(2), t\mathord:\Transl(2).\ \\
+      \hspace{1.2cm}\mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{real}\langle 1 \rangle \\
       \mathrm{area}\ [O]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
@@
 \end{example}
-
-% \subsection{Transformations Within and Between Frames}
-% \label{sec:runtime-rep-frames}
-
-% \begin{enumerate}
-% \item A translation with respect to some frame is simply a vector with
-%   respect to that frame's basis. We can apply it to points just by 
-% \item 
-% \end{enumerate}
-
-% Singleton types
-
-% Representing frames, a type representing translations:
-% \begin{displaymath}
-%   \mathsf{translation}\langle T \rangle
-% \end{displaymath}
-% And representing bases, as linear maps
-% \begin{displaymath}
-%   \mathsf{invLinearMap}\langle B \rangle
-% \end{displaymath}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% FIXME: write down all the groups, and the functions between them
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{plain}
-\bibliography{paper}
-
-\end{document}

### 2012-05-22 paper/old-examples.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

--- 2012-05-08 paper/paper.tex
@@
-\documentclass[natbib,preprint]{sigplanconf}
-
-\usepackage[usenames]{color}
-\definecolor{citationcolour}{rgb}{0,0.4,0.2}
-\definecolor{linkcolour}{rgb}{0,0,0.8}
-\usepackage{hyperref}
-\hypersetup{colorlinks=true,
-            urlcolor=linkcolour,
-            linkcolor=linkcolour,
-            citecolor=citationcolour,
-            pdftitle=Relational Parametricity for Algebraically-Indexed Types,
-            pdfauthor={Robert Atkey, Neil Ghani, Patricia Johann, Andrew Kennedy},
-            pdfkeywords={}}  
-\def\sectionautorefname{Section}
-\def\subsectionautorefname{Section}
-
-\title{Relational Parametricity for Algebraically Indexed Types}
-
-\authorinfo{Robert Atkey\and Neil Ghani\\\and Patricia Johann}
-           {University of Strathclyde}
-           {\{Robert.Atkey,Neil.Ghani,Patricia.Johann\}@cis.strath.ac.uk}
-
-\authorinfo{Andrew Kennedy}
-           {Microsoft Research}
-           {}
-
-\usepackage{mathpartir}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{stmaryrd}
-\usepackage[all]{xy}
-
-\newcommand{\sepbar}{\mathrel|}
-
-\newcommand{\Rel}{\mathrm{Rel}}
-\newcommand{\relArrow}{\mathrel{\widehat\to}}
-\newcommand{\relTimes}{\mathrel{\widehat\times}}
-\newcommand{\relSum}{\mathrel{\widehat+}}
-
-\newcommand{\setOfIntegers}{\mathbb{Z}}
-\newcommand{\setOfBooleans}{\{\tmTT, \tmFF\}}
-
-\newcommand{\id}{\mathrm{id}}
-
-\newcommand{\indexOp}[1]{\texttt{#1}}
-\newcommand{\idxTms}[2]{\mathrm{IdxTm}(#1 \vdash #2)}
-\newcommand{\idxTmAS}[1]{\mathrm{IdxTm}(#1)}
-
-\newcommand{\sortType}{\texttt{type}}
-\newcommand{\tyInt}{\texttt{int}}
-\newcommand{\tyBool}{\texttt{bool}}
-\newcommand{\tyUnit}{\texttt{unit}}
-\newcommand{\tyPrim}[2]{\texttt{#1}\langle #2 \rangle}
-\newcommand{\tyPrimNm}[1]{\texttt{#1}}
-
-\newcommand{\tyArr}{\to}
-\newcommand{\tyProduct}{\times}
-\newcommand{\tyX}[1]{\texttt{X}\langle #1 \rangle}
-\newcommand{\isType}{\textsf{ type}}
-\newcommand{\isCtxt}{\textsf{ ctxt}}
-\newcommand{\Ty}{\textsf{Ty}}
-\newcommand{\ty}{\textsf{ty}}
-\newcommand{\subst}[2]{\mathrm{Subst}(#1,#2)}
-
-\newcommand{\tmTT}{\texttt{tt}}
-\newcommand{\tmFF}{\texttt{ff}}
-
-\newcommand{\relEnv}[1]{\mathcal{#1}}
-\newcommand{\tySem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}}}
-\newcommand{\ctxtSem}[1]{\llbracket #1 \rrbracket^{\mathcal{C}}}
-\newcommand{\tmSem}[1]{\llbracket #1 \rrbracket^{\mathit{tm}}}
-\newcommand{\tyPrimSem}[1]{\llbracket \tyPrimNm{#1} \rrbracket^{\mathcal{T}_0}}
-\newcommand{\rsem}[3]{\llbracket #1 \rrbracket^{\mathcal{R}}_{#2}{#3}}
-\newcommand{\extends}[2]{\mathsf{ext}(#1,#2)}
-
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
-\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
-\newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
-\newcommand{\corref}[1]{\hyperref[#1]{Corollary \ref*{#1}}}
-\newcommand{\conref}[1]{\hyperref[#1]{Conjecture \ref*{#1}}}
-\newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-\end{abstract}
-
-\category{D.1.1}{Programming techniques}{Applicative (functional)
-  programming} \category{D.2.4}{Software Engineering}{Software/Program
-  Verification} \category{D.3.3}{Programming Languages}{Language
-  Constructs and Features---Data types and structures}
-
-\terms
-  Languages, Theory, Types
-
-\keywords
-  parametricity, dimension types, computational geometry
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-
-\begin{enumerate}
-\item Introduction using the area-of-a-triangle example
-\item Narrowing of the semantic gap by use of advanced type systems
-\item Importance of the use of symmetry in computer science,
-  mathematics and the physical sciences
-\item Development of the theory of parametricity
-\end{enumerate}
-
-\subsection{Contributions}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Background and Related Work}
-
-Reynolds (Parametricity), Wadler (Free Theorems).
-
-Kennedy (Units of Measure)
-
-Something about symmetry?
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Type System}
-
-We now introduce our type system. We parameterise our development by
-an arbitrary multi-sorted algebraic theory, in the sense of universal
-algebra. We present the precise form of the theories we allow in
-\autoref{sec:index-exp}, and define some auxiliary notions of
-substitution and terms-up-to-equality. Several examples of useful
-algebraic theories will be given in
-\autoref{sec:example-systems}. Once an algebraic theory has been
-fixed, we define the well-formed types and terms with respect to this
-theory in \autoref{sec:types-and-terms}.
-
-\subsection{Algebraic Theories for Indexing}\label{sec:index-exp}
-
-\begin{figure*}[t]
-  \centering
-  Well-sorted terms
-  \begin{mathpar}
-    \inferrule*
-    {i : s \in \Delta}
-    {\Delta \vdash x : s}
-
-    \inferrule*
-    {\indexOp{f} : s_1,...,s_n \to s \\
-      \Delta \vdash e_1 : s_1 \\
-      \dots \\
-      \Delta \vdash e_n : s_n}
-    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) : s}
-  \end{mathpar}
-
-  \medskip
-
-  Equational rules
-  \begin{mathpar}
-    \inferrule*
-    {e_1 \stackrel{ax}= e_2 : s \in \mathit{Ax}(\Delta)}
-    {\Delta \vdash e_1 \equiv e_2 : s}
-
-    \inferrule*
-    {\indexOp{f} : s_1,...,s_n \to s \\
-      \Delta \vdash e_1 \equiv e'_1 : s_1 \\
-      \dots \\
-      \Delta \vdash e_n \equiv e'_n : s_n}
-    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) \equiv \indexOp{f}(e'_1, ..., e'_n) : s}
-
-    \inferrule*
-    {\Delta \vdash e : s}
-    {\Delta \vdash e \equiv e : s}
-
-    \inferrule*
-    {\Delta \vdash e_1 \equiv e_2 : s}
-    {\Delta \vdash e_2 \equiv e_1 : s}
-
-    \inferrule*
-    {\Delta \vdash e_1 \equiv e_2 : s \\ \Delta \vdash e_2 \equiv e_3 : s}
-    {\Delta \vdash e_1 \equiv e_3 : s}
-  \end{mathpar}
-  \caption{Algebraic Theories for Indexing}
-  \label{fig:alg-theories}
-\end{figure*}
-
-We assume a set $\mathit{Sort}$ of \emph{index sorts}, and we will use
-$s, s_1, s_2, ...$ to range over elements of $\mathit{Sort}$. An index
-context $\Delta$ is list of index variable and sort pairs, $i : s$,
-and we use the symbol $\epsilon$ to denote the empty index context (we
-will also use this symbol below to denote the empty typing
-context). We use letters like $i$ and $j$, sometimes with subscripts,
-to range over index variables. We also assume a collection
-$\mathit{IndexOp}$ of sorted index operations $\indexOp{f} :
-s_1,...,s_n \to s$. From the index operations we generate the
-well-sorted index expressions by the rules in
-\autoref{fig:alg-theories}.
-
-For each index context $\Delta$, we assume a set $\mathit{Ax}(\Delta)$
-of axioms $e_1 \stackrel{ax}= e_2 : s$ where both $\Delta \vdash e_1 :
-s$ and $\Delta \vdash e_2 : s$. From this collection of axioms, we
-generate the equational theory of index expressions by the rules in
-\autoref{fig:alg-theories} by closing under congruence, reflexivity,
-symmetry and transitivity.
-
-% For a pair of index contexts $\Delta$ and $\Delta'$, a substitution
-% $\sigma \in \subst{\Delta}{\Delta'}$ is a function $\sigma$ mapping
-% variables in $\Delta$ to well-sorted index expressions in $\Delta'$.
-
-Index context morphisms $\sigma : \Delta \Rightarrow \Delta'$ consist
-of a sequence of terms, one for each element of $\Delta'$, all
-well-sorted in the index context $\Delta$.
-\begin{mathpar}
-  \inferrule*
-  { }
-  {\cdot : \Delta \Rightarrow \epsilon}
-
-  \inferrule*
-  {\sigma : \Delta \Rightarrow \Delta' \\ \Delta \vdash e : s}
-  {\sigma, e : \Delta \Rightarrow \Delta', i:s}
-\end{mathpar}
-Directly from the definition of index context morphisms, we can see
-that given an index context morphism $\sigma : \Delta \Rightarrow
-\Delta'$, then for any index variable $i : s \in \Delta'$ there is an
-index expression $\sigma(i)$ such that $\Delta \vdash \sigma(i) : s$.
-
-We write $\id_\Delta : \Delta \Rightarrow \Delta$ for the index
-context morphism comprised of the terms built from the variables of
-$\Delta$, in order. We write $\pi_{i\mathord:s} : \Delta, i : s
-\Rightarrow \Delta$ for the context morphism again comprised of the
-terms built from the variables of $\Delta$, but this time typed in a
-larger context. For an index context morphism $\sigma : \Delta \to
-\Delta'$ we write $\sigma_s : \Delta, i\mathord:s \Rightarrow \Delta',
-i\mathord:s$ for the index context morphism generated by lifting each
-term to the larger context $\Delta, i\mathord:s$, and mapping the
-variable $i\mathord:s$ to the term $i$.
-
-Given an index expression $\Delta' \vdash e : s$, we define
-re-indexing by an index context morphism $\sigma : \Delta \Rightarrow
-\Delta'$ by the following clauses:
-\begin{eqnarray*}
-  \sigma^*i & = & \sigma(i) \\
-  \sigma^*\indexOp{f}(e_1, ..., e_n) & = & \indexOp{f}(\sigma^*e_1,...,\sigma^*e_n)
-\end{eqnarray*}
-It is easy to check that these clauses preserve well-formedness of
-index expressions, so that we obtain $\Delta \vdash \sigma^*e :
-s$. Moreover re-indexing by index context morphisms preserves
-equality:
-\begin{lemma}\label{lem:indexeq-subst}
-  If $\Delta \vdash e_1 \equiv e_2 : s$, then for any $\sigma :
-  \Delta' \Rightarrow \Delta$, we have $\Delta' \vdash \sigma^*e_1
-  \equiv \sigma^*e_2 : s$.
-\end{lemma}
-
-Also, projection commutes with lifting of substitutions. This is used
-in the proof of \lemref{lem:tysubst-rel}.
-\begin{lemma}
-  $\sigma^* \circ \pi^*_{i\mathord:s} = \pi^*_{i\mathord:s} \circ \sigma^*_s$
-\end{lemma}
-
-For an index context $\Delta$ and sort $s$, we write
-$\idxTms{\Delta}{s}$ for the set of index expressions $e$ such that
-$\Delta \vdash e : s$, quotiented by the equivalence relation $\Delta
-\vdash - \equiv - : s$. In light of \lemref{lem:indexeq-subst}, we can
-see that re-indexing by an index context morphism $\sigma : \Delta
-\Rightarrow \Delta'$ can be used as a function $\sigma^* :
-\idxTms{\Delta'}{s} \to \idxTms{\Delta}{s}$, for any sort $s$.
-
-\subsection{Types and Terms}\label{sec:types-and-terms}
-
-\begin{figure*}[t]
-  \centering
-  Well-indexed Types
-  \begin{mathpar}
-    \inferrule* [right=TyPrimIdx]
-    {\tyPrimNm{X} \in \mathit{PrimType}(s) \\ \Delta \vdash e : s}
-    {\Delta \vdash \tyPrim{X}{e} : \sortType}
-
-    \inferrule* [right=TyPrimNoIdx]
-    {\tyPrimNm{X} \in \mathit{PrimType}(\cdot)}
-    {\Delta \vdash \tyPrimNm{X} : \sortType}
-
-    \inferrule* [right=TyUnit]
-    { }
-    {\Delta \vdash \tyUnit : \sortType}
-
-    \inferrule* [right=TyArr]
-    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
-    {\Delta \vdash A \tyArr B : \sortType}
-
-    \inferrule* [right=TyProd]
-    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
-    {\Delta \vdash A \tyProduct B : \sortType}
-
-    \inferrule* [right=TySum]
-    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
-    {\Delta \vdash A + B : \sortType}
-
-    \inferrule* [right=TyForall] %FIXME: macroize forall
-    {\Delta, i : s \vdash A : \sortType}
-    {\Delta \vdash \forall i \mathord: s. A : \sortType}
-  \end{mathpar}
-
-  \medskip
-  Equality of well-indexed types
-  \begin{mathpar}
-    \inferrule* [right=TyEqPrimIdx]
-    {\Delta \vdash e_1 \equiv e_2 : s}
-    {\Delta \vdash \tyPrim{X}{e_1} \equiv \tyPrim{X}{e_2} : \sortType}
-
-    \inferrule* [right=TyEqPrimNoIdx]
-    { }
-    {\Delta \vdash \tyPrimNm{X} \equiv \tyPrimNm{X} : \sortType}
-
-    \inferrule* [right=TyEqUnit]
-    { }
-    {\Delta \vdash \tyUnit \equiv \tyUnit : \sortType}
-
-    \inferrule* [right=TyEqArr]
-    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
-    {\Delta \vdash A \tyArr B \equiv A' \tyArr B' : \sortType}
-
-    \inferrule* [right=TyEqProd]
-    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
-    {\Delta \vdash A \tyProduct B \equiv A' \tyProduct B' : \sortType}
-
-    \inferrule* [right=TyEqSum]
-    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
-    {\Delta \vdash A + B \equiv A' + B' : \sortType}
-
-    \inferrule* [right=TyEqForall]
-    {\Delta, i : s \vdash A \equiv A' : \sortType}
-    {\Delta \vdash \forall i \mathord: s. A \equiv \forall i \mathord: s. A' : \sortType}
-  \end{mathpar}
-
-  \medskip
-  Well-indexed contexts
-  \begin{mathpar}
-    \inferrule*
-    { }
-    {\Delta \vdash \epsilon \isCtxt}
-
-    \inferrule*
-    {\Delta \vdash \Gamma \isCtxt \\ \Delta \vdash A : \sortType \\ x \not\in \Gamma}
-    {\Delta \vdash \Gamma, x : A \isCtxt}
-  \end{mathpar}
-  \caption{Well-indexed Types and Contexts}
-  \label{fig:types}
-\end{figure*}
-
-For each sort $s \in \mathit{Sort}$, we assume we are given a set of
-names of primitive types $\mathit{PrimType}(s)$. We assume that the
-sets $\mathit{PrimType}(s)$ are disjoint for non-equal sorts $s_1,
-s_2$. For example, in the units of measure example in
-\autoref{sec:units-example}, there is a single sort $\mathsf{unit}$,
-and $\mathit{PrimType}(\mathsf{unit}) = \{\tyPrimNm{num}\}$. For each
-primitive type $\tyPrimNm{X} \in \mathit{PrimType}(s)$, and index
-expression $\Delta \vdash e : s$, we form the type $\tyPrim{X}{e}$, by
-the \TirName{TyPrimIdx} rule in \autoref{fig:types}. Thus, in the
-units-of-measure example, the type $\tyPrim{num}{u}$ represents
-numbers measured by the units $u$.
-
-We also assume that there is an additional set of primitive types
-$\mathit{PrimType}(\cdot)$ that is disjoint from all the other sets of
-primitive types. These primitive types do not need to be paired with
-an index expression to form well-indexed types. Such types are created
-by the \TirName{TyPrimNoIdx} rule. An example of such a primitive type
-is the type $\tyPrimNm{real}$ of real-valued scalars in the Euclidean
-plane geometry example in \autoref{sec:2d-example}.
-
-The rules for generating the rest of the types of the system,
-including unit, product, sum, function and universal types, are given
-in \autoref{fig:types}. We use capital roman letters like $A$ and $B$
-to range over well-indexed types. Well-indexed types come with an
-equational theory, derived from the equational theory on index
-expressions. This equational theory is also presented in
-\autoref{fig:types}.
-
-We define substitution by an index context morphism $\sigma : \Delta
-\Rightarrow \Delta'$ on types by structural induction on the structure
-of the type: given a type $\Delta' \vdash A : \sortType$, we obtain a
-type $\Delta \vdash \sigma^*A : \sortType$. The only interesting case
-is for \TirName{TyPrimIdx}:
-\begin{displaymath}
-  \sigma^*\tyPrim{X}{e} = \tyPrim{X}{\sigma^*e}
-\end{displaymath}
-
-Well-formed contexts are defined to be lists of variables paired with
-well-formed types with respect to some index context $\Delta$, with
-the condition that no variable names are repeated. Formally, contexts
-are defined by the two rules shown in
-\autoref{fig:types}. Substitution by an index context morphism extends
-to contexts just by doing substitution on each individual type.
-
-The typing rules for our system define the well-typed terms with
-respect to an index context $\Delta$ and a type context $\Delta \vdash
-\Gamma \isCtxt$. The judgement $\Delta; \Gamma \vdash M : A$ is
-defined in \autoref{fig:terms}. The equational theory on types is
-incorporated into the type system via the rule \TirName{TyEq}, which
-allows for a term that is judged to have type $A$ to also have any
-other equal type $B$.
-\begin{figure*}[t]
-  \centering
-  \begin{mathpar}
-    \inferrule* [right=Var]
-    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
-    {\Delta; \Gamma \vdash x : A}
-
-    \inferrule* [right=TyEq]
-    {\Delta; \Gamma \vdash M : A \\ \Delta \vdash A \equiv B : \sortType}
-    {\Delta; \Gamma \vdash M : B}
-
-    \inferrule* [right=Unit]
-    { }
-    {\Delta; \Gamma \vdash * : 1}
-
-    \inferrule* [right=Pair]
-    {\Delta; \Gamma \vdash M : A \\
-      \Delta; \Gamma \vdash N : B}
-    {\Delta; \Gamma \vdash (M, N) : A \tyProduct B}
-
-    \inferrule* [right=Proj1]
-    {\Delta; \Gamma \vdash M : A \tyProduct B}
-    {\Delta; \Gamma \vdash \pi_1 M : A}
-
-    \inferrule* [right=Proj2]
-    {\Delta; \Gamma \vdash M : A \tyProduct B}
-    {\Delta; \Gamma \vdash \pi_2 M : B}
-
-    \inferrule* [right=Inl]
-    {\Delta; \Gamma \vdash M : A}
-    {\Delta; \Gamma \vdash \mathrm{inl}\ M : A + B}
-
-    \inferrule* [right=Inr]
-    {\Delta; \Gamma \vdash M : B}
-    {\Delta; \Gamma \vdash \mathrm{inr}\ M : A + B}
-
-    \inferrule* [right=Cond]
-    {\Delta; \Gamma \vdash M : A + B \\
-      \Delta; \Gamma, x : A \vdash N_1 : C \\
-      \Delta; \Gamma, y : B \vdash N_2 : C}
-    {\Delta; \Gamma \vdash \textrm{case}\ M\ \textrm{of}\ \textrm{inl}\ x.N_1; \textrm{inr}\ y.N_2 : C}
-
-    \inferrule* [right=Abs]
-    {\Delta; \Gamma, x : A \vdash M : B}
-    {\Delta; \Gamma \vdash \lambda x.M : A \tyArr B}
-
-    \inferrule* [right=App]
-    {\Delta; \Gamma \vdash M : A \tyArr B \\
-      \Delta; \Gamma \vdash N : A}
-    {\Delta; \Gamma \vdash M N : B}
-
-    \inferrule* [right=UnivAbs]
-    {\Delta, i : s; \pi_{i\mathord:s}^*\Gamma \vdash M : A}
-    {\Delta; \Gamma \vdash \Lambda i. M : \forall i\mathord:s. A}
-
-    \inferrule* [right=UnivApp]
-    {\Delta; \Gamma \vdash M : \forall i\mathord:s. A \\ \Delta \vdash e : s}
-    {\Delta; \Gamma \vdash M [e] : (\id_\Delta, e)^*A}
-  \end{mathpar}
-
-  \caption{Well-typed terms}
-  \label{fig:terms}
-\end{figure*}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 encoding of a natural deduction system?
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Semantics and Parametricity Theorem}
-
-There is an index-erasure semantics, and a relational semantics. The
-relational semantics allows us to derive free theorems about the
-index-erasure semantics.
-
-\subsection{Index Erasure Semantics}
-
-We now define a semantics of types and terms that ignores all indexing
-information. By defining this semantics we can easily see that the
-indexing information can be discarded before execution.
-
-\subsubsection{Index-Erasure Semantics of Types}
-
-For each primitive type $\tyPrimNm{X} \in \mathit{PrimType}(s)$, for
-some sort $s$, and for the primitive types $\tyPrimNm{X} \in
-\mathit{PrimType}(\cdot)$, we assume that $\tyPrimNm{X}$ is assigned a
-set $\tyPrimSem{X}$. We extend this to assign a set to every
-well-formed type by induction on the type structure. To interpret the
-universal quantifier, we simply ignore the indexing:
-\begin{eqnarray*}
-  \tySem{\tyUnit} & = & \{*\} \\
-  \tySem{\tyPrim{X}{e}} & = & \tyPrimSem{X} \\
-  \tySem{\tyPrim{X}} & = & \tyPrimSem{X} \\
-  \tySem{A \tyProduct B} & = & \tySem{A} \times \tySem{B} \\
-  \tySem{A \tyArr B} & = & \tySem{A} \to \tySem{B} \\
-  \tySem{A + B} & = & \tySem{A} + \tySem{B} \\
-  \tySem{\forall i\mathord:s. A} & = & \tySem{A}
-\end{eqnarray*}
-
-Since our definition has completely ignored the indexing, and type
-equality is defined as an extension of index equality, we can easily
-see that equal types have equal denotations in our index-erasure
-semantics:
-\begin{lemma}\label{lem:tyeq-erasure}
-  If $\Delta \vdash A \equiv B : \sortType$ then $\tySem{A} =
-  \tySem{B}$.
-\end{lemma}
-
-\begin{proof}
-  By induction on the derivation of $\Delta \vdash A \equiv B :
-  \sortType$:
-  \begin{description}
-  \item[Case \TirName{TyEqUnit}] In this case, $A = B = \tyUnit$, so
-    $\tySem{A} = \tySem{B}$.
-  \item[Case \TirName{TyEqPrimIdx}] We have $A = \tyPrim{X}{e_1}$ and
-    $B = \tyPrim{X}{e_2}$, for some $e_1$ and $e_2$ of the same
-    sort. The erasure semantics ignores the index arguments, so
-    $\tySem{A} = \tySem{\tyPrim{X}{e_1}} = \tyPrimSem{X} =
-    \tySem{\tyPrim{X}{e_2}} = \tySem{B}$, as required.
-  \item[Case \TirName{TyEqPrimNoIdx}] We have $A = \tyPrimNm{X}$ and
-    $B = \tyPrimNm{X}$. So $\tySem{A} = \tyPrimSem{X} = \tySem{B}$, as
-    required.
-  \item[Case \TirName{TyEqArr}] Here, $A = A_1 \tyArr A_2$ and $B =
-    B_1 \tyArr B_2$, and by the induction hypothesis we know that
-    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
-    $\tySem{A} = \tySem{A_1 \tyArr A_2} = \tySem{A_1} \to \tySem{A_2}
-    = \tySem{B_1} \to \tySem{B_2} = \tySem{B_1 \tyArr B_2} =
-    \tySem{B}$.
-  \item[Case \TirName{TyEqProd}] In this case, $A = A_1 \tyProduct
-    A_2$ and $B = B_1 \tyProduct B_2$, and by the induction hypothesis
-    we know that $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} =
-    \tySem{B_2}$. Hence $\tySem{A} = \tySem{A_1 \tyProduct A_2} =
-    \tySem{A_1} \times \tySem{A_2} = \tySem{B_1} \times \tySem{B_2} =
-    \tySem{B_1 \tyProduct B_2} = \tySem{B}$.
-  \item[Case \TirName{TyEqSum}] In this case, $A = A_1 + A_2$ and $B =
-    B_1 + B_2$, and by the induction hypothesis we know that
-    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
-    $\tySem{A} = \tySem{A_1 + A_2} = \tySem{A_1} + \tySem{A_2} =
-    \tySem{B_1} + \tySem{B_2} = \tySem{B_1 + B_2} = \tySem{B}$.
-  \item[Case \TirName{TyEqForall}] We have $A = \forall
-    i\mathord:s.A'$ and $B = \forall i\mathord:s.B'$. By the induction
-    hypothesis we know that $\tySem{A'} = \tySem{B'}$, and hence
-    $\tySem{A} = \tySem{\forall i\mathord:s. A'} = \tySem{A'} =
-    \tySem{B'} = \tySem{\forall i\mathord:s. B'} = \tySem{B}$.
-  \end{description}
-\end{proof}
-
-Moreover, the index-erasure semantics is insensitive to index
-substitutions.
-\begin{lemma}\label{lem:tysubst-erasure}
-  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then $\tySem{\sigma^*A} = \tySem{A}$.
-\end{lemma}
-
-\begin{proof}
-  By induction on the derivation of $\Delta \vdash A : \sortType$.
-  \begin{description}
-  \item[Case \TirName{TyUnit}] Directly from the definition of
-    substitution on types of the form $\tyUnit$, we have
-    $\tySem{\sigma^*\tyUnit} = \tySem{\tyUnit}$, as required.
-  \item[Case \TirName{TyPrimIdx}] Since the index-erasure semantics
-    ignores all index expessions, we have
-    $\tySem{\sigma^*\tyPrim{X}{e}} = \tySem{\tyPrim{X}{\sigma^*e}} =
-    \tyPrimSem{X} = \tySem{\tyPrim{X}{e}}$.
-  \item[Case \TirName{TyPrimNoIdx}] Similar to the previous case.
-  \item[Case \TirName{TyArr}] We have $A = A_1 \tyArr A_2$, and by the
-    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
-    \tySem{A_1}$ and $\tySem{\sigma^*B_1} = \tySem{B_1}$. Thus, by the
-    definition of substitution on arrow types, we have
-    $\tySem{\sigma^*(A_1 \tyArr A_2)} = \tySem{\sigma^*A_1 \tyArr
-      \sigma^*A_2} = \tySem{\sigma^*A_1} \to \tySem{\sigma^*A_2} =
-    \tySem{A_1} \to \tySem{A_2} = \tySem{A_1 \tyArr A_2}$, as
-    required.
-  \item[Case \TirName{TyProd}] We have $A = A_1 \tyProduct A_2$, and
-    by the induction hypothesis, we know that $\tySem{\sigma^*A_1} =
-    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
-    definition of substitution on product types, we have
-    $\tySem{\sigma^*(A_1 \tyProduct A_2)} = \tySem{\sigma^*A_1
-      \tyProduct \sigma^*A_2} = \tySem{\sigma^*A_1} \times
-    \tySem{\sigma^*A_2} = \tySem{A_1} \times \tySem{A_2} = \tySem{A_1
-      \tyProduct A_2}$, as required.
-  \item[Case \TirName{TySum}] We have $A = A_1 + A_2$, and by the
-    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
-    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
-    definition of substitution on sum types, we have
-    $\tySem{\sigma^*(A_1 + A_2)} = \tySem{\sigma^*A_1 + \sigma^*A_2} =
-    \tySem{\sigma^*A_1} + \tySem{\sigma^*A_2} = \tySem{A_1} +
-    \tySem{A_2} = \tySem{A_1 + A_2}$, as required.
-  \item[Case \TirName{TyForall}] We have $A = \forall i\mathord:s.A'$,
-    and by the induction hypothesis we know that, for all $\sigma'$,
-    $\tySem{\sigma'^*A'} = \tySem{A'}$. Therefore, from the definition
-    of substitution for $\forall$-types, we have
-    $\tySem{\sigma^*\forall i\mathord:s.A'} = \tySem{\forall
-      i\mathord:s.\sigma_s^{*}A'} = \tySem{\sigma_s^{*}A'} = \tySem{A'}
-    = \tySem{\forall i\mathord:s.A'}$, where we have relied on the
-    fact that the index-erasure semantics of types ignores the index
-    quantification.
-  \end{description}
-\end{proof}
-
-For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
-index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
-structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
-\ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
-one-element set.
-
-\subsubsection{Index-Erasure Semantics of Terms}
-
-For a well-formed term $\Delta; \Gamma \vdash e : A$, we define a
-function $\tmSem{e} : \ctxtSem{\Gamma} \to \tySem{A}$, completely
-ignoring the indexing information. In light of
-\lemref{lem:tyeq-erasure} and \lemref{lem:tysubst-erasure}, we may
-define this function directly on the syntax of well-typed terms,
-rather than on the typing derivations. We define $\tmSem{e}$ by the
-following clauses:
-\begin{eqnarray*}
-  \tmSem{x}\eta & = & \eta_x \\
-  \tmSem{(M,N)}\eta & = & (\tmSem{M}\eta, \tmSem{N}\eta) \\
-  \tmSem{\pi_1M}\eta & = & \pi_1(\tmSem{M}\eta) \\
-  \tmSem{\pi_2M}\eta & = & \pi_2(\tmSem{M}\eta) \\
-  \tmSem{\mathrm{inl}\ M}\eta & = & \mathrm{inl}\ (\tmSem{M}\eta) \\
-  \tmSem{\mathrm{inr}\ M}\eta & = & \mathrm{inr}\ (\tmSem{M}\eta) \\
-  \left\llbracket
-    \begin{array}{l}
-      \textrm{case}\ M\ \textrm{of}\\
-      \textrm{inl}\ x.N_1;\\
-      \textrm{inr}\ y.N_2
-    \end{array}\right\rrbracket^{\mathit{tm}}\eta & = &
-  \left\{
-    \begin{array}{ll}
-      \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
-      \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
-    \end{array}
-  \right. \\
-  \tmSem{\lambda x. M}\eta & = & \lambda v. \tmSem{M}(\eta, v) \\
-  \tmSem{M N}\eta & = & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
-  \tmSem{\Lambda i.\ M}\eta & = & \tmSem{M}\eta \\
-  \tmSem{M[e]}\eta & = & \tmSem{M}\eta
-\end{eqnarray*}
-In the above, we have written $\eta_x$ to denote the appropriate
-projection from $\eta$ to get the position corresponding to the
-variable $x$ in the context.
-
-\subsection{Relational Semantics}
-
-\subsubsection{Relation Environments}
-
-For a set $X$, let $\Rel(X)$ be the set of binary relations on
-$X$. For an index context $\Delta$, a \emph{relation environment} is a
-family of functions $\rho = \{ \rho_{\tyPrimNm{X}} :
-\idxTms{\Delta}{s} \to \Rel(\tyPrimSem{X}) \}_{s \in \mathit{Sort},
-  \tyPrimNm{X} \in \mathit{PrimType}(s)}$.
-
-We define our relational semantics of types with respect to a given
-collection of relation environments: we assume that, for each index
-context $\Delta$, we are given a set of relation environments
-$\relEnv{E}(\Delta)$. This family must satisfy the following two
-properties:
-\begin{enumerate}
-\item Closure under re-indexing by index context morphisms: if $\rho
-  \in \relEnv{E}(\Delta)$ and $\sigma : \Delta' \Rightarrow \Delta$,
-  then $\rho \circ \sigma^* \in \relEnv{E}(\Delta')$.
-\item If we have a pair of relation environments $\rho_1 \in
-  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
-  \relEnv{E}(\Delta)$, along with an index context morphism $\sigma :
-  \Delta' \Rightarrow \Delta$, such that the outer edges of the
-  following diagram commute, for all primitive types $\tyPrimNm{X} \in
-  \mathit{PrimType}(s)$:
-  \begin{displaymath}
-    \xymatrix{
-      {\idxTms{\Delta}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar[d]_{\sigma^*}
-      &
-      {\idxTms{\Delta,i\mathord:s'}{s}} \ar[d]^{\sigma_{s'}^*} \ar@/^/[rdd]^{\rho_1}
-      \\
-      {\idxTms{\Delta'}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2}
-      &
-      {\idxTms{\Delta',i\mathord:s'}{s}} \ar@{.>}[dr]_\rho
-      \\
-      &
-      &
-      {\Rel(X)}
-    }
-  \end{displaymath}
-  Then there exists a relation environment $\rho \in
-  \relEnv{E}(\Delta,i\mathord:s')$ such that the two triangles in the
-  bottom right of the diagram commute.
-\end{enumerate}
-
-Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
-the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
-additional index variable $i\mathord:s$ to be the following:
-\begin{displaymath}
-  \extends{\rho}{i\mathord:s} = \{ \rho' \in \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s} = \rho \}
-\end{displaymath}
-
-\subsubsection{Relational Semantics of Types}
-
-%FIXME: re-word this
-A given relation context $\rho \in \relEnv{E}(\Delta)$ is extended to
-a relation interpretation of any type $\Delta \vdash A : \sortType$ by
-induction its derivation. So we define a relation
-$\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-formed
-type $A$.
-\begin{eqnarray*}
-  \rsem{\tyUnit}{\relEnv{E}}\rho & = & \{(*,*)\} \\
-  \rsem{\tyPrim{X}{e}}{\relEnv{E}}\rho & = & \rho_{\tyPrimNm{X}}(e) \\
-  \rsem{A \tyArr B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{A \tyProduct B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{A + B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & = & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i:s} \}
-\end{eqnarray*}
-In this definition we have made use of the following two constructions
-on binary relations. If $R \in \Rel(X)$ and $S \in \Rel(Y)$, then $R
-\relArrow S \in \Rel(X \to Y)$ is defined as $\{ (f_1,f_2) \sepbar
-\forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S \}$. With the same
-assumptions on $R$ and $S$, the relation $R \relTimes S \in \Rel(X
-\times Y)$ is defined as $\{ ((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2)
-\in R \land (b_1,b_2) \in S \}$. FIXME: do $\relSum$.
-
-We extend the relational semantics to contexts of well-formed types by
-the following clauses:
-\begin{eqnarray*}
-  \rsem{\epsilon}{\relEnv{E}}{\rho} & = & \id_{\ctxtSem{\epsilon}} \\
-  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & = & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
-\end{eqnarray*}
-
-\subsubsection{Parametricity}
-
-This relies on \lemref{lem:tyeq-erasure} to make sure that it is
-well-typed.
-
-\begin{lemma}\label{lem:tyeq-rel}
-  If $\Delta \vdash A \equiv B : \sortType$, then for all $\rho \in
-  \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}\rho = \rsem{B}{\relEnv{E}}\rho$.
-\end{lemma}
-
-\begin{proof}
-  By induction on the derivation of $\Delta \vdash A \equiv B :
-  \sortType$.
-  \begin{description}
-  \item[Case \TirName{TyEqUnit}] In this case $A = B = \tyUnit$, so
-    the statement is trivially statisfied.
-  \item[Case \TirName{TyEqPrimIdx}] In this case we have $A =
-    \tyPrim{X}{e_1}$ and $B = \tyPrim{X}{e_2}$, with $\Delta \vdash
-    e_1 \equiv e_2 : s$. The index expressions $e_1$ and $e_2$ are in
-    the same equivalence class, so by the definition of relation
-    environments we have that $\rho_{\tyPrimNm{X}}(e_1) =
-    \rho_{\tyPrimNm{X}}(e_2)$ and hence
-    $\rsem{\tyPrim{X}{e_1}}{\relEnv{E}}\rho =
-    \rsem{\tyPrim{X}{e_2}}{\relEnv{E}}\rho$.
-  \item[Case \TirName{TyEqPrimNoIdx}] This case is similar to the case
-    for \TirName{TyEqUnit}.
-  \item[Case \TirName{TyEqArr}] We have $A = A_1 \tyArr A_2$ and $B =
-    B_1 \tyArr B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$
-    and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the induction
-    hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_2}{\relEnv{E}}\rho$. Thus we may reason as follows:
-    $\rsem{A_1 \tyArr A_2}{\relEnv{E}}\rho =
-    \rsem{A_1}{\relEnv{E}}\rho \relArrow \rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho \relArrow \rsem{B_2}{\relEnv{E}}\rho =
-    \rsem{B_1 \tyArr B_2}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{TyEqProd}] We have $A = A_1 \tyProduct A_2$ and
-    $B = B_1 \tyProduct B_2$, with $\Delta \vdash A_1 \equiv B_1 :
-    \sortType$ and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the
-    induction hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
-    $\rsem{A_1 \tyProduct A_2}{\relEnv{E}}\rho =
-    \rsem{A_1}{\relEnv{E}}\rho \relTimes \rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho \relTimes \rsem{B_2}{\relEnv{E}}\rho =
-    \rsem{B_1 \tyProduct B_2}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{TyEqSum}] We have $A = A_1 + A_2$ and $B = B_1 +
-    B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$ and $\Delta
-    \vdash A_2 \equiv B_2 : \sortType$. By the induction hypothesis,
-    we know that $\rsem{A_1}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
-    $\rsem{A_1 + A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}\rho
-    \relSum \rsem{A_2}{\relEnv{E}}\rho = \rsem{B_1}{\relEnv{E}}\rho
-    \relSum \rsem{B_2}{\relEnv{E}}\rho = \rsem{B_1 +
-      B_2}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{TyEqForall}] We have $A = \forall
-    i\mathord:s. A'$ and $B = \forall i\mathord:s.B'$, with $\Delta, i
-    : s \vdash A' \equiv B' : \sortType$. By the induction hypothesis,
-    we know that for all $\rho'$, $\rsem{A'}{\relEnv{E}}\rho' =
-    \rsem{B'}\rho'$. We reason as follows:
-    \begin{eqnarray*}
-      &    & (x_1,x_2) \in \rsem{\forall i\mathord:s. A'}{\relEnv{E}}\rho \\
-      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{A'}{\relEnv{E}}\rho' \\
-      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{B'}{\relEnv{E}}\rho' \\
-      &\iff& (x_1,x_2) \in \rsem{\forall i\mathord:s. B'}{\relEnv{E}}\rho
-    \end{eqnarray*}
-    Thus, by extensionality, we have $\rsem{\forall
-      i\mathord:s.A'}\rho \equiv \rsem{\forall
-      i\mathord:s.B'}{\relEnv{E}}\rho$, as required.
-  \end{description}
-\end{proof}
-
-This next lemma relies on \lemref{lem:tysubst-erasure} to make sure
-that it is well-typed.
-
-\begin{lemma}\label{lem:tysubst-rel}
-  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
-  $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Likewise,
-  if $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
-  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho = \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
-\end{lemma}
-
-\begin{proof}
-  The first part of the lemma statement is proved by induction on the
-  derivation of $\Delta \vdash A : \sortType$. We analyse each case in
-  turn:
-  \begin{description}
-  \item[Case \TirName{TyUnit}] Directly from the definition of
-    substitution on types of the form $\tyUnit$, and the fact that
-    $\rsem{\tyUnit}{\relEnv{E}}\rho_1 =
-    \rsem{\tyUnit}{\relEnv{E}}\rho_2$ for any pair $\rho_1,\rho_2$ of
-    relation environments, we have
-    $\rsem{\sigma^*\tyUnit}{\relEnv{E}}\rho =
-    \rsem{\tyUnit}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as required.
-  \item[Case \TirName{TyPrimIdx}] We have $A = \tyPrim{X}{e}$, therefore
-    $\rsem{\sigma^*A}{\relEnv{E}}\rho =
-    \rsem{\sigma^*\tyPrim{X}{e}}{\relEnv{E}}\rho =
-    \rsem{\tyPrim{X}{\sigma^*e}}{\relEnv{E}}\rho =
-    \rho_{\tyPrimNm{X}}(\sigma^*e) =
-    \rsem{\tyPrim{X}{e}}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as
-    required.
-  \item[Case \TirName{TyPrimNoIdx}] Similar to the case for
-    \TirName{TyUnit}.
-  \item[Case \TirName{TyArr}] In this case, $A = A_1 \tyArr A_2$ and
-    by the induction hypothesis, we know that $\rsem{\sigma^*A_1}{\relEnv{E}}\rho
-    = \rsem{A_1}{\relEnv{E}}{(\rho \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we reason as
-    follows: $\rsem{\sigma^*(A_1 \tyArr A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1
-      \tyArr \sigma^*A_2}{\relEnv{E}}\rho = \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relArrow
-    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho \circ \sigma^*)}
-    \relArrow \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)} = \rsem{A_1 \tyArr
-      A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as required.
-  \item[Case \TirName{TyProd}] In this case, $A = A_1 \tyProduct A_2$
-    and by the induction hypothesis, we know that
-    $\rsem{\sigma^*A_1}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we
-    reason as follows: $\rsem{\sigma^*(A_1 \tyProduct
-      A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1 \tyProduct
-      \sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relTimes
-    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)} \relTimes \rsem{A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)} = \rsem{A_1 \tyProduct A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)}$, as required.
-  \item[Case \TirName{TySum}] In this case, $A = A_1 + A_2$
-    and by the induction hypothesis, we know that
-    $\rsem{\sigma^*A_1}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we
-    reason as follows: $\rsem{\sigma^*(A_1 +
-      A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1 +
-      \sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relSum
-    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)} \relSum \rsem{A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)} = \rsem{A_1 + A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)}$, as required.
-  \item[Case \TirName{TyForall}] We have $A = \forall
-    i\mathord:s.A'$. We prove the required equation by demonstrating
-    two inclusions. From left-to-right, we know that
-    \begin{equation}
-      \label{eq:tysubst-rel-forall1}
-      (x_1,x_2) \in \rsem{\sigma^*\forall i\mathord:s. A'}{\relEnv{E}}\rho
-    \end{equation}
-    and we must show that $(x_1, x_2) \in \rsem{\forall
-      i\mathord:s. A'}{\relEnv{E}}{(\rho \circ \sigma^*)}$. From
-    (\ref{eq:tysubst-rel-forall1}) and the definition of substitution,
-    we know that
-    \begin{displaymath}
-      \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{\sigma_s^*A'}{\relEnv{E}}\rho'
-    \end{displaymath}
-    By the induction hypothesis, we can deduce that
-    \begin{equation}
-      \label{eq:tysubst-rel-forall2}
-      \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{A'}{\relEnv{E}}{(\rho' \circ \sigma_s^{*})}
-    \end{equation}
-    To show the required inclusion, we must show that for all $\rho_1
-    \in \extends{\rho \circ \sigma^*}{i:s}$, it is the case that
-    $(x_1,x_2) \in \rsem{A'}{\relEnv{E}}{\rho_1}$. We now make use of
-    the pushout property of our family $\relEnv{E}$ of relation
-    environments: we have a $\rho \in \relEnv{E}(\Delta)$ and $\rho_1
-    \in \relEnv{E}(\Delta, i:s)$ such that $\rho_1 \circ \pi_{i:s} =
-    \rho \circ \sigma^*$ (by the assumption that $\rho_1 \in
-    \extends{\rho \circ \sigma^*}{i:s}$), so we can obtain a $\rho'
-    \in \relEnv{E}(\Delta',i:s)$ such that $\rho' \circ \pi_{i:s} =
-    \rho$ (so $\rho' \in \extends{\rho}{i:s}$), and $\rho' \circ
-    \sigma^* = \rho_1$. Therefore we use
-    (\ref{eq:tysubst-rel-forall2}) to deduce that $(x_1,x_2) \in
-    \rsem{A'}{\relEnv{E}}{\rho_1}$, and the inclusion from
-    left-to-right is shown.
-
-    From right-to-left, we must now show
-    (\ref{eq:tysubst-rel-forall2}) under the assumption that
-    \begin{equation}
-      \label{eq:tysubst-rel-forall3}
-      \forall \rho_1 \in \extends{\rho \circ \sigma^*}{i:s}. (x_1,x_2) \in \rsem{A'}{\relEnv{E}}{\rho_1}.
-    \end{equation}
-    Given $\rho' \in \extends{\rho}{i:s}$, we let $\rho_1 = \rho'
-    \circ \sigma^*$. Now, $\rho_1 \circ \pi_{i:s} = \rho' \circ
-    \sigma^* \circ \pi_{i:s} = \rho' \circ \pi_{i:s} \circ \sigma_s^{*}
-    = \rho \circ \sigma_s^{*}$. Thus $\rho_1 \in \extends{\rho \circ
-      \sigma^*}{i:s}$ and we may use (\ref{eq:tysubst-rel-forall3}) to
-    deduce that $(x_1,x_2) \in \rsem{A'}{\relEnv{E}}{(\rho' \circ \sigma_s^{*})}$,
-    as required.
-  \end{description}
-  The second part of the lemma statement follows directly by induction
-  on the derivation of $\Delta \vdash \Gamma \isCtxt$ and the first
-  part. The step case of the induction is very similar to the
-  \TirName{TyProd} case above.
-\end{proof}
-
-\begin{theorem}[Parametricity]
-  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
-  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
-  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
-  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
-\end{theorem}
-
-\begin{proof}
-  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
-  three interesting cases follow:
-  \begin{description}
-  \item[Case \TirName{TyEq}] This case follows directly from
-    \lemref{lem:tyeq-rel}.
-  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
-    $A = \forall i\mathord:s. A'$. We know from the induction
-    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
-    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}\rho'$,
-    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho'$. We are given $\rho \in
-    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
-    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
-    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
-    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
-    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
-      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
-    \rsem{\pi_{i:s}^*\Gamma}\rho''$ by \lemref{lem:tysubst-rel}. So we
-    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}\rho''$, and thus
-    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
-      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
-      i.M'}\eta_2) \in \rsem{\forall
-      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
-    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
-    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
-    know that for all $\rho' \in \extends{\rho}{i:s}$,
-    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho'$. If we let $\rho' = \rho \circ
-    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
-    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
-    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
-    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
-    by \lemref{lem:tysubst-rel}, conclude that $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
-    as required.
-  \end{description}
-  The remaining cases are standard for proofs of the fundamental lemma
-  of logical relations for simply-typed systems.
-\end{proof}
 
@@
 $\idxTms{\Delta}{\mathsf{principal}}$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Extensions}
-
-\subsection{Inductive Types}
-
-\subsection{Type Operators}
-
-\subsection{Size indexing and Combinatorial Species}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{plainnat}
-\bibliography{paper}
-
-\end{document}

### 2012-05-22 paper/old-semantics.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

--- 2012-05-17 paper/paper.tex
@@
-\documentclass[natbib,preprint]{sigplanconf}
-
-\usepackage[usenames]{color}
-\definecolor{citationcolour}{rgb}{0,0.4,0.2}
-\definecolor{linkcolour}{rgb}{0,0,0.8}
-\usepackage{hyperref}
-\hypersetup{colorlinks=true,
-            urlcolor=linkcolour,
-            linkcolor=linkcolour,
-            citecolor=citationcolour,
-            pdftitle=Relational Parametricity for Algebraically-Indexed Types,
-            pdfauthor={Robert Atkey, Neil Ghani, Patricia Johann, Andrew Kennedy},
-            pdfkeywords={}}  
-\def\sectionautorefname{Section}
-\def\subsectionautorefname{Section}
-
-\title{Relational Parametricity for Algebraically Indexed Types}
-
-\authorinfo{Robert Atkey\and Neil Ghani\\\and Patricia Johann}
-           {University of Strathclyde}
-           {\{Robert.Atkey,Neil.Ghani,Patricia.Johann\}@cis.strath.ac.uk}
-
-\authorinfo{Andrew Kennedy}
-           {Microsoft Research}
-           {}
-
-\usepackage{mathpartir}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{stmaryrd}
-\usepackage[all]{xy}
-
-\newcommand{\sepbar}{\mathrel|}
-
-\newcommand{\Rel}{\mathrm{Rel}}
-\newcommand{\relArrow}{\mathrel{\widehat\to}}
-\newcommand{\relTimes}{\mathrel{\widehat\times}}
-\newcommand{\relSum}{\mathrel{\widehat+}}
-
-\newcommand{\setOfIntegers}{\mathbb{Z}}
-\newcommand{\setOfBooleans}{\{\tmTT, \tmFF\}}
-
-\newcommand{\id}{\mathrm{id}}
-
-\newcommand{\indexOp}[1]{\texttt{#1}}
-\newcommand{\idxTms}[2]{\mathrm{IdxTm}(#1 \vdash #2)}
-\newcommand{\idxTmAS}[1]{\mathrm{IdxTm}(#1)}
-
-\newcommand{\sortType}{\texttt{type}}
-\newcommand{\tyInt}{\texttt{int}}
-\newcommand{\tyBool}{\texttt{bool}}
-\newcommand{\tyUnit}{\texttt{unit}}
-\newcommand{\tyPrim}[2]{\texttt{#1}\langle #2 \rangle}
-\newcommand{\tyPrimNm}[1]{\texttt{#1}}
-
-\newcommand{\tyArr}{\to}
-\newcommand{\tyProduct}{\times}
-\newcommand{\tyX}[1]{\texttt{X}\langle #1 \rangle}
-\newcommand{\isType}{\textsf{ type}}
-\newcommand{\isCtxt}{\textsf{ ctxt}}
-\newcommand{\Ty}{\textsf{Ty}}
-\newcommand{\ty}{\textsf{ty}}
-\newcommand{\subst}[2]{\mathrm{Subst}(#1,#2)}
-
-\newcommand{\tmTT}{\texttt{tt}}
-\newcommand{\tmFF}{\texttt{ff}}
-
-\newcommand{\relEnv}[1]{\mathcal{#1}}
-\newcommand{\tySem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}}}
-\newcommand{\ctxtSem}[1]{\llbracket #1 \rrbracket^{\mathcal{C}}}
-\newcommand{\tmSem}[1]{\llbracket #1 \rrbracket^{\mathit{tm}}}
-\newcommand{\tyPrimSem}[1]{\llbracket \tyPrimNm{#1} \rrbracket^{\mathcal{T}_0}}
-\newcommand{\rsem}[3]{\llbracket #1 \rrbracket^{\mathcal{R}}_{#2}{#3}}
-\newcommand{\extends}[2]{\mathsf{ext}(#1,#2)}
-
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
-\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
-\newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
-\newcommand{\corref}[1]{\hyperref[#1]{Corollary \ref*{#1}}}
-\newcommand{\conref}[1]{\hyperref[#1]{Conjecture \ref*{#1}}}
-\newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-\end{abstract}
-
-\category{D.1.1}{Programming techniques}{Applicative (functional)
-  programming} \category{D.2.4}{Software Engineering}{Software/Program
-  Verification} \category{D.3.3}{Programming Languages}{Language
-  Constructs and Features---Data types and structures}
-
-\terms
-  Languages, Theory, Types
-
-\keywords
-  parametricity, dimension types, computational geometry
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-
-\begin{enumerate}
-\item Introduction using the area-of-a-triangle example
-\item Narrowing of the semantic gap by use of advanced type systems
-\item Importance of the use of symmetry in computer science,
-  mathematics and the physical sciences
-\item Development of the theory of parametricity
-\end{enumerate}
-
-\subsection{Contributions}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Background and Related Work}
-
-Reynolds (Parametricity), Wadler (Free Theorems).
-
-Kennedy (Units of Measure)
-
-Something about symmetry?
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Type System}
-
-We now introduce our type system. We parameterise our development by
-an arbitrary multi-sorted algebraic theory, in the sense of universal
-algebra. We present the precise form of the theories we allow in
-\autoref{sec:index-exp}, and define some auxiliary notions of
-substitution and terms-up-to-equality. Several examples of useful
-algebraic theories will be given in
-\autoref{sec:example-systems}. Once an algebraic theory has been
-fixed, we define the well-formed types and terms with respect to this
-theory in \autoref{sec:types-and-terms}.
-
-\subsection{Algebraic Theories for Indexing}\label{sec:index-exp}
-
-\begin{figure*}[t]
-  \centering
-  Well-sorted terms
-  \begin{mathpar}
-    \inferrule*
-    {i : s \in \Delta}
-    {\Delta \vdash x : s}
-
-    \inferrule*
-    {\indexOp{f} : s_1,...,s_n \to s \\
-      \Delta \vdash e_1 : s_1 \\
-      \dots \\
-      \Delta \vdash e_n : s_n}
-    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) : s}
-  \end{mathpar}
-
-  \medskip
-
-  Equational rules
-  \begin{mathpar}
-    \inferrule*
-    {e_1 \stackrel{ax}= e_2 : s \in \mathit{Ax}(\Delta)}
-    {\Delta \vdash e_1 \equiv e_2 : s}
-
-    \inferrule*
-    {\indexOp{f} : s_1,...,s_n \to s \\
-      \Delta \vdash e_1 \equiv e'_1 : s_1 \\
-      \dots \\
-      \Delta \vdash e_n \equiv e'_n : s_n}
-    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) \equiv \indexOp{f}(e'_1, ..., e'_n) : s}
-
-    \inferrule*
-    {\Delta \vdash e : s}
-    {\Delta \vdash e \equiv e : s}
-
-    \inferrule*
-    {\Delta \vdash e_1 \equiv e_2 : s}
-    {\Delta \vdash e_2 \equiv e_1 : s}
-
-    \inferrule*
-    {\Delta \vdash e_1 \equiv e_2 : s \\ \Delta \vdash e_2 \equiv e_3 : s}
-    {\Delta \vdash e_1 \equiv e_3 : s}
-  \end{mathpar}
-  \caption{Algebraic Theories for Indexing}
-  \label{fig:alg-theories}
-\end{figure*}
-
-We assume a set $\mathit{Sort}$ of \emph{index sorts}, and we will use
-$s, s_1, s_2, ...$ to range over elements of $\mathit{Sort}$. An index
-context $\Delta$ is list of index variable and sort pairs, $i : s$,
-and we use the symbol $\epsilon$ to denote the empty index context (we
-will also use this symbol below to denote the empty typing
-context). We use letters like $i$ and $j$, sometimes with subscripts,
-to range over index variables. We also assume a collection
-$\mathit{IndexOp}$ of sorted index operations $\indexOp{f} :
-s_1,...,s_n \to s$. From the index operations we generate the
-well-sorted index expressions by the rules in
-\autoref{fig:alg-theories}.
-
-For each index context $\Delta$, we assume a set $\mathit{Ax}(\Delta)$
-of axioms $e_1 \stackrel{ax}= e_2 : s$ where both $\Delta \vdash e_1 :
-s$ and $\Delta \vdash e_2 : s$. From this collection of axioms, we
-generate the equational theory of index expressions by the rules in
-\autoref{fig:alg-theories} by closing under congruence, reflexivity,
-symmetry and transitivity.
-
-% For a pair of index contexts $\Delta$ and $\Delta'$, a substitution
-% $\sigma \in \subst{\Delta}{\Delta'}$ is a function $\sigma$ mapping
-% variables in $\Delta$ to well-sorted index expressions in $\Delta'$.
-
-Index context morphisms $\sigma : \Delta \Rightarrow \Delta'$ consist
-of a sequence of terms, one for each element of $\Delta'$, all
-well-sorted in the index context $\Delta$.
-\begin{mathpar}
-  \inferrule*
-  { }
-  {\cdot : \Delta \Rightarrow \epsilon}
-
-  \inferrule*
-  {\sigma : \Delta \Rightarrow \Delta' \\ \Delta \vdash e : s}
-  {\sigma, e : \Delta \Rightarrow \Delta', i:s}
-\end{mathpar}
-Directly from the definition of index context morphisms, we can see
-that given an index context morphism $\sigma : \Delta \Rightarrow
-\Delta'$, then for any index variable $i : s \in \Delta'$ there is an
-index expression $\sigma(i)$ such that $\Delta \vdash \sigma(i) : s$.
-
-We write $\id_\Delta : \Delta \Rightarrow \Delta$ for the index
-context morphism comprised of the terms built from the variables of
-$\Delta$, in order. We write $\pi_{i\mathord:s} : \Delta, i : s
-\Rightarrow \Delta$ for the context morphism again comprised of the
-terms built from the variables of $\Delta$, but this time typed in a
-larger context. For an index context morphism $\sigma : \Delta \to
-\Delta'$ we write $\sigma_s : \Delta, i\mathord:s \Rightarrow \Delta',
-i\mathord:s$ for the index context morphism generated by lifting each
-term to the larger context $\Delta, i\mathord:s$, and mapping the
-variable $i\mathord:s$ to the term $i$.
-
-Given an index expression $\Delta' \vdash e : s$, we define
-re-indexing by an index context morphism $\sigma : \Delta \Rightarrow
-\Delta'$ by the following clauses:
-\begin{eqnarray*}
-  \sigma^*i & = & \sigma(i) \\
-  \sigma^*\indexOp{f}(e_1, ..., e_n) & = & \indexOp{f}(\sigma^*e_1,...,\sigma^*e_n)
-\end{eqnarray*}
-It is easy to check that these clauses preserve well-formedness of
-index expressions, so that we obtain $\Delta \vdash \sigma^*e :
-s$. Moreover re-indexing by index context morphisms preserves
-equality:
-\begin{lemma}\label{lem:indexeq-subst}
-  If $\Delta \vdash e_1 \equiv e_2 : s$, then for any $\sigma :
-  \Delta' \Rightarrow \Delta$, we have $\Delta' \vdash \sigma^*e_1
-  \equiv \sigma^*e_2 : s$.
-\end{lemma}
-
-Also, projection commutes with lifting of substitutions. This is used
-in the proof of \lemref{lem:tysubst-rel}.
-\begin{lemma}
-  $\sigma^* \circ \pi^*_{i\mathord:s} = \pi^*_{i\mathord:s} \circ \sigma^*_s$
-\end{lemma}
-
-For an index context $\Delta$ and sort $s$, we write
-$\idxTms{\Delta}{s}$ for the set of index expressions $e$ such that
-$\Delta \vdash e : s$, quotiented by the equivalence relation $\Delta
-\vdash - \equiv - : s$. In light of \lemref{lem:indexeq-subst}, we can
-see that re-indexing by an index context morphism $\sigma : \Delta
-\Rightarrow \Delta'$ can be used as a function $\sigma^* :
-\idxTms{\Delta'}{s} \to \idxTms{\Delta}{s}$, for any sort $s$.
-
-\subsection{Types and Terms}\label{sec:types-and-terms}
-
-\begin{figure*}[t]
-  \centering
-  Well-indexed Types
-  \begin{mathpar}
-    \inferrule* [right=TyPrimIdx]
-    {\tyPrimNm{X} \in \mathit{PrimType}(s) \\ \Delta \vdash e : s}
-    {\Delta \vdash \tyPrim{X}{e} : \sortType}
-
-    \inferrule* [right=TyPrimNoIdx]
-    {\tyPrimNm{X} \in \mathit{PrimType}(\cdot)}
-    {\Delta \vdash \tyPrimNm{X} : \sortType}
-
-    \inferrule* [right=TyUnit]
-    { }
-    {\Delta \vdash \tyUnit : \sortType}
-
-    \inferrule* [right=TyArr]
-    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
-    {\Delta \vdash A \tyArr B : \sortType}
-
-    \inferrule* [right=TyProd]
-    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
-    {\Delta \vdash A \tyProduct B : \sortType}
-
-    \inferrule* [right=TySum]
-    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
-    {\Delta \vdash A + B : \sortType}
-
-    \inferrule* [right=TyForall] %FIXME: macroize forall
-    {\Delta, i : s \vdash A : \sortType}
-    {\Delta \vdash \forall i \mathord: s. A : \sortType}
-  \end{mathpar}
-
-  \medskip
-  Equality of well-indexed types
-  \begin{mathpar}
-    \inferrule* [right=TyEqPrimIdx]
-    {\Delta \vdash e_1 \equiv e_2 : s}
-    {\Delta \vdash \tyPrim{X}{e_1} \equiv \tyPrim{X}{e_2} : \sortType}
-
-    \inferrule* [right=TyEqPrimNoIdx]
-    { }
-    {\Delta \vdash \tyPrimNm{X} \equiv \tyPrimNm{X} : \sortType}
-
-    \inferrule* [right=TyEqUnit]
-    { }
-    {\Delta \vdash \tyUnit \equiv \tyUnit : \sortType}
-
-    \inferrule* [right=TyEqArr]
-    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
-    {\Delta \vdash A \tyArr B \equiv A' \tyArr B' : \sortType}
-
-    \inferrule* [right=TyEqProd]
-    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
-    {\Delta \vdash A \tyProduct B \equiv A' \tyProduct B' : \sortType}
-
-    \inferrule* [right=TyEqSum]
-    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
-    {\Delta \vdash A + B \equiv A' + B' : \sortType}
-
-    \inferrule* [right=TyEqForall]
-    {\Delta, i : s \vdash A \equiv A' : \sortType}
-    {\Delta \vdash \forall i \mathord: s. A \equiv \forall i \mathord: s. A' : \sortType}
-  \end{mathpar}
-
-  \medskip
-  Well-indexed contexts
-  \begin{mathpar}
-    \inferrule*
-    { }
-    {\Delta \vdash \epsilon \isCtxt}
-
-    \inferrule*
-    {\Delta \vdash \Gamma \isCtxt \\ \Delta \vdash A : \sortType \\ x \not\in \Gamma}
-    {\Delta \vdash \Gamma, x : A \isCtxt}
-  \end{mathpar}
-  \caption{Well-indexed Types and Contexts}
-  \label{fig:types}
-\end{figure*}
-
-For each sort $s \in \mathit{Sort}$, we assume we are given a set of
-names of primitive types $\mathit{PrimType}(s)$. We assume that the
-sets $\mathit{PrimType}(s)$ are disjoint for non-equal sorts $s_1,
-s_2$. For example, in the units of measure example in
-\autoref{sec:units-example}, there is a single sort $\mathsf{unit}$,
-and $\mathit{PrimType}(\mathsf{unit}) = \{\tyPrimNm{num}\}$. For each
-primitive type $\tyPrimNm{X} \in \mathit{PrimType}(s)$, and index
-expression $\Delta \vdash e : s$, we form the type $\tyPrim{X}{e}$, by
-the \TirName{TyPrimIdx} rule in \autoref{fig:types}. Thus, in the
-units-of-measure example, the type $\tyPrim{num}{u}$ represents
-numbers measured by the units $u$.
-
-We also assume that there is an additional set of primitive types
-$\mathit{PrimType}(\cdot)$ that is disjoint from all the other sets of
-primitive types. These primitive types do not need to be paired with
-an index expression to form well-indexed types. Such types are created
-by the \TirName{TyPrimNoIdx} rule. An example of such a primitive type
-is the type $\tyPrimNm{real}$ of real-valued scalars in the Euclidean
-plane geometry example in \autoref{sec:2d-example}.
-
-The rules for generating the rest of the types of the system,
-including unit, product, sum, function and universal types, are given
-in \autoref{fig:types}. We use capital roman letters like $A$ and $B$
-to range over well-indexed types. Well-indexed types come with an
-equational theory, derived from the equational theory on index
-expressions. This equational theory is also presented in
-\autoref{fig:types}.
-
-We define substitution by an index context morphism $\sigma : \Delta
-\Rightarrow \Delta'$ on types by structural induction on the structure
-of the type: given a type $\Delta' \vdash A : \sortType$, we obtain a
-type $\Delta \vdash \sigma^*A : \sortType$. The only interesting case
-is for \TirName{TyPrimIdx}:
-\begin{displaymath}
-  \sigma^*\tyPrim{X}{e} = \tyPrim{X}{\sigma^*e}
-\end{displaymath}
-
-Well-formed contexts are defined to be lists of variables paired with
-well-formed types with respect to some index context $\Delta$, with
-the condition that no variable names are repeated. Formally, contexts
-are defined by the two rules shown in
-\autoref{fig:types}. Substitution by an index context morphism extends
-to contexts just by doing substitution on each individual type.
-
-The typing rules for our system define the well-typed terms with
-respect to an index context $\Delta$ and a type context $\Delta \vdash
-\Gamma \isCtxt$. The judgement $\Delta; \Gamma \vdash M : A$ is
-defined in \autoref{fig:terms}. The equational theory on types is
-incorporated into the type system via the rule \TirName{TyEq}, which
-allows for a term that is judged to have type $A$ to also have any
-other equal type $B$.
-\begin{figure*}[t]
-  \centering
-  \begin{mathpar}
-    \inferrule* [right=Var]
-    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
-    {\Delta; \Gamma \vdash x : A}
-
-    \inferrule* [right=TyEq]
-    {\Delta; \Gamma \vdash M : A \\ \Delta \vdash A \equiv B : \sortType}
-    {\Delta; \Gamma \vdash M : B}
-
-    \inferrule* [right=Unit]
-    { }
-    {\Delta; \Gamma \vdash * : 1}
-
-    \inferrule* [right=Pair]
-    {\Delta; \Gamma \vdash M : A \\
-      \Delta; \Gamma \vdash N : B}
-    {\Delta; \Gamma \vdash (M, N) : A \tyProduct B}
-
-    \inferrule* [right=Proj1]
-    {\Delta; \Gamma \vdash M : A \tyProduct B}
-    {\Delta; \Gamma \vdash \pi_1 M : A}
-
-    \inferrule* [right=Proj2]
-    {\Delta; \Gamma \vdash M : A \tyProduct B}
-    {\Delta; \Gamma \vdash \pi_2 M : B}
-
-    \inferrule* [right=Inl]
-    {\Delta; \Gamma \vdash M : A}
-    {\Delta; \Gamma \vdash \mathrm{inl}\ M : A + B}
-
-    \inferrule* [right=Inr]
-    {\Delta; \Gamma \vdash M : B}
-    {\Delta; \Gamma \vdash \mathrm{inr}\ M : A + B}
-
-    \inferrule* [right=Cond]
-    {\Delta; \Gamma \vdash M : A + B \\
-      \Delta; \Gamma, x : A \vdash N_1 : C \\
-      \Delta; \Gamma, y : B \vdash N_2 : C}
-    {\Delta; \Gamma \vdash \textrm{case}\ M\ \textrm{of}\ \textrm{inl}\ x.N_1; \textrm{inr}\ y.N_2 : C}
-
-    \inferrule* [right=Abs]
-    {\Delta; \Gamma, x : A \vdash M : B}
-    {\Delta; \Gamma \vdash \lambda x.M : A \tyArr B}
-
-    \inferrule* [right=App]
-    {\Delta; \Gamma \vdash M : A \tyArr B \\
-      \Delta; \Gamma \vdash N : A}
-    {\Delta; \Gamma \vdash M N : B}
-
-    \inferrule* [right=UnivAbs]
-    {\Delta, i : s; \pi_{i\mathord:s}^*\Gamma \vdash M : A}
-    {\Delta; \Gamma \vdash \Lambda i. M : \forall i\mathord:s. A}
-
-    \inferrule* [right=UnivApp]
-    {\Delta; \Gamma \vdash M : \forall i\mathord:s. A \\ \Delta \vdash e : s}
-    {\Delta; \Gamma \vdash M [e] : (\id_\Delta, e)^*A}
-  \end{mathpar}
-
-  \caption{Well-typed terms}
-  \label{fig:terms}
-\end{figure*}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Example Systems}\label{sec:example-systems}
-
-\subsection{Units of Measure}\label{sec:units-example}
-
-We now show that our general framework encompasses Kennedy's
-units-of-measure system \cite{kennedy97relational}. For our indexing
-algebraic theory, we have a single sort $\mathsf{unit}$ intended to
-represent composite units such as $\mathit{m}\ \mathit{s}^{-1}$ or
-$\mathit{kg}\ \mathit{m}^{-2}$. Units of measure form an abelian
-group, so we have three operations:
-\begin{mathpar}
-  1 : \mathsf{unit},
-
-  \cdot : \mathsf{unit}, \mathsf{unit} \to \mathsf{unit},
-
-  -^{-1} : \mathsf{unit} \to \mathsf{unit},
-\end{mathpar}
-subject to the normal abelian group axioms. We will elide the group
-multiplication operation and just write multiplication as
-juxtaposition: $u_1u_2$.
-
-We let $\mathit{PrimType}(\mathsf{unit}) = \{\tyPrimNm{num}$, and
-$\mathit{PrimType}(\cdot) = \{\tyPrimNm{rational}\}$. The intended
-reading is that the type $\tyPrim{num}{u_1u^2_2}$ represents
-quantities in units of $u_1u^2_2$, while values of type
-$\tyPrimNm{rational}$ represent dimensionless constants. We make a
-distinction between $\tyPrim{num}{1}$ and $\tyPrimNm{rational}$ for
-technical reasons arising from the relational interpretation below: we
-want the relation on $\tyPrim{num}{1}$ to be the identity relation,
-and the existence of $\mathit{toRat}$ and $\mathit{fromRat}$ forces
-this\footnote{Kennedy assumes that all relational environments set the
-  relation on $\tyPrim{num}{1}$ to be the identity relation, but we
-  cannot do this in general, because we do not necessarily have a
-  group structure in the indexing theory.}. To write programs using
-the $\tyPrimNm{num}$ type, we assume an typing context
-$\Gamma_{\textrm{units}}$ consisting of the following typed
-operations:
-\begin{eqnarray*}
-  \mathit{toRat} & : & \tyPrim{num}{1} \to \tyPrimNm{rational} \\
-  \mathit{fromRat} & : & \tyPrim{num}{1} \to \tyPrimNm{rational} \\
-  0 & : & \forall u. \tyPrim{num}{u} \\
-  1 & : & \tyPrim{num}{1} \\
-  + & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyPrim{num}{u} \\
-  - & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyPrim{num}{u} \\
-  * & : & \forall u_1 u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to \tyPrim{num}{u_1u_2} \\
-  / & : & \forall u_1 u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to \tyPrim{num}{u_1u_2^{-1}} + 1 \\
-  < & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyBool
-\end{eqnarray*}
-There are constants for zero and one, but only the constant for zero
-is polymorphic in the units. In \autoref{sec:units-semantics} below,
-we will following Kennedy by interpreting units of measure in terms of
-\emph{scalings} by rational numbers, where it will be apparent that
-the constant one cannot be unit-polymorphic: we can scale $0$ and
-still get $0$, but we cannot scale $1$ by an arbitrary rational number
-and expect to get $1$.
-
-\subsection{Euclidean Plane Geometry}\label{sec:2d-example}
-
-We now consider a type system for writing programs that are invariant
-under the translational symmetries of the Euclidean plane. We assume a
-single sort $\mathsf{translation}$. The sort $\mathsf{translation}$ is
-intended to represent translations in the plane, and comes equipped
-with abelian group structure that we write additively:
-\begin{mathpar}
-  0 : \mathsf{translation}
-
-  + : \mathsf{translation},\mathsf{translation} \to \mathsf{translation}
-
-  - : \mathsf{translation} \to \mathsf{translation}
-\end{mathpar}
-In indexing expressions involving variables of sort
-$\mathsf{translation}$, we use the shorthand $t_1 - t_2$ to stand for
-$t_1 + (-t_2)$.
-
-There are two primitive types:
-$\mathit{PrimType}(\mathsf{translation}) = \{\tyPrimNm{vec}\}$, and
-$\mathit{PrimType}(\cdot) = \{\tyPrimNm{real}\}$. The types
-$\tyPrim{vec}{t}$ represent two-dimensional vectors, indistinguishable
-up to some translation $t$. Values of type $\tyPrimNm{real}$ are just
-real numbers. To write programs using these types we assume a typing
-context $\Gamma_{\mathit{2D}}$ consisting of the following operations:
-\begin{eqnarray*}
-  + & : & \forall t_1 t_2\mathord:\mathsf{translation}.\tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2} \\
-  - & : & \forall t\mathord:\mathsf{translation}. \tyPrim{vec}{t} \to \tyPrim{vec}{-t} \\
-  0 & : & \tyPrim{vec}{0} \\
-  * & : & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0} \\
-  \mathrm{getX} & : & \tyPrim{vec}{0} \to \tyPrimNm{real} \\
-  \mathrm{getY} & : & \tyPrim{vec}{0} \to \tyPrimNm{real} \\
-  \mathrm{vec}  & : & \tyPrimNm{real} \to \tyPrimNm{real} \to \tyPrim{vec}{0}
-\end{eqnarray*}
-The first three operations express the abelian group structure of
-2-dimensional real vectors. Note how the indexing tracks the
-operations on vectors: e.g.~addition on vectors is reflected by
-addition on the indicies. The final four operations are operations
-that are not invariant under arbitrary translation: multiplication by
-a scalar, and projection from and constructon of vectors. For the
-purposes of examples, we also assume a standard set of arithmetic and
-other operations on real numbers.
-
-As an example of the use of the above operations, consider the
-following type of a function for computing the area of a triangle
-described by three points. Since the area of a triangle is invariant
-under translations that are applied to all three points, we can
-quantify over all translations $t$:
-\begin{displaymath}
-  \mathit{area} : \forall t\mathord:\mathsf{translation}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real}
-\end{displaymath}
-Since the result is invariant under translation, it safe to assume
-that one of the vertexes of the triangle is the origin, and only
-specify two points to describe the triangle. So we might expect that
-the type of $\mathit{area}$ is isomorphic to this type:
-\begin{displaymath}
-  \mathit{area'} : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrimNm{real}
-\end{displaymath}
-Given $\mathit{area'}$, we can define $\mathit{area}$ using the
-operations in $\Gamma_{\mathit{2D}}$:
-\begin{displaymath}
-  \mathit{area} = \Lambda t. \lambda v_1 v_2 v_3.\ \mathit{area'}\ (v_2 - v_1)\ (v_3 - v_1)
-\end{displaymath}
-and vice versa, we can define $\mathit{area'}$ in terms of
-$\mathit{area}$:
-\begin{displaymath}
-  \mathit{area'} = \lambda v_1 v_2. \mathit{area}\ [0]\ 0\ v_1\ v_2
-\end{displaymath}
-In \autoref{sec:2d-semantics} below, we will show that these two
-translations form a type isomorphism between the types of
-$\mathit{area}$ and $\mathit{area'}$.
-
-\paragraph{Alternative Transformation Groups?}
-We should also consider other groups of transformations for indexing:
-isometries, orthogonal transformations, rotations and reflections.
-
-\subsection{Security Typing and Non-interference}
-
-This example shows how our framework can accommodate non-interference
-properties common in security type systems. This example has a
-different flavour to the two previous examples because it does not
-concern invariance under the action of some group. We assume a single
-sort $\mathsf{principal}$, intended to represent (possibly composite)
-principals in a system. Principals are combined using the connectives
-of boolean logic:
-\begin{mathpar}
-  \top : \mathsf{principal}
-
-  \bot : \mathsf{principal}
-
-  \land : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}
-
-  \lor : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}
-
-  \Rightarrow : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}
-
-  \lnot : \mathsf{principal} \to \mathsf{principal}
-\end{mathpar}
-For the equational theory of principals, we assume all the axioms of
-boolean algebra.
-
-Let $\mathit{PrimType}(\mathsf{principal}) = \{\tyPrimNm{T}\}$. The
-intended reading of a type $\tyPrim{T}{p}$ for some principal $p$ is
-that the existence of a value of this type indicates that the
-principal $p$ is ``happy''. To write programs, we assume the following
-typing context $\Gamma_{\mathit{sec}}$ of operations:
-\begin{eqnarray*}
-  \mathrm{proj} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1 \land p_2} \to \tyPrim{T}{p_1} \\
-  \mathrm{combine} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1} \to \tyPrim{T}{p_2} \to \tyPrim{T}{p_1 \land p_2}
-\end{eqnarray*}
-
-To illustrate the use of this application of our framework, we define
-a type synonym, for each type $A$ and principal $p$:
-\begin{displaymath}
-  T_pA = \tyPrim{T}{p} \to A
-\end{displaymath}
-For every principal $p$, we can endow the types $T_p-$ with the
-structure of a monad. This is due to the fact that it is an instance
-of the ``environment'' (or ``reader'') monad. Explicitly, the
-$\mathit{return}$ and $>>=$ of these monads are given by:
-\begin{eqnarray*}
-  \mathit{return} & : & A \to T_pA \\
-  \mathit{return} & = & \lambda a\ t.\ a \\
-  \\
-  (>>=) & : & T_pA \to (A \to T_pB) \to T_pB \\
-  (>>=) & = & \lambda a\ f\ t.\ f\ (a\ t)\ t
-\end{eqnarray*}
-The intended reading of a value of type $T_pA$ is as a piece of data
-of type $A$ that can only be accessed if the principal $p$ is
-happy. The existence of such a monad should allow us to draw parallels
-between our system and the Dependency Core Calculus (DCC) system of
-Abadi \emph{et al.} \cite{abadi99core}. The definition of a DCC monad
-as an instance of an environment monad follows Tse and Zdancewic's
-interpretation of DCC in System F, where they use System F
-parametricity to ensure non-interference properties.
-
-There is also a coercion operator between instances of $T_pA$ for
-different principals. Reading the type in terms of happiness, this
-says that if we have a piece of data that is protected by $p$, and
-happiness of $q$ implies happiness of $p$, then this data can also be
-accessed by $q$.
-\begin{eqnarray*}
-  \mathit{coerce} & : & \tyPrim{T}{q \Rightarrow p} \to T_pA \to T_qA \\
-  \mathit{coerce} & = & \lambda t_{qp}\ a\ t_q.\ a\ (\mathrm{combine}\ t_{qp}\ t_q)
-\end{eqnarray*}
-Here we have used the fact that the result of the application of
-$\mathrm{combine}$ has type $\tyPrim{T}{(q \Rightarrow p) \land
-  q}$. By the laws of boolean algebras---which we have assumed for our
-indexing theory---this is equal to the type $\tyPrim{T}{p}$. Thus we
-get access to the data stored in $a$, which is protected by $p$.
-
-Now suppose we have a term $M$ with the following typing:
-\begin{displaymath}
-  p,q:\mathsf{principal}; t_{qp} : \tyPrim{T}{q \Rightarrow p}, x : T_q (1 + 1) \vdash M : T_p (1 + 1)
-\end{displaymath}
-Intuitively, given the operations $\Gamma_{\mathit{sec}}$ we have
-defined above, it should be the case that $M$ can only be equivalent
-to the program which is constantly $\lambda t. \mathrm{inl}(*)$ or
-$\lambda t. \mathrm{inr}(*)$, because there is no useful data to be
-gained from the $t_{qp}$ argument, and a value of $\tyPrim{T}{q}$
-cannot be generated in order to get access to the data within the $x$
-argument. The application of our framework that we have presented in
-this section has a ``non-interference'' property: the value of $x$
-cannot interfere with the output. In \autoref{sec:sec-semantics} we
-show that this is indeed the case.
-
-\paragraph{Alternative Logics?}
-Above, we assumed that the indexing theory of principals was the
-theory of boolean algebras. As a consequence, a value of type $T_{p
-  \lor \lnot p}A$ can always be accessed, due to the exlcuded middle
-property of boolean logic. An interesting question is: what happens we
-we consider alternative logics. If we take the theory of Heyting
-algebras for our theory of principals, then data typed as $T_{p \lor
-  \lnot p}A$ can only be accessed if either $p$ or $\lnot p$ can be
-demonstrated. Considering an alternative logical theory alters the
-non-interference property of the type system.
-
-\paragraph{Alternative Presentations?}
-In the above we presented the theory of boolean logic by means of the
-equational theory of boolean algebras. This isn't necessarily the most
-convenient way of presenting a logic. An alternative approach would be
-to consider a trivial equational theory with no axioms, and
-incorporate the rules of logic consequence as operations in
-$\Gamma_{\mathit{sec}}$. For example, as a Hilbert system:
-\begin{eqnarray*}
-  \mathit{MP} & : & \forall p,q.\ \tyPrim{T}{p \Rightarrow q} \to \tyPrim{T}{p} \to \tyPrim{T}{q} \\
-  \mathit{k} & : & \forall p,q.\ \tyPrim{T}{p \Rightarrow q \Rightarrow p} \\
-  \mathit{s} & : & \forall p,q,r.\ \tyPrim{T}{(p \Rightarrow q \Rightarrow r) \Rightarrow (p \Rightarrow q) \Rightarrow p \Rightarrow r}
-\end{eqnarray*}
-Plus axioms for the other connectives.  Or possibly some kind of
-encoding of a natural deduction system?
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   following diagram commute, for all primitive types $\tyPrimNm{X} \in
-  \mathit{PrimType}(s)$:
+  \mathit{PrimType}$:
   \begin{displaymath}
@@
       &
-      {\Rel(X)}
+      {\Rel(\tyPrimSem{X})}
     }
@@
 a relation interpretation of any type $\Delta \vdash A : \sortType$ by
-induction its derivation. So we define a relation
-$\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-formed
-type $A$.
+induction its derivation. Thus, we define a relation
+$\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-indexed
+type.
 \begin{eqnarray*}
   \rsem{\tyUnit}{\relEnv{E}}\rho & = & \{(*,*)\} \\
-  \rsem{\tyPrim{X}{e}}{\relEnv{E}}\rho & = & \rho_{\tyPrimNm{X}}(e) \\
+  \rsem{\tyPrim{X}{e_1,...,e_n}}{\relEnv{E}}\rho & = & \rho\ {\tyPrimNm{X}}\ (e_1,...,e_n) \\
   \rsem{A \tyArr B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
@@
 \end{lemma}
-
-\begin{theorem}[Parametricity]
-  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
-  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
-  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
-  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
-\end{theorem}
-
-\begin{proof}
-  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
-  three interesting cases follow:
-  \begin{description}
-  \item[Case \TirName{TyEq}] This case follows directly from
-    \lemref{lem:tyeq-rel}.
-  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
-    $A = \forall i\mathord:s. A'$. We know from the induction
-    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
-    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}\rho'$,
-    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho'$. We are given $\rho \in
-    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
-    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
-    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
-    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
-    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
-      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
-    \rsem{\pi_{i:s}^*\Gamma}\rho''$ by \lemref{lem:tysubst-rel}. So we
-    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}\rho''$, and thus
-    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
-      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
-      i.M'}\eta_2) \in \rsem{\forall
-      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
-    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
-    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
-    know that for all $\rho' \in \extends{\rho}{i:s}$,
-    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho'$. If we let $\rho' = \rho \circ
-    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
-    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
-    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
-    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
-    by \lemref{lem:tysubst-rel}, conclude that $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
-    as required.
-  \end{description}
-  The remaining cases are standard for proofs of the fundamental lemma
-  of logical relations for simply-typed systems.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Consequences of Parametricity}
-
-We now give instances of our semantic framework for each of the three
-examples from \autoref{sec:example-systems}. For each of the three
-examples, we defined an indexing theory, a collection of primitive
-types, and a collection of primitive operations $\Gamma_X$. Below, we
-assign the ``obvious'' index-erasure semantics to each of the
-primitive types and operations: so we have $\tyPrimSem{X}$ for each
-primitive type $\tyPrimNm{X}$, and a $\eta_X \in \ctxtSem{\Gamma_X}$.
-
-The key question we wish to answer below is whether there is a family
-of sets of relation environments $\{\relEnv{E}_X(\Delta)\}_{\Delta}$
-that satisfies the following completeness property identified by
-Kennedy.
-\begin{equation}\label{eq:kennedy-completeness}
-  (\forall \Delta. \forall \rho \in \relEnv{E}'(\Delta). (\eta_X, \eta_X) \in \rsem{\Gamma_X}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_X
-\end{equation}
-where the subset relation on families of sets of relation environments
-is defined as $\relEnv{E}_1 \subseteq \relEnv{E}_2$ if $\forall
-\Delta. \relEnv{E}_1(\Delta) \subseteq \relEnv{E}_2(\Delta)$.
-
-Such a completeness result allows us to derive inexpressibility
-results for a given instance of our general framework. Kennedy
-demonstrates this for the units of measure system by showing that a
-putative square root function of type $\forall u.\ \tyPrim{num}{u^2}
-\to \tyPrim{num}{u}$ can only ever be the constantly zero function (or
-non-terminating).
-
-\subsection{Units of Measure}\label{sec:units-semantics}
-
-For the units of measure example, we set $\tyPrimSem{num} =
-\tyPrimSem{rational} = \mathbb{Q}$, the set of rational numbers. We
-define $\eta_{\mathit{units}} \in \Gamma_{\mathit{units}}$ are defined
-to be the standard arithmetical and comparison operations on rational
-numbers, and $\mathit{toRat}$ and $\mathit{fromRat}$ are defined to be
-the identity functions. The division function (of type $/ : \forall
-u_1,u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to
-\tyPrim{num}{u_1u^{-1}_2} + 1$) returns $\mathrm{inr}(*)$ only when
-its second argument is zero.
-
-Kennedy defines the following family of sets of relational
-environments. Note that, since we have assumed the abelian group
-axioms, the set $\idxTms{\Delta}{\mathsf{unit}}$ of indexing
-expressions quotiented by the equational theory, is an abelian group.
-\begin{displaymath}
-  \relEnv{E}_{\mathit{units}}(\Delta) = \left\{ \rho^{G,h} \sepbar
-    \begin{array}{l}
-      G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{unit}} \\
-      h : G \to \mathbb{Q}^{>0} \\
-      h\textrm{ is a group homomorphism}
-    \end{array}
-  \right\}
-\end{displaymath}
-where $\mathbb{Q}^{>0} = \{ q \in \mathbb{Q} \sepbar q > 0 \}$ and
-\begin{displaymath}
-  \rho^{G,h}_{\tyPrimNm{num}}(e) = \left\{
-    \begin{array}{ll}
-      \{(0,0)\} & \textrm{if }e \not\in G \\
-      \{(r,h(e)r) \sepbar r \in \mathbb{Q} \} & \textrm{if }u \in G
-    \end{array}
-  \right.
-\end{displaymath}
-
-\begin{theorem}
-  The family of sets of relational environments
-  $\relEnv{E}_{\mathit{units}}$ satisfies the completeness property
-  \hyperref[eq:kennedy-completeness]{(\ref*{eq:kennedy-completeness})}
-  above. That is, for all $\relEnv{E}'$:
-  \begin{displaymath}
-    (\forall \Delta.\forall \rho \in \relEnv{E}'(\Delta).
-(\eta_{\mathit{units}}, \eta_{\mathit{units}}) \in \rsem{\Gamma_{\mathit{units}}}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_{\mathit{units}}
-  \end{displaymath}
-\end{theorem}
-
-\begin{proof}
-  (Following Kennedy).  From right to left, we have to check that for
-  each $\rho \in \relEnv{E}'(\Delta)$, where $\relEnv{E}' \subseteq
-  \relEnv{E}_{\mathit{units}}$, that each of the semantic
-  interpretations of the operations satisfies
-  $(\eta_{\mathit{units}}(\mathit{op}),
-  \eta_{\mathit{units}}(\mathit{op})) \in
-  \rsem{\Gamma(\mathit{op})}{\relEnv{E}'}\rho$. For instance,
-  multiplication has to satisfy: for all subgroups $G \subseteq
-  \idxTms{\Delta,u_1,u_2}{\mathsf{unit}}$ and all group homomorphisms
-  $h : G \to \mathbb{Q}^{>0}$, then $(q_1,q'_1) \in \rho^{G,h}(u_1)$
-  and $(q_2,q'_2) \in \rho^{G,h}(u_2)$ implies that $(q_1q_2,q'_1q'_2)
-  \in \rho^{G,h}(u_1u_2)$. There are four cases, depending on whether
-  $u_1 \in G$ and whether $u_2 \in G$. If $u_1 \not\in G$, then $q_1 =
-  q'_1 = 0$, so $q_1q_2 = 0$ and $q'_1q'_2 = 0$. Since $(0,0) \in
-  \rho^{G,h}(u')$ for any $u'$, $G$ and $h$, we are done. If $u_1 \in
-  G$ and $u_2 \in G$, then we need to show that $(q_1q_2,
-  h(u_1)q_1h(u_2)q_2) \in \rho^{G,h}(u_1u_2)$. By $h$ being a
-  homomorphism, we have $(q_1q_2, h(u_1)q_1h(u_2)q_2) = (q_1q_1,
-  h(u_1u_2)q_1q_2)$, which is in $\rho^{G,h}(u_1u_2)$. The cases for
-  the other operations are similar.
-
-  From left to right, we have the following properties of any
-  relational environment $\rho_{\tyPrimNm{num}} \in
-  \relEnv{E}'(\Delta)$, derived from the hypothesis that all the
-  operations in $\eta_{\mathit{units}}$ satisfy their types'
-  relational semantics.
-  \begin{enumerate}
-  \item\label{item:obsv-poly} By the existence of the function
-    $\mathit{toRat}$, the relation $\rho_{\tyPrimNm{num}}(1)$ is the
-    identity relation.
-  \item\label{item:zero-poly} For any unit expression $e$, $(0,0) \in
-    \rho_{\tyPrimNm{num}}(e)$.
-  \item \label{item:one-poly} $(1,1) \in \rho_{\tyPrimNm{num}}(1)$.
-  \item \label{item:mult-poly} For any unit expressions $e_1,e_2$,
-    such that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e_1)$ and
-    $(q_2,q'_2) \in \rho_{\tyPrimNm{num}}(e_2)$, we have $(q_1q_2,
-    q'_1q'_2) \in \rho_{\tyPrimNm{num}}(e_1e_2)$.
-  \item\label{item:div-poly} For any unit expressions $e_1,e_2$, such
-    that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e_1)$ and $(q_2,q'_2)
-    \in \rho_{\tyPrimNm{num}}(e_2)$, we have that either $q_2 = q'_2 =
-    0$, or we have that $(q_1/q_2,q'_1,q'_2) \in
-    \rho_{\tyPrimNm{num}}(e_1e_2^{-1})$.
-  \item\label{item:compare-poly} For any unit expression $e$, such
-    that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e)$ and $(q_2,q'_2) \in
-    \rho_{\tyPrimNm{num}}(e)$, we have $q_1 < q_2$ iff $q'_1 < q'_2$.
-  \end{enumerate}
-  We now show that if $(q,q') \in \rho_{\tyPrimNm{num}}(e)$ for some
-  unit expression $e$, then either $q = q' = 0$, or $q' = rq$ for some
-  $r \in \mathbb{Q}^{>0}$, determined by $e$. We know from
-  \ref{item:zero-poly} above that $q = q' = 0$ is a possibility. Now
-  consider what happens if $q \not= 0$: in this case, by
-  \ref{item:zero-poly} and \ref{item:compare-poly}, we know that $0 <
-  q$ iff $0 < q'$, so $q' \not= 0$, and $q'$ has the same sign as
-  $q$. So $q/q'$ exists and is in $\mathbb{Q}^{>0}$. Hence $q = rq'$
-  for $r = q/q'$. So now we know that if $(q,q') \in
-  \rho_{\tyPrimNm{num}}(e)$, and $q \not= 0$ or $q' \not= 0$ then
-  there exists an $r$ such that $q = rq'$.
-
-  We now show that the unit expression $e$ determines the value of
-  $r$. Assume that we have $(q_1,r_1q_1) \in \rho_{\tyPrimNm{num}}(e)$
-  and $(q_2,r_2q_2) \in \rho_{\tyPrimNm{num}}(e)$. By
-  \ref{item:div-poly}, we know that $(q_1/q_2, (r_1q_1)/(r_2q_2)) \in
-  \rho_{\tyPrimNm{num}}(1)$. By \ref{item:obsv-poly},
-  $\rho_{\tyPrimNm{num}}(1)$ is the equality relation, so we can
-  deduce that $r_1 = r_2$.
-
-  Now let $G$ be the subgroup of $\idxTms{\Delta}{\mathsf{unit}}$ such
-  that $e \in G$ implies that $\rho_{\tyPrimNm{num}}(e) \not=
-  \{(0,0)\}$. By the argument above, for each $e \in G$, there is a
-  unique $r \in \mathbb{Q}^{>0}$ such that $\rho_{\tyPrimNm{num}}(e) =
-  \{ (q,rq) \sepbar q \in \mathbb{Q} \}$. Thus there is a function $h
-  : G \to \mathbb{Q}^{>0}$. By \ref{item:one-poly} and
-  \ref{item:mult-poly}, $h$ is a group homomorphism. Thus we have
-  shown that $\rho_{\tyPrimNm{num}}$ has the form of
-  $\rho^{G,h}_{\tyPrimNm{num}}$ for some $G$ and $h$. Since
-  $\rho_{\tyPrimNm{num}}$ was an arbitrary element of
-  $\relEnv{E}'(\Delta)$, and $\Delta$ was arbitrary, we have shown
-  that $\relEnv{E}' \subseteq \relEnv{E}_{\mathit{units}}$.
-\end{proof}
-
-Some notes towards a generalisation:
-\begin{itemize}
-\item By inspecting Kennedy's proof, we can see that the operations of
-  addition and subtraction do not play any role, save for an
-  obligation to prove that they satisfy the relational semantics of
-  their types. This follows just by distributivity of addition over
-  multiplication.
-\item Kennedy's proof does not rely on any particular features of
-  $\mathbb{Q}$ other than the fact that it is an ordered field. By
-  inspecting the proof, it is possible to see that if we remove the
-  comparison operation $< : \forall u. \tyPrim{num}{u} \to
-  \tyPrim{num}{u} \to \tyPrimNm{bool}$ from $\Gamma_{\mathit{units}}$,
-  then the appropriate family of sets of relational environments uses
-  homomorphisms into the set of non-zero rationals:
-  \begin{displaymath}
-    \begin{array}{l}
-      \relEnv{E}_{\mathit{units}\backslash\{<\}}(\Delta) = \\
-      \quad\left\{ \rho^{G,h} \sepbar
-        \begin{array}{l}
-          G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{unit}} \\
-          h : G \to \mathbb{Q}^{\not=0} \\
-          h\textrm{ is a group homomorphism}
-        \end{array}
-      \right\}
-    \end{array}
-  \end{displaymath}
-  To complete the step in the proof where the comparison operation was
-  used, we use the division operation to show that if $(q,q') \in
-  \rho_{\tyPrimNm{num}}(e)$ for some $e$, and $q \not= 0$, then $q'
-  \not= 0$ (but they need not necessarily have the same sign).
-
-  Just as Kennedy's proof generalises to an arbitrary ordered field,
-  if we remove the comparison operation we can generalise to an
-  arbitrary field as the underlying semantics of $\tyPrimNm{num}$ and
-  $\tyPrimNm{rational}$.
-\end{itemize}
-
-\subsection{Euclidean Plane Geometry}\label{sec:2d-semantics}
-
-We set $\tyPrimSem{vec} = \mathbb{R}^2$, and $\tyPrimSem{real} =
-\mathbb{R}$. The semantic interpretation $\eta_{\mathit{2D}}$ of the
-operations in $\Gamma_{\mathit{2D}}$ consists of the standard vector
-zero, addition, negation and scaling in $\mathbb{R}^2$, along with
-projections to interpret $\mathit{getX}$ and $\mathit{getY}$, and
-pairing to interpret $\mathit{vec}$.
-
-Let
-\begin{displaymath}
-  \begin{array}{l}
-    \relEnv{E}_{\mathit{2D}}(\Delta) = \\
-    \quad\left\{ \rho^{G,h} \sepbar
-      \begin{array}{l}
-        G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{translation}} \\
-        h : G \to \mathbb{R}^2 \\
-        h\textrm{ is a group homomorphism}
-      \end{array}
-    \right\}
-  \end{array}
-\end{displaymath}
-where
-\begin{displaymath}
-  \rho^{G,h}_{\tyPrimNm{vec}}(t) = \left\{
-    \begin{array}{ll}
-      \{\} & \textrm{if }t \not\in G \\
-      \{(v,h(t) + v) \sepbar v \in \mathbb{R}^2 \} & \textrm{if }t \in G
-    \end{array}
-  \right.
-\end{displaymath}
-Note that, unlike the units of measure example above, the pair $(0,0)$
-need not appear in every relation. This is because we do not have a
-translation polymorphic constant like $0 : \forall u. \tyPrim{num}{u}$
-in the units of measure example.
-
-\begin{theorem}
-  The family of sets of relational environments
-  $\relEnv{E}_{\mathit{units}}$ satisfies the completeness property
-  \hyperref[eq:kennedy-completeness]{(\ref*{eq:kennedy-completeness})}
-  above. That is, for all $\relEnv{E}'$:
-  \begin{displaymath}
-    (\forall \Delta.\forall \rho \in \relEnv{E}'(\Delta).
-(\eta_{\mathit{2D}}, \eta_{\mathit{2D}}) \in \rsem{\Gamma_{\mathit{2D}}}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_{\mathit{2D}}
-  \end{displaymath}
-\end{theorem}
-
-\begin{proof}
-  From right to left, we must check that each operation satisfies the
-  relational interpretation of its type. This is very similar to the
-  case for the units of measure example above.
-
-  From left to right, we have the following properties of any
-  relational environment $\rho_{\tyPrimNm{vec}} \in
-  \relEnv{E}'(\Delta)$, derived from the hypothesis that all the
-  operations in $\eta_{\mathit{2D}}$ satisfy their types'
-  relational semantics.
-  \begin{enumerate}
-  \item\label{item:obsv-vec-poly} By the existence of the functions
-    $\mathit{getX}$ and $\mathit{getY}$, the relation
-    $\rho_{\tyPrimNm{vec}}(0)$ is the identity relation.
-  \item \label{item:zero-vec-poly} $(0,0) \in \rho_{\tyPrimNm{vec}}(0)$.
-  \item \label{item:add-vec-poly} For any translation expressions $t_1,t_2$,
-    such that $(v_1,v'_1) \in \rho_{\tyPrimNm{vec}}(t_1)$ and
-    $(v_2,v'_2) \in \rho_{\tyPrimNm{vec}}(t_2)$, we have $(v_1 + v_2,
-    v'_1 + v'_2) \in \rho_{\tyPrimNm{vec}}(t_1 + t_2)$.
-  \item\label{item:neg-vec-poly} For any translation expression $t$,
-    such that $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$, we have that
-    $(-v,-v') \in \rho_{\tyPrimNm{vec}}(-t)$.
-  \end{enumerate}
-  We now show that if $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$ for some
-  translation expression $t$, then $v' = w + v$ for some $w \in
-  \mathbb{R}^2$, where $w$ is determined by $t$. If we are given
-  $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$, then we can set $w = v' -
-  v$. This is uniquely determined by $t$: if we have $(v, w + v) \in
-  \rho_{\tyPrimNm{vec}}(t)$ and $(v, w' + v) \in
-  \rho_{\tyPrimNm{vec}}(t)$, then by \ref{item:add-vec-poly} and
-  \ref{item:neg-vec-poly}, we have $(v - v, (w + v) - (w' + v)) \in
-  \rho_{\tyPrimNm{vec}}(0)$. By \ref{item:obsv-vec-poly}, we know that
-  $\rho_{\tyPrimNm{vec}}(0)$ is the identity relation, so we can
-  deduce that $w = w'$.
-
-  Now define $G \subseteq \idxTms{\Delta}{\mathsf{translation}}$ to
-  have $t \in G$ if $\rho_{\tyPrimNm{vec}}(t) \not= \{\}$. Since the
-  translation factor $w$ is determined by $t$, we can define a
-  function $h : G \to \mathbb{R}^2$. By \ref{item:zero-vec-poly} and
-  \ref{item:add-vec-poly}, $h$ is a group homomorphism, and $(v,v')
-  \in \rho_{\tyPrimNm{vec}}$ iff $v = h(t) + v'$. Thus
-  $\rho_{\tyPrimNm{vec}}$ is of the form
-  $\rho^{G,h}_{\tyPrimNm{vec}}$, as required.
-\end{proof}
-
-Some more notes:
-\begin{enumerate}
-\item The proof is slightly simpler than the case for the units of
-  measure example above because the underlying semantics of
-  $\tyPrimNm{vec}$ forms an abelian group, so we can just use inverses
-  to determine the difference factor $w$, and we don't have to deal
-  with the possibility of division by zero.
-\item The proof above should generalise to any abelian group.
-\item Generalising to non-abelian groups does not look
-  straightforward: the step that shows that the translation factor $w$
-  is uniquely determined by the translation expression $t$ fails. We
-  end up with
-  \begin{displaymath}
-    0 = w + v + (-w') + (-v)
-  \end{displaymath}
-  from which it is not possible to conclude that $w = w'$. (Maybe
-  something using normal subgroups would work?)
-\item A slightly different example is obtained by considering
-  rotations rather than translations (we consider rotations as an
-  additive abelian group). In this case, we could have the following
-  typed operations:
-  \begin{eqnarray*}
-    0 & : & \forall r.\ \tyPrim{vec}{r} \\
-    + & : & \forall r.\ \tyPrim{vec}{r} \to \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
-    - & : & \forall r.\ \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
-    * & : & \forall r.\ \tyPrimNm{real} \to \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
-    \mathit{norm} & : & \forall r. \tyPrim{vec}{r} \to \tyPrimNm{real} \\
-    \mathit{angleOf} & : & \forall r. \tyPrim{vec}{r} \to \tyPrim{rot}{r} + 1 \\
-    \mathit{rotateBy} & : & \forall r_1 r_2. \tyPrim{vec}{r_1} \to \tyPrim{rot}{r_2} \to \tyPrim{vec}{r_1 + r_2} \\
-    0_{\mathit{rot}} & : & \tyPrim{rot}{0} \\
-    +_{\mathit{rot}} & : & \forall r_1 r_2. \tyPrim{rot}{r_1} \to \tyPrim{rot}{r_2} \to \tyPrim{rot}{r_1 + r_2} \\
-    -_{\mathit{rot}} & : & \forall r. \tyPrim{rot}{r} \to \tyPrim{rot}{-r}
-  \end{eqnarray*}
-  where $\tyPrimSem{vec} = \mathbb{R}^2$, as before, and
-  $\tyPrimSem{rot} = \mathrm{SO}(2)$, the group of rotations of the
-  plane. The operation $\mathit{angleOf}$ is intended to return
-  $\mathrm{inr}(*)$ when the provided vector is zero.
-
-  In this case, rotations form an abelian group, so we can re-use the
-  proof above to give a family of sets of relational environments that
-  is complete for the group operations on rotations. The fact that
-  rotations are linear admits the typing of the first four
-  operations. The existence of the $\mathit{angleOf}$ and
-  $\mathit{rotateBy}$ functions allow us to determine ``rotation
-  factors'' between arbitrary pairs of vectors.
-
-  I conjecture that the following definition of
-  $\relEnv{E}_{\mathit{2Drot}}$ satisfies the completeness criterion,
-  with the standard interpretation of the operations above.
-  \begin{displaymath}
-    \begin{array}{l}
-      \relEnv{E}_{\mathit{2Drot}}(\Delta) = \\
-      \quad\left\{ \rho^{G,h} \sepbar
-        \begin{array}{l}
-          G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{rotation}} \\
-          h : G \to \mathrm{SO}(2) \\
-          h\textrm{ is a group homomorphism}
-        \end{array}
-      \right\}
-    \end{array}
-  \end{displaymath}
-  where
-  \begin{displaymath}
-    \rho^{G,h}_{\tyPrimNm{vec}}(r) = \left\{
-      \begin{array}{ll}
-        \{(0,0)\} & \textrm{if }t \not\in G \\
-        \{(v,h(r) v) \sepbar v \in \mathbb{R}^2 \} & \textrm{if }t \in G
-      \end{array}
-    \right.
-  \end{displaymath}
-  and
-  \begin{displaymath}
-    \rho^{G,h}_{\tyPrimNm{rot}}(r) = \left\{
-      \begin{array}{ll}
-        \{\} & \textrm{if }t \not\in G \\
-        \{(s,h(r) s) \sepbar s \in \mathrm{SO}(2) \} & \textrm{if }t \in G
-      \end{array}
-    \right.
-  \end{displaymath}
-\end{enumerate}
-
-\subsection{Security Typing}\label{sec:sec-semantics}
-
-In this case, we let $\tyPrimSem{T} = \{*\}$, and the operations
-$\mathit{proj}$ and $\mathit{combine}$ are given the only possible
-interpretations.
-
-I conjecture that the follow definition works:
-\begin{displaymath}
-  \begin{array}{l}
-    \relEnv{E}_{\mathit{sec}}(\Delta) = \\
-    \quad\left\{ \rho^{h} \sepbar
-      \begin{array}{l}
-        h : \idxTms{\Delta}{\mathsf{principal}} \to \{\top,\bot\} \\
-        h\textrm{ is a boolean algebra homomorphism}
-      \end{array}
-    \right\}
-  \end{array}
-\end{displaymath}
-where
-\begin{displaymath}
-  \rho^{h}_{\tyPrimNm{T}}(p) = \left\{
-    \begin{array}{ll}
-      \{\} & \textrm{if }h(p) = \bot \\
-      \{(*,*) \} & \textrm{if }h(p) = \top
-    \end{array}
-  \right.
-\end{displaymath}
-
-One can compare this definition to the previous ones by considering
-the boolean algebra homomorphism as defining a sub-boolean algebra of
-$\idxTms{\Delta}{\mathsf{principal}}$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Extensions}
-
-\subsection{Inductive Types}
-
-\subsection{Type Operators}
-
-\subsection{Size indexing and Combinatorial Species}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{plainnat}
-\bibliography{paper}
-
-\appendix
-
-\section{Details of proofs}
-
-\begin{proof}\textbf{[Of \lemref{lem:tyeq-erasure}]}
-  By induction on the derivation of $\Delta \vdash A \equiv B :
-  \sortType$:
-  \begin{description}
-  \item[Case \TirName{TyEqUnit}] In this case, $A = B = \tyUnit$, so
-    $\tySem{A} = \tySem{B}$.
-  \item[Case \TirName{TyEqPrimIdx}] We have $A = \tyPrim{X}{e_1}$ and
-    $B = \tyPrim{X}{e_2}$, for some $e_1$ and $e_2$ of the same
-    sort. The erasure semantics ignores the index arguments, so
-    $\tySem{A} = \tySem{\tyPrim{X}{e_1}} = \tyPrimSem{X} =
-    \tySem{\tyPrim{X}{e_2}} = \tySem{B}$, as required.
-  \item[Case \TirName{TyEqPrimNoIdx}] We have $A = \tyPrimNm{X}$ and
-    $B = \tyPrimNm{X}$. So $\tySem{A} = \tyPrimSem{X} = \tySem{B}$, as
-    required.
-  \item[Case \TirName{TyEqArr}] Here, $A = A_1 \tyArr A_2$ and $B =
-    B_1 \tyArr B_2$, and by the induction hypothesis we know that
-    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
-    $\tySem{A} = \tySem{A_1 \tyArr A_2} = \tySem{A_1} \to \tySem{A_2}
-    = \tySem{B_1} \to \tySem{B_2} = \tySem{B_1 \tyArr B_2} =
-    \tySem{B}$.
-  \item[Case \TirName{TyEqProd}] In this case, $A = A_1 \tyProduct
-    A_2$ and $B = B_1 \tyProduct B_2$, and by the induction hypothesis
-    we know that $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} =
-    \tySem{B_2}$. Hence $\tySem{A} = \tySem{A_1 \tyProduct A_2} =
-    \tySem{A_1} \times \tySem{A_2} = \tySem{B_1} \times \tySem{B_2} =
-    \tySem{B_1 \tyProduct B_2} = \tySem{B}$.
-  \item[Case \TirName{TyEqSum}] In this case, $A = A_1 + A_2$ and $B =
-    B_1 + B_2$, and by the induction hypothesis we know that
-    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
-    $\tySem{A} = \tySem{A_1 + A_2} = \tySem{A_1} + \tySem{A_2} =
-    \tySem{B_1} + \tySem{B_2} = \tySem{B_1 + B_2} = \tySem{B}$.
-  \item[Case \TirName{TyEqForall}] We have $A = \forall
-    i\mathord:s.A'$ and $B = \forall i\mathord:s.B'$. By the induction
-    hypothesis we know that $\tySem{A'} = \tySem{B'}$, and hence
-    $\tySem{A} = \tySem{\forall i\mathord:s. A'} = \tySem{A'} =
-    \tySem{B'} = \tySem{\forall i\mathord:s. B'} = \tySem{B}$.
-  \end{description}
-\end{proof}
-
-\begin{proof}\textbf{[Of \lemref{lem:tysubst-erasure}]}
-  By induction on the derivation of $\Delta \vdash A : \sortType$.
-  \begin{description}
-  \item[Case \TirName{TyUnit}] Directly from the definition of
-    substitution on types of the form $\tyUnit$, we have
-    $\tySem{\sigma^*\tyUnit} = \tySem{\tyUnit}$, as required.
-  \item[Case \TirName{TyPrimIdx}] Since the index-erasure semantics
-    ignores all index expessions, we have
-    $\tySem{\sigma^*\tyPrim{X}{e}} = \tySem{\tyPrim{X}{\sigma^*e}} =
-    \tyPrimSem{X} = \tySem{\tyPrim{X}{e}}$.
-  \item[Case \TirName{TyPrimNoIdx}] Similar to the previous case.
-  \item[Case \TirName{TyArr}] We have $A = A_1 \tyArr A_2$, and by the
-    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
-    \tySem{A_1}$ and $\tySem{\sigma^*B_1} = \tySem{B_1}$. Thus, by the
-    definition of substitution on arrow types, we have
-    $\tySem{\sigma^*(A_1 \tyArr A_2)} = \tySem{\sigma^*A_1 \tyArr
-      \sigma^*A_2} = \tySem{\sigma^*A_1} \to \tySem{\sigma^*A_2} =
-    \tySem{A_1} \to \tySem{A_2} = \tySem{A_1 \tyArr A_2}$, as
-    required.
-  \item[Case \TirName{TyProd}] We have $A = A_1 \tyProduct A_2$, and
-    by the induction hypothesis, we know that $\tySem{\sigma^*A_1} =
-    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
-    definition of substitution on product types, we have
-    $\tySem{\sigma^*(A_1 \tyProduct A_2)} = \tySem{\sigma^*A_1
-      \tyProduct \sigma^*A_2} = \tySem{\sigma^*A_1} \times
-    \tySem{\sigma^*A_2} = \tySem{A_1} \times \tySem{A_2} = \tySem{A_1
-      \tyProduct A_2}$, as required.
-  \item[Case \TirName{TySum}] We have $A = A_1 + A_2$, and by the
-    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
-    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
-    definition of substitution on sum types, we have
-    $\tySem{\sigma^*(A_1 + A_2)} = \tySem{\sigma^*A_1 + \sigma^*A_2} =
-    \tySem{\sigma^*A_1} + \tySem{\sigma^*A_2} = \tySem{A_1} +
-    \tySem{A_2} = \tySem{A_1 + A_2}$, as required.
-  \item[Case \TirName{TyForall}] We have $A = \forall i\mathord:s.A'$,
-    and by the induction hypothesis we know that, for all $\sigma'$,
-    $\tySem{\sigma'^*A'} = \tySem{A'}$. Therefore, from the definition
-    of substitution for $\forall$-types, we have
-    $\tySem{\sigma^*\forall i\mathord:s.A'} = \tySem{\forall
-      i\mathord:s.\sigma_s^{*}A'} = \tySem{\sigma_s^{*}A'} = \tySem{A'}
-    = \tySem{\forall i\mathord:s.A'}$, where we have relied on the
-    fact that the index-erasure semantics of types ignores the index
-    quantification.
-  \end{description}
-\end{proof}
-
-\begin{proof}\textbf{[Of \lemref{lem:tyeq-rel}]}
-  By induction on the derivation of $\Delta \vdash A \equiv B :
-  \sortType$.
-  \begin{description}
-  \item[Case \TirName{TyEqUnit}] In this case $A = B = \tyUnit$, so
-    the statement is trivially statisfied.
-  \item[Case \TirName{TyEqPrimIdx}] In this case we have $A =
-    \tyPrim{X}{e_1}$ and $B = \tyPrim{X}{e_2}$, with $\Delta \vdash
-    e_1 \equiv e_2 : s$. The index expressions $e_1$ and $e_2$ are in
-    the same equivalence class, so by the definition of relation
-    environments we have that $\rho_{\tyPrimNm{X}}(e_1) =
-    \rho_{\tyPrimNm{X}}(e_2)$ and hence
-    $\rsem{\tyPrim{X}{e_1}}{\relEnv{E}}\rho =
-    \rsem{\tyPrim{X}{e_2}}{\relEnv{E}}\rho$.
-  \item[Case \TirName{TyEqPrimNoIdx}] This case is similar to the case
-    for \TirName{TyEqUnit}.
-  \item[Case \TirName{TyEqArr}] We have $A = A_1 \tyArr A_2$ and $B =
-    B_1 \tyArr B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$
-    and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the induction
-    hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_2}{\relEnv{E}}\rho$. Thus we may reason as follows:
-    $\rsem{A_1 \tyArr A_2}{\relEnv{E}}\rho =
-    \rsem{A_1}{\relEnv{E}}\rho \relArrow \rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho \relArrow \rsem{B_2}{\relEnv{E}}\rho =
-    \rsem{B_1 \tyArr B_2}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{TyEqProd}] We have $A = A_1 \tyProduct A_2$ and
-    $B = B_1 \tyProduct B_2$, with $\Delta \vdash A_1 \equiv B_1 :
-    \sortType$ and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the
-    induction hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
-    $\rsem{A_1 \tyProduct A_2}{\relEnv{E}}\rho =
-    \rsem{A_1}{\relEnv{E}}\rho \relTimes \rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho \relTimes \rsem{B_2}{\relEnv{E}}\rho =
-    \rsem{B_1 \tyProduct B_2}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{TyEqSum}] We have $A = A_1 + A_2$ and $B = B_1 +
-    B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$ and $\Delta
-    \vdash A_2 \equiv B_2 : \sortType$. By the induction hypothesis,
-    we know that $\rsem{A_1}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
-    $\rsem{A_1 + A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}\rho
-    \relSum \rsem{A_2}{\relEnv{E}}\rho = \rsem{B_1}{\relEnv{E}}\rho
-    \relSum \rsem{B_2}{\relEnv{E}}\rho = \rsem{B_1 +
-      B_2}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{TyEqForall}] We have $A = \forall
-    i\mathord:s. A'$ and $B = \forall i\mathord:s.B'$, with $\Delta, i
-    : s \vdash A' \equiv B' : \sortType$. By the induction hypothesis,
-    we know that for all $\rho'$, $\rsem{A'}{\relEnv{E}}\rho' =
-    \rsem{B'}\rho'$. We reason as follows:
-    \begin{eqnarray*}
-      &    & (x_1,x_2) \in \rsem{\forall i\mathord:s. A'}{\relEnv{E}}\rho \\
-      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{A'}{\relEnv{E}}\rho' \\
-      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{B'}{\relEnv{E}}\rho' \\
-      &\iff& (x_1,x_2) \in \rsem{\forall i\mathord:s. B'}{\relEnv{E}}\rho
-    \end{eqnarray*}
-    Thus, by extensionality, we have $\rsem{\forall
-      i\mathord:s.A'}\rho \equiv \rsem{\forall
-      i\mathord:s.B'}{\relEnv{E}}\rho$, as required.
-  \end{description}
-\end{proof}
-
-\begin{proof}\textbf{[Of \lemref{lem:tysubst-rel}]}
-  The first part of the lemma statement is proved by induction on the
-  derivation of $\Delta \vdash A : \sortType$. We analyse each case in
-  turn:
-  \begin{description}
-  \item[Case \TirName{TyUnit}] Directly from the definition of
-    substitution on types of the form $\tyUnit$, and the fact that
-    $\rsem{\tyUnit}{\relEnv{E}}\rho_1 =
-    \rsem{\tyUnit}{\relEnv{E}}\rho_2$ for any pair $\rho_1,\rho_2$ of
-    relation environments, we have
-    $\rsem{\sigma^*\tyUnit}{\relEnv{E}}\rho =
-    \rsem{\tyUnit}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as required.
-  \item[Case \TirName{TyPrimIdx}] We have $A = \tyPrim{X}{e}$, therefore
-    $\rsem{\sigma^*A}{\relEnv{E}}\rho =
-    \rsem{\sigma^*\tyPrim{X}{e}}{\relEnv{E}}\rho =
-    \rsem{\tyPrim{X}{\sigma^*e}}{\relEnv{E}}\rho =
-    \rho_{\tyPrimNm{X}}(\sigma^*e) =
-    \rsem{\tyPrim{X}{e}}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as
-    required.
-  \item[Case \TirName{TyPrimNoIdx}] Similar to the case for
-    \TirName{TyUnit}.
-  \item[Case \TirName{TyArr}] In this case, $A = A_1 \tyArr A_2$ and
-    by the induction hypothesis, we know that $\rsem{\sigma^*A_1}{\relEnv{E}}\rho
-    = \rsem{A_1}{\relEnv{E}}{(\rho \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we reason as
-    follows: $\rsem{\sigma^*(A_1 \tyArr A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1
-      \tyArr \sigma^*A_2}{\relEnv{E}}\rho = \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relArrow
-    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho \circ \sigma^*)}
-    \relArrow \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)} = \rsem{A_1 \tyArr
-      A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as required.
-  \item[Case \TirName{TyProd}] In this case, $A = A_1 \tyProduct A_2$
-    and by the induction hypothesis, we know that
-    $\rsem{\sigma^*A_1}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we
-    reason as follows: $\rsem{\sigma^*(A_1 \tyProduct
-      A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1 \tyProduct
-      \sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relTimes
-    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)} \relTimes \rsem{A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)} = \rsem{A_1 \tyProduct A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)}$, as required.
-  \item[Case \TirName{TySum}] In this case, $A = A_1 + A_2$
-    and by the induction hypothesis, we know that
-    $\rsem{\sigma^*A_1}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we
-    reason as follows: $\rsem{\sigma^*(A_1 +
-      A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1 +
-      \sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relSum
-    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)} \relSum \rsem{A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)} = \rsem{A_1 + A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)}$, as required.
-  \item[Case \TirName{TyForall}] We have $A = \forall
-    i\mathord:s.A'$. We prove the required equation by demonstrating
-    two inclusions. From left-to-right, we know that
-    \begin{equation}
-      \label{eq:tysubst-rel-forall1}
-      (x_1,x_2) \in \rsem{\sigma^*\forall i\mathord:s. A'}{\relEnv{E}}\rho
-    \end{equation}
-    and we must show that $(x_1, x_2) \in \rsem{\forall
-      i\mathord:s. A'}{\relEnv{E}}{(\rho \circ \sigma^*)}$. From
-    (\ref{eq:tysubst-rel-forall1}) and the definition of substitution,
-    we know that
-    \begin{displaymath}
-      \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{\sigma_s^*A'}{\relEnv{E}}\rho'
-    \end{displaymath}
-    By the induction hypothesis, we can deduce that
-    \begin{equation}
-      \label{eq:tysubst-rel-forall2}
-      \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{A'}{\relEnv{E}}{(\rho' \circ \sigma_s^{*})}
-    \end{equation}
-    To show the required inclusion, we must show that for all $\rho_1
-    \in \extends{\rho \circ \sigma^*}{i:s}$, it is the case that
-    $(x_1,x_2) \in \rsem{A'}{\relEnv{E}}{\rho_1}$. We now make use of
-    the pushout property of our family $\relEnv{E}$ of relation
-    environments: we have a $\rho \in \relEnv{E}(\Delta)$ and $\rho_1
-    \in \relEnv{E}(\Delta, i:s)$ such that $\rho_1 \circ \pi_{i:s} =
-    \rho \circ \sigma^*$ (by the assumption that $\rho_1 \in
-    \extends{\rho \circ \sigma^*}{i:s}$), so we can obtain a $\rho'
-    \in \relEnv{E}(\Delta',i:s)$ such that $\rho' \circ \pi_{i:s} =
-    \rho$ (so $\rho' \in \extends{\rho}{i:s}$), and $\rho' \circ
-    \sigma^* = \rho_1$. Therefore we use
-    (\ref{eq:tysubst-rel-forall2}) to deduce that $(x_1,x_2) \in
-    \rsem{A'}{\relEnv{E}}{\rho_1}$, and the inclusion from
-    left-to-right is shown.
-
-    From right-to-left, we must now show
-    (\ref{eq:tysubst-rel-forall2}) under the assumption that
-    \begin{equation}
-      \label{eq:tysubst-rel-forall3}
-      \forall \rho_1 \in \extends{\rho \circ \sigma^*}{i:s}. (x_1,x_2) \in \rsem{A'}{\relEnv{E}}{\rho_1}.
-    \end{equation}
-    Given $\rho' \in \extends{\rho}{i:s}$, we let $\rho_1 = \rho'
-    \circ \sigma^*$. Now, $\rho_1 \circ \pi_{i:s} = \rho' \circ
-    \sigma^* \circ \pi_{i:s} = \rho' \circ \pi_{i:s} \circ \sigma_s^{*}
-    = \rho \circ \sigma_s^{*}$. Thus $\rho_1 \in \extends{\rho \circ
-      \sigma^*}{i:s}$ and we may use (\ref{eq:tysubst-rel-forall3}) to
-    deduce that $(x_1,x_2) \in \rsem{A'}{\relEnv{E}}{(\rho' \circ \sigma_s^{*})}$,
-    as required.
-  \end{description}
-  The second part of the lemma statement follows directly by induction
-  on the derivation of $\Delta \vdash \Gamma \isCtxt$ and the first
-  part. The step case of the induction is very similar to the
-  \TirName{TyProd} case above.
-\end{proof}
-
-\end{document}

### 2012-05-22 paper/old-syntax.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

--- 2012-05-08 paper/paper.tex
@@
-\documentclass[natbib,preprint]{sigplanconf}
-
-\usepackage[usenames]{color}
-\definecolor{citationcolour}{rgb}{0,0.4,0.2}
-\definecolor{linkcolour}{rgb}{0,0,0.8}
-\usepackage{hyperref}
-\hypersetup{colorlinks=true,
-            urlcolor=linkcolour,
-            linkcolor=linkcolour,
-            citecolor=citationcolour,
-            pdftitle=Relational Parametricity for Algebraically-Indexed Types,
-            pdfauthor={Robert Atkey, Neil Ghani, Patricia Johann, Andrew Kennedy},
-            pdfkeywords={}}  
-\def\sectionautorefname{Section}
-\def\subsectionautorefname{Section}
-
-\title{Relational Parametricity for Algebraically Indexed Types}
-
-\authorinfo{Robert Atkey\and Neil Ghani\\\and Patricia Johann}
-           {University of Strathclyde}
-           {\{Robert.Atkey,Neil.Ghani,Patricia.Johann\}@cis.strath.ac.uk}
-
-\authorinfo{Andrew Kennedy}
-           {Microsoft Research}
-           {}
-
-\usepackage{mathpartir}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{stmaryrd}
-\usepackage[all]{xy}
-
-\newcommand{\sepbar}{\mathrel|}
-
-\newcommand{\Rel}{\mathrm{Rel}}
-\newcommand{\relArrow}{\mathrel{\widehat\to}}
-\newcommand{\relTimes}{\mathrel{\widehat\times}}
-\newcommand{\relSum}{\mathrel{\widehat+}}
-
-\newcommand{\setOfIntegers}{\mathbb{Z}}
-\newcommand{\setOfBooleans}{\{\tmTT, \tmFF\}}
-
-\newcommand{\id}{\mathrm{id}}
-
-\newcommand{\indexOp}[1]{\texttt{#1}}
-\newcommand{\idxTms}[2]{\mathrm{IdxTm}(#1 \vdash #2)}
-\newcommand{\idxTmAS}[1]{\mathrm{IdxTm}(#1)}
-
-\newcommand{\sortType}{\texttt{type}}
-\newcommand{\tyInt}{\texttt{int}}
-\newcommand{\tyBool}{\texttt{bool}}
-\newcommand{\tyUnit}{\texttt{unit}}
-\newcommand{\tyPrim}[2]{\texttt{#1}\langle #2 \rangle}
-\newcommand{\tyPrimNm}[1]{\texttt{#1}}
-
-\newcommand{\tyArr}{\to}
-\newcommand{\tyProduct}{\times}
-\newcommand{\tyX}[1]{\texttt{X}\langle #1 \rangle}
-\newcommand{\isType}{\textsf{ type}}
-\newcommand{\isCtxt}{\textsf{ ctxt}}
-\newcommand{\Ty}{\textsf{Ty}}
-\newcommand{\ty}{\textsf{ty}}
-\newcommand{\subst}[2]{\mathrm{Subst}(#1,#2)}
-
-\newcommand{\tmTT}{\texttt{tt}}
-\newcommand{\tmFF}{\texttt{ff}}
-
-\newcommand{\relEnv}[1]{\mathcal{#1}}
-\newcommand{\tySem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}}}
-\newcommand{\ctxtSem}[1]{\llbracket #1 \rrbracket^{\mathcal{C}}}
-\newcommand{\tmSem}[1]{\llbracket #1 \rrbracket^{\mathit{tm}}}
-\newcommand{\tyPrimSem}[1]{\llbracket \tyPrimNm{#1} \rrbracket^{\mathcal{T}_0}}
-\newcommand{\rsem}[3]{\llbracket #1 \rrbracket^{\mathcal{R}}_{#2}{#3}}
-\newcommand{\extends}[2]{\mathsf{ext}(#1,#2)}
-
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
-\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
-\newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
-\newcommand{\corref}[1]{\hyperref[#1]{Corollary \ref*{#1}}}
-\newcommand{\conref}[1]{\hyperref[#1]{Conjecture \ref*{#1}}}
-\newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-\end{abstract}
-
-\category{D.1.1}{Programming techniques}{Applicative (functional)
-  programming} \category{D.2.4}{Software Engineering}{Software/Program
-  Verification} \category{D.3.3}{Programming Languages}{Language
-  Constructs and Features---Data types and structures}
-
-\terms
-  Languages, Theory, Types
-
-\keywords
-  parametricity, dimension types, computational geometry
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-
-\begin{enumerate}
-\item Introduction using the area-of-a-triangle example
-\item Narrowing of the semantic gap by use of advanced type systems
-\item Importance of the use of symmetry in computer science,
-  mathematics and the physical sciences
-\item Development of the theory of parametricity
-\end{enumerate}
-
-\subsection{Contributions}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Background and Related Work}
-
-Reynolds (Parametricity), Wadler (Free Theorems).
-
-Kennedy (Units of Measure)
-
-Something about symmetry?
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 theory in \autoref{sec:types-and-terms}.
+
+Data:
+\begin{enumerate}
+\item A set $\mathit{Sort}$ of sorts
+\item A collection of primitive index operations $\mathit{IndexOp}$
+\item A map $\indexOpArity : \mathit{IndexOp} \to \mathit{Sort}^* \times \mathit{Sort}$
+\item A set $\mathit{PrimType}$ of primitive types
+\item A map $\primTyArity : \mathit{PrimType} \to \mathit{Sort}^*$
+\end{enumerate}
+
+Then generate:
+\begin{enumerate}
+\item Well-sorted index expressions
+\item Well-indexed types
+\end{enumerate}
+
+More data:
+\begin{enumerate}
+\item A set of primitive operations $\mathit{PrimOps}$
+\end{enumerate}
+
+This generates the well-typed terms.
+
+For the semantics:
+\begin{enumerate}
+\item A ``relational'' meaning for index contexts. Can this be made
+  separate? Make the point that it is non-compositional, which is
+  useful. 
+\item Underlying semantics for each type
+\item Assume that the underlying semantics for types is not affected
+  by the index environment (could it be?)
+\item For each index context and primitive type, we demand a
+  relational meaning for that type.
+\item Relational meanings are closed under some operations.
+\end{enumerate}
 
@@
   Well-indexed Types
-  \begin{mathpar}
-    \inferrule* [right=TyPrimIdx]
-    {\tyPrimNm{X} \in \mathit{PrimType}(s) \\ \Delta \vdash e : s}
-    {\Delta \vdash \tyPrim{X}{e} : \sortType}
-
-    \inferrule* [right=TyPrimNoIdx]
-    {\tyPrimNm{X} \in \mathit{PrimType}(\cdot)}
-    {\Delta \vdash \tyPrimNm{X} : \sortType}
-
-    \inferrule* [right=TyUnit]
-    { }
-    {\Delta \vdash \tyUnit : \sortType}
-
-    \inferrule* [right=TyArr]
-    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
-    {\Delta \vdash A \tyArr B : \sortType}
-
-    \inferrule* [right=TyProd]
-    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
-    {\Delta \vdash A \tyProduct B : \sortType}
-
-    \inferrule* [right=TySum]
-    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
-    {\Delta \vdash A + B : \sortType}
-
-    \inferrule* [right=TyForall] %FIXME: macroize forall
-    {\Delta, i : s \vdash A : \sortType}
-    {\Delta \vdash \forall i \mathord: s. A : \sortType}
-  \end{mathpar}
 
@@
   \begin{mathpar}
-    \inferrule* [right=TyEqPrimIdx]
-    {\Delta \vdash e_1 \equiv e_2 : s}
-    {\Delta \vdash \tyPrim{X}{e_1} \equiv \tyPrim{X}{e_2} : \sortType}
-
-    \inferrule* [right=TyEqPrimNoIdx]
-    { }
-    {\Delta \vdash \tyPrimNm{X} \equiv \tyPrimNm{X} : \sortType}
+    \inferrule* [right=TyEqPrim]
+    {\primTyArity(\tyPrimNm{X}) = (s_1,...,s_n) \\
+      \{\Delta \vdash e_i \equiv e'_i : s_i\}_{1\leq i \leq n}}
+    {\Delta \vdash \tyPrim{X}{e_1,...,e_n} \equiv \tyPrim{X}{e'_1,..,e'_n} : \sortType}
 
@@
 \end{figure*}
+
+The rules for building the types of a system in our generic framework
+are given in \autoref{fig:types}. The types of our system are built
+from a set of primitive types $\mathit{PrimType}$.
 
@@
 \end{figure*}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Example Systems}\label{sec:example-systems}
-
-\subsection{Units of Measure}\label{sec:units-example}
-
-We now show that our general framework encompasses Kennedy's
-units-of-measure system \cite{kennedy97relational}. For our indexing
-algebraic theory, we have a single sort $\mathsf{unit}$ intended to
-represent composite units such as $\mathit{m}\ \mathit{s}^{-1}$ or
-$\mathit{kg}\ \mathit{m}^{-2}$. Units of measure form an abelian
-group, so we have three operations:
-\begin{mathpar}
-  1 : \mathsf{unit},
-
-  \cdot : \mathsf{unit}, \mathsf{unit} \to \mathsf{unit},
-
-  -^{-1} : \mathsf{unit} \to \mathsf{unit},
-\end{mathpar}
-subject to the normal abelian group axioms. We will elide the group
-multiplication operation and just write multiplication as
-juxtaposition: $u_1u_2$.
-
-We let $\mathit{PrimType}(\mathsf{unit}) = \{\tyPrimNm{num}$, and
-$\mathit{PrimType}(\cdot) = \{\tyPrimNm{rational}\}$. The intended
-reading is that the type $\tyPrim{num}{u_1u^2_2}$ represents
-quantities in units of $u_1u^2_2$, while values of type
-$\tyPrimNm{rational}$ represent dimensionless constants. We make a
-distinction between $\tyPrim{num}{1}$ and $\tyPrimNm{rational}$ for
-technical reasons arising from the relational interpretation below: we
-want the relation on $\tyPrim{num}{1}$ to be the identity relation,
-and the existence of $\mathit{toRat}$ and $\mathit{fromRat}$ forces
-this\footnote{Kennedy assumes that all relational environments set the
-  relation on $\tyPrim{num}{1}$ to be the identity relation, but we
-  cannot do this in general, because we do not necessarily have a
-  group structure in the indexing theory.}. To write programs using
-the $\tyPrimNm{num}$ type, we assume an typing context
-$\Gamma_{\textrm{units}}$ consisting of the following typed
-operations:
-\begin{eqnarray*}
-  \mathit{toRat} & : & \tyPrim{num}{1} \to \tyPrimNm{rational} \\
-  \mathit{fromRat} & : & \tyPrim{num}{1} \to \tyPrimNm{rational} \\
-  0 & : & \forall u. \tyPrim{num}{u} \\
-  1 & : & \tyPrim{num}{1} \\
-  + & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyPrim{num}{u} \\
-  - & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyPrim{num}{u} \\
-  * & : & \forall u_1 u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to \tyPrim{num}{u_1u_2} \\
-  / & : & \forall u_1 u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to \tyPrim{num}{u_1u_2^{-1}} + 1 \\
-  < & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyBool
-\end{eqnarray*}
-There are constants for zero and one, but only the constant for zero
-is polymorphic in the units. In \autoref{sec:units-semantics} below,
-we will following Kennedy by interpreting units of measure in terms of
-\emph{scalings} by rational numbers, where it will be apparent that
-the constant one cannot be unit-polymorphic: we can scale $0$ and
-still get $0$, but we cannot scale $1$ by an arbitrary rational number
-and expect to get $1$.
-
-\subsection{Euclidean Plane Geometry}\label{sec:2d-example}
-
-We now consider a type system for writing programs that are invariant
-under the translational symmetries of the Euclidean plane. We assume a
-single sort $\mathsf{translation}$. The sort $\mathsf{translation}$ is
-intended to represent translations in the plane, and comes equipped
-with abelian group structure that we write additively:
-\begin{mathpar}
-  0 : \mathsf{translation}
-
-  + : \mathsf{translation},\mathsf{translation} \to \mathsf{translation}
-
-  - : \mathsf{translation} \to \mathsf{translation}
-\end{mathpar}
-In indexing expressions involving variables of sort
-$\mathsf{translation}$, we use the shorthand $t_1 - t_2$ to stand for
-$t_1 + (-t_2)$.
-
-There are two primitive types:
-$\mathit{PrimType}(\mathsf{translation}) = \{\tyPrimNm{vec}\}$, and
-$\mathit{PrimType}(\cdot) = \{\tyPrimNm{real}\}$. The types
-$\tyPrim{vec}{t}$ represent two-dimensional vectors, indistinguishable
-up to some translation $t$. Values of type $\tyPrimNm{real}$ are just
-real numbers. To write programs using these types we assume a typing
-context $\Gamma_{\mathit{2D}}$ consisting of the following operations:
-\begin{eqnarray*}
-  + & : & \forall t_1 t_2\mathord:\mathsf{translation}.\tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2} \\
-  - & : & \forall t\mathord:\mathsf{translation}. \tyPrim{vec}{t} \to \tyPrim{vec}{-t} \\
-  0 & : & \tyPrim{vec}{0} \\
-  * & : & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0} \\
-  \mathrm{getX} & : & \tyPrim{vec}{0} \to \tyPrimNm{real} \\
-  \mathrm{getY} & : & \tyPrim{vec}{0} \to \tyPrimNm{real} \\
-  \mathrm{vec}  & : & \tyPrimNm{real} \to \tyPrimNm{real} \to \tyPrim{vec}{0}
-\end{eqnarray*}
-The first three operations express the abelian group structure of
-2-dimensional real vectors. Note how the indexing tracks the
-operations on vectors: e.g.~addition on vectors is reflected by
-addition on the indicies. The final four operations are operations
-that are not invariant under arbitrary translation: multiplication by
-a scalar, and projection from and constructon of vectors. For the
-purposes of examples, we also assume a standard set of arithmetic and
-other operations on real numbers.
-
-As an example of the use of the above operations, consider the
-following type of a function for computing the area of a triangle
-described by three points. Since the area of a triangle is invariant
-under translations that are applied to all three points, we can
-quantify over all translations $t$:
-\begin{displaymath}
-  \mathit{area} : \forall t\mathord:\mathsf{translation}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real}
-\end{displaymath}
-Since the result is invariant under translation, it safe to assume
-that one of the vertexes of the triangle is the origin, and only
-specify two points to describe the triangle. So we might expect that
-the type of $\mathit{area}$ is isomorphic to this type:
-\begin{displaymath}
-  \mathit{area'} : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrimNm{real}
-\end{displaymath}
-Given $\mathit{area'}$, we can define $\mathit{area}$ using the
-operations in $\Gamma_{\mathit{2D}}$:
-\begin{displaymath}
-  \mathit{area} = \Lambda t. \lambda v_1 v_2 v_3.\ \mathit{area'}\ (v_2 - v_1)\ (v_3 - v_1)
-\end{displaymath}
-and vice versa, we can define $\mathit{area'}$ in terms of
-$\mathit{area}$:
-\begin{displaymath}
-  \mathit{area'} = \lambda v_1 v_2. \mathit{area}\ [0]\ 0\ v_1\ v_2
-\end{displaymath}
-In \autoref{sec:2d-semantics} below, we will show that these two
-translations form a type isomorphism between the types of
-$\mathit{area}$ and $\mathit{area'}$.
-
-\paragraph{Alternative Transformation Groups?}
-We should also consider other groups of transformations for indexing:
-isometries, orthogonal transformations, rotations and reflections.
-
-\subsection{Security Typing and Non-interference}
-
-This example shows how our framework can accommodate non-interference
-properties common in security type systems. This example has a
-different flavour to the two previous examples because it does not
-concern invariance under the action of some group. We assume a single
-sort $\mathsf{principal}$, intended to represent (possibly composite)
-principals in a system. Principals are combined using the connectives
-of boolean logic:
-\begin{mathpar}
-  \top : \mathsf{principal}
-
-  \bot : \mathsf{principal}
-
-  \land : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}
-
-  \lor : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}
-
-  \Rightarrow : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}
-
-  \lnot : \mathsf{principal} \to \mathsf{principal}
-\end{mathpar}
-For the equational theory of principals, we assume all the axioms of
-boolean algebra.
-
-Let $\mathit{PrimType}(\mathsf{principal}) = \{\tyPrimNm{T}\}$. The
-intended reading of a type $\tyPrim{T}{p}$ for some principal $p$ is
-that the existence of a value of this type indicates that the
-principal $p$ is ``happy''. To write programs, we assume the following
-typing context $\Gamma_{\mathit{sec}}$ of operations:
-\begin{eqnarray*}
-  \mathrm{proj} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1 \land p_2} \to \tyPrim{T}{p_1} \\
-  \mathrm{combine} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1} \to \tyPrim{T}{p_2} \to \tyPrim{T}{p_1 \land p_2}
-\end{eqnarray*}
-
-To illustrate the use of this application of our framework, we define
-a type synonym, for each type $A$ and principal $p$:
-\begin{displaymath}
-  T_pA = \tyPrim{T}{p} \to A
-\end{displaymath}
-For every principal $p$, we can endow the types $T_p-$ with the
-structure of a monad. This is due to the fact that it is an instance
-of the ``environment'' (or ``reader'') monad. Explicitly, the
-$\mathit{return}$ and $>>=$ of these monads are given by:
-\begin{eqnarray*}
-  \mathit{return} & : & A \to T_pA \\
-  \mathit{return} & = & \lambda a\ t.\ a \\
-  \\
-  (>>=) & : & T_pA \to (A \to T_pB) \to T_pB \\
-  (>>=) & = & \lambda a\ f\ t.\ f\ (a\ t)\ t
-\end{eqnarray*}
-The intended reading of a value of type $T_pA$ is as a piece of data
-of type $A$ that can only be accessed if the principal $p$ is
-happy. The existence of such a monad should allow us to draw parallels
-between our system and the Dependency Core Calculus (DCC) system of
-Abadi \emph{et al.} \cite{abadi99core}. The definition of a DCC monad
-as an instance of an environment monad follows Tse and Zdancewic's
-interpretation of DCC in System F, where they use System F
-parametricity to ensure non-interference properties.
-
-There is also a coercion operator between instances of $T_pA$ for
-different principals. Reading the type in terms of happiness, this
-says that if we have a piece of data that is protected by $p$, and
-happiness of $q$ implies happiness of $p$, then this data can also be
-accessed by $q$.
-\begin{eqnarray*}
-  \mathit{coerce} & : & \tyPrim{T}{q \Rightarrow p} \to T_pA \to T_qA \\
-  \mathit{coerce} & = & \lambda t_{qp}\ a\ t_q.\ a\ (\mathrm{combine}\ t_{qp}\ t_q)
-\end{eqnarray*}
-Here we have used the fact that the result of the application of
-$\mathrm{combine}$ has type $\tyPrim{T}{(q \Rightarrow p) \land
-  q}$. By the laws of boolean algebras---which we have assumed for our
-indexing theory---this is equal to the type $\tyPrim{T}{p}$. Thus we
-get access to the data stored in $a$, which is protected by $p$.
-
-Now suppose we have a term $M$ with the following typing:
-\begin{displaymath}
-  p,q:\mathsf{principal}; t_{qp} : \tyPrim{T}{q \Rightarrow p}, x : T_q (1 + 1) \vdash M : T_p (1 + 1)
-\end{displaymath}
-Intuitively, given the operations $\Gamma_{\mathit{sec}}$ we have
-defined above, it should be the case that $M$ can only be equivalent
-to the program which is constantly $\lambda t. \mathrm{inl}(*)$ or
-$\lambda t. \mathrm{inr}(*)$, because there is no useful data to be
-gained from the $t_{qp}$ argument, and a value of $\tyPrim{T}{q}$
-cannot be generated in order to get access to the data within the $x$
-argument. The application of our framework that we have presented in
-this section has a ``non-interference'' property: the value of $x$
-cannot interfere with the output. In \autoref{sec:sec-semantics} we
-show that this is indeed the case.
-
-\paragraph{Alternative Logics?}
-Above, we assumed that the indexing theory of principals was the
-theory of boolean algebras. As a consequence, a value of type $T_{p
-  \lor \lnot p}A$ can always be accessed, due to the exlcuded middle
-property of boolean logic. An interesting question is: what happens we
-we consider alternative logics. If we take the theory of Heyting
-algebras for our theory of principals, then data typed as $T_{p \lor
-  \lnot p}A$ can only be accessed if either $p$ or $\lnot p$ can be
-demonstrated. Considering an alternative logical theory alters the
-non-interference property of the type system.
-
-\paragraph{Alternative Presentations?}
-In the above we presented the theory of boolean logic by means of the
-equational theory of boolean algebras. This isn't necessarily the most
-convenient way of presenting a logic. An alternative approach would be
-to consider a trivial equational theory with no axioms, and
-incorporate the rules of logic consequence as operations in
-$\Gamma_{\mathit{sec}}$. For example, as a Hilbert system:
-\begin{eqnarray*}
-  \mathit{MP} & : & \forall p,q.\ \tyPrim{T}{p \Rightarrow q} \to \tyPrim{T}{p} \to \tyPrim{T}{q} \\
-  \mathit{k} & : & \forall p,q.\ \tyPrim{T}{p \Rightarrow q \Rightarrow p} \\
-  \mathit{s} & : & \forall p,q,r.\ \tyPrim{T}{(p \Rightarrow q \Rightarrow r) \Rightarrow (p \Rightarrow q) \Rightarrow p \Rightarrow r}
-\end{eqnarray*}
-Plus axioms for the other connectives.  Or possibly some kind of
-encoding of a natural deduction system?
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Semantics and Parametricity Theorem}
-
-There is an index-erasure semantics, and a relational semantics. The
-relational semantics allows us to derive free theorems about the
-index-erasure semantics.
-
-\subsection{Index Erasure Semantics}
-
-We now define a semantics of types and terms that ignores all indexing
-information. By defining this semantics we can easily see that the
-indexing information can be discarded before execution.
-
-\subsubsection{Index-Erasure Semantics of Types}
-
-For each primitive type $\tyPrimNm{X} \in \mathit{PrimType}(s)$, for
-some sort $s$, and for the primitive types $\tyPrimNm{X} \in
-\mathit{PrimType}(\cdot)$, we assume that $\tyPrimNm{X}$ is assigned a
-set $\tyPrimSem{X}$. We extend this to assign a set to every
-well-formed type by induction on the type structure. To interpret the
-universal quantifier, we simply ignore the indexing:
-\begin{eqnarray*}
-  \tySem{\tyUnit} & = & \{*\} \\
-  \tySem{\tyPrim{X}{e}} & = & \tyPrimSem{X} \\
-  \tySem{\tyPrim{X}} & = & \tyPrimSem{X} \\
-  \tySem{A \tyProduct B} & = & \tySem{A} \times \tySem{B} \\
-  \tySem{A \tyArr B} & = & \tySem{A} \to \tySem{B} \\
-  \tySem{A + B} & = & \tySem{A} + \tySem{B} \\
-  \tySem{\forall i\mathord:s. A} & = & \tySem{A}
-\end{eqnarray*}
-
-Since our definition has completely ignored the indexing, and type
-equality is defined as an extension of index equality, we can easily
-see that equal types have equal denotations in our index-erasure
-semantics:
-\begin{lemma}\label{lem:tyeq-erasure}
-  If $\Delta \vdash A \equiv B : \sortType$ then $\tySem{A} =
-  \tySem{B}$.
-\end{lemma}
-
-\begin{proof}
-  By induction on the derivation of $\Delta \vdash A \equiv B :
-  \sortType$:
-  \begin{description}
-  \item[Case \TirName{TyEqUnit}] In this case, $A = B = \tyUnit$, so
-    $\tySem{A} = \tySem{B}$.
-  \item[Case \TirName{TyEqPrimIdx}] We have $A = \tyPrim{X}{e_1}$ and
-    $B = \tyPrim{X}{e_2}$, for some $e_1$ and $e_2$ of the same
-    sort. The erasure semantics ignores the index arguments, so
-    $\tySem{A} = \tySem{\tyPrim{X}{e_1}} = \tyPrimSem{X} =
-    \tySem{\tyPrim{X}{e_2}} = \tySem{B}$, as required.
-  \item[Case \TirName{TyEqPrimNoIdx}] We have $A = \tyPrimNm{X}$ and
-    $B = \tyPrimNm{X}$. So $\tySem{A} = \tyPrimSem{X} = \tySem{B}$, as
-    required.
-  \item[Case \TirName{TyEqArr}] Here, $A = A_1 \tyArr A_2$ and $B =
-    B_1 \tyArr B_2$, and by the induction hypothesis we know that
-    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
-    $\tySem{A} = \tySem{A_1 \tyArr A_2} = \tySem{A_1} \to \tySem{A_2}
-    = \tySem{B_1} \to \tySem{B_2} = \tySem{B_1 \tyArr B_2} =
-    \tySem{B}$.
-  \item[Case \TirName{TyEqProd}] In this case, $A = A_1 \tyProduct
-    A_2$ and $B = B_1 \tyProduct B_2$, and by the induction hypothesis
-    we know that $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} =
-    \tySem{B_2}$. Hence $\tySem{A} = \tySem{A_1 \tyProduct A_2} =
-    \tySem{A_1} \times \tySem{A_2} = \tySem{B_1} \times \tySem{B_2} =
-    \tySem{B_1 \tyProduct B_2} = \tySem{B}$.
-  \item[Case \TirName{TyEqSum}] In this case, $A = A_1 + A_2$ and $B =
-    B_1 + B_2$, and by the induction hypothesis we know that
-    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
-    $\tySem{A} = \tySem{A_1 + A_2} = \tySem{A_1} + \tySem{A_2} =
-    \tySem{B_1} + \tySem{B_2} = \tySem{B_1 + B_2} = \tySem{B}$.
-  \item[Case \TirName{TyEqForall}] We have $A = \forall
-    i\mathord:s.A'$ and $B = \forall i\mathord:s.B'$. By the induction
-    hypothesis we know that $\tySem{A'} = \tySem{B'}$, and hence
-    $\tySem{A} = \tySem{\forall i\mathord:s. A'} = \tySem{A'} =
-    \tySem{B'} = \tySem{\forall i\mathord:s. B'} = \tySem{B}$.
-  \end{description}
-\end{proof}
-
-Moreover, the index-erasure semantics is insensitive to index
-substitutions.
-\begin{lemma}\label{lem:tysubst-erasure}
-  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then $\tySem{\sigma^*A} = \tySem{A}$.
-\end{lemma}
-
-\begin{proof}
-  By induction on the derivation of $\Delta \vdash A : \sortType$.
-  \begin{description}
-  \item[Case \TirName{TyUnit}] Directly from the definition of
-    substitution on types of the form $\tyUnit$, we have
-    $\tySem{\sigma^*\tyUnit} = \tySem{\tyUnit}$, as required.
-  \item[Case \TirName{TyPrimIdx}] Since the index-erasure semantics
-    ignores all index expessions, we have
-    $\tySem{\sigma^*\tyPrim{X}{e}} = \tySem{\tyPrim{X}{\sigma^*e}} =
-    \tyPrimSem{X} = \tySem{\tyPrim{X}{e}}$.
-  \item[Case \TirName{TyPrimNoIdx}] Similar to the previous case.
-  \item[Case \TirName{TyArr}] We have $A = A_1 \tyArr A_2$, and by the
-    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
-    \tySem{A_1}$ and $\tySem{\sigma^*B_1} = \tySem{B_1}$. Thus, by the
-    definition of substitution on arrow types, we have
-    $\tySem{\sigma^*(A_1 \tyArr A_2)} = \tySem{\sigma^*A_1 \tyArr
-      \sigma^*A_2} = \tySem{\sigma^*A_1} \to \tySem{\sigma^*A_2} =
-    \tySem{A_1} \to \tySem{A_2} = \tySem{A_1 \tyArr A_2}$, as
-    required.
-  \item[Case \TirName{TyProd}] We have $A = A_1 \tyProduct A_2$, and
-    by the induction hypothesis, we know that $\tySem{\sigma^*A_1} =
-    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
-    definition of substitution on product types, we have
-    $\tySem{\sigma^*(A_1 \tyProduct A_2)} = \tySem{\sigma^*A_1
-      \tyProduct \sigma^*A_2} = \tySem{\sigma^*A_1} \times
-    \tySem{\sigma^*A_2} = \tySem{A_1} \times \tySem{A_2} = \tySem{A_1
-      \tyProduct A_2}$, as required.
-  \item[Case \TirName{TySum}] We have $A = A_1 + A_2$, and by the
-    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
-    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
-    definition of substitution on sum types, we have
-    $\tySem{\sigma^*(A_1 + A_2)} = \tySem{\sigma^*A_1 + \sigma^*A_2} =
-    \tySem{\sigma^*A_1} + \tySem{\sigma^*A_2} = \tySem{A_1} +
-    \tySem{A_2} = \tySem{A_1 + A_2}$, as required.
-  \item[Case \TirName{TyForall}] We have $A = \forall i\mathord:s.A'$,
-    and by the induction hypothesis we know that, for all $\sigma'$,
-    $\tySem{\sigma'^*A'} = \tySem{A'}$. Therefore, from the definition
-    of substitution for $\forall$-types, we have
-    $\tySem{\sigma^*\forall i\mathord:s.A'} = \tySem{\forall
-      i\mathord:s.\sigma_s^{*}A'} = \tySem{\sigma_s^{*}A'} = \tySem{A'}
-    = \tySem{\forall i\mathord:s.A'}$, where we have relied on the
-    fact that the index-erasure semantics of types ignores the index
-    quantification.
-  \end{description}
-\end{proof}
-
-For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
-index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
-structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
-\ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
-one-element set.
-
-\subsubsection{Index-Erasure Semantics of Terms}
-
-For a well-formed term $\Delta; \Gamma \vdash e : A$, we define a
-function $\tmSem{e} : \ctxtSem{\Gamma} \to \tySem{A}$, completely
-ignoring the indexing information. In light of
-\lemref{lem:tyeq-erasure} and \lemref{lem:tysubst-erasure}, we may
-define this function directly on the syntax of well-typed terms,
-rather than on the typing derivations. We define $\tmSem{e}$ by the
-following clauses:
-\begin{eqnarray*}
-  \tmSem{x}\eta & = & \eta_x \\
-  \tmSem{(M,N)}\eta & = & (\tmSem{M}\eta, \tmSem{N}\eta) \\
-  \tmSem{\pi_1M}\eta & = & \pi_1(\tmSem{M}\eta) \\
-  \tmSem{\pi_2M}\eta & = & \pi_2(\tmSem{M}\eta) \\
-  \tmSem{\mathrm{inl}\ M}\eta & = & \mathrm{inl}\ (\tmSem{M}\eta) \\
-  \tmSem{\mathrm{inr}\ M}\eta & = & \mathrm{inr}\ (\tmSem{M}\eta) \\
-  \left\llbracket
-    \begin{array}{l}
-      \textrm{case}\ M\ \textrm{of}\\
-      \textrm{inl}\ x.N_1;\\
-      \textrm{inr}\ y.N_2
-    \end{array}\right\rrbracket^{\mathit{tm}}\eta & = &
-  \left\{
-    \begin{array}{ll}
-      \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
-      \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
-    \end{array}
-  \right. \\
-  \tmSem{\lambda x. M}\eta & = & \lambda v. \tmSem{M}(\eta, v) \\
-  \tmSem{M N}\eta & = & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
-  \tmSem{\Lambda i.\ M}\eta & = & \tmSem{M}\eta \\
-  \tmSem{M[e]}\eta & = & \tmSem{M}\eta
-\end{eqnarray*}
-In the above, we have written $\eta_x$ to denote the appropriate
-projection from $\eta$ to get the position corresponding to the
-variable $x$ in the context.
-
-\subsection{Relational Semantics}
-
-\subsubsection{Relation Environments}
-
-For a set $X$, let $\Rel(X)$ be the set of binary relations on
-$X$. For an index context $\Delta$, a \emph{relation environment} is a
-family of functions $\rho = \{ \rho_{\tyPrimNm{X}} :
-\idxTms{\Delta}{s} \to \Rel(\tyPrimSem{X}) \}_{s \in \mathit{Sort},
-  \tyPrimNm{X} \in \mathit{PrimType}(s)}$.
-
-We define our relational semantics of types with respect to a given
-collection of relation environments: we assume that, for each index
-context $\Delta$, we are given a set of relation environments
-$\relEnv{E}(\Delta)$. This family must satisfy the following two
-properties:
-\begin{enumerate}
-\item Closure under re-indexing by index context morphisms: if $\rho
-  \in \relEnv{E}(\Delta)$ and $\sigma : \Delta' \Rightarrow \Delta$,
-  then $\rho \circ \sigma^* \in \relEnv{E}(\Delta')$.
-\item If we have a pair of relation environments $\rho_1 \in
-  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
-  \relEnv{E}(\Delta)$, along with an index context morphism $\sigma :
-  \Delta' \Rightarrow \Delta$, such that the outer edges of the
-  following diagram commute, for all primitive types $\tyPrimNm{X} \in
-  \mathit{PrimType}(s)$:
-  \begin{displaymath}
-    \xymatrix{
-      {\idxTms{\Delta}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar[d]_{\sigma^*}
-      &
-      {\idxTms{\Delta,i\mathord:s'}{s}} \ar[d]^{\sigma_{s'}^*} \ar@/^/[rdd]^{\rho_1}
-      \\
-      {\idxTms{\Delta'}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2}
-      &
-      {\idxTms{\Delta',i\mathord:s'}{s}} \ar@{.>}[dr]_\rho
-      \\
-      &
-      &
-      {\Rel(X)}
-    }
-  \end{displaymath}
-  Then there exists a relation environment $\rho \in
-  \relEnv{E}(\Delta,i\mathord:s')$ such that the two triangles in the
-  bottom right of the diagram commute.
-\end{enumerate}
-
-Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
-the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
-additional index variable $i\mathord:s$ to be the following:
-\begin{displaymath}
-  \extends{\rho}{i\mathord:s} = \{ \rho' \in \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s} = \rho \}
-\end{displaymath}
-
-\subsubsection{Relational Semantics of Types}
-
-%FIXME: re-word this
-A given relation context $\rho \in \relEnv{E}(\Delta)$ is extended to
-a relation interpretation of any type $\Delta \vdash A : \sortType$ by
-induction its derivation. So we define a relation
-$\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-formed
-type $A$.
-\begin{eqnarray*}
-  \rsem{\tyUnit}{\relEnv{E}}\rho & = & \{(*,*)\} \\
-  \rsem{\tyPrim{X}{e}}{\relEnv{E}}\rho & = & \rho_{\tyPrimNm{X}}(e) \\
-  \rsem{A \tyArr B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{A \tyProduct B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{A + B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & = & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i:s} \}
-\end{eqnarray*}
-In this definition we have made use of the following two constructions
-on binary relations. If $R \in \Rel(X)$ and $S \in \Rel(Y)$, then $R
-\relArrow S \in \Rel(X \to Y)$ is defined as $\{ (f_1,f_2) \sepbar
-\forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S \}$. With the same
-assumptions on $R$ and $S$, the relation $R \relTimes S \in \Rel(X
-\times Y)$ is defined as $\{ ((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2)
-\in R \land (b_1,b_2) \in S \}$. FIXME: do $\relSum$.
-
-We extend the relational semantics to contexts of well-formed types by
-the following clauses:
-\begin{eqnarray*}
-  \rsem{\epsilon}{\relEnv{E}}{\rho} & = & \id_{\ctxtSem{\epsilon}} \\
-  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & = & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
-\end{eqnarray*}
-
-\subsubsection{Parametricity}
-
-This relies on \lemref{lem:tyeq-erasure} to make sure that it is
-well-typed.
-
-\begin{lemma}\label{lem:tyeq-rel}
-  If $\Delta \vdash A \equiv B : \sortType$, then for all $\rho \in
-  \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}\rho = \rsem{B}{\relEnv{E}}\rho$.
-\end{lemma}
-
-\begin{proof}
-  By induction on the derivation of $\Delta \vdash A \equiv B :
-  \sortType$.
-  \begin{description}
-  \item[Case \TirName{TyEqUnit}] In this case $A = B = \tyUnit$, so
-    the statement is trivially statisfied.
-  \item[Case \TirName{TyEqPrimIdx}] In this case we have $A =
-    \tyPrim{X}{e_1}$ and $B = \tyPrim{X}{e_2}$, with $\Delta \vdash
-    e_1 \equiv e_2 : s$. The index expressions $e_1$ and $e_2$ are in
-    the same equivalence class, so by the definition of relation
-    environments we have that $\rho_{\tyPrimNm{X}}(e_1) =
-    \rho_{\tyPrimNm{X}}(e_2)$ and hence
-    $\rsem{\tyPrim{X}{e_1}}{\relEnv{E}}\rho =
-    \rsem{\tyPrim{X}{e_2}}{\relEnv{E}}\rho$.
-  \item[Case \TirName{TyEqPrimNoIdx}] This case is similar to the case
-    for \TirName{TyEqUnit}.
-  \item[Case \TirName{TyEqArr}] We have $A = A_1 \tyArr A_2$ and $B =
-    B_1 \tyArr B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$
-    and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the induction
-    hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_2}{\relEnv{E}}\rho$. Thus we may reason as follows:
-    $\rsem{A_1 \tyArr A_2}{\relEnv{E}}\rho =
-    \rsem{A_1}{\relEnv{E}}\rho \relArrow \rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho \relArrow \rsem{B_2}{\relEnv{E}}\rho =
-    \rsem{B_1 \tyArr B_2}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{TyEqProd}] We have $A = A_1 \tyProduct A_2$ and
-    $B = B_1 \tyProduct B_2$, with $\Delta \vdash A_1 \equiv B_1 :
-    \sortType$ and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the
-    induction hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
-    $\rsem{A_1 \tyProduct A_2}{\relEnv{E}}\rho =
-    \rsem{A_1}{\relEnv{E}}\rho \relTimes \rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho \relTimes \rsem{B_2}{\relEnv{E}}\rho =
-    \rsem{B_1 \tyProduct B_2}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{TyEqSum}] We have $A = A_1 + A_2$ and $B = B_1 +
-    B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$ and $\Delta
-    \vdash A_2 \equiv B_2 : \sortType$. By the induction hypothesis,
-    we know that $\rsem{A_1}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
-    $\rsem{A_1 + A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}\rho
-    \relSum \rsem{A_2}{\relEnv{E}}\rho = \rsem{B_1}{\relEnv{E}}\rho
-    \relSum \rsem{B_2}{\relEnv{E}}\rho = \rsem{B_1 +
-      B_2}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{TyEqForall}] We have $A = \forall
-    i\mathord:s. A'$ and $B = \forall i\mathord:s.B'$, with $\Delta, i
-    : s \vdash A' \equiv B' : \sortType$. By the induction hypothesis,
-    we know that for all $\rho'$, $\rsem{A'}{\relEnv{E}}\rho' =
-    \rsem{B'}\rho'$. We reason as follows:
-    \begin{eqnarray*}
-      &    & (x_1,x_2) \in \rsem{\forall i\mathord:s. A'}{\relEnv{E}}\rho \\
-      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{A'}{\relEnv{E}}\rho' \\
-      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{B'}{\relEnv{E}}\rho' \\
-      &\iff& (x_1,x_2) \in \rsem{\forall i\mathord:s. B'}{\relEnv{E}}\rho
-    \end{eqnarray*}
-    Thus, by extensionality, we have $\rsem{\forall
-      i\mathord:s.A'}\rho \equiv \rsem{\forall
-      i\mathord:s.B'}{\relEnv{E}}\rho$, as required.
-  \end{description}
-\end{proof}
-
-This next lemma relies on \lemref{lem:tysubst-erasure} to make sure
-that it is well-typed.
-
-\begin{lemma}\label{lem:tysubst-rel}
-  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
-  $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Likewise,
-  if $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
-  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho = \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
-\end{lemma}
-
-\begin{proof}
-  The first part of the lemma statement is proved by induction on the
-  derivation of $\Delta \vdash A : \sortType$. We analyse each case in
-  turn:
-  \begin{description}
-  \item[Case \TirName{TyUnit}] Directly from the definition of
-    substitution on types of the form $\tyUnit$, and the fact that
-    $\rsem{\tyUnit}{\relEnv{E}}\rho_1 =
-    \rsem{\tyUnit}{\relEnv{E}}\rho_2$ for any pair $\rho_1,\rho_2$ of
-    relation environments, we have
-    $\rsem{\sigma^*\tyUnit}{\relEnv{E}}\rho =
-    \rsem{\tyUnit}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as required.
-  \item[Case \TirName{TyPrimIdx}] We have $A = \tyPrim{X}{e}$, therefore
-    $\rsem{\sigma^*A}{\relEnv{E}}\rho =
-    \rsem{\sigma^*\tyPrim{X}{e}}{\relEnv{E}}\rho =
-    \rsem{\tyPrim{X}{\sigma^*e}}{\relEnv{E}}\rho =
-    \rho_{\tyPrimNm{X}}(\sigma^*e) =
-    \rsem{\tyPrim{X}{e}}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as
-    required.
-  \item[Case \TirName{TyPrimNoIdx}] Similar to the case for
-    \TirName{TyUnit}.
-  \item[Case \TirName{TyArr}] In this case, $A = A_1 \tyArr A_2$ and
-    by the induction hypothesis, we know that $\rsem{\sigma^*A_1}{\relEnv{E}}\rho
-    = \rsem{A_1}{\relEnv{E}}{(\rho \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we reason as
-    follows: $\rsem{\sigma^*(A_1 \tyArr A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1
-      \tyArr \sigma^*A_2}{\relEnv{E}}\rho = \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relArrow
-    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho \circ \sigma^*)}
-    \relArrow \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)} = \rsem{A_1 \tyArr
-      A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as required.
-  \item[Case \TirName{TyProd}] In this case, $A = A_1 \tyProduct A_2$
-    and by the induction hypothesis, we know that
-    $\rsem{\sigma^*A_1}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we
-    reason as follows: $\rsem{\sigma^*(A_1 \tyProduct
-      A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1 \tyProduct
-      \sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relTimes
-    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)} \relTimes \rsem{A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)} = \rsem{A_1 \tyProduct A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)}$, as required.
-  \item[Case \TirName{TySum}] In this case, $A = A_1 + A_2$
-    and by the induction hypothesis, we know that
-    $\rsem{\sigma^*A_1}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we
-    reason as follows: $\rsem{\sigma^*(A_1 +
-      A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1 +
-      \sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relSum
-    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)} \relSum \rsem{A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)} = \rsem{A_1 + A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)}$, as required.
-  \item[Case \TirName{TyForall}] We have $A = \forall
-    i\mathord:s.A'$. We prove the required equation by demonstrating
-    two inclusions. From left-to-right, we know that
-    \begin{equation}
-      \label{eq:tysubst-rel-forall1}
-      (x_1,x_2) \in \rsem{\sigma^*\forall i\mathord:s. A'}{\relEnv{E}}\rho
-    \end{equation}
-    and we must show that $(x_1, x_2) \in \rsem{\forall
-      i\mathord:s. A'}{\relEnv{E}}{(\rho \circ \sigma^*)}$. From
-    (\ref{eq:tysubst-rel-forall1}) and the definition of substitution,
-    we know that
-    \begin{displaymath}
-      \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{\sigma_s^*A'}{\relEnv{E}}\rho'
-    \end{displaymath}
-    By the induction hypothesis, we can deduce that
-    \begin{equation}
-      \label{eq:tysubst-rel-forall2}
-      \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{A'}{\relEnv{E}}{(\rho' \circ \sigma_s^{*})}
-    \end{equation}
-    To show the required inclusion, we must show that for all $\rho_1
-    \in \extends{\rho \circ \sigma^*}{i:s}$, it is the case that
-    $(x_1,x_2) \in \rsem{A'}{\relEnv{E}}{\rho_1}$. We now make use of
-    the pushout property of our family $\relEnv{E}$ of relation
-    environments: we have a $\rho \in \relEnv{E}(\Delta)$ and $\rho_1
-    \in \relEnv{E}(\Delta, i:s)$ such that $\rho_1 \circ \pi_{i:s} =
-    \rho \circ \sigma^*$ (by the assumption that $\rho_1 \in
-    \extends{\rho \circ \sigma^*}{i:s}$), so we can obtain a $\rho'
-    \in \relEnv{E}(\Delta',i:s)$ such that $\rho' \circ \pi_{i:s} =
-    \rho$ (so $\rho' \in \extends{\rho}{i:s}$), and $\rho' \circ
-    \sigma^* = \rho_1$. Therefore we use
-    (\ref{eq:tysubst-rel-forall2}) to deduce that $(x_1,x_2) \in
-    \rsem{A'}{\relEnv{E}}{\rho_1}$, and the inclusion from
-    left-to-right is shown.
-
-    From right-to-left, we must now show
-    (\ref{eq:tysubst-rel-forall2}) under the assumption that
-    \begin{equation}
-      \label{eq:tysubst-rel-forall3}
-      \forall \rho_1 \in \extends{\rho \circ \sigma^*}{i:s}. (x_1,x_2) \in \rsem{A'}{\relEnv{E}}{\rho_1}.
-    \end{equation}
-    Given $\rho' \in \extends{\rho}{i:s}$, we let $\rho_1 = \rho'
-    \circ \sigma^*$. Now, $\rho_1 \circ \pi_{i:s} = \rho' \circ
-    \sigma^* \circ \pi_{i:s} = \rho' \circ \pi_{i:s} \circ \sigma_s^{*}
-    = \rho \circ \sigma_s^{*}$. Thus $\rho_1 \in \extends{\rho \circ
-      \sigma^*}{i:s}$ and we may use (\ref{eq:tysubst-rel-forall3}) to
-    deduce that $(x_1,x_2) \in \rsem{A'}{\relEnv{E}}{(\rho' \circ \sigma_s^{*})}$,
-    as required.
-  \end{description}
-  The second part of the lemma statement follows directly by induction
-  on the derivation of $\Delta \vdash \Gamma \isCtxt$ and the first
-  part. The step case of the induction is very similar to the
-  \TirName{TyProd} case above.
-\end{proof}
-
-\begin{theorem}[Parametricity]
-  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
-  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
-  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
-  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
-\end{theorem}
-
-\begin{proof}
-  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
-  three interesting cases follow:
-  \begin{description}
-  \item[Case \TirName{TyEq}] This case follows directly from
-    \lemref{lem:tyeq-rel}.
-  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
-    $A = \forall i\mathord:s. A'$. We know from the induction
-    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
-    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}\rho'$,
-    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho'$. We are given $\rho \in
-    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
-    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
-    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
-    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
-    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
-      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
-    \rsem{\pi_{i:s}^*\Gamma}\rho''$ by \lemref{lem:tysubst-rel}. So we
-    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}\rho''$, and thus
-    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
-      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
-      i.M'}\eta_2) \in \rsem{\forall
-      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
-    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
-    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
-    know that for all $\rho' \in \extends{\rho}{i:s}$,
-    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho'$. If we let $\rho' = \rho \circ
-    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
-    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
-    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
-    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
-    by \lemref{lem:tysubst-rel}, conclude that $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
-    as required.
-  \end{description}
-  The remaining cases are standard for proofs of the fundamental lemma
-  of logical relations for simply-typed systems.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Consequences of Parametricity}
-
-We now give instances of our semantic framework for each of the three
-examples from \autoref{sec:example-systems}. For each of the three
-examples, we defined an indexing theory, a collection of primitive
-types, and a collection of primitive operations $\Gamma_X$. Below, we
-assign the ``obvious'' index-erasure semantics to each of the
-primitive types and operations: so we have $\tyPrimSem{X}$ for each
-primitive type $\tyPrimNm{X}$, and a $\eta_X \in \ctxtSem{\Gamma_X}$.
-
-The key question we wish to answer below is whether there is a family
-of sets of relation environments $\{\relEnv{E}_X(\Delta)\}_{\Delta}$
-that satisfies the following completeness property identified by
-Kennedy.
-\begin{equation}\label{eq:kennedy-completeness}
-  (\forall \Delta. \forall \rho \in \relEnv{E}'(\Delta). (\eta_X, \eta_X) \in \rsem{\Gamma_X}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_X
-\end{equation}
-where the subset relation on families of sets of relation environments
-is defined as $\relEnv{E}_1 \subseteq \relEnv{E}_2$ if $\forall
-\Delta. \relEnv{E}_1(\Delta) \subseteq \relEnv{E}_2(\Delta)$.
-
-Such a completeness result allows us to derive inexpressibility
-results for a given instance of our general framework. Kennedy
-demonstrates this for the units of measure system by showing that a
-putative square root function of type $\forall u.\ \tyPrim{num}{u^2}
-\to \tyPrim{num}{u}$ can only ever be the constantly zero function (or
-non-terminating).
-
-\subsection{Units of Measure}\label{sec:units-semantics}
-
-For the units of measure example, we set $\tyPrimSem{num} =
-\tyPrimSem{rational} = \mathbb{Q}$, the set of rational numbers. We
-define $\eta_{\mathit{units}} \in \Gamma_{\mathit{units}}$ are defined
-to be the standard arithmetical and comparison operations on rational
-numbers, and $\mathit{toRat}$ and $\mathit{fromRat}$ are defined to be
-the identity functions. The division function (of type $/ : \forall
-u_1,u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to
-\tyPrim{num}{u_1u^{-1}_2} + 1$) returns $\mathrm{inr}(*)$ only when
-its second argument is zero.
-
-Kennedy defines the following family of sets of relational
-environments. Note that, since we have assumed the abelian group
-axioms, the set $\idxTms{\Delta}{\mathsf{unit}}$ of indexing
-expressions quotiented by the equational theory, is an abelian group.
-\begin{displaymath}
-  \relEnv{E}_{\mathit{units}}(\Delta) = \left\{ \rho^{G,h} \sepbar
-    \begin{array}{l}
-      G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{unit}} \\
-      h : G \to \mathbb{Q}^{>0} \\
-      h\textrm{ is a group homomorphism}
-    \end{array}
-  \right\}
-\end{displaymath}
-where $\mathbb{Q}^{>0} = \{ q \in \mathbb{Q} \sepbar q > 0 \}$ and
-\begin{displaymath}
-  \rho^{G,h}_{\tyPrimNm{num}}(e) = \left\{
-    \begin{array}{ll}
-      \{(0,0)\} & \textrm{if }e \not\in G \\
-      \{(r,h(e)r) \sepbar r \in \mathbb{Q} \} & \textrm{if }u \in G
-    \end{array}
-  \right.
-\end{displaymath}
-
-\begin{theorem}
-  The family of sets of relational environments
-  $\relEnv{E}_{\mathit{units}}$ satisfies the completeness property
-  \hyperref[eq:kennedy-completeness]{(\ref*{eq:kennedy-completeness})}
-  above. That is, for all $\relEnv{E}'$:
-  \begin{displaymath}
-    (\forall \Delta.\forall \rho \in \relEnv{E}'(\Delta).
-(\eta_{\mathit{units}}, \eta_{\mathit{units}}) \in \rsem{\Gamma_{\mathit{units}}}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_{\mathit{units}}
-  \end{displaymath}
-\end{theorem}
-
-\begin{proof}
-  (Following Kennedy).  From right to left, we have to check that for
-  each $\rho \in \relEnv{E}'(\Delta)$, where $\relEnv{E}' \subseteq
-  \relEnv{E}_{\mathit{units}}$, that each of the semantic
-  interpretations of the operations satisfies
-  $(\eta_{\mathit{units}}(\mathit{op}),
-  \eta_{\mathit{units}}(\mathit{op})) \in
-  \rsem{\Gamma(\mathit{op})}{\relEnv{E}'}\rho$. For instance,
-  multiplication has to satisfy: for all subgroups $G \subseteq
-  \idxTms{\Delta,u_1,u_2}{\mathsf{unit}}$ and all group homomorphisms
-  $h : G \to \mathbb{Q}^{>0}$, then $(q_1,q'_1) \in \rho^{G,h}(u_1)$
-  and $(q_2,q'_2) \in \rho^{G,h}(u_2)$ implies that $(q_1q_2,q'_1q'_2)
-  \in \rho^{G,h}(u_1u_2)$. There are four cases, depending on whether
-  $u_1 \in G$ and whether $u_2 \in G$. If $u_1 \not\in G$, then $q_1 =
-  q'_1 = 0$, so $q_1q_2 = 0$ and $q'_1q'_2 = 0$. Since $(0,0) \in
-  \rho^{G,h}(u')$ for any $u'$, $G$ and $h$, we are done. If $u_1 \in
-  G$ and $u_2 \in G$, then we need to show that $(q_1q_2,
-  h(u_1)q_1h(u_2)q_2) \in \rho^{G,h}(u_1u_2)$. By $h$ being a
-  homomorphism, we have $(q_1q_2, h(u_1)q_1h(u_2)q_2) = (q_1q_1,
-  h(u_1u_2)q_1q_2)$, which is in $\rho^{G,h}(u_1u_2)$. The cases for
-  the other operations are similar.
-
-  From left to right, we have the following properties of any
-  relational environment $\rho_{\tyPrimNm{num}} \in
-  \relEnv{E}'(\Delta)$, derived from the hypothesis that all the
-  operations in $\eta_{\mathit{units}}$ satisfy their types'
-  relational semantics.
-  \begin{enumerate}
-  \item\label{item:obsv-poly} By the existence of the function
-    $\mathit{toRat}$, the relation $\rho_{\tyPrimNm{num}}(1)$ is the
-    identity relation.
-  \item\label{item:zero-poly} For any unit expression $e$, $(0,0) \in
-    \rho_{\tyPrimNm{num}}(e)$.
-  \item \label{item:one-poly} $(1,1) \in \rho_{\tyPrimNm{num}}(1)$.
-  \item \label{item:mult-poly} For any unit expressions $e_1,e_2$,
-    such that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e_1)$ and
-    $(q_2,q'_2) \in \rho_{\tyPrimNm{num}}(e_2)$, we have $(q_1q_2,
-    q'_1q'_2) \in \rho_{\tyPrimNm{num}}(e_1e_2)$.
-  \item\label{item:div-poly} For any unit expressions $e_1,e_2$, such
-    that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e_1)$ and $(q_2,q'_2)
-    \in \rho_{\tyPrimNm{num}}(e_2)$, we have that either $q_2 = q'_2 =
-    0$, or we have that $(q_1/q_2,q'_1,q'_2) \in
-    \rho_{\tyPrimNm{num}}(e_1e_2^{-1})$.
-  \item\label{item:compare-poly} For any unit expression $e$, such
-    that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e)$ and $(q_2,q'_2) \in
-    \rho_{\tyPrimNm{num}}(e)$, we have $q_1 < q_2$ iff $q'_1 < q'_2$.
-  \end{enumerate}
-  We now show that if $(q,q') \in \rho_{\tyPrimNm{num}}(e)$ for some
-  unit expression $e$, then either $q = q' = 0$, or $q' = rq$ for some
-  $r \in \mathbb{Q}^{>0}$, determined by $e$. We know from
-  \ref{item:zero-poly} above that $q = q' = 0$ is a possibility. Now
-  consider what happens if $q \not= 0$: in this case, by
-  \ref{item:zero-poly} and \ref{item:compare-poly}, we know that $0 <
-  q$ iff $0 < q'$, so $q' \not= 0$, and $q'$ has the same sign as
-  $q$. So $q/q'$ exists and is in $\mathbb{Q}^{>0}$. Hence $q = rq'$
-  for $r = q/q'$. So now we know that if $(q,q') \in
-  \rho_{\tyPrimNm{num}}(e)$, and $q \not= 0$ or $q' \not= 0$ then
-  there exists an $r$ such that $q = rq'$.
-
-  We now show that the unit expression $e$ determines the value of
-  $r$. Assume that we have $(q_1,r_1q_1) \in \rho_{\tyPrimNm{num}}(e)$
-  and $(q_2,r_2q_2) \in \rho_{\tyPrimNm{num}}(e)$. By
-  \ref{item:div-poly}, we know that $(q_1/q_2, (r_1q_1)/(r_2q_2)) \in
-  \rho_{\tyPrimNm{num}}(1)$. By \ref{item:obsv-poly},
-  $\rho_{\tyPrimNm{num}}(1)$ is the equality relation, so we can
-  deduce that $r_1 = r_2$.
-
-  Now let $G$ be the subgroup of $\idxTms{\Delta}{\mathsf{unit}}$ such
-  that $e \in G$ implies that $\rho_{\tyPrimNm{num}}(e) \not=
-  \{(0,0)\}$. By the argument above, for each $e \in G$, there is a
-  unique $r \in \mathbb{Q}^{>0}$ such that $\rho_{\tyPrimNm{num}}(e) =
-  \{ (q,rq) \sepbar q \in \mathbb{Q} \}$. Thus there is a function $h
-  : G \to \mathbb{Q}^{>0}$. By \ref{item:one-poly} and
-  \ref{item:mult-poly}, $h$ is a group homomorphism. Thus we have
-  shown that $\rho_{\tyPrimNm{num}}$ has the form of
-  $\rho^{G,h}_{\tyPrimNm{num}}$ for some $G$ and $h$. Since
-  $\rho_{\tyPrimNm{num}}$ was an arbitrary element of
-  $\relEnv{E}'(\Delta)$, and $\Delta$ was arbitrary, we have shown
-  that $\relEnv{E}' \subseteq \relEnv{E}_{\mathit{units}}$.
-\end{proof}
-
-Some notes towards a generalisation:
-\begin{itemize}
-\item By inspecting Kennedy's proof, we can see that the operations of
-  addition and subtraction do not play any role, save for an
-  obligation to prove that they satisfy the relational semantics of
-  their types. This follows just by distributivity of addition over
-  multiplication.
-\item Kennedy's proof does not rely on any particular features of
-  $\mathbb{Q}$ other than the fact that it is an ordered field. By
-  inspecting the proof, it is possible to see that if we remove the
-  comparison operation $< : \forall u. \tyPrim{num}{u} \to
-  \tyPrim{num}{u} \to \tyPrimNm{bool}$ from $\Gamma_{\mathit{units}}$,
-  then the appropriate family of sets of relational environments uses
-  homomorphisms into the set of non-zero rationals:
-  \begin{displaymath}
-    \begin{array}{l}
-      \relEnv{E}_{\mathit{units}\backslash\{<\}}(\Delta) = \\
-      \quad\left\{ \rho^{G,h} \sepbar
-        \begin{array}{l}
-          G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{unit}} \\
-          h : G \to \mathbb{Q}^{\not=0} \\
-          h\textrm{ is a group homomorphism}
-        \end{array}
-      \right\}
-    \end{array}
-  \end{displaymath}
-  To complete the step in the proof where the comparison operation was
-  used, we use the division operation to show that if $(q,q') \in
-  \rho_{\tyPrimNm{num}}(e)$ for some $e$, and $q \not= 0$, then $q'
-  \not= 0$ (but they need not necessarily have the same sign).
-
-  Just as Kennedy's proof generalises to an arbitrary ordered field,
-  if we remove the comparison operation we can generalise to an
-  arbitrary field as the underlying semantics of $\tyPrimNm{num}$ and
-  $\tyPrimNm{rational}$.
-\end{itemize}
-
-\subsection{Euclidean Plane Geometry}\label{sec:2d-semantics}
-
-We set $\tyPrimSem{vec} = \mathbb{R}^2$, and $\tyPrimSem{real} =
-\mathbb{R}$. The semantic interpretation $\eta_{\mathit{2D}}$ of the
-operations in $\Gamma_{\mathit{2D}}$ consists of the standard vector
-zero, addition, negation and scaling in $\mathbb{R}^2$, along with
-projections to interpret $\mathit{getX}$ and $\mathit{getY}$, and
-pairing to interpret $\mathit{vec}$.
-
-Let
-\begin{displaymath}
-  \begin{array}{l}
-    \relEnv{E}_{\mathit{2D}}(\Delta) = \\
-    \quad\left\{ \rho^{G,h} \sepbar
-      \begin{array}{l}
-        G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{translation}} \\
-        h : G \to \mathbb{R}^2 \\
-        h\textrm{ is a group homomorphism}
-      \end{array}
-    \right\}
-  \end{array}
-\end{displaymath}
-where
-\begin{displaymath}
-  \rho^{G,h}_{\tyPrimNm{vec}}(t) = \left\{
-    \begin{array}{ll}
-      \{\} & \textrm{if }t \not\in G \\
-      \{(v,h(t) + v) \sepbar v \in \mathbb{R}^2 \} & \textrm{if }t \in G
-    \end{array}
-  \right.
-\end{displaymath}
-Note that, unlike the units of measure example above, the pair $(0,0)$
-need not appear in every relation. This is because we do not have a
-translation polymorphic constant like $0 : \forall u. \tyPrim{num}{u}$
-in the units of measure example.
-
-\begin{theorem}
-  The family of sets of relational environments
-  $\relEnv{E}_{\mathit{units}}$ satisfies the completeness property
-  \hyperref[eq:kennedy-completeness]{(\ref*{eq:kennedy-completeness})}
-  above. That is, for all $\relEnv{E}'$:
-  \begin{displaymath}
-    (\forall \Delta.\forall \rho \in \relEnv{E}'(\Delta).
-(\eta_{\mathit{2D}}, \eta_{\mathit{2D}}) \in \rsem{\Gamma_{\mathit{2D}}}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_{\mathit{2D}}
-  \end{displaymath}
-\end{theorem}
-
-\begin{proof}
-  From right to left, we must check that each operation satisfies the
-  relational interpretation of its type. This is very similar to the
-  case for the units of measure example above.
-
-  From left to right, we have the following properties of any
-  relational environment $\rho_{\tyPrimNm{vec}} \in
-  \relEnv{E}'(\Delta)$, derived from the hypothesis that all the
-  operations in $\eta_{\mathit{2D}}$ satisfy their types'
-  relational semantics.
-  \begin{enumerate}
-  \item\label{item:obsv-vec-poly} By the existence of the functions
-    $\mathit{getX}$ and $\mathit{getY}$, the relation
-    $\rho_{\tyPrimNm{vec}}(0)$ is the identity relation.
-  \item \label{item:zero-vec-poly} $(0,0) \in \rho_{\tyPrimNm{vec}}(0)$.
-  \item \label{item:add-vec-poly} For any translation expressions $t_1,t_2$,
-    such that $(v_1,v'_1) \in \rho_{\tyPrimNm{vec}}(t_1)$ and
-    $(v_2,v'_2) \in \rho_{\tyPrimNm{vec}}(t_2)$, we have $(v_1 + v_2,
-    v'_1 + v'_2) \in \rho_{\tyPrimNm{vec}}(t_1 + t_2)$.
-  \item\label{item:neg-vec-poly} For any translation expression $t$,
-    such that $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$, we have that
-    $(-v,-v') \in \rho_{\tyPrimNm{vec}}(-t)$.
-  \end{enumerate}
-  We now show that if $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$ for some
-  translation expression $t$, then $v' = w + v$ for some $w \in
-  \mathbb{R}^2$, where $w$ is determined by $t$. If we are given
-  $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$, then we can set $w = v' -
-  v$. This is uniquely determined by $t$: if we have $(v, w + v) \in
-  \rho_{\tyPrimNm{vec}}(t)$ and $(v, w' + v) \in
-  \rho_{\tyPrimNm{vec}}(t)$, then by \ref{item:add-vec-poly} and
-  \ref{item:neg-vec-poly}, we have $(v - v, (w + v) - (w' + v)) \in
-  \rho_{\tyPrimNm{vec}}(0)$. By \ref{item:obsv-vec-poly}, we know that
-  $\rho_{\tyPrimNm{vec}}(0)$ is the identity relation, so we can
-  deduce that $w = w'$.
-
-  Now define $G \subseteq \idxTms{\Delta}{\mathsf{translation}}$ to
-  have $t \in G$ if $\rho_{\tyPrimNm{vec}}(t) \not= \{\}$. Since the
-  translation factor $w$ is determined by $t$, we can define a
-  function $h : G \to \mathbb{R}^2$. By \ref{item:zero-vec-poly} and
-  \ref{item:add-vec-poly}, $h$ is a group homomorphism, and $(v,v')
-  \in \rho_{\tyPrimNm{vec}}$ iff $v = h(t) + v'$. Thus
-  $\rho_{\tyPrimNm{vec}}$ is of the form
-  $\rho^{G,h}_{\tyPrimNm{vec}}$, as required.
-\end{proof}
-
-Some more notes:
-\begin{enumerate}
-\item The proof is slightly simpler than the case for the units of
-  measure example above because the underlying semantics of
-  $\tyPrimNm{vec}$ forms an abelian group, so we can just use inverses
-  to determine the difference factor $w$, and we don't have to deal
-  with the possibility of division by zero.
-\item The proof above should generalise to any abelian group.
-\item Generalising to non-abelian groups does not look
-  straightforward: the step that shows that the translation factor $w$
-  is uniquely determined by the translation expression $t$ fails. We
-  end up with
-  \begin{displaymath}
-    0 = w + v + (-w') + (-v)
-  \end{displaymath}
-  from which it is not possible to conclude that $w = w'$. (Maybe
-  something using normal subgroups would work?)
-\item A slightly different example is obtained by considering
-  rotations rather than translations (we consider rotations as an
-  additive abelian group). In this case, we could have the following
-  typed operations:
-  \begin{eqnarray*}
-    0 & : & \forall r.\ \tyPrim{vec}{r} \\
-    + & : & \forall r.\ \tyPrim{vec}{r} \to \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
-    - & : & \forall r.\ \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
-    * & : & \forall r.\ \tyPrimNm{real} \to \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
-    \mathit{norm} & : & \forall r. \tyPrim{vec}{r} \to \tyPrimNm{real} \\
-    \mathit{angleOf} & : & \forall r. \tyPrim{vec}{r} \to \tyPrim{rot}{r} + 1 \\
-    \mathit{rotateBy} & : & \forall r_1 r_2. \tyPrim{vec}{r_1} \to \tyPrim{rot}{r_2} \to \tyPrim{vec}{r_1 + r_2} \\
-    0_{\mathit{rot}} & : & \tyPrim{rot}{0} \\
-    +_{\mathit{rot}} & : & \forall r_1 r_2. \tyPrim{rot}{r_1} \to \tyPrim{rot}{r_2} \to \tyPrim{rot}{r_1 + r_2} \\
-    -_{\mathit{rot}} & : & \forall r. \tyPrim{rot}{r} \to \tyPrim{rot}{-r}
-  \end{eqnarray*}
-  where $\tyPrimSem{vec} = \mathbb{R}^2$, as before, and
-  $\tyPrimSem{rot} = \mathrm{SO}(2)$, the group of rotations of the
-  plane. The operation $\mathit{angleOf}$ is intended to return
-  $\mathrm{inr}(*)$ when the provided vector is zero.
-
-  In this case, rotations form an abelian group, so we can re-use the
-  proof above to give a family of sets of relational environments that
-  is complete for the group operations on rotations. The fact that
-  rotations are linear admits the typing of the first four
-  operations. The existence of the $\mathit{angleOf}$ and
-  $\mathit{rotateBy}$ functions allow us to determine ``rotation
-  factors'' between arbitrary pairs of vectors.
-
-  I conjecture that the following definition of
-  $\relEnv{E}_{\mathit{2Drot}}$ satisfies the completeness criterion,
-  with the standard interpretation of the operations above.
-  \begin{displaymath}
-    \begin{array}{l}
-      \relEnv{E}_{\mathit{2Drot}}(\Delta) = \\
-      \quad\left\{ \rho^{G,h} \sepbar
-        \begin{array}{l}
-          G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{rotation}} \\
-          h : G \to \mathrm{SO}(2) \\
-          h\textrm{ is a group homomorphism}
-        \end{array}
-      \right\}
-    \end{array}
-  \end{displaymath}
-  where
-  \begin{displaymath}
-    \rho^{G,h}_{\tyPrimNm{vec}}(r) = \left\{
-      \begin{array}{ll}
-        \{(0,0)\} & \textrm{if }t \not\in G \\
-        \{(v,h(r) v) \sepbar v \in \mathbb{R}^2 \} & \textrm{if }t \in G
-      \end{array}
-    \right.
-  \end{displaymath}
-  and
-  \begin{displaymath}
-    \rho^{G,h}_{\tyPrimNm{rot}}(r) = \left\{
-      \begin{array}{ll}
-        \{\} & \textrm{if }t \not\in G \\
-        \{(s,h(r) s) \sepbar s \in \mathrm{SO}(2) \} & \textrm{if }t \in G
-      \end{array}
-    \right.
-  \end{displaymath}
-\end{enumerate}
-
-\subsection{Security Typing}\label{sec:sec-semantics}
-
-In this case, we let $\tyPrimSem{T} = \{*\}$, and the operations
-$\mathit{proj}$ and $\mathit{combine}$ are given the only possible
-interpretations.
-
-I conjecture that the follow definition works:
-\begin{displaymath}
-  \begin{array}{l}
-    \relEnv{E}_{\mathit{sec}}(\Delta) = \\
-    \quad\left\{ \rho^{h} \sepbar
-      \begin{array}{l}
-        h : \idxTms{\Delta}{\mathsf{principal}} \to \{\top,\bot\} \\
-        h\textrm{ is a boolean algebra homomorphism}
-      \end{array}
-    \right\}
-  \end{array}
-\end{displaymath}
-where
-\begin{displaymath}
-  \rho^{h}_{\tyPrimNm{T}}(p) = \left\{
-    \begin{array}{ll}
-      \{\} & \textrm{if }h(p) = \bot \\
-      \{(*,*) \} & \textrm{if }h(p) = \top
-    \end{array}
-  \right.
-\end{displaymath}
-
-One can compare this definition to the previous ones by considering
-the boolean algebra homomorphism as defining a sub-boolean algebra of
-$\idxTms{\Delta}{\mathsf{principal}}$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Extensions}
-
-\subsection{Inductive Types}
-
-\subsection{Type Operators}
-
-\subsection{Size indexing and Combinatorial Species}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{plainnat}
-\bibliography{paper}
-
-\end{document}

### 2012-05-22 paper/paper.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

--- 2012-05-08 paper/paper.tex
@@
 
+\newcommand{\GA}{\mathrm{GA}}
+\newcommand{\GL}{\mathrm{GL}}
+\newcommand{\Is}{\mathrm{Is}}
+\newcommand{\Orth}{\mathrm{O}}
+\newcommand{\SOrth}{\mathrm{SO}}
+\newcommand{\Transl}{\mathrm{T}}
+\newcommand{\Scale}{\mathrm{Scale}}
+
 \newcommand{\sepbar}{\mathrel|}
@@
 
+\newcommand{\SortSet}{\mathit{Sort}}
+\newcommand{\IndexOpSet}{\mathit{IndexOp}}
+\newcommand{\PrimTypeSet}{\mathit{PrimType}}
+\newcommand{\IndexAxiomSet}{\mathit{IndexAx}}
+
 \newcommand{\indexOp}[1]{\texttt{#1}}
@@
 
-\newcommand{\sortType}{\texttt{type}}
 \newcommand{\tyInt}{\texttt{int}}
@@
 \newcommand{\tyPrimNm}[1]{\texttt{#1}}
+\newcommand{\primTyArity}{\mathrm{tyArity}}
+\newcommand{\indexOpArity}{\mathrm{opArity}}
 
@@
 \newcommand{\tmSem}[1]{\llbracket #1 \rrbracket^{\mathit{tm}}}
-\newcommand{\tyPrimSem}[1]{\llbracket \tyPrimNm{#1} \rrbracket^{\mathcal{T}_0}}
+\newcommand{\tyPrimSem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}_0}}
 \newcommand{\rsem}[3]{\llbracket #1 \rrbracket^{\mathcal{R}}_{#2}{#3}}
@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{example}{Example}
 \newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
@@
 \newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}
+\newcommand{\statementref}[1]{\hyperref[#1]{Statement \ref*{#1}}}
+
+\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
+\newcommand{\isDefinedAs}{\stackrel{\mathit{def}}=}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-
-\begin{enumerate}
-\item Introduction using the area-of-a-triangle example
-\item Narrowing of the semantic gap by use of advanced type systems
-\item Importance of the use of symmetry in computer science,
-  mathematics and the physical sciences
-\item Development of the theory of parametricity
-\end{enumerate}
-
-\subsection{Contributions}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Background and Related Work}
-
-Reynolds (Parametricity), Wadler (Free Theorems).
-
-Kennedy (Units of Measure)
-
-Something about symmetry?
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Type System}
-
-We now introduce our type system. We parameterise our development by
-an arbitrary multi-sorted algebraic theory, in the sense of universal
-algebra. We present the precise form of the theories we allow in
-\autoref{sec:index-exp}, and define some auxiliary notions of
-substitution and terms-up-to-equality. Several examples of useful
-algebraic theories will be given in
-\autoref{sec:example-systems}. Once an algebraic theory has been
-fixed, we define the well-formed types and terms with respect to this
-theory in \autoref{sec:types-and-terms}.
-
-\subsection{Algebraic Theories for Indexing}\label{sec:index-exp}
-
-\begin{figure*}[t]
-  \centering
-  Well-sorted terms
-  \begin{mathpar}
-    \inferrule*
-    {i : s \in \Delta}
-    {\Delta \vdash x : s}
-
-    \inferrule*
-    {\indexOp{f} : s_1,...,s_n \to s \\
-      \Delta \vdash e_1 : s_1 \\
-      \dots \\
-      \Delta \vdash e_n : s_n}
-    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) : s}
-  \end{mathpar}
-
-  \medskip
-
-  Equational rules
-  \begin{mathpar}
-    \inferrule*
-    {e_1 \stackrel{ax}= e_2 : s \in \mathit{Ax}(\Delta)}
-    {\Delta \vdash e_1 \equiv e_2 : s}
-
-    \inferrule*
-    {\indexOp{f} : s_1,...,s_n \to s \\
-      \Delta \vdash e_1 \equiv e'_1 : s_1 \\
-      \dots \\
-      \Delta \vdash e_n \equiv e'_n : s_n}
-    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) \equiv \indexOp{f}(e'_1, ..., e'_n) : s}
-
-    \inferrule*
-    {\Delta \vdash e : s}
-    {\Delta \vdash e \equiv e : s}
-
-    \inferrule*
-    {\Delta \vdash e_1 \equiv e_2 : s}
-    {\Delta \vdash e_2 \equiv e_1 : s}
-
-    \inferrule*
-    {\Delta \vdash e_1 \equiv e_2 : s \\ \Delta \vdash e_2 \equiv e_3 : s}
-    {\Delta \vdash e_1 \equiv e_3 : s}
-  \end{mathpar}
-  \caption{Algebraic Theories for Indexing}
-  \label{fig:alg-theories}
-\end{figure*}
-
-We assume a set $\mathit{Sort}$ of \emph{index sorts}, and we will use
-$s, s_1, s_2, ...$ to range over elements of $\mathit{Sort}$. An index
-context $\Delta$ is list of index variable and sort pairs, $i : s$,
-and we use the symbol $\epsilon$ to denote the empty index context (we
-will also use this symbol below to denote the empty typing
-context). We use letters like $i$ and $j$, sometimes with subscripts,
-to range over index variables. We also assume a collection
-$\mathit{IndexOp}$ of sorted index operations $\indexOp{f} :
-s_1,...,s_n \to s$. From the index operations we generate the
-well-sorted index expressions by the rules in
-\autoref{fig:alg-theories}.
-
-For each index context $\Delta$, we assume a set $\mathit{Ax}(\Delta)$
-of axioms $e_1 \stackrel{ax}= e_2 : s$ where both $\Delta \vdash e_1 :
-s$ and $\Delta \vdash e_2 : s$. From this collection of axioms, we
-generate the equational theory of index expressions by the rules in
-\autoref{fig:alg-theories} by closing under congruence, reflexivity,
-symmetry and transitivity.
-
-% For a pair of index contexts $\Delta$ and $\Delta'$, a substitution
-% $\sigma \in \subst{\Delta}{\Delta'}$ is a function $\sigma$ mapping
-% variables in $\Delta$ to well-sorted index expressions in $\Delta'$.
-
-Index context morphisms $\sigma : \Delta \Rightarrow \Delta'$ consist
-of a sequence of terms, one for each element of $\Delta'$, all
-well-sorted in the index context $\Delta$.
-\begin{mathpar}
-  \inferrule*
-  { }
-  {\cdot : \Delta \Rightarrow \epsilon}
-
-  \inferrule*
-  {\sigma : \Delta \Rightarrow \Delta' \\ \Delta \vdash e : s}
-  {\sigma, e : \Delta \Rightarrow \Delta', i:s}
-\end{mathpar}
-Directly from the definition of index context morphisms, we can see
-that given an index context morphism $\sigma : \Delta \Rightarrow
-\Delta'$, then for any index variable $i : s \in \Delta'$ there is an
-index expression $\sigma(i)$ such that $\Delta \vdash \sigma(i) : s$.
-
-We write $\id_\Delta : \Delta \Rightarrow \Delta$ for the index
-context morphism comprised of the terms built from the variables of
-$\Delta$, in order. We write $\pi_{i\mathord:s} : \Delta, i : s
-\Rightarrow \Delta$ for the context morphism again comprised of the
-terms built from the variables of $\Delta$, but this time typed in a
-larger context. For an index context morphism $\sigma : \Delta \to
-\Delta'$ we write $\sigma_s : \Delta, i\mathord:s \Rightarrow \Delta',
-i\mathord:s$ for the index context morphism generated by lifting each
-term to the larger context $\Delta, i\mathord:s$, and mapping the
-variable $i\mathord:s$ to the term $i$.
-
-Given an index expression $\Delta' \vdash e : s$, we define
-re-indexing by an index context morphism $\sigma : \Delta \Rightarrow
-\Delta'$ by the following clauses:
-\begin{eqnarray*}
-  \sigma^*i & = & \sigma(i) \\
-  \sigma^*\indexOp{f}(e_1, ..., e_n) & = & \indexOp{f}(\sigma^*e_1,...,\sigma^*e_n)
-\end{eqnarray*}
-It is easy to check that these clauses preserve well-formedness of
-index expressions, so that we obtain $\Delta \vdash \sigma^*e :
-s$. Moreover re-indexing by index context morphisms preserves
-equality:
-\begin{lemma}\label{lem:indexeq-subst}
-  If $\Delta \vdash e_1 \equiv e_2 : s$, then for any $\sigma :
-  \Delta' \Rightarrow \Delta$, we have $\Delta' \vdash \sigma^*e_1
-  \equiv \sigma^*e_2 : s$.
-\end{lemma}
-
-Also, projection commutes with lifting of substitutions. This is used
-in the proof of \lemref{lem:tysubst-rel}.
-\begin{lemma}
-  $\sigma^* \circ \pi^*_{i\mathord:s} = \pi^*_{i\mathord:s} \circ \sigma^*_s$
-\end{lemma}
-
-For an index context $\Delta$ and sort $s$, we write
-$\idxTms{\Delta}{s}$ for the set of index expressions $e$ such that
-$\Delta \vdash e : s$, quotiented by the equivalence relation $\Delta
-\vdash - \equiv - : s$. In light of \lemref{lem:indexeq-subst}, we can
-see that re-indexing by an index context morphism $\sigma : \Delta
-\Rightarrow \Delta'$ can be used as a function $\sigma^* :
-\idxTms{\Delta'}{s} \to \idxTms{\Delta}{s}$, for any sort $s$.
-
-\subsection{Types and Terms}\label{sec:types-and-terms}
-
-\begin{figure*}[t]
-  \centering
-  Well-indexed Types
-  \begin{mathpar}
-    \inferrule* [right=TyPrimIdx]
-    {\tyPrimNm{X} \in \mathit{PrimType}(s) \\ \Delta \vdash e : s}
-    {\Delta \vdash \tyPrim{X}{e} : \sortType}
-
-    \inferrule* [right=TyPrimNoIdx]
-    {\tyPrimNm{X} \in \mathit{PrimType}(\cdot)}
-    {\Delta \vdash \tyPrimNm{X} : \sortType}
-
-    \inferrule* [right=TyUnit]
-    { }
-    {\Delta \vdash \tyUnit : \sortType}
-
-    \inferrule* [right=TyArr]
-    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
-    {\Delta \vdash A \tyArr B : \sortType}
-
-    \inferrule* [right=TyProd]
-    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
-    {\Delta \vdash A \tyProduct B : \sortType}
-
-    \inferrule* [right=TySum]
-    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
-    {\Delta \vdash A + B : \sortType}
-
-    \inferrule* [right=TyForall] %FIXME: macroize forall
-    {\Delta, i : s \vdash A : \sortType}
-    {\Delta \vdash \forall i \mathord: s. A : \sortType}
-  \end{mathpar}
-
-  \medskip
-  Equality of well-indexed types
-  \begin{mathpar}
-    \inferrule* [right=TyEqPrimIdx]
-    {\Delta \vdash e_1 \equiv e_2 : s}
-    {\Delta \vdash \tyPrim{X}{e_1} \equiv \tyPrim{X}{e_2} : \sortType}
-
-    \inferrule* [right=TyEqPrimNoIdx]
-    { }
-    {\Delta \vdash \tyPrimNm{X} \equiv \tyPrimNm{X} : \sortType}
-
-    \inferrule* [right=TyEqUnit]
-    { }
-    {\Delta \vdash \tyUnit \equiv \tyUnit : \sortType}
-
-    \inferrule* [right=TyEqArr]
-    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
-    {\Delta \vdash A \tyArr B \equiv A' \tyArr B' : \sortType}
-
-    \inferrule* [right=TyEqProd]
-    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
-    {\Delta \vdash A \tyProduct B \equiv A' \tyProduct B' : \sortType}
-
-    \inferrule* [right=TyEqSum]
-    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
-    {\Delta \vdash A + B \equiv A' + B' : \sortType}
-
-    \inferrule* [right=TyEqForall]
-    {\Delta, i : s \vdash A \equiv A' : \sortType}
-    {\Delta \vdash \forall i \mathord: s. A \equiv \forall i \mathord: s. A' : \sortType}
-  \end{mathpar}
-
-  \medskip
-  Well-indexed contexts
-  \begin{mathpar}
-    \inferrule*
-    { }
-    {\Delta \vdash \epsilon \isCtxt}
-
-    \inferrule*
-    {\Delta \vdash \Gamma \isCtxt \\ \Delta \vdash A : \sortType \\ x \not\in \Gamma}
-    {\Delta \vdash \Gamma, x : A \isCtxt}
-  \end{mathpar}
-  \caption{Well-indexed Types and Contexts}
-  \label{fig:types}
-\end{figure*}
-
-For each sort $s \in \mathit{Sort}$, we assume we are given a set of
-names of primitive types $\mathit{PrimType}(s)$. We assume that the
-sets $\mathit{PrimType}(s)$ are disjoint for non-equal sorts $s_1,
-s_2$. For example, in the units of measure example in
-\autoref{sec:units-example}, there is a single sort $\mathsf{unit}$,
-and $\mathit{PrimType}(\mathsf{unit}) = \{\tyPrimNm{num}\}$. For each
-primitive type $\tyPrimNm{X} \in \mathit{PrimType}(s)$, and index
-expression $\Delta \vdash e : s$, we form the type $\tyPrim{X}{e}$, by
-the \TirName{TyPrimIdx} rule in \autoref{fig:types}. Thus, in the
-units-of-measure example, the type $\tyPrim{num}{u}$ represents
-numbers measured by the units $u$.
-
-We also assume that there is an additional set of primitive types
-$\mathit{PrimType}(\cdot)$ that is disjoint from all the other sets of
-primitive types. These primitive types do not need to be paired with
-an index expression to form well-indexed types. Such types are created
-by the \TirName{TyPrimNoIdx} rule. An example of such a primitive type
-is the type $\tyPrimNm{real}$ of real-valued scalars in the Euclidean
-plane geometry example in \autoref{sec:2d-example}.
-
-The rules for generating the rest of the types of the system,
-including unit, product, sum, function and universal types, are given
-in \autoref{fig:types}. We use capital roman letters like $A$ and $B$
-to range over well-indexed types. Well-indexed types come with an
-equational theory, derived from the equational theory on index
-expressions. This equational theory is also presented in
-\autoref{fig:types}.
-
-We define substitution by an index context morphism $\sigma : \Delta
-\Rightarrow \Delta'$ on types by structural induction on the structure
-of the type: given a type $\Delta' \vdash A : \sortType$, we obtain a
-type $\Delta \vdash \sigma^*A : \sortType$. The only interesting case
-is for \TirName{TyPrimIdx}:
-\begin{displaymath}
-  \sigma^*\tyPrim{X}{e} = \tyPrim{X}{\sigma^*e}
-\end{displaymath}
-
-Well-formed contexts are defined to be lists of variables paired with
-well-formed types with respect to some index context $\Delta$, with
-the condition that no variable names are repeated. Formally, contexts
-are defined by the two rules shown in
-\autoref{fig:types}. Substitution by an index context morphism extends
-to contexts just by doing substitution on each individual type.
-
-The typing rules for our system define the well-typed terms with
-respect to an index context $\Delta$ and a type context $\Delta \vdash
-\Gamma \isCtxt$. The judgement $\Delta; \Gamma \vdash M : A$ is
-defined in \autoref{fig:terms}. The equational theory on types is
-incorporated into the type system via the rule \TirName{TyEq}, which
-allows for a term that is judged to have type $A$ to also have any
-other equal type $B$.
-\begin{figure*}[t]
-  \centering
-  \begin{mathpar}
-    \inferrule* [right=Var]
-    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
-    {\Delta; \Gamma \vdash x : A}
-
-    \inferrule* [right=TyEq]
-    {\Delta; \Gamma \vdash M : A \\ \Delta \vdash A \equiv B : \sortType}
-    {\Delta; \Gamma \vdash M : B}
-
-    \inferrule* [right=Unit]
-    { }
-    {\Delta; \Gamma \vdash * : 1}
-
-    \inferrule* [right=Pair]
-    {\Delta; \Gamma \vdash M : A \\
-      \Delta; \Gamma \vdash N : B}
-    {\Delta; \Gamma \vdash (M, N) : A \tyProduct B}
-
-    \inferrule* [right=Proj1]
-    {\Delta; \Gamma \vdash M : A \tyProduct B}
-    {\Delta; \Gamma \vdash \pi_1 M : A}
-
-    \inferrule* [right=Proj2]
-    {\Delta; \Gamma \vdash M : A \tyProduct B}
-    {\Delta; \Gamma \vdash \pi_2 M : B}
-
-    \inferrule* [right=Inl]
-    {\Delta; \Gamma \vdash M : A}
-    {\Delta; \Gamma \vdash \mathrm{inl}\ M : A + B}
-
-    \inferrule* [right=Inr]
-    {\Delta; \Gamma \vdash M : B}
-    {\Delta; \Gamma \vdash \mathrm{inr}\ M : A + B}
-
-    \inferrule* [right=Cond]
-    {\Delta; \Gamma \vdash M : A + B \\
-      \Delta; \Gamma, x : A \vdash N_1 : C \\
-      \Delta; \Gamma, y : B \vdash N_2 : C}
-    {\Delta; \Gamma \vdash \textrm{case}\ M\ \textrm{of}\ \textrm{inl}\ x.N_1; \textrm{inr}\ y.N_2 : C}
-
-    \inferrule* [right=Abs]
-    {\Delta; \Gamma, x : A \vdash M : B}
-    {\Delta; \Gamma \vdash \lambda x.M : A \tyArr B}
-
-    \inferrule* [right=App]
-    {\Delta; \Gamma \vdash M : A \tyArr B \\
-      \Delta; \Gamma \vdash N : A}
-    {\Delta; \Gamma \vdash M N : B}
-
-    \inferrule* [right=UnivAbs]
-    {\Delta, i : s; \pi_{i\mathord:s}^*\Gamma \vdash M : A}
-    {\Delta; \Gamma \vdash \Lambda i. M : \forall i\mathord:s. A}
-
-    \inferrule* [right=UnivApp]
-    {\Delta; \Gamma \vdash M : \forall i\mathord:s. A \\ \Delta \vdash e : s}
-    {\Delta; \Gamma \vdash M [e] : (\id_\Delta, e)^*A}
-  \end{mathpar}
-
-  \caption{Well-typed terms}
-  \label{fig:terms}
-\end{figure*}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Example Systems}\label{sec:example-systems}
-
-\subsection{Units of Measure}\label{sec:units-example}
-
-We now show that our general framework encompasses Kennedy's
-units-of-measure system \cite{kennedy97relational}. For our indexing
-algebraic theory, we have a single sort $\mathsf{unit}$ intended to
-represent composite units such as $\mathit{m}\ \mathit{s}^{-1}$ or
-$\mathit{kg}\ \mathit{m}^{-2}$. Units of measure form an abelian
-group, so we have three operations:
-\begin{mathpar}
-  1 : \mathsf{unit},
-
-  \cdot : \mathsf{unit}, \mathsf{unit} \to \mathsf{unit},
-
-  -^{-1} : \mathsf{unit} \to \mathsf{unit},
-\end{mathpar}
-subject to the normal abelian group axioms. We will elide the group
-multiplication operation and just write multiplication as
-juxtaposition: $u_1u_2$.
-
-We let $\mathit{PrimType}(\mathsf{unit}) = \{\tyPrimNm{num}$, and
-$\mathit{PrimType}(\cdot) = \{\tyPrimNm{rational}\}$. The intended
-reading is that the type $\tyPrim{num}{u_1u^2_2}$ represents
-quantities in units of $u_1u^2_2$, while values of type
-$\tyPrimNm{rational}$ represent dimensionless constants. We make a
-distinction between $\tyPrim{num}{1}$ and $\tyPrimNm{rational}$ for
-technical reasons arising from the relational interpretation below: we
-want the relation on $\tyPrim{num}{1}$ to be the identity relation,
-and the existence of $\mathit{toRat}$ and $\mathit{fromRat}$ forces
-this\footnote{Kennedy assumes that all relational environments set the
-  relation on $\tyPrim{num}{1}$ to be the identity relation, but we
-  cannot do this in general, because we do not necessarily have a
-  group structure in the indexing theory.}. To write programs using
-the $\tyPrimNm{num}$ type, we assume an typing context
-$\Gamma_{\textrm{units}}$ consisting of the following typed
-operations:
-\begin{eqnarray*}
-  \mathit{toRat} & : & \tyPrim{num}{1} \to \tyPrimNm{rational} \\
-  \mathit{fromRat} & : & \tyPrim{num}{1} \to \tyPrimNm{rational} \\
-  0 & : & \forall u. \tyPrim{num}{u} \\
-  1 & : & \tyPrim{num}{1} \\
-  + & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyPrim{num}{u} \\
-  - & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyPrim{num}{u} \\
-  * & : & \forall u_1 u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to \tyPrim{num}{u_1u_2} \\
-  / & : & \forall u_1 u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to \tyPrim{num}{u_1u_2^{-1}} + 1 \\
-  < & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyBool
-\end{eqnarray*}
-There are constants for zero and one, but only the constant for zero
-is polymorphic in the units. In \autoref{sec:units-semantics} below,
-we will following Kennedy by interpreting units of measure in terms of
-\emph{scalings} by rational numbers, where it will be apparent that
-the constant one cannot be unit-polymorphic: we can scale $0$ and
-still get $0$, but we cannot scale $1$ by an arbitrary rational number
-and expect to get $1$.
-
-\subsection{Euclidean Plane Geometry}\label{sec:2d-example}
-
-We now consider a type system for writing programs that are invariant
-under the translational symmetries of the Euclidean plane. We assume a
-single sort $\mathsf{translation}$. The sort $\mathsf{translation}$ is
-intended to represent translations in the plane, and comes equipped
-with abelian group structure that we write additively:
-\begin{mathpar}
-  0 : \mathsf{translation}
-
-  + : \mathsf{translation},\mathsf{translation} \to \mathsf{translation}
-
-  - : \mathsf{translation} \to \mathsf{translation}
-\end{mathpar}
-In indexing expressions involving variables of sort
-$\mathsf{translation}$, we use the shorthand $t_1 - t_2$ to stand for
-$t_1 + (-t_2)$.
-
-There are two primitive types:
-$\mathit{PrimType}(\mathsf{translation}) = \{\tyPrimNm{vec}\}$, and
-$\mathit{PrimType}(\cdot) = \{\tyPrimNm{real}\}$. The types
-$\tyPrim{vec}{t}$ represent two-dimensional vectors, indistinguishable
-up to some translation $t$. Values of type $\tyPrimNm{real}$ are just
-real numbers. To write programs using these types we assume a typing
-context $\Gamma_{\mathit{2D}}$ consisting of the following operations:
-\begin{eqnarray*}
-  + & : & \forall t_1 t_2\mathord:\mathsf{translation}.\tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2} \\
-  - & : & \forall t\mathord:\mathsf{translation}. \tyPrim{vec}{t} \to \tyPrim{vec}{-t} \\
-  0 & : & \tyPrim{vec}{0} \\
-  * & : & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0} \\
-  \mathrm{getX} & : & \tyPrim{vec}{0} \to \tyPrimNm{real} \\
-  \mathrm{getY} & : & \tyPrim{vec}{0} \to \tyPrimNm{real} \\
-  \mathrm{vec}  & : & \tyPrimNm{real} \to \tyPrimNm{real} \to \tyPrim{vec}{0}
-\end{eqnarray*}
-The first three operations express the abelian group structure of
-2-dimensional real vectors. Note how the indexing tracks the
-operations on vectors: e.g.~addition on vectors is reflected by
-addition on the indicies. The final four operations are operations
-that are not invariant under arbitrary translation: multiplication by
-a scalar, and projection from and constructon of vectors. For the
-purposes of examples, we also assume a standard set of arithmetic and
-other operations on real numbers.
-
-As an example of the use of the above operations, consider the
-following type of a function for computing the area of a triangle
-described by three points. Since the area of a triangle is invariant
-under translations that are applied to all three points, we can
-quantify over all translations $t$:
-\begin{displaymath}
-  \mathit{area} : \forall t\mathord:\mathsf{translation}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real}
-\end{displaymath}
-Since the result is invariant under translation, it safe to assume
-that one of the vertexes of the triangle is the origin, and only
-specify two points to describe the triangle. So we might expect that
-the type of $\mathit{area}$ is isomorphic to this type:
-\begin{displaymath}
-  \mathit{area'} : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrimNm{real}
-\end{displaymath}
-Given $\mathit{area'}$, we can define $\mathit{area}$ using the
-operations in $\Gamma_{\mathit{2D}}$:
-\begin{displaymath}
-  \mathit{area} = \Lambda t. \lambda v_1 v_2 v_3.\ \mathit{area'}\ (v_2 - v_1)\ (v_3 - v_1)
-\end{displaymath}
-and vice versa, we can define $\mathit{area'}$ in terms of
-$\mathit{area}$:
-\begin{displaymath}
-  \mathit{area'} = \lambda v_1 v_2. \mathit{area}\ [0]\ 0\ v_1\ v_2
-\end{displaymath}
-In \autoref{sec:2d-semantics} below, we will show that these two
-translations form a type isomorphism between the types of
-$\mathit{area}$ and $\mathit{area'}$.
-
-\paragraph{Alternative Transformation Groups?}
-We should also consider other groups of transformations for indexing:
-isometries, orthogonal transformations, rotations and reflections.
-
-\subsection{Security Typing and Non-interference}
-
-This example shows how our framework can accommodate non-interference
-properties common in security type systems. This example has a
-different flavour to the two previous examples because it does not
-concern invariance under the action of some group. We assume a single
-sort $\mathsf{principal}$, intended to represent (possibly composite)
-principals in a system. Principals are combined using the connectives
-of boolean logic:
-\begin{mathpar}
-  \top : \mathsf{principal}
-
-  \bot : \mathsf{principal}
-
-  \land : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}
-
-  \lor : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}
-
-  \Rightarrow : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}
-
-  \lnot : \mathsf{principal} \to \mathsf{principal}
-\end{mathpar}
-For the equational theory of principals, we assume all the axioms of
-boolean algebra.
-
-Let $\mathit{PrimType}(\mathsf{principal}) = \{\tyPrimNm{T}\}$. The
-intended reading of a type $\tyPrim{T}{p}$ for some principal $p$ is
-that the existence of a value of this type indicates that the
-principal $p$ is ``happy''. To write programs, we assume the following
-typing context $\Gamma_{\mathit{sec}}$ of operations:
-\begin{eqnarray*}
-  \mathrm{proj} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1 \land p_2} \to \tyPrim{T}{p_1} \\
-  \mathrm{combine} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1} \to \tyPrim{T}{p_2} \to \tyPrim{T}{p_1 \land p_2}
-\end{eqnarray*}
-
-To illustrate the use of this application of our framework, we define
-a type synonym, for each type $A$ and principal $p$:
-\begin{displaymath}
-  T_pA = \tyPrim{T}{p} \to A
-\end{displaymath}
-For every principal $p$, we can endow the types $T_p-$ with the
-structure of a monad. This is due to the fact that it is an instance
-of the ``environment'' (or ``reader'') monad. Explicitly, the
-$\mathit{return}$ and $>>=$ of these monads are given by:
-\begin{eqnarray*}
-  \mathit{return} & : & A \to T_pA \\
-  \mathit{return} & = & \lambda a\ t.\ a \\
-  \\
-  (>>=) & : & T_pA \to (A \to T_pB) \to T_pB \\
-  (>>=) & = & \lambda a\ f\ t.\ f\ (a\ t)\ t
-\end{eqnarray*}
-The intended reading of a value of type $T_pA$ is as a piece of data
-of type $A$ that can only be accessed if the principal $p$ is
-happy. The existence of such a monad should allow us to draw parallels
-between our system and the Dependency Core Calculus (DCC) system of
-Abadi \emph{et al.} \cite{abadi99core}. The definition of a DCC monad
-as an instance of an environment monad follows Tse and Zdancewic's
-interpretation of DCC in System F, where they use System F
-parametricity to ensure non-interference properties.
-
-There is also a coercion operator between instances of $T_pA$ for
-different principals. Reading the type in terms of happiness, this
-says that if we have a piece of data that is protected by $p$, and
-happiness of $q$ implies happiness of $p$, then this data can also be
-accessed by $q$.
-\begin{eqnarray*}
-  \mathit{coerce} & : & \tyPrim{T}{q \Rightarrow p} \to T_pA \to T_qA \\
-  \mathit{coerce} & = & \lambda t_{qp}\ a\ t_q.\ a\ (\mathrm{combine}\ t_{qp}\ t_q)
-\end{eqnarray*}
-Here we have used the fact that the result of the application of
-$\mathrm{combine}$ has type $\tyPrim{T}{(q \Rightarrow p) \land
-  q}$. By the laws of boolean algebras---which we have assumed for our
-indexing theory---this is equal to the type $\tyPrim{T}{p}$. Thus we
-get access to the data stored in $a$, which is protected by $p$.
-
-Now suppose we have a term $M$ with the following typing:
-\begin{displaymath}
-  p,q:\mathsf{principal}; t_{qp} : \tyPrim{T}{q \Rightarrow p}, x : T_q (1 + 1) \vdash M : T_p (1 + 1)
-\end{displaymath}
-Intuitively, given the operations $\Gamma_{\mathit{sec}}$ we have
-defined above, it should be the case that $M$ can only be equivalent
-to the program which is constantly $\lambda t. \mathrm{inl}(*)$ or
-$\lambda t. \mathrm{inr}(*)$, because there is no useful data to be
-gained from the $t_{qp}$ argument, and a value of $\tyPrim{T}{q}$
-cannot be generated in order to get access to the data within the $x$
-argument. The application of our framework that we have presented in
-this section has a ``non-interference'' property: the value of $x$
-cannot interfere with the output. In \autoref{sec:sec-semantics} we
-show that this is indeed the case.
-
-\paragraph{Alternative Logics?}
-Above, we assumed that the indexing theory of principals was the
-theory of boolean algebras. As a consequence, a value of type $T_{p
-  \lor \lnot p}A$ can always be accessed, due to the exlcuded middle
-property of boolean logic. An interesting question is: what happens we
-we consider alternative logics. If we take the theory of Heyting
-algebras for our theory of principals, then data typed as $T_{p \lor
-  \lnot p}A$ can only be accessed if either $p$ or $\lnot p$ can be
-demonstrated. Considering an alternative logical theory alters the
-non-interference property of the type system.
-
-\paragraph{Alternative Presentations?}
-In the above we presented the theory of boolean logic by means of the
-equational theory of boolean algebras. This isn't necessarily the most
-convenient way of presenting a logic. An alternative approach would be
-to consider a trivial equational theory with no axioms, and
-incorporate the rules of logic consequence as operations in
-$\Gamma_{\mathit{sec}}$. For example, as a Hilbert system:
-\begin{eqnarray*}
-  \mathit{MP} & : & \forall p,q.\ \tyPrim{T}{p \Rightarrow q} \to \tyPrim{T}{p} \to \tyPrim{T}{q} \\
-  \mathit{k} & : & \forall p,q.\ \tyPrim{T}{p \Rightarrow q \Rightarrow p} \\
-  \mathit{s} & : & \forall p,q,r.\ \tyPrim{T}{(p \Rightarrow q \Rightarrow r) \Rightarrow (p \Rightarrow q) \Rightarrow p \Rightarrow r}
-\end{eqnarray*}
-Plus axioms for the other connectives.  Or possibly some kind of
-encoding of a natural deduction system?
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Semantics and Parametricity Theorem}
-
-There is an index-erasure semantics, and a relational semantics. The
-relational semantics allows us to derive free theorems about the
-index-erasure semantics.
-
-\subsection{Index Erasure Semantics}
-
-We now define a semantics of types and terms that ignores all indexing
-information. By defining this semantics we can easily see that the
-indexing information can be discarded before execution.
-
-\subsubsection{Index-Erasure Semantics of Types}
-
-For each primitive type $\tyPrimNm{X} \in \mathit{PrimType}(s)$, for
-some sort $s$, and for the primitive types $\tyPrimNm{X} \in
-\mathit{PrimType}(\cdot)$, we assume that $\tyPrimNm{X}$ is assigned a
-set $\tyPrimSem{X}$. We extend this to assign a set to every
-well-formed type by induction on the type structure. To interpret the
-universal quantifier, we simply ignore the indexing:
-\begin{eqnarray*}
-  \tySem{\tyUnit} & = & \{*\} \\
-  \tySem{\tyPrim{X}{e}} & = & \tyPrimSem{X} \\
-  \tySem{\tyPrim{X}} & = & \tyPrimSem{X} \\
-  \tySem{A \tyProduct B} & = & \tySem{A} \times \tySem{B} \\
-  \tySem{A \tyArr B} & = & \tySem{A} \to \tySem{B} \\
-  \tySem{A + B} & = & \tySem{A} + \tySem{B} \\
-  \tySem{\forall i\mathord:s. A} & = & \tySem{A}
-\end{eqnarray*}
-
-Since our definition has completely ignored the indexing, and type
-equality is defined as an extension of index equality, we can easily
-see that equal types have equal denotations in our index-erasure
-semantics:
-\begin{lemma}\label{lem:tyeq-erasure}
-  If $\Delta \vdash A \equiv B : \sortType$ then $\tySem{A} =
-  \tySem{B}$.
-\end{lemma}
-
-\begin{proof}
-  By induction on the derivation of $\Delta \vdash A \equiv B :
-  \sortType$:
-  \begin{description}
-  \item[Case \TirName{TyEqUnit}] In this case, $A = B = \tyUnit$, so
-    $\tySem{A} = \tySem{B}$.
-  \item[Case \TirName{TyEqPrimIdx}] We have $A = \tyPrim{X}{e_1}$ and
-    $B = \tyPrim{X}{e_2}$, for some $e_1$ and $e_2$ of the same
-    sort. The erasure semantics ignores the index arguments, so
-    $\tySem{A} = \tySem{\tyPrim{X}{e_1}} = \tyPrimSem{X} =
-    \tySem{\tyPrim{X}{e_2}} = \tySem{B}$, as required.
-  \item[Case \TirName{TyEqPrimNoIdx}] We have $A = \tyPrimNm{X}$ and
-    $B = \tyPrimNm{X}$. So $\tySem{A} = \tyPrimSem{X} = \tySem{B}$, as
-    required.
-  \item[Case \TirName{TyEqArr}] Here, $A = A_1 \tyArr A_2$ and $B =
-    B_1 \tyArr B_2$, and by the induction hypothesis we know that
-    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
-    $\tySem{A} = \tySem{A_1 \tyArr A_2} = \tySem{A_1} \to \tySem{A_2}
-    = \tySem{B_1} \to \tySem{B_2} = \tySem{B_1 \tyArr B_2} =
-    \tySem{B}$.
-  \item[Case \TirName{TyEqProd}] In this case, $A = A_1 \tyProduct
-    A_2$ and $B = B_1 \tyProduct B_2$, and by the induction hypothesis
-    we know that $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} =
-    \tySem{B_2}$. Hence $\tySem{A} = \tySem{A_1 \tyProduct A_2} =
-    \tySem{A_1} \times \tySem{A_2} = \tySem{B_1} \times \tySem{B_2} =
-    \tySem{B_1 \tyProduct B_2} = \tySem{B}$.
-  \item[Case \TirName{TyEqSum}] In this case, $A = A_1 + A_2$ and $B =
-    B_1 + B_2$, and by the induction hypothesis we know that
-    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
-    $\tySem{A} = \tySem{A_1 + A_2} = \tySem{A_1} + \tySem{A_2} =
-    \tySem{B_1} + \tySem{B_2} = \tySem{B_1 + B_2} = \tySem{B}$.
-  \item[Case \TirName{TyEqForall}] We have $A = \forall
-    i\mathord:s.A'$ and $B = \forall i\mathord:s.B'$. By the induction
-    hypothesis we know that $\tySem{A'} = \tySem{B'}$, and hence
-    $\tySem{A} = \tySem{\forall i\mathord:s. A'} = \tySem{A'} =
-    \tySem{B'} = \tySem{\forall i\mathord:s. B'} = \tySem{B}$.
-  \end{description}
-\end{proof}
-
-Moreover, the index-erasure semantics is insensitive to index
-substitutions.
-\begin{lemma}\label{lem:tysubst-erasure}
-  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then $\tySem{\sigma^*A} = \tySem{A}$.
-\end{lemma}
-
-\begin{proof}
-  By induction on the derivation of $\Delta \vdash A : \sortType$.
-  \begin{description}
-  \item[Case \TirName{TyUnit}] Directly from the definition of
-    substitution on types of the form $\tyUnit$, we have
-    $\tySem{\sigma^*\tyUnit} = \tySem{\tyUnit}$, as required.
-  \item[Case \TirName{TyPrimIdx}] Since the index-erasure semantics
-    ignores all index expessions, we have
-    $\tySem{\sigma^*\tyPrim{X}{e}} = \tySem{\tyPrim{X}{\sigma^*e}} =
-    \tyPrimSem{X} = \tySem{\tyPrim{X}{e}}$.
-  \item[Case \TirName{TyPrimNoIdx}] Similar to the previous case.
-  \item[Case \TirName{TyArr}] We have $A = A_1 \tyArr A_2$, and by the
-    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
-    \tySem{A_1}$ and $\tySem{\sigma^*B_1} = \tySem{B_1}$. Thus, by the
-    definition of substitution on arrow types, we have
-    $\tySem{\sigma^*(A_1 \tyArr A_2)} = \tySem{\sigma^*A_1 \tyArr
-      \sigma^*A_2} = \tySem{\sigma^*A_1} \to \tySem{\sigma^*A_2} =
-    \tySem{A_1} \to \tySem{A_2} = \tySem{A_1 \tyArr A_2}$, as
-    required.
-  \item[Case \TirName{TyProd}] We have $A = A_1 \tyProduct A_2$, and
-    by the induction hypothesis, we know that $\tySem{\sigma^*A_1} =
-    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
-    definition of substitution on product types, we have
-    $\tySem{\sigma^*(A_1 \tyProduct A_2)} = \tySem{\sigma^*A_1
-      \tyProduct \sigma^*A_2} = \tySem{\sigma^*A_1} \times
-    \tySem{\sigma^*A_2} = \tySem{A_1} \times \tySem{A_2} = \tySem{A_1
-      \tyProduct A_2}$, as required.
-  \item[Case \TirName{TySum}] We have $A = A_1 + A_2$, and by the
-    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
-    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
-    definition of substitution on sum types, we have
-    $\tySem{\sigma^*(A_1 + A_2)} = \tySem{\sigma^*A_1 + \sigma^*A_2} =
-    \tySem{\sigma^*A_1} + \tySem{\sigma^*A_2} = \tySem{A_1} +
-    \tySem{A_2} = \tySem{A_1 + A_2}$, as required.
-  \item[Case \TirName{TyForall}] We have $A = \forall i\mathord:s.A'$,
-    and by the induction hypothesis we know that, for all $\sigma'$,
-    $\tySem{\sigma'^*A'} = \tySem{A'}$. Therefore, from the definition
-    of substitution for $\forall$-types, we have
-    $\tySem{\sigma^*\forall i\mathord:s.A'} = \tySem{\forall
-      i\mathord:s.\sigma_s^{*}A'} = \tySem{\sigma_s^{*}A'} = \tySem{A'}
-    = \tySem{\forall i\mathord:s.A'}$, where we have relied on the
-    fact that the index-erasure semantics of types ignores the index
-    quantification.
-  \end{description}
-\end{proof}
-
-For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
-index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
-structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
-\ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
-one-element set.
-
-\subsubsection{Index-Erasure Semantics of Terms}
-
-For a well-formed term $\Delta; \Gamma \vdash e : A$, we define a
-function $\tmSem{e} : \ctxtSem{\Gamma} \to \tySem{A}$, completely
-ignoring the indexing information. In light of
-\lemref{lem:tyeq-erasure} and \lemref{lem:tysubst-erasure}, we may
-define this function directly on the syntax of well-typed terms,
-rather than on the typing derivations. We define $\tmSem{e}$ by the
-following clauses:
-\begin{eqnarray*}
-  \tmSem{x}\eta & = & \eta_x \\
-  \tmSem{(M,N)}\eta & = & (\tmSem{M}\eta, \tmSem{N}\eta) \\
-  \tmSem{\pi_1M}\eta & = & \pi_1(\tmSem{M}\eta) \\
-  \tmSem{\pi_2M}\eta & = & \pi_2(\tmSem{M}\eta) \\
-  \tmSem{\mathrm{inl}\ M}\eta & = & \mathrm{inl}\ (\tmSem{M}\eta) \\
-  \tmSem{\mathrm{inr}\ M}\eta & = & \mathrm{inr}\ (\tmSem{M}\eta) \\
-  \left\llbracket
-    \begin{array}{l}
-      \textrm{case}\ M\ \textrm{of}\\
-      \textrm{inl}\ x.N_1;\\
-      \textrm{inr}\ y.N_2
-    \end{array}\right\rrbracket^{\mathit{tm}}\eta & = &
-  \left\{
-    \begin{array}{ll}
-      \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
-      \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
-    \end{array}
-  \right. \\
-  \tmSem{\lambda x. M}\eta & = & \lambda v. \tmSem{M}(\eta, v) \\
-  \tmSem{M N}\eta & = & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
-  \tmSem{\Lambda i.\ M}\eta & = & \tmSem{M}\eta \\
-  \tmSem{M[e]}\eta & = & \tmSem{M}\eta
-\end{eqnarray*}
-In the above, we have written $\eta_x$ to denote the appropriate
-projection from $\eta$ to get the position corresponding to the
-variable $x$ in the context.
-
-\subsection{Relational Semantics}
-
-\subsubsection{Relation Environments}
-
-For a set $X$, let $\Rel(X)$ be the set of binary relations on
-$X$. For an index context $\Delta$, a \emph{relation environment} is a
-family of functions $\rho = \{ \rho_{\tyPrimNm{X}} :
-\idxTms{\Delta}{s} \to \Rel(\tyPrimSem{X}) \}_{s \in \mathit{Sort},
-  \tyPrimNm{X} \in \mathit{PrimType}(s)}$.
-
-We define our relational semantics of types with respect to a given
-collection of relation environments: we assume that, for each index
-context $\Delta$, we are given a set of relation environments
-$\relEnv{E}(\Delta)$. This family must satisfy the following two
-properties:
-\begin{enumerate}
-\item Closure under re-indexing by index context morphisms: if $\rho
-  \in \relEnv{E}(\Delta)$ and $\sigma : \Delta' \Rightarrow \Delta$,
-  then $\rho \circ \sigma^* \in \relEnv{E}(\Delta')$.
-\item If we have a pair of relation environments $\rho_1 \in
-  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
-  \relEnv{E}(\Delta)$, along with an index context morphism $\sigma :
-  \Delta' \Rightarrow \Delta$, such that the outer edges of the
-  following diagram commute, for all primitive types $\tyPrimNm{X} \in
-  \mathit{PrimType}(s)$:
-  \begin{displaymath}
-    \xymatrix{
-      {\idxTms{\Delta}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar[d]_{\sigma^*}
-      &
-      {\idxTms{\Delta,i\mathord:s'}{s}} \ar[d]^{\sigma_{s'}^*} \ar@/^/[rdd]^{\rho_1}
-      \\
-      {\idxTms{\Delta'}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2}
-      &
-      {\idxTms{\Delta',i\mathord:s'}{s}} \ar@{.>}[dr]_\rho
-      \\
-      &
-      &
-      {\Rel(X)}
-    }
-  \end{displaymath}
-  Then there exists a relation environment $\rho \in
-  \relEnv{E}(\Delta,i\mathord:s')$ such that the two triangles in the
-  bottom right of the diagram commute.
-\end{enumerate}
-
-Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
-the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
-additional index variable $i\mathord:s$ to be the following:
-\begin{displaymath}
-  \extends{\rho}{i\mathord:s} = \{ \rho' \in \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s} = \rho \}
-\end{displaymath}
-
-\subsubsection{Relational Semantics of Types}
-
-%FIXME: re-word this
-A given relation context $\rho \in \relEnv{E}(\Delta)$ is extended to
-a relation interpretation of any type $\Delta \vdash A : \sortType$ by
-induction its derivation. So we define a relation
-$\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-formed
-type $A$.
-\begin{eqnarray*}
-  \rsem{\tyUnit}{\relEnv{E}}\rho & = & \{(*,*)\} \\
-  \rsem{\tyPrim{X}{e}}{\relEnv{E}}\rho & = & \rho_{\tyPrimNm{X}}(e) \\
-  \rsem{A \tyArr B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{A \tyProduct B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{A + B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & = & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i:s} \}
-\end{eqnarray*}
-In this definition we have made use of the following two constructions
-on binary relations. If $R \in \Rel(X)$ and $S \in \Rel(Y)$, then $R
-\relArrow S \in \Rel(X \to Y)$ is defined as $\{ (f_1,f_2) \sepbar
-\forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S \}$. With the same
-assumptions on $R$ and $S$, the relation $R \relTimes S \in \Rel(X
-\times Y)$ is defined as $\{ ((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2)
-\in R \land (b_1,b_2) \in S \}$. FIXME: do $\relSum$.
-
-We extend the relational semantics to contexts of well-formed types by
-the following clauses:
-\begin{eqnarray*}
-  \rsem{\epsilon}{\relEnv{E}}{\rho} & = & \id_{\ctxtSem{\epsilon}} \\
-  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & = & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
-\end{eqnarray*}
-
-\subsubsection{Parametricity}
-
-This relies on \lemref{lem:tyeq-erasure} to make sure that it is
-well-typed.
-
-\begin{lemma}\label{lem:tyeq-rel}
-  If $\Delta \vdash A \equiv B : \sortType$, then for all $\rho \in
-  \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}\rho = \rsem{B}{\relEnv{E}}\rho$.
-\end{lemma}
-
-\begin{proof}
-  By induction on the derivation of $\Delta \vdash A \equiv B :
-  \sortType$.
-  \begin{description}
-  \item[Case \TirName{TyEqUnit}] In this case $A = B = \tyUnit$, so
-    the statement is trivially statisfied.
-  \item[Case \TirName{TyEqPrimIdx}] In this case we have $A =
-    \tyPrim{X}{e_1}$ and $B = \tyPrim{X}{e_2}$, with $\Delta \vdash
-    e_1 \equiv e_2 : s$. The index expressions $e_1$ and $e_2$ are in
-    the same equivalence class, so by the definition of relation
-    environments we have that $\rho_{\tyPrimNm{X}}(e_1) =
-    \rho_{\tyPrimNm{X}}(e_2)$ and hence
-    $\rsem{\tyPrim{X}{e_1}}{\relEnv{E}}\rho =
-    \rsem{\tyPrim{X}{e_2}}{\relEnv{E}}\rho$.
-  \item[Case \TirName{TyEqPrimNoIdx}] This case is similar to the case
-    for \TirName{TyEqUnit}.
-  \item[Case \TirName{TyEqArr}] We have $A = A_1 \tyArr A_2$ and $B =
-    B_1 \tyArr B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$
-    and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the induction
-    hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_2}{\relEnv{E}}\rho$. Thus we may reason as follows:
-    $\rsem{A_1 \tyArr A_2}{\relEnv{E}}\rho =
-    \rsem{A_1}{\relEnv{E}}\rho \relArrow \rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho \relArrow \rsem{B_2}{\relEnv{E}}\rho =
-    \rsem{B_1 \tyArr B_2}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{TyEqProd}] We have $A = A_1 \tyProduct A_2$ and
-    $B = B_1 \tyProduct B_2$, with $\Delta \vdash A_1 \equiv B_1 :
-    \sortType$ and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the
-    induction hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
-    $\rsem{A_1 \tyProduct A_2}{\relEnv{E}}\rho =
-    \rsem{A_1}{\relEnv{E}}\rho \relTimes \rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho \relTimes \rsem{B_2}{\relEnv{E}}\rho =
-    \rsem{B_1 \tyProduct B_2}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{TyEqSum}] We have $A = A_1 + A_2$ and $B = B_1 +
-    B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$ and $\Delta
-    \vdash A_2 \equiv B_2 : \sortType$. By the induction hypothesis,
-    we know that $\rsem{A_1}{\relEnv{E}}\rho =
-    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
-    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
-    $\rsem{A_1 + A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}\rho
-    \relSum \rsem{A_2}{\relEnv{E}}\rho = \rsem{B_1}{\relEnv{E}}\rho
-    \relSum \rsem{B_2}{\relEnv{E}}\rho = \rsem{B_1 +
-      B_2}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{TyEqForall}] We have $A = \forall
-    i\mathord:s. A'$ and $B = \forall i\mathord:s.B'$, with $\Delta, i
-    : s \vdash A' \equiv B' : \sortType$. By the induction hypothesis,
-    we know that for all $\rho'$, $\rsem{A'}{\relEnv{E}}\rho' =
-    \rsem{B'}\rho'$. We reason as follows:
-    \begin{eqnarray*}
-      &    & (x_1,x_2) \in \rsem{\forall i\mathord:s. A'}{\relEnv{E}}\rho \\
-      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{A'}{\relEnv{E}}\rho' \\
-      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{B'}{\relEnv{E}}\rho' \\
-      &\iff& (x_1,x_2) \in \rsem{\forall i\mathord:s. B'}{\relEnv{E}}\rho
-    \end{eqnarray*}
-    Thus, by extensionality, we have $\rsem{\forall
-      i\mathord:s.A'}\rho \equiv \rsem{\forall
-      i\mathord:s.B'}{\relEnv{E}}\rho$, as required.
-  \end{description}
-\end{proof}
-
-This next lemma relies on \lemref{lem:tysubst-erasure} to make sure
-that it is well-typed.
-
-\begin{lemma}\label{lem:tysubst-rel}
-  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
-  $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Likewise,
-  if $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
-  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho = \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
-\end{lemma}
-
-\begin{proof}
-  The first part of the lemma statement is proved by induction on the
-  derivation of $\Delta \vdash A : \sortType$. We analyse each case in
-  turn:
-  \begin{description}
-  \item[Case \TirName{TyUnit}] Directly from the definition of
-    substitution on types of the form $\tyUnit$, and the fact that
-    $\rsem{\tyUnit}{\relEnv{E}}\rho_1 =
-    \rsem{\tyUnit}{\relEnv{E}}\rho_2$ for any pair $\rho_1,\rho_2$ of
-    relation environments, we have
-    $\rsem{\sigma^*\tyUnit}{\relEnv{E}}\rho =
-    \rsem{\tyUnit}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as required.
-  \item[Case \TirName{TyPrimIdx}] We have $A = \tyPrim{X}{e}$, therefore
-    $\rsem{\sigma^*A}{\relEnv{E}}\rho =
-    \rsem{\sigma^*\tyPrim{X}{e}}{\relEnv{E}}\rho =
-    \rsem{\tyPrim{X}{\sigma^*e}}{\relEnv{E}}\rho =
-    \rho_{\tyPrimNm{X}}(\sigma^*e) =
-    \rsem{\tyPrim{X}{e}}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as
-    required.
-  \item[Case \TirName{TyPrimNoIdx}] Similar to the case for
-    \TirName{TyUnit}.
-  \item[Case \TirName{TyArr}] In this case, $A = A_1 \tyArr A_2$ and
-    by the induction hypothesis, we know that $\rsem{\sigma^*A_1}{\relEnv{E}}\rho
-    = \rsem{A_1}{\relEnv{E}}{(\rho \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we reason as
-    follows: $\rsem{\sigma^*(A_1 \tyArr A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1
-      \tyArr \sigma^*A_2}{\relEnv{E}}\rho = \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relArrow
-    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho \circ \sigma^*)}
-    \relArrow \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)} = \rsem{A_1 \tyArr
-      A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$, as required.
-  \item[Case \TirName{TyProd}] In this case, $A = A_1 \tyProduct A_2$
-    and by the induction hypothesis, we know that
-    $\rsem{\sigma^*A_1}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we
-    reason as follows: $\rsem{\sigma^*(A_1 \tyProduct
-      A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1 \tyProduct
-      \sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relTimes
-    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)} \relTimes \rsem{A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)} = \rsem{A_1 \tyProduct A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)}$, as required.
-  \item[Case \TirName{TySum}] In this case, $A = A_1 + A_2$
-    and by the induction hypothesis, we know that
-    $\rsem{\sigma^*A_1}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)}$ and $\rsem{\sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{A_2}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Accordingly, we
-    reason as follows: $\rsem{\sigma^*(A_1 +
-      A_2)}{\relEnv{E}}\rho = \rsem{\sigma^*A_1 +
-      \sigma^*A_2}{\relEnv{E}}\rho =
-    \rsem{\sigma^*A_1}{\relEnv{E}}\rho \relSum
-    \rsem{\sigma^*A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}{(\rho
-      \circ \sigma^*)} \relSum \rsem{A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)} = \rsem{A_1 + A_2}{\relEnv{E}}{(\rho \circ
-      \sigma^*)}$, as required.
-  \item[Case \TirName{TyForall}] We have $A = \forall
-    i\mathord:s.A'$. We prove the required equation by demonstrating
-    two inclusions. From left-to-right, we know that
-    \begin{equation}
-      \label{eq:tysubst-rel-forall1}
-      (x_1,x_2) \in \rsem{\sigma^*\forall i\mathord:s. A'}{\relEnv{E}}\rho
-    \end{equation}
-    and we must show that $(x_1, x_2) \in \rsem{\forall
-      i\mathord:s. A'}{\relEnv{E}}{(\rho \circ \sigma^*)}$. From
-    (\ref{eq:tysubst-rel-forall1}) and the definition of substitution,
-    we know that
-    \begin{displaymath}
-      \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{\sigma_s^*A'}{\relEnv{E}}\rho'
-    \end{displaymath}
-    By the induction hypothesis, we can deduce that
-    \begin{equation}
-      \label{eq:tysubst-rel-forall2}
-      \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{A'}{\relEnv{E}}{(\rho' \circ \sigma_s^{*})}
-    \end{equation}
-    To show the required inclusion, we must show that for all $\rho_1
-    \in \extends{\rho \circ \sigma^*}{i:s}$, it is the case that
-    $(x_1,x_2) \in \rsem{A'}{\relEnv{E}}{\rho_1}$. We now make use of
-    the pushout property of our family $\relEnv{E}$ of relation
-    environments: we have a $\rho \in \relEnv{E}(\Delta)$ and $\rho_1
-    \in \relEnv{E}(\Delta, i:s)$ such that $\rho_1 \circ \pi_{i:s} =
-    \rho \circ \sigma^*$ (by the assumption that $\rho_1 \in
-    \extends{\rho \circ \sigma^*}{i:s}$), so we can obtain a $\rho'
-    \in \relEnv{E}(\Delta',i:s)$ such that $\rho' \circ \pi_{i:s} =
-    \rho$ (so $\rho' \in \extends{\rho}{i:s}$), and $\rho' \circ
-    \sigma^* = \rho_1$. Therefore we use
-    (\ref{eq:tysubst-rel-forall2}) to deduce that $(x_1,x_2) \in
-    \rsem{A'}{\relEnv{E}}{\rho_1}$, and the inclusion from
-    left-to-right is shown.
-
-    From right-to-left, we must now show
-    (\ref{eq:tysubst-rel-forall2}) under the assumption that
-    \begin{equation}
-      \label{eq:tysubst-rel-forall3}
-      \forall \rho_1 \in \extends{\rho \circ \sigma^*}{i:s}. (x_1,x_2) \in \rsem{A'}{\relEnv{E}}{\rho_1}.
-    \end{equation}
-    Given $\rho' \in \extends{\rho}{i:s}$, we let $\rho_1 = \rho'
-    \circ \sigma^*$. Now, $\rho_1 \circ \pi_{i:s} = \rho' \circ
-    \sigma^* \circ \pi_{i:s} = \rho' \circ \pi_{i:s} \circ \sigma_s^{*}
-    = \rho \circ \sigma_s^{*}$. Thus $\rho_1 \in \extends{\rho \circ
-      \sigma^*}{i:s}$ and we may use (\ref{eq:tysubst-rel-forall3}) to
-    deduce that $(x_1,x_2) \in \rsem{A'}{\relEnv{E}}{(\rho' \circ \sigma_s^{*})}$,
-    as required.
-  \end{description}
-  The second part of the lemma statement follows directly by induction
-  on the derivation of $\Delta \vdash \Gamma \isCtxt$ and the first
-  part. The step case of the induction is very similar to the
-  \TirName{TyProd} case above.
-\end{proof}
-
-\begin{theorem}[Parametricity]
-  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
-  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
-  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
-  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
-\end{theorem}
-
-\begin{proof}
-  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
-  three interesting cases follow:
-  \begin{description}
-  \item[Case \TirName{TyEq}] This case follows directly from
-    \lemref{lem:tyeq-rel}.
-  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
-    $A = \forall i\mathord:s. A'$. We know from the induction
-    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
-    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}\rho'$,
-    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho'$. We are given $\rho \in
-    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
-    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
-    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
-    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
-    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
-      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
-    \rsem{\pi_{i:s}^*\Gamma}\rho''$ by \lemref{lem:tysubst-rel}. So we
-    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}\rho''$, and thus
-    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
-      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
-      i.M'}\eta_2) \in \rsem{\forall
-      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
-    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
-    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
-    know that for all $\rho' \in \extends{\rho}{i:s}$,
-    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho'$. If we let $\rho' = \rho \circ
-    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
-    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
-    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
-    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
-    by \lemref{lem:tysubst-rel}, conclude that $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
-    as required.
-  \end{description}
-  The remaining cases are standard for proofs of the fundamental lemma
-  of logical relations for simply-typed systems.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Consequences of Parametricity}
-
-We now give instances of our semantic framework for each of the three
-examples from \autoref{sec:example-systems}. For each of the three
-examples, we defined an indexing theory, a collection of primitive
-types, and a collection of primitive operations $\Gamma_X$. Below, we
-assign the ``obvious'' index-erasure semantics to each of the
-primitive types and operations: so we have $\tyPrimSem{X}$ for each
-primitive type $\tyPrimNm{X}$, and a $\eta_X \in \ctxtSem{\Gamma_X}$.
-
-The key question we wish to answer below is whether there is a family
-of sets of relation environments $\{\relEnv{E}_X(\Delta)\}_{\Delta}$
-that satisfies the following completeness property identified by
-Kennedy.
-\begin{equation}\label{eq:kennedy-completeness}
-  (\forall \Delta. \forall \rho \in \relEnv{E}'(\Delta). (\eta_X, \eta_X) \in \rsem{\Gamma_X}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_X
-\end{equation}
-where the subset relation on families of sets of relation environments
-is defined as $\relEnv{E}_1 \subseteq \relEnv{E}_2$ if $\forall
-\Delta. \relEnv{E}_1(\Delta) \subseteq \relEnv{E}_2(\Delta)$.
-
-Such a completeness result allows us to derive inexpressibility
-results for a given instance of our general framework. Kennedy
-demonstrates this for the units of measure system by showing that a
-putative square root function of type $\forall u.\ \tyPrim{num}{u^2}
-\to \tyPrim{num}{u}$ can only ever be the constantly zero function (or
-non-terminating).
-
-\subsection{Units of Measure}\label{sec:units-semantics}
-
-For the units of measure example, we set $\tyPrimSem{num} =
-\tyPrimSem{rational} = \mathbb{Q}$, the set of rational numbers. We
-define $\eta_{\mathit{units}} \in \Gamma_{\mathit{units}}$ are defined
-to be the standard arithmetical and comparison operations on rational
-numbers, and $\mathit{toRat}$ and $\mathit{fromRat}$ are defined to be
-the identity functions. The division function (of type $/ : \forall
-u_1,u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to
-\tyPrim{num}{u_1u^{-1}_2} + 1$) returns $\mathrm{inr}(*)$ only when
-its second argument is zero.
-
-Kennedy defines the following family of sets of relational
-environments. Note that, since we have assumed the abelian group
-axioms, the set $\idxTms{\Delta}{\mathsf{unit}}$ of indexing
-expressions quotiented by the equational theory, is an abelian group.
-\begin{displaymath}
-  \relEnv{E}_{\mathit{units}}(\Delta) = \left\{ \rho^{G,h} \sepbar
-    \begin{array}{l}
-      G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{unit}} \\
-      h : G \to \mathbb{Q}^{>0} \\
-      h\textrm{ is a group homomorphism}
-    \end{array}
-  \right\}
-\end{displaymath}
-where $\mathbb{Q}^{>0} = \{ q \in \mathbb{Q} \sepbar q > 0 \}$ and
-\begin{displaymath}
-  \rho^{G,h}_{\tyPrimNm{num}}(e) = \left\{
-    \begin{array}{ll}
-      \{(0,0)\} & \textrm{if }e \not\in G \\
-      \{(r,h(e)r) \sepbar r \in \mathbb{Q} \} & \textrm{if }u \in G
-    \end{array}
-  \right.
-\end{displaymath}
-
-\begin{theorem}
-  The family of sets of relational environments
-  $\relEnv{E}_{\mathit{units}}$ satisfies the completeness property
-  \hyperref[eq:kennedy-completeness]{(\ref*{eq:kennedy-completeness})}
-  above. That is, for all $\relEnv{E}'$:
-  \begin{displaymath}
-    (\forall \Delta.\forall \rho \in \relEnv{E}'(\Delta).
-(\eta_{\mathit{units}}, \eta_{\mathit{units}}) \in \rsem{\Gamma_{\mathit{units}}}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_{\mathit{units}}
-  \end{displaymath}
-\end{theorem}
-
-\begin{proof}
-  (Following Kennedy).  From right to left, we have to check that for
-  each $\rho \in \relEnv{E}'(\Delta)$, where $\relEnv{E}' \subseteq
-  \relEnv{E}_{\mathit{units}}$, that each of the semantic
-  interpretations of the operations satisfies
-  $(\eta_{\mathit{units}}(\mathit{op}),
-  \eta_{\mathit{units}}(\mathit{op})) \in
-  \rsem{\Gamma(\mathit{op})}{\relEnv{E}'}\rho$. For instance,
-  multiplication has to satisfy: for all subgroups $G \subseteq
-  \idxTms{\Delta,u_1,u_2}{\mathsf{unit}}$ and all group homomorphisms
-  $h : G \to \mathbb{Q}^{>0}$, then $(q_1,q'_1) \in \rho^{G,h}(u_1)$
-  and $(q_2,q'_2) \in \rho^{G,h}(u_2)$ implies that $(q_1q_2,q'_1q'_2)
-  \in \rho^{G,h}(u_1u_2)$. There are four cases, depending on whether
-  $u_1 \in G$ and whether $u_2 \in G$. If $u_1 \not\in G$, then $q_1 =
-  q'_1 = 0$, so $q_1q_2 = 0$ and $q'_1q'_2 = 0$. Since $(0,0) \in
-  \rho^{G,h}(u')$ for any $u'$, $G$ and $h$, we are done. If $u_1 \in
-  G$ and $u_2 \in G$, then we need to show that $(q_1q_2,
-  h(u_1)q_1h(u_2)q_2) \in \rho^{G,h}(u_1u_2)$. By $h$ being a
-  homomorphism, we have $(q_1q_2, h(u_1)q_1h(u_2)q_2) = (q_1q_1,
-  h(u_1u_2)q_1q_2)$, which is in $\rho^{G,h}(u_1u_2)$. The cases for
-  the other operations are similar.
-
-  From left to right, we have the following properties of any
-  relational environment $\rho_{\tyPrimNm{num}} \in
-  \relEnv{E}'(\Delta)$, derived from the hypothesis that all the
-  operations in $\eta_{\mathit{units}}$ satisfy their types'
-  relational semantics.
-  \begin{enumerate}
-  \item\label{item:obsv-poly} By the existence of the function
-    $\mathit{toRat}$, the relation $\rho_{\tyPrimNm{num}}(1)$ is the
-    identity relation.
-  \item\label{item:zero-poly} For any unit expression $e$, $(0,0) \in
-    \rho_{\tyPrimNm{num}}(e)$.
-  \item \label{item:one-poly} $(1,1) \in \rho_{\tyPrimNm{num}}(1)$.
-  \item \label{item:mult-poly} For any unit expressions $e_1,e_2$,
-    such that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e_1)$ and
-    $(q_2,q'_2) \in \rho_{\tyPrimNm{num}}(e_2)$, we have $(q_1q_2,
-    q'_1q'_2) \in \rho_{\tyPrimNm{num}}(e_1e_2)$.
-  \item\label{item:div-poly} For any unit expressions $e_1,e_2$, such
-    that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e_1)$ and $(q_2,q'_2)
-    \in \rho_{\tyPrimNm{num}}(e_2)$, we have that either $q_2 = q'_2 =
-    0$, or we have that $(q_1/q_2,q'_1,q'_2) \in
-    \rho_{\tyPrimNm{num}}(e_1e_2^{-1})$.
-  \item\label{item:compare-poly} For any unit expression $e$, such
-    that $(q_1,q'_1) \in \rho_{\tyPrimNm{num}}(e)$ and $(q_2,q'_2) \in
-    \rho_{\tyPrimNm{num}}(e)$, we have $q_1 < q_2$ iff $q'_1 < q'_2$.
-  \end{enumerate}
-  We now show that if $(q,q') \in \rho_{\tyPrimNm{num}}(e)$ for some
-  unit expression $e$, then either $q = q' = 0$, or $q' = rq$ for some
-  $r \in \mathbb{Q}^{>0}$, determined by $e$. We know from
-  \ref{item:zero-poly} above that $q = q' = 0$ is a possibility. Now
-  consider what happens if $q \not= 0$: in this case, by
-  \ref{item:zero-poly} and \ref{item:compare-poly}, we know that $0 <
-  q$ iff $0 < q'$, so $q' \not= 0$, and $q'$ has the same sign as
-  $q$. So $q/q'$ exists and is in $\mathbb{Q}^{>0}$. Hence $q = rq'$
-  for $r = q/q'$. So now we know that if $(q,q') \in
-  \rho_{\tyPrimNm{num}}(e)$, and $q \not= 0$ or $q' \not= 0$ then
-  there exists an $r$ such that $q = rq'$.
-
-  We now show that the unit expression $e$ determines the value of
-  $r$. Assume that we have $(q_1,r_1q_1) \in \rho_{\tyPrimNm{num}}(e)$
-  and $(q_2,r_2q_2) \in \rho_{\tyPrimNm{num}}(e)$. By
-  \ref{item:div-poly}, we know that $(q_1/q_2, (r_1q_1)/(r_2q_2)) \in
-  \rho_{\tyPrimNm{num}}(1)$. By \ref{item:obsv-poly},
-  $\rho_{\tyPrimNm{num}}(1)$ is the equality relation, so we can
-  deduce that $r_1 = r_2$.
-
-  Now let $G$ be the subgroup of $\idxTms{\Delta}{\mathsf{unit}}$ such
-  that $e \in G$ implies that $\rho_{\tyPrimNm{num}}(e) \not=
-  \{(0,0)\}$. By the argument above, for each $e \in G$, there is a
-  unique $r \in \mathbb{Q}^{>0}$ such that $\rho_{\tyPrimNm{num}}(e) =
-  \{ (q,rq) \sepbar q \in \mathbb{Q} \}$. Thus there is a function $h
-  : G \to \mathbb{Q}^{>0}$. By \ref{item:one-poly} and
-  \ref{item:mult-poly}, $h$ is a group homomorphism. Thus we have
-  shown that $\rho_{\tyPrimNm{num}}$ has the form of
-  $\rho^{G,h}_{\tyPrimNm{num}}$ for some $G$ and $h$. Since
-  $\rho_{\tyPrimNm{num}}$ was an arbitrary element of
-  $\relEnv{E}'(\Delta)$, and $\Delta$ was arbitrary, we have shown
-  that $\relEnv{E}' \subseteq \relEnv{E}_{\mathit{units}}$.
-\end{proof}
-
-Some notes towards a generalisation:
-\begin{itemize}
-\item By inspecting Kennedy's proof, we can see that the operations of
-  addition and subtraction do not play any role, save for an
-  obligation to prove that they satisfy the relational semantics of
-  their types. This follows just by distributivity of addition over
-  multiplication.
-\item Kennedy's proof does not rely on any particular features of
-  $\mathbb{Q}$ other than the fact that it is an ordered field. By
-  inspecting the proof, it is possible to see that if we remove the
-  comparison operation $< : \forall u. \tyPrim{num}{u} \to
-  \tyPrim{num}{u} \to \tyPrimNm{bool}$ from $\Gamma_{\mathit{units}}$,
-  then the appropriate family of sets of relational environments uses
-  homomorphisms into the set of non-zero rationals:
-  \begin{displaymath}
-    \begin{array}{l}
-      \relEnv{E}_{\mathit{units}\backslash\{<\}}(\Delta) = \\
-      \quad\left\{ \rho^{G,h} \sepbar
-        \begin{array}{l}
-          G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{unit}} \\
-          h : G \to \mathbb{Q}^{\not=0} \\
-          h\textrm{ is a group homomorphism}
-        \end{array}
-      \right\}
-    \end{array}
-  \end{displaymath}
-  To complete the step in the proof where the comparison operation was
-  used, we use the division operation to show that if $(q,q') \in
-  \rho_{\tyPrimNm{num}}(e)$ for some $e$, and $q \not= 0$, then $q'
-  \not= 0$ (but they need not necessarily have the same sign).
-
-  Just as Kennedy's proof generalises to an arbitrary ordered field,
-  if we remove the comparison operation we can generalise to an
-  arbitrary field as the underlying semantics of $\tyPrimNm{num}$ and
-  $\tyPrimNm{rational}$.
-\end{itemize}
-
-\subsection{Euclidean Plane Geometry}\label{sec:2d-semantics}
-
-We set $\tyPrimSem{vec} = \mathbb{R}^2$, and $\tyPrimSem{real} =
-\mathbb{R}$. The semantic interpretation $\eta_{\mathit{2D}}$ of the
-operations in $\Gamma_{\mathit{2D}}$ consists of the standard vector
-zero, addition, negation and scaling in $\mathbb{R}^2$, along with
-projections to interpret $\mathit{getX}$ and $\mathit{getY}$, and
-pairing to interpret $\mathit{vec}$.
-
-Let
-\begin{displaymath}
-  \begin{array}{l}
-    \relEnv{E}_{\mathit{2D}}(\Delta) = \\
-    \quad\left\{ \rho^{G,h} \sepbar
-      \begin{array}{l}
-        G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{translation}} \\
-        h : G \to \mathbb{R}^2 \\
-        h\textrm{ is a group homomorphism}
-      \end{array}
-    \right\}
-  \end{array}
-\end{displaymath}
-where
-\begin{displaymath}
-  \rho^{G,h}_{\tyPrimNm{vec}}(t) = \left\{
-    \begin{array}{ll}
-      \{\} & \textrm{if }t \not\in G \\
-      \{(v,h(t) + v) \sepbar v \in \mathbb{R}^2 \} & \textrm{if }t \in G
-    \end{array}
-  \right.
-\end{displaymath}
-Note that, unlike the units of measure example above, the pair $(0,0)$
-need not appear in every relation. This is because we do not have a
-translation polymorphic constant like $0 : \forall u. \tyPrim{num}{u}$
-in the units of measure example.
-
-\begin{theorem}
-  The family of sets of relational environments
-  $\relEnv{E}_{\mathit{units}}$ satisfies the completeness property
-  \hyperref[eq:kennedy-completeness]{(\ref*{eq:kennedy-completeness})}
-  above. That is, for all $\relEnv{E}'$:
-  \begin{displaymath}
-    (\forall \Delta.\forall \rho \in \relEnv{E}'(\Delta).
-(\eta_{\mathit{2D}}, \eta_{\mathit{2D}}) \in \rsem{\Gamma_{\mathit{2D}}}{\relEnv{E}'}{\rho}) \Leftrightarrow \relEnv{E}'\subseteq \relEnv{E}_{\mathit{2D}}
-  \end{displaymath}
-\end{theorem}
-
-\begin{proof}
-  From right to left, we must check that each operation satisfies the
-  relational interpretation of its type. This is very similar to the
-  case for the units of measure example above.
-
-  From left to right, we have the following properties of any
-  relational environment $\rho_{\tyPrimNm{vec}} \in
-  \relEnv{E}'(\Delta)$, derived from the hypothesis that all the
-  operations in $\eta_{\mathit{2D}}$ satisfy their types'
-  relational semantics.
-  \begin{enumerate}
-  \item\label{item:obsv-vec-poly} By the existence of the functions
-    $\mathit{getX}$ and $\mathit{getY}$, the relation
-    $\rho_{\tyPrimNm{vec}}(0)$ is the identity relation.
-  \item \label{item:zero-vec-poly} $(0,0) \in \rho_{\tyPrimNm{vec}}(0)$.
-  \item \label{item:add-vec-poly} For any translation expressions $t_1,t_2$,
-    such that $(v_1,v'_1) \in \rho_{\tyPrimNm{vec}}(t_1)$ and
-    $(v_2,v'_2) \in \rho_{\tyPrimNm{vec}}(t_2)$, we have $(v_1 + v_2,
-    v'_1 + v'_2) \in \rho_{\tyPrimNm{vec}}(t_1 + t_2)$.
-  \item\label{item:neg-vec-poly} For any translation expression $t$,
-    such that $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$, we have that
-    $(-v,-v') \in \rho_{\tyPrimNm{vec}}(-t)$.
-  \end{enumerate}
-  We now show that if $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$ for some
-  translation expression $t$, then $v' = w + v$ for some $w \in
-  \mathbb{R}^2$, where $w$ is determined by $t$. If we are given
-  $(v,v') \in \rho_{\tyPrimNm{vec}}(t)$, then we can set $w = v' -
-  v$. This is uniquely determined by $t$: if we have $(v, w + v) \in
-  \rho_{\tyPrimNm{vec}}(t)$ and $(v, w' + v) \in
-  \rho_{\tyPrimNm{vec}}(t)$, then by \ref{item:add-vec-poly} and
-  \ref{item:neg-vec-poly}, we have $(v - v, (w + v) - (w' + v)) \in
-  \rho_{\tyPrimNm{vec}}(0)$. By \ref{item:obsv-vec-poly}, we know that
-  $\rho_{\tyPrimNm{vec}}(0)$ is the identity relation, so we can
-  deduce that $w = w'$.
-
-  Now define $G \subseteq \idxTms{\Delta}{\mathsf{translation}}$ to
-  have $t \in G$ if $\rho_{\tyPrimNm{vec}}(t) \not= \{\}$. Since the
-  translation factor $w$ is determined by $t$, we can define a
-  function $h : G \to \mathbb{R}^2$. By \ref{item:zero-vec-poly} and
-  \ref{item:add-vec-poly}, $h$ is a group homomorphism, and $(v,v')
-  \in \rho_{\tyPrimNm{vec}}$ iff $v = h(t) + v'$. Thus
-  $\rho_{\tyPrimNm{vec}}$ is of the form
-  $\rho^{G,h}_{\tyPrimNm{vec}}$, as required.
-\end{proof}
-
-Some more notes:
-\begin{enumerate}
-\item The proof is slightly simpler than the case for the units of
-  measure example above because the underlying semantics of
-  $\tyPrimNm{vec}$ forms an abelian group, so we can just use inverses
-  to determine the difference factor $w$, and we don't have to deal
-  with the possibility of division by zero.
-\item The proof above should generalise to any abelian group.
-\item Generalising to non-abelian groups does not look
-  straightforward: the step that shows that the translation factor $w$
-  is uniquely determined by the translation expression $t$ fails. We
-  end up with
-  \begin{displaymath}
-    0 = w + v + (-w') + (-v)
-  \end{displaymath}
-  from which it is not possible to conclude that $w = w'$. (Maybe
-  something using normal subgroups would work?)
-\item A slightly different example is obtained by considering
-  rotations rather than translations (we consider rotations as an
-  additive abelian group). In this case, we could have the following
-  typed operations:
-  \begin{eqnarray*}
-    0 & : & \forall r.\ \tyPrim{vec}{r} \\
-    + & : & \forall r.\ \tyPrim{vec}{r} \to \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
-    - & : & \forall r.\ \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
-    * & : & \forall r.\ \tyPrimNm{real} \to \tyPrim{vec}{r} \to \tyPrim{vec}{r} \\
-    \mathit{norm} & : & \forall r. \tyPrim{vec}{r} \to \tyPrimNm{real} \\
-    \mathit{angleOf} & : & \forall r. \tyPrim{vec}{r} \to \tyPrim{rot}{r} + 1 \\
-    \mathit{rotateBy} & : & \forall r_1 r_2. \tyPrim{vec}{r_1} \to \tyPrim{rot}{r_2} \to \tyPrim{vec}{r_1 + r_2} \\
-    0_{\mathit{rot}} & : & \tyPrim{rot}{0} \\
-    +_{\mathit{rot}} & : & \forall r_1 r_2. \tyPrim{rot}{r_1} \to \tyPrim{rot}{r_2} \to \tyPrim{rot}{r_1 + r_2} \\
-    -_{\mathit{rot}} & : & \forall r. \tyPrim{rot}{r} \to \tyPrim{rot}{-r}
-  \end{eqnarray*}
-  where $\tyPrimSem{vec} = \mathbb{R}^2$, as before, and
-  $\tyPrimSem{rot} = \mathrm{SO}(2)$, the group of rotations of the
-  plane. The operation $\mathit{angleOf}$ is intended to return
-  $\mathrm{inr}(*)$ when the provided vector is zero.
-
-  In this case, rotations form an abelian group, so we can re-use the
-  proof above to give a family of sets of relational environments that
-  is complete for the group operations on rotations. The fact that
-  rotations are linear admits the typing of the first four
-  operations. The existence of the $\mathit{angleOf}$ and
-  $\mathit{rotateBy}$ functions allow us to determine ``rotation
-  factors'' between arbitrary pairs of vectors.
-
-  I conjecture that the following definition of
-  $\relEnv{E}_{\mathit{2Drot}}$ satisfies the completeness criterion,
-  with the standard interpretation of the operations above.
-  \begin{displaymath}
-    \begin{array}{l}
-      \relEnv{E}_{\mathit{2Drot}}(\Delta) = \\
-      \quad\left\{ \rho^{G,h} \sepbar
-        \begin{array}{l}
-          G\textrm{ is a subgroup of }\idxTms{\Delta}{\mathsf{rotation}} \\
-          h : G \to \mathrm{SO}(2) \\
-          h\textrm{ is a group homomorphism}
-        \end{array}
-      \right\}
-    \end{array}
-  \end{displaymath}
-  where
-  \begin{displaymath}
-    \rho^{G,h}_{\tyPrimNm{vec}}(r) = \left\{
-      \begin{array}{ll}
-        \{(0,0)\} & \textrm{if }t \not\in G \\
-        \{(v,h(r) v) \sepbar v \in \mathbb{R}^2 \} & \textrm{if }t \in G
-      \end{array}
-    \right.
-  \end{displaymath}
-  and
-  \begin{displaymath}
-    \rho^{G,h}_{\tyPrimNm{rot}}(r) = \left\{
-      \begin{array}{ll}
-        \{\} & \textrm{if }t \not\in G \\
-        \{(s,h(r) s) \sepbar s \in \mathrm{SO}(2) \} & \textrm{if }t \in G
-      \end{array}
-    \right.
-  \end{displaymath}
-\end{enumerate}
-
-\subsection{Security Typing}\label{sec:sec-semantics}
-
-In this case, we let $\tyPrimSem{T} = \{*\}$, and the operations
-$\mathit{proj}$ and $\mathit{combine}$ are given the only possible
-interpretations.
-
-I conjecture that the follow definition works:
-\begin{displaymath}
-  \begin{array}{l}
-    \relEnv{E}_{\mathit{sec}}(\Delta) = \\
-    \quad\left\{ \rho^{h} \sepbar
-      \begin{array}{l}
-        h : \idxTms{\Delta}{\mathsf{principal}} \to \{\top,\bot\} \\
-        h\textrm{ is a boolean algebra homomorphism}
-      \end{array}
-    \right\}
-  \end{array}
-\end{displaymath}
-where
-\begin{displaymath}
-  \rho^{h}_{\tyPrimNm{T}}(p) = \left\{
-    \begin{array}{ll}
-      \{\} & \textrm{if }h(p) = \bot \\
-      \{(*,*) \} & \textrm{if }h(p) = \top
-    \end{array}
-  \right.
-\end{displaymath}
-
-One can compare this definition to the previous ones by considering
-the boolean algebra homomorphism as defining a sub-boolean algebra of
-$\idxTms{\Delta}{\mathsf{principal}}$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Extensions}
-
-\subsection{Inductive Types}
-
-\subsection{Type Operators}
-
-\subsection{Size indexing and Combinatorial Species}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
+\input{section-introduction.tex}
+\input{section-motivating-examples.tex}
+\input{section-background.tex}
+\input{section-general-framework.tex}
+\input{section-instantiations.tex}
+\input{section-discussion.tex}
 
@@
 
+\appendix
+
+\input{appendix-proofs.tex}
+
 \end{document}

### 2012-05-22 paper/section-background.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

\section{Background and Related Work}
\label{sec:background}

### 2012-05-22 paper/section-general-framework.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

\section{A General Framework}
\label{sec:a-general-framework}

We now develop a general framework for discussing
algebraically-indexed types and their relational interpretations. We
have chosen a particular order for our presentation: presenting the
semantics of types before we present the syntax of programs and their
typing relation. We have deliberately selected this order to emphasise
the importance of establishing what types \emph{mean} over the
language used to write programs.

We first present, in \autoref{sec:algebraically-indexed-types}, the
syntax of the language of algebraically-indexed types. Our general
framework is parameterised by a choice of multi-sorted algebraic
theory to be used for indexing types, and the collections of primitive
types and
operations. %FIXME: refer forwards to the names of the parameters
We define substitution of index terms, and what it means for two types
to be equal. Equality on types is generated from the chosen
multi-sorted algebraic theory.

We next present the semantics of types in
\autoref{sec:semantics-algebraically-indexed-types}. Types are first
interpreted without reference to the indexing terms: an
\emph{index-erasure} semantics. The index-erasure semantics provides
FIXME. Relational semantics...

Throughout this section, we use Kennedy's original units of measure
system as a running example in order to provide a concrete example of
our general framework, and also to demonstrate that we have in fact
generalised Kennedy's original development.

\subsection{Algebraically-Indexed Types}
\label{sec:algebraically-indexed-types}

The indexing terms and types of our system are derived from the
following data:
\begin{enumerate}
\item A collection $\SortSet$ of index sorts. We will use the
  meta-syntactic variables $s,s_1,s_2,...$ and so on to stand for
  arbitrary sorts taken from $\SortSet$.
\item A collection $\IndexOpSet$ of index operations, with an
  accompanying function $\indexOpArity : \IndexOpSet \to \SortSet^*
  \times \SortSet$ (we use the notation $A^*$ to denote the set of
  lists of elements of some set $A$). 
\item A collection $\PrimTypeSet$ of primitive types, with an
  accompanying function $\primTyArity : \PrimTypeSet \to
  \SortSet^*$.
\end{enumerate}

\begin{example}[Units of Measure]
  To instantiate our general framework as Kennedy's original units
  of measure system, we would take the following settings. For the
  collection of sorts we take $\SortSet = \{ \mathsf{unit} \}$. For
  the collection of index operations, we take $\mathit{IndexOp} = \{1,
  -\cdot-, -^{-1}\}$, with:
  \begin{eqnarray*}
    \indexOpArity{(1)} &=& ([],\mathsf{unit}) \\
    \indexOpArity{(-\cdot-)} & = & ([\mathsf{unit},\mathsf{unit}],\mathsf{unit}) \\
    \indexOpArity{(-^{-1})} & = & ([\mathsf{unit}],\mathsf{unit})
  \end{eqnarray*}
  The intended interpretation of these operations is group unit, group
  multiplication and group inverse, respectively. When we discuss
  equational theories for index expressions in
  \autoref{sec:type-equality} below, we will impose the abelian group
  laws on these operations.

  The primitive types for the units of measure example are taken to be
  $\PrimTypeSet = \{ \mathsf{num} \}$, with
  $\primTyArity(\mathsf{num}) = [\mathsf{unit}]$.
\end{example}

We also assume a countably infinite collection of index expression
variables. We will use meta-syntactic variables $i, i_1, i_2, ...$ and
so on to stand for index expression variables. \emph{Index contexts}
$\Delta = i_1 : s_1, ..., i_n : s_n$ are built from the index
expression variable/sort pairs, such that all the index variable names
are distinct.

\begin{figure*}[t]
  \centering
  \textbf{Index contexts}\\

  $\Delta = i_1 : s_1, ..., i_n : s_n$, where each $s_i \in \SortSet$ and no variable name is repeated.

  \bigskip

  \textbf{Well-sorted index expressions}
  \begin{mathpar}
    \inferrule* [right=IVar]
    {i : s \in \Delta}
    {\Delta \vdash i : s}

    \inferrule* [right=IOp]
    {\indexOp{f} \in \mathit{IndexOp} \\
      \indexOpArity(\indexOp{f}) = ([s_1,...,s_n], s) \\
      \{\Delta \vdash e_j : s_j\}_{1 \leq j \leq n}}
    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) : s}
  \end{mathpar}

  \bigskip

  \textbf{Well-indexed types}
  \begin{mathpar}
    \inferrule* [right=TyPrim]
    {\tyPrimNm{X} \in \mathit{PrimType} \\\\
      \primTyArity(\tyPrimNm{X}) = [s_1,...,s_n] \\
      \{\Delta \vdash e_j : s_j\}_{1\leq j \leq n}}
    {\Delta \vdash \tyPrim{X}{e_1,...,e_n} \isType}

    \inferrule* [right=TyUnit]
    { }
    {\Delta \vdash \tyUnit \isType}

    \inferrule* [right=TyArr]
    {\Delta \vdash A \isType \\ \Delta \vdash B \isType}
    {\Delta \vdash A \tyArr B \isType}

    \inferrule* [right=TyTuple]
    {\Delta \vdash A \isType \\ \Delta \vdash B \isType}
    {\Delta \vdash A \tyProduct B \isType}

    \inferrule* [right=TySum]
    {\Delta \vdash A \isType \\ \Delta \vdash B \isType}
    {\Delta \vdash A + B \isType}

    \inferrule* [right=TyForall] %FIXME: macroize forall
    {\Delta, i : s \vdash A \isType}
    {\Delta \vdash \forall i \mathord: s. A \isType}
  \end{mathpar}
  \caption{Index expressions and types}
  \label{fig:indexes-and-types}
\end{figure*}

Given the above data, the rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index terms $\Delta \vdash e : s$
and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are only well-indexed with
respect to some index context $\Delta$. The rules for well-sorted
index terms are particularly simple: either an index term is a
variable that appears in the context (rule \TirName{IVar}), or it is
an application of an index operation taken from $\IndexOpSet$ to other
index terms (rule \TirName{IOp}). The rules for well-indexed types
include the usual rules for constructing types of the simply-typed
$\lambda$-calculus with unit, sum and tuple types (rules
\TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
a primitive type $\tyPrimNm{X}$ and appropriately sorted index terms
$e_1,...,e_n$, the well-indexed type $\tyPrim{X}{e_1,...,e_n}$. The
rule \TirName{TyForall} permits the formation of universally
quantified types, where the universal quantification ranges over all
index terms of some sort.

\subsubsection{Simultaneous Substitutions of Index Terms}
\label{sec:simultaneous-substitution}

We now define simultaneous substitutions of index terms. Given a pair
of index contexts $\Delta$ and $\Delta' = i_1 \mathord: s_1, ..., i_n
\mathord: s_n$, a simultaneous substitution $\sigma : \Delta
\Rightarrow \Delta'$ is a sequence of terms $\sigma = (e_1,...,e_n)$
such that for each $j$ such that $1 \leq j \leq n$, $\Delta \vdash e_j
: s_j$. Given a simultaneous substitution $\sigma = (e_1,...,e_n) :
\Delta \Rightarrow \Delta'$ and a particular variable $i_j : s_j$ in
$\Delta'$, we write $\sigma(i_j)$ to stand for the index term $e_j$.

\begin{enumerate}
\item Define the action of simultaneous substitutions on index terms
  and types: $\sigma^*e$ and $\sigma^*A$. State that simultaneous
  substitution preserves well-sorted-ness and well-indexed-ness.
\item Define composition of simultaneous substitutions
\item Define lifting of simultaneous substitutions
\item Define projection simultaneous substitutions
\end{enumerate}

\begin{lemma}
  Let $\sigma : \Delta \Rightarrow \Delta'$ be a simultaneous
  substitution.
  \begin{enumerate}
  \item If $\Delta' \vdash e : s$, then $\Delta \vdash \sigma^*e : s$; and
  \item If $\Delta' \vdash A \isType$, then $\Delta \vdash \sigma^*A
    \isType$.
  \end{enumerate}
\end{lemma}

\subsubsection{Index Term Equality and Type Equality}
\label{sec:type-equality}

Much of the power of indexing types by the terms of an algebraic
theory comes from the equations of the theory, and the fact that these
equations are inherited by the types. In the units of measure example,
the well-indexed types $\tyPrim{X}{u_1\cdot u_2}$ and $\tyPrim{X}{u_2
  \cdot u_1}$ are considered equal by the type system, due to the
commutativity of the $-\cdot-$ operation in the theory of abelian
groups.

In our general framework, the equations between types are derived from
a set $\IndexAxiomSet$ of axioms $\Delta \vdash e \stackrel{ax}\equiv
e' : s$ between well-sorted index terms. The constraint that axioms
are well-sorted means that both $\Delta \vdash e : s$ and $\Delta
\vdash e' : s$ for all axioms $(\Delta \vdash e \stackrel{ax}\equiv e'
: s) \in \IndexAxiomSet$.

Given a set $\IndexAxiomSet$ of axioms, we generate the judgmental
equality between index terms $\Delta \vdash e \equiv e' : s$ by the
following rules, which allow us to use substitution instances of
axioms, and ensure that judgmental equality is a congruence relation
with respect to the index operations in $\IndexOpSet$:
\begin{mathpar}
  \inferrule*
  {(\Delta \vdash e \stackrel{ax}\equiv e' : s) \in \IndexAxiomSet \\
    \sigma : \Delta' \Rightarrow \Delta}
  {\Delta' \vdash \sigma^*e \equiv \sigma^*e' : s}

  \inferrule*
  {\Delta \vdash e_1 \equiv e'_1 : s_1 \\
    \dots \\
    \Delta \vdash e_n \equiv e'_n : s_n}
  {\Delta \vdash \indexOp{f}(e_1, ..., e_n) \equiv \indexOp{f}(e'_1, ..., e'_n) : s}
\end{mathpar}
We also assume the standard reflexivity, symmetry and transitivity
rules for equality.

\begin{example}[Units of Measure]
  For the units of measure system, seen as an instantiation of our
  general framework, the set $\IndexAxiomSet$ consists of the abelian
  group axioms:
  \begin{displaymath}
    \begin{array}{l}
      u : \mathsf{unit} \vdash u \cdot 1 \stackrel{ax}\equiv u : \mathsf{unit} \\
      u_1, u_2, u_3 : \mathsf{unit} \vdash u_1 \cdot (u_2 \cdot u_3) \stackrel{ax}\equiv (u_1 \cdot u_2) \cdot u_3 : \mathsf{unit} \\
      u_1, u_2 : \mathsf{unit} \vdash u_1 \cdot u_2 \stackrel{ax}\equiv u_2 \cdot u_1 : \mathsf{unit} \\
      u : \mathsf{unit} \vdash u \cdot u^{-1} \stackrel{ax}\equiv 1 : \mathsf{unit} \\
    \end{array}
  \end{displaymath}
\end{example}

The judgmental equality $\Delta \vdash e \equiv e' : s$ on index
terms generates the judgmental equality $\Delta \vdash A \equiv B
\isType$ on types. The basic rule generating judgmental equality on
types states that two applications of primitive types are equal if
their index term arguments are equal:
\begin{displaymath}
  \inferrule*
  {\{ \Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j \leq n}}
  {\Delta \vdash \tyPrim{X}{e_1,...,e_n} \equiv \tyPrim{X}{e'_1,...,e'_n} \isType}
\end{displaymath}
The rest of the rules for judgmental equality on types ensure that it
is a congruence relation on types, and that it is an equivalence
relation (i.e.~judgmental equality on types is reflexive, symmetric
and transitive). For example, for universally quantified types we have
the following congruence rule:
\begin{displaymath}
  \inferrule*
  {\Delta, i : s \vdash A \equiv B \isType}
  {\Delta \vdash \forall i\mathord:s.A \equiv \forall i\mathord:s.B \isType}
\end{displaymath}
The congruence rules for the other type formers are similar.

The judgmental equalities on index terms and on types are preserved by
substitution:
\begin{lemma}
  Let $\sigma : \Delta \Rightarrow \Delta'$ be a simultaneous
  substitution.
  \begin{enumerate}
  \item If $\Delta' \vdash e \equiv e' : s$ then $\Delta \vdash
    \sigma^*e \equiv \sigma^*e' : s$; and
  \item If $\Delta' \vdash A \equiv B \isType$ then $\Delta \vdash
    \sigma^*A \equiv \sigma^* B \isType$.
  \end{enumerate}
\end{lemma}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Semantics of Algebraically-Indexed Types}
\label{sec:semantics-algebraically-indexed-types}

Having defined the language of algebraically-indexed types that we
will consider, we now turn to the denotational semantics of these
types. In \autoref{sec:index-erasure-semantics} we define an
\emph{index-erasure} interpretation of types that interprets every
well-indexed type as a set, ignoring the indexing by algebraic
terms. In the index erasure semantics, the semantics of a well-indexed
type $\tyPrim{X}{e_1,...,e_n}$ is determined solely by the primitive
type $\tyPrimNm{X}$ and not by the index terms $e_1,...,e_n$. Building
on the index-erasure semantics, we define in
\autoref{sec:relational-semantics} the \emph{relational}
interpretation of types.

% For both of our semantics of types, we state and prove two properties:
% types that are syntactically equal have equal denotations, and that
% substitution of index terms is interpreted via composition. These
% properties ensure that our semantics of types is well-behaved.

\subsubsection{The Index-Erasure Interpretation of Types}
\label{sec:index-erasure-semantics}

For each primitive type $\tyPrimNm{X} \in \PrimTypeSet$, we assume
that $\tyPrimNm{X}$ is assigned a set $\tyPrimSem{\tyPrimNm{X}}$. We
extend this to assign a set to every well-formed type by induction on
the type structure by the following clauses.
\begin{displaymath}
  \begin{array}{cc}
    \begin{array}{l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
The interpretation of the unit type is a chosen one element set, and
the interpretations of the function, tuple and sum types is simply by
the corresponding construction on sets. The interpretation of the
universal quantifier simply ignores the indexing: the interpretation
of the type $\forall i\mathord:s.A$ is exactly the interpretation of
the type $A$.

\begin{example}[Units of Measure] Kennedy's original units of measure
  system seen as an instantiation of our system uses the assignment
  $\tyPrimSem{\mathsf{unit}} = \mathbb{Q}$, where $\mathbb{Q}$ is the
  set of rational numbers.
\end{example}

Since the definition of the index-erasure interpretation has
completely ignored the indexing, and type equality is defined as an
extension of index equality, it is straightforward to prove that equal
types have equal denotations when interpreted in the index-erasure
semantics, and that substitution of index terms has no effect on the
index-erasure interpretation of types:
\begin{lemma}\label{lem:tyeqsubst-erasure}
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$ then $\tySem{A} =
    \tySem{B}$; and
  \item If $\Delta \vdash A \isType$ and $\sigma : \Delta' \Rightarrow
    \Delta$, then $\tySem{\sigma^*A} = \tySem{A}$.
  \end{enumerate}
\end{lemma}

\subsubsection{The Relational Interpretation of Types}
\label{sec:relational-semantics}

We now define the relational semantics of each well-indexed type
$\Delta \vdash A \isType$ as some binary relation on the index-erasure
interpretation of $A$. The relational interpretation that we define
will be a \emph{logical relation}, so the relational interpretation of
the function type will be the standard definition relating two
functions if they take related inputs to related outputs. We also use
the standard binary relational interpretations for the unit, tuple and
sum types. (We define the standard relational interpretations formally
below.)

It remains to devise the relational interpretation of primitive types
instantiated with index terms ($\tyPrim{X}{e_1,...,e_n}$), and
universally quantified types ($\forall i\mathord:s. A$). In the
index-erasure semantics, we assumed an assignment of a set
$\tyPrimSem{\tyPrimNm{X}}$ to each primitive type $\tyPrimNm{X} \in
\PrimTypeSet$. For the relational semantics of types
$\tyPrim{X}{e_1,...,e_n}$, we will assume some assignment of a binary
relation on $\tyPrimSem{\tyPrimNm{X}}$ to each primitive type
$\tyPrimNm{X}$ \emph{and} sequence of appropriately sorted (in some
index context $\Delta$) index terms $e_1,...,e_n$. Unlike the index
erasure interpretation, we do not assume a fixed relational
interpretation of the primitive types: in order to interpret the
universal quantification type,

FIXME: explain that we must deal with the fact that we wish to
quantify over extensions to the index context when we interpret the
universal quantification type.

For a set $X$, let $\Rel(X)$ be the set of binary relations $R
\subseteq X \times X$ on $X$.

Given an index context $\Delta$, a relation environment is a
(dependently typed) function:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{X})
\end{displaymath}
that respects equality on index terms: for any $\tyPrimNm{X} \in
\PrimTypeSet$ and pair of simultaneous substitutions $\sigma : \Delta
\Rightarrow \primTyArity(\tyPrimNm{X})$ and $\sigma' : \Delta
\Rightarrow \primTyArity(\tyPrimNm{X})$ such that $\Delta \vdash
\sigma \equiv \sigma' : \primTyArity(\tyPrimNm{X})$, then $\rho\
\tyPrimNm{X}\ \sigma = \rho\ \tyPrimNm{X}\ \sigma'$.

\textbf{FIXME: define the composition of a relational environment with
  a simultaneous substitution}

\textbf{FIXME: reword this part} We define our relational semantics of
types with respect to a given collection of relation environments: we
assume that, for each index context $\Delta$, we are given a set of
relation environments $\relEnv{E}(\Delta)$. Note that this assignment
of sets of relation environments to index contexts is not assumed to
be compositional: it is not necessarily the case that the set of
relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is defined in
terms of $\relEnv{E}(\Delta_1)$ and $\relEnv{E}(\Delta_2)$. However,
it must satisfy the following two properties:
\begin{enumerate}
\item Closure under simultaneous substitution of index terms: if $\rho
  \in \relEnv{E}(\Delta)$ and $\sigma : \Delta' \Rightarrow \Delta$,
  then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
\item \textbf{FIXME: update the formulation of this condition} If we
  have a pair of relation environments $\rho_1 \in \relEnv{E}(\Delta',
  i\mathord:s')$ and $\rho_2 \in \relEnv{E}(\Delta)$, along with an
  index context morphism $\sigma : \Delta' \Rightarrow \Delta$, such
  that the outer edges of the following diagram commute, for all
  primitive types $\tyPrimNm{X} \in \mathit{PrimType}$:
  \begin{displaymath}
    \xymatrix{
      {\idxTms{\Delta}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar[d]_{\sigma^*}
      &
      {\idxTms{\Delta,i\mathord:s'}{s}} \ar[d]^{\sigma_{s'}^*} \ar@/^/[rdd]^{\rho_1}
      \\
      {\idxTms{\Delta'}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2}
      &
      {\idxTms{\Delta',i\mathord:s'}{s}} \ar@{.>}[dr]_\rho
      \\
      &
      &
      {\Rel(\tyPrimSem{\tyPrimNm{X}})}
    }
  \end{displaymath}
  Then there exists a relation environment $\rho \in
  \relEnv{E}(\Delta,i\mathord:s')$ such that the two triangles in the
  bottom right of the diagram commute.
\end{enumerate}

\begin{example}[Units of Measure]
  FIXME: give two examples of relational environments.
\end{example}

Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
additional index variable $i\mathord:s$ to be the following:
\begin{displaymath}
  \extends{\rho}{i\mathord:s} = \{ \rho' \in \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s} = \rho \}
\end{displaymath}

A given relation context $\rho \in \relEnv{E}(\Delta)$ is extended to
a relation interpretation of any type $\Delta \vdash A \isType$ by
induction its derivation. Thus, we define a relation
$\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-indexed
type.
\begin{eqnarray*}
  \rsem{\tyUnit}{\relEnv{E}}\rho & \isDefinedAs & \{(*,*)\} \\
  \rsem{\tyPrim{X}{e_1,...,e_n}}{\relEnv{E}}\rho & \isDefinedAs & \rho\ {\tyPrimNm{X}}\ (e_1,...,e_n) \\
  \rsem{A \tyArr B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A \tyProduct B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A + B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i:s} \}
\end{eqnarray*}
In this definition we have made use of the following three
constructions on binary relations. If $R \in \Rel(X)$ and $S \in
\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is defined as $\{
(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
\}$. With the same assumptions on $R$ and $S$, the relation $R
\relTimes S \in \Rel(X \times Y)$ is defined as $\{
((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
\}$. Finally, the relation $R \relSum S \in \Rel(X + Y)$ is defined as
$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.

The following lemma states that the relation interpretation of types
that we have defined in this section behaves well: the first part of
the lemma states that two types that are judgmentally equal are given
equal relational interpretations, and the second part states that
substitution of index terms in types can be interpreted by the
composition of relational environments with simultaneous
substitutions. Note that both parts of the lemma depend on
\lemref{lem:tyeqsubst-erasure} to be ``well-typed''.
\begin{lemma}\label{lem:tyeqsubst-relational}
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$, then for all $\rho \in
    \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}{\rho} =
    \rsem{B}{\relEnv{E}}{\rho}$;
  \item If $\Delta \vdash A \isType$ and $\sigma : \Delta' \Rightarrow
    \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
    $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
    \circ \sigma)$.
  \end{enumerate}
\end{lemma}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Well-typed Programs}
\label{sec:well-typed-programs}

The well-typed programs are defined in
\autoref{fig:programs}. \textbf{FIXME: define the erasure
  interpretation of terms}.

\begin{figure*}[t]
  \centering
  \begin{mathpar}
    \inferrule* [right=Var]
    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
    {\Delta; \Gamma \vdash x : A}

    \inferrule* [right=TyEq]
    {\Delta; \Gamma \vdash M : A \\ \Delta \vdash A \equiv B \isType}
    {\Delta; \Gamma \vdash M : B}

    \inferrule* [right=Unit]
    { }
    {\Delta; \Gamma \vdash * : 1}

    \inferrule* [right=Pair]
    {\Delta; \Gamma \vdash M : A \\
      \Delta; \Gamma \vdash N : B}
    {\Delta; \Gamma \vdash (M, N) : A \tyProduct B}

    \inferrule* [right=Proj1]
    {\Delta; \Gamma \vdash M : A \tyProduct B}
    {\Delta; \Gamma \vdash \pi_1 M : A}

    \inferrule* [right=Proj2]
    {\Delta; \Gamma \vdash M : A \tyProduct B}
    {\Delta; \Gamma \vdash \pi_2 M : B}

    \inferrule* [right=Inl]
    {\Delta; \Gamma \vdash M : A}
    {\Delta; \Gamma \vdash \mathrm{inl}\ M : A + B}

    \inferrule* [right=Inr]
    {\Delta; \Gamma \vdash M : B}
    {\Delta; \Gamma \vdash \mathrm{inr}\ M : A + B}

    \inferrule* [right=Cond]
    {\Delta; \Gamma \vdash M : A + B \\
      \Delta; \Gamma, x : A \vdash N_1 : C \\
      \Delta; \Gamma, y : B \vdash N_2 : C}
    {\Delta; \Gamma \vdash \textrm{case}\ M\ \textrm{of}\ \textrm{inl}\ x.N_1; \textrm{inr}\ y.N_2 : C}

    \inferrule* [right=Abs]
    {\Delta; \Gamma, x : A \vdash M : B}
    {\Delta; \Gamma \vdash \lambda x.M : A \tyArr B}

    \inferrule* [right=App]
    {\Delta; \Gamma \vdash M : A \tyArr B \\
      \Delta; \Gamma \vdash N : A}
    {\Delta; \Gamma \vdash M N : B}

    \inferrule* [right=UnivAbs]
    {\Delta, i : s; \pi_{i\mathord:s}^*\Gamma \vdash M : A}
    {\Delta; \Gamma \vdash \Lambda i. M : \forall i\mathord:s. A}

    \inferrule* [right=UnivApp]
    {\Delta; \Gamma \vdash M : \forall i\mathord:s. A \\ \Delta \vdash e : s}
    {\Delta; \Gamma \vdash M [e] : (\id_\Delta, e)^*A}
  \end{mathpar}

  \caption{Well-typed Programs}
  \label{fig:programs}
\end{figure*}

% For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
% index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
% structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
% \ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
% one-element set.

\begin{theorem}[Parametricity]
  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
\end{theorem}

\begin{proof}
  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
  three interesting cases follow:
  \begin{description}
  \item[Case \TirName{TyEq}] This case follows directly from
    \lemref{lem:tyeqsubst-relational}.
  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
    $A = \forall i\mathord:s. A'$. We know from the induction
    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}\rho'$,
    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho'$. We are given $\rho \in
    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
    \rsem{\pi_{i:s}^*\Gamma}\rho''$ by \lemref{lem:tysubst-rel}. So we
    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}\rho''$, and thus
    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
      i.M'}\eta_2) \in \rsem{\forall
      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
    know that for all $\rho' \in \extends{\rho}{i:s}$,
    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho'$. If we let $\rho' = \rho \circ
    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
    by \lemref{lem:tysubst-rel}, conclude that $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
    as required.
  \end{description}
  The remaining cases are standard for proofs of the fundamental lemma
  of logical relations for simply-typed systems.
\end{proof}

### 2012-05-22 paper/section-instantiations.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

\section{Instantiations of the General Framework}
\label{sec:instantiations}

\subsection{Multiplicative Group Examples}
\label{sec:multiplicative-examples}

\subsection{Additive Group Examples}
\label{sec:additive-examples}

\subsection{Logic Examples}
\label{sec:logic-examples}

### 2012-05-22 paper/section-introduction.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

\section{Introduction}
\label{sec:introduction}

There are three main points to get across:
\begin{enumerate}
\item Why algebraically indexed types?
\item Why relational parametricity?
\item Why study them together?
\end{enumerate}

Key lessons:

\paragraph{Relational meanings are induced by indexes}

\paragraph{Indexes can usefully have algebraic structure}

\paragraph{Relational meanings can be non-compositional}

\subsection{Contributions}
\label{sec:contributions}

### 2012-05-22 paper/section-motivating-examples.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

\section{Motivating Examples}
\label{sec:motivating-examples}

In this section we present three motivating examples that illustrate
why algebraically indexed types are an interesting object of study,
and why relational semantics is a useful and powerful way of studying
the semantics of algebraically indexed types. The common theme running
through each of the examples we present in this section is the use of
algebraic indexing to capture the invariance \emph{and} variance of
programs with respect to changes in their environment. The first
example that we consider is Kennedy's original units of measure
system.

\subsection{Units of Measure and Scale Invariance}
\label{sec:units-of-measure-example-intro}

FIXME:
\begin{enumerate}
\item Describe the units of measure system
\item Describe the concept of scale invariance
\item Describe the consequences of scale invariance
\end{enumerate}

\subsection{Vector Spaces and Frame Invariance}
\label{sec:vector-spaces-intro}

FIXME: integrate a cut-down version of the geometry-invariance note
here.

\subsection{Security Types and Satisfaction Invariance}
\label{sec:security-types-intro}

FIXME: re-do the security types example from the previous draft, but
making sure to re-phrase everything in terms of ``satisfaction''.

\subsection{Taking Stock}
\label{sec:taking-stock}

Main points to get across:
\begin{enumerate}
\item Algebraic indexing captures both the invariance properties as
  well as the variance properties of the action of programs.
\item Relational semantics provides a powerful way to connect the
  invariance and variance properties to types.
\item The assignment of relational semantics to algebraic indexed may
  not necessarily be compositional.
\end{enumerate}

### 2012-05-29 paper/section-general-framework.tex
Make the notation for inference rules with varying arity more consistent
Finish the description of simultaneous substitutions

@@
 : s_j$. Given a simultaneous substitution $\sigma = (e_1,...,e_n) :
-\Delta \Rightarrow \Delta'$ and a particular variable $i_j : s_j$ in
-$\Delta'$, we write $\sigma(i_j)$ to stand for the index term $e_j$.
+\Delta \Rightarrow \Delta'$ and a variable $i_j \mathord: s_j$ in
+$\Delta'$, we write $\sigma(i_j)$ for the index term $e_j$.
 
-\begin{enumerate}
-\item Define the action of simultaneous substitutions on index terms
-  and types: $\sigma^*e$ and $\sigma^*A$. State that simultaneous
-  substitution preserves well-sorted-ness and well-indexed-ness.
-\item Define composition of simultaneous substitutions
-\item Define lifting of simultaneous substitutions
-\item Define projection simultaneous substitutions
-\end{enumerate}
+For a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
+variable/type pair $i\mathord:s$ such that $i$ does not appear in
+$\Delta$, we define the \emph{projection} simultaneous substitution
+$\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
+$\pi_{i\mathord:s} = (i_1,...,i_n)$. Note that the subscript on
+$\pi_{i\mathord:s}$ is the variable/type pair that is being projected
+away.
 
+For a simultaneous substitution $\sigma : \Delta \Rightarrow \Delta'$
+where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$ and a
+variable/type pair $i\mathord:s$ such that $i$ does not appear in
+either $\Delta$ or $\Delta'$, then we can form the \emph{lifted}
+simultaneous substitution $\sigma_{i\mathord:s} = (\sigma(i_1), ...,
+\sigma(i_n), i) : \Delta,i\mathord:s \Rightarrow
+\Delta',i\mathord:s$. Note that we have implicity used the fact that
+well-sortedness of index terms is preserved by weakening.
+
+Given a simultaneous substitution $\sigma : \Delta \Rightarrow
+\Delta'$ and a well-sorted index term $\Delta' \vdash e : s$, we can
+apply $\sigma$ to $e$ to produce a new index term $\sigma^*e$. The new
+index term $\sigma^*e$ is defined by the following two clauses:
+\begin{mathpar}
+  \sigma^*i \isDefinedAs \sigma(i)
+
+  \sigma^*(\indexOp{f}(e_1,...,e_n)) \isDefinedAs \indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)
+\end{mathpar}
+Similarly, given a well-indexed type $\Delta' \vdash A \isType$, we
+can apply $\sigma$ to $A$ to produce a new type $\sigma^*A$. The
+application of a simultaneous substitution to a type is defined
+recursively by the following clauses:
+\begin{mathpar}
+  \sigma^*(\tyPrim{X}{e_1,...,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,...,\sigma^*e_n}
+
+  \sigma^*\tyUnit \isDefinedAs \tyUnit
+
+  \sigma^*(A \tyArr B) \isDefinedAs \sigma^*A \tyArr \sigma^*B
+
+  \sigma^*(A \tyProduct B) \isDefinedAs \sigma^*A \tyProduct \sigma^*B
+
+  \sigma^*(A + B) \isDefinedAs \sigma^*A + \sigma^*B
+
+  \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A
+\end{mathpar}
 \begin{lemma}
@@
 \end{lemma}
+
+The \emph{composition} of two simultaneous substitutions $\sigma :
+\Delta \Rightarrow \Delta'$ and $\sigma' : \Delta' \Rightarrow \Delta$
+is defined as $(\sigma^*e'_1, ..., \sigma^*e'_n)$, where $\sigma' =
+(e'_1,...,e'_n)$.
 
@@
   \inferrule*
-  {\Delta \vdash e_1 \equiv e'_1 : s_1 \\
-    \dots \\
-    \Delta \vdash e_n \equiv e'_n : s_n}
+  {\{\Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j\leq n}}
   {\Delta \vdash \indexOp{f}(e_1, ..., e_n) \equiv \indexOp{f}(e'_1, ..., e'_n) : s}
@@
 \end{lemma}
+
+We define a pair of simultaneous substitutions $\sigma, \sigma' :
+\Delta \Rightarrow \Delta'$ to be judgmentally equal if their
+component terms are judgmentally equal in the context $\Delta$: i.e.~
+if $\Delta \vdash e_j \equiv e'_j : s_j$, for all $j$. We write
+$\Delta \vdash \sigma \equiv \sigma' : \Delta'$ when two simultaneous
+substitutions are judgmentally equal.
 

### 2012-05-29 paper/section-general-framework.tex
Fix some texing errors in the proof of the abstraction theorem

@@
     \rsem{B}{\relEnv{E}}{\rho}$;
-  \item If $\Delta \vdash A \isType$ and $\sigma : \Delta' \Rightarrow
-    \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
+  \item If $\Delta' \vdash A \isType$ and $\sigma : \Delta \Rightarrow
+    \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
     $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
@@
     hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
-    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}\rho'$,
+    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho'}$,
     we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho'$. We are given $\rho \in
+    \rsem{A'}{\relEnv{E}}{\rho'}$. We are given $\rho \in
     \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
@@
       \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
-    \rsem{\pi_{i:s}^*\Gamma}\rho''$ by \lemref{lem:tysubst-rel}. So we
+    \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho''}$ by \lemref{lem:tysubst-rel}. So we
     can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}\rho''$, and thus
+    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}{\rho''}$, and thus
     $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
@@
     $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho'$. If we let $\rho' = \rho \circ
+    \rsem{A'}{\relEnv{E}}{\rho'}$. If we let $\rho' = \rho \circ
     (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,

### 2012-05-29 paper/section-general-framework.tex
Add the index-erasure semantics of programs, and tidy up the proof of the abstraction theorem

@@
 
-The well-typed programs are defined in
-\autoref{fig:programs}. \textbf{FIXME: define the erasure
-  interpretation of terms}.
+The well-typed programs are defined in \autoref{fig:programs}. FIXME:
+explain each of the rules.
 
@@
 
-% For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
-% index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
-% structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
-% \ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
-% one-element set.
+\subsubsection{The Index-Erasure Interpretation of Programs}
+\label{sec:erasure-semantics-programs}
 
-\begin{theorem}[Parametricity]
+For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
+index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
+structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
+\ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
+one-element set.
+
+% FIXME: do we need to prove something about the effect
+% of substitution on the erasure semantics of contexts?
+
+For a well-typed program $\Delta; \Gamma \vdash M : A$, we define the
+\emph{erasure interpretation} as a function $\tmSem{e} :
+\ctxtSem{\Gamma} \to \tySem{A}$, completely ignoring the indexing
+information. In light of \lemref{lem:tyeqsubst-erasure}, we define
+this function directly on the syntax of well-typed terms, rather than
+on the typing derivations. We define $\tmSem{e}$ by the following
+clauses:
+\begin{eqnarray*}
+  \tmSem{x}\eta & = & \eta_x \\
+  \tmSem{(M,N)}\eta & = & (\tmSem{M}\eta, \tmSem{N}\eta) \\
+  \tmSem{\pi_1M}\eta & = & \pi_1(\tmSem{M}\eta) \\
+  \tmSem{\pi_2M}\eta & = & \pi_2(\tmSem{M}\eta) \\
+  \tmSem{\mathrm{inl}\ M}\eta & = & \mathrm{inl}\ (\tmSem{M}\eta) \\
+  \tmSem{\mathrm{inr}\ M}\eta & = & \mathrm{inr}\ (\tmSem{M}\eta) \\
+  \left\llbracket
+    \begin{array}{l}
+      \textrm{case}\ M\ \textrm{of}\\
+      \textrm{inl}\ x.N_1;\\
+      \textrm{inr}\ y.N_2
+    \end{array}\right\rrbracket^{\mathit{tm}}\eta & = &
+  \left\{
+    \begin{array}{ll}
+      \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
+      \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
+    \end{array}
+  \right. \\
+  \tmSem{\lambda x. M}\eta & = & \lambda v. \tmSem{M}(\eta, v) \\
+  \tmSem{M N}\eta & = & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
+  \tmSem{\Lambda i.\ M}\eta & = & \tmSem{M}\eta \\
+  \tmSem{M[e]}\eta & = & \tmSem{M}\eta
+\end{eqnarray*}
+In the above, we have written $\eta_x$ to denote the appropriate
+projection from $\eta$ to get the position corresponding to the
+variable $x$ in the context.
+
+\subsubsection{The Abstraction Theorem}
+\label{sec:abstraction-theorem}
+
+We now state and prove the abstraction theorem for well-typed programs
+in our system. The abstraction theorem states that the type erasure
+interpretation of a well-typed program is related to itself by the
+relational interpretation of its type. From this result we will be
+able to deduce interesting properties of programs adhering to the
+algebraically-indexed typing discipline in
+\autoref{sec:instantiations}. % FIXME: spruce this up when the instantiations section is actually written.
+
+To state and prove the abstraction theorem for open terms, we first
+need to extend the relational interpretation of types to typing
+contexts. The relational interpretation of contexts is defined by the
+following clauses, where we have used the pairing operation
+$-\relTimes-$ defined in \autoref{sec:relational-semantics}.
+\begin{eqnarray*}
+  \rsem{\epsilon}{\relEnv{E}}{\rho} & = & \{(*,*)\} \\
+  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & = & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
+\end{eqnarray*}
+The relational interpretation of contexts inherits from the relational
+interpretation of types the property of interpretating the application
+of simultaneous substitutions as composition:
+\begin{lemma}\label{lem:ctxtsubst-rel}
+  If $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
+  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
+  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =
+  \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
+\end{lemma}
+
+\begin{theorem}[Abstraction]
   If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
@@
       \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
-    \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho''}$ by \lemref{lem:tysubst-rel}. So we
+    \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho''}$ by \lemref{lem:ctxtsubst-rel}. So we
     can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
@@
     \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
-    by \lemref{lem:tysubst-rel}, conclude that $(\tmSem{M'}\eta_1,
+    by \lemref{lem:ctxtsubst-rel}, conclude that $(\tmSem{M'}\eta_1,
     \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,

### 2012-05-29 paper/section-general-framework.tex
Add AUCTeX commands to deal with multifile document
Fill in some of the section on well-typed terms

@@
 
-The well-typed programs are defined in \autoref{fig:programs}. FIXME:
-explain each of the rules.
+We now present the rules for well-typed programs over the collection
+of types we generated in
+\autoref{sec:algebraically-indexed-types}. Each well-type program is
+assigned an index-erasure semantics, building on the index-erasure
+semantics of types we defined in
+\autoref{sec:index-erasure-semantics}. Our main result
+(\thmref{thm:abstraction}) is that the index-erasure semantics is
+every well-typed program is related to itself in the relational
+interpretation of its type: this is the abstraction theorem for every
+calculus in our general framework. In the next section we will use
+this theorem to derive several interesting results about the programs
+expressible in instances of our general framework.
 
+Well-typed programs are defined with respect to well-indexed typing
+contexts, which are in turn defined with respect to an index
+context. Well-indexed typing contexts with respect to an index context
+$\Delta$ are sequences of variable/type pairs with no repeated
+variable names such that each type is well-indexed with respect to
+$\Delta$. Formally, well-indexed typing contexts are defined by the
+following two rules:
+\begin{mathpar}
+  \inferrule*
+  { }
+  {\Delta \vdash \epsilon \isCtxt}
+
+  \inferrule*
+  {\Delta \vdash \Gamma \isCtxt \\ \Delta \vdash A \isType \\ x \not\in \Gamma}
+  {\Delta \vdash \Gamma, x : A \isCtxt}
+\end{mathpar}
+Substitution by a simultaneous substitution extends to typing contexts
+by performing substitution pointwise on each individual type.
+
+The typing rules for our system define the well-typed terms with
+respect to an index context $\Delta$ and a type context $\Delta \vdash
+\Gamma \isCtxt$. The judgement $\Delta; \Gamma \vdash M : A$ is
+defined in \autoref{fig:programs}. The equational theory on types is
+incorporated into the type system via the rule \TirName{TyEq}, which
+allows for a term that is judged to have type $A$ to also have any
+other equal type $B$.
 \begin{figure*}[t]
@@
 
-    \inferrule* [right=Cond]
+    \inferrule* [right=Case]
     {\Delta; \Gamma \vdash M : A + B \\
@@
 
-For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
-index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
-structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
-\ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
-one-element set.
-
 % FIXME: do we need to prove something about the effect
@@
 
-For a well-typed program $\Delta; \Gamma \vdash M : A$, we define the
-\emph{erasure interpretation} as a function $\tmSem{e} :
-\ctxtSem{\Gamma} \to \tySem{A}$, completely ignoring the indexing
-information. In light of \lemref{lem:tyeqsubst-erasure}, we define
-this function directly on the syntax of well-typed terms, rather than
-on the typing derivations. We define $\tmSem{e}$ by the following
-clauses:
-\begin{eqnarray*}
-  \tmSem{x}\eta & = & \eta_x \\
-  \tmSem{(M,N)}\eta & = & (\tmSem{M}\eta, \tmSem{N}\eta) \\
-  \tmSem{\pi_1M}\eta & = & \pi_1(\tmSem{M}\eta) \\
-  \tmSem{\pi_2M}\eta & = & \pi_2(\tmSem{M}\eta) \\
-  \tmSem{\mathrm{inl}\ M}\eta & = & \mathrm{inl}\ (\tmSem{M}\eta) \\
-  \tmSem{\mathrm{inr}\ M}\eta & = & \mathrm{inr}\ (\tmSem{M}\eta) \\
+We assign an index-erasure semantics to any well-indexed typing
+context $\Delta \vdash \Gamma \isCtxt$ by induction on its structure:
+$\ctxtSem{\epsilon} = \{*\}$ and $\ctxtSem{\Gamma, x : A} =
+\ctxtSem{\Gamma} \times \tySem{A}$. For a well-typed program $\Delta;
+\Gamma \vdash M : A$, we define the \emph{erasure interpretation} as a
+function $\tmSem{e} : \ctxtSem{\Gamma} \to \tySem{A}$, completely
+ignoring the indexing information. In light of
+\lemref{lem:tyeqsubst-erasure}, we define this function directly on
+the syntax of well-typed terms, rather than on the typing
+derivations. We define $\tmSem{e}$ by the following clauses:
+\begin{displaymath}
+  \begin{array}{@{\hspace{0em}}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l}
+    \tmSem{x}\eta & \isDefinedAs & \eta_x \\
+    \tmSem{(M,N)}\eta & \isDefinedAs & (\tmSem{M}\eta, \tmSem{N}\eta) \\
+    \tmSem{\pi_1M}\eta & \isDefinedAs & \pi_1(\tmSem{M}\eta) \\
+    \tmSem{\pi_2M}\eta & \isDefinedAs & \pi_2(\tmSem{M}\eta) \\
+    \tmSem{\mathrm{inl}\ M}\eta & \isDefinedAs & \mathrm{inl}\ (\tmSem{M}\eta) \\
+    \tmSem{\mathrm{inr}\ M}\eta & \isDefinedAs & \mathrm{inr}\ (\tmSem{M}\eta) \\
     \left\llbracket
@@
         \textrm{inr}\ y.N_2
-    \end{array}\right\rrbracket^{\mathit{tm}}\eta & = &
+      \end{array}\right\rrbracket^{\mathit{tm}}\eta & \isDefinedAs &
     \left\{
@@
     \right. \\
-  \tmSem{\lambda x. M}\eta & = & \lambda v. \tmSem{M}(\eta, v) \\
-  \tmSem{M N}\eta & = & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
-  \tmSem{\Lambda i.\ M}\eta & = & \tmSem{M}\eta \\
-  \tmSem{M[e]}\eta & = & \tmSem{M}\eta
-\end{eqnarray*}
+    \tmSem{\lambda x. M}\eta & \isDefinedAs & \lambda v. \tmSem{M}(\eta, v) \\
+    \tmSem{M N}\eta & \isDefinedAs & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
+    \tmSem{\Lambda i.\ M}\eta & \isDefinedAs & \tmSem{M}\eta \\
+    \tmSem{M[e]}\eta & \isDefinedAs & \tmSem{M}\eta    
+  \end{array}
+\end{displaymath}
 In the above, we have written $\eta_x$ to denote the appropriate
-projection from $\eta$ to get the position corresponding to the
-variable $x$ in the context.
+projection from $\eta$ to get the value in the environment
+corresponding to the variable $x$ in the context.
 
@@
 We now state and prove the abstraction theorem for well-typed programs
-in our system. The abstraction theorem states that the type erasure
-interpretation of a well-typed program is related to itself by the
-relational interpretation of its type. From this result we will be
-able to deduce interesting properties of programs adhering to the
-algebraically-indexed typing discipline in
-\autoref{sec:instantiations}. % FIXME: spruce this up when the instantiations section is actually written.
-
-To state and prove the abstraction theorem for open terms, we first
-need to extend the relational interpretation of types to typing
+in our system. To state and prove this theorem for open terms, we
+first need to extend the relational interpretation of types to typing
 contexts. The relational interpretation of contexts is defined by the
@@
 \begin{eqnarray*}
-  \rsem{\epsilon}{\relEnv{E}}{\rho} & = & \{(*,*)\} \\
-  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & = & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
+  \rsem{\epsilon}{\relEnv{E}}{\rho} & \isDefinedAs & \{(*,*)\} \\
+  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & \isDefinedAs & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
 \end{eqnarray*}
@@
 
-\begin{theorem}[Abstraction]
+\begin{theorem}[Abstraction]\label{thm:abstraction}
   If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
@@
 \end{proof}
+
+%%% Local Variables:
+%%% TeX-master: "paper"
+%%% End:

### 2012-05-29 paper/section-general-framework.tex
Fixes to some formatting, and change of '=' to \isDefinedAs, where appropriate

@@
 \begin{displaymath}
-  \begin{array}{cc}
+  \begin{array}{@{\hspace{0.2em}}c@{\hspace{0em}}c}
     \begin{array}{l}
@@
   system seen as an instantiation of our system uses the assignment
-  $\tyPrimSem{\mathsf{unit}} = \mathbb{Q}$, where $\mathbb{Q}$ is the
-  set of rational numbers.
+  $\tyPrimSem{\mathsf{unit}} \isDefinedAs \mathbb{Q}$, where
+  $\mathbb{Q}$ is the set of rational numbers.
 \end{example}
@@
 \begin{example}[Units of Measure]
-  FIXME: give two examples of relational environments.
+  FIXME: give two examples of sets of relational environments.
 \end{example}
 
+% FIXME: motivate this in terms of the semantics of forall types
 Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
@@
 \begin{displaymath}
-  \extends{\rho}{i\mathord:s} = \{ \rho' \in \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s} = \rho \}
+  \extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s} = \rho \}
 \end{displaymath}

### 2012-05-29 paper/section-general-framework.tex
Fix definition of simultaneous substitution composition
Fix some working

@@
 The \emph{composition} of two simultaneous substitutions $\sigma :
-\Delta \Rightarrow \Delta'$ and $\sigma' : \Delta' \Rightarrow \Delta$
-is defined as $(\sigma^*e'_1, ..., \sigma^*e'_n)$, where $\sigma' =
-(e'_1,...,e'_n)$.
+\Delta \Rightarrow \Delta'$ and $\sigma' : \Delta' \Rightarrow
+\Delta''$, where $\sigma' = (e'_1,...,e'_n)$, is defined as $\sigma'
+\circ \sigma \isDefinedAs (\sigma^*e'_1, ..., \sigma^*e'_n) : \Delta
+\Rightarrow \Delta''$.
 
@@
 
-Since the definition of the index-erasure interpretation has
-completely ignored the indexing, and type equality is defined as an
-extension of index equality, it is straightforward to prove that equal
-types have equal denotations when interpreted in the index-erasure
-semantics, and that substitution of index terms has no effect on the
-index-erasure interpretation of types:
+Since the index-erasure interpretation has completely ignores the
+index expressions and type equality is defined as an extension of
+index equality, it is straightforward to prove that equal types have
+equal denotations when interpreted in the index-erasure semantics, and
+that substitution of index terms has no effect on the index-erasure
+interpretation of types:
 \begin{lemma}\label{lem:tyeqsubst-erasure}

### 2012-06-01 paper/section-general-framework.tex
Nearly complete a full draft of the general framework section

@@
 \autoref{sec:semantics-algebraically-indexed-types}. Types are first
-interpreted without reference to the indexing terms: an
-\emph{index-erasure} semantics. The index-erasure semantics provides
-FIXME. Relational semantics...
+interpreted without reference to the indexing terms: the
+\emph{index-erasure} interpretation. The index-erasure interpretation
+provides an ``underlying'' semantics to each type (and term), which we
+think of as an abstraction of the actual operational behaviour of a
+term. By erasing all the index terms, we can immediately see that the
+run-time behaviour of programs cannot depend on index terms. We then
+define the \emph{relational interpretation}, which interprets every
+type as a binary relation on its index-erasure interpretation. The
+relational interpretation acts as ``commentary'' on the underlying
+interpretation, constraining the index-erasure interpretations that
+are allowed. In particular, we use the relational interpretation to
+enforce the invariance properties that we indentified in
+\autoref{sec:motivating-examples}.
 
@@
 
-We now define simultaneous substitutions of index terms. Given a pair
+It will technically convenient to express substitution of index terms
+in our framework in terms of simultaneous substitutions.  Given a pair
 of index contexts $\Delta$ and $\Delta' = i_1 \mathord: s_1, ..., i_n
-\mathord: s_n$, a simultaneous substitution $\sigma : \Delta
+\mathord: s_n$, a \emph{simultaneous substitution} $\sigma : \Delta
 \Rightarrow \Delta'$ is a sequence of terms $\sigma = (e_1,...,e_n)$
@@
 if $\Delta \vdash e_j \equiv e'_j : s_j$, for all $j$. We write
-$\Delta \vdash \sigma \equiv \sigma' : \Delta'$ when two simultaneous
-substitutions are judgmentally equal.
+$\Delta \vdash \sigma \equiv \sigma' \Rightarrow \Delta'$ when two
+simultaneous substitutions are judgmentally equal.
 
@@
 $\Delta \vdash A \isType$ as some binary relation on the index-erasure
-interpretation of $A$. The relational interpretation that we define
-will be a \emph{logical relation}, so the relational interpretation of
-the function type will be the standard definition relating two
-functions if they take related inputs to related outputs. We also use
-the standard binary relational interpretations for the unit, tuple and
-sum types. (We define the standard relational interpretations formally
-below.)
+interpretation of $A$. For a set $X$, we write $\Rel(X)$ for the set
+of binary relations $R \subseteq X \times X$ on $X$.
 
-It remains to devise the relational interpretation of primitive types
-instantiated with index terms ($\tyPrim{X}{e_1,...,e_n}$), and
-universally quantified types ($\forall i\mathord:s. A$). In the
-index-erasure semantics, we assumed an assignment of a set
-$\tyPrimSem{\tyPrimNm{X}}$ to each primitive type $\tyPrimNm{X} \in
-\PrimTypeSet$. For the relational semantics of types
-$\tyPrim{X}{e_1,...,e_n}$, we will assume some assignment of a binary
-relation on $\tyPrimSem{\tyPrimNm{X}}$ to each primitive type
-$\tyPrimNm{X}$ \emph{and} sequence of appropriately sorted (in some
-index context $\Delta$) index terms $e_1,...,e_n$. Unlike the index
-erasure interpretation, we do not assume a fixed relational
-interpretation of the primitive types: in order to interpret the
-universal quantification type,
+For the unit, tuple, sum and function types the we will define the
+relational interpretation as a standard logical relation. The
+relational interpretations of primitive types with index arguments and
+the universally quantified types requires an interpretation of index
+contexts. We assign interpretations to index contexts in terms of
+\emph{relational environments}, which we now define.
 
-FIXME: explain that we must deal with the fact that we wish to
-quantify over extensions to the index context when we interpret the
-universal quantification type.
+A relation environment for a context $\Delta$ is a
+(dependently typed) function that, for each primitive type
+$\tyPrimNm{X}$ and instantiation of its arguments $\sigma$, assigns a
+binary relation $\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure
+interpretation of $\tyPrimNm{X}$:
+\begin{displaymath}
+  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
+\end{displaymath}
+Every relation environment must respect index term equality: for any
+$\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous substitutions
+$\sigma : \Delta \Rightarrow \primTyArity(\tyPrimNm{X})$ and $\sigma'
+: \Delta \Rightarrow \primTyArity(\tyPrimNm{X})$ such that $\Delta
+\vdash \sigma \equiv \sigma' \Rightarrow \primTyArity(\tyPrimNm{X})$, % FIXME: make sure the notation for ss equality is consistent
+then $\rho\ \tyPrimNm{X}\ \sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
 
-For a set $X$, let $\Rel(X)$ be the set of binary relations $R
-\subseteq X \times X$ on $X$.
+% FIXME: an example?
 
-Given an index context $\Delta$, a relation environment is a
-(dependently typed) function:
-\begin{displaymath}
-  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{X})
-\end{displaymath}
-that respects equality on index terms: for any $\tyPrimNm{X} \in
-\PrimTypeSet$ and pair of simultaneous substitutions $\sigma : \Delta
-\Rightarrow \primTyArity(\tyPrimNm{X})$ and $\sigma' : \Delta
-\Rightarrow \primTyArity(\tyPrimNm{X})$ such that $\Delta \vdash
-\sigma \equiv \sigma' : \primTyArity(\tyPrimNm{X})$, then $\rho\
-\tyPrimNm{X}\ \sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
+We write $\mathrm{RelEnv}(\Delta)$ for the set of all relation
+environments for a context $\Delta$. Given a relational environment
+$\rho \in \mathrm{RelEnv}(\Delta')$ and a simultaneous substitution
+$\sigma : \Delta \Rightarrow \Delta'$, we can derive the composed
+relational environment (with an abuse of notation) $\rho \circ \sigma
+\in \mathrm{RelEnv}(\Delta)$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
+\sigma' = \rho\ \tyPrimNm{X}\ (\sigma \circ \sigma')$.
 
-\textbf{FIXME: define the composition of a relational environment with
-  a simultaneous substitution}
+% FIXME: fix the notation here for interpretation of index contexts:
+% should use some sort of semantic brackets
 
-\textbf{FIXME: reword this part} We define our relational semantics of
-types with respect to a given collection of relation environments: we
-assume that, for each index context $\Delta$, we are given a set of
-relation environments $\relEnv{E}(\Delta)$. Note that this assignment
-of sets of relation environments to index contexts is not assumed to
-be compositional: it is not necessarily the case that the set of
-relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is defined in
-terms of $\relEnv{E}(\Delta_1)$ and $\relEnv{E}(\Delta_2)$. However,
-it must satisfy the following two properties:
+As we stated above, index contexts $\Delta$ are interpreted as certain
+sets of relation environments, i.e.~ subsets of
+$\mathrm{RelEnv}(\Delta)$. Depending on the primitive operations that
+we assume for our system, we can have different sets of relational
+environments that enforce different invariants. For the general
+framework, we assume that, for each index context $\Delta$, we are
+given a set of relation environments $\relEnv{E}(\Delta)$. Note that
+this assignment of sets of relation environments to index contexts is
+not assumed to be compositional: it is not necessarily the case that
+the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
+defined in terms of $\relEnv{E}(\Delta_1)$ and
+$\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
+properties:
 \begin{enumerate}
@@
   then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
-\item \textbf{FIXME: update the formulation of this condition} If we
-  have a pair of relation environments $\rho_1 \in \relEnv{E}(\Delta',
-  i\mathord:s')$ and $\rho_2 \in \relEnv{E}(\Delta)$, along with an
-  index context morphism $\sigma : \Delta' \Rightarrow \Delta$, such
-  that the outer edges of the following diagram commute, for all
-  primitive types $\tyPrimNm{X} \in \mathit{PrimType}$:
+\item If we have a pair of relation environments $\rho_1 \in
+  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
+  \relEnv{E}(\Delta)$, along with an simultaneous substitution $\sigma
+  : \Delta' \Rightarrow \Delta$, such that the outer edges of the
+  following diagram commute, for all primitive types $\tyPrimNm{X} \in
+  \mathit{PrimType}$:
+  %FIXME: check that these compositions are the right way round
+  %FIXME: re-do this as a set of equations
+  %FIXME: give a better English description of why this is needed
   \begin{displaymath}
     \xymatrix{
-      {\idxTms{\Delta}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar[d]_{\sigma^*}
+      {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
       &
-      {\idxTms{\Delta,i\mathord:s'}{s}} \ar[d]^{\sigma_{s'}^*} \ar@/^/[rdd]^{\rho_1}
+      {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
       \\
-      {\idxTms{\Delta'}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2}
+      {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
       &
-      {\idxTms{\Delta',i\mathord:s'}{s}} \ar@{.>}[dr]_\rho
+      {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
       \\
@@
   Then there exists a relation environment $\rho \in
-  \relEnv{E}(\Delta,i\mathord:s')$ such that the two triangles in the
-  bottom right of the diagram commute.
+  \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
+  two triangles in the bottom right of the diagram commute.
+  % FIXME: mention that we already know that the square commutes
 \end{enumerate}
 
+Both of these conditions ensure that the relational interpretation of
+types that we define below behaves correctly with respect to
+simultaneous substitution of index terms
+(\lemref{lem:tyeqsubst-relational}, below). The second condition may
+seem mysterious at first, but it is essential for proving that the
+relational interpretation of universal quantification types behaves
+correctly with respect to application of simultaneous substitution.
+
 \begin{example}[Units of Measure]
-  FIXME: give two examples of sets of relational environments.
+  We give two examples of assignments of sets of relational
+  environments to index contexts in the units of measure example. The
+  second example is a refinment of the first. FIXME: some general
+  words about relational environments in these cases.
+  \begin{enumerate}
+  \item FIXME: assign a positive scaling factor to each variable in an
+    index context, extend homomorphically to all index terms, and then
+    use this to define a scaling relation
+  \item FIXME: given an index context, assume a subgroup and a
+    homomorphism into the scaling factors
+  \end{enumerate}
 \end{example}
 
-% FIXME: motivate this in terms of the semantics of forall types
 Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
 the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
-additional index variable $i\mathord:s$ to be the following:
-\begin{displaymath}
-  \extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s} = \rho \}
-\end{displaymath}
+additional index variable $i\mathord:s$ to be
+$\extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in
+\relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s}
+= \rho \}$. The set of extensions of a relational environment will be
+used in the interpretation of the universal quantification type.
 
-A given relation context $\rho \in \relEnv{E}(\Delta)$ is extended to
+A relational environment $\rho \in \relEnv{E}(\Delta)$ is extended to
 a relation interpretation of any type $\Delta \vdash A \isType$ by
-induction its derivation. Thus, we define a relation
-$\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-indexed
-type.
+induction on its derivation: % Thus, we define a relation
+% $\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-indexed
+% type.
 \begin{eqnarray*}
@@
   \rsem{A + B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i:s} \}
+  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i\mathord:s} \}
 \end{eqnarray*}
@@
 The relational interpretation of contexts inherits from the relational
-interpretation of types the property of interpretating the application
+interpretation of types the property of interpreting the application
 of simultaneous substitutions as composition:
 \begin{lemma}\label{lem:ctxtsubst-rel}
-  If $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
-  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
+  If $\Delta \vdash' \Gamma \isCtxt$ and $\sigma : \Delta \Rightarrow
+  \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
   $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =
-  \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
+  \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma)}$.
 \end{lemma}

### 2012-06-01 paper/geometry-twocolumn.tex
Half way through updating the geometry example

@@
-\section{Motivating Example: 2D Geometry}
+\section{Motivating Examples}
 
-\subsection{Coordinate-free Geometry and Invariance}
+\subsection{2-Dimensional Geometry and Invariance}
 
 When writing programs that manipulate $n$-dimensional geometric data,
-the basic data structure is the $n$-tuple of numbers. In the
+the basic data structure is the $n$-tuple of real numbers. In the
 two-dimensional case, tuples $\vec{v} = (x,y)$ can be called upon to
-represent both points -- as offsets from some origin -- and vectors --
-as offsets in their own right. Geometrically, the two notions of
-\emph{point} and \emph{vector} are very different despite their common
-representation. By representing a point concretely as an offset from
-some privileged origin, we inadvertently fix a particular origin,
-$(0,0)$. However, as made clear by Gallieo, many of the laws of
-physics are invariant under change of the frame of reference: if we
-change where we think the origin is, then we do not change the
-fundamental laws. Putting this in concrete terms, if we have two
-frames of reference, one with $(0,0)$ as the origin, and one with
-$(10,20)$ as the origin, then the tuple $(1,1)$ with respect to the
-first frame and $(11,21)$ with respect to the second frame represent
-the \emph{same} point. It is merely an artifact of our representation
-of points as pairs of numbers that these appear to be
-different. Therefore, when we write programs that manipulate data that
-represents points, then we ought to follow the general principle that
-our programs are invariant with respect to changes in the choice of
-origin.
+represent both \emph{points} -- offsets from some origin -- and
+\emph{vectors} -- offsets in their own right. The notions of point and
+vector are very different despite their common representation. By
+representing a point concretely as an offset from some origin, we
+inadvertently fix that origin as privileged.
+% However, as made clear by Gallieo, many of the laws of physics are
+% invariant under change of the frame of reference: if we change where
+% we think the origin is, then we do not change the fundamental laws.
+Putting this in concrete terms, if we have two frames of reference,
+one with $(0,0)$ as the origin, and one with $(10,20)$ as the origin,
+then the tuple $(1,1)$ with respect to the first frame and $(11,21)$
+with respect to the second frame represent the \emph{same} point. It
+is merely an artifact of our representation of points as pairs of
+numbers that these appear to be different. When we write programs that
+manipulate data that represents points, we ought to follow the general
+principle that our programs are invariant with respect to changes in
+the choice of origin. On the other hand, vectors are not intended to
+be regarded as invariant under change of choice of origin. Vectors
+represent offsets, or movements, and the vector $(0,0)$ always
+represents the zero offset.
+% However, they are intended to be invariant under change of basis: the
+% vector $(2,3)$, with respect to the basis $(1,0),(0,1)$, and the
+% vector $(1,1)$, with respect to the basis $(2,0),(0,3)$, should be
+% regarded as the same vector. Invariance under change of basis also
+% applies to the representation of points. We therefore refine the
+% principle of invariance under change of origin for points to also
+% include invariant under change of basis for points and vectors.
 
-Vectors are not intended to be regarded as invariant under change of
-choice of origin. Vectors represent offsets, or movements, and the
-vector $(0,0)$ always represents the zero offset. However, they are
-intended to be invariant under change of basis: the vector $(2,3)$,
-with respect to the basis $(1,0),(0,1)$, and the vector $(1,1)$, with
-respect to the basis $(2,0),(0,3)$, should be regarded as the same
-vector. Invariance under change of basis also applies to the
-representation of points. We therefore refine the principle of
-invariance under change of origin for points to also include invariant
-under change of basis for points and vectors. A particular choice of
-origin and basis is called a \emph{frame} (\cite{gallier11geometric},
-Chapter 2). In general, the technique of maintaining geometric
-invariants such as frame invariance has been referred to as
-\emph{coordinate-free geometry}: we should aim to think in terms of
-abstract concepts such as points and vectors, rather than their actual
-representation as tuples of numbers.
+% A particular choice of origin and basis is called a \emph{frame}
+% (\cite{gallier11geometric}, Chapter 2). In general, the technique of
+% maintaining geometric invariants such as frame invariance has been
+% referred to as \emph{coordinate-free geometry}: we should aim to
+% think in terms of abstract concepts such as points and vectors,
+% rather than their actual representation as tuples of numbers.
 
 Invariance under change of representation immediately recalls
-Reynolds' famous story of two professors teaching the theory of
+Reynolds' fable concerning two professors teaching the theory of
 complex numbers to students \cite{reynolds83types}. One professor
@@
 polar coordinates ($\alpha\cos\theta + i\alpha\sin\theta$). Happily,
-the two classes can interact because the theory of complex numbers is
-invariant with respect to the choice of representation. Reynolds
-formalises the idea of invariance with respect to changes of
-representation by using the notion of preservation of relations. If a
-binary relation $R$ relates two representations of a concept---for
-example the rectangular and polar representations of the same complex
-number are related---then a program that manipulates complex numbers
-at a level of abstraction above their specific representation should
-preserve the relation $R$. Concretely, let $f$ be a program taking
-complex numbers to complex numbers that is intended to respect
-invariance under change of representation. Then if $c$ is a complex
-number in rectangular form, and $c'$ is a complex number in polar form
-such that $R$ relates them, which we write as $(c,c') \in R$, then it
-should be the case that $(f(c), f(c')) \in R$.
+after learning the basic operations on complex numbers in the two
+representations, the two classes can interact because the theory of
+complex numbers is invariant with respect to the choice of
+representation. Reynolds formalises the idea of invariance with
+respect to changes of representation by using the notion of
+preservation of relations. If a binary relation $R$ relates two
+representations of a concept---for example the rectangular and polar
+representations of the same complex number are related---then a
+program that manipulates complex numbers at a level of abstraction
+above their specific representation should preserve the relation
+$R$. Concretely, let $f$ be a program taking complex numbers to
+complex numbers that is intended to respect invariance under change of
+representation. Then if $c$ is a complex number in rectangular form,
+and $c'$ is a complex number in polar form such that $R$ relates them,
+which we write as $(c,c') \in R$, then it should be the case that
+$(f(c), f(c')) \in R$.
 
 We now apply Reynolds' relational approach to defining invariance
-under changes of representation to the geometric setting. Following
+under changes of representation to our geometric setting. Following
 Reynolds, we express the fact that two tuples in $\mathbb{R}^2$ are
-representatives of the same point or vector by defining binary
-relations over $\mathbb{R}^2$. We wish to ensure that invariance with
-respect to all choices of frame is respected, so we parameterise our
-relations by the particular choice made. A neat way to represent a
-choice of frame is as an invertible affine map from $\mathbb{R}^2$ to
-$\mathbb{R}^2$. Affine maps are of the form $F(\vec{v})^\top =
-A\vec{v}^\top + b$, where $A$ is an invertible real $2 \mathord\times
-2$-matrix and $b$ is a vector in $\mathbb{R}^2$. The matrix $A$
-represents the change of basis, and $b$ represents the change of
-origin. The map $F$ tells us how to translate the representative of a
-point in our chosen frame to the representative of the same point in
-the ``standard frame'' with origin represented by $(0,0)$ and basis
-$(1,0),(0,1)$. The set of all such maps forms the general affine group
-$\GA(2)$.
-
-We define a $\GA(2)$-parameterised family of binary relations $\{
-R_{\texttt{pt}}(F) \subseteq \mathbb{R}^2 \times \mathbb{R}^2 \}_{F
-  \in \GA(2)}$. We intend that two tuples $\vec{v_1} = (x_1,y_1)$ and
-$\vec{v_2} = (x_2,y_2)$ are related for $F \in \GA(2)$, if, when the
-former is interpreted with respect to the standard frame and the
-latter is interpreted with respect to the frame described by $F$, the
-two tuples represent the same point.
+representatives of the same point under change of origin by defining a
+family of binary relations over $\mathbb{R}^2$. Our family of binary
+relations is parameterised by a particular choice of origin, and by
+quantifying over all choices of origin we will be able to ensure that
+our programs are invariant with respect to the particular choice of
+origin. We define a $\mathbb{R}^2$-parameterised family of binary
+relations $\{ R_{\vec{o}} \subseteq \mathbb{R}^2 \times \mathbb{R}^2
+\}_{\vec{o} \in \mathbb{R}^2}$ as follows:
 \begin{displaymath}
-  R_{\texttt{pt}}(F) = \{ (\vec{v_1}, \vec{v_2}) \sepbar F\vec{v_2} = \vec{v_1} \}
+  R_{\vec{o}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{o} \}
 \end{displaymath}
@@
 $\mathbb{R}^2$ and returns a single tuple in $\mathbb{R}^2$. We intend
-that all these tuples represent points in the same frame, but that the
-function itself should be invariant with respect to the choice of
-frame. We use Reynolds' idea of preservation of relations to formalise
-this: for any $F \in \GA(2)$ (i.e.~any choice of frame), the function
-$f$ should satisfy the following statement:
+that all these tuples represent points with respect to the same
+origin, but that the function itself should be invariant with respect
+to the choice of origin. We use Reynolds' idea of preservation of
+relations to formalise this: for any $\vec{o} \in \mathbb{R}^2$
+(i.e.~any choice of origin), the function $f$ should satisfy the
+following statement:
 \begin{equation}\label{eq:f-preserve-rel-frame}
-  \begin{array}{l}
-    \forall (\vec{v_1},\vec{v'_1}) \in R_{\texttt{pt}}(F), (\vec{v_2},\vec{v'_2}) \in R_{\texttt{pt}}(F).\\
-    \hspace{2cm}(f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\texttt{pt}}(F).
-  \end{array}
+  \forall (\vec{v_1},\vec{v'_1}) \in R_{\vec{o}}, (\vec{v_2},\vec{v'_2}) \in R_{\vec{o}}. (f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\vec{o}}
 \end{equation}
-Unfolding the definition of $R_{\texttt{pt}}$, we can see that this
-statement is equivalent to the following statement expressing
-invariance of $f$ with respect to changes of frame: for all $F \in
-\GA(2)$,
+Unfolding the definition of $R_{\vec{o}}$, this statement is
+equivalent to the following statement expressing invariance of $f$
+with respect to changes of origin: for all $\vec{o} \in \mathbb{R}^2$,
 \begin{equation}\label{eq:f-invariant-frame}
-  \forall \vec{v_1}, \vec{v_2}.\ f(F\vec{v_1},F\vec{v_2}) = F(f(\vec{v_1},\vec{v_2})).
+  \forall \vec{v_1}, \vec{v_2}.\ f(\vec{v_1} + \vec{o},\vec{v_2} + \vec{o}) = f(\vec{v_1},\vec{v_2}) + \vec{o}.
 \end{equation}
-So Reynolds' idea of preservation of relations, when instantiated with
-the family of relations $R_{\texttt{pt}}$, yields exactly the
-geometric property of invariance under change of frame.
+So Reynolds' preservation of relations, when instantiated with the
+family of relations $R$, yields exactly the geometric property of
+invariance under change of origin.
 
-An example function $f$ that satisfies
-\statementref{eq:f-invariant-frame} is the following function that
-computes a particular affine combination of two points by working
-directly on their coordinate representation:
-\begin{displaymath}
-  f(\vec{v_1}, \vec{v_2}) = \frac{1}{2}\vec{v_1} + \frac{1}{2}\vec{v_2}.
-\end{displaymath}
-In general, affine combinations of points $\lambda_1\vec{v_1} +
-\lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$, satisfy
-\statementref{eq:f-invariant-frame}. Affine combination is one of the
-fundamental building blocks of affine geometry -- the properties of
-points invariant under invertible affine maps
-(\cite{gallier11geometric}, Chapter 2). If we drop the condition that
-$\lambda_1 + \lambda_2 = 1$, then we are dealing with linear
-combinations of vectors, and we are no longer invariant with respect
-to changes of frame. However, linear combinations are invariant with
-respect to change of basis. We can represent changes of basis as
-linear invertible maps $B : \mathbb{R}^2 \to \mathbb{R}^2$. The set of
-all such maps forms the general linear group $\GL(2)$. We now define
-another family of relations $\{R_{\texttt{vec}}(B) \subseteq
-\mathbb{R}^2 \times \mathbb{R}^2 \}_{B \in \GL(2)}$ that relates two
-points up to change of basis:
-\begin{displaymath}
-  R_{\texttt{vec}}(B) = \{ (\vec{v_1},\vec{v_2}) \sepbar B\vec{v_2} = \vec{v_1} \}.
-\end{displaymath}
-Now the functions $f_{\lambda_1\lambda_2}(\vec{v_1},\vec{v_2}) =
-\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$, for arbitrary $\lambda_1$
-and $\lambda_2$, do preserve the relations $R_{\texttt{vec}}(B)$, for
-all $B \in \GL(2)$. Unfolding the definition of $R_{\texttt{vec}}(B)$
-in the analogous statement to \statementref{eq:f-preserve-rel-frame}
-for $R_{\texttt{vec}}$ instead of $R_{\texttt{pt}}$, we can see that
-preservation of the relations $R_{\texttt{vec}}$ characterises the
-functions that are invariant under change of basis. For all $B \in \GL(2)$, we have
-\begin{displaymath}
-  \forall \vec{v_1}, \vec{v_2}.\ f(B\vec{v_1},B\vec{v_2}) = B(f(\vec{v_1},\vec{v_2})),
-\end{displaymath}
-and this is exactly the property of invariance under change of basis
-we required above of programs manipulating vectors.
+% An example function $f$ that satisfies
+% \statementref{eq:f-invariant-frame} is the following function that
+% computes a particular affine combination of two points by working
+% directly on their coordinate representation:
+% \begin{displaymath}
+%   f(\vec{v_1}, \vec{v_2}) = \frac{1}{2}\vec{v_1} + \frac{1}{2}\vec{v_2}.
+% \end{displaymath}
+% In general, affine combinations of points $\lambda_1\vec{v_1} +
+% \lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$, satisfy
+% \statementref{eq:f-invariant-frame}. Affine combination is one of the
+% fundamental building blocks of affine geometry -- the properties of
+% points invariant under invertible affine maps
+% (\cite{gallier11geometric}, Chapter 2). If we drop the condition that
+% $\lambda_1 + \lambda_2 = 1$, then we are dealing with linear
+% combinations of vectors, and we are no longer invariant with respect
+% to changes of frame. However, linear combinations are invariant with
+% respect to change of basis. We can represent changes of basis as
+% linear invertible maps $B : \mathbb{R}^2 \to \mathbb{R}^2$. The set of
+% all such maps forms the general linear group $\GL(2)$. We now define
+% another family of relations $\{R_{\texttt{vec}}(B) \subseteq
+% \mathbb{R}^2 \times \mathbb{R}^2 \}_{B \in \GL(2)}$ that relates two
+% points up to change of basis:
+% \begin{displaymath}
+%   R_{\texttt{vec}}(B) = \{ (\vec{v_1},\vec{v_2}) \sepbar B\vec{v_2} = \vec{v_1} \}.
+% \end{displaymath}
+% Now the functions $f_{\lambda_1\lambda_2}(\vec{v_1},\vec{v_2}) =
+% \lambda_1\vec{v_1} + \lambda_2\vec{v_2}$, for arbitrary $\lambda_1$
+% and $\lambda_2$, do preserve the relations $R_{\texttt{vec}}(B)$, for
+% all $B \in \GL(2)$. Unfolding the definition of $R_{\texttt{vec}}(B)$
+% in the analogous statement to \statementref{eq:f-preserve-rel-frame}
+% for $R_{\texttt{vec}}$ instead of $R_{\texttt{pt}}$, we can see that
+% preservation of the relations $R_{\texttt{vec}}$ characterises the
+% functions that are invariant under change of basis. For all $B \in \GL(2)$, we have
+% \begin{displaymath}
+%   \forall \vec{v_1}, \vec{v_2}.\ f(B\vec{v_1},B\vec{v_2}) = B(f(\vec{v_1},\vec{v_2})),
+% \end{displaymath}
+% and this is exactly the property of invariance under change of basis
+% we required above of programs manipulating vectors.
 
-Note that the family of relations $R_{\texttt{vec}}$ is just
-$R_{\texttt{pt}}$ when restricted to elements of the group
-$\GL(2)$. In the type system we introduce in the next section, we
-combine points and vectors into the same data type. Whether it
-represents a point or a vector depends on the group of geometric
-transformations that we expect it to be invariant under.
+% Note that the family of relations $R_{\texttt{vec}}$ is just
+% $R_{\texttt{pt}}$ when restricted to elements of the group
+% $\GL(2)$. In the type system we introduce in the next section, we
+% combine points and vectors into the same data type. Whether it
+% represents a point or a vector depends on the group of geometric
+% transformations that we expect it to be invariant under.
 
-\subsection{A Type System for Geometric Invariance}
+FIXME: vectors are not invariant under change of origin, so they are
+represented by the relation $R_0$, which is just equality. Bxut they,
+and points are invariant under change of basis. However, as we shall
+see below, not all operations are invariant under all changes of
+basis. In particular the dot product is only invariant under
+orthogonal transformations.
+
+\subsection{A Type System for Origin Invariance}
 \label{sec:type-system-geom-intro}
@@
 discipline. In Reynolds' case, the typing discipline was the
-second-order polymorphic $\lambda$-calculus, also known as System F,
-where new types can be constructed by universal quantification over
-all types. Universal type quantification is commonly written $\forall
-\alpha. A$, which denotes a type that may be instantiated with any
-other type $B$ by replacing the type variable $\alpha$ with $B$ in
-$A$.
+polymorphic $\lambda$-calculus where new types can be constructed by
+universal quantification over all types. Universal type quantification
+is commonly written $\forall \alpha. A$, which denotes a type that may
+be instantiated with any other type $B$ by replacing the type variable
+$\alpha$ with $B$ in $A$.
 
@@
 the previous section (e.g., \statementref{eq:f-invariant-frame}), we
-did not quantify over all relations. Instead, we quantified over all
+did not quantify over all relations. Instead, we quantified over all choices of origin
+
 elements of a geometric group such as $\GA(2)$ or $\GL(2)$, and used
@@
 
-\subsection{Euclidean Geometry}
-\label{sec:euclidean-geom}
-
-\subsubsection{Dot Product}
-
-Euclidean geometry extends affine geometry with the operation of
-\emph{dot product} (also known as the \emph{inner product}) of two
-vectors, which produces a real number. The standard Euclidean dot
-product is defined on the coordinate representation in the following
-way:
-\begin{displaymath}
-  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2
-\end{displaymath}
-The dot product is used to define properties of vectors such as their
-\emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
-of the length of the vector $\vec{v}$, and the notion of
-orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
-if $\vec{v_1}\cdot\vec{v_2} = 0$.
-
-To incorporate the operation of dot product into our calculus, we must
-give it a type. It is not the case that the dot product is invariant
-under the action of any of the groups $\GA(2)$, $\GL(2)$ or
-$\Transl(2)$ that we have considered so far. In fact, the dot product
-is invariant under the subgroup $\Orth(2)$ of $\GL(2)$ of
-\emph{orthogonal} linear transformations. The group $\Orth(2)$
-consists of invertible linear maps whose determinant is either $1$ or
-$-1$. We introduce orthogonal transformations as a new sort
-$\Orth(2)$, and overload the multiplicative group operations syntax
-for inhabitants of $\Orth(2)$. We also assume a function $\iota_O$
-that takes $e : \Orth(2)$ to $\iota_O(e) : \GL(2)$.
-
-We may now assign the following type to the dot product primitive:
-\begin{displaymath}
-  (\cdot) : \forall O \mathord: \Orth(2).\ \mathsf{vec}\langle \iota_O(O) \ltimes 0 \rangle \to \mathsf{vec}\langle \iota_O(O) \ltimes 0 \rangle \to \mathsf{real}
-\end{displaymath}
-With dot product as a primitive operation, we can define several other
-useful derived operations. First, the norm of a vector, which in the
-case of standard Euclidean geometry is the square of the length of the
-vector. The norm is invariant under orthogonal changes of basis, but
-not arbitrary changes of basis. Hence, we quantify over all orthogonal
-transformations in the type of $\mathrm{norm}$:
-\begin{displaymath}
-  \begin{array}{l}
-    \mathrm{norm} : \forall O \mathord: \Orth(2).\ \mathsf{vec}\langle O \rangle \to \mathsf{real} \\
-    \mathrm{norm}\ [O]\ v = v \cdot v
-  \end{array}
-\end{displaymath}
-Given the norm function, we can compute the distance between a pair of
-points in a frame, provided that the frame has an orthogonal basis
-described by some orthogonal transformation $O$:
-\begin{displaymath}
-  \begin{array}{l}
-    \mathrm{distance} : \forall O \mathord: \Orth(2), t \mathord: \Transl(2).\ \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{real} \\
-    \mathrm{distance}\ [O]\ [t]\ p_1\ p_2 = \mathrm{sqrt}\ (\mathrm{norm}\ (\mathrm{offset}\ p_1\ p_2))
-  \end{array}
-\end{displaymath}
-In the definition of $\mathrm{distance}$, we have made use of a square
-root function on the reals $\mathrm{sqrt} : \mathsf{real} \to
-\mathsf{real}$.
-
-The dot product is also used to determine whether two vectors are
-orthogonal to one another. In the case of the standard dot product in
-$2$-dimensional space, two vectors are orthogonal if they are at right
-angles to one another.
-\begin{displaymath}
-  \begin{array}{l}
-    \mathrm{orthogonal} : \forall O \mathord: \Orth(2).\ \mathsf{vec}\langle O \rangle \to \mathsf{vec}\langle O \rangle \to \mathsf{bool} \\
-    \mathrm{orthogonal}\ [O]\ v_1\ v_2 = v_1 \cdot v_2 \equiv 0
-  \end{array}
-\end{displaymath}
-
-\subsubsection{Cross Product}
-
-\begin{figure}[t]
-  \centering
-  \begin{eqnarray*}
-    0   &:& \forall S \mathord:\Scale.\ \mathsf{real}\langle S \rangle \\
-    1   &:& \mathsf{real}\langle 1 \rangle \\
-    (+) &:& \forall S \mathord:\Scale.\ \mathsf{real}\langle S \rangle \to \mathsf{real}\langle S \rangle \to \mathsf{real}\langle S \rangle \\
-    (-) &:& \forall S \mathord:\Scale.\ \mathsf{real}\langle S \rangle \to \mathsf{real}\langle S \rangle \to \mathsf{real}\langle S \rangle \\
-    (*) &:& \forall S_1,S_2 \mathord:\Scale.\ \mathsf{real}\langle S_1 \rangle \to \mathsf{real}\langle S_2 \rangle \to \mathsf{real}\langle S_1S_2 \rangle \\
-    (/) &:& \forall S_1,S_2 \mathord:\Scale.\ \mathsf{real}\langle S_1 \rangle \to \mathsf{real}\langle S_2 \rangle \to \mathsf{real}\langle S_1S_2^{-1} \rangle + 1 \\
-    \mathrm{abs} &:& \forall S \mathord:\Scale.\ \mathsf{real}\langle S \rangle \to \mathsf{real}\langle |S| \rangle \\
-    \mathrm{sqrt} &:& \mathsf{real}\langle 1 \rangle \to \mathsf{real}\langle 1 \rangle
-  \end{eqnarray*}
-  \caption{Operations on scaled real numbers}
-  \label{fig:real-ops}
-\end{figure}
-
-The cross product is another operation that takes pairs of vectors and
-produces a real number. Algebraically, the cross product is defined on
-coordinate representations as follows:
-\begin{displaymath}
-  (x_1,y_1) \times (x_2,y_2) = x_1y_2 - x_2y_1
-\end{displaymath}
-Geometrically, the cross product of two vectors is the signed area of
-the parallelogram described by the pair of input vectors. It turns out
-that the value of the cross product of two vectors varies with the
-determinant of the matrix describing the basis of the vector space. We
-therefore augment our calculus with a new sort $\Scale$ of scale
-factors. Semantically, $\Scale$ ranges over the non-zero real numbers
-and forms an abelian group, which we write multiplicatively. We also
-add two new operations: determinant, $\det B$, which takes an
-inhabitant of $\GL(2)$ to its determinant in $\Scale$, and the
-operation of absolute value that takes scaling factors to scaling
-factors which we write as $|e|$. We also refine the type
-$\mathsf{real}$ of real numbers to take a parameter of sort $\Scale$:
-$\mathsf{real}\langle e \rangle$, where the old type $\mathsf{real}$
-is just $\mathsf{real}\langle 1 \rangle$. The collection of operations
-on real numbers with scaling factors is shown in
-\autoref{fig:real-ops}.
-
-With the additional sort $\Scale$, and the refined type of real
-numbers, we can state the type of the cross product.
-\begin{displaymath}
-  (\times) : \forall B \mathord: \GL(2).\ \mathsf{vec}\langle B \rangle \to \mathsf{vec}\langle B \rangle \to \mathsf{real}\langle \det B \rangle
-\end{displaymath}
-
-For $O : \Orth(2)$, the absolute value of the determinant is always
-$1$, so we assume the equation $|\det O| = 1$ to hold.
-
-\begin{example}
-  Using the structure we have defined in this subsection, we can
-  define the following function that computes the area of a
-  triangle. By the type of this function, we can see that the area of
-  a triangle is invariant under orthognal transformations and
-  arbitrary translations.
-  \begin{displaymath}
-    \begin{array}{l}
-      \mathrm{area} : \forall O\mathord:\Orth(2), t\mathord:\Transl(2).\ \\
-      \hspace{1.2cm}\mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{vec}\langle O \ltimes t \rangle \to \mathsf{real}\langle 1 \rangle \\
-      \mathrm{area}\ [O]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
-    \end{array}
-  \end{displaymath}
-  In the body of this function, we have performed the calculation in
-  several steps, each of which removes some of the symmetry described
-  by the types. First we have computed the two offset vectors $p_2 -
-  p_1$ and $p_3 - p_1$. These operations remove the effect of
-  translations on the result, in exactly the same way as the type
-  isomorphism we demonstrated above. We then compute the cross product
-  of the two vectors to determine the area of the parallelogram
-  described by the sides of the triangle. This value has type
-  $\mathsf{real}\langle \det O \rangle$. We know that the determinant
-  of an orthogonal transformation is always either $-1$ or $1$, so we
-  have a result which varies according to possible reflections of the
-  space. To remove this variation, we apply the $\mathrm{abs}$
-  function to leave us with a value of type $\mathsf{real}\langle
-  |\det O| \rangle$. This type is equal to the type
-  $\mathsf{real}\langle 1 \rangle$. Finally, we multiply by
-  $\frac{1}{2}$ to recover the area of the triangle, rather than the
-  whole parallelogram.
-\end{example}
+%%% Local Variables:
+%%% TeX-master: "paper"
+%%% End:

### 2012-06-05 paper/section-general-framework.tex
Minor changes to the discussion of the syntax of indexing theories

@@
 
-The indexing terms and types of our system are derived from the
-following data:
+The indexing terms and types of an instantiation of our general
+framework are derived from the following data:
 \begin{enumerate}
@@
   arbitrary sorts taken from $\SortSet$.
-\item A collection $\IndexOpSet$ of index operations, with an
-  accompanying function $\indexOpArity : \IndexOpSet \to \SortSet^*
-  \times \SortSet$ (we use the notation $A^*$ to denote the set of
-  lists of elements of some set $A$). 
-\item A collection $\PrimTypeSet$ of primitive types, with an
-  accompanying function $\primTyArity : \PrimTypeSet \to
-  \SortSet^*$.
+\item A collection $\IndexOpSet$ of index operations, with a function
+  $\indexOpArity : \IndexOpSet \to \SortSet^* \times \SortSet$ (we use
+  the notation $A^*$ to denote the set of lists of elements of some
+  set $A$).
+\item A collection $\PrimTypeSet$ of primitive types, with a function
+  $\primTyArity : \PrimTypeSet \to \SortSet^*$, describing the sorts
+  of the arguments of each primitive type.
 \end{enumerate}
+
+\begin{example}[Origin Invariance]
+  \fixme{Do this, then update the units of measure example}
+\end{example}
 
@@
 
-We also assume a countably infinite collection of index expression
-variables. We will use meta-syntactic variables $i, i_1, i_2, ...$ and
-so on to stand for index expression variables. \emph{Index contexts}
-$\Delta = i_1 : s_1, ..., i_n : s_n$ are built from the index
-expression variable/sort pairs, such that all the index variable names
-are distinct.
+We assume a countably infinite collection of index variable names $i,
+i_1, i_2,$ and so on. \emph{Index contexts} $\Delta = i_1 \mathord:
+s_1, ..., i_n \mathord: s_n$ are lists of index expression
+variable/sort pairs, such that all the index variable names are
+distinct.
 
@@
 
-  $\Delta = i_1 : s_1, ..., i_n : s_n$, where each $s_i \in \SortSet$ and no variable name is repeated.
+  $\Delta = i_1 \mathord: s_1, ..., i_n \mathord: s_n$, where each $s_i \in \SortSet$ and no variable name is repeated.
 
@@
 and well-indexed types $\Delta \vdash A \isType$. Since index
-variables may appear in types, types are only well-indexed with
-respect to some index context $\Delta$. The rules for well-sorted
+variables may appear in types, types are judged to be well-indexed
+with respect to an index context $\Delta$. The rules for well-sorted
 index terms are particularly simple: either an index term is a
@@
 
-It will technically convenient to express substitution of index terms
-in our framework in terms of simultaneous substitutions.  Given a pair
-of index contexts $\Delta$ and $\Delta' = i_1 \mathord: s_1, ..., i_n
-\mathord: s_n$, a \emph{simultaneous substitution} $\sigma : \Delta
-\Rightarrow \Delta'$ is a sequence of terms $\sigma = (e_1,...,e_n)$
-such that for each $j$ such that $1 \leq j \leq n$, $\Delta \vdash e_j
-: s_j$. Given a simultaneous substitution $\sigma = (e_1,...,e_n) :
-\Delta \Rightarrow \Delta'$ and a variable $i_j \mathord: s_j$ in
-$\Delta'$, we write $\sigma(i_j)$ for the index term $e_j$.
+It will be technically convenient to express substitution of index
+terms in our framework in terms of simultaneous substitutions.  Given
+a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord: s_1,
+..., i_n \mathord: s_n$, a \emph{simultaneous substitution} $\Delta
+\vdash \sigma \Rightarrow \Delta'$ is a sequence of terms $\sigma =
+(e_1,...,e_n)$ such that for each $j$ such that $1 \leq j \leq n$,
+$\Delta \vdash e_j : s_j$. Given a simultaneous substitution $\Delta
+\vdash \sigma = (e_1,...,e_n) \Rightarrow \Delta'$ and a variable $i_j
+\mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for the index term
+$e_j$.
 
-For a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
+Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
 variable/type pair $i\mathord:s$ such that $i$ does not appear in
@@
 $\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
-$\pi_{i\mathord:s} = (i_1,...,i_n)$. Note that the subscript on
-$\pi_{i\mathord:s}$ is the variable/type pair that is being projected
-away.
+$\pi_{i\mathord:s} = (i_1,...,i_n)$. The subscript on
+$\pi_{i\mathord:s}$ is the variable/type pair that is being discarded.
 
-For a simultaneous substitution $\sigma : \Delta \Rightarrow \Delta'$
-where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$ and a
+For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
+\Delta'$ where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$ and a
 variable/type pair $i\mathord:s$ such that $i$ does not appear in
 either $\Delta$ or $\Delta'$, then we can form the \emph{lifted}
-simultaneous substitution $\sigma_{i\mathord:s} = (\sigma(i_1), ...,
-\sigma(i_n), i) : \Delta,i\mathord:s \Rightarrow
+simultaneous substitution $\Delta,i\mathord:s \vdash
+\sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
 \Delta',i\mathord:s$. Note that we have implicity used the fact that
-well-sortedness of index terms is preserved by weakening.
+well-sortedness of index terms is preserved by addition of extra items
+to the context.
 
-Given a simultaneous substitution $\sigma : \Delta \Rightarrow
-\Delta'$ and a well-sorted index term $\Delta' \vdash e : s$, we can
-apply $\sigma$ to $e$ to produce a new index term $\sigma^*e$. The new
-index term $\sigma^*e$ is defined by the following two clauses:
+Application of a simultaneous substitution $\Delta \vdash \sigma
+\Rightarrow \Delta'$ to a well-sorted index expression $\Delta' \vdash
+e : s$ yields a well-sorted index expression $\Delta \vdash \sigma^*e
+: s$. The expression $\sigma^*e$ is defined by the following two
+clauses:
 \begin{mathpar}
@@
 application of a simultaneous substitution to a type is defined
-recursively by the following clauses:
+by the following clauses:
 \begin{mathpar}
@@
 
-The \emph{composition} of two simultaneous substitutions $\sigma :
-\Delta \Rightarrow \Delta'$ and $\sigma' : \Delta' \Rightarrow
-\Delta''$, where $\sigma' = (e'_1,...,e'_n)$, is defined as $\sigma'
-\circ \sigma \isDefinedAs (\sigma^*e'_1, ..., \sigma^*e'_n) : \Delta
-\Rightarrow \Delta''$.
+The \emph{composition} of two simultaneous substitutions $\Delta
+\vdash \sigma \Rightarrow \Delta'$ and $\Delta' \vdash \sigma'
+\Rightarrow \Delta''$, where $\sigma' = (e'_1,...,e'_n)$, is defined
+as $\Delta \vdash \sigma' \circ \sigma \isDefinedAs (\sigma^*e'_1,
+..., \sigma^*e'_n) \Rightarrow \Delta''$.
 

### 2012-06-05 paper/section-general-framework.tex
More updates to the general framework section

@@
 \begin{lemma}
-  Let $\sigma : \Delta \Rightarrow \Delta'$ be a simultaneous
+  Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
   substitution.
@@
 
-Much of the power of indexing types by the terms of an algebraic
-theory comes from the equations of the theory, and the fact that these
-equations are inherited by the types. In the units of measure example,
-the well-indexed types $\tyPrim{X}{u_1\cdot u_2}$ and $\tyPrim{X}{u_2
-  \cdot u_1}$ are considered equal by the type system, due to the
-commutativity of the $-\cdot-$ operation in the theory of abelian
-groups.
+\fixme{Consider updating this example} Much of the power of indexing
+types by the terms of an algebraic theory comes from the equations of
+the theory, and the fact that these equations are inherited by the
+types. In the units of measure example, the well-indexed types
+$\tyPrim{X}{u_1\cdot u_2}$ and $\tyPrim{X}{u_2 \cdot u_1}$ are
+considered equal by the type system, due to the commutativity of the
+$-\cdot-$ operation in the theory of abelian groups.
 
-In our general framework, the equations between types are derived from
+In the general framework, the equations between types are derived from
 a set $\IndexAxiomSet$ of axioms $\Delta \vdash e \stackrel{ax}\equiv
-e' : s$ between well-sorted index terms. The constraint that axioms
-are well-sorted means that both $\Delta \vdash e : s$ and $\Delta
-\vdash e' : s$ for all axioms $(\Delta \vdash e \stackrel{ax}\equiv e'
-: s) \in \IndexAxiomSet$.
+e' : s$. We assume that all axioms $(\Delta \vdash e
+\stackrel{ax}\equiv e' : s) \in \IndexAxiomSet$ are well-sorted:
+meaning that both $\Delta \vdash e : s$ and $\Delta \vdash e' : s$.
 
@@
   \inferrule*
-  {(\Delta \vdash e \stackrel{ax}\equiv e' : s) \in \IndexAxiomSet \\
-    \sigma : \Delta' \Rightarrow \Delta}
-  {\Delta' \vdash \sigma^*e \equiv \sigma^*e' : s}
+  {(\Delta' \vdash e \stackrel{ax}\equiv e' : s) \in \IndexAxiomSet \\
+    \Delta \vdash \sigma \Rightarrow \Delta'}
+  {\Delta \vdash \sigma^*e \equiv \sigma^*e' : s}
 
@@
 We also assume the standard reflexivity, symmetry and transitivity
-rules for equality.
+rules for the judgmental equality.
+
+\begin{example}[Origin Invariance]
+  FIXME
+\end{example}
 

### 2012-06-05 paper/section-general-framework.tex
Push the notation changes through, and add some notes

@@
 \label{sec:a-general-framework}
+
+\fixme{Update this introduction}
 
@@
 
+We write $\Delta \Rightarrow \Delta'$ for the set of all simultaneous
+substitutions $\sigma$ such that $\Delta \vdash \sigma \Rightarrow
+\Delta'$. When we discuss the relational interpretation of types
+below, it will be useful to think of any sequence of sorts as an index
+context. \fixme{flesh this out}
+
 Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
@@
 
-Having defined the language of algebraically-indexed types that we
-will consider, we now turn to the denotational semantics of these
-types. In \autoref{sec:index-erasure-semantics} we define an
+Having defined the language of algebraically-indexed types, we now
+turn to their denotational interpretation. In
+\autoref{sec:index-erasure-semantics} we define an
 \emph{index-erasure} interpretation of types that interprets every
-well-indexed type as a set, ignoring the indexing by algebraic
-terms. In the index erasure semantics, the semantics of a well-indexed
-type $\tyPrim{X}{e_1,...,e_n}$ is determined solely by the primitive
-type $\tyPrimNm{X}$ and not by the index terms $e_1,...,e_n$. Building
-on the index-erasure semantics, we define in
+well-indexed type as a set, ignoring the indexing expressions. Building on
+the index-erasure semantics, we define in
 \autoref{sec:relational-semantics} the \emph{relational}
@@
 
-For each primitive type $\tyPrimNm{X} \in \PrimTypeSet$, we assume
-that $\tyPrimNm{X}$ is assigned a set $\tyPrimSem{\tyPrimNm{X}}$. We
-extend this to assign a set to every well-formed type by induction on
-the type structure by the following clauses.
+The defining feature of the index erasure interpretation is that the
+semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
+determined solely by the primitive type $\tyPrimNm{X}$ and not by the
+index terms $e_1,...,e_n$. Accordingly, we assume that each primitive
+type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
+$\tyPrimSem{\tyPrimNm{X}}$. We extend this assignment to every
+well-indexed type by induction on the type structure:
 \begin{displaymath}
@@
 The interpretation of the unit type is a chosen one element set, and
-the interpretations of the function, tuple and sum types is simply by
-the corresponding construction on sets. The interpretation of the
-universal quantifier simply ignores the indexing: the interpretation
-of the type $\forall i\mathord:s.A$ is exactly the interpretation of
-the type $A$.
+the interpretations of the function, tuple and sum types are simply
+the corresponding constructions on sets. The interpretation of the
+universal quantifier ignores the indexing: the interpretation of the
+type $\forall i\mathord:s.A$ is exactly the interpretation of the type
+$A$.
+
+\begin{example}[Origin Invariance]
+  \fixme{do this}
+\end{example}
 
@@
   $\mathbb{Q}$ is the set of rational numbers.
+
+  FIXME: Also do the primitive operations
 \end{example}
 
-Since the index-erasure interpretation has completely ignores the
-index expressions and type equality is defined as an extension of
-index equality, it is straightforward to prove that equal types have
-equal denotations when interpreted in the index-erasure semantics, and
-that substitution of index terms has no effect on the index-erasure
-interpretation of types:
+% The index-erasure interpretation completely ignores index expressions
+% and type equality is defined as an extension of index
+% equality. Therefore, it is straightforward to prove that equal types
+% have equal denotations when interpreted in the index-erasure
+% semantics, and that substitution of index terms has no effect on the
+% index-erasure interpretation of types:
 \begin{lemma}\label{lem:tyeqsubst-erasure}
@@
     \tySem{B}$; and
-  \item If $\Delta \vdash A \isType$ and $\sigma : \Delta' \Rightarrow
-    \Delta$, then $\tySem{\sigma^*A} = \tySem{A}$.
+  \item If $\Delta' \vdash A \isType$ and $\Delta \vdash \sigma
+    \Rightarrow \Delta'$, then $\tySem{\sigma^*A} = \tySem{A}$.
   \end{enumerate}
@@
 
-For the unit, tuple, sum and function types the we will define the
+For the unit, tuple, sum and function types we will define the
 relational interpretation as a standard logical relation. The
@@
 contexts. We assign interpretations to index contexts in terms of
-\emph{relational environments}, which we now define.
+\emph{relational environments}, which we now define. \fixme{Backwards ref}
 
@@
 $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous substitutions
-$\sigma : \Delta \Rightarrow \primTyArity(\tyPrimNm{X})$ and $\sigma'
-: \Delta \Rightarrow \primTyArity(\tyPrimNm{X})$ such that $\Delta
-\vdash \sigma \equiv \sigma' \Rightarrow \primTyArity(\tyPrimNm{X})$, % FIXME: make sure the notation for ss equality is consistent
-then $\rho\ \tyPrimNm{X}\ \sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
+$\Delta \vdash \sigma \Rightarrow \primTyArity(\tyPrimNm{X})$ and
+$\Delta \vdash \sigma' \Rightarrow \primTyArity(\tyPrimNm{X})$ such
+that $\Delta \vdash \sigma \equiv \sigma' \Rightarrow
+\primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\ \sigma = \rho\
+\tyPrimNm{X}\ \sigma'$.
 
@@
 $\rho \in \mathrm{RelEnv}(\Delta')$ and a simultaneous substitution
-$\sigma : \Delta \Rightarrow \Delta'$, we can derive the composed
+$\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
 relational environment (with an abuse of notation) $\rho \circ \sigma
 \in \mathrm{RelEnv}(\Delta)$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
-\sigma' = \rho\ \tyPrimNm{X}\ (\sigma \circ \sigma')$.
+\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$. \fixme{Check this is the right way round}
 
@@
 $\mathrm{RelEnv}(\Delta)$. Depending on the primitive operations that
-we assume for our system, we can have different sets of relational
+we assume for our system, we may have different sets of relational
 environments that enforce different invariants. For the general
@@
 \item Closure under simultaneous substitution of index terms: if $\rho
-  \in \relEnv{E}(\Delta)$ and $\sigma : \Delta' \Rightarrow \Delta$,
-  then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
+  \in \relEnv{E}(\Delta)$ and $\Delta' \vdash \sigma \Rightarrow
+  \Delta$, then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
 \item If we have a pair of relation environments $\rho_1 \in
@@
 
-A relational environment $\rho \in \relEnv{E}(\Delta)$ is extended to
-a relation interpretation of any type $\Delta \vdash A \isType$ by
-induction on its derivation: % Thus, we define a relation
-% $\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-indexed
-% type.
+We assign a relational interpretation to all well-indexed types
+$\Delta \vdash A \isType$ by induction on their derivations,
+parameterised by relational environments $\rho \in
+\relEnv{E}(\Delta)$:
 \begin{eqnarray*}
@@
 \autoref{sec:index-erasure-semantics}. Our main result
-(\thmref{thm:abstraction}) is that the index-erasure semantics is
+(\thmref{thm:abstraction}) is that the index-erasure semantics of
 every well-typed program is related to itself in the relational
 interpretation of its type: this is the abstraction theorem for every
-calculus in our general framework. In the next section we will use
-this theorem to derive several interesting results about the programs
-expressible in instances of our general framework.
+calculus in our general framework. In the \fixme{next section} we will
+use this theorem to derive several interesting results about the
+programs expressible in instances of our general framework.
 
@@
 \end{mathpar}
-Substitution by a simultaneous substitution extends to typing contexts
-by performing substitution pointwise on each individual type.
+Application of simultaneous substitutions extends to typing contexts
+by applying the simultaneous substitution to each type.
 
@@
 respect to an index context $\Delta$ and a type context $\Delta \vdash
-\Gamma \isCtxt$. The judgement $\Delta; \Gamma \vdash M : A$ is
-defined in \autoref{fig:programs}. The equational theory on types is
+\Gamma \isCtxt$. The judgment $\Delta; \Gamma \vdash M : A$ is defined
+in \autoref{fig:programs}. The equational theory on types is
 incorporated into the type system via the rule \TirName{TyEq}, which

### 2012-06-05 paper/geometry-twocolumn.tex
Redo the geometry section to just cover the origin invariance

@@
 % rather than their actual representation as tuples of numbers.
+
+\textbf{FIXME:} talk about the standard way of separating the two
+concepts using abstract types.
 
@@
 
-FIXME: vectors are not invariant under change of origin, so they are
-represented by the relation $R_0$, which is just equality. Bxut they,
-and points are invariant under change of basis. However, as we shall
-see below, not all operations are invariant under all changes of
-basis. In particular the dot product is only invariant under
-orthogonal transformations.
+% FIXME: vectors are not invariant under change of origin, so they are
+% represented by the relation $R_0$, which is just equality. But they,
+% and points are invariant under change of basis. However, as we shall
+% see below, not all operations are invariant under all changes of
+% basis. In particular the dot product is only invariant under
+% orthogonal transformations.
 
@@
 
-In the last section, we saw how Reynolds' notion of preservation of
-relations can be used to define when it is the case that programs
-manipulating representatives of points and vectors make geometric
-sense. Reynolds further insight was to see how preservation of
-relations can automatically be established by following a typing
-discipline. In Reynolds' case, the typing discipline was the
-polymorphic $\lambda$-calculus where new types can be constructed by
-universal quantification over all types. Universal type quantification
-is commonly written $\forall \alpha. A$, which denotes a type that may
-be instantiated with any other type $B$ by replacing the type variable
+We have seen how Reynolds' notion of preservation of relations can be
+used to define when it is the case that programs manipulating
+representatives of points and vectors make geometric sense. Reynolds
+further insight was to see how preservation of relations can
+automatically be established by following a typing discipline. In
+Reynolds' case, the typing discipline was the polymorphic
+$\lambda$-calculus where new types can be constructed by universal
+quantification over all types. Universal type quantification is
+commonly written $\forall \alpha. A$, which denotes a type that may be
+instantiated with any other type $B$ by replacing the type variable
 $\alpha$ with $B$ in $A$.
@@
 the previous section (e.g., \statementref{eq:f-invariant-frame}), we
-did not quantify over all relations. Instead, we quantified over all choices of origin
+did not quantify over all relations. Instead, we quantified over all
+choices of origin and used the choice to select a relation from the
+family $\{R_{\vec{o}}\}$. Driven by this observation, we introduce
+quantification over choice of origin into the language of types.  We
+use the notation $\forall t \mathord: T_2. A$ for quantification over
+all 2-dimensional translations (i.e.~choices of origin) $t$. We refer
+to $T_2$ as the \emph{sort} of $t$.
 
-elements of a geometric group such as $\GA(2)$ or $\GL(2)$, and used
-those elements to select relations from the families $R_{\texttt{pt}}$
-and $R_{\texttt{vec}}$. Driven by this observation, we introduce
-quantification over elements of geometric groups into the language of
-types. We use the notation $\forall X \mathord: G. A$ for
-quantification over elements of geometric groups, where $G$ is the
-name of some geometric group, and $X$ is a variable name standing for
-an element of $G$. We refer to $G$ as the \emph{sort} of $X$. As we
-proceed, we will introduce the sorts representing geometric groups
-that we will be using in addition to $\GA(2)$ and $\GL(2)$.
+% As we proceed, we will introduce the sorts representing geometric
+% groups that we will be using in addition to $\GA(2)$ and $\GL(2)$.
 
-The sorts $\GA(2)$ and $\GL(2)$ represent groups, so we can combine
-their elements using the usual group operations. We write operations
-multiplicatively, using juxtaposition ($e_1e_2$) for multiplication,
-$e^{-1}$ for inverse and $1$ for the unit. We overload the same
-notation to stand for the group operations in both $\GA(2)$ and
-$\GL(2)$. We also regard expressions built from variables and the
-group operations up to the group axioms. So, for example,
-$F_1(F_2F_3)$ and $(F_1F_2)F_3$ are to be regarded as equivalent.
+The sort $T_2$ represents an abelian group, so we can combine its
+elements using the usual group operations. We write operations
+additively, using $e_1 + e_2$ for the group operation, $-e$ for
+inverse and $0$ for the unit.  We also regard expressions built from
+variables and the group operations up to the group axioms. So, for
+example, $e_1 + (e_2 + e_3)$ and $(e_1 + e_2) + e_3$ are to be
+regarded as equivalent.
 
@@
 relational interpretation of tuple types.) Therefore, we introduce a
-new type parameterised by expressions $e$ of sort $\GA(2)$ to
-represent points (and later, vectors): $\mathsf{vec}\langle e
-\rangle$. We have used the name $\mathsf{vec}$, even though we have
-taken pains to separate geometric points and vectors, to recall the
-computer science notion of vector as a sequence of values of
-homogeneous type which has a known length (in this case two).
+new type parameterised by expressions $e$ of sort $T_2$ to represent
+points (and later, vectors): $\mathsf{vec}\langle e \rangle$. We have
+used the name $\mathsf{vec}$, even though we have taken pains to
+separate geometric points and vectors, to recall the computer science
+notion of vector as a sequence of values of homogeneous type which has
+a known length (in this case two).
 
@@
 
-Equipped with quantification over the group $\GA(2)$ of invertible
-affine maps, and the parameterised types $\mathsf{vec}\langle \cdot
-\rangle$, we can now express geometric invariance properties as
-types. In the previous section, we gave examples of functions $f :
-\mathbb{R}^2 \times \mathbb{R}^2 \to \mathbb{R}^2$ that have the
-property that they preserve all invertible affine maps. We can now
-express this property as a type:
+Equipped with quantification over the group $T_2$ of invertible affine
+maps, and the parameterised types $\mathsf{vec}\langle \cdot \rangle$,
+we can now express origin invariance properties as types. In the
+previous section, we discussed functions $f : \mathbb{R}^2 \times
+\mathbb{R}^2 \to \mathbb{R}^2$ that have the property that they
+preserve all changes of origin. We can now express this property as a
+type:
 \begin{displaymath}
-  \mathrm{f} : \forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \times \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle
+  \mathrm{f} : \forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \times \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle
 \end{displaymath}
-We use the variable name $F$ for elements of $\GA(2)$ to suggest the
-intuition that this typing indicates that $\mathrm{f}$ takes a pair of
-points in the $F$rame $F$ and returns a point in the same frame.
+FIXME: forward references...
 
-\subsection{Affine Combination of Points}
-\label{sec:affine-combination}
+% \subsection{Affine Combination of Points}
+% \label{sec:affine-combination}
 
-As we stated in the previous section, the primitive operation of
-affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
-\lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
-affine combination can be interpreted as interpolation between the
-points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
-combination of points to our calculus, with the following typing and
-intended denotation.
-\begin{displaymath}
-  \begin{array}{l}
-    \mathrm{affineCombination} :\\
-    \hspace{1.5cm}\forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \\
-    \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ t\ (x_2,y_2) = \\
-    \hspace{3cm}((1-t)x_1 + tx_2, (1-t)y_1 + ty_2)
-  \end{array}
-\end{displaymath}
-By defining the primitive function $\mathrm{affineCombination}$ to
-take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
-we are taking the affine combination of two representatives of points,
-and not just an arbitrary linear combination. Note that the intended
-denotation in the second line does not take any argument corresponding
-to the universally quantified $F$ in the type. In our calculus,
-geometric group expressions are erased at runtime.
+% As we stated in the previous section, the primitive operation of
+% affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
+% \lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
+% affine combination can be interpreted as interpolation between the
+% points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
+% combination of points to our calculus, with the following typing and
+% intended denotation.
+% \begin{displaymath}
+%   \begin{array}{l}
+%     \mathrm{affineCombination} :\\
+%     \hspace{1.5cm}\forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \\
+%     \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ t\ (x_2,y_2) = \\
+%     \hspace{3cm}((1-t)x_1 + tx_2, (1-t)y_1 + ty_2)
+%   \end{array}
+% \end{displaymath}
+% By defining the primitive function $\mathrm{affineCombination}$ to
+% take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
+% we are taking the affine combination of two representatives of points,
+% and not just an arbitrary linear combination. Note that the intended
+% denotation in the second line does not take any argument corresponding
+% to the universally quantified $F$ in the type. In our calculus,
+% geometric group expressions are erased at runtime.
 
-% In Linear Algebra notation, we could have written the right-hand side
-% as $(1-t)\vec{v_1} + t\vec{v_2}$, where $\vec{v_1} = (x_1,y_1)$ and
-% $\vec{v_2} = (x_2,y_2)$.
+% % In Linear Algebra notation, we could have written the right-hand side
+% % as $(1-t)\vec{v_1} + t\vec{v_2}$, where $\vec{v_1} = (x_1,y_1)$ and
+% % $\vec{v_2} = (x_2,y_2)$.
 
-\begin{example}
-  With just the single primitive operation of affine combination, we
-  can already express some useful programs. For example, the
-  evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves with
-  two endpoints and a single control point) can be expressed using the
-  affine combination primitive and three steps of De Casteljau's
-  algorithm:
-  \begin{displaymath}
-    \begin{array}{l}
-      \mathrm{quadraticB\acute{e}zier} :\\
-      \hspace{0.5cm}\forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \\
-      \mathrm{quadraticB\acute{e}zier}\ [F]\ p_0\ p_1\ p_2\ t = \\
-      \quad
-      \begin{array}{l@{\hspace{-0.002cm}}l}
-        \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ t\ p_1) \\
-        & t \\
-        & (\mathrm{affineCombination}\ [F]\ p_1\ t\ p_2)
-      \end{array}
-    \end{array}
-  \end{displaymath}
-  For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
-  application $\mathrm{quadraticB\acute{e}zier}\ p_0\ p_1\ p_2\ t$ for
-  $t \in [0,1]$ gives us the point on the curve at ``time'' $t$. Note
-  that the type of $\mathrm{quadraticB\acute{e}zier}$ tells us
-  immediately that it preserves all invertible affine maps.
-  %FIXME: forward ref to ``free theorem''.
-\end{example}
+% \begin{example}
+%   With just the single primitive operation of affine combination, we
+%   can already express some useful programs. For example, the
+%   evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves with
+%   two endpoints and a single control point) can be expressed using the
+%   affine combination primitive and three steps of De Casteljau's
+%   algorithm:
+%   \begin{displaymath}
+%     \begin{array}{l}
+%       \mathrm{quadraticB\acute{e}zier} :\\
+%       \hspace{0.5cm}\forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \\
+%       \mathrm{quadraticB\acute{e}zier}\ [F]\ p_0\ p_1\ p_2\ t = \\
+%       \quad
+%       \begin{array}{l@{\hspace{-0.002cm}}l}
+%         \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ t\ p_1) \\
+%         & t \\
+%         & (\mathrm{affineCombination}\ [F]\ p_1\ t\ p_2)
+%       \end{array}
+%     \end{array}
+%   \end{displaymath}
+%   For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
+%   application $\mathrm{quadraticB\acute{e}zier}\ p_0\ p_1\ p_2\ t$ for
+%   $t \in [0,1]$ gives us the point on the curve at ``time'' $t$. Note
+%   that the type of $\mathrm{quadraticB\acute{e}zier}$ tells us
+%   immediately that it preserves all invertible affine maps.
+%   %FIXME: forward ref to ``free theorem''.
+% \end{example}
 
@@
 
-Affine combination only provides us with an operation on points. For
-the purposes of computational geometry, we are also interested in
-operations on the vectors representing offsets between points. We now
-examine the correct types to assign to the vector space operations of
-addition of vectors, negation of vectors, multiplication by a scalar
-and the zero vector.
+\textbf{FIXME}Affine combination only provides us with an operation on
+points. For the purposes of computational geometry, we are also
+interested in operations on the vectors representing offsets between
+points. We now examine the correct types to assign to the vector space
+operations of addition of vectors, negation of vectors, multiplication
+by a scalar and the zero vector.
 
-In our discussion in the previous section, we observed that vector
-operations are those that are preserved by change-of-basis operations
--- i.e.~under the action of elements of the group $\GL(2)$ of
-invertible linear maps. Since $\GL(2)$ is a subgroup of $\GA(2)$ we
-may form the type $\mathsf{vec}\langle e \rangle$ when $e$ is of sort
-$\GL(2)$ as well as $\GA(2)$. The obvious type assignment for vector
-addition is now the following:
+In our discussion in the previous section \textbf{?FIXME?}, we
+observed that vector operations are not invariant under change of
+origin. The obvious type assignment for vector addition is therefore:
 \begin{displaymath}
-  (+) : \forall B\mathord:\GL(2).\ \mathsf{vec}\langle B \rangle \to \mathsf{vec}\langle B \rangle \to \mathsf{vec}\langle B \rangle
+  (+) : \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle
 \end{displaymath}
@@
 when not appearing in infix position, following the Haskell
-syntax.) The use of $B$ for names of elements of $\GL(2)$ is meant to
-suggest that the type $\mathsf{vec}\langle B \rangle$ denotes vectors
-with respect to the $B$asis $B$.
+syntax.)
 
 However, it is possible to give vector addition a more precise type
-that describes its effect on the chosen origin as well as its
-invariance under change of basis. We now introduce a new sort
-$\Transl(2)$ of translations. Translations form an abelian group,
-which we write additively, using $+$,$-$ and $0$, and we use lower
-case $t$, $t_1$, etc.~for $t$ranslation variables. Invertible affine
-maps are constructed from invertible linear maps and translations, so
-given a $B : \GL(2)$ and a $t : \Transl(2)$, we can form an invertible
-affine map $B \ltimes t : \GA(2)$. Conversely, given $F : \GA(2)$, we
-have an invertible linear map $\pi_L(F) : \GL(2)$, and a translation
-$\pi_T(F) : \Transl(2)$. Thinking of elements of $\GA(2)$ as frames,
-$\pi_L$ returns the basis of the frame, and $\pi_T$ returns the
-origin.
-
-We can now give vector addition a more general type, which
-demonstrates how adding vectors has an effect on translations:
+that describes its effect on the chosen origin. 
+% We now introduce a new
+% sort $\Transl(2)$ of translations. Translations form an abelian group,
+% which we write additively, using $+$,$-$ and $0$, and we use lower
+% case $t$, $t_1$, etc.~for $t$ranslation variables. Invertible affine
+% maps are constructed from invertible linear maps and translations, so
+% given a $B : \GL(2)$ and a $t : \Transl(2)$, we can form an invertible
+% affine map $B \ltimes t : \GA(2)$. Conversely, given $F : \GA(2)$, we
+% have an invertible linear map $\pi_L(F) : \GL(2)$, and a translation
+% $\pi_T(F) : \Transl(2)$. Thinking of elements of $\GA(2)$ as frames,
+% $\pi_L$ returns the basis of the frame, and $\pi_T$ returns the
+% origin.
+%
+% We can now give vector addition a more general type, which
+% demonstrates how adding vectors has an effect on translations:
 \begin{displaymath}
-  \begin{array}{l}
-    (+) : \forall B \mathord: \GL(2), t_1, t_2 \mathord: T(2).\ \\
-    \quad\quad\quad\quad\mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2 \rangle \to \mathsf{vec}\langle B \ltimes (t_1 + t_2)\rangle
-\end{array}
+  (+) : \forall t_1, t_2 \mathord: T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2 \rangle \to \mathsf{vec}\langle t_1 + t_2 \rangle
 \end{displaymath}
 We can also negate vectors, yielding a vector which points in the
-opposite direction. Negation is invariant under invertible linear maps
-but negates translations:
+opposite direction. Negation negates translations:
 \begin{displaymath}
-  \mathrm{negate} : \forall B \mathord: \GL(2), t \mathord: T(2).\ \mathsf{vec}\langle B \ltimes t\rangle \to \mathsf{vec}\langle B\ltimes(-t)\rangle
+  \mathrm{negate} : \forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle -t \rangle
 \end{displaymath}
-With addition and negation of vectors, we can define the derived
+With the primitive operations of addition and negation of vectors, we can define the derived
 operation of vector subtraction with the following type:
@@
   \begin{array}{l}
-    (-) : \forall B \mathord:\GL(2), t_1,t_2 \mathord:T(2).\ \\
-    \hspace{1cm}\mathsf{vec}\langle B \ltimes t_1\rangle \to \mathsf{vec}\langle B \ltimes t_2\rangle \to \mathsf{vec}\langle B \ltimes (t_1 - t_2)\rangle \\
+    (-) : \forall t_1,t_2 \mathord:T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2\rangle \to \mathsf{vec}\langle t_1 - t_2 \rangle \\
     (-)\ [B]\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
@@
 
-Given two points in the same frame---i.e.~two values of type
-$\mathsf{vec}\langle F \rangle$---we can compute their offset, which
-is a vector expressed with respect to the basis part of $F$, as a
-special case of vector subtraction:
+Given two points that have been constrained to be invariant with
+respect to the same change of origin---i.e.~two values of type
+$\mathsf{vec}\langle t \rangle$---we can compute their offset, which
+is a vector expressed with respect to the null change of origin. The
+offset operation is a special case of vector subtraction:
 \begin{displaymath}
   \begin{array}{l}
-    \mathrm{offset} : \forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle \pi_L(F) \ltimes 0 \rangle \\
-    \mathrm{offset}\ [F]\ p_1\ p_2 = p_1 - p_2
+    \mathrm{offset} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle 0 \rangle \\
+    \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
   \end{array}
 \end{displaymath}
-The subtraction operation in the body of $\mathrm{offset}$ is
-instantiated with $B = \pi_L(F)$, $t_1 = \pi_T(F)$ and $t_2 =
-\pi_T(F)$. The translation $\pi_T(F)$ cancels itself out, and we are
-left with a vector in the basis $\pi_L(F)$ of the frame $F$.
-
 Similarly, the type of the vector addition operation can be
@@
   \begin{array}{l}
-    \mathrm{moveBy} : \forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle \pi_L(F) \ltimes 0 \rangle \to \mathsf{vec}\langle F \rangle \\
-    \mathrm{moveBy}\ [F]\ p\ v = p + v
+    \mathrm{moveBy} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle t \rangle \\
+    \mathrm{moveBy}\ [t]\ p\ v = p + v
   \end{array}
 \end{displaymath}
-The addition operation in the body of $\mathrm{moveBy}$ is
-instantiated with $B = \pi_L(F)$, $t_1 = \pi_T(F)$ and $t_2 =
-0$. Since $t_1 + t_2 = \pi_T(F) + 0 = \pi_T(F)$ in this case, we are
-left with a value of type $\mathsf{vec}\langle \pi_L(F) \ltimes
-\pi_T(F) \rangle$, which is equivalent to the type
-$\mathsf{vec}\langle F \rangle$.
-
 The types we assign to the remaining vector space primitives, the zero
@@
 effect on translations%\footnote{but they could...}
-. The zero vector has the following polymorphic type, indicating that
-for any basis $B$, there is a zero vector.
+. The zero vector has the following type.
 \begin{displaymath}
-  \mathrm{0} : \forall B \mathord: \GL(2).\ \mathsf{vec}\langle B \ltimes 0 \rangle
+  \mathrm{0} : \mathsf{vec}\langle 0 \rangle
 \end{displaymath}
-Intuitively, there cannot be a zero vector for any frame $F$, because
-this would entail picking a privileged origin. For multiplication of a
-vector by a scalar, we assign the following type:
+Intuitively, there cannot be a zero vector invariant under any change
+of origin, because this would entail picking a privileged origin. For
+multiplication of a vector by a scalar, we assign the following type,
+which is again not invariant under change of origin.
 \begin{displaymath}
-  (*) : \forall B \mathord: \GL(2).\ \mathsf{real} \to \mathsf{vec}\langle B \ltimes 0 \rangle \to \mathsf{vec}\langle B \ltimes 0 \rangle
+  (*) : \mathsf{real} \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle
 \end{displaymath}
-This type indicates that scaling a vector is preserved by change of
-basis.
 
 \begin{example}
-  The vector space primitives we have defined in this section, along
-  with the geometric invariance properties that follow from their
-  types, allow us to establish a useful type isomorphism. Consider
+  The vector space operators and the properties that follow from their
+  types allow us to establish a useful type isomorphism. Consider
   functions with types conforming to the following schema:
   \begin{displaymath}
-    \tau_{n,A} \isDefinedAs \forall F\mathord:\GA(2).\ \mathsf{vec}\langle F \rangle^{n+1} \to A[F]
+    \tau_{n} \isDefinedAs \forall t\mathord:T_2.\ \mathsf{vec}\langle t \rangle^{n+1} \to \mathsf{real}
   \end{displaymath}
@@
   denotes the $n$-ary tupling of $A$ with itself: i.e.~$A \times
-  ... \times A$, with $A$ repeated $n$ times. The notation $A[-]$
-  denotes a type with a ``hole'' that can be filled with an
-  expression. In this case, we require that $A[e]$ be a well-formed
-  type when $e$ has sort $\GA(2)$.
+  ... \times A$, with $A$ repeated $n$ times.
+% The notation $A[-]$
+%   denotes a type with a ``hole'' that can be filled with an
+%   expression. In this case, we require that $A[e]$ be a well-formed
+%   type when $e$ has sort $\GA(2)$.
 
-  Just by looking at the types $\tau_{n,A}$, we know that their
-  inhabitants will be invariant under change of frame, due to the
-  quantification over all $F$ in $\GA(2)$. Therefore, we may as well
-  pick one of the input points as the origin and assume that all the
-  other points are defined with respect to this chosen origin. This is
+  Just by looking at the types $\tau_{n}$, we know that their
+  inhabitants will be invariant under change of origin, due to the
+  quantification over all $t$ in $T_2$. Therefore, we may as well pick
+  one of the input points as the origin and assume that all the other
+  points are defined with respect to this chosen origin. This is
   similar to the common mathematical step of stating that, ``without
@@
 
-  The types $\tau_{n,A}$ are isomorphic to the corresponding types
-  $\sigma_{n,A}$, defined below. Inhabitants of the types
-  $\sigma_{n,A}$ take one fewer points (now seen as vector offsets
+  The types $\tau_{n}$ are isomorphic to the corresponding types
+  $\sigma_{n}$, defined below. Inhabitants of the types
+  $\sigma_{n}$ take one fewer points (now seen as vector offsets
   from some origin), but are invariant only under change of basis.
   \begin{displaymath}
-    \sigma_{n,A} \isDefinedAs \forall B\mathord:\GL(2).\ \mathsf{vec}\langle B \rangle^n \to A[B \ltimes 0]
+    \sigma_{n} \isDefinedAs \mathsf{vec}\langle 0 \rangle^n \to \mathsf{real}
   \end{displaymath}
-  The isomorphism between these two families of types can be witnessed
-  by the following two functions.
+  We demonstrate this isomorphism formally below in
+  \autoref{sec:FIXME}.
+
+  % The isomorphism between these two families of types can be witnessed
+  % by the following two functions.
+  % \begin{displaymath}
+  %   \begin{array}{l}
+  %     i_{n,A} : \tau_{n,A} \to \sigma_{n,A} \\
+  %     i_{n,A}\ f\ [B]\ (p_1, p_2, ..., p_n) = f\ [B \ltimes 0]\ (0, p_1, ..., p_n) \\
+  %     \\
+  %     i^{-1}_{n,A} : \sigma_{n,A} \to \tau_{n,A} \\
+  %     i^{-1}_{n,A}\ g\ [F]\ (p_0, ..., p_n) = g\ [\pi_L(F)]\ (p_1-p_0, ..., p_n-p_0)
+  %   \end{array}
+  % \end{displaymath}
+  % The direction defined by the function $i_{n,A}$ treats the inputs
+  % $p_1,...,p_n$ as vector offsets from the origin $0$. The direction
+  % witnessed by $i^{-1}_{n,A}$ picks the first point to act as the
+  % origin, and uses the operation of vector subtraction we defined
+  % above to turn each of the other points into offsets from this
+  % point. Note that this isomorphism is not unique: for each $n$ we can
+  % pick any of the $n+1$ inputs to act as the distinguished
+  % origin. Thus the families of types $\tau_{n,A}$ and $\sigma_{n,A}$
+  % are isomorphic, but not uniquely isomorphic.
+\end{example}
+
+\begin{example}
+  An inexpressible type
   \begin{displaymath}
-    \begin{array}{l}
-      i_{n,A} : \tau_{n,A} \to \sigma_{n,A} \\
-      i_{n,A}\ f\ [B]\ (p_1, p_2, ..., p_n) = f\ [B \ltimes 0]\ (0, p_1, ..., p_n) \\
-      \\
-      i^{-1}_{n,A} : \sigma_{n,A} \to \tau_{n,A} \\
-      i^{-1}_{n,A}\ g\ [F]\ (p_0, ..., p_n) = g\ [\pi_L(F)]\ (p_1-p_0, ..., p_n-p_0)
-    \end{array}
+    \forall t \mathord: T_2.\ \vec{t + t} \to \vec{t}
   \end{displaymath}
-  The direction defined by the function $i_{n,A}$ treats the inputs
-  $p_1,...,p_n$ as vector offsets from the origin $0$. The direction
-  witnessed by $i^{-1}_{n,A}$ picks the first point to act as the
-  origin, and uses the operation of vector subtraction we defined
-  above to turn each of the other points into offsets from this
-  point. Note that this isomorphism is not unique: for each $n$ we can
-  pick any of the $n+1$ inputs to act as the distinguished
-  origin. Thus the families of types $\tau_{n,A}$ and $\sigma_{n,A}$
-  are isomorphic, but not uniquely isomorphic.
+  FIXME: explain that it is possible to show that this type is
+  uninhabited with the given operations. We show this in general 
 \end{example}
+
+\subsection{Generalising from Origin Invariance}
+
+Points for consideration in terms of language design:
+\begin{itemize}
+\item Knowledge of the algebraic laws by the type checker are
+  essential for type-checking the examples above
+\item The relational interpretation gives us 
+\end{itemize}
 

### 2012-06-06 paper/geometry-twocolumn.tex
More updates to the motivating examples section

@@
 
-\subsection{2-Dimensional Geometry and Invariance}
+We motivate this paper's goal of investigating algebraically-indexed
+types and their relational interpretations by presenting several
+examples. Our main set of examples comprise a novel type system for
+writing programs that manipulate geometric data, while maintaining
+geometric invariants through the types. \fixme{and the information
+  flow examples}
 
-When writing programs that manipulate $n$-dimensional geometric data,
-the basic data structure is the $n$-tuple of real numbers. In the
-two-dimensional case, tuples $\vec{v} = (x,y)$ can be called upon to
-represent both \emph{points} -- offsets from some origin -- and
-\emph{vectors} -- offsets in their own right. The notions of point and
-vector are very different despite their common representation. By
-representing a point concretely as an offset from some origin, we
-inadvertently fix that origin as privileged.
-% However, as made clear by Gallieo, many of the laws of physics are
-% invariant under change of the frame of reference: if we change where
-% we think the origin is, then we do not change the fundamental laws.
-Putting this in concrete terms, if we have two frames of reference,
-one with $(0,0)$ as the origin, and one with $(10,20)$ as the origin,
-then the tuple $(1,1)$ with respect to the first frame and $(11,21)$
-with respect to the second frame represent the \emph{same} point. It
-is merely an artifact of our representation of points as pairs of
-numbers that these appear to be different. When we write programs that
-manipulate data that represents points, we ought to follow the general
-principle that our programs are invariant with respect to changes in
-the choice of origin. On the other hand, vectors are not intended to
-be regarded as invariant under change of choice of origin. Vectors
-represent offsets, or movements, and the vector $(0,0)$ always
-represents the zero offset.
+\subsection{2-Dimensional Geometry and Origin Invariance}
+
+When writing programs that manipulate geometric data, the basic data
+structure is the $n$-tuple of real numbers. In the 2-dimensional case,
+tuples $\vec{v} = (x,y)$ may be called upon to represent both
+\emph{points} -- offsets from some origin -- and \emph{vectors} --
+offsets in their own right. The notions of point and vector are very
+different despite their common representation. By representing a point
+concretely as an offset from an origin, we inadvertently fix that
+origin as privileged.
+
+Putting this in concrete terms, if we have two origins, $(0,0)$ and
+$(10,20)$, then the tuple $(1,1)$ with respect to the first origin and
+$(11,21)$ with respect to the second origin represent the \emph{same}
+point. It is merely an artifact of our representation of points as
+pairs of numbers that these appear to be different. When we write
+programs that manipulate data that represents points, we ought to
+follow the general principle that our programs are invariant with
+respect to changes in the choice of origin. On the other hand, vectors
+are not intended to be regarded as invariant under change of choice of
+origin. Vectors represent offsets, or movements, and the vector
+$(0,0)$ always represents the zero offset.
 % However, they are intended to be invariant under change of basis: the
@@
 
-\textbf{FIXME:} talk about the standard way of separating the two
-concepts using abstract types.
-
 Invariance under change of representation immediately recalls
 Reynolds' fable concerning two professors teaching the theory of
-complex numbers to students \cite{reynolds83types}. One professor
-chooses rectangular coordinates ($x + iy$), while the other chooses
-polar coordinates ($\alpha\cos\theta + i\alpha\sin\theta$). Happily,
-after learning the basic operations on complex numbers in the two
+complex numbers \cite{reynolds83types}. One professor chooses
+rectangular coordinates ($x + iy$), while the other chooses polar
+coordinates ($\alpha\cos\theta + i\alpha\sin\theta$). Happily, after
+learning the basic operations on complex numbers in the two
 representations, the two classes can interact because the theory of
@@
 above their specific representation should preserve the relation
-$R$. Concretely, let $f$ be a program taking complex numbers to
+$R$. For example, if $f$ be a program taking complex numbers to
 complex numbers that is intended to respect invariance under change of
@@
 We now apply Reynolds' relational approach to defining invariance
-under changes of representation to our geometric setting. Following
+under changes of representation to the geometric setting. Following
 Reynolds, we express the fact that two tuples in $\mathbb{R}^2$ are
@@
 
-We have seen how Reynolds' notion of preservation of relations can be
-used to define when it is the case that programs manipulating
-representatives of points and vectors make geometric sense. Reynolds
-further insight was to see how preservation of relations can
-automatically be established by following a typing discipline. In
-Reynolds' case, the typing discipline was the polymorphic
-$\lambda$-calculus where new types can be constructed by universal
-quantification over all types. Universal type quantification is
-commonly written $\forall \alpha. A$, which denotes a type that may be
-instantiated with any other type $B$ by replacing the type variable
-$\alpha$ with $B$ in $A$.
+Reynolds further insight was to see how preservation of relations by
+the denotational interpretations of programs can automatically be
+established by following a typing discipline. In Reynolds' case, the
+typing discipline was the polymorphic $\lambda$-calculus where new
+types can be constructed by universal quantification over all
+types. Universal type quantification is commonly written $\forall
+\alpha. A$, which denotes a type that may be instantiated with any
+other type $B$ by replacing the type variable $\alpha$ with $B$ in
+$A$.
 
@@
 
-Equipped with quantification over the group $T_2$ of invertible affine
-maps, and the parameterised types $\mathsf{vec}\langle \cdot \rangle$,
-we can now express origin invariance properties as types. In the
-previous section, we discussed functions $f : \mathbb{R}^2 \times
+Equipped with quantification over the group $T_2$ of changes of
+origin, and the parameterised types $\mathsf{vec}\langle \cdot
+\rangle$, we can now express origin invariance properties as types. In
+the previous section, we discussed functions $f : \mathbb{R}^2 \times
 \mathbb{R}^2 \to \mathbb{R}^2$ that have the property that they
@@
 
-% \subsection{Affine Combination of Points}
-% \label{sec:affine-combination}
+\subsection{Affine and Vector Operations}
+\label{sec:affine-vector-ops}
 
-% As we stated in the previous section, the primitive operation of
-% affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
-% \lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
-% affine combination can be interpreted as interpolation between the
-% points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
-% combination of points to our calculus, with the following typing and
-% intended denotation.
-% \begin{displaymath}
-%   \begin{array}{l}
-%     \mathrm{affineCombination} :\\
-%     \hspace{1.5cm}\forall F \mathord: \GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \\
-%     \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ t\ (x_2,y_2) = \\
-%     \hspace{3cm}((1-t)x_1 + tx_2, (1-t)y_1 + ty_2)
-%   \end{array}
-% \end{displaymath}
-% By defining the primitive function $\mathrm{affineCombination}$ to
-% take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
-% we are taking the affine combination of two representatives of points,
-% and not just an arbitrary linear combination. Note that the intended
-% denotation in the second line does not take any argument corresponding
-% to the universally quantified $F$ in the type. In our calculus,
-% geometric group expressions are erased at runtime.
+As we stated in the previous section, the primitive operation of
+affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
+\lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
+affine combination can be interpreted as interpolation between the
+points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
+combination of points to our calculus, with the following typing and
+intended denotation:
+\begin{displaymath}
+  \begin{array}{l}
+    \mathrm{affineCombination} :\forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{real} \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \\
+    \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ r\ (x_2,y_2) = \\
+    \hspace{3cm}((1-r)x_1 + tx_2, (1-r)y_1 + ty_2)
+  \end{array}
+\end{displaymath}
+By defining the primitive function $\mathrm{affineCombination}$ to
+take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
+we are taking the affine combination of two representatives of points,
+and not just an arbitrary linear combination. The intended denotation
+in the second line does not take any argument corresponding to the
+universally quantified $t$ in the type. Type-level variables are
+erased at runtime.
 
-% % In Linear Algebra notation, we could have written the right-hand side
-% % as $(1-t)\vec{v_1} + t\vec{v_2}$, where $\vec{v_1} = (x_1,y_1)$ and
-% % $\vec{v_2} = (x_2,y_2)$.
+% In Linear Algebra notation, we could have written the right-hand side
+% as $(1-t)\vec{v_1} + t\vec{v_2}$, where $\vec{v_1} = (x_1,y_1)$ and
+% $\vec{v_2} = (x_2,y_2)$.
 
-% \begin{example}
-%   With just the single primitive operation of affine combination, we
-%   can already express some useful programs. For example, the
-%   evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves with
-%   two endpoints and a single control point) can be expressed using the
-%   affine combination primitive and three steps of De Casteljau's
-%   algorithm:
-%   \begin{displaymath}
-%     \begin{array}{l}
-%       \mathrm{quadraticB\acute{e}zier} :\\
-%       \hspace{0.5cm}\forall F \mathord:\GA(2).\ \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{vec}\langle F \rangle \to \mathsf{real} \to \mathsf{vec}\langle F \rangle \\
-%       \mathrm{quadraticB\acute{e}zier}\ [F]\ p_0\ p_1\ p_2\ t = \\
-%       \quad
-%       \begin{array}{l@{\hspace{-0.002cm}}l}
-%         \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ t\ p_1) \\
-%         & t \\
-%         & (\mathrm{affineCombination}\ [F]\ p_1\ t\ p_2)
-%       \end{array}
-%     \end{array}
-%   \end{displaymath}
-%   For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
-%   application $\mathrm{quadraticB\acute{e}zier}\ p_0\ p_1\ p_2\ t$ for
-%   $t \in [0,1]$ gives us the point on the curve at ``time'' $t$. Note
-%   that the type of $\mathrm{quadraticB\acute{e}zier}$ tells us
-%   immediately that it preserves all invertible affine maps.
-%   %FIXME: forward ref to ``free theorem''.
-% \end{example}
+\begin{example}
+  The evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves
+  with two endpoints and a single control point) can be expressed
+  using the affine combination primitive and three steps of De
+  Casteljau's algorithm:
+  \begin{displaymath}
+    \begin{array}{@{}l}
+      \mathrm{quadB\acute{e}zier} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{real} \to \mathsf{vec}\langle t \rangle \\
+      \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\
+      \quad
+      \begin{array}{l@{\hspace{-0.002cm}}l}
+        \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ s\ p_1) \\
+        & s \\
+        & (\mathrm{affineCombination}\ [F]\ p_1\ s\ p_2)
+      \end{array}
+    \end{array}
+  \end{displaymath}
+  For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
+  application $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ for $s
+  \in [0,1]$ gives us the point on the curve at ``time'' $s$. Note
+  that the type of $\mathrm{quadB\acute{e}zier}$ tells us immediately
+  that it preserves all changes of origin
+\end{example}
 
-\subsection{Vector Space Operations}
-\label{sec:vector-space-ops}
+Affine combination only provides us with an operation on points. We
+are also interested in operations on the vectors representing offsets
+between points. We now examine the correct types to assign to the
+vector space operations of addition of vectors, negation of vectors,
+multiplication by a scalar and the zero vector. The typings of these
+operations will make use of the abelian group structure of the sort of
+translations $T_2$.
 
-\textbf{FIXME}Affine combination only provides us with an operation on
-points. For the purposes of computational geometry, we are also
-interested in operations on the vectors representing offsets between
-points. We now examine the correct types to assign to the vector space
-operations of addition of vectors, negation of vectors, multiplication
-by a scalar and the zero vector.
-
-In our discussion in the previous section \textbf{?FIXME?}, we
-observed that vector operations are not invariant under change of
-origin. The obvious type assignment for vector addition is therefore:
+Vector operations are not invariant under change of origin. The
+obvious type assignment for vector addition is therefore:
 \begin{displaymath}
@@
 that describes its effect on the chosen origin. 
-% We now introduce a new
-% sort $\Transl(2)$ of translations. Translations form an abelian group,
-% which we write additively, using $+$,$-$ and $0$, and we use lower
-% case $t$, $t_1$, etc.~for $t$ranslation variables. Invertible affine
-% maps are constructed from invertible linear maps and translations, so
-% given a $B : \GL(2)$ and a $t : \Transl(2)$, we can form an invertible
-% affine map $B \ltimes t : \GA(2)$. Conversely, given $F : \GA(2)$, we
-% have an invertible linear map $\pi_L(F) : \GL(2)$, and a translation
-% $\pi_T(F) : \Transl(2)$. Thinking of elements of $\GA(2)$ as frames,
-% $\pi_L$ returns the basis of the frame, and $\pi_T$ returns the
-% origin.
-%
-% We can now give vector addition a more general type, which
-% demonstrates how adding vectors has an effect on translations:
 \begin{displaymath}
@@
 We can also negate vectors, yielding a vector which points in the
-opposite direction. Negation negates translations:
+opposite direction. Negation negates translation arguments:
 \begin{displaymath}
@@
 \end{displaymath}
-With the primitive operations of addition and negation of vectors, we can define the derived
-operation of vector subtraction with the following type:
+With the primitive operations of addition and negation of vectors, we
+can define the derived operation of subtraction:
 \begin{displaymath}
@@
     (-) : \forall t_1,t_2 \mathord:T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2\rangle \to \mathsf{vec}\langle t_1 - t_2 \rangle \\
-    (-)\ [B]\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
+    (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
   \end{array}
@@
 \end{displaymath}
-Similarly, the type of the vector addition operation can be
-specialised to the case of moving a point by an offset represented by
-a vector:
+The type of the vector addition operation can be specialised to the
+case of moving a point by a vector:
 \begin{displaymath}
@@
 effect on translations%\footnote{but they could...}
-. The zero vector has the following type.
+. The zero vector has the type:
 \begin{displaymath}
@@
   types allow us to establish a useful type isomorphism. Consider
-  functions with types conforming to the following schema:
+  functions with types following the schema:
   \begin{displaymath}
-    \tau_{n} \isDefinedAs \forall t\mathord:T_2.\ \mathsf{vec}\langle t \rangle^{n+1} \to \mathsf{real}
+    \tau_{n} \isDefinedAs \forall t\mathord:T_2.\ \underbrace{\mathsf{vec}\langle t \rangle \to ... \mathsf{vec}\langle t \rangle}_{n+1\textrm{ times}} \to \mathsf{real}
   \end{displaymath}
-  We take $n$ to be a natural number, and $A^n$ for some type $A$
-  denotes the $n$-ary tupling of $A$ with itself: i.e.~$A \times
-  ... \times A$, with $A$ repeated $n$ times.
-% The notation $A[-]$
-%   denotes a type with a ``hole'' that can be filled with an
-%   expression. In this case, we require that $A[e]$ be a well-formed
-%   type when $e$ has sort $\GA(2)$.
-
   Just by looking at the types $\tau_{n}$, we know that their
@@
   The types $\tau_{n}$ are isomorphic to the corresponding types
-  $\sigma_{n}$, defined below. Inhabitants of the types
-  $\sigma_{n}$ take one fewer points (now seen as vector offsets
-  from some origin), but are invariant only under change of basis.
+  $\sigma_{n}$:
   \begin{displaymath}
-    \sigma_{n} \isDefinedAs \mathsf{vec}\langle 0 \rangle^n \to \mathsf{real}
+    \sigma_{n} \isDefinedAs \underbrace{\mathsf{vec}\langle 0 \rangle \to ... \mathsf{vec}\langle 0 \rangle}_{n\textrm{ times}} \to \mathsf{real}
   \end{displaymath}
-  We demonstrate this isomorphism formally below in
-  \autoref{sec:FIXME}.
+  We demonstrate this isomorphism formally in \autoref{sec:FIXME}.
 
@@
 
-\subsection{Generalising from Origin Invariance}
+\subsection{Generalising Origin Invariance}
+\label{sec:motivation-generalising}
 
-Points for consideration in terms of language design:
-\begin{itemize}
-\item Knowledge of the algebraic laws by the type checker are
-  essential for type-checking the examples above
-\item The relational interpretation gives us 
-\end{itemize}
+\begin{enumerate}
+\item Basis invariance
+\item Orthogonal transform invariance for dot product
+\item Rotational invariance (the type of the cross-product)
+\item Scale invariance (units of measure)
+\end{enumerate}
+
+In some cases, we can use the invariance as a programming tool
+(e.g. type isomorphisms), but at others we can only use it from the
+outside.
+
+\subsection{Information Flow}
+\label{sec:information-flow}
+
+\begin{enumerate}
+\item Think of information flow as passing around unforgable tokens
+  \begin{enumerate}
+  \item Give an example (given a piece of data protected by a token
+    and a token, can we extract the data?)
+  \item Need to make sure that tokens are unforgable
+  \end{enumerate}
+\item Could either enforce unforgability of tokens by some dynamic
+  means, or by static type-based means
+  \begin{enumerate}
+  \item Leave the dynamic method as vague
+  \item Ingredients of the static method: run-time tokens are just
+    dummy values ($*$), the token type is parameterised by logical
+    formulae, and the relational meaning is derived from an assignment
+    of boolean values to atoms.
+  \item Explain this as invariance under changes of satisfaction of
+    the participants. Relate to descriptions of logics as rules that
+    don't care about truth, only its preservation.
+  \end{enumerate}
+\item Cite Sabelfeld and Sands: ``A PER Model of Secure Information
+  Flow in Sequential Programs''
+\end{enumerate}
 

### 2012-06-07 paper/geometry-twocolumn.tex
Substantial changes to the motivating examples section

@@
 tuples $\vec{v} = (x,y)$ may be called upon to represent both
-\emph{points} -- offsets from some origin -- and \emph{vectors} --
-offsets in their own right. The notions of point and vector are very
-different despite their common representation. By representing a point
-concretely as an offset from an origin, we inadvertently fix that
-origin as privileged.
+\emph{points}---offsets from some origin---and
+\emph{vectors}---offsets in their own right. Points and vectors are
+very different despite their common representation.
+
+Distinguishing between points and vectors is the key feature of
+\emph{affine geometry} (see, for example, Gallier's book
+\cite{gallier11geometric}, Chapter 2). Traditionally in computational
+geometry libraries, the distinction between points and vectors has
+either been left to the programmer to maintain by themselves, or has
+been enforced through the use of separate abstract types for points
+and vectors (one might say that the standard mathematical formulation
+of affine spaces uses the abstract types approach). Here, we
+investigate a more sophisticated approach using types indexed by
+change of origin transformations, and think of the difference between
+points and vectors as a change of data representation.
 
@@
 programs that manipulate data that represents points, we ought to
-follow the general principle that our programs are invariant with
-respect to changes in the choice of origin. On the other hand, vectors
-are not intended to be regarded as invariant under change of choice of
-origin. Vectors represent offsets, or movements, and the vector
-$(0,0)$ always represents the zero offset.
-% However, they are intended to be invariant under change of basis: the
-% vector $(2,3)$, with respect to the basis $(1,0),(0,1)$, and the
-% vector $(1,1)$, with respect to the basis $(2,0),(0,3)$, should be
-% regarded as the same vector. Invariance under change of basis also
-% applies to the representation of points. We therefore refine the
-% principle of invariance under change of origin for points to also
-% include invariant under change of basis for points and vectors.
-
-% A particular choice of origin and basis is called a \emph{frame}
-% (\cite{gallier11geometric}, Chapter 2). In general, the technique of
-% maintaining geometric invariants such as frame invariance has been
-% referred to as \emph{coordinate-free geometry}: we should aim to
-% think in terms of abstract concepts such as points and vectors,
-% rather than their actual representation as tuples of numbers.
+ensure that our programs are invariant with respect to changes in the
+choice of origin. On the other hand, vectors are not intended to be
+regarded as invariant under change of origin. Vectors represent
+offsets, or movements, and the vector $(0,0)$ always represents the
+zero offset.
 
@@
 representation. Reynolds formalises the idea of invariance with
-respect to changes of representation by using the notion of
-preservation of relations. If a binary relation $R$ relates two
-representations of a concept---for example the rectangular and polar
-representations of the same complex number are related---then a
-program that manipulates complex numbers at a level of abstraction
-above their specific representation should preserve the relation
-$R$. For example, if $f$ be a program taking complex numbers to
-complex numbers that is intended to respect invariance under change of
-representation. Then if $c$ is a complex number in rectangular form,
-and $c'$ is a complex number in polar form such that $R$ relates them,
-which we write as $(c,c') \in R$, then it should be the case that
-$(f(c), f(c')) \in R$.
+respect to changes of representation as preservation of relations. If
+a binary relation $R$ relates two representations of a concept---for
+example the rectangular and polar representations of the same complex
+number are related---then a program that manipulates complex numbers
+at a level of abstraction above their specific representation should
+preserve the relation $R$. For example, let $f$ be a program taking
+complex numbers to complex numbers that is intended to respect
+invariance under change of representation. Then if $c$ is a complex
+number in rectangular form, and $c'$ is a complex number in polar form
+such that $R$ relates them, which we write as $(c,c') \in R$, then it
+should be the case that $(f(c), f(c')) \in R$.
 
-We now apply Reynolds' relational approach to defining invariance
-under changes of representation to the geometric setting. Following
-Reynolds, we express the fact that two tuples in $\mathbb{R}^2$ are
-representatives of the same point under change of origin by defining a
-family of binary relations over $\mathbb{R}^2$. Our family of binary
-relations is parameterised by a particular choice of origin, and by
-quantifying over all choices of origin we will be able to ensure that
-our programs are invariant with respect to the particular choice of
-origin. We define a $\mathbb{R}^2$-parameterised family of binary
-relations $\{ R_{\vec{o}} \subseteq \mathbb{R}^2 \times \mathbb{R}^2
-\}_{\vec{o} \in \mathbb{R}^2}$ as follows:
+We now apply Reynolds' relational approach to the geometric
+setting. We define a family of binary relations on $\mathbb{R}^2$
+which is indexed by changes of origin. Changes of origin are
+represented by vectors in $\mathbb{R}^2$, and form a group under
+addition. To emphasise the group structure, and the intended meaning
+as translations, we write the collection of all change of origin
+offsets as $\Transl_2$. By quantifying over all changes of origin we
+will be able to ensure that our programs are invariant with respect to
+the particular choice of origin. The members of the
+$\Transl_2$-indexed family of binary relations $\{ R_{\vec{t}}
+\subseteq \mathbb{R}^2 \times \mathbb{R}^2 \}_{\vec{t} \in \Transl_2}$
+are defined as follows:
 \begin{displaymath}
-  R_{\vec{o}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{o} \}
+  R_{\vec{t}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{t} \}
 \end{displaymath}
@@
 to the choice of origin. We use Reynolds' idea of preservation of
-relations to formalise this: for any $\vec{o} \in \mathbb{R}^2$
-(i.e.~any choice of origin), the function $f$ should satisfy the
-following statement:
+relations to formalise this: for any $\vec{t} \in \Transl_2$ (i.e.~any
+change of origin), the function $f$ should satisfy the following
+statement:
 \begin{equation}\label{eq:f-preserve-rel-frame}
-  \forall (\vec{v_1},\vec{v'_1}) \in R_{\vec{o}}, (\vec{v_2},\vec{v'_2}) \in R_{\vec{o}}. (f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\vec{o}}
+  \forall (\vec{v_1},\vec{v'_1}) \in R_{\vec{t}}, (\vec{v_2},\vec{v'_2}) \in R_{\vec{t}}. (f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\vec{t}}
 \end{equation}
-Unfolding the definition of $R_{\vec{o}}$, this statement is
+Unfolding the definition of $R_{\vec{t}}$, this statement is
 equivalent to the following statement expressing invariance of $f$
-with respect to changes of origin: for all $\vec{o} \in \mathbb{R}^2$,
+with respect to changes of origin: for all $\vec{t} \in \Transl_2$:
 \begin{equation}\label{eq:f-invariant-frame}
-  \forall \vec{v_1}, \vec{v_2}.\ f(\vec{v_1} + \vec{o},\vec{v_2} + \vec{o}) = f(\vec{v_1},\vec{v_2}) + \vec{o}.
+  \forall \vec{v_1}, \vec{v_2}.\ f(\vec{v_1} + \vec{t},\vec{v_2} + \vec{t}) = f(\vec{v_1},\vec{v_2}) + \vec{t}.
 \end{equation}
 So Reynolds' preservation of relations, when instantiated with the
-family of relations $R$, yields exactly the geometric property of
-invariance under change of origin.
+family of relations $\{R_{\vec{t}}\}$, yields exactly the geometric
+property of invariance under change of origin.
 
@@
 
-Reynolds further insight was to see how preservation of relations by
+Reynolds' further insight was to see how preservation of relations by
 the denotational interpretations of programs can automatically be
@@
 did not quantify over all relations. Instead, we quantified over all
-choices of origin and used the choice to select a relation from the
-family $\{R_{\vec{o}}\}$. Driven by this observation, we introduce
-quantification over choice of origin into the language of types.  We
-use the notation $\forall t \mathord: T_2. A$ for quantification over
-all 2-dimensional translations (i.e.~choices of origin) $t$. We refer
-to $T_2$ as the \emph{sort} of $t$.
+changes of origin and used the choice to select a relation from the
+family $\{R_{\vec{t}}\}$. Driven by this observation, we introduce
+quantification over changes of origin into the language of types.  We
+use the notation $\forall t \mathord: \SynTransl_2. A$ for
+quantification over all 2-dimensional translations (i.e.~choices of
+origin) $t$. We refer to $\SynTransl_2$ as the \emph{sort} of
+$t$. Note that we have used a different font to distinguish the
+semantic group $\Transl_2$ from the syntactic sort $\SynTransl_2$.
 
-% As we proceed, we will introduce the sorts representing geometric
-% groups that we will be using in addition to $\GA(2)$ and $\GL(2)$.
-
-The sort $T_2$ represents an abelian group, so we can combine its
-elements using the usual group operations. We write operations
+The sort $\SynTransl_2$ represents an abelian group, so we can combine
+its elements using the usual group operations. We write operations
 additively, using $e_1 + e_2$ for the group operation, $-e$ for
 inverse and $0$ for the unit.  We also regard expressions built from
-variables and the group operations up to the group axioms. So, for
+variables and the group operations up to the abelian group axioms. For
 example, $e_1 + (e_2 + e_3)$ and $(e_1 + e_2) + e_3$ are to be
@@
 function types, $A \to B$, sum types $A + B$, unit type
-$\mathsf{Unit}$ and tuple types $A \times B$. We also assume a
-primitive type $\mathsf{real}$, used to represent scalars. As we noted
-at the start of the previous section, the central data structure in
-geometric applications is the tuple of real numbers for representing
-points and vectors. However, we cannot simply express this as the type
-$\mathsf{real} \times \mathsf{real}$ because it will not have the
-correct relational interpretation. (Two elements of type
-$\mathsf{real}$ will be related if and only if they are equal, and
-hence two elements of $\mathsf{real} \times \mathsf{real}$ will be
-related if and only if they are equal, by Reynolds' definition of the
-relational interpretation of tuple types.) Therefore, we introduce a
-new type parameterised by expressions $e$ of sort $T_2$ to represent
-points (and later, vectors): $\mathsf{vec}\langle e \rangle$. We have
-used the name $\mathsf{vec}$, even though we have taken pains to
-separate geometric points and vectors, to recall the computer science
-notion of vector as a sequence of values of homogeneous type which has
-a known length (in this case two).
+$\tyPrimNm{unit}$ and tuple types $A \times B$. We also assume a
+primitive type $\tyPrimNm{real}$, used to represent scalars. As we
+noted at the start of the previous section, the central data structure
+in geometric applications is the tuple of real numbers for
+representing points and vectors. However, we cannot simply express
+this as the type $\tyPrimNm{real} \times \tyPrimNm{real}$ because it
+will not have the correct relational interpretation. (Two elements of
+type $\tyPrimNm{real}$ will be related if and only if they are equal,
+and hence two elements of $\tyPrimNm{real} \times \tyPrimNm{real}$
+will be related if and only if they are equal, by Reynolds' definition
+of the relational interpretation of tuple types.) Therefore, we
+introduce a new type parameterised by expressions $e$ of sort
+$\SynTransl_2$ to represent points \emph{and} vectors:
+$\tyPrim{vec}{e}$. We have used the name $\tyPrimNm{vec}$, even though
+we have taken pains to separate geometric points and vectors, to
+recall the computer science notion of vector as a sequence of values
+of homogeneous type which has a known length (in this case two).
 
-% FIXME: Denotational semantics of $\mathsf{vec}$? and transformation
-% quantification?
+We now sketch the interpretation of the types $\tyPrim{vec}{e}$ and
+$\forall t\mathord:\SynTransl_2.A$. In our calculus, each type has two
+interpretations: an index-erasure interpretation that ignores the
+indexing expression $e$, and a relational interpretation as a binary
+relation on the index-erasure interpretation. For now, we assume we
+are given some way of turning an expression $e$ of sort $\SynTransl_2$
+into an actual element $\sem{e}\rho$ of the group $\Transl_2$, under
+some environment $\rho$ which tells how to interpret the free
+variables. The (simplified) index-erasure and relation interpretations
+of the types $\tyPrim{vec}{e}$ are:
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
+    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2 \\
+    \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{t}\rho} = \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{t}\rho \}
+  \end{array}
+\end{displaymath}
+The index-erasure interpetation of the universal quantification
+$\forall t\mathord:\SynTransl_2.A$ is simply the interpretation of the
+underlying type $A$. The relational interpretation intersects the
+relational interpretations of the type $A$ under all extensions of the
+relational environment $\rho$:
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
+    \tySem{\forall t\mathord:\SynTransl_2.A} & \tySem{A} \\
+    \rsem{\forall t\mathord:\SynTransl_2.A}{}\rho & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i\mathord:\SynTransl_2} \}
+  \end{array}
+\end{displaymath}
+The exact definition of $\extends{\rho}{i\mathord:\SynTransl_2}$
+depends on which (potentially non-compositional) interpretation of
+relational environments that we take. For now, it will suffice to
+think of relational environments as simply mappings from free
+variables to the group $\Transl_2$, and extension as just adding an
+extra component to the tuple. The general index-erasure semantics will
+be given in \autoref{sec:index-erasure-semantics}. The full relational
+interpretation is in general more complex, due to the possibility of
+non-compositional interpretations of free index variables. The
+relational interpretation will be presented in
+\autoref{sec:relational-semantics}.
 
-Equipped with quantification over the group $T_2$ of changes of
-origin, and the parameterised types $\mathsf{vec}\langle \cdot
-\rangle$, we can now express origin invariance properties as types. In
-the previous section, we discussed functions $f : \mathbb{R}^2 \times
-\mathbb{R}^2 \to \mathbb{R}^2$ that have the property that they
-preserve all changes of origin. We can now express this property as a
-type:
+Equipped with quantification over the group $\SynTransl_2$ of changes
+of origin, and the indexed types $\tyPrim{vec}{\cdot}$, we can now
+express origin invariance properties as types. In the previous
+section, we discussed functions $f : \mathbb{R}^2 \times \mathbb{R}^2
+\to \mathbb{R}^2$ that have the property that they preserve all
+changes of origin. We can now express this property as a type:
 \begin{displaymath}
-  \mathrm{f} : \forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \times \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle
+  \mathrm{f} : \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}
 \end{displaymath}
-FIXME: forward references...
+Spelling out the relational interpretation of this type using the
+definitions above (and the standard relational interpretations for
+tuple and function types), we obtain
+\statementref{eq:f-invariant-frame}.
 
@@
 
-As we stated in the previous section, the primitive operation of
-affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
-\lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
-affine combination can be interpreted as interpolation between the
-points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
-combination of points to our calculus, with the following typing and
-intended denotation:
+Invariance under change of origin is the key feature of affine
+geometry, and the central operation of affine geometry is the affine
+combination of points: $\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$,
+where $\lambda_1 + \lambda_2 = 1$. Geometrically, affine combination
+can be interpreted as interpolation between the points represented by
+$\vec{v_1}$ and $\vec{v_2}$. We add affine combination of points to
+our calculus, with the following typing and intended denotation:
 \begin{displaymath}
   \begin{array}{l}
-    \mathrm{affineCombination} :\forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{real} \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \\
-    \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ r\ (x_2,y_2) = \\
-    \hspace{3cm}((1-r)x_1 + tx_2, (1-r)y_1 + ty_2)
+    \mathrm{affComb} :\forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\
+    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-t)\vec{v_1} + t\vec{v_2}
   \end{array}
 \end{displaymath}
-By defining the primitive function $\mathrm{affineCombination}$ to
-take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
-we are taking the affine combination of two representatives of points,
-and not just an arbitrary linear combination. The intended denotation
-in the second line does not take any argument corresponding to the
-universally quantified $t$ in the type. Type-level variables are
-erased at runtime.
-
-% In Linear Algebra notation, we could have written the right-hand side
-% as $(1-t)\vec{v_1} + t\vec{v_2}$, where $\vec{v_1} = (x_1,y_1)$ and
-% $\vec{v_2} = (x_2,y_2)$.
+It can be verified by hand that the intended denotation
+$\sem{\mathrm{affComb}}$ has the invariance property specified by its
+type.  By defining the primitive function $\mathrm{affComb}$ to take a
+single $\tyPrimNm{real}$ parameter $t$, we can easily ensure that we
+are taking the affine combination of two representatives of points,
+and not just an arbitrary linear combination.
 
@@
     \begin{array}{@{}l}
-      \mathrm{quadB\acute{e}zier} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{real} \to \mathsf{vec}\langle t \rangle \\
+      \mathrm{quadB\acute{e}zier} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \\
       \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\
-      \quad
-      \begin{array}{l@{\hspace{-0.002cm}}l}
-        \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ s\ p_1) \\
-        & s \\
-        & (\mathrm{affineCombination}\ [F]\ p_1\ s\ p_2)
-      \end{array}
+      \quad \mathrm{affComb}\ [F]\ (\mathrm{affComb}\ [F]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [F]\ p_1\ s\ p_2) \\
     \end{array}
@@
   that the type of $\mathrm{quadB\acute{e}zier}$ tells us immediately
-  that it preserves all changes of origin
+  that it preserves all changes of origin.
 \end{example}
@@
 multiplication by a scalar and the zero vector. The typings of these
-operations will make use of the abelian group structure of the sort of
-translations $T_2$.
+operations will make use of the abelian group structure of change of
+origin translations.
 
 Vector operations are not invariant under change of origin. The
-obvious type assignment for vector addition is therefore:
+obvious type for vector addition is therefore:
 \begin{displaymath}
-  (+) : \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle
+  (+) : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
 \end{displaymath}
@@
 \begin{displaymath}
-  (+) : \forall t_1, t_2 \mathord: T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2 \rangle \to \mathsf{vec}\langle t_1 + t_2 \rangle
+  (+) : \forall t_1, t_2 \mathord: \SynTransl_2.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
 \end{displaymath}
@@
 \begin{displaymath}
-  \mathrm{negate} : \forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle -t \rangle
+  \mathrm{negate} : \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
 \end{displaymath}
@@
   \begin{array}{l}
-    (-) : \forall t_1,t_2 \mathord:T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2\rangle \to \mathsf{vec}\langle t_1 - t_2 \rangle \\
+    (-) : \forall t_1,t_2 \mathord:\SynTransl_2.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2} \\
     (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
@@
 respect to the same change of origin---i.e.~two values of type
-$\mathsf{vec}\langle t \rangle$---we can compute their offset, which
-is a vector expressed with respect to the null change of origin. The
-offset operation is a special case of vector subtraction:
+$\tyPrim{vec}{t}$---we can compute their offset, which is a vector
+expressed with respect to the null change of origin. The offset
+operation is a special case of vector subtraction:
 \begin{displaymath}
   \begin{array}{l}
-    \mathrm{offset} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle 0 \rangle \\
+    \mathrm{offset} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{0} \\
     \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
@@
   \begin{array}{l}
-    \mathrm{moveBy} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle t \rangle \\
+    \mathrm{moveBy} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to \tyPrim{vec}{t} \\
     \mathrm{moveBy}\ [t]\ p\ v = p + v
@@
 vector and multiplication by a scalar, do not describe any interesting
-effect on translations%\footnote{but they could...}
-. The zero vector has the type:
+effect on translations:
 \begin{displaymath}
-  \mathrm{0} : \mathsf{vec}\langle 0 \rangle
+  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l}
+    \mathrm{0} & \tyPrim{vec}{0}\\
+    (*) & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
+  \end{array}
 \end{displaymath}
-Intuitively, there cannot be a zero vector invariant under any change
-of origin, because this would entail picking a privileged origin. For
-multiplication of a vector by a scalar, we assign the following type,
-which is again not invariant under change of origin.
-\begin{displaymath}
-  (*) : \mathsf{real} \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle
-\end{displaymath}
+
+% \fixme{If we had dependent types, could the affComb operation be
+%   derived from an appropriately dependently typed scaling operation?}
 
@@
   \begin{displaymath}
-    \tau_{n} \isDefinedAs \forall t\mathord:T_2.\ \underbrace{\mathsf{vec}\langle t \rangle \to ... \mathsf{vec}\langle t \rangle}_{n+1\textrm{ times}} \to \mathsf{real}
+    \tau_{n} \isDefinedAs \forall t\mathord:\SynTransl_2.\ \underbrace{\tyPrim{vec}{t} \to ... \tyPrim{vec}{t}}_{n+1\textrm{ times}} \to \tyPrimNm{real}
   \end{displaymath}
@@
   inhabitants will be invariant under change of origin, due to the
-  quantification over all $t$ in $T_2$. Therefore, we may as well pick
+  quantification over all $t$ in $\SynTransl_2$. Therefore, we may as well pick
   one of the input points as the origin and assume that all the other
@@
   \begin{displaymath}
-    \sigma_{n} \isDefinedAs \underbrace{\mathsf{vec}\langle 0 \rangle \to ... \mathsf{vec}\langle 0 \rangle}_{n\textrm{ times}} \to \mathsf{real}
+    \sigma_{n} \isDefinedAs \underbrace{\tyPrim{vec}{0} \to ... \tyPrim{vec}{0}}_{n\textrm{ times}} \to \tyPrimNm{real}
   \end{displaymath}
-  We demonstrate this isomorphism formally in \autoref{sec:FIXME}.
+  We demonstrate this isomorphism formally in
+  \autoref{sec:instantiations}, in a more general setting of types
+  indexed by abelian groups.
 
@@
   \begin{displaymath}
-    \forall t \mathord: T_2.\ \vec{t + t} \to \vec{t}
+    \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}
   \end{displaymath}
   FIXME: explain that it is possible to show that this type is
-  uninhabited with the given operations. We show this in general 
+  uninhabited with the given operations.
 \end{example}
 
-\subsection{Generalising Origin Invariance}
+\subsection{Change of Basis Invariance}
 \label{sec:motivation-generalising}
 
+From basic linear algebra, we know that the vector operations of
+addition, negation and scaling, while not invariant under change of
+origin, are invariant under change of basis. Following the same
+methodology as we did for origin invariance, we can express basis
+invariance by means of preservation of relations. For origin
+invariance, we quantified over members of the group $\SynTransl_2$ of
+$2$-dimensional translations. Members of this group represent changes
+of origin. Change of basis is accomplished by applying an invertible
+linear map, and the collection of all invertible linear maps on
+$\mathbb{R}^2$ forms the group $\GL_2$.
+
+\fixme{new relational interpretation of vectors}
+
+\paragraph{Affine Geometry} The combination of a change of basis
+transformation and a change of origin transformation is known as an
+\emph{affine transformation}. A way of describing affine geometry is
+as the collection of properties that are invariant under invertible
+affine transformations.
+
+\begin{eqnarray*}
+  \mathrm{affComb} & : &
+  \begin{array}[t]{@{}l}
+    \forall B \mathord: \SynGL_2, t \mathord: \SynTransl_2.\\
+    \hspace{0.2cm} \tyPrim{vec}{B,t} \to \tyPrimNm{real} \to \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,t}
+  \end{array}
+  \\
+  (+) & : &
+  \begin{array}[t]{@{}l}
+    \forall B \mathord: \SynGL_2, t_1,t_2 \mathord: \SynTransl_2. \\
+    \hspace{0.2cm}\tyPrim{vec}{B,t_1} \to \tyPrim{vec}{B,t_2} \to \tyPrim{vec}{B,t_1 + t_2}
+  \end{array}
+  \\
+  \mathrm{negate} & : & \forall B \mathord: \SynGL_2, t \mathord: \SynTransl_2.\ \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,-t} \\
+  0 & : & \forall B \mathord: \SynGL_2.\ \tyPrim{vec}{B,0} \\
+  (*) & : & \forall B \mathord: \SynGL_2.\ \tyPrimNm{real} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0}
+\end{eqnarray*}
+
+FIXME: we now have all the operations of affine geometry, along with
+the property that they are invariant under affine transformations.
+
+\paragraph{Euclidean Geometry} Euclidean geometry extends affine
+geometry with the operation of \emph{dot product} (also known as the
+\emph{inner product}) of two vectors, which produces a real
+number. The standard Euclidean dot product is defined on the
+coordinate representation in the following way:
+\begin{displaymath}
+  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2
+\end{displaymath}
+The dot product is used to define properties of vectors such as their
+\emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
+of the length of the vector $\vec{v}$, and the notion of
+orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
+if $\vec{v_1}\cdot\vec{v_2} = 0$.
+
+To incorporate the operation of dot product into our calculus, we must
+give it a type. It is not the case that the dot product is invariant
+under the action of either of the groups $\GL_2$ or $\Transl_2$ that
+we have considered so far. In fact, the dot product is invariant under
+the subgroup $\Orth_2$ of $\GL_2$ of \emph{orthogonal} linear
+transformations. The group $\Orth_2$ consists of invertible linear
+maps whose determinant is either $1$ or $-1$. We introduce orthogonal
+transformations as a new sort $\Orth_2$, and overload the
+multiplicative group operations syntax for inhabitants of
+$\Orth_2$. We also assume a function $\iota_O$ that takes $e :
+\Orth_2$ to $\iota_O(e) : \SynGL_2$. We may now assign the following type
+to the dot product:
+\begin{displaymath}
+  (\cdot) : \forall O \mathord: \Orth(2).\ \tyPrim{vec}{\iota_O(O), 0} \to \tyPrim{vec}{\iota_O(O), 0} \to \tyPrimNm{real}
+\end{displaymath}
+
+\begin{figure}[t]
+  \centering
+  \begin{eqnarray*}
+    0   &:& \forall s \mathord:\Scale.\ \tyPrim{real}{s} \\
+    1   &:& \tyPrim{real}{1} \\
+    (+) &:& \forall s \mathord:\Scale.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
+    (-) &:& \forall s \mathord:\Scale.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
+    (*) &:& \forall s_1,s_2 \mathord:\Scale.\ \tyPrim{real}{s_1} \to \tyPrim{real}{s_2} \to \tyPrim{real}{s_1s_2} \\
+    (/) &:&
+    \begin{array}[t]{@{}l}
+      \forall s_1,s_2 \mathord:\Scale.\ \tyPrim{real}{s_1} \to \tyPrim{real}{s_2} \\
+      \hspace{3.5cm}\to \tyPrim{real}{s_1s_2^{-1}} + \tyPrimNm{unit} \\
+    \end{array}\\
+    \mathrm{abs} &:& \forall s \mathord:\Scale.\ \tyPrim{real}{s} \to \tyPrim{real}{|s|} %\\
+%    \mathrm{sqrt} &:& \tyPrim{real}{1} \to \tyPrim{real}{1}
+  \end{eqnarray*}
+  \caption{Operations on scaled real numbers}
+  \label{fig:real-ops}
+\end{figure}
+
+The cross product is another operation that takes pairs of vectors and
+produces a real number. Algebraically, the cross product is defined on
+coordinate representations as follows:
+\begin{displaymath}
+  (x_1,y_1) \times (x_2,y_2) = x_1y_2 - x_2y_1
+\end{displaymath}
+Geometrically, the cross product of two vectors is the signed area of
+the parallelogram described by the pair of input vectors. It turns out
+that the value of the cross product of two vectors varies with the
+determinant of the matrix describing the basis of the vector space. We
+therefore augment our calculus with a new sort $\Scale$ of scale
+factors. Semantically, $\Scale$ ranges over the non-zero real numbers
+and forms an abelian group, which we write multiplicatively. We also
+add two new operations: determinant, $\det B$, which takes an
+inhabitant of $\SynGL_2$ to its determinant in $\Scale$, and the
+operation of absolute value that takes scaling factors to scaling
+factors which we write as $|e|$. We also refine the type
+$\tyPrimNm{real}$ of real numbers to take a parameter of sort
+$\Scale$: $\tyPrim{real}{e}$, where the old type $\tyPrimNm{real}$ is
+just $\tyPrim{real}{1}$. The collection of operations on real numbers
+with scaling factors is shown in \autoref{fig:real-ops}.
+
+With the additional sort $\Scale$, and the refined type of real
+numbers, we can state the type of the cross product.
+\begin{displaymath}
+  (\times) : \forall B \mathord: \SynGL_2.\ \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B}
+\end{displaymath}
+For orthogonal transformations $O : \Orth_2$, the absolute value of
+the determinant is always $1$, so we assume the equation $|\det O| =
+1$ to hold.
+
+\begin{example}
+  Using the structure we have defined in this subsection, we can
+  define the following function that computes the area of a
+  triangle. By the type of this function, we can see that the area of
+  a triangle is invariant under orthognal transformations and
+  arbitrary translations.
+  \begin{displaymath}
+    \begin{array}{@{}l}
+      \mathrm{area} : \forall O\mathord:\Orth_2, t\mathord:\SynTransl_2.\ \\
+      \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{real}{1} \\
+      \mathrm{area}\ [O]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
+    \end{array}
+  \end{displaymath}
+  In the body of this function, we have performed the calculation in
+  several steps, each of which removes some of the symmetry described
+  by the type of $\mathrm{area}$. First we have computed the two offset
+  vectors $p_2 - p_1$ and $p_3 - p_1$. These operations remove the
+  effect of translations on the result, in exactly the same way as the
+  type isomorphism we demonstrated above. We then compute the cross
+  product of the two vectors to determine the area of the
+  parallelogram described by the sides of the triangle. This value has
+  type $\mathsf{real}\langle \det O \rangle$. We know that the
+  determinant of an orthogonal transformation is always either $-1$ or
+  $1$, so we have a result which varies according to possible
+  reflections of the space. To remove this variation, we apply the
+  $\mathrm{abs}$ function to leave us with a value of type
+  $\mathsf{real}\langle |\det O| \rangle$. This type is equal to the
+  type $\mathsf{real}\langle 1 \rangle$. Finally, we multiply by
+  $\frac{1}{2}$ to recover the area of the triangle, rather than the
+  whole parallelogram.
+\end{example}
+
+% \begin{enumerate}
+% \item Basis invariance
+% \item Orthogonal transform invariance for dot product
+% \item Rotational invariance (the type of the cross-product)
+% \item Scale invariance (generalisation of units of measure)
+% \end{enumerate}
+
+% In some cases, we can use the invariance as a programming tool
+% (e.g. type isomorphisms), but at others we can only use it from the
+% outside.
+
+% \begin{enumerate}
+% \item Possibly do a $n$-body gravity simulator as an example?
+% \item Have a special sort and type for rotations. An operation for
+%   computing the rotation between two points, and for applying a
+%   rotation to a vector. This will allow for more free theorems...
+% \end{enumerate}
+
+\subsection{Scale Invariance and Dimension Analysis}
+\label{sec:scale-invariance}
+
 \begin{enumerate}
-\item Basis invariance
-\item Orthogonal transform invariance for dot product
-\item Rotational invariance (the type of the cross-product)
-\item Scale invariance (units of measure)
+\item Explicitly link scale invariance to units of measure system
 \end{enumerate}
-
-In some cases, we can use the invariance as a programming tool
-(e.g. type isomorphisms), but at others we can only use it from the
-outside.
 
@@
 \item Cite Sabelfeld and Sands: ``A PER Model of Secure Information
-  Flow in Sequential Programs''
+  Flow in Sequential Programs''. And also Tse and Zdancewic.
+\item Go a step further by considering equivalence relations on
+  word32, and their lattice structure. Link this to quantative
+  information flow. Will also need to include singleton types.
+\end{enumerate}
+
+\subsection{Basic Continuity Analysis}
+\label{sec:continuity-analysis}
+
+Doing epsilon-delta proofs!
+
+Have a new sort $\mathsf{R}^{>0}$, representing positive distances
+between real numbers.
+
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
+    \tySem{\tyPrim{real}{e}} & \mathbb{R} \\
+    \rsem{\tyPrim{real}{e}}{}\rho & \{ (x,x') \sepbar |x-x'| < \sem{e}\rho \}
+  \end{array}
+\end{displaymath}
+
+A continuous function has type:
+\begin{displaymath}
+  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
+\end{displaymath}
+
+\begin{enumerate}
+\item Need to give types to basic operations on real numbers
+\item Need to add existential types to the calculus
 \end{enumerate}

### 2012-06-07 paper/section-introduction.tex
Substantial changes to the motivating examples section

@@
 \label{sec:introduction}
+
+Possible angles of attack:
+\begin{itemize}
+\item Parametric polymorphic types allow us to prevent
+  over-specification of the behaviour of programs. For instance, the
+  type $\forall \alpha. [\alpha] \to [\alpha]$ is a generalisation of
+  the types $[\mathsf{int}] \to [\mathsf{int}]$ and $[\mathsf{char}]
+  \to [\mathsf{char}]$. Either of the latter two types over-specify
+  the behaviour of the function.
+\item There are other cases of programs that are over-specified. The
+  leading example we just below is of geometric programs that
+  manipulate coordinate data. Often, programs that manipulate
+  coordinate data are insensitive to geometric transformations. For
+  example, a program that computes the area of a triangle described by
+  three points is insensitive to translations or rotations applied to
+  all three points.
+\item 
+\end{itemize}
 
@@
 \label{sec:contributions}
+
+\begin{itemize}
+\item Formulation of indexed parametricity that talks about
+  terms, not as part of terms (FIXME)
+\item Therefore, it can act as an analysis of programs, which can be
+  imposed after the fact
+\item Allow for more refined relational interpretations in order to
+  derive non-definability results.
+\item 
+\end{itemize}
+
+%%% Local Variables:
+%%% TeX-master: "paper"
+%%% End:

### 2012-06-07 paper/section-motivating-examples.tex
Add AUCTeX commands to deal with multifile document
Substantial changes to the motivating examples section

@@
 
-FIXME:
+Kennedy's units of measure system enforces, through a static typing
+discipline, a semantic notion of unit correctness. Idea: units of
+measure $\Rightarrow$ scale invariance $\Rightarrow$ representation
+independence $\Rightarrow$ Reynolds-style parametricity.
+
+To illustrate the semantic formulation of unit correctness, we present
+a small example.  Let $f$ be a function that takes pairs of real
+numbers to a real number:
+\begin{displaymath}
+  f : \mathbb{R} \times \mathbb{R} \to \mathbb{R},
+\end{displaymath}
+and let us assume that the first input to the function should denote
+some quantity of distance, the second input should denote some
+quantity of time, and that the output should denote some quantity of
+distance over time. Intuitively, these unit constraints mean that $f$
+may only do something like divide the first argument by the second,
+and possibly multiply by some dimensionless constant. But how do we
+state such conditions formally in terms of $f$?
+
+One way to state unit correctness is in terms of preservation of
+scaling. If we scale all distance inputs by some factor $k_d$, and all
+time inputs by some factor $k_t$, then the output of $f$ should
+correspondingly scale by the factor $k_d/k_t$ (since the supposed
+units of the output is distance over time). In symbols:
+\begin{displaymath}
+  \forall k_d, k_t, x_d, x_t.\ f(k_dx_d, k_tx_t) = \frac{k_d}{k_t}f(x_d,x_t)
+\end{displaymath}
+
+Another way of looking at scale invariance is in terms of invariance
+under change of representation. The number $5$, with respect to
+
 \begin{enumerate}
-\item Describe the units of measure system
-\item Describe the concept of scale invariance
-\item Describe the consequences of scale invariance
+\item Semantic correctness for units of measure is enforced by scaling
+  invariance
+\item Systematise this by using Reynolds' idea that semantic type
+  abstraction is formulated as preservation of relations. In the units
+  of measure case, we assign a relational meaning to every unit
+  expression.
+\item Consequences...
 \end{enumerate}
 
-\subsection{Vector Spaces and Frame Invariance}
+\subsection{Affine Spaces, Vector Spaces and Frame Invariance}
 \label{sec:vector-spaces-intro}
+
+An affine space is a vector space that has ``forgotten its origin''.
 
@@
 making sure to re-phrase everything in terms of ``satisfaction''.
+
+% \subsection{Monadic Effects}
+% \label{sec:monadic-effects}
+
+% Maybe this will work...
 
@@
 \end{enumerate}
+
+%%% Local Variables:
+%%% TeX-master: "paper"
+%%% End:

### 2012-06-08 paper/geometry-twocolumn.tex
Latest changes

@@
 \section{Motivating Examples}
+\label{sec:motivating-examples}
 
@@
 When writing programs that manipulate geometric data, the basic data
-structure is the $n$-tuple of real numbers. In the 2-dimensional case,
-tuples $\vec{v} = (x,y)$ may be called upon to represent both
+structure is the $n$-tuple of numbers. In the 2-dimensional case,
+tuples $\vec{v} = (x,y)$ are called upon to represent both
 \emph{points}---offsets from some origin---and
@@
     \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2 \\
-    \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{t}\rho} = \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{t}\rho \}
+    \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{t}\rho} = \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
   \end{array}
 \end{displaymath}
-The index-erasure interpetation of the universal quantification
-$\forall t\mathord:\SynTransl_2.A$ is simply the interpretation of the
+The index-erasure interpetation of universal quantification $\forall
+t\mathord:\SynTransl_2.A$ is simply the interpretation of the
 underlying type $A$. The relational interpretation intersects the
@@
     \tySem{\forall t\mathord:\SynTransl_2.A} & \tySem{A} \\
-    \rsem{\forall t\mathord:\SynTransl_2.A}{}\rho & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i\mathord:\SynTransl_2} \}
+    \rsem{\forall t\mathord:\SynTransl_2.A}{}\rho & \bigcap\{ \rsem{A}{}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:\SynTransl_2} \}
   \end{array}
@@
 The exact definition of $\extends{\rho}{i\mathord:\SynTransl_2}$
-depends on which (potentially non-compositional) interpretation of
+depends on which (possibly non-compositional) interpretation of
 relational environments that we take. For now, it will suffice to
@@
     \mathrm{affComb} :\forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\
-    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-t)\vec{v_1} + t\vec{v_2}
+    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} + r\vec{v_2}
   \end{array}
@@
 $\sem{\mathrm{affComb}}$ has the invariance property specified by its
-type.  By defining the primitive function $\mathrm{affComb}$ to take a
-single $\tyPrimNm{real}$ parameter $t$, we can easily ensure that we
-are taking the affine combination of two representatives of points,
-and not just an arbitrary linear combination.
+type.
+% By defining the primitive function $\mathrm{affComb}$ to take a
+% single $\tyPrimNm{real}$ parameter $t$, we can easily ensure that we
+% are taking the affine combination of two representatives of points,
+% and not just an arbitrary linear combination.
 
@@
       \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\
-      \quad \mathrm{affComb}\ [F]\ (\mathrm{affComb}\ [F]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [F]\ p_1\ s\ p_2) \\
+      \quad \mathrm{affComb}\ [t]\ (\mathrm{affComb}\ [t]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [t]\ p_1\ s\ p_2) \\
     \end{array}
@@
 
-Affine combination only provides us with an operation on points. We
-are also interested in operations on the vectors representing offsets
-between points. We now examine the correct types to assign to the
-vector space operations of addition of vectors, negation of vectors,
-multiplication by a scalar and the zero vector. The typings of these
-operations will make use of the abelian group structure of change of
-origin translations.
+% Affine combination only provides us with an operation on points. We
+% are also interested in operations on the vectors representing offsets
+% between points. We now examine the correct types to assign to the
+% vector space operations of addition of vectors, negation of vectors,
+% multiplication by a scalar and the zero vector. The typings of these
+% operations will make use of the abelian group structure of change of
+% origin translations.
+
+\fixme{Need to make clear that the point is to make use of the
+  algebraic indexing structure}
 
@@
 methodology as we did for origin invariance, we can express basis
-invariance by means of preservation of relations. For origin
-invariance, we quantified over members of the group $\SynTransl_2$ of
-$2$-dimensional translations. Members of this group represent changes
-of origin. Change of basis is accomplished by applying an invertible
+invariance by means of preservation of relations indexed by changes of
+basis. Change of basis is accomplished by applying an invertible
 linear map, and the collection of all invertible linear maps on
@@
 \fixme{new relational interpretation of vectors}
+
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
+    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2 \\
+    \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \sem{e_B}\rho\vec{v} + \sem{e_t}\rho \}
+  \end{array}
+\end{displaymath}
 
@@
 
-FIXME: we now have all the operations of affine geometry, along with
-the property that they are invariant under affine transformations.
+\fixme{we now have all the operations of affine geometry, along with
+  the property that they are invariant under affine transformations.}
 
@@
 \end{displaymath}
-The dot product is used to define properties of vectors such as their
-\emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
-of the length of the vector $\vec{v}$, and the notion of
-orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
-if $\vec{v_1}\cdot\vec{v_2} = 0$.
+% The dot product is used to define properties of vectors such as their
+% \emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
+% of the length of the vector $\vec{v}$, and the notion of
+% orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
+% if $\vec{v_1}\cdot\vec{v_2} = 0$.
 
@@
 maps whose determinant is either $1$ or $-1$. We introduce orthogonal
-transformations as a new sort $\Orth_2$, and overload the
+transformations as a new sort $\SynOrth_2$, and overload the
 multiplicative group operations syntax for inhabitants of
-$\Orth_2$. We also assume a function $\iota_O$ that takes $e :
-\Orth_2$ to $\iota_O(e) : \SynGL_2$. We may now assign the following type
+$\SynOrth_2$. We also assume a function $\iota_O$ that takes $e :
+\SynOrth_2$ to $\iota_O(e) : \SynGL_2$. We may now assign the following type
 to the dot product:
 \begin{displaymath}
-  (\cdot) : \forall O \mathord: \Orth(2).\ \tyPrim{vec}{\iota_O(O), 0} \to \tyPrim{vec}{\iota_O(O), 0} \to \tyPrimNm{real}
+  (\cdot) : \forall O \mathord: \SynOrth(2).\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
 \end{displaymath}
@@
   \begin{eqnarray*}
-    0   &:& \forall s \mathord:\Scale.\ \tyPrim{real}{s} \\
+    0   &:& \forall s \mathord:\SynGL_1.\ \tyPrim{real}{s} \\
     1   &:& \tyPrim{real}{1} \\
-    (+) &:& \forall s \mathord:\Scale.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
-    (-) &:& \forall s \mathord:\Scale.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
-    (*) &:& \forall s_1,s_2 \mathord:\Scale.\ \tyPrim{real}{s_1} \to \tyPrim{real}{s_2} \to \tyPrim{real}{s_1s_2} \\
+    (+) &:& \forall s \mathord:\SynGL_1.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
+    (-) &:& \forall s \mathord:\SynGL_1.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
+    (*) &:& \forall s_1,s_2 \mathord:\SynGL_1.\ \tyPrim{real}{s_1} \to \tyPrim{real}{s_2} \to \tyPrim{real}{s_1s_2} \\
     (/) &:&
-    \begin{array}[t]{@{}l}
-      \forall s_1,s_2 \mathord:\Scale.\ \tyPrim{real}{s_1} \to \tyPrim{real}{s_2} \\
-      \hspace{3.5cm}\to \tyPrim{real}{s_1s_2^{-1}} + \tyPrimNm{unit} \\
+    \begin{array}[t]{@{}l@{}l}
+      \forall s_1,s_2 \mathord:\SynGL_1.\ \tyPrim{real}{s_1}\ & \to \tyPrim{real}{s_2} \\
+      &\to \tyPrim{real}{s_1s_2^{-1}} + \tyPrimNm{unit} \\
     \end{array}\\
-    \mathrm{abs} &:& \forall s \mathord:\Scale.\ \tyPrim{real}{s} \to \tyPrim{real}{|s|} %\\
+    \mathrm{abs} &:& \forall s \mathord:\SynGL_1.\ \tyPrim{real}{s} \to \tyPrim{real}{|s|} %\\
 %    \mathrm{sqrt} &:& \tyPrim{real}{1} \to \tyPrim{real}{1}
@@
 
-The cross product is another operation that takes pairs of vectors and
-produces a real number. Algebraically, the cross product is defined on
-coordinate representations as follows:
+The cross product of two vectors is defined on coordinate
+representations as follows:
 \begin{displaymath}
@@
 \end{displaymath}
-Geometrically, the cross product of two vectors is the signed area of
-the parallelogram described by the pair of input vectors. It turns out
+Geometrically, the cross product is the signed area of the
+parallelogram described by the pair of input vectors. It turns out
 that the value of the cross product of two vectors varies with the
 determinant of the matrix describing the basis of the vector space. We
-therefore augment our calculus with a new sort $\Scale$ of scale
-factors. Semantically, $\Scale$ ranges over the non-zero real numbers
-and forms an abelian group, which we write multiplicatively. We also
-add two new operations: determinant, $\det B$, which takes an
-inhabitant of $\SynGL_2$ to its determinant in $\Scale$, and the
-operation of absolute value that takes scaling factors to scaling
-factors which we write as $|e|$. We also refine the type
-$\tyPrimNm{real}$ of real numbers to take a parameter of sort
-$\Scale$: $\tyPrim{real}{e}$, where the old type $\tyPrimNm{real}$ is
-just $\tyPrim{real}{1}$. The collection of operations on real numbers
-with scaling factors is shown in \autoref{fig:real-ops}.
+therefore augment our calculus with a new sort $\SynGL_1$ of scale
+factors. Semantically, $\SynGL_1$ ranges over the non-zero real
+numbers and forms an abelian group, which we write
+multiplicatively. We also add two new operations: determinant, $\det
+B$, which takes an inhabitant of $\SynGL_2$ to its determinant in
+$\SynGL_1$, and the operation of absolute value that takes scaling
+factors to scaling factors which we write as $|e|$. We also refine the
+type $\tyPrimNm{real}$ of real numbers to take a parameter of sort
+$\SynGL_1$: $\tyPrim{real}{e}$, where the old type $\tyPrimNm{real}$
+is just $\tyPrim{real}{1}$. The collection of operations on real
+numbers with scaling factors is shown in \autoref{fig:real-ops}.
 
-With the additional sort $\Scale$, and the refined type of real
+With the additional sort $\SynGL_1$, and the refined type of real
 numbers, we can state the type of the cross product.
@@
 For orthogonal transformations $O : \Orth_2$, the absolute value of
-the determinant is always $1$, so we assume the equation $|\det O| =
-1$ to hold.
+the determinant is always $1$, so we assume the equation $|\det
+(\iota_O O)| = 1$ to hold.
 
@@
   triangle. By the type of this function, we can see that the area of
-  a triangle is invariant under orthognal transformations and
-  arbitrary translations.
+  a triangle is invariant under orthogonal transformations and
+  arbitrary translations \fixme{i.e. isometries}.
   \begin{displaymath}
     \begin{array}{@{}l}
-      \mathrm{area} : \forall O\mathord:\Orth_2, t\mathord:\SynTransl_2.\ \\
+      \mathrm{area} : \forall O\mathord:\SynOrth_2, t\mathord:\SynTransl_2.\ \\
       \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{real}{1} \\
@@
   several steps, each of which removes some of the symmetry described
-  by the type of $\mathrm{area}$. First we have computed the two offset
-  vectors $p_2 - p_1$ and $p_3 - p_1$. These operations remove the
-  effect of translations on the result, in exactly the same way as the
-  type isomorphism we demonstrated above. We then compute the cross
-  product of the two vectors to determine the area of the
-  parallelogram described by the sides of the triangle. This value has
-  type $\mathsf{real}\langle \det O \rangle$. We know that the
+  by the type of $\mathrm{area}$. First we have computed the two
+  offset vectors $p_2 - p_1$ and $p_3 - p_1$. These operations remove
+  the effect of translations on the result, in exactly the same way as
+  the type isomorphism we demonstrated above \fixme{ref}. We then
+  compute the cross product of the two vectors to determine the area
+  of the parallelogram described by the sides of the triangle. This
+  value has type $\tyPrim{real}{\det (\iota_O O)}$. We know that the
   determinant of an orthogonal transformation is always either $-1$ or
@@
   $\mathrm{abs}$ function to leave us with a value of type
-  $\mathsf{real}\langle |\det O| \rangle$. This type is equal to the
-  type $\mathsf{real}\langle 1 \rangle$. Finally, we multiply by
-  $\frac{1}{2}$ to recover the area of the triangle, rather than the
-  whole parallelogram.
+  $\tyPrim{real}{|\det (\iota_O O)|}$. This type is equal to the type
+  $\tyPrim{real}{1}$. Finally, we multiply by $\frac{1}{2}$ to recover
+  the area of the triangle, rather than the whole parallelogram.
 \end{example}
@@
 
+Need a function $\iota_1 : \SynGL_1 \to \SynGL_2$. Such that $\det (\iota_1 s) = s^2$
+
 \subsection{Information Flow}
 \label{sec:information-flow}
+
+Blah blah blah. Need to work this out in terms of representation indepdence.
 

### 2012-06-08 paper/paper.tex
Half way through updating the geometry example
Add \fixme newcommand
Substantial changes to the motivating examples section
Latest changes

@@
 \def\subsectionautorefname{Section}
+\def\subsubsectionautorefname{Section}
 
@@
 \newcommand{\Transl}{\mathrm{T}}
+\newcommand{\SynTransl}{\mathsf{T}}
+\newcommand{\SynGL}{\mathsf{GL}}
+\newcommand{\SynOrth}{\mathsf{O}}
 \newcommand{\Scale}{\mathrm{Scale}}
@@
 \newcommand{\ty}{\textsf{ty}}
-\newcommand{\subst}[2]{\mathrm{Subst}(#1,#2)}
 
@@
 \newtheorem{theorem}{Theorem}
-\newtheorem{example}{Example}
 \newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
@@
 \newcommand{\isDefinedAs}{\stackrel{\mathit{def}}=}
+
+\newtheoremstyle{examplestyle}
+  {\topsep}  % space above
+  {\topsep}  % space below
+  {\normalfont}% name of font to use in the body of the theorem
+  {0em}% measure of space to indent
+  {\bfseries}% name of head font
+  {.}% punctuation between head and body
+  {5pt plus 1pt minus 1pt}% space after theorem head; " " = normal interword space
+  {}% Manually specify head
+\theoremstyle{examplestyle}
+\newtheorem{example}{Example}
+
+\newtheoremstyle{restatementstyle}
+  {\topsep}  % space above
+  {\topsep}  % space below
+  {\itshape}% name of font to use in the body of the theorem
+  {0em}% measure of space to indent
+  {\bfseries}% name of head font
+  {.}% punctuation between head and body
+  {5pt plus 1pt minus 1pt}% space after theorem head; " " = normal interword space
+  {\thmname{#1} \thmnote{(Restatement of #3)}}% Manually specify head
+\theoremstyle{restatementstyle}
+\newtheorem{restateLemma}{Lemma}
+\newtheorem{restateTheorem}{Theorem}
+
+\newcommand{\fixme}[1]{\textbf{FIXME: #1}}
 
@@
 \input{section-introduction.tex}
-\input{section-motivating-examples.tex}
+%\input{section-motivating-examples.tex}
+\input{geometry-twocolumn.tex}
 \input{section-background.tex}

### 2012-06-08 paper/section-general-framework.tex
Latest changes

@@
 are allowed. In particular, we use the relational interpretation to
-enforce the invariance properties that we indentified in
+enforce the invariance properties that we identified in
 \autoref{sec:motivating-examples}.
@@
 \in \mathrm{RelEnv}(\Delta)$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
-\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$. \fixme{Check this is the right way round}
+\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$.
 
@@
   \relEnv{E}(\Delta)$, along with an simultaneous substitution $\sigma
-  : \Delta' \Rightarrow \Delta$, such that the outer edges of the
-  following diagram commute, for all primitive types $\tyPrimNm{X} \in
-  \mathit{PrimType}$:
-  %FIXME: check that these compositions are the right way round
-  %FIXME: re-do this as a set of equations
-  %FIXME: give a better English description of why this is needed
-  \begin{displaymath}
-    \xymatrix{
-      {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
-      &
-      {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
-      \\
-      {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
-      &
-      {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
-      \\
-      &
-      &
-      {\Rel(\tyPrimSem{\tyPrimNm{X}})}
-    }
-  \end{displaymath}
-  Then there exists a relation environment $\rho \in
-  \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
-  two triangles in the bottom right of the diagram commute.
+  : \Delta' \Rightarrow \Delta$, such that $\rho_1 \circ
+  \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists a $\rho
+  \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho \circ
+  \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ \pi_{i\mathord:s'} =
+  \rho_2$.
+
+  % such that the outer edges of the
+  % following diagram commute, for all primitive types $\tyPrimNm{X} \in
+  % \mathit{PrimType}$:
+  % %FIXME: check that these compositions are the right way round
+  % %FIXME: re-do this as a set of equations
+  % %FIXME: give a better English description of why this is needed
+  % \begin{displaymath}
+  %   \xymatrix{
+  %     {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
+  %     &
+  %     {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
+  %     \\
+  %     {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
+  %     &
+  %     {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
+  %     \\
+  %     &
+  %     &
+  %     {\Rel(\tyPrimSem{\tyPrimNm{X}})}
+  %   }
+  % \end{displaymath}
+  % Then there exists a relation environment $\rho \in
+  % \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
+  % two triangles in the bottom right of the diagram commute.
   % FIXME: mention that we already know that the square commutes
@@
 
-\begin{proof}
-  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
-  three interesting cases follow:
-  \begin{description}
-  \item[Case \TirName{TyEq}] This case follows directly from
-    \lemref{lem:tyeqsubst-relational}.
-  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
-    $A = \forall i\mathord:s. A'$. We know from the induction
-    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
-    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho'}$,
-    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}{\rho'}$. We are given $\rho \in
-    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
-    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
-    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
-    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
-    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
-      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
-    \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho''}$ by \lemref{lem:ctxtsubst-rel}. So we
-    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}{\rho''}$, and thus
-    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
-      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
-      i.M'}\eta_2) \in \rsem{\forall
-      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
-  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
-    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
-    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
-    know that for all $\rho' \in \extends{\rho}{i:s}$,
-    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}{\rho'}$. If we let $\rho' = \rho \circ
-    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
-    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
-    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
-    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
-    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
-    by \lemref{lem:ctxtsubst-rel}, conclude that $(\tmSem{M'}\eta_1,
-    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
-    as required.
-  \end{description}
-  The remaining cases are standard for proofs of the fundamental lemma
-  of logical relations for simply-typed systems.
-\end{proof}
-
 %%% Local Variables:

### 2012-06-12 paper/geometry-twocolumn.tex
Major changes
- Refined set of contributions
- "Complete" the motivating examples section
- Fill in some references for the background section
- "Complete" the general framework section
- Sketch the "consequences" section
- Sketch the "discussion" section

@@
 types and their relational interpretations by presenting several
-examples. Our main set of examples comprise a novel type system for
-writing programs that manipulate geometric data, while maintaining
-geometric invariants through the types. \fixme{and the information
-  flow examples}
+examples. Our leading example is a novel type system for writing
+programs that manipulate geometric data, while maintaining geometric
+invariants through the types. The other examples are enforcement of
+information flow properties via indexing by a logical theory
+(\autoref{sec:information-flow}), and incorporation of metric
+properties into types (\autoref{sec:continuity-analysis}).
 
@@
     \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2 \\
-    \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{t}\rho} = \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
+    \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{e}\rho} = \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
   \end{array}
@@
     \tySem{\forall t\mathord:\SynTransl_2.A} & \tySem{A} \\
-    \rsem{\forall t\mathord:\SynTransl_2.A}{}\rho & \bigcap\{ \rsem{A}{}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:\SynTransl_2} \}
+    \rsem{\forall t\mathord:\SynTransl_2.A}{}\rho & \bigcap\{ \rsem{A}{}{\rho'} \sepbar \rho' \in \extends{\rho}{t\mathord:\SynTransl_2} \}
   \end{array}
 \end{displaymath}
-The exact definition of $\extends{\rho}{i\mathord:\SynTransl_2}$
+The exact definition of $\extends{\rho}{t\mathord:\SynTransl_2}$
 depends on which (possibly non-compositional) interpretation of
@@
 think of relational environments as simply mappings from free
-variables to the group $\Transl_2$, and extension as just adding an
-extra component to the tuple. The general index-erasure semantics will
-be given in \autoref{sec:index-erasure-semantics}. The full relational
-interpretation is in general more complex, due to the possibility of
+variables of sort $\SynTransl_2$ to the group $\Transl_2$, and
+extension as just adding an extra component to the tuple. The general
+index-erasure semantics will be given in
+\autoref{sec:index-erasure-semantics}. The relational interpretation
+is in general more complex, due to the possibility of
 non-compositional interpretations of free index variables. The
@@
 
-Equipped with quantification over the group $\SynTransl_2$ of changes
-of origin, and the indexed types $\tyPrim{vec}{\cdot}$, we can now
-express origin invariance properties as types. In the previous
-section, we discussed functions $f : \mathbb{R}^2 \times \mathbb{R}^2
-\to \mathbb{R}^2$ that have the property that they preserve all
-changes of origin. We can now express this property as a type:
+In the previous section, we discussed functions $f : \mathbb{R}^2
+\times \mathbb{R}^2 \to \mathbb{R}^2$ that have the property that they
+preserve all changes of origin. We can now express this property as a
+type:
 \begin{displaymath}
@@
 
-\fixme{Need to make clear that the point is to make use of the
-  algebraic indexing structure}
-
 Vector operations are not invariant under change of origin. The
@@
 However, it is possible to give vector addition a more precise type
-that describes its effect on the chosen origin. 
+that shows how it is variant with change of origin:
 \begin{displaymath}
@@
 \end{displaymath}
-The type of the vector addition operation can be specialised to the
-case of moving a point by a vector:
+Note how the type of the result of $\mathrm{offset}$ simplifies from
+$\tyPrim{vec}{t-t}$ to $\tyPrim{vec}{0}$. The algebraic structure on
+the indexing theory induces type equalities.  The type of the vector
+addition operation can be specialised to the case of moving a point by
+a vector:
 \begin{displaymath}
@@
 \begin{displaymath}
-  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l}
-    \mathrm{0} & \tyPrim{vec}{0}\\
+  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{4em}}l@{\hspace{0.5em}:\hspace{0.5em}}l}
+    \mathrm{0} & \tyPrim{vec}{0}&
     (*) & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
@@
 
-\begin{example}
+\begin{example}\label{ex:type-iso}
   The vector space operators and the properties that follow from their
@@
 
-\begin{example}
-  An inexpressible type
-  \begin{displaymath}
-    \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}
-  \end{displaymath}
-  FIXME: explain that it is possible to show that this type is
-  uninhabited with the given operations.
+\begin{example}\label{ex:uninhabited-type}
+  To this point we have emphasised the properties that we can deduce
+  about programs from their types, so called ``free theorems''. By
+  refining our relational interpretations of types appropriately it is
+  also possible to show that certain types are uninhabited. For
+  example, the type $\forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t
+    + t} \to \tyPrim{vec}{t}$ does not have any inhabitants. We
+  demonstrate this in \autoref{sec:types-indexed-abelian-groups}. In
+  order to obtain such a result, we require a more sophisticated
+  relational interpretation of free indexing variables than the one we
+  sketched above.
 \end{example}
@@
 linear map, and the collection of all invertible linear maps on
-$\mathbb{R}^2$ forms the group $\GL_2$.
-
-\fixme{new relational interpretation of vectors}
-
+$\mathbb{R}^2$ forms the group $\GL_2$. We reflect in our language the
+group $\GL_2$ as a new indexing sort $\SynGL_2$, which has
+(non-abelian) group structure that we will write multiplicatively. We
+also extend the $\tyPrimNm{vec}$ type to be indexed by elements of
+$\SynGL_2$ as well as $\SynTransl_2$. Thus, $\tyPrim{vec}{B,t}$ is a
+vector which varies with change of basis $B$ and change of origin
+$t$. Formally, the index-erasure and (simplified) relational semantics
+of this type are as follows:
 \begin{displaymath}
@@
     \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2 \\
-    \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \sem{e_B}\rho\vec{v} + \sem{e_t}\rho \}
+    \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = (\sem{e_B}\rho)\vec{v} + \sem{e_t}\rho \}
   \end{array}
@@
 
-\paragraph{Affine Geometry} The combination of a change of basis
-transformation and a change of origin transformation is known as an
-\emph{affine transformation}. A way of describing affine geometry is
-as the collection of properties that are invariant under invertible
-affine transformations.
-
+\paragraph{Affine Geometry} The combination of an invertible linear
+map and a translation is an \emph{affine transformation}. We can now
+assign types to all the primitive affine and vector space operations
+that indicates how they they behave with respect to affine
+transformations:
 \begin{eqnarray*}
@@
 
-\fixme{we now have all the operations of affine geometry, along with
-  the property that they are invariant under affine transformations.}
-
 \paragraph{Euclidean Geometry} Euclidean geometry extends affine
 geometry with the operation of \emph{dot product} (also known as the
-\emph{inner product}) of two vectors, which produces a real
-number. The standard Euclidean dot product is defined on the
-coordinate representation in the following way:
+\emph{inner product}) of two vectors. The standard Euclidean dot
+product is defined on the coordinate representation in the following
+way:
 \begin{displaymath}
@@
 % if $\vec{v_1}\cdot\vec{v_2} = 0$.
-
 To incorporate the operation of dot product into our calculus, we must
-give it a type. It is not the case that the dot product is invariant
-under the action of either of the groups $\GL_2$ or $\Transl_2$ that
-we have considered so far. In fact, the dot product is invariant under
-the subgroup $\Orth_2$ of $\GL_2$ of \emph{orthogonal} linear
-transformations. The group $\Orth_2$ consists of invertible linear
-maps whose determinant is either $1$ or $-1$. We introduce orthogonal
-transformations as a new sort $\SynOrth_2$, and overload the
-multiplicative group operations syntax for inhabitants of
+assign it a type. The dot product is not invariant under the action of
+either of the groups $\GL_2$ and $\Transl_2$ that we have considered
+so far. However, the dot product is invariant under the subgroup
+$\Orth_2$ of $\GL_2$ of \emph{orthogonal} linear transformations:
+invertible linear maps whose determinant is either $1$ or $-1$. We
+introduce orthogonal transformations as a new sort $\SynOrth_2$, and
+overload the multiplicative group operations syntax for inhabitants of
 $\SynOrth_2$. We also assume a function $\iota_O$ that takes $e :
-\SynOrth_2$ to $\iota_O(e) : \SynGL_2$. We may now assign the following type
-to the dot product:
+\SynOrth_2$ to $\iota_O(e) : \SynGL_2$. We may now assign a type to
+the dot product, using quantification over $\SynOrth_2$:
 \begin{displaymath}
-  (\cdot) : \forall O \mathord: \SynOrth(2).\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
+  (\cdot) : \forall O \mathord: \SynOrth_2.\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
 \end{displaymath}
@@
 The cross product of two vectors is defined on coordinate
-representations as follows:
-\begin{displaymath}
-  (x_1,y_1) \times (x_2,y_2) = x_1y_2 - x_2y_1
-\end{displaymath}
+representations as $(x_1,y_1) \times (x_2,y_2) = x_1y_2 - x_2y_1$.
 Geometrically, the cross product is the signed area of the
-parallelogram described by the pair of input vectors. It turns out
-that the value of the cross product of two vectors varies with the
-determinant of the matrix describing the basis of the vector space. We
-therefore augment our calculus with a new sort $\SynGL_1$ of scale
-factors. Semantically, $\SynGL_1$ ranges over the non-zero real
-numbers and forms an abelian group, which we write
-multiplicatively. We also add two new operations: determinant, $\det
-B$, which takes an inhabitant of $\SynGL_2$ to its determinant in
-$\SynGL_1$, and the operation of absolute value that takes scaling
-factors to scaling factors which we write as $|e|$. We also refine the
-type $\tyPrimNm{real}$ of real numbers to take a parameter of sort
-$\SynGL_1$: $\tyPrim{real}{e}$, where the old type $\tyPrimNm{real}$
-is just $\tyPrim{real}{1}$. The collection of operations on real
-numbers with scaling factors is shown in \autoref{fig:real-ops}.
+parallelogram described by the pair of input vectors. Under change of
+basis by an invertible linear transformation $B$, the value of the
+cross product of two vectors varies with the determinant of $B$. This
+corresponds to the scaling of the plane by the change of basis
+transformation.
+
+We therefore augment our calculus with a new sort $\SynGL_1$ of scale
+factors (i.e.~$1$-dimensional invertible linear maps). Semantically,
+$\SynGL_1$ ranges over the non-zero real numbers and forms an abelian
+group, which we write multiplicatively. We also add two new
+operations: determinant, $\det B$, which takes an inhabitant of
+$\SynGL_2$ to its determinant in $\SynGL_1$, and the operation of
+absolute value that takes scaling factors to scaling factors which we
+write as $|e|$. We also refine the type $\tyPrimNm{real}$ of real
+numbers to be indexed by the sort $\SynGL_1$: $\tyPrim{real}{e}$,
+where the old type $\tyPrimNm{real}$ is just $\tyPrim{real}{1}$. The
+collection of operations on real numbers indexed by scaling factors is
+shown in \autoref{fig:real-ops}.
 
 With the additional sort $\SynGL_1$, and the refined type of real
-numbers, we can state the type of the cross product.
+numbers, we can assign a type to the cross product:
 \begin{displaymath}
@@
 \end{displaymath}
-For orthogonal transformations $O : \Orth_2$, the absolute value of
-the determinant is always $1$, so we assume the equation $|\det
-(\iota_O O)| = 1$ to hold.
+Since the absolute value of the determinant of an orthogonal
+transformation is always $1$, we assume $|\det (\iota_O O)| = 1$ to
+hold.
 
-\begin{example}
-  Using the structure we have defined in this subsection, we can
-  define the following function that computes the area of a
-  triangle. By the type of this function, we can see that the area of
-  a triangle is invariant under orthogonal transformations and
-  arbitrary translations \fixme{i.e. isometries}.
+\begin{example}\label{ex:area-of-triangle-1}
+  Using the operations we have defined in this subsection, we can
+  compute the area of a triangle. 
+% By the type of this function, we can
+%   see that the area of a triangle is invariant under orthogonal
+%   transformations and arbitrary translations. Such transformations are
+%   known as isometries: distance preserving maps.
+  \begin{displaymath}
+    \begin{array}{@{}l}
+      \mathrm{area} : \forall B\mathord:\SynGL_2, t\mathord:\SynTransl_2.\ \\
+      \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{|\det B|} \\
+      \mathrm{area}\ [B]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
+    \end{array}
+  \end{displaymath}
+  The calculation is performed in several steps, each of which removes
+  some of the symmetry described by the type of
+  $\mathrm{area}$. First, the two offset vectors $p_2 - p_1$ and $p_3
+  - p_1$ are computed. These operations remove the effect of
+  translations on the result, in exactly the same way as the type
+  isomorphism we introduced in \exref{ex:type-iso}. We then compute
+  the cross product of the two vectors to determine the area of the
+  parallelogram described by the sides of the triangle. This value has
+  type $\tyPrim{real}{\det B}$, and we have removed some of the
+  symmetry due to invertible linear maps. This value still varies with
+  the sign of the determinant, so we remove this symmetry by applying
+  the $\mathrm{abs}$ function to leave us with a value of type
+  $\tyPrim{real}{|\det B|}$. Finally, we multiply by the constant
+  $\frac{1}{2}$ to recover the area of the triangle, rather than the
+  whole parallelogram.
+
+  If we specialise $\mathrm{area}$ to just orthogonal transformations,
+  we obtain the following type, as a consequence of the equation
+  $|\det (\iota_OO)| = 1$:
   \begin{displaymath}
@@
       \mathrm{area} : \forall O\mathord:\SynOrth_2, t\mathord:\SynTransl_2.\ \\
-      \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{real}{1} \\
-      \mathrm{area}\ [O]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
+      \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{real}{1}
     \end{array}
   \end{displaymath}
-  In the body of this function, we have performed the calculation in
-  several steps, each of which removes some of the symmetry described
-  by the type of $\mathrm{area}$. First we have computed the two
-  offset vectors $p_2 - p_1$ and $p_3 - p_1$. These operations remove
-  the effect of translations on the result, in exactly the same way as
-  the type isomorphism we demonstrated above \fixme{ref}. We then
-  compute the cross product of the two vectors to determine the area
-  of the parallelogram described by the sides of the triangle. This
-  value has type $\tyPrim{real}{\det (\iota_O O)}$. We know that the
-  determinant of an orthogonal transformation is always either $-1$ or
-  $1$, so we have a result which varies according to possible
-  reflections of the space. To remove this variation, we apply the
-  $\mathrm{abs}$ function to leave us with a value of type
-  $\tyPrim{real}{|\det (\iota_O O)|}$. This type is equal to the type
-  $\tyPrim{real}{1}$. Finally, we multiply by $\frac{1}{2}$ to recover
-  the area of the triangle, rather than the whole parallelogram.
+  This type demonstrates that the area of a triangle is invariant
+  under orthogonal transformations and translations. Such combined
+  transformations are known as isometries---distance preserving maps.
 \end{example}
-
-% \begin{enumerate}
-% \item Basis invariance
-% \item Orthogonal transform invariance for dot product
-% \item Rotational invariance (the type of the cross-product)
-% \item Scale invariance (generalisation of units of measure)
-% \end{enumerate}
-
-% In some cases, we can use the invariance as a programming tool
-% (e.g. type isomorphisms), but at others we can only use it from the
-% outside.
 
@@
 
-\subsection{Scale Invariance and Dimension Analysis}
+\subsection{Scale Invariance and Dimensional Analysis}
 \label{sec:scale-invariance}
 
-\begin{enumerate}
-\item Explicitly link scale invariance to units of measure system
-\end{enumerate}
+Indexing types by scaling factors brings us to the original
+inspiration for the current work: Kennedy's interpretation of his
+units of measure type system via scaling invariance
+\cite{kennedy97relational}. Kennedy demonstrates how interpreting
+types in terms of scaling invariance brings the techniques of
+dimensional analysis \fixme{ref} to bear on programming. The types we
+have assigned to the real number arithmetic operations in
+\autoref{fig:real-ops} are exactly the types Kennedy assigns in his
+units of measure system, except for the absolute value operation. The
+effect of our assigned type is that, semantically, we are indexing by
+non-zero scaling factors, where Kennedy indexes by strictly positive
+scaling factors. We further examine the typing of the absolute value
+operation in \autoref{sec:types-indexed-abelian-groups}.
 
-Need a function $\iota_1 : \SynGL_1 \to \SynGL_2$. Such that $\det (\iota_1 s) = s^2$
+In our two-dimensional setting, we can make additions to Kennedy's
+one-dimensional scaling invariance. We add an operation $\iota_1$
+that, semantically, takes scale factors in $\GL_1$ to invertible
+linear maps in $\GL_2$: i.e.~takes numbers $s$ to matrices $\left(
+  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$.  This
+operation satisfies the equation $\det (\iota_1 s) = s^2$, indicating
+that if we scale the plane by $s$ in all directions, we scale areas by
+$s^2$. This is illustrated by the next example.
 
-\subsection{Information Flow}
+\begin{example}\label{ex:area-of-triangle-2}
+  Just as we specialised the type of $\mathrm{area}$ function to
+  orthogonal transformations in \exref{ex:area-of-triangle-1}, we can
+  also specialise $\mathrm{area}$'s type to scaling
+  transformations. This yields the type:
+  \begin{displaymath}
+    \begin{array}{@{}l}
+      \mathrm{area} : \forall s\mathord:\SynGL_1, t\mathord:\SynTransl_2.\ \\
+      \hspace{0.8cm} \tyPrim{vec}{\iota_1s, t} \to \tyPrim{vec}{\iota_1s, t} \to \tyPrim{vec}{\iota_1s, t} \to \tyPrim{real}{s^2}
+    \end{array}
+  \end{displaymath}
+  As expected, the area of a triangle varies with the square of
+  scalings of the plane, and this is reflected in the type.
+\end{example}
+
+Purely scaling linear maps, as generated by $\iota_1$, commute with
+all other invertible linear maps, and form the centre of the group
+$\GL_2$. Thus we add the equation $\iota_1(s)B = B\iota_1(s)$ to our
+calculus.
+
+If we keep track of scalings, we may assign more precise types to
+scalar multiplcation:
+\begin{displaymath}
+  (*) : \forall s \mathord: \SynGL_1, B \mathord: \SynGL_2.\ \tyPrim{real}{s} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{\iota_1(s)B,0}
+\end{displaymath}
+and dot product:
+\begin{displaymath}
+  \begin{array}{l@{}l}
+    (\cdot) : \forall s : \SynGL_1, O : \SynOrth_2.\ & \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\\
+    &\tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to \tyPrim{real}{s^2}
+  \end{array}
+\end{displaymath}
+
+\begin{example}
+  With the operations in \autoref{fig:real-ops}, it is not possible to
+  write a program with the following type that is not constantly zero:
+  \begin{displaymath}
+    \forall s : \SynGL_2.\ \tyPrim{real}{s^2} \to \tyPrim{real}{s}
+  \end{displaymath}
+  This was shown by Kennedy for his units of measure system
+  \cite{kennedy97relational}.  In particular, it is not possible to
+  write a square root function with the above type. The indefinability
+  of square root is similar to the uninhabited type from
+  \exref{ex:uninhabited-type}.
+
+  In \autoref{sec:types-indexed-abelian-groups} we revisit Kennedy's
+  proof and show that if we add square root as a primitive
+  operation---with the type above---then it is still not possible to
+  construct the cube root function.
+\end{example}
+
+\subsection{Logical Information Flow}
 \label{sec:information-flow}
 
-Blah blah blah. Need to work this out in terms of representation indepdence.
+An extreme example of invariance under change of representation is the
+tracking of information flow through programs. If a program does not
+depend upon a particular piece of information, then it will be
+invariant under \emph{all} changes of representation of this
+information. As thoroughly described by Sabelfeld and Sands
+\cite{sabelfeld01per}, information flow can be captured semantically
+through the use of partial equivalence relations (PERs). PERs form an
+instance of Reynolds' approach to abstraction and representation
+independence. Abadi, Banerjee, Heintze and Riecke \cite{abadi99core}
+built a \emph{Core Calculus for Dependency}, using a type system based
+around a security level indexed monad $T_lA$. Tse and Zdancewic
+\cite{tse04translating} showed that it is possible to translate Abadi
+et al.'s calculus into System F, translating the monadic type $T_lA$
+to $\alpha_l \to A$ for some free type variable $\alpha_l$, and making
+direct use of Reynolds' abstraction theorem to prove information flow
+properties. We now show how a refined variant of Tse and Zdancewic's
+translation can be expressed via algebraically-indexed types.
 
-\begin{enumerate}
-\item Think of information flow as passing around unforgable tokens
-  \begin{enumerate}
-  \item Give an example (given a piece of data protected by a token
-    and a token, can we extract the data?)
-  \item Need to make sure that tokens are unforgable
-  \end{enumerate}
-\item Could either enforce unforgability of tokens by some dynamic
-  means, or by static type-based means
-  \begin{enumerate}
-  \item Leave the dynamic method as vague
-  \item Ingredients of the static method: run-time tokens are just
-    dummy values ($*$), the token type is parameterised by logical
-    formulae, and the relational meaning is derived from an assignment
-    of boolean values to atoms.
-  \item Explain this as invariance under changes of satisfaction of
-    the participants. Relate to descriptions of logics as rules that
-    don't care about truth, only its preservation.
-  \end{enumerate}
-\item Cite Sabelfeld and Sands: ``A PER Model of Secure Information
-  Flow in Sequential Programs''. And also Tse and Zdancewic.
-\item Go a step further by considering equivalence relations on
-  word32, and their lattice structure. Link this to quantative
-  information flow. Will also need to include singleton types.
-\end{enumerate}
+We assume a single indexing sort $\mathsf{principal}$, intended to
+represent (possibly composite) principals in a system. Principals are
+combined using the connectives of boolean logic ($\top$, $\bot$,
+$\lor$, $\land$, $\lnot$). For the equational theory of principals, we
+assume all the axioms of boolean algebra, and semantically we
+interpret composite principals as boolean values: either true ($\top$)
+or false ($\bot$). We introduce a single primitive type, indexed by
+expressions of sort $\mathsf{principal}$, $\tyPrim{T}{e}$, with the
+following interpretations:
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
+    \tySem{\tyPrim{T}{e}} & \{*\} \\
+    \rsem{\tyPrim{T}{e}}{}\rho & \left\{
+      \begin{array}{ll}
+        \{(*,*)\} & \textrm{if }\sem{e}\rho = \top \\
+        \{\}      & \textrm{if }\sem{e}\rho = \bot
+      \end{array}
+      \right.
+  \end{array}
+\end{displaymath}
+The relational interpretation of $\tyPrim{T}{e}$ forces this type to
+only be inhabited when the boolean expression $e$ evaluates to true
+under the environment $\rho$. To write programs, we assume the
+following pair of primitive operations:
+\begin{eqnarray*}
+  \mathrm{proj} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1 \land p_2} \to \tyPrim{T}{p_1} \\
+  \mathrm{combine} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1} \to \tyPrim{T}{p_2} \to \tyPrim{T}{p_1 \land p_2}
+\end{eqnarray*}
+We can now adapt Tse and Zdancewic's translation of Abadi et al.'s
+monadic type to our setting. We define a type synonym, for each type
+$A$ and expression $e$ of sort $\mathsf{principal}$:
+\begin{displaymath}
+  T_eA = \tyPrim{T}{e} \to A
+\end{displaymath}
+For every principal $p$, we can endow the types $T_p-$ with the
+structure of a monad. This is due to the fact that it is an instance
+of the ``environment'' (or ``reader'') monad \fixme{ref}.
 
-\subsection{Basic Continuity Analysis}
+\begin{example}
+  Suppose we have a program with the following type:
+  \begin{displaymath}
+    \forall p,q\mathord:\mathsf{principal}. \tyPrim{T}{e} \to T_p(\tyPrimNm{unit} + \tyPrimNm{unit}) \to T_q(\tyPrimNm{unit} + \tyPrimNm{unit})
+  \end{displaymath}
+  where $e$ is some boolean algebra expression involving $p$ and
+  $q$. We will show in \autoref{sec:instantiations} that programs with
+  this type are non-constant (i.e.~depend on their input) if and only
+  if the formula $e \Rightarrow q \Rightarrow p$ is valid in boolean
+  logic.
+\end{example}
+
+% Go a step further by considering equivalence relations on word32, and
+% their lattice structure. Link this to quantative information
+% flow. Will also need to include singleton types.
+
+\subsection{Metric Spaces and Simple Continuity Analysis}
 \label{sec:continuity-analysis}
 
-Doing epsilon-delta proofs!
+Our final motivating example of algebraically indexed types and
+representation independence makes use of the metric space structure of
+the real numbers. In the geometry example, we used relational
+interpretations that related values via geometrical
+transformations. An alternative relational interpretation of the real
+numbers is to relate values that are close, in the metric space
+sense. By indexing types by a measure of closeness, we can actually
+state continuity of functions as a typing problem.
 
-Have a new sort $\mathsf{R}^{>0}$, representing positive distances
-between real numbers.
-
+We introduce an indexing sort $\mathsf{R}^{>0}$, representing positive
+distances between real numbers. Our type of real numbers is indexed by
+expressions of sort $\mathsf{R}^{>0}$, and we use the following
+index-erasure and relational interpretations:
 \begin{displaymath}
@@
 \end{displaymath}
-
-A continuous function has type:
+Thus, two real numbers are related if they only differ by the positive
+real number assigned to $e$. With this interpretation, and a
+straightforward interpretation of existential types, we can state the
+standard $\epsilon$-$\delta$ definition of continuity as a type:
 \begin{displaymath}
@@
 \end{displaymath}
+Chaudhuri, Gulwani and Lublinerman \cite{chaudhuri10continuity} have
+presented a program logic based approach to verifying the continuity
+of programs. Algebraically indexed types give us a way of expressing
+and verifying continuity properties of functions in a type based way.
 
-\begin{enumerate}
-\item Need to give types to basic operations on real numbers
-\item Need to add existential types to the calculus
-\end{enumerate}
+% \begin{eqnarray*}
+%   \underline{c} & : & \forall \epsilon\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon} \\
+%   (+) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
+%   (-) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
+%   (*) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1\epsilon_2}
+% \end{eqnarray*}
+
+% \begin{enumerate}
+% \item Need to give types to basic operations on real numbers
+% \item Need to add existential types to the calculus
+% \end{enumerate}
 

### 2012-06-12 paper/section-background.tex
Add AUCTeX commands to deal with multifile document
Some notes on background work
Major changes
- Refined set of contributions
- "Complete" the motivating examples section
- Fill in some references for the background section
- "Complete" the general framework section
- Sketch the "consequences" section
- Sketch the "discussion" section

@@
 \label{sec:background}
+
+\begin{enumerate}
+\item Reynolds \cite{reynolds83types}
+\item Kennedy \cite{kennedy97relational}
+\item Gallier (for information on geometry) \cite{gallier11geometric}
+\item Tse and Zdancewic, and Sabelfeld and Sands, for information flow \cite{sabelfeld01per}
+\item Chaudhuri, Gulwani and Lublinerman, for continuity analysis \cite{chaudhuri10continuity}
+\end{enumerate}
+
+We discuss further related work in \autoref{sec:discussion}.
+
+%%% Local Variables:
+%%% TeX-master: "paper"
+%%% End:

### 2012-06-12 paper/section-discussion.tex
Major changes
- Refined set of contributions
- "Complete" the motivating examples section
- Fill in some references for the background section
- "Complete" the general framework section
- Sketch the "consequences" section
- Sketch the "discussion" section

\section{Discussion}
\label{sec:discussion}

\begin{enumerate}
\item Dependent Types (allow for more precision, in some cases would
  subsume our general framework, reference Bernardy's work)
\item Type-indexed types (e.g.~type and effect systems)
\item Making use of the free theorems, either by internalisation or by
  deep embedding of our general calculus in a theorem prover.
\item Extension of the definability work. Is it possible to extend
  standard definability results for the $\lambda$-calculus to cover
  algebraically indexed types?
\item Discussion of the relationship between relational parametricity
  and the abstraction theorem we have demonstrated. Discussion of the
  identity extension property in this setting.
\item Extending the geometric example with linear transformations, or
  even higher-degree tensors.
\end{enumerate}

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:

### 2012-06-12 paper/section-general-framework.tex
Major changes
- Refined set of contributions
- "Complete" the motivating examples section
- Fill in some references for the background section
- "Complete" the general framework section
- Sketch the "consequences" section
- Sketch the "discussion" section

@@
 
-\fixme{Update this introduction}
-
-We now develop a general framework for discussing
-algebraically-indexed types and their relational interpretations. We
-have chosen a particular order for our presentation: presenting the
-semantics of types before we present the syntax of programs and their
-typing relation. We have deliberately selected this order to emphasise
-the importance of establishing what types \emph{mean} over the
-language used to write programs.
-
-We first present, in \autoref{sec:algebraically-indexed-types}, the
-syntax of the language of algebraically-indexed types. Our general
-framework is parameterised by a choice of multi-sorted algebraic
-theory to be used for indexing types, and the collections of primitive
-types and
-operations. %FIXME: refer forwards to the names of the parameters
-We define substitution of index terms, and what it means for two types
-to be equal. Equality on types is generated from the chosen
-multi-sorted algebraic theory.
-
-We next present the semantics of types in
-\autoref{sec:semantics-algebraically-indexed-types}. Types are first
-interpreted without reference to the indexing terms: the
-\emph{index-erasure} interpretation. The index-erasure interpretation
-provides an ``underlying'' semantics to each type (and term), which we
-think of as an abstraction of the actual operational behaviour of a
-term. By erasing all the index terms, we can immediately see that the
-run-time behaviour of programs cannot depend on index terms. We then
-define the \emph{relational interpretation}, which interprets every
-type as a binary relation on its index-erasure interpretation. The
-relational interpretation acts as ``commentary'' on the underlying
-interpretation, constraining the index-erasure interpretations that
-are allowed. In particular, we use the relational interpretation to
-enforce the invariance properties that we identified in
-\autoref{sec:motivating-examples}.
-
-Throughout this section, we use Kennedy's original units of measure
-system as a running example in order to provide a concrete example of
-our general framework, and also to demonstrate that we have in fact
-generalised Kennedy's original development.
+We now present a general framework for discussing
+algebraically-indexed types and representation independence. We define
+the syntax of algebraically-indexed types, then their semantics and
+finally we present a general programming language for
+algebraically-indexed types, with an abstraction theorem.
 
@@
 
-The indexing terms and types of an instantiation of our general
+The index expressions and types of an instantiation of our general
 framework are derived from the following data:
@@
 
-\begin{example}[Origin Invariance]
-  \fixme{Do this, then update the units of measure example}
+\begin{example}[Two-Dimensional Geometry]
+  The two-dimensional geometry system has a sort for each of the
+  geometric groups that we mentioned in
+  \autoref{sec:motivating-examples}. Thus, $\SortSet = \{\SynTransl_2,
+  \SynGL_2, \SynOrth_2, \SynGL_1 \}$. For the index operations, we
+  have additive group structure on $\SynTransl_2$, multiplicative
+  group structure on $\SynGL_1$, $\SynGL_2$ and $\SynOrth_2$,
+  injections from $\SynOrth_2$ and $\SynGL_1$ into $\SynGL_2$,
+  determinant and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
+  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, |\cdot| \}$, where
+  $G \in \{\SynGL_1, \SynGL_2, \SynOrth_1\}$, with the following
+  assignment of arities:
+  \begin{displaymath}
+    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
+      \indexOpArity(0) & ([], \SynTransl_2) &
+      \indexOpArity(1_G) & ([], G) \\
+      \indexOpArity(+) & ([\SynTransl_2, \SynTransl_2], \SynTransl_2) &
+      \indexOpArity(\cdot_G) & ([G,G],G) \\
+      \indexOpArity(-) & ([\SynTransl_2], \SynTransl_2) &
+      \indexOpArity(^{-1_G}) & ([G], G) \\
+      \indexOpArity(\iota_o) & ([\SynOrth_2], \SynGL_2) &
+      \indexOpArity(\iota_1) & ([\SynGL_1], \SynGL_2) \\
+      \indexOpArity(\det) & ([\SynGL_2], \SynGL_1) &
+      \indexOpArity(|\cdot|) & ([\SynGL_1], \SynGL_1)
+    \end{array}
+  \end{displaymath}
+  The intended interpretations of the top three pairs of operations
+  are group unit, group combination and group negation,
+  respectively. When we introduce equational theories for index
+  expressions in \autoref{sec:type-equality} below, we will impose the
+  (abelian) group laws. For this example, we have $\PrimTypeSet = \{
+  \tyPrimNm{vec}, \tyPrimNm{real} \}$, with
+  $\primTyArity(\tyPrimNm{vec}) = [\SynGL_2, \SynTransl_2]$ and
+  $\primTyArity(\tyPrimNm{real}) = [\SynGL_1]$.
 \end{example}
 
-\begin{example}[Units of Measure]
-  To instantiate our general framework as Kennedy's original units
-  of measure system, we would take the following settings. For the
-  collection of sorts we take $\SortSet = \{ \mathsf{unit} \}$. For
-  the collection of index operations, we take $\mathit{IndexOp} = \{1,
-  -\cdot-, -^{-1}\}$, with:
-  \begin{eqnarray*}
-    \indexOpArity{(1)} &=& ([],\mathsf{unit}) \\
-    \indexOpArity{(-\cdot-)} & = & ([\mathsf{unit},\mathsf{unit}],\mathsf{unit}) \\
-    \indexOpArity{(-^{-1})} & = & ([\mathsf{unit}],\mathsf{unit})
-  \end{eqnarray*}
-  The intended interpretation of these operations is group unit, group
-  multiplication and group inverse, respectively. When we discuss
-  equational theories for index expressions in
-  \autoref{sec:type-equality} below, we will impose the abelian group
-  laws on these operations.
+% \begin{example}[Logical Information Flow]
+%   \fixme{Do this}
+% \end{example}
 
-  The primitive types for the units of measure example are taken to be
-  $\PrimTypeSet = \{ \mathsf{num} \}$, with
-  $\primTyArity(\mathsf{num}) = [\mathsf{unit}]$.
-\end{example}
+% \begin{example}[Units of Measure]
+%   To instantiate our general framework as Kennedy's original units
+%   of measure system, we would take the following settings. For the
+%   collection of sorts we take $\SortSet = \{ \mathsf{unit} \}$. For
+%   the collection of index operations, we take $\mathit{IndexOp} = \{1,
+%   -\cdot-, -^{-1}\}$, with:
+%   \begin{eqnarray*}
+%     \indexOpArity{(1)} &=& ([],\mathsf{unit}) \\
+%     \indexOpArity{(-\cdot-)} & = & ([\mathsf{unit},\mathsf{unit}],\mathsf{unit}) \\
+%     \indexOpArity{(-^{-1})} & = & ([\mathsf{unit}],\mathsf{unit})
+%   \end{eqnarray*}
+%   The intended interpretation of these operations is group unit, group
+%   multiplication and group inverse, respectively. When we discuss
+%   equational theories for index expressions in
+%   \autoref{sec:type-equality} below, we will impose the abelian group
+%   laws on these operations.
+
+%   The primitive types for the units of measure example are taken to be
+%   $\PrimTypeSet = \{ \mathsf{num} \}$, with
+%   $\primTyArity(\mathsf{num}) = [\mathsf{unit}]$.
+% \end{example}
 
@@
 Given the above data, the rules in \autoref{fig:indexes-and-types}
-generate two judgements: well-sorted index terms $\Delta \vdash e : s$
-and well-indexed types $\Delta \vdash A \isType$. Since index
+generate two judgements: well-sorted index expressions $\Delta \vdash
+e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
 variables may appear in types, types are judged to be well-indexed
 with respect to an index context $\Delta$. The rules for well-sorted
-index terms are particularly simple: either an index term is a
-variable that appears in the context (rule \TirName{IVar}), or it is
-an application of an index operation taken from $\IndexOpSet$ to other
-index terms (rule \TirName{IOp}). The rules for well-indexed types
-include the usual rules for constructing types of the simply-typed
-$\lambda$-calculus with unit, sum and tuple types (rules
-\TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
+index expressions are particularly simple: either an index expression
+is a variable that appears in the context (rule \TirName{IVar}), or it
+is an application of an index operation taken from $\IndexOpSet$ to
+other index expressions (rule \TirName{IOp}). The rules for
+well-indexed types include the usual rules for constructing types of
+the simply-typed $\lambda$-calculus with unit, sum and tuple types
+(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
 \TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
-a primitive type $\tyPrimNm{X}$ and appropriately sorted index terms
-$e_1,...,e_n$, the well-indexed type $\tyPrim{X}{e_1,...,e_n}$. The
-rule \TirName{TyForall} permits the formation of universally
-quantified types, where the universal quantification ranges over all
-index terms of some sort.
+a primitive type $\tyPrimNm{X}$ and appropriately sorted index
+expressions $e_1,...,e_n$, the well-indexed type
+$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} permits the
+formation of universally quantified types, where the universal
+quantification ranges over all index expressions of some sort.
 
-\subsubsection{Simultaneous Substitutions of Index Terms}
+\subsubsection{Simultaneous Substitution of Index Expressions}
 \label{sec:simultaneous-substitution}
@@
 It will be technically convenient to express substitution of index
-terms in our framework in terms of simultaneous substitutions.  Given
-a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord: s_1,
-..., i_n \mathord: s_n$, a \emph{simultaneous substitution} $\Delta
-\vdash \sigma \Rightarrow \Delta'$ is a sequence of terms $\sigma =
-(e_1,...,e_n)$ such that for each $j$ such that $1 \leq j \leq n$,
-$\Delta \vdash e_j : s_j$. Given a simultaneous substitution $\Delta
-\vdash \sigma = (e_1,...,e_n) \Rightarrow \Delta'$ and a variable $i_j
-\mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for the index term
-$e_j$.
+expressions in our framework in terms of simultaneous substitutions.
+Given a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord:
+s_1, ..., i_n \mathord: s_n$, a \emph{simultaneous substitution}
+$\Delta \vdash \sigma \Rightarrow \Delta'$ is a sequence of
+expressions $\sigma = (e_1,...,e_n)$ such that for each $1 \leq j \leq
+n$, $\Delta \vdash e_j : s_j$. Given a simultaneous substitution
+$\Delta \vdash \sigma = (e_1,...,e_n) \Rightarrow \Delta'$ and a
+variable $i_j \mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for
+the index expression $e_j$.
+
+It will be useful to think of any sequence of sorts as an index
+context. In particular, we will make use of simultaneous substitutions
+of the form $\Delta \vdash \sigma \Rightarrow
+\primTyArity(\tyPrimNm{X})$, since these are exactly sequences of
+index arguments suitable for the primitive type $\tyPrimNm{X}$.
 
@@
 substitutions $\sigma$ such that $\Delta \vdash \sigma \Rightarrow
-\Delta'$. When we discuss the relational interpretation of types
-below, it will be useful to think of any sequence of sorts as an index
-context. \fixme{flesh this out}
-
-Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
-variable/type pair $i\mathord:s$ such that $i$ does not appear in
-$\Delta$, we define the \emph{projection} simultaneous substitution
-$\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
-$\pi_{i\mathord:s} = (i_1,...,i_n)$. The subscript on
-$\pi_{i\mathord:s}$ is the variable/type pair that is being discarded.
-
-For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
-\Delta'$ where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$ and a
-variable/type pair $i\mathord:s$ such that $i$ does not appear in
-either $\Delta$ or $\Delta'$, then we can form the \emph{lifted}
-simultaneous substitution $\Delta,i\mathord:s \vdash
-\sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
-\Delta',i\mathord:s$. Note that we have implicity used the fact that
-well-sortedness of index terms is preserved by addition of extra items
-to the context.
+\Delta'$. The \emph{identity} simultaneous substitution $\Delta \vdash
+\id_\Delta \Rightarrow \Delta$ is just the sequence of variables in
+$\Delta$: $\id_\Delta = (i_1,...,i_n)$ where $\Delta =
+i_1\mathord:s_1,...,i_n\mathord:s_n$. We will also define composition
+of simultaneous substitutions below, but first we must define the
+application of a simultaneous substitutions to index expressions.
 
@@
 
-\subsubsection{Index Term Equality and Type Equality}
+Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
+variable/type pair $i\mathord:s$ such that $i$ does not appear in
+$\Delta$, we define the \emph{projection} simultaneous substitution
+$\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
+$\pi_{i\mathord:s} = (i_1,...,i_n)$. The subscript on
+$\pi_{i\mathord:s}$ is the variable/type pair that is being discarded.
+
+For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
+\Delta'$ where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$ and a
+variable/type pair $i\mathord:s$ such that $i$ does not appear in
+either $\Delta$ or $\Delta'$, then we can form the \emph{lifted}
+simultaneous substitution $\Delta,i\mathord:s \vdash
+\sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
+\Delta',i\mathord:s$. Note that we have implicity used the fact that
+well-sortedness of index expressions is preserved by addition of extra
+items to the context.
+
+\subsubsection{Index Expression Equality and Type Equality}
 \label{sec:type-equality}
 
-\fixme{Consider updating this example} Much of the power of indexing
-types by the terms of an algebraic theory comes from the equations of
-the theory, and the fact that these equations are inherited by the
-types. In the units of measure example, the well-indexed types
-$\tyPrim{X}{u_1\cdot u_2}$ and $\tyPrim{X}{u_2 \cdot u_1}$ are
-considered equal by the type system, due to the commutativity of the
-$-\cdot-$ operation in the theory of abelian groups.
+Much of the power of indexing types by the expressions of an algebraic
+theory comes from the equations of the theory. In the two-dimensional
+geometry example, the types $\tyPrim{vec}{B,t_1 + t_2}$ and
+$\tyPrim{vec}{B, t_2 + t_1}$ are considered equal by the type system,
+due to the commutativity of the $+$ operation.
 
@@
 Given a set $\IndexAxiomSet$ of axioms, we generate the judgmental
-equality between index terms $\Delta \vdash e \equiv e' : s$ by the
-following rules, which allow us to use substitution instances of
+equality between index expressions $\Delta \vdash e \equiv e' : s$ by
+the following rules, which allow us to use substitution instances of
 axioms, and ensure that judgmental equality is a congruence relation
@@
 
-\begin{example}[Origin Invariance]
-  FIXME
+\begin{example}[Two-Dimensional Geometry]
+  Translations form an abelian group under addition, so we assume the
+  abelian group axioms for translations:
+  \begin{displaymath}
+    \begin{array}{l}
+      t : \SynTransl_2 \vdash t + 1 \stackrel{ax}\equiv t : \SynTransl_2 \\
+      t_1, t_2, t_3 : \SynTransl_2 \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl_2 \\
+      t : \SynTransl_2 \vdash t + (-t) \stackrel{ax}\equiv 1 : \SynTransl_2 \\
+      t_1, t_2 : \SynTransl_2 \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl_2 \\
+    \end{array}
+  \end{displaymath}
+  Similarly, the sort of scale factors $\SynGL_1$ forms an abelian
+  group under multiplication, and the sorts $\SynGL_2$ and
+  $\SynOrth_2$ form (non-abelian) multiplicative groups, so we assume
+  the appropriate axioms. We also assume that the operations $\iota_O,
+  \iota_1, \det$ and $|\cdot|$ are group homomorphisms, and that
+  expressions of the form $\iota_1(s)$ commute with group
+  multiplication in the sort $\SynGL_2$. The absolute value of the
+  determinant of orthgonal transformations is always $1$, so we also
+  assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
+  maps have a determined determinant: $\det(\iota_1(s))
+  \stackrel{ax}\equiv s \cdot s$.
 \end{example}
 
-\begin{example}[Units of Measure]
-  For the units of measure system, seen as an instantiation of our
-  general framework, the set $\IndexAxiomSet$ consists of the abelian
-  group axioms:
-  \begin{displaymath}
-    \begin{array}{l}
-      u : \mathsf{unit} \vdash u \cdot 1 \stackrel{ax}\equiv u : \mathsf{unit} \\
-      u_1, u_2, u_3 : \mathsf{unit} \vdash u_1 \cdot (u_2 \cdot u_3) \stackrel{ax}\equiv (u_1 \cdot u_2) \cdot u_3 : \mathsf{unit} \\
-      u_1, u_2 : \mathsf{unit} \vdash u_1 \cdot u_2 \stackrel{ax}\equiv u_2 \cdot u_1 : \mathsf{unit} \\
-      u : \mathsf{unit} \vdash u \cdot u^{-1} \stackrel{ax}\equiv 1 : \mathsf{unit} \\
-    \end{array}
-  \end{displaymath}
-\end{example}
+% \begin{example}[Logical Information Flow]
+%   \fixme{Do this}
+% \end{example}
+
+% \begin{example}[Units of Measure]
+%   For the units of measure system, seen as an instantiation of our
+%   general framework, the set $\IndexAxiomSet$ consists of the abelian
+%   group axioms:
+%   \begin{displaymath}
+%     \begin{array}{l}
+%       u : \mathsf{unit} \vdash u \cdot 1 \stackrel{ax}\equiv u : \mathsf{unit} \\
+%       u_1, u_2, u_3 : \mathsf{unit} \vdash u_1 \cdot (u_2 \cdot u_3) \stackrel{ax}\equiv (u_1 \cdot u_2) \cdot u_3 : \mathsf{unit} \\
+%       u_1, u_2 : \mathsf{unit} \vdash u_1 \cdot u_2 \stackrel{ax}\equiv u_2 \cdot u_1 : \mathsf{unit} \\
+%       u : \mathsf{unit} \vdash u \cdot u^{-1} \stackrel{ax}\equiv 1 : \mathsf{unit} \\
+%     \end{array}
+%   \end{displaymath}
+% \end{example}
 
 The judgmental equality $\Delta \vdash e \equiv e' : s$ on index
-terms generates the judgmental equality $\Delta \vdash A \equiv B
-\isType$ on types. The basic rule generating judgmental equality on
+expressions generates the judgmental equality $\Delta \vdash A \equiv
+B \isType$ on types. The basic rule generating judgmental equality on
 types states that two applications of primitive types are equal if
-their index term arguments are equal:
+their index expression arguments are equal:
 \begin{displaymath}
@@
 relation (i.e.~judgmental equality on types is reflexive, symmetric
-and transitive). For example, for universally quantified types we have
-the following congruence rule:
-\begin{displaymath}
-  \inferrule*
-  {\Delta, i : s \vdash A \equiv B \isType}
-  {\Delta \vdash \forall i\mathord:s.A \equiv \forall i\mathord:s.B \isType}
-\end{displaymath}
-The congruence rules for the other type formers are similar.
+and transitive).
+% For example, for universally quantified types we have
+% the following congruence rule:
+% \begin{displaymath}
+%   \inferrule*
+%   {\Delta, i : s \vdash A \equiv B \isType}
+%   {\Delta \vdash \forall i\mathord:s.A \equiv \forall i\mathord:s.B \isType}
+% \end{displaymath}
+% The congruence rules for the other type formers are similar.
 
-The judgmental equalities on index terms and on types are preserved by
-substitution:
 \begin{lemma}
@@
 
-We define a pair of simultaneous substitutions $\sigma, \sigma' :
-\Delta \Rightarrow \Delta'$ to be judgmentally equal if their
-component terms are judgmentally equal in the context $\Delta$: i.e.~
-if $\Delta \vdash e_j \equiv e'_j : s_j$, for all $j$. We write
+A pair of simultaneous substitutions $\Delta \vdash \sigma, \sigma'
+\Rightarrow \Delta'$ are defined to be judgmentally equal if their
+component expressions are judgmentally equal in the context $\Delta$:
+i.e.~ if $\Delta \vdash e_j \equiv e'_j : s_j$, for all $j$. We write
 $\Delta \vdash \sigma \equiv \sigma' \Rightarrow \Delta'$ when two
@@
 determined solely by the primitive type $\tyPrimNm{X}$ and not by the
-index terms $e_1,...,e_n$. Accordingly, we assume that each primitive
-type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
+index expressions $e_1,...,e_n$. Accordingly, we assume that each
+primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
 $\tyPrimSem{\tyPrimNm{X}}$. We extend this assignment to every
@@
 \begin{displaymath}
-  \begin{array}{@{\hspace{0.2em}}c@{\hspace{0em}}c}
-    \begin{array}{l}
+  \begin{array}{@{\hspace{0em}}c@{\hspace{0em}}c}
+    \begin{array}{@{}l}
       \tySem{\tyUnit} \isDefinedAs \{*\} \\
@@
 
-\begin{example}[Origin Invariance]
-  \fixme{do this}
-\end{example}
-
-\begin{example}[Units of Measure] Kennedy's original units of measure
-  system seen as an instantiation of our system uses the assignment
-  $\tyPrimSem{\mathsf{unit}} \isDefinedAs \mathbb{Q}$, where
-  $\mathbb{Q}$ is the set of rational numbers.
-
-  FIXME: Also do the primitive operations
-\end{example}
-
 % The index-erasure interpretation completely ignores index expressions
@@
 
+\begin{example}[Two-Dimensional Geometry]
+  The two-dimensional geometry instantiation of the general framework
+  uses the assignment $\tyPrimSem{\tyPrimNm{vec}} = \mathbb{R}^2$ and
+  $\tyPrimSem{\tyPrimNm{real}} = \mathbb{R}$.
+\end{example}
+
+% \begin{example}[Logical Information Flow]
+%   \fixme{do this}
+% \end{example}
+
 \subsubsection{The Relational Interpretation of Types}
@@
 contexts. We assign interpretations to index contexts in terms of
-\emph{relational environments}, which we now define. \fixme{Backwards ref}
+\emph{relational environments}, which we now define.
 
@@
 \end{displaymath}
-Every relation environment must respect index term equality: for any
-$\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous substitutions
-$\Delta \vdash \sigma \Rightarrow \primTyArity(\tyPrimNm{X})$ and
-$\Delta \vdash \sigma' \Rightarrow \primTyArity(\tyPrimNm{X})$ such
-that $\Delta \vdash \sigma \equiv \sigma' \Rightarrow
-\primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\ \sigma = \rho\
-\tyPrimNm{X}\ \sigma'$.
+Every relation environment must respect index expression equality: for
+any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
+substitutions $\Delta \vdash \sigma \Rightarrow
+\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
+\primTyArity(\tyPrimNm{X})$ such that $\Delta \vdash \sigma \equiv
+\sigma' \Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\
+\tyPrimNm{X}\ \sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
 
@@
 \begin{enumerate}
-\item Closure under simultaneous substitution of index terms: if $\rho
-  \in \relEnv{E}(\Delta)$ and $\Delta' \vdash \sigma \Rightarrow
-  \Delta$, then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
+\item Closure under simultaneous substitutions: if $\rho \in
+  \relEnv{E}(\Delta)$ and $\Delta' \vdash \sigma \Rightarrow \Delta$,
+  then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
 \item If we have a pair of relation environments $\rho_1 \in
@@
   \rho_2$.
-
   % such that the outer edges of the
@@
 types that we define below behaves correctly with respect to
-simultaneous substitution of index terms
-(\lemref{lem:tyeqsubst-relational}, below). The second condition may
-seem mysterious at first, but it is essential for proving that the
-relational interpretation of universal quantification types behaves
-correctly with respect to application of simultaneous substitution.
+simultaneous substitution of index expressions
+(\lemref{lem:tyeqsubst-relational}, below).
 
-\begin{example}[Units of Measure]
-  We give two examples of assignments of sets of relational
-  environments to index contexts in the units of measure example. The
-  second example is a refinment of the first. FIXME: some general
-  words about relational environments in these cases.
-  \begin{enumerate}
-  \item FIXME: assign a positive scaling factor to each variable in an
-    index context, extend homomorphically to all index terms, and then
-    use this to define a scaling relation
-  \item FIXME: given an index context, assume a subgroup and a
-    homomorphism into the scaling factors
-  \end{enumerate}
-\end{example}
+% The second condition may
+% seem mysterious at first, but it is essential for proving that the
+% relational interpretation of universal quantification types behaves
+% correctly with respect to application of simultaneous substitution.
+
+% \begin{example}
+%   The construction of relational environments is the most subtle part
+%   of our development. \fixme{complete this part}
+% \end{example}
+
+% \begin{example}[Units of Measure]
+%   We give two examples of assignments of sets of relational
+%   environments to index contexts in the units of measure example. The
+%   second example is a refinment of the first. FIXME: some general
+%   words about relational environments in these cases.
+%   \begin{enumerate}
+%   \item FIXME: assign a positive scaling factor to each variable in an
+%     index context, extend homomorphically to all index expressions,
+%     and then use this to define a scaling relation
+%   \item FIXME: given an index context, assume a subgroup and a
+%     homomorphism into the scaling factors
+%   \end{enumerate}
+% \end{example}
 
@@
 
-The following lemma states that the relation interpretation of types
-that we have defined in this section behaves well: the first part of
-the lemma states that two types that are judgmentally equal are given
-equal relational interpretations, and the second part states that
-substitution of index terms in types can be interpreted by the
-composition of relational environments with simultaneous
-substitutions. Note that both parts of the lemma depend on
-\lemref{lem:tyeqsubst-erasure} to be ``well-typed''.
+% The following lemma states that the relation interpretation of types
+% that we have defined in this section behaves well: the first part of
+% the lemma states that two types that are judgmentally equal are given
+% equal relational interpretations, and the second part states that
+% substitution of index expressions in types can be interpreted by the
+% composition of relational environments with simultaneous
+% substitutions.
 \begin{lemma}\label{lem:tyeqsubst-relational}
@@
     \rsem{B}{\relEnv{E}}{\rho}$;
-  \item If $\Delta' \vdash A \isType$ and $\sigma : \Delta \Rightarrow
-    \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
+  \item If $\Delta' \vdash A \isType$ and $\Delta \vdash \sigma
+    \Rightarrow \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
     $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
@@
 \end{lemma}
+
+Note that both parts of \lemref{lem:tyeqsubst-relational} depend on
+\lemref{lem:tyeqsubst-erasure} to be ``well-typed''.
 
@@
 interpretation of its type: this is the abstraction theorem for every
-calculus in our general framework. In the \fixme{next section} we will
-use this theorem to derive several interesting results about the
-programs expressible in instances of our general framework.
+calculus in our general framework.
 
@@
 
-The typing rules for our system define the well-typed terms with
+The typing rules for our system define the well-typed programs with
 respect to an index context $\Delta$ and a type context $\Delta \vdash
@@
 incorporated into the type system via the rule \TirName{TyEq}, which
-allows for a term that is judged to have type $A$ to also have any
+allows for a program that is judged to have type $A$ to also have any
 other equal type $B$.
@@
 \lemref{lem:tyeqsubst-erasure}, we define this function directly on
-the syntax of well-typed terms, rather than on the typing
+the syntax of well-typed programs, rather than on the typing
 derivations. We define $\tmSem{e}$ by the following clauses:
@@
 We now state and prove the abstraction theorem for well-typed programs
-in our system. To state and prove this theorem for open terms, we
+in our system. To state and prove this theorem for open programs, we
 first need to extend the relational interpretation of types to typing
 contexts. The relational interpretation of contexts is defined by the
-following clauses, where we have used the pairing operation
-$-\relTimes-$ defined in \autoref{sec:relational-semantics}.
-\begin{eqnarray*}
-  \rsem{\epsilon}{\relEnv{E}}{\rho} & \isDefinedAs & \{(*,*)\} \\
-  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & \isDefinedAs & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
-\end{eqnarray*}
-The relational interpretation of contexts inherits from the relational
-interpretation of types the property of interpreting the application
-of simultaneous substitutions as composition:
+following clauses, where we have used the relational lifting of
+cartesian product $-\relTimes-$ defined in
+\autoref{sec:relational-semantics}.
+\begin{displaymath}
+  \begin{array}{@{}l@{\hspace{0.1em}\isDefinedAs\hspace{0.1em}}l@{\hspace{1.5em}}l@{\hspace{0.1em}\isDefinedAs\hspace{0.1em}}l}
+    \rsem{\epsilon}{\relEnv{E}}{\rho} & \{(*,*)\} &
+    \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
+  \end{array}
+\end{displaymath}
+% The relational interpretation of contexts inherits from the relational
+% interpretation of types the property of interpreting the application
+% of simultaneous substitutions as composition:
 \begin{lemma}\label{lem:ctxtsubst-rel}

### 2012-06-12 paper/section-instantiations.tex
Add AUCTeX commands to deal with multifile document
Substantial changes to the motivating examples section
Major changes
- Refined set of contributions
- "Complete" the motivating examples section
- Fill in some references for the background section
- "Complete" the general framework section
- Sketch the "consequences" section
- Sketch the "discussion" section

@@
-\section{Instantiations of the General Framework}
+\section{Consequences of the Abstraction Theorem}
 \label{sec:instantiations}
 
-\subsection{Multiplicative Group Examples}
-\label{sec:multiplicative-examples}
+\fixme{explain the general setup, with a set of operations for each
+  instantiation.}
 
-\subsection{Additive Group Examples}
-\label{sec:additive-examples}
+\subsection{Theorems for Free}
+\label{sec:theorems-for-free}
 
-\subsection{Logic Examples}
-\label{sec:logic-examples}
+Wadler's influential ``Theorems for Free!''  \cite{wadler89theorems}
+emphasised a particular aspect of Reynolds' theory of relational
+parametricity: the fact that we can read off theorems about a program
+simply by looking at the relational interpretation of its type. We now
+state some free theorems that are derivable in each of the three main
+examples we introduced in \autoref{sec:motivating-examples}.
+
+\subsubsection{Two-Dimensional Geometry}
+\label{sec:geometry-consequences}
+
+(In)variance properties
+
+\subsubsection{Logical Information Flow}
+\label{sec:information-flow-consequences}
+
+Non-interference properties
+
+\subsubsection{Metric Spaces}
+\label{sec:metric-space-consequences}
+
+Continuity via $\epsilon$-$\delta$ proofs.
+
+\subsection{Types Indexed by Abelian Groups}
+\label{sec:types-indexed-abelian-groups}
+
+\begin{enumerate}
+\item Two cases: when we have a total ``division'' operation, and when
+  we don't.
+\item Type isomorphisms, based on Smith Normal Form
+\item Non-definability results, based on the non-compositional
+  relational environments
+\item Alteration of the non-definability results in the light of
+  addition of a square root operation
+\item Mention the variation we can obtain by changing some types,
+  e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
+\end{enumerate}
+
+%%% Local Variables:
+%%% TeX-master: "paper"
+%%% End:

### 2012-06-12 paper/section-introduction.tex
Major changes
- Refined set of contributions
- "Complete" the motivating examples section
- Fill in some references for the background section
- "Complete" the general framework section
- Sketch the "consequences" section
- Sketch the "discussion" section

--- 2012-05-22 paper/section-introduction.tex
@@
 
-There are three main points to get across:
-\begin{enumerate}
-\item Why algebraically indexed types?
-\item Why relational parametricity?
-\item Why study them together?
-\end{enumerate}
+% Possible angles of attack:
+% \begin{itemize}
+% \item Parametric polymorphic types allow us to prevent
+%   over-specification of the behaviour of programs. For instance, the
+%   type $\forall \alpha. [\alpha] \to [\alpha]$ is a generalisation of
+%   the types $[\mathsf{int}] \to [\mathsf{int}]$ and $[\mathsf{char}]
+%   \to [\mathsf{char}]$. Either of the latter two types over-specify
+%   the behaviour of the function.
+% \item There are other cases of programs that are over-specified. The
+%   leading example we just below is of geometric programs that
+%   manipulate coordinate data. Often, programs that manipulate
+%   coordinate data are insensitive to geometric transformations. For
+%   example, a program that computes the area of a triangle described by
+%   three points is insensitive to translations or rotations applied to
+%   all three points.
+% \item 
+% \end{itemize}
+
+% There are three main points to get across:
+% \begin{enumerate}
+% \item Why algebraically indexed types?
+% \item Why relational parametricity?
+% \item Why study them together?
+% \end{enumerate}
 
@@
 \label{sec:contributions}
+
+\begin{itemize}
+\item Formulation of a general class of type systems that can either
+  be used as programming languages in their own right, or as the
+  targets of type-based analyses.
+\item A collection of compelling examples of algebraically indexed
+  types, including a novel type system for geometry, a refined type
+  system for information flow, based on logic, and a simple type
+  system for basic continuity analysis.
+\item Deduction of useful free theorems in each of our main examples.x
+\item A refined relational interpretation in order to derive
+  non-definability results. Fixing a minor problem in Kennedy's work.
+\end{itemize}
+
+%%% Local Variables:
+%%% TeX-master: "paper"
+%%% End:

### 2012-06-12 paper/section-general-framework.tex
Add existential types, and extend the proof of the substitution lemma

@@
     \inferrule* [right=TyForall] %FIXME: macroize forall
-    {\Delta, i : s \vdash A \isType}
+    {\Delta, i \mathord: s \vdash A \isType}
     {\Delta \vdash \forall i \mathord: s. A \isType}
+
+    \inferrule* [right=TyExist]
+    {\Delta, i \mathord: s \vdash A \isType}
+    {\Delta \vdash \exists i \mathord: s. A \isType}
   \end{mathpar}
@@
 formation of universally quantified types, where the universal
-quantification ranges over all index expressions of some sort.
+quantification ranges over all index expressions of some
+sort. Existential types are constructed using the \TirName{TyExist}
+rule.
 
@@
   \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A
+
+  \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:S. \sigma_{i\mathord:s}^*A
 \end{mathpar}
@@
       \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
-      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A}
+      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A} \\
+      \tySem{\exists i\mathord:s. A} \isDefinedAs \tySem{A}
     \end{array}
@@
 the interpretations of the function, tuple and sum types are simply
-the corresponding constructions on sets. The interpretation of the
-universal quantifier ignores the indexing: the interpretation of the
-type $\forall i\mathord:s.A$ is exactly the interpretation of the type
-$A$.
+the corresponding constructions on sets. The interpretations of the
+universal and existential quantifiers ignore the indexing: the
+interpretation of the type $\forall i\mathord:s.A$ is exactly the
+interpretation of the type $A$, and likewise for $\exists
+i\mathord:s.A$.
 
@@
 = \rho \}$. The set of extensions of a relational environment will be
-used in the interpretation of the universal quantification type.
+used in the interpretation of the universal and existential type
+formers.
 
@@
   \rsem{A + B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i\mathord:s} \}
+  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:s} \} \\
+  \rsem{\exists i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcup\{ \rsem{A}{\relEnv{E}}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:s} \}
 \end{eqnarray*}

### 2012-06-14 paper/geometry-twocolumn.tex
Macro-ize the names of the groups with dimensions, so we can change notation easily if need be

@@
 as translations, we write the collection of all change of origin
-offsets as $\Transl_2$. By quantifying over all changes of origin we
+offsets as $\Transl{2}$. By quantifying over all changes of origin we
 will be able to ensure that our programs are invariant with respect to
 the particular choice of origin. The members of the
-$\Transl_2$-indexed family of binary relations $\{ R_{\vec{t}}
-\subseteq \mathbb{R}^2 \times \mathbb{R}^2 \}_{\vec{t} \in \Transl_2}$
+$\Transl{2}$-indexed family of binary relations $\{ R_{\vec{t}}
+\subseteq \mathbb{R}^2 \times \mathbb{R}^2 \}_{\vec{t} \in \Transl{2}}$
 are defined as follows:
@@
 to the choice of origin. We use Reynolds' idea of preservation of
-relations to formalise this: for any $\vec{t} \in \Transl_2$ (i.e.~any
+relations to formalise this: for any $\vec{t} \in \Transl{2}$ (i.e.~any
 change of origin), the function $f$ should satisfy the following
@@
 equivalent to the following statement expressing invariance of $f$
-with respect to changes of origin: for all $\vec{t} \in \Transl_2$:
+with respect to changes of origin: for all $\vec{t} \in \Transl{2}$:
 \begin{equation}\label{eq:f-invariant-frame}
@@
 quantification over changes of origin into the language of types.  We
-use the notation $\forall t \mathord: \SynTransl_2. A$ for
+use the notation $\forall t \mathord: \SynTransl{2}. A$ for
 quantification over all 2-dimensional translations (i.e.~choices of
-origin) $t$. We refer to $\SynTransl_2$ as the \emph{sort} of
+origin) $t$. We refer to $\SynTransl{2}$ as the \emph{sort} of
 $t$. Note that we have used a different font to distinguish the
-semantic group $\Transl_2$ from the syntactic sort $\SynTransl_2$.
+semantic group $\Transl{2}$ from the syntactic sort $\SynTransl{2}$.
 
-The sort $\SynTransl_2$ represents an abelian group, so we can combine
+The sort $\SynTransl{2}$ represents an abelian group, so we can combine
 its elements using the usual group operations. We write operations
@@
 introduce a new type parameterised by expressions $e$ of sort
-$\SynTransl_2$ to represent points \emph{and} vectors:
+$\SynTransl{2}$ to represent points \emph{and} vectors:
 $\tyPrim{vec}{e}$. We have used the name $\tyPrimNm{vec}$, even though
@@
 We now sketch the interpretation of the types $\tyPrim{vec}{e}$ and
-$\forall t\mathord:\SynTransl_2.A$. In our calculus, each type has two
+$\forall t\mathord:\SynTransl{2}.A$. In our calculus, each type has two
 interpretations: an index-erasure interpretation that ignores the
@@
 relation on the index-erasure interpretation. For now, we assume we
-are given some way of turning an expression $e$ of sort $\SynTransl_2$
-into an actual element $\sem{e}\rho$ of the group $\Transl_2$, under
+are given some way of turning an expression $e$ of sort $\SynTransl{2}$
+into an actual element $\sem{e}\rho$ of the group $\Transl{2}$, under
 some environment $\rho$ which tells how to interpret the free
@@
 The index-erasure interpetation of universal quantification $\forall
-t\mathord:\SynTransl_2.A$ is simply the interpretation of the
+t\mathord:\SynTransl{2}.A$ is simply the interpretation of the
 underlying type $A$. The relational interpretation intersects the
@@
   \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\forall t\mathord:\SynTransl_2.A} & \tySem{A} \\
-    \rsem{\forall t\mathord:\SynTransl_2.A}{}\rho & \bigcap\{ \rsem{A}{}{\rho'} \sepbar \rho' \in \extends{\rho}{t\mathord:\SynTransl_2} \}
+    \tySem{\forall t\mathord:\SynTransl{2}.A} & \tySem{A} \\
+    \rsem{\forall t\mathord:\SynTransl{2}.A}{}\rho & \bigcap\{ \rsem{A}{}{\rho'} \sepbar \rho' \in \extends{\rho}{t\mathord:\SynTransl{2}} \}
   \end{array}
 \end{displaymath}
-The exact definition of $\extends{\rho}{t\mathord:\SynTransl_2}$
+The exact definition of $\extends{\rho}{t\mathord:\SynTransl{2}}$
 depends on which (possibly non-compositional) interpretation of
@@
 think of relational environments as simply mappings from free
-variables of sort $\SynTransl_2$ to the group $\Transl_2$, and
+variables of sort $\SynTransl{2}$ to the group $\Transl{2}$, and
 extension as just adding an extra component to the tuple. The general
@@
 \begin{displaymath}
-  \mathrm{f} : \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}
+  \mathrm{f} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}
 \end{displaymath}
@@
   \begin{array}{l}
-    \mathrm{affComb} :\forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\
+    \mathrm{affComb} :\forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\
     \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} + r\vec{v_2}
@@
     \begin{array}{@{}l}
-      \mathrm{quadB\acute{e}zier} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \\
+      \mathrm{quadB\acute{e}zier} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \\
       \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\
@@
 \begin{displaymath}
-  (+) : \forall t_1, t_2 \mathord: \SynTransl_2.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
+  (+) : \forall t_1, t_2 \mathord: \SynTransl{2}.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
 \end{displaymath}
@@
 \begin{displaymath}
-  \mathrm{negate} : \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
+  \mathrm{negate} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
 \end{displaymath}
@@
   \begin{array}{l}
-    (-) : \forall t_1,t_2 \mathord:\SynTransl_2.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2} \\
+    (-) : \forall t_1,t_2 \mathord:\SynTransl{2}.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2} \\
     (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
@@
   \begin{array}{l}
-    \mathrm{offset} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{0} \\
+    \mathrm{offset} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{0} \\
     \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
@@
   \begin{array}{l}
-    \mathrm{moveBy} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to \tyPrim{vec}{t} \\
+    \mathrm{moveBy} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to \tyPrim{vec}{t} \\
     \mathrm{moveBy}\ [t]\ p\ v = p + v
@@
   \begin{displaymath}
-    \tau_{n} \isDefinedAs \forall t\mathord:\SynTransl_2.\ \underbrace{\tyPrim{vec}{t} \to ... \tyPrim{vec}{t}}_{n+1\textrm{ times}} \to \tyPrimNm{real}
+    \tau_{n} \isDefinedAs \forall t\mathord:\SynTransl{2}.\ \underbrace{\tyPrim{vec}{t} \to ... \tyPrim{vec}{t}}_{n+1\textrm{ times}} \to \tyPrimNm{real}
   \end{displaymath}
@@
   inhabitants will be invariant under change of origin, due to the
-  quantification over all $t$ in $\SynTransl_2$. Therefore, we may as well pick
+  quantification over all $t$ in $\SynTransl{2}$. Therefore, we may as well pick
   one of the input points as the origin and assume that all the other
@@
   also possible to show that certain types are uninhabited. For
-  example, the type $\forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t
+  example, the type $\forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t
     + t} \to \tyPrim{vec}{t}$ does not have any inhabitants. We
@@
 linear map, and the collection of all invertible linear maps on
-$\mathbb{R}^2$ forms the group $\GL_2$. We reflect in our language the
-group $\GL_2$ as a new indexing sort $\SynGL_2$, which has
+$\mathbb{R}^2$ forms the group $\GL{2}$. We reflect in our language the
+group $\GL{2}$ as a new indexing sort $\SynGL{2}$, which has
 (non-abelian) group structure that we will write multiplicatively. We
 also extend the $\tyPrimNm{vec}$ type to be indexed by elements of
-$\SynGL_2$ as well as $\SynTransl_2$. Thus, $\tyPrim{vec}{B,t}$ is a
+$\SynGL{2}$ as well as $\SynTransl{2}$. Thus, $\tyPrim{vec}{B,t}$ is a
 vector which varies with change of basis $B$ and change of origin
@@
   \begin{array}[t]{@{}l}
-    \forall B \mathord: \SynGL_2, t \mathord: \SynTransl_2.\\
+    \forall B \mathord: \SynGL{2}, t \mathord: \SynTransl{2}.\\
     \hspace{0.2cm} \tyPrim{vec}{B,t} \to \tyPrimNm{real} \to \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,t}
@@
   \begin{array}[t]{@{}l}
-    \forall B \mathord: \SynGL_2, t_1,t_2 \mathord: \SynTransl_2. \\
+    \forall B \mathord: \SynGL{2}, t_1,t_2 \mathord: \SynTransl{2}. \\
     \hspace{0.2cm}\tyPrim{vec}{B,t_1} \to \tyPrim{vec}{B,t_2} \to \tyPrim{vec}{B,t_1 + t_2}
@@
   \\
-  \mathrm{negate} & : & \forall B \mathord: \SynGL_2, t \mathord: \SynTransl_2.\ \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,-t} \\
-  0 & : & \forall B \mathord: \SynGL_2.\ \tyPrim{vec}{B,0} \\
-  (*) & : & \forall B \mathord: \SynGL_2.\ \tyPrimNm{real} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0}
+  \mathrm{negate} & : & \forall B \mathord: \SynGL{2}, t \mathord: \SynTransl{2}.\ \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,-t} \\
+  0 & : & \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \\
+  (*) & : & \forall B \mathord: \SynGL{2}.\ \tyPrimNm{real} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0}
 \end{eqnarray*}
@@
 assign it a type. The dot product is not invariant under the action of
-either of the groups $\GL_2$ and $\Transl_2$ that we have considered
+either of the groups $\GL{2}$ and $\Transl{2}$ that we have considered
 so far. However, the dot product is invariant under the subgroup
-$\Orth_2$ of $\GL_2$ of \emph{orthogonal} linear transformations:
+$\Orth{2}$ of $\GL{2}$ of \emph{orthogonal} linear transformations:
 invertible linear maps whose determinant is either $1$ or $-1$. We
-introduce orthogonal transformations as a new sort $\SynOrth_2$, and
+introduce orthogonal transformations as a new sort $\SynOrth{2}$, and
 overload the multiplicative group operations syntax for inhabitants of
-$\SynOrth_2$. We also assume a function $\iota_O$ that takes $e :
-\SynOrth_2$ to $\iota_O(e) : \SynGL_2$. We may now assign a type to
-the dot product, using quantification over $\SynOrth_2$:
+$\SynOrth{2}$. We also assume a function $\iota_O$ that takes $e :
+\SynOrth{2}$ to $\iota_O(e) : \SynGL{2}$. We may now assign a type to
+the dot product, using quantification over $\SynOrth{2}$:
 \begin{displaymath}
-  (\cdot) : \forall O \mathord: \SynOrth_2.\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
+  (\cdot) : \forall O \mathord: \SynOrth{2}.\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
 \end{displaymath}
@@
   \begin{eqnarray*}
-    0   &:& \forall s \mathord:\SynGL_1.\ \tyPrim{real}{s} \\
+    0   &:& \forall s \mathord:\SynGL{1}.\ \tyPrim{real}{s} \\
     1   &:& \tyPrim{real}{1} \\
-    (+) &:& \forall s \mathord:\SynGL_1.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
-    (-) &:& \forall s \mathord:\SynGL_1.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
-    (*) &:& \forall s_1,s_2 \mathord:\SynGL_1.\ \tyPrim{real}{s_1} \to \tyPrim{real}{s_2} \to \tyPrim{real}{s_1s_2} \\
+    (+) &:& \forall s \mathord:\SynGL{1}.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
+    (-) &:& \forall s \mathord:\SynGL{1}.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
+    (*) &:& \forall s_1,s_2 \mathord:\SynGL{1}.\ \tyPrim{real}{s_1} \to \tyPrim{real}{s_2} \to \tyPrim{real}{s_1s_2} \\
     (/) &:&
     \begin{array}[t]{@{}l@{}l}
-      \forall s_1,s_2 \mathord:\SynGL_1.\ \tyPrim{real}{s_1}\ & \to \tyPrim{real}{s_2} \\
+      \forall s_1,s_2 \mathord:\SynGL{1}.\ \tyPrim{real}{s_1}\ & \to \tyPrim{real}{s_2} \\
       &\to \tyPrim{real}{s_1s_2^{-1}} + \tyPrimNm{unit} \\
     \end{array}\\
-    \mathrm{abs} &:& \forall s \mathord:\SynGL_1.\ \tyPrim{real}{s} \to \tyPrim{real}{|s|} %\\
+    \mathrm{abs} &:& \forall s \mathord:\SynGL{1}.\ \tyPrim{real}{s} \to \tyPrim{real}{|s|} %\\
 %    \mathrm{sqrt} &:& \tyPrim{real}{1} \to \tyPrim{real}{1}
@@
 
-We therefore augment our calculus with a new sort $\SynGL_1$ of scale
+We therefore augment our calculus with a new sort $\SynGL{1}$ of scale
 factors (i.e.~$1$-dimensional invertible linear maps). Semantically,
-$\SynGL_1$ ranges over the non-zero real numbers and forms an abelian
+$\SynGL{1}$ ranges over the non-zero real numbers and forms an abelian
 group, which we write multiplicatively. We also add two new
 operations: determinant, $\det B$, which takes an inhabitant of
-$\SynGL_2$ to its determinant in $\SynGL_1$, and the operation of
+$\SynGL{2}$ to its determinant in $\SynGL{1}$, and the operation of
 absolute value that takes scaling factors to scaling factors which we
 write as $|e|$. We also refine the type $\tyPrimNm{real}$ of real
-numbers to be indexed by the sort $\SynGL_1$: $\tyPrim{real}{e}$,
+numbers to be indexed by the sort $\SynGL{1}$: $\tyPrim{real}{e}$,
 where the old type $\tyPrimNm{real}$ is just $\tyPrim{real}{1}$. The
@@
 
-With the additional sort $\SynGL_1$, and the refined type of real
+With the additional sort $\SynGL{1}$, and the refined type of real
 numbers, we can assign a type to the cross product:
 \begin{displaymath}
-  (\times) : \forall B \mathord: \SynGL_2.\ \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B}
+  (\times) : \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B}
 \end{displaymath}
@@
     \begin{array}{@{}l}
-      \mathrm{area} : \forall B\mathord:\SynGL_2, t\mathord:\SynTransl_2.\ \\
+      \mathrm{area} : \forall B\mathord:\SynGL{2}, t\mathord:\SynTransl{2}.\ \\
       \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{|\det B|} \\
@@
     \begin{array}{@{}l}
-      \mathrm{area} : \forall O\mathord:\SynOrth_2, t\mathord:\SynTransl_2.\ \\
+      \mathrm{area} : \forall O\mathord:\SynOrth{2}, t\mathord:\SynTransl{2}.\ \\
       \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{real}{1}
@@
 one-dimensional scaling invariance. We add an operation $\iota_1$
-that, semantically, takes scale factors in $\GL_1$ to invertible
-linear maps in $\GL_2$: i.e.~takes numbers $s$ to matrices $\left(
+that, semantically, takes scale factors in $\GL{1}$ to invertible
+linear maps in $\GL{2}$: i.e.~takes numbers $s$ to matrices $\left(
   \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$.  This
@@
     \begin{array}{@{}l}
-      \mathrm{area} : \forall s\mathord:\SynGL_1, t\mathord:\SynTransl_2.\ \\
+      \mathrm{area} : \forall s\mathord:\SynGL{1}, t\mathord:\SynTransl{2}.\ \\
       \hspace{0.8cm} \tyPrim{vec}{\iota_1s, t} \to \tyPrim{vec}{\iota_1s, t} \to \tyPrim{vec}{\iota_1s, t} \to \tyPrim{real}{s^2}
@@
 all other invertible linear maps, and form the centre of the group
-$\GL_2$. Thus we add the equation $\iota_1(s)B = B\iota_1(s)$ to our
+$\GL{2}$. Thus we add the equation $\iota_1(s)B = B\iota_1(s)$ to our
 calculus.
@@
 \begin{displaymath}
-  (*) : \forall s \mathord: \SynGL_1, B \mathord: \SynGL_2.\ \tyPrim{real}{s} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{\iota_1(s)B,0}
+  (*) : \forall s \mathord: \SynGL{1}, B \mathord: \SynGL{2}.\ \tyPrim{real}{s} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{\iota_1(s)B,0}
 \end{displaymath}
@@
   \begin{array}{l@{}l}
-    (\cdot) : \forall s : \SynGL_1, O : \SynOrth_2.\ & \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\\
+    (\cdot) : \forall s \mathord: \SynGL{1}, O \mathord: \SynOrth{2}.\ & \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\\
     &\tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to \tyPrim{real}{s^2}
@@
   \begin{displaymath}
-    \forall s : \SynGL_2.\ \tyPrim{real}{s^2} \to \tyPrim{real}{s}
+    \forall s \mathord: \SynGL{2}.\ \tyPrim{real}{s^2} \to \tyPrim{real}{s}
   \end{displaymath}

### 2012-06-14 paper/section-general-framework.tex
Add program syntax for dealing with existential types, and extend the proof of the abstraction theorem
Small fixes
Macro-ize the names of the groups with dimensions, so we can change notation easily if need be

@@
   geometric groups that we mentioned in
-  \autoref{sec:motivating-examples}. Thus, $\SortSet = \{\SynTransl_2,
-  \SynGL_2, \SynOrth_2, \SynGL_1 \}$. For the index operations, we
-  have additive group structure on $\SynTransl_2$, multiplicative
-  group structure on $\SynGL_1$, $\SynGL_2$ and $\SynOrth_2$,
-  injections from $\SynOrth_2$ and $\SynGL_1$ into $\SynGL_2$,
+  \autoref{sec:motivating-examples}. Thus, $\SortSet = \{\SynTransl{2},
+  \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$. For the index operations, we
+  have additive group structure on $\SynTransl{2}$, multiplicative
+  group structure on $\SynGL{1}$, $\SynGL{2}$ and $\SynOrth{2}$,
+  injections from $\SynOrth{2}$ and $\SynGL{1}$ into $\SynGL{2}$,
   determinant and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
   1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, |\cdot| \}$, where
-  $G \in \{\SynGL_1, \SynGL_2, \SynOrth_1\}$, with the following
+  $G \in \{\SynGL{1}, \SynGL{2}, \SynOrth{2}\}$, with the following
   assignment of arities:
@@
     \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
-      \indexOpArity(0) & ([], \SynTransl_2) &
+      \indexOpArity(0) & ([], \SynTransl{2}) &
       \indexOpArity(1_G) & ([], G) \\
-      \indexOpArity(+) & ([\SynTransl_2, \SynTransl_2], \SynTransl_2) &
+      \indexOpArity(+) & ([\SynTransl{2}, \SynTransl{2}], \SynTransl{2}) &
       \indexOpArity(\cdot_G) & ([G,G],G) \\
-      \indexOpArity(-) & ([\SynTransl_2], \SynTransl_2) &
+      \indexOpArity(-) & ([\SynTransl{2}], \SynTransl{2}) &
       \indexOpArity(^{-1_G}) & ([G], G) \\
-      \indexOpArity(\iota_o) & ([\SynOrth_2], \SynGL_2) &
-      \indexOpArity(\iota_1) & ([\SynGL_1], \SynGL_2) \\
-      \indexOpArity(\det) & ([\SynGL_2], \SynGL_1) &
-      \indexOpArity(|\cdot|) & ([\SynGL_1], \SynGL_1)
+      \indexOpArity(\iota_o) & ([\SynOrth{2}], \SynGL{2}) &
+      \indexOpArity(\iota_1) & ([\SynGL{1}], \SynGL{2}) \\
+      \indexOpArity(\det) & ([\SynGL{2}], \SynGL{1}) &
+      \indexOpArity(|\cdot|) & ([\SynGL{1}], \SynGL{1})
     \end{array}
@@
   \tyPrimNm{vec}, \tyPrimNm{real} \}$, with
-  $\primTyArity(\tyPrimNm{vec}) = [\SynGL_2, \SynTransl_2]$ and
-  $\primTyArity(\tyPrimNm{real}) = [\SynGL_1]$.
+  $\primTyArity(\tyPrimNm{vec}) = [\SynGL{2}, \SynTransl{2}]$ and
+  $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
 \end{example}
@@
     \begin{array}{l}
-      t : \SynTransl_2 \vdash t + 1 \stackrel{ax}\equiv t : \SynTransl_2 \\
-      t_1, t_2, t_3 : \SynTransl_2 \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl_2 \\
-      t : \SynTransl_2 \vdash t + (-t) \stackrel{ax}\equiv 1 : \SynTransl_2 \\
-      t_1, t_2 : \SynTransl_2 \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl_2 \\
+      t : \SynTransl{2} \vdash t + 1 \stackrel{ax}\equiv t : \SynTransl{2} \\
+      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
+      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 1 : \SynTransl{2} \\
+      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
     \end{array}
   \end{displaymath}
-  Similarly, the sort of scale factors $\SynGL_1$ forms an abelian
-  group under multiplication, and the sorts $\SynGL_2$ and
-  $\SynOrth_2$ form (non-abelian) multiplicative groups, so we assume
+  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
+  group under multiplication, and the sorts $\SynGL{2}$ and
+  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
   the appropriate axioms. We also assume that the operations $\iota_O,
@@
   expressions of the form $\iota_1(s)$ commute with group
-  multiplication in the sort $\SynGL_2$. The absolute value of the
+  multiplication in the sort $\SynGL{2}$. The absolute value of the
   determinant of orthgonal transformations is always $1$, so we also
@@
 $\extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in
-\relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s}
-= \rho \}$. The set of extensions of a relational environment will be
+\relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi_{i\mathord:s} =
+\rho \}$. The set of extensions of a relational environment will be
 used in the interpretation of the universal and existential type
@@
 Note that both parts of \lemref{lem:tyeqsubst-relational} depend on
-\lemref{lem:tyeqsubst-erasure} to be ``well-typed''.
+the corresponding parts of \lemref{lem:tyeqsubst-erasure} to be
+``well-typed''.
 
@@
     {\Delta; \Gamma \vdash M [e] : (\id_\Delta, e)^*A}
+
+    \inferrule* [right=ExPack]
+    {\Delta; \Gamma \vdash M : (\id_\Delta, e)^*A \\ \Delta, i\mathord:s \vdash A \isType}
+    {\Delta; \Gamma \vdash \mathrm{pack}\ [e]\ M : \exists i\mathord:s. A}
+
+    \inferrule* [right=ExUnpack]
+    {\Delta; \Gamma \vdash M : \exists i\mathord:s. A \\
+      \Delta, i\mathord:s; \pi_{i\mathord:s}^*\Gamma, x : A \vdash N : \pi_{i\mathord:s}^*B}
+    {\Delta; \Gamma \vdash \mathrm{unpack}\ [i]\ x = M\ \mathrm{in}\ N : B}
   \end{mathpar}
@@
     \tmSem{\Lambda i.\ M}\eta & \isDefinedAs & \tmSem{M}\eta \\
-    \tmSem{M[e]}\eta & \isDefinedAs & \tmSem{M}\eta    
+    \tmSem{M[e]}\eta & \isDefinedAs & \tmSem{M}\eta \\
+    \tmSem{\mathrm{pack}\ [e]\ M}\eta & \isDefinedAs & \tmSem{M}\eta \\
+    \tmSem{\mathrm{unpack}\ [i]\ x = M\ \mathrm{in}\ N}\eta & \isDefinedAs & \tmSem{N}(\eta, \tmSem{M}\eta)
   \end{array}

### 2012-06-14 paper/geometry-twocolumn.tex
Changes to Section 2
- Move the information flow and metric space examples to after the
  general framework (now in section-further-examples.tex)

- Write a new introduction to section 2 to highlight the structure of
  the geometry type system

- Small fixes throughout the geometry text

@@
-\section{Motivating Examples}
+\section{Geometry via Algebraically Indexed Types}
 \label{sec:motivating-examples}
@@
 We motivate this paper's goal of investigating algebraically-indexed
-types and their relational interpretations by presenting several
-examples. Our leading example is a novel type system for writing
-programs that manipulate geometric data, while maintaining geometric
-invariants through the types. The other examples are enforcement of
-information flow properties via indexing by a logical theory
-(\autoref{sec:information-flow}), and incorporation of metric
-properties into types (\autoref{sec:continuity-analysis}).
+types and their relational interpretations by presenting a novel type
+system for writing programs that manipulate two-dimensional geometric
+data. Geometry is rich with operations that are invariant under
+transformation: affine operations are invariant under change of
+origin, vector space operations are invariant under change of basis,
+and the Euclidean dot product is invariant under orthogonal changes of
+basis. Moreover, some operations are interestingly variant under
+transformations: for example, the cross product varies with scalings
+of the plane. We incorporate this extra information about our
+geometric primitives into the type system by means of algebraically
+indexed types.
 
-\subsection{2-Dimensional Geometry and Origin Invariance}
+\subsection{Origin Invariance and Representation Independence}
 
@@
 $(11,21)$ with respect to the second origin represent the \emph{same}
-point. It is merely an artifact of our representation of points as
+point because they have the same displacement from their respective
+origins. It is merely an artifact of our representation of points as
 pairs of numbers that these appear to be different. When we write
@@
 regarded as invariant under change of origin. Vectors represent
-offsets, or movements, and the vector $(0,0)$ always represents the
-zero offset.
+offsets and the vector $(0,0)$ always represents the zero offset.
 
 Invariance under change of representation immediately recalls
-Reynolds' fable concerning two professors teaching the theory of
-complex numbers \cite{reynolds83types}. One professor chooses
-rectangular coordinates ($x + iy$), while the other chooses polar
-coordinates ($\alpha\cos\theta + i\alpha\sin\theta$). Happily, after
-learning the basic operations on complex numbers in the two
-representations, the two classes can interact because the theory of
-complex numbers is invariant with respect to the choice of
-representation. Reynolds formalises the idea of invariance with
-respect to changes of representation as preservation of relations. If
-a binary relation $R$ relates two representations of a concept---for
-example the rectangular and polar representations of the same complex
-number are related---then a program that manipulates complex numbers
-at a level of abstraction above their specific representation should
-preserve the relation $R$. For example, let $f$ be a program taking
-complex numbers to complex numbers that is intended to respect
-invariance under change of representation. Then if $c$ is a complex
-number in rectangular form, and $c'$ is a complex number in polar form
-such that $R$ relates them, which we write as $(c,c') \in R$, then it
-should be the case that $(f(c), f(c')) \in R$.
+Reynolds' fable about two professors teaching the theory of complex
+numbers \cite{reynolds83types}. One professor chooses rectangular
+coordinates ($x + iy$), while the other chooses polar coordinates
+($\alpha\cos\theta + i\alpha\sin\theta$). Happily, after learning the
+basic operations on complex numbers in the two representations, the
+two classes can interact because the theory of complex numbers is
+invariant with respect to the choice of representation. Reynolds
+formalises the idea of invariance with respect to changes of
+representation as preservation of relations. If a binary relation $R$
+relates two representations of a concept---for example the rectangular
+and polar representations of the same complex number are
+related---then a program that manipulates complex numbers at a level
+of abstraction above their specific representation should preserve the
+relation $R$. For example, let $f$ be a program taking complex numbers
+to complex numbers that respects invariance under change of
+representation. Then if $c$ is a complex number in rectangular form,
+and $c'$ is a complex number in polar form such that $R$ relates them,
+which we write as $(c,c') \in R$, then it is the case that $(f(c),
+f(c')) \in R$.
 
@@
 represented by vectors in $\mathbb{R}^2$, and form a group under
-addition. To emphasise the group structure, and the intended meaning
-as translations, we write the collection of all change of origin
-offsets as $\Transl{2}$. By quantifying over all changes of origin we
-will be able to ensure that our programs are invariant with respect to
-the particular choice of origin. The members of the
-$\Transl{2}$-indexed family of binary relations $\{ R_{\vec{t}}
-\subseteq \mathbb{R}^2 \times \mathbb{R}^2 \}_{\vec{t} \in \Transl{2}}$
-are defined as follows:
+addition. To emphasise the intended meaning as translations, we write
+the collection of all change of origin offsets as $\Transl{2}$. By
+quantifying over all changes of origin we will be able to ensure that
+our programs are invariant with respect to the particular choice of
+origin. The members of the $\Transl{2}$-indexed family of binary
+relations $\{ R_{\vec{t}} \subseteq \mathbb{R}^2 \times \mathbb{R}^2
+\}_{\vec{t} \in \Transl{2}}$ are defined as follows:
 \begin{displaymath}
@@
 
-\subsection{A Type System for Origin Invariance}
+\subsection{A Type System for Change of Origin Invariance}
 \label{sec:type-system-geom-intro}
@@
 of the relational interpretation of tuple types.) Therefore, we
-introduce a new type parameterised by expressions $e$ of sort
+introduce a new type indexed by expressions $e$ of sort
 $\SynTransl{2}$ to represent points \emph{and} vectors:
-$\tyPrim{vec}{e}$. We have used the name $\tyPrimNm{vec}$, even though
-we have taken pains to separate geometric points and vectors, to
-recall the computer science notion of vector as a sequence of values
-of homogeneous type which has a known length (in this case two).
+$\tyPrim{vec}{e}$. The indexing argument $e$ represents the possible
+amount of displacement by change of origin this point may have been
+subjected to. We have used the name $\tyPrimNm{vec}$, even though we
+have taken pains to separate geometric points and vectors, to recall
+the computer science notion of vector as a sequence of values of
+homogeneous type which has a known length (in this case two).
 
@@
 \end{displaymath}
-We can also negate vectors, yielding a vector which points in the
-opposite direction. Negation negates translation arguments:
+Intuitively, this type says that if the first input vector has been
+displaced by $t_1$ and the second by $t_2$, then their sum is
+displaced by $t_1 + t_2$. We can also negate vectors, yielding a
+vector which points in the opposite direction. Negation negates
+translation arguments:
 \begin{displaymath}
@@
   also possible to show that certain types are uninhabited. For
-  example, the type $\forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t
-    + t} \to \tyPrim{vec}{t}$ does not have any inhabitants. We
-  demonstrate this in \autoref{sec:types-indexed-abelian-groups}. In
-  order to obtain such a result, we require a more sophisticated
-  relational interpretation of free indexing variables than the one we
-  sketched above.
+  example, the type $\forall t \mathord: \SynTransl{2}.\
+  \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ does not have any
+  inhabitants. Intuitively, this is because we have no way of removing
+  the extra $t$ in $\tyPrim{vec}{t+t}$ using our operations. We
+  demonstrate this indefinability result in
+  \autoref{sec:types-indexed-abelian-groups}. In order to obtain this
+  result, we require a more sophisticated relational interpretation of
+  free translation variables than the one we sketched above.
 \end{example}
@@
 linear map, and the collection of all invertible linear maps on
-$\mathbb{R}^2$ forms the group $\GL{2}$. We reflect in our language the
-group $\GL{2}$ as a new indexing sort $\SynGL{2}$, which has
-(non-abelian) group structure that we will write multiplicatively. We
-also extend the $\tyPrimNm{vec}$ type to be indexed by elements of
-$\SynGL{2}$ as well as $\SynTransl{2}$. Thus, $\tyPrim{vec}{B,t}$ is a
-vector which varies with change of basis $B$ and change of origin
-$t$. Formally, the index-erasure and (simplified) relational semantics
-of this type are as follows:
+$\mathbb{R}^2$ forms the \emph{General Linear} group $\GL{2}$. We
+reflect in our language the group $\GL{2}$ as a new indexing sort
+$\SynGL{2}$, which has (non-abelian) group structure that we will
+write multiplicatively. We also extend the $\tyPrimNm{vec}$ type to be
+indexed by elements of $\SynGL{2}$ as well as $\SynTransl{2}$. Thus,
+$\tyPrim{vec}{B,t}$ is a vector which varies with change of basis $B$
+and change of origin $t$. Formally, the index-erasure and (simplified)
+relational semantics of this type are as follows:
 \begin{displaymath}
@@
   compute the area of a triangle. 
-% By the type of this function, we can
-%   see that the area of a triangle is invariant under orthogonal
-%   transformations and arbitrary translations. Such transformations are
-%   known as isometries: distance preserving maps.
   \begin{displaymath}
@@
   If we specialise $\mathrm{area}$ to just orthogonal transformations,
-  we obtain the following type, as a consequence of the equation
-  $|\det (\iota_OO)| = 1$:
+  we obtain the following type, as a consequence of assuming the
+  equation $|\det (\iota_OO)| = 1$:
   \begin{displaymath}
@@
   under orthogonal transformations and translations. Such combined
-  transformations are known as isometries---distance preserving maps.
+  transformations are isometries---distance preserving maps.
 \end{example}
@@
 effect of our assigned type is that, semantically, we are indexing by
-non-zero scaling factors, where Kennedy indexes by strictly positive
+non-zero scaling factors, whereas Kennedy indexes by strictly positive
 scaling factors. We further examine the typing of the absolute value
@@
 operation satisfies the equation $\det (\iota_1 s) = s^2$, indicating
-that if we scale the plane by $s$ in all directions, we scale areas by
-$s^2$. This is illustrated by the next example.
+that if we scale the plane by $s$ in both directions, we scale areas
+by $s^2$. This is illustrated by the next example.
 
 \begin{example}\label{ex:area-of-triangle-2}
-  Just as we specialised the type of $\mathrm{area}$ function to
+  Just as we specialised the type of the $\mathrm{area}$ function to
   orthogonal transformations in \exref{ex:area-of-triangle-1}, we can
@@
 
-Purely scaling linear maps, as generated by $\iota_1$, commute with
-all other invertible linear maps, and form the centre of the group
-$\GL{2}$. Thus we add the equation $\iota_1(s)B = B\iota_1(s)$ to our
-calculus.
+Linear maps of the form $\left(
+  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$, as
+generated by $\iota_1$, commute with all other invertible linear
+maps. (These maps are all the elements of $\GL{2}$ that commute with
+everything, and form the \emph{centre} of $\GL{2}$.) Therefore we
+assume $(\iota_1 s)B = B(\iota_1 s)$ to hold.
 
@@
   operation---with the type above---then it is still not possible to
-  construct the cube root function.
+  construct the cube root function. The indefinability of cube root is
+  of interest due to its relevance to the classical problem of
+  trisecting an angle by ruler and compass constructions.
 \end{example}
-
-\subsection{Logical Information Flow}
-\label{sec:information-flow}
-
-An extreme example of invariance under change of representation is the
-tracking of information flow through programs. If a program does not
-depend upon a particular piece of information, then it will be
-invariant under \emph{all} changes of representation of this
-information. As thoroughly described by Sabelfeld and Sands
-\cite{sabelfeld01per}, information flow can be captured semantically
-through the use of partial equivalence relations (PERs). PERs form an
-instance of Reynolds' approach to abstraction and representation
-independence. Abadi, Banerjee, Heintze and Riecke \cite{abadi99core}
-built a \emph{Core Calculus for Dependency}, using a type system based
-around a security level indexed monad $T_lA$. Tse and Zdancewic
-\cite{tse04translating} showed that it is possible to translate Abadi
-et al.'s calculus into System F, translating the monadic type $T_lA$
-to $\alpha_l \to A$ for some free type variable $\alpha_l$, and making
-direct use of Reynolds' abstraction theorem to prove information flow
-properties. We now show how a refined variant of Tse and Zdancewic's
-translation can be expressed via algebraically-indexed types.
-
-We assume a single indexing sort $\mathsf{principal}$, intended to
-represent (possibly composite) principals in a system. Principals are
-combined using the connectives of boolean logic ($\top$, $\bot$,
-$\lor$, $\land$, $\lnot$). For the equational theory of principals, we
-assume all the axioms of boolean algebra, and semantically we
-interpret composite principals as boolean values: either true ($\top$)
-or false ($\bot$). We introduce a single primitive type, indexed by
-expressions of sort $\mathsf{principal}$, $\tyPrim{T}{e}$, with the
-following interpretations:
-\begin{displaymath}
-  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\tyPrim{T}{e}} & \{*\} \\
-    \rsem{\tyPrim{T}{e}}{}\rho & \left\{
-      \begin{array}{ll}
-        \{(*,*)\} & \textrm{if }\sem{e}\rho = \top \\
-        \{\}      & \textrm{if }\sem{e}\rho = \bot
-      \end{array}
-      \right.
-  \end{array}
-\end{displaymath}
-The relational interpretation of $\tyPrim{T}{e}$ forces this type to
-only be inhabited when the boolean expression $e$ evaluates to true
-under the environment $\rho$. To write programs, we assume the
-following pair of primitive operations:
-\begin{eqnarray*}
-  \mathrm{proj} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1 \land p_2} \to \tyPrim{T}{p_1} \\
-  \mathrm{combine} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1} \to \tyPrim{T}{p_2} \to \tyPrim{T}{p_1 \land p_2}
-\end{eqnarray*}
-We can now adapt Tse and Zdancewic's translation of Abadi et al.'s
-monadic type to our setting. We define a type synonym, for each type
-$A$ and expression $e$ of sort $\mathsf{principal}$:
-\begin{displaymath}
-  T_eA = \tyPrim{T}{e} \to A
-\end{displaymath}
-For every principal $p$, we can endow the types $T_p-$ with the
-structure of a monad. This is due to the fact that it is an instance
-of the ``environment'' (or ``reader'') monad \fixme{ref}.
-
-\begin{example}
-  Suppose we have a program with the following type:
-  \begin{displaymath}
-    \forall p,q\mathord:\mathsf{principal}. \tyPrim{T}{e} \to T_p(\tyPrimNm{unit} + \tyPrimNm{unit}) \to T_q(\tyPrimNm{unit} + \tyPrimNm{unit})
-  \end{displaymath}
-  where $e$ is some boolean algebra expression involving $p$ and
-  $q$. We will show in \autoref{sec:instantiations} that programs with
-  this type are non-constant (i.e.~depend on their input) if and only
-  if the formula $e \Rightarrow q \Rightarrow p$ is valid in boolean
-  logic.
-\end{example}
-
-% Go a step further by considering equivalence relations on word32, and
-% their lattice structure. Link this to quantative information
-% flow. Will also need to include singleton types.
-
-\subsection{Metric Spaces and Simple Continuity Analysis}
-\label{sec:continuity-analysis}
-
-Our final motivating example of algebraically indexed types and
-representation independence makes use of the metric space structure of
-the real numbers. In the geometry example, we used relational
-interpretations that related values via geometrical
-transformations. An alternative relational interpretation of the real
-numbers is to relate values that are close, in the metric space
-sense. By indexing types by a measure of closeness, we can actually
-state continuity of functions as a typing problem.
-
-We introduce an indexing sort $\mathsf{R}^{>0}$, representing positive
-distances between real numbers. Our type of real numbers is indexed by
-expressions of sort $\mathsf{R}^{>0}$, and we use the following
-index-erasure and relational interpretations:
-\begin{displaymath}
-  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\tyPrim{real}{e}} & \mathbb{R} \\
-    \rsem{\tyPrim{real}{e}}{}\rho & \{ (x,x') \sepbar |x-x'| < \sem{e}\rho \}
-  \end{array}
-\end{displaymath}
-Thus, two real numbers are related if they only differ by the positive
-real number assigned to $e$. With this interpretation, and a
-straightforward interpretation of existential types, we can state the
-standard $\epsilon$-$\delta$ definition of continuity as a type:
-\begin{displaymath}
-  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
-\end{displaymath}
-Chaudhuri, Gulwani and Lublinerman \cite{chaudhuri10continuity} have
-presented a program logic based approach to verifying the continuity
-of programs. Algebraically indexed types give us a way of expressing
-and verifying continuity properties of functions in a type based way.
-
-% \begin{eqnarray*}
-%   \underline{c} & : & \forall \epsilon\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon} \\
-%   (+) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
-%   (-) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
-%   (*) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1\epsilon_2}
-% \end{eqnarray*}
-
-% \begin{enumerate}
-% \item Need to give types to basic operations on real numbers
-% \item Need to add existential types to the calculus
-% \end{enumerate}
 

### 2012-06-14 paper/section-further-examples.tex
Changes to Section 2
- Move the information flow and metric space examples to after the
  general framework (now in section-further-examples.tex)

- Write a new introduction to section 2 to highlight the structure of
  the geometry type system

- Small fixes throughout the geometry text

\section{Further Examples}
\label{sec:further-examples}

\subsection{Logical Information Flow}
\label{sec:information-flow}

An extreme example of invariance under change of representation is the
tracking of information flow through programs. If a program does not
depend upon a particular piece of information, then it will be
invariant under \emph{all} changes of representation of this
information. As thoroughly described by Sabelfeld and Sands
\cite{sabelfeld01per}, information flow can be captured semantically
through the use of partial equivalence relations (PERs). PERs form an
instance of Reynolds' approach to abstraction and representation
independence. Abadi, Banerjee, Heintze and Riecke \cite{abadi99core}
built a \emph{Core Calculus for Dependency}, using a type system based
around a security level indexed monad $T_lA$. Tse and Zdancewic
\cite{tse04translating} showed that it is possible to translate Abadi
et al.'s calculus into System F, translating the monadic type $T_lA$
to $\alpha_l \to A$ for some free type variable $\alpha_l$, and making
direct use of Reynolds' abstraction theorem to prove information flow
properties. We now show how a refined variant of Tse and Zdancewic's
translation can be expressed via algebraically-indexed types.

We assume a single indexing sort $\mathsf{principal}$, intended to
represent (possibly composite) principals in a system. Principals are
combined using the connectives of boolean logic ($\top$, $\bot$,
$\lor$, $\land$, $\lnot$). For the equational theory of principals, we
assume all the axioms of boolean algebra, and semantically we
interpret composite principals as boolean values: either true ($\top$)
or false ($\bot$). We introduce a single primitive type, indexed by
expressions of sort $\mathsf{principal}$, $\tyPrim{T}{e}$, with the
following interpretations:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{T}{e}} & \{*\} \\
    \rsem{\tyPrim{T}{e}}{}\rho & \left\{
      \begin{array}{ll}
        \{(*,*)\} & \textrm{if }\sem{e}\rho = \top \\
        \{\}      & \textrm{if }\sem{e}\rho = \bot
      \end{array}
      \right.
  \end{array}
\end{displaymath}
The relational interpretation of $\tyPrim{T}{e}$ forces this type to
only be inhabited when the boolean expression $e$ evaluates to true
under the environment $\rho$. To write programs, we assume the
following pair of primitive operations:
\begin{eqnarray*}
  \mathrm{proj} & : & \forall p, q\mathord:\mathsf{principal}.\ \tyPrim{T}{p \land q} \to \tyPrim{T}{p} \\
  \mathrm{combine} & : & \forall p, q\mathord:\mathsf{principal}.\ \tyPrim{T}{p} \to \tyPrim{T}{q} \to \tyPrim{T}{p \land q}
\end{eqnarray*}
We can now adapt Tse and Zdancewic's translation of Abadi et al.'s
monadic type to our setting. We define a type synonym, for each type
$A$ and expression $e$ of sort $\mathsf{principal}$:
\begin{displaymath}
  T_eA = \tyPrim{T}{e} \to A
\end{displaymath}
For every principal $e$, we can endow the types $T_e-$ with the
structure of a monad. This is due to the fact that it is an instance
of the ``environment'' (or ``reader'') monad \cite{jones95functional}.

\begin{example}
  Suppose we have a program with the following type:
  \begin{displaymath}
    \forall p,q\mathord:\mathsf{principal}. \tyPrim{T}{e} \to T_p(\tyPrimNm{unit} + \tyPrimNm{unit}) \to T_q(\tyPrimNm{unit} + \tyPrimNm{unit})
  \end{displaymath}
  where $e$ is some boolean algebra expression involving $p$ and
  $q$. We will show in \autoref{sec:instantiations} that programs with
  this type are non-constant (i.e.~depend on their input) if and only
  if the formula $e \Rightarrow q \Rightarrow p$ is valid in boolean
  logic.
\end{example}

% Go a step further by considering equivalence relations on word32, and
% their lattice structure. Link this to quantative information
% flow. Will also need to include singleton types.

\subsection{Metric Spaces and Simple Continuity Analysis}
\label{sec:continuity-analysis}

Our final motivating example of algebraically indexed types and
representation independence makes use of the metric space structure of
the real numbers. In the geometry example, we used relational
interpretations that related values via geometrical
transformations. An alternative relational interpretation of the real
numbers is to relate values that are close, in the metric space
sense. By indexing types by a measure of closeness, we can actually
state continuity of functions as a typing problem.

We introduce an indexing sort $\mathsf{R}^{>0}$, representing positive
distances between real numbers. Our type of real numbers is indexed by
expressions of sort $\mathsf{R}^{>0}$, and we use the following
index-erasure and relational interpretations:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{real}{e}} & \mathbb{R} \\
    \rsem{\tyPrim{real}{e}}{}\rho & \{ (x,x') \sepbar |x-x'| < \sem{e}\rho \}
  \end{array}
\end{displaymath}
Thus, two real numbers are related if they only differ by the positive
real number assigned to $e$. With this interpretation, and a
straightforward interpretation of existential types, we can state the
standard $\epsilon$-$\delta$ definition of continuity as a type:
\begin{displaymath}
  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
\end{displaymath}
Chaudhuri, Gulwani and Lublinerman \cite{chaudhuri10continuity} have
presented a program logic based approach to verifying the continuity
of programs. Algebraically indexed types give us a way of expressing
and verifying continuity properties of functions in a type based way.

% \begin{eqnarray*}
%   \underline{c} & : & \forall \epsilon\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon} \\
%   (+) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
%   (-) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
%   (*) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1\epsilon_2}
% \end{eqnarray*}

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:

### 2012-06-15 paper/section-introduction.tex
Paragraphs in intro

@@
 \label{sec:introduction}
+Types express properties of programs. The statement $\mathit{b}: \tyBool$ tells
+us that $\mathit{b}$ has value $\mathtt{true}$ or $\mathtt{false}$;
+the statement $\mathit{pos}:\tyReal\tyProduct\tyReal$ says that
+$\mathit{pos}$ is a pair whose components are real numbers. More interestingly,
+the statement $\mathit{f} : \forall X. \tyPrim{list}X\tyArr\tyPrim{list}X$ tells us that
+if we apply $\mathit{f}$ to a list whose elements satisfy some property $\psi$, 
+then the result is a list whose elements also satisfy $\psi$.
+
+Unary predicates only go so far, and for polymorphic types
+particularly do not express everything that follows from possessing that type. 
+Imagine a function $\mathit{oddRev}$ that reverses any list whose first element
+is an odd number, otherwise behaving as the identity. It certainly
+preserves properties of the elements -- but truly parametric
+polymorphism should tell us more.
+
+To obtain stronger properties we must \emph{relate} multiple
+executions of the same code. The statement $\mathit{f}:\forall
+X. \tyPrim{list}X\tyArr\tyPrim{list}X$ then says that any change in
+representation of the list elements supplied to $\mathit{f}$ is
+reflected in the elements in the result.  Formally, for any function
+$g$, we have $\mathit{f}(\mathrm{map}\:\mathit{g}\:\mathit{x}) =
+\mathrm{map}\:\mathit{g}\:(\mathit{f}\:\mathit{x})$.  This is the
+essence of \emph{relational parametricity} as formulated by
+Reynolds~\cite{reynolds83types}. Its consequences are striking.  It
+immediately tells us that $\mathit{oddRev}$ cannot be given a
+polymorphic type: simply consider the change-of-representation given
+by $\lambda x. x+1$.  But more profoundly it can be used to prove blah
+blah (initial algebras, term reps, non-inhabitation, etc.).
+
+The relational interpretation of polymorphism gives us a principle of
+\emph{invariance}, namely that generic code is invariant under changes of
+representation. It is central to information hiding, abstract data
+types, encapsulation and modularity -- at its broadest, software
+engineering!  The idea of invariance is also ubiquitous in mathematics
+and physics.  The area of a triangle is invariant with respect to
+isometries of the Euclidean plane; the determinant of a matrix is
+invariant under changes of basis; and Newton's laws are the same in all
+inertial frames.  Typically, transformations in mathematics and
+physics have interesting structure; for example, translations in the Euclidean plane
+form an Abelian group.
+
+Inspired by this connection, we study relational parametricity for 
+types that are \emph{indexed} by attributes that themselves have algebraic structure.
+For example, in geometry, points in 2-space can be indexed by a co-ordinate frame;
+in physical modelling, quantities are indexed by units-of-measure;
+in information-flow security, computations are indexed by principals. 
+Types that are polymorphic over such indices induce invariance properties. 
+A function that computes the area of a triangle might have simple type
+\[
+\mathit{area} : \tyPrimNm{vec}\tyArr\tyPrimNm{vec}\tyArr\tyPrimNm{vec}\tyArr\tyReal
+\]
+but by assigning a polymorphic type
+\[
+\mathrm{area} : \forall i\mathord:\SynSE{2}.
+  \tyPrim{vec}{i} \to \tyPrim{vec}{i} \to \tyPrim{vec}{i} \to \tyReal \\
+\]
+we can deduce that it is invariant under direct isometries (rigid body motions) as 
+represented by the index $i$ on the type of vectors.
+\todo{polymorphism imposes abstraction barrier; talk about co-ordinate free geometry}
 

### 2012-06-17 paper/geometry-twocolumn.tex

@@
 Note how the type of the result of $\mathrm{offset}$ simplifies from
-$\tyPrim{vec}{t-t}$ to $\tyPrim{vec}{0}$. The algebraic structure on
+$\tyPrim{vec}{t-t}$ to $\tyPrim{vec}{0}$: the algebraic structure on
 the indexing theory induces type equalities.  The type of the vector
@@
 $\Orth{2}$ of $\GL{2}$ of \emph{orthogonal} linear transformations:
-invertible linear maps whose determinant is either $1$ or $-1$. We
-introduce orthogonal transformations as a new sort $\SynOrth{2}$, and
-overload the multiplicative group operations syntax for inhabitants of
-$\SynOrth{2}$. We also assume a function $\iota_O$ that takes $e :
-\SynOrth{2}$ to $\iota_O(e) : \SynGL{2}$. We may now assign a type to
-the dot product, using quantification over $\SynOrth{2}$:
+invertible linear maps whose matrix representation's transpose is
+equal to their inverse. We introduce orthogonal transformations as a
+new sort $\SynOrth{2}$, and overload the multiplicative group
+operations syntax for inhabitants of $\SynOrth{2}$. We also assume a
+function $\iota_O$ that takes $e : \SynOrth{2}$ to $\iota_O(e) :
+\SynGL{2}$. We may now assign a type to the dot product, using
+quantification over $\SynOrth{2}$:
 \begin{displaymath}

### 2012-06-17 paper/section-further-examples.tex

@@
 
+\subsection{Monoid Indexed Types}
+\label{sec:monoid-indexed-types}
+
+Primitive type is a monoid, and we index by a commutative monoid. An
+example of this is strings with tainting information, or bags, or sets
+of stuff. This gives a slightly different information flow analysis to
+the logical information flow above. In particular, we can now count
+uses of input data.
+
+Primitive operations $\Gamma_{\mathit{Mon}}$:
+\begin{displaymath}
+  \begin{array}{c@{\hspace{0.5em}:\hspace{0.5em}}l}
+    1 & \tyPrim{M}{1} \\
+    (\cdot) & \forall a,b.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{a\cdot b}
+  \end{array}
+\end{displaymath}
+
+For the index-erasure interpretation, we assume a commutative monoid
+$\mathcal{M}$. The primitive type $\tyPrimNm{M}$ is interpreted as
+$\tySem{\tyPrim{M}{e}} = \mathcal{M}$. The primitive operations in
+$\Gamma_{\mathit{Mon}}$ are interpreted just as the unit and multiplication of
+the monoid $\mathcal{M}$.
+
+% Need the following definitions: syntatic group/monoid arising from
+% the syntax, and a neater way of defining relational
+% environments. Also, a neat way of formally introducing a load of
+% primitive operations.
+
+For each index context $\relEnv{E}(\Delta)$, the set
+$\relEnv{E}(\Delta)$ of relational environments is defined to be $\{
+\rho_{(M,h)} \sepbar M\textrm{ is a submonoid of }(\Delta \Rightarrow
+\mathsf{M})/\equiv, h\textrm{ is a monoid homomorphism }M \to
+Z(\mathcal{M}) \}$, where
+\begin{displaymath}
+  \rho_{(M,h)}\ \tyPrimNm{M}\ (e) = \left\{
+    \begin{array}{ll}
+      \{ \} & \textrm{if}\ e\not\in M \\
+      \{ (x, h(e)\cdot x) \sepbar x \in \mathcal{M} \} & \textrm{if}\ e\in M
+    \end{array}
+  \right.
+\end{displaymath}
+Note the mapping to the centre of the monoid $\mathcal{M}$.
+
+\begin{lemma}
+  For all $\Delta$ and $\rho \in \relEnv{E}(\Delta)$,
+  $(\eta_{M},\eta_{M}) \in \rsem{\Gamma_{\mathit{Mon}}}{\relEnv{E}}\rho$.
+\end{lemma}
+
+\begin{theorem}
+  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
+  natural numbers $A$ and $m$-by-$1$ matricies of natural numbers $b$,
+  there exists a program $M$ such that
+  \begin{equation}
+    \label{eq:fo-monoid-type}
+    \Gamma_{\mathit{Mon}} \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
+  \end{equation}
+  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e =
+  i_1^{b_1}\dots i_m^{b_m}$, if and only if there is a solution
+  $\vec{x} = (x_1,\cdots,x_n)$ in the natural numbers to the equation
+  $A x^\top = b$.
+\end{theorem}
+
+\begin{proof}
+  (If) The program $\Lambda i_1\dots i_m.\ \lambda z_1\dots z_n.\
+  z_1^{x_1}\dots z_n^{x_n}$ satisfies the typing judgment
+  (\ref{eq:fo-monoid-type}).
+
+  (Only if) Assume that there is a program $M$ satisfying the typing
+  judgment (\ref{eq:fo-monoid-type}). By \thmref{thm:abstraction} we
+  know that for all relational environments $\rho \in
+  \relEnv{E}(i_1,\dots,i_m)$, and all $z_j,z'_j \in \mathcal{M}$,
+  \begin{displaymath}
+    \begin{array}{l}
+      (z_1,z'_1) \in \rho\ \tyPrimNm{M}\ (e_1) \land \dots \land (z_n,z'_n) \in \rho\ \tyPrimNm{M}\ (e_n) \Rightarrow \\
+      \quad (\tmSem{M}z_1\dots z_n, \tmSem{M}z'_1\dots z'_n) \in \rho\ \tyPrimNm{M}\ (e)
+    \end{array}
+  \end{displaymath}
+  We select the relational environments
+  \begin{displaymath}
+    \rho\ \tyPrimNm{M}\ (e) = \left\{
+      \begin{array}{ll}
+        \{(z,z) \sepbar z \in \mathcal{M} \} &
+        \begin{array}[t]{@{}l}
+          \textrm{if}\ e = e_1^{x_1}\dots e_n^{x_n} \\
+          \textrm{ for some }x_1,\dots,x_n \in \mathbb{N}
+        \end{array}
+        \\
+        \{\} & \textrm{otherwise}
+      \end{array}
+    \right.
+  \end{displaymath}
+  Now, for any $z_1,\dots,z_m \in \mathcal{M}$, we know that, for all
+  $j$, $(z_j,z_j) \in \rho\ \tyPrimNm{M}\ (e_j)$, so we know that
+  $(\tySem{M}z_1\dots z_n, \tySem{M}z_1 \dots z_n) \in \rho\
+  \tyPrimNm{M}\ (e)$. Thus there exist $x_1,...,x_n$ such that $e =
+  e_1^{x_1}\dots e_n^{x_n}$. But we also know that $e =
+  i_1^{b_1}...i_m^{b_m}$. By the cancellation property of free
+  monoids, we learn that $x_1,...,x_n$ is the solution we need.
+\end{proof}
+
+We can further assume that there are other operations that distribute
+over the multiplication of the monoid, for example semi-ring
+operations (e.g.~relational algebra).
+
+Examples
+\begin{enumerate}
+\item Finite sets of stuff.
+\item Let $\mathcal{M} = \Sigma^* \times
+  \mathcal{P}_{\mathit{fin}}(C)$, where $\Sigma$ is some alphabet, and
+  $C$ is some set of taint markers. The centre of this monoid is
+  $\{([], c) \sepbar c \in \mathcal{P}_{\mathit{fin}}(C) \}$.
+\end{enumerate}
+
 %%% Local Variables:

### 2012-06-21 paper/section-general-framework.tex
Tighten up the general framework section

@@
   geometric groups that we mentioned in
-  \autoref{sec:motivating-examples}. Thus, $\SortSet = \{\SynTransl{2},
-  \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$. For the index operations, we
-  have additive group structure on $\SynTransl{2}$, multiplicative
-  group structure on $\SynGL{1}$, $\SynGL{2}$ and $\SynOrth{2}$,
-  injections from $\SynOrth{2}$ and $\SynGL{1}$ into $\SynGL{2}$,
-  determinant and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
-  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, |\cdot| \}$, where
-  $G \in \{\SynGL{1}, \SynGL{2}, \SynOrth{2}\}$, with the following
-  assignment of arities:
+  \autoref{sec:motivating-examples}. Thus, $\SortSet =
+  \{\SynTransl{2}, \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$. For the
+  index operations, we have additive group structure on
+  $\SynTransl{2}$, multiplicative group structure on $\SynGL{1}$,
+  $\SynGL{2}$ and $\SynOrth{2}$, injections from $\SynOrth{2}$ and
+  $\SynGL{1}$ into $\SynGL{2}$, determinant, and absolute value. Thus,
+  $\IndexOpSet = \{ 0, +, -, 1_G, -\cdot_G-, -^{-1_G}, \iota_O,
+  \iota_1, \det, |\cdot| \}$, where $G \in \{\SynGL{1}, \SynGL{2},
+  \SynOrth{2}\}$, with the following assignment of arities:
   \begin{displaymath}
@@
 We assume a countably infinite collection of index variable names $i,
-i_1, i_2,$ and so on. \emph{Index contexts} $\Delta = i_1 \mathord:
-s_1, ..., i_n \mathord: s_n$ are lists of index expression
-variable/sort pairs, such that all the index variable names are
-distinct.
+i_1, i_2,\ldots$. \emph{Index contexts} $\Delta = i_1 \mathord: s_1,
+..., i_n \mathord: s_n$ are lists of variable/sort pairs, such that
+all the variable names are distinct.
 
@@
   \centering
-  \textbf{Index contexts}\\
+  % \textbf{Index contexts}\\
 
-  $\Delta = i_1 \mathord: s_1, ..., i_n \mathord: s_n$, where each $s_i \in \SortSet$ and no variable name is repeated.
+  % $\Delta = i_1 \mathord: s_1, ..., i_n \mathord: s_n$, where each $s_i \in \SortSet$ and no variable name is repeated.
 
-  \bigskip
+  % \bigskip
 
@@
 sort. Existential types are constructed using the \TirName{TyExist}
-rule.
+rule, and allow for abstraction by hiding. %FIXME: think about the introduction of existentials more, perhaps forward ref to their use
 
@@
 variable $i_j \mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for
-the index expression $e_j$.
+the index expression $e_j$. We write $\Delta \Rightarrow \Delta'$ for
+the set of all simultaneous substitutions $\sigma$ such that $\Delta
+\vdash \sigma \Rightarrow \Delta'$.
 
@@
 \primTyArity(\tyPrimNm{X})$, since these are exactly sequences of
-index arguments suitable for the primitive type $\tyPrimNm{X}$.
-
-We write $\Delta \Rightarrow \Delta'$ for the set of all simultaneous
+index arguments suitable for the primitive type $\tyPrimNm{X}$. With
+further abuse of notation, we write $\Delta \Rightarrow
+\primTyArity(\tyPrimNm{X})$ for the set of all simultaneous
 substitutions $\sigma$ such that $\Delta \vdash \sigma \Rightarrow
-\Delta'$. The \emph{identity} simultaneous substitution $\Delta \vdash
-\id_\Delta \Rightarrow \Delta$ is just the sequence of variables in
-$\Delta$: $\id_\Delta = (i_1,...,i_n)$ where $\Delta =
-i_1\mathord:s_1,...,i_n\mathord:s_n$. We will also define composition
-of simultaneous substitutions below, but first we must define the
-application of a simultaneous substitutions to index expressions.
-
-Application of a simultaneous substitution $\Delta \vdash \sigma
-\Rightarrow \Delta'$ to a well-sorted index expression $\Delta' \vdash
-e : s$ yields a well-sorted index expression $\Delta \vdash \sigma^*e
-: s$. The expression $\sigma^*e$ is defined by the following two
-clauses:
-\begin{mathpar}
-  \sigma^*i \isDefinedAs \sigma(i)
-
-  \sigma^*(\indexOp{f}(e_1,...,e_n)) \isDefinedAs \indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)
-\end{mathpar}
-Similarly, given a well-indexed type $\Delta' \vdash A \isType$, we
-can apply $\sigma$ to $A$ to produce a new type $\sigma^*A$. The
-application of a simultaneous substitution to a type is defined
-by the following clauses:
-\begin{mathpar}
-  \sigma^*(\tyPrim{X}{e_1,...,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,...,\sigma^*e_n}
-
-  \sigma^*\tyUnit \isDefinedAs \tyUnit
-
-  \sigma^*(A \tyArr B) \isDefinedAs \sigma^*A \tyArr \sigma^*B
-
-  \sigma^*(A \tyProduct B) \isDefinedAs \sigma^*A \tyProduct \sigma^*B
-
-  \sigma^*(A + B) \isDefinedAs \sigma^*A + \sigma^*B
-
-  \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A
-
-  \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:S. \sigma_{i\mathord:s}^*A
-\end{mathpar}
-\begin{lemma}
-  Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
-  substitution.
-  \begin{enumerate}
-  \item If $\Delta' \vdash e : s$, then $\Delta \vdash \sigma^*e : s$; and
-  \item If $\Delta' \vdash A \isType$, then $\Delta \vdash \sigma^*A
-    \isType$.
-  \end{enumerate}
-\end{lemma}
-
-The \emph{composition} of two simultaneous substitutions $\Delta
-\vdash \sigma \Rightarrow \Delta'$ and $\Delta' \vdash \sigma'
-\Rightarrow \Delta''$, where $\sigma' = (e'_1,...,e'_n)$, is defined
-as $\Delta \vdash \sigma' \circ \sigma \isDefinedAs (\sigma^*e'_1,
-..., \sigma^*e'_n) \Rightarrow \Delta''$.
-
-Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
-variable/type pair $i\mathord:s$ such that $i$ does not appear in
-$\Delta$, we define the \emph{projection} simultaneous substitution
-$\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
-$\pi_{i\mathord:s} = (i_1,...,i_n)$. The subscript on
-$\pi_{i\mathord:s}$ is the variable/type pair that is being discarded.
+\primTyArity(\tyPrimNm{X})$.
 
 For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
-\Delta'$ where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$ and a
-variable/type pair $i\mathord:s$ such that $i$ does not appear in
+\Delta'$, where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
+variable/sort pair $i\mathord:s$ such that $i$ does not appear in
 either $\Delta$ or $\Delta'$, then we can form the \emph{lifted}
@@
 
+Application of a simultaneous substitution $\Delta \vdash \sigma
+\Rightarrow \Delta'$ to a well-sorted index expression $\Delta' \vdash
+e : s$ yields a well-sorted index expression $\Delta \vdash \sigma^*e
+: s$. The expression $\sigma^*e$ is defined on variables as $\sigma^*i
+\isDefinedAs \sigma(i)$, and on operation symbols as
+$\sigma^*(\indexOp{f}(e_1,...,e_n)) \isDefinedAs
+\indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)$.  Similarly, given a
+well-indexed type $\Delta' \vdash A \isType$, we can apply $\sigma$ to
+$A$ to produce a new well-indexed type $\Delta \vdash \sigma^*A
+\isType$. The key clauses defining $\sigma^*A$ are for primitive types
+and the universal and existential quantifiers:
+\begin{displaymath}
+  \begin{array}{c}
+    \sigma^*(\tyPrim{X}{e_1,...,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,...,\sigma^*e_n}
+    \\
+    \begin{array}{c@{\hspace{2em}}c}
+      \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A
+      &
+      \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:S. \sigma_{i\mathord:s}^*A
+    \end{array}
+  \end{array}
+\end{displaymath}
+% \begin{lemma}
+%   Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
+%   substitution.
+%   \begin{enumerate}
+%   \item If $\Delta' \vdash e : s$, then $\Delta \vdash \sigma^*e : s$; and
+%   \item If $\Delta' \vdash A \isType$, then $\Delta \vdash \sigma^*A
+%     \isType$.
+%   \end{enumerate}
+% \end{lemma}
+
+The \emph{identity} simultaneous substitution $\Delta \vdash
+\id_\Delta \Rightarrow \Delta$ is just the sequence of variables in
+$\Delta$: $\id_\Delta = (i_1,...,i_n)$ where $\Delta =
+i_1\mathord:s_1,...,i_n\mathord:s_n$. The \emph{composition} of two
+simultaneous substitutions $\Delta \vdash \sigma \Rightarrow \Delta'$
+and $\Delta' \vdash \sigma' \Rightarrow \Delta''$, where $\sigma' =
+(e'_1,...,e'_n)$, is defined as $\Delta \vdash \sigma' \circ \sigma
+\isDefinedAs (\sigma^*e'_1, ..., \sigma^*e'_n) \Rightarrow \Delta''$.
+
+Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
+variable/sort pair $i\mathord:s$ such that $i$ does not appear in
+$\Delta$, we define the \emph{projection} simultaneous substitution
+$\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
+$\pi_{i\mathord:s} = (i_1,...,i_n)$. The subscript on
+$\pi_{i\mathord:s}$ is the variable/sort pair that is being discarded.
+
 \subsubsection{Index Expression Equality and Type Equality}
@@
 theory comes from the equations of the theory. In the two-dimensional
-geometry example, the types $\tyPrim{vec}{B,t_1 + t_2}$ and
-$\tyPrim{vec}{B, t_2 + t_1}$ are considered equal by the type system,
-due to the commutativity of the $+$ operation.
+geometry example of \autoref{sec:motivating-examples}, the types
+$\tyPrim{vec}{B,t_1 + t_2}$ and $\tyPrim{vec}{B, t_2 + t_1}$ are
+considered equal by the type system, due to the commutativity of the
+$+$ operation.
 
@@
 e' : s$. We assume that all axioms $(\Delta \vdash e
-\stackrel{ax}\equiv e' : s) \in \IndexAxiomSet$ are well-sorted:
-meaning that both $\Delta \vdash e : s$ and $\Delta \vdash e' : s$.
+\stackrel{ax}\equiv e' : s) \in \IndexAxiomSet$ are well-sorted,
+meaning that both $\Delta \vdash e : s$ and $\Delta \vdash e' : s$
+hold.
 
-Given a set $\IndexAxiomSet$ of axioms, we generate the judgmental
-equality between index expressions $\Delta \vdash e \equiv e' : s$ by
-the following rules, which allow us to use substitution instances of
-axioms, and ensure that judgmental equality is a congruence relation
+Given a set $\IndexAxiomSet$ of axioms, we generate the equality
+judgment between index expressions $\Delta \vdash e \equiv e' : s$ by
+the following rules. The first allows us to use substitution instances
+of axioms. The second ensures that equality is a congruence relation
 with respect to the index operations in $\IndexOpSet$:
@@
 We also assume the standard reflexivity, symmetry and transitivity
-rules for the judgmental equality.
+rules for the equality judgment.
 
 \begin{example}[Two-Dimensional Geometry]
+  \fixme{Tie this example back to section 2}
   Translations form an abelian group under addition, so we assume the
@@
   multiplication in the sort $\SynGL{2}$. The absolute value of the
-  determinant of orthgonal transformations is always $1$, so we also
+  determinant of an orthgonal transformation is always $1$, so we also
   assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
@@
 
-The judgmental equality $\Delta \vdash e \equiv e' : s$ on index
-expressions generates the judgmental equality $\Delta \vdash A \equiv
-B \isType$ on types. The basic rule generating judgmental equality on
+The equality judgment $\Delta \vdash e \equiv e' : s$ on index
+expressions generates the equality judgment $\Delta \vdash A \equiv
+B \isType$ on types. The basic rule generating equality judgments on
 types states that two applications of primitive types are equal if
@@
 \end{displaymath}
-The rest of the rules for judgmental equality on types ensure that it
-is a congruence relation on types, and that it is an equivalence
-relation (i.e.~judgmental equality on types is reflexive, symmetric
-and transitive).
+The rest of the rules for equality on types ensure that it is a
+congruence relation on types, and that it is an equivalence relation
+(i.e.,~equality on types is reflexive, symmetric and transitive).
 % For example, for universally quantified types we have
@@
 
-\begin{lemma}
-  Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
-  substitution.
-  \begin{enumerate}
-  \item If $\Delta' \vdash e \equiv e' : s$ then $\Delta \vdash
-    \sigma^*e \equiv \sigma^*e' : s$; and
-  \item If $\Delta' \vdash A \equiv B \isType$ then $\Delta \vdash
-    \sigma^*A \equiv \sigma^* B \isType$.
-  \end{enumerate}
-\end{lemma}
+% FIXME: think about integrating this lemma into the running text
+% \begin{lemma}
+%   Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
+%   substitution.
+%   \begin{enumerate}
+%   \item If $\Delta' \vdash e \equiv e' : s$ then $\Delta \vdash
+%     \sigma^*e \equiv \sigma^*e' : s$; and
+%   \item If $\Delta' \vdash A \equiv B \isType$ then $\Delta \vdash
+%     \sigma^*A \equiv \sigma^* B \isType$.
+%   \end{enumerate}
+% \end{lemma}
 
-A pair of simultaneous substitutions $\Delta \vdash \sigma, \sigma'
-\Rightarrow \Delta'$ are defined to be judgmentally equal if their
-component expressions are judgmentally equal in the context $\Delta$:
-i.e.~ if $\Delta \vdash e_j \equiv e'_j : s_j$, for all $j$. We write
-$\Delta \vdash \sigma \equiv \sigma' \Rightarrow \Delta'$ when two
-simultaneous substitutions are judgmentally equal.
+A pair of simultaneous substitutions $\Delta \vdash \sigma \Rightarrow
+\Delta'$ and $\Delta \vdash \sigma' \Rightarrow \Delta'$ are defined
+to be equal if their component expressions are equal in the context
+$\Delta$: i.e.,~ if $\Delta \vdash e_j \equiv e'_j : s_j$, for all
+$j$. We write $\Delta \vdash \sigma \equiv \sigma' \Rightarrow
+\Delta'$ when two simultaneous substitutions are equal.
 
@@
 
+\fixme{Make sure that the semantics is motivated}
 Having defined the language of algebraically-indexed types, we now
@@
 relational interpretations of primitive types with index arguments and
-the universally quantified types requires an interpretation of index
-contexts. We assign interpretations to index contexts in terms of
-\emph{relational environments}, which we now define.
+the universally quantified types require an interpretation of index
+contexts. We assign interpretations to index contexts in terms of sets
+of \emph{relational environments}, which we now define.
 
-A relation environment for a context $\Delta$ is a
-(dependently typed) function that, for each primitive type
-$\tyPrimNm{X}$ and instantiation of its arguments $\sigma$, assigns a
-binary relation $\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure
-interpretation of $\tyPrimNm{X}$:
+A relation environment for a context $\Delta$ is a (dependently typed)
+function $\rho$ that, for each primitive type $\tyPrimNm{X}$ and
+instantiation of its arguments $\sigma$, assigns a binary relation
+$\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure interpretation of
+$\tyPrimNm{X}$:
 \begin{displaymath}
@@
 \primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
-\primTyArity(\tyPrimNm{X})$ such that $\Delta \vdash \sigma \equiv
-\sigma' \Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\
-\tyPrimNm{X}\ \sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
+\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
+\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
+\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
 
@@
 As we stated above, index contexts $\Delta$ are interpreted as certain
-sets of relation environments, i.e.~ subsets of
+sets of relation environments, i.e.,~subsets of
 $\mathrm{RelEnv}(\Delta)$. Depending on the primitive operations that
@@
   \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
-  \relEnv{E}(\Delta)$, along with an simultaneous substitution $\sigma
-  : \Delta' \Rightarrow \Delta$, such that $\rho_1 \circ
+  \relEnv{E}(\Delta)$, along with a simultaneous substitution $\Delta'
+  \vdash \sigma \Rightarrow \Delta$, such that $\rho_1 \circ
   \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists a $\rho
@@
 
-Both of these conditions ensure that the relational interpretation of
-types that we define below behaves correctly with respect to
-simultaneous substitution of index expressions
-(\lemref{lem:tyeqsubst-relational}, below).
+These conditions ensure that the relational interpretation of types
+that we define below behaves correctly with respect to simultaneous
+substitution of index expressions (see
+\lemref{lem:tyeqsubst-relational}, below).
 
@@
 
-Note that both parts of \lemref{lem:tyeqsubst-relational} depend on
-the corresponding parts of \lemref{lem:tyeqsubst-erasure} to be
-``well-typed''.
+Note that the equations in both parts of
+\lemref{lem:tyeqsubst-relational} are well-typed by virtue of the
+corresponding parts of \lemref{lem:tyeqsubst-erasure}.
 
@@
 We now present the rules for well-typed programs over the collection
-of types we generated in
-\autoref{sec:algebraically-indexed-types}. Each well-type program is
-assigned an index-erasure semantics, building on the index-erasure
-semantics of types we defined in
+of types we defined in \autoref{sec:algebraically-indexed-types}. Each
+well-type program is assigned an index-erasure semantics, building on
+the index-erasure semantics of types we defined in
 \autoref{sec:index-erasure-semantics}. Our main result
@@
 interpretation of its type: this is the abstraction theorem for every
-calculus in our general framework.
+calculus that instantiates our general framework.
 
@@
     \inferrule* [right=UnivAbs]
-    {\Delta, i : s; \pi_{i\mathord:s}^*\Gamma \vdash M : A}
+    {\Delta, i \mathord: s; \pi_{i\mathord:s}^*\Gamma \vdash M : A}
     {\Delta; \Gamma \vdash \Lambda i. M : \forall i\mathord:s. A}
@@
 \Gamma \vdash M : A$, we define the \emph{erasure interpretation} as a
-function $\tmSem{e} : \ctxtSem{\Gamma} \to \tySem{A}$, completely
-ignoring the indexing information. In light of
+function $\tmSem{M} : \ctxtSem{\Gamma} \to \tySem{A}$, that completely
+ignores the indexing information. In light of
 \lemref{lem:tyeqsubst-erasure}, we define this function directly on
 the syntax of well-typed programs, rather than on the typing
-derivations. We define $\tmSem{e}$ by the following clauses:
+derivations. The definition of $\tmSem{M}$ is completely standard,
+except for the clauses for universal and existential types:
 \begin{displaymath}
   \begin{array}{@{\hspace{0em}}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l}
-    \tmSem{x}\eta & \isDefinedAs & \eta_x \\
-    \tmSem{(M,N)}\eta & \isDefinedAs & (\tmSem{M}\eta, \tmSem{N}\eta) \\
-    \tmSem{\pi_1M}\eta & \isDefinedAs & \pi_1(\tmSem{M}\eta) \\
-    \tmSem{\pi_2M}\eta & \isDefinedAs & \pi_2(\tmSem{M}\eta) \\
-    \tmSem{\mathrm{inl}\ M}\eta & \isDefinedAs & \mathrm{inl}\ (\tmSem{M}\eta) \\
-    \tmSem{\mathrm{inr}\ M}\eta & \isDefinedAs & \mathrm{inr}\ (\tmSem{M}\eta) \\
-    \left\llbracket
-      \begin{array}{l}
-        \textrm{case}\ M\ \textrm{of}\\
-        \textrm{inl}\ x.N_1;\\
-        \textrm{inr}\ y.N_2
-      \end{array}\right\rrbracket^{\mathit{tm}}\eta & \isDefinedAs &
-    \left\{
-      \begin{array}{ll}
-        \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
-        \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
-      \end{array}
-    \right. \\
-    \tmSem{\lambda x. M}\eta & \isDefinedAs & \lambda v. \tmSem{M}(\eta, v) \\
-    \tmSem{M N}\eta & \isDefinedAs & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
     \tmSem{\Lambda i.\ M}\eta & \isDefinedAs & \tmSem{M}\eta \\
@@
 \end{displaymath}
-In the above, we have written $\eta_x$ to denote the appropriate
-projection from $\eta$ to get the value in the environment
-corresponding to the variable $x$ in the context.
 
@@
 in our system. To state and prove this theorem for open programs, we
-first need to extend the relational interpretation of types to typing
+must extend the relational interpretation of types to typing
 contexts. The relational interpretation of contexts is defined by the
@@
 \begin{lemma}\label{lem:ctxtsubst-rel}
-  If $\Delta \vdash' \Gamma \isCtxt$ and $\sigma : \Delta \Rightarrow
-  \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
+  If $\Delta' \vdash \Gamma \isCtxt$ and $\Delta \vdash \sigma
+  \Rightarrow \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
   $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =

### 2012-06-22 paper/section-further-examples.tex
Some unsorted notes on the examples

@@
-\section{Further Examples}
+\section{More Algebraically Indexed Types}
 \label{sec:further-examples}
+
+\subsection{Singleton Types}
+\label{sec:singleton-types}
+
+Singleton types differ from the ``variation'' types that we have
+considered so far because they nail down exactly the underlying value
+instead of describing how it varies.
+
+Indexing theory contains all the integers as constants, along with
+$+$, $-$ and $*$. Normal arithmetic laws for these
+operations. Program-level operations:
+\begin{displaymath}
+  \begin{array}{@{}c@{\hspace{0.5em}:\hspace{0.5em}}l}
+    \underline{z} & \tyPrim{int}{z} \\ % FIXME: only need zero and one
+    (+) & \forall z_1,z_2\mathord:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_1 + z_2} \\
+    (-) & \forall z_1,z_2\mathord:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_1 - z_2} \\
+    (*) & \forall z_1,z_2\mathord:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_1 * z_2}
+  \end{array}
+\end{displaymath}
+
+\begin{enumerate}
+\item Take integers as a useful example
+\item The relational interpretation is simply that two things are
+  related if they are both equal to the indexing value (and each
+  other).
+\item Types are extremely precise... Can this be quantified?
+\item Indefinability result by considering sub-algebras? Is this a
+  general method? Can't write:
+  \begin{displaymath}
+    \forall z_1,z_2,z_3.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_3}
+  \end{displaymath}
+  because there is no way of getting $z_3$ from $z_1$ and $z_2$. Can
+  prove this using the relational environments. Is there a nice
+  matrix-style way of characterising the allowable things? I guess it
+  is harder when there are two operations lying around (multiplication
+  and addition here). Maybe we could restrict to multiplication by
+  constants? Then we could possibly characterise the definable
+  elements by a matrix representation?
+\end{enumerate}
+
+\subsection{Predicate Types}
+\label{sec:predicate-types}
+
+\begin{enumerate}
+\item Let the indexing theory be predicates on the underlying value
+\item With the laws of boolean algebra
+\item Singleton types are a special case (i.e. with equality to a fixed value)
+\item But also have $\land$, $\lor$, $\lnot$ (and derived $\Rightarrow$)
+\item A pair of things is related if they both satisfy the predicate
+  (and are equal?)
+\item Need a special type for comparisons?
+  \begin{displaymath}
+    (<) : \forall z_1,z_2:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to (\tyPrim{int}{< z_1} \times 
+  \end{displaymath}
+\end{enumerate}
 
@@
 \label{sec:information-flow}
+
+\begin{enumerate}
+\item This is now a special case of the preceeding subsection
+\item Also makes use of the relational aspect, the previous stuff
+  could make do with unary logical predicates as the interpretations
+  of types, but the information flow properties make use of the
+  relational interpretation.
+\end{enumerate}
+
+\fixme{
+  \begin{enumerate}
+  \item Make the definition of control of information flow clearer
+  \item Make the example a bit more compelling
+  \item Put in some examples of composite principals formed from
+    boolean logic, and the use of boolean algebra to do reasoning in
+    the types. List the laws of boolean algebra, and the laws of
+    Heyting algebra. Need to explain the difference between 
+  \item Possibly use singleton types to be able to do more interesting
+    security policies?
+  \item Put in the formal statement of the information flow property
+    and the proof.
+  \end{enumerate}
+}
 
@@
 \cite{sabelfeld01per}, information flow can be captured semantically
-through the use of partial equivalence relations (PERs). PERs form an
-instance of Reynolds' approach to abstraction and representation
-independence. Abadi, Banerjee, Heintze and Riecke \cite{abadi99core}
-built a \emph{Core Calculus for Dependency}, using a type system based
-around a security level indexed monad $T_lA$. Tse and Zdancewic
-\cite{tse04translating} showed that it is possible to translate Abadi
-et al.'s calculus into System F, translating the monadic type $T_lA$
-to $\alpha_l \to A$ for some free type variable $\alpha_l$, and making
-direct use of Reynolds' abstraction theorem to prove information flow
-properties. We now show how a refined variant of Tse and Zdancewic's
-translation can be expressed via algebraically-indexed types.
+through the use of partial equivalence relations (PERs). Interpreting
+types as PERs forms an instance of Reynolds' approach to abstraction
+and representation independence. Abadi, Banerjee, Heintze and Riecke
+\cite{abadi99core} built a \emph{Core Calculus for Dependency}, using
+a type system based around a security level indexed monad $T_lA$. Tse
+and Zdancewic \cite{tse04translating} showed that it is possible to
+translate Abadi et al.'s calculus into System F, translating the
+monadic type $T_lA$ to $\alpha_l \to A$ for some free type variable
+$\alpha_l$, and making direct use of Reynolds' abstraction theorem to
+prove information flow properties. We now show how a refined variant
+of Tse and Zdancewic's translation can be expressed via
+algebraically-indexed types.
 
@@
 
+\begin{enumerate}
+\item Treat the information flow thing as a type isomorphism, in the
+  same way that Tse and Zdancewic do
+  \begin{displaymath}
+    \begin{array}{l}
+    \forall \vec{p}\mathord:\mathsf{principal}.\ T_{e_1}\tyPrimNm{bool} \to T_{e_2}\tyPrimNm{bool}\\
+    \hspace{4em}\cong \\
+    \left\{
+      \begin{array}{ll}
+        \tyPrimNm{bool} \tyArr \tyPrimNm{bool} & \textrm{if }e_2 \vdash e_1 \\
+        \tyPrimNm{bool} & \textrm{if }e_2 \not\vdash e_1
+      \end{array}
+    \right.
+  \end{array}
+  \end{displaymath}
+  where $\tyPrimNm{bool} = \tyUnit + \tyUnit$.
+\item mention the issue with bogus tokens if we allow for a
+  call-by-name style of non-termination
+\item if we axiomatise intuitionistic logic instead of
+  boolean/classical logic, we get a different type isomorphism.
+\item The isomorphism relies upon $e_1 \vdash e_2 \Leftrightarrow e_1
+  \lor e_2 = e_1$.
+\item Relevant logic? Look at the Wikipedia page to see how to do
+  residuated (commutative) monoids on lattices with just
+  equations. With this, we will be able to do substructural logics
+  too. As long as we know that a particular axiomatisation is sound
+  and complete for some reasoning system, then we can state and prove
+  the non-interference isomorphism above.
+\item Compare to Tse and Zdancewic's addition of axioms to the logic.
+\item Internalising equality? The logical structure can then be used
+  on types as well as everything else. Need the special rules for
+  equality... (see the Fomega paper).
+\end{enumerate}
+
 % Go a step further by considering equivalence relations on word32, and
@@
 
-% \begin{eqnarray*}
-%   \underline{c} & : & \forall \epsilon\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon} \\
-%   (+) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
-%   (-) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
-%   (*) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1\epsilon_2}
-% \end{eqnarray*}
+\begin{displaymath}
+  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
+    \underline{c} & \forall \epsilon\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon} \\
+    (+) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
+    (-) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
+    (*) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1\epsilon_2}
+  \end{array}
+\end{displaymath}
+
+\begin{enumerate}
+\item State the operations and equations (rational powers, addition
+  and multiplication (remember: positive reals))
+\item Derive subtyping by addition of $0$
+\item Do the probability monad as a special case, with a specific
+  relational interpretation that relates things according to the
+  differential privacy notion. Need to think about the
+  measure-theoretic problems...
+\item Work out the details of the continuity example, especially the
+  implementation of an ADT for continuous functions that can handle
+  polynomials.
+\end{enumerate}
+
+\subsection{Abelian Group Indexed Types}
+\label{sec:abelian-group-indexed-types}
+
+\newcommand{\Grp}{\mathit{Grp}}
+
+Primitive operations $\Gamma_\Grp$:
+\begin{displaymath}
+  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
+    1 & \tyPrim{G}{1} \\
+    (\cdot) & \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to \tyPrim{G}{a b} \\
+    \ ^{-1} & \forall a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}
+  \end{array}
+\end{displaymath}
+
+For the index-erasure interpretation, we assume a group $(G, 1, \cdot,
+\ ^{-1})$. Let $Z(G)$ be the centre of the $G$: i.e.,~the set of
+elements that commute with every other element in $G$. We set
+$\tyPrimSem{G} = G$. The primitive operations in $\Gamma_\Grp$ are
+interpreted just as the unit, multiplication and inverse operations
+of $G$. For each index context $\Delta$, the set
+$\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be
+$\{ \rho_{(H,h)} \sepbar H\textrm{ is a subgroup of }(\Delta
+\Rightarrow \mathsf{G})/\equiv, h\textrm{ is a group homomorphism }H
+\to Z(G) \}$, where
+\begin{displaymath}
+  \rho_{(H,h)}\ \tyPrimNm{G}\ (e) = \left\{
+    \begin{array}{ll}
+      \{ \} & \textrm{if}\ e\not\in H \\
+      \{ (x, h(e)\cdot x) \sepbar x \in G \} & \textrm{if}\ e\in H
+    \end{array}
+  \right.
+\end{displaymath}
+
+\begin{lemma}
+  For all $\Delta$ and $\rho \in \relEnv{E}_\Grp(\Delta)$,
+  $(\eta_\Grp,\eta_\Grp) \in \rsem{\Gamma_\Grp}{\relEnv{E}}\rho$.
+\end{lemma}
+
+\begin{theorem}
+  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
+  natural numbers $A$ and $m$-by-$1$ matricies of integers $b$,
+  there exists a program $M$ such that
+  \begin{equation}
+    \label{eq:fo-group-type}
+    \Gamma_\Grp \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
+  \end{equation}
+  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e = i_1^{b_1}\dots
+  i_m^{b_m}$, if and only if there is a solution $\vec{x} =
+  (x_1,\cdots,x_n)$ in the integers to the equation $A x^\top = b$.
+\end{theorem}
+
+\fixme{Define shorthand notation $x^z$}
+
+\begin{proof}
+  (If) The program $\Lambda i_1\dots i_m.\ \lambda g_1\dots g_n.\
+  g_1^{x_1}\dots g_n^{x_n}$ satisfies the typing judgment
+  (\ref{eq:fo-group-type}).
+
+  (Only if) Assume that there is a program $M$ satisfying the typing
+  judgment (\ref{eq:fo-group-type}). By \thmref{thm:abstraction} we
+  know that for all relational environments $\rho \in
+  \relEnv{E}_\Grp(i_1,\dots,i_m)$, and all $g_j,g'_j \in G$,
+  \begin{displaymath}
+    \begin{array}{l}
+      (g_1,g'_1) \in \rho\ \tyPrimNm{G}\ (e_1) \land \dots \land (g_n,g'_n) \in \rho\ \tyPrimNm{G}\ (e_n) \Rightarrow \\
+      \quad (\tmSem{M}g_1\dots g_n, \tmSem{M}g'_1\dots g'_n) \in \rho\ \tyPrimNm{G}\ (e)
+    \end{array}
+  \end{displaymath}
+  We select the relational environments
+  \begin{displaymath}
+    \rho\ \tyPrimNm{G}\ (e) = \left\{
+      \begin{array}{ll}
+        \{(g,g) \sepbar g \in \mathcal{G} \} &
+        \begin{array}[t]{@{}l}
+          \textrm{if}\ e = e_1^{x_1}\dots e_n^{x_n} \\
+          \textrm{ for some }x_1,\dots,x_n \in \mathbb{Z}
+        \end{array}
+        \\
+        \{\} & \textrm{otherwise}
+      \end{array}
+    \right.
+  \end{displaymath}
+  Now, for any $g_1,\dots,g_m \in \mathcal{G}$, we know that, for all
+  $j$, $(g_j,g_j) \in \rho\ \tyPrimNm{G}\ (e_j)$, so we know that
+  $(\tySem{M}g_1\dots g_n, \tySem{M}g_1 \dots g_n) \in \rho\
+  \tyPrimNm{G}\ (e)$. Thus there exist integers $x_1,...,x_n$ such
+  that $e = e_1^{x_1}\dots e_n^{x_n}$. But we also know that $e =
+  i_1^{b_1}...i_m^{b_m}$. By the cancellation property of free groups,
+  we learn that $x_1,...,x_n$ is the solution we need.
+\end{proof}
+
+\begin{enumerate}
+\item Can add any other operations that distribute over the group
+  multiplication.
+\item What about partial division operations? Make division just
+  return $0$.
+\item Adding square root as an operation. Need to check that all the
+  other operations preserve the new relational environments. Square
+  root is also supported by the plain relational environments, so the
+  type isomorphism result still holds.
+\end{enumerate}
+
+\subsubsection{Type isomorphisms}
+\label{sec:abelian-group-type-isos}
+
+\begin{enumerate}
+\item This only needs the simpler relational environments, but needs
+  the group used for the index-erasure semantics to be abelian.
+\item One direction is easy, but the other requires the use of the
+  free theorem for the type.
+\item There are two version: one with some fixed ground type as the
+  result, and one with the same type as the result, but with the same
+  parameter.
+\item Integrates into the Smith Normal Form thing too, since
+  invertible integer-valued matrices can be turned into type
+  isomorphisms.
+\item Partiality?
+\end{enumerate}
+
+\fixme{Define what a type isomorphism is}
+
+\begin{theorem}
+  For all natural numbers $n$, the types
+  \begin{displaymath}
+    \tau_{n} \isDefinedAs \forall a\mathord:\mathsf{G}.\ \underbrace{\tyPrim{G}{a} \to ... \to \tyPrim{G}{a}}_{n+1\textrm{ times}} \to \tyPrim{G}{a}
+  \end{displaymath}
+  and
+  \begin{displaymath}
+    \sigma_{n} \isDefinedAs \underbrace{\tyPrim{G}{1} \to ... \tyPrim{G}{1}}_{n\textrm{ times}} \to \tyPrim{G}{1}
+  \end{displaymath}
+  are isomorphic.
+\end{theorem}
+
+\begin{proof}
+  We define programs $\Gamma_\Grp \vdash M : \tau_n \to \sigma_n$ and
+  $\Gamma_\Grp \vdash M^{-1} : \sigma_n \to \tau_n$ as follows:
+  \begin{displaymath}
+    M = \lambda f. \lambda g_1 \dots g_n.\ f\ [1]\ 1\ g_1\ \dots\ g_n
+  \end{displaymath}
+  and
+  \begin{displaymath}
+    M^{-1} = \lambda f. \Lambda a.\ \lambda g_0 \dots g_n.\ (f\ (g_1 g^{-1}_0)\ \dots\ (g_n g^{-1}_0)) g_0
+  \end{displaymath}
+  In one direction, the proof that these are mutually inverse is
+  straightfoward, using only the fact that $G$ is a group. For any
+  program $\Gamma_\Grp \vdash f : \sigma_n$, we have
+  \begin{displaymath}
+    \begin{array}{cl}
+        & \tmSem{M\ (M^{-1}\ f)}\eta_\Grp \\
+      = & (\tmSem{M}\eta_{\Grp})\ (\lambda g_0 \dots g_n.\ (\tmSem{f}\eta_\Grp\ (g_1 g^{-1}_0)\ \dots\ (g_n g^{-1}_0)) g_0) \\
+      = & \lambda g_1\dots g_n.\ \tmSem{f}\eta_\Grp\ (g_1 1^{-1})\ \dots\ (g_n 1^{-1})) 1) \\
+      = & \lambda g_1\dots g_n.\ \tmSem{f}\eta_\Grp\ g_1\ \dots\ g_n \\
+      = & \tmSem{f}\eta_\Grp
+    \end{array}
+  \end{displaymath}
+  as required. In the other direction, we need to use the abstraction
+  theorem. For any program $\Gamma_\Grp \vdash f : \tau_n$, we have
+  \begin{displaymath}
+    \begin{array}{cl}
+        & \tmSem{M^{-1}\ (M\ f)}\eta_\Grp \\
+      = & (\tmSem{M^{-1}}\eta_\Grp)\ (\lambda g_1\dots g_n.\ (\tmSem{f}\eta_\Grp\ 1\ g_1\ \dots\ g_n)) \\
+      = & \lambda g_0\dots g_n.\ (\tmSem{f}\eta_\Grp\ 1\ (g_1g^{-1}_0)\ \dots\ (g_ng^{-1}_0))g_0
+    \end{array}
+  \end{displaymath}
+  By \thmref{thm:abstraction}, we know that the following holds of
+  $\tmSem{f}$:
+  \begin{displaymath}
+    \forall a.\ \forall g_0 \dots g_n.\ \tmSem{f}\eta_\Grp\ (g_0a)\ \dots\ (g_na) = (\tmSem{f}\eta_\Grp\ g_0\ \dots\ g_n)a
+  \end{displaymath}
+  As a special case, we set $a = g^{-1}_0$, and learn that
+  \begin{displaymath}
+    \forall g_0 \dots g_n.\ \tmSem{f}\eta_\Grp\ 1\ (g_1g^{-1}_0) \dots\ (g_ng^{-1}_0) = (\tmSem{f}\eta_\Grp\ g_0\ \dots\ g_n)g^{-1}_0.
+  \end{displaymath}
+  By functional extensionality, we are done.
+\end{proof}
 
@@
 \label{sec:monoid-indexed-types}
+
+\newcommand{\Mon}{\textit{Mon}}
 
@@
 \begin{displaymath}
-  \begin{array}{c@{\hspace{0.5em}:\hspace{0.5em}}l}
+  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
     1 & \tyPrim{M}{1} \\
@@
 
-For the index-erasure interpretation, we assume a commutative monoid
+For the index-erasure interpretation, we assume a monoid
 $\mathcal{M}$. The primitive type $\tyPrimNm{M}$ is interpreted as
 $\tySem{\tyPrim{M}{e}} = \mathcal{M}$. The primitive operations in
-$\Gamma_{\mathit{Mon}}$ are interpreted just as the unit and multiplication of
-the monoid $\mathcal{M}$.
+$\Gamma_{\mathit{Mon}}$ are interpreted just as the unit and
+multiplication of the monoid $\mathcal{M}$.
 
@@
 
-For each index context $\relEnv{E}(\Delta)$, the set
-$\relEnv{E}(\Delta)$ of relational environments is defined to be $\{
-\rho_{(M,h)} \sepbar M\textrm{ is a submonoid of }(\Delta \Rightarrow
-\mathsf{M})/\equiv, h\textrm{ is a monoid homomorphism }M \to
-Z(\mathcal{M}) \}$, where
+For each index context $\Delta$, the set
+$\relEnv{E}_\Mon(\Delta)$ of relational environments is
+defined to be $\{ \rho_{(M,h)} \sepbar M\textrm{ is a submonoid of
+}(\Delta \Rightarrow \mathsf{M})/\equiv, h\textrm{ is a monoid
+  homomorphism }M \to Z(\mathcal{M}) \}$, where
 \begin{displaymath}
@@
 \end{displaymath}
-Note the mapping to the centre of the monoid $\mathcal{M}$.
+Note the mapping to the centre of the monoid
+$\mathcal{M}$. \fixme{introduce the centre notation earlier on}
 
 \begin{lemma}
-  For all $\Delta$ and $\rho \in \relEnv{E}(\Delta)$,
-  $(\eta_{M},\eta_{M}) \in \rsem{\Gamma_{\mathit{Mon}}}{\relEnv{E}}\rho$.
+  For all $\Delta$ and $\rho \in \relEnv{E}_\Mon(\Delta)$,
+  $(\eta_\Mon,\eta_\Mon) \in \rsem{\Gamma_\Mon}{\relEnv{E}}\rho$.
 \end{lemma}
@@
     \label{eq:fo-monoid-type}
-    \Gamma_{\mathit{Mon}} \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
+    \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
   \end{equation}

### 2012-07-03 paper/section-further-examples.tex
Additions to the general framework, and fill in some of the examples
- Add a subsection on the construction of relational environments (unfinished)
- Fill in the monoid indexed types section
- Rearrange the example sections a bit

@@
-\section{More Algebraically Indexed Types}
-\label{sec:further-examples}
+\section{Monoid Indexed Types}
+\label{sec:monoid-indexed-types}
 
-\subsection{Singleton Types}
+\newcommand{\Mon}{\textit{Mon}}
+
+In the last section we looked at types indexed by abelian groups. We
+now examine an instantiation of our general framework with a monoid
+type indexed by elements of a commutative monoid. Our main example of
+monoid indexed types is strings indexed by ``taint'' markers: when
+strings are appended their taint markers are combined. Thanks to the
+information in the indexes, we are able to prove an indefinability
+result (\thmref{thm:monoid-indefinability}).
+
+We assume a single indexing sort $\mathsf{M}$, with the monoid
+operations of unit and multiplication along with the commutative
+monoid axioms. There is a single primitive type $\tyPrimNm{M}$, with
+$\primTyArity(\tyPrimNm{M}) = [\mathsf{M}]$. For the index-erasure
+semantics, we assume some (not necessarily commutative) monoid $M$,
+which we also write multiplicatively. Thus, we set
+$\tyPrimSem{\tyPrimNm{M}} = M$. The collection of primitive operations
+$\Gamma_{\mathit{Mon}}$ is simply the two monoid operations:
+\begin{displaymath}
+  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{2em}}r@{\hspace{0.5em}:\hspace{0.5em}}l}
+    1 & \tyPrim{M}{1}, &
+    (\cdot) & \forall a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{a b}
+  \end{array}
+\end{displaymath}
+which we interpret using the monoid structure of $M$, giving an
+environment $\eta_\Mon \in \ctxtSem{\Gamma_\Mon}$.
+
+For the sets of relational environments, we make use of the
+construction in \autoref{sec:constr-rel-env}, setting the chosen model
+$\mathcal{M}$ to be the \emph{centre} $Z(M)$ of $M$ (recall that the
+centre of a monoid is the set of all elements that commute with every
+other element). We then set $R_{\tyPrimNm{M}}(m) = \{ (n,mn) \sepbar n
+\in M \}$ and $R^\bullet_{\tyPrimNm{M}} = \emptyset$. By setting the
+model $\mathcal{M}$ to only be the elements of $M$ that commute with
+everything else, we ensure that the following lemma holds:
+\begin{lemma}
+  For all $\Delta$ and $\rho \in \relEnv{E}_\Mon(\Delta)$,
+  $(\eta_\Mon,\eta_\Mon) \in \rsem{\Gamma_\Mon}{\relEnv{E}}\rho$.
+\end{lemma}
+
+\newcommand{\mscup}{\stackrel{ms}\cup}
+
+\begin{example}
+  By selecting a suitable monoid $M$, we can simulate a programming
+  language that dynamically annotates strings with taint markers. Let
+  $\Sigma$ be a set of characters from some alphabet, and let $T$ be a
+  set of taint markers. Set $M = \Sigma^* \times
+  \mathcal{MS}_{\mathit{fin}}(T)$, where $\mathcal{MS}_{\mathit{fin}}$
+  denotes the finite power set operator. Since both $\Sigma^*$ and
+  $\mathcal{MS}_{\mathit{fin}}(T)$ are monoids (the latter with
+  multiset union as the monoid multiplication, which we write as
+  $\mscup$), $M$ is itself a monoid. The centre of $M$, $Z(M)$,
+  consists of all elements of the form $([], t)$ for multisets of
+  taint markers.
+
+  As an example of a free theorem for this instantiation of our
+  general framework, consider programs of the following type:
+  \begin{displaymath}
+    ; \Gamma_\Mon \vdash M : \forall a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{aa}
+  \end{displaymath}
+  From this type we can derive the following free theorem about
+  $M$. For all $a, b \in \mathcal{MS}_{\mathit{fin}}(T)$, $(s_1,t_1),
+  (s_2,t_2) \in \Sigma^*$,
+  \begin{displaymath}
+    \begin{array}{l}
+      \pi_2(\tmSem{M}\ (s_1,t_1 \mscup a)\ (s_2,t_2 \mscup b)) = \\
+      \hspace{2em}\pi_2(\tmSem{M}\ (s_1,t_1)\ (s_2,t_2)) \mscup a \mscup a
+    \end{array}
+  \end{displaymath}
+  We read this result as follows: if we taint the first input by $a$
+  and the second by $b$, then the taint of the output varies according
+  to $M$'s type. In this case the output contains two copies of $a$
+  and none of $b$.
+\end{example}
+
+% if $b$ does not appear in the output, then it cannot have been used in
+% the program. Turn this into a theorem...
+
+% If forall i1...,in. M<e0> \to ... \to M<em> -> M<e>
+% and e0 is not a factor of e, then we can get rid of the first input
+
+% It ought to be invariant under its first input... Set R^\bullet to
+% be the everywhere relation....
+
+% prove this by using a relational environment that excludes
+
+\begin{theorem}
+  \begin{equation}
+    \label{eq:fo-monoid-type-insensitive}
+    ; \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
+  \end{equation}
+
+\end{theorem}
+
+\begin{theorem}
+  \label{thm:monoid-indefinability}
+  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
+  natural numbers $A$ and $m$-by-$1$ matricies of natural numbers $b$,
+  there exists a program $M$ such that
+  \begin{displaymath}
+    \label{eq:fo-monoid-type}
+    ; \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
+  \end{displaymath}
+  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e =
+  i_1^{b_1}\dots i_m^{b_m}$, if and only if there is a solution
+  $\vec{x} = (x_1,\cdots,x_n)$ in the natural numbers to the equation
+  $A x^\top = b$.
+\end{theorem}
+
+\begin{example}
+  By \thmref{thm:monoid-indefinability}, it is impossible to write a
+  program with the type $\forall m_1,m_2 \mathord: \mathsf{M}.\
+  \tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_1}$, showing that it is
+  impossible to convert a doubly tainted value to a singly tainted
+  one.
+
+  In contrast to the case of abelian group indexed types, the lack of
+  inverses in the monoid indexed setting means that the type $\forall
+  m_1,m_2\mathord:\mathsf{M}.\ \tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_2}
+  \to \tyPrim{M}{m_1}$ is also uninhabited. There are no naturals
+  $a_1,a_2$ such that $(m_1m_2)^{a_1}m_2^{a_2} \equiv m_1$.
+\end{example}
+
+% \fixme{Finish this} We can further assume that there are other
+% operations that distribute over the multiplication of the monoid, for
+% example semi-ring operations (e.g.~relational algebra).
+
+\section{Metric Spaces and Simple Continuity Analysis}
+\label{sec:continuity-analysis}
+
+Our final motivating example of algebraically indexed types and
+representation independence makes use of the metric space structure of
+the real numbers. In the geometry example, we used relational
+interpretations that related values via geometrical
+transformations. An alternative relational interpretation of the real
+numbers is to relate values that are close, in the metric space
+sense. By indexing types by a measure of closeness, we can actually
+state continuity of functions as a typing problem.
+
+We introduce an indexing sort $\mathsf{R}^{>0}$, representing positive
+distances between real numbers. Our type of real numbers is indexed by
+expressions of sort $\mathsf{R}^{>0}$, and we use the following
+index-erasure and relational interpretations:
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
+    \tySem{\tyPrim{real}{e}} & \mathbb{R} \\
+    \rsem{\tyPrim{real}{e}}{}\rho & \{ (x,x') \sepbar |x-x'| < \sem{e}\rho \}
+  \end{array}
+\end{displaymath}
+Thus, two real numbers are related if they only differ by the positive
+real number assigned to $e$. With this interpretation, and a
+straightforward interpretation of existential types, we can state the
+standard $\epsilon$-$\delta$ definition of continuity as a type:
+\begin{displaymath}
+  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
+\end{displaymath}
+Chaudhuri, Gulwani and Lublinerman \cite{chaudhuri10continuity} have
+presented a program logic based approach to verifying the continuity
+of programs. Algebraically indexed types give us a way of expressing
+and verifying continuity properties of functions in a type based way.
+
+\begin{displaymath}
+  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
+    \underline{c} & \forall \epsilon\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon} \\
+    (+) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
+    (-) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
+    (*) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1\epsilon_2}
+  \end{array}
+\end{displaymath}
+
+\begin{enumerate}
+\item State the operations and equations (rational powers, addition
+  and multiplication (remember: positive reals))
+\item Derive subtyping by addition of $0$
+\item Do the probability monad as a special case, with a specific
+  relational interpretation that relates things according to the
+  differential privacy notion. Need to think about the
+  measure-theoretic problems...
+\item Work out the details of the continuity example, especially the
+  implementation of an ADT for continuous functions that can handle
+  polynomials.
+\end{enumerate}
+
+\section{Singleton Types}
 \label{sec:singleton-types}
+
+\newcommand{\Sing}{\mathit{Sing}}
 
@@
 \begin{enumerate}
-\item Take integers as a useful example
-\item The relational interpretation is simply that two things are
-  related if they are both equal to the indexing value (and each
-  other).
-\item Types are extremely precise... Can this be quantified?
-\item Indefinability result by considering sub-algebras? Is this a
-  general method? Can't write:
-  \begin{displaymath}
-    \forall z_1,z_2,z_3.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_3}
-  \end{displaymath}
-  because there is no way of getting $z_3$ from $z_1$ and $z_2$. Can
-  prove this using the relational environments. Is there a nice
-  matrix-style way of characterising the allowable things? I guess it
-  is harder when there are two operations lying around (multiplication
-  and addition here). Maybe we could restrict to multiplication by
-  constants? Then we could possibly characterise the definable
-  elements by a matrix representation?
+\item Given an algebraic theory, and a model, we can define a
+  singleton type system.
+\item Need to carefully define this in general, and give the
+  appropriate relational environments
+\item Give the integers (freely generated monoid) as an example
 \end{enumerate}
 
-\subsection{Predicate Types}
-\label{sec:predicate-types}
+\begin{theorem}
+  There exists a program
+  \begin{displaymath}
+    -; \Gamma_\Sing \vdash M : \forall i_1,...,i_m\mathord:\mathsf{S}.\ \tyPrim{S}{e_1} \to ... \to \tyPrim{S}{e_n} \to \tyPrim{S}{e}
+  \end{displaymath}
+  if and only if there exists an index expression
+  $i'_1\mathord:\mathsf{S}, ..., i'_n\mathord:\mathsf{S} \vdash e' :
+  \mathsf{S}$ such that $(e_1,...,e_n)^*e' \equiv e$.
+\end{theorem}
 
-\begin{enumerate}
-\item Let the indexing theory be predicates on the underlying value
-\item With the laws of boolean algebra
-\item Singleton types are a special case (i.e. with equality to a fixed value)
-\item But also have $\land$, $\lor$, $\lnot$ (and derived $\Rightarrow$)
-\item A pair of things is related if they both satisfy the predicate
-  (and are equal?)
-\item Need a special type for comparisons?
-  \begin{displaymath}
-    (<) : \forall z_1,z_2:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to (\tyPrim{int}{< z_1} \times 
-  \end{displaymath}
-\end{enumerate}
+\begin{proof}
+  (If) Use the index expression $e'$ to define $M$, replacing each
+  index operation with its program-level counterpart.
 
-\subsection{Logical Information Flow}
+  (Only if) \fixme{use a special relational environment...}
+\end{proof}
+
+% \subsection{Predicate Types}
+% \label{sec:predicate-types}
+
+% \begin{enumerate}
+% \item Let the indexing theory be predicates on the underlying value
+% \item With the laws of boolean algebra
+% \item Singleton types are a special case (i.e. with equality to a fixed value)
+% \item But also have $\land$, $\lor$, $\lnot$ (and derived $\Rightarrow$)
+% \item A pair of things is related if they both satisfy the predicate
+%   (and are equal?)
+% \item Need a special type for comparisons?
+%   \begin{displaymath}
+%     (<) : \forall z_1,z_2:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to (\tyPrim{int}{< z_1} \times 
+%   \end{displaymath}
+% \end{enumerate}
+
+\section{Logical Information Flow}
 \label{sec:information-flow}
@@
 % flow. Will also need to include singleton types.
-
-\subsection{Metric Spaces and Simple Continuity Analysis}
-\label{sec:continuity-analysis}
-
-Our final motivating example of algebraically indexed types and
-representation independence makes use of the metric space structure of
-the real numbers. In the geometry example, we used relational
-interpretations that related values via geometrical
-transformations. An alternative relational interpretation of the real
-numbers is to relate values that are close, in the metric space
-sense. By indexing types by a measure of closeness, we can actually
-state continuity of functions as a typing problem.
-
-We introduce an indexing sort $\mathsf{R}^{>0}$, representing positive
-distances between real numbers. Our type of real numbers is indexed by
-expressions of sort $\mathsf{R}^{>0}$, and we use the following
-index-erasure and relational interpretations:
-\begin{displaymath}
-  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\tyPrim{real}{e}} & \mathbb{R} \\
-    \rsem{\tyPrim{real}{e}}{}\rho & \{ (x,x') \sepbar |x-x'| < \sem{e}\rho \}
-  \end{array}
-\end{displaymath}
-Thus, two real numbers are related if they only differ by the positive
-real number assigned to $e$. With this interpretation, and a
-straightforward interpretation of existential types, we can state the
-standard $\epsilon$-$\delta$ definition of continuity as a type:
-\begin{displaymath}
-  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
-\end{displaymath}
-Chaudhuri, Gulwani and Lublinerman \cite{chaudhuri10continuity} have
-presented a program logic based approach to verifying the continuity
-of programs. Algebraically indexed types give us a way of expressing
-and verifying continuity properties of functions in a type based way.
-
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    \underline{c} & \forall \epsilon\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon} \\
-    (+) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
-    (-) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
-    (*) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1\epsilon_2}
-  \end{array}
-\end{displaymath}
-
-\begin{enumerate}
-\item State the operations and equations (rational powers, addition
-  and multiplication (remember: positive reals))
-\item Derive subtyping by addition of $0$
-\item Do the probability monad as a special case, with a specific
-  relational interpretation that relates things according to the
-  differential privacy notion. Need to think about the
-  measure-theoretic problems...
-\item Work out the details of the continuity example, especially the
-  implementation of an ADT for continuous functions that can handle
-  polynomials.
-\end{enumerate}
 
@@
 
-\subsection{Monoid Indexed Types}
-\label{sec:monoid-indexed-types}
-
-\newcommand{\Mon}{\textit{Mon}}
-
-Primitive type is a monoid, and we index by a commutative monoid. An
-example of this is strings with tainting information, or bags, or sets
-of stuff. This gives a slightly different information flow analysis to
-the logical information flow above. In particular, we can now count
-uses of input data.
-
-Primitive operations $\Gamma_{\mathit{Mon}}$:
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    1 & \tyPrim{M}{1} \\
-    (\cdot) & \forall a,b.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{a\cdot b}
-  \end{array}
-\end{displaymath}
-
-For the index-erasure interpretation, we assume a monoid
-$\mathcal{M}$. The primitive type $\tyPrimNm{M}$ is interpreted as
-$\tySem{\tyPrim{M}{e}} = \mathcal{M}$. The primitive operations in
-$\Gamma_{\mathit{Mon}}$ are interpreted just as the unit and
-multiplication of the monoid $\mathcal{M}$.
-
-% Need the following definitions: syntatic group/monoid arising from
-% the syntax, and a neater way of defining relational
-% environments. Also, a neat way of formally introducing a load of
-% primitive operations.
-
-For each index context $\Delta$, the set
-$\relEnv{E}_\Mon(\Delta)$ of relational environments is
-defined to be $\{ \rho_{(M,h)} \sepbar M\textrm{ is a submonoid of
-}(\Delta \Rightarrow \mathsf{M})/\equiv, h\textrm{ is a monoid
-  homomorphism }M \to Z(\mathcal{M}) \}$, where
-\begin{displaymath}
-  \rho_{(M,h)}\ \tyPrimNm{M}\ (e) = \left\{
-    \begin{array}{ll}
-      \{ \} & \textrm{if}\ e\not\in M \\
-      \{ (x, h(e)\cdot x) \sepbar x \in \mathcal{M} \} & \textrm{if}\ e\in M
-    \end{array}
-  \right.
-\end{displaymath}
-Note the mapping to the centre of the monoid
-$\mathcal{M}$. \fixme{introduce the centre notation earlier on}
-
-\begin{lemma}
-  For all $\Delta$ and $\rho \in \relEnv{E}_\Mon(\Delta)$,
-  $(\eta_\Mon,\eta_\Mon) \in \rsem{\Gamma_\Mon}{\relEnv{E}}\rho$.
-\end{lemma}
-
-\begin{theorem}
-  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
-  natural numbers $A$ and $m$-by-$1$ matricies of natural numbers $b$,
-  there exists a program $M$ such that
-  \begin{equation}
-    \label{eq:fo-monoid-type}
-    \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
-  \end{equation}
-  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e =
-  i_1^{b_1}\dots i_m^{b_m}$, if and only if there is a solution
-  $\vec{x} = (x_1,\cdots,x_n)$ in the natural numbers to the equation
-  $A x^\top = b$.
-\end{theorem}
-
-\begin{proof}
-  (If) The program $\Lambda i_1\dots i_m.\ \lambda z_1\dots z_n.\
-  z_1^{x_1}\dots z_n^{x_n}$ satisfies the typing judgment
-  (\ref{eq:fo-monoid-type}).
-
-  (Only if) Assume that there is a program $M$ satisfying the typing
-  judgment (\ref{eq:fo-monoid-type}). By \thmref{thm:abstraction} we
-  know that for all relational environments $\rho \in
-  \relEnv{E}(i_1,\dots,i_m)$, and all $z_j,z'_j \in \mathcal{M}$,
-  \begin{displaymath}
-    \begin{array}{l}
-      (z_1,z'_1) \in \rho\ \tyPrimNm{M}\ (e_1) \land \dots \land (z_n,z'_n) \in \rho\ \tyPrimNm{M}\ (e_n) \Rightarrow \\
-      \quad (\tmSem{M}z_1\dots z_n, \tmSem{M}z'_1\dots z'_n) \in \rho\ \tyPrimNm{M}\ (e)
-    \end{array}
-  \end{displaymath}
-  We select the relational environments
-  \begin{displaymath}
-    \rho\ \tyPrimNm{M}\ (e) = \left\{
-      \begin{array}{ll}
-        \{(z,z) \sepbar z \in \mathcal{M} \} &
-        \begin{array}[t]{@{}l}
-          \textrm{if}\ e = e_1^{x_1}\dots e_n^{x_n} \\
-          \textrm{ for some }x_1,\dots,x_n \in \mathbb{N}
-        \end{array}
-        \\
-        \{\} & \textrm{otherwise}
-      \end{array}
-    \right.
-  \end{displaymath}
-  Now, for any $z_1,\dots,z_m \in \mathcal{M}$, we know that, for all
-  $j$, $(z_j,z_j) \in \rho\ \tyPrimNm{M}\ (e_j)$, so we know that
-  $(\tySem{M}z_1\dots z_n, \tySem{M}z_1 \dots z_n) \in \rho\
-  \tyPrimNm{M}\ (e)$. Thus there exist $x_1,...,x_n$ such that $e =
-  e_1^{x_1}\dots e_n^{x_n}$. But we also know that $e =
-  i_1^{b_1}...i_m^{b_m}$. By the cancellation property of free
-  monoids, we learn that $x_1,...,x_n$ is the solution we need.
-\end{proof}
-
-We can further assume that there are other operations that distribute
-over the multiplication of the monoid, for example semi-ring
-operations (e.g.~relational algebra).
-
-Examples
-\begin{enumerate}
-\item Finite sets of stuff.
-\item Let $\mathcal{M} = \Sigma^* \times
-  \mathcal{P}_{\mathit{fin}}(C)$, where $\Sigma$ is some alphabet, and
-  $C$ is some set of taint markers. The centre of this monoid is
-  $\{([], c) \sepbar c \in \mathcal{P}_{\mathit{fin}}(C) \}$.
-\end{enumerate}
-
 %%% Local Variables:

### 2012-07-03 paper/section-general-framework.tex
Additions to the general framework, and fill in some of the examples
- Add a subsection on the construction of relational environments (unfinished)
- Fill in the monoid indexed types section
- Rearrange the example sections a bit

@@
 \begin{example}[Two-Dimensional Geometry]
+  \label{ex:two-dim-geo-operations}
   The two-dimensional geometry system has a sort for each of the
@@
 
-% \begin{example}[Logical Information Flow]
-%   \fixme{Do this}
-% \end{example}
-
-% \begin{example}[Units of Measure]
-%   To instantiate our general framework as Kennedy's original units
-%   of measure system, we would take the following settings. For the
-%   collection of sorts we take $\SortSet = \{ \mathsf{unit} \}$. For
-%   the collection of index operations, we take $\mathit{IndexOp} = \{1,
-%   -\cdot-, -^{-1}\}$, with:
-%   \begin{eqnarray*}
-%     \indexOpArity{(1)} &=& ([],\mathsf{unit}) \\
-%     \indexOpArity{(-\cdot-)} & = & ([\mathsf{unit},\mathsf{unit}],\mathsf{unit}) \\
-%     \indexOpArity{(-^{-1})} & = & ([\mathsf{unit}],\mathsf{unit})
-%   \end{eqnarray*}
-%   The intended interpretation of these operations is group unit, group
-%   multiplication and group inverse, respectively. When we discuss
-%   equational theories for index expressions in
-%   \autoref{sec:type-equality} below, we will impose the abelian group
-%   laws on these operations.
-
-%   The primitive types for the units of measure example are taken to be
-%   $\PrimTypeSet = \{ \mathsf{num} \}$, with
-%   $\primTyArity(\mathsf{num}) = [\mathsf{unit}]$.
-% \end{example}
-
 We assume a countably infinite collection of index variable names $i,
@@
 
+  {\small
   \textbf{Well-sorted index expressions}
@@
 
-    \inferrule* [right=TyExist]
+    \inferrule* [right=TyEx]
     {\Delta, i \mathord: s \vdash A \isType}
     {\Delta \vdash \exists i \mathord: s. A \isType}
-  \end{mathpar}
+  \end{mathpar}}
   \caption{Index expressions and types}
@@
 quantification ranges over all index expressions of some
-sort. Existential types are constructed using the \TirName{TyExist}
+sort. Existential types are constructed using the \TirName{TyEx}
 rule, and allow for abstraction by hiding. %FIXME: think about the introduction of existentials more, perhaps forward ref to their use
@@
 \begin{example}[Two-Dimensional Geometry]
-  \fixme{Tie this example back to section 2}
-  Translations form an abelian group under addition, so we assume the
-  abelian group axioms for translations:
+  \label{ex:two-dim-geo-axioms}
+  In \autoref{sec:motivating-examples}, we assumed that various
+  equational axioms hold for the indexing expressions standing for
+  elements of geometric groups. We can now make this assumption formal
+  in our general framework. Semantically, translations form an abelian
+  group under addition, so we assume the abelian group axioms for
+  translations:
   \begin{displaymath}
@@
 \end{example}
-
-% \begin{example}[Logical Information Flow]
-%   \fixme{Do this}
-% \end{example}
-
-% \begin{example}[Units of Measure]
-%   For the units of measure system, seen as an instantiation of our
-%   general framework, the set $\IndexAxiomSet$ consists of the abelian
-%   group axioms:
-%   \begin{displaymath}
-%     \begin{array}{l}
-%       u : \mathsf{unit} \vdash u \cdot 1 \stackrel{ax}\equiv u : \mathsf{unit} \\
-%       u_1, u_2, u_3 : \mathsf{unit} \vdash u_1 \cdot (u_2 \cdot u_3) \stackrel{ax}\equiv (u_1 \cdot u_2) \cdot u_3 : \mathsf{unit} \\
-%       u_1, u_2 : \mathsf{unit} \vdash u_1 \cdot u_2 \stackrel{ax}\equiv u_2 \cdot u_1 : \mathsf{unit} \\
-%       u : \mathsf{unit} \vdash u \cdot u^{-1} \stackrel{ax}\equiv 1 : \mathsf{unit} \\
-%     \end{array}
-%   \end{displaymath}
-% \end{example}
 
@@
 \end{displaymath}
-The interpretation of the unit type is a chosen one element set, and
-the interpretations of the function, tuple and sum types are simply
-the corresponding constructions on sets. The interpretations of the
+% The interpretation of the unit type is a chosen one element set, and
+% the interpretations of the function, tuple and sum types are simply
+% the corresponding constructions on sets. 
+The interpretations of the
 universal and existential quantifiers ignore the indexing: the
@@
 
-% \begin{example}[Logical Information Flow]
-%   \fixme{do this}
-% \end{example}
-
 \subsubsection{The Relational Interpretation of Types}
@@
 contexts. We assign interpretations to index contexts in terms of sets
-of \emph{relational environments}, which we now define.
+of \emph{relational environments}.
 
-A relation environment for a context $\Delta$ is a (dependently typed)
+The definition of relational environments is the most subtle part of
+our development. The straightforward approach would be to assume some
+interpretation of the indexing expressions, where sorts are
+interpreted as sets and index operations as functions, and then define
+the relational interpretation of a primtive type
+$\tyPrim{X}{e_1,...,e_n}$ in terms of the interpretations of
+$e_1,...,e_n$. This approach suffices to prove free theorems and type
+isomorphisms for instantiations of our general framework. However, in
+order to prove non-definability results (like, for example, the one in
+\exref{ex:uninhabited-type}), we choose a more subtle interpretation
+of index contexts and primitive types that can account more precisely
+for the definable programs permitted by the algebraic indexing. In
+this section, we set out the general requirements that we require of
+relational environments in order for the abstraction theorem to
+hold. In \autoref{sec:constr-rel-env}, we give a general construction
+of relational environments that will allow us to prove several
+non-definability results.
+
+A relational environment for a context $\Delta$ is a (dependently typed)
 function $\rho$ that, for each primitive type $\tyPrimNm{X}$ and
@@
 
-% FIXME: an example?
-
 We write $\mathrm{RelEnv}(\Delta)$ for the set of all relation
@@
 
-% FIXME: fix the notation here for interpretation of index contexts:
-% should use some sort of semantic brackets
-
 As we stated above, index contexts $\Delta$ are interpreted as certain
@@
 we assume for our system, we may have different sets of relational
-environments that enforce different invariants. For the general
-framework, we assume that, for each index context $\Delta$, we are
-given a set of relation environments $\relEnv{E}(\Delta)$. Note that
-this assignment of sets of relation environments to index contexts is
-not assumed to be compositional: it is not necessarily the case that
-the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
-defined in terms of $\relEnv{E}(\Delta_1)$ and
-$\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
-properties:
+environments that enforce different invariants. We require that the
+sets of relational environments are \emph{valid} in the following
+sense:
+
+\begin{definition}
+  \label{defn:valid-rel-env-family}
+  A family of sets of relational environments $\relEnv{E}(\Delta)
+  \subseteq \mathrm{RelEnv}(\Delta)$ is \emph{valid} if the following
+  two properties are satisfied:
   \begin{enumerate}
@@
     \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
-  \relEnv{E}(\Delta)$, along with a simultaneous substitution $\Delta'
-  \vdash \sigma \Rightarrow \Delta$, such that $\rho_1 \circ
-  \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists a $\rho
-  \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho \circ
-  \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ \pi_{i\mathord:s'} =
-  \rho_2$.
+    \relEnv{E}(\Delta)$, along with a simultaneous substitution
+    $\Delta' \vdash \sigma \Rightarrow \Delta$, such that $\rho_1
+    \circ \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists
+    a $\rho \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho
+    \circ \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ
+    \pi_{i\mathord:s'} = \rho_2$.
     % such that the outer edges of the
@@
     % \mathit{PrimType}$:
-  % %FIXME: check that these compositions are the right way round
-  % %FIXME: re-do this as a set of equations
-  % %FIXME: give a better English description of why this is needed
+    % % FIXME: check that these compositions are the right way round
+    % % FIXME: re-do this as a set of equations
+    % % FIXME: give a better English description of why this is needed
     % \begin{displaymath}
@@
   \end{enumerate}
+\end{definition}
+
+% For the general
+% framework, we assume that, for each index context $\Delta$, we are
+% given a set of relation environments $\relEnv{E}(\Delta)$. Note that
+% this assignment of sets of relation environments to index contexts is
+% not assumed to be compositional: it is not necessarily the case that
+% the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
+% defined in terms of $\relEnv{E}(\Delta_1)$ and
+% $\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
+% properties:
 
@@
 % correctly with respect to application of simultaneous substitution.
-
-% \begin{example}
-%   The construction of relational environments is the most subtle part
-%   of our development. \fixme{complete this part}
-% \end{example}
-
-% \begin{example}[Units of Measure]
-%   We give two examples of assignments of sets of relational
-%   environments to index contexts in the units of measure example. The
-%   second example is a refinment of the first. FIXME: some general
-%   words about relational environments in these cases.
-%   \begin{enumerate}
-%   \item FIXME: assign a positive scaling factor to each variable in an
-%     index context, extend homomorphically to all index expressions,
-%     and then use this to define a scaling relation
-%   \item FIXME: given an index context, assume a subgroup and a
-%     homomorphism into the scaling factors
-%   \end{enumerate}
-% \end{example}
 
@@
 
+\subsubsection{A General Construction of Relational Environments}
+\label{sec:constr-rel-env}
+
+\newcommand{\Free}{\mathrm{Free}}
+\newcommand{\semSort}[1]{\llbracket #1 \rrbracket^{\mathcal{S}}}
+\newcommand{\semIndexExp}[1]{\llbracket #1 \rrbracket^{\mathcal{I}}}
+
+We now give a general construction of valid families of sets of
+relational environments that will allow us to derive useful
+consequences of the abstraction theorem, including free theorem, type
+isomorphisms and non-definability results. We make use of some
+concepts of universal algebra, which we recall here. \fixme{Desperately needs some intuition: do so in terms of clumpiness}
+
+\paragraph{Some Universal Algebra}
+For each sort $s \in \SortSet$ we assume a carrier set
+$\semSort{s}$. We extend this assignment to index contexts and other
+sequences of sorts (e.g., type arities) using the cartesian product of
+sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} = \semSort{s_1}
+\times ... \times \semSort{s_n}$. For each index operation $\texttt{f}
+\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) = ([s_1,...,s_n],s)$,
+we assume a function $\semIndexExp{\texttt{f}} : \semSort{s_1,...,s_n}
+\to \semSort{s}$. For each well-sorted index expression $\Delta \vdash
+e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
+\semSort{s}$ by recursion on the structure of $e$. We further assume
+that for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
+\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$. A
+\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
+\semIndexExp{\cdot}_{\mathcal{M}})$ such that all the axioms in
+$\IndexAxiomSet$ are satisfied.
+
+% A sub-model $\mathcal{N}$ of a given
+% model $\mathcal{M}$ (written $\mathcal{N} \subseteq \mathcal{M}$) is a
+% model such that $\semSort{s}_{\mathcal{N}} \subseteq
+% \semSort{s}_{\mathcal{M}}$ for all $s \in \SortSet$.
+
+\fixme{Define free models for any sort-indexed collection of sets}
+Given an index context $\Delta$, the \emph{free} model on $\Delta$,
+$\Free(\Delta)$, has $\semSort{s}_{\Free(\Delta)} = \{e \sepbar \Delta
+\vdash e : s\}/\mathord\equiv$: the set of expressions of sort $s$ in
+the indexing context $\Delta$, quotiented by the equality
+judgment. The free model has
+$\semIndexExp{\texttt{f}}_{\Free(\Delta)}(e_1,..,e_n) =
+\texttt{f}(e_1,...,e_n)$.
+
+Given a pair of models $\mathcal{M}_1 = (\semSort{\cdot}_1,
+\semIndexExp{\cdot}_1)$ and $\mathcal{M}_2 = (\semSort{\cdot}_2,
+\semIndexExp{\cdot}_2)$, a \emph{homomorphism} $h : \mathcal{M}_1 \to
+\mathcal{M}_2$ is a family of functions $\{ h_s : \semSort{s}_1 \to
+\semSort{s}_2 \}_{s \in \SortSet}$ such that for each $\texttt{f} \in
+\IndexOpSet$, we have $h_s(\semIndexExp{\texttt{f}}_1 (x_1,...,x_n)) =
+\semIndexExp{\texttt{f}}_2(h_{s_1}x_1, ..., h_{s_n}x_n)$.
+
+\paragraph{Relational Environments} \fixme{Say some stuff in english to make sure that the construction is understandable}
+
+To assign a relational meaning to
+each of the primitive types we assume a some model $\mathcal{M}$ of
+the indexing theory along with a function $R_{\tyPrimNm{X}} :
+\semSort{\primTyArity(\tyPrimNm{X})}_{\mathcal{M}} \to
+\Rel(\tyPrimSem{\tyPrimNm{X}})$ for each primitive type
+$\tyPrimNm{X}$, and a constant $R^\bullet_{\tyPrimNm{X}} \in
+\Rel(\tyPrimSem{\tyPrimNm{X}})$. We set
+$\relEnv{E}_{\mathcal{M},R}(\Delta) = \{ \rho^{\mathcal{A},h} \sepbar \mathcal{A}
+\subseteq \Free(\Delta), h : \mathcal{A} \to \mathcal{M} \}$, where:
+\begin{displaymath}
+  \rho^{\mathcal{A},h}\ \tyPrimNm{X}\ (e_1,...,e_n) = \left\{
+    \begin{array}{ll}
+      R^\bullet_{\tyPrimNm{X}} & \textrm{if any }e_i \not\in \mathcal{A} \\
+      R_{\tyPrimNm{X}}(h(e_1),...,h(e_n)) & \textrm{if all }e_i \in \mathcal{A}
+    \end{array}
+  \right.
+\end{displaymath}
+
+\begin{lemma}
+  The family $\relEnv{E}_{\mathcal{M},R}$ of sets of relation
+  environments is valid (c.f.~\defref{defn:valid-rel-env-family}).
+\end{lemma}
+
+% \fixme{Note that we can recover the basic approach by restricting the
+%   sub-algebras to just the whole thing. Does this work in general?
+%   What about arbitrary collections of sub-algebras?}
+
+\begin{example}[Two-dimensional Geometry] We define the ``standard''
+  model of the indexing theory for the two-dimensional geometry
+  example as follows. Each of the sorts are interpreted just as their
+  semantic counterparts:
+  \begin{displaymath}
+    \begin{array}{@{}llll}
+      \semSort{\SynTransl{2}} = \Transl{2}
+      &
+      \semSort{\SynGL{2}} = \GL{2}
+      &
+      \semSort{\SynOrth{2}} = \Orth{2}
+      &
+      \semSort{\SynGL{1}} = \GL{1}
+    \end{array}
+  \end{displaymath}
+  Each of the index operations (e.g.,~the group structure and
+  determinant) are interpreted by the intended semantic
+  operations. These clearly satisfy the axioms we stated in
+  \exref{ex:two-dim-geo-axioms}. The assignment of relations for the
+  primitive types $\tyPrimNm{vec}$ and $\tyPrimNm{real}$ are given by
+  the following definitions:
+  \begin{displaymath}
+    \begin{array}{ll}
+      R^{\bullet}_{\tyPrimNm{vec}} = \{( \vec{0}, \vec{0} ) \}&
+      R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
+      R^{\bullet}_{\tyPrimNm{real}} = \{(0,0)\} &
+      R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
+    \end{array}
+  \end{displaymath}
+  The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
+  are as may be expected given the informal definitions in
+  \autoref{sec:motivating-examples}. We have set
+  $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ to
+  relate $\vec{0}$ and $0$ to themselves to account for the
+  polymorphic zero constants we have assumed.
+\end{example}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Well-typed Programs}
+\subsection{Well-typed Programs and the Abstraction Theorem}
 \label{sec:well-typed-programs}
@@
   \centering
+  {\tiny
   \begin{mathpar}
@@
     {\Delta; \Gamma \vdash M : (\id_\Delta, e)^*A \\ \Delta, i\mathord:s \vdash A \isType}
-    {\Delta; \Gamma \vdash \mathrm{pack}\ [e]\ M : \exists i\mathord:s. A}
+    {\Delta; \Gamma \vdash \langle[e], M\rangle: \exists i\mathord:s. A}
 
@@
       \Delta, i\mathord:s; \pi_{i\mathord:s}^*\Gamma, x : A \vdash N : \pi_{i\mathord:s}^*B}
-    {\Delta; \Gamma \vdash \mathrm{unpack}\ [i]\ x = M\ \mathrm{in}\ N : B}
-  \end{mathpar}
+    {\Delta; \Gamma \vdash \mathrm{let}\langle[i],x\rangle = M\ \mathrm{in}\ N : B}
+  \end{mathpar}}
 
@@
 
-\subsubsection{The Index-Erasure Interpretation of Programs}
-\label{sec:erasure-semantics-programs}
+\paragraph{Index-Erasure Interpretation of Programs}
 
@@
 \begin{displaymath}
-  \begin{array}{@{\hspace{0em}}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l}
-    \tmSem{\Lambda i.\ M}\eta & \isDefinedAs & \tmSem{M}\eta \\
+  \begin{array}{@{}l}
+    \begin{array}{@{}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l@{\hspace{2em}}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l@{}}
+      \tmSem{\Lambda i.\ M}\eta & \isDefinedAs & \tmSem{M}\eta
+      &
       \tmSem{M[e]}\eta & \isDefinedAs & \tmSem{M}\eta \\
-    \tmSem{\mathrm{pack}\ [e]\ M}\eta & \isDefinedAs & \tmSem{M}\eta \\
-    \tmSem{\mathrm{unpack}\ [i]\ x = M\ \mathrm{in}\ N}\eta & \isDefinedAs & \tmSem{N}(\eta, \tmSem{M}\eta)
+      \tmSem{\langle [e], M\rangle}\eta & \isDefinedAs & \tmSem{M}\eta
+    \end{array} \\
+    \tmSem{\mathrm{let}\langle[i], x\rangle = M\ \mathrm{in}\ N}\eta \isDefinedAs \tmSem{N}(\eta, \tmSem{M}\eta)
   \end{array}
@@
 
-\subsubsection{The Abstraction Theorem}
-\label{sec:abstraction-theorem}
+\paragraph{The Abstraction Theorem}
+%\label{sec:abstraction-theorem}
 
-We now state and prove the abstraction theorem for well-typed programs
-in our system. To state and prove this theorem for open programs, we
-must extend the relational interpretation of types to typing
-contexts. The relational interpretation of contexts is defined by the
-following clauses, where we have used the relational lifting of
-cartesian product $-\relTimes-$ defined in
-\autoref{sec:relational-semantics}.
+We now state the abstraction theorem for well-typed programs. To state
+and prove this theorem for open programs, we must extend the
+relational interpretation of types to typing contexts. The relational
+interpretation of contexts is defined by the following clauses, where
+we have used the relational lifting of cartesian product $-\relTimes-$
+defined in \autoref{sec:relational-semantics}.
 \begin{displaymath}

### 2012-07-03 paper/section-instantiations.tex
Additions to the general framework, and fill in some of the examples
- Add a subsection on the construction of relational environments (unfinished)
- Fill in the monoid indexed types section
- Rearrange the example sections a bit

--- 2012-05-22 paper/section-instantiations.tex
@@
-\section{Instantiations of the General Framework}
+\section{Geometric Consequences of Abstraction}
 \label{sec:instantiations}
 
-\subsection{Multiplicative Group Examples}
-\label{sec:multiplicative-examples}
+\fixme{explain the general setup, with a set of operations for each
+  instantiation.}
 
-\subsection{Additive Group Examples}
-\label{sec:additive-examples}
+\begin{lemma}
+  The thing about $\Gamma_\Geom$.
+\end{lemma}
 
-\subsection{Logic Examples}
-\label{sec:logic-examples}
+\subsection{Theorems for Free}
+\label{sec:theorems-for-free}
+
+Consider the type of the area of a triangle function we defined in
+\exref{ex:area-of-triangle-1}:
+\begin{displaymath}
+  \begin{array}{@{}l}
+    \mathrm{area} : \forall B\mathord:\SynGL{2}, t\mathord:\SynTransl{2}.\ \\
+    \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{|\det B|}
+  \end{array}
+\end{displaymath}
+By \thmref{thm:abstraction}, we can derive the following free theorem. For all $B \in \GL{2}$, $\vec{t} \in \Transl{2}$, and $\vec{x}, \vec{y}, \vec{z} \in \mathbb{R}^2$, we have 
+\begin{displaymath}
+  |\det B|(\tmSem{\mathrm{area}}\ \vec{x}\ \vec{y}\ \vec{z}) = \tmSem{\mathrm{area}}\ (B\vec{x} + \vec{t})\ (B\vec{y} + \vec{t})\ (B\vec{z} + \vec{t})
+\end{displaymath}
+Thus, directly from the type of the $\mathrm{area}$ function, we can
+see that its index-erasure semantics is (a) invariant under
+translations, and (b) if the inputs are subjected to a linear
+transformation $B$, the output varies with the absolute value of the
+determinant of $B$.
+
+% Wadler's influential ``Theorems for Free!''  \cite{wadler89theorems}
+% emphasised a particular aspect of Reynolds' theory of relational
+% parametricity: the fact that we can read off theorems about a program
+% simply by looking at the relational interpretation of its type. We now
+% state some free theorems that are derivable in each of the three main
+% examples we introduced in \autoref{sec:motivating-examples}.
+
+\subsection{Isomorphisms for Types Indexed by Abelian Groups}
+\label{sec:types-indexed-abelian-groups}
+
+\begin{enumerate}
+\item Two cases: when we have a total ``division'' operation, and when
+  we don't.
+\item Type isomorphisms, based on Smith Normal Form
+% \item Non-definability results, based on the non-compositional
+%   relational environments
+% \item Alteration of the non-definability results in the light of
+%   addition of a square root operation
+% \item Mention the variation we can obtain by changing some types,
+%   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
+\end{enumerate}
+
+\fixme{Copy the stuff over from the other file}
+
+\subsection{Indefinability}
+
+\fixme{Copy the stuff over from the other file}
+
+%%% Local Variables:
+%%% TeX-master: "paper"
+%%% End:

### 2012-07-03 paper/section-introduction.tex
intro

--- 2012-06-12 paper/section-introduction.tex
@@
 \label{sec:introduction}
+The best way we know of describing the semantics of parametric
+polymorphism is \emph{relational parametricity}, whose central result
+is Reynolds' Abstraction Theorem~\cite{reynolds83types}. Its striking
+consequences include the well-known ``theorems for free'' for
+polymorphic types~\cite{wadler89theorems}, non-inhabitation results, and precise
+correspondences between System F encodings and algebraic datatypes~\cite{PittsAM:parpoe},
+abstract data types, and, most recently, higher-order encodings of
+binder syntax~\cite{syntaxforfree}.
+
+Relational parametricity is in essence a principle of
+\emph{invariance}: the behaviour of polymorphic code is invariant
+under changes to data representation. Invariance results also abound in
+mathematics and physics. The area of a triangle is invariant with
+respect to isometries of the Euclidean plane; the determinant of a
+matrix is invariant under changes of basis; and Newton's laws are the
+same in all inertial frames. Typically, transformations in mathematics
+and physics have interesting structure; for example, translations in
+the Euclidean plane form an Abelian group.
+
+%Types express properties of programs. The statement $\mathit{b}: \tyBool$ tells
+%us that $\mathit{b}$ has value $\mathtt{true}$ or $\mathtt{false}$;
+%the statement $\mathit{pos}:\tyReal\tyProduct\tyReal$ says that
+%$\mathit{pos}$ is a pair whose components are real numbers. More interestingly,
+%the statement $\mathit{f} : \forall X. \tyPrim{list}X\tyArr\tyPrim{list}X$ tells us that
+%if we apply $\mathit{f}$ to a list whose elements satisfy some property $\psi$, 
+%then the result is a list whose elements also satisfy $\psi$.
+
+%Unary predicates only go so far, and for polymorphic types
+%particularly do not express everything that follows from possessing that type. 
+%Imagine a function $\mathit{oddRev}$ that reverses any list whose first element
+%is an odd number, otherwise behaving as the identity. It certainly
+%preserves properties of the elements -- but truly parametric
+%polymorphism should tell us more.
+
+%To obtain stronger properties we must \emph{relate} multiple
+%executions of the same code. The statement $\mathit{f}:\forall
+%X. \tyPrim{list}X\tyArr\tyPrim{list}X$ then says that any change in
+%representation of the list elements supplied to $\mathit{f}$ is
+%reflected in the elements in the result.  Formally, for any function
+%$g$, we have $\mathit{f}(\mathrm{map}\:\mathit{g}\:\mathit{x}) =
+%\mathrm{map}\:\mathit{g}\:(\mathit{f}\:\mathit{x})$.  This is the
+%essence of \emph{relational parametricity} as formulated by
+%Reynolds~\cite{reynolds83types}. Its consequences are striking.  It
+%immediately tells us that $\mathit{oddRev}$ cannot be given a
+%polymorphic type: simply consider the change-of-representation given
+%by $\lambda x. x+1$.  But more profoundly it can be used to prove blah
+%blah (initial algebras, term reps, non-inhabitation, etc.).
+
+Inspired by this connection, we study relational parametricity for
+types that are \emph{indexed} by attributes that themselves have
+algebraic structure.  For example, in computational geometry, points
+in 2-space can be indexed by a co-ordinate frame; in physical
+modelling, quantities are indexed by units-of-measure; in
+information-flow security, computations are indexed by principals.
+Types that are polymorphic over such indices induce invariance
+properties.
+
+A function that computes the area of a triangle might
+have simple type
+\[
+\mathit{areaTri} : \tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal
+\]
+but instead we can assign it a polymorphic type
+\[
+\mathrm{areaTri} : \forall\alpha\mathord:\SynTransl{2}.
+  \tyPrim{vec}{\alpha} \times \tyPrim{vec}{\alpha} \times \tyPrim{vec}{\alpha} \to \tyReal \\
+\]
+in which the index $\alpha$ can be thought of as a `translation frame' that is
+preserved by the vector arguments, from which we can deduce that
+\[
+\mathrm{areaTri}\;(t + v_1, t + v_2, t + v_3) = 
+\mathrm{areaTri}\;(v_1, v_2, v_3)
+\]
+for any translation $t\in\Transl{2}$.
+
+In more complex examples the invariants \emph{compose} in various
+ways; moreover, the compositions satisfy algebraic laws. Consider a function
+that computes the area of a circle given its radius; by ascribing it
+the polymorphic type
+\[
+\mathrm{areaCircle} : \forall\alpha\mathord:\SynScal.\tyPrim{real}{\alpha}\to
+\tyPrim{real}{\alpha\cdot\alpha}
+\]
+we have captured that its area varies as the square of its radius, that is, for any $s>0$,
+\[
+\mathrm{areaCircle}\;(s\cdot r) = s^2\cdot \mathrm{areaCircle}\;r.
+\]
+Here $\alpha$ can be interpreted as the \emph{units} of the argument to
+$\mathrm{areaCircle}$, and `$\cdot$' composes units by forming their
+product. To complete the system, we would add an inverse operation and
+identity unit, together with the algebraic laws of an Abelian group,
+permitting, for example, the identification of
+$\tyPrim{real}{\alpha\cdot\alpha^{-1}}$ with $\tyPrim{real}{1}$.
+
+In his original paper, Reynolds asserted that ``type structure is a
+syntactic discipline for enforcing levels of abstraction''. 
+We see something analogous here: supposing that all primitive operations are
+given types that reflect their behaviour under translation, there is no way
+to `break' this property, for example, for $\mathrm{areaTri}$ to depend
+on the actual coordinates of its inputs. In fact, through types alone we can
+derive so-called \emph{coordinate-free} geometry.
+
+The invariance properties listed above can be seen as `theorems for
+free'~\cite{wadler89theorems} obtained from types alone. We can
+further observe that the abstraction afforded by polymorphic indexed
+types induces interesting type \emph{isomorphisms}.  The type of
+$\mathrm{areaCircle}$ above is in fact isomorphic to
+$\tyPrim{real}{1}$, that is, a dimensionless constant. A moment's
+thought reveals why: what possible unary functions can be constructed whose
+outputs scale as the square of the scaling of their inputs? Answer:
+just those functions of the form~$\lambda x. k x^2$ for some constant~$k$.
+(In this case, of course, we expect that~$k = \pi$.)
 

### 2012-07-04 paper/paper.tex
Major changes
- Refined set of contributions
- "Complete" the motivating examples section
- Fill in some references for the background section
- "Complete" the general framework section
- Sketch the "consequences" section
- Sketch the "discussion" section
Macro-ize the names of the groups with dimensions, so we can change notation easily if need be
Changes to Section 2
- Move the information flow and metric space examples to after the
  general framework (now in section-further-examples.tex)

- Write a new introduction to section 2 to highlight the structure of
  the geometry type system

- Sm

@@
 
-\title{Relational Parametricity for Algebraically Indexed Types}
+\title{Abstraction and Invariance for Algebraically Indexed Types}
 
-\authorinfo{Robert Atkey\and Neil Ghani\\\and Patricia Johann}
+\authorinfo{Robert Atkey\and Patricia Johann}
            {University of Strathclyde}
-           {\{Robert.Atkey,Neil.Ghani,Patricia.Johann\}@cis.strath.ac.uk}
+           {\{Robert.Atkey,Patricia.Johann\}@strath.ac.uk}
 
 \authorinfo{Andrew Kennedy}
-           {Microsoft Research}
-           {}
+           {Microsoft Research Cambridge}
+           {akenn@microsoft.com}
 
@@
 
-\newcommand{\GA}{\mathrm{GA}}
-\newcommand{\GL}{\mathrm{GL}}
-\newcommand{\Is}{\mathrm{Is}}
-\newcommand{\Orth}{\mathrm{O}}
-\newcommand{\SOrth}{\mathrm{SO}}
-\newcommand{\Transl}{\mathrm{T}}
-\newcommand{\SynTransl}{\mathsf{T}}
-\newcommand{\SynGL}{\mathsf{GL}}
-\newcommand{\SynOrth}{\mathsf{O}}
-\newcommand{\Scale}{\mathrm{Scale}}
+\newcommand{\todo}[1]{[\textbf{TODO}: #1]}
+
+\newcommand{\GL}[1]{\mathrm{GL}_#1}
+\newcommand{\SynGL}[1]{\mathsf{GL}_#1}
+\newcommand{\SE}[1]{\mathsf{SE}_#1}
+\newcommand{\SynSE}[1]{\mathsf{SE}_#1}
+\newcommand{\Orth}[1]{\mathrm{O}_#1}
+\newcommand{\SynOrth}[1]{\mathsf{O}_#1}
+\newcommand{\Transl}[1]{\mathrm{T}_#1}
+\newcommand{\SynTransl}[1]{\mathsf{T}_#1}
+\newcommand{\Scal}{\mathrm{Scal}}
+\newcommand{\SynScal}{\mathsf{Scal}}
+
+% \newcommand{\GLtwo}{\mathrm{GL}_2}
+% \newcommand{\SynGLtwo}{\mathsf{GL}_2}
+% \newcommand{\GLone}{\mathrm{GL}_1}
+% \newcommand{\SynGLone}{\mathsf{GL}_1}
 
@@
 \newcommand{\tyUnit}{\texttt{unit}}
-\newcommand{\tyPrim}[2]{\texttt{#1}\langle #2 \rangle}
+\newcommand{\tyReal}{\texttt{real}}
+\newcommand{\tyPrim}[2]{\textup{\texttt{#1}}\langle #2 \rangle}
 \newcommand{\tyPrimNm}[1]{\texttt{#1}}
@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{definition}{Definition}
 \newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
 \newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
+\newcommand{\defref}[1]{\hyperref[#1]{Definition \ref*{#1}}}
 \newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
@@
 \newcommand{\isDefinedAs}{\stackrel{\mathit{def}}=}
+
+\newcommand{\Geom}{\mathit{Geom}}
 
@@
 \begin{abstract}
+  Reynolds' relational parametricity provides a powerful way to reason
+  about programs in terms of invariance under changes of data
+  representation. A dazzling array of applications of Reynolds' theory
+  exists, exploiting invariance to yield ``theorems for free'',
+  uninhabitation results, and encodings of algebraic datatypes.
 
+  Outside computer science, invariance under change is a common theme
+  running through many areas mathematics and physics. For example, the
+  area of a triangle is not altered by rotation or flipping. If we
+  scale a triangle, then we scale its area, maintaining an invariant
+  relationship between the triangle and its area. The transformations
+  that properties are invariant under are often organised into groups,
+  the algebraic structure reflecting the composability and
+  invertibility of transformations.
+
+  In this paper, we investigate programming languages whose types are
+  indexed by algebraic structures such as groups of geometric
+  transformations. Other examples include types indexed by
+  principals--for information flow security--and types indexed by
+  distances--for analysis of analytic continuity properties. Following
+  Reynolds, we prove a general abstraction theorem that covers all
+  these instances. Consequences of the abstraction theorem include
+  free theorems expressing invariance properties of programs, type
+  isomorphisms based on invariance properties, and non-definability
+  results, indicating when certain algebraically indexed types are
+  uninhabited or only inhabited by trivial programs.
 \end{abstract}
@@
 \keywords
-  parametricity, dimension types, computational geometry
+  parametricity, units of measure, dimensional analysis, invariance, computational geometry
 
@@
 \input{section-introduction.tex}
-%\input{section-motivating-examples.tex}
 \input{geometry-twocolumn.tex}
-\input{section-background.tex}
 \input{section-general-framework.tex}
 \input{section-instantiations.tex}
+\input{section-further-examples.tex}
 \input{section-discussion.tex}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{plainnat}
+\bibliographystyle{abbrvnat}
 \bibliography{paper}
 
+\newpage
 \appendix

### 2012-07-04 paper/paper.tex

@@
   relationship between the triangle and its area. The transformations
-  that properties are invariant under are often organised into groups,
-  the algebraic structure reflecting the composability and
-  invertibility of transformations.
+  under which properties are invariant are often organised into
+  groups, with the algebraic structure reflecting the composability
+  and invertibility of transformations.
 

### 2012-07-05 paper/section-general-framework.tex
Better explanation of the construction of relational environments

@@
 order to prove non-definability results (like, for example, the one in
-\exref{ex:uninhabited-type}), we choose a more subtle interpretation
-of index contexts and primitive types that can account more precisely
-for the definable programs permitted by the algebraic indexing. In
-this section, we set out the general requirements that we require of
-relational environments in order for the abstraction theorem to
-hold. In \autoref{sec:constr-rel-env}, we give a general construction
-of relational environments that will allow us to prove several
-non-definability results.
+\exref{ex:uninhabited-type}), we choose a more refined interpretation
+that accounts more precisely for the definable programs permitted by
+the algebraic indexing. In this section, we set out the general
+requirements that we require of relational environments in order for
+the abstraction theorem to hold. In \autoref{sec:constr-rel-env}, we
+give a general construction of relational environments that will allow
+us to prove several non-definability results.
 
@@
 corresponding parts of \lemref{lem:tyeqsubst-erasure}.
-
-\subsubsection{A General Construction of Relational Environments}
-\label{sec:constr-rel-env}
-
-\newcommand{\Free}{\mathrm{Free}}
-\newcommand{\semSort}[1]{\llbracket #1 \rrbracket^{\mathcal{S}}}
-\newcommand{\semIndexExp}[1]{\llbracket #1 \rrbracket^{\mathcal{I}}}
-
-We now give a general construction of valid families of sets of
-relational environments that will allow us to derive useful
-consequences of the abstraction theorem, including free theorem, type
-isomorphisms and non-definability results. We make use of some
-concepts of universal algebra, which we recall here. \fixme{Desperately needs some intuition: do so in terms of clumpiness}
-
-\paragraph{Some Universal Algebra}
-For each sort $s \in \SortSet$ we assume a carrier set
-$\semSort{s}$. We extend this assignment to index contexts and other
-sequences of sorts (e.g., type arities) using the cartesian product of
-sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} = \semSort{s_1}
-\times ... \times \semSort{s_n}$. For each index operation $\texttt{f}
-\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) = ([s_1,...,s_n],s)$,
-we assume a function $\semIndexExp{\texttt{f}} : \semSort{s_1,...,s_n}
-\to \semSort{s}$. For each well-sorted index expression $\Delta \vdash
-e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
-\semSort{s}$ by recursion on the structure of $e$. We further assume
-that for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
-\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$. A
-\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
-\semIndexExp{\cdot}_{\mathcal{M}})$ such that all the axioms in
-$\IndexAxiomSet$ are satisfied.
-
-% A sub-model $\mathcal{N}$ of a given
-% model $\mathcal{M}$ (written $\mathcal{N} \subseteq \mathcal{M}$) is a
-% model such that $\semSort{s}_{\mathcal{N}} \subseteq
-% \semSort{s}_{\mathcal{M}}$ for all $s \in \SortSet$.
-
-\fixme{Define free models for any sort-indexed collection of sets}
-Given an index context $\Delta$, the \emph{free} model on $\Delta$,
-$\Free(\Delta)$, has $\semSort{s}_{\Free(\Delta)} = \{e \sepbar \Delta
-\vdash e : s\}/\mathord\equiv$: the set of expressions of sort $s$ in
-the indexing context $\Delta$, quotiented by the equality
-judgment. The free model has
-$\semIndexExp{\texttt{f}}_{\Free(\Delta)}(e_1,..,e_n) =
-\texttt{f}(e_1,...,e_n)$.
-
-Given a pair of models $\mathcal{M}_1 = (\semSort{\cdot}_1,
-\semIndexExp{\cdot}_1)$ and $\mathcal{M}_2 = (\semSort{\cdot}_2,
-\semIndexExp{\cdot}_2)$, a \emph{homomorphism} $h : \mathcal{M}_1 \to
-\mathcal{M}_2$ is a family of functions $\{ h_s : \semSort{s}_1 \to
-\semSort{s}_2 \}_{s \in \SortSet}$ such that for each $\texttt{f} \in
-\IndexOpSet$, we have $h_s(\semIndexExp{\texttt{f}}_1 (x_1,...,x_n)) =
-\semIndexExp{\texttt{f}}_2(h_{s_1}x_1, ..., h_{s_n}x_n)$.
-
-\paragraph{Relational Environments} \fixme{Say some stuff in english to make sure that the construction is understandable}
-
-To assign a relational meaning to
-each of the primitive types we assume a some model $\mathcal{M}$ of
-the indexing theory along with a function $R_{\tyPrimNm{X}} :
-\semSort{\primTyArity(\tyPrimNm{X})}_{\mathcal{M}} \to
-\Rel(\tyPrimSem{\tyPrimNm{X}})$ for each primitive type
-$\tyPrimNm{X}$, and a constant $R^\bullet_{\tyPrimNm{X}} \in
-\Rel(\tyPrimSem{\tyPrimNm{X}})$. We set
-$\relEnv{E}_{\mathcal{M},R}(\Delta) = \{ \rho^{\mathcal{A},h} \sepbar \mathcal{A}
-\subseteq \Free(\Delta), h : \mathcal{A} \to \mathcal{M} \}$, where:
-\begin{displaymath}
-  \rho^{\mathcal{A},h}\ \tyPrimNm{X}\ (e_1,...,e_n) = \left\{
-    \begin{array}{ll}
-      R^\bullet_{\tyPrimNm{X}} & \textrm{if any }e_i \not\in \mathcal{A} \\
-      R_{\tyPrimNm{X}}(h(e_1),...,h(e_n)) & \textrm{if all }e_i \in \mathcal{A}
-    \end{array}
-  \right.
-\end{displaymath}
-
-\begin{lemma}
-  The family $\relEnv{E}_{\mathcal{M},R}$ of sets of relation
-  environments is valid (c.f.~\defref{defn:valid-rel-env-family}).
-\end{lemma}
-
-% \fixme{Note that we can recover the basic approach by restricting the
-%   sub-algebras to just the whole thing. Does this work in general?
-%   What about arbitrary collections of sub-algebras?}
-
-\begin{example}[Two-dimensional Geometry] We define the ``standard''
-  model of the indexing theory for the two-dimensional geometry
-  example as follows. Each of the sorts are interpreted just as their
-  semantic counterparts:
-  \begin{displaymath}
-    \begin{array}{@{}llll}
-      \semSort{\SynTransl{2}} = \Transl{2}
-      &
-      \semSort{\SynGL{2}} = \GL{2}
-      &
-      \semSort{\SynOrth{2}} = \Orth{2}
-      &
-      \semSort{\SynGL{1}} = \GL{1}
-    \end{array}
-  \end{displaymath}
-  Each of the index operations (e.g.,~the group structure and
-  determinant) are interpreted by the intended semantic
-  operations. These clearly satisfy the axioms we stated in
-  \exref{ex:two-dim-geo-axioms}. The assignment of relations for the
-  primitive types $\tyPrimNm{vec}$ and $\tyPrimNm{real}$ are given by
-  the following definitions:
-  \begin{displaymath}
-    \begin{array}{ll}
-      R^{\bullet}_{\tyPrimNm{vec}} = \{( \vec{0}, \vec{0} ) \}&
-      R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
-      R^{\bullet}_{\tyPrimNm{real}} = \{(0,0)\} &
-      R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
-    \end{array}
-  \end{displaymath}
-  The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
-  are as may be expected given the informal definitions in
-  \autoref{sec:motivating-examples}. We have set
-  $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ to
-  relate $\vec{0}$ and $0$ to themselves to account for the
-  polymorphic zero constants we have assumed.
-\end{example}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{A General Construction of Relational Environments}
+\label{sec:constr-rel-env}
+
+\newcommand{\Free}{\mathrm{Free}}
+\newcommand{\semSort}[1]{\llbracket #1 \rrbracket^{\mathcal{S}}}
+\newcommand{\semIndexExp}[1]{\llbracket #1 \rrbracket^{\mathcal{I}}}
+
+We now give a general construction of valid families of sets of
+relational environments (in the sense of
+\defref{defn:valid-rel-env-family}) that will allow us to derive
+useful consequences of \thmref{thm:abstraction}. In particular, we
+wish to show that certain types are uninhabited. To motivate this,
+consider the following type, which makes use of vectors indexed by
+translations, as in \autoref{sec:affine-vector-ops}:
+\begin{displaymath}
+  \forall t\mathord:\SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}
+\end{displaymath}
+In \exref{ex:uninhabited-type} we claimed that this type is
+uninhabited. Intuitively, it is clear why: given the operations on
+vectors we presented in \autoref{sec:affine-vector-ops}, there is no
+way of cancelling the additional $t$ in the argument type to get the
+result type. More generally, the operations on vectors only allow us
+to combine indexes according to the group operations. Thus, if the
+index of the result type of a function does not lie within the free
+abelian group generated by the indexes of the input types, then there
+is no way of writing a program with the given type. We now develop
+this idea to generate a family of sets of relational environments that
+is valid in the sense of \defref{defn:valid-rel-env-family}. The key
+idea is to assign different relational interpretations to types with
+indexes that lie outside some fintely generated of index expressions.
+
+\paragraph{Some Universal Algebra}
+We make use of some concepts of universal algebra, which now recall.
+
+A \emph{model} for our indexing expressions and their equality is
+defined as follows. For each sort $s \in \SortSet$ we assume a carrier
+set $\semSort{s}$. We extend this assignment to index contexts and
+other sequences of sorts (e.g., type arities) using the cartesian
+product of sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} =
+\semSort{s_1} \times ... \times \semSort{s_n}$. For each index
+operation $\texttt{f} \in \IndexOpSet$ with $\indexOpArity(\texttt{f})
+= ([s_1,...,s_n],s)$, we assume a function $\semIndexExp{\texttt{f}} :
+\semSort{s_1,...,s_n} \to \semSort{s}$. For each well-sorted index
+expression $\Delta \vdash e : s$, we assign a function
+$\semIndexExp{e} : \semSort{\Delta} \to \semSort{s}$ by recursion on
+the structure of $e$. We further assume that for each axiom $\Delta
+\vdash e \stackrel{ax}\equiv e' : s \in \IndexAxiomSet$, we have
+$\semIndexExp{e} = \semIndexExp{e'}$. A \emph{model} is a pair
+$\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
+\semIndexExp{\cdot}_{\mathcal{M}})$ such that all the axioms in
+$\IndexAxiomSet$ are satisfied.
+
+Given a pair of models $\mathcal{M}_1 = (\semSort{\cdot}_1,
+\semIndexExp{\cdot}_1)$ and $\mathcal{M}_2 = (\semSort{\cdot}_2,
+\semIndexExp{\cdot}_2)$, a \emph{homomorphism} $h : \mathcal{M}_1 \to
+\mathcal{M}_2$ is a family of functions $\{ h_s : \semSort{s}_1 \to
+\semSort{s}_2 \}_{s \in \SortSet}$ such that for each $\texttt{f} \in
+\IndexOpSet$, we have $h_s(\semIndexExp{\texttt{f}}_1 (x_1,...,x_n)) =
+\semIndexExp{\texttt{f}}_2(h_{s_1}x_1, ..., h_{s_n}x_n)$.
+
+Let $A = \{ A_s \}_{s \in \SortSet}$ be a family of finite sets
+indexed by sorts. We construct an index context $\Delta_A$ which has
+variables $a\mathord:s$ for each $a \in A_s$. The \emph{free} model
+over $A$, denoted $\Free(A)$, is defined by setting
+$\semSort{s}_{\Free(A)} = \{e \sepbar \Delta_A \vdash e :
+s\}/\mathord\equiv$, where we quotient by the equality judgment
+defined in \autoref{sec:type-equality}. The free model has
+$\semIndexExp{\texttt{f}}_{\Free(A)}(e_1,..,e_n) =
+\texttt{f}(e_1,...,e_n)$.
+
+\paragraph{Relational Environments} We now fix some model
+$\mathcal{M}$ that will provide the underlying interpretation of index
+expressions. As we indicated above, we wish to discriminate between
+index expressions that are generated from some finite set of ``input''
+expressions and those that are not. Accordingly, we assume that for
+each primitive type $\tyPrimNm{X}$ we have two ways of assigning a
+relational meaning to that type. When the index arguments of the type
+are within the generated set, we will use a function $R_{\tyPrimNm{X}}
+\in \semSort{\primTyArity(\tyPrimNm{X})}_{\mathcal{M}} \to
+\Rel(\tyPrimSem{\tyPrimNm{X}})$ that assigns a relational meaning to
+every element in the interpretation of the type's argument sorts. When
+the index arguments are not within the generated set we default to a
+constant relational meaning $R^\bullet_{\tyPrimNm{X}} \in
+\Rel(\tyPrimSem{\tyPrimNm{X}})$.
+
+We now use $R_{\tyPrimNm{X}}$ and $R^\bullet_{\tyPrimNm{X}}$ to
+construct the relational environments that we are interested in.  Fix
+an index context $\Delta$. Let $A = \{ A_s \}_{s \in \SortSet}$ be a
+sort-indexed family of finite sets of expressions such that if $e \in
+A_s$, then $\Delta \vdash e : s$. Let $h : \Free(A) \to \mathcal{M}$
+be a homomorphism. For each $A$ and $h$ we have a relational
+environment for the context $\Delta$:
+\begin{displaymath}
+  \rho^A_h\ \tyPrimNm{X}\ (e_1,...,e_n) = \left\{
+    \begin{array}{@{}l@{\hspace{0.5em}}l}
+      R^\bullet_{\tyPrimNm{X}} & \textrm{if any }e_i \not\in \Free(A) \\
+      R_{\tyPrimNm{X}}(h(e_1),...,h(e_n)) & \textrm{if all }e_i \in \Free(A)
+    \end{array}
+  \right.
+\end{displaymath}
+Collecting all these relational environments together yields a family
+of relational environments, parameterised by our choice of model,
+$\mathcal{M}$, and relational interpretations of primitive types
+$R^\bullet,R$:
+\begin{displaymath}
+  \relEnv{E}_{\mathcal{M},R,R^\bullet}(\Delta) = \{ \rho^A_h \sepbar \forall s.A_s \subseteq \semSort{s}_{\Free(\Delta)}, h : \Free(A) \to \mathcal{M} \}
+\end{displaymath}
+
+\begin{theorem}
+  The family $\relEnv{E}_{\mathcal{M},R,R^\bullet}$ of sets of
+  relation environments is valid in the sense of
+  \defref{defn:valid-rel-env-family}.
+\end{theorem}
+
+Note that we still need to prove for each instantiation of our general
+framework that the program-level primitive operations preserve all the
+relations, in order to fulfil the hypothesis of
+\thmref{thm:abstraction}. \fixme{This is the content of lemma X, lemma
+  Y, lemma Z, below}.
+
 %%% Local Variables:

### 2012-07-05 paper/section-further-examples.tex
Edit the monoid example a bit

@@
 
-In the last section we looked at types indexed by abelian groups. We
-now examine an instantiation of our general framework with a monoid
-type indexed by elements of a commutative monoid. Our main example of
-monoid indexed types is strings indexed by ``taint'' markers: when
-strings are appended their taint markers are combined. Thanks to the
-information in the indexes, we are able to prove an indefinability
-result (\thmref{thm:monoid-indefinability}).
+We have investigated types indexed by abelian group expressions, whose
+elements are taken from some group. We now switch from abelian groups
+to commutative monoids.  Our main example is strings indexed by
+``taint'' markers: when strings are appended their taint markers are
+combined. Thanks to the information in the indexes, we are able to
+prove an indefinability result (\thmref{thm:monoid-indefinability}).
 
-We assume a single indexing sort $\mathsf{M}$, with the monoid
-operations of unit and multiplication along with the commutative
-monoid axioms. There is a single primitive type $\tyPrimNm{M}$, with
-$\primTyArity(\tyPrimNm{M}) = [\mathsf{M}]$. For the index-erasure
-semantics, we assume some (not necessarily commutative) monoid $M$,
-which we also write multiplicatively. Thus, we set
-$\tyPrimSem{\tyPrimNm{M}} = M$. The collection of primitive operations
+We assume a single sort $\mathsf{M}$, with the monoid operations of
+unit and multiplication, and the commutative monoid axioms. There is a
+single primitive type $\tyPrimNm{M}$, with $\primTyArity(\tyPrimNm{M})
+= [\mathsf{M}]$. For the index-erasure semantics, we assume some (not
+necessarily commutative) monoid $M$, which we also write
+multiplicatively. The collection of primitive operations
 $\Gamma_{\mathit{Mon}}$ is simply the two monoid operations:
@@
 
-\newcommand{\mscup}{\stackrel{ms}\cup}
-
 \begin{example}
   By selecting a suitable monoid $M$, we can simulate a programming
-  language that dynamically annotates strings with taint markers. Let
-  $\Sigma$ be a set of characters from some alphabet, and let $T$ be a
-  set of taint markers. Set $M = \Sigma^* \times
+  language that annotates strings with taint markers. Let $\Sigma$ be
+  a set of characters from some alphabet, and let $T$ be a set of
+  taint markers. Set $M = \Sigma^* \times
   \mathcal{MS}_{\mathit{fin}}(T)$, where $\mathcal{MS}_{\mathit{fin}}$
@@
   multiset union as the monoid multiplication, which we write as
-  $\mscup$), $M$ is itself a monoid. The centre of $M$, $Z(M)$,
-  consists of all elements of the form $([], t)$ for multisets of
-  taint markers.
+  $\uplus$), $M$ is itself a monoid. The centre of $M$, $Z(M)$,
+  consists of all elements of the form $([], t)$.
 
@@
   \begin{displaymath}
-    ; \Gamma_\Mon \vdash M : \forall a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{aa}
+    ; \Gamma_\Mon \vdash N : \forall a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{aa}
   \end{displaymath}
   From this type we can derive the following free theorem about
-  $M$. For all $a, b \in \mathcal{MS}_{\mathit{fin}}(T)$, $(s_1,t_1),
+  $N$. For all $a, b \in \mathcal{MS}_{\mathit{fin}}(T)$, $(s_1,t_1),
   (s_2,t_2) \in \Sigma^*$,
@@
     \begin{array}{l}
-      \pi_2(\tmSem{M}\ (s_1,t_1 \mscup a)\ (s_2,t_2 \mscup b)) = \\
-      \hspace{2em}\pi_2(\tmSem{M}\ (s_1,t_1)\ (s_2,t_2)) \mscup a \mscup a
+      \pi_2(\tmSem{N}\ (s_1,t_1 \uplus a)\ (s_2,t_2 \uplus b)) = \\
+      \hspace{2em}\pi_2(\tmSem{N}\ (s_1,t_1)\ (s_2,t_2)) \uplus a \uplus a
     \end{array}
   \end{displaymath}
-  We read this result as follows: if we taint the first input by $a$
-  and the second by $b$, then the taint of the output varies according
-  to $M$'s type. In this case the output contains two copies of $a$
-  and none of $b$.
+  If we taint the first input by $a$ and the second by $b$, then two
+  copies of the taint $a$ will flow to the output, and none of taint
+  $b$.
 \end{example}
@@
 
-\begin{theorem}
-  \begin{equation}
-    \label{eq:fo-monoid-type-insensitive}
-    ; \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
-  \end{equation}
+% \begin{theorem}
+%   \begin{equation}
+%     \label{eq:fo-monoid-type-insensitive}
+%     ; \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
+%   \end{equation}
+%   If all solutions to the matrix equation below have a row of zeros at
+%   position $i$, then $\tmSem{M}$ is insensitive to 
+% \end{theorem}
 
-\end{theorem}
+% as long as $e_1$ is not derivable from the other indexes, and the
+% output one is, then the program is insensitive to its first argument:
+% if $e_1 \not\in \Free(\{e_2,...,e_n\})$ and $e \in
+% \Free(\{e_2,...,e_n\})$, then $M$ is insensitive to its first
+% argument.
+
+% \begin{proof}
+%   Use the relational environment generated by the set $\{e_2, ...,
+%   e_n\}$. Then the relational interpretation of the first argument's
+%   type will be the total relation, and the remainder will be in the
+%   identity relation. So for all $x_1,x_1' \in M$, and $x_2,...,x_n \in
+%   M$ we have $\tmSem{M}\eta_\Mon\ x_1\ x_2...x_n = \tmSem{M}\eta_\Mon\
+%   x'_1\ x_2...x_n$.
+% \end{proof}
+
+% Doesn't apply if we have:
+% \begin{displaymath}
+%   \forall a,b.\ \tyPrim{M}{a} \to \tyPrim{M}{a} \to \tyPrim{M}{b}
+% \end{displaymath}
+% but it ought to work... probably ought to state it in terms of minimal
+% set of generators of the input: if there is a set of $e_i$ that
+% generates the result type's index, then each of the inputs that it
+% doesn't generate are not used by the function. Need the restrict to
+% ones it doesn't generate because of the following:
+% \begin{displaymath}
+%   \forall a.\ \tyPrim{M}{a} \to \tyPrim{M}{a} \to \tyPrim{M}{a}
+% \end{displaymath}
+% we don't know which of the inputs will be used...
+
+% Problem: what about
+% \begin{displaymath}
+%   \forall a.\ \tyPrim{M}{a} \to \tyPrim{M}{a^2} \to \tyPrim{M}{a^2}
+% \end{displaymath}
+% The set $\{a^2\}$ generates the output, and doesn't generate $a$, but
+% a program of this type could easily use either of the inputs...
+
+% Relational environment is:
+% \begin{displaymath}
+%   \rho\ \tyPrimNm{M}\ (e) = \left\{
+%     \begin{array}{ll}
+%       \id_M & \textrm{if }e = a^2 \\
+%       M \times M & \textrm{otherwise}
+%     \end{array}
+%   \right.
+% \end{displaymath}
+
+% In general, we are just setting $R^\bullet_{\tyPrimNm{M}} = M \times
+% M$. However, this is not supported by the monoid multiplication
+% operation: if the two inputs are not in generated set, the output
+% still might be, and is expected to be equal... So we need some notion
+% of \emph{not} being generated by the set... free structures that avoid
+% some argument?
+
+% If we restrict to just variables, then we might be OK... otherwise,
+% I'm stuck.
+
+% The multiplication operation does not satisfy this spec. The
+% generating set needs to be closed under decomposition....
+
+% so we need to say that, somehow, the redundant arguments' indexes
+% never contribute anything to the result type's index.
+
+% Maybe: if an $e_i$ is not present in any of the minimal sets of
+% generators that produces the output type, then it is redundant.
+
+% Let $S = \{ G \subseteq \{e_1,...,e_n\} \sepbar e \in \Free(G) \}$.
+
+% $S' = $ minimal elements of $S$: all the elements of $S$ that do not
+% have anything beneath them in the subset ordering. Then if some $e_j$
+% is not present in any of these,
 
@@
   natural numbers $A$ and $m$-by-$1$ matricies of natural numbers $b$,
-  there exists a program $M$ such that
-  \begin{displaymath}
-    \label{eq:fo-monoid-type}
+  there exists a program:
+  \begin{equation}
     ; \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
-  \end{displaymath}
-  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e =
-  i_1^{b_1}\dots i_m^{b_m}$, if and only if there is a solution
-  $\vec{x} = (x_1,\cdots,x_n)$ in the natural numbers to the equation
-  $A x^\top = b$.
+  \end{equation}
+  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e = i_1^{b_1}\dots
+  i_m^{b_m}$, iff there is a solution $\vec{x} = (x_1,...,x_n)$ in the
+  natural numbers to $A x^\top = b$.
 \end{theorem}

### 2012-07-06 paper/paper.tex
Metric space example, and splitting of sections into files
- fill out the metric space example, and correct 'continuity' to be 'uniform continuity'
  - still need to flesh out the connection with differential privacy
- split individual examples into separate files
- move the abelian group proofs into the appendix, pending fixing up
- tiny changes to the singleton types and logical types sections

@@
 \newcommand{\relSum}{\mathrel{\widehat+}}
+\newcommand{\Eq}{\mathrm{Eq}}
 
@@
   principals--for information flow security--and types indexed by
-  distances--for analysis of analytic continuity properties. Following
-  Reynolds, we prove a general abstraction theorem that covers all
-  these instances. Consequences of the abstraction theorem include
-  free theorems expressing invariance properties of programs, type
-  isomorphisms based on invariance properties, and non-definability
-  results, indicating when certain algebraically indexed types are
-  uninhabited or only inhabited by trivial programs.
+  distances--for analysis of analytic uniform continuity
+  properties. Following Reynolds, we prove a general abstraction
+  theorem that covers all these instances. Consequences of the
+  abstraction theorem include free theorems expressing invariance
+  properties of programs, type isomorphisms based on invariance
+  properties, and non-definability results, indicating when certain
+  algebraically indexed types are uninhabited or only inhabited by
+  trivial programs.
 \end{abstract}
@@
 \input{section-further-examples.tex}
+\input{section-monoid.tex}
+\input{section-singleton.tex}
+\input{section-metric.tex}
 \input{section-discussion.tex}

### 2012-07-06 paper/section-monoid.tex
Metric space example, and splitting of sections into files
- fill out the metric space example, and correct 'continuity' to be 'uniform continuity'
  - still need to flesh out the connection with differential privacy
- split individual examples into separate files
- move the abelian group proofs into the appendix, pending fixing up
- tiny changes to the singleton types and logical types sections

--- 2012-07-05 paper/section-further-examples.tex
@@
 
-We have investigated types indexed by abelian group expressions, whose
-elements are taken from some group. We now switch from abelian groups
-to commutative monoids.  Our main example is strings indexed by
+Above, we investigated types indexed by abelian group expressions,
+whose elements are taken from some group. We now switch from abelian
+groups to commutative monoids.  Our main example is strings indexed by
 ``taint'' markers: when strings are appended their taint markers are
@@
 
-\section{Metric Spaces and Simple Continuity Analysis}
-\label{sec:continuity-analysis}
-
-Our final motivating example of algebraically indexed types and
-representation independence makes use of the metric space structure of
-the real numbers. In the geometry example, we used relational
-interpretations that related values via geometrical
-transformations. An alternative relational interpretation of the real
-numbers is to relate values that are close, in the metric space
-sense. By indexing types by a measure of closeness, we can actually
-state continuity of functions as a typing problem.
-
-We introduce an indexing sort $\mathsf{R}^{>0}$, representing positive
-distances between real numbers. Our type of real numbers is indexed by
-expressions of sort $\mathsf{R}^{>0}$, and we use the following
-index-erasure and relational interpretations:
-\begin{displaymath}
-  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\tyPrim{real}{e}} & \mathbb{R} \\
-    \rsem{\tyPrim{real}{e}}{}\rho & \{ (x,x') \sepbar |x-x'| < \sem{e}\rho \}
-  \end{array}
-\end{displaymath}
-Thus, two real numbers are related if they only differ by the positive
-real number assigned to $e$. With this interpretation, and a
-straightforward interpretation of existential types, we can state the
-standard $\epsilon$-$\delta$ definition of continuity as a type:
-\begin{displaymath}
-  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
-\end{displaymath}
-Chaudhuri, Gulwani and Lublinerman \cite{chaudhuri10continuity} have
-presented a program logic based approach to verifying the continuity
-of programs. Algebraically indexed types give us a way of expressing
-and verifying continuity properties of functions in a type based way.
-
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    \underline{c} & \forall \epsilon\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon} \\
-    (+) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
-    (-) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
-    (*) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1\epsilon_2}
-  \end{array}
-\end{displaymath}
-
-\begin{enumerate}
-\item State the operations and equations (rational powers, addition
-  and multiplication (remember: positive reals))
-\item Derive subtyping by addition of $0$
-\item Do the probability monad as a special case, with a specific
-  relational interpretation that relates things according to the
-  differential privacy notion. Need to think about the
-  measure-theoretic problems...
-\item Work out the details of the continuity example, especially the
-  implementation of an ADT for continuous functions that can handle
-  polynomials.
-\end{enumerate}
-
-\section{Singleton Types}
-\label{sec:singleton-types}
-
-\newcommand{\Sing}{\mathit{Sing}}
-
-Singleton types differ from the ``variation'' types that we have
-considered so far because they nail down exactly the underlying value
-instead of describing how it varies.
-
-Indexing theory contains all the integers as constants, along with
-$+$, $-$ and $*$. Normal arithmetic laws for these
-operations. Program-level operations:
-\begin{displaymath}
-  \begin{array}{@{}c@{\hspace{0.5em}:\hspace{0.5em}}l}
-    \underline{z} & \tyPrim{int}{z} \\ % FIXME: only need zero and one
-    (+) & \forall z_1,z_2\mathord:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_1 + z_2} \\
-    (-) & \forall z_1,z_2\mathord:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_1 - z_2} \\
-    (*) & \forall z_1,z_2\mathord:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_1 * z_2}
-  \end{array}
-\end{displaymath}
-
-\begin{enumerate}
-\item Given an algebraic theory, and a model, we can define a
-  singleton type system.
-\item Need to carefully define this in general, and give the
-  appropriate relational environments
-\item Give the integers (freely generated monoid) as an example
-\end{enumerate}
-
-\begin{theorem}
-  There exists a program
-  \begin{displaymath}
-    -; \Gamma_\Sing \vdash M : \forall i_1,...,i_m\mathord:\mathsf{S}.\ \tyPrim{S}{e_1} \to ... \to \tyPrim{S}{e_n} \to \tyPrim{S}{e}
-  \end{displaymath}
-  if and only if there exists an index expression
-  $i'_1\mathord:\mathsf{S}, ..., i'_n\mathord:\mathsf{S} \vdash e' :
-  \mathsf{S}$ such that $(e_1,...,e_n)^*e' \equiv e$.
-\end{theorem}
-
-\begin{proof}
-  (If) Use the index expression $e'$ to define $M$, replacing each
-  index operation with its program-level counterpart.
-
-  (Only if) \fixme{use a special relational environment...}
-\end{proof}
-
-% \subsection{Predicate Types}
-% \label{sec:predicate-types}
-
-% \begin{enumerate}
-% \item Let the indexing theory be predicates on the underlying value
-% \item With the laws of boolean algebra
-% \item Singleton types are a special case (i.e. with equality to a fixed value)
-% \item But also have $\land$, $\lor$, $\lnot$ (and derived $\Rightarrow$)
-% \item A pair of things is related if they both satisfy the predicate
-%   (and are equal?)
-% \item Need a special type for comparisons?
-%   \begin{displaymath}
-%     (<) : \forall z_1,z_2:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to (\tyPrim{int}{< z_1} \times 
-%   \end{displaymath}
-% \end{enumerate}
-
-\section{Logical Information Flow}
-\label{sec:information-flow}
-
-\begin{enumerate}
-\item This is now a special case of the preceeding subsection
-\item Also makes use of the relational aspect, the previous stuff
-  could make do with unary logical predicates as the interpretations
-  of types, but the information flow properties make use of the
-  relational interpretation.
-\end{enumerate}
-
-\fixme{
-  \begin{enumerate}
-  \item Make the definition of control of information flow clearer
-  \item Make the example a bit more compelling
-  \item Put in some examples of composite principals formed from
-    boolean logic, and the use of boolean algebra to do reasoning in
-    the types. List the laws of boolean algebra, and the laws of
-    Heyting algebra. Need to explain the difference between 
-  \item Possibly use singleton types to be able to do more interesting
-    security policies?
-  \item Put in the formal statement of the information flow property
-    and the proof.
-  \end{enumerate}
-}
-
-An extreme example of invariance under change of representation is the
-tracking of information flow through programs. If a program does not
-depend upon a particular piece of information, then it will be
-invariant under \emph{all} changes of representation of this
-information. As thoroughly described by Sabelfeld and Sands
-\cite{sabelfeld01per}, information flow can be captured semantically
-through the use of partial equivalence relations (PERs). Interpreting
-types as PERs forms an instance of Reynolds' approach to abstraction
-and representation independence. Abadi, Banerjee, Heintze and Riecke
-\cite{abadi99core} built a \emph{Core Calculus for Dependency}, using
-a type system based around a security level indexed monad $T_lA$. Tse
-and Zdancewic \cite{tse04translating} showed that it is possible to
-translate Abadi et al.'s calculus into System F, translating the
-monadic type $T_lA$ to $\alpha_l \to A$ for some free type variable
-$\alpha_l$, and making direct use of Reynolds' abstraction theorem to
-prove information flow properties. We now show how a refined variant
-of Tse and Zdancewic's translation can be expressed via
-algebraically-indexed types.
-
-We assume a single indexing sort $\mathsf{principal}$, intended to
-represent (possibly composite) principals in a system. Principals are
-combined using the connectives of boolean logic ($\top$, $\bot$,
-$\lor$, $\land$, $\lnot$). For the equational theory of principals, we
-assume all the axioms of boolean algebra, and semantically we
-interpret composite principals as boolean values: either true ($\top$)
-or false ($\bot$). We introduce a single primitive type, indexed by
-expressions of sort $\mathsf{principal}$, $\tyPrim{T}{e}$, with the
-following interpretations:
-\begin{displaymath}
-  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\tyPrim{T}{e}} & \{*\} \\
-    \rsem{\tyPrim{T}{e}}{}\rho & \left\{
-      \begin{array}{ll}
-        \{(*,*)\} & \textrm{if }\sem{e}\rho = \top \\
-        \{\}      & \textrm{if }\sem{e}\rho = \bot
-      \end{array}
-      \right.
-  \end{array}
-\end{displaymath}
-The relational interpretation of $\tyPrim{T}{e}$ forces this type to
-only be inhabited when the boolean expression $e$ evaluates to true
-under the environment $\rho$. To write programs, we assume the
-following pair of primitive operations:
-\begin{eqnarray*}
-  \mathrm{proj} & : & \forall p, q\mathord:\mathsf{principal}.\ \tyPrim{T}{p \land q} \to \tyPrim{T}{p} \\
-  \mathrm{combine} & : & \forall p, q\mathord:\mathsf{principal}.\ \tyPrim{T}{p} \to \tyPrim{T}{q} \to \tyPrim{T}{p \land q}
-\end{eqnarray*}
-We can now adapt Tse and Zdancewic's translation of Abadi et al.'s
-monadic type to our setting. We define a type synonym, for each type
-$A$ and expression $e$ of sort $\mathsf{principal}$:
-\begin{displaymath}
-  T_eA = \tyPrim{T}{e} \to A
-\end{displaymath}
-For every principal $e$, we can endow the types $T_e-$ with the
-structure of a monad. This is due to the fact that it is an instance
-of the ``environment'' (or ``reader'') monad \cite{jones95functional}.
-
-\begin{example}
-  Suppose we have a program with the following type:
-  \begin{displaymath}
-    \forall p,q\mathord:\mathsf{principal}. \tyPrim{T}{e} \to T_p(\tyPrimNm{unit} + \tyPrimNm{unit}) \to T_q(\tyPrimNm{unit} + \tyPrimNm{unit})
-  \end{displaymath}
-  where $e$ is some boolean algebra expression involving $p$ and
-  $q$. We will show in \autoref{sec:instantiations} that programs with
-  this type are non-constant (i.e.~depend on their input) if and only
-  if the formula $e \Rightarrow q \Rightarrow p$ is valid in boolean
-  logic.
-\end{example}
-
-\begin{enumerate}
-\item Treat the information flow thing as a type isomorphism, in the
-  same way that Tse and Zdancewic do
-  \begin{displaymath}
-    \begin{array}{l}
-    \forall \vec{p}\mathord:\mathsf{principal}.\ T_{e_1}\tyPrimNm{bool} \to T_{e_2}\tyPrimNm{bool}\\
-    \hspace{4em}\cong \\
-    \left\{
-      \begin{array}{ll}
-        \tyPrimNm{bool} \tyArr \tyPrimNm{bool} & \textrm{if }e_2 \vdash e_1 \\
-        \tyPrimNm{bool} & \textrm{if }e_2 \not\vdash e_1
-      \end{array}
-    \right.
-  \end{array}
-  \end{displaymath}
-  where $\tyPrimNm{bool} = \tyUnit + \tyUnit$.
-\item mention the issue with bogus tokens if we allow for a
-  call-by-name style of non-termination
-\item if we axiomatise intuitionistic logic instead of
-  boolean/classical logic, we get a different type isomorphism.
-\item The isomorphism relies upon $e_1 \vdash e_2 \Leftrightarrow e_1
-  \lor e_2 = e_1$.
-\item Relevant logic? Look at the Wikipedia page to see how to do
-  residuated (commutative) monoids on lattices with just
-  equations. With this, we will be able to do substructural logics
-  too. As long as we know that a particular axiomatisation is sound
-  and complete for some reasoning system, then we can state and prove
-  the non-interference isomorphism above.
-\item Compare to Tse and Zdancewic's addition of axioms to the logic.
-\item Internalising equality? The logical structure can then be used
-  on types as well as everything else. Need the special rules for
-  equality... (see the Fomega paper).
-\end{enumerate}
-
-% Go a step further by considering equivalence relations on word32, and
-% their lattice structure. Link this to quantative information
-% flow. Will also need to include singleton types.
-
-\subsection{Abelian Group Indexed Types}
-\label{sec:abelian-group-indexed-types}
-
-\newcommand{\Grp}{\mathit{Grp}}
-
-Primitive operations $\Gamma_\Grp$:
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    1 & \tyPrim{G}{1} \\
-    (\cdot) & \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to \tyPrim{G}{a b} \\
-    \ ^{-1} & \forall a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}
-  \end{array}
-\end{displaymath}
-
-For the index-erasure interpretation, we assume a group $(G, 1, \cdot,
-\ ^{-1})$. Let $Z(G)$ be the centre of the $G$: i.e.,~the set of
-elements that commute with every other element in $G$. We set
-$\tyPrimSem{G} = G$. The primitive operations in $\Gamma_\Grp$ are
-interpreted just as the unit, multiplication and inverse operations
-of $G$. For each index context $\Delta$, the set
-$\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be
-$\{ \rho_{(H,h)} \sepbar H\textrm{ is a subgroup of }(\Delta
-\Rightarrow \mathsf{G})/\equiv, h\textrm{ is a group homomorphism }H
-\to Z(G) \}$, where
-\begin{displaymath}
-  \rho_{(H,h)}\ \tyPrimNm{G}\ (e) = \left\{
-    \begin{array}{ll}
-      \{ \} & \textrm{if}\ e\not\in H \\
-      \{ (x, h(e)\cdot x) \sepbar x \in G \} & \textrm{if}\ e\in H
-    \end{array}
-  \right.
-\end{displaymath}
-
-\begin{lemma}
-  For all $\Delta$ and $\rho \in \relEnv{E}_\Grp(\Delta)$,
-  $(\eta_\Grp,\eta_\Grp) \in \rsem{\Gamma_\Grp}{\relEnv{E}}\rho$.
-\end{lemma}
-
-\begin{theorem}
-  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
-  natural numbers $A$ and $m$-by-$1$ matricies of integers $b$,
-  there exists a program $M$ such that
-  \begin{equation}
-    \label{eq:fo-group-type}
-    \Gamma_\Grp \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
-  \end{equation}
-  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e = i_1^{b_1}\dots
-  i_m^{b_m}$, if and only if there is a solution $\vec{x} =
-  (x_1,\cdots,x_n)$ in the integers to the equation $A x^\top = b$.
-\end{theorem}
-
-\fixme{Define shorthand notation $x^z$}
-
-\begin{proof}
-  (If) The program $\Lambda i_1\dots i_m.\ \lambda g_1\dots g_n.\
-  g_1^{x_1}\dots g_n^{x_n}$ satisfies the typing judgment
-  (\ref{eq:fo-group-type}).
-
-  (Only if) Assume that there is a program $M$ satisfying the typing
-  judgment (\ref{eq:fo-group-type}). By \thmref{thm:abstraction} we
-  know that for all relational environments $\rho \in
-  \relEnv{E}_\Grp(i_1,\dots,i_m)$, and all $g_j,g'_j \in G$,
-  \begin{displaymath}
-    \begin{array}{l}
-      (g_1,g'_1) \in \rho\ \tyPrimNm{G}\ (e_1) \land \dots \land (g_n,g'_n) \in \rho\ \tyPrimNm{G}\ (e_n) \Rightarrow \\
-      \quad (\tmSem{M}g_1\dots g_n, \tmSem{M}g'_1\dots g'_n) \in \rho\ \tyPrimNm{G}\ (e)
-    \end{array}
-  \end{displaymath}
-  We select the relational environments
-  \begin{displaymath}
-    \rho\ \tyPrimNm{G}\ (e) = \left\{
-      \begin{array}{ll}
-        \{(g,g) \sepbar g \in \mathcal{G} \} &
-        \begin{array}[t]{@{}l}
-          \textrm{if}\ e = e_1^{x_1}\dots e_n^{x_n} \\
-          \textrm{ for some }x_1,\dots,x_n \in \mathbb{Z}
-        \end{array}
-        \\
-        \{\} & \textrm{otherwise}
-      \end{array}
-    \right.
-  \end{displaymath}
-  Now, for any $g_1,\dots,g_m \in \mathcal{G}$, we know that, for all
-  $j$, $(g_j,g_j) \in \rho\ \tyPrimNm{G}\ (e_j)$, so we know that
-  $(\tySem{M}g_1\dots g_n, \tySem{M}g_1 \dots g_n) \in \rho\
-  \tyPrimNm{G}\ (e)$. Thus there exist integers $x_1,...,x_n$ such
-  that $e = e_1^{x_1}\dots e_n^{x_n}$. But we also know that $e =
-  i_1^{b_1}...i_m^{b_m}$. By the cancellation property of free groups,
-  we learn that $x_1,...,x_n$ is the solution we need.
-\end{proof}
-
-\begin{enumerate}
-\item Can add any other operations that distribute over the group
-  multiplication.
-\item What about partial division operations? Make division just
-  return $0$.
-\item Adding square root as an operation. Need to check that all the
-  other operations preserve the new relational environments. Square
-  root is also supported by the plain relational environments, so the
-  type isomorphism result still holds.
-\end{enumerate}
-
-\subsubsection{Type isomorphisms}
-\label{sec:abelian-group-type-isos}
-
-\begin{enumerate}
-\item This only needs the simpler relational environments, but needs
-  the group used for the index-erasure semantics to be abelian.
-\item One direction is easy, but the other requires the use of the
-  free theorem for the type.
-\item There are two version: one with some fixed ground type as the
-  result, and one with the same type as the result, but with the same
-  parameter.
-\item Integrates into the Smith Normal Form thing too, since
-  invertible integer-valued matrices can be turned into type
-  isomorphisms.
-\item Partiality?
-\end{enumerate}
-
-\fixme{Define what a type isomorphism is}
-
-\begin{theorem}
-  For all natural numbers $n$, the types
-  \begin{displaymath}
-    \tau_{n} \isDefinedAs \forall a\mathord:\mathsf{G}.\ \underbrace{\tyPrim{G}{a} \to ... \to \tyPrim{G}{a}}_{n+1\textrm{ times}} \to \tyPrim{G}{a}
-  \end{displaymath}
-  and
-  \begin{displaymath}
-    \sigma_{n} \isDefinedAs \underbrace{\tyPrim{G}{1} \to ... \tyPrim{G}{1}}_{n\textrm{ times}} \to \tyPrim{G}{1}
-  \end{displaymath}
-  are isomorphic.
-\end{theorem}
-
-\begin{proof}
-  We define programs $\Gamma_\Grp \vdash M : \tau_n \to \sigma_n$ and
-  $\Gamma_\Grp \vdash M^{-1} : \sigma_n \to \tau_n$ as follows:
-  \begin{displaymath}
-    M = \lambda f. \lambda g_1 \dots g_n.\ f\ [1]\ 1\ g_1\ \dots\ g_n
-  \end{displaymath}
-  and
-  \begin{displaymath}
-    M^{-1} = \lambda f. \Lambda a.\ \lambda g_0 \dots g_n.\ (f\ (g_1 g^{-1}_0)\ \dots\ (g_n g^{-1}_0)) g_0
-  \end{displaymath}
-  In one direction, the proof that these are mutually inverse is
-  straightfoward, using only the fact that $G$ is a group. For any
-  program $\Gamma_\Grp \vdash f : \sigma_n$, we have
-  \begin{displaymath}
-    \begin{array}{cl}
-        & \tmSem{M\ (M^{-1}\ f)}\eta_\Grp \\
-      = & (\tmSem{M}\eta_{\Grp})\ (\lambda g_0 \dots g_n.\ (\tmSem{f}\eta_\Grp\ (g_1 g^{-1}_0)\ \dots\ (g_n g^{-1}_0)) g_0) \\
-      = & \lambda g_1\dots g_n.\ \tmSem{f}\eta_\Grp\ (g_1 1^{-1})\ \dots\ (g_n 1^{-1})) 1) \\
-      = & \lambda g_1\dots g_n.\ \tmSem{f}\eta_\Grp\ g_1\ \dots\ g_n \\
-      = & \tmSem{f}\eta_\Grp
-    \end{array}
-  \end{displaymath}
-  as required. In the other direction, we need to use the abstraction
-  theorem. For any program $\Gamma_\Grp \vdash f : \tau_n$, we have
-  \begin{displaymath}
-    \begin{array}{cl}
-        & \tmSem{M^{-1}\ (M\ f)}\eta_\Grp \\
-      = & (\tmSem{M^{-1}}\eta_\Grp)\ (\lambda g_1\dots g_n.\ (\tmSem{f}\eta_\Grp\ 1\ g_1\ \dots\ g_n)) \\
-      = & \lambda g_0\dots g_n.\ (\tmSem{f}\eta_\Grp\ 1\ (g_1g^{-1}_0)\ \dots\ (g_ng^{-1}_0))g_0
-    \end{array}
-  \end{displaymath}
-  By \thmref{thm:abstraction}, we know that the following holds of
-  $\tmSem{f}$:
-  \begin{displaymath}
-    \forall a.\ \forall g_0 \dots g_n.\ \tmSem{f}\eta_\Grp\ (g_0a)\ \dots\ (g_na) = (\tmSem{f}\eta_\Grp\ g_0\ \dots\ g_n)a
-  \end{displaymath}
-  As a special case, we set $a = g^{-1}_0$, and learn that
-  \begin{displaymath}
-    \forall g_0 \dots g_n.\ \tmSem{f}\eta_\Grp\ 1\ (g_1g^{-1}_0) \dots\ (g_ng^{-1}_0) = (\tmSem{f}\eta_\Grp\ g_0\ \dots\ g_n)g^{-1}_0.
-  \end{displaymath}
-  By functional extensionality, we are done.
-\end{proof}
-
 %%% Local Variables:

### 2012-07-06 paper/section-info-flow.tex
Rename the logical information flow file

--- 2012-06-22 paper/section-further-examples.tex
@@
-\section{More Algebraically Indexed Types}
-\label{sec:further-examples}
-
-\subsection{Singleton Types}
-\label{sec:singleton-types}
-
-Singleton types differ from the ``variation'' types that we have
-considered so far because they nail down exactly the underlying value
-instead of describing how it varies.
-
-Indexing theory contains all the integers as constants, along with
-$+$, $-$ and $*$. Normal arithmetic laws for these
-operations. Program-level operations:
-\begin{displaymath}
-  \begin{array}{@{}c@{\hspace{0.5em}:\hspace{0.5em}}l}
-    \underline{z} & \tyPrim{int}{z} \\ % FIXME: only need zero and one
-    (+) & \forall z_1,z_2\mathord:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_1 + z_2} \\
-    (-) & \forall z_1,z_2\mathord:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_1 - z_2} \\
-    (*) & \forall z_1,z_2\mathord:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_1 * z_2}
-  \end{array}
-\end{displaymath}
-
-\begin{enumerate}
-\item Take integers as a useful example
-\item The relational interpretation is simply that two things are
-  related if they are both equal to the indexing value (and each
-  other).
-\item Types are extremely precise... Can this be quantified?
-\item Indefinability result by considering sub-algebras? Is this a
-  general method? Can't write:
-  \begin{displaymath}
-    \forall z_1,z_2,z_3.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_3}
-  \end{displaymath}
-  because there is no way of getting $z_3$ from $z_1$ and $z_2$. Can
-  prove this using the relational environments. Is there a nice
-  matrix-style way of characterising the allowable things? I guess it
-  is harder when there are two operations lying around (multiplication
-  and addition here). Maybe we could restrict to multiplication by
-  constants? Then we could possibly characterise the definable
-  elements by a matrix representation?
-\end{enumerate}
-
-\subsection{Predicate Types}
-\label{sec:predicate-types}
-
-\begin{enumerate}
-\item Let the indexing theory be predicates on the underlying value
-\item With the laws of boolean algebra
-\item Singleton types are a special case (i.e. with equality to a fixed value)
-\item But also have $\land$, $\lor$, $\lnot$ (and derived $\Rightarrow$)
-\item A pair of things is related if they both satisfy the predicate
-  (and are equal?)
-\item Need a special type for comparisons?
-  \begin{displaymath}
-    (<) : \forall z_1,z_2:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to (\tyPrim{int}{< z_1} \times 
-  \end{displaymath}
-\end{enumerate}
-
-\subsection{Logical Information Flow}
+\section{Logical Information Flow}
 \label{sec:information-flow}
@@
 \begin{enumerate}
-\item This is now a special case of the preceeding subsection
+\item Make sure the logic is at least a meet semi-lattice
+\item 
+\end{enumerate}
+
+\begin{enumerate}
 \item Also makes use of the relational aspect, the previous stuff
@@
   relational interpretation.
+\item Can also obtain a type non-inhabitation result based on logical
+  derivability
 \end{enumerate}
@@
 under the environment $\rho$. To write programs, we assume the
-following pair of primitive operations:
-\begin{eqnarray*}
-  \mathrm{proj} & : & \forall p, q\mathord:\mathsf{principal}.\ \tyPrim{T}{p \land q} \to \tyPrim{T}{p} \\
-  \mathrm{combine} & : & \forall p, q\mathord:\mathsf{principal}.\ \tyPrim{T}{p} \to \tyPrim{T}{q} \to \tyPrim{T}{p \land q}
-\end{eqnarray*}
+following primitive operations:
+\begin{displaymath}
+  \begin{array}{l}
+  \begin{array}{@{}c@{\hspace{2em}}c}
+    \mathrm{truth} : \tyPrim{T}{\top} &
+    \mathrm{and}   : \forall p, q\mathord:\mathsf{prn}.\ \tyPrim{T}{p} \to \tyPrim{T}{q} \to \tyPrim{T}{p \land q}
+  \end{array} \\
+  \mathrm{coerce} : \forall p, q\mathord:\mathsf{prn}.\ \tyPrim{T}{p \land q} \to \tyPrim{T}{p}
+\end{array}
+\end{displaymath}
 We can now adapt Tse and Zdancewic's translation of Abadi et al.'s
@@
 
-\subsection{Metric Spaces and Simple Continuity Analysis}
-\label{sec:continuity-analysis}
-
-Our final motivating example of algebraically indexed types and
-representation independence makes use of the metric space structure of
-the real numbers. In the geometry example, we used relational
-interpretations that related values via geometrical
-transformations. An alternative relational interpretation of the real
-numbers is to relate values that are close, in the metric space
-sense. By indexing types by a measure of closeness, we can actually
-state continuity of functions as a typing problem.
-
-We introduce an indexing sort $\mathsf{R}^{>0}$, representing positive
-distances between real numbers. Our type of real numbers is indexed by
-expressions of sort $\mathsf{R}^{>0}$, and we use the following
-index-erasure and relational interpretations:
-\begin{displaymath}
-  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\tyPrim{real}{e}} & \mathbb{R} \\
-    \rsem{\tyPrim{real}{e}}{}\rho & \{ (x,x') \sepbar |x-x'| < \sem{e}\rho \}
-  \end{array}
-\end{displaymath}
-Thus, two real numbers are related if they only differ by the positive
-real number assigned to $e$. With this interpretation, and a
-straightforward interpretation of existential types, we can state the
-standard $\epsilon$-$\delta$ definition of continuity as a type:
-\begin{displaymath}
-  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
-\end{displaymath}
-Chaudhuri, Gulwani and Lublinerman \cite{chaudhuri10continuity} have
-presented a program logic based approach to verifying the continuity
-of programs. Algebraically indexed types give us a way of expressing
-and verifying continuity properties of functions in a type based way.
-
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    \underline{c} & \forall \epsilon\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon} \\
-    (+) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
-    (-) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
-    (*) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1\epsilon_2}
-  \end{array}
-\end{displaymath}
-
-\begin{enumerate}
-\item State the operations and equations (rational powers, addition
-  and multiplication (remember: positive reals))
-\item Derive subtyping by addition of $0$
-\item Do the probability monad as a special case, with a specific
-  relational interpretation that relates things according to the
-  differential privacy notion. Need to think about the
-  measure-theoretic problems...
-\item Work out the details of the continuity example, especially the
-  implementation of an ADT for continuous functions that can handle
-  polynomials.
-\end{enumerate}
-
-\subsection{Abelian Group Indexed Types}
-\label{sec:abelian-group-indexed-types}
-
-\newcommand{\Grp}{\mathit{Grp}}
-
-Primitive operations $\Gamma_\Grp$:
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    1 & \tyPrim{G}{1} \\
-    (\cdot) & \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to \tyPrim{G}{a b} \\
-    \ ^{-1} & \forall a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}
-  \end{array}
-\end{displaymath}
-
-For the index-erasure interpretation, we assume a group $(G, 1, \cdot,
-\ ^{-1})$. Let $Z(G)$ be the centre of the $G$: i.e.,~the set of
-elements that commute with every other element in $G$. We set
-$\tyPrimSem{G} = G$. The primitive operations in $\Gamma_\Grp$ are
-interpreted just as the unit, multiplication and inverse operations
-of $G$. For each index context $\Delta$, the set
-$\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be
-$\{ \rho_{(H,h)} \sepbar H\textrm{ is a subgroup of }(\Delta
-\Rightarrow \mathsf{G})/\equiv, h\textrm{ is a group homomorphism }H
-\to Z(G) \}$, where
-\begin{displaymath}
-  \rho_{(H,h)}\ \tyPrimNm{G}\ (e) = \left\{
-    \begin{array}{ll}
-      \{ \} & \textrm{if}\ e\not\in H \\
-      \{ (x, h(e)\cdot x) \sepbar x \in G \} & \textrm{if}\ e\in H
-    \end{array}
-  \right.
-\end{displaymath}
-
-\begin{lemma}
-  For all $\Delta$ and $\rho \in \relEnv{E}_\Grp(\Delta)$,
-  $(\eta_\Grp,\eta_\Grp) \in \rsem{\Gamma_\Grp}{\relEnv{E}}\rho$.
-\end{lemma}
-
-\begin{theorem}
-  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
-  natural numbers $A$ and $m$-by-$1$ matricies of integers $b$,
-  there exists a program $M$ such that
-  \begin{equation}
-    \label{eq:fo-group-type}
-    \Gamma_\Grp \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
-  \end{equation}
-  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e = i_1^{b_1}\dots
-  i_m^{b_m}$, if and only if there is a solution $\vec{x} =
-  (x_1,\cdots,x_n)$ in the integers to the equation $A x^\top = b$.
-\end{theorem}
-
-\fixme{Define shorthand notation $x^z$}
-
-\begin{proof}
-  (If) The program $\Lambda i_1\dots i_m.\ \lambda g_1\dots g_n.\
-  g_1^{x_1}\dots g_n^{x_n}$ satisfies the typing judgment
-  (\ref{eq:fo-group-type}).
-
-  (Only if) Assume that there is a program $M$ satisfying the typing
-  judgment (\ref{eq:fo-group-type}). By \thmref{thm:abstraction} we
-  know that for all relational environments $\rho \in
-  \relEnv{E}_\Grp(i_1,\dots,i_m)$, and all $g_j,g'_j \in G$,
-  \begin{displaymath}
-    \begin{array}{l}
-      (g_1,g'_1) \in \rho\ \tyPrimNm{G}\ (e_1) \land \dots \land (g_n,g'_n) \in \rho\ \tyPrimNm{G}\ (e_n) \Rightarrow \\
-      \quad (\tmSem{M}g_1\dots g_n, \tmSem{M}g'_1\dots g'_n) \in \rho\ \tyPrimNm{G}\ (e)
-    \end{array}
-  \end{displaymath}
-  We select the relational environments
-  \begin{displaymath}
-    \rho\ \tyPrimNm{G}\ (e) = \left\{
-      \begin{array}{ll}
-        \{(g,g) \sepbar g \in \mathcal{G} \} &
-        \begin{array}[t]{@{}l}
-          \textrm{if}\ e = e_1^{x_1}\dots e_n^{x_n} \\
-          \textrm{ for some }x_1,\dots,x_n \in \mathbb{Z}
-        \end{array}
-        \\
-        \{\} & \textrm{otherwise}
-      \end{array}
-    \right.
-  \end{displaymath}
-  Now, for any $g_1,\dots,g_m \in \mathcal{G}$, we know that, for all
-  $j$, $(g_j,g_j) \in \rho\ \tyPrimNm{G}\ (e_j)$, so we know that
-  $(\tySem{M}g_1\dots g_n, \tySem{M}g_1 \dots g_n) \in \rho\
-  \tyPrimNm{G}\ (e)$. Thus there exist integers $x_1,...,x_n$ such
-  that $e = e_1^{x_1}\dots e_n^{x_n}$. But we also know that $e =
-  i_1^{b_1}...i_m^{b_m}$. By the cancellation property of free groups,
-  we learn that $x_1,...,x_n$ is the solution we need.
-\end{proof}
-
-\begin{enumerate}
-\item Can add any other operations that distribute over the group
-  multiplication.
-\item What about partial division operations? Make division just
-  return $0$.
-\item Adding square root as an operation. Need to check that all the
-  other operations preserve the new relational environments. Square
-  root is also supported by the plain relational environments, so the
-  type isomorphism result still holds.
-\end{enumerate}
-
-\subsubsection{Type isomorphisms}
-\label{sec:abelian-group-type-isos}
-
-\begin{enumerate}
-\item This only needs the simpler relational environments, but needs
-  the group used for the index-erasure semantics to be abelian.
-\item One direction is easy, but the other requires the use of the
-  free theorem for the type.
-\item There are two version: one with some fixed ground type as the
-  result, and one with the same type as the result, but with the same
-  parameter.
-\item Integrates into the Smith Normal Form thing too, since
-  invertible integer-valued matrices can be turned into type
-  isomorphisms.
-\item Partiality?
-\end{enumerate}
-
-\fixme{Define what a type isomorphism is}
-
-\begin{theorem}
-  For all natural numbers $n$, the types
-  \begin{displaymath}
-    \tau_{n} \isDefinedAs \forall a\mathord:\mathsf{G}.\ \underbrace{\tyPrim{G}{a} \to ... \to \tyPrim{G}{a}}_{n+1\textrm{ times}} \to \tyPrim{G}{a}
-  \end{displaymath}
-  and
-  \begin{displaymath}
-    \sigma_{n} \isDefinedAs \underbrace{\tyPrim{G}{1} \to ... \tyPrim{G}{1}}_{n\textrm{ times}} \to \tyPrim{G}{1}
-  \end{displaymath}
-  are isomorphic.
-\end{theorem}
-
-\begin{proof}
-  We define programs $\Gamma_\Grp \vdash M : \tau_n \to \sigma_n$ and
-  $\Gamma_\Grp \vdash M^{-1} : \sigma_n \to \tau_n$ as follows:
-  \begin{displaymath}
-    M = \lambda f. \lambda g_1 \dots g_n.\ f\ [1]\ 1\ g_1\ \dots\ g_n
-  \end{displaymath}
-  and
-  \begin{displaymath}
-    M^{-1} = \lambda f. \Lambda a.\ \lambda g_0 \dots g_n.\ (f\ (g_1 g^{-1}_0)\ \dots\ (g_n g^{-1}_0)) g_0
-  \end{displaymath}
-  In one direction, the proof that these are mutually inverse is
-  straightfoward, using only the fact that $G$ is a group. For any
-  program $\Gamma_\Grp \vdash f : \sigma_n$, we have
-  \begin{displaymath}
-    \begin{array}{cl}
-        & \tmSem{M\ (M^{-1}\ f)}\eta_\Grp \\
-      = & (\tmSem{M}\eta_{\Grp})\ (\lambda g_0 \dots g_n.\ (\tmSem{f}\eta_\Grp\ (g_1 g^{-1}_0)\ \dots\ (g_n g^{-1}_0)) g_0) \\
-      = & \lambda g_1\dots g_n.\ \tmSem{f}\eta_\Grp\ (g_1 1^{-1})\ \dots\ (g_n 1^{-1})) 1) \\
-      = & \lambda g_1\dots g_n.\ \tmSem{f}\eta_\Grp\ g_1\ \dots\ g_n \\
-      = & \tmSem{f}\eta_\Grp
-    \end{array}
-  \end{displaymath}
-  as required. In the other direction, we need to use the abstraction
-  theorem. For any program $\Gamma_\Grp \vdash f : \tau_n$, we have
-  \begin{displaymath}
-    \begin{array}{cl}
-        & \tmSem{M^{-1}\ (M\ f)}\eta_\Grp \\
-      = & (\tmSem{M^{-1}}\eta_\Grp)\ (\lambda g_1\dots g_n.\ (\tmSem{f}\eta_\Grp\ 1\ g_1\ \dots\ g_n)) \\
-      = & \lambda g_0\dots g_n.\ (\tmSem{f}\eta_\Grp\ 1\ (g_1g^{-1}_0)\ \dots\ (g_ng^{-1}_0))g_0
-    \end{array}
-  \end{displaymath}
-  By \thmref{thm:abstraction}, we know that the following holds of
-  $\tmSem{f}$:
-  \begin{displaymath}
-    \forall a.\ \forall g_0 \dots g_n.\ \tmSem{f}\eta_\Grp\ (g_0a)\ \dots\ (g_na) = (\tmSem{f}\eta_\Grp\ g_0\ \dots\ g_n)a
-  \end{displaymath}
-  As a special case, we set $a = g^{-1}_0$, and learn that
-  \begin{displaymath}
-    \forall g_0 \dots g_n.\ \tmSem{f}\eta_\Grp\ 1\ (g_1g^{-1}_0) \dots\ (g_ng^{-1}_0) = (\tmSem{f}\eta_\Grp\ g_0\ \dots\ g_n)g^{-1}_0.
-  \end{displaymath}
-  By functional extensionality, we are done.
-\end{proof}
-
-\subsection{Monoid Indexed Types}
-\label{sec:monoid-indexed-types}
-
-\newcommand{\Mon}{\textit{Mon}}
-
-Primitive type is a monoid, and we index by a commutative monoid. An
-example of this is strings with tainting information, or bags, or sets
-of stuff. This gives a slightly different information flow analysis to
-the logical information flow above. In particular, we can now count
-uses of input data.
-
-Primitive operations $\Gamma_{\mathit{Mon}}$:
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    1 & \tyPrim{M}{1} \\
-    (\cdot) & \forall a,b.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{a\cdot b}
-  \end{array}
-\end{displaymath}
-
-For the index-erasure interpretation, we assume a monoid
-$\mathcal{M}$. The primitive type $\tyPrimNm{M}$ is interpreted as
-$\tySem{\tyPrim{M}{e}} = \mathcal{M}$. The primitive operations in
-$\Gamma_{\mathit{Mon}}$ are interpreted just as the unit and
-multiplication of the monoid $\mathcal{M}$.
-
-% Need the following definitions: syntatic group/monoid arising from
-% the syntax, and a neater way of defining relational
-% environments. Also, a neat way of formally introducing a load of
-% primitive operations.
-
-For each index context $\Delta$, the set
-$\relEnv{E}_\Mon(\Delta)$ of relational environments is
-defined to be $\{ \rho_{(M,h)} \sepbar M\textrm{ is a submonoid of
-}(\Delta \Rightarrow \mathsf{M})/\equiv, h\textrm{ is a monoid
-  homomorphism }M \to Z(\mathcal{M}) \}$, where
-\begin{displaymath}
-  \rho_{(M,h)}\ \tyPrimNm{M}\ (e) = \left\{
-    \begin{array}{ll}
-      \{ \} & \textrm{if}\ e\not\in M \\
-      \{ (x, h(e)\cdot x) \sepbar x \in \mathcal{M} \} & \textrm{if}\ e\in M
-    \end{array}
-  \right.
-\end{displaymath}
-Note the mapping to the centre of the monoid
-$\mathcal{M}$. \fixme{introduce the centre notation earlier on}
-
-\begin{lemma}
-  For all $\Delta$ and $\rho \in \relEnv{E}_\Mon(\Delta)$,
-  $(\eta_\Mon,\eta_\Mon) \in \rsem{\Gamma_\Mon}{\relEnv{E}}\rho$.
-\end{lemma}
-
-\begin{theorem}
-  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
-  natural numbers $A$ and $m$-by-$1$ matricies of natural numbers $b$,
-  there exists a program $M$ such that
-  \begin{equation}
-    \label{eq:fo-monoid-type}
-    \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
-  \end{equation}
-  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e =
-  i_1^{b_1}\dots i_m^{b_m}$, if and only if there is a solution
-  $\vec{x} = (x_1,\cdots,x_n)$ in the natural numbers to the equation
-  $A x^\top = b$.
-\end{theorem}
-
-\begin{proof}
-  (If) The program $\Lambda i_1\dots i_m.\ \lambda z_1\dots z_n.\
-  z_1^{x_1}\dots z_n^{x_n}$ satisfies the typing judgment
-  (\ref{eq:fo-monoid-type}).
-
-  (Only if) Assume that there is a program $M$ satisfying the typing
-  judgment (\ref{eq:fo-monoid-type}). By \thmref{thm:abstraction} we
-  know that for all relational environments $\rho \in
-  \relEnv{E}(i_1,\dots,i_m)$, and all $z_j,z'_j \in \mathcal{M}$,
-  \begin{displaymath}
-    \begin{array}{l}
-      (z_1,z'_1) \in \rho\ \tyPrimNm{M}\ (e_1) \land \dots \land (z_n,z'_n) \in \rho\ \tyPrimNm{M}\ (e_n) \Rightarrow \\
-      \quad (\tmSem{M}z_1\dots z_n, \tmSem{M}z'_1\dots z'_n) \in \rho\ \tyPrimNm{M}\ (e)
-    \end{array}
-  \end{displaymath}
-  We select the relational environments
-  \begin{displaymath}
-    \rho\ \tyPrimNm{M}\ (e) = \left\{
-      \begin{array}{ll}
-        \{(z,z) \sepbar z \in \mathcal{M} \} &
-        \begin{array}[t]{@{}l}
-          \textrm{if}\ e = e_1^{x_1}\dots e_n^{x_n} \\
-          \textrm{ for some }x_1,\dots,x_n \in \mathbb{N}
-        \end{array}
-        \\
-        \{\} & \textrm{otherwise}
-      \end{array}
-    \right.
-  \end{displaymath}
-  Now, for any $z_1,\dots,z_m \in \mathcal{M}$, we know that, for all
-  $j$, $(z_j,z_j) \in \rho\ \tyPrimNm{M}\ (e_j)$, so we know that
-  $(\tySem{M}z_1\dots z_n, \tySem{M}z_1 \dots z_n) \in \rho\
-  \tyPrimNm{M}\ (e)$. Thus there exist $x_1,...,x_n$ such that $e =
-  e_1^{x_1}\dots e_n^{x_n}$. But we also know that $e =
-  i_1^{b_1}...i_m^{b_m}$. By the cancellation property of free
-  monoids, we learn that $x_1,...,x_n$ is the solution we need.
-\end{proof}
-
-We can further assume that there are other operations that distribute
-over the multiplication of the monoid, for example semi-ring
-operations (e.g.~relational algebra).
-
-Examples
-\begin{enumerate}
-\item Finite sets of stuff.
-\item Let $\mathcal{M} = \Sigma^* \times
-  \mathcal{P}_{\mathit{fin}}(C)$, where $\Sigma$ is some alphabet, and
-  $C$ is some set of taint markers. The centre of this monoid is
-  $\{([], c) \sepbar c \in \mathcal{P}_{\mathit{fin}}(C) \}$.
-\end{enumerate}
-
 %%% Local Variables:

### 2012-07-06 paper/section-introduction.tex
Introduction, contributions

@@
 algebraic structure.  For example, in computational geometry, points
-in 2-space can be indexed by a co-ordinate frame; in physical
+in 2-space can be indexed by a coordinate frame; in physical
 modelling, quantities are indexed by units-of-measure; in
@@
 Types that are polymorphic over such indices induce invariance
-properties.
+properties, and introduce a useful abstraction barrier.
 
-A function that computes the area of a triangle might
-have simple type
-\[
-\mathit{areaTri} : \tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal
-\]
-but instead we can assign it a polymorphic type
+\paragraph{Invariance}
+A function $\mathit{areaTri}$ that computes the area of a triangle might
+have simple type $\tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal$
+but we can assign it a more expresive polymorphic type
 \[
@@
 \]
-Here $\alpha$ can be interpreted as the \emph{units} of the argument to
-$\mathrm{areaCircle}$, and `$\cdot$' composes units by forming their
-product. To complete the system, we would add an inverse operation and
-identity unit, together with the algebraic laws of an Abelian group,
-permitting, for example, the identification of
+Here $\alpha$ can be interpreted as the \emph{units of measure} of the argument
+to $\mathrm{areaCircle}$, and `$\cdot$' composes units by forming
+their product. To complete the system, we would add an inverse
+operation and identity unit-of-measure, together with the algebraic laws of
+Abelian groups, permitting, for example, the identification of
 $\tyPrim{real}{\alpha\cdot\alpha^{-1}}$ with $\tyPrim{real}{1}$.
 
-In his original paper, Reynolds asserted that ``type structure is a
-syntactic discipline for enforcing levels of abstraction''. 
-We see something analogous here: supposing that all primitive operations are
-given types that reflect their behaviour under translation, there is no way
-to `break' this property, for example, for $\mathrm{areaTri}$ to depend
-on the actual coordinates of its inputs. In fact, through types alone we can
-derive so-called \emph{coordinate-free} geometry.
+\paragraph{Abstraction}
+In his original paper on parametricity, Reynolds asserted that \emph{type structure is
+  a syntactic discipline for enforcing levels of abstraction}.  We see
+something analogous here: supposing that all primitive operations are
+given types that reflect their behaviour under translation, there is
+no way to `break' this property, for example, for $\mathrm{areaTri}$
+to depend on the actual coordinates of its inputs. Furthermore, the
+distinction between points and vectors that is often enforced through
+abstract data types~\cite{CGAL} is here captured by the indices
+instead. For example, the operation that takes two points and computes
+their vector `difference' can be assigned the type
+$\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\times\tyPrim{vec}\alpha\to\tyPrim{vec}0$,
+reflecting the invariance of the result (a pure vector) under
+translations of the point arguments. In essence, through types alone
+we can derive so-called \emph{coordinate-free} geometry~\cite{CFGADT}.
 
@@
 just those functions of the form~$\lambda x. k x^2$ for some constant~$k$.
-(In this case, of course, we expect that~$k = \pi$.)
+In this case, of course, we expect that~$k = \pi$.
+
+\paragraph{Relational parametricity}
+In order to derive such invariance and abstraction properties of types, we
+adopt the techniques of relational parametricity. Over an underlying index-erasure semantics
+we construct binary
+relations parameterized by an environment $\rho$ that describes how values of primitive type
+are related according to their indices. 
+For example, values $v$ and
+$w$ of type $\tyPrim{real}\alpha$ are related when $v$ `scales to' $w$
+according to an interpretation of $\alpha$ (that is, $w=\rho(\alpha)\cdot v$).
+Values of polymorphic type are related exactly when they are related for
+all possible interpretations of the quantified variable. For example, 
+values $v$ and $w$ of type $\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\to
+\tyPrim{vec}\alpha$ are related when they are related at type 
+$\tyPrim{vec}\alpha\to\tyPrim{vec}\alpha$ for all 
+translations $t\in\Transl{2}$ associated with $\alpha$.
+
+As it happens, these examples can be expressed using unary functions
+rather than binary relations: we can write functions that scale, or
+translate, one value to another according to their type, even for
+higher-order function types. Also note that the interpretation of
+indices can be built up compositionally, by recursion over the
+syntactic structure. For example, we have $\rho(\alpha\cdot\alpha) =
+\rho(\alpha)\cdot\rho(\alpha)$ where product of units-of-measure is interpreted by
+product of scale factors.
+
+For more sophisticated applications we need the fully generality of
+binary relations. First, as is also the case with type polymorphism,
+it provides the appropriate meaning for nested quantifiers in types
+such as
+$(\forall\alpha\mathord:\SynScal.\tyPrim{real}\alpha\to\tyPrim{real}\alpha)\to\tyBool$.
+More interestingly, we sometimes use relations that are not built up
+in a compositional way -- we can, for instance, use such relations to
+prove that there are no non-trivial functions of type
+$\forall\alpha\mathord:\SynScal.\tyPrim{real}{\alpha\cdot\alpha}\to\tyPrim{real}\alpha$.
+Finally, there are interesting applications in which even the basic
+relations are not functional. Consider a type system in which the
+index in $\tyPrim{real}\alpha$ is interpreted not as a
+unit-of-measure, but as a measure of \emph{closeness}. Two values $x$
+and $y$ of this type are related if $|x-y| < \rho(\alpha)$, in which
+$\rho(\alpha)$ is a positive real number.  Rather beautifully, the
+standard notion of continuous functions can then be expressed as a
+type:
+\begin{displaymath}
+  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
+\end{displaymath}
 
@@
 
-Key lessons:
-
-\paragraph{Relational meanings are induced by indexes}
-
-\paragraph{Indexes can usefully have algebraic structure}
-
-\paragraph{Relational meanings can be non-compositional}
-
 \subsection{Contributions}
@@
 
+We make the following contributions. 
 \begin{itemize}
-\item Formulation of a general class of type systems that can either
-  be used as programming languages in their own right, or as the
-  targets of type-based analyses.
-\item A collection of compelling examples of algebraically indexed
-  types, including a novel type system for geometry, a refined type
+\item 
+We present a collection of compelling examples of algebraically indexed
+types, including a novel type system for geometry, a type system for
+tracking taint marking (or provenance) of strings, a refined type
 system for information flow, based on logic, and a simple type
 system for basic continuity analysis.
-\item Deduction of useful free theorems in each of our main examples.x
-\item A refined relational interpretation in order to derive
-  non-definability results. Fixing a minor problem in Kennedy's work.
+\item 
+We formulate a general class of type systems that can either be
+used as programming languages in their own right, or as the targets of
+type-based analyses. The type system consists of the usual type
+constructors together with a collection of indexed primitive types,
+universal and existential quantification over the indices, and a
+multi-sorted equational theory for indices. 
+\item
+We describe a relational semantics for the type system and prove an analogue
+of Reynolds' Abstraction Theorem. We pin down precise conditions on sets of
+relation environments for which the theorem holds.
+\item
+For each of our main examples we deduce free theorems that are consequences of
+the Abstraction Theorem, prove non-definability results, and derive interesting
+type isomorphisms.
 \end{itemize}
+We have fully formalized the framework and Abstraction Theorem using the Coq proof assistant,
+making use of strongly-typed term representations throughout~\cite{TypedSyntax}.
 

### 2012-07-06 paper/section-instantiations.tex

@@
   instantiation.}
+
+\begin{example}[Two-dimensional Geometry] We define the ``standard''
+  model of the indexing theory for the two-dimensional geometry
+  example as follows. Each of the sorts are interpreted just as their
+  semantic counterparts:
+  \begin{displaymath}
+    \begin{array}{@{}llll}
+      \semSort{\SynTransl{2}} = \Transl{2}
+      &
+      \semSort{\SynGL{2}} = \GL{2}
+      &
+      \semSort{\SynOrth{2}} = \Orth{2}
+      &
+      \semSort{\SynGL{1}} = \GL{1}
+    \end{array}
+  \end{displaymath}
+  Each of the index operations (e.g.,~the group structure and
+  determinant) are interpreted by the intended semantic
+  operations. These clearly satisfy the axioms we stated in
+  \exref{ex:two-dim-geo-axioms}. The assignment of relations for the
+  primitive types $\tyPrimNm{vec}$ and $\tyPrimNm{real}$ are given by
+  the following definitions:
+  \begin{displaymath}
+    \begin{array}{ll}
+      R^{\bullet}_{\tyPrimNm{vec}} = \{( \vec{0}, \vec{0} ) \}&
+      R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
+      R^{\bullet}_{\tyPrimNm{real}} = \{(0,0)\} &
+      R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
+    \end{array}
+  \end{displaymath}
+  The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
+  are as may be expected given the informal definitions we gave in
+  \autoref{sec:motivating-examples}. We have set
+  $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ to
+  relate $\vec{0}$ and $0$ to themselves to account for the
+  polymorphic zero constants we have assumed.
+\end{example}
 

### 2012-07-06 paper/section-metric.tex

\section{Distance-Indexed Types}
\label{sec:metric-types}

\newcommand{\Metric}{\mathit{Met}}
\newcommand{\metricSort}{\mathsf{R}^{>0}}

The previous two sections have presented examples where the relational
interpretations of primitive types relates a pair of elements by some
transformation if applying the transformation to the first element
yields the second. Thus, the free theorems that we derived directly
take the form of ``invariance'' properties, where some equation holds
between two terms. In this section, we examine another instantiation
of our general framework that relates values when they are within a
certain distance. The free theorems that we obtain inform us of the
effect that programs have on the distances between values. For
example, a program $M$ of type
\begin{displaymath}
  \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2}
\end{displaymath}
must satisfy the property that for all $\epsilon_1, \epsilon_2 > 0$
and $x, x', y, y' \in \mathbb{R}$,
\begin{displaymath} % FIXME: include \eta_\Metric
  \begin{array}{l}
    \textrm{if }|x - x'| < \epsilon_1 \textrm{ and } |y - y'| < \epsilon_2 \textrm{ then}\\
    \hspace{2em}|\tmSem{M}\ x\ y - \tmSem{M}\ x'\ y'| < \epsilon_1 + \epsilon_2
  \end{array}
\end{displaymath}
A type system with a relational interpretation that tracks distances
between values has previously been investigated by Reed and Pierce
\cite{reed10distance} in the setting of differential privacy. Their
system uses a linear type discipline to ensure that all programs are
non-expansive (i.e., the distance between the outputs is no greater
than the distance between the inputs). Here, we instantiate the
framework from \autoref{sec:a-general-framework} in order to express
non-expansivity, and more, through algebraically indexed types.

\paragraph{Instantiation of the General Framework}
We assume a single indexing sort $\metricSort$, intended to represent
positive, non-zero real numbers. For the index operations, we assume
the operations $\min, \max, +$ and multiplication by constant
reals. There is a single primitive type $\tyPrimNm{real}$, with
$\primTyArity(\tyPrimNm{real}) = [\metricSort]$. The primitive
operations $\Gamma_{\Metric}$ are as follows, where $c$ stands for
arbitrary real-valued constants.
\begin{displaymath}
  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
    \underline{c} & \forall \epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \\
    (+) & \forall \epsilon_1, \epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
    (-) & \forall \epsilon_1, \epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
    \underline{c} * & \forall \epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \to \tyPrim{real}{c\epsilon} \\
    \mathrm{coerce} & \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\max(\epsilon_1,\epsilon_2)}
  \end{array}
\end{displaymath}
We assume that the index-erasure semantics of the $\tyPrimNm{real}$
type is just the set $\mathbb{R}$, so all except the last operation
have straightforward interpretations. The $\mathrm{coerce}$ operation
is interpreted just as the identity function. The index-erasure
interpretations of the primitive operations are collected together
into an environment $\eta_\Metric \in \ctxtSem{\Gamma_\Mon}$.

For the families of sets of relational environments
$\relEnv{E}_\Metric$, we use the general construction in
\autoref{sec:constr-rel-env} with the following interpretations of the
primitive type $\tyPrimNm{real}$:
\begin{displaymath} %FIXME: define \Eq earlier on
  \begin{array}{l@{\hspace{2em}}l}
    R^\bullet_{\tyPrimNm{real}} = \Eq_\mathbb{R} &
    R_{\tyPrimNm{real}}(\epsilon) = \{ (x,x') \sepbar |x - x'| < \epsilon \}
  \end{array}
\end{displaymath}

\begin{lemma}
  For all $\Delta$ and $\rho \in \relEnv{E}_\Metric(\Delta)$,
  $(\eta_\Metric, \eta_\Metric) \in
  \rsem{\Gamma}{\relEnv{E}_\Metric}{\rho}$.
\end{lemma}

\begin{example}[Uniform Continuity]
  By making use of the existential types in our framework, we can
  state the standard $\epsilon$-$\delta$ definition of
  \emph{uniform} continuity as a type:
  \begin{displaymath}
    \forall \epsilon \mathord: \metricSort.\ \exists \delta\mathord: \metricSort.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
  \end{displaymath}
  Given a program $M$ with this type, we obtain from
  \thmref{thm:abstraction} a free theorem, which is exactly the
  definition of uniform continuity:
  \begin{displaymath}
    \forall \epsilon \mathord> 0. \exists \delta \mathord> 0. \forall x,x'. |x - x'| < \delta \Rightarrow |\tmSem{M}\ x - \tmSem{M}\ x'| < \epsilon
  \end{displaymath}
  Note that this definition differs from the $\epsilon$-$\delta$
  definition of continuity by the order of quantification. For
  continuity, the $\forall x$ is before the $\exists \delta$, so the
  distance $\delta$ may depend on the value of $x$. We suspect that to
  express standard continuity as a type would require some form of
  type dependency. Chaudhuri, Gulwani and Lublinerman
  \cite{chaudhuri10continuity} have presented a program logic based
  approach to verifying the continuity of programs.
\end{example}

% \begin{example}[Differential Privacy]
%   Relative error? Probability monads?

%   Don't have space to develop this fully...

%   $\circ \tyPrimNm{real}$ -- probability distributions on the real numbers

%   Need to think about this...
%   \begin{displaymath}
%     \mathsf{noise} : \tyPrim{$\circ$real}{1}
%   \end{displaymath}
% \end{example}

% \begin{mathpar}
%   % commutativity
%   \min(\epsilon_1, \epsilon_2) = \min(\epsilon_2, \epsilon_1)

%   \max(\epsilon_1, \epsilon_2) = \max(\epsilon_2, \epsilon_1)

%   % associativity
%   \max(\epsilon_1, \max(\epsilon_2, \epsilon_3)) = \max(\max(\epsilon_1, \epsilon_2), \epsilon_3)

%   \min(\epsilon_1, \min(\epsilon_2, \epsilon_3)) = \min(\min(\epsilon_1, \epsilon_2), \epsilon_3)

%   % absorption
%   \max(\epsilon_1, \min(\epsilon_1, \epsilon_2)) = \epsilon_1

%   \min(\epsilon_1, \max(\epsilon_1, \epsilon_2)) = \epsilon_1
% \end{mathpar}

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:

### 2012-07-06 paper/section-singleton.tex

--- 2012-07-03 paper/section-further-examples.tex
@@
-\section{Monoid Indexed Types}
-\label{sec:monoid-indexed-types}
-
-\newcommand{\Mon}{\textit{Mon}}
-
-In the last section we looked at types indexed by abelian groups. We
-now examine an instantiation of our general framework with a monoid
-type indexed by elements of a commutative monoid. Our main example of
-monoid indexed types is strings indexed by ``taint'' markers: when
-strings are appended their taint markers are combined. Thanks to the
-information in the indexes, we are able to prove an indefinability
-result (\thmref{thm:monoid-indefinability}).
-
-We assume a single indexing sort $\mathsf{M}$, with the monoid
-operations of unit and multiplication along with the commutative
-monoid axioms. There is a single primitive type $\tyPrimNm{M}$, with
-$\primTyArity(\tyPrimNm{M}) = [\mathsf{M}]$. For the index-erasure
-semantics, we assume some (not necessarily commutative) monoid $M$,
-which we also write multiplicatively. Thus, we set
-$\tyPrimSem{\tyPrimNm{M}} = M$. The collection of primitive operations
-$\Gamma_{\mathit{Mon}}$ is simply the two monoid operations:
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{2em}}r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    1 & \tyPrim{M}{1}, &
-    (\cdot) & \forall a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{a b}
-  \end{array}
-\end{displaymath}
-which we interpret using the monoid structure of $M$, giving an
-environment $\eta_\Mon \in \ctxtSem{\Gamma_\Mon}$.
-
-For the sets of relational environments, we make use of the
-construction in \autoref{sec:constr-rel-env}, setting the chosen model
-$\mathcal{M}$ to be the \emph{centre} $Z(M)$ of $M$ (recall that the
-centre of a monoid is the set of all elements that commute with every
-other element). We then set $R_{\tyPrimNm{M}}(m) = \{ (n,mn) \sepbar n
-\in M \}$ and $R^\bullet_{\tyPrimNm{M}} = \emptyset$. By setting the
-model $\mathcal{M}$ to only be the elements of $M$ that commute with
-everything else, we ensure that the following lemma holds:
-\begin{lemma}
-  For all $\Delta$ and $\rho \in \relEnv{E}_\Mon(\Delta)$,
-  $(\eta_\Mon,\eta_\Mon) \in \rsem{\Gamma_\Mon}{\relEnv{E}}\rho$.
-\end{lemma}
-
-\newcommand{\mscup}{\stackrel{ms}\cup}
-
-\begin{example}
-  By selecting a suitable monoid $M$, we can simulate a programming
-  language that dynamically annotates strings with taint markers. Let
-  $\Sigma$ be a set of characters from some alphabet, and let $T$ be a
-  set of taint markers. Set $M = \Sigma^* \times
-  \mathcal{MS}_{\mathit{fin}}(T)$, where $\mathcal{MS}_{\mathit{fin}}$
-  denotes the finite power set operator. Since both $\Sigma^*$ and
-  $\mathcal{MS}_{\mathit{fin}}(T)$ are monoids (the latter with
-  multiset union as the monoid multiplication, which we write as
-  $\mscup$), $M$ is itself a monoid. The centre of $M$, $Z(M)$,
-  consists of all elements of the form $([], t)$ for multisets of
-  taint markers.
-
-  As an example of a free theorem for this instantiation of our
-  general framework, consider programs of the following type:
-  \begin{displaymath}
-    ; \Gamma_\Mon \vdash M : \forall a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{aa}
-  \end{displaymath}
-  From this type we can derive the following free theorem about
-  $M$. For all $a, b \in \mathcal{MS}_{\mathit{fin}}(T)$, $(s_1,t_1),
-  (s_2,t_2) \in \Sigma^*$,
-  \begin{displaymath}
-    \begin{array}{l}
-      \pi_2(\tmSem{M}\ (s_1,t_1 \mscup a)\ (s_2,t_2 \mscup b)) = \\
-      \hspace{2em}\pi_2(\tmSem{M}\ (s_1,t_1)\ (s_2,t_2)) \mscup a \mscup a
-    \end{array}
-  \end{displaymath}
-  We read this result as follows: if we taint the first input by $a$
-  and the second by $b$, then the taint of the output varies according
-  to $M$'s type. In this case the output contains two copies of $a$
-  and none of $b$.
-\end{example}
-
-% if $b$ does not appear in the output, then it cannot have been used in
-% the program. Turn this into a theorem...
-
-% If forall i1...,in. M<e0> \to ... \to M<em> -> M<e>
-% and e0 is not a factor of e, then we can get rid of the first input
-
-% It ought to be invariant under its first input... Set R^\bullet to
-% be the everywhere relation....
-
-% prove this by using a relational environment that excludes
-
-\begin{theorem}
-  \begin{equation}
-    \label{eq:fo-monoid-type-insensitive}
-    ; \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
-  \end{equation}
-
-\end{theorem}
-
-\begin{theorem}
-  \label{thm:monoid-indefinability}
-  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
-  natural numbers $A$ and $m$-by-$1$ matricies of natural numbers $b$,
-  there exists a program $M$ such that
-  \begin{displaymath}
-    \label{eq:fo-monoid-type}
-    ; \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
-  \end{displaymath}
-  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e =
-  i_1^{b_1}\dots i_m^{b_m}$, if and only if there is a solution
-  $\vec{x} = (x_1,\cdots,x_n)$ in the natural numbers to the equation
-  $A x^\top = b$.
-\end{theorem}
-
-\begin{example}
-  By \thmref{thm:monoid-indefinability}, it is impossible to write a
-  program with the type $\forall m_1,m_2 \mathord: \mathsf{M}.\
-  \tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_1}$, showing that it is
-  impossible to convert a doubly tainted value to a singly tainted
-  one.
-
-  In contrast to the case of abelian group indexed types, the lack of
-  inverses in the monoid indexed setting means that the type $\forall
-  m_1,m_2\mathord:\mathsf{M}.\ \tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_2}
-  \to \tyPrim{M}{m_1}$ is also uninhabited. There are no naturals
-  $a_1,a_2$ such that $(m_1m_2)^{a_1}m_2^{a_2} \equiv m_1$.
-\end{example}
-
-% \fixme{Finish this} We can further assume that there are other
-% operations that distribute over the multiplication of the monoid, for
-% example semi-ring operations (e.g.~relational algebra).
-
-\section{Metric Spaces and Simple Continuity Analysis}
-\label{sec:continuity-analysis}
-
-Our final motivating example of algebraically indexed types and
-representation independence makes use of the metric space structure of
-the real numbers. In the geometry example, we used relational
-interpretations that related values via geometrical
-transformations. An alternative relational interpretation of the real
-numbers is to relate values that are close, in the metric space
-sense. By indexing types by a measure of closeness, we can actually
-state continuity of functions as a typing problem.
-
-We introduce an indexing sort $\mathsf{R}^{>0}$, representing positive
-distances between real numbers. Our type of real numbers is indexed by
-expressions of sort $\mathsf{R}^{>0}$, and we use the following
-index-erasure and relational interpretations:
-\begin{displaymath}
-  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\tyPrim{real}{e}} & \mathbb{R} \\
-    \rsem{\tyPrim{real}{e}}{}\rho & \{ (x,x') \sepbar |x-x'| < \sem{e}\rho \}
-  \end{array}
-\end{displaymath}
-Thus, two real numbers are related if they only differ by the positive
-real number assigned to $e$. With this interpretation, and a
-straightforward interpretation of existential types, we can state the
-standard $\epsilon$-$\delta$ definition of continuity as a type:
-\begin{displaymath}
-  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
-\end{displaymath}
-Chaudhuri, Gulwani and Lublinerman \cite{chaudhuri10continuity} have
-presented a program logic based approach to verifying the continuity
-of programs. Algebraically indexed types give us a way of expressing
-and verifying continuity properties of functions in a type based way.
-
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    \underline{c} & \forall \epsilon\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon} \\
-    (+) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
-    (-) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
-    (*) & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1\epsilon_2}
-  \end{array}
-\end{displaymath}
-
-\begin{enumerate}
-\item State the operations and equations (rational powers, addition
-  and multiplication (remember: positive reals))
-\item Derive subtyping by addition of $0$
-\item Do the probability monad as a special case, with a specific
-  relational interpretation that relates things according to the
-  differential privacy notion. Need to think about the
-  measure-theoretic problems...
-\item Work out the details of the continuity example, especially the
-  implementation of an ADT for continuous functions that can handle
-  polynomials.
-\end{enumerate}
-
 \section{Singleton Types}
@@
 \newcommand{\Sing}{\mathit{Sing}}
+
+\begin{enumerate}
+\item Introduce the general idea of singleton types as a way of
+  reifying information held in the indexes into the ``programming''
+  world.
+\item Helps maintain a separation between run-time and compile time
+  data.
+\end{enumerate}
 
@@
 
+\begin{lemma}
+  For all $\Delta$ and $\rho \in \relEnv{E}_\Sing(\Delta)$, $(\eta_\Sing, \eta_\Sing) \in \rsem{\Gamma}{\relEnv{E}_\Sing}{\rho}$.
+\end{lemma}
+
+Due to the precision of the index expressions, we obtain a very
+precise characterisation of the definable programs under the singleton
+typing discipline:
 \begin{theorem}
@@
 
-\section{Logical Information Flow}
-\label{sec:information-flow}
-
-\begin{enumerate}
-\item This is now a special case of the preceeding subsection
-\item Also makes use of the relational aspect, the previous stuff
-  could make do with unary logical predicates as the interpretations
-  of types, but the information flow properties make use of the
-  relational interpretation.
-\end{enumerate}
-
-\fixme{
-  \begin{enumerate}
-  \item Make the definition of control of information flow clearer
-  \item Make the example a bit more compelling
-  \item Put in some examples of composite principals formed from
-    boolean logic, and the use of boolean algebra to do reasoning in
-    the types. List the laws of boolean algebra, and the laws of
-    Heyting algebra. Need to explain the difference between 
-  \item Possibly use singleton types to be able to do more interesting
-    security policies?
-  \item Put in the formal statement of the information flow property
-    and the proof.
-  \end{enumerate}
-}
-
-An extreme example of invariance under change of representation is the
-tracking of information flow through programs. If a program does not
-depend upon a particular piece of information, then it will be
-invariant under \emph{all} changes of representation of this
-information. As thoroughly described by Sabelfeld and Sands
-\cite{sabelfeld01per}, information flow can be captured semantically
-through the use of partial equivalence relations (PERs). Interpreting
-types as PERs forms an instance of Reynolds' approach to abstraction
-and representation independence. Abadi, Banerjee, Heintze and Riecke
-\cite{abadi99core} built a \emph{Core Calculus for Dependency}, using
-a type system based around a security level indexed monad $T_lA$. Tse
-and Zdancewic \cite{tse04translating} showed that it is possible to
-translate Abadi et al.'s calculus into System F, translating the
-monadic type $T_lA$ to $\alpha_l \to A$ for some free type variable
-$\alpha_l$, and making direct use of Reynolds' abstraction theorem to
-prove information flow properties. We now show how a refined variant
-of Tse and Zdancewic's translation can be expressed via
-algebraically-indexed types.
-
-We assume a single indexing sort $\mathsf{principal}$, intended to
-represent (possibly composite) principals in a system. Principals are
-combined using the connectives of boolean logic ($\top$, $\bot$,
-$\lor$, $\land$, $\lnot$). For the equational theory of principals, we
-assume all the axioms of boolean algebra, and semantically we
-interpret composite principals as boolean values: either true ($\top$)
-or false ($\bot$). We introduce a single primitive type, indexed by
-expressions of sort $\mathsf{principal}$, $\tyPrim{T}{e}$, with the
-following interpretations:
-\begin{displaymath}
-  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\tyPrim{T}{e}} & \{*\} \\
-    \rsem{\tyPrim{T}{e}}{}\rho & \left\{
-      \begin{array}{ll}
-        \{(*,*)\} & \textrm{if }\sem{e}\rho = \top \\
-        \{\}      & \textrm{if }\sem{e}\rho = \bot
-      \end{array}
-      \right.
-  \end{array}
-\end{displaymath}
-The relational interpretation of $\tyPrim{T}{e}$ forces this type to
-only be inhabited when the boolean expression $e$ evaluates to true
-under the environment $\rho$. To write programs, we assume the
-following pair of primitive operations:
-\begin{eqnarray*}
-  \mathrm{proj} & : & \forall p, q\mathord:\mathsf{principal}.\ \tyPrim{T}{p \land q} \to \tyPrim{T}{p} \\
-  \mathrm{combine} & : & \forall p, q\mathord:\mathsf{principal}.\ \tyPrim{T}{p} \to \tyPrim{T}{q} \to \tyPrim{T}{p \land q}
-\end{eqnarray*}
-We can now adapt Tse and Zdancewic's translation of Abadi et al.'s
-monadic type to our setting. We define a type synonym, for each type
-$A$ and expression $e$ of sort $\mathsf{principal}$:
-\begin{displaymath}
-  T_eA = \tyPrim{T}{e} \to A
-\end{displaymath}
-For every principal $e$, we can endow the types $T_e-$ with the
-structure of a monad. This is due to the fact that it is an instance
-of the ``environment'' (or ``reader'') monad \cite{jones95functional}.
-
-\begin{example}
-  Suppose we have a program with the following type:
-  \begin{displaymath}
-    \forall p,q\mathord:\mathsf{principal}. \tyPrim{T}{e} \to T_p(\tyPrimNm{unit} + \tyPrimNm{unit}) \to T_q(\tyPrimNm{unit} + \tyPrimNm{unit})
-  \end{displaymath}
-  where $e$ is some boolean algebra expression involving $p$ and
-  $q$. We will show in \autoref{sec:instantiations} that programs with
-  this type are non-constant (i.e.~depend on their input) if and only
-  if the formula $e \Rightarrow q \Rightarrow p$ is valid in boolean
-  logic.
-\end{example}
-
-\begin{enumerate}
-\item Treat the information flow thing as a type isomorphism, in the
-  same way that Tse and Zdancewic do
-  \begin{displaymath}
-    \begin{array}{l}
-    \forall \vec{p}\mathord:\mathsf{principal}.\ T_{e_1}\tyPrimNm{bool} \to T_{e_2}\tyPrimNm{bool}\\
-    \hspace{4em}\cong \\
-    \left\{
-      \begin{array}{ll}
-        \tyPrimNm{bool} \tyArr \tyPrimNm{bool} & \textrm{if }e_2 \vdash e_1 \\
-        \tyPrimNm{bool} & \textrm{if }e_2 \not\vdash e_1
-      \end{array}
-    \right.
-  \end{array}
-  \end{displaymath}
-  where $\tyPrimNm{bool} = \tyUnit + \tyUnit$.
-\item mention the issue with bogus tokens if we allow for a
-  call-by-name style of non-termination
-\item if we axiomatise intuitionistic logic instead of
-  boolean/classical logic, we get a different type isomorphism.
-\item The isomorphism relies upon $e_1 \vdash e_2 \Leftrightarrow e_1
-  \lor e_2 = e_1$.
-\item Relevant logic? Look at the Wikipedia page to see how to do
-  residuated (commutative) monoids on lattices with just
-  equations. With this, we will be able to do substructural logics
-  too. As long as we know that a particular axiomatisation is sound
-  and complete for some reasoning system, then we can state and prove
-  the non-interference isomorphism above.
-\item Compare to Tse and Zdancewic's addition of axioms to the logic.
-\item Internalising equality? The logical structure can then be used
-  on types as well as everything else. Need the special rules for
-  equality... (see the Fomega paper).
-\end{enumerate}
-
-% Go a step further by considering equivalence relations on word32, and
-% their lattice structure. Link this to quantative information
-% flow. Will also need to include singleton types.
-
-\subsection{Abelian Group Indexed Types}
-\label{sec:abelian-group-indexed-types}
-
-\newcommand{\Grp}{\mathit{Grp}}
-
-Primitive operations $\Gamma_\Grp$:
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    1 & \tyPrim{G}{1} \\
-    (\cdot) & \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to \tyPrim{G}{a b} \\
-    \ ^{-1} & \forall a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}
-  \end{array}
-\end{displaymath}
-
-For the index-erasure interpretation, we assume a group $(G, 1, \cdot,
-\ ^{-1})$. Let $Z(G)$ be the centre of the $G$: i.e.,~the set of
-elements that commute with every other element in $G$. We set
-$\tyPrimSem{G} = G$. The primitive operations in $\Gamma_\Grp$ are
-interpreted just as the unit, multiplication and inverse operations
-of $G$. For each index context $\Delta$, the set
-$\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be
-$\{ \rho_{(H,h)} \sepbar H\textrm{ is a subgroup of }(\Delta
-\Rightarrow \mathsf{G})/\equiv, h\textrm{ is a group homomorphism }H
-\to Z(G) \}$, where
-\begin{displaymath}
-  \rho_{(H,h)}\ \tyPrimNm{G}\ (e) = \left\{
-    \begin{array}{ll}
-      \{ \} & \textrm{if}\ e\not\in H \\
-      \{ (x, h(e)\cdot x) \sepbar x \in G \} & \textrm{if}\ e\in H
-    \end{array}
-  \right.
-\end{displaymath}
-
-\begin{lemma}
-  For all $\Delta$ and $\rho \in \relEnv{E}_\Grp(\Delta)$,
-  $(\eta_\Grp,\eta_\Grp) \in \rsem{\Gamma_\Grp}{\relEnv{E}}\rho$.
-\end{lemma}
-
-\begin{theorem}
-  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
-  natural numbers $A$ and $m$-by-$1$ matricies of integers $b$,
-  there exists a program $M$ such that
-  \begin{equation}
-    \label{eq:fo-group-type}
-    \Gamma_\Grp \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
-  \end{equation}
-  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e = i_1^{b_1}\dots
-  i_m^{b_m}$, if and only if there is a solution $\vec{x} =
-  (x_1,\cdots,x_n)$ in the integers to the equation $A x^\top = b$.
-\end{theorem}
-
-\fixme{Define shorthand notation $x^z$}
-
-\begin{proof}
-  (If) The program $\Lambda i_1\dots i_m.\ \lambda g_1\dots g_n.\
-  g_1^{x_1}\dots g_n^{x_n}$ satisfies the typing judgment
-  (\ref{eq:fo-group-type}).
-
-  (Only if) Assume that there is a program $M$ satisfying the typing
-  judgment (\ref{eq:fo-group-type}). By \thmref{thm:abstraction} we
-  know that for all relational environments $\rho \in
-  \relEnv{E}_\Grp(i_1,\dots,i_m)$, and all $g_j,g'_j \in G$,
-  \begin{displaymath}
-    \begin{array}{l}
-      (g_1,g'_1) \in \rho\ \tyPrimNm{G}\ (e_1) \land \dots \land (g_n,g'_n) \in \rho\ \tyPrimNm{G}\ (e_n) \Rightarrow \\
-      \quad (\tmSem{M}g_1\dots g_n, \tmSem{M}g'_1\dots g'_n) \in \rho\ \tyPrimNm{G}\ (e)
-    \end{array}
-  \end{displaymath}
-  We select the relational environments
-  \begin{displaymath}
-    \rho\ \tyPrimNm{G}\ (e) = \left\{
-      \begin{array}{ll}
-        \{(g,g) \sepbar g \in \mathcal{G} \} &
-        \begin{array}[t]{@{}l}
-          \textrm{if}\ e = e_1^{x_1}\dots e_n^{x_n} \\
-          \textrm{ for some }x_1,\dots,x_n \in \mathbb{Z}
-        \end{array}
-        \\
-        \{\} & \textrm{otherwise}
-      \end{array}
-    \right.
-  \end{displaymath}
-  Now, for any $g_1,\dots,g_m \in \mathcal{G}$, we know that, for all
-  $j$, $(g_j,g_j) \in \rho\ \tyPrimNm{G}\ (e_j)$, so we know that
-  $(\tySem{M}g_1\dots g_n, \tySem{M}g_1 \dots g_n) \in \rho\
-  \tyPrimNm{G}\ (e)$. Thus there exist integers $x_1,...,x_n$ such
-  that $e = e_1^{x_1}\dots e_n^{x_n}$. But we also know that $e =
-  i_1^{b_1}...i_m^{b_m}$. By the cancellation property of free groups,
-  we learn that $x_1,...,x_n$ is the solution we need.
-\end{proof}
-
-\begin{enumerate}
-\item Can add any other operations that distribute over the group
-  multiplication.
-\item What about partial division operations? Make division just
-  return $0$.
-\item Adding square root as an operation. Need to check that all the
-  other operations preserve the new relational environments. Square
-  root is also supported by the plain relational environments, so the
-  type isomorphism result still holds.
-\end{enumerate}
-
-\subsubsection{Type isomorphisms}
-\label{sec:abelian-group-type-isos}
-
-\begin{enumerate}
-\item This only needs the simpler relational environments, but needs
-  the group used for the index-erasure semantics to be abelian.
-\item One direction is easy, but the other requires the use of the
-  free theorem for the type.
-\item There are two version: one with some fixed ground type as the
-  result, and one with the same type as the result, but with the same
-  parameter.
-\item Integrates into the Smith Normal Form thing too, since
-  invertible integer-valued matrices can be turned into type
-  isomorphisms.
-\item Partiality?
-\end{enumerate}
-
-\fixme{Define what a type isomorphism is}
-
-\begin{theorem}
-  For all natural numbers $n$, the types
-  \begin{displaymath}
-    \tau_{n} \isDefinedAs \forall a\mathord:\mathsf{G}.\ \underbrace{\tyPrim{G}{a} \to ... \to \tyPrim{G}{a}}_{n+1\textrm{ times}} \to \tyPrim{G}{a}
-  \end{displaymath}
-  and
-  \begin{displaymath}
-    \sigma_{n} \isDefinedAs \underbrace{\tyPrim{G}{1} \to ... \tyPrim{G}{1}}_{n\textrm{ times}} \to \tyPrim{G}{1}
-  \end{displaymath}
-  are isomorphic.
-\end{theorem}
-
-\begin{proof}
-  We define programs $\Gamma_\Grp \vdash M : \tau_n \to \sigma_n$ and
-  $\Gamma_\Grp \vdash M^{-1} : \sigma_n \to \tau_n$ as follows:
-  \begin{displaymath}
-    M = \lambda f. \lambda g_1 \dots g_n.\ f\ [1]\ 1\ g_1\ \dots\ g_n
-  \end{displaymath}
-  and
-  \begin{displaymath}
-    M^{-1} = \lambda f. \Lambda a.\ \lambda g_0 \dots g_n.\ (f\ (g_1 g^{-1}_0)\ \dots\ (g_n g^{-1}_0)) g_0
-  \end{displaymath}
-  In one direction, the proof that these are mutually inverse is
-  straightfoward, using only the fact that $G$ is a group. For any
-  program $\Gamma_\Grp \vdash f : \sigma_n$, we have
-  \begin{displaymath}
-    \begin{array}{cl}
-        & \tmSem{M\ (M^{-1}\ f)}\eta_\Grp \\
-      = & (\tmSem{M}\eta_{\Grp})\ (\lambda g_0 \dots g_n.\ (\tmSem{f}\eta_\Grp\ (g_1 g^{-1}_0)\ \dots\ (g_n g^{-1}_0)) g_0) \\
-      = & \lambda g_1\dots g_n.\ \tmSem{f}\eta_\Grp\ (g_1 1^{-1})\ \dots\ (g_n 1^{-1})) 1) \\
-      = & \lambda g_1\dots g_n.\ \tmSem{f}\eta_\Grp\ g_1\ \dots\ g_n \\
-      = & \tmSem{f}\eta_\Grp
-    \end{array}
-  \end{displaymath}
-  as required. In the other direction, we need to use the abstraction
-  theorem. For any program $\Gamma_\Grp \vdash f : \tau_n$, we have
-  \begin{displaymath}
-    \begin{array}{cl}
-        & \tmSem{M^{-1}\ (M\ f)}\eta_\Grp \\
-      = & (\tmSem{M^{-1}}\eta_\Grp)\ (\lambda g_1\dots g_n.\ (\tmSem{f}\eta_\Grp\ 1\ g_1\ \dots\ g_n)) \\
-      = & \lambda g_0\dots g_n.\ (\tmSem{f}\eta_\Grp\ 1\ (g_1g^{-1}_0)\ \dots\ (g_ng^{-1}_0))g_0
-    \end{array}
-  \end{displaymath}
-  By \thmref{thm:abstraction}, we know that the following holds of
-  $\tmSem{f}$:
-  \begin{displaymath}
-    \forall a.\ \forall g_0 \dots g_n.\ \tmSem{f}\eta_\Grp\ (g_0a)\ \dots\ (g_na) = (\tmSem{f}\eta_\Grp\ g_0\ \dots\ g_n)a
-  \end{displaymath}
-  As a special case, we set $a = g^{-1}_0$, and learn that
-  \begin{displaymath}
-    \forall g_0 \dots g_n.\ \tmSem{f}\eta_\Grp\ 1\ (g_1g^{-1}_0) \dots\ (g_ng^{-1}_0) = (\tmSem{f}\eta_\Grp\ g_0\ \dots\ g_n)g^{-1}_0.
-  \end{displaymath}
-  By functional extensionality, we are done.
-\end{proof}
-
 %%% Local Variables:

### 2012-07-06 paper/section-general-framework.tex
Definitions of contextual equivalence, semantic equivalence and the soundness property

@@
 
+We use our index-erasure semantics to define when a pair of programs
+are contextually equivalent with respect to syntactically defined
+applicative contexts.
+
+\begin{definition}\label{defn:ctxt-equiv}
+  Let $\Gamma_{\mathit{ops}}$ be a typing context with no free index
+  variables that describes the types of the primitive operations, and
+  let $\eta_{\mathit{ops}} \in \ctxtSem{\Gamma}$ be an interpretation
+  of these primitive operations.
+
+  Two programs $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 : A$
+  and $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_2 : A$ are
+  \emph{contextually equivalent} ($\Delta; \Gamma_{\mathit{ops}},
+  \Gamma \vdash M_1 \stackrel{ctx}\approx M_2 : A$) if for all
+  programs $-;\Gamma_{\mathit{ops}} \vdash T : (\forall \Delta. \Gamma
+  \to A) \to \tyPrimNm{unit} + \tyPrimNm{unit}$, it is the case that
+  $\tmSem{T\ (\Lambda \Delta. \lambda \Gamma. M_1)}\eta_{\mathit{ops}}
+  = \tmSem{T\ (\Lambda \Delta. \lambda
+    \Gamma. M_2)}\eta_{\mathit{ops}}$.
+\end{definition}
+
 \paragraph{The Abstraction Theorem}
@@
   $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
+\end{theorem}
+
+We use the relational interpretation to define the notion of semantic
+equivalence, which is a sound approximation of contextual equivalence,
+as a consequence of \thmref{thm:abstraction}.
+\begin{definition}
+  Let $\Gamma_{\mathit{ops}}$ and $\eta_{\mathit{ops}}$ be a context
+  of primitive operations and its interpretation as in
+  \defref{defn:ctxt-equiv}. Let $\relEnv{E}_{\mathit{ops}}$ be a
+  family of sets of relational environments such that for all $\Delta$
+  and $\rho \in \relEnv{E}_{\mathit{ops}}(\Delta)$,
+  $(\eta_{\mathit{ops}}, \eta_{\mathit{ops}}) \in
+  \rsem{\Gamma}{\relEnv{E}_{\mathit{ops}}}\rho$.
+
+  Two programs $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 : A$
+  and $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_2 : A$ are
+  \emph{semantically equal} ($\Delta; \Gamma_{\mathit{ops}}, \Gamma
+  \vdash M_1 \stackrel{sem}\sim M_2 : A$) if for all $\rho \in
+  \relEnv{E}_{\mathit{ops}}$, and all $(\eta_1,\eta_2) \in
+  \rsem{\Gamma}{\relEnv{E}_{\mathit{ops}}}\rho$, we have
+  $(\tmSem{M_1}(\eta_{\mathit{ops}},
+  \eta_1),\tmSem{M_2}(\eta_{\mathit{ops}}, \eta_2)) \in
+  \rsem{A}{\relEnv{E}_{\mathit{ops}}}\rho$.
+\end{definition}
+
+\begin{theorem}\label{thm:soundness}
+  $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 \stackrel{sem}\sim M_2 : A$ implies
+  $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 \stackrel{ctx}\approx M_2 : A$
 \end{theorem}

### 2012-07-06 paper/section-general-framework.tex
Incorporate the well-indexed context rules into the figure

@@
 contexts, which are in turn defined with respect to an index
-context. Well-indexed typing contexts with respect to an index context
-$\Delta$ are sequences of variable/type pairs with no repeated
-variable names such that each type is well-indexed with respect to
-$\Delta$. Formally, well-indexed typing contexts are defined by the
-following two rules:
-\begin{mathpar}
-  \inferrule*
-  { }
-  {\Delta \vdash \epsilon \isCtxt}
-
-  \inferrule*
-  {\Delta \vdash \Gamma \isCtxt \\ \Delta \vdash A \isType \\ x \not\in \Gamma}
-  {\Delta \vdash \Gamma, x : A \isCtxt}
-\end{mathpar}
-Application of simultaneous substitutions extends to typing contexts
-by applying the simultaneous substitution to each type.
+context. Well-indexed typing contexts $\Delta \vdash \Gamma \isCtxt$
+with respect to an index context $\Delta$ are sequences of
+variable/type pairs with no repeated variable names such that each
+type is well-indexed with respect to $\Delta$. Formally, well-indexed
+typing contexts are defined by the two rules in
+\autoref{fig:programs}.  Application of simultaneous substitutions
+extends to typing contexts by applying the simultaneous substitution
+to each type.
 
@@
     \begin{mathpar}
+      \inferrule*
+      { }
+      {\Delta \vdash \epsilon \isCtxt}
+
+      \inferrule*
+      {\Delta \vdash \Gamma \isCtxt \\ \Delta \vdash A \isType \\ x \not\in \Gamma}
+      {\Delta \vdash \Gamma, x : A \isCtxt}
+
       \inferrule* [right=Var]
@@
 
-  \caption{Well-typed Programs}
+  \caption{Well-indexed Contexts and Well-typed Programs}
   \label{fig:programs}

### 2012-07-06 paper/section-info-flow.tex
Updates to the logical information flow section

@@
 
-\begin{enumerate}
-\item Make sure the logic is at least a meet semi-lattice
-\item 
-\end{enumerate}
+\newcommand{\infoflow}{\mathit{Log}}
 
-\begin{enumerate}
-\item Also makes use of the relational aspect, the previous stuff
-  could make do with unary logical predicates as the interpretations
-  of types, but the information flow properties make use of the
-  relational interpretation.
-\item Can also obtain a type non-inhabitation result based on logical
-  derivability
-\end{enumerate}
+We now apply our general framework to types that are indexed by
+logical propositions. By including a primitive type that represents
+logical truth, we can recover--through a construction due to Tse and
+Zdancewic \cite{tse04translating}--strong information flow properties
+of programs. As a result of our general framework being parameterised
+by the choice of equational theory, we can alter the logic that we use
+for reasoning about type equality, and hence alter the information
+flow properties of the system.
 
-\fixme{
-  \begin{enumerate}
-  \item Make the definition of control of information flow clearer
-  \item Make the example a bit more compelling
-  \item Put in some examples of composite principals formed from
-    boolean logic, and the use of boolean algebra to do reasoning in
-    the types. List the laws of boolean algebra, and the laws of
-    Heyting algebra. Need to explain the difference between 
-  \item Possibly use singleton types to be able to do more interesting
-    security policies?
-  \item Put in the formal statement of the information flow property
-    and the proof.
-  \end{enumerate}
-}
+We first recall the concept of information flow. A function $f : A
+\times B \to C$ is said to not allow information to flow from its
+second argument to the output if for all $b,b' \in B$ and all $a \in
+A$, $f(a,b) = f(a,b')$. If we think of the $B$ argument as
+representing high-security information, then we have stated that $f$
+does not allow the high-security input to flow to the low security
+output. Information flow can be seen as a kind of invariance property
+of programs, and so our relational interpretation of types is well
+tailored to proving this kind of property.
 
-An extreme example of invariance under change of representation is the
-tracking of information flow through programs. If a program does not
-depend upon a particular piece of information, then it will be
-invariant under \emph{all} changes of representation of this
-information. As thoroughly described by Sabelfeld and Sands
+As was thoroughly described by Sabelfeld and Sands
 \cite{sabelfeld01per}, information flow can be captured semantically
-through the use of partial equivalence relations (PERs). Interpreting
-types as PERs forms an instance of Reynolds' approach to abstraction
-and representation independence. Abadi, Banerjee, Heintze and Riecke
-\cite{abadi99core} built a \emph{Core Calculus for Dependency}, using
-a type system based around a security level indexed monad $T_lA$. Tse
-and Zdancewic \cite{tse04translating} showed that it is possible to
-translate Abadi et al.'s calculus into System F, translating the
-monadic type $T_lA$ to $\alpha_l \to A$ for some free type variable
-$\alpha_l$, and making direct use of Reynolds' abstraction theorem to
-prove information flow properties. We now show how a refined variant
-of Tse and Zdancewic's translation can be expressed via
-algebraically-indexed types.
+through the use of partial equivalence relations (PERs). Abadi,
+Banerjee, Heintze and Riecke \cite{abadi99core} built a \emph{Core
+  Calculus for Dependency}, using a type system based around a
+security level indexed monad $T_lA$, using PERs to prove the
+information flow properties. Tse and Zdancewic \cite{tse04translating}
+showed that it is possible to translate Abadi et al.'s calculus into
+System F, translating the monadic type $T_lA$ to $\alpha_l \to A$ for
+some free type variable $\alpha_l$, and making direct use of Reynolds'
+abstraction theorem to prove information flow properties. For example,
+if we use the type variable $\alpha_H$ to represent high-level
+information, then the non-interference property of the function $f$
+above could be expressed by the following System F type:
+\begin{displaymath}
+  A \to (\alpha_H \to B) \to C
+\end{displaymath}
+If a program cannot generate a value of type $\alpha_H$, then it
+cannot access the value of $B$, and hence is insensitive to the actual
+value. Relationships between security levels are captured by
+postulating functions $\alpha_{l_1} \to \alpha_{l_2}$ whenever $l_1$
+is a lower security level than $l_2$.
 
-We assume a single indexing sort $\mathsf{principal}$, intended to
-represent (possibly composite) principals in a system. Principals are
-combined using the connectives of boolean logic ($\top$, $\bot$,
-$\lor$, $\land$, $\lnot$). For the equational theory of principals, we
-assume all the axioms of boolean algebra, and semantically we
-interpret composite principals as boolean values: either true ($\top$)
-or false ($\bot$). We introduce a single primitive type, indexed by
-expressions of sort $\mathsf{principal}$, $\tyPrim{T}{e}$, with the
-following interpretations:
+By making use of algebraically indexed types, we are able to refine
+Tse and Zdancewic's translation by replacing the use of type variables
+$\alpha_l$ with a primitive type $\tyPrim{T}{e}$ of representations of
+the truth of a logical proposition $e$. The relationships between
+security levels are now replaced by logical entailment, with the
+consequence that we only have functions of type $\tyPrim{T}{e_1} \to
+\tyPrim{T}{e_2}$ when $e_1$ entails $e_2$. Inspired by the PER
+interpretation of information flow mentioned above, we instantiate our
+general relational framework to interpret $\tyPrimNm{T}{e}$ as the
+identity relation when $e$ is true, and the empty relation when $e$ is
+false. In doing so, we can recover the information flow properties of
+Abadi et al. and Tse and Zdancewic.
+
+\paragraph{Instantiation of the General Framework}
+For this instantiation of our framework, we assume a single indexing
+sort $\mathsf{prop}$. For now, we assume that this sort supports
+meet-semilattice structure: we have a binary operation $\land$ and a
+constant $\top$, where $\land$ is commutative, associative and has
+$\top$ as a unit, and $\land$ is idempotent: so $p \land p =
+p$. Recall that we can define a partial order structure on any
+meet-semilattice by setting $p \leq q$ when $p = p \land q$. We take
+an arbitrary meet-semilattive $L$ for our chosen model.
+
+% Below, we will consider additional structure on the sort
+% $\mathsf{prop}$ and the model $L$
+
+We assume a single primitive type $\tyPrimNm{T}$, with
+$\primTyArity(\tyPrimNm{T}) = [\mathsf{prop}]$ and index-erasure
+semantics $\tyPrimSem{\tyPrimNm{X}} = \{ * \}$. Thus values of type
+$\tyPrimNm{T}{e}$ have no real run-time content; their meaning is
+given by the relational semantics. For the relational environments
+$\relEnv{E}_\infoflow$, we use the general construction in
+\autoref{sec:constr-rel-env}, with our chosen model $L$ and the
+following relational interpretations:
 \begin{displaymath}
-  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\tyPrim{T}{e}} & \{*\} \\
-    \rsem{\tyPrim{T}{e}}{}\rho & \left\{
-      \begin{array}{ll}
-        \{(*,*)\} & \textrm{if }\sem{e}\rho = \top \\
-        \{\}      & \textrm{if }\sem{e}\rho = \bot
-      \end{array}
-      \right.
+  \begin{array}{l@{\hspace{2em}}l}
+    R^\bullet_{\tyPrimNm{T}} = \emptyset &
+    R_{\tyPrimNm{T}}(e) = \{ (*,*) \sepbar e = \top \}
   \end{array}
 \end{displaymath}
-The relational interpretation of $\tyPrim{T}{e}$ forces this type to
-only be inhabited when the boolean expression $e$ evaluates to true
-under the environment $\rho$. To write programs, we assume the
-following primitive operations:
+The primitive operations $\Gamma_\infoflow$ for this instantiation
+reflect the meet-semilattice structure of the indexes:
 \begin{displaymath}
@@
     \mathrm{truth} : \tyPrim{T}{\top} &
-    \mathrm{and}   : \forall p, q\mathord:\mathsf{prn}.\ \tyPrim{T}{p} \to \tyPrim{T}{q} \to \tyPrim{T}{p \land q}
+    \mathrm{and}   : \forall p, q\mathord:\mathsf{prop}.\ \tyPrim{T}{p} \to \tyPrim{T}{q} \to \tyPrim{T}{p \land q}
   \end{array} \\
-  \mathrm{coerce} : \forall p, q\mathord:\mathsf{prn}.\ \tyPrim{T}{p \land q} \to \tyPrim{T}{p}
+  \mathrm{up} : \forall p, q\mathord:\mathsf{prop}.\ \tyPrim{T}{p \land q} \to \tyPrim{T}{p}
 \end{array}
 \end{displaymath}
-We can now adapt Tse and Zdancewic's translation of Abadi et al.'s
-monadic type to our setting. We define a type synonym, for each type
-$A$ and expression $e$ of sort $\mathsf{principal}$:
+The combination of the \TirName{TyEq} rule and the primitive
+$\mathrm{up}$ operation allow for logical entailment to be reflected
+in programs. Each of the primitive operations has a trivial
+interpretation, due the index-erasure interpretation of
+$\tyPrim{T}{e}$ as a one-element set, giving an environment
+$\eta_\infoflow \in \ctxtSem{\Gamma}$. Less trivially, we have this
+lemma:
+\begin{lemma}\label{lem:environments-information-flow}
+  For all $\Delta$ and $\rho \in \relEnv{E}_\infoflow$, $(\eta_\infoflow, \eta_\infoflow) \in \rsem{\Gamma_\infoflow}{\relEnv{E}_\infoflow}{\rho}$.
+\end{lemma}
+
+It is also possible to think of more complicated logics, as long as
+they include
+
+\begin{theorem}
+  Inexpressibility result
+\end{theorem}
+
+\paragraph{Application to Information Flow}
+We now think of logical expressions as ``composite principals''. That
+is, propositional variables represent various principals that are
+combined using the logical connectives. We interpret ``truth'' for
+principals as stating that that principal is satisfied with the
+current state of affairs, and that
+
+We adapt Tse and
+Zdancewic's translation of Abadi et al.'s monadic type to our
+setting. We define a type synonym, for each type $A$ and expression
+$e$ of sort $\mathsf{principal}$:
 \begin{displaymath}
@@
 
+For any extension of the logical theory...
+
 \begin{example}
-  Suppose we have a program with the following type:
-  \begin{displaymath}
-    \forall p,q\mathord:\mathsf{principal}. \tyPrim{T}{e} \to T_p(\tyPrimNm{unit} + \tyPrimNm{unit}) \to T_q(\tyPrimNm{unit} + \tyPrimNm{unit})
-  \end{displaymath}
-  where $e$ is some boolean algebra expression involving $p$ and
-  $q$. We will show in \autoref{sec:instantiations} that programs with
-  this type are non-constant (i.e.~depend on their input) if and only
-  if the formula $e \Rightarrow q \Rightarrow p$ is valid in boolean
-  logic.
+  Different notions of entailment lead to different notions of
+  non-interference.
 \end{example}
 
-\begin{enumerate}
-\item Treat the information flow thing as a type isomorphism, in the
-  same way that Tse and Zdancewic do
-  \begin{displaymath}
-    \begin{array}{l}
-    \forall \vec{p}\mathord:\mathsf{principal}.\ T_{e_1}\tyPrimNm{bool} \to T_{e_2}\tyPrimNm{bool}\\
-    \hspace{4em}\cong \\
-    \left\{
-      \begin{array}{ll}
-        \tyPrimNm{bool} \tyArr \tyPrimNm{bool} & \textrm{if }e_2 \vdash e_1 \\
-        \tyPrimNm{bool} & \textrm{if }e_2 \not\vdash e_1
-      \end{array}
-    \right.
-  \end{array}
-  \end{displaymath}
-  where $\tyPrimNm{bool} = \tyUnit + \tyUnit$.
-\item mention the issue with bogus tokens if we allow for a
-  call-by-name style of non-termination
-\item if we axiomatise intuitionistic logic instead of
-  boolean/classical logic, we get a different type isomorphism.
-\item The isomorphism relies upon $e_1 \vdash e_2 \Leftrightarrow e_1
-  \lor e_2 = e_1$.
-\item Relevant logic? Look at the Wikipedia page to see how to do
-  residuated (commutative) monoids on lattices with just
-  equations. With this, we will be able to do substructural logics
-  too. As long as we know that a particular axiomatisation is sound
-  and complete for some reasoning system, then we can state and prove
-  the non-interference isomorphism above.
-\item Compare to Tse and Zdancewic's addition of axioms to the logic.
-\item Internalising equality? The logical structure can then be used
-  on types as well as everything else. Need the special rules for
-  equality... (see the Fomega paper).
-\end{enumerate}
+\begin{theorem}
+  Type non-inhabitation result
+\end{theorem}
+
+\begin{theorem}
+  Type isomorphism theorem, making sure to emphasi
+\end{theorem}
+
+% In the previous two sections, we looked at algebraically-indexed types
+% whose relational interpretations used functional relations, with the
+% consequence that the derived free theorems express invariance
+% properties.
+
+% In this section we examine a type
+
+% We now look at another application of algebraically indexed types that
+% does not involve the kinds of functional relationships between values
+% that we have looked at the previous two sections. Rather, we use the
+% relational interpretation to restrict the elements of types
+
+% \begin{enumerate}
+% \item Make sure the logic is at least a meet semi-lattice. We have a
+%   single indexing sort $\mathsf{prn}$. There is a binary operation
+%   $\land$ which is associative and commutative and has unit $\top$,
+%   and also the binary operation is idempotent: $p \land p = p$.
+% \item Have a single primitive type $\tyPrimNm{T}$, with
+%   $\primTyArity(\tyPrimNm{T}) = [\mathsf{prn}]$. The index-erasure
+%   semantics of this type interprets it simply as 
+% \end{enumerate}
+
+% \begin{enumerate}
+% \item Also makes use of the relational aspect, the previous stuff
+%   could make do with unary logical predicates as the interpretations
+%   of types, but the information flow properties make use of the
+%   relational interpretation.
+% \item Can also obtain a type non-inhabitation result based on logical
+%   derivability
+% \end{enumerate}
+
+% \fixme{
+%   \begin{enumerate}
+%   \item Make the definition of control of information flow clearer
+%   \item Make the example a bit more compelling
+%   \item Put in some examples of composite principals formed from
+%     boolean logic, and the use of boolean algebra to do reasoning in
+%     the types. List the laws of boolean algebra, and the laws of
+%     Heyting algebra. Need to explain the difference between 
+%   \item Possibly use singleton types to be able to do more interesting
+%     security policies?
+%   \item Put in the formal statement of the information flow property
+%     and the proof.
+%   \end{enumerate}
+% }
+
+% An extreme example of invariance under change of representation is the
+% tracking of information flow through programs. If a program does not
+% depend upon a particular piece of information, then it will be
+% invariant under \emph{all} changes of representation of this
+% information. As thoroughly described by Sabelfeld and Sands
+% \cite{sabelfeld01per}, information flow can be captured semantically
+% through the use of partial equivalence relations (PERs). Interpreting
+% types as PERs forms an instance of Reynolds' approach to abstraction
+% and representation independence. Abadi, Banerjee, Heintze and Riecke
+% \cite{abadi99core} built a \emph{Core Calculus for Dependency}, using
+% a type system based around a security level indexed monad $T_lA$. Tse
+% and Zdancewic \cite{tse04translating} showed that it is possible to
+% translate Abadi et al.'s calculus into System F, translating the
+% monadic type $T_lA$ to $\alpha_l \to A$ for some free type variable
+% $\alpha_l$, and making direct use of Reynolds' abstraction theorem to
+% prove information flow properties. We now show how a refined variant
+% of Tse and Zdancewic's translation can be expressed via
+% algebraically-indexed types.
+
+% We assume a single indexing sort $\mathsf{principal}$, intended to
+% represent (possibly composite) principals in a system. Principals are
+% combined using the connectives of boolean logic ($\top$, $\bot$,
+% $\lor$, $\land$, $\lnot$). For the equational theory of principals, we
+% assume all the axioms of boolean algebra, and semantically we
+% interpret composite principals as boolean values: either true ($\top$)
+% or false ($\bot$). We introduce a single primitive type, indexed by
+% expressions of sort $\mathsf{principal}$, $\tyPrim{T}{e}$, with the
+% following interpretations:
+% \begin{displaymath}
+%   \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
+%     \tySem{\tyPrim{T}{e}} & \{*\} \\
+%     \rsem{\tyPrim{T}{e}}{}\rho & \left\{
+%       \begin{array}{ll}
+%         \{(*,*)\} & \textrm{if }\sem{e}\rho = \top \\
+%         \{\}      & \textrm{if }\sem{e}\rho = \bot
+%       \end{array}
+%       \right.
+%   \end{array}
+% \end{displaymath}
+% The relational interpretation of $\tyPrim{T}{e}$ forces this type to
+% only be inhabited when the boolean expression $e$ evaluates to true
+% under the environment $\rho$. To write programs,
+
+% \begin{example}
+%   Suppose we have a program with the following type:
+%   \begin{displaymath}
+%     \forall p,q\mathord:\mathsf{principal}. \tyPrim{T}{e} \to T_p(\tyPrimNm{unit} + \tyPrimNm{unit}) \to T_q(\tyPrimNm{unit} + \tyPrimNm{unit})
+%   \end{displaymath}
+%   where $e$ is some boolean algebra expression involving $p$ and
+%   $q$. We will show in \autoref{sec:instantiations} that programs with
+%   this type are non-constant (i.e.~depend on their input) if and only
+%   if the formula $e \Rightarrow q \Rightarrow p$ is valid in boolean
+%   logic.
+% \end{example}
+
+% \begin{enumerate}
+% \item Treat the information flow thing as a type isomorphism, in the
+%   same way that Tse and Zdancewic do
+%   \begin{displaymath}
+%     \begin{array}{l}
+%     \forall \vec{p}\mathord:\mathsf{principal}.\ T_{e_1}\tyPrimNm{bool} \to T_{e_2}\tyPrimNm{bool}\\
+%     \hspace{4em}\cong \\
+%     \left\{
+%       \begin{array}{ll}
+%         \tyPrimNm{bool} \tyArr \tyPrimNm{bool} & \textrm{if }e_2 \vdash e_1 \\
+%         \tyPrimNm{bool} & \textrm{if }e_2 \not\vdash e_1
+%       \end{array}
+%     \right.
+%   \end{array}
+%   \end{displaymath}
+%   where $\tyPrimNm{bool} = \tyUnit + \tyUnit$.
+% \item mention the issue with bogus tokens if we allow for a
+%   call-by-name style of non-termination
+% \item if we axiomatise intuitionistic logic instead of
+%   boolean/classical logic, we get a different type isomorphism.
+% \item The isomorphism relies upon $e_1 \vdash e_2 \Leftrightarrow e_1
+%   \lor e_2 = e_1$.
+% \item Relevant logic? Look at the Wikipedia page to see how to do
+%   residuated (commutative) monoids on lattices with just
+%   equations. With this, we will be able to do substructural logics
+%   too. As long as we know that a particular axiomatisation is sound
+%   and complete for some reasoning system, then we can state and prove
+%   the non-interference isomorphism above.
+% \item Compare to Tse and Zdancewic's addition of axioms to the logic.
+% \item Internalising equality? The logical structure can then be used
+%   on types as well as everything else. Need the special rules for
+%   equality... (see the Fomega paper).
+% \end{enumerate}
 

### 2012-07-08 paper/section-instantiations.tex
Geometric isomorphisms

@@
 \label{sec:instantiations}
+We now instantiate the general framework of the previous section with the indexing theory
+of \autoref{sec:motivating-examples}, and present
+more general and formally-justified theorems for free, type isomorphisms, and non-definability
+results.
 
-\fixme{explain the general setup, with a set of operations for each
-  instantiation.}
-
-\begin{example}[Two-dimensional Geometry] We define the ``standard''
+%\begin{example}[Two-dimensional Geometry] 
+We define the ``standard''
   model of the indexing theory for the two-dimensional geometry
@@
   polymorphic zero constants we have assumed.
-\end{example}
+%\end{example}
 
-\begin{lemma}
-  The thing about $\Gamma_\Geom$.
-\end{lemma}
+%\begin{lemma}
+%  The thing about $\Gamma_\Geom$.
+%\end{lemma}
 
@@
 
-\subsection{Isomorphisms for Types Indexed by Abelian Groups}
+\subsection{Type Isomorphisms}
 \label{sec:types-indexed-abelian-groups}
+We say that two types $A$ and $B$ are isomorphic, and write $A \cong B$, if there exist maps
+between them that are mutually inverse with respect to contextual equivalence. That is, 
+there are terms $\vdash I : A \to B$ and $\vdash J : B\to A$ such
+that
+\[
+x:A\vdash J(I(x))\stackrel{ctx}\approx x : A
+\text{ and }y:B\vdash I(J(y))\stackrel{ctx}\approx y : B.
+\]
+It is straightforward to prove that $\cong$ is a congruence relation on types. We
+can also derive isomorphisms that are independent of the indexing theory, such as
+$A\times B \cong B \times A$, and $\forall i\mathord:s.A\to B\cong A\to\forall i\mathord:s.B$ for $i$ not free in $A$.
 
-\begin{enumerate}
-\item Two cases: when we have a total ``division'' operation, and when
-  we don't.
-\item Type isomorphisms, based on Smith Normal Form
+Types indexed by abelian groups induce a particularly rich theory of
+type isomorphisms; previous work on units-of-measure~\cite{kennedy97relational} relates
+these to Buckingham's theorem from dimensional analysis. Here we consider the additive 
+abelian groups of translations and the multiplicative abelian group of scalings.
+
+\paragraph{Translations} 
+Consider first the group $\Transl{2}$ of translations as presented in
+\autoref{sec:type-system-geom-intro} and \autoref{sec:affine-vector-ops}. 
+Given any quantifier-free type $A$ we can prove the following isomorphism
+\begin{eqnarray*}
+\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{t}\to A
+& \cong &
+A[0/t]
+\end{eqnarray*}
+where $A[0/t]$ denotes substitution of the identity translation $0$
+for index variable $t$ in type $A$.  Intuitively, the isomorphism
+expresses the equivalence between \emph{coordinate-free} geometry, as
+expressed by a polymorphic type whose first argument can be thought of
+as representing the \emph{origin} with respect to which $A$ is
+specified, and a \emph{coordinate-based} geometry, as expressed by the
+type $A[0/t]$, in which the origin is fixed to be $(0,0)$. (It is sometimes said
+that an affine space is a vector space that's forgotten its origin; we have captured
+this in a type isomorphism.)
+
+\exref{ex:type-iso} is a special case of this isomorphism; another instance is the
+type of vector addition:
+\begin{eqnarray*}
+\forall t_1,t_2\mathord:\SynTransl{2}.\tyPrim{vec}{t_1}\to\tyPrim{vec}{t_2}\to\tyPrim{vec}{t_1+t_2}\\
+\cong\quad
+\forall t_2\mathord:\SynTransl{2}.\tyPrim{vec}{t_2}\to\tyPrim{vec}{t_2}\quad\cong\quad
+\tyPrim{vec}{0}
+\end{eqnarray*}
+
+The maps associated with this isomorphism are as follows:
+\newcommand{\trans}[3]{\mathrm{tr}_{#1}^{{#2}/{#3}}}
+\begin{eqnarray*}
+I = \lambda x. x\ [0]\ 0 & \text{ and } &
+J = \lambda y. \Lambda t.\ \lambda v. \trans A v t(y)
+\end{eqnarray*}
+where $\mathrm{tr}$ is defined as follows:
+\begin{eqnarray*}
+\trans{1} v t(x) & = & x\\
+\trans{A_1\times A_2} v t(p) & = & (\trans{A_1} v t(\pi_1\ p),\trans{A_2} v t(\pi_2\ p))\\
+\trans{A_1 + A_2} v t(x) & = & \textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\trans{A_1} v t(y)); \textrm{inr}\ y.\textrm{inr}(\trans{A_2} v t(y))\\
+\trans{A_1\to A_2} v t(f) & = & \lambda x. (\trans{A_2} v t(f\ (\trans{A_1}{-v} t(x))))\\
+\trans{\tyPrim{vec}{e}} v t (x) & = & x + kv\text{ where }\vdash e \equiv k t + e' : \SynTransl{2}\text{ for some }k\in\setOfIntegers, t \text{ and }e'\\
+\end{eqnarray*}
+To show $I(J(y))$ is contextually equivalent to $y$ is an easy induction on $A$ that
+makes use of the fact that the zero vector $0$ is an identity for vector addition. 
+To show $J(I(x))$ is contextually equivalent to $x$ we must apply the abstraction theorem,
+making use of the fact that $\mathrm{tr}$ is actually a term-based functional representation
+of an instance of the relational interpretation:
+\[
+\rsem A {t \mapsto v} {} = \{ (x, \tmSem{\trans A v t (x)}) \mid x \in \tySem A\}.
+\]
+
+\paragraph{Scalings}
+Now consider the group $\GL{1}$ of scalings. This time, we can treat the real-valued
+argument to a function as a `unit of measure' with which to scale the result. The small
+fly in the ointment is that the argument might be zero, and this cannot be used for scaling.
+Nevertheless, we can obtain the following slightly more complicated isomorphism:
+\begin{eqnarray*}
+\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A
+& \cong &
+A[1/s]\times \forall s\mathord:\SynGL{1}.A
+\end{eqnarray*}
+The type of $\mathrm{areaCircle}$ from the introduction is one instance:
+\begin{eqnarray*}
+\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s\cdot s}\\
+\quad\cong\quad \tyPrim{real}{1}\times(\forall s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s})\\
+\quad\cong\quad \tyPrim{real}{1}\times 1\quad\cong\quad \tyPrim{real}{1}
+\end{eqnarray*}
+where the isomorphism $\forall s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s}\cong 1$ is
+proved using the abstraction theorem.
+
+%\begin{enumerate}
+%\item Two cases: when we have a total ``division'' operation, and when
+%  we don't.
+%\item Type isomorphisms, based on Smith Normal Form
 % \item Non-definability results, based on the non-compositional
@@
 %   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
-\end{enumerate}
-
-\fixme{Copy the stuff over from the other file}
+%\end{enumerate}
 

### 2012-07-09 paper/geometry-twocolumn.tex
Tighten up the general framework section
Patty's changes

--- 2012-06-14 paper/geometry-twocolumn.tex
@@
 
-We motivate this paper's goal of investigating algebraically-indexed
-types and their relational interpretations by presenting a novel type
-system for writing programs that manipulate two-dimensional geometric
-data. Geometry is rich with operations that are invariant under
-transformation: affine operations are invariant under change of
-origin, vector space operations are invariant under change of basis,
-and the Euclidean dot product is invariant under orthogonal changes of
-basis. Moreover, some operations are interestingly variant under
-transformations: for example, the cross product varies with scalings
-of the plane. We incorporate this extra information about our
-geometric primitives into the type system by means of algebraically
-indexed types.
+We motivate our investigation of algebraically indexed types and their
+relational interpretations by developing a novel type system for
+programs that manipulate two-dimensional geometric data. Geometry is
+rich with operations that are invariant under transformation: affine
+operations are invariant under change of origin, vector space
+operations are invariant under change of basis, and dot product is
+invariant under orthogonal changes of basis. On the other hand, some
+geometric operations are interestingly variant under transformation.
+For example, cross products vary with scalings of the plane. We
+can incorporate (in)variance information about geometric primitives
+into type systems via algebraically indexed types.
 
-\subsection{Origin Invariance and Representation Independence}
+\subsection{Origin Invariance and Representation Independence}\label{sec:oiri}
 
-When writing programs that manipulate geometric data, the basic data
-structure is the $n$-tuple of numbers. In the 2-dimensional case,
-tuples $\vec{v} = (x,y)$ are called upon to represent both
-\emph{points}---offsets from some origin---and
-\emph{vectors}---offsets in their own right. Points and vectors are
-very different despite their common representation.
+The $n$-tuple of numbers is the basic data structure for programs that
+manipulate geometric data. Pairs, i.e., $2$-tuples, 
+%The basic data structure used in programs that manipulate geometric data is
+%the $n$-tuple of numbers. In the 2-dimensional case, tuples 
+$\vec{v} = (x,y)$ serve double duty, representing
+%are called upon to represent both 
+both \emph{points}---offsets from some origin---and
+\emph{vectors}---offsets in their own right.  Despite their common
+representation, points and vectors are very different, and
+distinguishing between them is the key feature of \emph{affine
+  geometry} (see, for example, Chapter 2 of
+\cite{gallier11geometric}). Nevertheless, computational geometry
+libraries traditionally either leave it to the programmer to maintain
+the distinction between points and vectors, or else use different
+abstract types for points and vectors to enforce it.
+%(one might say that the standard mathematical formulation of affine
+%spaces uses the abstract types approach). 
+In this paper we investigate a more sophisticated approach based on 
+types indexed by change of origin transformations.
 
-Distinguishing between points and vectors is the key feature of
-\emph{affine geometry} (see, for example, Gallier's book
-\cite{gallier11geometric}, Chapter 2). Traditionally in computational
-geometry libraries, the distinction between points and vectors has
-either been left to the programmer to maintain by themselves, or has
-been enforced through the use of separate abstract types for points
-and vectors (one might say that the standard mathematical formulation
-of affine spaces uses the abstract types approach). Here, we
-investigate a more sophisticated approach using types indexed by
-change of origin transformations, and think of the difference between
-points and vectors as a change of data representation.
-
-Putting this in concrete terms, if we have two origins, $(0,0)$ and
-$(10,20)$, then the tuple $(1,1)$ with respect to the first origin and
-$(11,21)$ with respect to the second origin represent the \emph{same}
-point because they have the same displacement from their respective
-origins. It is merely an artifact of our representation of points as
-pairs of numbers that these appear to be different. When we write
-programs that manipulate data that represents points, we ought to
-ensure that our programs are invariant with respect to changes in the
-choice of origin. On the other hand, vectors are not intended to be
-regarded as invariant under change of origin. Vectors represent
-offsets and the vector $(0,0)$ always represents the zero offset.
+This approach regards the difference between points and vectors as a
+change of data representation. For example, if $(0,0)$ and $(10,20)$
+are two origins, then the tuple $(1,1)$ with respect to $(0,0)$ the
+tuple $(11,21)$ with respect to $(10,20)$ represent \emph{the same
+  point} because they have the same displacement from these two
+origins, respectively.
+%It is merely an artifact of the representation of points as pairs of
+%numbers that they appear to be different.
+This suggests that programs that manipulate %data representing points
+should be invariant with respect to changes in the choice of
+origin. Program sthat manipulate vectors, on the other hand, should
+not be invariant under change of origin. Different vectors represent
+different offsets, and the vector $(0,0)$ always represents the zero
+offset.
 
@@
 Reynolds' fable about two professors teaching the theory of complex
-numbers \cite{reynolds83types}. One professor chooses rectangular
-coordinates ($x + iy$), while the other chooses polar coordinates
-($\alpha\cos\theta + i\alpha\sin\theta$). Happily, after learning the
-basic operations on complex numbers in the two representations, the
-two classes can interact because the theory of complex numbers is
-invariant with respect to the choice of representation. Reynolds
-formalises the idea of invariance with respect to changes of
-representation as preservation of relations. If a binary relation $R$
-relates two representations of a concept---for example the rectangular
-and polar representations of the same complex number are
-related---then a program that manipulates complex numbers at a level
-of abstraction above their specific representation should preserve the
-relation $R$. For example, let $f$ be a program taking complex numbers
-to complex numbers that respects invariance under change of
-representation. Then if $c$ is a complex number in rectangular form,
-and $c'$ is a complex number in polar form such that $R$ relates them,
-which we write as $(c,c') \in R$, then it is the case that $(f(c),
-f(c')) \in R$.
+numbers \cite{reynolds83types}. One professor represents complex
+numbers using rectangular coordinates ($x + iy$), while the other
+represents them using polar coordinates ($\alpha\cos\theta +
+i\alpha\sin\theta$). Happily, after learning the basic operations on
+complex numbers in the two representations, the two classes can
+interact because the theory of complex numbers is invariant under the
+choice of representation. Reynolds formalises the idea of invariance
+under changes of representation as preservation of relations.  For
+example, if a binary relation $R$ relates the rectangular and polar
+representations of complex numbers, then a program that manipulates
+complex numbers at a level of abstraction above their specific
+representation should preserve $R$. %More precisely, if
+%$f$ is a program that is invariant under the choice of representation
+%of complex numbers,
+%%taking complex numbers to complex numbers that respects invariance
+%%under change of representation. Then 
+%$c$ is a complex number in rectangular form, $c'$ is a complex number
+%in polar form, and $R$ is a relation that relates the rectangluar and
+%polar representations of complex numbers, then $(c,c') \in R$ implies
+%$(f(c), f(c')) \in R$.
 
-We now apply Reynolds' relational approach to the geometric
-setting. We define a family of binary relations on $\mathbb{R}^2$
-which is indexed by changes of origin. Changes of origin are
-represented by vectors in $\mathbb{R}^2$, and form a group under
-addition. To emphasise the intended meaning as translations, we write
-the collection of all change of origin offsets as $\Transl{2}$. By
-quantifying over all changes of origin we will be able to ensure that
-our programs are invariant with respect to the particular choice of
-origin. The members of the $\Transl{2}$-indexed family of binary
-relations $\{ R_{\vec{t}} \subseteq \mathbb{R}^2 \times \mathbb{R}^2
-\}_{\vec{t} \in \Transl{2}}$ are defined as follows:
-\begin{displaymath}
-  R_{\vec{t}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{t} \}
-\end{displaymath}
-
-Now consider a function $f$ that, for example, takes two tuples in
+Reynolds' relational approach can be applied in the geometric setting
+to show how quantifying over all changes of origin ensures the
+invariance of programs under any particular choice of origin. For
+this, we first define a family of binary relations on $\mathbb{R}^2$
+that is indexed by changes of origin. Changes of origin are
+represented by vectors in $\mathbb{R}^2$, and form a group
+$\Transl{2}$ of translations under addition. The $\Transl{2}$-indexed
+family of binary relations $\{ R_{\vec{t}} \subseteq \mathbb{R}^2
+\times \mathbb{R}^2 \}_{\vec{t} \in \Transl{2}}$ is then defined by
+%\begin{displaymath}
+$R_{\vec{t}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{t} \}$.
+%\end{displaymath}
+We then consider a function $f$ that takes as input two tuples in
 $\mathbb{R}^2$ and returns a single tuple in $\mathbb{R}^2$. We intend
-that all these tuples represent points with respect to the same
-origin, but that the function itself should be invariant with respect
-to the choice of origin. We use Reynolds' idea of preservation of
-relations to formalise this: for any $\vec{t} \in \Transl{2}$ (i.e.~any
-change of origin), the function $f$ should satisfy the following
-statement:
+that the tuples all represent points with respect to the same origin,
+and that $f$ is invariant under the choice of origin.  Reynolds'
+relational approach formalises this intention precisely: for any
+$\vec{t} \in \Transl{2}$: % (i.e.~any change of origin), the function
+%$f$
+%should satisfy the following statement:
 \begin{equation}\label{eq:f-preserve-rel-frame}
-  \forall (\vec{v_1},\vec{v'_1}) \in R_{\vec{t}}, (\vec{v_2},\vec{v'_2}) \in R_{\vec{t}}. (f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\vec{t}}
+  \forall (\vec{v_1},\vec{v'_1}) \in R_{\vec{t}},
+  (\vec{v_2},\vec{v'_2}) \in R_{\vec{t}}. (f(\vec{v_1}, \vec{v_2}),
+  f(\vec{v'_1}, \vec{v'_2})) \in R_{\vec{t}}
 \end{equation}
-Unfolding the definition of $R_{\vec{t}}$, this statement is
-equivalent to the following statement expressing invariance of $f$
-with respect to changes of origin: for all $\vec{t} \in \Transl{2}$:
+Unfolding the definition of $R_{\vec{t}}$ gives the equivalent formulation:
+%\statementref{eq:f-preserve-rel-frame} is equivalent to: 
+%for all $\vec{t} \in \Transl{2}$,
 \begin{equation}\label{eq:f-invariant-frame}
-  \forall \vec{v_1}, \vec{v_2}.\ f(\vec{v_1} + \vec{t},\vec{v_2} + \vec{t}) = f(\vec{v_1},\vec{v_2}) + \vec{t}.
+  \forall \vec{v_1}, \vec{v_2}.\ f(\vec{v_1} + \vec{t},\vec{v_2} +
+  \vec{t}) = f(\vec{v_1},\vec{v_2}) + \vec{t}.
 \end{equation}
-So Reynolds' preservation of relations, when instantiated with the
+Thus, Reynolds' preservation of relations, when instantiated with the
 family of relations $\{R_{\vec{t}}\}$, yields exactly the geometric
@@
 
-Reynolds' further insight was to see how preservation of relations by
-the denotational interpretations of programs can automatically be
-established by following a typing discipline. In Reynolds' case, the
-typing discipline was the polymorphic $\lambda$-calculus where new
-types can be constructed by universal quantification over all
-types. Universal type quantification is commonly written $\forall
-\alpha. A$, which denotes a type that may be instantiated with any
-other type $B$ by replacing the type variable $\alpha$ with $B$ in
-$A$.
-
+Reynolds also showed how a type discipline can be used to establish
+that (the denotational interpretations of) programs preserve
+relations. For Reynolds, the type discipline of interest was that of
+the polymorphic $\lambda$-calculus, which supports the construction of
+new types by universal quantification over types.
+%, in which new types can be
+%constructed by universal quantification over all types.
 In terms of relations, Reynolds interprets universal quantification
 over types as quantification over binary relations between denotations
-of types. In the statements of geometric invariance that we stated in
-the previous section (e.g., \statementref{eq:f-invariant-frame}), we
-did not quantify over all relations. Instead, we quantified over all
-changes of origin and used the choice to select a relation from the
-family $\{R_{\vec{t}}\}$. Driven by this observation, we introduce
-quantification over changes of origin into the language of types.  We
-use the notation $\forall t \mathord: \SynTransl{2}. A$ for
-quantification over all 2-dimensional translations (i.e.~choices of
-origin) $t$. We refer to $\SynTransl{2}$ as the \emph{sort} of
-$t$. Note that we have used a different font to distinguish the
-semantic group $\Transl{2}$ from the syntactic sort $\SynTransl{2}$.
+of types. By contrast, in our statements of geometric invariance in
+\autoref{sec:type-system-geom-intro} we did not quantify over all
+relations, but instead quantified over all changes of origin and used
+a specific choice of origin to select a relation from the family
+$\{R_{\vec{t}}\}$. This suggests introducing quantification over
+changes of origin into the language of types. We use the notation
+$\forall t \mathord: \SynTransl{2}. A$ for quantification over all
+2-dimensional translations (i.e.,~choices of origin) $t$, and refer to
+$\SynTransl{2}$ as the \emph{sort} of $t$. Note the difference in
+fonts used to distinguish the semantic group $\Transl{2}$ from the
+syntactic sort $\SynTransl{2}$. We use a similar convention
+below, too.
 
-The sort $\SynTransl{2}$ represents an abelian group, so we can combine
-its elements using the usual group operations. We write operations
-additively, using $e_1 + e_2$ for the group operation, $-e$ for
-inverse and $0$ for the unit.  We also regard expressions built from
-variables and the group operations up to the abelian group axioms. For
-example, $e_1 + (e_2 + e_3)$ and $(e_1 + e_2) + e_3$ are to be
-regarded as equivalent.
+Since the sort $\SynTransl{2}$ represents an abelian group, we can
+combine its elements using the usual group operations. We write
+operations additively, using $e_1 + e_2$ for the group operation, $-e$
+for inverse and $0$ for the unit.  We also regard expressions built
+from variables and the group operations up to the abelian group
+axioms. For example, we regard $e_1 + (e_2 + e_3)$ and $(e_1 + e_2) +
+e_3$ as equivalent.
 
-Our language of types also includes the common type constructors for
-function types, $A \to B$, sum types $A + B$, unit type
-$\tyPrimNm{unit}$ and tuple types $A \times B$. We also assume a
-primitive type $\tyPrimNm{real}$, used to represent scalars. As we
-noted at the start of the previous section, the central data structure
-in geometric applications is the tuple of real numbers for
-representing points and vectors. However, we cannot simply express
-this as the type $\tyPrimNm{real} \times \tyPrimNm{real}$ because it
-will not have the correct relational interpretation. (Two elements of
-type $\tyPrimNm{real}$ will be related if and only if they are equal,
-and hence two elements of $\tyPrimNm{real} \times \tyPrimNm{real}$
-will be related if and only if they are equal, by Reynolds' definition
-of the relational interpretation of tuple types.) Therefore, we
-introduce a new type indexed by expressions $e$ of sort
-$\SynTransl{2}$ to represent points \emph{and} vectors:
-$\tyPrim{vec}{e}$. The indexing argument $e$ represents the possible
-amount of displacement by change of origin this point may have been
-subjected to. We have used the name $\tyPrimNm{vec}$, even though we
-have taken pains to separate geometric points and vectors, to recall
-the computer science notion of vector as a sequence of values of
-homogeneous type which has a known length (in this case two).
+Our language of types includes the unit type $\tyPrimNm{unit}$ and,
+for all types $A$ and $B$, the function type $A \to B$, the sum type
+$A + B$, and the tuple type $A \times B$. We also assume a primitive
+type $\tyPrimNm{real}$, used to represent scalars. Although tuples of
+real numbers represent points and vectors in geometric applications,
+we cannot express this via the type
+%central data structure in geometric applications is the tuple of real
+%numbers for representing points and vectors, we cannot simply express
+%this as the type 
+$\tyPrimNm{real} \times \tyPrimNm{real}$. Indeed,
+%because this type does not have the correct relational
+%interpretation. (
+two elements of type $\tyPrimNm{real}$ are related if and only if they
+are equal and, by Reynolds' interpretation of tuple types, two
+elements of $\tyPrimNm{real} \times \tyPrimNm{real}$ are also related
+if and only if they are equal. But since this does not give the
+correct relational interpretations for points and vectors, we
+introduce a new type $\tyPrim{vec}{e}$, indexed by expressions $e$ of
+sort $\SynTransl{2}$, to represent them.  The index $e$ represents the
+displacement by change of origin of a point of this type.  Although we
+have taken pains to distinguish geometric points and vectors, we use
+the name $\tyPrimNm{vec}$ for both to recall the computer science
+notion of vector as a homogeneous sequence of values with a known
+length (in this case, 2).
 
-We now sketch the interpretation of the types $\tyPrim{vec}{e}$ and
-$\forall t\mathord:\SynTransl{2}.A$. In our calculus, each type has two
-interpretations: an index-erasure interpretation that ignores the
-indexing expression $e$, and a relational interpretation as a binary
-relation on the index-erasure interpretation. For now, we assume we
-are given some way of turning an expression $e$ of sort $\SynTransl{2}$
-into an actual element $\sem{e}\rho$ of the group $\Transl{2}$, under
-some environment $\rho$ which tells how to interpret the free
+As is standard for parametricity, every type has two interpretations:
+an index-erasure interpretation that ignores the indexing expression,
+and a relational interpretation as a binary relation on the
+index-erasure interpretation. To give such interpretations for the
+types $\tyPrim{vec}{e}$ and $\forall t\mathord:\SynTransl{2}.A$,
+we assume for now that we can map each expression $e$ of sort
+$\SynTransl{2}$ to an element $\sem{e}\rho$ of the group $\Transl{2}$
+using some environment $\rho$ that interprets $e$'s free
 variables. The (simplified) index-erasure and relation interpretations
-of the types $\tyPrim{vec}{e}$ are:
+of $\tyPrim{vec}{e}$ are:
 \begin{displaymath}
   \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2 \\
-    \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{e}\rho} = \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
+    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2
+    \\ \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{e}\rho} = \{
+    (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
   \end{array}
 \end{displaymath}
-The index-erasure interpetation of universal quantification $\forall
-t\mathord:\SynTransl{2}.A$ is simply the interpretation of the
-underlying type $A$. The relational interpretation intersects the
-relational interpretations of the type $A$ under all extensions of the
-relational environment $\rho$:
+The index-erasure and relational interpetations of 
+$\forall t\mathord:\SynTransl{2}.A$ are:
+%is simply the interpretation of the
+%underlying type $A$. The relational interpretation intersects the
+%relational interpretations of the type $A$ under all extensions of the
+%relational environment $\rho$:
 \begin{displaymath}
   \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\forall t\mathord:\SynTransl{2}.A} & \tySem{A} \\
-    \rsem{\forall t\mathord:\SynTransl{2}.A}{}\rho & \bigcap\{ \rsem{A}{}{\rho'} \sepbar \rho' \in \extends{\rho}{t\mathord:\SynTransl{2}} \}
+    \tySem{\forall t\mathord:\SynTransl{2}.A} & \tySem{A}
+    \\ \rsem{\forall t\mathord:\SynTransl{2}.A}{}\rho & \bigcap\{
+    \rsem{A}{}{\rho'} \sepbar \rho' \in
+    \extends{\rho}{t\mathord:\SynTransl{2}} \}
   \end{array}
@@
 depends on which (possibly non-compositional) interpretation of
-relational environments that we take. For now, it will suffice to
-think of relational environments as simply mappings from free
-variables of sort $\SynTransl{2}$ to the group $\Transl{2}$, and
-extension as just adding an extra component to the tuple. The general
-index-erasure semantics will be given in
-\autoref{sec:index-erasure-semantics}. The relational interpretation
-is in general more complex, due to the possibility of
-non-compositional interpretations of free index variables. The
-relational interpretation will be presented in
-\autoref{sec:relational-semantics}.
+relational environments we use. For now we can think of environments
+as mappings from free variables of sort $\SynTransl{2}$ to the group
+$\Transl{2}$, and of extension as adding extra components to
+mappings. The index-erasure and relational interpretations are given
+formally in Sections~\ref{sec:index-erasure-semantics}
+%the relational interpretation is in general more complex, due to the
+%possibility of non-compositional interpretations of free index
+%variables. The relational interpretation will be presented in
+and~\ref{sec:relational-semantics}. %, respectively.
 
-In the previous section, we discussed functions $f : \mathbb{R}^2
-\times \mathbb{R}^2 \to \mathbb{R}^2$ that have the property that they
-preserve all changes of origin. We can now express this property as a
-type:
-\begin{displaymath}
-  \mathrm{f} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}
-\end{displaymath}
+At the end of \autoref{sec:oiri} we considered functions $f :
+\mathbb{R}^2 \times \mathbb{R}^2 \to \mathbb{R}^2$ that preserve all
+changes of origin. This property of $f$ can be expressed in terms of
+types by
+%\begin{displaymath}
+$  \mathrm{f} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t}
+  \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}$.
+%\end{displaymath}
 Spelling out the relational interpretation of this type using the
-definitions above (and the standard relational interpretations for
-tuple and function types), we obtain
-\statementref{eq:f-invariant-frame}.
+definitions above and the standard relational interpretations for
+tuple and function types, we recover
+\statementref{eq:f-invariant-frame} exactly.
 
@@
 Invariance under change of origin is the key feature of affine
-geometry, and the central operation of affine geometry is the affine
-combination of points: $\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$,
-where $\lambda_1 + \lambda_2 = 1$. Geometrically, affine combination
-can be interpreted as interpolation between the points represented by
-$\vec{v_1}$ and $\vec{v_2}$. We add affine combination of points to
-our calculus, with the following typing and intended denotation:
+geometry, whose central operation is the affine combination of points:
+$\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$, where $\lambda_1 +
+\lambda_2 = 1$.  Geometrically, this can be interpreted as
+interpolation between the points represented by $\vec{v_1}$ and
+$\vec{v_2}$.  We add affine combination of points to our calculus as
+follows: % typing and intended denotation:for
 \begin{displaymath}
   \begin{array}{l}
-    \mathrm{affComb} :\forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\
-    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} + r\vec{v_2}
+    \mathrm{affComb} :\forall t \mathord:
+    \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to
+    \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\ 
+    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} +
+    r\vec{v_2} 
 \end{array}
 \end{displaymath}
-It can be verified by hand that the intended denotation
+It can be verified by hand that %the intended denotation
 $\sem{\mathrm{affComb}}$ has the invariance property specified by its
-type.
+type. {\bf Vague. Say which invariance property.}
 % By defining the primitive function $\mathrm{affComb}$ to take a
@@
   with two endpoints and a single control point) can be expressed
-  using the affine combination primitive and three steps of De
-  Casteljau's algorithm:
+  using the affine combination primitive as follows: % and three steps
+  % of De Casteljau's algorithm:
   \begin{displaymath}
     \begin{array}{@{}l}
-      \mathrm{quadB\acute{e}zier} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \\
-      \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\
-      \quad \mathrm{affComb}\ [t]\ (\mathrm{affComb}\ [t]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [t]\ p_1\ s\ p_2) \\
+      \mathrm{quadB\acute{e}zier} : \forall t
+      \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to
+      \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t}
+      \\ \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\ \quad
+      \mathrm{affComb}\ [t]\ (\mathrm{affComb}\ [t]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [t]\ p_1\ s\ p_2)
+      \\
     \end{array}
   \end{displaymath}
-  For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
-  application $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ for $s
-  \in [0,1]$ gives us the point on the curve at ``time'' $s$. Note
-  that the type of $\mathrm{quadB\acute{e}zier}$ tells us immediately
-  that it preserves all changes of origin.
+  For two endpoints $p_0$ and $p_2$, a control point $p_1$, and $s \in
+  [0,1]$, an application
+  $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ gives the point on
+  the curve at ``time'' $s$.  The type of
+  $\mathrm{quadB\acute{e}zier}$ immediately tells us that it preserves
+  all changes of origin.
 \end{example}
-
 % Affine combination only provides us with an operation on points. We
@@
 
-Vector operations are not invariant under change of origin. The
-obvious type for vector addition is therefore:
+The obvious type for vector addition is
+%\begin{displaymath}
+$  (+) : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}$.
+%\end{displaymath}
+%(we write binary operators intended to be used infix in parentheses
+%when not appearing in infix position, following the Haskell
+%syntax.)
+But we can reflect the fact that $(+)$ is not invariant under change
+of origin by giving it a more precise type that reflects how it varies
+with change of origin:
 \begin{displaymath}
-  (+) : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
-\end{displaymath}
-(we write binary operators intended to be used infix in parentheses
-when not appearing in infix position, following the Haskell
-syntax.)
-
-However, it is possible to give vector addition a more precise type
-that shows how it is variant with change of origin:
-\begin{displaymath}
-  (+) : \forall t_1, t_2 \mathord: \SynTransl{2}.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
+  (+) : \forall t_1, t_2 \mathord: \SynTransl{2}.\ \tyPrim{vec}{t_1}
+  \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
 \end{displaymath}
@@
 \begin{displaymath}
-  \mathrm{negate} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
+  \mathrm{negate} : \forall t \mathord:
+  \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
 \end{displaymath}
-With the primitive operations of addition and negation of vectors, we
-can define the derived operation of subtraction:
+Finally, with the primitive operations of addition and negation of
+vectors we can define the derived operation of subtraction:
 \begin{displaymath}
   \begin{array}{l}
-    (-) : \forall t_1,t_2 \mathord:\SynTransl{2}.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2} \\
-    (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
+    (-) : \forall t_1,t_2 \mathord:\SynTransl{2}.\ \tyPrim{vec}{t_1}
+    \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2}
+    \\ (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
   \end{array}
@@
 
-Given two points that have been constrained to be invariant with
-respect to the same change of origin---i.e.~two values of type
-$\tyPrim{vec}{t}$---we can compute their offset, which is a vector
-expressed with respect to the null change of origin. The offset
-operation is a special case of vector subtraction:
+Given two points that are invariant with respect to the same change of
+origin---i.e.~two values of type $\tyPrim{vec}{t}$---we can use
+subtraction to compute their offset:
 \begin{displaymath}
   \begin{array}{l}
-    \mathrm{offset} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{0} \\
-    \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
+    \mathrm{offset} : \forall t
+    \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to
+    \tyPrim{vec}{0} \\ \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
   \end{array}
 \end{displaymath}
-Note how the type of the result of $\mathrm{offset}$ simplifies from
-$\tyPrim{vec}{t-t}$ to $\tyPrim{vec}{0}$. The algebraic structure on
-the indexing theory induces type equalities.  The type of the vector
-addition operation can be specialised to the case of moving a point by
-a vector:
+The result is a vector expressed with respect to the null change of
+origin: note how the algebraic structure on the indexing theory
+induces type equalities that can be used to simplify the type of the
+result of $\mathrm{offset}$ from $\tyPrim{vec}{t-t}$ to
+$\tyPrim{vec}{0}$. The type of $(+)$ can also be specialised to the
+case of moving a point by a vector:
 \begin{displaymath}
   \begin{array}{l}
-    \mathrm{moveBy} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to \tyPrim{vec}{t} \\
-    \mathrm{moveBy}\ [t]\ p\ v = p + v
+    \mathrm{moveBy} : \forall t
+    \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to
+    \tyPrim{vec}{t} \\ \mathrm{moveBy}\ [t]\ p\ v = p + v
   \end{array}
 \end{displaymath}
-The types we assign to the remaining vector space primitives, the zero
-vector and multiplication by a scalar, do not describe any interesting
-effect on translations:
-\begin{displaymath}
-  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{4em}}l@{\hspace{0.5em}:\hspace{0.5em}}l}
-    \mathrm{0} & \tyPrim{vec}{0}&
-    (*) & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
-  \end{array}
-\end{displaymath}
+The types we assign to the remaining vector space primitives, namely
+$\mathrm{0} : \tyPrim{vec}{0}$ for
+the zero vector and $(*) : \tyPrimNm{real} \to
+    \tyPrim{vec}{0} \to \tyPrim{vec}{0}$ for
+multiplication by a scalar, do not describe any
+interesting effects on translations.
+%\begin{displaymath}
+%  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{4em}}l@{\hspace{0.5%em}:\hspace{0.5em}}l}
+%    \mathrm{0} & \tyPrim{vec}{0}& (*) & \tyPrimNm{real} \to
+%    \tyPrim{vec}{0} \to \tyPrim{vec}{0}
+%  \end{array}
+%\end{displaymath}
 
@@
   \begin{displaymath}
-    \tau_{n} \isDefinedAs \forall t\mathord:\SynTransl{2}.\ \underbrace{\tyPrim{vec}{t} \to ... \tyPrim{vec}{t}}_{n+1\textrm{ times}} \to \tyPrimNm{real}
+    \tau_{n} \isDefinedAs \forall
+    t\mathord:\SynTransl{2}.\ \underbrace{\tyPrim{vec}{t} \to ... \to
+      \tyPrim{vec}{t}}_{n+1\textrm{ times}} \to \tyPrimNm{real}
   \end{displaymath}
   Just by looking at the types $\tau_{n}$, we know that their
-  inhabitants will be invariant under change of origin, due to the
-  quantification over all $t$ in $\SynTransl{2}$. Therefore, we may as well pick
-  one of the input points as the origin and assume that all the other
-  points are defined with respect to this chosen origin. This is
-  similar to the common mathematical step of stating that, ``without
-  loss of generality'', we may pick some point in a description of a
-  problem to be the origin, as long as the problem statement is
-  invariant under translation.
-
-  The types $\tau_{n}$ are isomorphic to the corresponding types
-  $\sigma_{n}$:
+  inhabitants will be invariant under change of origin because of the
+  quantification over all $t$ in $\SynTransl{2}$. So we may as well
+  choose one of the input points as the origin and assume that all the
+  other points are defined with respect to it.  This formalises the
+  common mathematical practice of stating that ``without loss of
+  generality'' we can take some point in a description of a problem to
+  be the origin provided the problem statement is invariant under
+  translation. Each type $\tau_{n}$ is isomorphic to the coresponding
+  type $\sigma_{n}$:
   \begin{displaymath}
-    \sigma_{n} \isDefinedAs \underbrace{\tyPrim{vec}{0} \to ... \tyPrim{vec}{0}}_{n\textrm{ times}} \to \tyPrimNm{real}
+    \sigma_{n} \isDefinedAs \underbrace{\tyPrim{vec}{0} \to
+      ... \tyPrim{vec}{0}}_{n\textrm{ times}} \to \tyPrimNm{real}
   \end{displaymath}
-  We demonstrate this isomorphism formally in
-  \autoref{sec:instantiations}, in a more general setting of types
+  We demonstrate these isomorphisms formally in
+  \autoref{sec:instantiations}, in the more general setting of types
   indexed by abelian groups.
-
   % The isomorphism between these two families of types can be witnessed
@@
 \begin{example}\label{ex:uninhabited-type}
-  To this point we have emphasised the properties that we can deduce
-  about programs from their types, so called ``free theorems''. By
-  refining our relational interpretations of types appropriately it is
-  also possible to show that certain types are uninhabited. For
-  example, the type $\forall t \mathord: \SynTransl{2}.\
-  \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ does not have any
-  inhabitants. Intuitively, this is because we have no way of removing
-  the extra $t$ in $\tyPrim{vec}{t+t}$ using our operations. We
-  demonstrate this indefinability result in
-  \autoref{sec:types-indexed-abelian-groups}. In order to obtain this
-  result, we require a more sophisticated relational interpretation of
-  free translation variables than the one we sketched above.
+  So far we have emphasised the derivation of properties, or ``free
+  theorems'', of programs from their types. But using more refined
+  relational interpretations of types we can also show that certain
+  types are uninhabited. For example, the type $\forall t \mathord:
+  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ has no
+  inhabitants. Intuitively, this is because there is no way to remove
+  the extra $t$ in $\tyPrim{vec}{t+t}$ using the vector operations. We
+  formalise this indefinability result in
+  \autoref{sec:types-indexed-abelian-groups} using a more
+  sophisticated relational interpretation of free translation
+  variables than the one we sketched above.
 \end{example}
@@
 
-From basic linear algebra, we know that the vector operations of
-addition, negation and scaling, while not invariant under change of
-origin, are invariant under change of basis. Following the same
-methodology as we did for origin invariance, we can express basis
-invariance by means of preservation of relations indexed by changes of
-basis. Change of basis is accomplished by applying an invertible
-linear map, and the collection of all invertible linear maps on
-$\mathbb{R}^2$ forms the \emph{General Linear} group $\GL{2}$. We
-reflect in our language the group $\GL{2}$ as a new indexing sort
-$\SynGL{2}$, which has (non-abelian) group structure that we will
-write multiplicatively. We also extend the $\tyPrimNm{vec}$ type to be
-indexed by elements of $\SynGL{2}$ as well as $\SynTransl{2}$. Thus,
-$\tyPrim{vec}{B,t}$ is a vector which varies with change of basis $B$
+Although vector addition, negation, and scaling are not invariant
+under change of origin, they are invariant under change of basis. As
+with origin invariance, we can express basis invariance as
+preservation of relations indexed by changes of basis. Change of basis
+is achieved by applying an invertible linear map, and the collection of
+all such maps on $\mathbb{R}^2$ forms the \emph{General Linear} group
+$\GL{2}$ which we represent in our language by a new indexing sort
+$\SynGL{2}$ with (non-abelian) group structure that we will write
+multiplicatively. We then extend $\tyPrimNm{vec}$ to allow indices of
+sort $\SynGL{2}$, as well as $\SynTransl{2}$, so that
+$\tyPrim{vec}{B,t}$ is a vector that varies with change of basis $B$
 and change of origin $t$. Formally, the index-erasure and (simplified)
-relational semantics of this type are as follows:
+relational semantics of $\tyPrim{vec}{B,t}$ are given by:
 \begin{displaymath}
   \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2 \\
-    \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = (\sem{e_B}\rho)\vec{v} + \sem{e_t}\rho \}
+    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2
+    \\ \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'})
+    \sepbar \vec{v'} = (\sem{e_B}\rho)\vec{v} + \sem{e_t}\rho \}
   \end{array}
@@
 
-\paragraph{Affine Geometry} The combination of an invertible linear
-map and a translation is an \emph{affine transformation}. We can now
-assign types to all the primitive affine and vector space operations
-that indicates how they they behave with respect to affine
+\paragraph{Affine Geometry} An \emph{affine transformation} 
+an invertible linear map together with a translation. We can assign
+types to all the primitive affine and vector space operations
+indicating how they they behave with respect to affine
 transformations:
@@
     \forall B \mathord: \SynGL{2}, t \mathord: \SynTransl{2}.\\
-    \hspace{0.2cm} \tyPrim{vec}{B,t} \to \tyPrimNm{real} \to \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,t}
+    \hspace{0.2cm} \tyPrim{vec}{B,t} \to \tyPrimNm{real} \to
+    \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,t}
   \end{array}
@@
     \forall B \mathord: \SynGL{2}, t_1,t_2 \mathord: \SynTransl{2}. \\
-    \hspace{0.2cm}\tyPrim{vec}{B,t_1} \to \tyPrim{vec}{B,t_2} \to \tyPrim{vec}{B,t_1 + t_2}
+    \hspace{0.2cm}\tyPrim{vec}{B,t_1} \to \tyPrim{vec}{B,t_2} \to
+    \tyPrim{vec}{B,t_1 + t_2}
   \end{array}
@@
 \paragraph{Euclidean Geometry} Euclidean geometry extends affine
-geometry with the operation of \emph{dot product} (also known as the
-\emph{inner product}) of two vectors. The standard Euclidean dot
-product is defined on the coordinate representation in the following
-way:
-\begin{displaymath}
-  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2
-\end{displaymath}
+geometry with the \emph{dot product}, or \emph{inner product},
+operation of two vectors. The dot product is defined 
+%on the coordinate representation 
+by
+%\begin{displaymath}
+$  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2$.
+%\end{displaymath}
 % The dot product is used to define properties of vectors such as their
@@
 % if $\vec{v_1}\cdot\vec{v_2} = 0$.
-To incorporate the operation of dot product into our calculus, we must
-assign it a type. The dot product is not invariant under the action of
-either of the groups $\GL{2}$ and $\Transl{2}$ that we have considered
-so far. However, the dot product is invariant under the subgroup
-$\Orth{2}$ of $\GL{2}$ of \emph{orthogonal} linear transformations:
-invertible linear maps whose determinant is either $1$ or $-1$. We
-introduce orthogonal transformations as a new sort $\SynOrth{2}$, and
-overload the multiplicative group operations syntax for inhabitants of
-$\SynOrth{2}$. We also assume a function $\iota_O$ that takes $e :
-\SynOrth{2}$ to $\iota_O(e) : \SynGL{2}$. We may now assign a type to
-the dot product, using quantification over $\SynOrth{2}$:
+To %incorporate the operation of dot product into our calculus, we
+must assign it a type we note that, although dot product is not
+invariant under $\GL{2}$ or $\Transl{2}$, it
+% that we have considered so far. However,e  the dot product 
+is invariant under the subgroup $\Orth{2}$ of $\GL{2}$ of
+\emph{orthogonal} linear transformations, i.e., the subgroup of
+invertible linear maps whose matrix representations' transposes are
+equal to their inverses. We thus introduce a new sort $\SynOrth{2}$ of
+orthogonal transformations, and overload the multiplicative group
+operations for inhabitants of $\SynOrth{2}$. Further assuming a
+function $\iota_O$ that takes $e : \SynOrth{2}$ to $\iota_O(e) :
+\SynGL{2}$ we assign dot product this type:
+%a type to the dot product, using quantification over $\SynOrth{2}$:
 \begin{displaymath}
@@
 parallelogram described by the pair of input vectors.  Under change of
-basis by an invertible linear transformation $B$, the value of the
-cross product of two vectors varies with the determinant of $B$. This
-corresponds to the scaling of the plane by the change of basis
-transformation.
-
-We therefore augment our calculus with a new sort $\SynGL{1}$ of scale
-factors (i.e.~$1$-dimensional invertible linear maps). Semantically,
+basis by an invertible linear transformation $B$, the cross product of
+two vectors varies with the determinant of $B$. This corresponds to
+scaling the plane by the change of basis transformation, so we augment
+our calculus with a new sort $\SynGL{1}$ of scale factors
+(i.e.,~$1$-dimensional invertible linear maps). Semantically,
 $\SynGL{1}$ ranges over the non-zero real numbers and forms an abelian
-group, which we write multiplicatively. We also add two new
-operations: determinant, $\det B$, which takes an inhabitant of
-$\SynGL{2}$ to its determinant in $\SynGL{1}$, and the operation of
-absolute value that takes scaling factors to scaling factors which we
-write as $|e|$. We also refine the type $\tyPrimNm{real}$ of real
-numbers to be indexed by the sort $\SynGL{1}$: $\tyPrim{real}{e}$,
-where the old type $\tyPrimNm{real}$ is just $\tyPrim{real}{1}$. The
-collection of operations on real numbers indexed by scaling factors is
-shown in \autoref{fig:real-ops}.
-
-With the additional sort $\SynGL{1}$, and the refined type of real
-numbers, we can assign a type to the cross product:
+group which we write multiplicatively. We also add two new operations:
+determinant, $\det B$, which takes an inhabitant of $\SynGL{2}$ to its
+determinant in $\SynGL{1}$, and absolute value, $|e|$, which takes
+scaling factors to scaling factors. We also refine the type
+$\tyPrimNm{real}$ of real numbers so that it is indexed by the sort
+$\SynGL{1}$: $\tyPrim{real}{e}$. The old type $\tyPrimNm{real}$ is
+then just $\tyPrim{real}{1}$, and the full collection of operations on
+real numbers indexed by scaling factors is shown in
+\autoref{fig:real-ops}. We can thus assign cross product the type:
+%With the additional sort $\SynGL{1}$ and the refined type of real
+%numbers, we can assign a type to the cross product:
 \begin{displaymath}
-  (\times) : \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B}
+  (\times) : \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \to
+  \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B} 
 \end{displaymath}
@@
 \begin{example}\label{ex:area-of-triangle-1}
-  Using the operations we have defined in this subsection, we can
-  compute the area of a triangle. 
+  We can use the operations of this subsection to
+  compute the area of a triangle. We have:
   \begin{displaymath}
     \begin{array}{@{}l}
-      \mathrm{area} : \forall B\mathord:\SynGL{2}, t\mathord:\SynTransl{2}.\ \\
-      \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{|\det B|} \\
-      \mathrm{area}\ [B]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
+      \mathrm{area} : \forall B\mathord:\SynGL{2},
+      t\mathord:\SynTransl{2}.\ \\
+      \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to
+      \tyPrim{vec}{B, t} \to \tyPrim{real}{|\det B|}
+      \\ \mathrm{area}\ [B]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} *
+      \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
     \end{array}
@@
   - p_1$ are computed. These operations remove the effect of
-  translations on the result, in exactly the same way as the type
-  isomorphism we introduced in \exref{ex:type-iso}. We then compute
-  the cross product of the two vectors to determine the area of the
-  parallelogram described by the sides of the triangle. This value has
-  type $\tyPrim{real}{\det B}$, and we have removed some of the
-  symmetry due to invertible linear maps. This value still varies with
-  the sign of the determinant, so we remove this symmetry by applying
-  the $\mathrm{abs}$ function to leave us with a value of type
-  $\tyPrim{real}{|\det B|}$. Finally, we multiply by the constant
-  $\frac{1}{2}$ to recover the area of the triangle, rather than the
-  whole parallelogram.
-
-  If we specialise $\mathrm{area}$ to just orthogonal transformations,
-  we obtain the following type, as a consequence of assuming the
-  equation $|\det (\iota_OO)| = 1$:
+  translations on the result in exactly the same way as the type
+  isomorphism in \exref{ex:type-iso}. Next, we compute the cross
+  product of the two vectors is computed, which gives the area of the
+  parallelogram described by the sides of the triangle and has type
+  $\tyPrim{real}{\det B}$. This removes some of the symmetry due to
+  invertible linear maps, but the cross product still varies with the
+  sign of the determinant. We remove this symmetry as well using
+  $\mathrm{abs}$. This gives a value of type $\tyPrim{real}{|\det B|}$
+  which we multiply by $\frac{1}{2}$ to recover the area of the
+  triangle rather than that of the whole parallelogram. If we
+  specialise $\mathrm{area}$ to just orthogonal transformations, the
+  assumption $|\det (\iota_OO)| = 1$ gives the following type:
   \begin{displaymath}
     \begin{array}{@{}l}
-      \mathrm{area} : \forall O\mathord:\SynOrth{2}, t\mathord:\SynTransl{2}.\ \\
-      \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{real}{1}
+      \mathrm{area} : \forall O\mathord:\SynOrth{2},
+      t\mathord:\SynTransl{2}.\ \\
+      \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to
+      \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to
+      \tyPrim{real}{1}
     \end{array}
   \end{displaymath}
-  This type demonstrates that the area of a triangle is invariant
-  under orthogonal transformations and translations. Such combined
-  transformations are isometries---distance preserving maps.
+  This type shows that the area of a triangle is invariant under
+  orthogonal transformations and translations. Combinations of such
+  transformations are {\em isometries}, i.e., distance preserving
+  maps.
 \end{example}
@@
 units of measure type system via scaling invariance
-\cite{kennedy97relational}. Kennedy demonstrates how interpreting
-types in terms of scaling invariance brings the techniques of
-dimensional analysis \fixme{ref} to bear on programming. The types we
-have assigned to the real number arithmetic operations in
-\autoref{fig:real-ops} are exactly the types Kennedy assigns in his
-units of measure system, except for the absolute value operation. The
-effect of our assigned type is that, semantically, we are indexing by
-non-zero scaling factors, whereas Kennedy indexes by strictly positive
-scaling factors. We further examine the typing of the absolute value
-operation in \autoref{sec:types-indexed-abelian-groups}.
+\cite{kennedy97relational}. Kennedy shows how interpreting types in
+terms of scaling invariance brings the techniques of dimensional
+analysis \fixme{ref} to bear on programming. The types of the real
+number arithmetic operations in \autoref{fig:real-ops} are exactly the
+types Kennedy assigns in his units of measure system, except for that
+of the absolute value operation. Semantically, our type indexes by
+%our assigned type is that, semantically, we are indexing by
+non-zero scaling factors, whereas Kennedy's indexes by strictly
+positive ones. We examine the typing of the absolute value operation
+further in \autoref{sec:types-indexed-abelian-groups}.
 
-In our two-dimensional setting, we can make additions to Kennedy's
-one-dimensional scaling invariance. We add an operation $\iota_1$
-that, semantically, takes scale factors in $\GL{1}$ to invertible
-linear maps in $\GL{2}$: i.e.~takes numbers $s$ to matrices $\left(
+In our two-dimensional setting we can add to Kennedy's one-dimensional
+scaling invariance an operation $\iota_1$ that, semantically, takes
+scale factors in $\GL{1}$ to invertible linear maps in $\GL{2}$,
+i.e., takes numbers $s$ to matrices $\left(
   \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$.  This
 operation satisfies the equation $\det (\iota_1 s) = s^2$, indicating
-that if we scale the plane by $s$ in both directions, we scale areas
-by $s^2$. This is illustrated by the next example.
+that scaling the plane by $s$ in both directions scales areas by
+$s^2$. %, as we now illustrate.
 
@@
 generated by $\iota_1$, commute with all other invertible linear
-maps. (These maps are all the elements of $\GL{2}$ that commute with
-everything, and form the \emph{centre} of $\GL{2}$.) Therefore we
-assume $(\iota_1 s)B = B(\iota_1 s)$ to hold.
-
-If we keep track of scalings, we may assign more precise types to
-scalar multiplcation:
-\begin{displaymath}
-  (*) : \forall s \mathord: \SynGL{1}, B \mathord: \SynGL{2}.\ \tyPrim{real}{s} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{\iota_1(s)B,0}
-\end{displaymath}
-and dot product:
+maps. We thus require
+$(\iota_1 s)B = B(\iota_1 s)$ to hold.
+The scaling maps $\left(
+  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$
+are precisely the elements of $\GL{2}$ that commute with all others;
+these form the \emph{centre} of $\GL{2}$. If we keep track of
+scalings, then we can assign the following more precise types to
+scalar multiplcation and dot product:
 \begin{displaymath}
 \begin{array}{l@{}l}
+  (*) : \forall s \mathord: \SynGL{1}, B \mathord:
+  \SynGL{2}.\ & \tyPrim{real}{s} \to \tyPrim{vec}{B,0} \to
+  \tyPrim{vec}{\iota_1(s)B,0}\\
+%\end{displaymath}
+%and dot product:
+%\begin{displaymath}
+%  \begin{array}{l@{}l}
     (\cdot) : \forall s \mathord: \SynGL{1}, O \mathord: \SynOrth{2}.\ & \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\\
@@
   \cite{kennedy97relational}.  In particular, it is not possible to
-  write a square root function with the above type. The indefinability
-  of square root is similar to the uninhabited type from
-  \exref{ex:uninhabited-type}.
+  write a square root function with the above type.  The
+  indefinability of square root is similar to the uninhabitation of the
+  type in \exref{ex:uninhabited-type}.
 
   In \autoref{sec:types-indexed-abelian-groups} we revisit Kennedy's
-  proof and show that if we add square root as a primitive
+  proof and show that even if we add square root as a primitive
   operation---with the type above---then it is still not possible to
   construct the cube root function. The indefinability of cube root is
-  of interest due to its relevance to the classical problem of
-  trisecting an angle by ruler and compass constructions.
+  related to the impossibility of trisecting an arbitrary angle by
+  ruler and compass constructions.
+%  is of interest due to its relevance to the classical problem of
+%  trisecting an angle by ruler and compass constructions.
 \end{example}

### 2012-07-09 paper/paper.tex
Rename the logical information flow file
Introduction, contributions
Definitions of contextual equivalence, semantic equivalence and the soundness property
Patty's changes

@@
-\documentclass[natbib,preprint]{sigplanconf}
+\documentclass[preprint]{sigplanconf}
 
 \usepackage[usenames]{color}
+\usepackage[numbers]{natbib}
 \definecolor{citationcolour}{rgb}{0,0.4,0.2}
@@
 \newcommand{\tyPrim}[2]{\textup{\texttt{#1}}\langle #2 \rangle}
-\newcommand{\tyPrimNm}[1]{\texttt{#1}}
+\newcommand{\tyPrimNm}[1]{\textup{\texttt{#1}}}
 \newcommand{\primTyArity}{\mathrm{tyArity}}
@@
   representation. A dazzling array of applications of Reynolds' theory
-  exists, exploiting invariance to yield ``theorems for free'',
+  exists, exploiting invariance to yield ``free theorems'',
   uninhabitation results, and encodings of algebraic datatypes.
-
-  Outside computer science, invariance under change is a common theme
-  running through many areas mathematics and physics. For example, the
-  area of a triangle is not altered by rotation or flipping. If we
-  scale a triangle, then we scale its area, maintaining an invariant
-  relationship between the triangle and its area. The transformations
-  under which properties are invariant are often organised into
-  groups, with the algebraic structure reflecting the composability
-  and invertibility of transformations.
+  Outside computer science, invariance is a common theme running
+  through many areas mathematics and physics. For example, the area of
+  a triangle is unaltered by rotation or flipping. If we scale a
+  triangle, then we scale its area, maintaining an invariant
+  relationship between the two. The transformations under which
+  properties are invariant are often organised into groups, with the
+  algebraic structure reflecting the composability and invertibility
+  of transformations.
 
@@
   properties. Following Reynolds, we prove a general abstraction
-  theorem that covers all these instances. Consequences of the
+  theorem that covers all these instances. Consequences of our
   abstraction theorem include free theorems expressing invariance
   properties of programs, type isomorphisms based on invariance
-  properties, and non-definability results, indicating when certain
+  properties, and non-definability results indicating when certain
   algebraically indexed types are uninhabited or only inhabited by
-  trivial programs.
+  trivial programs.  Our results have been fully formalised in Coq.
 \end{abstract}
@@
 \input{section-instantiations.tex}
-\input{section-further-examples.tex}
+\input{section-info-flow.tex}
 \input{section-monoid.tex}

### 2012-07-09 paper/section-general-framework.tex
Patty's changes

--- 2012-07-06 paper/section-general-framework.tex
@@
 
-We now present a general framework for discussing
-algebraically-indexed types and representation independence. We define
-the syntax of algebraically-indexed types, then their semantics and
-finally we present a general programming language for
-algebraically-indexed types, with an abstraction theorem.
+%In this section 
+We now present our framework for algebraically
+indexed types. % and representation independence. %We define the syntax
+%and semantics of algebraically indexed types, then their semantics
+%and finally we present a general programming language for
+%algebraically indexed types, together with an abstraction theorem.
 
@@
 \begin{enumerate}
-\item A collection $\SortSet$ of index sorts. We will use the
-  meta-syntactic variables $s,s_1,s_2,...$ and so on to stand for
-  arbitrary sorts taken from $\SortSet$.
+\item A collection $\SortSet$ of index sorts. We use the
+  meta-syntactic variables $s,s_1,s_2,...$ for arbitrary sorts taken
+  from $\SortSet$.
 \item A collection $\IndexOpSet$ of index operations, with a function
-  $\indexOpArity : \IndexOpSet \to \SortSet^* \times \SortSet$ (we use
+  $\indexOpArity : \IndexOpSet \to \SortSet^* \times \SortSet$. (We use
   the notation $A^*$ to denote the set of lists of elements of some
-  set $A$).
+  set $A$.)
 \item A collection $\PrimTypeSet$ of primitive types, with a function
-  $\primTyArity : \PrimTypeSet \to \SortSet^*$, describing the sorts
-  of the arguments of each primitive type.
+  $\primTyArity : \PrimTypeSet \to \SortSet^*$. %, describing the sorts
+%  of the arguments of each primitive type.
 \end{enumerate}
@@
   The two-dimensional geometry system has a sort for each of the
-  geometric groups that we mentioned in
-  \autoref{sec:motivating-examples}. Thus, $\SortSet =
-  \{\SynTransl{2}, \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$. For the
-  index operations, we have additive group structure on
-  $\SynTransl{2}$, multiplicative group structure on $\SynGL{1}$,
-  $\SynGL{2}$ and $\SynOrth{2}$, injections from $\SynOrth{2}$ and
-  $\SynGL{1}$ into $\SynGL{2}$, determinant, and absolute value. Thus,
-  $\IndexOpSet = \{ 0, +, -, 1_G, -\cdot_G-, -^{-1_G}, \iota_O,
-  \iota_1, \det, |\cdot| \}$, where $G \in \{\SynGL{1}, \SynGL{2},
-  \SynOrth{2}\}$, with the following assignment of arities:
+  geometric groups mentioned in \autoref{sec:motivating-examples}, so
+  $\SortSet = \{\SynTransl{2}, \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$.
+  %For the index operations, 
+  We have additive group structure on $\SynTransl{2}$, multiplicative
+  group structure on $\SynGL{1}$, $\SynGL{2}$, and $\SynOrth{2}$,
+  injections from $\SynOrth{2}$ and $\SynGL{1}$ into $\SynGL{2}$,
+  determinant, and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
+  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, |\cdot| \}$, where
+  $G \in \{\SynGL{1}, \SynGL{2}, \SynOrth{2}\}$, and
+  %  with the following assignment of arities:
   \begin{displaymath}
@@
   \end{displaymath}
-  The intended interpretations of the top three pairs of operations
-  are group unit, group combination and group negation,
-  respectively. When we introduce equational theories for index
-  expressions in \autoref{sec:type-equality} below, we will impose the
-  (abelian) group laws. For this example, we have $\PrimTypeSet = \{
-  \tyPrimNm{vec}, \tyPrimNm{real} \}$, with
-  $\primTyArity(\tyPrimNm{vec}) = [\SynGL{2}, \SynTransl{2}]$ and
+%  The intended interpretations of the top three pairs of operations
+%  are group unit, group combination and group negation,
+%  respectively. 
+%  When we discuss equational theories on index expressions 
+%in \autoref{sec:type-equality} 
+%we will impose the (abelian) group laws. For this example, 
+We also have $\PrimTypeSet = \{ \tyPrimNm{vec},
+  \tyPrimNm{real} \}$, with $\primTyArity(\tyPrimNm{vec}) =
+           [\SynGL{2}, \SynTransl{2}]$ and
            $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
@@
 i_1, i_2,\ldots$. \emph{Index contexts} $\Delta = i_1 \mathord: s_1,
-..., i_n \mathord: s_n$ are lists of variable/sort pairs, such that
+..., i_n \mathord: s_n$ are lists of variable/sort pairs such that
 all the variable names are distinct.
-
 \begin{figure*}[t]
@@
 \end{figure*}
-
-Given the above data, the rules in \autoref{fig:indexes-and-types}
+%Given the above data, 
+The rules in \autoref{fig:indexes-and-types}
 generate two judgements: well-sorted index expressions $\Delta \vdash
@@
 with respect to an index context $\Delta$. The rules for well-sorted
-index expressions are particularly simple: either an index expression
-is a variable that appears in the context (rule \TirName{IVar}), or it
-is an application of an index operation taken from $\IndexOpSet$ to
-other index expressions (rule \TirName{IOp}). The rules for
-well-indexed types include the usual rules for constructing types of
+index expressions are particularly simple. %: either an index expression
+%is a variable that appears in the context (rule \TirName{IVar}), or it
+%is an application of an index operation taken from $\IndexOpSet$ to
+%other index expressions (rule \TirName{IOp}). 
+The rules for well-indexed types include the usual ones
+%rules 
+for
+%constructing types of 
 the simply-typed $\lambda$-calculus with unit, sum and tuple types
 (rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
-\TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
-a primitive type $\tyPrimNm{X}$ and appropriately sorted index
-expressions $e_1,...,e_n$, the well-indexed type
-$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} permits the
-formation of universally quantified types, where the universal
+\TirName{TySum}). The rule \TirName{TyPrim} %allows us to 
+forms, from a primitive type $\tyPrimNm{X}$ and appropriately sorted
+index expressions $e_1,...,e_n$, the well-indexed type
+$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} 
+%permits the formation of 
+forms universally quantified types, where the universal
 quantification ranges over all index expressions of some
-sort. Existential types are constructed using the \TirName{TyEx}
-rule, and allow for abstraction by hiding. %FIXME: think about the introduction of existentials more, perhaps forward ref to their use
+sort. Existential types, %are constructed 
+formed using the \TirName{TyEx} rule,
+%and 
+allow for abstraction by hiding. %FIXME: think about the
+%introduction of existentials more, perhaps forward ref to their use
 
@@
 
-It will be technically convenient to express substitution of index
-expressions in our framework in terms of simultaneous substitutions.
+It %will be technically 
+is convenient to express substitution of index
+expressions %in our framework 
+in terms of simultaneous substitutions.
 Given a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord:
@@
 $\Delta \vdash \sigma \Rightarrow \Delta'$ is a sequence of
-expressions $\sigma = (e_1,...,e_n)$ such that for each $1 \leq j \leq
-n$, $\Delta \vdash e_j : s_j$. Given a simultaneous substitution
+expressions $\sigma = (e_1,...,e_n)$ such that $\Delta \vdash e_j :
+s_j$ for all $1 \leq j \leq n$. Given a simultaneous substitution
 $\Delta \vdash \sigma = (e_1,...,e_n) \Rightarrow \Delta'$ and a
@@
 \vdash \sigma \Rightarrow \Delta'$.
-
-It will be useful to think of any sequence of sorts as an index
+%
+%It will be useful to 
+We can think of any sequence of sorts as an index
 context. In particular, we will make use of simultaneous substitutions
@@
 \primTyArity(\tyPrimNm{X})$, since these are exactly sequences of
-index arguments suitable for the primitive type $\tyPrimNm{X}$. With
+index arguments suitable for the primitive type $\tyPrimNm{X}$. By
 further abuse of notation, we write $\Delta \Rightarrow
@@
 variable/sort pair $i\mathord:s$ such that $i$ does not appear in
-either $\Delta$ or $\Delta'$, then we can form the \emph{lifted}
+either $\Delta$ or $\Delta'$, we can form the \emph{lifted}
 simultaneous substitution $\Delta,i\mathord:s \vdash
 \sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
-\Delta',i\mathord:s$. Note that we have implicity used the fact that
-well-sortedness of index expressions is preserved by addition of extra
-items to the context.
-
+\Delta',i\mathord:s$. %Note that we have implicity used the fact that
+%well-sortedness of index expressions is preserved by addition of extra
+%items to the context.
+%
 Application of a simultaneous substitution $\Delta \vdash \sigma
@@
 $\sigma^*(\indexOp{f}(e_1,...,e_n)) \isDefinedAs
-\indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)$.  Similarly, given a
-well-indexed type $\Delta' \vdash A \isType$, we can apply $\sigma$ to
-$A$ to produce a new well-indexed type $\Delta \vdash \sigma^*A
+\indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)$.  %Similarly, 
+Given
+%a well-indexed type 
+$\Delta' \vdash A \isType$, we 
+%can apply $\sigma$ to $A$ to produce a new well-indexed type 
+we have $\Delta \vdash \sigma^*A
 \isType$. The key clauses defining $\sigma^*A$ are for primitive types
@@
 % \end{lemma}
-
 The \emph{identity} simultaneous substitution $\Delta \vdash
-\id_\Delta \Rightarrow \Delta$ is just the sequence of variables in
-$\Delta$: $\id_\Delta = (i_1,...,i_n)$ where $\Delta =
-i_1\mathord:s_1,...,i_n\mathord:s_n$. The \emph{composition} of two
-simultaneous substitutions $\Delta \vdash \sigma \Rightarrow \Delta'$
+\id_\Delta \Rightarrow \Delta$ is %just the sequence of variables in
+%$\Delta$: 
+$\id_\Delta = (i_1,...,i_n)$ where $\Delta =
+i_1\mathord:s_1,...,i_n\mathord:s_n$. The \emph{composition} of 
+%two simultaneous substitutions 
+$\Delta \vdash \sigma \Rightarrow \Delta'$
 and $\Delta' \vdash \sigma' \Rightarrow \Delta''$, where $\sigma' =
@@
 \isDefinedAs (\sigma^*e'_1, ..., \sigma^*e'_n) \Rightarrow \Delta''$.
-
+%
 Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
@@
 $\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
-$\pi_{i\mathord:s} = (i_1,...,i_n)$. The subscript on
-$\pi_{i\mathord:s}$ is the variable/sort pair that is being discarded.
+$\pi_{i\mathord:s} = (i_1,...,i_n)$. %The subscript on
+%$\pi_{i\mathord:s}$ is the variable/sort pair that is being discarded.
 
@@
 Much of the power of indexing types by the expressions of an algebraic
-theory comes from the equations of the theory. In the two-dimensional
-geometry example of \autoref{sec:motivating-examples}, the types
+theory comes from the equations of the theory. 
+%In the %two-dimensional geometry example of 
+For example, in 
+\autoref{sec:motivating-examples} the types
 $\tyPrim{vec}{B,t_1 + t_2}$ and $\tyPrim{vec}{B, t_2 + t_1}$ are
-considered equal by the type system, due to the commutativity of the
-$+$ operation.
-
+considered equal by the type system
+%due to the commutativity of the $+$ operation.
+because $+$ is commutative.
+%
 In the general framework, the equations between types are derived from
 a set $\IndexAxiomSet$ of axioms $\Delta \vdash e \stackrel{ax}\equiv
-e' : s$. We assume that all axioms $(\Delta \vdash e
-\stackrel{ax}\equiv e' : s) \in \IndexAxiomSet$ are well-sorted,
-meaning that both $\Delta \vdash e : s$ and $\Delta \vdash e' : s$
+e' : s$ that %We assume that all axioms $(\Delta \vdash e
+%\stackrel{ax}\equiv e' : s)$ % \in \IndexAxiomSet$ 
+are well-sorted, in the sense that both $\Delta \vdash e : s$ and $\Delta \vdash e' : s$
 hold.
@@
   \label{ex:two-dim-geo-axioms}
-  In \autoref{sec:motivating-examples}, we assumed that various
-  equational axioms hold for the indexing expressions standing for
-  elements of geometric groups. We can now make this assumption formal
-  in our general framework. Semantically, translations form an abelian
-  group under addition, so we assume the abelian group axioms for
-  translations:
+  In \autoref{sec:motivating-examples} we assumed 
+  %that 
+  various equational axioms %hold 
+  for indexing expressions standing for
+  elements of geometric groups. Assuming the abelian group axioms for
+  translations we can formalise this in our %general 
+  framework:
+  %We can now make this assumption formal
+  %in our general framework. Semantically, translations form an abelian
+  %group under addition, so we assume the abelian group axioms for
+  %translations:
   \begin{displaymath}
@@
 B \isType$ on types. The basic rule generating equality judgments on
-types states that two applications of primitive types are equal if
-their index expression arguments are equal:
+types equates %states that %two 
+applications of primitive types %are equal 
+if
+their %index expression 
+arguments are equal:
 \begin{displaymath}
@@
 The rest of the rules for equality on types ensure that it is a
-congruence relation on types, and that it is an equivalence relation
-(i.e.,~equality on types is reflexive, symmetric and transitive).
+congruence relation 
+and an equivalence relation.
+%on types. %and that it is an equivalence relation
+%(i.e.,~equality on types is reflexive, symmetric and transitive).
 % For example, for universally quantified types we have
@@
 
-A pair of simultaneous substitutions $\Delta \vdash \sigma \Rightarrow
+%A pair of 
+The simultaneous substitutions $\Delta \vdash \sigma \Rightarrow
 \Delta'$ and $\Delta \vdash \sigma' \Rightarrow \Delta'$ are defined
@@
 
-\fixme{Make sure that the semantics is motivated}
-Having defined the language of algebraically-indexed types, we now
-turn to their denotational interpretation. In
-\autoref{sec:index-erasure-semantics} we define an
-\emph{index-erasure} interpretation of types that interprets every
-well-indexed type as a set, ignoring the indexing expressions. Building on
-the index-erasure semantics, we define in
-\autoref{sec:relational-semantics} the \emph{relational}
-interpretation of types.
+%\fixme{Make sure that the semantics is motivated}
+%Having defined the language of algebraically-indexed types, we now
+%turn to their denotational interpretation. 
+%We now give the denotational interpretation of algebraically indexed
+%types.  In \autoref{sec:index-erasure-semantics} we define an
+%\emph{index-erasure} interpretation of types that interprets every
+%well-indexed type as a set, ignoring the indexing
+%expressions. %Building on the index-erasure semantics, we 
+%We then define in
+%\autoref{sec:relational-semantics} the relational
+%interpretation of types.
+
+\autoref{sec:index-erasure-semantics} gives an \emph{index-erasure}
+interpretation of types, interpreting every well-indexed type as a
+set, ignoring the indexing expressions.
+\autoref{sec:relational-semantics} gives the relational interpretation
+of types.
 
@@
 
-The defining feature of the index erasure interpretation is that the
+The defining feature of the index erasure interpretation is that 
 semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
 determined solely by the primitive type $\tyPrimNm{X}$ and not by the
-index expressions $e_1,...,e_n$. Accordingly, we assume that each
+index expressions $e_1,...,e_n$. %Accordingly, 
+We thus assume each
 primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
-$\tyPrimSem{\tyPrimNm{X}}$. We extend this assignment to every
-well-indexed type by induction on the type structure:
+$\tyPrimSem{\tyPrimNm{X}}$ and %We 
+extend this assignment to %every
+well-indexed types by: % induction on the type structure:
 \begin{displaymath}
@@
 % the corresponding constructions on sets. 
-The interpretations of the
-universal and existential quantifiers ignore the indexing: the
-interpretation of the type $\forall i\mathord:s.A$ is exactly the
-interpretation of the type $A$, and likewise for $\exists
-i\mathord:s.A$.
+Note that 
+the interpretations of the
+universal and existential quantifiers 
+do indeed
+ignore the indexing. %: the
+%interpretation of the type $\forall i\mathord:s.A$ is exactly the
+%interpretation of the type $A$, and likewise for $\exists
+%i\mathord:s.A$.
+We then have:
 
@@
 \begin{example}[Two-Dimensional Geometry]
-  The two-dimensional geometry instantiation of the general framework
+  Example~\autoref{ex:two-dim-geo-axioms}
+%  The two-dimensional geometry instantiation of the general framework
   uses the assignment $\tyPrimSem{\tyPrimNm{vec}} = \mathbb{R}^2$ and
@@
 
-We now define the relational semantics of each well-indexed type
-$\Delta \vdash A \isType$ as some binary relation on the index-erasure
-interpretation of $A$. For a set $X$, we write $\Rel(X)$ for the set
-of binary relations $R \subseteq X \times X$ on $X$.
+%We now define 
+The relational semantics of the well-indexed type
+$\Delta \vdash A \isType$ is a binary relation on the index-erasure
+interpretation of $A$. We write $\Rel(X)$ for the set
+of binary relations $R \subseteq X \times X$ on the set $X$.
 
-For the unit, tuple, sum and function types we will define the
+For the unit, tuple, sum and function types we define the
 relational interpretation as a standard logical relation. The
@@
 the universally quantified types require an interpretation of index
-contexts. We assign interpretations to index contexts in terms of sets
+contexts. %We assign interpretations to index contexts 
+These are given in terms of sets
 of \emph{relational environments}.
@@
 The definition of relational environments is the most subtle part of
-our development. The straightforward approach would be to assume some
-interpretation of the indexing expressions, where sorts are
-interpreted as sets and index operations as functions, and then define
-the relational interpretation of a primtive type
-$\tyPrim{X}{e_1,...,e_n}$ in terms of the interpretations of
-$e_1,...,e_n$. This approach suffices to prove free theorems and type
-isomorphisms for instantiations of our general framework. However, in
-order to prove non-definability results (like, for example, the one in
-\exref{ex:uninhabited-type}), we choose a more refined interpretation
+our framework. The straightforward approach is to assume an
+interpretation of indexing expressions in which sorts are
+%interpreted as 
+sets and index operations are functions, and then define the
+relational interpretation of a primtive type $\tyPrim{X}{e_1,...,e_n}$
+in terms of the interpretations of $e_1,...,e_n$. This approach
+suffices to prove free theorems and type isomorphisms for
+instantiations of our general framework. However, to prove
+non-definability results, like the one in
+\exref{ex:uninhabited-type}, we need a more refined interpretation
 that accounts more precisely for the definable programs permitted by
-the algebraic indexing. In this section, we set out the general
-requirements that we require of relational environments in order for
+the algebraic indexing. In this section, we identify the 
+properties we require of relational environments in order for
 the abstraction theorem to hold. In \autoref{sec:constr-rel-env}, we
-give a general construction of relational environments that will allow
-us to prove several non-definability results.
+give a general construction of relational environments and use it 
+to prove several non-definability results.
 
-A relational environment for a context $\Delta$ is a (dependently typed)
+A relational environment for a context $\Delta$ is a %(dependently typed)
 function $\rho$ that, for each primitive type $\tyPrimNm{X}$ and
@@
 \end{displaymath}
-Every relation environment must respect index expression equality: for
-any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
-substitutions $\Delta \vdash \sigma \Rightarrow
-\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
+%Every relation environment must respect index expression equality: for
+%any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
+%substitutions $\Delta \vdash \sigma \Rightarrow
+%\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
+%\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
+%\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
+%\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
+Relation environments must respect index expression equality: for
+all $\tyPrimNm{X} \in \PrimTypeSet$,
+$\Delta \vdash \sigma \Rightarrow
+\primTyArity(\tyPrimNm{X})$, and $\Delta \vdash \sigma' \Rightarrow
 \primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
@@
 \sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
-
+%
 We write $\mathrm{RelEnv}(\Delta)$ for the set of all relation
-environments for a context $\Delta$. Given a relational environment
-$\rho \in \mathrm{RelEnv}(\Delta')$ and a simultaneous substitution
+environments for a context $\Delta$. Given %a relational environment
+$\rho \in \mathrm{RelEnv}(\Delta')$ and %a simultaneous substitution
 $\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
@@
 As we stated above, index contexts $\Delta$ are interpreted as certain
-sets of relation environments, i.e.,~subsets of
-$\mathrm{RelEnv}(\Delta)$. Depending on the primitive operations that
-we assume for our system, we may have different sets of relational
-environments that enforce different invariants. We require that the
-sets of relational environments are \emph{valid} in the following
-sense:
+%sets of relation environments, i.e.,~
+subsets of $\mathrm{RelEnv}(\Delta)$. Depending on the primitive
+operations that we assume for our system, we may have different sets
+of relational environments that enforce different invariants. We
+require that the sets of relational environments are \emph{valid} in
+the following sense:
 
@@
     then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
-  \item If we have a pair of relation environments $\rho_1 \in
-    \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
-    \relEnv{E}(\Delta)$, along with a simultaneous substitution
-    $\Delta' \vdash \sigma \Rightarrow \Delta$, such that $\rho_1
-    \circ \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists
-    a $\rho \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho
-    \circ \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ
-    \pi_{i\mathord:s'} = \rho_2$.
+%  \item If we have a pair of relation environments $\rho_1 \in
+%    \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
+%    \relEnv{E}(\Delta)$, along with a simultaneous substitution
+%    $\Delta' \vdash \sigma \Rightarrow \Delta$, such that $\rho_1
+%    \circ \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists
+%    a $\rho \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho
+%    \circ \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ
+%    \pi_{i\mathord:s'} = \rho_2$.
+  \item If $\rho_1 \in \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2
+    \in \relEnv{E}(\Delta)$, and $\Delta' \vdash \sigma \Rightarrow
+    \Delta$ is such that $\rho_1 \circ \pi_{i\mathord:s'} = \rho_2
+    \circ \sigma$, then there exists a $\rho \in \relEnv{E}(\Delta,
+    i\mathord:s')$ such that $\rho \circ \sigma_{i\mathord:s'} =
+    \rho_1$ and $\rho \circ \pi_{i\mathord:s'} = \rho_2$.
     % such that the outer edges of the
@@
 \end{definition}
-
+%
 % For the general
@@
 % properties:
-
-These conditions ensure that the relational interpretation of types
-that we define below behaves correctly with respect to simultaneous
+%
+\noindent
+These conditions ensure that the relational interpretation of types we
+define below behaves correctly with respect to simultaneous
 substitution of index expressions (see
@@
 % correctly with respect to application of simultaneous substitution.
-
-Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
-the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
-additional index variable $i\mathord:s$ to be
-$\extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in
+%
+Given a relation environment $\rho \in \relEnv{E}(\Delta)$, %we define
+the set of extensions %$\extends{\rho}{i\mathord:s}$ 
+of $\rho$ by %an
+%additional index variable 
+$i\mathord:s$ %to be 
+is $\extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in
 \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi_{i\mathord:s} =
 \rho \}$. The set of extensions of a relational environment will be
-used in the interpretation of the universal and existential type
-formers.
-
-We assign a relational interpretation to all well-indexed types
-$\Delta \vdash A \isType$ by induction on their derivations,
-parameterised by relational environments $\rho \in
-\relEnv{E}(\Delta)$:
+used in interpretating the universal and existential type formers.  We
+assign a relational interpretation to all well-indexed types $\Delta
+\vdash A \isType$ by induction on their derivations, parameterised by
+%relational environments 
+$\rho \in \relEnv{E}(\Delta)$:
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-In this definition we have made use of the following three
-constructions on binary relations. If $R \in \Rel(X)$ and $S \in
-\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is defined as $\{
-(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
-\}$. With the same assumptions on $R$ and $S$, the relation $R
-\relTimes S \in \Rel(X \times Y)$ is defined as $\{
-((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
-\}$. Finally, the relation $R \relSum S \in \Rel(X + Y)$ is defined as
+In this definition we have made use of the following %three
+constructions on binary relations: if $R \in \Rel(X)$ and $S \in
+\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is %defined as 
+$\{(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
+\}$, and %. With the same assumptions on $R$ and $S$, the relation 
+$R
+\relTimes S \in \Rel(X \times Y)$ is %defined as 
+$\{((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
+\}$, and %. Finally, the relation 
+$R \relSum S \in \Rel(X + Y)$ is %defined as
 $\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
@@
 \end{lemma}
-
+\noindent
 Note that the equations in both parts of
@@
 
-We now present the rules for well-typed programs over the collection
-of types we defined in \autoref{sec:algebraically-indexed-types}. Each
+We now present the rules for well-typed programs over the 
+%collection of 
+types %we defined 
+in \autoref{sec:algebraically-indexed-types}. Each
 well-type program is assigned an index-erasure semantics, building on
-the index-erasure semantics of types we defined in
-\autoref{sec:index-erasure-semantics}. Our main result
+the index-erasure semantics of types %we defined 
+in \autoref{sec:index-erasure-semantics}. Our main result
 (\thmref{thm:abstraction}) is that the index-erasure semantics of
@@
 variable names such that each type is well-indexed with respect to
-$\Delta$. Formally, well-indexed typing contexts are defined by the
-following two rules:
+$\Delta$. Formally, well-indexed typing contexts are 
+given by
+%defined by the following two rules:
 \begin{mathpar}
@@
 
-The typing rules for our system define the well-typed programs with
-respect to an index context $\Delta$ and a type context $\Delta \vdash
+%The typing rules for our framework define 
+Well-typed programs 
+are defined
+with respect to an index context $\Delta$ and a type context $\Delta \vdash
 \Gamma \isCtxt$. The judgment $\Delta; \Gamma \vdash M : A$ is defined
@@
 incorporated into the type system via the rule \TirName{TyEq}, which
-allows for a program that is judged to have type $A$ to also have any
-other equal type $B$.
+allows for a program that has %is judged to have 
+type $A$ to also have any
+%other 
+equal type $B$ as well.
 \begin{figure*}[t]
@@
 We assign an index-erasure semantics to any well-indexed typing
-context $\Delta \vdash \Gamma \isCtxt$ by induction on its structure:
+context $\Delta \vdash \Gamma \isCtxt$ by induction:
 $\ctxtSem{\epsilon} = \{*\}$ and $\ctxtSem{\Gamma, x : A} =
@@
 \Gamma \vdash M : A$, we define the \emph{erasure interpretation} as a
-function $\tmSem{M} : \ctxtSem{\Gamma} \to \tySem{A}$, that completely
+function $\tmSem{M} : \ctxtSem{\Gamma} \to \tySem{A}$ that completely
 ignores the indexing information. In light of
-\lemref{lem:tyeqsubst-erasure}, we define this function directly on
-the syntax of well-typed programs, rather than on the typing
+\lemref{lem:tyeqsubst-erasure}, %we define this function is defined
+directly on the syntax of well-typed programs, rather than on typing
 derivations. The definition of $\tmSem{M}$ is completely standard,
@@
 We now state the abstraction theorem for well-typed programs. To state
-and prove this theorem for open programs, we must extend the
+and prove this theorem for open programs, we extend the
 relational interpretation of types to typing contexts. The relational
-interpretation of contexts is defined by the following clauses, where
-we have used the relational lifting of cartesian product $-\relTimes-$
-defined in \autoref{sec:relational-semantics}.
+interpretation of contexts is defined by: % the following clauses: %, where
+%we have used the relational lifting of cartesian product $-\relTimes-$
+%defined in \autoref{sec:relational-semantics}.
 \begin{displaymath}
@@
 We now give a general construction of valid families of sets of
-relational environments (in the sense of
-\defref{defn:valid-rel-env-family}) that will allow us to derive
-useful consequences of \thmref{thm:abstraction}. In particular, we
-wish to show that certain types are uninhabited. To motivate this,
-consider the following type, which makes use of vectors indexed by
-translations, as in \autoref{sec:affine-vector-ops}:
-\begin{displaymath}
-  \forall t\mathord:\SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}
-\end{displaymath}
+relational environments 
+%(in the sense of \defref{defn:valid-rel-env-family}) 
+that allow us to derive
+useful consequences of \thmref{thm:abstraction}, e.g.,
+%in particular, 
+%we wish to show 
+that certain types are uninhabited. To %motivate 
+this end, consider the
+%following 
+type $\forall t\mathord:\SynTransl{2}.\ \tyPrim{vec}{t + t} \to
+\tyPrim{vec}{t}$ involving vectors indexed by translations.
+% as in \autoref{sec:affine-vector-ops}.
+%\begin{displaymath}
+%  \forall t\mathord:\SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}
+%\end{displaymath}
 In \exref{ex:uninhabited-type} we claimed that this type is
@@
 way of cancelling the additional $t$ in the argument type to get the
-result type. More generally, the operations on vectors only allow us
-to combine indexes according to the group operations. Thus, if the
-index of the result type of a function does not lie within the free
-abelian group generated by the indexes of the input types, then there
-is no way of writing a program with the given type. We now develop
-this idea to generate a family of sets of relational environments that
-is valid in the sense of \defref{defn:valid-rel-env-family}. The key
-idea is to assign different relational interpretations to types with
-indexes that lie outside some fintely generated of index expressions.
+result type. More generally, our operations on vectors 
+only allow us to combine indices according to the group operations. Thus,
+if the index of the result type of a function does not lie within the
+free abelian group generated by the indices of the input types, then
+there is no way to write a program with this %given 
+type. We now
+develop this idea to generate a family of sets of relational
+environments that is valid in the sense of
+\defref{defn:valid-rel-env-family}. The key idea is to assign
+different relational interpretations to types with indices that lie
+outside some fintely generated group of index expressions.
 
@@
 We make use of some concepts of universal algebra, which now recall.
-
 A \emph{model} for our indexing expressions and their equality is
-defined as follows. For each sort $s \in \SortSet$ we assume a carrier
+defined as follows. To each sort $s \in \SortSet$ we assign a carrier
 set $\semSort{s}$. We extend this assignment to index contexts and
@@
 $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
-\semIndexExp{\cdot}_{\mathcal{M}})$ such that all the axioms in
-$\IndexAxiomSet$ are satisfied.
+\semIndexExp{\cdot}_{\mathcal{M}})$ 
+%such that all the axioms in
+satisfying all the axioms in
+$\IndexAxiomSet$.
+% are satisfied.
 
@@
 \paragraph{Relational Environments} We now fix some model
-$\mathcal{M}$ that will provide the underlying interpretation of index
-expressions. As we indicated above, we wish to discriminate between
-index expressions that are generated from some finite set of ``input''
-expressions and those that are not. Accordingly, we assume that for
-each primitive type $\tyPrimNm{X}$ we have two ways of assigning a
-relational meaning to that type. When the index arguments of the type
-are within the generated set, we will use a function $R_{\tyPrimNm{X}}
-\in \semSort{\primTyArity(\tyPrimNm{X})}_{\mathcal{M}} \to
+$\mathcal{M}$ to provide provide the underlying interpretation of
+index expressions. %As we indicated above, we wish 
+To discriminate
+between index expressions that are generated from some finite set of
+``input'' expressions and those that are not, %. Accordingly, 
+we assume that for each primitive type $\tyPrimNm{X}$ we have two ways of
+assigning a relational meaning to that type. When the index arguments
+of the type are within the generated set, we use a function
+$R_{\tyPrimNm{X}} \in
+\semSort{\primTyArity(\tyPrimNm{X})}_{\mathcal{M}} \to
 \Rel(\tyPrimSem{\tyPrimNm{X}})$ that assigns a relational meaning to
@@
 \Rel(\tyPrimSem{\tyPrimNm{X}})$.
-
-We now use $R_{\tyPrimNm{X}}$ and $R^\bullet_{\tyPrimNm{X}}$ to
-construct the relational environments that we are interested in.  Fix
-an index context $\Delta$. Let $A = \{ A_s \}_{s \in \SortSet}$ be a
+%We now use $R_{\tyPrimNm{X}}$ and $R^\bullet_{\tyPrimNm{X}}$ to
+%construct the relational environments that we are interested in.  
+Now, fix
+an index context $\Delta$, and let $A = \{ A_s \}_{s \in \SortSet}$ be a
 sort-indexed family of finite sets of expressions such that if $e \in
@@
 
-Note that we still need to prove for each instantiation of our general
-framework that the program-level primitive operations preserve all the
-relations, in order to fulfil the hypothesis of
-\thmref{thm:abstraction}. \fixme{This is the content of lemma X, lemma
-  Y, lemma Z, below}.
+To satisfy the hypothesis of \thmref{thm:abstraction} we still need to
+prove, for each instantiation of our general framework, that the
+program-level primitive operations preserve all the relations.
+\fixme{This is the content of lemma X, lemma Y, lemma Z, below}.
 

### 2012-07-09 paper/section-introduction.tex
Patty's changes

@@
 is Reynolds' Abstraction Theorem~\cite{reynolds83types}. Its striking
-consequences include the well-known ``theorems for free'' for
-polymorphic types~\cite{wadler89theorems}, non-inhabitation results, and precise
-correspondences between System F encodings and algebraic datatypes~\cite{PittsAM:parpoe},
-abstract data types, and, most recently, higher-order encodings of
-binder syntax~\cite{syntaxforfree}.
+consequences include the well-known ``free theorems'' for polymorphic
+types~\cite{wadler89theorems}, non-inhabitation results, and precise
+correspondences between System F encodings and algebraic
+datatypes~\cite{PittsAM:parpoe}, abstract data types, and, most
+recently, higher-order encodings of binder
+syntax~\cite{syntaxforfree}.
 
@@
 \emph{invariance}: the behaviour of polymorphic code is invariant
-under changes to data representation. Invariance results also abound in
+under changes of data representation. Invariance results also abound in
 mathematics and physics. The area of a triangle is invariant with
@@
 and physics have interesting structure; for example, translations in
-the Euclidean plane form an Abelian group.
+the Euclidean plane form an abelian group.
 
-%Types express properties of programs. The statement $\mathit{b}: \tyBool$ tells
-%us that $\mathit{b}$ has value $\mathtt{true}$ or $\mathtt{false}$;
-%the statement $\mathit{pos}:\tyReal\tyProduct\tyReal$ says that
-%$\mathit{pos}$ is a pair whose components are real numbers. More interestingly,
-%the statement $\mathit{f} : \forall X. \tyPrim{list}X\tyArr\tyPrim{list}X$ tells us that
-%if we apply $\mathit{f}$ to a list whose elements satisfy some property $\psi$, 
+%Types express properties of programs. The statement $\mathit{b}:
+%\tyBool$ tells us that $\mathit{b}$ has value $\mathtt{true}$ or
+%$\mathtt{false}$; the statement
+%$\mathit{pos}:\tyReal\tyProduct\tyReal$ says that $\mathit{pos}$ is a
+%pair whose components are real numbers. More interestingly, the
+%statement $\mathit{f} : \forall
+%X. \tyPrim{list}X\tyArr\tyPrim{list}X$ tells us that if we apply
+%$\mathit{f}$ to a list whose elements satisfy some property $\psi$,
 %then the result is a list whose elements also satisfy $\psi$.
@@
 %Unary predicates only go so far, and for polymorphic types
-%particularly do not express everything that follows from possessing that type. 
-%Imagine a function $\mathit{oddRev}$ that reverses any list whose first element
-%is an odd number, otherwise behaving as the identity. It certainly
-%preserves properties of the elements -- but truly parametric
-%polymorphism should tell us more.
+%particularly do not express everything that follows from possessing
+%that type.  Imagine a function $\mathit{oddRev}$ that reverses any
+%list whose first element is an odd number, otherwise behaving as the
+%identity. It certainly preserves properties of the elements -- but
+%truly parametric polymorphism should tell us more.
 
@@
 %polymorphic type: simply consider the change-of-representation given
-%by $\lambda x. x+1$.  But more profoundly it can be used to prove blah
-%blah (initial algebras, term reps, non-inhabitation, etc.).
+%by $\lambda x. x+1$.  But more profoundly it can be used to prove
+%blah blah (initial algebras, term reps, non-inhabitation, etc.).
 
-Inspired by this connection, we study relational parametricity for
-types that are \emph{indexed} by attributes that themselves have
-algebraic structure.  For example, in computational geometry, points
-in 2-space can be indexed by a coordinate frame; in physical
-modelling, quantities are indexed by units-of-measure; in
-information-flow security, computations are indexed by principals.
-Types that are polymorphic over such indices induce invariance
-properties, and introduce a useful abstraction barrier.
+In this paper we show how to capture, and generalise, invariance under
+groups of translations using
+%Inspired by this connection, we study 
+relational parametricity for types that are indexed by attributes with
+algebraic structure. For example, in computational geometry, points in
+2-space can be indexed by a coordinate frame; in physical modelling,
+quantities are indexed by units-of-measure; in information-flow
+security, computations are indexed by principals.  Types that are
+polymorphic over such indices induce invariance properties and
+abstraction barriers beyond those introduced by their unindexed
+versions, as we now illustrate.
 
 \paragraph{Invariance}
-A function $\mathit{areaTri}$ that computes the area of a triangle might
-have simple type $\tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal$
-but we can assign it a more expresive polymorphic type
+A function $\mathit{areaTri}$ that computes the area of a triangle
+can be assigned the type:
+$\tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal$.
+But we can also assign it the following more expressive polymorphic
+type:
 \[
@@
 \]
-in which the index $\alpha$ can be thought of as a `translation frame' that is
-preserved by the vector arguments, from which we can deduce that
+in which the index $\alpha$ is thought of as a ``translation frame''
+that is preserved by the vector arguments. From this we can deduce
+that
 \[
@@
 In more complex examples the invariants \emph{compose} in various
-ways; moreover, the compositions satisfy algebraic laws. Consider a function
-that computes the area of a circle given its radius; by ascribing it
-the polymorphic type
+ways, with the compositions satisfy algebraic laws. For example, 
+we can assign a function that computes the area of a circle given its
+radius the following polymorphic type:
 \[
@@
 \]
-we have captured that its area varies as the square of its radius, that is, for any $s>0$,
-\[
-\mathrm{areaCircle}\;(s\cdot r) = s^2\cdot \mathrm{areaCircle}\;r.
-\]
-Here $\alpha$ can be interpreted as the \emph{units of measure} of the argument
-to $\mathrm{areaCircle}$, and `$\cdot$' composes units by forming
-their product. To complete the system, we would add an inverse
-operation and identity unit-of-measure, together with the algebraic laws of
-Abelian groups, permitting, for example, the identification of
-$\tyPrim{real}{\alpha\cdot\alpha^{-1}}$ with $\tyPrim{real}{1}$.
+This captures the fact that the area of a circle varies as the square
+of its radius, i.e., for any $s>0$, $\mathrm{areaCircle}\;(s\cdot r) =
+s^2\cdot \mathrm{areaCircle}\;r$.  Here, $\alpha$ is interpreted as
+the \emph{units of measure} of the argument to $\mathrm{areaCircle}$,
+and `$\cdot$' composes units using the product. We can also add an
+inverse operation and identity unit of measure, and then impose the
+algebraic laws of abelian groups. This permits identification of, for
+example, $\tyPrim{real}{\alpha\cdot\alpha^{-1}}$ with the type
+$\tyPrim{real}{1}$ of dimensionless constants.
 
 \paragraph{Abstraction}
-In his original paper on parametricity, Reynolds asserted that \emph{type structure is
-  a syntactic discipline for enforcing levels of abstraction}.  We see
-something analogous here: supposing that all primitive operations are
-given types that reflect their behaviour under translation, there is
-no way to `break' this property, for example, for $\mathrm{areaTri}$
-to depend on the actual coordinates of its inputs. Furthermore, the
-distinction between points and vectors that is often enforced through
-abstract data types~\cite{CGAL} is here captured by the indices
-instead. For example, the operation that takes two points and computes
-their vector `difference' can be assigned the type
+In his original paper on parametricity, Reynolds asserted that
+\emph{type structure is a syntactic discipline for enforcing levels of
+  abstraction}.  We see something analogous here: if all primitive
+operations are given types that reflect their behaviour under
+translation, then there is no way to `break' this property. For
+example, there is no way that $\mathrm{areaTri}$ can depend on the
+actual coordinates of its inputs. Furthermore, the distinction between
+points and vectors that is often enforced through abstract data
+types~\cite{CGAL} is captured here by indices instead. For example,
+the operation that takes two points and computes their vector
+`difference' can be assigned the type
 $\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\times\tyPrim{vec}\alpha\to\tyPrim{vec}0$,
 reflecting the invariance of the result (a pure vector) under
-translations of the point arguments. In essence, through types alone
-we can derive so-called \emph{coordinate-free} geometry~\cite{CFGADT}.
+translations of the point arguments. As a result through types alone
+we can, in essence, derive so-called \emph{coordinate-free}
+geometry~\cite{CFGADT}.
 
-The invariance properties listed above can be seen as `theorems for
-free'~\cite{wadler89theorems} obtained from types alone. We can
-further observe that the abstraction afforded by polymorphic indexed
-types induces interesting type \emph{isomorphisms}.  The type of
-$\mathrm{areaCircle}$ above is in fact isomorphic to
-$\tyPrim{real}{1}$, that is, a dimensionless constant. A moment's
-thought reveals why: what possible unary functions can be constructed whose
-outputs scale as the square of the scaling of their inputs? Answer:
-just those functions of the form~$\lambda x. k x^2$ for some constant~$k$.
-In this case, of course, we expect that~$k = \pi$.
+The invariance properties discussed above can be seen as ``free
+theorems''~\cite{wadler89theorems}, but the abstraction afforded by
+polymorphic indexed types can also induce interesting type
+\emph{isomorphisms}.  The type of $\mathrm{areaCircle}$ above is in
+fact isomorphic to $\tyPrim{real}{1}$. A moment's thought reveals why:
+what possible unary functions can be constructed whose outputs scale
+as the square of the scaling of their inputs?  Answer: just those
+functions of the form~$\lambda x. k x^2$ for some constant~$k$.  In
+this case, of course, we expect that~$k = \pi$.
 
 \paragraph{Relational parametricity}
-In order to derive such invariance and abstraction properties of types, we
-adopt the techniques of relational parametricity. Over an underlying index-erasure semantics
-we construct binary
-relations parameterized by an environment $\rho$ that describes how values of primitive type
-are related according to their indices. 
-For example, values $v$ and
-$w$ of type $\tyPrim{real}\alpha$ are related when $v$ `scales to' $w$
-according to an interpretation of $\alpha$ (that is, $w=\rho(\alpha)\cdot v$).
-Values of polymorphic type are related exactly when they are related for
-all possible interpretations of the quantified variable. For example, 
-values $v$ and $w$ of type $\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\to
+To derive such invariance and abstraction properties of types, we
+adopt the techniques of relational parametricity. Over an underlying
+index-erasure semantics we construct binary relations parameterised by
+an environment $\rho$ that describes how values of primitive type are
+related according to their indices.  For example, values $v$ and $w$
+of type $\tyPrim{real}\alpha$ are related when $v$ ``scales to'' $w$
+according to an interpretation of $\alpha$ (i.e., $w=\rho(\alpha)\cdot
+v$).  Values of polymorphic type are related exactly when they are
+related for all possible interpretations of the quantified
+variable. For example, values $v$ and $w$ of type
+$\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\to
 \tyPrim{vec}\alpha$ are related when they are related at type
-$\tyPrim{vec}\alpha\to\tyPrim{vec}\alpha$ for all 
-translations $t\in\Transl{2}$ associated with $\alpha$.
+$\tyPrim{vec}\alpha\to\tyPrim{vec}\alpha$ for all translations
+$t\in\Transl{2}$ associated with $\alpha$.
 
-As it happens, these examples can be expressed using unary functions
-rather than binary relations: we can write functions that scale, or
-translate, one value to another according to their type, even for
-higher-order function types. Also note that the interpretation of
+\vspace*{0.1in}
+
+As it happens, the above examples can all be expressed using unary
+functions rather than binary relations: we can write functions that
+scale, or translate, one value to another according to their type,
+even for higher-order function types. Moreover, the interpretation of
 indices can be built up compositionally, by recursion over the
 syntactic structure. For example, we have $\rho(\alpha\cdot\alpha) =
-\rho(\alpha)\cdot\rho(\alpha)$ where product of units-of-measure is interpreted by
-product of scale factors.
-
-For more sophisticated applications we need the fully generality of
-binary relations. First, as is also the case with type polymorphism,
-it provides the appropriate meaning for nested quantifiers in types
-such as
+\rho(\alpha)\cdot\rho(\alpha)$ where product of units-of-measure is
+interpreted by product of scale factors. For more sophisticated
+applications, however, we need the fully generality of binary
+relations. First, as is also the case with type polymorphism, it
+provides the appropriate meaning for nested quantifiers in types such
+as
 $(\forall\alpha\mathord:\SynScal.\tyPrim{real}\alpha\to\tyPrim{real}\alpha)\to\tyBool$.
-More interestingly, we sometimes use relations that are not built up
-in a compositional way -- we can, for instance, use such relations to
-prove that there are no non-trivial functions of type
+More interestingly, particular applications sometimes require
+relations that are not built compositionally---we can, for instance,
+use such relations to prove that there are no non-trivial functions of
+type
 $\forall\alpha\mathord:\SynScal.\tyPrim{real}{\alpha\cdot\alpha}\to\tyPrim{real}\alpha$.
 Finally, there are interesting applications in which even the basic
-relations are not functional. Consider a type system in which the
-index in $\tyPrim{real}\alpha$ is interpreted not as a
-unit-of-measure, but as a measure of \emph{closeness}. Two values $x$
-and $y$ of this type are related if $|x-y| < \rho(\alpha)$, in which
-$\rho(\alpha)$ is a positive real number.  Rather beautifully, the
-standard notion of continuous functions can then be expressed as a
-type:
+relations are not functional. In a type system in which the index in
+$\tyPrim{real}\alpha$ is interpreted not as a unit of measure, but as
+a measure of \emph{closeness}, two values $x$ and $y$ of this type are
+related if $|x-y| < \rho(\alpha)$ for a positive real number
+$\rho(\alpha)$.  Rather beautifully, the standard notion of continuous
+functions can then be expressed as a type:
 \begin{displaymath}
@@
 
-We make the following contributions. 
+This paper makes the following specific contributions:
 \begin{itemize}
 \item 
-We present a collection of compelling examples of algebraically indexed
-types, including a novel type system for geometry, a type system for
-tracking taint marking (or provenance) of strings, a refined type
-system for information flow, based on logic, and a simple type
-system for basic continuity analysis.
+We present a collection of compelling examples of algebraically
+indexed types, including a novel type system for geometry, a type
+system for tracking taint marking (or provenance) of strings, a
+refined type system for information flow based on logic, and a simple
+type system for basic continuity analysis.
 \item 
-We formulate a general class of type systems that can either be
-used as programming languages in their own right, or as the targets of
-type-based analyses. The type system consists of the usual type
-constructors together with a collection of indexed primitive types,
-universal and existential quantification over the indices, and a
-multi-sorted equational theory for indices. 
+We formulate a type system that can either be used as a programming
+language in its own right, or as the target of type-based
+analyses. The type system consists of the usual type constructors
+together with a collection of indexed primitive types, universal and
+existential quantification over the indices, and a multi-sorted
+equational theory for indices.
 \item
-We describe a relational semantics for the type system and prove an analogue
-of Reynolds' Abstraction Theorem. We pin down precise conditions on sets of
-relation environments for which the theorem holds.
+We describe a relational semantics for the type system and prove an
+analogue of Reynolds' Abstraction Theorem. We give precise conditions
+on sets of relation environments under which our Abstraction Theorem
+holds.
 \item
-For each of our main examples we deduce free theorems that are consequences of
-the Abstraction Theorem, prove non-definability results, and derive interesting
-type isomorphisms.
+For each of our main examples we deduce free theorems that are
+consequences of our Abstraction Theorem, prove non-definability
+results, and derive interesting type isomorphisms.
 \end{itemize}
-We have fully formalized the framework and Abstraction Theorem using the Coq proof assistant,
-making use of strongly-typed term representations throughout~\cite{TypedSyntax}.
+We have fully formalised our framework and Abstraction Theorem in Coq,
+using strongly-typed term representations
+throughout~\cite{TypedSyntax}.
 

### 2012-07-09 paper/geometry-twocolumn.tex
Add caveats about quantifer-free ness

@@
   \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ has no
-  inhabitants. Intuitively, this is because there is no way to remove
-  the extra $t$ in $\tyPrim{vec}{t+t}$ using the vector operations. We
-  formalise this indefinability result in
-  \autoref{sec:types-indexed-abelian-groups} using a more
-  sophisticated relational interpretation of free translation
-  variables than the one we sketched above.
+  quantifier-free inhabitants\footnote{We define quantifier-free
+    programs in \thmref{thm:abstraction}.}. Intuitively, this is
+  because there is no way to remove the extra $t$ in
+  $\tyPrim{vec}{t+t}$ using the vector operations. We formalise this
+  indefinability result in \autoref{sec:types-indexed-abelian-groups-indef}
+  using a more sophisticated relational interpretation of free
+  translation variables than the one we sketched above.
 \end{example}
@@
   With the operations in \autoref{fig:real-ops}, it is not possible to
-  write a program with the following type that is not constantly zero:
+  write a quantifier-free program with the following type that is not
+  constantly zero:
   \begin{displaymath}
@@
 
-  In \autoref{sec:types-indexed-abelian-groups} we revisit Kennedy's
-  proof and show that even if we add square root as a primitive
-  operation---with the type above---then it is still not possible to
-  construct the cube root function. The indefinability of cube root is
-  related to the impossibility of trisecting an arbitrary angle by
-  ruler and compass constructions.
+  In \autoref{sec:types-indexed-abelian-groups-indef} we revisit
+  Kennedy's proof and show that even if we add square root as a
+  primitive operation---with the type above---then it is still not
+  possible to construct the cube root function. The indefinability of
+  cube root is related to the impossibility of trisecting an arbitrary
+  angle by ruler and compass constructions.
 %  is of interest due to its relevance to the classical problem of

### 2012-07-09 paper/section-introduction.tex
trivial word change
Change continuity to uniform continuity

@@
 relations that are not built compositionally---we can, for instance,
-use such relations to prove that there are no non-trivial functions of
+use such relations to prove that there are no non-trivial programs of
 type
@@
 related if $|x-y| < \rho(\alpha)$ for a positive real number
-$\rho(\alpha)$.  Rather beautifully, the standard notion of continuous
-functions can then be expressed as a type:
+$\rho(\alpha)$.  Rather beautifully, the standard notion of uniform
+continuity can then be expressed as a type:
 \begin{displaymath}
@@
 refined type system for information flow based on logic, and a simple
-type system for basic continuity analysis.
+type system for basic uniform continuity analysis.
 \item 

### 2012-07-09 paper/section-general-framework.tex

@@
 %introduction of existentials more, perhaps forward ref to their use
+We say that a type is \emph{quantifier-free} if it does not contain
+ether universal or existential quantification.
 
@@
 \begin{example}[Two-Dimensional Geometry]
-  Example~\autoref{ex:two-dim-geo-axioms}
+  \exref{ex:two-dim-geo-axioms}
 %  The two-dimensional geometry instantiation of the general framework
@@
 in terms of the interpretations of $e_1,...,e_n$. This approach
-suffices to prove free theorems and type isomorphisms for
-instantiations of our general framework. However, to prove
-non-definability results, like the one in
+suffices to prove free theorems and type isomorphisms. However, to
+prove non-definability results, like the one in
 \exref{ex:uninhabited-type}, we need a more refined interpretation
-that accounts more precisely for the definable programs permitted by
-the algebraic indexing. In this section, we identify the 
-properties we require of relational environments in order for
-the abstraction theorem to hold. In \autoref{sec:constr-rel-env}, we
-give a general construction of relational environments and use it 
-to prove several non-definability results.
+that accounts for the definable programs permitted by the algebraic
+indexing. Here, we identify the properties we require of relational
+environments for the abstraction theorem to hold. In
+\autoref{sec:constr-rel-env}, we give two generic constructions of
+relational environments.
 
@@
 environments for a context $\Delta$. Given %a relational environment
-$\rho \in \mathrm{RelEnv}(\Delta')$ and %a simultaneous substitution
+$\rho \in \mathrm{RelEnv}(\Delta)$ and %a simultaneous substitution
 $\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
 relational environment (with an abuse of notation) $\rho \circ \sigma
-\in \mathrm{RelEnv}(\Delta)$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
+\in \mathrm{RelEnv}(\Delta')$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
 \sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$.
@@
 operations that we assume for our system, we may have different sets
-of relational environments that enforce different invariants. We
-require that the sets of relational environments are \emph{valid} in
-the following sense:
+of relational environments that enforce different invariants. 
+% We
+% require that the sets of relational environments are \emph{valid} in
+% the following sense:
 
@@
   A family of sets of relational environments $\relEnv{E}(\Delta)
-  \subseteq \mathrm{RelEnv}(\Delta)$ is \emph{valid} if the following
+  \subseteq \mathrm{RelEnv}(\Delta)$ is \emph{substitutive} if these
   two properties are satisfied:
@@
   \item If $\rho_1 \in \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2
-    \in \relEnv{E}(\Delta)$, and $\Delta' \vdash \sigma \Rightarrow
-    \Delta$ is such that $\rho_1 \circ \pi_{i\mathord:s'} = \rho_2
+    \in \relEnv{E}(\Delta)$, and $\Delta \vdash \sigma \Rightarrow
+    \Delta'$ is such that $\rho_1 \circ \pi_{i\mathord:s'} = \rho_2
     \circ \sigma$, then there exists a $\rho \in \relEnv{E}(\Delta,
@@
   \end{enumerate}
+  A family $\relEnv{E}$ is \emph{pre-substitutive} if only the first
+  point holds.
 \end{definition}
@@
     \rsem{B}{\relEnv{E}}{\rho}$;
-  \item If $\Delta' \vdash A \isType$ and $\Delta \vdash \sigma
-    \Rightarrow \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
+  \item If $\Delta' \vdash A \isType$ and either $\relEnv{E}$ is
+    substitutive or $\relEnv{E}$ is pre-substitutive and $A$ is
+    quantifier-free, then for all $\Delta \vdash \sigma \Rightarrow
+    \Delta'$ and $\rho \in \relEnv{E}(\Delta)$,
     $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
@@
   \centering
-  {\tiny
+  {\small
   \begin{mathpar}
@@
     \inferrule* [right=Case]
-    {\Delta; \Gamma \vdash M : A + B \\
-      \Delta; \Gamma, x : A \vdash N_1 : C \\
+    {\Delta; \Gamma \vdash M : A + B \\\\
+      \Delta; \Gamma, x : A \vdash N_1 : C \\\\
       \Delta; \Gamma, y : B \vdash N_2 : C}
@@
     \inferrule* [right=ExUnpack]
-    {\Delta; \Gamma \vdash M : \exists i\mathord:s. A \\
+    {\Delta; \Gamma \vdash M : \exists i\mathord:s. A \\\\
       \Delta, i\mathord:s; \pi_{i\mathord:s}^*\Gamma, x : A \vdash N : \pi_{i\mathord:s}^*B}
@@
 \begin{theorem}[Abstraction]\label{thm:abstraction}
-  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
-  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
-  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
-  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
+  Assume $\relEnv{E}$ is substitutive. If $\Delta; \Gamma \vdash M :
+  A$, then for all $\rho \in \relEnv{E}(\Delta)$ and $\eta_1, \eta_2
+  \in \ctxtSem{\Gamma}$ such that $(\eta_1, \eta_2) \in
+  \rsem{\Gamma}{\relEnv{E}}\rho$, we have $(\tmSem{M}\eta_1,
+  \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
+
+  If $\relEnv{E}$ is pre-substitutive, then the above restricts to the
+  case when $\Gamma$ only contains types of the form $\forall
+  \vec{i}\mathord:\vec{s}. B$, where $B$ is quantifier-free, $A$ is
+  quantifier-free and $M$ contains no uses of \TirName{UnivAbs},
+  \TirName{ExPack} or \TirName{ExUnpack} and all uses of
+  \TirName{UnivApp} are fully applied.
 \end{theorem}
 
-We use the relational interpretation to define the notion of semantic
-equivalence, which is a sound approximation of contextual equivalence,
-as a consequence of \thmref{thm:abstraction}.
+We refer to programs satisfying the conditions in the second part of
+\thmref{thm:abstraction} as being in the \emph{quantifier-free}
+fragment of our framework. These conditions are very similar to those
+used in the fragment of System F covered by Hindley-Milner type
+inference.
+
+We use the relational interpretation to define semantic equivalence,
+which is a sound approximation of contextual equivalence, as a
+consequence of \thmref{thm:abstraction}.
 \begin{definition}
@@
   \defref{defn:ctxt-equiv}. Let $\relEnv{E}_{\mathit{ops}}$ be a
-  family of sets of relational environments such that for all $\Delta$
-  and $\rho \in \relEnv{E}_{\mathit{ops}}(\Delta)$,
+  substitutive family of sets of relational environments such that for
+  all $\Delta$ and $\rho \in \relEnv{E}_{\mathit{ops}}(\Delta)$,
   $(\eta_{\mathit{ops}}, \eta_{\mathit{ops}}) \in
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{A General Construction of Relational Environments}
+\subsection{Construction of Relational Environments}
 \label{sec:constr-rel-env}
 
+\newcommand{\Gen}{\mathrm{Gen}}
 \newcommand{\Free}{\mathrm{Free}}
@@
 
-We now give a general construction of valid families of sets of
-relational environments 
-%(in the sense of \defref{defn:valid-rel-env-family}) 
-that allow us to derive
-useful consequences of \thmref{thm:abstraction}, e.g.,
-%in particular, 
-%we wish to show 
-that certain types are uninhabited. To %motivate 
-this end, consider the
-%following 
-type $\forall t\mathord:\SynTransl{2}.\ \tyPrim{vec}{t + t} \to
-\tyPrim{vec}{t}$ involving vectors indexed by translations.
-% as in \autoref{sec:affine-vector-ops}.
-%\begin{displaymath}
-%  \forall t\mathord:\SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}
-%\end{displaymath}
+We give two constructions of families of sets of relational
+environments that allow us to derive useful consequences of
+\thmref{thm:abstraction}.
+
+\paragraph{Compositional Relational Environments}
+Our compositional relational environments are defined in terms of
+models. To each sort $s \in \SortSet$ we assign a carrier set
+$\semSort{s}$. We extend this assignment to index contexts and other
+sequences of sorts (e.g., type arities) using the cartesian product of
+sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} = \semSort{s_1}
+\times ... \times \semSort{s_n}$. For each index operation $\texttt{f}
+\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) = ([s_1,...,s_n],s)$,
+we assume a function $\semIndexExp{\texttt{f}} : \semSort{s_1,...,s_n}
+\to \semSort{s}$. For each well-sorted index expression $\Delta \vdash
+e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
+\semSort{s}$ by recursion on the structure of $e$. We further assume
+that for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
+\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$. A
+\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
+\semIndexExp{\cdot}_{\mathcal{M}})$ satisfying all the axioms in
+$\IndexAxiomSet$.
+
+Now fix a model $\mathcal{M}$ and assume, for each primitive type
+$\tyPrimNm{X}$, a relational interpretation parameterised by elements
+from the model: $R_{\tyPrimNm{X}} :
+\semSort{\primTyArity(\tyPrimNm{X})}_{\mathcal{M}} \to
+\Rel(\tyPrimSem{\tyPrimNm{X}})$. Our family of sets of relational
+environments is $\relEnv{E}_{\mathcal{M},R}(\Delta) = \{ \rho^\delta
+\sepbar \delta \in \semSort{\Delta}_{\mathcal{M}} \}$, where
+$\rho^\delta\ \tyPrimNm{X}\ (e_1,...,e_n) =
+R_{\tyPrimNm{X}}(\semIndexExp{e_1}_{\mathcal{M}}\delta, ...,
+\semIndexExp{e_n}_{\mathcal{M}}\delta)$.
+
+\begin{theorem}
+  The families $\relEnv{E}_{\mathcal{M},R}$ are substitutive.
+\end{theorem}
+
+To satisfy the hypothesis of \thmref{thm:abstraction} we still need to
+prove, for each instantiation of our general framework, that the
+program-level primitive operations preserve all the relations.
+\fixme{This is the content of lemma X, lemma Y, lemma Z, below}.
+
+\paragraph{Non-compositional Relational Environments}
+We now construct relational environments that can be used to prove
+non-definability results. To this end, consider the typing
+$t\mathord:\SynTransl{2}; \Gamma_{\Geom} \vdash M : \tyPrim{vec}{t +
+  t} \to \tyPrim{vec}{t}$ involving vectors indexed by translations.
 In \exref{ex:uninhabited-type} we claimed that this type is
@@
 way of cancelling the additional $t$ in the argument type to get the
-result type. More generally, our operations on vectors 
-only allow us to combine indices according to the group operations. Thus,
-if the index of the result type of a function does not lie within the
-free abelian group generated by the indices of the input types, then
-there is no way to write a program with this %given 
-type. We now
-develop this idea to generate a family of sets of relational
-environments that is valid in the sense of
-\defref{defn:valid-rel-env-family}. The key idea is to assign
-different relational interpretations to types with indices that lie
-outside some fintely generated group of index expressions.
-
-\paragraph{Some Universal Algebra}
-We make use of some concepts of universal algebra, which now recall.
-A \emph{model} for our indexing expressions and their equality is
-defined as follows. To each sort $s \in \SortSet$ we assign a carrier
-set $\semSort{s}$. We extend this assignment to index contexts and
-other sequences of sorts (e.g., type arities) using the cartesian
-product of sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} =
-\semSort{s_1} \times ... \times \semSort{s_n}$. For each index
-operation $\texttt{f} \in \IndexOpSet$ with $\indexOpArity(\texttt{f})
-= ([s_1,...,s_n],s)$, we assume a function $\semIndexExp{\texttt{f}} :
-\semSort{s_1,...,s_n} \to \semSort{s}$. For each well-sorted index
-expression $\Delta \vdash e : s$, we assign a function
-$\semIndexExp{e} : \semSort{\Delta} \to \semSort{s}$ by recursion on
-the structure of $e$. We further assume that for each axiom $\Delta
-\vdash e \stackrel{ax}\equiv e' : s \in \IndexAxiomSet$, we have
-$\semIndexExp{e} = \semIndexExp{e'}$. A \emph{model} is a pair
-$\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
-\semIndexExp{\cdot}_{\mathcal{M}})$ 
-%such that all the axioms in
-satisfying all the axioms in
-$\IndexAxiomSet$.
-% are satisfied.
-
-Given a pair of models $\mathcal{M}_1 = (\semSort{\cdot}_1,
-\semIndexExp{\cdot}_1)$ and $\mathcal{M}_2 = (\semSort{\cdot}_2,
-\semIndexExp{\cdot}_2)$, a \emph{homomorphism} $h : \mathcal{M}_1 \to
-\mathcal{M}_2$ is a family of functions $\{ h_s : \semSort{s}_1 \to
-\semSort{s}_2 \}_{s \in \SortSet}$ such that for each $\texttt{f} \in
-\IndexOpSet$, we have $h_s(\semIndexExp{\texttt{f}}_1 (x_1,...,x_n)) =
-\semIndexExp{\texttt{f}}_2(h_{s_1}x_1, ..., h_{s_n}x_n)$.
-
-Let $A = \{ A_s \}_{s \in \SortSet}$ be a family of finite sets
-indexed by sorts. We construct an index context $\Delta_A$ which has
-variables $a\mathord:s$ for each $a \in A_s$. The \emph{free} model
-over $A$, denoted $\Free(A)$, is defined by setting
-$\semSort{s}_{\Free(A)} = \{e \sepbar \Delta_A \vdash e :
-s\}/\mathord\equiv$, where we quotient by the equality judgment
-defined in \autoref{sec:type-equality}. The free model has
-$\semIndexExp{\texttt{f}}_{\Free(A)}(e_1,..,e_n) =
-\texttt{f}(e_1,...,e_n)$.
-
-\paragraph{Relational Environments} We now fix some model
-$\mathcal{M}$ to provide provide the underlying interpretation of
-index expressions. %As we indicated above, we wish 
-To discriminate
-between index expressions that are generated from some finite set of
-``input'' expressions and those that are not, %. Accordingly, 
-we assume that for each primitive type $\tyPrimNm{X}$ we have two ways of
-assigning a relational meaning to that type. When the index arguments
-of the type are within the generated set, we use a function
-$R_{\tyPrimNm{X}} \in
-\semSort{\primTyArity(\tyPrimNm{X})}_{\mathcal{M}} \to
-\Rel(\tyPrimSem{\tyPrimNm{X}})$ that assigns a relational meaning to
-every element in the interpretation of the type's argument sorts. When
-the index arguments are not within the generated set we default to a
-constant relational meaning $R^\bullet_{\tyPrimNm{X}} \in
-\Rel(\tyPrimSem{\tyPrimNm{X}})$.
-%We now use $R_{\tyPrimNm{X}}$ and $R^\bullet_{\tyPrimNm{X}}$ to
-%construct the relational environments that we are interested in.  
-Now, fix
-an index context $\Delta$, and let $A = \{ A_s \}_{s \in \SortSet}$ be a
-sort-indexed family of finite sets of expressions such that if $e \in
-A_s$, then $\Delta \vdash e : s$. Let $h : \Free(A) \to \mathcal{M}$
-be a homomorphism. For each $A$ and $h$ we have a relational
-environment for the context $\Delta$:
+result type. More generally, our operations on vectors only allow us
+to combine indices according to the group operations. Thus, if the
+index of the result type of a function does not lie within the free
+abelian group generated by the indices of the input types, then there
+is no way to write a program with this %given
+type. We now develop this idea to generate a family of sets of
+relational environments. The key idea is to assign different
+relational interpretations to types with indices that lie outside some
+finitely generated collection of index expressions.
+
+These relational environments are non-compositional in the sense that
+elements of $\relEnv{E}(\Delta,\Delta')$ are not only constructed from
+elements of $\relEnv{E}(\Delta)$ and $\relEnv{E}(\Delta')$. As a
+consequence of their non-compositionality, these relational
+environments will only be \emph{pre-substitutive} in the sense of
+\defref{defn:valid-rel-env-family}. Therefore our non-definability
+results are subject to the conditions in \thmref{thm:abstraction}.
+
+Given a set $S$ of index expressions that are all well-indexed in some
+index context $\Delta$, define $\Gen(S)$ to be the set of terms that
+are generated by the elements of $S$ and the primitive index
+operations. We close under index expression equivalence to get the
+sets $\Gen^\equiv(S) = \{ e \sepbar \exists s.\ \exists e' \in
+\Gen(S).\ \Delta \vdash e \equiv e' : s \}$.
+
+We construct relational environments parameterised by sets of terms
+$S$. We assign the equality relation to primitive types whose index
+arguments are generated up to equivalence by $S$. For terms whose
+index arguments are not in $\Gen^\equiv(S)$, we require a default
+binary relation $R^\bullet_{\tyPrimNm{X}} \subseteq
+\Eq_{\tyPrimSem{\tyPrimNm{S}}}$. We define the non-compositional
+family of sets of relational environments as
+$\relEnv{E}_{R^\bullet}(\Delta) = \{ \rho^S \sepbar \forall e \in S.\
+\exists s. \Delta \vdash e : s \}$, where
 \begin{displaymath}
-  \rho^A_h\ \tyPrimNm{X}\ (e_1,...,e_n) = \left\{
-    \begin{array}{@{}l@{\hspace{0.5em}}l}
-      R^\bullet_{\tyPrimNm{X}} & \textrm{if any }e_i \not\in \Free(A) \\
-      R_{\tyPrimNm{X}}(h(e_1),...,h(e_n)) & \textrm{if all }e_i \in \Free(A)
+  \rho^S\ \tyPrimNm{X}\ (e_1,...,e_n) = \left\{
+    \begin{array}{ll}
+      \mathrm{Eq}_{\tyPrimSem{\tyPrimNm{X}}} & \textrm{if all}\ e_i \in \Gen^\equiv(S) \\
+      R^\bullet_{\tyPrimNm{X}} & \textrm{otherwise}
     \end{array}
@@
 \end{displaymath}
-Collecting all these relational environments together yields a family
-of relational environments, parameterised by our choice of model,
-$\mathcal{M}$, and relational interpretations of primitive types
-$R^\bullet,R$:
-\begin{displaymath}
-  \relEnv{E}_{\mathcal{M},R,R^\bullet}(\Delta) = \{ \rho^A_h \sepbar \forall s.A_s \subseteq \semSort{s}_{\Free(\Delta)}, h : \Free(A) \to \mathcal{M} \}
-\end{displaymath}
 
 \begin{theorem}
-  The family $\relEnv{E}_{\mathcal{M},R,R^\bullet}$ of sets of
-  relation environments is valid in the sense of
-  \defref{defn:valid-rel-env-family}.
+  The families $\relEnv{E}_{R^\bullet}$ are pre-substitutive.
 \end{theorem}
 
-To satisfy the hypothesis of \thmref{thm:abstraction} we still need to
-prove, for each instantiation of our general framework, that the
-program-level primitive operations preserve all the relations.
-\fixme{This is the content of lemma X, lemma Y, lemma Z, below}.
+% FIXME: look forward to the applications
 

### 2012-07-09 paper/section-metric.tex
Patty's changes to the metric section

@@
 effect that programs have on the distances between values. For
-example, a program $M$ of type
+example, a program $M$ of type:
 \begin{displaymath}
@@
 A type system with a relational interpretation that tracks distances
-between values has previously been investigated by Reed and Pierce
+between values has %previously 
+been investigated by Reed and Pierce
 \cite{reed10distance} in the setting of differential privacy. Their
 system uses a linear type discipline to ensure that all programs are
-non-expansive (i.e., the distance between the outputs is no greater
-than the distance between the inputs). Here, we instantiate the
-framework from \autoref{sec:a-general-framework} in order to express
-non-expansivity, and more, through algebraically indexed types.
+non-expansive (i.e., the distance between the outputs is no more greater
+than the distance between the inputs). Here, we instantiate our general
+framework %from \autoref{sec:a-general-framework} %in order 
+to express
+non-expansivity, and more, via algebraically indexed types.
 
 \paragraph{Instantiation of the General Framework}
-We assume a single indexing sort $\metricSort$, intended to represent
-positive, non-zero real numbers. For the index operations, we assume
-the operations $\min, \max, +$ and multiplication by constant
-reals. There is a single primitive type $\tyPrimNm{real}$, with
-$\primTyArity(\tyPrimNm{real}) = [\metricSort]$. The primitive
-operations $\Gamma_{\Metric}$ are as follows, where $c$ stands for
-arbitrary real-valued constants.
+We assume a single indexing sort $\metricSort$ %, intended to 
+to represent positive, non-zero real numbers. For the index
+operations, we assume the operations $\min, \max, +$ and
+multiplication by constant reals. There is a single primitive type
+$\tyPrimNm{real}$ with $\primTyArity(\tyPrimNm{real}) =
+[\metricSort]$. The primitive operations $\Gamma_{\Metric}$ are as
+follows, where $c$ stands for arbitrary real-valued constants:
 \begin{displaymath}
@@
 \autoref{sec:constr-rel-env} with the following interpretations of the
-primitive type $\tyPrimNm{real}$:
-\begin{displaymath} %FIXME: define \Eq earlier on
-  \begin{array}{l@{\hspace{2em}}l}
-    R^\bullet_{\tyPrimNm{real}} = \Eq_\mathbb{R} &
-    R_{\tyPrimNm{real}}(\epsilon) = \{ (x,x') \sepbar |x - x'| < \epsilon \}
-  \end{array}
-\end{displaymath}
-
+primitive type $\tyPrimNm{real}$: $R^\bullet_{\tyPrimNm{real}} = \Eq_\mathbb{R}$ and $R_{\tyPrimNm{real}}(\epsilon) = \{ (x,x') \sepbar |x - x'| < \epsilon\}$.
+%\begin{displaymath} %FIXME: define \Eq earlier on
+%  \begin{array}{l@{\hspace{2em}}l}
+%    R^\bullet_{\tyPrimNm{real}} = \Eq_\mathbb{R} &
+%    R_{\tyPrimNm{real}}(\epsilon) = \{ (x,x') \sepbar |x - x'| < \epsilon \}
+%  \end{array}
+%\end{displaymath}
+We have:
 \begin{lemma}
@@
 \begin{example}[Uniform Continuity]
-  By making use of the existential types in our framework, we can
-  state the standard $\epsilon$-$\delta$ definition of
+  %By making use of 
+Using %the 
+existential types, % in our framework, 
+we can
+  state the standard 
+$\epsilon$-$\delta$ definition of
   \emph{uniform} continuity as a type:
@@
   \end{displaymath}
-  Note that this definition differs from the $\epsilon$-$\delta$
-  definition of continuity by the order of quantification. For
-  continuity, the $\forall x$ is before the $\exists \delta$, so the
-  distance $\delta$ may depend on the value of $x$. We suspect that to
+  This definition differs from the $\epsilon$-$\delta$
+  definition of (regular) continuity by the order of quantification: 
+  %For continuity, 
+  there, $\forall x$ comes before $\exists \delta$, so 
+  the distance 
+  $\delta$ may depend on %the value of 
+  $x$. We suspect that to
   express standard continuity as a type would require some form of
   type dependency. Chaudhuri, Gulwani and Lublinerman
-  \cite{chaudhuri10continuity} have presented a program logic based
+  \cite{chaudhuri10continuity} have given a program logic based
   approach to verifying the continuity of programs.

### 2012-07-09 paper/section-monoid.tex
Patty's changes to the monoid section

@@
 
-Above, we investigated types indexed by abelian group expressions,
+Above we investigated types indexed by abelian group expressions,
 whose elements are taken from some group. We now switch from abelian
@@
 ``taint'' markers: when strings are appended their taint markers are
-combined. Thanks to the information in the indexes, we are able to
+combined. Thanks to the information in the indices, we are able to
 prove an indefinability result (\thmref{thm:monoid-indefinability}).
 
-We assume a single sort $\mathsf{M}$, with the monoid operations of
-unit and multiplication, and the commutative monoid axioms. There is a
-single primitive type $\tyPrimNm{M}$, with $\primTyArity(\tyPrimNm{M})
+For this, we assume a single sort $\mathsf{M}$, with the monoid operations of
+unit and multiplication and the commutative monoid axioms, and a % There is a
+single primitive type $\tyPrimNm{M}$ with $\primTyArity(\tyPrimNm{M})
 = [\mathsf{M}]$. For the index-erasure semantics, we assume some (not
-necessarily commutative) monoid $M$, which we also write
+necessarily commutative) monoid $M$, which we 
+also write
 multiplicatively. The collection of primitive operations
-$\Gamma_{\mathit{Mon}}$ is simply the two monoid operations:
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{2em}}r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    1 & \tyPrim{M}{1}, &
-    (\cdot) & \forall a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{a b}
-  \end{array}
-\end{displaymath}
-which we interpret using the monoid structure of $M$, giving an
-environment $\eta_\Mon \in \ctxtSem{\Gamma_\Mon}$.
+$\Gamma_{\mathit{Mon}}$ %is simply 
+has the two monoid operations
+%\begin{displaymath}
+ % \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{2em}}r@{\hspace{0.5em}:\hspace{0.5em}}l}
+%    1 & \tyPrim{M}{1}, &
+%    (\cdot) & \forall a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{a b}
+%  \end{array}
+%\end{displaymath}
+$1 : \tyPrim{M}{1}$ and $(\cdot) : \forall
+a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to
+\tyPrim{M}{a b}$, which we interpret using the monoid structure of
+$M$, giving an environment $\eta_\Mon \in \ctxtSem{\Gamma_\Mon}$.
 
-For the sets of relational environments, we make use of the
+For the sets of relational environments, we use %make use of 
+the
 construction in \autoref{sec:constr-rel-env}, setting the chosen model
-$\mathcal{M}$ to be the \emph{centre} $Z(M)$ of $M$ (recall that the
-centre of a monoid is the set of all elements that commute with every
-other element). We then set $R_{\tyPrimNm{M}}(m) = \{ (n,mn) \sepbar n
-\in M \}$ and $R^\bullet_{\tyPrimNm{M}} = \emptyset$. By setting the
-model $\mathcal{M}$ to only be the elements of $M$ that commute with
-everything else, we ensure that the following lemma holds:
+$\mathcal{M}$ to be the centre $Z(M)$ of $M$. % (recall that the
+%centre of a monoid is the set of all elements that commute with every
+%other element). 
+Setting $R_{\tyPrimNm{M}}(m) = \{ (n,mn) \sepbar n
+\in M \}$ and $R^\bullet_{\tyPrimNm{M}} = \emptyset$
+%By setting the model $\mathcal{M}$ to be only the elements of $M$ that
+%commute with everything else, we ensure that the following lemma
+%holds:
+gives:
 \begin{lemma}
@@
   multiset union as the monoid multiplication, which we write as
-  $\uplus$), $M$ is itself a monoid. The centre of $M$, $Z(M)$,
+  $\uplus$), $M$ is itself a monoid. The centre $Z(M)$ of $M$, 
   consists of all elements of the form $([], t)$.
@@
   \begin{equation}
-    ; \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
+    ; \Gamma_\Mon \vdash M : \forall
+    i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to
+    \tyPrim{M}{e_n} \to \tyPrim{M}{e}
   \end{equation}
   where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e = i_1^{b_1}\dots
-  i_m^{b_m}$, iff there is a solution $\vec{x} = (x_1,...,x_n)$ in the
+  i_m^{b_m}$ iff there is a solution $\vec{x} = (x_1,...,x_n)$ in the
   natural numbers to $A x^\top = b$.
@@
   program with the type $\forall m_1,m_2 \mathord: \mathsf{M}.\
-  \tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_1}$, showing that it is
+  \tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_1}$, and is thus % shows that it is
   impossible to convert a doubly tainted value to a singly tainted
   one.
-
-  In contrast to the case of abelian group indexed types, the lack of
-  inverses in the monoid indexed setting means that the type $\forall
-  m_1,m_2\mathord:\mathsf{M}.\ \tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_2}
-  \to \tyPrim{M}{m_1}$ is also uninhabited. There are no naturals
-  $a_1,a_2$ such that $(m_1m_2)^{a_1}m_2^{a_2} \equiv m_1$.
+%
+  By contrast with the situation for abelian group indexed types, the
+  lack of inverses in the monoid indexed setting means that the type
+  $\forall m_1,m_2\mathord:\mathsf{M}.\ \tyPrim{M}{m_1m_2} \to
+  \tyPrim{M}{m_2} \to \tyPrim{M}{m_1}$ is also uninhabited since there
+  are no naturals $a_1,a_2$ such that $(m_1m_2)^{a_1}m_2^{a_2} \equiv
+  m_1$.
 \end{example}

### 2012-07-09 paper/section-monoid.tex
Fixes to the monoid section to account for the restrictions on non-definability results

@@
 
-For the sets of relational environments, we use %make use of 
-the
+For the relational environments, we use %make use of 
+the compositional
 construction in \autoref{sec:constr-rel-env}, setting the chosen model
@@
 Setting $R_{\tyPrimNm{M}}(m) = \{ (n,mn) \sepbar n
-\in M \}$ and $R^\bullet_{\tyPrimNm{M}} = \emptyset$
+\in M \}$
 %By setting the model $\mathcal{M}$ to be only the elements of $M$ that
@@
 gives:
-\begin{lemma}
+\begin{lemma}\label{lem:monoid-ops-related}
   For all $\Delta$ and $\rho \in \relEnv{E}_\Mon(\Delta)$,
@@
   natural numbers $A$ and $m$-by-$1$ matricies of natural numbers $b$,
-  there exists a program:
+  there exists an $M$ in the quantifier-free fragment, with
   \begin{equation}
-    ; \Gamma_\Mon \vdash M : \forall
-    i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to
-    \tyPrim{M}{e_n} \to \tyPrim{M}{e}
+    i_1,\dots,i_m\mathord:\mathsf{M}; \Gamma_\Mon \vdash M : \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
   \end{equation}
@@
 
+The proof of this theorem uses the non-compositional relational
+environments from \autoref{sec:constr-rel-env}, with
+$R^\bullet_{\tyPrimNm{M}} = \emptyset$.
+
 \begin{example}
   By \thmref{thm:monoid-indefinability}, it is impossible to write a
-  program with the type $\forall m_1,m_2 \mathord: \mathsf{M}.\
-  \tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_1}$, and is thus % shows that it is
+  program with the type $\tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_1}$, and is thus % shows that it is
   impossible to convert a doubly tainted value to a singly tainted
@@
   lack of inverses in the monoid indexed setting means that the type
-  $\forall m_1,m_2\mathord:\mathsf{M}.\ \tyPrim{M}{m_1m_2} \to
-  \tyPrim{M}{m_2} \to \tyPrim{M}{m_1}$ is also uninhabited since there
-  are no naturals $a_1,a_2$ such that $(m_1m_2)^{a_1}m_2^{a_2} \equiv
-  m_1$.
+  $\tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_2} \to \tyPrim{M}{m_1}$ is also
+  uninhabited since there are no naturals $a_1,a_2$ such that
+  $(m_1m_2)^{a_1}m_2^{a_2} \equiv m_1$.
 \end{example}

### 2012-07-09 paper/section-monoid.tex
Remove stray equation number in the monoid section

@@
   there exists an $M$ in the quantifier-free fragment, with
-  \begin{equation}
+  \begin{displaymath}
     i_1,\dots,i_m\mathord:\mathsf{M}; \Gamma_\Mon \vdash M : \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
-  \end{equation}
+  \end{displaymath}
   where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e = i_1^{b_1}\dots

### 2012-07-09 paper/section-instantiations.tex
Small fixes in intro; fixed error in AG axioms of vector addition; new type isomorphisms maps

--- 2012-07-03 paper/section-instantiations.tex
@@
 \label{sec:instantiations}
+We now instantiate our general framework %of the previous section 
+with
+the indexing theory of \autoref{sec:motivating-examples}, and present
+more general and formally-justified theorems for free, type
+isomorphisms, and non-definability results.
 
-\fixme{explain the general setup, with a set of operations for each
-  instantiation.}
+%\begin{example}[Two-dimensional Geometry]
 
-\begin{lemma}
-  The thing about $\Gamma_\Geom$.
-\end{lemma}
+We define the ``standard'' model of the indexing theory for the
+two-dimensional geometry example as follows. Each of the sorts is
+interpreted just as its semantic counterpart: $\semSort{\SynTransl{2}}
+= \Transl{2}$, $ \semSort{\SynGL{2}} = \GL{2}$, $
+\semSort{\SynOrth{2}} = \Orth{2}$, and $ \semSort{\SynGL{1}} =
+\GL{1}$.
+%  \begin{displaymath}
+%    \begin{array}{@{}llll}
+%      \semSort{\SynTransl{2}} = \Transl{2}
+%      &
+%      \semSort{\SynGL{2}} = \GL{2}
+%      &
+%      \semSort{\SynOrth{2}} = \Orth{2}
+%      &
+%      \semSort{\SynGL{1}} = \GL{1}
+%    \end{array}
+%  \end{displaymath}
+  Each of the index operations (e.g.,~the group structure and
+  determinant) is interpreted by the intended semantic operation, and
+  clearly satisfy the axioms in \exref{ex:two-dim-geo-axioms}. The
+  assignment of relations for the primitive types $\tyPrimNm{vec}$ and
+  $\tyPrimNm{real}$ is given by: % the following definitions:
+  \begin{displaymath}
+    \begin{array}{ll}
+      R^{\bullet}_{\tyPrimNm{vec}} = \{( \vec{0}, \vec{0} ) \}&
+      R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
+      R^{\bullet}_{\tyPrimNm{real}} = \{(0,0)\} &
+      R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
+    \end{array}
+  \end{displaymath}
+  The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
+  %are as expected. % given the informal definitions we gave in
+formalise the informal ones from~\autoref{sec:motivating-examples}, and 
+  the definitions of %We have set
+  $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ 
+  %to relate $\vec{0}$ and $0$ to themselves to 
+  account for the
+  polymorphic zero constants we have assumed.
+%\end{example}
+
+%\begin{lemma}
+%  The thing about $\Gamma_\Geom$.
+%\end{lemma}
 
@@
 
-Consider the type of the area of a triangle function we defined in
+Consider the type of the area %of a triangle 
+function from
 \exref{ex:area-of-triangle-1}:
@@
 
-\subsection{Isomorphisms for Types Indexed by Abelian Groups}
+\subsection{Type Isomorphisms}
 \label{sec:types-indexed-abelian-groups}
+We say that two types $A$ and $B$ are isomorphic, and write $A \cong
+B$, if there exist maps between them that are mutually inverse with
+respect to contextual equivalence, i.e., if there are terms $\vdash I
+: A \to B$ and $\vdash J : B\to A$ such that $x:A\vdash
+J(I(x))\stackrel{ctx}\approx x : A$ and $y:B\vdash
+I(J(y))\stackrel{ctx}\approx y : B$.
+%\[
+%x:A\vdash J(I(x))\stackrel{ctx}\approx x : A
+%\text{ and }y:B\vdash I(J(y))\stackrel{ctx}\approx y : B.
+%\]
+That $\cong$ is a congruence relation on types is straightforward. We
+can also derive isomorphisms that are independent of the indexing
+theory, such as $A\times B \cong B \times A$, and $\forall
+i\mathord:s.A\to B\cong A\to\forall i\mathord:s.B$ for $i$ not free in
+$A$.
 
-\begin{enumerate}
-\item Two cases: when we have a total ``division'' operation, and when
-  we don't.
-\item Type isomorphisms, based on Smith Normal Form
+Types indexed by abelian groups induce a particularly rich theory of
+type isomorphisms; previous work on
+units-of-measure~\cite{kennedy97relational} relates these to
+Buckingham's theorem from dimensional analysis. Here we consider the
+additive abelian groups of translations and the multiplicative abelian
+group of scalings.
+
+\paragraph{Translations} 
+Consider first the group $\Transl{2}$ of translations from
+\autoref{sec:type-system-geom-intro} and 
+\autoref{sec:affine-vector-ops}. For any quantifier-free type $A$
+we have: % the isomorphism:
+\begin{eqnarray*}
+\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{t}\to A
+& \cong &
+A[0/t]
+\end{eqnarray*}
+where $A[0/t]$ denotes substitution of the identity translation $0$
+for index variable $t$ in type $A$. This isomorphism formalises the
+equivalence between \emph{coordinate-free} geometry, expressed by a
+polymorphic type whose first argument can be thought of as the
+\emph{origin} with respect to which $A$ is specified, and a
+\emph{coordinate-based} geometry, expressed by the type $A[0/t]$, in
+which the origin is fixed at $(0,0)$. It is sometimes said that an
+affine space is a vector space that's forgotten its origin; we have
+captured this in a type isomorphism.
+
+\exref{ex:type-iso} is a special case of this isomorphism; another instance is the
+type of vector addition:
+\begin{eqnarray*}
+\forall t_1\mathord:\SynTransl{2}.\forall t_2\mathord:\SynTransl{2}.\tyPrim{vec}{t_1}\to\tyPrim{vec}{t_2}\to\tyPrim{vec}{t_1+t_2}\\
+\cong\quad
+\forall t_1\mathord:\SynTransl{2}.\tyPrim{vec}{t_1}\to\tyPrim{vec}{t_1}\quad\cong\quad
+\tyPrim{vec}{0}
+\end{eqnarray*}
+\noindent
+\newcommand{\transup}[1]{\uparrow_{#1}}
+\newcommand{\transdn}[1]{\downarrow_{#1}}
+The maps associated with this isomorphism are 
+$I = \lambda x. x\ [0]\ 0$ and $J = \lambda y. \Lambda t.\lambda v. \transdn A (y)$,
+%\begin{eqnarray*}
+%I = \lambda x. x\ [0]\ 0 & \text{ and } &
+%J = \lambda y. \Lambda t.\ \lambda v. \trans A v t(y)
+%\end{eqnarray*}
+where terms $\transdn A : A\to A[0/t]$ and $\transup A : A[0/t] \to A$ are defined by induction over the syntax of $A$:
+\begin{eqnarray*}
+\transdn{1} & = & \lambda x. x\\ 
+\transdn{A_1\times A_2} & = & \lambda p. (\transdn{A_1}(\pi_1 p),\transdn{A_2}(\pi_2 p))\\ 
+\transdn{A_1 + A_2} & = & \lambda x.
+\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn{A_1}y)); \textrm{inr}\ y.\textrm{inr}(\transdn{A_2}y)\\ 
+\transdn{A_1\to A_2} & = & \lambda f. \lambda x. (\transdn{A_2} (f(\transup{A_1}(x))))\\ 
+\transdn{\tyPrim{vec}{e}} & = & \lambda x. x - kv\\ 
+\transup{\tyPrim{vec}{e}} & = & \lambda x. x + kv\\ & & \text{ if }e \equiv k t + e'\text{ for
+  some }k\in\setOfIntegers \text{ and }e'\text{ with }t\notin e'\\
+\end{eqnarray*}
+To show $I(J(y))$ is contextually equivalent to $y$ is an easy
+induction on $A$ using the fact that the vector $0$
+is an identity for vector addition. To show $J(I(x))$ is contextually
+equivalent to $x$ we must apply the abstraction theorem and use
+the fact that $\tmSem{\transup A}$ and $\tmSem{\transdn A}$ 
+considered as relations are simply instances of $\rsem A{}{}$.
+%\[
+%$\rsem A {t \mapsto v} {} = \{ (x, \tmSem{\transup A (x)}) \mid x \in
+%\tySem A\}$.
+%\]
+
+\paragraph{Scalings}
+Now consider the group $\GL{1}$ of scalings. This time, we can treat
+the real-valued argument to a function as a `unit of measure' with
+which to scale the result. 
+%The small fly in the ointment is that the
+Although the argument might be zero, and this cannot be used for scaling,
+%Nevertheless, 
+we still can obtain the following slightly more complicated
+isomorphism:
+\begin{eqnarray*}
+\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A
+& \cong &
+A[1/s]\times \forall s\mathord:\SynGL{1}.A
+\end{eqnarray*}
+The type of $\mathrm{areaCircle}$ from the introduction is one instance:
+\begin{eqnarray*}
+\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s\cdot
+  s}\\ \quad\cong\quad \tyPrim{real}{1}\times(\forall
+s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s})\\ \quad\cong\quad
+\tyPrim{real}{1}\times 1\quad\cong\quad \tyPrim{real}{1}
+\end{eqnarray*}
+%where %the isomorphism 
+The abstraction theorem is used to prove
+$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s}\cong 1$.
+% is proved using
+%the abstraction theorem.
+
+%\begin{enumerate}
+%\item Two cases: when we have a total ``division'' operation, and when
+%  we don't.
+%\item Type isomorphisms, based on Smith Normal Form
 % \item Non-definability results, based on the non-compositional

### 2012-07-09 paper/section-introduction.tex
Small fixes in intro; fixed error in AG axioms of vector addition; new type isomorphisms maps

@@
 
-In more complex examples the invariants \emph{compose} in various
-ways, with the compositions satisfy algebraic laws. For example, 
+Transformations typically \emph{compose} in various
+ways, and the compositions satisfy algebraic laws. For example, 
 we can assign a function that computes the area of a circle given its
@@
 \[
-\mathrm{areaCircle} : \forall\alpha\mathord:\SynScal.\tyPrim{real}{\alpha}\to
+\mathrm{areaCircle} : \forall\alpha\mathord:\SynGL{1}.\tyPrim{real}{\alpha}\to
 \tyPrim{real}{\alpha\cdot\alpha}
@@
 This captures the fact that the area of a circle varies as the square
-of its radius, i.e., for any $s>0$, $\mathrm{areaCircle}\;(s\cdot r) =
-s^2\cdot \mathrm{areaCircle}\;r$.  Here, $\alpha$ is interpreted as
+of its radius, i.e., $\mathrm{areaCircle}\;(s\cdot r) =
+s^2\cdot \mathrm{areaCircle}\;r$ for any $s\neq 0$.  Here, $\alpha$ is interpreted as
 the \emph{units of measure} of the argument to $\mathrm{areaCircle}$,
 and `$\cdot$' composes units using the product. We can also add an
-inverse operation and identity unit of measure, and then impose the
+inverse operation and identity unit of measure $1$, and then impose the
 algebraic laws of abelian groups. This permits identification of, for
@@
 the operation that takes two points and computes their vector
-`difference' can be assigned the type
+difference can be assigned the type
 $\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\times\tyPrim{vec}\alpha\to\tyPrim{vec}0$,
@@
 variable. For example, values $v$ and $w$ of type
-$\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\to
+$\forall\alpha\mathord:\SynTransl{2}.\tyPrim{vec}\alpha\to
 \tyPrim{vec}\alpha$ are related when they are related at type
@@
 indices can be built up compositionally, by recursion over the
-syntactic structure. For example, we have $\rho(\alpha\cdot\alpha) =
-\rho(\alpha)\cdot\rho(\alpha)$ where product of units-of-measure is
+syntactic structure. For example, we have $\rho(\alpha\cdot\beta) =
+\rho(\alpha)\cdot\rho(\beta)$ where product of units-of-measure is
 interpreted by product of scale factors. For more sophisticated
 applications, however, we need the fully generality of binary
-relations. First, as is also the case with type polymorphism, it
+relations. As is the case with type polymorphism, it
 provides the appropriate meaning for nested quantifiers in types such
 as
-$(\forall\alpha\mathord:\SynScal.\tyPrim{real}\alpha\to\tyPrim{real}\alpha)\to\tyBool$.
+$(\forall\alpha\mathord:\SynGL{1}.\tyPrim{real}\alpha\to\tyPrim{real}\alpha)\to\tyBool$.
 More interestingly, particular applications sometimes require
@@
 type
-$\forall\alpha\mathord:\SynScal.\tyPrim{real}{\alpha\cdot\alpha}\to\tyPrim{real}\alpha$.
+$\forall\alpha\mathord:\SynGL{1}.\tyPrim{real}{\alpha\cdot\alpha}\to\tyPrim{real}\alpha$.
 Finally, there are interesting applications in which even the basic
@@
 analogue of Reynolds' Abstraction Theorem. We give precise conditions
-on sets of relation environments under which our Abstraction Theorem
+on sets of relation environments under which our abstraction theorem
 holds.
@@
 For each of our main examples we deduce free theorems that are
-consequences of our Abstraction Theorem, prove non-definability
+consequences of our abstraction theorem, prove non-definability
 results, and derive interesting type isomorphisms.
@@
 using strongly-typed term representations
-throughout~\cite{TypedSyntax}.
+throughout~\cite{TypedSyntax}, available from
+\\\url{github.com/bobatkey/algebraically-indexed-types/tree/master/coq}
 

### 2012-07-09 paper/section-metric.tex
Couple of small fixes to the metric section

@@
     \underline{c} * & \forall \epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \to \tyPrim{real}{c\epsilon} \\
-    \mathrm{coerce} & \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\max(\epsilon_1,\epsilon_2)}
+    \mathrm{up} & \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\max(\epsilon_1,\epsilon_2)}
   \end{array}
@@
 type is just the set $\mathbb{R}$, so all except the last operation
-have straightforward interpretations. The $\mathrm{coerce}$ operation
-is interpreted just as the identity function. The index-erasure
+have straightforward interpretations. The $\mathrm{up}$ operation is
+interpreted just as the identity function. The index-erasure
 interpretations of the primitive operations are collected together
@@
 For the families of sets of relational environments
-$\relEnv{E}_\Metric$, we use the general construction in
-\autoref{sec:constr-rel-env} with the following interpretations of the
-primitive type $\tyPrimNm{real}$: $R^\bullet_{\tyPrimNm{real}} = \Eq_\mathbb{R}$ and $R_{\tyPrimNm{real}}(\epsilon) = \{ (x,x') \sepbar |x - x'| < \epsilon\}$.
+$\relEnv{E}_\Metric$, we use the compositional construction in
+\autoref{sec:constr-rel-env}, interpreting the primitive type
+$\tyPrimNm{real}$ with $R_{\tyPrimNm{real}}(\epsilon) = \{ (x,x')
+\sepbar |x - x'| < \epsilon\}$.
 %\begin{displaymath} %FIXME: define \Eq earlier on
@@
 %\end{displaymath}
-We have:
 \begin{lemma}

### 2012-07-09 paper/section-monoid.tex
Minor spacing changes to the monoid section

@@
 
-\begin{example}
 By selecting a suitable monoid $M$, we can simulate a programming
-  language that annotates strings with taint markers. Let $\Sigma$ be
-  a set of characters from some alphabet, and let $T$ be a set of
-  taint markers. Set $M = \Sigma^* \times
-  \mathcal{MS}_{\mathit{fin}}(T)$, where $\mathcal{MS}_{\mathit{fin}}$
-  denotes the finite power set operator. Since both $\Sigma^*$ and
-  $\mathcal{MS}_{\mathit{fin}}(T)$ are monoids (the latter with
-  multiset union as the monoid multiplication, which we write as
-  $\uplus$), $M$ is itself a monoid. The centre $Z(M)$ of $M$, 
-  consists of all elements of the form $([], t)$.
+language that annotates strings with taint markers. Let $\Sigma$ be a
+set of characters from some alphabet, and let $T$ be a set of taint
+markers. Set $M = \Sigma^* \times \mathcal{MS}_{\mathit{fin}}(T)$,
+where $\mathcal{MS}_{\mathit{fin}}$ denotes the finite power set
+operator. Since both $\Sigma^*$ and $\mathcal{MS}_{\mathit{fin}}(T)$
+are monoids (the latter with multiset union as the monoid
+multiplication, which we write as $\uplus$), $M$ is itself a
+monoid. The centre $Z(M)$ of $M$, consists of all elements of the form
+$([], t)$.
 
-  As an example of a free theorem for this instantiation of our
-  general framework, consider programs of the following type:
+As an example of a free theorem for this instantiation of our general
+framework, consider programs of the following type:
 \begin{displaymath}
@@
 \end{displaymath}
-  From this type we can derive the following free theorem about
-  $N$. For all $a, b \in \mathcal{MS}_{\mathit{fin}}(T)$, $(s_1,t_1),
-  (s_2,t_2) \in \Sigma^*$,
+From this type we can derive the following free theorem about $N$. For
+all $a, b \in \mathcal{MS}_{\mathit{fin}}(T)$, $(s_1,t_1), (s_2,t_2)
+\in \Sigma^*$,
 \begin{displaymath}
@@
 $b$.
-\end{example}
 
@@
 environments from \autoref{sec:constr-rel-env}, with
-$R^\bullet_{\tyPrimNm{M}} = \emptyset$.
-
-\begin{example}
-  By \thmref{thm:monoid-indefinability}, it is impossible to write a
-  program with the type $\tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_1}$, and is thus % shows that it is
-  impossible to convert a doubly tainted value to a singly tainted
-  one.
+$R^\bullet_{\tyPrimNm{M}} = \emptyset$. From the theorem we can see
+that it is impossible to write a program with the type
+$\tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_1}$, and is thus impossible to
+convert a doubly tainted value to a singly tainted one.
 %
@@
 $(m_1m_2)^{a_1}m_2^{a_2} \equiv m_1$.
-\end{example}
 

### 2012-07-09 paper/section-info-flow.tex

--- 2012-07-06 paper/section-info-flow.tex
@@
 
-\begin{enumerate}
-\item Make sure the logic is at least a meet semi-lattice
-\item 
-\end{enumerate}
+\newcommand{\infoflow}{\mathit{Log}}
 
-\begin{enumerate}
-\item Also makes use of the relational aspect, the previous stuff
-  could make do with unary logical predicates as the interpretations
-  of types, but the information flow properties make use of the
-  relational interpretation.
-\item Can also obtain a type non-inhabitation result based on logical
-  derivability
-\end{enumerate}
+We now apply our general framework to types that are indexed by
+logical propositions. By including a primitive type that represents
+logical truth, we can recover--through a construction due to Tse and
+Zdancewic \cite{tse04translating}--strong information flow properties
+of programs. As a result of our general framework being parameterised
+by the choice of equational theory, we can alter the logic that we use
+for reasoning about type equality, and hence alter the information
+flow properties of the system.
 
-\fixme{
-  \begin{enumerate}
-  \item Make the definition of control of information flow clearer
-  \item Make the example a bit more compelling
-  \item Put in some examples of composite principals formed from
-    boolean logic, and the use of boolean algebra to do reasoning in
-    the types. List the laws of boolean algebra, and the laws of
-    Heyting algebra. Need to explain the difference between 
-  \item Possibly use singleton types to be able to do more interesting
-    security policies?
-  \item Put in the formal statement of the information flow property
-    and the proof.
-  \end{enumerate}
-}
+We first recall the concept of information flow. A function $f : A
+\times B \to C$ is said to not allow information to flow from its
+second argument to the output if for all $b,b' \in B$ and all $a \in
+A$, $f(a,b) = f(a,b')$. If we think of the $B$ argument as
+representing high-security information, then we have stated that $f$
+does not allow the high-security input to flow to the low security
+output. Information flow can be seen as a kind of invariance property
+of programs, and so our relational interpretation of types is well
+tailored to proving this kind of property.
 
-An extreme example of invariance under change of representation is the
-tracking of information flow through programs. If a program does not
-depend upon a particular piece of information, then it will be
-invariant under \emph{all} changes of representation of this
-information. As thoroughly described by Sabelfeld and Sands
-\cite{sabelfeld01per}, information flow can be captured semantically
-through the use of partial equivalence relations (PERs). Interpreting
-types as PERs forms an instance of Reynolds' approach to abstraction
-and representation independence. Abadi, Banerjee, Heintze and Riecke
-\cite{abadi99core} built a \emph{Core Calculus for Dependency}, using
-a type system based around a security level indexed monad $T_lA$. Tse
-and Zdancewic \cite{tse04translating} showed that it is possible to
-translate Abadi et al.'s calculus into System F, translating the
+As described by Sabelfeld and Sands \cite{sabelfeld01per}, information
+flow can be captured semantically by partial equivalence relations
+(PERs). Abadi, Banerjee, Heintze and Riecke \cite{abadi99core} built a
+\emph{Core Calculus for Dependency}, using a type system based around
+a security level indexed monad $T_lA$, using PERs to prove the
+information flow properties. Tse and Zdancewic \cite{tse04translating}
+translated Abadi et al.'s calculus into System F, translating the
 monadic type $T_lA$ to $\alpha_l \to A$ for some free type variable
-$\alpha_l$, and making direct use of Reynolds' abstraction theorem to
-prove information flow properties. We now show how a refined variant
-of Tse and Zdancewic's translation can be expressed via
-algebraically-indexed types.
+$\alpha_l$, and using Reynolds' abstraction theorem to prove
+information flow properties. For example, if the type variable
+$\alpha_H$ represents high-level information, then the
+non-interference property of the function $f$ could be expressed by
+the System F type $A \to (\alpha_H \to B) \to C$. If a program cannot
+generate a value of type $\alpha_H$, then it cannot access the value
+of $B$, and hence is insensitive to the actual value. Relationships
+between security levels are captured by postulating functions
+$\alpha_{l_1} \to \alpha_{l_2}$ whenever $l_1$ is a lower security
+level than $l_2$.
 
-We assume a single indexing sort $\mathsf{principal}$, intended to
-represent (possibly composite) principals in a system. Principals are
-combined using the connectives of boolean logic ($\top$, $\bot$,
-$\lor$, $\land$, $\lnot$). For the equational theory of principals, we
-assume all the axioms of boolean algebra, and semantically we
-interpret composite principals as boolean values: either true ($\top$)
-or false ($\bot$). We introduce a single primitive type, indexed by
-expressions of sort $\mathsf{principal}$, $\tyPrim{T}{e}$, with the
-following interpretations:
-\begin{displaymath}
-  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
-    \tySem{\tyPrim{T}{e}} & \{*\} \\
-    \rsem{\tyPrim{T}{e}}{}\rho & \left\{
-      \begin{array}{ll}
-        \{(*,*)\} & \textrm{if }\sem{e}\rho = \top \\
-        \{\}      & \textrm{if }\sem{e}\rho = \bot
-      \end{array}
-      \right.
-  \end{array}
-\end{displaymath}
-The relational interpretation of $\tyPrim{T}{e}$ forces this type to
-only be inhabited when the boolean expression $e$ evaluates to true
-under the environment $\rho$. To write programs, we assume the
-following primitive operations:
+Using algebraically indexed types, we refine Tse and Zdancewic's
+translation by replacing each type variable $\alpha_l$ with a
+primitive type $\tyPrim{T}{e}$ of representations of the truth of a
+logical proposition $e$. The relationships between security levels are
+now replaced by logical entailment, so we only have functions of type
+$\tyPrim{T}{\phi} \to \tyPrim{T}{\psi}$ when $\phi$ entails
+$\psi$. Inspired by the PER interpretation of information flow, we
+instantiate our general relational framework to interpret
+$\tyPrimNm{T}{e}$ as the identity relation when $e$ is true, and the
+empty relation when $e$ is false. In doing so, we can recover the
+information flow properties of Abadi et al. and Tse and Zdancewic.
+
+\paragraph{Instantiation of the General Framework}
+We assume a single indexing sort $\mathsf{prop}$ and assume the
+operations and equations of boolean algebra. Thus we have constants
+$\top$, $\bot$ and binary operators $\land$, $\lor$ with the axioms of
+a bounded lattice, and a unary complementation operator $\lnot$. We
+will use $\phi, \psi$ to stand for index expressions of sort
+$\mathsf{prop}$. We use an equational presentation to fit with our
+general framework, but note that we can define an order on index
+expressions as $\phi \leq \psi$ when $\phi = \phi \land \psi$.
+
+We have a single primitive type $\tyPrimNm{T}$, with
+$\primTyArity(\tyPrimNm{T}) = [\mathsf{prop}]$ and index-erasure
+semantics $\tyPrimSem{\tyPrimNm{X}} = \{ * \}$. Thus values of type
+$\tyPrim{T}{\phi}$ have no run-time content; their only meaning is
+given by the relational semantics. For the relational environments
+$\relEnv{E}_\infoflow$, we use the compositional construction in
+\autoref{sec:constr-rel-env}, with an arbitrary boolean algebra $L$
+and $R_{\tyPrimNm{T}}(x) = \{ (*,*) \sepbar x = \top \}$.  The
+primitive operations $\Gamma_\infoflow$ reflect logical consequence:
 \begin{displaymath}
@@
     \mathrm{truth} : \tyPrim{T}{\top} &
-    \mathrm{and}   : \forall p, q\mathord:\mathsf{prn}.\ \tyPrim{T}{p} \to \tyPrim{T}{q} \to \tyPrim{T}{p \land q}
+    \mathrm{and}   : \forall p, q\mathord:\mathsf{prop}.\ \tyPrim{T}{p} \to \tyPrim{T}{q} \to \tyPrim{T}{p \land q}
   \end{array} \\
-  \mathrm{coerce} : \forall p, q\mathord:\mathsf{prn}.\ \tyPrim{T}{p \land q} \to \tyPrim{T}{p}
+  \mathrm{up} : \forall p, q\mathord:\mathsf{prop}.\ \tyPrim{T}{p \land q} \to \tyPrim{T}{p}
 \end{array}
 \end{displaymath}
-We can now adapt Tse and Zdancewic's translation of Abadi et al.'s
-monadic type to our setting. We define a type synonym, for each type
-$A$ and expression $e$ of sort $\mathsf{principal}$:
+The combination of the \TirName{TyEq} rule and the primitive
+$\mathrm{up}$ operation allow for logical entailment to be reflected
+in programs: if we have $\phi \leq \psi$ and $M : \tyPrim{T}{\phi}$
+then $\mathrm{up}\ M : \tyPrim{T}{\psi}$. Each of the primitive
+operations has a trivial interpretation, due the index-erasure
+interpretation of $\tyPrim{T}{\phi}$ as a one-element set, giving an
+environment $\eta_\infoflow \in \ctxtSem{\Gamma}$. Less trivially, we
+have this lemma:
+\begin{lemma}\label{lem:environments-information-flow}
+  For all $\Delta$ and $\rho \in \relEnv{E}_\infoflow$, $(\eta_\infoflow, \eta_\infoflow) \in \rsem{\Gamma_\infoflow}{\relEnv{E}_\infoflow}{\rho}$.
+\end{lemma}
+
+\paragraph{Information Flow} We think of logical expressions as
+``composite principals''. That is, propositional variables represent
+atomic principals that are combined using the logical connectives. We
+interpret ``truth'' for principals as stating that a principal is true
+when satisfied with the current state of affairs. Thus a relationship
+$\phi \leq \psi$ indicates that satisfaction of the composite
+principal $\phi$ implies satisfaction of the composite principal
+$\psi$. In terms of security levels, the ordering is reversed: if a
+high security principal is satisfied, then all of their underlings
+must also be satisfied.
+
+To study information flow, we adapt Tse and Zdancewic's translation of
+Abadi et al.'s monadic type to our setting. We define a type
+abbreviation $T_\phi A = \tyPrim{T}{\phi} \to A$, where $A$ is a type
+and $\phi$ is an expression of sort $\mathsf{prop}$. For every $\phi$,
+we can endow the types $T_\phi-$ with the structure of a monad. This
+is due to the fact that it is an instance of the ``environment'' (or
+``reader'') monad \cite{jones95functional}. We read the types
+$T_\phi A$ as data of type $A$ ``protected'' by the principal
+$\phi$.
+
+As a consequence of the relational interpretation given above, it
+follows that if we have an index expression $\phi$ that is interpreted
+as some value other than $\top$ in a relational environment $\rho$,
+then for all $x,x' \in \tySem{T_\phi\tyPrimNm{bool}}$, we have $(x,x')
+\in \rsem{T_\phi\tyPrimNm{bool}}{\relEnv{E}_\infoflow}{\rho}$. In
+terms of the reading above, if a principal is dissatisfied
+(i.e.,~$\phi \not= \top$), then data protected by this principal is
+indistinguishable from any other data. Thus a program cannot get
+access to the exact value. From this and \thmref{thm:abstraction}, we
+obtain the following information flow result:
+
+\begin{theorem}
+  Let $\phi$ and $\psi$ be index expressions of sort $\mathsf{prop}$
+  in some indexing context $\Delta$, such that $\psi \not\leq
+  \phi$. Then for all programs
   \begin{displaymath}
-  T_eA = \tyPrim{T}{e} \to A
+    \Delta; \Gamma_\infoflow,\Gamma \vdash M : T_\phi\tyPrimNm{bool} \to T_\psi\tyPrimNm{bool}
   \end{displaymath}
-For every principal $e$, we can endow the types $T_e-$ with the
-structure of a monad. This is due to the fact that it is an instance
-of the ``environment'' (or ``reader'') monad \cite{jones95functional}.
-
-\begin{example}
-  Suppose we have a program with the following type:
-  \begin{displaymath}
-    \forall p,q\mathord:\mathsf{principal}. \tyPrim{T}{e} \to T_p(\tyPrimNm{unit} + \tyPrimNm{unit}) \to T_q(\tyPrimNm{unit} + \tyPrimNm{unit})
-  \end{displaymath}
-  where $e$ is some boolean algebra expression involving $p$ and
-  $q$. We will show in \autoref{sec:instantiations} that programs with
-  this type are non-constant (i.e.~depend on their input) if and only
-  if the formula $e \Rightarrow q \Rightarrow p$ is valid in boolean
-  logic.
-\end{example}
-
-\begin{enumerate}
-\item Treat the information flow thing as a type isomorphism, in the
-  same way that Tse and Zdancewic do
-  \begin{displaymath}
-    \begin{array}{l}
-    \forall \vec{p}\mathord:\mathsf{principal}.\ T_{e_1}\tyPrimNm{bool} \to T_{e_2}\tyPrimNm{bool}\\
-    \hspace{4em}\cong \\
-    \left\{
-      \begin{array}{ll}
-        \tyPrimNm{bool} \tyArr \tyPrimNm{bool} & \textrm{if }e_2 \vdash e_1 \\
-        \tyPrimNm{bool} & \textrm{if }e_2 \not\vdash e_1
-      \end{array}
-    \right.
-  \end{array}
-  \end{displaymath}
-  where $\tyPrimNm{bool} = \tyUnit + \tyUnit$.
-\item mention the issue with bogus tokens if we allow for a
-  call-by-name style of non-termination
-\item if we axiomatise intuitionistic logic instead of
-  boolean/classical logic, we get a different type isomorphism.
-\item The isomorphism relies upon $e_1 \vdash e_2 \Leftrightarrow e_1
-  \lor e_2 = e_1$.
-\item Relevant logic? Look at the Wikipedia page to see how to do
-  residuated (commutative) monoids on lattices with just
-  equations. With this, we will be able to do substructural logics
-  too. As long as we know that a particular axiomatisation is sound
-  and complete for some reasoning system, then we can state and prove
-  the non-interference isomorphism above.
-\item Compare to Tse and Zdancewic's addition of axioms to the logic.
-\item Internalising equality? The logical structure can then be used
-  on types as well as everything else. Need the special rules for
-  equality... (see the Fomega paper).
-\end{enumerate}
-
-% Go a step further by considering equivalence relations on word32, and
-% their lattice structure. Link this to quantative information
-% flow. Will also need to include singleton types.
+  and all terms $N_1,N_2$ of type $T_\phi\tyPrimNm{bool}$, $M\ N_1
+  \stackrel{ctx}\approx M\ N_2$.  Thus there is no information flow
+  from $M$'s input to its output.
+\end{theorem}
+This result also holds if we move to logics other than boolean
+logic. For example, if our equational theory models intuitionistic
+logic by taking the axioms of Heyting algebras, then the same non-flow
+property for programs of type $T_{p\lor\lnot p}\tyPrimNm{bool} \to
+T_\top\tyPrimNm{bool}$ holds, due to the lack of excluded middle. If
+we take linear logic, then programs of type $T_{p\otimes
+  p}\tyPrimNm{bool} \to T_p\tyPrimNm{bool}$ have no information flow
+from their input, due to non-provability of $p \vdash p \otimes p$.
 

### 2012-07-09 paper/section-metric.tex

@@
 %\end{displaymath}
-\begin{lemma}
+\begin{lemma}\label{lem:metric-environments}
   For all $\Delta$ and $\rho \in \relEnv{E}_\Metric(\Delta)$,
@@
 
-\begin{example}[Uniform Continuity]
+\paragraph{Uniform continuity}
   %By making use of 
@@
 existential types, % in our framework,
-we can
-  state the standard 
-$\epsilon$-$\delta$ definition of
+we can state the standard $\epsilon$-$\delta$ definition of
 \emph{uniform} continuity as a type:
@@
 \end{displaymath}
-  This definition differs from the $\epsilon$-$\delta$
-  definition of (regular) continuity by the order of quantification: 
+This definition differs from the $\epsilon$-$\delta$ definition of
+(regular) continuity by the order of quantification:
   %For continuity, 
-  there, $\forall x$ comes before $\exists \delta$, so 
-  the distance 
+there, $\forall x$ comes before $\exists \delta$, so the distance
 $\delta$ may depend on %the value of
-  $x$. We suspect that to
-  express standard continuity as a type would require some form of
-  type dependency. Chaudhuri, Gulwani and Lublinerman
-  \cite{chaudhuri10continuity} have given a program logic based
-  approach to verifying the continuity of programs.
-\end{example}
+$x$. We suspect that to express standard continuity as a type would
+require some form of type dependency. Chaudhuri, Gulwani and
+Lublinerman \cite{chaudhuri10continuity} have given a program logic
+based approach to verifying the continuity of programs.
 

### 2012-07-09 paper/section-monoid.tex

@@
 that it is impossible to write a program with the type
-$\tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_1}$, and is thus impossible to
+$\tyPrim{M}{ab} \to \tyPrim{M}{a}$, and is thus impossible to
 convert a doubly tainted value to a singly tainted one.
 %
-By contrast with the situation for abelian group indexed types, the
-lack of inverses in the monoid indexed setting means that the type
-$\tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_2} \to \tyPrim{M}{m_1}$ is also
-uninhabited since there are no naturals $a_1,a_2$ such that
-$(m_1m_2)^{a_1}m_2^{a_2} \equiv m_1$.
+By contrast with abelian group indexed types, the lack of inverses in
+the monoid indexed setting means that the type $\tyPrim{M}{ab} \to
+\tyPrim{M}{b} \to \tyPrim{M}{a}$ is also uninhabited since there are
+no naturals $x_1,x_2$ such that $(ab)^{x_1}b^{x_2} \equiv a$.
 

### 2012-07-09 paper/section-general-framework.tex
Remove a fixme
minor changes
Add a label on the semantic equality definition

@@
 consequence of \thmref{thm:abstraction}.
-\begin{definition}
+\begin{definition}\label{def:semantic-equality}
   Let $\Gamma_{\mathit{ops}}$ and $\eta_{\mathit{ops}}$ be a context
@@
 prove, for each instantiation of our general framework, that the
-program-level primitive operations preserve all the relations.
-\fixme{This is the content of lemma X, lemma Y, lemma Z, below}.
+program-level primitive operations preserve all the relations. This is
+the content of \lemref{lem:geom-environments},
+\lemref{lem:monoid-ops-related},
+\lemref{lem:environments-information-flow} and
+\lemref{lem:metric-environments}.
 

### 2012-07-09 paper/geometry-twocolumn.tex
Tidy up the two outstanding FIXMEs

@@
 It can be verified by hand that %the intended denotation
-$\sem{\mathrm{affComb}}$ has the invariance property specified by its
-type. {\bf Vague. Say which invariance property.}
+$\sem{\mathrm{affComb}}$ is invariant under all changes of origin, as
+dictated by its type.
 % By defining the primitive function $\mathrm{affComb}$ to take a
@@
 terms of scaling invariance brings the techniques of dimensional
-analysis \fixme{ref} to bear on programming. The types of the real
-number arithmetic operations in \autoref{fig:real-ops} are exactly the
-types Kennedy assigns in his units of measure system, except for that
-of the absolute value operation. Semantically, our type indexes by
+analysis to bear on programming. The types of the real number
+arithmetic operations in \autoref{fig:real-ops} are exactly the types
+Kennedy assigns in his units of measure system, except for that of the
+absolute value operation. Semantically, our type indexes by
 %our assigned type is that, semantically, we are indexing by

### 2012-07-09 paper/paper.tex
Updates to paper.tex
- Make bibliography smaller
- remove unneeded loading of xypic
- fix PDF metadata
- update list of keywords
- add things to bibliography

@@
 \usepackage[usenames]{color}
-\usepackage[numbers]{natbib}
 \definecolor{citationcolour}{rgb}{0,0.4,0.2}
@@
             pdftitle=Relational Parametricity for Algebraically-Indexed Types,
-            pdfauthor={Robert Atkey, Neil Ghani, Patricia Johann, Andrew Kennedy},
+            pdfauthor={Robert Atkey, Patricia Johann, Andrew Kennedy},
             pdfkeywords={}}  
@@
 \usepackage{stmaryrd}
-\usepackage[all]{xy}
 
@@
 
-\keywords
-  parametricity, units of measure, dimensional analysis, invariance, computational geometry
+\keywords parametricity, units of measure, dimensional analysis,
+invariance, computational geometry, information flow, metric types, uniform continuity
 
@@
 \input{section-instantiations.tex}
+\input{section-monoid.tex}
 \input{section-info-flow.tex}
-\input{section-monoid.tex}
-\input{section-singleton.tex}
+%\input{section-singleton.tex}
 \input{section-metric.tex}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{abbrvnat}
-\bibliography{paper}
 
-\newpage
-\appendix
+\begin{thebibliography}{12}
+\softraggedright
 
-\input{appendix-proofs.tex}
+\bibitem{abadi99core} 
+M. Abadi, A. Banerjee, N. Heintze, and J. G. Riecke. A Core Calculus
+of Dependency. Proceedings, Symposium on Principles of Programming
+Languages, pp.147-160, 1999.
+
+\bibitem{syntaxforfree}
+R. Atkey.  Syntax for Free: Representing Syntax with Binding Using
+Parametricity.  Proceedings, Typed Lambda Calculi and Applications,
+pp. 35-49, 2009.
+
+\bibitem{TypedSyntax}
+N.~Benton, C.-K.~Hur, A.~J.~Kennedy, C.~McBride.
+Strongly Typed Term Representations in {Coq}.
+{\em Journal of Automated Reasoning} 49(2), pp. 141-159, 2012.
+
+\bibitem{benton06reading}
+N.~Benton, A.~Kennedy, M.~Hofmann, L.~Beringer.
+Reading, Writing and Relations.
+Proceedings, Asian Symposium on Programming Languages and Systems,
+pp. 114-130, 2006
+
+\bibitem{bernardy12proofs} 
+J.-P.~Bernardy, P.~Jansson, R.~Paterson.
+Proofs for free: Parametricity for dependent types.
+{\em Journal of Functional Programming} 22(02), pp. 107-152, 2012.
+
+\bibitem{CGAL}
+Computational Geometry Algorithms Library (CGAL): User and Reference
+Manual.  Available at {\tt http://www.cgal.org}.
+
+\bibitem{chaudhuri10continuity}
+S. Chaudhuri, S. Gulwani, and R. Lublinerman.  Continuity Analysis of
+Programs.  Proceedings, Symposium on Principles of Programming
+Languages, pp. 57-70, 2010.
+
+\bibitem{gallier11geometric}
+J. Gallier. {\em Geometric Methods and Applications, For Computer
+Science and Engineering. Springer Science+Business Media, LLC}, 2011.
+
+\bibitem{harrison09without}
+J.~Harrison.
+Without Loss of Generality.
+Proceedings, Theorem Proving in Higher Order Logics,
+pp. 43-59, 2009
+
+\bibitem{hofmann08correctness}
+M. Hofmann. Correctness of effect-based program
+transformations. Formal Logical Methods for System Security and
+Correctness, pp. 149-173, 2008
+
+\bibitem{jones95functional}
+M. P. Jones.  Functional Programming with Overloading and Higher-Order
+Polymorphism. Proceedings, Advanced Functional Programming,
+pp. 97-136, 1995.
+
+\bibitem{kennedy97relational} 
+A. Kennedy. Relational Parametricity and Units of Measure.
+Proceedings, Symposium on Principles of Programming Languages,
+pp. 442-455, 1997.
+
+\bibitem{CFGADT}
+S.~Mann, N.~Litke, T.~DeRose.  A Coordinate Free Geometry {ADT}.
+Technical Report CS-97-15, University of Waterloo, 1997.
+
+\bibitem{PittsAM:parpoe}
+A. M. Pitts.  Parametric Polymorphism and Operational Equivalence.
+{\em Mathematical Structures in Computer Science} 10, pp. 321-359,
+2000.
+
+\bibitem{reed10distance}
+J. Reed and B. C. Pierce.  Distance Makes the Types Grow Stronger.
+Proceeding, International Conference on Functional Programming,
+pp. 157-169, 2010.
+
+\bibitem{reynolds83types}
+J. C. Reynolds. Types, Abstraction and Parametric Polymorphism.
+{\em Information Processing} 83, pp. 513-523, 1983.
+
+\bibitem{sabelfeld01per} 
+A. Sabelfeld and D. Sands.  A Per Model of Secure Information Flow in
+Sequential Programs.  {\em Higher-Order and Symbolic Computation} 14
+(1), pp. 59-91, 2001.
+
+\bibitem{tse04translating}
+S. Tse and S. Zdancewic.  Translating Dependency into Parametricity.
+Proceedings, International Conference on Functional Programming pp.
+115-125, 2004.
+
+\bibitem{wadler89theorems}
+P. Wadler.  Theorems for Free!.  Proceedings, International Conference
+on Functional Programming Languages and Computer Architecture,
+pp. 347-359, 1989.
+
+\end{thebibliography}
+
+%\bibliographystyle{plainnat}
+%\renewcommand{\bibfont}{\normalsize}
+%\bibliography{paper}
+
+% \newpage
+% \appendix
+
+% \input{appendix-proofs.tex}
 

### 2012-07-09 paper/section-discussion.tex
Discussion section

@@
 
-\begin{enumerate}
-\item Dependent Types (allow for more precision, in some cases would
-  subsume our general framework, reference Bernardy's work)
-\item Type-indexed types (e.g.~type and effect systems)
-\item Making use of the free theorems, either by internalisation or by
-  deep embedding of our general calculus in a theorem prover.
-\item Extension of the definability work. Is it possible to extend
-  standard definability results for the $\lambda$-calculus to cover
-  algebraically indexed types?
-\item Discussion of the relationship between relational parametricity
-  and the abstraction theorem we have demonstrated. Discussion of the
-  identity extension property in this setting.
-\item Extending the geometric example with linear transformations, or
-  even higher-degree tensors.
-\end{enumerate}
+We have presented a general framework for algebraically indexed types
+and instantiated it to yield novel type systems for geometry, logical
+information flow and distance-indexed types. Our framework further
+demonstrates the power of relational reasoning about typed
+programs. From \thmref{thm:abstraction}, we have derived interesting
+free theorems, type isomorphisms and non-definability results. We
+conclude with some observations and suggestions for further work.
+
+\paragraph{Further Applications and Extensions} We have covered
+several applications of algebraically indexed types in this paper, but
+there are undoubtedly many more. Geometry for dimensions greater than
+two is an obvious candidate, as well as systems that are invariant
+under different geometric groups (e.g.,~the Poincar\'{e} group for
+relativity). Mathematical Physics is particularly rich in theories
+that have some notion of invariance, and it will be exciting to pin
+down the precise connections between these and type systems for which
+analogues of Reynolds' abstraction theorem hold. Geometric theorem
+proving is another possible application area. Harrison
+\cite{harrison09without} comments on the pervasive useful of
+invariance properties in this area. Programs in our framework
+automatically satisfy invariance properties, removing the need for
+ad-hoc proofs of these facts.
+
+In \autoref{sec:metric-types} we presented a type system with
+distance-indexed types, and noted the similarity with Reed and
+Pierce's system for non-expansivity \cite{reed10distance}. Exploring
+the relationship--particularly with respect to probability
+monads--also looks promising.
+
+Type and effect analyses use types indexed by effect annotations that
+have algebraic structure (e.g.~sets of read/write effect labels, with
+the structure of an idempotent monoid). Benton et al.~have used
+relational interpretations to prove effect-dependent equivalences
+\cite{benton06reading}. We expect that an extension of our framework
+with type-indexed types should be able to express their effect-indexed
+monads and prove their equivalences.
+
+Extending our framework with type dependency would also allow for
+further applications. For example, we could consider a type of lists
+of length $n$, indexed by elements of the permutation group
+$S_n$. Bernardy et al.~have presented a general framework for
+relational reasoning and an abstraction theorem for dependent types
+\cite{bernardy12proofs}. However, they work with pure type systems,
+which define type equality via untyped rewriting, so it is not
+immediately obvious how to integrate arbitrary equational theories
+into their framework.
+
+\paragraph{Extension of the non-definability results} Our
+non-definabilty results are currently restricted to the quantifier
+free fragment of our framework, as defined in the statement of
+\thmref{thm:abstraction}. This is a consequence of the
+non-compositionality of the relational environments that we use to
+prove non-definabilty results. Lifting this restriction is a key item
+of future work.
+
+\paragraph{Semantic Equality} In general, the semantic equality
+$\stackrel{sem}\sim$ in \defref{def:semantic-equality} is not an
+equivalence relation. If we assume that the interpretations of all the
+primitive types are partial equivalence relations then semantic
+equality is indeed an equivalence relation. However, this excludes the
+geometry and distance-indexed examples. More generally, we can
+consider relational interpretations that are \emph{difunctional}. (A
+relation is difunctional if whenever $(x,y)$, $(x',y')$ and $(x,y')$
+are in the relation then so is $(x',y)$.) Difunctionality is weaker
+than being a PER, but still suffices to prove that semantic equality
+is an equivalence relation. Hofmann \cite{hofmann08correctness} has
+used difunctional relations in the setting of effect
+analyses. Difunctionality covers all our examples except
+distance-indexed types. Note that for both PERs and difunctional
+relations we need to close the relational interpretation of
+existential types under the appropriate property to ensure that all
+types are interpreted as PERs/difunctional relations. For
+distance-indexed types it is possible that a new notion of
+equivalence, based on closeness, is required.
 

### 2012-07-09 paper/section-discussion.tex
Grammatical fixes to the discussion section

@@
 We have presented a general framework for algebraically indexed types
-and instantiated it to yield novel type systems for geometry, logical
-information flow and distance-indexed types. Our framework further
-demonstrates the power of relational reasoning about typed
-programs. From \thmref{thm:abstraction}, we have derived interesting
-free theorems, type isomorphisms and non-definability results. We
-conclude with some observations and suggestions for further work.
+and instantiated it to yield novel type systems for geometry, taint
+tracking, logical information flow and distance-indexed types. Our
+framework further demonstrates the power of relational reasoning about
+typed programs. From \thmref{thm:abstraction}, we have derived
+interesting free theorems, type isomorphisms and non-definability
+results. We conclude with some observations and suggestions for
+further work.
 
@@
 analogues of Reynolds' abstraction theorem hold. Geometric theorem
-proving is another possible application area. Harrison
-\cite{harrison09without} comments on the pervasive useful of
+proving is another possible application. Harrison
+\cite{harrison09without} comments on the pervasive usefulness of
 invariance properties in this area. Programs in our framework

### 2012-07-09 paper/paper.tex
Minor fixes to the bibliography
Fix title in PDF metadata

@@
             citecolor=citationcolour,
-            pdftitle=Relational Parametricity for Algebraically-Indexed Types,
+            pdftitle=Abstraction and Invariance for Algebraically-Indexed Types,
             pdfauthor={Robert Atkey, Patricia Johann, Andrew Kennedy},
@@
 Proceedings, Asian Symposium on Programming Languages and Systems,
-pp. 114-130, 2006
+pp. 114-130, 2006.
 
@@
 Proceedings, Theorem Proving in Higher Order Logics,
-pp. 43-59, 2009
+pp. 43-59, 2009.
 
 \bibitem{hofmann08correctness}
-M. Hofmann. Correctness of effect-based program
+M.~Hofmann. Correctness of effect-based program
 transformations. Formal Logical Methods for System Security and
-Correctness, pp. 149-173, 2008
+Correctness, pp. 149-173, 2008.
 
 \bibitem{jones95functional}
-M. P. Jones.  Functional Programming with Overloading and Higher-Order
+M.~P.~Jones.  Functional Programming with Overloading and Higher-Order
 Polymorphism. Proceedings, Advanced Functional Programming,
@@
 \bibitem{kennedy97relational} 
-A. Kennedy. Relational Parametricity and Units of Measure.
+A.~Kennedy. Relational Parametricity and Units of Measure.
 Proceedings, Symposium on Principles of Programming Languages,
@@
 \bibitem{PittsAM:parpoe}
-A. M. Pitts.  Parametric Polymorphism and Operational Equivalence.
+A.~M.~Pitts.  Parametric Polymorphism and Operational Equivalence.
 {\em Mathematical Structures in Computer Science} 10, pp. 321-359,
@@
 \bibitem{reed10distance}
-J. Reed and B. C. Pierce.  Distance Makes the Types Grow Stronger.
+J.~Reed and B.~C.~Pierce.  Distance Makes the Types Grow Stronger.
 Proceeding, International Conference on Functional Programming,
@@
 \bibitem{reynolds83types}
-J. C. Reynolds. Types, Abstraction and Parametric Polymorphism.
+J.~C.~Reynolds. Types, Abstraction and Parametric Polymorphism.
 {\em Information Processing} 83, pp. 513-523, 1983.
@@
 \bibitem{sabelfeld01per} 
-A. Sabelfeld and D. Sands.  A Per Model of Secure Information Flow in
+A.~Sabelfeld and D.~Sands.  A Per Model of Secure Information Flow in
 Sequential Programs.  {\em Higher-Order and Symbolic Computation} 14
@@
 \bibitem{tse04translating}
-S. Tse and S. Zdancewic.  Translating Dependency into Parametricity.
+S.~Tse and S.~Zdancewic.  Translating Dependency into Parametricity.
 Proceedings, International Conference on Functional Programming pp.
@@
 \bibitem{wadler89theorems}
-P. Wadler.  Theorems for Free!.  Proceedings, International Conference
+P.~Wadler.  Theorems for Free!.  Proceedings, International Conference
 on Functional Programming Languages and Computer Architecture,

### 2012-07-09 paper/section-introduction.tex
Lead in from units-of-measure; comment about omitted cases in isomorphism proof sketch

@@
 
-In this paper we show how to capture, and generalise, invariance under
-groups of translations using
-%Inspired by this connection, we study 
-relational parametricity for types that are indexed by attributes with
-algebraic structure. For example, in computational geometry, points in
-2-space can be indexed by a coordinate frame; in physical modelling,
-quantities are indexed by units-of-measure; in information-flow
-security, computations are indexed by principals.  Types that are
+In previous work the third author studied relational parametricity for
+the types parameterized by units-of-measure, whose invariance 
+properties relate to changes of units, or \emph{scaling}~\cite{kennedy97relational}. 
+In this paper we investigate the general case of types indexed by attributes
+with algebraic structure. For example, in computational geometry, points in
+2-space can be indexed by a coordinate frame; in information-flow
+security, computations are indexed by principals; in differential privacy, types
+can be indexed by `distance'. Types that are
 polymorphic over such indices induce invariance properties and

### 2012-07-09 paper/geometry-twocolumn.tex
Lots of little typesetting fixes and typos

@@
       \mathrm{quadB\acute{e}zier} : \forall t
-      \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to
-      \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t}
+      \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \mathord\to \tyPrim{vec}{t}\mathord\to
+      \tyPrim{vec}{t} \mathord\to \tyPrimNm{real} \mathord\to \tyPrim{vec}{t}
       \\ \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\ \quad
@@
   be the origin provided the problem statement is invariant under
-  translation. Each type $\tau_{n}$ is isomorphic to the coresponding
+  translation. Each type $\tau_{n}$ is isomorphic to the corresponding
   type $\sigma_{n}$:
@@
   quantifier-free inhabitants\footnote{We define quantifier-free
-    programs in \thmref{thm:abstraction}.}. Intuitively, this is
-  because there is no way to remove the extra $t$ in
-  $\tyPrim{vec}{t+t}$ using the vector operations. We formalise this
-  indefinability result in \autoref{sec:types-indexed-abelian-groups-indef}
-  using a more sophisticated relational interpretation of free
-  translation variables than the one we sketched above.
+    programs in the statement of
+    \thmref{thm:abstraction}.}. Intuitively, this is because there is
+  no way to remove the extra $t$ in $\tyPrim{vec}{t+t}$ using the
+  vector operations. We formalise this indefinability result in
+  \autoref{sec:types-indexed-abelian-groups-indef} using a more
+  sophisticated relational interpretation of free translation
+  variables than the one we sketched above.
 \end{example}
@@
 \begin{displaymath}
-\begin{array}{l@{}l}
+\begin{array}{@{}l@{}l}
   (*) : \forall s \mathord: \SynGL{1}, B \mathord:

### 2012-07-09 paper/section-general-framework.tex
Lots of little typesetting fixes and typos

@@
     \begin{array}{l}
-      t : \SynTransl{2} \vdash t + 1 \stackrel{ax}\equiv t : \SynTransl{2} \\
+      t : \SynTransl{2} \vdash t + 0 \stackrel{ax}\equiv t : \SynTransl{2} \\
       t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
-      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 1 : \SynTransl{2} \\
+      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 0 : \SynTransl{2} \\
       t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
@@
 \begin{displaymath}
-  \begin{array}{@{\hspace{0em}}c@{\hspace{0em}}c}
+  \begin{array}{@{}c@{\hspace{0em}}c@{}}
     \begin{array}{@{}l}
@@
 
-We use the relational interpretation to define semantic equivalence,
-which is a sound approximation of contextual equivalence, as a
-consequence of \thmref{thm:abstraction}.
+Semantic equivalence of programs is defined in terms of the relational
+interpretation. As a consequence of \thmref{thm:abstraction}, semantic
+equivalence is a sound approximation of contextual equivalence.
 \begin{definition}\label{def:semantic-equality}
@@
 
-These relational environments are non-compositional in the sense that
-elements of $\relEnv{E}(\Delta,\Delta')$ are not only constructed from
-elements of $\relEnv{E}(\Delta)$ and $\relEnv{E}(\Delta')$. As a
-consequence of their non-compositionality, these relational
-environments will only be \emph{pre-substitutive} in the sense of
-\defref{defn:valid-rel-env-family}. Therefore our non-definability
-results are subject to the conditions in \thmref{thm:abstraction}.
+These relational environments are non-compositional: elements of
+$\relEnv{E}(\Delta,\Delta')$ are not purely constructed from elements
+of $\relEnv{E}(\Delta)$ and $\relEnv{E}(\Delta')$. As a consequence,
+these relational environments will only be
+\emph{pre-substitutive}. Therefore our non-definability results are
+subject to the quantifier-free conditions in \thmref{thm:abstraction}.
 

### 2012-07-09 paper/section-info-flow.tex
Lots of little typesetting fixes and typos

@@
 
-Using algebraically indexed types, we refine Tse and Zdancewic's
-translation by replacing each type variable $\alpha_l$ with a
-primitive type $\tyPrim{T}{e}$ of representations of the truth of a
-logical proposition $e$. The relationships between security levels are
-now replaced by logical entailment, so we only have functions of type
+Using algebraic indexing, we refine Tse and Zdancewic's translation by
+replacing each type variable $\alpha_l$ with a primitive type
+$\tyPrim{T}{e}$ of representations of the truth of a logical
+proposition $e$. The relationships between security levels are now
+replaced by logical entailment, so we only have functions of type
 $\tyPrim{T}{\phi} \to \tyPrim{T}{\psi}$ when $\phi$ entails
@@
 instantiate our general relational framework to interpret
-$\tyPrimNm{T}{e}$ as the identity relation when $e$ is true, and the
-empty relation when $e$ is false. In doing so, we can recover the
+$\tyPrim{T}{\phi}$ as the identity relation when $\phi$ is true, and
+the empty relation when $\phi$ is false. By doing so, we recover the
 information flow properties of Abadi et al. and Tse and Zdancewic.
@@
 will use $\phi, \psi$ to stand for index expressions of sort
-$\mathsf{prop}$. We use an equational presentation to fit with our
-general framework, but note that we can define an order on index
-expressions as $\phi \leq \psi$ when $\phi = \phi \land \psi$.
+$\mathsf{prop}$. We use an equational presentation of boolean algebras
+to fit with our general framework, but note that we can define an
+order on index expressions as $\phi \leq \psi$ when $\phi = \phi \land
+\psi$.
 
@@
 
-To study information flow, we adapt Tse and Zdancewic's translation of
-Abadi et al.'s monadic type to our setting. We define a type
-abbreviation $T_\phi A = \tyPrim{T}{\phi} \to A$, where $A$ is a type
-and $\phi$ is an expression of sort $\mathsf{prop}$. For every $\phi$,
-we can endow the types $T_\phi-$ with the structure of a monad. This
-is due to the fact that it is an instance of the ``environment'' (or
-``reader'') monad \cite{jones95functional}. We read the types
-$T_\phi A$ as data of type $A$ ``protected'' by the principal
-$\phi$.
+We adapt Tse and Zdancewic's translation of Abadi et al.'s monadic
+type to our setting. We define a type abbreviation $T_\phi A =
+\tyPrim{T}{\phi} \to A$, where $A$ is a type and $\phi$ is an
+expression of sort $\mathsf{prop}$. For every $\phi$, we can endow the
+types $T_\phi-$ with the structure of a monad. This is due to the fact
+that it is an instance of the ``environment'' (or ``reader'') monad
+\cite{jones95functional}. We read the types $T_\phi A$ as data of type
+$A$ ``protected'' by the principal $\phi$.
 
@@
 then for all $x,x' \in \tySem{T_\phi\tyPrimNm{bool}}$, we have $(x,x')
-\in \rsem{T_\phi\tyPrimNm{bool}}{\relEnv{E}_\infoflow}{\rho}$. In
-terms of the reading above, if a principal is dissatisfied
-(i.e.,~$\phi \not= \top$), then data protected by this principal is
-indistinguishable from any other data. Thus a program cannot get
-access to the exact value. From this and \thmref{thm:abstraction}, we
-obtain the following information flow result:
+\in \rsem{T_\phi\tyPrimNm{bool}}{\relEnv{E}_\infoflow}{\rho}$. Thus if
+a principal is dissatisfied (i.e.,~$\phi \not= \top$), then data
+protected by this principal is indistinguishable from any other
+data. Thus a program cannot get access to the exact value. From this
+and \thmref{thm:abstraction}, we obtain the following information flow
+result:
 

### 2012-07-09 paper/section-metric.tex
Lots of little typesetting fixes and typos

@@
 
-The previous two sections have presented examples where the relational
-interpretations of primitive types relates a pair of elements by some
-transformation if applying the transformation to the first element
-yields the second. Thus, the free theorems that we derived directly
-take the form of ``invariance'' properties, where some equation holds
-between two terms. In this section, we examine another instantiation
-of our general framework that relates values when they are within a
-certain distance. The free theorems that we obtain inform us of the
-effect that programs have on the distances between values. For
-example, a program $M$ of type:
+\autoref{sec:instantiations} and \autoref{sec:monoid-indexed-types}
+presented examples where the relational interpretations of primitive
+types relates a pair of elements by some transformation if applying
+the transformation to the first element yields the second. Thus, the
+free theorems that we derived directly take the form of ``invariance''
+properties, where some equation holds between two terms. In this
+section, we examine another instantiation of our general framework
+that relates values when they are within a certain distance. The free
+theorems that we obtain inform us of the effect that programs have on
+the distances between values. For example, a program $M$ of type:
 \begin{displaymath}

### 2012-07-09 paper/section-monoid.tex
Lots of little typesetting fixes and typos

@@
   \label{thm:monoid-indefinability}
-  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
-  natural numbers $A$ and $m$-by-$1$ matricies of natural numbers $b$,
-  there exists an $M$ in the quantifier-free fragment, with
+  Let $m$, $n$ be natural numbers. For all $n$-by-$m$-matrices of
+  natural numbers $A$ and $m$-by-$1$ matrices of natural numbers $b$,
+  there is a quantifier-free program $N$, with
   \begin{displaymath}
-    i_1,\dots,i_m\mathord:\mathsf{M}; \Gamma_\Mon \vdash M : \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
+    i_1,\dots,i_m\mathord:\mathsf{M}; \Gamma_\Mon \vdash N : \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
   \end{displaymath}

### 2012-07-09 paper/section-instantiations.tex

@@
 %\end{eqnarray*}
-where terms $\transdn A : A\to A[0/t]$ and $\transup A : A[0/t] \to A$ are defined by induction over the syntax of $A$:
+where terms $\transdn A : A\to A[0/t]$ and $\transup A : A[0/t] \to A$ are defined by induction over the syntax of $A$ (cases omitted for $\transup A$ have definition symmetric to $\transdn A$):
 \begin{eqnarray*}

### 2012-07-09 paper/section-general-framework.tex
Save a bit of space in the framework section

@@
 judgment between index expressions $\Delta \vdash e \equiv e' : s$ by
-the following rules. The first allows us to use substitution instances
-of axioms. The second ensures that equality is a congruence relation
-with respect to the index operations in $\IndexOpSet$:
-\begin{mathpar}
+a set of rules. The following rule allows us to use substitution
+instances of axioms.
+\begin{displaymath}
   \inferrule*
@@
   {\Delta \vdash \sigma^*e \equiv \sigma^*e' : s}
-
-  \inferrule*
-  {\{\Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j\leq n}}
-  {\Delta \vdash \indexOp{f}(e_1, ..., e_n) \equiv \indexOp{f}(e'_1, ..., e'_n) : s}
-\end{mathpar}
-We also assume the standard reflexivity, symmetry and transitivity
-rules for the equality judgment.
+\end{displaymath}
+We also assume the standard congruence, reflexivity, symmetry and
+transitivity rules for the equality judgment.
 

### 2012-07-10 paper/section-instantiations.tex
Inhabitation of AG

@@
 The maps associated with this isomorphism are 
-$I = \lambda x. x\ [0]\ 0$ and $J = \lambda y. \Lambda t.\lambda v. \transdn A (y)$,
+$I = \lambda x. x\ [0]\ 0$ and $J = \lambda y. \Lambda t.\lambda v. \transup A (y)$,
 %\begin{eqnarray*}
@@
 \subsection{Indefinability}
+\label{sec:types-indexed-abelian-groups-indef}
+\newcommand{\Grp}{\mathit{Grp}}
+For proving non-definability results, we
+consider the general case of a sort $\mathsf{G}$ with the equational theory of abelian groups.
+We assume a group-indexed primitive type $\tyPrimNm{G}$ and
+set of primitive operations typed by $\Gamma_\Grp$ that reflect the group structure at the value level:
+\begin{displaymath}
+  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
+    1 & \tyPrim{G}{1} \\
+    (\cdot) & \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to \tyPrim{G}{a\cdot b} \\
+    \ ^{-1} & \forall a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}
+  \end{array}
+\end{displaymath}
+One instance is the sort $\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ operations.
 
-\fixme{Copy the stuff over from the other file}
+For index-erasure interpretation, we assume a (not necessarily abelian) group $(G, 1, \cdot,
+\ ^{-1})$. Let $Z(G)$ be the centre of the $G$: i.e.,~the set of
+elements that commute with every other element in $G$. We set
+$\tyPrimSem{G} = G$. The primitive operations in $\Gamma_\Grp$ are
+interpreted just as the unit, multiplication and inverse operations
+of $G$. For each index context $\Delta$, the set
+$\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be
+$\{ \rho_{(H,h)} \sepbar H\textrm{ is a subgroup of }(\Delta
+\Rightarrow \mathsf{G})/\equiv, h\textrm{ is a group homomorphism }H
+\to Z(G) \}$, where
+\begin{displaymath}
+  \rho_{(H,h)}\ \tyPrimNm{G}\ (e) = \left\{
+    \begin{array}{ll}
+      \{ \} & \textrm{if}\ e\not\in H \\
+      \{ (x, h(e)\oplus x) \sepbar x \in G \} & \textrm{if}\ e\in H
+    \end{array}
+  \right.
+\end{displaymath}
+
+\begin{lemma}
+  For all $\Delta$ and $\rho \in \relEnv{E}_\Grp(\Delta)$,
+  $(\eta_\Grp,\eta_\Grp) \in \rsem{\Gamma_\Grp}{\relEnv{E}}\rho$.
+\end{lemma}
+
+We write $i^n$ for $n$-fold iteration of the group operation.
+\begin{theorem}
+  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matrices of
+  natural numbers $A$ and $m$-by-$1$ matricies of integers $b$,
+  there exists a program $M$ such that
+  \begin{equation}
+    \label{eq:fo-group-type}
+    \Gamma_\Grp \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
+  \end{equation}
+  where $e_j = i_1^{A_{j1}}\cdots i_m^{A_{jm}}$ and $e = i_1^{b_1}\cdots i_m^{b_m}$, if and only if there is a solution $\vec{x}$
+  in integers to the equation $A \vec x = b$.
+\end{theorem}
+The uninhabited type  $\forall t \mathord:
+  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$
+from \exref{ex:uninhabited-type} is an instance of this theorem.
+
+For types indexed by the scaling group $\SynGL{1}$ the situation is a little more complex, as
+the index-erasure interpretation of $\tyPrim{real}{s}$ is larger than the group $\GL{1}$, as it contains
+zero. We can show, nonetheless, that $\forall s \mathord:
+  \SynGL{1}.\ \tyPrim{real}{s \circ s} \to \tyPrim{real}{s}\cong 1$, reflecting 
+the fact that any term of this type must be the function $\lambda x.0$.
+
+%\begin{proof}
+%  (If) The program $\Lambda i_1\dots i_m.\ \lambda g_1\dots g_n.\
+%  g_1^{x_1}\dots g_n^{x_n}$ satisfies the typing judgment
+%  (\ref{eq:fo-group-type}).
+
+%  (Only if) Assume that there is a program $M$ satisfying the typing
+%  judgment (\ref{eq:fo-group-type}). By \thmref{thm:abstraction} we
+%  know that for all relational environments $\rho \in
+%  \relEnv{E}_\Grp(i_1,\dots,i_m)$, and all $g_j,g'_j \in G$,
+%  \begin{displaymath}
+%    \begin{array}{l}
+%      (g_1,g'_1) \in \rho\ \tyPrimNm{G}\ (e_1) \land \dots \land (g_n,g'_n) \in \rho\ \tyPrimNm{G}\ (e_n) \Rightarrow \\
+%      \quad (\tmSem{M}g_1\dots g_n, \tmSem{M}g'_1\dots g'_n) \in \rho\ \tyPrimNm{G}\ (e)
+%    \end{array}
+%  \end{displaymath}
+%  We select the relational environments
+%  \begin{displaymath}
+%    \rho\ \tyPrimNm{G}\ (e) = \left\{
+%      \begin{array}{ll}
+%        \{(g,g) \sepbar g \in \mathcal{G} \} &
+%        \begin{array}[t]{@{}l}
+%          \textrm{if}\ e = e_1^{x_1}\dots e_n^{x_n} \\
+%          \textrm{ for some }x_1,\dots,x_n \in \mathbb{Z}
+%        \end{array}
+%        \\
+%        \{\} & \textrm{otherwise}
+%      \end{array}
+%    \right.
+%  \end{displaymath}
+%  Now, for any $g_1,\dots,g_m \in \mathcal{G}$, we know that, for all
+%  $j$, $(g_j,g_j) \in \rho\ \tyPrimNm{G}\ (e_j)$, so we know that
+%  $(\tySem{M}g_1\dots g_n, \tySem{M}g_1 \dots g_n) \in \rho\
+%  \tyPrimNm{G}\ (e)$. Thus there exist integers $x_1,...,x_n$ such
+%  that $e = e_1^{x_1}\dots e_n^{x_n}$. But we also know that $e =
+%  i_1^{b_1}...i_m^{b_m}$. By the cancellation property of free groups,
+%  we learn that $x_1,...,x_n$ is the solution we need.
+%\end{proof}
 

### 2012-07-10 paper/section-instantiations.tex
minor formatting

@@
 \noindent
-\newcommand{\transup}[1]{\uparrow_{#1}}
-\newcommand{\transdn}[1]{\downarrow_{#1}}
+\newcommand{\transup}[1]{\mathord\uparrow_{#1}}
+\newcommand{\transdn}[1]{\mathord\downarrow_{#1}}
 The maps associated with this isomorphism are 
@@
 \transdn{A_1 + A_2} & = & \lambda x.
-\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn{A_1}y)); \textrm{inr}\ y.\textrm{inr}(\transdn{A_2}y)\\ 
-\transdn{A_1\to A_2} & = & \lambda f. \lambda x. (\transdn{A_2} (f(\transup{A_1}(x))))\\ 
+\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn{A_1}y); \textrm{inr}\ y.\textrm{inr}(\transdn{A_2}y)\\ 
+\transdn{A_1\to A_2} & = & \lambda f. \lambda x. \transdn{A_2} (f(\transup{A_1}x))\\ 
 \transdn{\tyPrim{vec}{e}} & = & \lambda x. x - kv\\ 
 \transup{\tyPrim{vec}{e}} & = & \lambda x. x + kv\\ & & \text{ if }e \equiv k t + e'\text{ for
-  some }k\in\setOfIntegers \text{ and }e'\text{ with }t\notin e'\\
+  some }k\in\setOfIntegers \text{ and }e'\text{ with }t\notin e'
 \end{eqnarray*}
@@
 from \exref{ex:uninhabited-type} is an instance of this theorem.
-
-For types indexed by the scaling group $\SynGL{1}$ the situation is a little more complex, as
-the index-erasure interpretation of $\tyPrim{real}{s}$ is larger than the group $\GL{1}$, as it contains
-zero. We can show, nonetheless, that $\forall s \mathord:
-  \SynGL{1}.\ \tyPrim{real}{s \circ s} \to \tyPrim{real}{s}\cong 1$, reflecting 
-the fact that any term of this type must be the function $\lambda x.0$.
+For types indexed by the scaling group $\SynGL{1}$ the situation is more subtle, as
+the index-erasure interpretation of $\tyPrim{real}{s}$ is group $\GL{1}$ together with zero.
+We can show, nonetheless, that $\forall s \mathord:
+  \SynGL{1}.\ \tyPrim{real}{s \cdot s} \to \tyPrim{real}{s}\cong 1$, reflecting 
+the fact that any term of this type must be $\lambda x.0$.
 

### 2012-07-10 paper/section-instantiations.tex
Space saving

@@
 \newcommand{\transdn}[1]{\mathord\downarrow_{#1}}
-The maps associated with this isomorphism are 
+The maps associated with the isomorphism are 
 $I = \lambda x. x\ [0]\ 0$ and $J = \lambda y. \Lambda t.\lambda v. \transup A (y)$,
@@
 where terms $\transdn A : A\to A[0/t]$ and $\transup A : A[0/t] \to A$ are defined by induction over the syntax of $A$ (cases omitted for $\transup A$ have definition symmetric to $\transdn A$):
-\begin{eqnarray*}
-\transdn{1} & = & \lambda x. x\\ 
-\transdn{A_1\times A_2} & = & \lambda p. (\transdn{A_1}(\pi_1 p),\transdn{A_2}(\pi_2 p))\\ 
-\transdn{A_1 + A_2} & = & \lambda x.
+\begin{align*}
+\transdn{1} &= \lambda x. x \qquad
+\transdn{A_1\times A_2} = \lambda p. (\transdn{A_1}(\pi_1 p),\transdn{A_2}(\pi_2 p))\\ 
+\transdn{A_1 + A_2} &= \lambda x.
 \textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn{A_1}y); \textrm{inr}\ y.\textrm{inr}(\transdn{A_2}y)\\ 
-\transdn{A_1\to A_2} & = & \lambda f. \lambda x. \transdn{A_2} (f(\transup{A_1}x))\\ 
-\transdn{\tyPrim{vec}{e}} & = & \lambda x. x - kv\\ 
-\transup{\tyPrim{vec}{e}} & = & \lambda x. x + kv\\ & & \text{ if }e \equiv k t + e'\text{ for
+\transdn{A_1\to A_2} &= \lambda f. \lambda x. \transdn{A_2} (f(\transup{A_1}x))\\ 
+\transdn{\tyPrim{vec}{e}} &= \lambda x. x - kv \qquad
+\transup{\tyPrim{vec}{e}} = \lambda x. x + kv\\ 
+\text{ if }e & \equiv k t + e' \text{ for
   some }k\in\setOfIntegers \text{ and }e'\text{ with }t\notin e'
-\end{eqnarray*}
+\end{align*}
 To show $I(J(y))$ is contextually equivalent to $y$ is an easy
@@
 \paragraph{Scalings}
-Now consider the group $\GL{1}$ of scalings. This time, we can treat
-the real-valued argument to a function as a `unit of measure' with
+For the group $\GL{1}$ of scalings we can treat
+a real-valued argument to a function as a unit-of-measure with
 which to scale the result. 
@@
 We assume a group-indexed primitive type $\tyPrimNm{G}$ and
-set of primitive operations typed by $\Gamma_\Grp$ that reflect the group structure at the value level:
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    1 & \tyPrim{G}{1} \\
-    (\cdot) & \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to \tyPrim{G}{a\cdot b} \\
-    \ ^{-1} & \forall a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}
-  \end{array}
-\end{displaymath}
-One instance is the sort $\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ operations.
+set of primitive operations typed by $\Gamma_\Grp$ that reflect the group structure at the value level,
+with identity $1:\tyPrim{G}{1}$, operation
+$(\cdot): \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to \tyPrim{G}{a\cdot b}$
+and inverse $\ ^{-1}: \forall a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$.
+One instance is the sort~$\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ operations.
 

### 2012-07-10 paper/geometry-twocolumn.tex
Patty's changes to shorten the text

@@
     programs in the statement of
-    \thmref{thm:abstraction}.}. Intuitively, this is because there is
-  no way to remove the extra $t$ in $\tyPrim{vec}{t+t}$ using the
+    \thmref{thm:abstraction}.}. Intuitively, this is because we cannot
+  remove the extra $t$ in $\tyPrim{vec}{t+t}$ using the
   vector operations. We formalise this indefinability result in
@@
   sophisticated relational interpretation of free translation
-  variables than the one we sketched above.
+  variables than the one sketched above.
 \end{example}

### 2012-07-10 paper/section-discussion.tex
Patty's changes to shorten the text

--- 2012-07-09 paper/section-discussion.tex
@@
 
-We have presented a general framework for algebraically indexed types
-and instantiated it to yield novel type systems for geometry, logical
-information flow and distance-indexed types. Our framework further
-demonstrates the power of relational reasoning about typed
-programs. From \thmref{thm:abstraction}, we have derived interesting
-free theorems, type isomorphisms and non-definability results. We
-conclude with some observations and suggestions for further work.
+We presented a general framework for algebraically indexed types
+and instantiated it to yield novel type systems for geometry, taint
+tracking, logical information flow and distance-indexed types. Our
+framework further demonstrates the power of relational reasoning about
+typed programs. From \thmref{thm:abstraction}, we derived
+interesting free theorems, type isomorphisms and non-definability
+results. We conclude with some observations and suggestions for
+further work.
 
@@
 there are undoubtedly many more. Geometry for dimensions greater than
-two is an obvious candidate, as well as systems that are invariant
+two is an obvious candidate, as are systems that are invariant
 under different geometric groups (e.g.,~the Poincar\'{e} group for
@@
 down the precise connections between these and type systems for which
-analogues of Reynolds' abstraction theorem hold. Geometric theorem
-proving is another possible application area. Harrison
-\cite{harrison09without} comments on the pervasive useful of
+%analogues of Reynolds' 
+an abstraction theorem holds. Geometric theorem
+proving is another %possible 
+application. Harrison
+\cite{harrison09without} comments on the pervasiveness of % usefulness of
 invariance properties in this area. Programs in our framework
 automatically satisfy invariance properties, removing the need for
-ad-hoc proofs of these facts.
+{\em ad hoc} proofs. % of these facts.
 
@@
 
-Type and effect analyses use types indexed by effect annotations that
-have algebraic structure (e.g.~sets of read/write effect labels, with
-the structure of an idempotent monoid). Benton et al.~have used
-relational interpretations to prove effect-dependent equivalences
-\cite{benton06reading}. We expect that an extension of our framework
-with type-indexed types should be able to express their effect-indexed
-monads and prove their equivalences.
+Type and effect analyses use types indexed by effect annotations with
+algebraic structure (e.g.,~sets of read/write effect labels with an
+idempotent monoid structure). Benton {\em et al.}~ have used relational
+interpretations to prove effect-dependent equivalences
+\cite{benton06reading}. An extension of our framework with
+type-indexed types should be able to express their effect-indexed monads
+and prove their equivalences.
 

### 2012-07-10 paper/section-general-framework.tex
Inhabitation of AG
Space saving
Patty's changes to shorten the text

@@
 
-  \bigskip
+  \smallskip
 
@@
 judgment between index expressions $\Delta \vdash e \equiv e' : s$ by
-a set of rules. The following rule allows us to use substitution
-instances of axioms.
+a set of rules. The following rule lets us use substitution
+instances of axioms:
 \begin{displaymath}
@@
 uninhabited. Intuitively, it is clear why: given the operations on
-vectors we presented in \autoref{sec:affine-vector-ops}, there is no
-way of cancelling the additional $t$ in the argument type to get the
-result type. More generally, our operations on vectors only allow us
-to combine indices according to the group operations. Thus, if the
-index of the result type of a function does not lie within the free
-abelian group generated by the indices of the input types, then there
-is no way to write a program with this %given
-type. We now develop this idea to generate a family of sets of
-relational environments. The key idea is to assign different
-relational interpretations to types with indices that lie outside some
-finitely generated collection of index expressions.
+vectors in \autoref{sec:affine-vector-ops}, there is no way of
+cancelling the additional $t$ in the argument type to get the result
+type. More generally, the vector operations only allow us to combine
+indices according to the group operations, so if the index of the
+result type of a function lies outside the free abelian group
+generated by the indices of the input types, then we cannot write a
+program of this type. 
+We now develop this idea to generate a
+family of sets of relational environments. The key idea is to assign
+different relational interpretations to types with indices that lie
+outside some finitely generated collection of index expressions.
 

### 2012-07-10 paper/section-info-flow.tex
Patty's changes to shorten the text

@@
 $\tyPrim{T}{\phi} \to \tyPrim{T}{\psi}$ when $\phi$ entails
-$\psi$. Inspired by the PER interpretation of information flow, we
-instantiate our general relational framework to interpret
-$\tyPrim{T}{\phi}$ as the identity relation when $\phi$ is true, and
-the empty relation when $\phi$ is false. By doing so, we recover the
-information flow properties of Abadi et al. and Tse and Zdancewic.
+$\psi$. %Inspired by the PER interpretation of information flow, 
+We instantiate our general relational framework to interpret
+$\tyPrim{T}{\phi}$ as the identity relation if $\phi$ is true and the
+empty relation if $\phi$ is false. This recovers the information flow
+properties of Abadi {\em et al.} and Tse and Zdancewic.
 

### 2012-07-10 paper/section-instantiations.tex
Patty's changes to shorten the text

--- 2012-07-10 paper/section-instantiations.tex
@@
 \label{sec:types-indexed-abelian-groups-indef}
-\newcommand{\Grp}{\mathit{Grp}}
-For proving non-definability results, we
-consider the general case of a sort $\mathsf{G}$ with the equational theory of abelian groups.
-We assume a group-indexed primitive type $\tyPrimNm{G}$ and
-set of primitive operations typed by $\Gamma_\Grp$ that reflect the group structure at the value level:
-\begin{displaymath}
-  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
-    1 & \tyPrim{G}{1} \\
-    (\cdot) & \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to \tyPrim{G}{a\cdot b} \\
-    \ ^{-1} & \forall a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}
-  \end{array}
-\end{displaymath}
-One instance is the sort $\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ operations.
+\newcommand{\Grp}{\mathit{Grp}} To prove non-definability results,
+consider %the general case of 
+a sort $\mathsf{G}$ with the equational
+theory of abelian groups.  We assume a group-indexed primitive type
+$\tyPrimNm{G}$ and set of primitive operations typed by $\Gamma_\Grp$
+that reflect the group structure at the value level, namely $1 :
+\tyPrim{G}{1}$, $ (\cdot) : \forall
+a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to
+\tyPrim{G}{a\cdot b}$, and $\ ^{-1} : \forall
+a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$.
+%\begin{displaymath} \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
+%  1 & \tyPrim{G}{1} \\ (\cdot) & \forall
+%  a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to
+%  \ty%Prim{G}{a\cdot b} \\ \ ^{-1} & \forall
+%  a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to
+%  \tyPrim{G}{a^{-1}} \end{array} \end{displaymath}
+One instance is the sort $\SynTransl{2}$ with $0$, $(+)$ and
+$\mathrm{negate}$ operations.
 
-For index-erasure interpretation, we assume a (not necessarily abelian) group $(G, 1, \cdot,
-\ ^{-1})$. Let $Z(G)$ be the centre of the $G$: i.e.,~the set of
-elements that commute with every other element in $G$. We set
-$\tyPrimSem{G} = G$. The primitive operations in $\Gamma_\Grp$ are
-interpreted just as the unit, multiplication and inverse operations
-of $G$. For each index context $\Delta$, the set
+For index-erasure interpretation, we assume a (not necessarily
+abelian) group $(G, 1, \cdot, \ ^{-1})$. Let $Z(G)$ be the centre of
+the $G$ %: i.e.,~the set of elements that commute with every other
+%element in $G$. We set 
+and $\tyPrimSem{G} = G$. The primitive operations
+in $\Gamma_\Grp$ are interpreted just as the unit, multiplication and
+inverse operations of $G$. For each index context $\Delta$, the set
 $\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be
@@
 
-We write $i^n$ for $n$-fold iteration of the group operation.
+Writing $i^n$ for $n$-fold iteration of the group operation, we have:
 \begin{theorem}

### 2012-07-10 paper/section-introduction.tex
Lots of little typesetting fixes and typos
Patty's changes to shorten the text

@@
 $\rho(\alpha)$.  Rather beautifully, the standard notion of uniform
-continuity can then be expressed as a type:
-\begin{displaymath}
-  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
-\end{displaymath}
+continuity can then be expressed as %a type:
+%\begin{displaymath}
+ $ \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
+%\end{displaymath}
 
@@
 using strongly-typed term representations
-throughout~\cite{TypedSyntax}, available from
-\\\url{github.com/bobatkey/algebraically-indexed-types/tree/master/coq}
+throughout~\cite{TypedSyntax}. The formalisation is available from
+\\{\small \url{https://github.com/bobatkey/algebraically-indexed-types}}
 

### 2012-07-10 paper/section-metric.tex
Patty's changes to shorten the text

@@
 existential types, % in our framework,
-we can state the standard $\epsilon$-$\delta$ definition of
-\emph{uniform} continuity as a type:
-\begin{displaymath}
-  \forall \epsilon \mathord: \metricSort.\ \exists \delta\mathord: \metricSort.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
-\end{displaymath}
-Given a program $M$ with this type, we obtain from
-\thmref{thm:abstraction} a free theorem, which is exactly the
-definition of uniform continuity:
+%we can state 
+the standard $\epsilon$-$\delta$ definition of
+\emph{uniform} continuity can be expressed as %a type:
+%\begin{displaymath}
+$  \forall \epsilon \mathord: \metricSort.\ \exists \delta\mathord: \metricSort.\\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
+%\end{displaymath}
+For any program $M$ of this type, %we obtain from
+\thmref{thm:abstraction} gives a free theorem that is exactly 
+%the definition of 
+uniform continuity:
 \begin{displaymath}

### 2012-07-10 paper/section-instantiations.tex

--- 2012-07-10 paper/section-instantiations.tex
@@
 \label{sec:types-indexed-abelian-groups-indef}
+
 \newcommand{\Grp}{\mathit{Grp}}
@@
 
-For index-erasure interpretation, we assume a (not necessarily abelian) group $(G, 1, \cdot,
-\ ^{-1})$. Let $Z(G)$ be the centre of the $G$: i.e.,~the set of
-elements that commute with every other element in $G$. We set
-$\tyPrimSem{G} = G$. The primitive operations in $\Gamma_\Grp$ are
-interpreted just as the unit, multiplication and inverse operations
-of $G$. For each index context $\Delta$, the set
+For index-erasure interpretation, we assume a (not necessarily
+abelian) group $(G, 1, \cdot, \ ^{-1})$. Let $Z(G)$ be the centre of
+the $G$ %: i.e.,~the set of elements that commute with every other
+%element in $G$. We set 
+and $\tyPrimSem{G} = G$. The primitive operations
+in $\Gamma_\Grp$ are interpreted just as the unit, multiplication and
+inverse operations of $G$. For each index context $\Delta$, the set
 $\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be

### 2012-07-10 paper/geometry-twocolumn.tex
Fix typesetting of absolute value operation

@@
     \end{array}\\
-    \mathrm{abs} &:& \forall s \mathord:\SynGL{1}.\ \tyPrim{real}{s} \to \tyPrim{real}{|s|} %\\
+    \mathrm{abs} &:& \forall s \mathord:\SynGL{1}.\ \tyPrim{real}{s} \to \tyPrim{real}{\abs{s}} %\\
 %    \mathrm{sqrt} &:& \tyPrim{real}{1} \to \tyPrim{real}{1}
@@
 determinant, $\det B$, which takes an inhabitant of $\SynGL{2}$ to its
-determinant in $\SynGL{1}$, and absolute value, $|e|$, which takes
+determinant in $\SynGL{1}$, and absolute value, $\abs{e}$, which takes
 scaling factors to scaling factors. We also refine the type
@@
 Since the absolute value of the determinant of an orthogonal
-transformation is always $1$, we assume $|\det (\iota_O O)| = 1$ to
+transformation is always $1$, we assume $\abs{\det (\iota_O O)} = 1$ to
 hold.
@@
       \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to
-      \tyPrim{vec}{B, t} \to \tyPrim{real}{|\det B|}
+      \tyPrim{vec}{B, t} \to \tyPrim{real}{\abs{\det B}}
       \\ \mathrm{area}\ [B]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} *
@@
   sign of the determinant. We remove this symmetry as well using
-  $\mathrm{abs}$. This gives a value of type $\tyPrim{real}{|\det B|}$
+  $\mathrm{abs}$. This gives a value of type $\tyPrim{real}{\abs{\det B}}$
   which we multiply by $\frac{1}{2}$ to recover the area of the

### 2012-07-10 paper/section-general-framework.tex
Fix typesetting of absolute value operation
Rearrange the construction of non-compositional relational environments a bit to shorten it

@@
   determinant, and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
-  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, |\cdot| \}$, where
+  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, \abs{\cdot} \}$, where
   $G \in \{\SynGL{1}, \SynGL{2}, \SynOrth{2}\}$, and
@@
       \indexOpArity(\det) & ([\SynGL{2}], \SynGL{1}) &
-      \indexOpArity(|\cdot|) & ([\SynGL{1}], \SynGL{1})
+      \indexOpArity(\abs{\cdot}) & ([\SynGL{1}], \SynGL{1})
     \end{array}
@@
   the appropriate axioms. We also assume that the operations $\iota_O,
-  \iota_1, \det$ and $|\cdot|$ are group homomorphisms, and that
+  \iota_1, \det$ and $\abs{\cdot}$ are group homomorphisms, and that
   expressions of the form $\iota_1(s)$ commute with group
@@
   determinant of an orthgonal transformation is always $1$, so we also
-  assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
+  assume $\abs{\det(\iota_OO)} \stackrel{ax}\equiv 1$. Likewise, scaling
   maps have a determined determinant: $\det(\iota_1(s))
@@
 environments that allow us to derive useful consequences of
-\thmref{thm:abstraction}.
+\thmref{thm:abstraction}. To satisfy the hypothesis of
+\thmref{thm:abstraction} we still need to prove, for each
+instantiation of our general framework, that the program-level
+primitive operations preserve all the relations in our
+constructions. This is the content of Lemmas
+\ref{lem:geom-environments-1}, \ref{lem:geom-environments-2},
+\ref{lem:monoid-ops-related}, \ref{lem:environments-information-flow}
+and \ref{lem:metric-environments}.
 
@@
 
-To satisfy the hypothesis of \thmref{thm:abstraction} we still need to
-prove, for each instantiation of our general framework, that the
-program-level primitive operations preserve all the relations. This is
-the content of \lemref{lem:geom-environments},
-\lemref{lem:monoid-ops-related},
-\lemref{lem:environments-information-flow} and
-\lemref{lem:metric-environments}.
-
 \paragraph{Non-compositional Relational Environments}
-We now construct relational environments that can be used to prove
+We now construct relational environments that can be used for
 non-definability results. To this end, consider the typing
@@
 In \exref{ex:uninhabited-type} we claimed that this type is
-uninhabited. Intuitively, it is clear why: given the operations on
+uninhabited. Intuitively, it is clear why: with the operations on
 vectors in \autoref{sec:affine-vector-ops}, there is no way of
@@
 generated by the indices of the input types, then we cannot write a
-program of this type. 
-We now develop this idea to generate a
-family of sets of relational environments. The key idea is to assign
-different relational interpretations to types with indices that lie
-outside some finitely generated collection of index expressions.
-
-These relational environments are non-compositional: elements of
-$\relEnv{E}(\Delta,\Delta')$ are not purely constructed from elements
-of $\relEnv{E}(\Delta)$ and $\relEnv{E}(\Delta')$. As a consequence,
-these relational environments will only be
-\emph{pre-substitutive}. Therefore our non-definability results are
-subject to the quantifier-free conditions in \thmref{thm:abstraction}.
+program of this type.  The key idea is to assign different relational
+interpretations to types with indices that lie outside some generated
+collection of index expressions.
 
@@
 
-We construct relational environments parameterised by sets of terms
-$S$. We assign the equality relation to primitive types whose index
-arguments are generated up to equivalence by $S$. For terms whose
-index arguments are not in $\Gen^\equiv(S)$, we require a default
-binary relation $R^\bullet_{\tyPrimNm{X}} \subseteq
-\Eq_{\tyPrimSem{\tyPrimNm{S}}}$. We define the non-compositional
-family of sets of relational environments as
-$\relEnv{E}_{R^\bullet}(\Delta) = \{ \rho^S \sepbar \forall e \in S.\
-\exists s. \Delta \vdash e : s \}$, where
+Relational environments are parameterised by sets of terms $S$:
 \begin{displaymath}
@@
 \end{displaymath}
+Thus we assign the equality relation to primitive types whose index
+arguments are generated up to equivalence by $S$. For types whose
+indices are not in $\Gen^\equiv(S)$, we require a binary relation
+$R^\bullet_{\tyPrimNm{X}} \in \Rel(\tyPrimSem{\tyPrimNm{X}})$. We set
+$\relEnv{E}_{R^\bullet}(\Delta) = \{ \rho^S \sepbar \forall e \in S.\
+\exists s. \Delta \vdash e : s \}$.
+
+We have not been able to prove that this family of sets of relational
+interpretations is substitutive. Therefore our non-definability
+results are subject to the quantifier-free conditions in
+\thmref{thm:abstraction}.
+
+% Hence we must make do with the
+% following theorem:
+
+% These relational environments are non-compositional: elements of
+% $\relEnv{E}(\Delta,\Delta')$ are not purely constructed from elements
+% of $\relEnv{E}(\Delta)$ and $\relEnv{E}(\Delta')$. As a consequence,
+% these relational environments will only be
+% \emph{pre-substitutive}.
 

### 2012-07-10 paper/section-instantiations.tex
Fix typesetting of absolute value operation
Make geometry section more consistent with rest of paper

@@
 the indexing theory of \autoref{sec:motivating-examples}, and present
-more general and formally-justified theorems for free, type
+more general and formally-justified free theorems, type
 isomorphisms, and non-definability results.
@@
 
-We define the ``standard'' model of the indexing theory for the
-two-dimensional geometry example as follows. Each of the sorts is
-interpreted just as its semantic counterpart: $\semSort{\SynTransl{2}}
-= \Transl{2}$, $ \semSort{\SynGL{2}} = \GL{2}$, $
-\semSort{\SynOrth{2}} = \Orth{2}$, and $ \semSort{\SynGL{1}} =
-\GL{1}$.
+We define the model of the indexing theory for the two-dimensional
+geometry instantiation as follows. Each of the sorts is interpreted just as
+its semantic counterpart: $\semSort{\SynTransl{2}} = \Transl{2}$, $
+\semSort{\SynGL{2}} = \GL{2}$, $ \semSort{\SynOrth{2}} = \Orth{2}$,
+and $ \semSort{\SynGL{1}} = \GL{1}$.
 %  \begin{displaymath}
@@
 determinant) is interpreted by the intended semantic operation, and
-  clearly satisfy the axioms in \exref{ex:two-dim-geo-axioms}. The
-  assignment of relations for the primitive types $\tyPrimNm{vec}$ and
-  $\tyPrimNm{real}$ is given by: % the following definitions:
-  \begin{displaymath}
-    \begin{array}{ll}
-      R^{\bullet}_{\tyPrimNm{vec}} = \{( \vec{0}, \vec{0} ) \}&
-      R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
-      R^{\bullet}_{\tyPrimNm{real}} = \{(0,0)\} &
-      R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
-    \end{array}
-  \end{displaymath}
-  The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
-  %are as expected. % given the informal definitions we gave in
-formalise the informal ones from~\autoref{sec:motivating-examples}, and 
-  the definitions of %We have set
-  $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ 
-  %to relate $\vec{0}$ and $0$ to themselves to 
-  account for the
-  polymorphic zero constants we have assumed.
+clearly satisfy the axioms in \exref{ex:two-dim-geo-axioms}. For the
+relational interpretations of $\tyPrim{vec}{B,t}$ and
+$\tyPrim{real}{s}$, we use the compositional construction from
+\autoref{sec:constr-rel-env}, with $R_{\tyPrimNm{vec}}(B,\vec{t}) = \{
+(\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \}$ and
+$R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}$. The
+types of the primitive vector operations presented in
+\autoref{sec:motivating-examples} are collected into a context
+$\Gamma_\Geom$ with a straightforward interpretation $\eta_\Geom \in
+\ctxtSem{\Gamma_\Geom}$.
+  % \begin{displaymath}
+  %   \begin{array}{l}
+  %     R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
+  %     R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
+  %   \end{array}
+  % \end{displaymath}
+%   The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
+%   %are as expected. % given the informal definitions we gave in
+% formalise the informal ones from~\autoref{sec:motivating-examples}, and 
+%   the definitions of %We have set
+%   $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ 
+%   %to relate $\vec{0}$ and $0$ to themselves to 
+%   account for the
+%   polymorphic zero constants we have assumed.
 %\end{example}
 
-%\begin{lemma}
-%  The thing about $\Gamma_\Geom$.
-%\end{lemma}
+\begin{lemma}\label{lem:geom-environments-1}
+  For all $\Delta$ and $\rho \in \relEnv{E}_\Geom(\Delta)$, $(\eta_\Geom,\eta_\Geom) \in \rsem{\Gamma_\Geom}{\relEnv{E}_\Geom}\rho$.
+\end{lemma}
 
-\subsection{Theorems for Free}
+\subsection{Free Theorems}
 \label{sec:theorems-for-free}
@@
     \mathrm{area} : \forall B\mathord:\SynGL{2}, t\mathord:\SynTransl{2}.\ \\
-    \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{|\det B|}
+    \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{\abs{\det B}}
   \end{array}
@@
 \begin{displaymath}
-  |\det B|(\tmSem{\mathrm{area}}\ \vec{x}\ \vec{y}\ \vec{z}) = \tmSem{\mathrm{area}}\ (B\vec{x} + \vec{t})\ (B\vec{y} + \vec{t})\ (B\vec{z} + \vec{t})
+  \abs{\det B}(\tmSem{\mathrm{area}}\ \vec{x}\ \vec{y}\ \vec{z}) = \tmSem{\mathrm{area}}\ (B\vec{x} + \vec{t})\ (B\vec{y} + \vec{t})\ (B\vec{z} + \vec{t})
 \end{displaymath}
@@
 which the origin is fixed at $(0,0)$. It is sometimes said that an
-affine space is a vector space that's forgotten its origin; we have
+affine space is a vector space that has forgotten its origin; we have
 captured this in a type isomorphism.
@@
 \begin{align*}
-\transdn{1} &= \lambda x. x \qquad
+\transdn{\tyPrimNm{unit}} &= \lambda x. x \qquad
 \transdn{A_1\times A_2} = \lambda p. (\transdn{A_1}(\pi_1 p),\transdn{A_2}(\pi_2 p))\\ 
@@
 
-\subsection{Indefinability}
+\subsection{Non-definability}
 \label{sec:types-indexed-abelian-groups-indef}
 
-\newcommand{\Grp}{\mathit{Grp}}
-For proving non-definability results, we
-consider the general case of a sort $\mathsf{G}$ with the equational theory of abelian groups.
-We assume a group-indexed primitive type $\tyPrimNm{G}$ and
-set of primitive operations typed by $\Gamma_\Grp$ that reflect the group structure at the value level,
-with identity $1:\tyPrim{G}{1}$, operation
-$(\cdot): \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to \tyPrim{G}{a\cdot b}$
-and inverse $\ ^{-1}: \forall a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$.
-One instance is the sort~$\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ operations.
+\newcommand{\Grp}{\mathit{Gr}}
 
-For index-erasure interpretation, we assume a (not necessarily
-abelian) group $(G, 1, \cdot, \ ^{-1})$. Let $Z(G)$ be the centre of
-the $G$ %: i.e.,~the set of elements that commute with every other
-%element in $G$. We set 
-and $\tyPrimSem{G} = G$. The primitive operations
-in $\Gamma_\Grp$ are interpreted just as the unit, multiplication and
-inverse operations of $G$. For each index context $\Delta$, the set
-$\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be
-$\{ \rho_{(H,h)} \sepbar H\textrm{ is a subgroup of }(\Delta
-\Rightarrow \mathsf{G})/\equiv, h\textrm{ is a group homomorphism }H
-\to Z(G) \}$, where
-\begin{displaymath}
-  \rho_{(H,h)}\ \tyPrimNm{G}\ (e) = \left\{
-    \begin{array}{ll}
-      \{ \} & \textrm{if}\ e\not\in H \\
-      \{ (x, h(e)\oplus x) \sepbar x \in G \} & \textrm{if}\ e\in H
-    \end{array}
-  \right.
-\end{displaymath}
+For proving non-definability results, we consider the general case of
+a sort $\mathsf{G}$ with the equational theory of abelian groups.  We
+assume a group-indexed primitive type $\tyPrimNm{G}$ and set of
+primitive operations typed by $\Gamma_\Grp$ that reflect the group
+structure at the value level: identity $1:\tyPrim{G}{1}$, operation
+$(\cdot): \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to
+\tyPrim{G}{b} \to \tyPrim{G}{a\cdot b}$ and inverse $\ ^{-1}: \forall
+a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$. For the
+index-erasure semantics, we assume some (not-necessarily abelian)
+group $G$ and $\tyPrimSem{\tyPrimNm{G}} = G$.  The interpretation
+$\eta_\Grp$ of the primitive operations in $\Gamma_\Grp$ is just the
+unit, multiplication and inverse operations of $G$. One instance is the
+sort~$\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ on
+$\tyPrim{vec}{t}$.
 
-\begin{lemma}
+We use the construction of non-compositional relational environments
+from \autoref{sec:constr-rel-env}, with $R^\bullet_{\tyPrimNm{G}} =
+\emptyset$, to give a pre-substitutive family $\relEnv{E}_\Grp$. This
+gives us the following lemma and theorem.
+
+\begin{lemma}\label{lem:geom-environments-2}
   For all $\Delta$ and $\rho \in \relEnv{E}_\Grp(\Delta)$,
-  $(\eta_\Grp,\eta_\Grp) \in \rsem{\Gamma_\Grp}{\relEnv{E}}\rho$.
+  $(\eta_\Grp,\eta_\Grp) \in \rsem{\Gamma_\Grp}{\relEnv{E}_\Grp}\rho$.
 \end{lemma}
+
+% For index-erasure interpretation, we assume a (not necessarily
+% abelian) group $(G, 1, \cdot, \ ^{-1})$. Let $Z(G)$ be the centre of
+% the $G$ %: i.e.,~the set of elements that commute with every other
+% %element in $G$. We set 
+% and $\tyPrimSem{G} = G$. For each index context $\Delta$, the set
+% $\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be
+% $\{ \rho_{(H,h)} \sepbar H\textrm{ is a subgroup of }(\Delta
+% \Rightarrow \mathsf{G})/\equiv, h\textrm{ is a group homomorphism }H
+% \to Z(G) \}$, where
+% \begin{displaymath}
+%   \rho_{(H,h)}\ \tyPrimNm{G}\ (e) = \left\{
+%     \begin{array}{ll}
+%       \{ \} & \textrm{if}\ e\not\in H \\
+%       \{ (x, h(e)\oplus x) \sepbar x \in G \} & \textrm{if}\ e\in H
+%     \end{array}
+%   \right.
+% \end{displaymath}
 
@@
   natural numbers $A$ and $m$-by-$1$ matricies of integers $b$,
-  there exists a program $M$ such that
-  \begin{equation}
-    \label{eq:fo-group-type}
-    \Gamma_\Grp \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
-  \end{equation}
-  where $e_j = i_1^{A_{j1}}\cdots i_m^{A_{jm}}$ and $e = i_1^{b_1}\cdots i_m^{b_m}$, if and only if there is a solution $\vec{x}$
-  in integers to the equation $A \vec x = b$.
+  there exists a quantifier-free program $M$ such that
+  \begin{displaymath}
+    i_1,\dots,i_m\mathord:\mathsf{G}; \Gamma_\Grp \vdash M : \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
+  \end{displaymath}
+  where $e_j = i_1^{A_{j1}}\cdots i_m^{A_{jm}}$ and $e = i_1^{b_1}\cdots i_m^{b_m}$, iff there is a solution $\vec{x}$
+  in the integers to the equation $A \vec x = b$.
 \end{theorem}
-The uninhabited type  $\forall t \mathord:
-  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$
-from \exref{ex:uninhabited-type} is an instance of this theorem.
-For types indexed by the scaling group $\SynGL{1}$ the situation is more subtle, as
-the index-erasure interpretation of $\tyPrim{real}{s}$ is group $\GL{1}$ together with zero.
-We can show, nonetheless, that $\forall s \mathord:
-  \SynGL{1}.\ \tyPrim{real}{s \cdot s} \to \tyPrim{real}{s}\cong 1$, reflecting 
-the fact that any term of this type must be $\lambda x.0$.
+The uninhabited type $\forall t \mathord: \SynTransl{2}.\
+\tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ from
+\exref{ex:uninhabited-type} is an instance of this theorem.  For types
+indexed by the scaling group $\SynGL{1}$ the situation is more subtle,
+as we have assumed that the type $\tyPrim{real}{s}$ has a polymorphic zero
+(see \autoref{fig:real-ops}). We can show, nonetheless, that $\forall
+s \mathord: \SynGL{1}.\ \tyPrim{real}{s^2} \to \tyPrim{real}{s}$ is
+only inhabited by the constantly zero function by setting
+$R^\bullet_{\tyPrimNm{real}} = \{(0,0)\}$. If we augment our sort
+$\SynGL{1}$ with a square root operation $s^{1/2}$, then it is
+possible to add a primitive square root function and still be able to
+prove that the cube root type $\forall s \mathord: \SynGL{1}.\
+\tyPrim{real}{s^3} \to \tyPrim{real}{s}$ is only inhabited by the
+constantly zero function. This is a consequence of the impossibility
+of generating $s$ from $s^3$ with only the abelian group operations
+and square root.
 

### 2012-07-10 paper/geometry-twocolumn.tex
Read-through changes
- Fix missing word in abstract
- Replace "Abstraction Theorem" with "abstraction theorem" throughout
- Remove forward reference to discussion of absolute value
- Fix a few typos in the framework section
- attribute the definition of contextual equivalence in terms of den. sem. to Hofmann
- make sure it is always "relational environments", not "relation"
- Rename the "compositional" and "non-compositional" relational semantics to "model-based" and "definability" respectively
- Remove the statement of the lemma about substitution for relational interp of contexts, to save sp

@@
 For example, cross products vary with scalings of the plane. We
-can incorporate (in)variance information about geometric primitives
+incorporate (in)variance information about geometric primitives
 into type systems via algebraically indexed types.
@@
 distinguishing between them is the key feature of \emph{affine
-  geometry} (see, for example, Chapter 2 of
+  geometry} (see, for example, Chapter 2 of Gallier's book
 \cite{gallier11geometric}). Nevertheless, computational geometry
@@
 %numbers that they appear to be different.
-This suggests that programs that manipulate %data representing points
-should be invariant with respect to changes in the choice of
-origin. Program sthat manipulate vectors, on the other hand, should
+This suggests that programs that manipulate points %data representing points
+should be invariant with respect to changes of
+origin. Programs that manipulate vectors, on the other hand, should
 not be invariant under change of origin. Different vectors represent
@@
 \end{displaymath}
-The index-erasure and relational interpetations of 
+The index-erasure and relational interpretations of 
 $\forall t\mathord:\SynTransl{2}.A$ are:
@@
   isomorphism in \exref{ex:type-iso}. Next, we compute the cross
-  product of the two vectors is computed, which gives the area of the
+  product of the two vectors, which gives the area of the
   parallelogram described by the sides of the triangle and has type
@@
 non-zero scaling factors, whereas Kennedy's indexes by strictly
-positive ones. We examine the typing of the absolute value operation
-further in \autoref{sec:types-indexed-abelian-groups}.
+positive ones.
 
@@
   In \autoref{sec:types-indexed-abelian-groups-indef} we revisit
-  Kennedy's proof and show that even if we add square root as a
+  Kennedy's result and show that even if we add square root as a
   primitive operation---with the type above---then it is still not

### 2012-07-10 paper/paper.tex
Lots of little typesetting fixes and typos
Fix typesetting of absolute value operation
Make geometry section more consistent with rest of paper
Read-through changes
- Fix missing word in abstract
- Replace "Abstraction Theorem" with "abstraction theorem" throughout
- Remove forward reference to discussion of absolute value
- Fix a few typos in the framework section
- attribute the definition of contextual equivalence in terms of den. sem. to Hofmann
- make sure it is always "relational environments", not "relation"
- Rename the "compositional" and "non-compositional" relational semantics to "m

@@
            {University of Strathclyde}
-           {\{Robert.Atkey,Patricia.Johann\}@strath.ac.uk}
+           {$\{$Robert.Atkey,Patricia.Johann$\}$@strath.ac.uk}
 
@@
 
+\newcommand{\abs}[1]{\lvert #1 \rvert}
+
 \newcommand{\GL}[1]{\mathrm{GL}_#1}
@@
 \newcommand{\SynScal}{\mathsf{Scal}}
-
-% \newcommand{\GLtwo}{\mathrm{GL}_2}
-% \newcommand{\SynGLtwo}{\mathsf{GL}_2}
-% \newcommand{\GLone}{\mathrm{GL}_1}
-% \newcommand{\SynGLone}{\mathsf{GL}_1}
 
@@
 
-\newcommand{\indexOp}[1]{\texttt{#1}}
+\newcommand{\indexOp}[1]{\textup{\texttt{#1}}}
 \newcommand{\idxTms}[2]{\mathrm{IdxTm}(#1 \vdash #2)}
@@
 \newcommand{\tyX}[1]{\texttt{X}\langle #1 \rangle}
-\newcommand{\isType}{\textsf{ type}}
-\newcommand{\isCtxt}{\textsf{ ctxt}}
+\newcommand{\isType}{\textup{\textsf{ type}}}
+\newcommand{\isCtxt}{\textup{\textsf{ ctxt}}}
 \newcommand{\Ty}{\textsf{Ty}}
@@
 
-\newcommand{\Geom}{\mathit{Geom}}
+\newcommand{\Geom}{\mathit{2D}}
 
@@
   Outside computer science, invariance is a common theme running
-  through many areas mathematics and physics. For example, the area of
+  through many areas of mathematics and physics. For example, the area of
   a triangle is unaltered by rotation or flipping. If we scale a
@@
 Proofs for free: Parametricity for dependent types.
-{\em Journal of Functional Programming} 22(02), pp. 107-152, 2012.
+{\em Journal of Functional Programming} 22(2), pp. 107-152, 2012.
 
@@
 A.~M.~Pitts.  Parametric Polymorphism and Operational Equivalence.
-{\em Mathematical Structures in Computer Science} 10, pp. 321-359,
+{\em Mathematical Structures in Computer Science} 10(3), pp. 321-359,
 2000.
@@
 J.~Reed and B.~C.~Pierce.  Distance Makes the Types Grow Stronger.
-Proceeding, International Conference on Functional Programming,
+Proceedings, International Conference on Functional Programming,
 pp. 157-169, 2010.

### 2012-07-10 paper/section-introduction.tex
Read-through changes
- Fix missing word in abstract
- Replace "Abstraction Theorem" with "abstraction theorem" throughout
- Remove forward reference to discussion of absolute value
- Fix a few typos in the framework section
- attribute the definition of contextual equivalence in terms of den. sem. to Hofmann
- make sure it is always "relational environments", not "relation"
- Rename the "compositional" and "non-compositional" relational semantics to "model-based" and "definability" respectively
- Remove the statement of the lemma about substitution for relational interp of contexts, to save sp

@@
 polymorphism is \emph{relational parametricity}, whose central result
-is Reynolds' Abstraction Theorem~\cite{reynolds83types}. Its striking
+is Reynolds' abstraction theorem~\cite{reynolds83types}. Its striking
 consequences include the well-known ``free theorems'' for polymorphic
@@
 with algebraic structure. For example, in computational geometry, points in
-2-space can be indexed by a coordinate frame; in information-flow
+the plane can be indexed by a coordinate frame; in information-flow
 security, computations are indexed by principals; in differential privacy, types
@@
 \]
-for any translation $t\in\Transl{2}$.
+for any translation $t$ in the group $\Transl{2}$ of translations.
 
@@
 refined type system for information flow based on logic, and a simple
-type system for basic uniform continuity analysis.
+type system with distance-indexed types.
 \item 
@@
 We describe a relational semantics for the type system and prove an
-analogue of Reynolds' Abstraction Theorem. We give precise conditions
+analogue of Reynolds' abstraction theorem. We give precise conditions
 on sets of relation environments under which our abstraction theorem
@@
 \end{itemize}
-We have fully formalised our framework and Abstraction Theorem in Coq,
+We have fully formalised our framework and abstraction theorem in Coq,
 using strongly-typed term representations

### 2012-07-10 paper/section-metric.tex
Read-through changes
- Fix missing word in abstract
- Replace "Abstraction Theorem" with "abstraction theorem" throughout
- Remove forward reference to discussion of absolute value
- Fix a few typos in the framework section
- attribute the definition of contextual equivalence in terms of den. sem. to Hofmann
- make sure it is always "relational environments", not "relation"
- Rename the "compositional" and "non-compositional" relational semantics to "model-based" and "definability" respectively
- Remove the statement of the lemma about substitution for relational interp of contexts, to save sp

@@
 must satisfy the property that for all $\epsilon_1, \epsilon_2 > 0$
-and $x, x', y, y' \in \mathbb{R}$,
-\begin{displaymath} % FIXME: include \eta_\Metric
+and $x, x', y, y' \in \mathbb{R}$:
+\begin{displaymath}
   \begin{array}{l}
     \textrm{if }|x - x'| < \epsilon_1 \textrm{ and } |y - y'| < \epsilon_2 \textrm{ then}\\
-    \hspace{2em}|\tmSem{M}\ x\ y - \tmSem{M}\ x'\ y'| < \epsilon_1 + \epsilon_2
+    \hspace{2em}|\tmSem{M}\ \eta_\Metric\ x\ y - \tmSem{M}\ \eta_\Metric\ x'\ y'| < \epsilon_1 + \epsilon_2
   \end{array}
 \end{displaymath}
-A type system with a relational interpretation that tracks distances
-between values has %previously 
-been investigated by Reed and Pierce
-\cite{reed10distance} in the setting of differential privacy. Their
-system uses a linear type discipline to ensure that all programs are
-non-expansive (i.e., the distance between the outputs is no more greater
-than the distance between the inputs). Here, we instantiate our general
-framework %from \autoref{sec:a-general-framework} %in order 
-to express
-non-expansivity, and more, via algebraically indexed types.
 
@@
 For the families of sets of relational environments
-$\relEnv{E}_\Metric$, we use the compositional construction in
+$\relEnv{E}_\Metric$, we use the model-based construction in
 \autoref{sec:constr-rel-env}, interpreting the primitive type
@@
 
+\paragraph{Function Sensitivity}
+A type system with a relational interpretation that tracks distances
+between values has %previously
+been investigated by Reed and Pierce \cite{reed10distance} in the
+setting of differential privacy. Their system uses a linear type
+discipline to ensure that all programs are of $c$-sensitive (i.e., the
+distance between the outputs is no more greater than $c$ times the
+distance between the inputs, for some constant $c$). We can express
+their central concept of $c$-sensitivity (for functions on the reals)
+as an algebraically-indexed type: $\forall
+\epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \to
+\tyPrim{real}{\frac{1}{c}\epsilon}$. Investigating the precise
+connection between their system and ours is left to future work.
+
 % \begin{example}[Differential Privacy]

### 2012-07-10 paper/paper.tex
Cardelli ref in bib

@@
   Outside computer science, invariance is a common theme running
-  through many areas of mathematics and physics. For example, the area of
+  through many areas mathematics and physics. For example, the area of
   a triangle is unaltered by rotation or flipping. If we scale a
@@
 Proofs for free: Parametricity for dependent types.
-{\em Journal of Functional Programming} 22(2), pp. 107-152, 2012.
+{\em Journal of Functional Programming} 22(02), pp. 107-152, 2012.
+
+\bibitem{Cardelli}
+L.~Cardelli, P.~Gardner.
+Processes in Space.
+{\em Programs, Proofs, Processes}: Proc, 6th Conference on Computability in Europe, pp 78-87, LNCS vol. 6158, 2010.
 
@@
 A.~M.~Pitts.  Parametric Polymorphism and Operational Equivalence.
-{\em Mathematical Structures in Computer Science} 10(3), pp. 321-359,
+{\em Mathematical Structures in Computer Science} 10, pp. 321-359,
 2000.
@@
 J.~Reed and B.~C.~Pierce.  Distance Makes the Types Grow Stronger.
-Proceedings, International Conference on Functional Programming,
+Proceeding, International Conference on Functional Programming,
 pp. 157-169, 2010.

### 2012-07-10 paper/section-discussion.tex

@@
 
-In \autoref{sec:metric-types} we presented a type system with
-distance-indexed types, and noted the similarity with Reed and
-Pierce's system for non-expansivity \cite{reed10distance}. Exploring
-the relationship--particularly with respect to probability
-monads--also looks promising.
+% In \autoref{sec:metric-types} we presented a type system with
+% distance-indexed types, and noted the similarity with Reed and
+% Pierce's system for non-expansivity \cite{reed10distance}. Exploring
+% the relationship--particularly with respect to probability
+% monads--also looks promising.
 
@@
 algebraic structure (e.g.,~sets of read/write effect labels with an
-idempotent monoid structure). Benton {\em et al.}~ have used relational
+idempotent monoid structure). Benton {\em et al.}~have used relational
 interpretations to prove effect-dependent equivalences
@@
 of length $n$, indexed by elements of the permutation group
-$S_n$. Bernardy et al.~have presented a general framework for
+$S_n$. Bernardy {\em et al.}~have presented a general framework for
 relational reasoning and an abstraction theorem for dependent types
@@
 distance-indexed types it is possible that a new notion of
-equivalence, based on closeness, is required.
+equivalence based on closeness is required.
 

### 2012-07-10 paper/section-general-framework.tex

@@
 (rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
-\TirName{TySum}). The rule \TirName{TyPrim} %allows us to 
+\TirName{TySum}). 
+We use $\tyPrimNm{bool}$ as an abbreviation for $\tyPrimNm{unit} + \tyPrimNm{unit}$.
+The rule \TirName{TyPrim} %allows us to 
 forms, from a primitive type $\tyPrimNm{X}$ and appropriately sorted
@@
 %a well-indexed type 
-$\Delta' \vdash A \isType$, we 
+$\Delta' \vdash A \isType$, 
 %can apply $\sigma$ to $A$ to produce a new well-indexed type 
@@
 $\Delta$, we define the \emph{projection} simultaneous substitution
-$\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
+$\Delta, i\mathord:s \vdash \pi_{i\mathord:s} \Rightarrow \Delta$ as
 $\pi_{i\mathord:s} = (i_1,...,i_n)$. %The subscript on
@@
 %\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
-Relation environments must respect index expression equality: for
+Relational environments must respect index expression equality: for
 all $\tyPrimNm{X} \in \PrimTypeSet$,
@@
 %
-We write $\mathrm{RelEnv}(\Delta)$ for the set of all relation
+We write $\mathrm{RelEnv}(\Delta)$ for the set of all relational
 environments for a context $\Delta$. Given %a relational environment
@@
 
-As we stated above, index contexts $\Delta$ are interpreted as certain
+We interpret index contexts $\Delta$ as certain
 %sets of relation environments, i.e.,~
@@
 %
-Given a relation environment $\rho \in \relEnv{E}(\Delta)$, %we define
+Given a relational environment $\rho \in \relEnv{E}(\Delta)$, %we define
 the set of extensions %$\extends{\rho}{i\mathord:s}$ 
@@
 \rho \}$. The set of extensions of a relational environment will be
-used in interpretating the universal and existential type formers.  We
+used in interpreting the universal and existential type formers.  We
 assign a relational interpretation to all well-indexed types $\Delta
@@
 interpretation of its type: this is the abstraction theorem for every
-calculus that instantiates our general framework.
+instantiation of our general framework.
 
@@
 ignores the indexing information. In light of
-\lemref{lem:tyeqsubst-erasure}, %we define this function is defined
+\lemref{lem:tyeqsubst-erasure}, we do this %we define this function is defined
 directly on the syntax of well-typed programs, rather than on typing
@@
 are contextually equivalent with respect to syntactically defined
-applicative contexts.
+contexts, following Hofmann \cite{hofmann08correctness}.
 
@@
   programs $-;\Gamma_{\mathit{ops}} \vdash T : (\forall \Delta. \Gamma
-  \to A) \to \tyPrimNm{unit} + \tyPrimNm{unit}$, it is the case that
-  $\tmSem{T\ (\Lambda \Delta. \lambda \Gamma. M_1)}\eta_{\mathit{ops}}
-  = \tmSem{T\ (\Lambda \Delta. \lambda
-    \Gamma. M_2)}\eta_{\mathit{ops}}$.
+  \to A) \to \tyPrimNm{bool}$, it is the case that $\tmSem{T\ (\Lambda
+    \Delta. \lambda \Gamma. M_1)}\eta_{\mathit{ops}} = \tmSem{T\
+    (\Lambda \Delta. \lambda \Gamma. M_2)}\eta_{\mathit{ops}}$.
 \end{definition}
@@
 % of simultaneous substitutions as composition:
-\begin{lemma}\label{lem:ctxtsubst-rel}
-  If $\Delta' \vdash \Gamma \isCtxt$ and $\Delta \vdash \sigma
-  \Rightarrow \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
-  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =
-  \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma)}$.
-\end{lemma}
+% \begin{lemma}\label{lem:ctxtsubst-rel}
+%   If $\Delta' \vdash \Gamma \isCtxt$ and $\Delta \vdash \sigma
+%   \Rightarrow \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
+%   $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =
+%   \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma)}$.
+% \end{lemma}
 
@@
 
-We give two constructions of families of sets of relational
-environments that allow us to derive useful consequences of
+We give two constructions of families $\relEnv{E}$ of sets of
+relational environments that allow us to derive useful consequences of
 \thmref{thm:abstraction}. To satisfy the hypothesis of
@@
 
-\paragraph{Compositional Relational Environments}
-Our compositional relational environments are defined in terms of
+\paragraph{Model-based Relational Environments}
+Our model-based relational environments are defined in terms of
 models. To each sort $s \in \SortSet$ we assign a carrier set
@@
 
-\paragraph{Non-compositional Relational Environments}
+\paragraph{Definability Relational Environments}
 We now construct relational environments that can be used for

### 2012-07-10 paper/section-info-flow.tex

@@
 information flow properties. Tse and Zdancewic \cite{tse04translating}
-translated Abadi et al.'s calculus into System F, translating the
+translated Abadi \emph{et al.}'s calculus into System F, translating the
 monadic type $T_lA$ to $\alpha_l \to A$ for some free type variable
@@
 replacing each type variable $\alpha_l$ with a primitive type
-$\tyPrim{T}{e}$ of representations of the truth of a logical
-proposition $e$. The relationships between security levels are now
+$\tyPrim{T}{\phi}$ of representations of the truth of a logical
+proposition $\phi$. The relationships between security levels are now
 replaced by logical entailment, so we only have functions of type
@@
 $\tyPrim{T}{\phi}$ as the identity relation if $\phi$ is true and the
-empty relation if $\phi$ is false. This recovers the information flow
-properties of Abadi {\em et al.} and Tse and Zdancewic.
+empty relation if $\phi$ is false. We shall see that this recovers the
+information flow properties of Abadi {\em et al.} and Tse and
+Zdancewic.
 
@@
 given by the relational semantics. For the relational environments
-$\relEnv{E}_\infoflow$, we use the compositional construction in
+$\relEnv{E}_\infoflow$, we use the model-based construction in
 \autoref{sec:constr-rel-env}, with an arbitrary boolean algebra $L$
@@
 \paragraph{Information Flow} We think of logical expressions as
-``composite principals''. That is, propositional variables represent
+``composite principals''. That is, propositional variables representing
 atomic principals that are combined using the logical connectives. We
@@
 $\psi$. In terms of security levels, the ordering is reversed: if a
-high security principal is satisfied, then all of their underlings
+high security principal is satisfied, then all of their subordinates
 must also be satisfied.
 
-We adapt Tse and Zdancewic's translation of Abadi et al.'s monadic
+We adapt Tse and Zdancewic's translation of Abadi \emph{et al.}'s monadic
 type to our setting. We define a type abbreviation $T_\phi A =
@@
 protected by this principal is indistinguishable from any other
-data. Thus a program cannot get access to the exact value. From this
+data, and a program cannot get access to the exact value. From this
 and \thmref{thm:abstraction}, we obtain the following information flow
@@
 \end{theorem}
-This result also holds if we move to logics other than boolean
+Note that if $\psi \leq \phi$, then it is always possible to write the
+identity function with this type, using the $\mathrm{up}$ operation.
+The theorem also holds if we move to logics other than boolean
 logic. For example, if our equational theory models intuitionistic

### 2012-07-10 paper/section-instantiations.tex

@@
 relational interpretations of $\tyPrim{vec}{B,t}$ and
-$\tyPrim{real}{s}$, we use the compositional construction from
+$\tyPrim{real}{s}$, we use the model-based construction from
 \autoref{sec:constr-rel-env}, with $R_{\tyPrimNm{vec}}(B,\vec{t}) = \{
@@
 
-Consider the type of the area %of a triangle 
+Consider the type of the triangle area %of a triangle 
 function from
@@
 assume a group-indexed primitive type $\tyPrimNm{G}$ and set of
-primitive operations typed by $\Gamma_\Grp$ that reflect the group
-structure at the value level: identity $1:\tyPrim{G}{1}$, operation
-$(\cdot): \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to
-\tyPrim{G}{b} \to \tyPrim{G}{a\cdot b}$ and inverse $\ ^{-1}: \forall
-a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$. For the
-index-erasure semantics, we assume some (not-necessarily abelian)
-group $G$ and $\tyPrimSem{\tyPrimNm{G}} = G$.  The interpretation
-$\eta_\Grp$ of the primitive operations in $\Gamma_\Grp$ is just the
-unit, multiplication and inverse operations of $G$. One instance is the
+primitive operations typed by $\Gamma_\Grp$ that include the group
+identity $1:\tyPrim{G}{1}$, operation $(\cdot): \forall
+a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to
+\tyPrim{G}{a\cdot b}$ and inverse $\ ^{-1}: \forall
+a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$. The
+context $\Gamma_\Grp$ may also include members with types of the form
+$\forall \vec{a}\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to ...
+\tyPrim{G}{e_n} \to \tyPrim{G}{e}$ as long as the result index $e$ is
+generated by the set $\{e_1,...,e_n\}$. One instance is the
 sort~$\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ on
-$\tyPrim{vec}{t}$.
+$\tyPrim{vec}{t}$. For the purposes of non-definability results, we
+may choose an arbitrary index-erasure semantics that supports the
+primitive operations.
 
-We use the construction of non-compositional relational environments
+We use the construction of definability relational environments
 from \autoref{sec:constr-rel-env}, with $R^\bullet_{\tyPrimNm{G}} =
@@
   Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matrices of
-  natural numbers $A$ and $m$-by-$1$ matricies of integers $b$,
-  there exists a quantifier-free program $M$ such that
+  natural numbers $A$ and $m$-by-$1$ matricies of integers $\vec{b}$,
+  there exists a quantifier-free program $M$ such that:
   \begin{displaymath}
@@
   where $e_j = i_1^{A_{j1}}\cdots i_m^{A_{jm}}$ and $e = i_1^{b_1}\cdots i_m^{b_m}$, iff there is a solution $\vec{x}$
-  in the integers to the equation $A \vec x = b$.
+  in the integers to the equation $A \vec x = \vec{b}$.
 \end{theorem}

### 2012-07-10 paper/section-monoid.tex

@@
 
-Above we investigated types indexed by abelian group expressions,
+Above, we investigated types indexed by abelian group expressions,
 whose elements are taken from some group. We now switch from abelian
@@
 For the relational environments, we use %make use of 
-the compositional
+the model-based
 construction in \autoref{sec:constr-rel-env}, setting the chosen model
@@
   Let $m$, $n$ be natural numbers. For all $n$-by-$m$-matrices of
-  natural numbers $A$ and $m$-by-$1$ matrices of natural numbers $b$,
-  there is a quantifier-free program $N$, with
+  natural numbers $A$ and $m$-by-$1$ matrices of natural numbers $\vec{b}$,
+  there is a quantifier-free program $N$, with:
   \begin{displaymath}
@@
   where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e = i_1^{b_1}\dots
-  i_m^{b_m}$ iff there is a solution $\vec{x} = (x_1,...,x_n)$ in the
-  natural numbers to $A x^\top = b$.
+  i_m^{b_m}$, iff there is a solution $\vec{x}$ in the
+  natural numbers to the equation $A \vec{x} = \vec{b}$.
 \end{theorem}
 
-The proof of this theorem uses the non-compositional relational
+The proof of this theorem uses the definability relational
 environments from \autoref{sec:constr-rel-env}, with

### 2012-07-10 paper/section-discussion.tex
cardelli description

@@
 %analogues of Reynolds' 
-an abstraction theorem holds. Geometric theorem
+an abstraction theorem holds. 
+Cardelli and Gardner describe a process calculus that builds in 3-d affine geometry~\cite{Cardelli},
+proving that process behaviour is invariant under affine transformations. Distinguishing points from vectors
+provides the appropriate abstraction barrier, and the geometric group is determined by inspecting
+term syntax. It would be interesting to recast their language in terms of our indexed types to
+obtain purely type-based invariance theorems.
+
+Geometric theorem
 proving is another %possible 
@@
 $\stackrel{sem}\sim$ in \defref{def:semantic-equality} is not an
-equivalence relation. If we assume that the interpretations of all the
+equivalence relation, but if the interpretations of all 
 primitive types are partial equivalence relations then semantic

### 2012-07-10 paper/section-general-framework.tex
cardelli description

@@
 \begin{lemma}\label{lem:tyeqsubst-relational}
+\par
   \begin{enumerate}
@@
 in \autoref{sec:algebraically-indexed-types}. Each
-well-type program is assigned an index-erasure semantics, building on
+well-typed program is assigned an index-erasure semantics, building on
 the index-erasure semantics of types %we defined 
@@
 
-\begin{theorem}\label{thm:soundness}
+\begin{theorem}[Soundness]\label{thm:soundness}
   $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 \stackrel{sem}\sim M_2 : A$ implies

### 2012-07-10 paper/paper.tex
POPL ICFP abbreviation

--- 2012-07-10 paper/paper.tex
@@
 M. Abadi, A. Banerjee, N. Heintze, and J. G. Riecke. A Core Calculus
-of Dependency. Proceedings, Symposium on Principles of Programming
-Languages, pp.147-160, 1999.
+of Dependency. Proceedings, POPL, pp.147-160, 1999.
 
@@
 
+\bibitem{Cardelli}
+L.~Cardelli, P.~Gardner.
+Processes in Space.
+{\em Programs, Proofs, Processes}: Proc, 6th Conference on Computability in Europe, pp 78-87, LNCS vol. 6158, 2010.
+
 \bibitem{CGAL}
@@
 S. Chaudhuri, S. Gulwani, and R. Lublinerman.  Continuity Analysis of
-Programs.  Proceedings, Symposium on Principles of Programming
-Languages, pp. 57-70, 2010.
+Programs.  Proceedings, POPL, pp. 57-70, 2010.
 
 \bibitem{gallier11geometric}
-J. Gallier. {\em Geometric Methods and Applications, For Computer
-Science and Engineering. Springer Science+Business Media, LLC}, 2011.
+J. Gallier. {\em Geometric Methods and Applications For Computer
+Science and Engineering}. Springer, 2011.
 
@@
 A.~Kennedy. Relational Parametricity and Units of Measure.
-Proceedings, Symposium on Principles of Programming Languages,
-pp. 442-455, 1997.
+Proceedings, POPL, pp. 442-455, 1997.
 
@@
 J.~Reed and B.~C.~Pierce.  Distance Makes the Types Grow Stronger.
-Proceedings, International Conference on Functional Programming,
-pp. 157-169, 2010.
+Proceedings, ICFP, pp. 157-169, 2010.
 
@@
 \bibitem{sabelfeld01per} 
-A.~Sabelfeld and D.~Sands.  A Per Model of Secure Information Flow in
+A.~Sabelfeld and D.~Sands.  A PER Model of Secure Information Flow in
 Sequential Programs.  {\em Higher-Order and Symbolic Computation} 14
@@
 S.~Tse and S.~Zdancewic.  Translating Dependency into Parametricity.
-Proceedings, International Conference on Functional Programming pp.
+Proceedings, ICFP,  pp.
 115-125, 2004.

### 2012-07-11 paper/section-general-framework.tex
Fix some small mistakes

@@
   \label{defn:valid-rel-env-family}
-  A family of sets of relational environments $\relEnv{E}(\Delta)
-  \subseteq \mathrm{RelEnv}(\Delta)$ is \emph{substitutive} if these
-  two properties are satisfied:
+  An index context-indexed family of sets of relational environments
+  $\{ \relEnv{E}(\Delta) \subseteq \mathrm{RelEnv}(\Delta) \}_\Delta$
+  is \emph{substitutive} if:
   \begin{enumerate}
-  \item Closure under simultaneous substitutions: if $\rho \in
-    \relEnv{E}(\Delta)$ and $\Delta' \vdash \sigma \Rightarrow \Delta$,
-    then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
+  \item it is closed under composition with simultaneous substitution:
+    if $\rho \in \relEnv{E}(\Delta)$ and $\Delta \vdash \sigma
+    \Rightarrow \Delta'$, then $\rho \circ \sigma \in
+    \relEnv{E}(\Delta')$; and
 %  \item If we have a pair of relation environments $\rho_1 \in
@@
 %    \pi_{i\mathord:s'} = \rho_2$.
-  \item If $\rho_1 \in \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2
+  \item if $\rho_1 \in \relEnv{E}(\Delta', i\mathord:s)$ and $\rho_2
     \in \relEnv{E}(\Delta)$, and $\Delta \vdash \sigma \Rightarrow
-    \Delta'$ is such that $\rho_1 \circ \pi_{i\mathord:s'} = \rho_2
+    \Delta'$ is such that $\rho_1 \circ \pi_{i\mathord:s} = \rho_2
     \circ \sigma$, then there exists a $\rho \in \relEnv{E}(\Delta,
-    i\mathord:s')$ such that $\rho \circ \sigma_{i\mathord:s'} =
-    \rho_1$ and $\rho \circ \pi_{i\mathord:s'} = \rho_2$.
+    i\mathord:s)$ such that $\rho \circ \sigma_{i\mathord:s} =
+    \rho_1$ and $\rho \circ \pi_{i\mathord:s} = \rho_2$.
     % such that the outer edges of the
@@
   variables that describes the types of the primitive operations, and
-  let $\eta_{\mathit{ops}} \in \ctxtSem{\Gamma}$ be an interpretation
+  let $\eta_{\mathit{ops}} \in \ctxtSem{\Gamma_{\mathit{ops}}}$ be an interpretation
   of these primitive operations.
@@
   \Gamma \vdash M_1 \stackrel{ctx}\approx M_2 : A$) if for all
-  programs $-;\Gamma_{\mathit{ops}} \vdash T : (\forall \Delta. \Gamma
+  programs $;\Gamma_{\mathit{ops}} \vdash T : (\forall \Delta. \Gamma
   \to A) \to \tyPrimNm{bool}$, it is the case that $\tmSem{T\ (\Lambda
@@
   $(\eta_{\mathit{ops}}, \eta_{\mathit{ops}}) \in
-  \rsem{\Gamma}{\relEnv{E}_{\mathit{ops}}}\rho$.
+  \rsem{\Gamma_{\mathit{ops}}}{\relEnv{E}_{\mathit{ops}}}\rho$.
 
@@
 non-definability results. To this end, consider the typing
-$t\mathord:\SynTransl{2}; \Gamma_{\Geom} \vdash M : \tyPrim{vec}{t +
-  t} \to \tyPrim{vec}{t}$ involving vectors indexed by translations.
+$t\mathord:\SynTransl{2}; \Gamma \vdash M : \tyPrim{vec}{t + t} \to
+\tyPrim{vec}{t}$, where $\Gamma$ contains the primitive operations on
+vectors indexed by translations from \autoref{sec:affine-vector-ops}.
 In \exref{ex:uninhabited-type} we claimed that this type is
@@
 program of this type.  The key idea is to assign different relational
-interpretations to types with indices that lie outside some generated
+interpretations to types with indices not in some generated
 collection of index expressions.

### 2012-07-11 paper/section-introduction.tex
Fix some small mistakes

@@
 This captures the fact that the area of a circle varies as the square
-of its radius, i.e., $\mathrm{areaCircle}\;(s\cdot r) =
-s^2\cdot \mathrm{areaCircle}\;r$ for any $s\neq 0$.  Here, $\alpha$ is interpreted as
-the \emph{units of measure} of the argument to $\mathrm{areaCircle}$,
-and `$\cdot$' composes units using the product. We can also add an
-inverse operation and identity unit of measure $1$, and then impose the
-algebraic laws of abelian groups. This permits identification of, for
-example, $\tyPrim{real}{\alpha\cdot\alpha^{-1}}$ with the type
+of its radius, i.e., $\mathrm{areaCircle}\;(s\cdot r) = s^2\cdot
+\mathrm{areaCircle}\;r$ for any $s\neq 0$ (i.e., any $s$ in the
+one-dimensional general linear group $\GL{1}$).  Here, $\alpha$ is
+interpreted as the \emph{units of measure} of the argument to
+$\mathrm{areaCircle}$, and `$\cdot$' composes units using the
+product. We can also add an inverse operation and identity unit of
+measure $1$, and then impose the algebraic laws of abelian
+groups. This permits identification of, for example,
+$\tyPrim{real}{\alpha\cdot\alpha^{-1}}$ with the type
 $\tyPrim{real}{1}$ of dimensionless constants.
@@
 difference can be assigned the type
-$\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\times\tyPrim{vec}\alpha\to\tyPrim{vec}0$,
+$\forall\alpha\mathord:\SynTransl{2}.\tyPrim{vec}\alpha\times\tyPrim{vec}\alpha\to\tyPrim{vec}0$,
 reflecting the invariance of the result (a pure vector) under

### 2012-07-11 paper/section-metric.tex
Fix some small mistakes

@@
 setting of differential privacy. Their system uses a linear type
-discipline to ensure that all programs are of $c$-sensitive (i.e., the
-distance between the outputs is no more greater than $c$ times the
+discipline to ensure that all programs are $c$-sensitive (i.e., the
+distance between the outputs is no greater than $c$ times the
 distance between the inputs, for some constant $c$). We can express

### 2012-07-11 paper/section-discussion.tex
Final small changes

--- 2012-07-10 paper/section-discussion.tex
@@
 %analogues of Reynolds' 
-an abstraction theorem holds. Geometric theorem
+an abstraction theorem holds. 
+Cardelli and Gardner describe a process calculus that builds in 3D affine geometry~\cite{Cardelli},
+proving that process behaviour is invariant under affine transformations. Distinguishing points from vectors
+provides the appropriate abstraction barrier, and the geometric group is determined by inspecting
+term syntax. It would be interesting to recast their language in terms of our indexed types to
+obtain purely type-based invariance theorems.
+
+Geometric theorem
 proving is another %possible 
@@
 $\stackrel{sem}\sim$ in \defref{def:semantic-equality} is not an
-equivalence relation. If we assume that the interpretations of all the
+equivalence relation. If the interpretations of all 
 primitive types are partial equivalence relations then semantic

### 2012-07-11 paper/section-info-flow.tex
Final small changes

@@
 replacing each type variable $\alpha_l$ with a primitive type
-$\tyPrim{T}{\phi}$ of representations of the truth of a logical
-proposition $\phi$. The relationships between security levels are now
-replaced by logical entailment, so we only have functions of type
-$\tyPrim{T}{\phi} \to \tyPrim{T}{\psi}$ when $\phi$ entails
-$\psi$. %Inspired by the PER interpretation of information flow,
+$\tyPrim{T}{\phi_l}$ of representations of the truth of a logical
+proposition $\phi_l$ that stands for the security level $l$. The
+relationships between security levels are now replaced by logical
+entailment, so we only have functions of type $\tyPrim{T}{\phi_{l_1}} \to
+\tyPrim{T}{\phi_{l_2}}$ when $\phi_{l_1}$ entails
+$\phi_{l_2}$. %Inspired by the PER interpretation of information flow,
 We instantiate our general relational framework to interpret
@@
 then $\mathrm{up}\ M : \tyPrim{T}{\psi}$. Each of the primitive
-operations has a trivial interpretation, due the index-erasure
+operations has a trivial interpretation, due to the index-erasure
 interpretation of $\tyPrim{T}{\phi}$ as a one-element set, giving an

### 2012-07-11 paper/section-metric.tex
Final small changes

@@
 presented examples where the relational interpretations of primitive
-types relates a pair of elements by some transformation if applying
-the transformation to the first element yields the second. Thus, the
-free theorems that we derived directly take the form of ``invariance''
-properties, where some equation holds between two terms. In this
-section, we examine another instantiation of our general framework
-that relates values when they are within a certain distance. The free
-theorems that we obtain inform us of the effect that programs have on
-the distances between values. For example, a program $M$ of type:
+types relate pairs of elements by some transformation if applying the
+transformation to the first element of a pair yields the second. Thus,
+the free theorems that we derived directly take the form of
+``invariance'' properties, where some equation holds between two
+terms. In this section, we examine another instantiation of our
+general framework that relates values when they are within a certain
+distance. The free theorems that we obtain inform us of the effect
+that programs have on the distances between values. For example, a
+program $M$ of type:
 \begin{displaymath}

### 2012-10-09 paper/section-discussion.tex
Split syn into exp and ty

@@
 
-\paragraph{Extension of the non-definability results} Our
-non-definabilty results are currently restricted to the quantifier
-free fragment of our framework, as defined in the statement of
-\thmref{thm:abstraction}. This is a consequence of the
-non-compositionality of the relational environments that we use to
-prove non-definabilty results. Lifting this restriction is a key item
-of future work.
-
 \paragraph{Semantic Equality} In general, the semantic equality

### 2012-10-09 paper/section-general-framework.tex
Split syn into exp and ty

@@
 The definition of relational environments is the most subtle part of
-our framework. The straightforward approach is to assume an
-interpretation of indexing expressions in which sorts are
+our framework.
+
+%The straightforward approach is to assume an
+%interpretation of indexing expressions in which sorts are
 %interpreted as 
-sets and index operations are functions, and then define the
-relational interpretation of a primtive type $\tyPrim{X}{e_1,...,e_n}$
-in terms of the interpretations of $e_1,...,e_n$. This approach
-suffices to prove free theorems and type isomorphisms. However, to
-prove non-definability results, like the one in
-\exref{ex:uninhabited-type}, we need a more refined interpretation
-that accounts for the definable programs permitted by the algebraic
-indexing. Here, we identify the properties we require of relational
-environments for the abstraction theorem to hold. In
-\autoref{sec:constr-rel-env}, we give two generic constructions of
-relational environments.
+%sets and index operations are functions, and then define the
+%relational interpretation of a primtive type $\tyPrim{X}{e_1,...,e_n}$
+%in terms of the interpretations of $e_1,...,e_n$. This approach
+%suffices to prove free theorems and type isomorphisms. However, to
+%prove non-definability results, like the one in
+%\exref{ex:uninhabited-type}, we need a more refined interpretation
+%that accounts for the definable programs permitted by the algebraic
+%indexing. Here, we identify the properties we require of relational
+%environments for the abstraction theorem to hold. In
+%\autoref{sec:constr-rel-env}, we give two generic constructions of
+%relational environments.
 
@@
   \end{enumerate}
-  A family $\relEnv{E}$ is \emph{pre-substitutive} if only the first
-  point holds.
 \end{definition}
@@
     \rsem{B}{\relEnv{E}}{\rho}$;
-  \item If $\Delta' \vdash A \isType$ and either $\relEnv{E}$ is
-    substitutive or $\relEnv{E}$ is pre-substitutive and $A$ is
-    quantifier-free, then for all $\Delta \vdash \sigma \Rightarrow
+  \item If $\Delta' \vdash A \isType$ and $\relEnv{E}$ is
+    substitutive, then for all $\Delta \vdash \sigma \Rightarrow
     \Delta'$ and $\rho \in \relEnv{E}(\Delta)$,
@@
   \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
-
-  If $\relEnv{E}$ is pre-substitutive, then the above restricts to the
-  case when $\Gamma$ only contains types of the form $\forall
-  \vec{i}\mathord:\vec{s}. B$, where $B$ is quantifier-free, $A$ is
-  quantifier-free and $M$ contains no uses of \TirName{UnivAbs},
-  \TirName{ExPack} or \TirName{ExUnpack} and all uses of
-  \TirName{UnivApp} are fully applied.
 \end{theorem}
@@
 
-We give two constructions of families $\relEnv{E}$ of sets of
+We give construction of families $\relEnv{E}$ of sets of
 relational environments that allow us to derive useful consequences of
-\thmref{thm:abstraction}. To satisfy the hypothesis of
+\thmref{thm:abstraction}. 
+To satisfy the hypothesis of
 \thmref{thm:abstraction} we still need to prove, for each
@@
 
-\paragraph{Model-based Relational Environments}
-Our model-based relational environments are defined in terms of
+%\paragraph{Model-based Relational Environments}
+Our relational environments are defined in terms of
 models. To each sort $s \in \SortSet$ we assign a carrier set
@@
 
-\paragraph{Definability Relational Environments}
-We now construct relational environments that can be used for
-non-definability results. To this end, consider the typing
-$t\mathord:\SynTransl{2}; \Gamma \vdash M : \tyPrim{vec}{t + t} \to
-\tyPrim{vec}{t}$, where $\Gamma$ contains the primitive operations on
-vectors indexed by translations from \autoref{sec:affine-vector-ops}.
-In \exref{ex:uninhabited-type} we claimed that this type is
-uninhabited. Intuitively, it is clear why: with the operations on
-vectors in \autoref{sec:affine-vector-ops}, there is no way of
-cancelling the additional $t$ in the argument type to get the result
-type. More generally, the vector operations only allow us to combine
-indices according to the group operations, so if the index of the
-result type of a function lies outside the free abelian group
-generated by the indices of the input types, then we cannot write a
-program of this type.  The key idea is to assign different relational
-interpretations to types with indices not in some generated
-collection of index expressions.
-
-Given a set $S$ of index expressions that are all well-indexed in some
-index context $\Delta$, define $\Gen(S)$ to be the set of terms that
-are generated by the elements of $S$ and the primitive index
-operations. We close under index expression equivalence to get the
-sets $\Gen^\equiv(S) = \{ e \sepbar \exists s.\ \exists e' \in
-\Gen(S).\ \Delta \vdash e \equiv e' : s \}$.
-
-Relational environments are parameterised by sets of terms $S$:
-\begin{displaymath}
-  \rho^S\ \tyPrimNm{X}\ (e_1,...,e_n) = \left\{
-    \begin{array}{ll}
-      \mathrm{Eq}_{\tyPrimSem{\tyPrimNm{X}}} & \textrm{if all}\ e_i \in \Gen^\equiv(S) \\
-      R^\bullet_{\tyPrimNm{X}} & \textrm{otherwise}
-    \end{array}
-  \right.
-\end{displaymath}
-Thus we assign the equality relation to primitive types whose index
-arguments are generated up to equivalence by $S$. For types whose
-indices are not in $\Gen^\equiv(S)$, we require a binary relation
-$R^\bullet_{\tyPrimNm{X}} \in \Rel(\tyPrimSem{\tyPrimNm{X}})$. We set
-$\relEnv{E}_{R^\bullet}(\Delta) = \{ \rho^S \sepbar \forall e \in S.\
-\exists s. \Delta \vdash e : s \}$.
-
-We have not been able to prove that this family of sets of relational
-interpretations is substitutive. Therefore our non-definability
-results are subject to the quantifier-free conditions in
-\thmref{thm:abstraction}.
-
-% Hence we must make do with the
-% following theorem:
-
-% These relational environments are non-compositional: elements of
-% $\relEnv{E}(\Delta,\Delta')$ are not purely constructed from elements
-% of $\relEnv{E}(\Delta)$ and $\relEnv{E}(\Delta')$. As a consequence,
-% these relational environments will only be
-% \emph{pre-substitutive}.
-
-\begin{theorem}
-  The families $\relEnv{E}_{R^\bullet}$ are pre-substitutive.
-\end{theorem}
-
-% FIXME: look forward to the applications
-
 %%% Local Variables:

### 2012-10-16 paper/section-general-framework.tex
New framework setup

@@
+\newcommand{\Gen}{\mathrm{Gen}}
+\newcommand{\Free}{\mathrm{Free}}
+\newcommand{\semSort}[2]{\llbracket #1 \rrbracket^{\mathcal{#2}}}
+\newcommand{\semIndexExp}[2]{\llbracket #1 \rrbracket^{\mathcal{#2}}}
+\newcommand{\semPrim}[3]{\llbracket #1 \rrbacket^{\mathcal{#2}}_{\mathcal{#3}}}
+\newcommand{\opsCtxt}{\Gamma_{\mathit{ops}}}
+\newcommand{\opsEnv}{\eta_{\mathit{ops}}}
 \section{A General Framework}
@@
 %In this section 
-We now present our framework for algebraically
-indexed types. % and representation independence. %We define the syntax
+We now present our framework for algebraically indexed types and its
+relational interpretation.  We define the syntax of algebraically
+indexed types (\autoref{sec:algebraically-indexed-types}) and a syntax
+for terms in a general programming language for algebraically indexed
+types (\autoref{sec:well-typed-programs}). We give an index-erasure
+semantics to types and terms (\autoref{sec:erasure-semantics}), and
+based on this semantics define a notion of contextual equivalence. We
+then introduce a relational semantics for types parameterised by an appropriate
+`model' of the algebraic theory (\autoref{sec:relational-semantics}),
+prove the central Abstraction theorem
+(\autoref{sec:abstraction-theorem}), and use the relational semantics
+to define a notion of semantic equivalence which we prove is a sound
+approximation of contextual equivalence.
+
 %and semantics of algebraically indexed types, then their semantics
@@
 \item A collection $\SortSet$ of index sorts. We use the
-  meta-syntactic variables $s,s_1,s_2,...$ for arbitrary sorts taken
+  meta-syntactic variables $s,s_1,s_2,\ldots$ for arbitrary sorts taken
   from $\SortSet$.
@@
 
-\begin{example}[Two-Dimensional Geometry]
+\begin{example*}[Geometry: syntax]
   \label{ex:two-dim-geo-operations}
@@
            $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
-\end{example}
+\qed
+\end{example*}
 
 We assume a countably infinite collection of index variable names $i,
-i_1, i_2,\ldots$. \emph{Index contexts} $\Delta = i_1 \mathord: s_1,
-..., i_n \mathord: s_n$ are lists of variable/sort pairs such that
+i_1, i_2$, \emph{etc}. \emph{Index contexts} $\Delta = i_1 \mathord: s_1,
+\ldots, i_n \mathord: s_n$ are lists of variable/sort pairs such that
 all the variable names are distinct.
@@
     {\indexOp{f} \in \mathit{IndexOp} \\
-      \indexOpArity(\indexOp{f}) = ([s_1,...,s_n], s) \\
+      \indexOpArity(\indexOp{f}) = ([s_1,\ldots,s_n], s) \\
       \{\Delta \vdash e_j : s_j\}_{1 \leq j \leq n}}
-    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) : s}
+    {\Delta \vdash \indexOp{f}(e_1, \ldots, e_n) : s}
   \end{mathpar}
@@
     {\tyPrimNm{X} \in \mathit{PrimType} \\\\
-      \primTyArity(\tyPrimNm{X}) = [s_1,...,s_n] \\
+      \primTyArity(\tyPrimNm{X}) = [s_1,\ldots,s_n] \\
       \{\Delta \vdash e_j : s_j\}_{1\leq j \leq n}}
-    {\Delta \vdash \tyPrim{X}{e_1,...,e_n} \isType}
+    {\Delta \vdash \tyPrim{X}{e_1,\ldots,e_n} \isType}
 
@@
 forms, from a primitive type $\tyPrimNm{X}$ and appropriately sorted
-index expressions $e_1,...,e_n$, the well-indexed type
-$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} 
+index expressions $e_1,\ldots,e_n$, the well-indexed type
+$\tyPrim{X}{e_1,\ldots,e_n}$. The rule \TirName{TyForall} 
 %permits the formation of 
@@
 %introduction of existentials more, perhaps forward ref to their use
-We say that a type is \emph{quantifier-free} if it does not contain
-ether universal or existential quantification.
 
@@
 Given a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord:
-s_1, ..., i_n \mathord: s_n$, a \emph{simultaneous substitution}
+s_1, \ldots, i_n \mathord: s_n$, a \emph{simultaneous substitution}
 $\Delta \vdash \sigma \Rightarrow \Delta'$ is a sequence of
-expressions $\sigma = (e_1,...,e_n)$ such that $\Delta \vdash e_j :
+expressions $\sigma = (e_1,\ldots,e_n)$ such that $\Delta \vdash e_j :
 s_j$ for all $1 \leq j \leq n$. Given a simultaneous substitution
-$\Delta \vdash \sigma = (e_1,...,e_n) \Rightarrow \Delta'$ and a
+$\Delta \vdash \sigma = (e_1,\ldots,e_n) \Rightarrow \Delta'$ and a
 variable $i_j \mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for
@@
 For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
-\Delta'$, where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
+\Delta'$, where $\Delta' = i_1\mathord:s_1,\ldots,i_n\mathord:s_n$, and a
 variable/sort pair $i\mathord:s$ such that $i$ does not appear in
@@
 simultaneous substitution $\Delta,i\mathord:s \vdash
-\sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
+\sigma_{i\mathord:s} = (\sigma(i_1), \ldots, \sigma(i_n), i) \Rightarrow
 \Delta',i\mathord:s$. %Note that we have implicity used the fact that
@@
 \isDefinedAs \sigma(i)$, and on operation symbols as
-$\sigma^*(\indexOp{f}(e_1,...,e_n)) \isDefinedAs
-\indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)$.  %Similarly, 
+$\sigma^*(\indexOp{f}(e_1,\ldots,e_n)) \isDefinedAs
+\indexOp{f}(\sigma^*e_1, \ldots, \sigma^*e_n)$.  %Similarly, 
 Given
@@
   \begin{array}{c}
-    \sigma^*(\tyPrim{X}{e_1,...,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,...,\sigma^*e_n}
+    \sigma^*(\tyPrim{X}{e_1,\ldots,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,\ldots,\sigma^*e_n}
     \\
@@
 %$\Delta$: 
-$\id_\Delta = (i_1,...,i_n)$ where $\Delta =
-i_1\mathord:s_1,...,i_n\mathord:s_n$. The \emph{composition} of 
+$\id_\Delta = (i_1,\ldots,i_n)$ where $\Delta =
+i_1\mathord:s_1,\ldots,i_n\mathord:s_n$. The \emph{composition} of 
 %two simultaneous substitutions 
@@
 and $\Delta' \vdash \sigma' \Rightarrow \Delta''$, where $\sigma' =
-(e'_1,...,e'_n)$, is defined as $\Delta \vdash \sigma' \circ \sigma
-\isDefinedAs (\sigma^*e'_1, ..., \sigma^*e'_n) \Rightarrow \Delta''$.
+(e'_1,\ldots,e'_n)$, is defined as $\Delta \vdash \sigma' \circ \sigma
+\isDefinedAs (\sigma^*e'_1, \ldots, \sigma^*e'_n) \Rightarrow \Delta''$.
 %
-Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
+Given a context $\Delta = i_1\mathord:s_1,\ldots,i_n\mathord:s_n$, and a
 variable/sort pair $i\mathord:s$ such that $i$ does not appear in
@@
 $\Delta, i\mathord:s \vdash \pi_{i\mathord:s} \Rightarrow \Delta$ as
-$\pi_{i\mathord:s} = (i_1,...,i_n)$. %The subscript on
+$\pi_{i\mathord:s} = (i_1,\ldots,i_n)$. %The subscript on
 %$\pi_{i\mathord:s}$ is the variable/sort pair that is being discarded.
@@
 
-\begin{example}[Two-Dimensional Geometry]
+\begin{example*}[Geometry: axioms]
   \label{ex:two-dim-geo-axioms}
@@
   \stackrel{ax}\equiv s \cdot s$.
-\end{example}
+\qed
+\end{example*}
 
@@
   {\{ \Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j \leq n}}
-  {\Delta \vdash \tyPrim{X}{e_1,...,e_n} \equiv \tyPrim{X}{e'_1,...,e'_n} \isType}
+  {\Delta \vdash \tyPrim{X}{e_1,\ldots,e_n} \equiv \tyPrim{X}{e'_1,\ldots,e'_n} \isType}
 \end{displaymath}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Semantics of Algebraically-Indexed Types}
-\label{sec:semantics-algebraically-indexed-types}
-
-%\fixme{Make sure that the semantics is motivated}
-%Having defined the language of algebraically-indexed types, we now
-%turn to their denotational interpretation. 
-%We now give the denotational interpretation of algebraically indexed
-%types.  In \autoref{sec:index-erasure-semantics} we define an
-%\emph{index-erasure} interpretation of types that interprets every
-%well-indexed type as a set, ignoring the indexing
-%expressions. %Building on the index-erasure semantics, we 
-%We then define in
-%\autoref{sec:relational-semantics} the relational
-%interpretation of types.
-
-\autoref{sec:index-erasure-semantics} gives an \emph{index-erasure}
-interpretation of types, interpreting every well-indexed type as a
-set, ignoring the indexing expressions.
-\autoref{sec:relational-semantics} gives the relational interpretation
-of types.
-
-% For both of our semantics of types, we state and prove two properties:
-% types that are syntactically equal have equal denotations, and that
-% substitution of index terms is interpreted via composition. These
-% properties ensure that our semantics of types is well-behaved.
-
-\subsubsection{The Index-Erasure Interpretation of Types}
-\label{sec:index-erasure-semantics}
-
-The defining feature of the index erasure interpretation is that 
-semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
-determined solely by the primitive type $\tyPrimNm{X}$ and not by the
-index expressions $e_1,...,e_n$. %Accordingly, 
-We thus assume each
-primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
-$\tyPrimSem{\tyPrimNm{X}}$ and %We 
-extend this assignment to %every
-well-indexed types by: % induction on the type structure:
-\begin{displaymath}
-  \begin{array}{@{}c@{\hspace{0em}}c@{}}
-    \begin{array}{@{}l}
-      \tySem{\tyUnit} \isDefinedAs \{*\} \\
-      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
-      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
-    \end{array}
-    &
-    \begin{array}{l}
-      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
-      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
-      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A} \\
-      \tySem{\exists i\mathord:s. A} \isDefinedAs \tySem{A}
-    \end{array}
-  \end{array}
-\end{displaymath}
-% The interpretation of the unit type is a chosen one element set, and
-% the interpretations of the function, tuple and sum types are simply
-% the corresponding constructions on sets. 
-Note that 
-the interpretations of the
-universal and existential quantifiers 
-do indeed
-ignore the indexing. %: the
-%interpretation of the type $\forall i\mathord:s.A$ is exactly the
-%interpretation of the type $A$, and likewise for $\exists
-%i\mathord:s.A$.
-We then have:
-
-% The index-erasure interpretation completely ignores index expressions
-% and type equality is defined as an extension of index
-% equality. Therefore, it is straightforward to prove that equal types
-% have equal denotations when interpreted in the index-erasure
-% semantics, and that substitution of index terms has no effect on the
-% index-erasure interpretation of types:
-\begin{lemma}\label{lem:tyeqsubst-erasure}
-  \begin{enumerate}
-  \item If $\Delta \vdash A \equiv B \isType$ then $\tySem{A} =
-    \tySem{B}$; and
-  \item If $\Delta' \vdash A \isType$ and $\Delta \vdash \sigma
-    \Rightarrow \Delta'$, then $\tySem{\sigma^*A} = \tySem{A}$.
-  \end{enumerate}
-\end{lemma}
-
-\begin{example}[Two-Dimensional Geometry]
-  \exref{ex:two-dim-geo-axioms}
-%  The two-dimensional geometry instantiation of the general framework
-  uses the assignment $\tyPrimSem{\tyPrimNm{vec}} = \mathbb{R}^2$ and
-  $\tyPrimSem{\tyPrimNm{real}} = \mathbb{R}$.
-\end{example}
-
-\subsubsection{The Relational Interpretation of Types}
-\label{sec:relational-semantics}
-
-%We now define 
-The relational semantics of the well-indexed type
-$\Delta \vdash A \isType$ is a binary relation on the index-erasure
-interpretation of $A$. We write $\Rel(X)$ for the set
-of binary relations $R \subseteq X \times X$ on the set $X$.
-
-For the unit, tuple, sum and function types we define the
-relational interpretation as a standard logical relation. The
-relational interpretations of primitive types with index arguments and
-the universally quantified types require an interpretation of index
-contexts. %We assign interpretations to index contexts 
-These are given in terms of sets
-of \emph{relational environments}.
-
-The definition of relational environments is the most subtle part of
-our framework.
-
-%The straightforward approach is to assume an
-%interpretation of indexing expressions in which sorts are
-%interpreted as 
-%sets and index operations are functions, and then define the
-%relational interpretation of a primtive type $\tyPrim{X}{e_1,...,e_n}$
-%in terms of the interpretations of $e_1,...,e_n$. This approach
-%suffices to prove free theorems and type isomorphisms. However, to
-%prove non-definability results, like the one in
-%\exref{ex:uninhabited-type}, we need a more refined interpretation
-%that accounts for the definable programs permitted by the algebraic
-%indexing. Here, we identify the properties we require of relational
-%environments for the abstraction theorem to hold. In
-%\autoref{sec:constr-rel-env}, we give two generic constructions of
-%relational environments.
-
-A relational environment for a context $\Delta$ is a %(dependently typed)
-function $\rho$ that, for each primitive type $\tyPrimNm{X}$ and
-instantiation of its arguments $\sigma$, assigns a binary relation
-$\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure interpretation of
-$\tyPrimNm{X}$:
-\begin{displaymath}
-  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
-\end{displaymath}
-%Every relation environment must respect index expression equality: for
-%any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
-%substitutions $\Delta \vdash \sigma \Rightarrow
-%\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
-%\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
-%\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
-%\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
-Relational environments must respect index expression equality: for
-all $\tyPrimNm{X} \in \PrimTypeSet$,
-$\Delta \vdash \sigma \Rightarrow
-\primTyArity(\tyPrimNm{X})$, and $\Delta \vdash \sigma' \Rightarrow
-\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
-\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
-\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
-%
-We write $\mathrm{RelEnv}(\Delta)$ for the set of all relational
-environments for a context $\Delta$. Given %a relational environment
-$\rho \in \mathrm{RelEnv}(\Delta)$ and %a simultaneous substitution
-$\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
-relational environment (with an abuse of notation) $\rho \circ \sigma
-\in \mathrm{RelEnv}(\Delta')$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
-\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$.
-
-We interpret index contexts $\Delta$ as certain
-%sets of relation environments, i.e.,~
-subsets of $\mathrm{RelEnv}(\Delta)$. Depending on the primitive
-operations that we assume for our system, we may have different sets
-of relational environments that enforce different invariants. 
-% We
-% require that the sets of relational environments are \emph{valid} in
-% the following sense:
-
-\begin{definition}
-  \label{defn:valid-rel-env-family}
-  An index context-indexed family of sets of relational environments
-  $\{ \relEnv{E}(\Delta) \subseteq \mathrm{RelEnv}(\Delta) \}_\Delta$
-  is \emph{substitutive} if:
-  \begin{enumerate}
-  \item it is closed under composition with simultaneous substitution:
-    if $\rho \in \relEnv{E}(\Delta)$ and $\Delta \vdash \sigma
-    \Rightarrow \Delta'$, then $\rho \circ \sigma \in
-    \relEnv{E}(\Delta')$; and
-%  \item If we have a pair of relation environments $\rho_1 \in
-%    \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
-%    \relEnv{E}(\Delta)$, along with a simultaneous substitution
-%    $\Delta' \vdash \sigma \Rightarrow \Delta$, such that $\rho_1
-%    \circ \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists
-%    a $\rho \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho
-%    \circ \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ
-%    \pi_{i\mathord:s'} = \rho_2$.
-  \item if $\rho_1 \in \relEnv{E}(\Delta', i\mathord:s)$ and $\rho_2
-    \in \relEnv{E}(\Delta)$, and $\Delta \vdash \sigma \Rightarrow
-    \Delta'$ is such that $\rho_1 \circ \pi_{i\mathord:s} = \rho_2
-    \circ \sigma$, then there exists a $\rho \in \relEnv{E}(\Delta,
-    i\mathord:s)$ such that $\rho \circ \sigma_{i\mathord:s} =
-    \rho_1$ and $\rho \circ \pi_{i\mathord:s} = \rho_2$.
-    % such that the outer edges of the
-    % following diagram commute, for all primitive types $\tyPrimNm{X} \in
-    % \mathit{PrimType}$:
-    % % FIXME: check that these compositions are the right way round
-    % % FIXME: re-do this as a set of equations
-    % % FIXME: give a better English description of why this is needed
-    % \begin{displaymath}
-    %   \xymatrix{
-    %   {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
-    %   &
-    %   {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
-    %   \\
-    %   {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
-    %   &
-    %   {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
-    %   \\
-    %   &
-    %   &
-    %   {\Rel(\tyPrimSem{\tyPrimNm{X}})}
-    % }
-    % \end{displaymath}
-    % Then there exists a relation environment $\rho \in
-    % \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
-    % two triangles in the bottom right of the diagram commute.
-    % FIXME: mention that we already know that the square commutes
-  \end{enumerate}
-\end{definition}
-%
-% For the general
-% framework, we assume that, for each index context $\Delta$, we are
-% given a set of relation environments $\relEnv{E}(\Delta)$. Note that
-% this assignment of sets of relation environments to index contexts is
-% not assumed to be compositional: it is not necessarily the case that
-% the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
-% defined in terms of $\relEnv{E}(\Delta_1)$ and
-% $\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
-% properties:
-%
-\noindent
-These conditions ensure that the relational interpretation of types we
-define below behaves correctly with respect to simultaneous
-substitution of index expressions (see
-\lemref{lem:tyeqsubst-relational}, below).
-
-% The second condition may
-% seem mysterious at first, but it is essential for proving that the
-% relational interpretation of universal quantification types behaves
-% correctly with respect to application of simultaneous substitution.
-%
-Given a relational environment $\rho \in \relEnv{E}(\Delta)$, %we define
-the set of extensions %$\extends{\rho}{i\mathord:s}$ 
-of $\rho$ by %an
-%additional index variable 
-$i\mathord:s$ %to be 
-is $\extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in
-\relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi_{i\mathord:s} =
-\rho \}$. The set of extensions of a relational environment will be
-used in interpreting the universal and existential type formers.  We
-assign a relational interpretation to all well-indexed types $\Delta
-\vdash A \isType$ by induction on their derivations, parameterised by
-%relational environments 
-$\rho \in \relEnv{E}(\Delta)$:
-\begin{eqnarray*}
-  \rsem{\tyUnit}{\relEnv{E}}\rho & \isDefinedAs & \{(*,*)\} \\
-  \rsem{\tyPrim{X}{e_1,...,e_n}}{\relEnv{E}}\rho & \isDefinedAs & \rho\ {\tyPrimNm{X}}\ (e_1,...,e_n) \\
-  \rsem{A \tyArr B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{A \tyProduct B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{A + B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
-  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:s} \} \\
-  \rsem{\exists i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcup\{ \rsem{A}{\relEnv{E}}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:s} \}
-\end{eqnarray*}
-In this definition we have made use of the following %three
-constructions on binary relations: if $R \in \Rel(X)$ and $S \in
-\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is %defined as 
-$\{(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
-\}$, and %. With the same assumptions on $R$ and $S$, the relation 
-$R
-\relTimes S \in \Rel(X \times Y)$ is %defined as 
-$\{((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
-\}$, and %. Finally, the relation 
-$R \relSum S \in \Rel(X + Y)$ is %defined as
-$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
-\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.
-
-% The following lemma states that the relation interpretation of types
-% that we have defined in this section behaves well: the first part of
-% the lemma states that two types that are judgmentally equal are given
-% equal relational interpretations, and the second part states that
-% substitution of index expressions in types can be interpreted by the
-% composition of relational environments with simultaneous
-% substitutions.
-\begin{lemma}\label{lem:tyeqsubst-relational}
-\par
-  \begin{enumerate}
-  \item If $\Delta \vdash A \equiv B \isType$, then for all $\rho \in
-    \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}{\rho} =
-    \rsem{B}{\relEnv{E}}{\rho}$;
-  \item If $\Delta' \vdash A \isType$ and $\relEnv{E}$ is
-    substitutive, then for all $\Delta \vdash \sigma \Rightarrow
-    \Delta'$ and $\rho \in \relEnv{E}(\Delta)$,
-    $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
-    \circ \sigma)$.
-  \end{enumerate}
-\end{lemma}
-\noindent
-Note that the equations in both parts of
-\lemref{lem:tyeqsubst-relational} are well-typed by virtue of the
-corresponding parts of \lemref{lem:tyeqsubst-erasure}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Well-typed Programs and the Abstraction Theorem}
+\subsection{Well-typed Programs}
 \label{sec:well-typed-programs}
@@
 We now present the rules for well-typed programs over the 
-%collection of 
-types %we defined 
-in \autoref{sec:algebraically-indexed-types}. Each
-well-typed program is assigned an index-erasure semantics, building on
-the index-erasure semantics of types %we defined 
-in \autoref{sec:index-erasure-semantics}. Our main result
-(\thmref{thm:abstraction}) is that the index-erasure semantics of
-every well-typed program is related to itself in the relational
-interpretation of its type: this is the abstraction theorem for every
-instantiation of our general framework.
+collection of 
+types we defined 
+in \autoref{sec:algebraically-indexed-types}.
+
+%Each
+%well-typed program is assigned an index-erasure semantics, building on
+%the index-erasure semantics of types %we defined 
+%in \autoref{sec:index-erasure-semantics}. Our main result
+%(\thmref{thm:abstraction}) is that the index-erasure semantics of
+%every well-typed program is related to itself in the relational
+%interpretation of its type: this is the abstraction theorem for every
+%instantiation of our general framework.
 
@@
 
-\paragraph{Index-Erasure Interpretation of Programs}
+For any particular theory we assume that there is a closed
+typing context $\opsCtxt$ that describes the types of the primitive operations.
+\begin{example*}[Geometry: operations]
+We assume $\opsCtxt$ collects together the types of primitive operations as described in \autoref{sec:motivation-generalising}.
+\end{example*}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Index-Erasure Semantics}
+\label{sec:erasure-semantics}
+%\fixme{Make sure that the semantics is motivated}
+Having defined the language of algebraically-indexed types and a
+programs, we turn to their denotational interpretation.  We first define an
+\emph{index-erasure} interpretation of types and terms that interprets
+every well-indexed type as a set, ignoring the indexing
+expressions. %Building on the index-erasure semantics, we
+%We then define in
+%\autoref{sec:relational-semantics} the relational
+%interpretation of types.
+
+% For both of our semantics of types, we state and prove two properties:
+% types that are syntactically equal have equal denotations, and that
+% substitution of index terms is interpreted via composition. These
+% properties ensure that our semantics of types is well-behaved.
+
+\paragraph{Interpretation of Types.}
+%\label{sec:index-erasure-semantics}
+
+The defining feature of the index erasure interpretation is that 
+semantics of a well-indexed type $\tyPrim{X}{e_1,\ldots,e_n}$ is
+determined solely by the primitive type $\tyPrimNm{X}$ and not by the
+index expressions $e_1,\ldots,e_n$. %Accordingly, 
+We thus assume each
+primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
+$\tyPrimSem{\tyPrimNm{X}}$ and %We 
+extend this assignment to %every
+well-indexed types by induction on the type structure:
+\begin{displaymath}
+  \begin{array}{@{}c@{\hspace{0em}}c@{}}
+    \begin{array}{@{}l}
+      \tySem{\tyUnit} \isDefinedAs \{*\} \\
+      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
+      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
+    \end{array}
+    &
+    \begin{array}{l}
+      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
+      \tySem{\tyPrim{X}{e_1,\ldots,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
+      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A} \\
+      \tySem{\exists i\mathord:s. A} \isDefinedAs \tySem{A}
+    \end{array}
+  \end{array}
+\end{displaymath}
+Note that the interpretations of the universal and existential
+quantifiers do indeed ignore the indexing. (We will overload the
+notation $\left|\cdot\right|$ for all index-erasure interpretations, 
+and reserve notation $\llbracket\cdot\rrbracket$ for the
+index-observing relational semantics defined later.)
+% The interpretation of the unit type is a chosen one element set, and
+% the interpretations of the function, tuple and sum types are simply
+% the corresponding constructions on sets. 
+%: the
+%interpretation of the type $\forall i\mathord:s.A$ is exactly the
+%interpretation of the type $A$, and likewise for $\exists
+%i\mathord:s.A$.
+We then have:
+
+% The index-erasure interpretation completely ignores index expressions
+% and type equality is defined as an extension of index
+% equality. Therefore, it is straightforward to prove that equal types
+% have equal denotations when interpreted in the index-erasure
+% semantics, and that substitution of index terms has no effect on the
+% index-erasure interpretation of types:
+\begin{lemma}\label{lem:tyeqsubst-erasure}
+  \begin{enumerate}
+  \item If $\Delta \vdash A \equiv B \isType$ then $\tySem{A} =
+    \tySem{B}$; and
+  \item If $\Delta' \vdash A \isType$ and $\Delta \vdash \sigma
+    \Rightarrow \Delta'$, then $\tySem{\sigma^*A} = \tySem{A}$.
+  \end{enumerate}
+\end{lemma}
+
+\paragraph{Interpretation of Programs.}
 
@@
 \lemref{lem:tyeqsubst-erasure}, we do this %we define this function is defined
-directly on the syntax of well-typed programs, rather than on typing
+directly on the syntax of well-typed terms, rather than on typing
 derivations. The definition of $\tmSem{M}$ is completely standard,
@@
 
+For any particular theory we assume that there is an
+interpretation of the primitive operations $\opsEnv \in \ctxtSem{\opsCtxt}$.
+
+\begin{example*}[Geometry: interpretation]
+  The two-dimensional geometry instantiation of the general framework
+  uses the assignment $\tyPrimSem{\tyPrimNm{vec}} = \mathbb{R}^2$ and
+  $\tyPrimSem{\tyPrimNm{real}} = \mathbb{R}$.
+  We assume that $\opsEnv$ interprets the operations described in \autoref{sec:motivation-generalising}
+  in the standard way, so for example, for  blah blah
+\qed
+\end{example*}
+
+\paragraph{Contextual Equivalence.}
 We use our index-erasure semantics to define when a pair of programs
@@
 
-\begin{definition}\label{defn:ctxt-equiv}
-  Let $\Gamma_{\mathit{ops}}$ be a typing context with no free index
-  variables that describes the types of the primitive operations, and
-  let $\eta_{\mathit{ops}} \in \ctxtSem{\Gamma_{\mathit{ops}}}$ be an interpretation
-  of these primitive operations.
-
+\newcommand{\ctxteq}[5]{{#1};{#2}\vdash{#3}\approx{#4}:{#5}}
+\begin{definition}[Contextual Equivalence]\label{defn:ctxt-equiv}
 Two programs $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 : A$
 and $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_2 : A$ are
-  \emph{contextually equivalent} ($\Delta; \Gamma_{\mathit{ops}},
-  \Gamma \vdash M_1 \stackrel{ctx}\approx M_2 : A$) if for all
-  programs $;\Gamma_{\mathit{ops}} \vdash T : (\forall \Delta. \Gamma
-  \to A) \to \tyPrimNm{bool}$, it is the case that $\tmSem{T\ (\Lambda
-    \Delta. \lambda \Gamma. M_1)}\eta_{\mathit{ops}} = \tmSem{T\
+\emph{contextually equivalent}, written $\ctxteq\Delta\Gamma{M_1}{M_2}A$, if 
+  for all
+  programs $\cdot;\Gamma_{\mathit{ops}} \vdash P : (\forall \Delta. \Gamma
+  \to A) \to \tyPrimNm{bool}$, it is the case that $\tmSem{P\ (\Lambda
+    \Delta. \lambda \Gamma. M_1)}\eta_{\mathit{ops}} = \tmSem{P\
     (\Lambda \Delta. \lambda \Gamma. M_2)}\eta_{\mathit{ops}}$.
@@
 
-\paragraph{The Abstraction Theorem}
+\subsection{The Relational Interpretation of Types}
+\label{sec:relational-semantics}
+
+%We now define 
+The relational semantics of the well-indexed type
+$\Delta \vdash A \isType$ is a binary relation on the index-erasure
+interpretation of $A$. We write $\Rel(X)$ for the set
+of binary relations $R \subseteq X \times X$ on the set $X$.
+
+For the unit, tuple, sum and function types we define the
+relational interpretation as a standard logical relation. The
+relational interpretations of primitive types with index arguments and
+the universally quantified types require an interpretation of index
+contexts.
+
+A \emph{relational environment} $\rho$ assigns to each index variable $i:s$
+in the context $\Delta$ a value drawn from an interpretation of
+the sort $s$ that soundly models the equational theory associated with $s$. We call
+such an interpretation a \emph{model}: it assigns to each index operation in the equational 
+theory a corresponding operation in the interpretation, so that index expressions 
+can be interpreted by recursion on their structure.
+
+For example, the sort $\SynTransl{2}$ of translations can be modelled by
+any abelian group. An obvious candidate here is the additive group over 
+$\mathbb R^2$, which we will use to obtain invariance under translation;
+but we could use the additive group $\mathbb Q$, or even a finite group such as the 
+two-element group $\mathbb Z_2$.
+
+\paragraph{Models.}
+Formally, a \emph{model} $\mathcal{M}$ assigns to each sort $s \in \SortSet$ a
+carrier set $\semSort s M$, and assigns to each index operation $\texttt{f}
+\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) =
+([s_1,\ldots,s_n],s)$, a function $\semIndexExp{\texttt{f}}M :
+\semSort{s_1}M\times\cdots\times\semSort{s_n}M \to \semSort{s}M$.
+An index context $\Delta = (i_1\mathord:s_1,\ldots,i_n\mathord:s_n)$ is
+interpreted as cartesian product \emph{i.e.} $\semSort{\Delta}M \isDefinedAs \semSort{s_1}M
+\times\cdots \times \semSort{s_n}M$. 
+For each well-sorted index expression $\Delta \vdash
+e : s$, we assign a function $\semIndexExp{e}M : \semSort{\Delta}M \to
+\semSort{s}M$ by recursion on the structure of~$e$:
+\[
+\begin{array}{rcl}
+\semIndexExp{i}M \rho & \isDefinedAs & \rho(i)\\
+\semIndexExp{\indexOp{f}(e_1, \ldots, e_n)}M \rho & \isDefinedAs & \semIndexExp{\indexOp{f}}M(
+\semIndexExp{e_1}M\rho,\ldots,\semIndexExp{e_n}M\rho)
+\end{array}
+\]
+
+Finally, a model $\mathcal{M}$
+must be sound, that is, for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
+\IndexAxiomSet$, we have $\semIndexExp{e}M = \semIndexExp{e'}M$.
+
+\begin{example*}[Geometry: affine model]
+\todo{Give example}
+\end{example*}
+
+Given $\rho \in \semSort{\Delta}M$ and a simultaneous substitution
+$\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
+relational environment (with an abuse of notation) $\rho \circ \sigma
+\in \semSort{\Delta'}M$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
+\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$.
+\todo{This is wrong: fix!}
+
+%A
+%\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
+%\semIndexExp{\cdot}_{\mathcal{M}})$ satisfying all the axioms in
+%$\IndexAxiomSet$.
+
+\paragraph{Relational interpretation of primitive types.}
+Having fixed a model $\mathcal{M}$ we next choose a relational interpretation $\mathcal{R}$ of
+primitive types: for each each primitive type
+$\tyPrimNm{X}$, a relational interpretation parameterised by elements
+from the model: $\mathcal{R}_{\tyPrimNm{X}} :
+\semSort{\primTyArity(\tyPrimNm{X})}M \to \Rel(\tyPrimSem{\tyPrimNm{X}})$.
+
+\begin{example*}[Geometry: change of basis interpretation]
+\todo{Give example}
+\end{example*}
+
+\paragraph{Relational interpretation of types.}
+We assign a relational interpretation to all well-indexed types $\Delta
+\vdash A \isType$ by induction on their derivations, parameterised by
+relational environments 
+$\rho \in \semSort{\Delta}M$:
+\begin{eqnarray*}
+  \rsem{\Delta \vdash A \isType}M R & : & \semSort\Delta M \to \Rel(\tyPrimSem{A})\\
+  \rsem{\tyUnit}M R \rho & \isDefinedAs & \{(*,*)\} \\
+  \rsem{\tyPrim{X}{e_1,\ldots,e_n}} M R\rho & \isDefinedAs & 
+  \mathcal{R}_{\tyPrimNm{X}} (\semIndexExp{e_1}M\delta, \ldots,
+  \semIndexExp{e_n}M\delta)\\
+  \rsem{A \tyArr B}M R \rho & \isDefinedAs & \rsem{A}M R \rho \relArrow \rsem{B} M R \rho \\
+  \rsem{A \tyProduct B} M R \rho & \isDefinedAs & \rsem{A} M R \rho \relTimes \rsem{B} M R \rho \\
+  \rsem{A + B}M R \rho & \isDefinedAs & \rsem{A} M R \rho \relSum \rsem{B} M R \rho \\
+  \rsem{\forall i\mathord:s.A} M R \rho & \isDefinedAs & 
+  \bigcap\{ \rsem{A} M R {(m,\rho)} \sepbar m \in \semSort{s}M \} \\
+  \rsem{\exists i\mathord:s.A} M R \rho & \isDefinedAs & 
+  \bigcup\{ \rsem{A} M R {(m,\rho)} \sepbar m \in \semSort{s}M \}
+\end{eqnarray*}
+In this definition we have used the following standard %three
+constructions on binary relations: if $R \in \Rel(X)$ and $S \in
+\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is %defined as 
+$\{(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
+\}$, and %. With the same assumptions on $R$ and $S$, the relation 
+$R
+\relTimes S \in \Rel(X \times Y)$ is %defined as 
+$\{((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
+\}$, and %. Finally, the relation 
+$R \relSum S \in \Rel(X + Y)$ is %defined as
+$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
+\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.
+
+The following lemma states that the relation interpretation of types
+that we have defined in this section behaves well: the first part of
+the lemma states that two types that are judgmentally equal are given
+equal relational interpretations, and the second part states that
+substitution of index expressions in types can be interpreted by the
+composition of relational environments with simultaneous
+substitutions.
+\begin{lemma}\label{lem:tyeqsubst-relational}
+\par
+  \begin{enumerate}
+  \item If $\Delta \vdash A \equiv B \isType$, then $\rsem{A} M R =  \rsem{B} M R$;
+  \item If $\Delta' \vdash A \isType$ then for all $\Delta \vdash \sigma \Rightarrow
+    \Delta'$ and $\rho \in \semSort{\Delta}M$,
+    $\rsem{\sigma^*A} M R \rho = \rsem{A} M R(\rho \circ \sigma)$.
+  \end{enumerate}
+\end{lemma}
+\noindent
+Note that the equations in both parts of
+\lemref{lem:tyeqsubst-relational} are well-typed by virtue of the
+corresponding parts of \lemref{lem:tyeqsubst-erasure}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{The Abstraction Theorem and Semantic Equivalence}
+\label{sec:abstraction-theorem}
+
+Our main result
+(\thmref{thm:abstraction}) is that the index-erasure semantics of
+every well-typed program is related to itself in the relational
+interpretation of its type: this is the abstraction theorem for every
+instantiation of our general framework.
+
+\paragraph{The Abstraction Theorem.}
 %\label{sec:abstraction-theorem}
@@
   \begin{array}{@{}l@{\hspace{0.1em}\isDefinedAs\hspace{0.1em}}l@{\hspace{1.5em}}l@{\hspace{0.1em}\isDefinedAs\hspace{0.1em}}l}
-    \rsem{\epsilon}{\relEnv{E}}{\rho} & \{(*,*)\} &
-    \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
+    \rsem{\epsilon} M R{\rho} & \{(*,*)\} &
+    \rsem{\Gamma, x : A} M R{\rho} & \rsem{\Gamma} M R\rho \relTimes \rsem{A} M R\rho
   \end{array}
@@
 \begin{theorem}[Abstraction]\label{thm:abstraction}
-  Assume $\relEnv{E}$ is substitutive. If $\Delta; \Gamma \vdash M :
-  A$, then for all $\rho \in \relEnv{E}(\Delta)$ and $\eta_1, \eta_2
+  If $\Delta; \Gamma \vdash M :
+  A$, then for all $\rho \in \semSort{\Delta}M$ and $\eta_1, \eta_2
   \in \ctxtSem{\Gamma}$ such that $(\eta_1, \eta_2) \in
-  \rsem{\Gamma}{\relEnv{E}}\rho$, we have $(\tmSem{M}\eta_1,
-  \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
+  \rsem{\Gamma} M R\rho$, we have $(\tmSem{M}\eta_1,
+  \tmSem{M}\eta_2) \in \rsem{A} M R\rho$.
 \end{theorem}
-
-We refer to programs satisfying the conditions in the second part of
-\thmref{thm:abstraction} as being in the \emph{quantifier-free}
-fragment of our framework. These conditions are very similar to those
-used in the fragment of System F covered by Hindley-Milner type
-inference.
+\begin{proof}
+By induction on the typing derivation, making use of
+Lemma~\ref{lem:tyeqsubst-relational} (part 1) for rule \TirName{TyEq},
+and Lemma~\ref{lem:tyeqsubst-relational} (part 2) for the quantifier introduction
+and elimination rules. The details can be found in the Coq development.
+\end{proof}
 
@@
 equivalence is a sound approximation of contextual equivalence.
-\begin{definition}\label{def:semantic-equality}
+
   Let $\Gamma_{\mathit{ops}}$ and $\eta_{\mathit{ops}}$ be a context
   of primitive operations and its interpretation as in
-  \defref{defn:ctxt-equiv}. Let $\relEnv{E}_{\mathit{ops}}$ be a
-  substitutive family of sets of relational environments such that for
-  all $\Delta$ and $\rho \in \relEnv{E}_{\mathit{ops}}(\Delta)$,
+  \defref{defn:ctxt-equiv}. Let $\mathcal{M}$ be a model and $\mathcal{R}$  
+  a relational interpretation such that for
+  all $\Delta$ and $\rho \in \semSort{\Delta}M$,
   $(\eta_{\mathit{ops}}, \eta_{\mathit{ops}}) \in
-  \rsem{\Gamma_{\mathit{ops}}}{\relEnv{E}_{\mathit{ops}}}\rho$.
+  \rsem{\Gamma_{\mathit{ops}}} M R \rho$.
+\newcommand{\semeq}[7]{{#3};\Gamma_{\mathit{ops}},{#4}\models^{\mathcal{#1}}_{\mathcal{#2}}{#5}\sim{#6}:{#7}}
 
+\begin{definition}[Semantic equivalence]\label{def:semantic-equality}
   Two programs $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 : A$
   and $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_2 : A$ are
-  \emph{semantically equal} ($\Delta; \Gamma_{\mathit{ops}}, \Gamma
-  \vdash M_1 \stackrel{sem}\sim M_2 : A$) if for all $\rho \in
-  \relEnv{E}_{\mathit{ops}}$, and all $(\eta_1,\eta_2) \in
-  \rsem{\Gamma}{\relEnv{E}_{\mathit{ops}}}\rho$, we have
+  \emph{semantically equal} with respect to $\mathcal{M}$ and $\mathcal{R}$,
+  written $\semeq M R \Delta\Gamma{M_1}{M_2} A$, if for all $\rho \in
+  \semSort{\Delta}M$, and all $(\eta_1,\eta_2) \in
+  \rsem{\Gamma} M R \rho$, we have
   $(\tmSem{M_1}(\eta_{\mathit{ops}},
   \eta_1),\tmSem{M_2}(\eta_{\mathit{ops}}, \eta_2)) \in
-  \rsem{A}{\relEnv{E}_{\mathit{ops}}}\rho$.
+  \rsem{A} M R\rho$.
 \end{definition}
@@
 \begin{theorem}[Soundness]\label{thm:soundness}
-  $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 \stackrel{sem}\sim M_2 : A$ implies
-  $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 \stackrel{ctx}\approx M_2 : A$
+For any model $\mathcal{M}$ and interpretation $\mathcal{R}$, 
+if $\semeq M R \Delta\Gamma {M_1}{M_2} A$ then
+$\ctxteq \Delta \Gamma {M_1}{M_2}A$.
 \end{theorem}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Construction of Relational Environments}
-\label{sec:constr-rel-env}
-
-\newcommand{\Gen}{\mathrm{Gen}}
-\newcommand{\Free}{\mathrm{Free}}
-\newcommand{\semSort}[1]{\llbracket #1 \rrbracket^{\mathcal{S}}}
-\newcommand{\semIndexExp}[1]{\llbracket #1 \rrbracket^{\mathcal{I}}}
-
-We give construction of families $\relEnv{E}$ of sets of
-relational environments that allow us to derive useful consequences of
-\thmref{thm:abstraction}. 
-To satisfy the hypothesis of
-\thmref{thm:abstraction} we still need to prove, for each
-instantiation of our general framework, that the program-level
-primitive operations preserve all the relations in our
-constructions. This is the content of Lemmas
-\ref{lem:geom-environments-1}, \ref{lem:geom-environments-2},
-\ref{lem:monoid-ops-related}, \ref{lem:environments-information-flow}
-and \ref{lem:metric-environments}.
-
-%\paragraph{Model-based Relational Environments}
-Our relational environments are defined in terms of
-models. To each sort $s \in \SortSet$ we assign a carrier set
-$\semSort{s}$. We extend this assignment to index contexts and other
-sequences of sorts (e.g., type arities) using the cartesian product of
-sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} = \semSort{s_1}
-\times ... \times \semSort{s_n}$. For each index operation $\texttt{f}
-\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) = ([s_1,...,s_n],s)$,
-we assume a function $\semIndexExp{\texttt{f}} : \semSort{s_1,...,s_n}
-\to \semSort{s}$. For each well-sorted index expression $\Delta \vdash
-e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
-\semSort{s}$ by recursion on the structure of $e$. We further assume
-that for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
-\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$. A
-\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
-\semIndexExp{\cdot}_{\mathcal{M}})$ satisfying all the axioms in
-$\IndexAxiomSet$.
-
-Now fix a model $\mathcal{M}$ and assume, for each primitive type
-$\tyPrimNm{X}$, a relational interpretation parameterised by elements
-from the model: $R_{\tyPrimNm{X}} :
-\semSort{\primTyArity(\tyPrimNm{X})}_{\mathcal{M}} \to
-\Rel(\tyPrimSem{\tyPrimNm{X}})$. Our family of sets of relational
-environments is $\relEnv{E}_{\mathcal{M},R}(\Delta) = \{ \rho^\delta
-\sepbar \delta \in \semSort{\Delta}_{\mathcal{M}} \}$, where
-$\rho^\delta\ \tyPrimNm{X}\ (e_1,...,e_n) =
-R_{\tyPrimNm{X}}(\semIndexExp{e_1}_{\mathcal{M}}\delta, ...,
-\semIndexExp{e_n}_{\mathcal{M}}\delta)$.
-
-\begin{theorem}
-  The families $\relEnv{E}_{\mathcal{M},R}$ are substitutive.
-\end{theorem}
 

### 2012-10-16 paper/section-instantiations.tex
Final small changes
New framework setup

@@
 
-We define the model of the indexing theory for the two-dimensional
+We define the model $\mathcal{M}$ of the indexing theory for the two-dimensional
 geometry instantiation as follows. Each of the sorts is interpreted just as
-its semantic counterpart: $\semSort{\SynTransl{2}} = \Transl{2}$, $
-\semSort{\SynGL{2}} = \GL{2}$, $ \semSort{\SynOrth{2}} = \Orth{2}$,
-and $ \semSort{\SynGL{1}} = \GL{1}$.
+its semantic counterpart: $\semSort{\SynTransl{2}}M = \Transl{2}$, $
+\semSort{\SynGL{2}}M = \GL{2}$, $ \semSort{\SynOrth{2}}M = \Orth{2}$,
+and $ \semSort{\SynGL{1}}M = \GL{1}$.
 %  \begin{displaymath}
@@
 clearly satisfy the axioms in \exref{ex:two-dim-geo-axioms}. For the
-relational interpretations of $\tyPrim{vec}{B,t}$ and
-$\tyPrim{real}{s}$, we use the model-based construction from
-\autoref{sec:constr-rel-env}, with $R_{\tyPrimNm{vec}}(B,\vec{t}) = \{
+relational interpretations $\mathcal{R}$ of $\tyPrim{vec}{B,t}$ and
+$\tyPrim{real}{s}$, we use $\mathcal{R}_{\tyPrimNm{vec}}(B,\vec{t}) = \{
 (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \}$ and
-$R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}$. The
+$\mathcal{R}_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}$. The
 types of the primitive vector operations presented in
@@
 \begin{lemma}\label{lem:geom-environments-1}
-  For all $\Delta$ and $\rho \in \relEnv{E}_\Geom(\Delta)$, $(\eta_\Geom,\eta_\Geom) \in \rsem{\Gamma_\Geom}{\relEnv{E}_\Geom}\rho$.
+  For all $\Delta$ and $\rho \in \semSort{\Delta}M$, $(\eta_\Geom,\eta_\Geom) \in \rsem{\Gamma_\Geom}{\relEnv{E}_\Geom}\rho$.
 \end{lemma}
@@
 respect to contextual equivalence, i.e., if there are terms $\vdash I
-: A \to B$ and $\vdash J : B\to A$ such that $x:A\vdash
-J(I(x))\stackrel{ctx}\approx x : A$ and $y:B\vdash
-I(J(y))\stackrel{ctx}\approx y : B$.
+: A \to B$ and $\vdash J : B\to A$ such that $\ctxteq{\cdot}{x:A}{J(I(x))}{x}{A}$ 
+and $\ctxteq{\cdot}{y:B}{I(J(y))} y B$.
 %\[
@@
 $\tyPrim{vec}{t}$. For the purposes of non-definability results, we
-may choose an arbitrary index-erasure semantics that supports the
-primitive operations.
+may choose an arbitrary index-erasure semantics with $\eta_\Grp \in
+\ctxtSem{\Gamma_\Grp}$.
 

### 2012-10-17 paper/section-general-framework.tex
More cleanup

@@
 \newcommand{\Free}{\mathrm{Free}}
-\newcommand{\semSort}[2]{\llbracket #1 \rrbracket^{\mathcal{#2}}}
-\newcommand{\semIndexExp}[2]{\llbracket #1 \rrbracket^{\mathcal{#2}}}
-\newcommand{\semPrim}[3]{\llbracket #1 \rrbacket^{\mathcal{#2}}_{\mathcal{#3}}}
+\newcommand{\semSort}[1]{\llbracket #1 \rrbracket}
+\newcommand{\semIndexExp}[1]{\llbracket #1 \rrbracket}
+\newcommand{\semPrimType}[1]{\llbracket #1 \rrbracket}
 \newcommand{\opsCtxt}{\Gamma_{\mathit{ops}}}
 \newcommand{\opsEnv}{\eta_{\mathit{ops}}}
+\newcommand{\ctxteq}[5]{{#1};{#2}\vdash{#3}\approx{#4}:{#5}}
+\newcommand{\semeq}[5]{{#1};{#2}\models{#3}\sim{#4}:{#5}}
 \section{A General Framework}
@@
 prove the central Abstraction theorem
-(\autoref{sec:abstraction-theorem}), and use the relational semantics
-to define a notion of semantic equivalence which we prove is a sound
-approximation of contextual equivalence.
+and use the relational semantics
+to define a notion of semantic equivalence that soundly approximates 
+contextual equivalence (\autoref{sec:abstraction-theorem}).
 
+We will use a type system for affine geometry as a running example
+throughout, so that by the end of the section we have prepared enough
+syntactic and semantic gadgets to let us prove invariance
+and abstraction properties for geometric examples in
+\autoref{sec:instantiations}.
 %and semantics of algebraically indexed types, then their semantics
@@
 \item A collection $\PrimTypeSet$ of primitive types, with a function
-  $\primTyArity : \PrimTypeSet \to \SortSet^*$. %, describing the sorts
-%  of the arguments of each primitive type.
+  $\primTyArity : \PrimTypeSet \to \SortSet^*$, describing the sorts
+  of the arguments of each primitive type.
 \end{enumerate}
@@
   \end{displaymath}
-%  The intended interpretations of the top three pairs of operations
-%  are group unit, group combination and group negation,
-%  respectively. 
-%  When we discuss equational theories on index expressions 
-%in \autoref{sec:type-equality} 
-%we will impose the (abelian) group laws. For this example, 
-We also have $\PrimTypeSet = \{ \tyPrimNm{vec},
+  The intended interpretations of the top three pairs of operations
+  are group unit, group combination and group negation,
+  respectively. 
+  When we discuss equational theories on index expressions 
+in \autoref{sec:type-equality} 
+we will impose the (abelian) group laws. For this example, 
+we also have $\PrimTypeSet = \{ \tyPrimNm{vec},
   \tyPrimNm{real} \}$, with $\primTyArity(\tyPrimNm{vec}) =
@@
 with respect to an index context $\Delta$. The rules for well-sorted
-index expressions are particularly simple. %: either an index expression
-%is a variable that appears in the context (rule \TirName{IVar}), or it
-%is an application of an index operation taken from $\IndexOpSet$ to
-%other index expressions (rule \TirName{IOp}). 
+index expressions are particularly simple: either an index expression
+is a variable that appears in the context (rule \TirName{IVar}), or it
+is an application of an index operation taken from $\IndexOpSet$ to
+other index expressions (rule \TirName{IOp}). 
 The rules for well-indexed types include the usual ones
@@
 
-\subsubsection{Simultaneous Substitution of Index Expressions}
+\subsubsection{Substitution of Index Expressions}
 \label{sec:simultaneous-substitution}
@@
 Given a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord:
-s_1, \ldots, i_n \mathord: s_n$, a \emph{simultaneous substitution}
+s_1, \ldots, i_n \mathord: s_n$, a (simultaneous) \emph{substitution}
 $\Delta \vdash \sigma \Rightarrow \Delta'$ is a sequence of
 expressions $\sigma = (e_1,\ldots,e_n)$ such that $\Delta \vdash e_j :
-s_j$ for all $1 \leq j \leq n$. Given a simultaneous substitution
+s_j$ for all $1 \leq j \leq n$. Given a substitution
 $\Delta \vdash \sigma = (e_1,\ldots,e_n) \Rightarrow \Delta'$ and a
@@
 the index expression $e_j$. We write $\Delta \Rightarrow \Delta'$ for
-the set of all simultaneous substitutions $\sigma$ such that $\Delta
+the set of all substitutions $\sigma$ such that $\Delta
 \vdash \sigma \Rightarrow \Delta'$.
@@
 We can think of any sequence of sorts as an index
-context. In particular, we will make use of simultaneous substitutions
+context. In particular, we will make use of substitutions
 of the form $\Delta \vdash \sigma \Rightarrow
@@
 further abuse of notation, we write $\Delta \Rightarrow
-\primTyArity(\tyPrimNm{X})$ for the set of all simultaneous
+\primTyArity(\tyPrimNm{X})$ for the set of all 
 substitutions $\sigma$ such that $\Delta \vdash \sigma \Rightarrow
@@
 
-For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
+For a substitution $\Delta \vdash \sigma \Rightarrow
 \Delta'$, where $\Delta' = i_1\mathord:s_1,\ldots,i_n\mathord:s_n$, and a
@@
 either $\Delta$ or $\Delta'$, we can form the \emph{lifted}
-simultaneous substitution $\Delta,i\mathord:s \vdash
+substitution $\Delta,i\mathord:s \vdash
 \sigma_{i\mathord:s} = (\sigma(i_1), \ldots, \sigma(i_n), i) \Rightarrow
@@
 %
-Application of a simultaneous substitution $\Delta \vdash \sigma
+Application of a substitution $\Delta \vdash \sigma
 \Rightarrow \Delta'$ to a well-sorted index expression $\Delta' \vdash
@@
 % \end{lemma}
-The \emph{identity} simultaneous substitution $\Delta \vdash
+The \emph{identity} substitution $\Delta \vdash
 \id_\Delta \Rightarrow \Delta$ is %just the sequence of variables in
@@
 i_1\mathord:s_1,\ldots,i_n\mathord:s_n$. The \emph{composition} of 
-%two simultaneous substitutions 
+two substitutions 
 $\Delta \vdash \sigma \Rightarrow \Delta'$
@@
 variable/sort pair $i\mathord:s$ such that $i$ does not appear in
-$\Delta$, we define the \emph{projection} simultaneous substitution
+$\Delta$, we define the \emph{projection} substitution
 $\Delta, i\mathord:s \vdash \pi_{i\mathord:s} \Rightarrow \Delta$ as
@@
 %A pair of 
-The simultaneous substitutions $\Delta \vdash \sigma \Rightarrow
+The substitutions $\Delta \vdash \sigma \Rightarrow
 \Delta'$ and $\Delta \vdash \sigma' \Rightarrow \Delta'$ are defined
-to be equal if their component expressions are equal in the context
+to be equal, and written  $\Delta \vdash \sigma \equiv \sigma' \Rightarrow
+\Delta'$, if their component expressions are equal in the context
 $\Delta$: i.e.,~ if $\Delta \vdash e_j \equiv e'_j : s_j$, for all
-$j$. We write $\Delta \vdash \sigma \equiv \sigma' \Rightarrow
-\Delta'$ when two simultaneous substitutions are equal.
+$j$.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Well-typed Programs}
+\subsection{Well-typed terms}
 \label{sec:well-typed-programs}
 
-We now present the rules for well-typed programs over the 
+We now present the rules for well-typed terms over the 
 collection of 
@@
 
-Well-typed programs are defined with respect to well-indexed typing
+Well-typed terms are defined with respect to well-indexed typing
 contexts, which are in turn defined with respect to an index
@@
 \end{mathpar}
-Application of simultaneous substitutions extends to typing contexts
-by applying the simultaneous substitution to each type.
+Application of substitutions extends to typing contexts
+by applying the substitution to each type.
 
 %The typing rules for our framework define 
-Well-typed programs 
+Well-typed terms
 are defined
@@
 incorporated into the type system via the rule \TirName{TyEq}, which
-allows for a program that has %is judged to have 
+allows a term that has %is judged to have 
 type $A$ to also have any
@@
 
-  \caption{Well-typed Programs}
+  \caption{Well-typed terms}
   \label{fig:programs}
@@
 We assume $\opsCtxt$ collects together the types of primitive operations as described in \autoref{sec:motivation-generalising}.
+\todo{Write these down}.
+\qed
 \end{example*}
@@
 %\fixme{Make sure that the semantics is motivated}
-Having defined the language of algebraically-indexed types and a
-programs, we turn to their denotational interpretation.  We first define an
+Having defined the syntax of algebraically-indexed types and terms,
+we turn to their denotational interpretation.  We first define an
 \emph{index-erasure} interpretation of types and terms that interprets
 every well-indexed type as a set, ignoring the indexing
-expressions. %Building on the index-erasure semantics, we
+expressions, and which interprets open terms as functions that map environments
+to final values.
+%Building on the index-erasure semantics, we
 %We then define in
@@
 \end{displaymath}
-Note that the interpretations of the universal and existential
-quantifiers do indeed ignore the indexing. (We will overload the
+(We will overload the
 notation $\left|\cdot\right|$ for all index-erasure interpretations, 
@@
 %i\mathord:s.A$.
-We then have:
 
-% The index-erasure interpretation completely ignores index expressions
-% and type equality is defined as an extension of index
-% equality. Therefore, it is straightforward to prove that equal types
-% have equal denotations when interpreted in the index-erasure
-% semantics, and that substitution of index terms has no effect on the
-% index-erasure interpretation of types:
+The index-erasure interpretation completely ignores index expressions
+and quantifiers, and type equality is defined as an extension of index
+equality. Therefore, it is straightforward to prove that equal types
+have equal denotations when interpreted in the index-erasure
+semantics, and that substitution of index terms has no effect on the
+index-erasure interpretation of types:
 \begin{lemma}\label{lem:tyeqsubst-erasure}
@@
 
-\paragraph{Interpretation of Programs.}
+\paragraph{Interpretation of Terms.}
 
@@
 $\ctxtSem{\epsilon} = \{*\}$ and $\ctxtSem{\Gamma, x : A} =
-\ctxtSem{\Gamma} \times \tySem{A}$. For a well-typed program $\Delta;
+\ctxtSem{\Gamma} \times \tySem{A}$. For a well-typed term $\Delta;
 \Gamma \vdash M : A$, we define the \emph{erasure interpretation} as a
@@
   We assume that $\opsEnv$ interprets the operations described in \autoref{sec:motivation-generalising}
-  in the standard way, so for example, for  blah blah
+  in the standard way. \todo{Expand}. 
 \qed
@@
 \paragraph{Contextual Equivalence.}
-We use our index-erasure semantics to define when a pair of programs
+We use our index-erasure semantics to define when a pair of terms
 are contextually equivalent with respect to syntactically defined
@@
 
-\newcommand{\ctxteq}[5]{{#1};{#2}\vdash{#3}\approx{#4}:{#5}}
 \begin{definition}[Contextual Equivalence]\label{defn:ctxt-equiv}
-Two programs $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 : A$
-and $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_2 : A$ are
-\emph{contextually equivalent}, written $\ctxteq\Delta\Gamma{M_1}{M_2}A$, if 
+Two terms $\Delta; \opsCtxt, \Gamma \vdash M_1, M_2 : A$
+are \emph{contextually equivalent}, written $\ctxteq\Delta\Gamma{M_1}{M_2}A$, if 
   for all
-  programs $\cdot;\Gamma_{\mathit{ops}} \vdash P : (\forall \Delta. \Gamma
-  \to A) \to \tyPrimNm{bool}$, it is the case that $\tmSem{P\ (\Lambda
-    \Delta. \lambda \Gamma. M_1)}\eta_{\mathit{ops}} = \tmSem{P\
-    (\Lambda \Delta. \lambda \Gamma. M_2)}\eta_{\mathit{ops}}$.
+  contexts $\cdot;\opsCtxt \vdash C : (\forall \Delta. \Gamma
+  \to A) \to \tyPrimNm{bool}$, it is the case that $\tmSem{C\ (\Lambda
+    \Delta. \lambda \Gamma. M_1)}\opsEnv = \tmSem{C\
+    (\Lambda \Delta. \lambda \Gamma. M_2)}\opsEnv$.
 \end{definition}
@@
 \paragraph{Models.}
-Formally, a \emph{model} $\mathcal{M}$ assigns to each sort $s \in \SortSet$ a
-carrier set $\semSort s M$, and assigns to each index operation $\texttt{f}
+A \emph{model} 
+%$\mathcal{M}$ 
+assigns to each sort $s \in \SortSet$ a
+carrier set $\semSort s$, and assigns to each operation $\texttt{f}
 \in \IndexOpSet$ with $\indexOpArity(\texttt{f}) =
-([s_1,\ldots,s_n],s)$, a function $\semIndexExp{\texttt{f}}M :
-\semSort{s_1}M\times\cdots\times\semSort{s_n}M \to \semSort{s}M$.
+([s_1,\ldots,s_n],s)$, a function $\semIndexExp{\texttt{f}} :
+\semSort{s_1}\times\cdots\times\semSort{s_n} \to \semSort{s}$.
+
 An index context $\Delta = (i_1\mathord:s_1,\ldots,i_n\mathord:s_n)$ is
-interpreted as cartesian product \emph{i.e.} $\semSort{\Delta}M \isDefinedAs \semSort{s_1}M
-\times\cdots \times \semSort{s_n}M$. 
+interpreted as cartesian product \emph{i.e.} $\semSort{\Delta} \isDefinedAs \semSort{s_1}
+\times\cdots \times \semSort{s_n}$. 
 For each well-sorted index expression $\Delta \vdash
-e : s$, we assign a function $\semIndexExp{e}M : \semSort{\Delta}M \to
-\semSort{s}M$ by recursion on the structure of~$e$:
-\[
-\begin{array}{rcl}
-\semIndexExp{i}M \rho & \isDefinedAs & \rho(i)\\
-\semIndexExp{\indexOp{f}(e_1, \ldots, e_n)}M \rho & \isDefinedAs & \semIndexExp{\indexOp{f}}M(
-\semIndexExp{e_1}M\rho,\ldots,\semIndexExp{e_n}M\rho)
-\end{array}
-\]
+e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
+\semSort{s}$ by recursion on the structure of~$e$:
+\begin{align*}
+\semIndexExp{i} \rho & \isDefinedAs \rho(i) &
+\semIndexExp{\indexOp{f}(e_1, \ldots, e_n)} \rho & \isDefinedAs \semIndexExp{\indexOp{f}}(
+\semIndexExp{e_1}\rho,\ldots,\semIndexExp{e_n}\rho)
+\end{align*}
 
-Finally, a model $\mathcal{M}$
+Finally, a model
+% $\mathcal{M}$
 must be sound, that is, for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
-\IndexAxiomSet$, we have $\semIndexExp{e}M = \semIndexExp{e'}M$.
+\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$.
 
 \begin{example*}[Geometry: affine model]
-\todo{Give example}
+We define the model
+% $\mathcal{M}$ 
+of the indexing theory for the two-dimensional
+geometry example as follows. Each of the sorts is interpreted just as
+its semantic counterpart: 
+%$\semSort{\SynTransl{2}}M = \Transl{2}$, $
+%\semSort{\SynGL{2}}M = \GL{2}$, $ \semSort{\SynOrth{2}}M = \Orth{2}$,
+%and $ \semSort{\SynGL{1}}M = \GL{1}$.
+\begin{align*}
+\semSort{\SynTransl{2}} & = \Transl{2}
+&\semSort{\SynGL{2}} & = \GL{2} &
+\semSort{\SynOrth{2}} & = \Orth{2}
+&\semSort{\SynGL{1}} & = \GL{1}
+\end{align*}
+Each of the index operations (e.g.,~the group structure and
+determinant) is interpreted by the intended semantic operation, and
+clearly satisfy the axioms in \exref{ex:two-dim-geo-axioms}. 
+\qed
 \end{example*}
 
-Given $\rho \in \semSort{\Delta}M$ and a simultaneous substitution
-$\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
-relational environment (with an abuse of notation) $\rho \circ \sigma
-\in \semSort{\Delta'}M$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
-\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$.
-\todo{This is wrong: fix!}
+Given $\rho \in \semSort{\Delta}$ and a substitution
+$\Delta \vdash \sigma \Rightarrow \Delta'$ with $\sigma = (e_1,\ldots,e_n)$, 
+we can derive the composed
+relational environment $\rho \circ \sigma
+\in \semSort{\Delta'}$ as $\rho \circ \sigma \isDefinedAs (\semIndexExp{e_1}\rho,\ldots,\semIndexExp{e_n}\rho)$.
 
@@
 \paragraph{Relational interpretation of primitive types.}
-Having fixed a model $\mathcal{M}$ we next choose a relational interpretation $\mathcal{R}$ of
+Having fixed a model
+% $\mathcal{M}$ 
+we next choose a relational interpretation 
+%$\mathcal{R}$ 
+of
 primitive types: for each each primitive type
 $\tyPrimNm{X}$, a relational interpretation parameterised by elements
-from the model: $\mathcal{R}_{\tyPrimNm{X}} :
-\semSort{\primTyArity(\tyPrimNm{X})}M \to \Rel(\tyPrimSem{\tyPrimNm{X}})$.
+from the model: $\semPrimType{\tyPrimNm{X}} :
+\semSort{\primTyArity(\tyPrimNm{X})} \to \Rel(\tyPrimSem{\tyPrimNm{X}})$.
 
 \begin{example*}[Geometry: change of basis interpretation]
-\todo{Give example}
+Given the model of geometric groups described above,
+for the relational interpretation  of $\tyPrim{vec}{B,t}$ and
+$\tyPrim{real}{s}$, we use $\semPrimType{\tyPrimNm{vec}}(B,\vec{t}) = \{
+(\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \}$ and
+$\semPrimType{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}$.
+\qed
 \end{example*}
@@
 relational environments 
-$\rho \in \semSort{\Delta}M$:
+$\rho \in \semSort{\Delta}$:
 \begin{eqnarray*}
-  \rsem{\Delta \vdash A \isType}M R & : & \semSort\Delta M \to \Rel(\tyPrimSem{A})\\
-  \rsem{\tyUnit}M R \rho & \isDefinedAs & \{(*,*)\} \\
-  \rsem{\tyPrim{X}{e_1,\ldots,e_n}} M R\rho & \isDefinedAs & 
-  \mathcal{R}_{\tyPrimNm{X}} (\semIndexExp{e_1}M\delta, \ldots,
-  \semIndexExp{e_n}M\delta)\\
-  \rsem{A \tyArr B}M R \rho & \isDefinedAs & \rsem{A}M R \rho \relArrow \rsem{B} M R \rho \\
-  \rsem{A \tyProduct B} M R \rho & \isDefinedAs & \rsem{A} M R \rho \relTimes \rsem{B} M R \rho \\
-  \rsem{A + B}M R \rho & \isDefinedAs & \rsem{A} M R \rho \relSum \rsem{B} M R \rho \\
-  \rsem{\forall i\mathord:s.A} M R \rho & \isDefinedAs & 
-  \bigcap\{ \rsem{A} M R {(m,\rho)} \sepbar m \in \semSort{s}M \} \\
-  \rsem{\exists i\mathord:s.A} M R \rho & \isDefinedAs & 
-  \bigcup\{ \rsem{A} M R {(m,\rho)} \sepbar m \in \semSort{s}M \}
+%  \rsem{\Delta \vdash A \isType}M R & : & \semSort\Delta  \to \Rel(\tyPrimSem{A})\\
+  \rsem{\tyUnit} \rho & \isDefinedAs & \{(*,*)\} \\
+  \rsem{\tyPrim{X}{e_1,\ldots,e_n}}\rho & \isDefinedAs & 
+  \semPrimType{\tyPrimNm{X}} (\semIndexExp{e_1}\rho, \ldots,
+  \semIndexExp{e_n}\rho)\\
+  \rsem{A \tyArr B} \rho & \isDefinedAs & \rsem{A}\rho \relArrow \rsem{B} \rho \\
+  \rsem{A \tyProduct B}\rho & \isDefinedAs & \rsem{A} \rho \relTimes \rsem{B} \rho \\
+  \rsem{A + B} \rho & \isDefinedAs & \rsem{A} \rho \relSum \rsem{B} \rho \\
+  \rsem{\forall i\mathord:s.A} \rho & \isDefinedAs & 
+  \bigcap\{ \rsem{A} {(m,\rho)} \sepbar m \in \semSort{s} \} \\
+  \rsem{\exists i\mathord:s.A} \rho & \isDefinedAs & 
+  \bigcup\{ \rsem{A} {(m,\rho)} \sepbar m \in \semSort{s} \}
 \end{eqnarray*}
@@
 
-The following lemma states that the relation interpretation of types
+The following lemma states that the relational interpretation of types
 that we have defined in this section behaves well: the first part of
@@
 substitution of index expressions in types can be interpreted by the
-composition of relational environments with simultaneous
+composition of relational environments with 
 substitutions.
@@
   \begin{enumerate}
-  \item If $\Delta \vdash A \equiv B \isType$, then $\rsem{A} M R =  \rsem{B} M R$;
+  \item If $\Delta \vdash A \equiv B \isType$, then $\rsem{A} =  \rsem{B}$;
   \item If $\Delta' \vdash A \isType$ then for all $\Delta \vdash \sigma \Rightarrow
-    \Delta'$ and $\rho \in \semSort{\Delta}M$,
-    $\rsem{\sigma^*A} M R \rho = \rsem{A} M R(\rho \circ \sigma)$.
+    \Delta'$ and $\rho \in \semSort{\Delta}$,
+    $\rsem{\sigma^*A} \rho = \rsem{A}(\rho \circ \sigma)$.
   \end{enumerate}
@@
 (\thmref{thm:abstraction}) is that the index-erasure semantics of
-every well-typed program is related to itself in the relational
+every well-typed term is related to itself in the relational
 interpretation of its type: this is the abstraction theorem for every
@@
 
-We now state the abstraction theorem for well-typed programs. To state
-and prove this theorem for open programs, we extend the
+We now state the abstraction theorem for well-typed terms. To state
+and prove this theorem for open terms, we extend the
 relational interpretation of types to typing contexts. The relational
@@
   \begin{array}{@{}l@{\hspace{0.1em}\isDefinedAs\hspace{0.1em}}l@{\hspace{1.5em}}l@{\hspace{0.1em}\isDefinedAs\hspace{0.1em}}l}
-    \rsem{\epsilon} M R{\rho} & \{(*,*)\} &
-    \rsem{\Gamma, x : A} M R{\rho} & \rsem{\Gamma} M R\rho \relTimes \rsem{A} M R\rho
+    \rsem{\epsilon} {\rho} & \{(*,*)\} &
+    \rsem{\Gamma, x : A} {\rho} & \rsem{\Gamma} \rho \relTimes \rsem{A} \rho
   \end{array}
 \end{displaymath}
-% The relational interpretation of contexts inherits from the relational
-% interpretation of types the property of interpreting the application
-% of simultaneous substitutions as composition:
-% \begin{lemma}\label{lem:ctxtsubst-rel}
-%   If $\Delta' \vdash \Gamma \isCtxt$ and $\Delta \vdash \sigma
-%   \Rightarrow \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
-%   $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =
-%   \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma)}$.
-% \end{lemma}
 
+The relational interpretation of contexts inherits from the relational
+interpretation of types the property of interpreting the application
+of substitutions as composition:
+\begin{lemma}\label{lem:ctxtsubst-rel}
+  If $\Delta' \vdash \Gamma \isCtxt$ and $\Delta \vdash \sigma
+  \Rightarrow \Delta'$, then for all $\rho \in \semSort{\Delta}$,
+we have 
+  $\rsem{\sigma^*\Gamma} \rho =
+  \rsem{\Gamma}(\rho \circ \sigma)$.
+\end{lemma}
+
+Given a particular choice of model and relational interpretation of
+primitive types, we can then prove the following.
 \begin{theorem}[Abstraction]\label{thm:abstraction}
   If $\Delta; \Gamma \vdash M :
-  A$, then for all $\rho \in \semSort{\Delta}M$ and $\eta_1, \eta_2
+  A$, then for all $\rho \in \semSort{\Delta}$ and $\eta_1, \eta_2
   \in \ctxtSem{\Gamma}$ such that $(\eta_1, \eta_2) \in
-  \rsem{\Gamma} M R\rho$, we have $(\tmSem{M}\eta_1,
-  \tmSem{M}\eta_2) \in \rsem{A} M R\rho$.
+  \rsem{\Gamma} \rho$, we have $(\tmSem{M}\eta_1,
+  \tmSem{M}\eta_2) \in \rsem{A} \rho$.
 \end{theorem}
@@
 By induction on the typing derivation, making use of
-Lemma~\ref{lem:tyeqsubst-relational} (part 1) for rule \TirName{TyEq},
-and Lemma~\ref{lem:tyeqsubst-relational} (part 2) for the quantifier introduction
-and elimination rules. The details can be found in the Coq development.
+Lemma~\autoref{lem:tyeqsubst-relational} (part 1) for rule \TirName{TyEq},
+Lemma~\autoref{lem:tyeqsubst-relational} (part 2) for rules \TirName{UnivApp} and \TirName{ExPack},
+and Lemma~\autoref{lem:ctxtsubst-rel} for rules \TirName{UnivAbs} and \TirName{ExUnpack}. 
+The details can be found in the Coq development.
 \end{proof}
 
-Semantic equivalence of programs is defined in terms of the relational
+Semantic equivalence of terms is defined in terms of the relational
 interpretation. As a consequence of \thmref{thm:abstraction}, semantic
@@
 
-  Let $\Gamma_{\mathit{ops}}$ and $\eta_{\mathit{ops}}$ be a context
-  of primitive operations and its interpretation as in
-  \defref{defn:ctxt-equiv}. Let $\mathcal{M}$ be a model and $\mathcal{R}$  
-  a relational interpretation such that for
-  all $\Delta$ and $\rho \in \semSort{\Delta}M$,
-  $(\eta_{\mathit{ops}}, \eta_{\mathit{ops}}) \in
-  \rsem{\Gamma_{\mathit{ops}}} M R \rho$.
-\newcommand{\semeq}[7]{{#3};\Gamma_{\mathit{ops}},{#4}\models^{\mathcal{#1}}_{\mathcal{#2}}{#5}\sim{#6}:{#7}}
+  Let $\opsCtxt$ be the context of primitive operations and $\opsEnv$  its interpretation as in
+  \defref{defn:ctxt-equiv}. Fix a model and relational interpretation of primitive types so that
+  $(\opsEnv, \opsEnv) \in  \rsem{\opsCtxt} *$.
 
 \begin{definition}[Semantic equivalence]\label{def:semantic-equality}
-  Two programs $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 : A$
-  and $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_2 : A$ are
-  \emph{semantically equal} with respect to $\mathcal{M}$ and $\mathcal{R}$,
-  written $\semeq M R \Delta\Gamma{M_1}{M_2} A$, if for all $\rho \in
-  \semSort{\Delta}M$, and all $(\eta_1,\eta_2) \in
-  \rsem{\Gamma} M R \rho$, we have
-  $(\tmSem{M_1}(\eta_{\mathit{ops}},
-  \eta_1),\tmSem{M_2}(\eta_{\mathit{ops}}, \eta_2)) \in
-  \rsem{A} M R\rho$.
+  Two terms $\Delta; \opsCtxt, \Gamma \vdash M_1 , M_2 : A$
+  are
+  \emph{semantically equal},
+  written $\semeq \Delta\Gamma{M_1}{M_2} A$, if for all $\rho \in
+  \semSort{\Delta}$, and all $(\eta_1,\eta_2) \in
+  \rsem{\Gamma} \rho$, we have
+  $(\tmSem{M_1}(\opsEnv,
+  \eta_1),\tmSem{M_2}(\opsEnv, \eta_2)) \in
+  \rsem{A}\rho$.
 \end{definition}
@@
 \begin{theorem}[Soundness]\label{thm:soundness}
-For any model $\mathcal{M}$ and interpretation $\mathcal{R}$, 
-if $\semeq M R \Delta\Gamma {M_1}{M_2} A$ then
+If $\semeq \Delta\Gamma {M_1}{M_2} A$ then
 $\ctxteq \Delta \Gamma {M_1}{M_2}A$.

### 2012-10-17 paper/section-instantiations.tex
More cleanup

@@
 \label{sec:types-indexed-abelian-groups}
-We say that two types $A$ and $B$ are isomorphic, and write $A \cong
+We say that two types $A$ and $B$ are isomorphic, and write $\Delta\vdash A \cong
 B$, if there exist maps between them that are mutually inverse with
-respect to contextual equivalence, i.e., if there are terms $\vdash I
-: A \to B$ and $\vdash J : B\to A$ such that $\ctxteq{\cdot}{x:A}{J(I(x))}{x}{A}$ 
-and $\ctxteq{\cdot}{y:B}{I(J(y))} y B$.
+respect to contextual equivalence, i.e., if there are terms $\Delta\vdash I
+: A \to B$ and $\Delta\vdash J : B\to A$ such that $\ctxteq{\Delta}{x:A}{J(I(x))}{x}{A}$ 
+and $\ctxteq{\Delta}{y:B}{I(J(y))} y B$.
 %\[

### 2012-10-18 paper/section-general-framework.tex
Definability examples, with proofs

@@
 semantics to types and terms (\autoref{sec:erasure-semantics}), and
-based on this semantics define a notion of contextual equivalence. We
+based on this semantics define notions of contextual equivalence and
+type isomorphism. We
 then introduce a relational semantics for types parameterised by an appropriate
@@
 
+\paragraph{Type Isomorphism.}
+We say that well-indexed types $\Delta \vdash A \isType$ and $\Delta\vdash B \isType$
+are isomorphic, and write $\Delta\vdash A \cong
+B$, if there exist maps between them that are mutually inverse with
+respect to contextual equivalence, i.e., if there are terms $\Delta\vdash I
+: A \to B$ and $\Delta\vdash J : B\to A$ such that $\ctxteq{\Delta}{x:A}{J(I(x))}{x}{A}$ 
+and $\ctxteq{\Delta}{y:B}{I(J(y))} y B$.
+%\[
+%x:A\vdash J(I(x))\stackrel{ctx}\approx x : A
+%\text{ and }y:B\vdash I(J(y))\stackrel{ctx}\approx y : B.
+%\]
+That $\cong$ is a congruence with respect to the type formation rules
+of \autoref{fig:indexes-and-types} is straightforward. We can also
+derive isomorphisms that are independent of the indexing theory, such
+as $\Delta\vdash A\times B \cong B \times A$, and $\Delta\vdash\forall
+i\mathord:s.(A\to B)\cong A\to\forall i\mathord:s.B$ for $i$ not free in $A$.
+
 \subsection{The Relational Interpretation of Types}

### 2012-10-18 paper/section-instantiations.tex
Definability examples, with proofs

--- 2012-10-16 paper/section-instantiations.tex
@@
 \label{sec:instantiations}
-We now instantiate our general framework %of the previous section 
-with
+We now instantiate our general framework %of the previous section with
 the indexing theory of \autoref{sec:motivating-examples}, and present
-more general and formally-justified free theorems, type
-isomorphisms, and non-definability results.
+more general and formally-justified free theorems, type isomorphisms,
+and non-definability results.  For the free theorems and isomorphisms,
+we use the model and relational interpretation of primitive types
+described in \autoref{sec:relational-semantics}, namely that of
+affine transformations for vectors and scaling for scalars.
 
 %\begin{example}[Two-dimensional Geometry]
-
-We define the model $\mathcal{M}$ of the indexing theory for the two-dimensional
-geometry instantiation as follows. Each of the sorts is interpreted just as
-its semantic counterpart: $\semSort{\SynTransl{2}}M = \Transl{2}$, $
-\semSort{\SynGL{2}}M = \GL{2}$, $ \semSort{\SynOrth{2}}M = \Orth{2}$,
-and $ \semSort{\SynGL{1}}M = \GL{1}$.
-%  \begin{displaymath}
-%    \begin{array}{@{}llll}
-%      \semSort{\SynTransl{2}} = \Transl{2}
-%      &
-%      \semSort{\SynGL{2}} = \GL{2}
-%      &
-%      \semSort{\SynOrth{2}} = \Orth{2}
-%      &
-%      \semSort{\SynGL{1}} = \GL{1}
-%    \end{array}
-%  \end{displaymath}
-Each of the index operations (e.g.,~the group structure and
-determinant) is interpreted by the intended semantic operation, and
-clearly satisfy the axioms in \exref{ex:two-dim-geo-axioms}. For the
-relational interpretations $\mathcal{R}$ of $\tyPrim{vec}{B,t}$ and
-$\tyPrim{real}{s}$, we use $\mathcal{R}_{\tyPrimNm{vec}}(B,\vec{t}) = \{
-(\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \}$ and
-$\mathcal{R}_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}$. The
-types of the primitive vector operations presented in
-\autoref{sec:motivating-examples} are collected into a context
-$\Gamma_\Geom$ with a straightforward interpretation $\eta_\Geom \in
-\ctxtSem{\Gamma_\Geom}$.
-  % \begin{displaymath}
-  %   \begin{array}{l}
-  %     R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
-  %     R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
-  %   \end{array}
-  % \end{displaymath}
-%   The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
-%   %are as expected. % given the informal definitions we gave in
-% formalise the informal ones from~\autoref{sec:motivating-examples}, and 
-%   the definitions of %We have set
-%   $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ 
-%   %to relate $\vec{0}$ and $0$ to themselves to 
-%   account for the
-%   polymorphic zero constants we have assumed.
-%\end{example}
-
-\begin{lemma}\label{lem:geom-environments-1}
-  For all $\Delta$ and $\rho \in \semSort{\Delta}M$, $(\eta_\Geom,\eta_\Geom) \in \rsem{\Gamma_\Geom}{\relEnv{E}_\Geom}\rho$.
-\end{lemma}
 
@@
 \label{sec:types-indexed-abelian-groups}
-We say that two types $A$ and $B$ are isomorphic, and write $A \cong
-B$, if there exist maps between them that are mutually inverse with
-respect to contextual equivalence, i.e., if there are terms $\vdash I
-: A \to B$ and $\vdash J : B\to A$ such that $\ctxteq{\cdot}{x:A}{J(I(x))}{x}{A}$ 
-and $\ctxteq{\cdot}{y:B}{I(J(y))} y B$.
-%\[
-%x:A\vdash J(I(x))\stackrel{ctx}\approx x : A
-%\text{ and }y:B\vdash I(J(y))\stackrel{ctx}\approx y : B.
-%\]
-That $\cong$ is a congruence relation on types is straightforward. We
-can also derive isomorphisms that are independent of the indexing
-theory, such as $A\times B \cong B \times A$, and $\forall
-i\mathord:s.A\to B\cong A\to\forall i\mathord:s.B$ for $i$ not free in
-$A$.
-
 Types indexed by abelian groups induce a particularly rich theory of
@@
 \begin{eqnarray*}
-\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s\cdot
-  s}\\ \quad\cong\quad \tyPrim{real}{1}\times(\forall
-s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s})\\ \quad\cong\quad
+\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s^2}\\ \quad\cong\quad \tyPrim{real}{1}\times(\forall
+s\mathord:\SynGL{1}.\tyPrim{real}{s^2})\\ \quad\cong\quad
 \tyPrim{real}{1}\times 1\quad\cong\quad \tyPrim{real}{1}
@@
 The abstraction theorem is used to prove
-$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s}\cong 1$.
+$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s^2}\cong 1$.
 % is proved using
@@
 \newcommand{\Grp}{\mathit{Gr}}
+To prove non-definability results, the model and relational interpretation
+used in the previous two sections are not sufficient.
+This is true even for simple scalars with invariance under scaling.
+Consider the type 
+$\forall s \mathord: \SynGL{1}.\tyPrim{real}{s^2} \to \tyPrim{real}{s}$. 
+There are many functions $f$ in $\mathbb R \to \mathbb R$
+that satisfy its relational interpretation, i.e. for which $f(k^2\cdot x) = k\cdot f(x)$ for 
+any $k\neq 0$.
+Consider 
+\[
+f(x) = \begin{cases}\sqrt x & \text{if }x > 0\\0&\text{otherwise}\end{cases}
+\]
+for instance. Yet this type contains only the constant zero function, a fact that
+we can prove using a surprisingly simple model!
+
+\begin{example}
+If  $\opsCtxt \vdash f : \forall s \mathord: \SynGL{1}.\
+\tyPrim{real}{s^2} \to \tyPrim{real}{s}$ 
+then $\tmSem f = \lambda x.0$.
+To show this, take the model
+\[
+\begin{array}{l}
+\semSort{\SynGL{1}} \isDefinedAs \{0,1\}\qquad
+\semIndexExp{1} \isDefinedAs 0 \qquad
+\semIndexExp{{}^{-1}}  \isDefinedAs \mathrm{id}\\
+\semIndexExp{\cdot} \isDefinedAs \lambda(x,y).(x+y)\mathrel{\mathrm{mod}} 2
+\qquad
+\semIndexExp{|-|} \isDefinedAs \mathrm{id},
+
+\end{array}
+\]
+in other words, the additive abelian group $(\mathbb Z_2, +)$. 
+Now set
+\[
+\semPrimType{\tyPrimNm{real}}(z) = 
+  \begin{cases} \Delta_{\mathbb R} & \text{if }z=0\\
+                      {(0,0)} & \text{if }z=1
+\end{cases}
+\]
+It's easy to check that $(\opsEnv,\opsEnv)\in\rsem{\opsCtxt}*$.
+
+You can think of $z$ as tracking whether or not exponents on scale
+parameters are divisible by 2: all the primitive operations in
+\autoref{fig:real-ops} produce results with even exponents if their
+inputs have even exponents. But the type under consideration here does
+not.
+
+By \autoref{thm:abstraction} we have that $(f,f)\in\rsem{\forall s
+  \mathord: \SynGL{1}.\ \tyPrim{real}{s^2} \to \tyPrim{real}{s}} *$.
+Expanding the definitions, this means that for any
+$z\in\semSort{\SynGL{1}}$ and any
+$(x,y)\in\semPrimType{\tyPrimNm{real}}((z+z)\mathrel\textrm{mod}2)$, it is the case that
+$(f(x),f(y))\in\semPrimType{\tyPrimNm{real}}(z)$. Set $z = 1$.  
+Then we have that if $x=y$ then $f(x)=f(y)=0$.
+\qed
+\end{example}
+
+The particular model and relational interpretation used here is devised only to
+rule out the `square root type' (it does not rule out a `cube root type', for instance); 
+nevertheless, it is possible to devise craftier interpretations that serve up both invariance
+and more general non-definability results. For example, we can set
+\begin{align*}
+\semSort{\SynGL{1}} & \isDefinedAs \mathbb \mathbb \GL{1} \times \mathbb Q &
+\semIndexExp{\cdot} & \isDefinedAs (\cdot,+) & 
+\semIndexExp{1} & \isDefinedAs (1,0) \\
+\semIndexExp{{}^{-1}} & \isDefinedAs \mathrm{({}^{-1},-)} &
+\semIndexExp{|\cdot|} & \isDefinedAs \mathrm{(|\cdot|,\mathrm{id})}
+\end{align*}
+and
+\[
+\semPrimType{\tyPrimNm{real}}(k,q) = 
+  {\begin{cases} \{(x,k x) \sepbar x \in \mathbb R \} & \text{if }q \in \mathbb Z\\
+                      {(0,0)} & \text{otherwise}
+  \end{cases}}
+\]
+
+\begin{example}
+Suppose we extend the operations $\opsCtxt$ of \autoref{fig:real-ops}
+and corresponding $\opsEnv$ with a square root operation typed by
+\[
+\mathrm{sqrt} : \forall s\mathord: \SynGL{1}.\tyPrim{real}{s^2} \to \tyPrim{real}{s}.
+\]
+If  $\opsCtxt \vdash f : \forall s \mathord: \SynGL{1}.\
+\tyPrim{real}{s^3} \to \tyPrim{real}{s}$ 
+then $\tmSem f = \lambda x.0$.
+To show this, take the model
+\begin{align*}
+\semSort{\SynGL{1}} & \isDefinedAs \mathbb Q &
+\semIndexExp{\cdot} & \isDefinedAs + & 
+\semIndexExp{1} & \isDefinedAs 0 &
+\semIndexExp{{}^{-1}} & \isDefinedAs \mathrm{-} &
+\semIndexExp{|\cdot|} & \isDefinedAs \mathrm{id},
+\end{align*}
+in other words, the additive abelian group $(\mathbb Q, +)$. 
+Now set
+\[
+\semPrimType{\tyPrimNm{real}}(q) =
+\begin{cases} \Delta_{\mathbb R} & \text{if }\exists n, 2^n q \in\mathbb Z\\
+\{(0,0)\} & \text{otherwise}
+\end{cases}
+\]
+It's easy to check that $(\opsEnv,\opsEnv)\in\rsem{\opsCtxt}*$.
+
+Here $q$ is tracking the exponents on scale parameters, with
+$\semPrimType{\tyPrimNm{real}}(q)$ accepting only the zero value
+unless the denominator of $q$ is a power of two (a so-called
+\emph{dyadic} number).  
+Applying the abstraction theorem and expanding, we have that for
+any
+$q\in\mathbb Q$ and any
+$(x,y)\in\semPrimType{\tyPrimNm{real}}(3q)$, it is the case that
+$(f(x),f(y))\in\semPrimType{\tyPrimNm{real}}(q)$. Set $q = \frac 1 3$.  
+Then we have that if $x=y$ then $f(x)=f(y)=0$.
+\qed
+\end{example}
 
@@
 \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ from
-\exref{ex:uninhabited-type} is an instance of this theorem.  For types
-indexed by the scaling group $\SynGL{1}$ the situation is more subtle,
-as we have assumed that the type $\tyPrim{real}{s}$ has a polymorphic zero
-(see \autoref{fig:real-ops}). We can show, nonetheless, that $\forall
-s \mathord: \SynGL{1}.\ \tyPrim{real}{s^2} \to \tyPrim{real}{s}$ is
-only inhabited by the constantly zero function by setting
-$R^\bullet_{\tyPrimNm{real}} = \{(0,0)\}$. If we augment our sort
-$\SynGL{1}$ with a square root operation $s^{1/2}$, then it is
-possible to add a primitive square root function and still be able to
-prove that the cube root type $\forall s \mathord: \SynGL{1}.\
-\tyPrim{real}{s^3} \to \tyPrim{real}{s}$ is only inhabited by the
-constantly zero function. This is a consequence of the impossibility
-of generating $s$ from $s^3$ with only the abelian group operations
-and square root.
-
+\exref{ex:uninhabited-type} is an instance of this theorem.  
 %\begin{proof}

### 2012-10-22 paper/section-info-flow.tex
Update the Logical Information Flow section to the new notation and rel. envs.

@@
 $\tyPrim{T}{\phi}$ have no run-time content; their only meaning is
-given by the relational semantics. For the relational environments
-$\relEnv{E}_\infoflow$, we use the model-based construction in
-\autoref{sec:constr-rel-env}, with an arbitrary boolean algebra $L$
-and $R_{\tyPrimNm{T}}(x) = \{ (*,*) \sepbar x = \top \}$.  The
-primitive operations $\Gamma_\infoflow$ reflect logical consequence:
+given by the relational semantics. For the model of the indexing
+theory, we take an arbitrary boolean algebra $L$. The relational
+interpretation of the truth representation type is
+$\rsem{\tyPrimNm{T}}(x) = \{ (*,*) \sepbar x = \top \}$, where $\top$
+is the top element of the boolean algebra $L$. The primitive
+operations $\Gamma_\infoflow$ reflect logical consequence:
 \begin{displaymath}
@@
 \begin{lemma}\label{lem:environments-information-flow}
-  For all $\Delta$ and $\rho \in \relEnv{E}_\infoflow$, $(\eta_\infoflow, \eta_\infoflow) \in \rsem{\Gamma_\infoflow}{\relEnv{E}_\infoflow}{\rho}$.
+  For all $\Delta$ and $\rho \in \rsem{\Delta}$, $(\eta_\infoflow, \eta_\infoflow) \in \rsem{\Gamma_\infoflow}\rho$.
 \end{lemma}
@@
 then for all $x,x' \in \tySem{T_\phi\tyPrimNm{bool}}$, we have $(x,x')
-\in \rsem{T_\phi\tyPrimNm{bool}}{\relEnv{E}_\infoflow}{\rho}$. Thus if
-a principal is dissatisfied (i.e.,~$\phi \not= \top$), then data
-protected by this principal is indistinguishable from any other
-data, and a program cannot get access to the exact value. From this
-and \thmref{thm:abstraction}, we obtain the following information flow
+\in \rsem{T_\phi\tyPrimNm{bool}}{\rho}$. Thus if a principal is
+dissatisfied (i.e.,~$\phi \not= \top$), then data protected by this
+principal is indistinguishable from any other data, and a program
+cannot get access to the exact value. From this and
+\thmref{thm:abstraction}, we obtain the following information flow
 result:

### 2012-10-22 paper/section-metric.tex
Update the distance-indexed types section to new rel. envs.

@@
   \begin{array}{l}
-    \textrm{if }|x - x'| < \epsilon_1 \textrm{ and } |y - y'| < \epsilon_2 \textrm{ then}\\
-    \hspace{2em}|\tmSem{M}\ \eta_\Metric\ x\ y - \tmSem{M}\ \eta_\Metric\ x'\ y'| < \epsilon_1 + \epsilon_2
+    \textrm{if }\abs{x - x'} < \epsilon_1 \textrm{ and } \abs{y - y'} < \epsilon_2 \textrm{ then}\\
+    \hspace{2em}\abs{\tmSem{M}\ \eta_\Metric\ x\ y - \tmSem{M}\ \eta_\Metric\ x'\ y'} < \epsilon_1 + \epsilon_2
   \end{array}
@@
 
-For the families of sets of relational environments
-$\relEnv{E}_\Metric$, we use the model-based construction in
-\autoref{sec:constr-rel-env}, interpreting the primitive type
-$\tyPrimNm{real}$ with $R_{\tyPrimNm{real}}(\epsilon) = \{ (x,x')
-\sepbar |x - x'| < \epsilon\}$.
-%\begin{displaymath} %FIXME: define \Eq earlier on
-%  \begin{array}{l@{\hspace{2em}}l}
-%    R^\bullet_{\tyPrimNm{real}} = \Eq_\mathbb{R} &
-%    R_{\tyPrimNm{real}}(\epsilon) = \{ (x,x') \sepbar |x - x'| < \epsilon \}
-%  \end{array}
-%\end{displaymath}
+For the relational intepretation we construct a model of the indexing
+theory by interpreting $\metricSort$ with strictly positive real
+numbers. We set $\rsem{\tyPrimNm{real}}(\epsilon) = \{ (x,x') \sepbar
+\abs{x - x'} < \epsilon \}$.
 \begin{lemma}\label{lem:metric-environments}
-  For all $\Delta$ and $\rho \in \relEnv{E}_\Metric(\Delta)$,
-  $(\eta_\Metric, \eta_\Metric) \in
-  \rsem{\Gamma}{\relEnv{E}_\Metric}{\rho}$.
+  For all $\Delta$ and $\rho \in \rsem{\Delta}$, $(\eta_\Metric,
+  \eta_\Metric) \in \rsem{\Gamma}{\rho}$.
 \end{lemma}
@@
 \begin{displaymath}
-  \forall \epsilon \mathord> 0. \exists \delta \mathord> 0. \forall x,x'. |x - x'| < \delta \Rightarrow |\tmSem{M}\ x - \tmSem{M}\ x'| < \epsilon
+  \forall \epsilon \mathord> 0. \exists \delta \mathord> 0. \forall x,x'. abs{x - x'} < \delta \Rightarrow \abs{\tmSem{M}\ x - \tmSem{M}\ x'} < \epsilon
 \end{displaymath}

### 2012-10-22 paper/section-instantiations.tex
Isomorphism example

@@
 \autoref{sec:type-system-geom-intro} and 
-\autoref{sec:affine-vector-ops}. For any quantifier-free type $A$
-we have: % the isomorphism:
-\begin{eqnarray*}
-\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{t}\to A
-& \cong &
-A[0/t]
-\end{eqnarray*}
+\autoref{sec:affine-vector-ops}.
+
+\begin{example}[Geometry: change of origin]\label{ex:simpletrans}
+We prove the isomorphism
+$\Delta\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to\tyPrim{vec}{e,t}) \cong
+\tyPrim{vec}{e,0}$ where $e$ is an expression of sort $\SynGL{2}$ not containing $t$.
+Let 
+\begin{align*}
+X & \isDefinedAs\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to\tyPrim{vec}{e,t}\\
+Y & \isDefinedAs\tyPrim{vec}{e,0}\\
+I & \isDefinedAs\lambda f\mathord:X.f_0\,0_0\\
+J & \isDefinedAs\lambda v\mathord:Y.\Lambda t\mathord:\SynTransl{2}.\lambda w\mathord:\tyPrim{vec}{e,t}.w +_{e,t,0} v
+\end{align*}
+Unfolding definitions gives that $\tmSem{I(J(v))}(v,\opsEnv)$ = $0+v$
+for any $v\in\mathbb R ^ 2$.  Because $0$ is the identity for vector
+addition it follows that $\ctxteq{\Delta}{v\mathord:Y}{I(J(v))}{v}Y$.
+To show $\ctxteq{\Delta}{f\mathord:X}{J(I(f))}{f}{X}$ we appeal to
+\autoref{thm:soundness} and reason using the relational semantics.  It
+suffices to show $\semeq{\Delta}{f\mathord:X}{J(I(f))}{f}{X}$; that
+is, given $\rho\in\semSort\Delta$, for any $f$ and $f'$ with
+$(f,f')\in\rsem X\rho$, it is the case that 
+$(\tmSem{I(J(f))}(f,\opsEnv), f')\in\rsem X\rho$.
+Expanding the premise, and setting $B = \semIndexExp{e}\rho$, we have
+\begin{eqnarray}\label{eqn:transpremise}
+\forall t\in \mathbb R ^ 2,
+\forall v\in \mathbb R ^ 2,
+f'(Bv + t) = B(f(v)) + t
+\end{eqnarray}
+Expanding the conclusion, we have to show that for any
+$t'\in\mathbb R^2$ and $v'\in\mathbb R^2$, it is the case that
+$f'(Bv' + t') = B(v' + f(0)) + t'$.
+By instantiating \autoref{eqn:transpremise} thrice, with~$v=t=0$, with~$v=v'$ and~$t=t'$,
+and with (crucially)~$v=v'$ and~$t=-Bv'$, and applying a little algebra, we obtain
+the desired result.
+\qed
+\end{example}
+
+We can generalise this isomorphism to any quantifier-free (but possibly open) type
+of the form $\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A$. 
+To be precise, we have the following isomorphism:
+\[
+\Delta\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A)
+\cong A[0/t]
+\]
 where $A[0/t]$ denotes substitution of the identity translation $0$
@@
 
-\exref{ex:type-iso} is a special case of this isomorphism; another instance is the
+\exref{ex:type-iso} and \exref{ex:simpletrans} are special cases of this isomorphism; another instance is the
 type of vector addition:
-\begin{eqnarray*}
-\forall t_1\mathord:\SynTransl{2}.\forall t_2\mathord:\SynTransl{2}.\tyPrim{vec}{t_1}\to\tyPrim{vec}{t_2}\to\tyPrim{vec}{t_1+t_2}\\
+\[
+\begin{array}{l}
+\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}, t_2\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_2}\to\tyPrim{vec}{B,t_1+t_2}\\
 \cong\quad
-\forall t_1\mathord:\SynTransl{2}.\tyPrim{vec}{t_1}\to\tyPrim{vec}{t_1}\quad\cong\quad
-\tyPrim{vec}{0}
-\end{eqnarray*}
+\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_1}\\
+\cong\quad
+\forall B\mathord:\SynGL{2}.\tyPrim{vec}{B,0}
+\end{array}
+\]
 \noindent
-\newcommand{\transup}[1]{\mathord\uparrow_{#1}}
-\newcommand{\transdn}[1]{\mathord\downarrow_{#1}}
+\newcommand{\transup}[3]{\mathord\uparrow_{#1}^{#2\mathord:#3}}
+\newcommand{\transdn}[3]{\mathord\downarrow_{#1}^{#2\mathord:#3}}
 The maps associated with the isomorphism are 
-$I = \lambda x. x\ [0]\ 0$ and $J = \lambda y. \Lambda t.\lambda v. \transup A (y)$,
+$I = \lambda f. f_0\ 0_0$ and $J = \lambda y. \Lambda t.\lambda v. \transup A t v (y)$,
 %\begin{eqnarray*}
@@
 %\end{eqnarray*}
-where terms $\transdn A : A\to A[0/t]$ and $\transup A : A[0/t] \to A$ are defined by induction over the syntax of $A$ (cases omitted for $\transup A$ have definition symmetric to $\transdn A$):
+where terms $\transdn A t v : A\to A[0/t]$ and $\transup A t v : A[0/t] \to A$ are defined by induction over the syntax of $A$ (cases omitted for $\transup A t v$ have definition symmetric to $\transdn A t v$):
 \begin{align*}
-\transdn{\tyPrimNm{unit}} &= \lambda x. x \qquad
-\transdn{A_1\times A_2} = \lambda p. (\transdn{A_1}(\pi_1 p),\transdn{A_2}(\pi_2 p))\\ 
-\transdn{A_1 + A_2} &= \lambda x.
-\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn{A_1}y); \textrm{inr}\ y.\textrm{inr}(\transdn{A_2}y)\\ 
-\transdn{A_1\to A_2} &= \lambda f. \lambda x. \transdn{A_2} (f(\transup{A_1}x))\\ 
-\transdn{\tyPrim{vec}{e}} &= \lambda x. x - kv \qquad
-\transup{\tyPrim{vec}{e}} = \lambda x. x + kv\\ 
+\transdn{\tyPrimNm{unit}} t v &= \lambda x. x \qquad
+\transdn{A_1\times A_2} t v = \lambda p. (\transdn{A_1} t v(\pi_1 p),\transdn{A_2}t v(\pi_2 p))\\ 
+\transdn{A_1 + A_2} t v &= \lambda x.
+\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn {A_1}t v y); \textrm{inr}\ y.\textrm{inr}(\transdn {A_2} t v y)\\ 
+\transdn{A_1\to A_2} t v &= \lambda f. \lambda x. \transdn{A_2} t v (f(\transup {A_1} t v x))\\ 
+\transdn{\tyPrim{vec}{e}} t v &= \lambda x. x - kv \qquad
+\transup{\tyPrim{vec}{e}} t v = \lambda x. x + kv\\ 
 \text{ if }e & \equiv k t + e' \text{ for
@@
 equivalent to $x$ we must apply the abstraction theorem and use
-the fact that $\tmSem{\transup A}$ and $\tmSem{\transdn A}$ 
+the fact that $\tmSem{\transup A t v}$ and $\tmSem{\transdn A t v}$ 
 considered as relations are simply instances of $\rsem A{}{}$.
@@
   Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matrices of
-  natural numbers $A$ and $m$-by-$1$ matricies of integers $\vec{b}$,
+  natural numbers $A$ and $m$-by-$1$ matrices of integers $\vec{b}$,
   there exists a quantifier-free program $M$ such that:

### 2012-10-23 paper/section-instantiations.tex
Continued work on general isomorphism

@@
 \begin{align*}
-X & \isDefinedAs\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to\tyPrim{vec}{e,t}\\
+X & \isDefinedAs\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to\tyPrim{vec}{e,t}
+&
 Y & \isDefinedAs\tyPrim{vec}{e,0}\\
-I & \isDefinedAs\lambda f\mathord:X.f_0\,0_0\\
-J & \isDefinedAs\lambda v\mathord:Y.\Lambda t\mathord:\SynTransl{2}.\lambda w\mathord:\tyPrim{vec}{e,t}.w +_{e,t,0} v
+I & \isDefinedAs\lambda f\mathord:X.f\,[0]\,(0\,[e])\\
+J & \isDefinedAs\lambda v\mathord:Y.\Lambda t\mathord:\SynTransl{2}.\lambda w\mathord:\tyPrim{vec}{e,t}.v +[e,0,t] w
 \end{align*}
-Unfolding definitions gives that $\tmSem{I(J(v))}(v,\opsEnv)$ = $0+v$
+Unfolding definitions gives that $\tmSem{I(J(v))}(v,\opsEnv)$ = $v+0$
 for any $v\in\mathbb R ^ 2$.  Because $0$ is the identity for vector
@@
 $(f,f')\in\rsem X\rho$, it is the case that 
-$(\tmSem{I(J(f))}(f,\opsEnv), f')\in\rsem X\rho$.
+$(\tmSem{J(I(f))}(f,\opsEnv), f')\in\rsem X\rho$.
 Expanding the premise, and setting $B = \semIndexExp{e}\rho$, we have
@@
 \begin{array}{l}
-\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}, t_2\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_2}\to\tyPrim{vec}{B,t_1+t_2}\\
+\forall B\mathord:\SynGL{2}, t_1
+%\mathord:\SynTransl{2}
+, t_2\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_2}\to\tyPrim{vec}{B,t_1\mathord+t_2}\\
 \cong\quad
@@
 \newcommand{\transdn}[3]{\mathord\downarrow_{#1}^{#2\mathord:#3}}
-The maps associated with the isomorphism are 
-$I = \lambda f. f_0\ 0_0$ and $J = \lambda y. \Lambda t.\lambda v. \transup A t v (y)$,
-%\begin{eqnarray*}
-%I = \lambda x. x\ [0]\ 0 & \text{ and } &
-%J = \lambda y. \Lambda t.\ \lambda v. \trans A v t(y)
-%\end{eqnarray*}
-where terms $\transdn A t v : A\to A[0/t]$ and $\transup A t v : A[0/t] \to A$ are defined by induction over the syntax of $A$ (cases omitted for $\transup A t v$ have definition symmetric to $\transdn A t v$):
+
+\begin{example}
+Let 
 \begin{align*}
-\transdn{\tyPrimNm{unit}} t v &= \lambda x. x \qquad
-\transdn{A_1\times A_2} t v = \lambda p. (\transdn{A_1} t v(\pi_1 p),\transdn{A_2}t v(\pi_2 p))\\ 
+X & \isDefinedAs \forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A
+& 
+Y & \isDefinedAs A[0/t]\\
+I & \isDefinedAs \lambda f\mathord: X. f\ [0]\ (0\ [e])
+& 
+J & \isDefinedAs \lambda y\mathord: Y. \Lambda t.\lambda v\mathord:\tyPrim{vec}{e,t}. \transup A t v (y)
+\end{align*}
+where terms 
+\begin{align*}
+&
+\Delta,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{e,t}\vdash \transup A t v : A[0/t] \to A
+\\\text{and } &
+\Delta,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{e,t}\vdash \transdn A t v : A\to A[0/t]
+\end{align*}
+are defined by induction over the syntax of $A$. For brevity we omit types on binders
+and note that the omitted cases for $\transup A t v$ 
+have definition symmetric to $\transdn A t v$:
+\begin{align*}
+\transdn{\tyPrimNm{unit}} t v &= \lambda x. x
+\qquad 
+\transdn{\tyPrim{real}{e}} t v = \lambda x. x
+\\
+\transdn{A_1\times A_2}t v &= \lambda p. (\transdn{A_1}t v(\pi_1 p),\transdn{A_2}t v(\pi_2 p))
+\\
 \transdn{A_1 + A_2} t v &= \lambda x.
-\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn {A_1}t v y); \textrm{inr}\ y.\textrm{inr}(\transdn {A_2} t v y)\\ 
-\transdn{A_1\to A_2} t v &= \lambda f. \lambda x. \transdn{A_2} t v (f(\transup {A_1} t v x))\\ 
-\transdn{\tyPrim{vec}{e}} t v &= \lambda x. x - kv \qquad
-\transup{\tyPrim{vec}{e}} t v = \lambda x. x + kv\\ 
-\text{ if }e & \equiv k t + e' \text{ for
-  some }k\in\setOfIntegers \text{ and }e'\text{ with }t\notin e'
+\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn {A_1}t v y); \textrm{inr}\ y.\textrm{inr}(\transdn {A_2} t v y)
+\\ 
+\transdn{A_1\to A_2} t v &= \lambda f. \lambda x. \transdn{A_2} t v (f(\transup {A_1} t v x))
+\end{align*}
+\begin{align*}
+\transdn{\tyPrim{vec}{B,e}} t v &= \lambda x. \transdn{e}  t v (x) 
+&
+\transup{\tyPrim{vec}{B,e}} t v &= \lambda x. \transup{e} t v (x)
+\\
+\transdn{0} t v &= \lambda w.w
+&
+\transup{0} t v &= \lambda w.w
+\\ 
+\transdn{e_1+e_2} t v &= \lambda w.\transdn{e_1} t v(\transdn{e_2} t v(w))
+&
+\transup{e_1+e_2} t v &= \lambda w.\transup{e_1} t v(\transup{e_2} t v(w))
+\\
+\transdn{-e} t v &= \lambda w.\transup{e} t v (w)
+&
+\transup{-e} t v &= \lambda w.\transdn{e} t v (w)
+\\
+\transdn{t} t v &= \lambda w.w + v 
+&
+\transup{t} t v &= \lambda w.w - v
+\\
+\transdn{t'} t v &= \lambda w.w\quad(t'\neq t)
+&
+\transup{t'} t v &= \lambda w.w\quad(t'\neq t)
 \end{align*}
@@
 considered as relations are simply instances of $\rsem A{}{}$.
+\qed
 %\[
@@
 %the abstraction theorem.
-
+\end{example}
 %\begin{enumerate}

### 2012-10-23 paper/paper.tex
Final small changes
New framework setup
More cleanup
acknowledgements, popl copyright etc

@@
-\documentclass[preprint]{sigplanconf}
+\documentclass{sigplanconf}
 
@@
 \newcommand{\relEnv}[1]{\mathcal{#1}}
-\newcommand{\tySem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}}}
-\newcommand{\ctxtSem}[1]{\llbracket #1 \rrbracket^{\mathcal{C}}}
-\newcommand{\tmSem}[1]{\llbracket #1 \rrbracket^{\mathit{tm}}}
-\newcommand{\tyPrimSem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}_0}}
-\newcommand{\rsem}[3]{\llbracket #1 \rrbracket^{\mathcal{R}}_{#2}{#3}}
+%\newcommand{\tySem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}}}
+%\newcommand{\ctxtSem}[1]{\llbracket #1 \rrbracket^{\mathcal{C}}}
+%\newcommand{\tmSem}[1]{\llbracket #1 \rrbracket^{\mathit{tm}}}
+%\newcommand{\tyPrimSem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}_0}}
+\newcommand{\tySem}[1]{\left| #1\right|}
+\newcommand{\ctxtSem}[1]{\left| #1\right|}
+\newcommand{\tmSem}[1]{\left| #1\right|}
+\newcommand{\tyPrimSem}[1]{\left| #1\right|}
+\newcommand{\rsem}[1]{\llbracket #1 \rrbracket}
 \newcommand{\extends}[2]{\mathsf{ext}(#1,#2)}
@@
 \newtheorem{example}{Example}
+\newtheorem*{example*}{Example}
 
@@
 \begin{document}
+
+\conferenceinfo{POPL'13,} {January 23--25, 2013, Rome, Italy.}
+\CopyrightYear{2013}
+\copyrightdata{978-1-4503-1832-7/13/01}
 
@@
 
+\section*{Acknowledgements}
+The authors would like to thank Nick Benton, John Reppy and Martin Will for 
+illuminating discussions on geometry and types.
+
 \begin{thebibliography}{12}
@@
 J.-P.~Bernardy, P.~Jansson, R.~Paterson.
-Proofs for free: Parametricity for dependent types.
+Proofs for Free: Parametricity for Dependent Types.
 {\em Journal of Functional Programming} 22(2), pp. 107-152, 2012.
@@
 \bibitem{hofmann08correctness}
-M.~Hofmann. Correctness of effect-based program
-transformations. Formal Logical Methods for System Security and
+M.~Hofmann. Correctness of Effect-based Program
+Transformations. Formal Logical Methods for System Security and
 Correctness, pp. 149-173, 2008.

### 2012-10-23 paper/section-instantiations.tex
Revamped section on general non-definability results

@@
 
-\begin{example}
+\begin{example}\label{ex:sqrt-root-nondef}
 If  $\opsCtxt \vdash f : \forall s \mathord: \SynGL{1}.\
@@
 
-\begin{example}
+\begin{example}\label{ex:cube-root-nondef}
 Suppose we extend the operations $\opsCtxt$ of \autoref{fig:real-ops}
@@
 
-For proving non-definability results, we consider the general case of
-a sort $\mathsf{G}$ with the equational theory of abelian groups. We
-assume a group-indexed primitive type $\tyPrimNm{G}$ and set of
-primitive operations typed by $\Gamma_\Grp$ that include the group
-identity $1:\tyPrim{G}{1}$, operation $(\cdot): \forall
-a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to
-\tyPrim{G}{a\cdot b}$ and inverse $\ ^{-1}: \forall
-a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$. The
-context $\Gamma_\Grp$ may also include members with types of the form
-$\forall \vec{a}\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to ...
-\tyPrim{G}{e_n} \to \tyPrim{G}{e}$ as long as the result index $e$ is
-generated by the set $\{e_1,...,e_n\}$. One instance is the
-sort~$\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ on
-$\tyPrim{vec}{t}$. For the purposes of non-definability results, we
-may choose an arbitrary index-erasure semantics with $\eta_\Grp \in
-\ctxtSem{\Gamma_\Grp}$.
+% For proving non-definability results, we consider the general case of
+% a sort $\mathsf{G}$ with the equational theory of abelian groups. We
+% assume a group-indexed primitive type $\tyPrimNm{G}$ and set of
+% primitive operations typed by $\Gamma_\Grp$ that include the group
+% identity $1:\tyPrim{G}{1}$, operation $(\cdot): \forall
+% a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to
+% \tyPrim{G}{a\cdot b}$ and inverse $\ ^{-1}: \forall
+% a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$. The
+% context $\Gamma_\Grp$ may also include members with types of the form
+% $\forall \vec{a}\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to ...
+% \tyPrim{G}{e_n} \to \tyPrim{G}{e}$ as long as the result index $e$ is
+% generated by the set $\{e_1,...,e_n\}$. One instance is the
+% sort~$\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ on
+% $\tyPrim{vec}{t}$. For the purposes of non-definability results, we
+% may choose an arbitrary index-erasure semantics with $\eta_\Grp \in
+% \ctxtSem{\Gamma_\Grp}$.
 
-We use the construction of definability relational environments
-from \autoref{sec:constr-rel-env}, with $R^\bullet_{\tyPrimNm{G}} =
-\emptyset$, to give a pre-substitutive family $\relEnv{E}_\Grp$. This
-gives us the following lemma and theorem.
+% We use the construction of definability relational environments
+% from \autoref{sec:constr-rel-env}, with $R^\bullet_{\tyPrimNm{G}} =
+% \emptyset$, to give a pre-substitutive family $\relEnv{E}_\Grp$. This
+% gives us the following lemma and theorem.
 
-\begin{lemma}\label{lem:geom-environments-2}
-  For all $\Delta$ and $\rho \in \relEnv{E}_\Grp(\Delta)$,
-  $(\eta_\Grp,\eta_\Grp) \in \rsem{\Gamma_\Grp}{\relEnv{E}_\Grp}\rho$.
-\end{lemma}
+% \begin{lemma}\label{lem:geom-environments-2}
+%   For all $\Delta$ and $\rho \in \relEnv{E}_\Grp(\Delta)$,
+%   $(\eta_\Grp,\eta_\Grp) \in \rsem{\Gamma_\Grp}{\relEnv{E}_\Grp}\rho$.
+% \end{lemma}
 
@@
 
-Writing $i^n$ for $n$-fold iteration of the group operation, we have:
-\begin{theorem}
-  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matrices of
-  natural numbers $A$ and $m$-by-$1$ matrices of integers $\vec{b}$,
-  there exists a quantifier-free program $M$ such that:
-  \begin{displaymath}
-    i_1,\dots,i_m\mathord:\mathsf{G}; \Gamma_\Grp \vdash M : \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
-  \end{displaymath}
-  where $e_j = i_1^{A_{j1}}\cdots i_m^{A_{jm}}$ and $e = i_1^{b_1}\cdots i_m^{b_m}$, iff there is a solution $\vec{x}$
-  in the integers to the equation $A \vec x = \vec{b}$.
-\end{theorem}
-The uninhabited type $\forall t \mathord: \SynTransl{2}.\
-\tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ from
-\exref{ex:uninhabited-type} is an instance of this theorem.  
+% Writing $i^n$ for $n$-fold iteration of the group operation, we have:
+% \begin{theorem}
+%   Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matrices of
+%   natural numbers $A$ and $m$-by-$1$ matrices of integers $\vec{b}$,
+%   there exists a quantifier-free program $M$ such that:
+%   \begin{displaymath}
+%     i_1,\dots,i_m\mathord:\mathsf{G}; \Gamma_\Grp \vdash M : \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
+%   \end{displaymath}
+%   where $e_j = i_1^{A_{j1}}\cdots i_m^{A_{jm}}$ and $e = i_1^{b_1}\cdots i_m^{b_m}$, iff there is a solution $\vec{x}$
+%   in the integers to the equation $A \vec x = \vec{b}$.
+% \end{theorem}
+% The uninhabited type $\forall t \mathord: \SynTransl{2}.\
+% \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ from
+% \exref{ex:uninhabited-type} is an instance of this theorem.  
 %\begin{proof}

### 2012-10-23 paper/geometry-twocolumn.tex
Update the geometry section

@@
 rich with operations that are invariant under transformation: affine
-operations are invariant under change of origin, vector space
-operations are invariant under change of basis, and dot product is
-invariant under orthogonal changes of basis. On the other hand, some
+operations are invariant under change of origin
+(\autoref{sec:affine-vector-ops}), vector space operations are
+invariant under change of basis, and dot product is invariant under
+orthogonal changes of basis
+(\autoref{sec:motivation-generalising}). On the other hand, some
 geometric operations are interestingly variant under transformation.
-For example, cross products vary with scalings of the plane. We
-incorporate (in)variance information about geometric primitives
-into type systems via algebraically indexed types.
+For example, cross products vary with scalings of the plane
+(\autoref{sec:scale-invariance}). We incorporate (in)variance
+information about geometric primitives into type systems via
+algebraically indexed types.
 
@@
 index-erasure interpretation. To give such interpretations for the
-types $\tyPrim{vec}{e}$ and $\forall t\mathord:\SynTransl{2}.A$,
-we assume for now that we can map each expression $e$ of sort
+types $\tyPrim{vec}{e}$ and $\forall t\mathord:\SynTransl{2}.A$, we
+assume for now that we can map each expression $e$ of sort
 $\SynTransl{2}$ to an element $\sem{e}\rho$ of the group $\Transl{2}$
 using some environment $\rho$ that interprets $e$'s free
-variables. The (simplified) index-erasure and relation interpretations
-of $\tyPrim{vec}{e}$ are:
+variables. The index-erasure and relational interpretations of
+$\tyPrim{vec}{e}$ are:
 \begin{displaymath}
@@
     \\ \rsem{\forall t\mathord:\SynTransl{2}.A}{}\rho & \bigcap\{
-    \rsem{A}{}{\rho'} \sepbar \rho' \in
-    \extends{\rho}{t\mathord:\SynTransl{2}} \}
+    \rsem{A}{}{(\rho,\vec{t})} \sepbar \vec{t} \in \Transl{2} \}
   \end{array}
 \end{displaymath}
-The exact definition of $\extends{\rho}{t\mathord:\SynTransl{2}}$
-depends on which (possibly non-compositional) interpretation of
-relational environments we use. For now we can think of environments
-as mappings from free variables of sort $\SynTransl{2}$ to the group
-$\Transl{2}$, and of extension as adding extra components to
-mappings. The index-erasure and relational interpretations are given
-formally in Sections~\ref{sec:index-erasure-semantics}
+The index-erasure and relational interpretations are given
+formally in Sections~\ref{sec:erasure-semantics}
 %the relational interpretation is in general more complex, due to the
@@
 $\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$, where $\lambda_1 +
-\lambda_2 = 1$.  Geometrically, this can be interpreted as
-interpolation between the points represented by $\vec{v_1}$ and
-$\vec{v_2}$.  We add affine combination of points to our calculus as
+\lambda_2 = 1$.  Geometrically, this can be interpreted as describing
+all the points on the unique line through the points represented by
+$\vec{v_1}$ and $\vec{v_2}$ (assuming $\vec{v_1} \not= \vec{v_2}$).
+We add affine combination of points to our calculus as
 follows: % typing and intended denotation:for
@@
   \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ has no
-  quantifier-free inhabitants\footnote{We define quantifier-free
-    programs in the statement of
-    \thmref{thm:abstraction}.}. Intuitively, this is because we cannot
-  remove the extra $t$ in $\tyPrim{vec}{t+t}$ using the
-  vector operations. We formalise this indefinability result in
-  \autoref{sec:types-indexed-abelian-groups-indef} using a more
-  sophisticated relational interpretation of free translation
-  variables than the one sketched above.
+  inhabitants. Intuitively, this is because we cannot remove the extra
+  $t$ in $\tyPrim{vec}{t+t}$ using the vector operations. We formalise
+  this non-definability result in \autoref{sec:general-nondef} using a
+  specialised relational interpretation.
 \end{example}
@@
 preservation of relations indexed by changes of basis. Change of basis
-is achieved by applying an invertible linear map, and the collection of
-all such maps on $\mathbb{R}^2$ forms the \emph{General Linear} group
-$\GL{2}$ which we represent in our language by a new indexing sort
-$\SynGL{2}$ with (non-abelian) group structure that we will write
+is achieved by applying an invertible linear map, and the collection
+of all such maps on $\mathbb{R}^2$ forms the \emph{General Linear}
+group $\GL{2}$ which we represent in our language by a new indexing
+sort $\SynGL{2}$ with (non-abelian) group structure that we will write
 multiplicatively. We then extend $\tyPrimNm{vec}$ to allow indices of
@@
 $\tyPrim{vec}{B,t}$ is a vector that varies with change of basis $B$
-and change of origin $t$. Formally, the index-erasure and (simplified)
-relational semantics of $\tyPrim{vec}{B,t}$ are given by:
+and change of origin $t$. Formally, the index-erasure and relational
+semantics of $\tyPrim{vec}{B,t}$ are given by:
 \begin{displaymath}
@@
 
-\paragraph{Affine Geometry} An \emph{affine transformation} 
+\paragraph{Affine Geometry} An \emph{affine transformation} is
 an invertible linear map together with a translation. We can assign
@@
 To %incorporate the operation of dot product into our calculus, we
-must assign it a type we note that, although dot product is not
+assign it a type we note that, although dot product is not
 invariant under $\GL{2}$ or $\Transl{2}$, it
@@
 orthogonal transformations, and overload the multiplicative group
-operations for inhabitants of $\SynOrth{2}$. Further assuming a
-function $\iota_O$ that takes $e : \SynOrth{2}$ to $\iota_O(e) :
+operations for inhabitants of $\SynOrth{2}$. Further assuming an
+injection $\iota_O$ that takes $e : \SynOrth{2}$ to $\iota_O(e) :
 \SynGL{2}$ we assign dot product this type:
@@
 Since the absolute value of the determinant of an orthogonal
-transformation is always $1$, we assume $\abs{\det (\iota_O O)} = 1$ to
-hold.
+transformation is always $1$, we assume $\abs{\det (\iota_O O)} = 1$
+to hold for any $O \in \SynOrth{2}$.
 
@@
 scalings, then we can assign the following more precise types to
-scalar multiplcation and dot product:
+scalar multiplication and dot product:
 \begin{displaymath}
@@
   With the operations in \autoref{fig:real-ops}, it is not possible to
-  write a quantifier-free program with the following type that is not
-  constantly zero:
+  write a term with the following type that is not constantly zero:
   \begin{displaymath}
@@
   write a square root function with the above type.  The
-  indefinability of square root is similar to the uninhabitation of the
+  non-definability of square root is similar to the uninhabitation of the
   type in \exref{ex:uninhabited-type}.
@@
   primitive operation---with the type above---then it is still not
-  possible to construct the cube root function. The indefinability of
+  possible to construct the cube root function. The non-definability of
   cube root is related to the impossibility of trisecting an arbitrary

### 2012-10-23 paper/paper.tex
Revamped section on general non-definability results
Add Kenneth to the list of acknowledgements.

@@
 \input{section-instantiations.tex}
-\input{section-monoid.tex}
+\input{section-general-nondef.tex}
 \input{section-info-flow.tex}
@@
 \section*{Acknowledgements}
-The authors would like to thank Nick Benton, John Reppy and Martin Will for 
-illuminating discussions on geometry and types.
+The authors would like to thank Nick Benton, Kenneth MacKenzie, John
+Reppy and Martin Will for illuminating discussions on geometry and
+types.
 

### 2012-10-23 paper/paper.tex
Add the EPSRC grant number
Add reference to Shikuma and Igarashi's paper pointing out the error in Tse and Zdancewic's paper
various

@@
 \newtheorem{definition}{Definition}
-\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
-\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
-\newcommand{\defref}[1]{\hyperref[#1]{Definition \ref*{#1}}}
-\newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
-\newcommand{\corref}[1]{\hyperref[#1]{Corollary \ref*{#1}}}
-\newcommand{\conref}[1]{\hyperref[#1]{Conjecture \ref*{#1}}}
-\newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}
-\newcommand{\statementref}[1]{\hyperref[#1]{Statement \ref*{#1}}}
+\newcommand{\lemref}[1]{\hyperref[#1]{Lemma~\ref*{#1}}}
+\newcommand{\thmref}[1]{\hyperref[#1]{Theorem~\ref*{#1}}}
+\newcommand{\defref}[1]{\hyperref[#1]{Definition~\ref*{#1}}}
+\newcommand{\propref}[1]{\hyperref[#1]{Proposition~\ref*{#1}}}
+\newcommand{\corref}[1]{\hyperref[#1]{Corollary~\ref*{#1}}}
+\newcommand{\conref}[1]{\hyperref[#1]{Conjecture~\ref*{#1}}}
+\newcommand{\exref}[1]{\hyperref[#1]{Example~\ref*{#1}}}
+\newcommand{\statementref}[1]{\hyperref[#1]{Statement~\ref*{#1}}}
 
@@
   exists, exploiting invariance to yield ``free theorems'',
-  uninhabitation results, and encodings of algebraic datatypes.
+  non-inhabitation results, and encodings of algebraic datatypes.
   Outside computer science, invariance is a common theme running
@@
 Reppy and Martin Will for illuminating discussions on geometry and
-types.
+types. Atkey and Johann were supported by EPSRC grant EP/G068917/1.
 
@@
 
+\bibitem{shikuma08proving}
+N.~Shikuma and A.Igarahsi.  Proving Noninterference by a Fully
+Complete Translation to the Simply Typed lambda-calculus.  {\em
+    Logical Methods in Computer Science} 4(3), 2008.
+
 \bibitem{tse04translating}

### 2012-10-23 paper/section-instantiations.tex
various

@@
 \end{align*}
-Unfolding definitions gives that $\tmSem{I(J(v))}(v,\opsEnv)$ = $v+0$
-for any $v\in\mathbb R ^ 2$.  Because $0$ is the identity for vector
+Unfolding definitions gives that $\tmSem{I(J(v))}(\opsEnv,\vec{v})$ = $\vec{v}+(0,0)$
+for any $\vec{v}\in\mathbb R ^ 2$.  Because $(0,0)$ is the identity for vector
 addition it follows that $\ctxteq{\Delta}{v\mathord:Y}{I(J(v))}{v}Y$.
@@
 $(f,f')\in\rsem X\rho$, it is the case that 
-$(\tmSem{J(I(f))}(f,\opsEnv), f')\in\rsem X\rho$.
+$(\tmSem{J(I(f))}(\opsEnv,f), f')\in\rsem X\rho$.
 Expanding the premise, and setting $B = \semIndexExp{e}\rho$, we have
 \begin{eqnarray}\label{eqn:transpremise}
-\forall t\in \mathbb R ^ 2,
-\forall v\in \mathbb R ^ 2,
-f'(Bv + t) = B(f(v)) + t
+\forall \vec{t}\in \mathbb R ^ 2,
+\forall \vec{v}\in \mathbb R ^ 2,
+f'(B\vec{v} + \vec{t}) = B(f(\vec{v})) + \vec{t}
 \end{eqnarray}
 Expanding the conclusion, we have to show that for any
-$t'\in\mathbb R^2$ and $v'\in\mathbb R^2$, it is the case that
-$f'(Bv' + t') = B(v' + f(0)) + t'$.
-By instantiating \autoref{eqn:transpremise} thrice, with~$v=t=0$, with~$v=v'$ and~$t=t'$,
-and with (crucially)~$v=v'$ and~$t=-Bv'$, and applying a little algebra, we obtain
+$\vec{t_0}\in\mathbb R^2$ and $\vec{v_0}\in\mathbb R^2$, it is the case that
+$f'(B\vec{v_0} + \vec{t_0}) = B(\vec{v_0} + f(0)) + \vec{t_0}$.
+By instantiating \autoref{eqn:transpremise} thrice, with~$\vec{v}=\vec{t}=(0,0)$, with~$\vec{v}=\vec{v_0}$ and~$\vec{t}=\vec{t_0}$,
+and with (crucially)~$\vec{v}=\vec{v_0}$ and~$\vec{t}=-B\vec{v_0}$, and applying a little algebra, we obtain
 the desired result.
@@
 
-We can generalise this isomorphism to any quantifier-free (but possibly open) type
-of the form $\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A$. 
+We can generalise this isomorphism to type
+of the form $\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A$
+in which $A$ contains no quantifiers. 
 To be precise, we have the following isomorphism:
@@
 
-\exref{ex:type-iso} and \exref{ex:simpletrans} are special cases of this isomorphism; another instance is the
-type of vector addition:
-\[
-\begin{array}{l}
-\forall B\mathord:\SynGL{2}, t_1
-%\mathord:\SynTransl{2}
-, t_2\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_2}\to\tyPrim{vec}{B,t_1\mathord+t_2}\\
-\cong\quad
-\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_1}\\
-\cong\quad
-\forall B\mathord:\SynGL{2}.\tyPrim{vec}{B,0}
-\end{array}
-\]
-\noindent
 \newcommand{\transup}[3]{\mathord\uparrow_{#1}^{#2\mathord:#3}}
@@
 
-\begin{example}
+\begin{lemma}\label{lem:transiso}
+Let $A$ be a quantifier-free type. Then
+\[
+\Delta\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A)
+\cong A[0/t]
+\]
+\end{lemma}
+\begin{proof}
 Let 
@@
 considered as relations are simply instances of $\rsem A{}{}$.
-\qed
+\end{proof}
 %\[
@@
 
+\exref{ex:type-iso} and \exref{ex:simpletrans} are special cases of
+this isomorphism; another instance is the type of vector addition:
+\begin{align*}
+\forall B\mathord:\SynGL{2}, t_1, t_2\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_2}\to\tyPrim{vec}{B,t_1\mathord+t_2}\\
+\cong\quad\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_1}
+\quad (\text{by~\lemref{lem:transiso}})\\
+\cong\quad\forall B\mathord:\SynGL{2}.\tyPrim{vec}{B,0} \quad (\text{by~\lemref{lem:transiso}})\\
+\cong\quad\tyUnit\quad(\text{by abstraction theorem})
+\end{align*}
+In other words, the vector addition operation is the only inhabitant of its type!
 \paragraph{Scalings}
@@
 isomorphism:
-\begin{eqnarray*}
-\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A
-& \cong &
-A[1/s]\times \forall s\mathord:\SynGL{1}.A
-\end{eqnarray*}
+\begin{eqnarray}
+\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A &
+\cong & [1/s]\times \forall s\mathord:\SynGL{1}.A\label{eqn:scaleiso}
+\end{eqnarray}
 The type of $\mathrm{areaCircle}$ from the introduction is one instance:
-\begin{eqnarray*}
-\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s^2}\\ \quad\cong\quad \tyPrim{real}{1}\times(\forall
-s\mathord:\SynGL{1}.\tyPrim{real}{s^2})\\ \quad\cong\quad
-\tyPrim{real}{1}\times 1\quad\cong\quad \tyPrim{real}{1}
-\end{eqnarray*}
+\begin{align*}
+\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s^2}\\ 
+\cong\quad\tyPrim{real}{1}\times(\forall
+s\mathord:\SynGL{1}.\tyPrim{real}{s^2}) \quad(\text{by }\ref{eqn:scaleiso})\\ 
+\cong\quad\tyPrim{real}{1}\times \tyUnit \quad (\text{by abstraction theorem})\\
+\cong\quad\tyPrim{real}{1}\quad (\text{trivially})
+\end{align*}
 %where %the isomorphism 
-The abstraction theorem is used to prove
-$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s^2}\cong 1$.
+%The abstraction theorem is used to prove
+%$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s^2}\cong \tyUnit$.
 % is proved using
 %the abstraction theorem.
-\end{example}
 %\begin{enumerate}
@@
   {\begin{cases} \{(x,k x) \sepbar x \in \mathbb R \} & \text{if }q \in \mathbb Z\\
-                      {(0,0)} & \text{otherwise}
+                      \{(0,0)\} & \text{otherwise}
   \end{cases}}
@@
 Here $q$ is tracking the exponents on scale parameters, with
-$\semPrimType{\tyPrimNm{real}}(q)$ accepting only the zero value
+$\semPrimType{\tyPrimNm{real}}(q)$ relating only zero to itself
 unless the denominator of $q$ is a power of two (a so-called

### 2012-10-23 paper/section-introduction.tex
various

@@
 In previous work the third author studied relational parametricity for
-the types parameterized by units-of-measure, whose invariance 
+types parameterized by units-of-measure, whose invariance 
 properties relate to changes of units, or \emph{scaling}~\cite{kennedy97relational}. 
@@
 \[
-\mathrm{areaTri} : \forall\alpha\mathord:\SynTransl{2}.
-  \tyPrim{vec}{\alpha} \times \tyPrim{vec}{\alpha} \times \tyPrim{vec}{\alpha} \to \tyReal \\
+\mathrm{areaTri} : \forall t\mathord:\SynTransl{2}.
+  \tyPrim{vec}{t} \times \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyReal \\
 \]
-in which the index $\alpha$ is thought of as a ``translation frame''
-that is preserved by the vector arguments. From this we can deduce
-that
+This type expresses the fact that if each of the arguments to $\mathrm{areaTri}$
+is translated by the same vector, then the result remains the same,
+that is, it is \emph{invariant} under translation. Formally, for any 
+vector $\vec t$,
 \[
-\mathrm{areaTri}\;(t + v_1, t + v_2, t + v_3) = 
-\mathrm{areaTri}\;(v_1, v_2, v_3)
+\mathrm{areaTri}\;(\vec{t} + \vec{v_1}, \vec{t} + \vec{v_2}, \vec{t} + \vec{v_3}) = 
+\mathrm{areaTri}\;(\vec{v_1}, \vec{v_2}, \vec{v_3})
 \]
-for any translation $t$ in the group $\Transl{2}$ of translations.
 
@@
 \[
-\mathrm{areaCircle} : \forall\alpha\mathord:\SynGL{1}.\tyPrim{real}{\alpha}\to
-\tyPrim{real}{\alpha\cdot\alpha}
+\mathrm{areaCircle} : \forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to
+\tyPrim{real}{s\cdot s}
 \]
 This captures the fact that the area of a circle varies as the square
-of its radius, i.e., $\mathrm{areaCircle}\;(s\cdot r) = s^2\cdot
-\mathrm{areaCircle}\;r$ for any $s\neq 0$ (i.e., any $s$ in the
-one-dimensional general linear group $\GL{1}$).  Here, $\alpha$ is
-interpreted as the \emph{units of measure} of the argument to
+of its radius, i.e., $\mathrm{areaCircle}(k r) = k^2\cdot
+\mathrm{areaCircle}(r)$ for any $k\neq 0$ (i.e., any $k$ in the
+one-dimensional general linear group $\GL{1}$).  Here, $s$ 
+can be interpreted as the \emph{units of measure} of the argument to
 $\mathrm{areaCircle}$, and `$\cdot$' composes units using the
@@
 groups. This permits identification of, for example,
-$\tyPrim{real}{\alpha\cdot\alpha^{-1}}$ with the type
+$\tyPrim{real}{s\cdot s^{-1}}$ with the type
 $\tyPrim{real}{1}$ of dimensionless constants.
@@
 We present a collection of compelling examples of algebraically
-indexed types, including a novel type system for geometry, a type
-system for tracking taint marking (or provenance) of strings, a
+indexed types, including a novel type system for geometry, a
 refined type system for information flow based on logic, and a simple

### 2012-10-24 paper/geometry-twocolumn.tex
small changes

--- 2012-07-10 paper/geometry-twocolumn.tex
@@
 
-The $n$-tuple of numbers is the basic data structure for programs that
-manipulate geometric data. Pairs, i.e., $2$-tuples, 
-%The basic data structure used in programs that manipulate geometric data is
-%the $n$-tuple of numbers. In the 2-dimensional case, tuples 
+The basic data structure used in programs that manipulate geometric data is
+the $n$-tuple of numbers. In the 2-dimensional case, tuples 
 $\vec{v} = (x,y)$ serve double duty, representing
@@
 mappings. The index-erasure and relational interpretations are given
-formally in Sections~\ref{sec:index-erasure-semantics}
+formally in \autoref{sec:erasure-semantics}
 %the relational interpretation is in general more complex, due to the
@@
 %variables. The relational interpretation will be presented in
-and~\ref{sec:relational-semantics}. %, respectively.
+and~\autoref{sec:relational-semantics}. %, respectively.
 

### 2012-10-24 paper/section-general-framework.tex
various
small changes

@@
 By induction on the typing derivation, making use of
-Lemma~\autoref{lem:tyeqsubst-relational} (part 1) for rule \TirName{TyEq},
-Lemma~\autoref{lem:tyeqsubst-relational} (part 2) for rules \TirName{UnivApp} and \TirName{ExPack},
-and Lemma~\autoref{lem:ctxtsubst-rel} for rules \TirName{UnivAbs} and \TirName{ExUnpack}. 
+Lemma~\ref{lem:tyeqsubst-relational} (part 1) for rule \TirName{TyEq},
+Lemma~\ref{lem:tyeqsubst-relational} (part 2) for rules \TirName{UnivApp} and \TirName{ExPack},
+and Lemma~\ref{lem:ctxtsubst-rel} for rules \TirName{UnivAbs} and \TirName{ExUnpack}. 
 The details can be found in the Coq development.

### 2012-10-24 paper/section-introduction.tex
small changes

--- 2012-07-11 paper/section-introduction.tex
@@
 \end{itemize}
-We have fully formalised our framework and abstraction theorem in Coq,
+We have fully formalised our framework and most examples in Coq,
 using strongly-typed term representations

### 2012-10-24 paper/section-discussion.tex

@@
 
-We presented a general framework for algebraically indexed types
-and instantiated it to yield novel type systems for geometry, taint
-tracking, logical information flow and distance-indexed types. Our
-framework further demonstrates the power of relational reasoning about
-typed programs. From \thmref{thm:abstraction}, we derived
-interesting free theorems, type isomorphisms and non-definability
-results. We conclude with some observations and suggestions for
-further work.
+We presented a general framework for algebraically indexed types and
+instantiated it to yield novel type systems for geometry, logical
+information flow and distance-indexed types. Our framework further
+demonstrates the power of relational reasoning about typed
+programs. From \thmref{thm:abstraction}, we derived interesting free
+theorems, type isomorphisms and non-definability results. We conclude
+with some observations and suggestions for further work.
 

### 2012-10-24 paper/section-general-nondef.tex

\section{Non-definability for the General Framework}
\label{sec:general-nondef}

The non-definability results in the previous section required the use
of specially crafted models and relational interpretations. It is
reasonable to ask whether there is a general method for constructing
suitable models and relational interpretations to prove
non-definability results. In this section, we show that this is
possible for a large class of first-order types.

In \exref{ex:sqrt-root-nondef}, we showed that the type $\forall s
\mathord: \SynGL{1}.~\tyPrim{real}{s^2} \to \tyPrim{real}{s}$ only has
trivial inhabitants. Intuitively, this is because the index of the
result type ($s$) cannot be obtained from the index of the input type
($s^2$) using the abelian group operations and axioms. This
observation can be used to give a sufficient condition for
non-inhabitation for any instantiation of our general framework for
types of the form:
\begin{displaymath}
  \forall i_1\mathord:s_1,\ldots,i_m\mathord:s_m.~\tyPrim{X}{e_1} \to \ldots \to \tyPrim{X}{e_n} \to \tyPrim{X}{e}
\end{displaymath}
Roughly speaking, if this type is inhabited, then it must be the case
that the index expression $e$ can be generated from the set of index
expressions $\{e_1,...,e_n\}$.

We assume that we are working with an instantiation of the general
framework from \autoref{sec:a-general-framework} with a closed typing
context $\Gamma_{\mathit{ops}}$ describing the types of the primitive
operations and a chosen index-erasure semantics. We will use a special
relational interpretation built from the syntax of the indexing
expressions to prove our general non-definability result. For
simplicity, we assume that there is only one primitive type,
$\tyPrimNm{X}$, but the technique we describe here extends to the
general case. We also assume that $\tySem{\tyPrimNm{X}}$ is non-empty.

To state our general non-definability condition, we need to define the
set of index expressions generated by some finite set of index
expressions. Given a set $S$ of index expressions that are all
well-indexed in some index context $\Delta$, we define
$\mathrm{Gen}_0(S)$ to be the set of expressions that are built from
the elements of $S$ and the primitive index operations. To account for
the equations between indexing terms, we close under index expression
equivalence to get the set $\Gen(S) = \{ e \sepbar \exists s.\ \exists
e' \in \mathrm{Gen}_0(S).\ \Delta \vdash e \equiv e' : s \}$. Now a
type is \emph{well-generated} if it is closed and of the form:
\begin{displaymath}
  \forall i_1\mathord:s_1,\ldots,i_m\mathord:s_m.~\tyPrim{X}{e_1} \to \ldots \to \tyPrim{X}{e_n} \to \tyPrim{X}{e}
\end{displaymath}
and $e \in \Gen(\{e_1,...,e_n\})$.
\begin{theorem}\label{thm:general-nondef}
  Assume that the members of $\Gamma_{\mathit{ops}}$ are all of
  well-generated types. If there exists an $M$ with typing:
  \begin{displaymath}
    \Gamma_{\mathit{ops}} \vdash M : \forall i_1\mathord:s_1,...,i_m\mathord:s_m.~\tyPrim{X}{e_1} \to ... \to \tyPrim{X}{e_n} \to \tyPrim{X}{e}
  \end{displaymath}
  then $e \in \Gen(\{e_1,...,e_n\})$.
\end{theorem}
This theorem is usually applied in the contrapositive: if $e \not\in
\Gen(\{e_1,...,e_n\})$ then no such $M$ can exist.
\begin{proof}
  Let $\Delta = i_1\mathord:s_1,...,i_m\mathord:s_m$ be the index
  context constructed from the universally quantified type variables
  in the type of $M$.  To interpret the index expressions, we use the
  free model over the variables in $\Delta$ constructed from the
  syntax. This model assigns to each sort $s$ the set of index
  expressions $\{ e \sepbar \Delta \vdash e : s \}/\mathord\equiv$,
  quotiented by index expression equality. Index operations are
  interpreted by the corresponding syntactic operation on equivalence
  classes: $\semIndexExp{\texttt{\textup{f}}}([e_1],...,[e_n]) =
  [\texttt{\textup{f}}(e_1,...,e_n)]$.

  We take the relational interpretation of the primitive type
  $\tyPrimNm{X}$ as:
  \begin{displaymath}
    \semPrimType{\tyPrimNm{X}}(e) = \{ (x,x) \sepbar x \in \tySem{\tyPrimNm{X}} \land e \in \mathrm{Gen}(\{e_1,...,e_n\}) \}
  \end{displaymath}
  It is straightforward to check that for any index-erasure
  interpretation of the primitive operations $\eta_{\mathit{ops}} \in
  \Gamma_{\mathit{ops}}$, we have $(\eta_{\mathit{ops}},
  \eta_{\mathit{ops}}) \in \rsem{\Gamma_{\mathit{ops}}}*$ due to the
  fact that all members of $\Gamma_{\mathit{ops}}$ have well-generated
  types. Hence, by \thmref{thm:abstraction}, we know that for all
  terms (i.e. elements of the free model over $\Delta$)
  $e'_1\mathord:s_1,...,e'_m\mathord:s_m$:
  \begin{displaymath}%FIXME: check order of index environments
    \begin{array}{@{}l}
      \forall (x_1,x_1') \in \rsem{\tyPrim{X}{e_1}}(e'_1,...,e'_m), ..., \\
      \quad(x_n, x'_n) \in \rsem{\tyPrim{X}{e_n}}(e'_1,...,e'_m). \\
      \quad\quad (\tmSem{M} \eta_{\mathit{ops}}~x_1 ~...~x_n, \tmSem{M}\eta_{\mathit{ops}}~x'_1~...~x'_n) \in \rsem{\tyPrim{X}{e}}(e'_1,...,e'_m)
    \end{array}
  \end{displaymath}
  By setting $e'_j = i_j$, and using an arbitrary element $x \in
  \tySem{\tyPrimNm{X}}$ (which we have assumed non-empty), we have,
  for all $k$, $(x,x) \in \rsem{\tyPrim{X}{e_k}}(i_1,...,i_m)$ since
  $e_k$ is a member of the set of terms we are using to generate
  terms. Hence, $(\tmSem{M} \eta_{\mathit{ops}}~x...x,
  \tmSem{M}\eta_{\mathit{ops}}~x...x) \in \rsem{\tyPrimNm{X}}(e)$ and
  so $e \in \Gen(\{e_1,...,e_n\})$.
\end{proof}

% When applying \thmref{thm:general-nondef}, it is useful to observe
% that the free model constructed from the index variables $i_1,...,i_m$
% often has a well-known structure. Moreover, the set $\Gen(S)$ is
% always a sub-model of the free model. These observations can make it
% easy to apply the theorem in specific cases.

\paragraph{Application to \exref{ex:uninhabited-type}}
\thmref{thm:general-nondef} can be directly applied to show that the
type $\forall t\mathord:\SynTransl{2}.~\tyPrim{vec}{t + t} \to
\tyPrim{vec}{t}$ has no inhabitants. The free model over the single
index variable $t$ is (isomorphic to) the integers, and the sub-model
generated by the index expression $t + t$ corresponds to the even
integers. The result now follows simply because $1$ (i.e. the
interpreration of $t$) is not an even number.

\paragraph{Abelian Group Indexed Types} Kennedy
\cite{kennedy97relational} gave a general characterisation of
definability at first-order in the case of abelian group indexing in
terms of integer solutions to a set of linear equations. Specialising
\thmref{thm:general-nondef} to the case of abelian group indexing
yields Kennedy's characterisation.

\paragraph{Polymorphic Constants} \thmref{thm:general-nondef} does not
apply in the case when we have polymorphic constants. This is the case
with the polymorphic $0 : \forall
s\mathord:\SynGL{1}.~\tyPrim{real}{s}$ in
\autoref{fig:real-ops}. \thmref{thm:general-nondef} does not apply
because the index expression $s$ is not generated by the empty set:
$0$'s type is not well-generated. Nevertheless, it is easy to adapt
the proof of \thmref{thm:general-nondef} to handle a polymorphic
constant like $0$ by setting the relational interpretation of
$\tyPrimNm{X}$ to be:
\begin{displaymath}
  \rsem{X}(e) = \{ (x,x) \sepbar x \in \tySem{\tyPrimNm{X}} \land (x = 0 \lor e \in \Gen(\{e_1,...,e_n\})) \}
\end{displaymath}
The conclusion of the theorem now states that either we have $e \in
\Gen(\{e_1,...,e_n\})$ \emph{or} $\tmSem{M}$ is the constant $0$
function. This extended theorem can now be use to recreate the proof
given above in \exref{ex:sqrt-root-nondef}.

\paragraph{Adding Index Operations} \thmref{thm:general-nondef} also
does not directly apply in the case of \exref{ex:cube-root-nondef},
again because assumption that the types of the primitive operations
are all well-generated is not satisfied. In this case, the assumed
square root operation has type $\forall
s\mathord:\SynGL{1}.~\tyPrim{real}{s^2} \to \tyPrim{real}{s}$, and as
we observed at the start of this section, $s$ is not in the set
generated by $s^2$. However, to enable the application of the theorem,
we can assume an additional index operation $-^{1/2}$, acting
like square root at the index level. Now the free model over a single
generator is exactly the dyadic rationals, and the generated sub-model
consists of the dyadic rationals of the form $\frac{3k}{2^n}$.

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:

### 2012-10-24 paper/section-info-flow.tex

@@
 information flow properties. Tse and Zdancewic \cite{tse04translating}
-translated Abadi \emph{et al.}'s calculus into System F, translating the
-monadic type $T_lA$ to $\alpha_l \to A$ for some free type variable
-$\alpha_l$, and using Reynolds' abstraction theorem to prove
-information flow properties. For example, if the type variable
-$\alpha_H$ represents high-level information, then the
-non-interference property of the function $f$ could be expressed by
-the System F type $A \to (\alpha_H \to B) \to C$. If a program cannot
-generate a value of type $\alpha_H$, then it cannot access the value
-of $B$, and hence is insensitive to the actual value. Relationships
-between security levels are captured by postulating functions
-$\alpha_{l_1} \to \alpha_{l_2}$ whenever $l_1$ is a lower security
-level than $l_2$.
+translated Abadi \emph{et al.}'s calculus into System F\footnote{Tse
+  and Zdancewic's translation did not satisfy all the properties that
+  they claimed, as pointed out by Shikuma and Igarashi
+  \cite{shikuma08proving}. However, this problem is not relevant to
+  our discussion here.} translating the monadic type $T_lA$ to
+$\alpha_l \to A$ for some free type variable $\alpha_l$, and using
+Reynolds' abstraction theorem to prove information flow
+properties. For example, if the type variable $\alpha_H$ represents
+high-level information, then the non-interference property of the
+function $f$ could be expressed by the System F type $A \to (\alpha_H
+\to B) \to C$. If a program cannot generate a value of type
+$\alpha_H$, then it cannot access the value of $B$, and hence is
+insensitive to the actual value. Relationships between security levels
+are captured by postulating functions $\alpha_{l_1} \to \alpha_{l_2}$
+whenever $l_1$ is a lower security level than $l_2$.
 
@@
 principal is indistinguishable from any other data, and a program
-cannot get access to the exact value. From this and
+cannot get access to the exact value. From this observation, and
 \thmref{thm:abstraction}, we obtain the following information flow
@@
   in some indexing context $\Delta$, such that $\psi \not\leq
-  \phi$. Then for all programs
+  \phi$. Then for all terms:
   \begin{displaymath}

### 2012-10-24 paper/section-metric.tex
Fix minor typo in the distance-indexed types
Revamped section on general non-definability results

@@
 
-\autoref{sec:instantiations} and \autoref{sec:monoid-indexed-types}
-presented examples where the relational interpretations of primitive
-types relate pairs of elements by some transformation if applying the
+The type system for geometry we discussed in
+\autoref{sec:motivating-examples} and \autoref{sec:instantiations}
+made use of a relational interpretation of primitive types that
+relates pairs of elements by some transformation if applying the
 transformation to the first element of a pair yields the second. Thus,
@@
 interpretations of the primitive operations are collected together
-into an environment $\eta_\Metric \in \ctxtSem{\Gamma_\Mon}$.
+into an environment $\eta_\Metric \in \ctxtSem{\Gamma_\Metric}$.
 
@@
 \begin{displaymath}
-  \forall \epsilon \mathord> 0. \exists \delta \mathord> 0. \forall x,x'. abs{x - x'} < \delta \Rightarrow \abs{\tmSem{M}\ x - \tmSem{M}\ x'} < \epsilon
+  \forall \epsilon \mathord> 0. \exists \delta \mathord> 0. \forall x,x'. \abs{x - x'} \mathord< \delta \Rightarrow \abs{\tmSem{M}\eta_\Metric~x - \tmSem{M}\eta_\Metric~x'} \mathord< \epsilon
 \end{displaymath}

### 2012-10-24 paper/section-instantiations.tex
small changes
iso

@@
 
-We can generalise this isomorphism to type
-of the form $\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A$
-in which $A$ contains no quantifiers. 
-To be precise, we have the following isomorphism:
-\[
-\Delta\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A)
-\cong A[0/t]
-\]
-where $A[0/t]$ denotes substitution of the identity translation $0$
+We can generalise this isomorphism substantially, proving
+that for any quantifier-free type $A$, 
+the type $\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A$
+is isomorphic to $A[0/t]$, where $A[0/t]$ 
+denotes substitution of the identity translation $0$
 for index variable $t$ in type $A$. This isomorphism formalises the

### 2012-10-24 paper/section-instantiations.tex
Iso, correct statement

@@
 \autoref{sec:affine-vector-ops}. 
+\newcommand{\DeltaWithB}{\Delta,B\mathord:\SynGL{2}}
 
-\begin{example}[Geometry: change of origin]\label{ex:simpletrans}
-We prove the isomorphism
-$\Delta\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to\tyPrim{vec}{e,t}) \cong
-\tyPrim{vec}{e,0}$ where $e$ is an expression of sort $\SynGL{2}$ not containing $t$.
+\begin{example}[Geometry: wlog]\label{ex:simpletrans}
+We prove that
+\[\DeltaWithB\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to\tyPrim{vec}{B,t}) \cong
+\tyPrim{vec}{B,0}.
+\]
 Let 
 \begin{align*}
-X & \isDefinedAs\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to\tyPrim{vec}{e,t}
-&
-Y & \isDefinedAs\tyPrim{vec}{e,0}\\
-I & \isDefinedAs\lambda f\mathord:X.f\,[0]\,(0\,[e])\\
-J & \isDefinedAs\lambda v\mathord:Y.\Lambda t\mathord:\SynTransl{2}.\lambda w\mathord:\tyPrim{vec}{e,t}.v +[e,0,t] w
+X & \isDefinedAs\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to\tyPrim{vec}{B,t}
+\quad\text{and}\quad Y \isDefinedAs\tyPrim{vec}{B,0}\\
+I & \isDefinedAs\lambda f\mathord:X.f\,[0]\,(0\,[B])\\
+J & \isDefinedAs\lambda v\mathord:Y.\Lambda t\mathord:\SynTransl{2}.\lambda w\mathord:\tyPrim{vec}{B,t}.v +[B,0,t] w
 \end{align*}
@@
 for any $\vec{v}\in\mathbb R ^ 2$.  Because $(0,0)$ is the identity for vector
-addition it follows that $\ctxteq{\Delta}{v\mathord:Y}{I(J(v))}{v}Y$.
-To show $\ctxteq{\Delta}{f\mathord:X}{J(I(f))}{f}{X}$ we appeal to
+addition it follows that $\ctxteq{\DeltaWithB}{v\mathord:Y}{I(J(v))}{v}Y$.
+To show $\ctxteq{\DeltaWithB}{f\mathord:X}{J(I(f))}{f}{X}$ we appeal to
 \autoref{thm:soundness} and reason using the relational semantics.  It
-suffices to show $\semeq{\Delta}{f\mathord:X}{J(I(f))}{f}{X}$; that
-is, given $\rho\in\semSort\Delta$, for any $f$ and $f'$ with
+suffices to show $\semeq{\DeltaWithB}{f\mathord:X}{J(I(f))}{f}{X}$; that
+is, given $\rho\in\semSort\DeltaWithB$, for any $f$ and $f'$ with
 $(f,f')\in\rsem X\rho$, it is the case that 
 $(\tmSem{J(I(f))}(\opsEnv,f), f')\in\rsem X\rho$.
-Expanding the premise, and setting $B = \semIndexExp{e}\rho$, we have
+Expanding the premise, we have
 \begin{eqnarray}\label{eqn:transpremise}
@@
 \forall \vec{v}\in \mathbb R ^ 2,
-f'(B\vec{v} + \vec{t}) = B(f(\vec{v})) + \vec{t}
+f'(\rho(B)\vec{v} + \vec{t}) = \rho(B)(f(\vec{v})) + \vec{t}
 \end{eqnarray}
@@
 $\vec{t_0}\in\mathbb R^2$ and $\vec{v_0}\in\mathbb R^2$, it is the case that
-$f'(B\vec{v_0} + \vec{t_0}) = B(\vec{v_0} + f(0)) + \vec{t_0}$.
+$f'(\rho(B)\vec{v_0} + \vec{t_0}) = \rho(B)(\vec{v_0} + f(0)) + \vec{t_0}$.
 By instantiating \autoref{eqn:transpremise} thrice, with~$\vec{v}=\vec{t}=(0,0)$, with~$\vec{v}=\vec{v_0}$ and~$\vec{t}=\vec{t_0}$,
-and with (crucially)~$\vec{v}=\vec{v_0}$ and~$\vec{t}=-B\vec{v_0}$, and applying a little algebra, we obtain
+and with (crucially)~$\vec{v}=\vec{v_0}$ and~$\vec{t}=-\rho(B)\vec{v_0}$, and applying a little algebra, we obtain
 the desired result.
@@
 that for any quantifier-free type $A$, 
-the type $\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A$
+the type $\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to A$
 is isomorphic to $A[0/t]$, where $A[0/t]$ 
@@
 \begin{lemma}\label{lem:transiso}
-Let $A$ be a quantifier-free type. Then
+Let $A$ be a quantifier-free type in which every
+occurrence of $\tyPrimNm{vec}$ is of the form $\tyPrim{vec}{B,e}$ for some $e$. Then
 \[
-\Delta\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A)
+\DeltaWithB\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to A)
 \cong A[0/t]
@@
 \begin{align*}
-X & \isDefinedAs \forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A
+X & \isDefinedAs \forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to A
 & 
 Y & \isDefinedAs A[0/t]\\
-I & \isDefinedAs \lambda f\mathord: X. f\ [0]\ (0\ [e])
+I & \isDefinedAs \lambda f\mathord: X. f\ [0]\ (0\ [B])
 & 
-J & \isDefinedAs \lambda y\mathord: Y. \Lambda t.\lambda v\mathord:\tyPrim{vec}{e,t}. \transup A t v (y)
+J & \isDefinedAs \lambda y\mathord: Y. \Lambda t.\lambda v\mathord:\tyPrim{vec}{B,t}. \transup A t v (y)
 \end{align*}
@@
 &
-\Delta,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{e,t}\vdash \transup A t v : A[0/t] \to A
+\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transup A t v : A[0/t] \to A
 \\\text{and } &
-\Delta,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{e,t}\vdash \transdn A t v : A\to A[0/t]
+\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transdn A t v : A\to A[0/t]
 \end{align*}
 are defined by induction over the syntax of $A$. For brevity we omit types on binders
-and note that the omitted cases for $\transup A t v$ 
+and indices on primitives; also note that the omitted cases for $\transup A t v$ 
 have definition symmetric to $\transdn A t v$:
@@
 &
-\transup{t} t v &= \lambda w.w - v
+\transup{t} t v &= \lambda w.w + (\mathrm{negate}\, v)
 \\
@@
 To show $I(J(y))$ is contextually equivalent to $y$ is an easy
-induction on $A$ using the fact that the vector $0$
+induction on $A$ using the fact that the vector $(0,0)$
 is an identity for vector addition. To show $J(I(x))$ is contextually
-equivalent to $x$ we must apply the abstraction theorem and use
-the fact that $\tmSem{\transup A t v}$ and $\tmSem{\transdn A t v}$ 
-considered as relations are simply instances of $\rsem A{}{}$.
+equivalent to $x$ we first prove (by induction on $A$) 
+that for any $\rho\in\semSort{\DeltaWithB}$, and any $\vec v\in\mathbb R^2$, 
+the graph of the function given by $\tmSem{\transdn A t v}(\opsEnv,\vec v)$
+is just  $\rsem A (\rho,\vec v)$, and conversely, the graph of $\tmSem{\transup A t v}(\opsEnv,\vec v)$
+is ${\rsem A (\rho,\vec v)}^{\mathrm{op}}$.
 \end{proof}
-%\[
-%$\rsem A {t \mapsto v} {} = \{ (x, \tmSem{\transup A (x)}) \mid x \in
-%\tySem A\}$.
-%\]
 
 \exref{ex:type-iso} and \exref{ex:simpletrans} are special cases of
-this isomorphism; another instance is the type of vector addition:
+this isomorphism. Another instance is the type of vector addition:
 \begin{align*}
@@
 $\forall s \mathord: \SynGL{1}.\tyPrim{real}{s^2} \to \tyPrim{real}{s}$. 
-There are many functions $f$ in $\mathbb R \to \mathbb R$
+There are many functions~$f: \mathbb R \to \mathbb R$
 that satisfy its relational interpretation, i.e. for which $f(k^2\cdot x) = k\cdot f(x)$ for 

### 2012-10-24 paper/section-general-framework.tex
floor notation

@@
 \begin{example*}[Geometry: operations]
-We assume $\opsCtxt$ collects together the types of primitive operations as described in \autoref{sec:motivation-generalising}.
-\todo{Write these down}.
+For geometry $\opsCtxt$ would collect together the types of primitive operations as 
+listed in \autoref{fig:real-ops} and \autoref{fig:vec-ops}.
 \qed
@@
 (We will overload the
-notation $\left|\cdot\right|$ for all index-erasure interpretations, 
+notation $\tmSem\cdot$ for all index-erasure interpretations, 
 and reserve notation $\llbracket\cdot\rrbracket$ for the
@@
   $\tyPrimSem{\tyPrimNm{real}} = \mathbb{R}$.
-  We assume that $\opsEnv$ interprets the operations described in \autoref{sec:motivation-generalising}
-  in the standard way. \todo{Expand}. 
+  We assume that $\opsEnv$ gives the usual interpretation to scalar
+  and vector operations from
+  \autoref{fig:real-ops} and \autoref{fig:vec-ops}.
 \qed
@@
   \rsem{\forall i\mathord:s.A} \rho & \isDefinedAs & 
-  \bigcap\{ \rsem{A} {(m,\rho)} \sepbar m \in \semSort{s} \} \\
+  \bigcap\{ \rsem{A} {(\rho,m)} \sepbar m \in \semSort{s} \} \\
   \rsem{\exists i\mathord:s.A} \rho & \isDefinedAs & 
-  \bigcup\{ \rsem{A} {(m,\rho)} \sepbar m \in \semSort{s} \}
+  \bigcup\{ \rsem{A} {(\rho,m)} \sepbar m \in \semSort{s} \}
 \end{eqnarray*}

### 2012-10-24 paper/section-instantiations.tex
finally proved iso

@@
 \end{align*}
-Unfolding definitions gives that $\tmSem{I(J(v))}(\opsEnv,\vec{v})$ = $\vec{v}+(0,0)$
-for any $\vec{v}\in\mathbb R ^ 2$.  Because $(0,0)$ is the identity for vector
+Unfolding definitions gives that $\tmSem{I(J(v))}(\opsEnv,\vec{v})$ = $\vec{v}+\vec{0}$
+for any $\vec{v}\in\mathbb R ^ 2$.  Because $\vec{0}$ is the identity for vector
 addition it follows that $\ctxteq{\DeltaWithB}{v\mathord:Y}{I(J(v))}{v}Y$.
@@
 $f'(\rho(B)\vec{v_0} + \vec{t_0}) = \rho(B)(\vec{v_0} + f(0)) + \vec{t_0}$.
-By instantiating \autoref{eqn:transpremise} thrice, with~$\vec{v}=\vec{t}=(0,0)$, with~$\vec{v}=\vec{v_0}$ and~$\vec{t}=\vec{t_0}$,
+By instantiating \autoref{eqn:transpremise} thrice, with~$\vec{v}=\vec{t}=\vec{0}$, with~$\vec{v}=\vec{v_0}$ and~$\vec{t}=\vec{t_0}$,
 and with (crucially)~$\vec{v}=\vec{v_0}$ and~$\vec{t}=-\rho(B)\vec{v_0}$, and applying a little algebra, we obtain
@@
 \newcommand{\transdn}[3]{\mathord\downarrow_{#1}^{#2\mathord:#3}}
+\newcommand{\transtrans}[3]{\mathrm{tr}_{#1}^{#2\mathord:#3}}
 
 \begin{lemma}\label{lem:transiso}
-Let $A$ be a quantifier-free type in which every
-occurrence of $\tyPrimNm{vec}$ is of the form $\tyPrim{vec}{B,e}$ for some $e$. Then
+Suppose that $\DeltaWithB,t\mathord:\SynTransl{2}\vdash A \isType$, 
+$A$ contains no quantifiers and every occurrence of $\tyPrimNm{vec}$ is of the form $\tyPrim{vec}{B,?}$. Then
 \[
@@
 X & \isDefinedAs \forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to A
-& 
-Y & \isDefinedAs A[0/t]\\
-I & \isDefinedAs \lambda f\mathord: X. f\ [0]\ (0\ [B])
-& 
+\quad\text{and}\quad Y \isDefinedAs A[0/t]\\
+I & \isDefinedAs \lambda f\mathord: X. f\ [0]\ (0\ [B])\\
 J & \isDefinedAs \lambda y\mathord: Y. \Lambda t.\lambda v\mathord:\tyPrim{vec}{B,t}. \transup A t v (y)
@@
 \begin{align*}
-&
-\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transup A t v : A[0/t] \to A
-\\\text{and } &
-\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transdn A t v : A\to A[0/t]
+&\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transup A t v : A[0/t] \to A\\
+&\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transdn A t v : A\to A[0/t]\\
+&\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transup e t v : \tyPrim{vec}{B,e[0/t]}\to\tyPrim{vec}{B,e}\\
+&\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transdn e t v : \tyPrim{vec}{B,e}\to\tyPrim{vec}{B,e[0/t]}
 \end{align*}
-are defined by induction over the syntax of $A$. For brevity we omit types on binders
-and indices on primitives; also note that the omitted cases for $\transup A t v$ 
-have definition symmetric to $\transdn A t v$:
+are defined by induction over the structure of $A$ and $e$. For brevity we omit types on binders;
+also note that the omitted cases for $\transdn A t v$ 
+have definition symmetric to $\transup A t v$:
 \begin{align*}
-\transdn{\tyPrimNm{unit}} t v &= \lambda x. x
+\transup{\tyPrimNm{unit}} t v &= \lambda x. x
 \qquad 
-\transdn{\tyPrim{real}{e}} t v = \lambda x. x
+\transup{\tyPrim{real}{e}} t v = \lambda x. x
 \\
-\transdn{A_1\times A_2}t v &= \lambda p. (\transdn{A_1}t v(\pi_1 p),\transdn{A_2}t v(\pi_2 p))
+\transup{A_1\times A_2}t v &= \lambda p. (\transup{A_1}t v(\pi_1 p),\transup{A_2}t v(\pi_2 p))
 \\
-\transdn{A_1 + A_2} t v &= \lambda x.
-\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn {A_1}t v y); \textrm{inr}\ y.\textrm{inr}(\transdn {A_2} t v y)
+\transup{A_1 + A_2} t v &= \lambda x.
+\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transup {A_1}t v y); \textrm{inr}\ y.\textrm{inr}(\transup {A_2} t v y)
 \\ 
-\transdn{A_1\to A_2} t v &= \lambda f. \lambda x. \transdn{A_2} t v (f(\transup {A_1} t v x))
+\transup{A_1\to A_2} t v &= \lambda f. \lambda x. \transup{A_2} t v (f(\transdn {A_1} t v x))
 \end{align*}
 \begin{align*}
-\transdn{\tyPrim{vec}{B,e}} t v &= \lambda x. \transdn{e}  t v (x) 
+\transup{\tyPrim{vec}{B,e}} t v &= \lambda x. \transdn{e}  t v (x) 
 &
-\transup{\tyPrim{vec}{B,e}} t v &= \lambda x. \transup{e} t v (x)
+\transdn{\tyPrim{vec}{B,e}} t v &= \lambda x. \transup{e} t v (x)
 \\
+\transup{0} t v &= \lambda w.w
+&
 \transdn{0} t v &= \lambda w.w
+\\ 
+\transup{e_1+e_2} t v &= \lambda w.\transup{e_1} t v(\transup{e_2} t v(w))
 &
-\transup{0} t v &= \lambda w.w
+\transdn{e_1+e_2} t v &= \lambda w.\transdn{e_1} t v(\transdn{e_2} t v(w))
 \\
-\transdn{e_1+e_2} t v &= \lambda w.\transdn{e_1} t v(\transdn{e_2} t v(w))
+\transup{-e} t v &= \lambda w.\transdn{e} t v (w)
 &
-\transup{e_1+e_2} t v &= \lambda w.\transup{e_1} t v(\transup{e_2} t v(w))
+\transdn{-e} t v &= \lambda w.\transup{e} t v (w)
 \\
-\transdn{-e} t v &= \lambda w.\transup{e} t v (w)
+\transup{t} t v &= \lambda w.w + v 
 &
-\transup{-e} t v &= \lambda w.\transdn{e} t v (w)
+\transdn{t} t v &= \lambda w.w + (\mathrm{negate}\, v)
 \\
-\transdn{t} t v &= \lambda w.w + v 
+\transup{t'} t v &= \lambda w.w\quad(t'\neq t)
 &
-\transup{t} t v &= \lambda w.w + (\mathrm{negate}\, v)
-\\
 \transdn{t'} t v &= \lambda w.w\quad(t'\neq t)
-&
-\transup{t'} t v &= \lambda w.w\quad(t'\neq t)
 \end{align*}
-To show $I(J(y))$ is contextually equivalent to $y$ is an easy
-induction on $A$ using the fact that the vector $(0,0)$
-is an identity for vector addition. To show $J(I(x))$ is contextually
-equivalent to $x$ we first prove (by induction on $A$) 
-that for any $\rho\in\semSort{\DeltaWithB}$, and any $\vec v\in\mathbb R^2$, 
-the graph of the function given by $\tmSem{\transdn A t v}(\opsEnv,\vec v)$
-is just  $\rsem A (\rho,\vec v)$, and conversely, the graph of $\tmSem{\transup A t v}(\opsEnv,\vec v)$
-is ${\rsem A (\rho,\vec v)}^{\mathrm{op}}$.
+
+Unfolding definitions, we deduce that
+\[
+\tmSem{I(J(y))}(\opsEnv,y) = \tmSem{\transup A t v}(\opsEnv,\vec 0)(y)
+\]
+for any $y \in \tmSem{A[0/t]}$. It's easy to prove by induction on $A$
+that $\tmSem{\transup A t v}(\opsEnv,\vec 0)(y) = y$, and so
+$\ctxteq\DeltaWithB{y\mathord:Y}{J(I(y))}y Y$ follows.
+
+For the opposite direction of the isomorphism, we appeal to 
+\autoref{thm:soundness} and so it suffices to show
+$\semeq{\DeltaWithB}{f\mathord:X}{J(I(f))}{f}{X}$; that
+is, given $\rho\in\semSort\DeltaWithB$, for any $f$ and $f'$ with
+$(f,f')\in\rsem X\rho$, it is the case that 
+$(\tmSem{J(I(f))}(\opsEnv,f), f')\in\rsem X\rho$.
+Expanding the premise, we have
+\begin{eqnarray}\label{eqn:isopremise}
+\forall \vec{t}\in \mathbb R ^ 2,
+\forall \vec{v}\in \mathbb R ^ 2,
+(f(\vec{v}),f'(\rho(B)\vec{v} + \vec{t}))\in\rsem A (\rho,\vec{t})
+\end{eqnarray}
+Expanding the conclusion, we have to show for any
+$\vec{t_0}\in\mathbb R^2$ and $\vec{v_0}\in\mathbb R^2$, it is the case that
+$(\tmSem{J(I(f))}(\opsEnv,f)(\vec{v_0}),f'(\rho(B)\vec{v_0} + \vec{t_0}))\in\rsem A (\rho,\vec{t_0})$.
+Unfolding the definitions of $J$ and $I$ and expanding the erasure semantics, 
+this amounts to showing that
+\[
+\forall \vec{t_0},\vec{v_0}\in \mathbb R ^ 2,
+(\tmSem{\transup A t v}(\opsEnv,\vec{v_0})(f\ \vec 0),f'(\rho(B)\vec{v_0} + \vec{t_0}))\in\rsem A (\rho,\vec{t_0}).
+\]
+In order to make progress we prove first by induction on $A$ that for any $\vec v$ and $\vec w$, 
+the meaning of $\transup A t v$ can be characterized by
+\begin{align*}
+(\tmSem{\transup A t v}(\opsEnv,\vec{v})(x),y)&\in\rsem A (\rho,\vec{w})
+\\\Leftrightarrow 
+(x,y)&\in\rsem A (\rho,\rho(B)\vec{v} + \vec{w})
+\end{align*}
+Thus our goal becomes
+\[
+\forall \vec{t_0},\vec{v_0}\in \mathbb R ^ 2,
+(f\ \vec 0,f'(\rho(B)\vec{v_0} + \vec{t_0}))\in\rsem A (\rho,\rho(B)\vec{v_0} + \vec{t_0}).
+\]
+This is obtained immediately by instantiating \autoref{eqn:isopremise}
+with $\vec{v}=\vec 0$ and $\vec{t}=\rho(B)\vec{v_0} + \vec{t_0}$.
 \end{proof}
@@
 \cong\quad\forall B\mathord:\SynGL{2}.\tyPrim{vec}{B,0} \quad (\text{by~\lemref{lem:transiso}})\\
-\cong\quad\tyUnit\quad(\text{by abstraction theorem})
+\cong\quad\tyUnit\quad(\text{by \autoref{thm:abstraction}})
 \end{align*}
@@
 \cong\quad\tyPrim{real}{1}\times(\forall
-s\mathord:\SynGL{1}.\tyPrim{real}{s^2}) \quad(\text{by }\ref{eqn:scaleiso})\\ 
-\cong\quad\tyPrim{real}{1}\times \tyUnit \quad (\text{by abstraction theorem})\\
+s\mathord:\SynGL{1}.\tyPrim{real}{s^2}) \quad(\text{by \autoref{eqn:scaleiso}})\\ 
+\cong\quad\tyPrim{real}{1}\times \tyUnit \quad (\text{by \autoref{thm:abstraction}})\\
 \cong\quad\tyPrim{real}{1}\quad (\text{trivially})

### 2012-10-24 paper/geometry-twocolumn.tex
summary figure in geometry
Text about figure 2

--- 2012-10-23 paper/geometry-twocolumn.tex
@@
 these form the \emph{centre} of $\GL{2}$. If we keep track of
-scalings, then we can assign the following more precise types to
-scalar multiplication and dot product:
-\begin{displaymath}
-\begin{array}{@{}l@{}l}
-  (*) : \forall s \mathord: \SynGL{1}, B \mathord:
-  \SynGL{2}.\ & \tyPrim{real}{s} \to \tyPrim{vec}{B,0} \to
-  \tyPrim{vec}{\iota_1(s)B,0}\\
-%\end{displaymath}
-%and dot product:
-%\begin{displaymath}
-%  \begin{array}{l@{}l}
-    (\cdot) : \forall s \mathord: \SynGL{1}, O \mathord: \SynOrth{2}.\ & \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\\
-    &\tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to \tyPrim{real}{s^2}
-  \end{array}
-\end{displaymath}
+scalings, then we can assign the more precise types to
+scalar multiplication and dot product. These
+ are shown in Figure~\ref{fig:vec-ops}, which summarises the most general types 
+of all the vector operations that we have described.
 
@@
 
+\begin{figure}[t]
+  \centering
+  \begin{align*}
+    0 & : \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \\
+    (+) & :
+    \begin{array}[t]{@{}l}
+      \forall B \mathord: \SynGL{2}, t_1,t_2 \mathord: \SynTransl{2}. \\
+      \tyPrim{vec}{B,t_1} \to \tyPrim{vec}{B,t_2} \to
+      \tyPrim{vec}{B,t_1 + t_2}
+    \end{array}
+    \\
+    \mathrm{negate} & : \forall B \mathord: \SynGL{2}, t \mathord: \SynTransl{2}.\ \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,-t} \\
+  (*) & : 
+  \begin{array}[t]{@{}l}
+    \forall s \mathord: \SynGL{1}, B \mathord:\SynGL{2}.\ \\
+    \tyPrim{real}{s} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{\iota_1(s)B,0}
+  \end{array}\\
+    \mathrm{affComb} &:
+    \begin{array}[t]{@{}l}\forall t \mathord:
+    \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to
+    \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\ 
+    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} +
+    r\vec{v_2} 
+    \end{array}
+    (\cdot) & :
+\begin{array}[t]{@{}l}
+    \forall s \mathord: \SynGL{1}, O \mathord: \SynOrth{2}.\ \\
+    \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to \tyPrim{real}{s^2}
+\end{array}\\
+  (\times) & : \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \to
+  \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B} 
+\end{align*}
+  \caption{Operations on vectors}
+  \label{fig:vec-ops}
+\end{figure}
+
 %%% Local Variables:

### 2012-10-24 paper/section-instantiations.tex
Text about figure 2
silly me, simplified example

@@
 $f'(\rho(B)\vec{v_0} + \vec{t_0}) = \rho(B)(\vec{v_0} + f(0)) + \vec{t_0}$.
-By instantiating \autoref{eqn:transpremise} thrice, with~$\vec{v}=\vec{t}=\vec{0}$, with~$\vec{v}=\vec{v_0}$ and~$\vec{t}=\vec{t_0}$,
-and with (crucially)~$\vec{v}=\vec{v_0}$ and~$\vec{t}=-\rho(B)\vec{v_0}$, and applying a little algebra, we obtain
-the desired result.
+By instantiating \autoref{eqn:transpremise} with
+$\vec{v}=\vec{0}$ and~$\vec{t}=\rho(B)\vec{v_0}+\vec{t_0}$ and applying a little algebra we obtain this result.
 \qed
@@
   \begin{cases} \Delta_{\mathbb R} & \text{if }z=0\\
-                      {(0,0)} & \text{if }z=1
+                      \{(0,0)\} & \text{if }z=1
 \end{cases}

### 2012-10-25 paper/section-introduction.tex
Add a paper-overview picture

@@
 
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+  \node at (-3,1) [rectangle,style={align=center}] {Algebraic \\ Theory};
+  \node at (-2,1) {$+$};
+  \node at (-0.5,1) [rectangle,style={align=center}] {Index-Erasure \\ Interpretation};
+  \node at (1,1) {$+$};
+  \node at (2.6,1) [rectangle,style={align=center}] {Relational \\ Interpretation};
+
+  \draw plot (-4,0.8) .. controls (-4,0) and (-0.4,1) .. (-0.4,0)
+             -- (-0.6,0) -- (0,-0.3) -- (0.6,0) --
+             (0.4,0) .. controls (0.4,1) and (4,0) .. (4,0.8);
+
+  \node at (0,-0.7) [rectangle,style={align=center}] {Abstraction Theorem (\thmref{thm:abstraction})};
+
+  \draw plot (-0.8,-1) .. controls (-0.8,-1.3) and (-3.5,-1.5) .. (-3.5,-2)
+             -- (-3.7,-2) -- (-3.1,-2.3) -- (-2.5,-2) -- 
+             (-2.7,-2) .. controls (-2.7,-1.5) and (-0.8,-1.6) .. (-0.7,-1.5)
+                       .. controls (-0.8,-1.6) and (-0.5,-1.5) ..
+             (-0.4,-2) -- (-0.6,-2) -- (0,-2.3) -- (0.6,-2) --
+             (0.4,-2) .. controls (0.5,-1.5) and (0.8,-1.6) .. (0.7,-1.5)
+                      .. controls (0.8,-1.6) and (2.7,-1.5) ..
+             (2.7,-2) -- (2.5,-2) -- (3.1,-2.3) -- (3.7,-2) -- (3.5,-2)
+                      .. controls (3.5,-1.5) and (0.8,-1.3) .. (0.8,-1);
+
+  \node at (-3.1,-2.8) [rectangle,style={align=center}] {Free \\ Theorems};
+  \node at (0,-2.8) [rectangle,style={align=center}] {Type \\ Isomorphisms};
+  \node at (3.1,-2.8) [rectangle,style={align=center}] {Non-definability};
+\end{tikzpicture}
+  \caption{Summary of the Paper}
+  \label{fig:summary}
+\end{figure}
+
 This paper makes the following specific contributions:

### 2012-10-25 paper/section-introduction.tex
alphas

--- 2012-10-24 paper/section-introduction.tex
@@
 difference can be assigned the type
-$\forall\alpha\mathord:\SynTransl{2}.\tyPrim{vec}\alpha\times\tyPrim{vec}\alpha\to\tyPrim{vec}0$,
+$\forall t\mathord:\SynTransl{2}.\tyPrim{vec}t\times\tyPrim{vec}t\to\tyPrim{vec}0$,
 reflecting the invariance of the result (a pure vector) under
@@
 variable. For example, values $v$ and $w$ of type
-$\forall\alpha\mathord:\SynTransl{2}.\tyPrim{vec}\alpha\to
-\tyPrim{vec}\alpha$ are related when they are related at type
-$\tyPrim{vec}\alpha\to\tyPrim{vec}\alpha$ for all translations
-$t\in\Transl{2}$ associated with $\alpha$.
+$\forall t\mathord:\SynTransl{2}.\tyPrim{vec} t\to
+\tyPrim{vec} t$ are related when they are related at type
+$\tyPrim{vec} t\to\tyPrim{vec}t$ for all translations
+$\vec t\in\Transl{2}$ associated with $t$.
 
@@
 as
-$(\forall\alpha\mathord:\SynGL{1}.\tyPrim{real}\alpha\to\tyPrim{real}\alpha)\to\tyBool$.
+$(\forall s\mathord:\SynGL{1}.\tyPrim{real} s\to\tyPrim{real} s)\to\tyBool$.
 More interestingly, particular applications sometimes require
@@
 type
-$\forall\alpha\mathord:\SynGL{1}.\tyPrim{real}{\alpha\cdot\alpha}\to\tyPrim{real}\alpha$.
+$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s}\to\tyPrim{real}s$.
 Finally, there are interesting applications in which even the basic
 relations are not functional. In a type system in which the index in
-$\tyPrim{real}\alpha$ is interpreted not as a unit of measure, but as
+$\tyPrim{real}s$ is interpreted not as a unit of measure, but as
 a measure of \emph{closeness}, two values $x$ and $y$ of this type are
-related if $|x-y| < \rho(\alpha)$ for a positive real number
-$\rho(\alpha)$.  Rather beautifully, the standard notion of uniform
+related if $|x-y| < \rho(s)$ for a positive real number
+$\rho(s)$.  Rather beautifully, the standard notion of uniform
 continuity can then be expressed as %a type:

### 2012-10-25 paper/section-general-nondef.tex

@@
 of specially crafted models and relational interpretations. It is
-reasonable to ask whether there is a general method for constructing
-suitable models and relational interpretations to prove
+reasonable to ask whether or not there is a general method for
+constructing suitable models and relational interpretations to prove
 non-definability results. In this section, we show that this is
-possible for a large class of first-order types.
+possible for a large class of first-order types in any instance of our
+general framework.
 
@@
 observation can be used to give a sufficient condition for
-non-inhabitation for any instantiation of our general framework for
-types of the form:
+non-inhabitation for types of the form:
 \begin{displaymath}
@@
 This theorem is usually applied in the contrapositive: if $e \not\in
-\Gen(\{e_1,...,e_n\})$ then no such $M$ can exist.
+\Gen(\{e_1,...,e_n\})$ then no such $M$ can exist. Note that if
+$\Gamma_{\mathit{ops}}$ contains operations corresponding to each of
+the index-level operations (as, for example, in
+\autoref{fig:real-ops}), then this theorems yields a characterisation
+of definable terms, since the construction of $e$ from $e_1,...,e_n$
+can be replicated at the term level.
 \begin{proof}
@@
       \quad(x_n, x'_n) \in \rsem{\tyPrim{X}{e_n}}(e'_1,...,e'_m). \\
-      \quad\quad (\tmSem{M} \eta_{\mathit{ops}}~x_1 ~...~x_n, \tmSem{M}\eta_{\mathit{ops}}~x'_1~...~x'_n) \in \rsem{\tyPrim{X}{e}}(e'_1,...,e'_m)
+      \quad (\tmSem{M} \eta_{\mathit{ops}}~x_1 ~...~x_n, \tmSem{M}\eta_{\mathit{ops}}~x'_1~...~x'_n) \in \rsem{\tyPrim{X}{e}}(e'_1,...,e'_m)
     \end{array}
@@
   $e_k$ is a member of the set of terms we are using to generate
-  terms. Hence, $(\tmSem{M} \eta_{\mathit{ops}}~x...x,
+  terms. Now, $(\tmSem{M} \eta_{\mathit{ops}}~x...x,
   \tmSem{M}\eta_{\mathit{ops}}~x...x) \in \rsem{\tyPrimNm{X}}(e)$ and
@@
 integers. The result now follows simply because $1$ (i.e. the
-interpreration of $t$) is not an even number.
+interpretation of $t$) is not an even number.
 
@@
 \Gen(\{e_1,...,e_n\})$ \emph{or} $\tmSem{M}$ is the constant $0$
-function. This extended theorem can now be use to recreate the proof
-given above in \exref{ex:sqrt-root-nondef}.
+function. This extended theorem can now be used to give an alternative
+proof for \exref{ex:sqrt-root-nondef}. As this example illustrates,
+there may be many different models that can be used to prove a
+non-definability result.
 
@@
 generated by $s^2$. However, to enable the application of the theorem,
-we can assume an additional index operation $-^{1/2}$, acting
-like square root at the index level. Now the free model over a single
-generator is exactly the dyadic rationals, and the generated sub-model
-consists of the dyadic rationals of the form $\frac{3k}{2^n}$.
+we can assume an additional index operation $-^{1/2}$, acting like
+square root at the index level. Now the free model produced in the
+proof of the theorem is isomorphic to the dyadic numbers with addition
+and halving, and the generated sub-model consists of the dyadic
+rationals of the form $\frac{3k}{2^n}$. Again there is a diversity of
+models that can be used to prove a single non-definability result.
 

### 2012-10-25 paper/section-discussion.tex
intro, contributions

@@
 automatically satisfy invariance properties, removing the need for
-{\em ad hoc} proofs. % of these facts.
+{\em ad hoc} proofs of these facts.
 
@@
 \paragraph{Semantic Equality} In general, the semantic equality
-$\stackrel{sem}\sim$ in \defref{def:semantic-equality} is not an
+in \defref{def:semantic-equality} is not an
 equivalence relation. If the interpretations of all 

### 2012-10-25 paper/section-introduction.tex
intro, contributions

@@
 \emph{invariance}: the behaviour of polymorphic code is invariant
-under changes of data representation. Invariance results also abound in
+under changes of data representation. For example, the type
+$\forall\alpha.\tyPrim{list}\alpha\to\tyPrim{list}\alpha$
+tells us that any transformation applied to elements of
+the input list will be reflected by the same transformation applied
+to elements of the result. 
+Invariance results also abound in
 mathematics and physics. The area of a triangle is invariant with
@@
 matrix is invariant under changes of basis; and Newton's laws are the
-same in all inertial frames. Typically, transformations in mathematics
-and physics have interesting structure; for example, translations in
+same in all inertial frames. Typically, the 
+transformations under which invariants are preserved have interesting structure: 
+for example, translations in
 the Euclidean plane form an abelian group.
 
-%Types express properties of programs. The statement $\mathit{b}:
-%\tyBool$ tells us that $\mathit{b}$ has value $\mathtt{true}$ or
-%$\mathtt{false}$; the statement
-%$\mathit{pos}:\tyReal\tyProduct\tyReal$ says that $\mathit{pos}$ is a
-%pair whose components are real numbers. More interestingly, the
-%statement $\mathit{f} : \forall
-%X. \tyPrim{list}X\tyArr\tyPrim{list}X$ tells us that if we apply
-%$\mathit{f}$ to a list whose elements satisfy some property $\psi$,
-%then the result is a list whose elements also satisfy $\psi$.
+Inspired by this connection, we study type systems that
+capture rich invariants through types indexed by attributes with
+algebraic structure.  For example, in computational geometry, points
+in the plane can be indexed by attributes representing affine
+transformations; in information-flow security, computations can be
+indexed by principals; in differential privacy, types can be indexed
+by `distance'. Types that are polymorphic over such indices induce
+invariance properties and abstraction barriers beyond those introduced
+by their unindexed versions, as we shall illustrate.  This generalises
+previous work by the third author on types parameterized by
+units-of-measure, whose invariance properties relate to changes of
+units, or \emph{scaling}~\cite{kennedy97relational}.
 
-%Unary predicates only go so far, and for polymorphic types
-%particularly do not express everything that follows from possessing
-%that type.  Imagine a function $\mathit{oddRev}$ that reverses any
-%list whose first element is an odd number, otherwise behaving as the
-%identity. It certainly preserves properties of the elements -- but
-%truly parametric polymorphism should tell us more.
+Our motivations for studying algebraically-indexed types are
+threefold. First, we believe that, as with
+units-of-measure~\cite{fsharp}, practical programming language
+extensions will follow. For example, in computational geometry and
+graphics, attributes on points, vectors, and other geometric types
+could be used to prevent the mixing of different coordinate systems,
+or `frames'. Second, type-based static analyses can be based on
+indexed types, for example, in effect systems~\cite{benton06reading},
+and, more speculatively, in continuity
+analysis~\cite{chaudhuri10continuity}.  Finally, we believe that
+expressing algebraic invariants through types has the potential to
+offer slick proof techniques for mechanized
+mathematics. Harrison has applied the
+invariance properties of geometric primitives to create elegant proofs
+in geometry, based on `without loss of
+generality' principles~\cite{harrison09without}. The invariance properties are expressed 
+and propagated using ad-hoc tactics; our types offer a more principled
+means of achieving the same end, and the `wlog' principle itself
+is expressed through type isomorphisms.
 
-%To obtain stronger properties we must \emph{relate} multiple
-%executions of the same code. The statement $\mathit{f}:\forall
-%X. \tyPrim{list}X\tyArr\tyPrim{list}X$ then says that any change in
-%representation of the list elements supplied to $\mathit{f}$ is
-%reflected in the elements in the result.  Formally, for any function
-%$g$, we have $\mathit{f}(\mathrm{map}\:\mathit{g}\:\mathit{x}) =
-%\mathrm{map}\:\mathit{g}\:(\mathit{f}\:\mathit{x})$.  This is the
-%essence of \emph{relational parametricity} as formulated by
-%Reynolds~\cite{reynolds83types}. Its consequences are striking.  It
-%immediately tells us that $\mathit{oddRev}$ cannot be given a
-%polymorphic type: simply consider the change-of-representation given
-%by $\lambda x. x+1$.  But more profoundly it can be used to prove
-%blah blah (initial algebras, term reps, non-inhabitation, etc.).
-
-In previous work the third author studied relational parametricity for
-types parameterized by units-of-measure, whose invariance 
-properties relate to changes of units, or \emph{scaling}~\cite{kennedy97relational}. 
-In this paper we investigate the general case of types indexed by attributes
-with algebraic structure. For example, in computational geometry, points in
-the plane can be indexed by a coordinate frame; in information-flow
-security, computations are indexed by principals; in differential privacy, types
-can be indexed by `distance'. Types that are
-polymorphic over such indices induce invariance properties and
-abstraction barriers beyond those introduced by their unindexed
-versions, as we now illustrate.
+We follow the mantra \emph{semantics first, syntax later} in studying
+types with algebraic structure. We have not yet built a practical
+programming language that supports algebraically-indexed types; nor
+have we designed type checking, type inference, or static analysis
+algorithms.  But when we do so, the semantics will guide us. The fact
+that zero is polymorphic in units-of-measure (it can be given type $\forall
+u.\tyPrim{real}u$) whereas other constants are dimensionless (having
+type $\tyPrim{real}1$) is justified by the invariance properties
+induced by the types: zero is invariant under scaling, other constants
+are not. For less trivial constants and operations, the appropriate types are
+not so apparent, as we shall see, but invariance properties expressed by
+the semantics guide us in assigning appropriate types. Semantics does not lie!
 
 \paragraph{Invariance}
+To illustrate type-induced invariance properties, consider
+2-dimensional geometry. 
 A function $\mathit{areaTri}$ that computes the area of a triangle
@@
 of its radius, i.e., $\mathrm{areaCircle}(k r) = k^2\cdot
-\mathrm{areaCircle}(r)$ for any $k\neq 0$ (i.e., any $k$ in the
-one-dimensional general linear group $\GL{1}$).  Here, $s$ 
+\mathrm{areaCircle}(r)$ for any $k\neq 0$ (the `sorts' $\SynTransl{2}$ and
+$\SynGL{1}$ will be explained later).  Here, $s$ 
 can be interpreted as the \emph{units of measure} of the argument to
@@
 related according to their indices.  For example, values $v$ and $w$
-of type $\tyPrim{real}\alpha$ are related when $v$ ``scales to'' $w$
-according to an interpretation of $\alpha$ (i.e., $w=\rho(\alpha)\cdot
+of type $\tyPrim{real}s$ are related when $v$ ``scales to'' $w$
+according to an interpretation of $s$ (i.e., $w=\rho(s)\cdot
 v$).  Values of polymorphic type are related exactly when they are
@@
 
-\vspace*{0.1in}
-
-As it happens, the above examples can all be expressed using unary
-functions rather than binary relations: we can write functions that
-scale, or translate, one value to another according to their type,
-even for higher-order function types. Moreover, the interpretation of
-indices can be built up compositionally, by recursion over the
-syntactic structure. For example, we have $\rho(\alpha\cdot\beta) =
-\rho(\alpha)\cdot\rho(\beta)$ where product of units-of-measure is
-interpreted by product of scale factors. For more sophisticated
-applications, however, we need the fully generality of binary
-relations. As is the case with type polymorphism, it
-provides the appropriate meaning for nested quantifiers in types such
-as
-$(\forall s\mathord:\SynGL{1}.\tyPrim{real} s\to\tyPrim{real} s)\to\tyBool$.
-More interestingly, particular applications sometimes require
-relations that are not built compositionally---we can, for instance,
-use such relations to prove that there are no non-trivial programs of
-type
-$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s}\to\tyPrim{real}s$.
-Finally, there are interesting applications in which even the basic
-relations are not functional. In a type system in which the index in
+As it happens, the relational interpretations given above are
+functional, relating one value uniquely to another. Other applications
+make use of primitive relations that are not simple functions.
+For example, in a type system in which the index in
 $\tyPrim{real}s$ is interpreted not as a unit of measure, but as
@@
 We describe a relational semantics for the type system and prove an
-analogue of Reynolds' abstraction theorem. We give precise conditions
-on sets of relation environments under which our abstraction theorem
-holds.
+analogue of Reynolds' abstraction theorem, for a given \emph{model}
+of index sorts and relational interpretation of primitive types.
+We prove that the semantics soundly approximates contextual equivalence.
 \item
@@
 results, and derive interesting type isomorphisms.
+\item
+We improve on the earlier semantics of
+units-of-measure~\cite{kennedy97relational} in a number of ways.  By
+extending the language of units with an `absolute value' operation, we
+can give more precise types and obtain more general invariance
+properties.  The relational interpretation for units is both simpler
+and more flexible, and we derive slicker proofs of non-definability,
+and new results.  Our notion of type isomorphism is stronger than
+previously, being based on contextual equivalence.
 \end{itemize}

### 2012-10-25 paper/geometry-twocolumn.tex
Various little fixes to Sections 1,2,4 and 5
Section 1:
- Make the diagram in Figure 1 a little tighter, and add a reference
  in the text (in the fourth bullet point in the contributions).

Section 2:
- Introduce to the \floor and semantic brackets explicitly in Section
  2.

- Change the brackets around the intended semantics of affComb

- Insert a line break after affComb in Figure 3 and remove the
  semantics of affComb.

Section 4:
- Fix a small typo in the type isomorphism for scalings ("A[1/s]"
  instead of just "[1/s]")

Section 5:
- Two small wording improvements

@@
 and a relational interpretation as a binary relation on the
-index-erasure interpretation. To give such interpretations for the
+index-erasure interpretation. We denote the index-erasure and
+relational interpretations with the notations $\tySem{\cdot}$ and
+$\rsem{\cdot}{}$ respectively. To give such interpretations for the
 types $\tyPrim{vec}{e}$ and $\forall t\mathord:\SynTransl{2}.A$, we
@@
     \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\ 
-    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} +
+    \tmSem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} +
     r\vec{v_2} 
@@
 \end{displaymath}
-It can be verified by hand that %the intended denotation
-$\sem{\mathrm{affComb}}$ is invariant under all changes of origin, as
+It can be verified by hand that the index-erasure interpretation %the intended denotation
+$\tmSem{\mathrm{affComb}}$ is invariant under all changes of origin, as
 dictated by its type.
@@
 scalar multiplication and dot product. These
- are shown in Figure~\ref{fig:vec-ops}, which summarises the most general types 
+ are shown in \autoref{fig:vec-ops}, which summarises the most general types 
 of all the vector operations that we have described.
@@
     \begin{array}[t]{@{}l}\forall t \mathord:
-    \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to
-    \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\ 
-    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} +
-    r\vec{v_2} 
-    \end{array}
+    \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{real}{1} \to
+    \tyPrim{vec}{t} \to \tyPrim{vec}{t}
+    \end{array} \\
     (\cdot) & :
 \begin{array}[t]{@{}l}
-    \forall s \mathord: \SynGL{1}, O \mathord: \SynOrth{2}.\ \\
-    \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to \tyPrim{real}{s^2}
+    \forall s \mathord: \SynGL{1}, O \mathord: \SynOrth{2}.\ \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\\
+    \hspace{2cm}\tyPrim{vec}{\iota_1(s)\iota_O(O),0}\to \tyPrim{real}{s^2}
 \end{array}\\

### 2012-10-25 paper/section-general-nondef.tex
Various little fixes to Sections 1,2,4 and 5
Section 1:
- Make the diagram in Figure 1 a little tighter, and add a reference
  in the text (in the fourth bullet point in the contributions).

Section 2:
- Introduce to the \floor and semantic brackets explicitly in Section
  2.

- Change the brackets around the intended semantics of affComb

- Insert a line break after affComb in Figure 3 and remove the
  semantics of affComb.

Section 4:
- Fix a small typo in the type isomorphism for scalings ("A[1/s]"
  instead of just "[1/s]")

Section 5:
- Two small wording improvements

@@
   for all $k$, $(x,x) \in \rsem{\tyPrim{X}{e_k}}(i_1,...,i_m)$ since
-  $e_k$ is a member of the set of terms we are using to generate
-  terms. Now, $(\tmSem{M} \eta_{\mathit{ops}}~x...x,
+  each $e_k$ is a member of the set we are using to generate
+  terms. Now $(\tmSem{M} \eta_{\mathit{ops}}~x...x,
   \tmSem{M}\eta_{\mathit{ops}}~x...x) \in \rsem{\tyPrimNm{X}}(e)$ and
@@
 \paragraph{Abelian Group Indexed Types} Kennedy
-\cite{kennedy97relational} gave a general characterisation of
+\cite{kennedy97relational} has given a general characterisation of
 definability at first-order in the case of abelian group indexing in

### 2012-10-25 paper/section-introduction.tex
Various little fixes to Sections 1,2,4 and 5
Section 1:
- Make the diagram in Figure 1 a little tighter, and add a reference
  in the text (in the fourth bullet point in the contributions).

Section 2:
- Introduce to the \floor and semantic brackets explicitly in Section
  2.

- Change the brackets around the intended semantics of affComb

- Insert a line break after affComb in Figure 3 and remove the
  semantics of affComb.

Section 4:
- Fix a small typo in the type isomorphism for scalings ("A[1/s]"
  instead of just "[1/s]")

Section 5:
- Two small wording improvements

@@
   \begin{tikzpicture}
-  \node at (-3,1) [rectangle,style={align=center}] {Algebraic \\ Theory};
-  \node at (-2,1) {$+$};
-  \node at (-0.5,1) [rectangle,style={align=center}] {Index-Erasure \\ Interpretation};
-  \node at (1,1) {$+$};
-  \node at (2.6,1) [rectangle,style={align=center}] {Relational \\ Interpretation};
+  \node at (-3,0.8) [rectangle,style={align=center}] {Algebraic \\ Theory};
+  \node at (-2,0.8) {$+$};
+  \node at (-0.5,0.8) [rectangle,style={align=center}] {Index-Erasure \\ Interpretation};
+  \node at (1,0.8) {$+$};
+  \node at (2.6,0.8) [rectangle,style={align=center}] {Relational \\ Interpretation};
 
-  \draw plot (-4,0.8) .. controls (-4,0) and (-0.4,1) .. (-0.4,0)
-             -- (-0.6,0) -- (0,-0.3) -- (0.6,0) --
-             (0.4,0) .. controls (0.4,1) and (4,0) .. (4,0.8);
+  \draw plot (-4,0.6) .. controls (-4,-0.2) and (-0.4,0.8) .. (-0.4,-0.2)
+             -- (-0.6,-0.2) -- (0,-0.5) -- (0.6,-0.2) --
+             (0.4,-0.2) .. controls (0.4,0.8) and (4,-0.2) .. (4,0.6);
 
@@
 consequences of our abstraction theorem, prove non-definability
-results, and derive interesting type isomorphisms.
+results, and derive interesting type isomorphisms. \autoref{fig:summary}
+illustrates the central position of our analogue of Reynolds' Abstraction
+Theorem in these results.
 \item

### 2012-10-26 paper/geometry-twocolumn.tex
Put commas after all the 'i.e.'s

@@
 Given two points that are invariant with respect to the same change of
-origin---i.e.~two values of type $\tyPrim{vec}{t}$---we can use
+origin---i.e.,~two values of type $\tyPrim{vec}{t}$---we can use
 subtraction to compute their offset:

### 2012-10-26 paper/section-general-framework.tex
Put commas after all the 'i.e.'s

@@
 An index context $\Delta = (i_1\mathord:s_1,\ldots,i_n\mathord:s_n)$ is
-interpreted as cartesian product \emph{i.e.} $\semSort{\Delta} \isDefinedAs \semSort{s_1}
+interpreted as cartesian product, i.e., $\semSort{\Delta} \isDefinedAs \semSort{s_1}
 \times\cdots \times \semSort{s_n}$. 

### 2012-10-26 paper/section-general-nondef.tex
Put commas after all the 'i.e.'s

@@
   types. Hence, by \thmref{thm:abstraction}, we know that for all
-  terms (i.e. elements of the free model over $\Delta$)
+  terms (i.e., elements of the free model over $\Delta$)
   $e'_1\mathord:s_1,...,e'_m\mathord:s_m$:
@@
 generated by the index expression $t + t$ corresponds to the even
-integers. The result now follows simply because $1$ (i.e. the
+integers. The result now follows simply because $1$ (i.e., the
 interpretation of $t$) is not an even number.

### 2012-10-26 paper/paper.tex
floor notation
finally proved iso
Add a paper-overview picture
intro, contributions
Remove hyphen in 'algebraically-indexed' everywhere

@@
             citecolor=citationcolour,
-            pdftitle=Abstraction and Invariance for Algebraically-Indexed Types,
+            pdftitle=Abstraction and Invariance for Algebraically Indexed Types,
             pdfauthor={Robert Atkey, Patricia Johann, Andrew Kennedy},
@@
 \usepackage{stmaryrd}
+\usepackage{pgf}
+\usepackage{tikz}
+\usepgflibrary{shapes.geometric}
+\usepgflibrary{shapes.misc}
+\usepgflibrary{shapes.multipart}
+\usetikzlibrary{shapes.geometric}
+\usetikzlibrary{shapes.misc}
+\usetikzlibrary{shapes.multipart}
 
@@
 \newcommand{\relEnv}[1]{\mathcal{#1}}
-%\newcommand{\tySem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}}}
-%\newcommand{\ctxtSem}[1]{\llbracket #1 \rrbracket^{\mathcal{C}}}
-%\newcommand{\tmSem}[1]{\llbracket #1 \rrbracket^{\mathit{tm}}}
-%\newcommand{\tyPrimSem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}_0}}
-\newcommand{\tySem}[1]{\left| #1\right|}
-\newcommand{\ctxtSem}[1]{\left| #1\right|}
-\newcommand{\tmSem}[1]{\left| #1\right|}
-\newcommand{\tyPrimSem}[1]{\left| #1\right|}
+\newcommand{\tySem}[1]{\left\lfloor #1 \right\rfloor}
+\newcommand{\ctxtSem}[1]{\left\lfloor #1 \right\rfloor}
+\newcommand{\tmSem}[1]{\left\lfloor{#1}\right\rfloor}
+\newcommand{\tyPrimSem}[1]{\left\lfloor #1 \right\rfloor}
+%\newcommand{\tySem}[1]{\left| #1\right|}
+%\newcommand{\ctxtSem}[1]{\left| #1\right|}
+%\newcommand{\tmSem}[1]{\left| #1\right|}
+%\newcommand{\tyPrimSem}[1]{\left| #1\right|}
 \newcommand{\rsem}[1]{\llbracket #1 \rrbracket}
@@
 \bibitem{kennedy97relational} 
-A.~Kennedy. Relational Parametricity and Units of Measure.
+A.~J.~Kennedy. Relational Parametricity and Units of Measure.
 Proceedings, POPL, pp. 442-455, 1997.
+
+\bibitem{fsharp}
+A.~J.~Kennedy. Types for Units-of-Measure: Theory and Practice.
+Central European Functional Programming school (CEFP), pp. 268--305, LNCS vol. 6299, 2010.
 

### 2012-10-26 paper/section-general-framework.tex
Remove hyphen in 'algebraically-indexed' everywhere

@@
 %\fixme{Make sure that the semantics is motivated}
-Having defined the syntax of algebraically-indexed types and terms,
+Having defined the syntax of algebraically indexed types and terms,
 we turn to their denotational interpretation.  We first define an
@@
 An index context $\Delta = (i_1\mathord:s_1,\ldots,i_n\mathord:s_n)$ is
-interpreted as cartesian product, i.e., $\semSort{\Delta} \isDefinedAs \semSort{s_1}
+interpreted as cartesian product,x i.e., $\semSort{\Delta} \isDefinedAs \semSort{s_1}
 \times\cdots \times \semSort{s_n}$. 

### 2012-10-26 paper/section-introduction.tex
Remove hyphen in 'algebraically-indexed' everywhere

@@
 
-Our motivations for studying algebraically-indexed types are
+Our motivations for studying algebraically indexed types are
 threefold. First, we believe that, as with
@@
 types with algebraic structure. We have not yet built a practical
-programming language that supports algebraically-indexed types; nor
+programming language that supports algebraically indexed types; nor
 have we designed type checking, type inference, or static analysis

### 2012-10-26 paper/section-metric.tex
Refine statements of \eta_ops self-related-ness lemmas in Sections 6 and 7
Remove hyphen in 'algebraically-indexed' everywhere

@@
 \begin{lemma}\label{lem:metric-environments}
-  For all $\Delta$ and $\rho \in \rsem{\Delta}$, $(\eta_\Metric,
-  \eta_\Metric) \in \rsem{\Gamma}{\rho}$.
+  $(\eta_\Metric, \eta_\Metric) \in \rsem{\Gamma_\Metric}{}*$.
 \end{lemma}
@@
 their central concept of $c$-sensitivity (for functions on the reals)
-as an algebraically-indexed type: $\forall
+as an algebraically indexed type: $\forall
 \epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \to

### 2012-10-26 paper/geometry-twocolumn.tex
Patty's fixes to Section 2

@@
 change of data representation. For example, if $(0,0)$ and $(10,20)$
-are two origins, then the tuple $(1,1)$ with respect to $(0,0)$ the
+are two origins, then the tuple $(1,1)$ with respect to $(0,0)$ and the
 tuple $(11,21)$ with respect to $(10,20)$ represent \emph{the same
@@
 and that $f$ is invariant under the choice of origin.  Reynolds'
-relational approach formalises this intention precisely: for any
+relational approach formalises this intention precisely. For any
 $\vec{t} \in \Transl{2}$: % (i.e.~any change of origin), the function
@@
 \end{equation}
-Unfolding the definition of $R_{\vec{t}}$ gives the equivalent formulation:
+Unfolding the definition of $R_{\vec{t}}$ gives the equivalent formulation,
 %\statementref{eq:f-preserve-rel-frame} is equivalent to: 
-%for all $\vec{t} \in \Transl{2}$,
+again for all $\vec{t} \in \Transl{2}$:
 \begin{equation}\label{eq:f-invariant-frame}
@@
 of types. By contrast, in our statements of geometric invariance in
-\autoref{sec:type-system-geom-intro} we did not quantify over all
+\autoref{sec:oiri} we did not quantify over all
 relations, but instead quantified over all changes of origin and used
@@
 %\begin{displaymath}
-$  \mathrm{f} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t}
+$  \mathit{f} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t}
   \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}$.
@@
 of all such maps on $\mathbb{R}^2$ forms the \emph{General Linear}
-group $\GL{2}$ which we represent in our language by a new indexing
+group $\GL{2}$, which we represent in our language by a new indexing
 sort $\SynGL{2}$ with (non-abelian) group structure that we will write

### 2012-10-26 paper/section-instantiations.tex
Various little fixes to Sections 1,2,4 and 5
Section 1:
- Make the diagram in Figure 1 a little tighter, and add a reference
  in the text (in the fourth bullet point in the contributions).

Section 2:
- Introduce to the \floor and semantic brackets explicitly in Section
  2.

- Change the brackets around the intended semantics of affComb

- Insert a line break after affComb in Figure 3 and remove the
  semantics of affComb.

Section 4:
- Fix a small typo in the type isomorphism for scalings ("A[1/s]"
  instead of just "[1/s]")

Section 5:
- Two small wording improvements
Put commas after all

@@
 that $\tmSem{\transup A t v}(\opsEnv,\vec 0)(y) = y$, and so
-$\ctxteq\DeltaWithB{y\mathord:Y}{J(I(y))}y Y$ follows.
+$\ctxteq\DeltaWithB{y\mathord:Y}{I(J(y))}y Y$ follows.
 
@@
 \forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A &
-\cong & [1/s]\times \forall s\mathord:\SynGL{1}.A\label{eqn:scaleiso}
+\cong & A[1/s]\times \forall s\mathord:\SynGL{1}.A\label{eqn:scaleiso}
 \end{eqnarray}
@@
 There are many functions~$f: \mathbb R \to \mathbb R$
-that satisfy its relational interpretation, i.e. for which $f(k^2\cdot x) = k\cdot f(x)$ for 
+that satisfy its relational interpretation, i.e., for which $f(k^2\cdot x) = k\cdot f(x)$ for 
 any $k\neq 0$.
@@
 
-You can think of $z$ as tracking whether or not exponents on scale
+We think of $z$ as tracking whether or not exponents on scale
 parameters are divisible by 2: all the primitive operations in
@@
 
-Here $q$ is tracking the exponents on scale parameters, with
+Here, $q$ is tracking the exponents on scale parameters, with
 $\semPrimType{\tyPrimNm{real}}(q)$ relating only zero to itself

### 2012-10-26 paper/section-general-nondef.tex
Integrate Patty's fixes into Section 5

@@
 the index-level operations (as, for example, in
-\autoref{fig:real-ops}), then this theorems yields a characterisation
+\autoref{fig:real-ops}), then this theorem yields a characterisation
 of definable terms, since the construction of $e$ from $e_1,...,e_n$
@@
   free model over the variables in $\Delta$ constructed from the
-  syntax. This model assigns to each sort $s$ the set of index
-  expressions $\{ e \sepbar \Delta \vdash e : s \}/\mathord\equiv$,
+  syntax. This model assigns to each sort $s$ the set $\{ e \sepbar
+  \Delta \vdash e : s \}/\mathord\equiv$ of index expressions
   quotiented by index expression equality. Index operations are
@@
   \Gamma_{\mathit{ops}}$, we have $(\eta_{\mathit{ops}},
-  \eta_{\mathit{ops}}) \in \rsem{\Gamma_{\mathit{ops}}}*$ due to the
-  fact that all members of $\Gamma_{\mathit{ops}}$ have well-generated
+  \eta_{\mathit{ops}}) \in \rsem{\Gamma_{\mathit{ops}}}*$ because 
+  all members of $\Gamma_{\mathit{ops}}$ have well-generated
   types. Hence, by \thmref{thm:abstraction}, we know that for all

### 2012-10-26 paper/section-general-framework.tex
Integrate Patty's comments on Section 3
- fix \iota_o -> \iota_O

- fix add explanation of \forall \Delta notation just before defn on
  ctxt equivalence.
Pedantic fixes to the typing rules
Some more fixes from Patty to Sections 3 and 4

@@
       \indexOpArity(^{-1_G}) & ([G], G) \\
-      \indexOpArity(\iota_o) & ([\SynOrth{2}], \SynGL{2}) &
+      \indexOpArity(\iota_O) & ([\SynOrth{2}], \SynGL{2}) &
       \indexOpArity(\iota_1) & ([\SynGL{1}], \SynGL{2}) \\
@@
       &
-      \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:S. \sigma_{i\mathord:s}^*A
+      \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:s. \sigma_{i\mathord:s}^*A
     \end{array}
@@
   determinant of an orthgonal transformation is always $1$, so we also
-  assume $\abs{\det(\iota_OO)} \stackrel{ax}\equiv 1$. Likewise, scaling
-  maps have a determined determinant: $\det(\iota_1(s))
-  \stackrel{ax}\equiv s \cdot s$.
-\qed
+  assume $\abs{\det(\iota_OO)} \stackrel{ax}\equiv 1$. Similarly,
+  scaling maps have a determinant expressible in terms of other
+  operations: $\det(\iota_1(s)) \stackrel{ax}\equiv s \cdot s$. We
+  also assume the axiom $|s^2| \stackrel{ax}\equiv s^2$.  \qed
 \end{example*}
@@
     \inferrule* [right=Unit]
-    { }
+    {\Delta \vdash \Gamma \isCtxt}
     {\Delta; \Gamma \vdash * : 1}
@@
     \inferrule* [right=ExPack]
-    {\Delta; \Gamma \vdash M : (\id_\Delta, e)^*A \\ \Delta, i\mathord:s \vdash A \isType}
+    {\Delta; \Gamma \vdash M : (\id_\Delta, e)^*A \\ \Delta \vdash e : s \\ \Delta, i\mathord:s \vdash A \isType}
     {\Delta; \Gamma \vdash \langle[e], M\rangle: \exists i\mathord:s. A}
@@
 \paragraph{Contextual Equivalence.}
-We use our index-erasure semantics to define when a pair of terms
-are contextually equivalent with respect to syntactically defined
-contexts, following Hofmann \cite{hofmann08correctness}.
-
+We use our index-erasure semantics to define when a pair of terms are
+contextually equivalent with respect to syntactically defined
+contexts, following Hofmann \cite{hofmann08correctness}.  Given an
+index context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, we write
+$\forall \Delta.A$ for $\forall i_1\mathord:s_1.\dots\forall
+i_n\mathord:s_n.A$, and similarly for $\Lambda\Delta.M$ and $\lambda
+\Gamma.M$.
 \begin{definition}[Contextual Equivalence]\label{defn:ctxt-equiv}
@@
 determinant) is interpreted by the intended semantic operation, and
-clearly satisfy the axioms in \exref{ex:two-dim-geo-axioms}. 
+clearly satisfies the axioms in \exref{ex:two-dim-geo-axioms}. 
 \qed
@@
 
-  Let $\opsCtxt$ be the context of primitive operations and $\opsEnv$  its interpretation as in
-  \defref{defn:ctxt-equiv}. Fix a model and relational interpretation of primitive types so that
+  Let $\opsCtxt$ be the context of primitive operations and $\opsEnv$  its interpretation 
+%as in
+%  \defref{defn:ctxt-equiv}.
+Fix a model and relational interpretation of primitive types so that
   $(\opsEnv, \opsEnv) \in  \rsem{\opsCtxt} *$.

### 2012-10-26 paper/section-instantiations.tex
Some more fixes from Patty to Sections 3 and 4
Remove hyphens from unit(s)-of-measure

@@
 \label{sec:instantiations}
-We now instantiate our general framework %of the previous section with
+We now instantiate our general framework %of the previous section 
+with
 the indexing theory of \autoref{sec:motivating-examples}, and present
@@
 type isomorphisms; previous work on
-units-of-measure~\cite{kennedy97relational} relates these to
+units of measure~\cite{kennedy97relational} relates these to
 Buckingham's theorem from dimensional analysis. Here we consider the
@@
 For the group $\GL{1}$ of scalings we can treat
-a real-valued argument to a function as a unit-of-measure with
+a real-valued argument to a function as a unit of measure with
 which to scale the result. 

### 2012-10-26 paper/section-introduction.tex
Remove hyphens from unit(s)-of-measure

@@
 previous work by the third author on types parameterized by
-units-of-measure, whose invariance properties relate to changes of
+units of measure, whose invariance properties relate to changes of
 units, or \emph{scaling}~\cite{kennedy97relational}.
@@
 threefold. First, we believe that, as with
-units-of-measure~\cite{fsharp}, practical programming language
+units of measure~\cite{fsharp}, practical programming language
 extensions will follow. For example, in computational geometry and
@@
 algorithms.  But when we do so, the semantics will guide us. The fact
-that zero is polymorphic in units-of-measure (it can be given type $\forall
+that zero is polymorphic in units of measure (it can be given type $\forall
 u.\tyPrim{real}u$) whereas other constants are dimensionless (having
@@
 We improve on the earlier semantics of
-units-of-measure~\cite{kennedy97relational} in a number of ways.  By
+units of measure~\cite{kennedy97relational} in a number of ways.  By
 extending the language of units with an `absolute value' operation, we

### 2012-10-26 paper/paper.tex
Capitalise Abstraction Theorem everywhere

@@
   distances--for analysis of analytic uniform continuity
-  properties. Following Reynolds, we prove a general abstraction
-  theorem that covers all these instances. Consequences of our
-  abstraction theorem include free theorems expressing invariance
+  properties. Following Reynolds, we prove a general Abstraction
+  Theorem that covers all these instances. Consequences of our
+  Abstraction Theorem include free theorems expressing invariance
   properties of programs, type isomorphisms based on invariance

### 2012-10-26 paper/section-general-framework.tex
Capitalise Abstraction Theorem everywhere

@@
 `model' of the algebraic theory (\autoref{sec:relational-semantics}),
-prove the central Abstraction theorem
+prove the central Abstraction Theorem
 and use the relational semantics
@@
 every well-typed term is related to itself in the relational
-interpretation of its type: this is the abstraction theorem for every
+interpretation of its type: this is the Abstraction Theorem for every
 instantiation of our general framework.
@@
 
-We now state the abstraction theorem for well-typed terms. To state
+We now state the Abstraction Theorem for well-typed terms. To state
 and prove this theorem for open terms, we extend the

### 2012-10-26 paper/section-info-flow.tex
Refine statements of \eta_ops self-related-ness lemmas in Sections 6 and 7
Capitalise Abstraction Theorem everywhere

@@
 $\alpha_l \to A$ for some free type variable $\alpha_l$, and using
-Reynolds' abstraction theorem to prove information flow
+Reynolds' Abstraction Theorem to prove information flow
 properties. For example, if the type variable $\alpha_H$ represents
@@
 \begin{lemma}\label{lem:environments-information-flow}
-  For all $\Delta$ and $\rho \in \rsem{\Delta}$, $(\eta_\infoflow, \eta_\infoflow) \in \rsem{\Gamma_\infoflow}\rho$.
+  $(\eta_\infoflow, \eta_\infoflow) \in \rsem{\Gamma_\infoflow}{}*$.
 \end{lemma}

### 2012-10-26 paper/section-instantiations.tex
Capitalise Abstraction Theorem everywhere

@@
 \emph{dyadic} number).  
-Applying the abstraction theorem and expanding, we have that for
+Applying the Abstraction Theorem and expanding, we have that for
 any

### 2012-10-26 paper/section-introduction.tex
Add hyperlink to Abstraction Theorem in the introduction text
Capitalise Abstraction Theorem everywhere

@@
 polymorphism is \emph{relational parametricity}, whose central result
-is Reynolds' abstraction theorem~\cite{reynolds83types}. Its striking
+is Reynolds' Abstraction Theorem~\cite{reynolds83types}. Its striking
 consequences include the well-known ``free theorems'' for polymorphic
@@
 We describe a relational semantics for the type system and prove an
-analogue of Reynolds' abstraction theorem, for a given \emph{model}
+analogue of Reynolds' Abstraction Theorem, for a given \emph{model}
 of index sorts and relational interpretation of primitive types.
@@
 For each of our main examples we deduce free theorems that are
-consequences of our abstraction theorem, prove non-definability
+consequences of our Abstraction Theorem, prove non-definability
 results, and derive interesting type isomorphisms. \autoref{fig:summary}
 illustrates the central position of our analogue of Reynolds' Abstraction
-Theorem in these results.
+Theorem (\thmref{thm:abstraction}) in these results.
 \item

### 2012-10-26 paper/section-general-framework.tex
Tiny fixes

@@
 primitive types: for each each primitive type
-$\tyPrimNm{X}$, a relational interpretation parameterised by elements
+$\tyPrimNm{X}$, its relational interpretation is parameterised by elements
 from the model: $\semPrimType{\tyPrimNm{X}} :

### 2012-10-28 paper/section-introduction.tex
Reordered intro a little; added paper structure section

@@
 Inspired by this connection, we study type systems that
-capture rich invariants through types indexed by attributes with
+capture rich invariants in types indexed by attributes with
 algebraic structure.  For example, in computational geometry, points
@@
 
-Our motivations for studying algebraically indexed types are
-threefold. First, we believe that, as with
-units of measure~\cite{fsharp}, practical programming language
-extensions will follow. For example, in computational geometry and
-graphics, attributes on points, vectors, and other geometric types
-could be used to prevent the mixing of different coordinate systems,
-or `frames'. Second, type-based static analyses can be based on
-indexed types, for example, in effect systems~\cite{benton06reading},
-and, more speculatively, in continuity
-analysis~\cite{chaudhuri10continuity}.  Finally, we believe that
-expressing algebraic invariants through types has the potential to
-offer slick proof techniques for mechanized
-mathematics. Harrison has applied the
-invariance properties of geometric primitives to create elegant proofs
-in geometry, based on `without loss of
-generality' principles~\cite{harrison09without}. The invariance properties are expressed 
-and propagated using ad-hoc tactics; our types offer a more principled
-means of achieving the same end, and the `wlog' principle itself
-is expressed through type isomorphisms.
-
-We follow the mantra \emph{semantics first, syntax later} in studying
-types with algebraic structure. We have not yet built a practical
-programming language that supports algebraically indexed types; nor
-have we designed type checking, type inference, or static analysis
-algorithms.  But when we do so, the semantics will guide us. The fact
-that zero is polymorphic in units of measure (it can be given type $\forall
-u.\tyPrim{real}u$) whereas other constants are dimensionless (having
-type $\tyPrim{real}1$) is justified by the invariance properties
-induced by the types: zero is invariant under scaling, other constants
-are not. For less trivial constants and operations, the appropriate types are
-not so apparent, as we shall see, but invariance properties expressed by
-the semantics guide us in assigning appropriate types. Semantics does not lie!
-
 \paragraph{Invariance}
 To illustrate type-induced invariance properties, consider
-2-dimensional geometry. 
-A function $\mathit{areaTri}$ that computes the area of a triangle
-can be assigned the type:
+two-dimensional geometry. 
+In a conventional type system, a function $\mathit{areaTri}$ that computes the area of a triangle
+might be assigned the type:
 $\tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal$.
-But we can also assign it the following more expressive polymorphic
+But in our proposed system we can assign it the following more expressive polymorphic
 type:
@@
 
+\paragraph{Motivations}
+Our motivations for studying algebraically indexed types are
+threefold. First, we believe that, as with
+units of measure~\cite{fsharp}, practical programming language
+extensions will follow. For example, in computational geometry and
+graphics, attributes on points, vectors, and other geometric types
+could be used to prevent the mixing of different coordinate systems,
+or `frames'. Second, type-based static analyses can be based on
+indexed types, for example, in effect systems~\cite{benton06reading},
+and, more speculatively, in continuity
+analysis~\cite{chaudhuri10continuity}.  Finally, we believe that
+expressing algebraic invariants through types has the potential to
+offer slick proof techniques for mechanized
+mathematics. Harrison has applied the
+invariance properties of geometric primitives to create elegant proofs
+in geometry, based on `without loss of
+generality' principles~\cite{harrison09without}. The invariance properties are expressed 
+and propagated using ad-hoc tactics; our types offer a more principled
+means of achieving the same end, and the `wlog' principle itself
+is expressed through type isomorphisms.
+
+We follow the mantra \emph{semantics first, syntax later} in studying
+types with algebraic structure. We have not yet built a practical
+programming language that supports algebraically indexed types; nor
+have we designed type checking, type inference, or static analysis
+algorithms.  But when we do so, the semantics will guide us. The fact
+that zero is polymorphic in units of measure (it can be given type $\forall
+u.\tyPrim{real}u$) whereas other constants are dimensionless (having
+type $\tyPrim{real}1$) is justified by the invariance properties
+induced by the types: zero is invariant under scaling, other constants
+are not. For less trivial constants and operations, the appropriate types are
+not so apparent, as we shall see, but invariance properties expressed by
+the semantics guide us in assigning appropriate types. Semantics does not lie!
+
 % Possible angles of attack:
@@
 For each of our main examples we deduce free theorems that are
-consequences of our Abstraction Theorem, prove non-definability
-results, and derive interesting type isomorphisms. \autoref{fig:summary}
+consequences of our Abstraction Theorem, prove specific non-definability
+results, and derive interesting type isomorphisms. 
+For a large class of first-order types we give a general method for 
+constructing suitable models to prove non-definability results. 
+\autoref{fig:summary}
 illustrates the central position of our analogue of Reynolds' Abstraction
 Theorem (\thmref{thm:abstraction}) in these results.
-\item
+\end{itemize}
 We improve on the earlier semantics of
@@
 and new results.  Our notion of type isomorphism is stronger than
-previously, being based on contextual equivalence.
-\end{itemize}
+before, being based on contextual equivalence.
+
 We have fully formalised our framework and most examples in Coq,
@@
 \\{\small \url{https://github.com/bobatkey/algebraically-indexed-types}}
+\subsection{Structure of paper}
+Our paper is structured as follows. In \autoref{sec:motivating-examples}
+we present an extended case study of two-dimensional geometry, with semi-formal
+description of types and results.
+In \autoref{sec:a-general-framework} we describe fully formally a general framework
+for algebraically-indexed types, and prove the Abstraction Theorem and
+soundness for semantic equivalence. \autoref{sec:instantiations} presents a number
+of applications of the theory to 2-d geometry, including free theorems, type isomorphisms
+and non-definability. \autoref{sec:general-nondef} develops a more general technique 
+for proving non-definability results. \autoref{sec:information-flow} presents
+the application of algebraically indexed types to information flow security, and
+\autoref{sec:metric-types} applies it to types indexed by distances.
+Finally, \autoref{sec:discussion} discusses related work and future plans.
 

### 2012-10-29 paper/paper.tex
Tiny fixes
Make the references more consistent

@@
 R. Atkey.  Syntax for Free: Representing Syntax with Binding Using
-Parametricity.  Proceedings, Typed Lambda Calculi and Applications,
+Parametricity.  Proceedings, TLCA,
 pp. 35-49, 2009.
@@
 Reading, Writing and Relations.
-Proceedings, Asian Symposium on Programming Languages and Systems,
+Proceedings, APLAS,
 pp. 114-130, 2006.
@@
 Processes in Space.
-{\em Programs, Proofs, Processes}: Proc, 6th Conference on Computability in Europe, pp 78-87, LNCS vol. 6158, 2010.
+{\em Programs, Proofs, Processes}: Proceedings, CiE, pp. 78-87, 
+%LNCS vol. 6158, 
+2010.
 
@@
 Without Loss of Generality.
-Proceedings, Theorem Proving in Higher Order Logics,
+Proceedings, TPHOLs,
 pp. 43-59, 2009.
@@
 M.~P.~Jones.  Functional Programming with Overloading and Higher-Order
-Polymorphism. Proceedings, Advanced Functional Programming,
+Polymorphism. Proceedings, AFP,
 pp. 97-136, 1995.
@@
 \bibitem{wadler89theorems}
-P.~Wadler.  Theorems for Free!.  Proceedings, International Conference
-on Functional Programming Languages and Computer Architecture,
+P.~Wadler.  Theorems for Free!.  Proceedings, FPCA,
 pp. 347-359, 1989.

### 2012-10-29 paper/section-general-framework.tex
Fix a minor typo

@@
 An index context $\Delta = (i_1\mathord:s_1,\ldots,i_n\mathord:s_n)$ is
-interpreted as cartesian product,x i.e., $\semSort{\Delta} \isDefinedAs \semSort{s_1}
+interpreted as cartesian product, i.e., $\semSort{\Delta} \isDefinedAs \semSort{s_1}
 \times\cdots \times \semSort{s_n}$. 

### 2012-10-29 paper/section-general-framework.tex
Rename 'relational environments' to 'index environments'

@@
 
-A \emph{relational environment} $\rho$ assigns to each index variable $i:s$
+An \emph{index environment} $\rho$ assigns to each index variable $i:s$
 in the context $\Delta$ a value drawn from an interpretation of
@@
 we can derive the composed
-relational environment $\rho \circ \sigma
+index environment $\rho \circ \sigma
 \in \semSort{\Delta'}$ as $\rho \circ \sigma \isDefinedAs (\semIndexExp{e_1}\rho,\ldots,\semIndexExp{e_n}\rho)$.
@@
 \vdash A \isType$ by induction on their derivations, parameterised by
-relational environments 
+index environments 
 $\rho \in \semSort{\Delta}$:
@@
 substitution of index expressions in types can be interpreted by the
-composition of relational environments with 
+composition of index environments with 
 substitutions.

### 2012-10-29 paper/section-info-flow.tex
Rename 'relational environments' to 'index environments'

@@
 follows that if we have an index expression $\phi$ that is interpreted
-as some value other than $\top$ in a relational environment $\rho$,
+as some value other than $\top$ in an index environment $\rho$,
 then for all $x,x' \in \tySem{T_\phi\tyPrimNm{bool}}$, we have $(x,x')

### 2012-10-29 paper/section-general-framework.tex
Add a little bit of explanation about the relational interpretation

@@
 \end{eqnarray*}
-In this definition we have used the following standard %three
+In this definition, the relational interpretation of an application of
+a primitive type $\rsem{\tyPrim{X}{e_1,\ldots,e_n}}\rho$ is built from
+the relational interpretation of the primitive type,
+$\rsem{\tyPrimNm{X}}$, and the interpretation of the index terms
+$e_1,...,e_n$ in the index environment $\rho$. Universal and
+existential quantification are interpretated by the set-theoretic
+intersection and union, respectively, over all extensions of the index
+environment.
+
+We have also used the following standard %three
 constructions on binary relations: if $R \in \Rel(X)$ and $S \in
@@
 \}$, and %. With the same assumptions on $R$ and $S$, the relation
-$R
-\relTimes S \in \Rel(X \times Y)$ is %defined as 
+$R \relTimes S \in \Rel(X \times Y)$ is %defined as
 $\{((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S

### 2012-10-29 paper/section-general-framework.tex
Fix spelling error

@@
 $e_1,...,e_n$ in the index environment $\rho$. Universal and
-existential quantification are interpretated by the set-theoretic
+existential quantification are interpreted by the set-theoretic
 intersection and union, respectively, over all extensions of the index

### 2012-10-29 paper/section-introduction.tex
Fix font of 'areaTri' in introduction

@@
 two-dimensional geometry. 
-In a conventional type system, a function $\mathit{areaTri}$ that computes the area of a triangle
+In a conventional type system, a function $\mathrm{areaTri}$ that computes the area of a triangle
 might be assigned the type:

### 2012-10-29 paper/popl049-atkey.tex
All-in-one tex file submitted to the publisher, and the PDF and PS files

\documentclass{sigplanconf}

\usepackage[usenames]{color}
\definecolor{citationcolour}{rgb}{0,0.4,0.2}
\definecolor{linkcolour}{rgb}{0,0,0.8}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
            urlcolor=linkcolour,
            linkcolor=linkcolour,
            citecolor=citationcolour,
            pdftitle=Abstraction and Invariance for Algebraically Indexed Types,
            pdfauthor={Robert Atkey, Patricia Johann, Andrew Kennedy},
            pdfkeywords={}}  
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
\def\subsubsectionautorefname{Section}

\title{Abstraction and Invariance for Algebraically Indexed Types}

\authorinfo{Robert Atkey\and Patricia Johann}
           {University of Strathclyde}
           {$\{$Robert.Atkey,Patricia.Johann$\}$@strath.ac.uk}

\authorinfo{Andrew Kennedy}
           {Microsoft Research Cambridge}
           {akenn@microsoft.com}

\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{pgf}
\usepackage{tikz}
\usepgflibrary{shapes.geometric}
\usepgflibrary{shapes.misc}
\usepgflibrary{shapes.multipart}
\usetikzlibrary{shapes.geometric}
\usetikzlibrary{shapes.misc}
\usetikzlibrary{shapes.multipart}

\newcommand{\todo}[1]{[\textbf{TODO}: #1]}

\newcommand{\abs}[1]{\lvert #1 \rvert}

\newcommand{\GL}[1]{\mathrm{GL}_#1}
\newcommand{\SynGL}[1]{\mathsf{GL}_#1}
\newcommand{\SE}[1]{\mathsf{SE}_#1}
\newcommand{\SynSE}[1]{\mathsf{SE}_#1}
\newcommand{\Orth}[1]{\mathrm{O}_#1}
\newcommand{\SynOrth}[1]{\mathsf{O}_#1}
\newcommand{\Transl}[1]{\mathrm{T}_#1}
\newcommand{\SynTransl}[1]{\mathsf{T}_#1}
\newcommand{\Scal}{\mathrm{Scal}}
\newcommand{\SynScal}{\mathsf{Scal}}

\newcommand{\sepbar}{\mathrel|}

\newcommand{\Rel}{\mathrm{Rel}}
\newcommand{\relArrow}{\mathrel{\widehat\to}}
\newcommand{\relTimes}{\mathrel{\widehat\times}}
\newcommand{\relSum}{\mathrel{\widehat+}}
\newcommand{\Eq}{\mathrm{Eq}}

\newcommand{\setOfIntegers}{\mathbb{Z}}
\newcommand{\setOfBooleans}{\{\tmTT, \tmFF\}}

\newcommand{\id}{\mathrm{id}}

\newcommand{\SortSet}{\mathit{Sort}}
\newcommand{\IndexOpSet}{\mathit{IndexOp}}
\newcommand{\PrimTypeSet}{\mathit{PrimType}}
\newcommand{\IndexAxiomSet}{\mathit{IndexAx}}

\newcommand{\indexOp}[1]{\textup{\texttt{#1}}}
\newcommand{\idxTms}[2]{\mathrm{IdxTm}(#1 \vdash #2)}
\newcommand{\idxTmAS}[1]{\mathrm{IdxTm}(#1)}

\newcommand{\tyInt}{\texttt{int}}
\newcommand{\tyBool}{\texttt{bool}}
\newcommand{\tyUnit}{\texttt{unit}}
\newcommand{\tyReal}{\texttt{real}}
\newcommand{\tyPrim}[2]{\textup{\texttt{#1}}\langle #2 \rangle}
\newcommand{\tyPrimNm}[1]{\textup{\texttt{#1}}}
\newcommand{\primTyArity}{\mathrm{tyArity}}
\newcommand{\indexOpArity}{\mathrm{opArity}}

\newcommand{\tyArr}{\to}
\newcommand{\tyProduct}{\times}
\newcommand{\tyX}[1]{\texttt{X}\langle #1 \rangle}
\newcommand{\isType}{\textup{\textsf{ type}}}
\newcommand{\isCtxt}{\textup{\textsf{ ctxt}}}
\newcommand{\Ty}{\textsf{Ty}}
\newcommand{\ty}{\textsf{ty}}

\newcommand{\tmTT}{\texttt{tt}}
\newcommand{\tmFF}{\texttt{ff}}

\newcommand{\relEnv}[1]{\mathcal{#1}}
\newcommand{\tySem}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\ctxtSem}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\tmSem}[1]{\left\lfloor{#1}\right\rfloor}
\newcommand{\tyPrimSem}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\rsem}[1]{\llbracket #1 \rrbracket}
\newcommand{\extends}[2]{\mathsf{ext}(#1,#2)}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newcommand{\lemref}[1]{\hyperref[#1]{Lemma~\ref*{#1}}}
\newcommand{\thmref}[1]{\hyperref[#1]{Theorem~\ref*{#1}}}
\newcommand{\defref}[1]{\hyperref[#1]{Definition~\ref*{#1}}}
\newcommand{\propref}[1]{\hyperref[#1]{Proposition~\ref*{#1}}}
\newcommand{\corref}[1]{\hyperref[#1]{Corollary~\ref*{#1}}}
\newcommand{\conref}[1]{\hyperref[#1]{Conjecture~\ref*{#1}}}
\newcommand{\exref}[1]{\hyperref[#1]{Example~\ref*{#1}}}
\newcommand{\statementref}[1]{\hyperref[#1]{Statement~\ref*{#1}}}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\isDefinedAs}{\stackrel{\mathit{def}}=}

\newcommand{\Geom}{\mathit{2D}}

\newtheoremstyle{examplestyle}
  {\topsep}  % space above
  {\topsep}  % space below
  {\normalfont}% name of font to use in the body of the theorem
  {0em}% measure of space to indent
  {\bfseries}% name of head font
  {.}% punctuation between head and body
  {5pt plus 1pt minus 1pt}% space after theorem head; " " = normal interword space
  {}% Manually specify head
\theoremstyle{examplestyle}
\newtheorem{example}{Example}
\newtheorem*{example*}{Example}

\newtheoremstyle{restatementstyle}
  {\topsep}  % space above
  {\topsep}  % space below
  {\itshape}% name of font to use in the body of the theorem
  {0em}% measure of space to indent
  {\bfseries}% name of head font
  {.}% punctuation between head and body
  {5pt plus 1pt minus 1pt}% space after theorem head; " " = normal interword space
  {\thmname{#1} \thmnote{(Restatement of #3)}}% Manually specify head
\theoremstyle{restatementstyle}
\newtheorem{restateLemma}{Lemma}
\newtheorem{restateTheorem}{Theorem}

\newcommand{\fixme}[1]{\textbf{FIXME: #1}}

\begin{document}

\conferenceinfo{POPL'13,} {January 23--25, 2013, Rome, Italy.}
\CopyrightYear{2013}
\copyrightdata{978-1-4503-1832-7/13/01}

\maketitle

\begin{abstract}
  Reynolds' relational parametricity provides a powerful way to reason
  about programs in terms of invariance under changes of data
  representation. A dazzling array of applications of Reynolds' theory
  exists, exploiting invariance to yield ``free theorems'',
  non-inhabitation results, and encodings of algebraic datatypes.
  Outside computer science, invariance is a common theme running
  through many areas of mathematics and physics. For example, the area of
  a triangle is unaltered by rotation or flipping. If we scale a
  triangle, then we scale its area, maintaining an invariant
  relationship between the two. The transformations under which
  properties are invariant are often organised into groups, with the
  algebraic structure reflecting the composability and invertibility
  of transformations.

  In this paper, we investigate programming languages whose types are
  indexed by algebraic structures such as groups of geometric
  transformations. Other examples include types indexed by
  principals--for information flow security--and types indexed by
  distances--for analysis of analytic uniform continuity
  properties. Following Reynolds, we prove a general Abstraction
  Theorem that covers all these instances. Consequences of our
  Abstraction Theorem include free theorems expressing invariance
  properties of programs, type isomorphisms based on invariance
  properties, and non-definability results indicating when certain
  algebraically indexed types are uninhabited or only inhabited by
  trivial programs.  We have fully formalised our framework and most
  examples in Coq.
\end{abstract}

\category{D.1.1}{Programming techniques}{Applicative (functional)
  programming} \category{D.2.4}{Software Engineering}{Software/Program
  Verification} \category{D.3.3}{Programming Languages}{Language
  Constructs and Features---Data types and structures}

\terms
  Languages, Theory, Types

\keywords parametricity, units of measure, dimensional analysis,
invariance, computational geometry, information flow, metric types, uniform continuity

%%%%%%%%%%% Springe nach section-introduction.tex
%%%---------- open: section-introduction.tex
\section{Introduction}
\label{sec:introduction}
The best way we know of describing the semantics of parametric
polymorphism is \emph{relational parametricity}, whose central result
is Reynolds' Abstraction Theorem~\cite{reynolds83types}. Its striking
consequences include the well-known ``free theorems'' for polymorphic
types~\cite{wadler89theorems}, non-inhabitation results, and precise
correspondences between System F encodings and algebraic
datatypes~\cite{PittsAM:parpoe}, abstract data types, and, most
recently, higher-order encodings of binder
syntax~\cite{syntaxforfree}.

Relational parametricity is in essence a principle of
\emph{invariance}: the behaviour of polymorphic code is invariant
under changes of data representation. For example, the type
$\forall\alpha.\tyPrim{list}\alpha\to\tyPrim{list}\alpha$
tells us that any transformation applied to elements of
the input list will be reflected by the same transformation applied
to elements of the result. 
Invariance results also abound in
mathematics and physics. The area of a triangle is invariant with
respect to isometries of the Euclidean plane; the determinant of a
matrix is invariant under changes of basis; and Newton's laws are the
same in all inertial frames. Typically, the 
transformations under which invariants are preserved have interesting structure: 
for example, translations in
the Euclidean plane form an abelian group.

Inspired by this connection, we study type systems that
capture rich invariants in types indexed by attributes with
algebraic structure.  For example, in computational geometry, points
in the plane can be indexed by attributes representing affine
transformations; in information-flow security, computations can be
indexed by principals; in differential privacy, types can be indexed
by `distance'. Types that are polymorphic over such indices induce
invariance properties and abstraction barriers beyond those introduced
by their unindexed versions, as we shall illustrate.  This generalises
previous work by the third author on types parameterized by
units of measure, whose invariance properties relate to changes of
units, or \emph{scaling}~\cite{kennedy97relational}.

\paragraph{Invariance}
To illustrate type-induced invariance properties, consider
two-dimensional geometry. 
In a conventional type system, a function $\mathrm{areaTri}$ that computes the area of a triangle
might be assigned the type:
$\tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal$.
But in our proposed system we can assign it the following more expressive polymorphic
type:
\[
\mathrm{areaTri} : \forall t\mathord:\SynTransl{2}.
  \tyPrim{vec}{t} \times \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyReal \\
\]
This type expresses the fact that if each of the arguments to $\mathrm{areaTri}$
is translated by the same vector, then the result remains the same,
that is, it is \emph{invariant} under translation. Formally, for any 
vector $\vec t$,
\[
\mathrm{areaTri}\;(\vec{t} + \vec{v_1}, \vec{t} + \vec{v_2}, \vec{t} + \vec{v_3}) = 
\mathrm{areaTri}\;(\vec{v_1}, \vec{v_2}, \vec{v_3})
\]

Transformations typically \emph{compose} in various
ways, and the compositions satisfy algebraic laws. For example, 
we can assign a function that computes the area of a circle given its
radius the following polymorphic type:
\[
\mathrm{areaCircle} : \forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to
\tyPrim{real}{s\cdot s}
\]
This captures the fact that the area of a circle varies as the square
of its radius, i.e., $\mathrm{areaCircle}(k r) = k^2\cdot
\mathrm{areaCircle}(r)$ for any $k\neq 0$ (the `sorts' $\SynTransl{2}$ and
$\SynGL{1}$ will be explained later).  Here, $s$ 
can be interpreted as the \emph{units of measure} of the argument to
$\mathrm{areaCircle}$, and `$\cdot$' composes units using the
product. We can also add an inverse operation and identity unit of
measure $1$, and then impose the algebraic laws of abelian
groups. This permits identification of, for example,
$\tyPrim{real}{s\cdot s^{-1}}$ with the type
$\tyPrim{real}{1}$ of dimensionless constants.

\paragraph{Abstraction}
In his original paper on parametricity, Reynolds asserted that
\emph{type structure is a syntactic discipline for enforcing levels of
  abstraction}.  We see something analogous here: if all primitive
operations are given types that reflect their behaviour under
translation, then there is no way to `break' this property. For
example, there is no way that $\mathrm{areaTri}$ can depend on the
actual coordinates of its inputs. Furthermore, the distinction between
points and vectors that is often enforced through abstract data
types~\cite{CGAL} is captured here by indices instead. For example,
the operation that takes two points and computes their vector
difference can be assigned the type
$\forall t\mathord:\SynTransl{2}.\tyPrim{vec}t\times\tyPrim{vec}t\to\tyPrim{vec}0$,
reflecting the invariance of the result (a pure vector) under
translations of the point arguments. As a result through types alone
we can, in essence, derive so-called \emph{coordinate-free}
geometry~\cite{CFGADT}.

The invariance properties discussed above can be seen as ``free
theorems''~\cite{wadler89theorems}, but the abstraction afforded by
polymorphic indexed types can also induce interesting type
\emph{isomorphisms}.  The type of $\mathrm{areaCircle}$ above is in
fact isomorphic to $\tyPrim{real}{1}$. A moment's thought reveals why:
what possible unary functions can be constructed whose outputs scale
as the square of the scaling of their inputs?  Answer: just those
functions of the form~$\lambda x. k x^2$ for some constant~$k$.  In
this case, of course, we expect that~$k = \pi$.

\paragraph{Relational parametricity}
To derive such invariance and abstraction properties of types, we
adopt the techniques of relational parametricity. Over an underlying
index-erasure semantics we construct binary relations parameterised by
an environment $\rho$ that describes how values of primitive type are
related according to their indices.  For example, values $v$ and $w$
of type $\tyPrim{real}s$ are related when $v$ ``scales to'' $w$
according to an interpretation of $s$ (i.e., $w=\rho(s)\cdot
v$).  Values of polymorphic type are related exactly when they are
related for all possible interpretations of the quantified
variable. For example, values $v$ and $w$ of type
$\forall t\mathord:\SynTransl{2}.\tyPrim{vec} t\to
\tyPrim{vec} t$ are related when they are related at type
$\tyPrim{vec} t\to\tyPrim{vec}t$ for all translations
$\vec t\in\Transl{2}$ associated with $t$.

As it happens, the relational interpretations given above are
functional, relating one value uniquely to another. Other applications
make use of primitive relations that are not simple functions.
For example, in a type system in which the index in
$\tyPrim{real}s$ is interpreted not as a unit of measure, but as
a measure of \emph{closeness}, two values $x$ and $y$ of this type are
related if $|x-y| < \rho(s)$ for a positive real number
$\rho(s)$.  Rather beautifully, the standard notion of uniform
continuity can then be expressed as %a type:
%\begin{displaymath}
 $ \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
%\end{displaymath}

\paragraph{Motivations}
Our motivations for studying algebraically indexed types are
threefold. First, we believe that, as with
units of measure~\cite{fsharp}, practical programming language
extensions will follow. For example, in computational geometry and
graphics, attributes on points, vectors, and other geometric types
could be used to prevent the mixing of different coordinate systems,
or `frames'. Second, type-based static analyses can be based on
indexed types, for example, in effect systems~\cite{benton06reading},
and, more speculatively, in continuity
analysis~\cite{chaudhuri10continuity}.  Finally, we believe that
expressing algebraic invariants through types has the potential to
offer slick proof techniques for mechanized
mathematics. Harrison has applied the
invariance properties of geometric primitives to create elegant proofs
in geometry, based on `without loss of
generality' principles~\cite{harrison09without}. The invariance properties are expressed 
and propagated using ad-hoc tactics; our types offer a more principled
means of achieving the same end, and the `wlog' principle itself
is expressed through type isomorphisms.

We follow the mantra \emph{semantics first, syntax later} in studying
types with algebraic structure. We have not yet built a practical
programming language that supports algebraically indexed types; nor
have we designed type checking, type inference, or static analysis
algorithms.  But when we do so, the semantics will guide us. The fact
that zero is polymorphic in units of measure (it can be given type $\forall
u.\tyPrim{real}u$) whereas other constants are dimensionless (having
type $\tyPrim{real}1$) is justified by the invariance properties
induced by the types: zero is invariant under scaling, other constants
are not. For less trivial constants and operations, the appropriate types are
not so apparent, as we shall see, but invariance properties expressed by
the semantics guide us in assigning appropriate types. Semantics does not lie!

% Possible angles of attack:
% \begin{itemize}
% \item Parametric polymorphic types allow us to prevent
%   over-specification of the behaviour of programs. For instance, the
%   type $\forall \alpha. [\alpha] \to [\alpha]$ is a generalisation of
%   the types $[\mathsf{int}] \to [\mathsf{int}]$ and $[\mathsf{char}]
%   \to [\mathsf{char}]$. Either of the latter two types over-specify
%   the behaviour of the function.
% \item There are other cases of programs that are over-specified. The
%   leading example we just below is of geometric programs that
%   manipulate coordinate data. Often, programs that manipulate
%   coordinate data are insensitive to geometric transformations. For
%   example, a program that computes the area of a triangle described by
%   three points is insensitive to translations or rotations applied to
%   all three points.
% \item 
% \end{itemize}

% There are three main points to get across:
% \begin{enumerate}
% \item Why algebraically indexed types?
% \item Why relational parametricity?
% \item Why study them together?
% \end{enumerate}

\subsection{Contributions}
\label{sec:contributions}

\begin{figure}
  \centering
  \begin{tikzpicture}
  \node at (-3,0.8) [rectangle,style={align=center}] {Algebraic \\ Theory};
  \node at (-2,0.8) {$+$};
  \node at (-0.5,0.8) [rectangle,style={align=center}] {Index-Erasure \\ Interpretation};
  \node at (1,0.8) {$+$};
  \node at (2.6,0.8) [rectangle,style={align=center}] {Relational \\ Interpretation};

  \draw plot (-4,0.6) .. controls (-4,-0.2) and (-0.4,0.8) .. (-0.4,-0.2)
             -- (-0.6,-0.2) -- (0,-0.5) -- (0.6,-0.2) --
             (0.4,-0.2) .. controls (0.4,0.8) and (4,-0.2) .. (4,0.6);

  \node at (0,-0.7) [rectangle,style={align=center}] {Abstraction Theorem (\thmref{thm:abstraction})};

  \draw plot (-0.8,-1) .. controls (-0.8,-1.3) and (-3.5,-1.5) .. (-3.5,-2)
             -- (-3.7,-2) -- (-3.1,-2.3) -- (-2.5,-2) -- 
             (-2.7,-2) .. controls (-2.7,-1.5) and (-0.8,-1.6) .. (-0.7,-1.5)
                       .. controls (-0.8,-1.6) and (-0.5,-1.5) ..
             (-0.4,-2) -- (-0.6,-2) -- (0,-2.3) -- (0.6,-2) --
             (0.4,-2) .. controls (0.5,-1.5) and (0.8,-1.6) .. (0.7,-1.5)
                      .. controls (0.8,-1.6) and (2.7,-1.5) ..
             (2.7,-2) -- (2.5,-2) -- (3.1,-2.3) -- (3.7,-2) -- (3.5,-2)
                      .. controls (3.5,-1.5) and (0.8,-1.3) .. (0.8,-1);

  \node at (-3.1,-2.8) [rectangle,style={align=center}] {Free \\ Theorems};
  \node at (0,-2.8) [rectangle,style={align=center}] {Type \\ Isomorphisms};
  \node at (3.1,-2.8) [rectangle,style={align=center}] {Non-definability};
\end{tikzpicture}
  \caption{Summary of the Paper}
  \label{fig:summary}
\end{figure}

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a
refined type system for information flow based on logic, and a simple
type system with distance-indexed types.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' Abstraction Theorem, for a given \emph{model}
of index sorts and relational interpretation of primitive types.
We prove that the semantics soundly approximates contextual equivalence.
\item
For each of our main examples we deduce free theorems that are
consequences of our Abstraction Theorem, prove specific non-definability
results, and derive interesting type isomorphisms. 
For a large class of first-order types we give a general method for 
constructing suitable models to prove non-definability results. 
\autoref{fig:summary}
illustrates the central position of our analogue of Reynolds' Abstraction
Theorem (\thmref{thm:abstraction}) in these results.
\end{itemize}
We improve on the earlier semantics of
units of measure~\cite{kennedy97relational} in a number of ways.  By
extending the language of units with an `absolute value' operation, we
can give more precise types and obtain more general invariance
properties.  The relational interpretation for units is both simpler
and more flexible, and we derive slicker proofs of non-definability,
and new results.  Our notion of type isomorphism is stronger than
before, being based on contextual equivalence.

We have fully formalised our framework and most examples in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}. The formalisation is available from
\\{\small \url{https://github.com/bobatkey/algebraically-indexed-types}}
\subsection{Structure of paper}
Our paper is structured as follows. In \autoref{sec:motivating-examples}
we present an extended case study of two-dimensional geometry, with semi-formal
description of types and results.
In \autoref{sec:a-general-framework} we describe fully formally a general framework
for algebraically-indexed types, and prove the Abstraction Theorem and
soundness for semantic equivalence. \autoref{sec:instantiations} presents a number
of applications of the theory to 2-d geometry, including free theorems, type isomorphisms
and non-definability. \autoref{sec:general-nondef} develops a more general technique 
for proving non-definability results. \autoref{sec:information-flow} presents
the application of algebraically indexed types to information flow security, and
\autoref{sec:metric-types} applies it to types indexed by distances.
Finally, \autoref{sec:discussion} discusses related work and future plans.

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:
%%%---------- close: section-introduction.tex
%%%%%%%%%%% Springe nach geometry-twocolumn.tex
%%%---------- open: geometry-twocolumn.tex
\section{Geometry via Algebraically Indexed Types}
\label{sec:motivating-examples}

We motivate our investigation of algebraically indexed types and their
relational interpretations by developing a novel type system for
programs that manipulate two-dimensional geometric data. Geometry is
rich with operations that are invariant under transformation: affine
operations are invariant under change of origin
(\autoref{sec:affine-vector-ops}), vector space operations are
invariant under change of basis, and dot product is invariant under
orthogonal changes of basis
(\autoref{sec:motivation-generalising}). On the other hand, some
geometric operations are interestingly variant under transformation.
For example, cross products vary with scalings of the plane
(\autoref{sec:scale-invariance}). We incorporate (in)variance
information about geometric primitives into type systems via
algebraically indexed types.

\subsection{Origin Invariance and Representation Independence}\label{sec:oiri}

The basic data structure used in programs that manipulate geometric data is
the $n$-tuple of numbers. In the 2-dimensional case, tuples 
$\vec{v} = (x,y)$ serve double duty, representing
%are called upon to represent both 
both \emph{points}---offsets from some origin---and
\emph{vectors}---offsets in their own right.  Despite their common
representation, points and vectors are very different, and
distinguishing between them is the key feature of \emph{affine
  geometry} (see, for example, Chapter 2 of Gallier's book
\cite{gallier11geometric}). Nevertheless, computational geometry
libraries traditionally either leave it to the programmer to maintain
the distinction between points and vectors, or else use different
abstract types for points and vectors to enforce it.
%(one might say that the standard mathematical formulation of affine
%spaces uses the abstract types approach). 
In this paper we investigate a more sophisticated approach based on 
types indexed by change of origin transformations.

This approach regards the difference between points and vectors as a
change of data representation. For example, if $(0,0)$ and $(10,20)$
are two origins, then the tuple $(1,1)$ with respect to $(0,0)$ and the
tuple $(11,21)$ with respect to $(10,20)$ represent \emph{the same
  point} because they have the same displacement from these two
origins, respectively.
%It is merely an artifact of the representation of points as pairs of
%numbers that they appear to be different.
This suggests that programs that manipulate points %data representing points
should be invariant with respect to changes of
origin. Programs that manipulate vectors, on the other hand, should
not be invariant under change of origin. Different vectors represent
different offsets, and the vector $(0,0)$ always represents the zero
offset.

Invariance under change of representation immediately recalls
Reynolds' fable about two professors teaching the theory of complex
numbers \cite{reynolds83types}. One professor represents complex
numbers using rectangular coordinates ($x + iy$), while the other
represents them using polar coordinates ($\alpha\cos\theta +
i\alpha\sin\theta$). Happily, after learning the basic operations on
complex numbers in the two representations, the two classes can
interact because the theory of complex numbers is invariant under the
choice of representation. Reynolds formalises the idea of invariance
under changes of representation as preservation of relations.  For
example, if a binary relation $R$ relates the rectangular and polar
representations of complex numbers, then a program that manipulates
complex numbers at a level of abstraction above their specific
representation should preserve $R$. %More precisely, if
%$f$ is a program that is invariant under the choice of representation
%of complex numbers,
%%taking complex numbers to complex numbers that respects invariance
%%under change of representation. Then 
%$c$ is a complex number in rectangular form, $c'$ is a complex number
%in polar form, and $R$ is a relation that relates the rectangluar and
%polar representations of complex numbers, then $(c,c') \in R$ implies
%$(f(c), f(c')) \in R$.

Reynolds' relational approach can be applied in the geometric setting
to show how quantifying over all changes of origin ensures the
invariance of programs under any particular choice of origin. For
this, we first define a family of binary relations on $\mathbb{R}^2$
that is indexed by changes of origin. Changes of origin are
represented by vectors in $\mathbb{R}^2$, and form a group
$\Transl{2}$ of translations under addition. The $\Transl{2}$-indexed
family of binary relations $\{ R_{\vec{t}} \subseteq \mathbb{R}^2
\times \mathbb{R}^2 \}_{\vec{t} \in \Transl{2}}$ is then defined by
%\begin{displaymath}
$R_{\vec{t}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{t} \}$.
%\end{displaymath}
We then consider a function $f$ that takes as input two tuples in
$\mathbb{R}^2$ and returns a single tuple in $\mathbb{R}^2$. We intend
that the tuples all represent points with respect to the same origin,
and that $f$ is invariant under the choice of origin.  Reynolds'
relational approach formalises this intention precisely. For any
$\vec{t} \in \Transl{2}$: % (i.e.~any change of origin), the function
%$f$
%should satisfy the following statement:
\begin{equation}\label{eq:f-preserve-rel-frame}
  \forall (\vec{v_1},\vec{v'_1}) \in R_{\vec{t}},
  (\vec{v_2},\vec{v'_2}) \in R_{\vec{t}}. (f(\vec{v_1}, \vec{v_2}),
  f(\vec{v'_1}, \vec{v'_2})) \in R_{\vec{t}}
\end{equation}
Unfolding the definition of $R_{\vec{t}}$ gives the equivalent formulation,
%\statementref{eq:f-preserve-rel-frame} is equivalent to: 
again for all $\vec{t} \in \Transl{2}$:
\begin{equation}\label{eq:f-invariant-frame}
  \forall \vec{v_1}, \vec{v_2}.\ f(\vec{v_1} + \vec{t},\vec{v_2} +
  \vec{t}) = f(\vec{v_1},\vec{v_2}) + \vec{t}.
\end{equation}
Thus, Reynolds' preservation of relations, when instantiated with the
family of relations $\{R_{\vec{t}}\}$, yields exactly the geometric
property of invariance under change of origin.

% An example function $f$ that satisfies
% \statementref{eq:f-invariant-frame} is the following function that
% computes a particular affine combination of two points by working
% directly on their coordinate representation:
% \begin{displaymath}
%   f(\vec{v_1}, \vec{v_2}) = \frac{1}{2}\vec{v_1} + \frac{1}{2}\vec{v_2}.
% \end{displaymath}
% In general, affine combinations of points $\lambda_1\vec{v_1} +
% \lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$, satisfy
% \statementref{eq:f-invariant-frame}. Affine combination is one of the
% fundamental building blocks of affine geometry -- the properties of
% points invariant under invertible affine maps
% (\cite{gallier11geometric}, Chapter 2). If we drop the condition that
% $\lambda_1 + \lambda_2 = 1$, then we are dealing with linear
% combinations of vectors, and we are no longer invariant with respect
% to changes of frame. However, linear combinations are invariant with
% respect to change of basis. We can represent changes of basis as
% linear invertible maps $B : \mathbb{R}^2 \to \mathbb{R}^2$. The set of
% all such maps forms the general linear group $\GL(2)$. We now define
% another family of relations $\{R_{\texttt{vec}}(B) \subseteq
% \mathbb{R}^2 \times \mathbb{R}^2 \}_{B \in \GL(2)}$ that relates two
% points up to change of basis:
% \begin{displaymath}
%   R_{\texttt{vec}}(B) = \{ (\vec{v_1},\vec{v_2}) \sepbar B\vec{v_2} = \vec{v_1} \}.
% \end{displaymath}
% Now the functions $f_{\lambda_1\lambda_2}(\vec{v_1},\vec{v_2}) =
% \lambda_1\vec{v_1} + \lambda_2\vec{v_2}$, for arbitrary $\lambda_1$
% and $\lambda_2$, do preserve the relations $R_{\texttt{vec}}(B)$, for
% all $B \in \GL(2)$. Unfolding the definition of $R_{\texttt{vec}}(B)$
% in the analogous statement to \statementref{eq:f-preserve-rel-frame}
% for $R_{\texttt{vec}}$ instead of $R_{\texttt{pt}}$, we can see that
% preservation of the relations $R_{\texttt{vec}}$ characterises the
% functions that are invariant under change of basis. For all $B \in \GL(2)$, we have
% \begin{displaymath}
%   \forall \vec{v_1}, \vec{v_2}.\ f(B\vec{v_1},B\vec{v_2}) = B(f(\vec{v_1},\vec{v_2})),
% \end{displaymath}
% and this is exactly the property of invariance under change of basis
% we required above of programs manipulating vectors.

% Note that the family of relations $R_{\texttt{vec}}$ is just
% $R_{\texttt{pt}}$ when restricted to elements of the group
% $\GL(2)$. In the type system we introduce in the next section, we
% combine points and vectors into the same data type. Whether it
% represents a point or a vector depends on the group of geometric
% transformations that we expect it to be invariant under.

% FIXME: vectors are not invariant under change of origin, so they are
% represented by the relation $R_0$, which is just equality. But they,
% and points are invariant under change of basis. However, as we shall
% see below, not all operations are invariant under all changes of
% basis. In particular the dot product is only invariant under
% orthogonal transformations.

\subsection{A Type System for Change of Origin Invariance}
\label{sec:type-system-geom-intro}

Reynolds also showed how a type discipline can be used to establish
that (the denotational interpretations of) programs preserve
relations. For Reynolds, the type discipline of interest was that of
the polymorphic $\lambda$-calculus, which supports the construction of
new types by universal quantification over types.
%, in which new types can be
%constructed by universal quantification over all types.
In terms of relations, Reynolds interprets universal quantification
over types as quantification over binary relations between denotations
of types. By contrast, in our statements of geometric invariance in
\autoref{sec:oiri} we did not quantify over all
relations, but instead quantified over all changes of origin and used
a specific choice of origin to select a relation from the family
$\{R_{\vec{t}}\}$. This suggests introducing quantification over
changes of origin into the language of types. We use the notation
$\forall t \mathord: \SynTransl{2}. A$ for quantification over all
2-dimensional translations (i.e.,~choices of origin) $t$, and refer to
$\SynTransl{2}$ as the \emph{sort} of $t$. Note the difference in
fonts used to distinguish the semantic group $\Transl{2}$ from the
syntactic sort $\SynTransl{2}$. We use a similar convention
below, too.

Since the sort $\SynTransl{2}$ represents an abelian group, we can
combine its elements using the usual group operations. We write
operations additively, using $e_1 + e_2$ for the group operation, $-e$
for inverse and $0$ for the unit.  We also regard expressions built
from variables and the group operations up to the abelian group
axioms. For example, we regard $e_1 + (e_2 + e_3)$ and $(e_1 + e_2) +
e_3$ as equivalent.

Our language of types includes the unit type $\tyPrimNm{unit}$ and,
for all types $A$ and $B$, the function type $A \to B$, the sum type
$A + B$, and the tuple type $A \times B$. We also assume a primitive
type $\tyPrimNm{real}$, used to represent scalars. Although tuples of
real numbers represent points and vectors in geometric applications,
we cannot express this via the type
%central data structure in geometric applications is the tuple of real
%numbers for representing points and vectors, we cannot simply express
%this as the type 
$\tyPrimNm{real} \times \tyPrimNm{real}$. Indeed,
%because this type does not have the correct relational
%interpretation. (
two elements of type $\tyPrimNm{real}$ are related if and only if they
are equal and, by Reynolds' interpretation of tuple types, two
elements of $\tyPrimNm{real} \times \tyPrimNm{real}$ are also related
if and only if they are equal. But since this does not give the
correct relational interpretations for points and vectors, we
introduce a new type $\tyPrim{vec}{e}$, indexed by expressions $e$ of
sort $\SynTransl{2}$, to represent them.  The index $e$ represents the
displacement by change of origin of a point of this type.  Although we
have taken pains to distinguish geometric points and vectors, we use
the name $\tyPrimNm{vec}$ for both to recall the computer science
notion of vector as a homogeneous sequence of values with a known
length (in this case, 2).

As is standard for parametricity, every type has two interpretations:
an index-erasure interpretation that ignores the indexing expression,
and a relational interpretation as a binary relation on the
index-erasure interpretation. We denote the index-erasure and
relational interpretations with the notations $\tySem{\cdot}$ and
$\rsem{\cdot}{}$ respectively. To give such interpretations for the
types $\tyPrim{vec}{e}$ and $\forall t\mathord:\SynTransl{2}.A$, we
assume for now that we can map each expression $e$ of sort
$\SynTransl{2}$ to an element $\sem{e}\rho$ of the group $\Transl{2}$
using some environment $\rho$ that interprets $e$'s free
variables. The index-erasure and relational interpretations of
$\tyPrim{vec}{e}$ are:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2
    \\ \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{e}\rho} = \{
    (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
  \end{array}
\end{displaymath}
The index-erasure and relational interpretations of 
$\forall t\mathord:\SynTransl{2}.A$ are:
%is simply the interpretation of the
%underlying type $A$. The relational interpretation intersects the
%relational interpretations of the type $A$ under all extensions of the
%relational environment $\rho$:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\forall t\mathord:\SynTransl{2}.A} & \tySem{A}
    \\ \rsem{\forall t\mathord:\SynTransl{2}.A}{}\rho & \bigcap\{
    \rsem{A}{}{(\rho,\vec{t})} \sepbar \vec{t} \in \Transl{2} \}
  \end{array}
\end{displaymath}
The index-erasure and relational interpretations are given
formally in Sections~\ref{sec:erasure-semantics}
%the relational interpretation is in general more complex, due to the
%possibility of non-compositional interpretations of free index
%variables. The relational interpretation will be presented in
and~\ref{sec:relational-semantics}. %, respectively.

At the end of \autoref{sec:oiri} we considered functions $f :
\mathbb{R}^2 \times \mathbb{R}^2 \to \mathbb{R}^2$ that preserve all
changes of origin. This property of $f$ can be expressed in terms of
types by
%\begin{displaymath}
$  \mathit{f} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t}
  \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}$.
%\end{displaymath}
Spelling out the relational interpretation of this type using the
definitions above and the standard relational interpretations for
tuple and function types, we recover
\statementref{eq:f-invariant-frame} exactly.

\subsection{Affine and Vector Operations}
\label{sec:affine-vector-ops}

Invariance under change of origin is the key feature of affine
geometry, whose central operation is the affine combination of points:
$\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$, where $\lambda_1 +
\lambda_2 = 1$.  Geometrically, this can be interpreted as describing
all the points on the unique line through the points represented by
$\vec{v_1}$ and $\vec{v_2}$ (assuming $\vec{v_1} \not= \vec{v_2}$).
We add affine combination of points to our calculus as
follows: % typing and intended denotation:for
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affComb} :\forall t \mathord:
    \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to
    \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\ 
    \tmSem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} +
    r\vec{v_2} 
\end{array}
\end{displaymath}
It can be verified by hand that the index-erasure interpretation %the intended denotation
$\tmSem{\mathrm{affComb}}$ is invariant under all changes of origin, as
dictated by its type.
% By defining the primitive function $\mathrm{affComb}$ to take a
% single $\tyPrimNm{real}$ parameter $t$, we can easily ensure that we
% are taking the affine combination of two representatives of points,
% and not just an arbitrary linear combination.

\begin{example}
  The evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves
  with two endpoints and a single control point) can be expressed
  using the affine combination primitive as follows: % and three steps
  % of De Casteljau's algorithm:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{quadB\acute{e}zier} : \forall t
      \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \mathord\to \tyPrim{vec}{t}\mathord\to
      \tyPrim{vec}{t} \mathord\to \tyPrimNm{real} \mathord\to \tyPrim{vec}{t}
      \\ \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\ \quad
      \mathrm{affComb}\ [t]\ (\mathrm{affComb}\ [t]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [t]\ p_1\ s\ p_2)
      \\
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$, a control point $p_1$, and $s \in
  [0,1]$, an application
  $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ gives the point on
  the curve at ``time'' $s$.  The type of
  $\mathrm{quadB\acute{e}zier}$ immediately tells us that it preserves
  all changes of origin.
\end{example}
% Affine combination only provides us with an operation on points. We
% are also interested in operations on the vectors representing offsets
% between points. We now examine the correct types to assign to the
% vector space operations of addition of vectors, negation of vectors,
% multiplication by a scalar and the zero vector. The typings of these
% operations will make use of the abelian group structure of change of
% origin translations.

The obvious type for vector addition is
%\begin{displaymath}
$  (+) : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}$.
%\end{displaymath}
%(we write binary operators intended to be used infix in parentheses
%when not appearing in infix position, following the Haskell
%syntax.)
But we can reflect the fact that $(+)$ is not invariant under change
of origin by giving it a more precise type that reflects how it varies
with change of origin:
\begin{displaymath}
  (+) : \forall t_1, t_2 \mathord: \SynTransl{2}.\ \tyPrim{vec}{t_1}
  \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
\end{displaymath}
Intuitively, this type says that if the first input vector has been
displaced by $t_1$ and the second by $t_2$, then their sum is
displaced by $t_1 + t_2$. We can also negate vectors, yielding a
vector which points in the opposite direction. Negation negates
translation arguments:
\begin{displaymath}
  \mathrm{negate} : \forall t \mathord:
  \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
\end{displaymath}
Finally, with the primitive operations of addition and negation of
vectors we can define the derived operation of subtraction:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall t_1,t_2 \mathord:\SynTransl{2}.\ \tyPrim{vec}{t_1}
    \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2}
    \\ (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
  \end{array}
\end{displaymath}

Given two points that are invariant with respect to the same change of
origin---i.e.,~two values of type $\tyPrim{vec}{t}$---we can use
subtraction to compute their offset:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall t
    \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to
    \tyPrim{vec}{0} \\ \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
The result is a vector expressed with respect to the null change of
origin: note how the algebraic structure on the indexing theory
induces type equalities that can be used to simplify the type of the
result of $\mathrm{offset}$ from $\tyPrim{vec}{t-t}$ to
$\tyPrim{vec}{0}$. The type of $(+)$ can also be specialised to the
case of moving a point by a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall t
    \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to
    \tyPrim{vec}{t} \\ \mathrm{moveBy}\ [t]\ p\ v = p + v
  \end{array}
\end{displaymath}
The types we assign to the remaining vector space primitives, namely
$\mathrm{0} : \tyPrim{vec}{0}$ for
the zero vector and $(*) : \tyPrimNm{real} \to
    \tyPrim{vec}{0} \to \tyPrim{vec}{0}$ for
multiplication by a scalar, do not describe any
interesting effects on translations.
%\begin{displaymath}
%  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{4em}}l@{\hspace{0.5%em}:\hspace{0.5em}}l}
%    \mathrm{0} & \tyPrim{vec}{0}& (*) & \tyPrimNm{real} \to
%    \tyPrim{vec}{0} \to \tyPrim{vec}{0}
%  \end{array}
%\end{displaymath}

% \fixme{If we had dependent types, could the affComb operation be
%   derived from an appropriately dependently typed scaling operation?}

\begin{example}\label{ex:type-iso}
  The vector space operators and the properties that follow from their
  types allow us to establish a useful type isomorphism. Consider
  functions with types following the schema:
  \begin{displaymath}
    \tau_{n} \isDefinedAs \forall
    t\mathord:\SynTransl{2}.\ \underbrace{\tyPrim{vec}{t} \to ... \to
      \tyPrim{vec}{t}}_{n+1\textrm{ times}} \to \tyPrimNm{real}
  \end{displaymath}
  Just by looking at the types $\tau_{n}$, we know that their
  inhabitants will be invariant under change of origin because of the
  quantification over all $t$ in $\SynTransl{2}$. So we may as well
  choose one of the input points as the origin and assume that all the
  other points are defined with respect to it.  This formalises the
  common mathematical practice of stating that ``without loss of
  generality'' we can take some point in a description of a problem to
  be the origin provided the problem statement is invariant under
  translation. Each type $\tau_{n}$ is isomorphic to the corresponding
  type $\sigma_{n}$:
  \begin{displaymath}
    \sigma_{n} \isDefinedAs \underbrace{\tyPrim{vec}{0} \to
      ... \tyPrim{vec}{0}}_{n\textrm{ times}} \to \tyPrimNm{real}
  \end{displaymath}
  We demonstrate these isomorphisms formally in
  \autoref{sec:instantiations}, in the more general setting of types
  indexed by abelian groups.
  % The isomorphism between these two families of types can be witnessed
  % by the following two functions.
  % \begin{displaymath}
  %   \begin{array}{l}
  %     i_{n,A} : \tau_{n,A} \to \sigma_{n,A} \\
  %     i_{n,A}\ f\ [B]\ (p_1, p_2, ..., p_n) = f\ [B \ltimes 0]\ (0, p_1, ..., p_n) \\
  %     \\
  %     i^{-1}_{n,A} : \sigma_{n,A} \to \tau_{n,A} \\
  %     i^{-1}_{n,A}\ g\ [F]\ (p_0, ..., p_n) = g\ [\pi_L(F)]\ (p_1-p_0, ..., p_n-p_0)
  %   \end{array}
  % \end{displaymath}
  % The direction defined by the function $i_{n,A}$ treats the inputs
  % $p_1,...,p_n$ as vector offsets from the origin $0$. The direction
  % witnessed by $i^{-1}_{n,A}$ picks the first point to act as the
  % origin, and uses the operation of vector subtraction we defined
  % above to turn each of the other points into offsets from this
  % point. Note that this isomorphism is not unique: for each $n$ we can
  % pick any of the $n+1$ inputs to act as the distinguished
  % origin. Thus the families of types $\tau_{n,A}$ and $\sigma_{n,A}$
  % are isomorphic, but not uniquely isomorphic.
\end{example}

\begin{example}\label{ex:uninhabited-type}
  So far we have emphasised the derivation of properties, or ``free
  theorems'', of programs from their types. But using more refined
  relational interpretations of types we can also show that certain
  types are uninhabited. For example, the type $\forall t \mathord:
  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ has no
  inhabitants. Intuitively, this is because we cannot remove the extra
  $t$ in $\tyPrim{vec}{t+t}$ using the vector operations. We formalise
  this non-definability result in \autoref{sec:general-nondef} using a
  specialised relational interpretation.
\end{example}

\subsection{Change of Basis Invariance}
\label{sec:motivation-generalising}

Although vector addition, negation, and scaling are not invariant
under change of origin, they are invariant under change of basis. As
with origin invariance, we can express basis invariance as
preservation of relations indexed by changes of basis. Change of basis
is achieved by applying an invertible linear map, and the collection
of all such maps on $\mathbb{R}^2$ forms the \emph{General Linear}
group $\GL{2}$, which we represent in our language by a new indexing
sort $\SynGL{2}$ with (non-abelian) group structure that we will write
multiplicatively. We then extend $\tyPrimNm{vec}$ to allow indices of
sort $\SynGL{2}$, as well as $\SynTransl{2}$, so that
$\tyPrim{vec}{B,t}$ is a vector that varies with change of basis $B$
and change of origin $t$. Formally, the index-erasure and relational
semantics of $\tyPrim{vec}{B,t}$ are given by:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2
    \\ \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'})
    \sepbar \vec{v'} = (\sem{e_B}\rho)\vec{v} + \sem{e_t}\rho \}
  \end{array}
\end{displaymath}

\paragraph{Affine Geometry} An \emph{affine transformation} is
an invertible linear map together with a translation. We can assign
types to all the primitive affine and vector space operations
indicating how they they behave with respect to affine
transformations:
\begin{eqnarray*}
  \mathrm{affComb} & : &
  \begin{array}[t]{@{}l}
    \forall B \mathord: \SynGL{2}, t \mathord: \SynTransl{2}.\\
    \hspace{0.2cm} \tyPrim{vec}{B,t} \to \tyPrimNm{real} \to
    \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,t}
  \end{array}
  \\
  (+) & : &
  \begin{array}[t]{@{}l}
    \forall B \mathord: \SynGL{2}, t_1,t_2 \mathord: \SynTransl{2}. \\
    \hspace{0.2cm}\tyPrim{vec}{B,t_1} \to \tyPrim{vec}{B,t_2} \to
    \tyPrim{vec}{B,t_1 + t_2}
  \end{array}
  \\
  \mathrm{negate} & : & \forall B \mathord: \SynGL{2}, t \mathord: \SynTransl{2}.\ \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,-t} \\
  0 & : & \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \\
  (*) & : & \forall B \mathord: \SynGL{2}.\ \tyPrimNm{real} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0}
\end{eqnarray*}

\paragraph{Euclidean Geometry} Euclidean geometry extends affine
geometry with the \emph{dot product}, or \emph{inner product},
operation of two vectors. The dot product is defined 
%on the coordinate representation 
by
%\begin{displaymath}
$  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2$.
%\end{displaymath}
% The dot product is used to define properties of vectors such as their
% \emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
% of the length of the vector $\vec{v}$, and the notion of
% orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
% if $\vec{v_1}\cdot\vec{v_2} = 0$.
To %incorporate the operation of dot product into our calculus, we
assign it a type we note that, although dot product is not
invariant under $\GL{2}$ or $\Transl{2}$, it
% that we have considered so far. However,e  the dot product 
is invariant under the subgroup $\Orth{2}$ of $\GL{2}$ of
\emph{orthogonal} linear transformations, i.e., the subgroup of
invertible linear maps whose matrix representations' transposes are
equal to their inverses. We thus introduce a new sort $\SynOrth{2}$ of
orthogonal transformations, and overload the multiplicative group
operations for inhabitants of $\SynOrth{2}$. Further assuming an
injection $\iota_O$ that takes $e : \SynOrth{2}$ to $\iota_O(e) :
\SynGL{2}$ we assign dot product this type:
%a type to the dot product, using quantification over $\SynOrth{2}$:
\begin{displaymath}
  (\cdot) : \forall O \mathord: \SynOrth{2}.\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
\end{displaymath}

\begin{figure}[t]
  \centering
  \begin{eqnarray*}
    0   &:& \forall s \mathord:\SynGL{1}.\ \tyPrim{real}{s} \\
    1   &:& \tyPrim{real}{1} \\
    (+) &:& \forall s \mathord:\SynGL{1}.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
    (-) &:& \forall s \mathord:\SynGL{1}.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
    (*) &:& \forall s_1,s_2 \mathord:\SynGL{1}.\ \tyPrim{real}{s_1} \to \tyPrim{real}{s_2} \to \tyPrim{real}{s_1s_2} \\
    (/) &:&
    \begin{array}[t]{@{}l@{}l}
      \forall s_1,s_2 \mathord:\SynGL{1}.\ \tyPrim{real}{s_1}\ & \to \tyPrim{real}{s_2} \\
      &\to \tyPrim{real}{s_1s_2^{-1}} + \tyPrimNm{unit} \\
    \end{array}\\
    \mathrm{abs} &:& \forall s \mathord:\SynGL{1}.\ \tyPrim{real}{s} \to \tyPrim{real}{\abs{s}} %\\
%    \mathrm{sqrt} &:& \tyPrim{real}{1} \to \tyPrim{real}{1}
  \end{eqnarray*}
  \caption{Operations on scaled real numbers}
  \label{fig:real-ops}
\end{figure}

The cross product of two vectors is defined on coordinate
representations as $(x_1,y_1) \times (x_2,y_2) = x_1y_2 - x_2y_1$.
Geometrically, the cross product is the signed area of the
parallelogram described by the pair of input vectors.  Under change of
basis by an invertible linear transformation $B$, the cross product of
two vectors varies with the determinant of $B$. This corresponds to
scaling the plane by the change of basis transformation, so we augment
our calculus with a new sort $\SynGL{1}$ of scale factors
(i.e.,~$1$-dimensional invertible linear maps). Semantically,
$\SynGL{1}$ ranges over the non-zero real numbers and forms an abelian
group which we write multiplicatively. We also add two new operations:
determinant, $\det B$, which takes an inhabitant of $\SynGL{2}$ to its
determinant in $\SynGL{1}$, and absolute value, $\abs{e}$, which takes
scaling factors to scaling factors. We also refine the type
$\tyPrimNm{real}$ of real numbers so that it is indexed by the sort
$\SynGL{1}$: $\tyPrim{real}{e}$. The old type $\tyPrimNm{real}$ is
then just $\tyPrim{real}{1}$, and the full collection of operations on
real numbers indexed by scaling factors is shown in
\autoref{fig:real-ops}. We can thus assign cross product the type:
%With the additional sort $\SynGL{1}$ and the refined type of real
%numbers, we can assign a type to the cross product:
\begin{displaymath}
  (\times) : \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \to
  \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B} 
\end{displaymath}
Since the absolute value of the determinant of an orthogonal
transformation is always $1$, we assume $\abs{\det (\iota_O O)} = 1$
to hold for any $O \in \SynOrth{2}$.

\begin{example}\label{ex:area-of-triangle-1}
  We can use the operations of this subsection to
  compute the area of a triangle. We have:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall B\mathord:\SynGL{2},
      t\mathord:\SynTransl{2}.\ \\
      \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to
      \tyPrim{vec}{B, t} \to \tyPrim{real}{\abs{\det B}}
      \\ \mathrm{area}\ [B]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} *
      \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
    \end{array}
  \end{displaymath}
  The calculation is performed in several steps, each of which removes
  some of the symmetry described by the type of
  $\mathrm{area}$. First, the two offset vectors $p_2 - p_1$ and $p_3
  - p_1$ are computed. These operations remove the effect of
  translations on the result in exactly the same way as the type
  isomorphism in \exref{ex:type-iso}. Next, we compute the cross
  product of the two vectors, which gives the area of the
  parallelogram described by the sides of the triangle and has type
  $\tyPrim{real}{\det B}$. This removes some of the symmetry due to
  invertible linear maps, but the cross product still varies with the
  sign of the determinant. We remove this symmetry as well using
  $\mathrm{abs}$. This gives a value of type $\tyPrim{real}{\abs{\det B}}$
  which we multiply by $\frac{1}{2}$ to recover the area of the
  triangle rather than that of the whole parallelogram. If we
  specialise $\mathrm{area}$ to just orthogonal transformations, the
  assumption $|\det (\iota_OO)| = 1$ gives the following type:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall O\mathord:\SynOrth{2},
      t\mathord:\SynTransl{2}.\ \\
      \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to
      \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to
      \tyPrim{real}{1}
    \end{array}
  \end{displaymath}
  This type shows that the area of a triangle is invariant under
  orthogonal transformations and translations. Combinations of such
  transformations are {\em isometries}, i.e., distance preserving
  maps.
\end{example}

% \begin{enumerate}
% \item Possibly do a $n$-body gravity simulator as an example?
% \item Have a special sort and type for rotations. An operation for
%   computing the rotation between two points, and for applying a
%   rotation to a vector. This will allow for more free theorems...
% \end{enumerate}

\subsection{Scale Invariance and Dimensional Analysis}
\label{sec:scale-invariance}

Indexing types by scaling factors brings us to the original
inspiration for the current work: Kennedy's interpretation of his
units of measure type system via scaling invariance
\cite{kennedy97relational}. Kennedy shows how interpreting types in
terms of scaling invariance brings the techniques of dimensional
analysis to bear on programming. The types of the real number
arithmetic operations in \autoref{fig:real-ops} are exactly the types
Kennedy assigns in his units of measure system, except for that of the
absolute value operation. Semantically, our type indexes by
%our assigned type is that, semantically, we are indexing by
non-zero scaling factors, whereas Kennedy's indexes by strictly
positive ones.

In our two-dimensional setting we can add to Kennedy's one-dimensional
scaling invariance an operation $\iota_1$ that, semantically, takes
scale factors in $\GL{1}$ to invertible linear maps in $\GL{2}$,
i.e., takes numbers $s$ to matrices $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$.  This
operation satisfies the equation $\det (\iota_1 s) = s^2$, indicating
that scaling the plane by $s$ in both directions scales areas by
$s^2$. %, as we now illustrate.

\begin{example}\label{ex:area-of-triangle-2}
  Just as we specialised the type of the $\mathrm{area}$ function to
  orthogonal transformations in \exref{ex:area-of-triangle-1}, we can
  also specialise $\mathrm{area}$'s type to scaling
  transformations. This yields the type:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall s\mathord:\SynGL{1}, t\mathord:\SynTransl{2}.\ \\
      \hspace{0.8cm} \tyPrim{vec}{\iota_1s, t} \to \tyPrim{vec}{\iota_1s, t} \to \tyPrim{vec}{\iota_1s, t} \to \tyPrim{real}{s^2}
    \end{array}
  \end{displaymath}
  As expected, the area of a triangle varies with the square of
  scalings of the plane, and this is reflected in the type.
\end{example}

Linear maps of the form $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$, as
generated by $\iota_1$, commute with all other invertible linear
maps. We thus require
$(\iota_1 s)B = B(\iota_1 s)$ to hold.
The scaling maps $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$
are precisely the elements of $\GL{2}$ that commute with all others;
these form the \emph{centre} of $\GL{2}$. If we keep track of
scalings, then we can assign the more precise types to
scalar multiplication and dot product. These
 are shown in \autoref{fig:vec-ops}, which summarises the most general types 
of all the vector operations that we have described.

\begin{example}
  With the operations in \autoref{fig:real-ops}, it is not possible to
  write a term with the following type that is not constantly zero:
  \begin{displaymath}
    \forall s \mathord: \SynGL{2}.\ \tyPrim{real}{s^2} \to \tyPrim{real}{s}
  \end{displaymath}
  This was shown by Kennedy for his units of measure system
  \cite{kennedy97relational}.  In particular, it is not possible to
  write a square root function with the above type.  The
  non-definability of square root is similar to the uninhabitation of the
  type in \exref{ex:uninhabited-type}.

  In \autoref{sec:types-indexed-abelian-groups-indef} we revisit
  Kennedy's result and show that even if we add square root as a
  primitive operation---with the type above---then it is still not
  possible to construct the cube root function. The non-definability of
  cube root is related to the impossibility of trisecting an arbitrary
  angle by ruler and compass constructions.
%  is of interest due to its relevance to the classical problem of
%  trisecting an angle by ruler and compass constructions.
\end{example}

\begin{figure}[t]
  \centering
  \begin{align*}
    0 & : \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \\
    (+) & :
    \begin{array}[t]{@{}l}
      \forall B \mathord: \SynGL{2}, t_1,t_2 \mathord: \SynTransl{2}. \\
      \tyPrim{vec}{B,t_1} \to \tyPrim{vec}{B,t_2} \to
      \tyPrim{vec}{B,t_1 + t_2}
    \end{array}
    \\
    \mathrm{negate} & : \forall B \mathord: \SynGL{2}, t \mathord: \SynTransl{2}.\ \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,-t} \\
  (*) & : 
  \begin{array}[t]{@{}l}
    \forall s \mathord: \SynGL{1}, B \mathord:\SynGL{2}.\ \\
    \tyPrim{real}{s} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{\iota_1(s)B,0}
  \end{array}\\
    \mathrm{affComb} &:
    \begin{array}[t]{@{}l}\forall B \mathord: \SynGL{2}, t \mathord:
    \SynTransl{2}.\\ \tyPrim{vec}{B,t} \to \tyPrim{real}{1} \to
    \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,t}
    \end{array} \\
    (\cdot) & :
\begin{array}[t]{@{}l}
    \forall s \mathord: \SynGL{1}, O \mathord: \SynOrth{2}.\ \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\\
    \hspace{2cm}\tyPrim{vec}{\iota_1(s)\iota_O(O),0}\to \tyPrim{real}{s^2}
\end{array}\\
  (\times) & : \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \to
  \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B} 
\end{align*}
  \caption{Operations on vectors}
  \label{fig:vec-ops}
\end{figure}

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:

%%%---------- close: geometry-twocolumn.tex
%%%%%%%%%%% Springe nach section-general-framework.tex
%%%---------- open: section-general-framework.tex
\newcommand{\Gen}{\mathrm{Gen}}
\newcommand{\Free}{\mathrm{Free}}
\newcommand{\semSort}[1]{\llbracket #1 \rrbracket}
\newcommand{\semIndexExp}[1]{\llbracket #1 \rrbracket}
\newcommand{\semPrimType}[1]{\llbracket #1 \rrbracket}
\newcommand{\opsCtxt}{\Gamma_{\mathit{ops}}}
\newcommand{\opsEnv}{\eta_{\mathit{ops}}}
\newcommand{\ctxteq}[5]{{#1};{#2}\vdash{#3}\approx{#4}:{#5}}
\newcommand{\semeq}[5]{{#1};{#2}\models{#3}\sim{#4}:{#5}}
\section{A General Framework}
\label{sec:a-general-framework}

%In this section 
We now present our framework for algebraically indexed types and its
relational interpretation.  We define the syntax of algebraically
indexed types (\autoref{sec:algebraically-indexed-types}) and a syntax
for terms in a general programming language for algebraically indexed
types (\autoref{sec:well-typed-programs}). We give an index-erasure
semantics to types and terms (\autoref{sec:erasure-semantics}), and
based on this semantics define notions of contextual equivalence and
type isomorphism. We
then introduce a relational semantics for types parameterised by an appropriate
`model' of the algebraic theory (\autoref{sec:relational-semantics}),
prove the central Abstraction Theorem
and use the relational semantics
to define a notion of semantic equivalence that soundly approximates 
contextual equivalence (\autoref{sec:abstraction-theorem}).

We will use a type system for affine geometry as a running example
throughout, so that by the end of the section we have prepared enough
syntactic and semantic gadgets to let us prove invariance
and abstraction properties for geometric examples in
\autoref{sec:instantiations}.
%and semantics of algebraically indexed types, then their semantics
%and finally we present a general programming language for
%algebraically indexed types, together with an abstraction theorem.

\subsection{Algebraically-Indexed Types}
\label{sec:algebraically-indexed-types}

The index expressions and types of an instantiation of our general
framework are derived from the following data:
\begin{enumerate}
\item A collection $\SortSet$ of index sorts. We use the
  meta-syntactic variables $s,s_1,s_2,\ldots$ for arbitrary sorts taken
  from $\SortSet$.
\item A collection $\IndexOpSet$ of index operations, with a function
  $\indexOpArity : \IndexOpSet \to \SortSet^* \times \SortSet$. (We use
  the notation $A^*$ to denote the set of lists of elements of some
  set $A$.)
\item A collection $\PrimTypeSet$ of primitive types, with a function
  $\primTyArity : \PrimTypeSet \to \SortSet^*$, describing the sorts
  of the arguments of each primitive type.
\end{enumerate}

\begin{example*}[Geometry: syntax]
  \label{ex:two-dim-geo-operations}
  The two-dimensional geometry system has a sort for each of the
  geometric groups mentioned in \autoref{sec:motivating-examples}, so
  $\SortSet = \{\SynTransl{2}, \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$.
  %For the index operations, 
  We have additive group structure on $\SynTransl{2}$, multiplicative
  group structure on $\SynGL{1}$, $\SynGL{2}$, and $\SynOrth{2}$,
  injections from $\SynOrth{2}$ and $\SynGL{1}$ into $\SynGL{2}$,
  determinant, and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, \abs{\cdot} \}$, where
  $G \in \{\SynGL{1}, \SynGL{2}, \SynOrth{2}\}$, and
  %  with the following assignment of arities:
  \begin{displaymath}
    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
      \indexOpArity(0) & ([], \SynTransl{2}) &
      \indexOpArity(1_G) & ([], G) \\
      \indexOpArity(+) & ([\SynTransl{2}, \SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(\cdot_G) & ([G,G],G) \\
      \indexOpArity(-) & ([\SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(^{-1_G}) & ([G], G) \\
      \indexOpArity(\iota_O) & ([\SynOrth{2}], \SynGL{2}) &
      \indexOpArity(\iota_1) & ([\SynGL{1}], \SynGL{2}) \\
      \indexOpArity(\det) & ([\SynGL{2}], \SynGL{1}) &
      \indexOpArity(\abs{\cdot}) & ([\SynGL{1}], \SynGL{1})
    \end{array}
  \end{displaymath}
  The intended interpretations of the top three pairs of operations
  are group unit, group combination and group negation,
  respectively. 
  When we discuss equational theories on index expressions 
in \autoref{sec:type-equality} 
we will impose the (abelian) group laws. For this example, 
we also have $\PrimTypeSet = \{ \tyPrimNm{vec},
  \tyPrimNm{real} \}$, with $\primTyArity(\tyPrimNm{vec}) =
           [\SynGL{2}, \SynTransl{2}]$ and
           $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
\qed
\end{example*}

We assume a countably infinite collection of index variable names $i,
i_1, i_2$, \emph{etc}. \emph{Index contexts} $\Delta = i_1 \mathord: s_1,
\ldots, i_n \mathord: s_n$ are lists of variable/sort pairs such that
all the variable names are distinct.
\begin{figure*}[t]
  \centering
  % \textbf{Index contexts}\\

  % $\Delta = i_1 \mathord: s_1, ..., i_n \mathord: s_n$, where each $s_i \in \SortSet$ and no variable name is repeated.

  % \bigskip

  {\small
  \textbf{Well-sorted index expressions}
  \begin{mathpar}
    \inferrule* [right=IVar]
    {i : s \in \Delta}
    {\Delta \vdash i : s}

    \inferrule* [right=IOp]
    {\indexOp{f} \in \mathit{IndexOp} \\
      \indexOpArity(\indexOp{f}) = ([s_1,\ldots,s_n], s) \\
      \{\Delta \vdash e_j : s_j\}_{1 \leq j \leq n}}
    {\Delta \vdash \indexOp{f}(e_1, \ldots, e_n) : s}
  \end{mathpar}

  \smallskip

  \textbf{Well-indexed types}
  \begin{mathpar}
    \inferrule* [right=TyPrim]
    {\tyPrimNm{X} \in \mathit{PrimType} \\\\
      \primTyArity(\tyPrimNm{X}) = [s_1,\ldots,s_n] \\
      \{\Delta \vdash e_j : s_j\}_{1\leq j \leq n}}
    {\Delta \vdash \tyPrim{X}{e_1,\ldots,e_n} \isType}

    \inferrule* [right=TyUnit]
    { }
    {\Delta \vdash \tyUnit \isType}

    \inferrule* [right=TyArr]
    {\Delta \vdash A \isType \\ \Delta \vdash B \isType}
    {\Delta \vdash A \tyArr B \isType}

    \inferrule* [right=TyTuple]
    {\Delta \vdash A \isType \\ \Delta \vdash B \isType}
    {\Delta \vdash A \tyProduct B \isType}

    \inferrule* [right=TySum]
    {\Delta \vdash A \isType \\ \Delta \vdash B \isType}
    {\Delta \vdash A + B \isType}

    \inferrule* [right=TyForall] %FIXME: macroize forall
    {\Delta, i \mathord: s \vdash A \isType}
    {\Delta \vdash \forall i \mathord: s. A \isType}

    \inferrule* [right=TyEx]
    {\Delta, i \mathord: s \vdash A \isType}
    {\Delta \vdash \exists i \mathord: s. A \isType}
  \end{mathpar}}
  \caption{Index expressions and types}
  \label{fig:indexes-and-types}
\end{figure*}
%Given the above data, 
The rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple: either an index expression
is a variable that appears in the context (rule \TirName{IVar}), or it
is an application of an index operation taken from $\IndexOpSet$ to
other index expressions (rule \TirName{IOp}). 
The rules for well-indexed types include the usual ones
%rules 
for
%constructing types of 
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). 
We use $\tyPrimNm{bool}$ as an abbreviation for $\tyPrimNm{unit} + \tyPrimNm{unit}$.
The rule \TirName{TyPrim} %allows us to 
forms, from a primitive type $\tyPrimNm{X}$ and appropriately sorted
index expressions $e_1,\ldots,e_n$, the well-indexed type
$\tyPrim{X}{e_1,\ldots,e_n}$. The rule \TirName{TyForall} 
%permits the formation of 
forms universally quantified types, where the universal
quantification ranges over all index expressions of some
sort. Existential types, %are constructed 
formed using the \TirName{TyEx} rule,
%and 
allow for abstraction by hiding. %FIXME: think about the
%introduction of existentials more, perhaps forward ref to their use

\subsubsection{Substitution of Index Expressions}
\label{sec:simultaneous-substitution}

It %will be technically 
is convenient to express substitution of index
expressions %in our framework 
in terms of simultaneous substitutions.
Given a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord:
s_1, \ldots, i_n \mathord: s_n$, a (simultaneous) \emph{substitution}
$\Delta \vdash \sigma \Rightarrow \Delta'$ is a sequence of
expressions $\sigma = (e_1,\ldots,e_n)$ such that $\Delta \vdash e_j :
s_j$ for all $1 \leq j \leq n$. Given a substitution
$\Delta \vdash \sigma = (e_1,\ldots,e_n) \Rightarrow \Delta'$ and a
variable $i_j \mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for
the index expression $e_j$. We write $\Delta \Rightarrow \Delta'$ for
the set of all substitutions $\sigma$ such that $\Delta
\vdash \sigma \Rightarrow \Delta'$.
%
%It will be useful to 
We can think of any sequence of sorts as an index
context. In particular, we will make use of substitutions
of the form $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$, since these are exactly sequences of
index arguments suitable for the primitive type $\tyPrimNm{X}$. By
further abuse of notation, we write $\Delta \Rightarrow
\primTyArity(\tyPrimNm{X})$ for the set of all 
substitutions $\sigma$ such that $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$.

For a substitution $\Delta \vdash \sigma \Rightarrow
\Delta'$, where $\Delta' = i_1\mathord:s_1,\ldots,i_n\mathord:s_n$, and a
variable/sort pair $i\mathord:s$ such that $i$ does not appear in
either $\Delta$ or $\Delta'$, we can form the \emph{lifted}
substitution $\Delta,i\mathord:s \vdash
\sigma_{i\mathord:s} = (\sigma(i_1), \ldots, \sigma(i_n), i) \Rightarrow
\Delta',i\mathord:s$. %Note that we have implicity used the fact that
%well-sortedness of index expressions is preserved by addition of extra
%items to the context.
%
Application of a substitution $\Delta \vdash \sigma
\Rightarrow \Delta'$ to a well-sorted index expression $\Delta' \vdash
e : s$ yields a well-sorted index expression $\Delta \vdash \sigma^*e
: s$. The expression $\sigma^*e$ is defined on variables as $\sigma^*i
\isDefinedAs \sigma(i)$, and on operation symbols as
$\sigma^*(\indexOp{f}(e_1,\ldots,e_n)) \isDefinedAs
\indexOp{f}(\sigma^*e_1, \ldots, \sigma^*e_n)$.  %Similarly, 
Given
%a well-indexed type 
$\Delta' \vdash A \isType$, 
%can apply $\sigma$ to $A$ to produce a new well-indexed type 
we have $\Delta \vdash \sigma^*A
\isType$. The key clauses defining $\sigma^*A$ are for primitive types
and the universal and existential quantifiers:
\begin{displaymath}
  \begin{array}{c}
    \sigma^*(\tyPrim{X}{e_1,\ldots,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,\ldots,\sigma^*e_n}
    \\
    \begin{array}{c@{\hspace{2em}}c}
      \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A
      &
      \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:s. \sigma_{i\mathord:s}^*A
    \end{array}
  \end{array}
\end{displaymath}
% \begin{lemma}
%   Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
%   substitution.
%   \begin{enumerate}
%   \item If $\Delta' \vdash e : s$, then $\Delta \vdash \sigma^*e : s$; and
%   \item If $\Delta' \vdash A \isType$, then $\Delta \vdash \sigma^*A
%     \isType$.
%   \end{enumerate}
% \end{lemma}
The \emph{identity} substitution $\Delta \vdash
\id_\Delta \Rightarrow \Delta$ is %just the sequence of variables in
%$\Delta$: 
$\id_\Delta = (i_1,\ldots,i_n)$ where $\Delta =
i_1\mathord:s_1,\ldots,i_n\mathord:s_n$. The \emph{composition} of 
two substitutions 
$\Delta \vdash \sigma \Rightarrow \Delta'$
and $\Delta' \vdash \sigma' \Rightarrow \Delta''$, where $\sigma' =
(e'_1,\ldots,e'_n)$, is defined as $\Delta \vdash \sigma' \circ \sigma
\isDefinedAs (\sigma^*e'_1, \ldots, \sigma^*e'_n) \Rightarrow \Delta''$.
%
Given a context $\Delta = i_1\mathord:s_1,\ldots,i_n\mathord:s_n$, and a
variable/sort pair $i\mathord:s$ such that $i$ does not appear in
$\Delta$, we define the \emph{projection} substitution
$\Delta, i\mathord:s \vdash \pi_{i\mathord:s} \Rightarrow \Delta$ as
$\pi_{i\mathord:s} = (i_1,\ldots,i_n)$. %The subscript on
%$\pi_{i\mathord:s}$ is the variable/sort pair that is being discarded.

\subsubsection{Index Expression Equality and Type Equality}
\label{sec:type-equality}

Much of the power of indexing types by the expressions of an algebraic
theory comes from the equations of the theory. 
%In the %two-dimensional geometry example of 
For example, in 
\autoref{sec:motivating-examples} the types
$\tyPrim{vec}{B,t_1 + t_2}$ and $\tyPrim{vec}{B, t_2 + t_1}$ are
considered equal by the type system
%due to the commutativity of the $+$ operation.
because $+$ is commutative.
%
In the general framework, the equations between types are derived from
a set $\IndexAxiomSet$ of axioms $\Delta \vdash e \stackrel{ax}\equiv
e' : s$ that %We assume that all axioms $(\Delta \vdash e
%\stackrel{ax}\equiv e' : s)$ % \in \IndexAxiomSet$ 
are well-sorted, in the sense that both $\Delta \vdash e : s$ and $\Delta \vdash e' : s$
hold.

Given a set $\IndexAxiomSet$ of axioms, we generate the equality
judgment between index expressions $\Delta \vdash e \equiv e' : s$ by
a set of rules. The following rule lets us use substitution
instances of axioms:
\begin{displaymath}
  \inferrule*
  {(\Delta' \vdash e \stackrel{ax}\equiv e' : s) \in \IndexAxiomSet \\
    \Delta \vdash \sigma \Rightarrow \Delta'}
  {\Delta \vdash \sigma^*e \equiv \sigma^*e' : s}
\end{displaymath}
We also assume the standard congruence, reflexivity, symmetry and
transitivity rules for the equality judgment.

\begin{example*}[Geometry: axioms]
  \label{ex:two-dim-geo-axioms}
  In \autoref{sec:motivating-examples} we assumed 
  %that 
  various equational axioms %hold 
  for indexing expressions standing for
  elements of geometric groups. Assuming the abelian group axioms for
  translations we can formalise this in our %general 
  framework:
  %We can now make this assumption formal
  %in our general framework. Semantically, translations form an abelian
  %group under addition, so we assume the abelian group axioms for
  %translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl{2} \vdash t + 0 \stackrel{ax}\equiv t : \SynTransl{2} \\
      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 0 : \SynTransl{2} \\
      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
  group under multiplication, and the sorts $\SynGL{2}$ and
  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $\abs{\cdot}$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL{2}$. The absolute value of the
  determinant of an orthgonal transformation is always $1$, so we also
  assume $\abs{\det(\iota_OO)} \stackrel{ax}\equiv 1$. Similarly,
  scaling maps have a determinant expressible in terms of other
  operations: $\det(\iota_1(s)) \stackrel{ax}\equiv s \cdot s$. We
  also assume the axiom $|s^2| \stackrel{ax}\equiv s^2$.  \qed
\end{example*}

The equality judgment $\Delta \vdash e \equiv e' : s$ on index
expressions generates the equality judgment $\Delta \vdash A \equiv
B \isType$ on types. The basic rule generating equality judgments on
types equates %states that %two 
applications of primitive types %are equal 
if
their %index expression 
arguments are equal:
\begin{displaymath}
  \inferrule*
  {\{ \Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j \leq n}}
  {\Delta \vdash \tyPrim{X}{e_1,\ldots,e_n} \equiv \tyPrim{X}{e'_1,\ldots,e'_n} \isType}
\end{displaymath}
The rest of the rules for equality on types ensure that it is a
congruence relation 
and an equivalence relation.
%on types. %and that it is an equivalence relation
%(i.e.,~equality on types is reflexive, symmetric and transitive).
% For example, for universally quantified types we have
% the following congruence rule:
% \begin{displaymath}
%   \inferrule*
%   {\Delta, i : s \vdash A \equiv B \isType}
%   {\Delta \vdash \forall i\mathord:s.A \equiv \forall i\mathord:s.B \isType}
% \end{displaymath}
% The congruence rules for the other type formers are similar.

% FIXME: think about integrating this lemma into the running text
% \begin{lemma}
%   Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
%   substitution.
%   \begin{enumerate}
%   \item If $\Delta' \vdash e \equiv e' : s$ then $\Delta \vdash
%     \sigma^*e \equiv \sigma^*e' : s$; and
%   \item If $\Delta' \vdash A \equiv B \isType$ then $\Delta \vdash
%     \sigma^*A \equiv \sigma^* B \isType$.
%   \end{enumerate}
% \end{lemma}

%A pair of 
The substitutions $\Delta \vdash \sigma \Rightarrow
\Delta'$ and $\Delta \vdash \sigma' \Rightarrow \Delta'$ are defined
to be equal, and written  $\Delta \vdash \sigma \equiv \sigma' \Rightarrow
\Delta'$, if their component expressions are equal in the context
$\Delta$: i.e.,~ if $\Delta \vdash e_j \equiv e'_j : s_j$, for all
$j$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Well-typed terms}
\label{sec:well-typed-programs}

We now present the rules for well-typed terms over the 
collection of 
types we defined 
in \autoref{sec:algebraically-indexed-types}.

%Each
%well-typed program is assigned an index-erasure semantics, building on
%the index-erasure semantics of types %we defined 
%in \autoref{sec:index-erasure-semantics}. Our main result
%(\thmref{thm:abstraction}) is that the index-erasure semantics of
%every well-typed program is related to itself in the relational
%interpretation of its type: this is the abstraction theorem for every
%instantiation of our general framework.

Well-typed terms are defined with respect to well-indexed typing
contexts, which are in turn defined with respect to an index
context. Well-indexed typing contexts with respect to an index context
$\Delta$ are sequences of variable/type pairs with no repeated
variable names such that each type is well-indexed with respect to
$\Delta$. Formally, well-indexed typing contexts are 
given by
%defined by the following two rules:
\begin{mathpar}
  \inferrule*
  { }
  {\Delta \vdash \epsilon \isCtxt}

  \inferrule*
  {\Delta \vdash \Gamma \isCtxt \\ \Delta \vdash A \isType \\ x \not\in \Gamma}
  {\Delta \vdash \Gamma, x : A \isCtxt}
\end{mathpar}
Application of substitutions extends to typing contexts
by applying the substitution to each type.

%The typing rules for our framework define 
Well-typed terms
are defined
with respect to an index context $\Delta$ and a type context $\Delta \vdash
\Gamma \isCtxt$. The judgment $\Delta; \Gamma \vdash M : A$ is defined
in \autoref{fig:programs}. The equational theory on types is
incorporated into the type system via the rule \TirName{TyEq}, which
allows a term that has %is judged to have 
type $A$ to also have any
%other 
equal type $B$ as well.
\begin{figure*}[t]
  \centering
  {\small
  \begin{mathpar}
    \inferrule* [right=Var]
    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
    {\Delta; \Gamma \vdash x : A}

    \inferrule* [right=TyEq]
    {\Delta; \Gamma \vdash M : A \\ \Delta \vdash A \equiv B \isType}
    {\Delta; \Gamma \vdash M : B}

    \inferrule* [right=Unit]
    {\Delta \vdash \Gamma \isCtxt}
    {\Delta; \Gamma \vdash * : 1}

    \inferrule* [right=Pair]
    {\Delta; \Gamma \vdash M : A \\
      \Delta; \Gamma \vdash N : B}
    {\Delta; \Gamma \vdash (M, N) : A \tyProduct B}

    \inferrule* [right=Proj1]
    {\Delta; \Gamma \vdash M : A \tyProduct B}
    {\Delta; \Gamma \vdash \pi_1 M : A}

    \inferrule* [right=Proj2]
    {\Delta; \Gamma \vdash M : A \tyProduct B}
    {\Delta; \Gamma \vdash \pi_2 M : B}

    \inferrule* [right=Inl]
    {\Delta; \Gamma \vdash M : A}
    {\Delta; \Gamma \vdash \mathrm{inl}\ M : A + B}

    \inferrule* [right=Inr]
    {\Delta; \Gamma \vdash M : B}
    {\Delta; \Gamma \vdash \mathrm{inr}\ M : A + B}

    \inferrule* [right=Case]
    {\Delta; \Gamma \vdash M : A + B \\\\
      \Delta; \Gamma, x : A \vdash N_1 : C \\\\
      \Delta; \Gamma, y : B \vdash N_2 : C}
    {\Delta; \Gamma \vdash \textrm{case}\ M\ \textrm{of}\ \textrm{inl}\ x.N_1; \textrm{inr}\ y.N_2 : C}

    \inferrule* [right=Abs]
    {\Delta; \Gamma, x : A \vdash M : B}
    {\Delta; \Gamma \vdash \lambda x.M : A \tyArr B}

    \inferrule* [right=App]
    {\Delta; \Gamma \vdash M : A \tyArr B \\
      \Delta; \Gamma \vdash N : A}
    {\Delta; \Gamma \vdash M N : B}

    \inferrule* [right=UnivAbs]
    {\Delta, i \mathord: s; \pi_{i\mathord:s}^*\Gamma \vdash M : A}
    {\Delta; \Gamma \vdash \Lambda i. M : \forall i\mathord:s. A}

    \inferrule* [right=UnivApp]
    {\Delta; \Gamma \vdash M : \forall i\mathord:s. A \\ \Delta \vdash e : s}
    {\Delta; \Gamma \vdash M [e] : (\id_\Delta, e)^*A}

    \inferrule* [right=ExPack]
    {\Delta; \Gamma \vdash M : (\id_\Delta, e)^*A \\ \Delta \vdash e : s \\ \Delta, i\mathord:s \vdash A \isType}
    {\Delta; \Gamma \vdash \langle[e], M\rangle: \exists i\mathord:s. A}

    \inferrule* [right=ExUnpack]
    {\Delta; \Gamma \vdash M : \exists i\mathord:s. A \\\\
      \Delta, i\mathord:s; \pi_{i\mathord:s}^*\Gamma, x : A \vdash N : \pi_{i\mathord:s}^*B}
    {\Delta; \Gamma \vdash \mathrm{let}\langle[i],x\rangle = M\ \mathrm{in}\ N : B}
  \end{mathpar}}

  \caption{Well-typed terms}
  \label{fig:programs}
\end{figure*}

For any particular theory we assume that there is a closed
typing context $\opsCtxt$ that describes the types of the primitive operations.
\begin{example*}[Geometry: operations]
For geometry $\opsCtxt$ would collect together the types of primitive operations as 
listed in \autoref{fig:real-ops} and \autoref{fig:vec-ops}.
\qed
\end{example*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Index-Erasure Semantics}
\label{sec:erasure-semantics}
%\fixme{Make sure that the semantics is motivated}
Having defined the syntax of algebraically indexed types and terms,
we turn to their denotational interpretation.  We first define an
\emph{index-erasure} interpretation of types and terms that interprets
every well-indexed type as a set, ignoring the indexing
expressions, and which interprets open terms as functions that map environments
to final values.
%Building on the index-erasure semantics, we
%We then define in
%\autoref{sec:relational-semantics} the relational
%interpretation of types.

% For both of our semantics of types, we state and prove two properties:
% types that are syntactically equal have equal denotations, and that
% substitution of index terms is interpreted via composition. These
% properties ensure that our semantics of types is well-behaved.

\paragraph{Interpretation of Types.}
%\label{sec:index-erasure-semantics}

The defining feature of the index erasure interpretation is that 
semantics of a well-indexed type $\tyPrim{X}{e_1,\ldots,e_n}$ is
determined solely by the primitive type $\tyPrimNm{X}$ and not by the
index expressions $e_1,\ldots,e_n$. %Accordingly, 
We thus assume each
primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
$\tyPrimSem{\tyPrimNm{X}}$ and %We 
extend this assignment to %every
well-indexed types by induction on the type structure:
\begin{displaymath}
  \begin{array}{@{}c@{\hspace{0em}}c@{}}
    \begin{array}{@{}l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,\ldots,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A} \\
      \tySem{\exists i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
(We will overload the
notation $\tmSem\cdot$ for all index-erasure interpretations, 
and reserve notation $\llbracket\cdot\rrbracket$ for the
index-observing relational semantics defined later.)
% The interpretation of the unit type is a chosen one element set, and
% the interpretations of the function, tuple and sum types are simply
% the corresponding constructions on sets. 
%: the
%interpretation of the type $\forall i\mathord:s.A$ is exactly the
%interpretation of the type $A$, and likewise for $\exists
%i\mathord:s.A$.

The index-erasure interpretation completely ignores index expressions
and quantifiers, and type equality is defined as an extension of index
equality. Therefore, it is straightforward to prove that equal types
have equal denotations when interpreted in the index-erasure
semantics, and that substitution of index terms has no effect on the
index-erasure interpretation of types:
\begin{lemma}\label{lem:tyeqsubst-erasure}
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$ then $\tySem{A} =
    \tySem{B}$; and
  \item If $\Delta' \vdash A \isType$ and $\Delta \vdash \sigma
    \Rightarrow \Delta'$, then $\tySem{\sigma^*A} = \tySem{A}$.
  \end{enumerate}
\end{lemma}

\paragraph{Interpretation of Terms.}

% FIXME: do we need to prove something about the effect
% of substitution on the erasure semantics of contexts?

We assign an index-erasure semantics to any well-indexed typing
context $\Delta \vdash \Gamma \isCtxt$ by induction:
$\ctxtSem{\epsilon} = \{*\}$ and $\ctxtSem{\Gamma, x : A} =
\ctxtSem{\Gamma} \times \tySem{A}$. For a well-typed term $\Delta;
\Gamma \vdash M : A$, we define the \emph{erasure interpretation} as a
function $\tmSem{M} : \ctxtSem{\Gamma} \to \tySem{A}$ that completely
ignores the indexing information. In light of
\lemref{lem:tyeqsubst-erasure}, we do this %we define this function is defined
directly on the syntax of well-typed terms, rather than on typing
derivations. The definition of $\tmSem{M}$ is completely standard,
except for the clauses for universal and existential types:
\begin{displaymath}
  \begin{array}{@{}l}
    \begin{array}{@{}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l@{\hspace{2em}}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l@{}}
      \tmSem{\Lambda i.\ M}\eta & \isDefinedAs & \tmSem{M}\eta
      &
      \tmSem{M[e]}\eta & \isDefinedAs & \tmSem{M}\eta \\
      \tmSem{\langle [e], M\rangle}\eta & \isDefinedAs & \tmSem{M}\eta
    \end{array} \\
    \tmSem{\mathrm{let}\langle[i], x\rangle = M\ \mathrm{in}\ N}\eta \isDefinedAs \tmSem{N}(\eta, \tmSem{M}\eta)
  \end{array}
\end{displaymath}

For any particular theory we assume that there is an
interpretation of the primitive operations $\opsEnv \in \ctxtSem{\opsCtxt}$.

\begin{example*}[Geometry: interpretation]
  The two-dimensional geometry instantiation of the general framework
  uses the assignment $\tyPrimSem{\tyPrimNm{vec}} = \mathbb{R}^2$ and
  $\tyPrimSem{\tyPrimNm{real}} = \mathbb{R}$.
  We assume that $\opsEnv$ gives the usual interpretation to scalar
  and vector operations from
  \autoref{fig:real-ops} and \autoref{fig:vec-ops}.
\qed
\end{example*}

\paragraph{Contextual Equivalence.}
We use our index-erasure semantics to define when a pair of terms are
contextually equivalent with respect to syntactically defined
contexts, following Hofmann \cite{hofmann08correctness}.  Given an
index context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, we write
$\forall \Delta.A$ for $\forall i_1\mathord:s_1.\dots\forall
i_n\mathord:s_n.A$, and similarly for $\Lambda\Delta.M$ and $\lambda
\Gamma.M$.
\begin{definition}[Contextual Equivalence]\label{defn:ctxt-equiv}
Two terms $\Delta; \opsCtxt, \Gamma \vdash M_1, M_2 : A$
are \emph{contextually equivalent}, written $\ctxteq\Delta\Gamma{M_1}{M_2}A$, if 
  for all
  contexts $\cdot;\opsCtxt \vdash C : (\forall \Delta. \Gamma
  \to A) \to \tyPrimNm{bool}$, it is the case that $\tmSem{C\ (\Lambda
    \Delta. \lambda \Gamma. M_1)}\opsEnv = \tmSem{C\
    (\Lambda \Delta. \lambda \Gamma. M_2)}\opsEnv$.
\end{definition}

\paragraph{Type Isomorphism.}
We say that well-indexed types $\Delta \vdash A \isType$ and $\Delta\vdash B \isType$
are isomorphic, and write $\Delta\vdash A \cong
B$, if there exist maps between them that are mutually inverse with
respect to contextual equivalence, i.e., if there are terms $\Delta\vdash I
: A \to B$ and $\Delta\vdash J : B\to A$ such that $\ctxteq{\Delta}{x:A}{J(I(x))}{x}{A}$ 
and $\ctxteq{\Delta}{y:B}{I(J(y))} y B$.
%\[
%x:A\vdash J(I(x))\stackrel{ctx}\approx x : A
%\text{ and }y:B\vdash I(J(y))\stackrel{ctx}\approx y : B.
%\]
That $\cong$ is a congruence with respect to the type formation rules
of \autoref{fig:indexes-and-types} is straightforward. We can also
derive isomorphisms that are independent of the indexing theory, such
as $\Delta\vdash A\times B \cong B \times A$, and $\Delta\vdash\forall
i\mathord:s.(A\to B)\cong A\to\forall i\mathord:s.B$ for $i$ not free in $A$.

\subsection{The Relational Interpretation of Types}
\label{sec:relational-semantics}

%We now define 
The relational semantics of the well-indexed type
$\Delta \vdash A \isType$ is a binary relation on the index-erasure
interpretation of $A$. We write $\Rel(X)$ for the set
of binary relations $R \subseteq X \times X$ on the set $X$.

For the unit, tuple, sum and function types we define the
relational interpretation as a standard logical relation. The
relational interpretations of primitive types with index arguments and
the universally quantified types require an interpretation of index
contexts.

An \emph{index environment} $\rho$ assigns to each index variable $i:s$
in the context $\Delta$ a value drawn from an interpretation of
the sort $s$ that soundly models the equational theory associated with $s$. We call
such an interpretation a \emph{model}: it assigns to each index operation in the equational 
theory a corresponding operation in the interpretation, so that index expressions 
can be interpreted by recursion on their structure.

For example, the sort $\SynTransl{2}$ of translations can be modelled by
any abelian group. An obvious candidate here is the additive group over 
$\mathbb R^2$, which we will use to obtain invariance under translation;
but we could use the additive group $\mathbb Q$, or even a finite group such as the 
two-element group $\mathbb Z_2$.

\paragraph{Models.}
A \emph{model} 
%$\mathcal{M}$ 
assigns to each sort $s \in \SortSet$ a
carrier set $\semSort s$, and assigns to each operation $\texttt{f}
\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) =
([s_1,\ldots,s_n],s)$, a function $\semIndexExp{\texttt{f}} :
\semSort{s_1}\times\cdots\times\semSort{s_n} \to \semSort{s}$.

An index context $\Delta = (i_1\mathord:s_1,\ldots,i_n\mathord:s_n)$ is
interpreted as cartesian product, i.e., $\semSort{\Delta} \isDefinedAs \semSort{s_1}
\times\cdots \times \semSort{s_n}$. 
For each well-sorted index expression $\Delta \vdash
e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
\semSort{s}$ by recursion on the structure of~$e$:
\begin{align*}
\semIndexExp{i} \rho & \isDefinedAs \rho(i) &
\semIndexExp{\indexOp{f}(e_1, \ldots, e_n)} \rho & \isDefinedAs \semIndexExp{\indexOp{f}}(
\semIndexExp{e_1}\rho,\ldots,\semIndexExp{e_n}\rho)
\end{align*}

Finally, a model
% $\mathcal{M}$
must be sound, that is, for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$.

\begin{example*}[Geometry: affine model]
We define the model
% $\mathcal{M}$ 
of the indexing theory for the two-dimensional
geometry example as follows. Each of the sorts is interpreted just as
its semantic counterpart: 
%$\semSort{\SynTransl{2}}M = \Transl{2}$, $
%\semSort{\SynGL{2}}M = \GL{2}$, $ \semSort{\SynOrth{2}}M = \Orth{2}$,
%and $ \semSort{\SynGL{1}}M = \GL{1}$.
\begin{align*}
\semSort{\SynTransl{2}} & = \Transl{2}
&\semSort{\SynGL{2}} & = \GL{2} &
\semSort{\SynOrth{2}} & = \Orth{2}
&\semSort{\SynGL{1}} & = \GL{1}
\end{align*}
Each of the index operations (e.g.,~the group structure and
determinant) is interpreted by the intended semantic operation, and
clearly satisfies the axioms in \exref{ex:two-dim-geo-axioms}. 
\qed
\end{example*}

Given $\rho \in \semSort{\Delta}$ and a substitution
$\Delta \vdash \sigma \Rightarrow \Delta'$ with $\sigma = (e_1,\ldots,e_n)$, 
we can derive the composed
index environment $\rho \circ \sigma
\in \semSort{\Delta'}$ as $\rho \circ \sigma \isDefinedAs (\semIndexExp{e_1}\rho,\ldots,\semIndexExp{e_n}\rho)$.

%A
%\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
%\semIndexExp{\cdot}_{\mathcal{M}})$ satisfying all the axioms in
%$\IndexAxiomSet$.

\paragraph{Relational interpretation of primitive types.}
Having fixed a model
% $\mathcal{M}$ 
we next choose a relational interpretation 
%$\mathcal{R}$ 
of
primitive types: for each each primitive type
$\tyPrimNm{X}$, its relational interpretation is parameterised by elements
from the model: $\semPrimType{\tyPrimNm{X}} :
\semSort{\primTyArity(\tyPrimNm{X})} \to \Rel(\tyPrimSem{\tyPrimNm{X}})$.

\begin{example*}[Geometry: change of basis interpretation]
Given the model of geometric groups described above,
for the relational interpretation  of $\tyPrim{vec}{B,t}$ and
$\tyPrim{real}{s}$, we use $\semPrimType{\tyPrimNm{vec}}(B,\vec{t}) = \{
(\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \}$ and
$\semPrimType{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}$.
\qed
\end{example*}

\paragraph{Relational interpretation of types.}
We assign a relational interpretation to all well-indexed types $\Delta
\vdash A \isType$ by induction on their derivations, parameterised by
index environments 
$\rho \in \semSort{\Delta}$:
\begin{eqnarray*}
%  \rsem{\Delta \vdash A \isType}M R & : & \semSort\Delta  \to \Rel(\tyPrimSem{A})\\
  \rsem{\tyUnit} \rho & \isDefinedAs & \{(*,*)\} \\
  \rsem{\tyPrim{X}{e_1,\ldots,e_n}}\rho & \isDefinedAs & 
  \semPrimType{\tyPrimNm{X}} (\semIndexExp{e_1}\rho, \ldots,
  \semIndexExp{e_n}\rho)\\
  \rsem{A \tyArr B} \rho & \isDefinedAs & \rsem{A}\rho \relArrow \rsem{B} \rho \\
  \rsem{A \tyProduct B}\rho & \isDefinedAs & \rsem{A} \rho \relTimes \rsem{B} \rho \\
  \rsem{A + B} \rho & \isDefinedAs & \rsem{A} \rho \relSum \rsem{B} \rho \\
  \rsem{\forall i\mathord:s.A} \rho & \isDefinedAs & 
  \bigcap\{ \rsem{A} {(\rho,m)} \sepbar m \in \semSort{s} \} \\
  \rsem{\exists i\mathord:s.A} \rho & \isDefinedAs & 
  \bigcup\{ \rsem{A} {(\rho,m)} \sepbar m \in \semSort{s} \}
\end{eqnarray*}
In this definition, the relational interpretation of an application of
a primitive type $\rsem{\tyPrim{X}{e_1,\ldots,e_n}}\rho$ is built from
the relational interpretation of the primitive type,
$\rsem{\tyPrimNm{X}}$, and the interpretation of the index terms
$e_1,...,e_n$ in the index environment $\rho$. Universal and
existential quantification are interpreted by the set-theoretic
intersection and union, respectively, over all extensions of the index
environment.

We have also used the following standard %three
constructions on binary relations: if $R \in \Rel(X)$ and $S \in
\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is %defined as
$\{(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
\}$, and %. With the same assumptions on $R$ and $S$, the relation
$R \relTimes S \in \Rel(X \times Y)$ is %defined as
$\{((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
\}$, and %. Finally, the relation
$R \relSum S \in \Rel(X + Y)$ is %defined as
$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.

The following lemma states that the relational interpretation of types
that we have defined in this section behaves well: the first part of
the lemma states that two types that are judgmentally equal are given
equal relational interpretations, and the second part states that
substitution of index expressions in types can be interpreted by the
composition of index environments with 
substitutions.
\begin{lemma}\label{lem:tyeqsubst-relational}
\par
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$, then $\rsem{A} =  \rsem{B}$;
  \item If $\Delta' \vdash A \isType$ then for all $\Delta \vdash \sigma \Rightarrow
    \Delta'$ and $\rho \in \semSort{\Delta}$,
    $\rsem{\sigma^*A} \rho = \rsem{A}(\rho \circ \sigma)$.
  \end{enumerate}
\end{lemma}
\noindent
Note that the equations in both parts of
\lemref{lem:tyeqsubst-relational} are well-typed by virtue of the
corresponding parts of \lemref{lem:tyeqsubst-erasure}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The Abstraction Theorem and Semantic Equivalence}
\label{sec:abstraction-theorem}

Our main result
(\thmref{thm:abstraction}) is that the index-erasure semantics of
every well-typed term is related to itself in the relational
interpretation of its type: this is the Abstraction Theorem for every
instantiation of our general framework.

\paragraph{The Abstraction Theorem.}
%\label{sec:abstraction-theorem}

We now state the Abstraction Theorem for well-typed terms. To state
and prove this theorem for open terms, we extend the
relational interpretation of types to typing contexts. The relational
interpretation of contexts is defined by: % the following clauses: %, where
%we have used the relational lifting of cartesian product $-\relTimes-$
%defined in \autoref{sec:relational-semantics}.
\begin{displaymath}
  \begin{array}{@{}l@{\hspace{0.1em}\isDefinedAs\hspace{0.1em}}l@{\hspace{1.5em}}l@{\hspace{0.1em}\isDefinedAs\hspace{0.1em}}l}
    \rsem{\epsilon} {\rho} & \{(*,*)\} &
    \rsem{\Gamma, x : A} {\rho} & \rsem{\Gamma} \rho \relTimes \rsem{A} \rho
  \end{array}
\end{displaymath}

The relational interpretation of contexts inherits from the relational
interpretation of types the property of interpreting the application
of substitutions as composition:
\begin{lemma}\label{lem:ctxtsubst-rel}
  If $\Delta' \vdash \Gamma \isCtxt$ and $\Delta \vdash \sigma
  \Rightarrow \Delta'$, then for all $\rho \in \semSort{\Delta}$,
we have 
  $\rsem{\sigma^*\Gamma} \rho =
  \rsem{\Gamma}(\rho \circ \sigma)$.
\end{lemma}

Given a particular choice of model and relational interpretation of
primitive types, we can then prove the following.
\begin{theorem}[Abstraction]\label{thm:abstraction}
  If $\Delta; \Gamma \vdash M :
  A$, then for all $\rho \in \semSort{\Delta}$ and $\eta_1, \eta_2
  \in \ctxtSem{\Gamma}$ such that $(\eta_1, \eta_2) \in
  \rsem{\Gamma} \rho$, we have $(\tmSem{M}\eta_1,
  \tmSem{M}\eta_2) \in \rsem{A} \rho$.
\end{theorem}
\begin{proof}
By induction on the typing derivation, making use of
Lemma~\ref{lem:tyeqsubst-relational} (part 1) for rule \TirName{TyEq},
Lemma~\ref{lem:tyeqsubst-relational} (part 2) for rules \TirName{UnivApp} and \TirName{ExPack},
and Lemma~\ref{lem:ctxtsubst-rel} for rules \TirName{UnivAbs} and \TirName{ExUnpack}. 
The details can be found in the Coq development.
\end{proof}

Semantic equivalence of terms is defined in terms of the relational
interpretation. As a consequence of \thmref{thm:abstraction}, semantic
equivalence is a sound approximation of contextual equivalence.

  Let $\opsCtxt$ be the context of primitive operations and $\opsEnv$  its interpretation 
%as in
%  \defref{defn:ctxt-equiv}.
Fix a model and relational interpretation of primitive types so that
  $(\opsEnv, \opsEnv) \in  \rsem{\opsCtxt} *$.

\begin{definition}[Semantic equivalence]\label{def:semantic-equality}
  Two terms $\Delta; \opsCtxt, \Gamma \vdash M_1 , M_2 : A$
  are
  \emph{semantically equal},
  written $\semeq \Delta\Gamma{M_1}{M_2} A$, if for all $\rho \in
  \semSort{\Delta}$, and all $(\eta_1,\eta_2) \in
  \rsem{\Gamma} \rho$, we have
  $(\tmSem{M_1}(\opsEnv,
  \eta_1),\tmSem{M_2}(\opsEnv, \eta_2)) \in
  \rsem{A}\rho$.
\end{definition}

\begin{theorem}[Soundness]\label{thm:soundness}
If $\semeq \Delta\Gamma {M_1}{M_2} A$ then
$\ctxteq \Delta \Gamma {M_1}{M_2}A$.
\end{theorem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:
%%%---------- close: section-general-framework.tex
%%%%%%%%%%% Springe nach section-instantiations.tex
%%%---------- open: section-instantiations.tex
\section{Geometric Consequences of Abstraction}
\label{sec:instantiations}
We now instantiate our general framework %of the previous section 
with
the indexing theory of \autoref{sec:motivating-examples}, and present
more general and formally-justified free theorems, type isomorphisms,
and non-definability results.  For the free theorems and isomorphisms,
we use the model and relational interpretation of primitive types
described in \autoref{sec:relational-semantics}, namely that of
affine transformations for vectors and scaling for scalars.

%\begin{example}[Two-dimensional Geometry]

\subsection{Free Theorems}
\label{sec:theorems-for-free}

Consider the type of the triangle area %of a triangle 
function from
\exref{ex:area-of-triangle-1}:
\begin{displaymath}
  \begin{array}{@{}l}
    \mathrm{area} : \forall B\mathord:\SynGL{2}, t\mathord:\SynTransl{2}.\ \\
    \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{\abs{\det B}}
  \end{array}
\end{displaymath}
By \thmref{thm:abstraction}, we can derive the following free theorem. For all $B \in \GL{2}$, $\vec{t} \in \Transl{2}$, and $\vec{x}, \vec{y}, \vec{z} \in \mathbb{R}^2$, we have 
\begin{displaymath}
  \abs{\det B}(\tmSem{\mathrm{area}}\ \vec{x}\ \vec{y}\ \vec{z}) = \tmSem{\mathrm{area}}\ (B\vec{x} + \vec{t})\ (B\vec{y} + \vec{t})\ (B\vec{z} + \vec{t})
\end{displaymath}
Thus, directly from the type of the $\mathrm{area}$ function, we can
see that its index-erasure semantics is (a) invariant under
translations, and (b) if the inputs are subjected to a linear
transformation $B$, the output varies with the absolute value of the
determinant of $B$.

% Wadler's influential ``Theorems for Free!''  \cite{wadler89theorems}
% emphasised a particular aspect of Reynolds' theory of relational
% parametricity: the fact that we can read off theorems about a program
% simply by looking at the relational interpretation of its type. We now
% state some free theorems that are derivable in each of the three main
% examples we introduced in \autoref{sec:motivating-examples}.

\subsection{Type Isomorphisms}
\label{sec:types-indexed-abelian-groups}
Types indexed by abelian groups induce a particularly rich theory of
type isomorphisms; previous work on
units of measure~\cite{kennedy97relational} relates these to
Buckingham's theorem from dimensional analysis. Here we consider the
additive abelian groups of translations and the multiplicative abelian
group of scalings.

\paragraph{Translations} 
Consider first the group $\Transl{2}$ of translations from
\autoref{sec:type-system-geom-intro} and 
\autoref{sec:affine-vector-ops}. 
\newcommand{\DeltaWithB}{\Delta,B\mathord:\SynGL{2}}

\begin{example}[Geometry: wlog]\label{ex:simpletrans}
We prove that
\[\DeltaWithB\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to\tyPrim{vec}{B,t}) \cong
\tyPrim{vec}{B,0}.
\]
Let 
\begin{align*}
X & \isDefinedAs\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to\tyPrim{vec}{B,t}
\quad\text{and}\quad Y \isDefinedAs\tyPrim{vec}{B,0}\\
I & \isDefinedAs\lambda f\mathord:X.f\,[0]\,(0\,[B])\\
J & \isDefinedAs\lambda v\mathord:Y.\Lambda t\mathord:\SynTransl{2}.\lambda w\mathord:\tyPrim{vec}{B,t}.v +[B,0,t] w
\end{align*}
Unfolding definitions gives that $\tmSem{I(J(v))}(\opsEnv,\vec{v})$ = $\vec{v}+\vec{0}$
for any $\vec{v}\in\mathbb R ^ 2$.  Because $\vec{0}$ is the identity for vector
addition it follows that $\ctxteq{\DeltaWithB}{v\mathord:Y}{I(J(v))}{v}Y$.
To show $\ctxteq{\DeltaWithB}{f\mathord:X}{J(I(f))}{f}{X}$ we appeal to
\autoref{thm:soundness} and reason using the relational semantics.  It
suffices to show $\semeq{\DeltaWithB}{f\mathord:X}{J(I(f))}{f}{X}$; that
is, given $\rho\in\semSort\DeltaWithB$, for any $f$ and $f'$ with
$(f,f')\in\rsem X\rho$, it is the case that 
$(\tmSem{J(I(f))}(\opsEnv,f), f')\in\rsem X\rho$.
Expanding the premise, we have
\begin{eqnarray}\label{eqn:transpremise}
\forall \vec{t}\in \mathbb R ^ 2,
\forall \vec{v}\in \mathbb R ^ 2,
f'(\rho(B)\vec{v} + \vec{t}) = \rho(B)(f(\vec{v})) + \vec{t}
\end{eqnarray}
Expanding the conclusion, we have to show that for any
$\vec{t_0}\in\mathbb R^2$ and $\vec{v_0}\in\mathbb R^2$, it is the case that
$f'(\rho(B)\vec{v_0} + \vec{t_0}) = \rho(B)(\vec{v_0} + f(0)) + \vec{t_0}$.
By instantiating \autoref{eqn:transpremise} with
$\vec{v}=\vec{0}$ and~$\vec{t}=\rho(B)\vec{v_0}+\vec{t_0}$ and applying a little algebra we obtain this result.
\qed
\end{example}

We can generalise this isomorphism substantially, proving
that for any quantifier-free type $A$, 
the type $\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to A$
is isomorphic to $A[0/t]$, where $A[0/t]$ 
denotes substitution of the identity translation $0$
for index variable $t$ in type $A$. This isomorphism formalises the
equivalence between \emph{coordinate-free} geometry, expressed by a
polymorphic type whose first argument can be thought of as the
\emph{origin} with respect to which $A$ is specified, and a
\emph{coordinate-based} geometry, expressed by the type $A[0/t]$, in
which the origin is fixed at $(0,0)$. It is sometimes said that an
affine space is a vector space that has forgotten its origin; we have
captured this in a type isomorphism.

\newcommand{\transup}[3]{\mathord\uparrow_{#1}^{#2\mathord:#3}}
\newcommand{\transdn}[3]{\mathord\downarrow_{#1}^{#2\mathord:#3}}
\newcommand{\transtrans}[3]{\mathrm{tr}_{#1}^{#2\mathord:#3}}

\begin{lemma}\label{lem:transiso}
Suppose that $\DeltaWithB,t\mathord:\SynTransl{2}\vdash A \isType$, 
$A$ contains no quantifiers and every occurrence of $\tyPrimNm{vec}$ is of the form $\tyPrim{vec}{B,?}$. Then
\[
\DeltaWithB\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to A)
\cong A[0/t]
\]
\end{lemma}
\begin{proof}
Let 
\begin{align*}
X & \isDefinedAs \forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to A
\quad\text{and}\quad Y \isDefinedAs A[0/t]\\
I & \isDefinedAs \lambda f\mathord: X. f\ [0]\ (0\ [B])\\
J & \isDefinedAs \lambda y\mathord: Y. \Lambda t.\lambda v\mathord:\tyPrim{vec}{B,t}. \transup A t v (y)
\end{align*}
where terms 
\begin{align*}
&\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transup A t v : A[0/t] \to A\\
&\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transdn A t v : A\to A[0/t]\\
&\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transup e t v : \tyPrim{vec}{B,e[0/t]}\to\tyPrim{vec}{B,e}\\
&\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transdn e t v : \tyPrim{vec}{B,e}\to\tyPrim{vec}{B,e[0/t]}
\end{align*}
are defined by induction over the structure of $A$ and $e$. For brevity we omit types on binders;
also note that the omitted cases for $\transdn A t v$ 
have definition symmetric to $\transup A t v$:
\begin{align*}
\transup{\tyPrimNm{unit}} t v &= \lambda x. x
\qquad 
\transup{\tyPrim{real}{e}} t v = \lambda x. x
\\
\transup{A_1\times A_2}t v &= \lambda p. (\transup{A_1}t v(\pi_1 p),\transup{A_2}t v(\pi_2 p))
\\
\transup{A_1 + A_2} t v &= \lambda x.
\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transup {A_1}t v y); \textrm{inr}\ y.\textrm{inr}(\transup {A_2} t v y)
\\ 
\transup{A_1\to A_2} t v &= \lambda f. \lambda x. \transup{A_2} t v (f(\transdn {A_1} t v x))
\end{align*}
\begin{align*}
\transup{\tyPrim{vec}{B,e}} t v &= \lambda x. \transdn{e}  t v (x) 
&
\transdn{\tyPrim{vec}{B,e}} t v &= \lambda x. \transup{e} t v (x)
\\
\transup{0} t v &= \lambda w.w
&
\transdn{0} t v &= \lambda w.w
\\ 
\transup{e_1+e_2} t v &= \lambda w.\transup{e_1} t v(\transup{e_2} t v(w))
&
\transdn{e_1+e_2} t v &= \lambda w.\transdn{e_1} t v(\transdn{e_2} t v(w))
\\
\transup{-e} t v &= \lambda w.\transdn{e} t v (w)
&
\transdn{-e} t v &= \lambda w.\transup{e} t v (w)
\\
\transup{t} t v &= \lambda w.w + v 
&
\transdn{t} t v &= \lambda w.w + (\mathrm{negate}\, v)
\\
\transup{t'} t v &= \lambda w.w\quad(t'\neq t)
&
\transdn{t'} t v &= \lambda w.w\quad(t'\neq t)
\end{align*}

Unfolding definitions, we deduce that
\[
\tmSem{I(J(y))}(\opsEnv,y) = \tmSem{\transup A t v}(\opsEnv,\vec 0)(y)
\]
for any $y \in \tmSem{A[0/t]}$. It's easy to prove by induction on $A$
that $\tmSem{\transup A t v}(\opsEnv,\vec 0)(y) = y$, and so
$\ctxteq\DeltaWithB{y\mathord:Y}{I(J(y))}y Y$ follows.

For the opposite direction of the isomorphism, we appeal to 
\autoref{thm:soundness} and so it suffices to show
$\semeq{\DeltaWithB}{f\mathord:X}{J(I(f))}{f}{X}$; that
is, given $\rho\in\semSort\DeltaWithB$, for any $f$ and $f'$ with
$(f,f')\in\rsem X\rho$, it is the case that 
$(\tmSem{J(I(f))}(\opsEnv,f), f')\in\rsem X\rho$.
Expanding the premise, we have
\begin{eqnarray}\label{eqn:isopremise}
\forall \vec{t}\in \mathbb R ^ 2,
\forall \vec{v}\in \mathbb R ^ 2,
(f(\vec{v}),f'(\rho(B)\vec{v} + \vec{t}))\in\rsem A (\rho,\vec{t})
\end{eqnarray}
Expanding the conclusion, we have to show for any
$\vec{t_0}\in\mathbb R^2$ and $\vec{v_0}\in\mathbb R^2$, it is the case that
$(\tmSem{J(I(f))}(\opsEnv,f)(\vec{v_0}),f'(\rho(B)\vec{v_0} + \vec{t_0}))\in\rsem A (\rho,\vec{t_0})$.
Unfolding the definitions of $J$ and $I$ and expanding the erasure semantics, 
this amounts to showing that
\[
\forall \vec{t_0},\vec{v_0}\in \mathbb R ^ 2,
(\tmSem{\transup A t v}(\opsEnv,\vec{v_0})(f\ \vec 0),f'(\rho(B)\vec{v_0} + \vec{t_0}))\in\rsem A (\rho,\vec{t_0}).
\]
In order to make progress we prove first by induction on $A$ that for any $\vec v$ and $\vec w$, 
the meaning of $\transup A t v$ can be characterized by
\begin{align*}
(\tmSem{\transup A t v}(\opsEnv,\vec{v})(x),y)&\in\rsem A (\rho,\vec{w})
\\\Leftrightarrow 
(x,y)&\in\rsem A (\rho,\rho(B)\vec{v} + \vec{w})
\end{align*}
Thus our goal becomes
\[
\forall \vec{t_0},\vec{v_0}\in \mathbb R ^ 2,
(f\ \vec 0,f'(\rho(B)\vec{v_0} + \vec{t_0}))\in\rsem A (\rho,\rho(B)\vec{v_0} + \vec{t_0}).
\]
This is obtained immediately by instantiating \autoref{eqn:isopremise}
with $\vec{v}=\vec 0$ and $\vec{t}=\rho(B)\vec{v_0} + \vec{t_0}$.
\end{proof}

\exref{ex:type-iso} and \exref{ex:simpletrans} are special cases of
this isomorphism. Another instance is the type of vector addition:
\begin{align*}
\forall B\mathord:\SynGL{2}, t_1, t_2\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_2}\to\tyPrim{vec}{B,t_1\mathord+t_2}\\
\cong\quad\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_1}
\quad (\text{by~\lemref{lem:transiso}})\\
\cong\quad\forall B\mathord:\SynGL{2}.\tyPrim{vec}{B,0} \quad (\text{by~\lemref{lem:transiso}})\\
\cong\quad\tyUnit\quad(\text{by \autoref{thm:abstraction}})
\end{align*}
In other words, the vector addition operation is the only inhabitant of its type!
\paragraph{Scalings}
For the group $\GL{1}$ of scalings we can treat
a real-valued argument to a function as a unit of measure with
which to scale the result. 
%The small fly in the ointment is that the
Although the argument might be zero, and this cannot be used for scaling,
%Nevertheless, 
we still can obtain the following slightly more complicated
isomorphism:
\begin{eqnarray}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A &
\cong & A[1/s]\times \forall s\mathord:\SynGL{1}.A\label{eqn:scaleiso}
\end{eqnarray}
The type of $\mathrm{areaCircle}$ from the introduction is one instance:
\begin{align*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s^2}\\ 
\cong\quad\tyPrim{real}{1}\times(\forall
s\mathord:\SynGL{1}.\tyPrim{real}{s^2}) \quad(\text{by \autoref{eqn:scaleiso}})\\ 
\cong\quad\tyPrim{real}{1}\times \tyUnit \quad (\text{by \autoref{thm:abstraction}})\\
\cong\quad\tyPrim{real}{1}\quad (\text{trivially})
\end{align*}
%where %the isomorphism 
%The abstraction theorem is used to prove
%$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s^2}\cong \tyUnit$.
% is proved using
%the abstraction theorem.
%\begin{enumerate}
%\item Two cases: when we have a total ``division'' operation, and when
%  we don't.
%\item Type isomorphisms, based on Smith Normal Form
% \item Non-definability results, based on the non-compositional
%   relational environments
% \item Alteration of the non-definability results in the light of
%   addition of a square root operation
% \item Mention the variation we can obtain by changing some types,
%   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
%\end{enumerate}

\subsection{Non-definability}
\label{sec:types-indexed-abelian-groups-indef}

\newcommand{\Grp}{\mathit{Gr}}
To prove non-definability results, the model and relational interpretation
used in the previous two sections are not sufficient.
This is true even for simple scalars with invariance under scaling.
Consider the type 
$\forall s \mathord: \SynGL{1}.\tyPrim{real}{s^2} \to \tyPrim{real}{s}$. 
There are many functions~$f: \mathbb R \to \mathbb R$
that satisfy its relational interpretation, i.e., for which $f(k^2\cdot x) = k\cdot f(x)$ for 
any $k\neq 0$.
Consider 
\[
f(x) = \begin{cases}\sqrt x & \text{if }x > 0\\0&\text{otherwise}\end{cases}
\]
for instance. Yet this type contains only the constant zero function, a fact that
we can prove using a surprisingly simple model!

\begin{example}\label{ex:sqrt-root-nondef}
If  $\opsCtxt \vdash f : \forall s \mathord: \SynGL{1}.\
\tyPrim{real}{s^2} \to \tyPrim{real}{s}$ 
then $\tmSem f = \lambda x.0$.
To show this, take the model
\[
\begin{array}{l}
\semSort{\SynGL{1}} \isDefinedAs \{0,1\}\qquad
\semIndexExp{1} \isDefinedAs 0 \qquad
\semIndexExp{{}^{-1}}  \isDefinedAs \mathrm{id}\\
\semIndexExp{\cdot} \isDefinedAs \lambda(x,y).(x+y)\mathrel{\mathrm{mod}} 2
\qquad
\semIndexExp{|-|} \isDefinedAs \mathrm{id},

\end{array}
\]
in other words, the additive abelian group $(\mathbb Z_2, +)$. 
Now set
\[
\semPrimType{\tyPrimNm{real}}(z) = 
  \begin{cases} \Delta_{\mathbb R} & \text{if }z=0\\
                      \{(0,0)\} & \text{if }z=1
\end{cases}
\]
It's easy to check that $(\opsEnv,\opsEnv)\in\rsem{\opsCtxt}*$.

We think of $z$ as tracking whether or not exponents on scale
parameters are divisible by 2: all the primitive operations in
\autoref{fig:real-ops} produce results with even exponents if their
inputs have even exponents. But the type under consideration here does
not.

By \autoref{thm:abstraction} we have that $(f,f)\in\rsem{\forall s
  \mathord: \SynGL{1}.\ \tyPrim{real}{s^2} \to \tyPrim{real}{s}} *$.
Expanding the definitions, this means that for any
$z\in\semSort{\SynGL{1}}$ and any
$(x,y)\in\semPrimType{\tyPrimNm{real}}((z+z)\mathrel\textrm{mod}2)$, it is the case that
$(f(x),f(y))\in\semPrimType{\tyPrimNm{real}}(z)$. Set $z = 1$.  
Then we have that if $x=y$ then $f(x)=f(y)=0$.
\qed
\end{example}

The particular model and relational interpretation used here is devised only to
rule out the `square root type' (it does not rule out a `cube root type', for instance); 
nevertheless, it is possible to devise craftier interpretations that serve up both invariance
and more general non-definability results. For example, we can set
\begin{align*}
\semSort{\SynGL{1}} & \isDefinedAs \mathbb \mathbb \GL{1} \times \mathbb Q &
\semIndexExp{\cdot} & \isDefinedAs (\cdot,+) & 
\semIndexExp{1} & \isDefinedAs (1,0) \\
\semIndexExp{{}^{-1}} & \isDefinedAs \mathrm{({}^{-1},-)} &
\semIndexExp{|\cdot|} & \isDefinedAs \mathrm{(|\cdot|,\mathrm{id})}
\end{align*}
and
\[
\semPrimType{\tyPrimNm{real}}(k,q) = 
  {\begin{cases} \{(x,k x) \sepbar x \in \mathbb R \} & \text{if }q \in \mathbb Z\\
                      \{(0,0)\} & \text{otherwise}
  \end{cases}}
\]

\begin{example}\label{ex:cube-root-nondef}
Suppose we extend the operations $\opsCtxt$ of \autoref{fig:real-ops}
and corresponding $\opsEnv$ with a square root operation typed by
\[
\mathrm{sqrt} : \forall s\mathord: \SynGL{1}.\tyPrim{real}{s^2} \to \tyPrim{real}{s}.
\]
If  $\opsCtxt \vdash f : \forall s \mathord: \SynGL{1}.\
\tyPrim{real}{s^3} \to \tyPrim{real}{s}$ 
then $\tmSem f = \lambda x.0$.
To show this, take the model
\begin{align*}
\semSort{\SynGL{1}} & \isDefinedAs \mathbb Q &
\semIndexExp{\cdot} & \isDefinedAs + & 
\semIndexExp{1} & \isDefinedAs 0 &
\semIndexExp{{}^{-1}} & \isDefinedAs \mathrm{-} &
\semIndexExp{|\cdot|} & \isDefinedAs \mathrm{id},
\end{align*}
in other words, the additive abelian group $(\mathbb Q, +)$. 
Now set
\[
\semPrimType{\tyPrimNm{real}}(q) =
\begin{cases} \Delta_{\mathbb R} & \text{if }\exists n, 2^n q \in\mathbb Z\\
\{(0,0)\} & \text{otherwise}
\end{cases}
\]
It's easy to check that $(\opsEnv,\opsEnv)\in\rsem{\opsCtxt}*$.

Here, $q$ is tracking the exponents on scale parameters, with
$\semPrimType{\tyPrimNm{real}}(q)$ relating only zero to itself
unless the denominator of $q$ is a power of two (a so-called
\emph{dyadic} number).  
Applying the Abstraction Theorem and expanding, we have that for
any
$q\in\mathbb Q$ and any
$(x,y)\in\semPrimType{\tyPrimNm{real}}(3q)$, it is the case that
$(f(x),f(y))\in\semPrimType{\tyPrimNm{real}}(q)$. Set $q = \frac 1 3$.  
Then we have that if $x=y$ then $f(x)=f(y)=0$.
\qed
\end{example}

% For proving non-definability results, we consider the general case of
% a sort $\mathsf{G}$ with the equational theory of abelian groups. We
% assume a group-indexed primitive type $\tyPrimNm{G}$ and set of
% primitive operations typed by $\Gamma_\Grp$ that include the group
% identity $1:\tyPrim{G}{1}$, operation $(\cdot): \forall
% a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to
% \tyPrim{G}{a\cdot b}$ and inverse $\ ^{-1}: \forall
% a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$. The
% context $\Gamma_\Grp$ may also include members with types of the form
% $\forall \vec{a}\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to ...
% \tyPrim{G}{e_n} \to \tyPrim{G}{e}$ as long as the result index $e$ is
% generated by the set $\{e_1,...,e_n\}$. One instance is the
% sort~$\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ on
% $\tyPrim{vec}{t}$. For the purposes of non-definability results, we
% may choose an arbitrary index-erasure semantics with $\eta_\Grp \in
% \ctxtSem{\Gamma_\Grp}$.

% We use the construction of definability relational environments
% from \autoref{sec:constr-rel-env}, with $R^\bullet_{\tyPrimNm{G}} =
% \emptyset$, to give a pre-substitutive family $\relEnv{E}_\Grp$. This
% gives us the following lemma and theorem.

% \begin{lemma}\label{lem:geom-environments-2}
%   For all $\Delta$ and $\rho \in \relEnv{E}_\Grp(\Delta)$,
%   $(\eta_\Grp,\eta_\Grp) \in \rsem{\Gamma_\Grp}{\relEnv{E}_\Grp}\rho$.
% \end{lemma}

% For index-erasure interpretation, we assume a (not necessarily
% abelian) group $(G, 1, \cdot, \ ^{-1})$. Let $Z(G)$ be the centre of
% the $G$ %: i.e.,~the set of elements that commute with every other
% %element in $G$. We set 
% and $\tyPrimSem{G} = G$. For each index context $\Delta$, the set
% $\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be
% $\{ \rho_{(H,h)} \sepbar H\textrm{ is a subgroup of }(\Delta
% \Rightarrow \mathsf{G})/\equiv, h\textrm{ is a group homomorphism }H
% \to Z(G) \}$, where
% \begin{displaymath}
%   \rho_{(H,h)}\ \tyPrimNm{G}\ (e) = \left\{
%     \begin{array}{ll}
%       \{ \} & \textrm{if}\ e\not\in H \\
%       \{ (x, h(e)\oplus x) \sepbar x \in G \} & \textrm{if}\ e\in H
%     \end{array}
%   \right.
% \end{displaymath}

% Writing $i^n$ for $n$-fold iteration of the group operation, we have:
% \begin{theorem}
%   Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matrices of
%   natural numbers $A$ and $m$-by-$1$ matrices of integers $\vec{b}$,
%   there exists a quantifier-free program $M$ such that:
%   \begin{displaymath}
%     i_1,\dots,i_m\mathord:\mathsf{G}; \Gamma_\Grp \vdash M : \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
%   \end{displaymath}
%   where $e_j = i_1^{A_{j1}}\cdots i_m^{A_{jm}}$ and $e = i_1^{b_1}\cdots i_m^{b_m}$, iff there is a solution $\vec{x}$
%   in the integers to the equation $A \vec x = \vec{b}$.
% \end{theorem}
% The uninhabited type $\forall t \mathord: \SynTransl{2}.\
% \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ from
% \exref{ex:uninhabited-type} is an instance of this theorem.  
%\begin{proof}
%  (If) The program $\Lambda i_1\dots i_m.\ \lambda g_1\dots g_n.\
%  g_1^{x_1}\dots g_n^{x_n}$ satisfies the typing judgment
%  (\ref{eq:fo-group-type}).

%  (Only if) Assume that there is a program $M$ satisfying the typing
%  judgment (\ref{eq:fo-group-type}). By \thmref{thm:abstraction} we
%  know that for all relational environments $\rho \in
%  \relEnv{E}_\Grp(i_1,\dots,i_m)$, and all $g_j,g'_j \in G$,
%  \begin{displaymath}
%    \begin{array}{l}
%      (g_1,g'_1) \in \rho\ \tyPrimNm{G}\ (e_1) \land \dots \land (g_n,g'_n) \in \rho\ \tyPrimNm{G}\ (e_n) \Rightarrow \\
%      \quad (\tmSem{M}g_1\dots g_n, \tmSem{M}g'_1\dots g'_n) \in \rho\ \tyPrimNm{G}\ (e)
%    \end{array}
%  \end{displaymath}
%  We select the relational environments
%  \begin{displaymath}
%    \rho\ \tyPrimNm{G}\ (e) = \left\{
%      \begin{array}{ll}
%        \{(g,g) \sepbar g \in \mathcal{G} \} &
%        \begin{array}[t]{@{}l}
%          \textrm{if}\ e = e_1^{x_1}\dots e_n^{x_n} \\
%          \textrm{ for some }x_1,\dots,x_n \in \mathbb{Z}
%        \end{array}
%        \\
%        \{\} & \textrm{otherwise}
%      \end{array}
%    \right.
%  \end{displaymath}
%  Now, for any $g_1,\dots,g_m \in \mathcal{G}$, we know that, for all
%  $j$, $(g_j,g_j) \in \rho\ \tyPrimNm{G}\ (e_j)$, so we know that
%  $(\tySem{M}g_1\dots g_n, \tySem{M}g_1 \dots g_n) \in \rho\
%  \tyPrimNm{G}\ (e)$. Thus there exist integers $x_1,...,x_n$ such
%  that $e = e_1^{x_1}\dots e_n^{x_n}$. But we also know that $e =
%  i_1^{b_1}...i_m^{b_m}$. By the cancellation property of free groups,
%  we learn that $x_1,...,x_n$ is the solution we need.
%\end{proof}

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:
%%%---------- close: section-instantiations.tex
%%%%%%%%%%% Springe nach section-general-nondef.tex
%%%---------- open: section-general-nondef.tex
\section{Non-definability for the General Framework}
\label{sec:general-nondef}

The non-definability results in the previous section required the use
of specially crafted models and relational interpretations. It is
reasonable to ask whether or not there is a general method for
constructing suitable models and relational interpretations to prove
non-definability results. In this section, we show that this is
possible for a large class of first-order types in any instance of our
general framework.

In \exref{ex:sqrt-root-nondef}, we showed that the type $\forall s
\mathord: \SynGL{1}.~\tyPrim{real}{s^2} \to \tyPrim{real}{s}$ only has
trivial inhabitants. Intuitively, this is because the index of the
result type ($s$) cannot be obtained from the index of the input type
($s^2$) using the abelian group operations and axioms. This
observation can be used to give a sufficient condition for
non-inhabitation for types of the form:
\begin{displaymath}
  \forall i_1\mathord:s_1,\ldots,i_m\mathord:s_m.~\tyPrim{X}{e_1} \to \ldots \to \tyPrim{X}{e_n} \to \tyPrim{X}{e}
\end{displaymath}
Roughly speaking, if this type is inhabited, then it must be the case
that the index expression $e$ can be generated from the set of index
expressions $\{e_1,...,e_n\}$.

We assume that we are working with an instantiation of the general
framework from \autoref{sec:a-general-framework} with a closed typing
context $\Gamma_{\mathit{ops}}$ describing the types of the primitive
operations and a chosen index-erasure semantics. We will use a special
relational interpretation built from the syntax of the indexing
expressions to prove our general non-definability result. For
simplicity, we assume that there is only one primitive type,
$\tyPrimNm{X}$, but the technique we describe here extends to the
general case. We also assume that $\tySem{\tyPrimNm{X}}$ is non-empty.

To state our general non-definability condition, we need to define the
set of index expressions generated by some finite set of index
expressions. Given a set $S$ of index expressions that are all
well-indexed in some index context $\Delta$, we define
$\mathrm{Gen}_0(S)$ to be the set of expressions that are built from
the elements of $S$ and the primitive index operations. To account for
the equations between indexing terms, we close under index expression
equivalence to get the set $\Gen(S) = \{ e \sepbar \exists s.\ \exists
e' \in \mathrm{Gen}_0(S).\ \Delta \vdash e \equiv e' : s \}$. Now a
type is \emph{well-generated} if it is closed and of the form:
\begin{displaymath}
  \forall i_1\mathord:s_1,\ldots,i_m\mathord:s_m.~\tyPrim{X}{e_1} \to \ldots \to \tyPrim{X}{e_n} \to \tyPrim{X}{e}
\end{displaymath}
and $e \in \Gen(\{e_1,...,e_n\})$.
\begin{theorem}\label{thm:general-nondef}
  Assume that the members of $\Gamma_{\mathit{ops}}$ are all of
  well-generated types. If there exists an $M$ with typing:
  \begin{displaymath}
    \Gamma_{\mathit{ops}} \vdash M : \forall i_1\mathord:s_1,...,i_m\mathord:s_m.~\tyPrim{X}{e_1} \to ... \to \tyPrim{X}{e_n} \to \tyPrim{X}{e}
  \end{displaymath}
  then $e \in \Gen(\{e_1,...,e_n\})$.
\end{theorem}
This theorem is usually applied in the contrapositive: if $e \not\in
\Gen(\{e_1,...,e_n\})$ then no such $M$ can exist. Note that if
$\Gamma_{\mathit{ops}}$ contains operations corresponding to each of
the index-level operations (as, for example, in
\autoref{fig:real-ops}), then this theorem yields a characterisation
of definable terms, since the construction of $e$ from $e_1,...,e_n$
can be replicated at the term level.
\begin{proof}
  Let $\Delta = i_1\mathord:s_1,...,i_m\mathord:s_m$ be the index
  context constructed from the universally quantified type variables
  in the type of $M$.  To interpret the index expressions, we use the
  free model over the variables in $\Delta$ constructed from the
  syntax. This model assigns to each sort $s$ the set $\{ e \sepbar
  \Delta \vdash e : s \}/\mathord\equiv$ of index expressions
  quotiented by index expression equality. Index operations are
  interpreted by the corresponding syntactic operation on equivalence
  classes: $\semIndexExp{\texttt{\textup{f}}}([e_1],...,[e_n]) =
  [\texttt{\textup{f}}(e_1,...,e_n)]$.

  We take the relational interpretation of the primitive type
  $\tyPrimNm{X}$ as:
  \begin{displaymath}
    \semPrimType{\tyPrimNm{X}}(e) = \{ (x,x) \sepbar x \in \tySem{\tyPrimNm{X}} \land e \in \mathrm{Gen}(\{e_1,...,e_n\}) \}
  \end{displaymath}
  It is straightforward to check that for any index-erasure
  interpretation of the primitive operations $\eta_{\mathit{ops}} \in
  \Gamma_{\mathit{ops}}$, we have $(\eta_{\mathit{ops}},
  \eta_{\mathit{ops}}) \in \rsem{\Gamma_{\mathit{ops}}}*$ because 
  all members of $\Gamma_{\mathit{ops}}$ have well-generated
  types. Hence, by \thmref{thm:abstraction}, we know that for all
  terms (i.e., elements of the free model over $\Delta$)
  $e'_1\mathord:s_1,...,e'_m\mathord:s_m$:
  \begin{displaymath}%FIXME: check order of index environments
    \begin{array}{@{}l}
      \forall (x_1,x_1') \in \rsem{\tyPrim{X}{e_1}}(e'_1,...,e'_m), ..., \\
      \quad(x_n, x'_n) \in \rsem{\tyPrim{X}{e_n}}(e'_1,...,e'_m). \\
      \quad (\tmSem{M} \eta_{\mathit{ops}}~x_1 ~...~x_n, \tmSem{M}\eta_{\mathit{ops}}~x'_1~...~x'_n) \in \rsem{\tyPrim{X}{e}}(e'_1,...,e'_m)
    \end{array}
  \end{displaymath}
  By setting $e'_j = i_j$, and using an arbitrary element $x \in
  \tySem{\tyPrimNm{X}}$ (which we have assumed non-empty), we have,
  for all $k$, $(x,x) \in \rsem{\tyPrim{X}{e_k}}(i_1,...,i_m)$ since
  each $e_k$ is a member of the set we are using to generate
  terms. Now $(\tmSem{M} \eta_{\mathit{ops}}~x...x,
  \tmSem{M}\eta_{\mathit{ops}}~x...x) \in \rsem{\tyPrimNm{X}}(e)$ and
  so $e \in \Gen(\{e_1,...,e_n\})$.
\end{proof}

% When applying \thmref{thm:general-nondef}, it is useful to observe
% that the free model constructed from the index variables $i_1,...,i_m$
% often has a well-known structure. Moreover, the set $\Gen(S)$ is
% always a sub-model of the free model. These observations can make it
% easy to apply the theorem in specific cases.

\paragraph{Application to \exref{ex:uninhabited-type}}
\thmref{thm:general-nondef} can be directly applied to show that the
type $\forall t\mathord:\SynTransl{2}.~\tyPrim{vec}{t + t} \to
\tyPrim{vec}{t}$ has no inhabitants. The free model over the single
index variable $t$ is (isomorphic to) the integers, and the sub-model
generated by the index expression $t + t$ corresponds to the even
integers. The result now follows simply because $1$ (i.e., the
interpretation of $t$) is not an even number.

\paragraph{Abelian Group Indexed Types} Kennedy
\cite{kennedy97relational} has given a general characterisation of
definability at first-order in the case of abelian group indexing in
terms of integer solutions to a set of linear equations. Specialising
\thmref{thm:general-nondef} to the case of abelian group indexing
yields Kennedy's characterisation.

\paragraph{Polymorphic Constants} \thmref{thm:general-nondef} does not
apply in the case when we have polymorphic constants. This is the case
with the polymorphic $0 : \forall
s\mathord:\SynGL{1}.~\tyPrim{real}{s}$ in
\autoref{fig:real-ops}. \thmref{thm:general-nondef} does not apply
because the index expression $s$ is not generated by the empty set:
$0$'s type is not well-generated. Nevertheless, it is easy to adapt
the proof of \thmref{thm:general-nondef} to handle a polymorphic
constant like $0$ by setting the relational interpretation of
$\tyPrimNm{X}$ to be:
\begin{displaymath}
  \rsem{X}(e) = \{ (x,x) \sepbar x \in \tySem{\tyPrimNm{X}} \land (x = 0 \lor e \in \Gen(\{e_1,...,e_n\})) \}
\end{displaymath}
The conclusion of the theorem now states that either we have $e \in
\Gen(\{e_1,...,e_n\})$ \emph{or} $\tmSem{M}$ is the constant $0$
function. This extended theorem can now be used to give an alternative
proof for \exref{ex:sqrt-root-nondef}. As this example illustrates,
there may be many different models that can be used to prove a
non-definability result.

\paragraph{Adding Index Operations} \thmref{thm:general-nondef} also
does not directly apply in the case of \exref{ex:cube-root-nondef},
again because assumption that the types of the primitive operations
are all well-generated is not satisfied. In this case, the assumed
square root operation has type $\forall
s\mathord:\SynGL{1}.~\tyPrim{real}{s^2} \to \tyPrim{real}{s}$, and as
we observed at the start of this section, $s$ is not in the set
generated by $s^2$. However, to enable the application of the theorem,
we can assume an additional index operation $-^{1/2}$, acting like
square root at the index level. Now the free model produced in the
proof of the theorem is isomorphic to the dyadic numbers with addition
and halving, and the generated sub-model consists of the dyadic
rationals of the form $\frac{3k}{2^n}$. Again there is a diversity of
models that can be used to prove a single non-definability result.

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:
%%%---------- close: section-general-nondef.tex
%%%%%%%%%%% Springe nach section-info-flow.tex
%%%---------- open: section-info-flow.tex
\section{Logical Information Flow}
\label{sec:information-flow}

\newcommand{\infoflow}{\mathit{Log}}

We now apply our general framework to types that are indexed by
logical propositions. By including a primitive type that represents
logical truth, we can recover--through a construction due to Tse and
Zdancewic \cite{tse04translating}--strong information flow properties
of programs. As a result of our general framework being parameterised
by the choice of equational theory, we can alter the logic that we use
for reasoning about type equality, and hence alter the information
flow properties of the system.

We first recall the concept of information flow. A function $f : A
\times B \to C$ is said to not allow information to flow from its
second argument to the output if for all $b,b' \in B$ and all $a \in
A$, $f(a,b) = f(a,b')$. If we think of the $B$ argument as
representing high-security information, then we have stated that $f$
does not allow the high-security input to flow to the low security
output. Information flow can be seen as a kind of invariance property
of programs, and so our relational interpretation of types is well
tailored to proving this kind of property.

As described by Sabelfeld and Sands \cite{sabelfeld01per}, information
flow can be captured semantically by partial equivalence relations
(PERs). Abadi, Banerjee, Heintze and Riecke \cite{abadi99core} built a
\emph{Core Calculus for Dependency}, using a type system based around
a security level indexed monad $T_lA$, using PERs to prove the
information flow properties. Tse and Zdancewic \cite{tse04translating}
translated Abadi \emph{et al.}'s calculus into System F\footnote{Tse
  and Zdancewic's translation did not satisfy all the properties that
  they claimed, as pointed out by Shikuma and Igarashi
  \cite{shikuma08proving}. However, this problem is not relevant to
  our discussion here.} translating the monadic type $T_lA$ to
$\alpha_l \to A$ for some free type variable $\alpha_l$, and using
Reynolds' Abstraction Theorem to prove information flow
properties. For example, if the type variable $\alpha_H$ represents
high-level information, then the non-interference property of the
function $f$ could be expressed by the System F type $A \to (\alpha_H
\to B) \to C$. If a program cannot generate a value of type
$\alpha_H$, then it cannot access the value of $B$, and hence is
insensitive to the actual value. Relationships between security levels
are captured by postulating functions $\alpha_{l_1} \to \alpha_{l_2}$
whenever $l_1$ is a lower security level than $l_2$.

Using algebraic indexing, we refine Tse and Zdancewic's translation by
replacing each type variable $\alpha_l$ with a primitive type
$\tyPrim{T}{\phi_l}$ of representations of the truth of a logical
proposition $\phi_l$ that stands for the security level $l$. The
relationships between security levels are now replaced by logical
entailment, so we only have functions of type $\tyPrim{T}{\phi_{l_1}} \to
\tyPrim{T}{\phi_{l_2}}$ when $\phi_{l_1}$ entails
$\phi_{l_2}$. %Inspired by the PER interpretation of information flow,
We instantiate our general relational framework to interpret
$\tyPrim{T}{\phi}$ as the identity relation if $\phi$ is true and the
empty relation if $\phi$ is false. We shall see that this recovers the
information flow properties of Abadi {\em et al.} and Tse and
Zdancewic.

\paragraph{Instantiation of the General Framework}
We assume a single indexing sort $\mathsf{prop}$ and assume the
operations and equations of boolean algebra. Thus we have constants
$\top$, $\bot$ and binary operators $\land$, $\lor$ with the axioms of
a bounded lattice, and a unary complementation operator $\lnot$. We
will use $\phi, \psi$ to stand for index expressions of sort
$\mathsf{prop}$. We use an equational presentation of boolean algebras
to fit with our general framework, but note that we can define an
order on index expressions as $\phi \leq \psi$ when $\phi = \phi \land
\psi$.

We have a single primitive type $\tyPrimNm{T}$, with
$\primTyArity(\tyPrimNm{T}) = [\mathsf{prop}]$ and index-erasure
semantics $\tyPrimSem{\tyPrimNm{X}} = \{ * \}$. Thus values of type
$\tyPrim{T}{\phi}$ have no run-time content; their only meaning is
given by the relational semantics. For the model of the indexing
theory, we take an arbitrary boolean algebra $L$. The relational
interpretation of the truth representation type is
$\rsem{\tyPrimNm{T}}(x) = \{ (*,*) \sepbar x = \top \}$, where $\top$
is the top element of the boolean algebra $L$. The primitive
operations $\Gamma_\infoflow$ reflect logical consequence:
\begin{displaymath}
  \begin{array}{l}
  \begin{array}{@{}c@{\hspace{2em}}c}
    \mathrm{truth} : \tyPrim{T}{\top} &
    \mathrm{and}   : \forall p, q\mathord:\mathsf{prop}.\ \tyPrim{T}{p} \to \tyPrim{T}{q} \to \tyPrim{T}{p \land q}
  \end{array} \\
  \mathrm{up} : \forall p, q\mathord:\mathsf{prop}.\ \tyPrim{T}{p \land q} \to \tyPrim{T}{p}
\end{array}
\end{displaymath}
The combination of the \TirName{TyEq} rule and the primitive
$\mathrm{up}$ operation allow for logical entailment to be reflected
in programs: if we have $\phi \leq \psi$ and $M : \tyPrim{T}{\phi}$
then $\mathrm{up}\ M : \tyPrim{T}{\psi}$. Each of the primitive
operations has a trivial interpretation, due to the index-erasure
interpretation of $\tyPrim{T}{\phi}$ as a one-element set, giving an
environment $\eta_\infoflow \in \ctxtSem{\Gamma}$. Less trivially, we
have this lemma:
\begin{lemma}\label{lem:environments-information-flow}
  $(\eta_\infoflow, \eta_\infoflow) \in \rsem{\Gamma_\infoflow}{}*$.
\end{lemma}

\paragraph{Information Flow} We think of logical expressions as
``composite principals''. That is, propositional variables representing
atomic principals that are combined using the logical connectives. We
interpret ``truth'' for principals as stating that a principal is true
when satisfied with the current state of affairs. Thus a relationship
$\phi \leq \psi$ indicates that satisfaction of the composite
principal $\phi$ implies satisfaction of the composite principal
$\psi$. In terms of security levels, the ordering is reversed: if a
high security principal is satisfied, then all of their subordinates
must also be satisfied.

We adapt Tse and Zdancewic's translation of Abadi \emph{et al.}'s monadic
type to our setting. We define a type abbreviation $T_\phi A =
\tyPrim{T}{\phi} \to A$, where $A$ is a type and $\phi$ is an
expression of sort $\mathsf{prop}$. For every $\phi$, we can endow the
types $T_\phi-$ with the structure of a monad. This is due to the fact
that it is an instance of the ``environment'' (or ``reader'') monad
\cite{jones95functional}. We read the types $T_\phi A$ as data of type
$A$ ``protected'' by the principal $\phi$.

As a consequence of the relational interpretation given above, it
follows that if we have an index expression $\phi$ that is interpreted
as some value other than $\top$ in an index environment $\rho$,
then for all $x,x' \in \tySem{T_\phi\tyPrimNm{bool}}$, we have $(x,x')
\in \rsem{T_\phi\tyPrimNm{bool}}{\rho}$. Thus if a principal is
dissatisfied (i.e.,~$\phi \not= \top$), then data protected by this
principal is indistinguishable from any other data, and a program
cannot get access to the exact value. From this observation, and
\thmref{thm:abstraction}, we obtain the following information flow
result:

\begin{theorem}
  Let $\phi$ and $\psi$ be index expressions of sort $\mathsf{prop}$
  in some indexing context $\Delta$, such that $\psi \not\leq
  \phi$. Then for all terms:
  \begin{displaymath}
    \Delta; \Gamma_\infoflow,\Gamma \vdash M : T_\phi\tyPrimNm{bool} \to T_\psi\tyPrimNm{bool}
  \end{displaymath}
  and all terms $N_1,N_2$ of type $T_\phi\tyPrimNm{bool}$, $M\ N_1
  \stackrel{ctx}\approx M\ N_2$.  Thus there is no information flow
  from $M$'s input to its output.
\end{theorem}
Note that if $\psi \leq \phi$, then it is always possible to write the
identity function with this type, using the $\mathrm{up}$ operation.
The theorem also holds if we move to logics other than boolean
logic. For example, if our equational theory models intuitionistic
logic by taking the axioms of Heyting algebras, then the same non-flow
property for programs of type $T_{p\lor\lnot p}\tyPrimNm{bool} \to
T_\top\tyPrimNm{bool}$ holds, due to the lack of excluded middle. If
we take linear logic, then programs of type $T_{p\otimes
  p}\tyPrimNm{bool} \to T_p\tyPrimNm{bool}$ have no information flow
from their input, due to non-provability of $p \vdash p \otimes p$.

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:
%%%---------- close: section-info-flow.tex
%%%%%%%%%%% Springe nach section-metric.tex
%%%---------- open: section-metric.tex
\section{Distance-Indexed Types}
\label{sec:metric-types}

\newcommand{\Metric}{\mathit{Met}}
\newcommand{\metricSort}{\mathsf{R}^{>0}}

The type system for geometry we discussed in
\autoref{sec:motivating-examples} and \autoref{sec:instantiations}
made use of a relational interpretation of primitive types that
relates pairs of elements by some transformation if applying the
transformation to the first element of a pair yields the second. Thus,
the free theorems that we derived directly take the form of
``invariance'' properties, where some equation holds between two
terms. In this section, we examine another instantiation of our
general framework that relates values when they are within a certain
distance. The free theorems that we obtain inform us of the effect
that programs have on the distances between values. For example, a
program $M$ of type:
\begin{displaymath}
  \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2}
\end{displaymath}
must satisfy the property that for all $\epsilon_1, \epsilon_2 > 0$
and $x, x', y, y' \in \mathbb{R}$:
\begin{displaymath}
  \begin{array}{l}
    \textrm{if }\abs{x - x'} < \epsilon_1 \textrm{ and } \abs{y - y'} < \epsilon_2 \textrm{ then}\\
    \hspace{2em}\abs{\tmSem{M}\ \eta_\Metric\ x\ y - \tmSem{M}\ \eta_\Metric\ x'\ y'} < \epsilon_1 + \epsilon_2
  \end{array}
\end{displaymath}

\paragraph{Instantiation of the General Framework}
We assume a single indexing sort $\metricSort$ %, intended to 
to represent positive, non-zero real numbers. For the index
operations, we assume the operations $\min, \max, +$ and
multiplication by constant reals. There is a single primitive type
$\tyPrimNm{real}$ with $\primTyArity(\tyPrimNm{real}) =
[\metricSort]$. The primitive operations $\Gamma_{\Metric}$ are as
follows, where $c$ stands for arbitrary real-valued constants:
\begin{displaymath}
  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
    \underline{c} & \forall \epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \\
    (+) & \forall \epsilon_1, \epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
    (-) & \forall \epsilon_1, \epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
    \underline{c} * & \forall \epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \to \tyPrim{real}{c\epsilon} \\
    \mathrm{up} & \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\max(\epsilon_1,\epsilon_2)}
  \end{array}
\end{displaymath}
We assume that the index-erasure semantics of the $\tyPrimNm{real}$
type is just the set $\mathbb{R}$, so all except the last operation
have straightforward interpretations. The $\mathrm{up}$ operation is
interpreted just as the identity function. The index-erasure
interpretations of the primitive operations are collected together
into an environment $\eta_\Metric \in \ctxtSem{\Gamma_\Metric}$.

For the relational intepretation we construct a model of the indexing
theory by interpreting $\metricSort$ with strictly positive real
numbers. We set $\rsem{\tyPrimNm{real}}(\epsilon) = \{ (x,x') \sepbar
\abs{x - x'} < \epsilon \}$.
\begin{lemma}\label{lem:metric-environments}
  $(\eta_\Metric, \eta_\Metric) \in \rsem{\Gamma_\Metric}{}*$.
\end{lemma}

\paragraph{Uniform continuity}
  %By making use of 
Using %the
existential types, % in our framework,
%we can state 
the standard $\epsilon$-$\delta$ definition of
\emph{uniform} continuity can be expressed as %a type:
%\begin{displaymath}
$  \forall \epsilon \mathord: \metricSort.\ \exists \delta\mathord: \metricSort.\\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
%\end{displaymath}
For any program $M$ of this type, %we obtain from
\thmref{thm:abstraction} gives a free theorem that is exactly 
%the definition of 
uniform continuity:
\begin{displaymath}
  \forall \epsilon \mathord> 0. \exists \delta \mathord> 0. \forall x,x'. \abs{x - x'} \mathord< \delta \Rightarrow \abs{\tmSem{M}\eta_\Metric~x - \tmSem{M}\eta_\Metric~x'} \mathord< \epsilon
\end{displaymath}
This definition differs from the $\epsilon$-$\delta$ definition of
(regular) continuity by the order of quantification:
  %For continuity, 
there, $\forall x$ comes before $\exists \delta$, so the distance
$\delta$ may depend on %the value of
$x$. We suspect that to express standard continuity as a type would
require some form of type dependency. Chaudhuri, Gulwani and
Lublinerman \cite{chaudhuri10continuity} have given a program logic
based approach to verifying the continuity of programs.

\paragraph{Function Sensitivity}
A type system with a relational interpretation that tracks distances
between values has %previously
been investigated by Reed and Pierce \cite{reed10distance} in the
setting of differential privacy. Their system uses a linear type
discipline to ensure that all programs are $c$-sensitive (i.e., the
distance between the outputs is no greater than $c$ times the
distance between the inputs, for some constant $c$). We can express
their central concept of $c$-sensitivity (for functions on the reals)
as an algebraically indexed type: $\forall
\epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \to
\tyPrim{real}{\frac{1}{c}\epsilon}$. Investigating the precise
connection between their system and ours is left to future work.

% \begin{example}[Differential Privacy]
%   Relative error? Probability monads?

%   Don't have space to develop this fully...

%   $\circ \tyPrimNm{real}$ -- probability distributions on the real numbers

%   Need to think about this...
%   \begin{displaymath}
%     \mathsf{noise} : \tyPrim{$\circ$real}{1}
%   \end{displaymath}
% \end{example}

% \begin{mathpar}
%   % commutativity
%   \min(\epsilon_1, \epsilon_2) = \min(\epsilon_2, \epsilon_1)

%   \max(\epsilon_1, \epsilon_2) = \max(\epsilon_2, \epsilon_1)

%   % associativity
%   \max(\epsilon_1, \max(\epsilon_2, \epsilon_3)) = \max(\max(\epsilon_1, \epsilon_2), \epsilon_3)

%   \min(\epsilon_1, \min(\epsilon_2, \epsilon_3)) = \min(\min(\epsilon_1, \epsilon_2), \epsilon_3)

%   % absorption
%   \max(\epsilon_1, \min(\epsilon_1, \epsilon_2)) = \epsilon_1

%   \min(\epsilon_1, \max(\epsilon_1, \epsilon_2)) = \epsilon_1
% \end{mathpar}

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:

%%%---------- close: section-metric.tex
%%%%%%%%%%% Springe nach section-discussion.tex
%%%---------- open: section-discussion.tex
\section{Discussion}
\label{sec:discussion}

We presented a general framework for algebraically indexed types and
instantiated it to yield novel type systems for geometry, logical
information flow and distance-indexed types. Our framework further
demonstrates the power of relational reasoning about typed
programs. From \thmref{thm:abstraction}, we derived interesting free
theorems, type isomorphisms and non-definability results. We conclude
with some observations and suggestions for further work.

\paragraph{Further Applications and Extensions} We have covered
several applications of algebraically indexed types in this paper, but
there are undoubtedly many more. Geometry for dimensions greater than
two is an obvious candidate, as are systems that are invariant
under different geometric groups (e.g.,~the Poincar\'{e} group for
relativity). Mathematical Physics is particularly rich in theories
that have some notion of invariance, and it will be exciting to pin
down the precise connections between these and type systems for which
%analogues of Reynolds' 
an Abstraction Theorem holds. 
Cardelli and Gardner describe a process calculus that builds in 3D affine geometry~\cite{Cardelli},
proving that process behaviour is invariant under affine transformations. Distinguishing points from vectors
provides the appropriate abstraction barrier, and the geometric group is determined by inspecting
term syntax. It would be interesting to recast their language in terms of our indexed types to
obtain purely type-based invariance theorems.

Geometric theorem
proving is another %possible 
application. Harrison
\cite{harrison09without} comments on the pervasiveness of % usefulness of
invariance properties in this area. Programs in our framework
automatically satisfy invariance properties, removing the need for
{\em ad hoc} proofs of these facts.

% In \autoref{sec:metric-types} we presented a type system with
% distance-indexed types, and noted the similarity with Reed and
% Pierce's system for non-expansivity \cite{reed10distance}. Exploring
% the relationship--particularly with respect to probability
% monads--also looks promising.

Type and effect analyses use types indexed by effect annotations with
algebraic structure (e.g.,~sets of read/write effect labels with an
idempotent monoid structure). Benton {\em et al.}~have used relational
interpretations to prove effect-dependent equivalences
\cite{benton06reading}. An extension of our framework with
type-indexed types should be able to express their effect-indexed monads
and prove their equivalences.

Extending our framework with type dependency would also allow for
further applications. For example, we could consider a type of lists
of length $n$, indexed by elements of the permutation group
$S_n$. Bernardy {\em et al.}~have presented a general framework for
relational reasoning and an Abstraction Theorem for dependent types
\cite{bernardy12proofs}. However, they work with pure type systems,
which define type equality via untyped rewriting, so it is not
immediately obvious how to integrate arbitrary equational theories
into their framework.

\paragraph{Semantic Equality} In general, the semantic equality
in \defref{def:semantic-equality} is not an
equivalence relation. If the interpretations of all 
primitive types are partial equivalence relations then semantic
equality is indeed an equivalence relation. However, this excludes the
geometry and distance-indexed examples. More generally, we can
consider relational interpretations that are \emph{difunctional}. (A
relation is difunctional if whenever $(x,y)$, $(x',y')$ and $(x,y')$
are in the relation then so is $(x',y)$.) Difunctionality is weaker
than being a PER, but still suffices to prove that semantic equality
is an equivalence relation. Hofmann \cite{hofmann08correctness} has
used difunctional relations in the setting of effect
analyses. Difunctionality covers all our examples except
distance-indexed types. Note that for both PERs and difunctional
relations we need to close the relational interpretation of
existential types under the appropriate property to ensure that all
types are interpreted as PERs/difunctional relations. For
distance-indexed types it is possible that a new notion of
equivalence based on closeness is required.

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:
%%%---------- close: section-discussion.tex

\section*{Acknowledgements}
The authors would like to thank Nick Benton, Kenneth MacKenzie, John
Reppy and Martin Will for illuminating discussions on geometry and
types. Atkey and Johann were supported by EPSRC grant EP/G068917/1.

\begin{thebibliography}{12}
\softraggedright

\bibitem{abadi99core} 
M. Abadi, A. Banerjee, N. Heintze, and J. G. Riecke. A Core Calculus
of Dependency. Proceedings, POPL, pp.147-160, 1999.

\bibitem{syntaxforfree}
R. Atkey.  Syntax for Free: Representing Syntax with Binding Using
Parametricity.  Proceedings, TLCA,
pp. 35-49, 2009.

\bibitem{TypedSyntax}
N.~Benton, C.-K.~Hur, A.~J.~Kennedy, C.~McBride.
Strongly Typed Term Representations in {Coq}.
{\em Journal of Automated Reasoning} 49(2), pp. 141-159, 2012.

\bibitem{benton06reading}
N.~Benton, A.~Kennedy, M.~Hofmann, L.~Beringer.
Reading, Writing and Relations.
Proceedings, APLAS,
pp. 114-130, 2006.

\bibitem{bernardy12proofs} 
J.-P.~Bernardy, P.~Jansson, R.~Paterson.
Proofs for Free: Parametricity for Dependent Types.
{\em Journal of Functional Programming} 22(2), pp. 107-152, 2012.

\bibitem{Cardelli}
L.~Cardelli, P.~Gardner.
Processes in Space.
{\em Programs, Proofs, Processes}: Proceedings, CiE, pp. 78-87, 
2010.

\bibitem{CGAL}
Computational Geometry Algorithms Library (CGAL): User and Reference
Manual.  Available at {\tt http://www.cgal.org}.

\bibitem{chaudhuri10continuity}
S. Chaudhuri, S. Gulwani, and R. Lublinerman.  Continuity Analysis of
Programs.  Proceedings, POPL, pp. 57-70, 2010.

\bibitem{gallier11geometric}
J. Gallier. {\em Geometric Methods and Applications For Computer
Science and Engineering}. Springer, 2011.

\bibitem{harrison09without}
J.~Harrison.
Without Loss of Generality.
Proceedings, TPHOLs,
pp. 43-59, 2009.

\bibitem{hofmann08correctness}
M.~Hofmann. Correctness of Effect-based Program
Transformations. Formal Logical Methods for System Security and
Correctness, pp. 149-173, 2008.

\bibitem{jones95functional}
M.~P.~Jones.  Functional Programming with Overloading and Higher-Order
Polymorphism. Proceedings, AFP,
pp. 97-136, 1995.

\bibitem{kennedy97relational} 
A.~J.~Kennedy. Relational Parametricity and Units of Measure.
Proceedings, POPL, pp. 442-455, 1997.

\bibitem{fsharp}
A.~J.~Kennedy. Types for Units-of-Measure: Theory and Practice.
Central European Functional Programming school (CEFP), pp. 268--305, LNCS vol. 6299, 2010.

\bibitem{CFGADT}
S.~Mann, N.~Litke, T.~DeRose.  A Coordinate Free Geometry {ADT}.
Technical Report CS-97-15, University of Waterloo, 1997.

\bibitem{PittsAM:parpoe}
A.~M.~Pitts.  Parametric Polymorphism and Operational Equivalence.
{\em Mathematical Structures in Computer Science} 10(3), pp. 321-359,
2000.

\bibitem{reed10distance}
J.~Reed and B.~C.~Pierce.  Distance Makes the Types Grow Stronger.
Proceedings, ICFP, pp. 157-169, 2010.

\bibitem{reynolds83types}
J.~C.~Reynolds. Types, Abstraction and Parametric Polymorphism.
{\em Information Processing} 83, pp. 513-523, 1983.

\bibitem{sabelfeld01per} 
A.~Sabelfeld and D.~Sands.  A PER Model of Secure Information Flow in
Sequential Programs.  {\em Higher-Order and Symbolic Computation} 14
(1), pp. 59-91, 2001.

\bibitem{shikuma08proving}
N.~Shikuma and A.Igarahsi.  Proving Noninterference by a Fully
Complete Translation to the Simply Typed lambda-calculus.  {\em
    Logical Methods in Computer Science} 4(3), 2008.

\bibitem{tse04translating}
S.~Tse and S.~Zdancewic.  Translating Dependency into Parametricity.
Proceedings, ICFP,  pp.
115-125, 2004.

\bibitem{wadler89theorems}
P.~Wadler.  Theorems for Free!.  Proceedings, FPCA,
pp. 347-359, 1989.

\end{thebibliography}

\end{document}

### 2013-07-05 mechanics/paper.tex
Outline of Mechanics paper

\documentclass[preprint]{sigplanconf}

\newcommand{\sepbar}{\mathrel|}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{stmaryrd}

\title{From Parametricity to Conservation Laws, via Noether's Theorem}

\authorinfo{}{}{}
% \authorinfo{Robert Atkey}
%            {}
%            {bob.atkey@gmail.com}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{property}{Property}
\newtheorem{proposition}{Proposition}
\newcommand{\lemref}[1]{\hyperref[#1]{Lemma~\ref*{#1}}}
\newcommand{\thmref}[1]{\hyperref[#1]{Theorem~\ref*{#1}}}
\newcommand{\defref}[1]{\hyperref[#1]{Definition~\ref*{#1}}}
\newcommand{\propref}[1]{\hyperref[#1]{Proposition~\ref*{#1}}}
\newcommand{\corref}[1]{\hyperref[#1]{Corollary~\ref*{#1}}}
\newcommand{\conref}[1]{\hyperref[#1]{Conjecture~\ref*{#1}}}
\newcommand{\exref}[1]{\hyperref[#1]{Example~\ref*{#1}}}
\newcommand{\statementref}[1]{\hyperref[#1]{Statement~\ref*{#1}}}

\newcommand{\smoothrightarrow}{\stackrel{C^\infty}\longrightarrow}
\newcommand{\typeOfCartSp}[1]{\lbag #1 \rbag}
\newcommand{\elab}[1]{\lfloor #1 \rfloor}

\newtheoremstyle{examplestyle}
  {\topsep}  % space above
  {\topsep}  % space below
  {\normalfont}% name of font to use in the body of the theorem
  {0em}% measure of space to indent
  {\bfseries}% name of head font
  {.}% punctuation between head and body
  {5pt plus 1pt minus 1pt}% space after theorem head; " " = normal interword space
  {}% Manually specify head
\theoremstyle{examplestyle}
\newtheorem{example}{Example}
\newtheorem*{example*}{Example}

\newcommand{\fomega}{F$\omega$}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Rel}{\mathsf{Rel}}
\newcommand{\Pow}{\mathcal{P}}
\newcommand{\frel}[1]{\langle #1 \rangle}
\newcommand{\coq}[1]{\textsf{#1}}
\newcommand{\CIC}{\textsc{CIC}}

\newcommand{\semKU}[1]{\llbracket #1 \rrbracket^O}
\newcommand{\semKR}[1]{\llbracket #1 \rrbracket^R}
\newcommand{\semKI}[1]{\llbracket #1 \rrbracket^{\mathrm{id}}}

\begin{document}

\maketitle

\begin{abstract}
  Invariance is of paramount importance in programming languages and
  in physics. In programming languages, John Reynolds' theory of
  relational parametricity demonstrates that parametric polymorphic
  programs are invariant under change of data representation, a
  property that yields ``free'' theorems about programs just from
  their types. In physics, Emmy Noether showed that if the action of a
  physical system is invariant under change of coordinates, then the
  physical system has a conserved quantity: a quantity that remains
  constant for all time. Knowledge of conserved quantities can reveal
  deep properties of physical systems. For example, the conservation
  of energy, which by Noether's theorem is a consequence of a system's
  invariance under time-shifting.

  In this paper, we link Reynolds' relational parametricity with
  Noether's theorem for deriving conserved quantities. We propose an
  extension of System F$\omega$ with new kinds, types and term
  constants for writing programs that describe classical mechanical
  systems in terms of their Lagrangians. We show, by constructing a
  relationally parametric model of our extension of F$\omega$, that
  relational parametricity is enough to satisfy the hypotheses of
  Noether's theorem, and so to derive conserved quantities for free,
  directly from the polymorphic types of Lagrangians expressed in our
  system.
\end{abstract}

\section{Introduction}

\begin{enumerate}
\item Symmetry in programming and physics
  \begin{enumerate}
  \item In programming: replacement of one data type by another
  \item In physics: invariance under change of unobservable
    properties. FIXME: check this.
  \item What unifies these, at a conceptual level? Inability to
    distinguish between situations, given the tools provided. In both
    cases, this leads to interesting properties of such systems. In
    programming, we get ``free theorems''. In Physics, via Noether's
    theorem, we get ``conservation laws''.
  \item In this paper, we show that we can define a programming
    language for classical mechanics, based on System F$\omega$, where
    the abstraction theorem directly fulfils the hypothesis of
    Noether's theorem, allowing us to derive conservation laws.
  \end{enumerate}
\item Parametricity: polymorphic programs have symmetry properties
  \begin{enumerate}
  \item Example of a free theorem
  \item Example of a geometric free theorem (the area of a triangle
    example)
  \end{enumerate}
\item Noether's theorem: symmetry properties of descriptions of
  systems lead to conservation laws
\item Let's put the two together
  \begin{enumerate}
  \item Need invariance properties of smooth functions
  \item So we extend System F$\omega$ with 
  \end{enumerate}
\item We present the classical mechanics and Noether's theorem first,
  since it is likely to be more unfamiliar to readers.
\end{enumerate}

We restate the results of Atkey, Johann and Kennedy in the more
general setting of System F$\omega$... which is good because it allows
for a deeper interpretation of what the unit kinds are. Also, we can
introduce useful stuff like nat-indexed types without too much fuss.

\paragraph{Summary of the Paper}

\begin{itemize}
\item FIXME: Recall the necessary background for Noether's theorem
  (Section~\ref{sec:conservation-laws-from-symmetry}).
\item FIXME: Show how to get symmetry properties from types
  (Section~\ref{sec:symmetry-from-types}).
\item FIXME: In Section~\ref{sec:types-for-classical-mech}, we
  construct a type system for classical mechanics,
\item FIXME: Section~\ref{sec:examples} we present examples of using
  our extension of System F$\omega$ for writing invariant Lagrangians
  that describe physical systems.
\item FIXME: field theory
\end{itemize}

\section{Conservation Laws from Symmetry}
\label{sec:conservation-laws-from-symmetry}

Make sure to state up front the necessary background required. To
understand: basic(ish) calculus. We will be using the Calculus of
Variations, but only by stating the main results. We go for standard
physics notation, rather than anything more complicated.

\subsection{The Principle of Stationary Action}

Introduce by stating Newton's second law, and saying that it can be
derived from the principle of stationary action.

Analytical mechanics, Lagrangian and Hamiltonian mechanics. Basically
different mathematical reformulations of Newtonian mechanics that
yield new insights into the underlying structure of the kinds of
physical systems studied in classical mechanics.

\paragraph{Lagrangian and Action}

Lagrangian mechanics reformulates Newtonian mechanics in terms of

Two points: the Langrangian and generalised coordinates. Do
generalised coordinates as a secondary thing.

\begin{displaymath}
  \mathcal{S}[q;a;b] = \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
\end{displaymath}

\begin{example}
  A single particle under a potential field. Set the potential field
  to be (downward) gravity.
\end{example}

\paragraph{The Principle of Stationary Action}

The principle of stationary action

\begin{displaymath}
  \frac{d}{dt}\frac{\partial L}{\partial \dot{q_i}} - \frac{\partial L}{\partial q_i} = 0
\end{displaymath}

Explain the different derivatives going on here

Example: Newton's second law...

\subsection{Noether's Theorem} Noether's theorem provides us with deep
insights into the properties of the solutions of the Euler-Lagrange
equations. FIXME: motivate invariance.

\paragraph{Invariance of the Action} Let
\begin{displaymath}
  \mathcal{S}[q;a;b] = \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
\end{displaymath}
be the action of some physical system described by the Lagrangian
$L$. Assume a differentiable invertible function $\tau : \mathbb{R}
\to \mathbb{R}$ that transforms time in some way and a function
$\sigma : \mathbb{R} \times \mathbb{R}^n \to \mathbb{R}^n$ that
transforms the vector of generalised coordinates $\vec{q}$ at time $t$
to another vector of generalised coordinates $\sigma(t,\vec{q})$.

We say that the action $\mathcal{S}$ is invariant under the
transformations $\tau$ and $\sigma$ if it is the case that for all
$a$, $b$ and $q$ that:
\begin{displaymath}
  \int_a^b L(t,q,\dot{q}) \mathit{dt} = \int_{\tau(a)}^{\tau(b)} L(t', \sigma(\tau^{-1}(t'), q(t')), \mathbf{FIXME}) \mathit{dt'}
\end{displaymath}
FIXME: Remark on the use of $t$ and $t'$ in the integrals. FIXME:
state that all we are doing is transforming the time and the path by
the transformations $\tau$ and $\sigma$ respectively. State why it
will be useful for the

Remark on how to give a simpler version of the invariance principle by
applying some simple transformations of integrals (see notes).

\begin{example}
  Continue a simple example from above to show that its action is
  invariant under (e.g.) time translation.
\end{example}

\paragraph{Noether's Theorem} Noether's theorem applies to actions
that are \emph{continuously} invariant. FIXME: not quite right: write
out the maths first.

(a) parameterise $\tau$ and $\sigma$ by a real value $\epsilon$ such
that we get a continuous symmetry of the action; (b) state Noether's
theorem as we will use it.

\begin{theorem}
  Noether's theorem
\end{theorem}

\begin{example}
  Continue the simple example from above to show that the invariance
  under time translation yields conservation of energy.
\end{example}

State that we will do more examples later on, after we discuss how to
derive symmetry from types.

\section{Symmetry from Types}

The first part here will be at least partly taken from Atkey
2012. Work out a way of emphasising that we do not need to do anything
to extend the basic underlying semantic structures involved: reflexive
graphs are enough. This presentation differs from Atkey 2012 in that
we emphasise the reflexive graph structure more, in order to be able
to better expose the connection with the groupoid kinds in the section
below.

\subsection{Relational Parametricity for System F$\omega$}
\label{sec:refl-graphs-for-fomega}

Reflexive graphs give us a way of capturing the higher-dimensional
structure of types in a

Relational parametricity (Reynolds, BFSS), reflexive graphs
(Robinson/Rosolini, Hasegawa, me).

Explain the semantic setting. Explain the ``property'' thing.

\paragraph{Syntax of System F$\omega$} Present the basic syntax of the
kinds, types and terms of (predicative) System F$\omega$.

Small kinds. We will extend the kinds, types and terms later to
encompass the necessary geometrical constructs for classical
mechanics.

\begin{displaymath}
  \kappa ::= * \sepbar \kappa_1 \to \kappa_2 \sepbar \kappa_1 \times \kappa_2 \sepbar \cdots
\end{displaymath}

\begin{mathpar}
  \inferrule*
  {\kappa_1~\mathrm{small} \\ \kappa_2~\mathrm{small}}
  {\kappa_1 \to \kappa_2~\mathrm{small}}

  \inferrule*
  {\kappa_1~\mathrm{small} \\ \kappa_2~\mathrm{small}}
  {\kappa_1 \times \kappa_2~\mathrm{small}}
\end{mathpar}

Type equality... $\Theta \vdash A \equiv B : \kappa$

\begin{figure}[t]
  \centering
  \begin{mathpar}
    \inferrule*
    {\alpha : \kappa \in \Theta}
    {\Theta \vdash \alpha : \kappa}

    \inferrule*
    {\Theta, \alpha : \kappa_1 \vdash A : \kappa_2}
    {\Theta \vdash \lambda \alpha : \kappa_1.\ A : \kappa_1 \to \kappa_2}

    \inferrule*
    {\Theta \vdash F : \kappa_1 \to \kappa_2 \\ \Theta \vdash A : \kappa_1}
    {\Theta \vdash F A : \kappa_2}

    \inferrule*
    {\Theta \vdash A : \kappa_1 \\ \Theta \vdash B : \kappa_2}
    {\Theta \vdash \langle A, B \rangle : \kappa_1 \times \kappa_2}

    \inferrule* [right=${i \in \{1,2\}}$]
    {\Theta \vdash A : \kappa_1 \times \kappa_2}
    {\Theta \vdash \pi_i~A : \kappa_i}

    \inferrule*
    {\Theta \vdash A : * \\ \Theta \vdash B : *}
    {\Theta \vdash A \times B : *}

    \inferrule*
    {\Theta \vdash A : * \\ \Theta \vdash B : *}
    {\Theta \vdash A \to B : *}

    \inferrule*
    {\Theta, \alpha : \kappa \vdash A : * \\ \kappa\textrm{ small}}
    {\Theta \vdash \forall\alpha\mathord:\kappa.\ A : *}
  \end{mathpar}
  \caption{Types and their Kinds}
  \label{fig:types}
\end{figure}

\begin{figure}[t]
  \centering
  \begin{mathpar}
    \inferrule*
    {x : A \in \Gamma}
    {\Theta \sepbar \Gamma \vdash x : A}

    \inferrule*
    {\Theta \sepbar \Gamma \vdash e : A \\ \Theta \vdash A \equiv B : *}
    {\Theta \sepbar \Gamma \vdash e : B}
  \end{mathpar}
  \begin{mathpar}
    \inferrule*
    {\Theta \sepbar \Gamma \vdash e_1 : A_1 \\\\
      \Theta \sepbar \Gamma \vdash e_2 : A_2}
    {\Theta \sepbar \Gamma \vdash (e_1, e_2) : A_1 \times A_2}

    \inferrule* [right=${i \in \{1,2\}}$]
    {\Theta \sepbar \Gamma \vdash e : A_1 \times A_2}
    {\Theta \sepbar \Gamma \vdash \pi_i e : A_i}
  \end{mathpar}
  \begin{mathpar}
    \inferrule*
    {\Theta \sepbar \Gamma, x : A \vdash e : B}
    {\Theta \sepbar \Gamma \vdash \lambda x : A.\ e : A \to B}

    \inferrule*
    {\Theta \sepbar \Gamma \vdash e_1 : A \to B \\
      \Theta \sepbar \Gamma \vdash e_2 : A}
    {\Theta \sepbar \Gamma \vdash e_1 e_2 : B}

    \inferrule*
    {\Theta, \alpha : \kappa \sepbar \Gamma \vdash e : A \\ \alpha \not\in \mathit{fv}(\Gamma)}
    {\Theta \sepbar \Gamma \vdash \Lambda \alpha \mathord: \kappa.\ e : \forall \alpha \mathord:\kappa. A}

    \inferrule*
    {\Theta \sepbar \Gamma \vdash e : \forall \alpha\mathord:\kappa. A \\ \Theta \vdash B : \kappa}
    {\Theta \sepbar \Gamma \vdash e\ [B] : A\{B/\alpha\}}
  \end{mathpar}
  \caption{Terms and their Types}
  \label{fig:terms}
\end{figure}

\paragraph{Reflexive Graphs and the Interpretation of Kinds} We will
interpret every kind $\kappa$ is as a reflexive graph, which we now
define. A \emph{reflexive graph} is a triple $(O, R, \mathrm{id})$,
where $O$ is a large set of objects, $R : O \times O \to \Set$ assigns
a small set of directed `edges' to each pair of objects, and
$\mathrm{id} : \forall o \in O.~R(o,o)$ assigns a distinguished
`identity' edge from every object to itself. We think of the edges of
a reflexive graph as abstract ``relations'' between the
objects. Indeed, in the interpretation of the kind of types, $*$,
below, the edges will be exactly relations.

A \emph{small} reflexive graph is a reflexive graph $(O, R,
\mathrm{id})$ where $O$ is a small set of objects. We use small
reflexive graphs as the semantic interpretation of small kinds.

The interpretation of kinds as reflexive graphs, and small kinds as
small reflexive graphs is a key property of our semantics that we will
maintain as we add additional kinds in Sections XXX below.

\begin{property}\label{property:semantic-kinds}
  Each kind $\kappa$ is interpreted as a reflexive graph
  $\sem{\kappa}$. If $\kappa~\mathrm{small}$, then $\sem{\kappa}$ is a
  small reflexive graph.
\end{property}

An appealing interpretation of reflexive graphs is as ``categories
without composition''. Following this intuition, we define morphisms
of reflexive graphs as ``functors'', without the preservation of
composition condition. A \emph{morphism of reflexive graphs}
$(O_1,R_1,\mathrm{id}_1)$ and $(O_2, R_2, \mathrm{id}_2)$ is a pair of
mappings $f : O_1 \to O_2$ and $r : \forall o, o' \in
O_1.~R_1(o,o') \to R_2(f~o, f~o')$ such that identities are
preserved: $r~o~o~(\mathrm{id}_1~o) = \mathrm{id}_2~(f~o)$. We use
morphisms of reflexive graphs below to interpret well-kinded types.

We will use the notation $-^O$, $-^R$ and $-^{\mathrm{id}}$ for the
first, second and third projections out of tuples representing
reflexive graphs. Similarly, we use $-^f$ and $-^r$ for the first and
second projections out of tuples representing reflexive graph
morphisms.

We now define the interpretations of the basic kinds of System
F$\omega$ we defined above, making sure that we maintain
Property~\ref{property:semantic-kinds}. At base kind, the collection
of objects is simply the type of all small sets; edges between $A$ and
$B$ are binary relations on $A$ and $B$ (i.e., subsets of $A \times
B$); and the distinguished identity edge is exactly the equality
relation:
\begin{displaymath}
  \sem{*} = (\Set, \Rel, \equiv)
\end{displaymath}
The reflexive graph $\sem{*}$ is not small, due to the collection of
all small sets $\Set$ not forming a small set.

For higher kinds $\kappa_1 \to \kappa_2$, the collection of objects
consists of reflexive graph morphisms from the interpretation of
$\kappa_1$ to the interpretation of $\kappa_2$; the edges between
morphisms $(f, r)$ and $(f', r')$ are edge transformers; and the
distinguished identity relation for $(f,r)$ is just $r$:
\begin{displaymath}
  \begin{array}{@{}l}
    \sem{\kappa_1 \to \kappa_2} = \\
    \begin{array}[t]{l@{}l@{}}
      (& \{ (f,r) \sepbar (f, r) : \sem{\kappa_1} \to \sem{\kappa_2} \},\\
      &((f,r),(f',r')) \mapsto \forall o,o'. \semKR{\kappa_1}(o,o') \to \semKR{\kappa_2}(f~o, f'~o'), \\
      &(f,r) \mapsto r\ )
    \end{array}
  \end{array}
\end{displaymath}
By the assumption that our collection of small sets is closed under
the formation of function spaces and set comprehension, if
$\sem{\kappa_1}$ and $\sem{\kappa_2}$ are small reflexive graphs, then
so is $\sem{\kappa_1 \to \kappa_2}$.

Product kinds are interpreted by taking the product of their
interpretations as reflexive graphs:
\begin{displaymath}
  \begin{array}{@{}l}
    \sem{\kappa_1 \times \kappa_2} = \\
    \begin{array}[t]{l@{}l}
      (& \semKU{\kappa_1} \times \semKU{\kappa_2}, \\
      & ((o_1,o_2),(o_1',o_2')) \mapsto \semKR{\kappa_1}(o_1,o_1') \times \semKR{\kappa_2}(o_2,o_2'), \\
      & (o_1,o_2) \mapsto (\semKI{\kappa_1}(o_1), \semKI{\kappa_2}(o_2)))
    \end{array}
  \end{array}
\end{displaymath}
The objects of the interpretation of a product kind is the product of
the underlying collections of objects of the two parts, and the
relational component is simply the product of the relational
components. This naturally leads to the identity component being
defined as the tuple of the identity components of the two
parts. Again, by assumption that our collection of small sets is
closed under products, if $\sem{\kappa_1}$ and $\sem{\kappa_2}$ are
small reflexive graphs, then so is $\sem{\kappa_1 \times
  \kappa_2}$. Product kinds generalise to the interpretation of
kinding contexts $\Theta = \alpha_1 : \kappa_1, ..., \alpha_n :
\kappa_n$, which are interpreted as the product of the reflexive graph
interpretations of $\kappa_1, ..., \kappa_n$:
\begin{displaymath}
  \begin{array}{l}
    \sem{\alpha_1 : \kappa_1, ..., \alpha_n : \kappa_n} = \\
    \begin{array}[t]{l@{}l}
      (&\semKU{\kappa_1} \times ... \times \semKU{\kappa_n}, \\
      &(\theta,\theta') \mapsto \semKR{\kappa_1}(\pi_1\theta, \pi_1\theta') \times ... \times \semKR{\kappa_n}(\pi_n\theta, \pi_n\theta'), \\
      &\theta \mapsto (\semKI{\kappa_1}(\pi_1 \theta), ..., \semKI{\kappa_n}(\pi_n \theta)))
    \end{array}
  \end{array}
\end{displaymath}

In Sections~\ref{sec:discrete-kinds} and~\ref{sec:groupoid-kinds}
below, we will extend System F$\omega$ with additional kinds, and
assign them reflexive graph interpretations, making sure that we
maintain Property~\ref{property:semantic-kinds}.

\begin{figure*}[t]
  \begin{displaymath}
    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
      \sem{\Theta \vdash \alpha_i : \kappa_i}^f \theta & = & \pi_i \theta \\
      \sem{\Theta \vdash \alpha_i : \kappa_i}^r\theta\theta' \rho & = & \pi_i \rho \\
      \\
      \sem{\Theta \vdash \lambda \alpha \mathord: \kappa_1.~A : \kappa_1 \to \kappa_2}^f \theta & = & (
      \begin{array}[t]{@{}l}
        \lambda o \in \semKU{\kappa_1}.\ \sem{A}^f(\theta, o), 
        \lambda o, o' \in \semKU{\kappa_1}, r \in \semKR{\kappa_1}(o,o').\ \sem{A}^r(\theta,o)(\theta,o')(\semKI{\Theta}\theta, r))
      \end{array}\\
      \sem{\Theta \vdash \lambda \alpha \mathord: \kappa_1.~A : \kappa_1 \to \kappa_2}^r \theta\theta' \rho & = &
      \lambda o,o' \in \semKU{\kappa_1}, r \in \semKR{\kappa_1}(o,o').\ \sem{A}^r(\theta,o)(\theta',o')(\rho, r) \\
      \\
      \sem{\Theta \vdash F A : \kappa_2}^f \theta & = & \pi_1 (\sem{F}^f\ \theta)\ (\sem{A}^f\ \theta)\\
      \sem{\Theta \vdash F A : \kappa_2}^r\theta\theta'\rho & = & \sem{F}^r\theta\theta'\rho\ (\sem{A}^f \theta)(\sem{A}^f \theta')(\sem{A}^r\theta\theta'\rho) \\
      \\
      \sem{\Theta \vdash \langle A, B \rangle : \kappa_1 \times \kappa_2}^f \theta & = & (\sem{A}^f \theta, \sem{B}^f \theta) \\
      \sem{\Theta \vdash \langle A, B \rangle : \kappa_1 \times \kappa_2}^r \theta \theta' \rho & = & (\sem{A}^r \theta \theta' \rho, \sem{B}^r \theta \theta' \rho) \\
      \\
      \sem{\Theta \vdash \pi_i A : \kappa_i}^f \theta & = & \pi_i (\sem{A}^f \theta) \\
      \sem{\Theta \vdash \pi_i A : \kappa_i}^r \theta \theta' \rho & = & \pi_i (\sem{A}^r \theta \theta' \rho)
    \end{array}
  \end{displaymath}
  \caption{Interpretation of type-level $\lambda$-calculus as reflexive graph morphisms}
  \label{fig:type-level-lambda-calc}
\end{figure*}

\begin{figure*}[t]
  \begin{displaymath}
    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
      \sem{\Theta \vdash A \to B : *}^f \theta & = & \sem{A}^f\theta \to \sem{B}^f\theta \\
      \sem{\Theta \vdash A \to B : *}^r\theta\theta'\rho & = &
      \{ (f,f') \sepbar \forall (a,a') \in \sem{A}^r\theta\theta'\rho.\ (f~a, f'~a') \in \sem{B}^r\theta\theta'\rho \} \\
      \\
      \sem{\Theta \vdash A \times B : *}^f \theta & = & \sem{A}^f\theta \times \sem{B}^f\theta \\
      \sem{\Theta \vdash A \times B : *}^r\theta\theta'\rho & = &
      \{ ((a,b),(a',b')) \sepbar (a,a') \in \sem{A}^r\theta\theta'\rho, (b,b') \in \sem{B}^r\theta\theta'\rho \} \\
      \\
      \sem{\Theta \vdash \forall \alpha\mathord:\kappa.~A : *}^f \theta &=& \{
      \begin{array}[t]{@{}l}
        x \in (\forall o \in \semKU{\kappa}.\ \sem{A}^f(\theta, o)) \sepbar
        \forall o,o',r \in \semKR{\kappa}(o,o').\ (x~o, x~o') \in \sem{A}^r(\theta,o)(\theta,o')(\semKI{\Theta} \theta, r) \}
      \end{array} \\
      \sem{\Theta \vdash \forall \alpha\mathord:\kappa.~A : *}^r\theta\theta'\rho &=& \{ (x,x') \sepbar \forall o, o', r \in \semKR{\kappa}(o,o').\ (x~o, x'~o') \in \sem{A}^r(\theta,o)(\theta',o')(\rho, r) \}
    \end{array}
  \end{displaymath}
  \caption{Interpretation of basic types as reflexive graph morphisms}
  \label{fig:basic-type-interpretation}
\end{figure*}

\paragraph{Interpretation of Types} Well-kinded types $\Theta \vdash A
: \kappa$ from Figure~\ref{fig:types} are interpreted as reflexive
graph morphisms $\sem{A} : \sem{\Theta} \to \sem{\kappa}$. We sum this
up as a property of our semantics:
\begin{property}\label{property:semantic-types}
  Each well-kinded type $\Theta \vdash A : \kappa$ is interpreted as a
  reflexive graph morphism $\sem{A} : \sem{\Theta} \to \sem{\kappa}$,
  such that if $\Theta \vdash A \equiv B : \kappa$, then $\sem{A} =
  \sem{B}$.
\end{property}
%
% That is, every well-kinded type is interpreted as a pair $(\sem{A}^f,
% \sem{A}^r)$ of functions $\sem{A}^f : \sem{\Theta}^O \to
% \sem{\kappa}^O$ and $\sem{A}^r : \forall o, o'.~\sem{\Theta}^R(o,o')
% \to \sem{\kappa}^R(\sem{A}^f~o, \sem{A}^f~o')$, such that identity
% edges are preserved.
%
The interpretation of the $\lambda$-calculus fragment (i.e.,
variables, $\lambda$-abstraction and application, and products) of the
language of well-kinded types is displayed in
Figure~\ref{fig:type-level-lambda-calc}. The interpretations are
unsurprising given the reflexive graph interpretation of the kinds
$\kappa_1 \to \kappa_2$ and $\kappa_1 \times \kappa_2$ we gave above.

Figure~\ref{fig:basic-type-interpretation} shows the interpretations
of the basic type constructors for function and product types, and
universal quantification. Each of these constructs builds an object of
kind $*$, so the object-level interpretation is a small set in $\Set$,
and the relation-level interpretation is an actual relation. In the
cases of the function and product types, the object-level
interpretation is just as set-theoretic function and product
respectively, and the relation-level interpretation uses the standard
logical relations interpretations of these type
constructors. Universal quantification, $\forall
\alpha\mathord:\kappa.~A$, is interpreted at the object level by
taking the dependent product over the objects of the interpretation of
$\kappa$ (this product exists because we have stipulated that $\kappa$
must be small), and then restricting to those elements of the
dependent product that preserve relations. This restriction is
required for this interpretation to preserve identity edges, and so be
a reflexive graph morphism. The relation-level interpretation of
universal quantification is the standard relational interpretation of
such types, albeit here generalised to kinds interpreted as arbitrary
reflexive graphs.

Well-kinded typing contexts $\Theta \vdash \Gamma$ are interpreted as
reflexive graph morphisms $\sem{\Gamma} : \sem{\Theta} \to \sem{*}$ by
taking the product of the interpretations of their constituent types,
similar to the interpretation of the product types $A \times B$.

We will extend basic System F$\omega$ with additional types and type
equalities in Sections~\ref{sec:discrete-kinds},
\ref{sec:groupoid-kinds}, and~\ref{sec:types-for-classical-mech},
below. These new types will also be assigned interpretations as
reflexive graph morphisms, and we will ensure that
Property~\ref{property:semantic-types} is maintained.

\paragraph{Interpretation of Terms} We omit the straightforward and
relatively uninteresting interpretation of well-typed terms $\Theta
\sepbar \Gamma \vdash e : A$, and just state that there is a
well-defined function interpreting each well-typed term, with the
property that it takes related environments to related results:

\begin{property}
  For all well-typed terms $\Theta \sepbar \Gamma \vdash e : A$ there
  is a function $\sem{e} \in (\forall \theta \in \semKU{\Theta}.\
  \sem{\Gamma}^f\theta \to \sem{A}^f\theta)$, such that, for all
  $\theta, \theta' \in \semKU{\Theta}$, $\rho \in
  \semKR{\Theta}(\theta,\theta')$, $\gamma \in \sem{\Gamma}^f\theta$
  and $\gamma' \in \sem{\Gamma}^f\theta'$, if $(\gamma, \gamma') \in
  \sem{\Gamma}^r\theta\theta'\rho$ then $(\sem{e}\theta\gamma,
  \sem{e}\theta'\gamma') \in \sem{A}^r\theta\theta'\rho$.  Moreover,
  this interpretation is sound for the $\beta\eta$ equational theory
  of terms.
\end{property}

\subsection{Discrete Kinds}
\label{sec:discrete-kinds}

In Section~\ref{sec:refl-graphs-for-fomega}, we only had a single base
kind: the kind $*$ of proper types.

Natural numbers as a discrete kind. The type of length-indexed
vectors.

\begin{displaymath}
  \kappa ::= \cdots \sepbar \mathsf{nat}
\end{displaymath}

\begin{displaymath}
  \inferrule*
  { }
  {\mathsf{nat}~\mathrm{small}}
\end{displaymath}

\begin{displaymath}
  \sem{\mathsf{nat}} = (\mathbb{N}, (n_1,n_2) \mapsto \{ * \sepbar n_1 = n_2 \}, n \mapsto *)
\end{displaymath}

Make the point that any set can be used as the interpretation of a
kind. Also, the utility of singleton types.

\subsection{Group(oid) Kinds}
\label{sec:groupoid-kinds}

There is a full and faithful functor from the category of groupoids to
the category of reflexive graphs, that preserves smallness.

\begin{enumerate}
\item Each (matrix) group gives a kind, which we can use as parameter
  to a type.
\item The collection of cartesian spaces, with diffeomorphisms as the
  edges.
\item Do units of measure as an example (include one of the results
  from that paper, e.g., type iso)
\item Mention groupoid kinds, and link to Hofmann and Streicher. Do
  finite sets and bijections as an example.
\end{enumerate}

\paragraph{Why not Groupoids Everywhere?} Explain why we aren't using
groupoids as kinds: don't need to, doesn't directly generalise
relational parametricity, see Robinson's paper. Distance-indexed
types.

\section{A Type System for Classical Mechanics}
\label{sec:types-for-classical-mech}

FIXME: rewrite this para to refer back concretely

We have a way of getting symmetry properties from types, and a way of
using symmetry properties to derive conservation laws for physical
systems. We now put the two together by building an applied version
System F$\omega$ with new kind-, type- and term-level constants that
will deliver us the exact symmetry properties we require to be able to
apply Noether's theorem.

\subsection{Extending System F$\omega$ for Classical Mechanics}

Let us enumerate our requirements for an extension of System F$\omega$
for writing invariant Lagrangians\footnote{Make sure that I say this
  is the plan back in Sec 2}.

\begin{enumerate}
\item Kind of cartesian spaces, parameterised by a matrix group and a
  translation group
\item Smooth functions between spaces: motivate this by looking at the
  requirements of the Euler-Lagrange equations.
\item Primitives for dealing with the geometry of the spaces (inner
  product, vector space operations, cross product(?), trigonometric
  functions (at least sin and cos)).\footnote{FIXME: fix the notation of $\forall$}
  \begin{displaymath}
    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
      \sin &:& \forall z \mathord: \mathsf{Z}.~\mathbb{R}\langle 2\pi \cdot z, 1 \rangle \smoothrightarrow \mathbb{R} \langle 0, 1 \rangle \\
      \cos &:& \forall z \mathord: \mathsf{Z}.~\mathbb{R}\langle 2\pi \cdot z, 1 \rangle \smoothrightarrow \mathbb{R} \langle 0, 1 \rangle \\
    \end{array}
  \end{displaymath}
\end{enumerate}

\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~\mathbb{R}\langle t, 1 \rangle \smoothrightarrow \mathbb{R}\langle 0, \mathit{exp}~t \rangle
  \end{array}
\end{displaymath}

Group homomorphisms:
\begin{itemize}
\item $\mathit{exp} : \mathsf{T}(1) \to \mathsf{GL}(1)$
\item $\mathit{scale}_n : \mathsf{GL}(1) \to \mathsf{GL}(n)$
\item $\mathit{ortho}_n : \mathsf{O}(n) \to \mathsf{GL}(n)$
\item $\underline{c}\cdot- : \mathsf{Z} \to \mathsf{T}(1)$
\end{itemize}
Group operations on translations (for all $n$), and on
$\mathsf{GL}(1)$ (no abs, because it is not smooth).

\begin{displaymath}
  \kappa ::= \cdots \sepbar \mathsf{CartSp} \sepbar \mathsf{T}(n) \sepbar \mathsf{GL}(n) \sepbar \mathsf{O}(n) \sepbar \mathsf{Z} \sepbar XXX
\end{displaymath}

\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    (\smoothrightarrow) &:& \mathsf{CartSp} \to \mathsf{CartSp} \to * \\
    \mathbb{R}^n &:& \mathsf{T}(n) \to \mathsf{GL}(n) \to \mathsf{CartSp} \\
    (\times) &:& \mathsf{CartSp} \to \mathsf{CartSp} \to \mathsf{CartSp}
  \end{array}
\end{displaymath}

\begin{itemize}
\item $\mathsf{CartSp}$ is the kind of cartesian spaces
  ($\mathbb{R}^n$, for some $n$) and diffeomorphisms between them,
  with the standard definition of diffeomorphism on these spaces. The
  identity edge is just the identity function.
\item $\mathbb{R}^n~t~g$ is the cartesian space with the
  diffeomorphism defined by the action of the translation and matrix
  group.
\item $X \times Y$ takes the product of two cartesian spaces, and the
  corresponding product of the diffeomorphisms.
\item $X \smoothrightarrow Y$ consists of the infinitely
  differentiable functions from $X$ to $Y$ that preserve the
  diffeomorphisms.
\end{itemize}

\begin{enumerate}
\item Need coercions between the different groups, as required by the
  examples.
\item Need operations on vectors, e.g., the usual linear space ones,
  and inner product. Perhaps also the cross product one.
\item What operations on the smooth function space do we need?
  composition, products, identities.
\end{enumerate}

\begin{figure*}[t]
  \centering
  \textbf{Administrative rules}
  \begin{mathpar}
    \inferrule*
    {x : \typeOfCartSp{X} \in \Gamma}
    {\Theta \sepbar \Gamma; \Delta \vdash x : X}

    \inferrule*
    {x : X \in \Delta}
    {\Theta \sepbar \Gamma; \Delta \vdash x : X}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : X \\
      \Theta \sepbar \Gamma; \Delta, x : X \vdash e_2 : Y}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{let}~x = e_1~\mathrm{in}~e_2 : Y}

    \inferrule*
    {\Theta \sepbar \Gamma \vdash e_1 : C^\infty(X, Y) \\\\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : X}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 (e_2) : Y}
  \end{mathpar}
  \textbf{Vector space operations, and dot product}
  \begin{mathpar}
    \inferrule*
    {\vec{x} \in \mathbb{R}^n}
    {\Theta \sepbar \Gamma; \Delta \vdash \vec{x} : \mathbb{R}^n\langle 1, 0 \rangle}

    \inferrule*
    {\Theta \vdash g : \mathsf{GL}(n)}
    {\Theta \sepbar \Gamma; \Delta \vdash 0 : \mathbb{R}^n\langle g, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle g, t_1 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle g, t_2 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 + e_2 : \mathbb{R}^n\langle g, t_1 + t_2 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle g, t_1 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle g, t_2 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 - e_2 : \mathbb{R}^n\langle g, t_1 - t_2 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}\langle g_1, 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle g_2, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 * e_2 : \mathbb{R}^n\langle \mathit{scale}_n(g_1)g_2, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle \mathit{scale}_n(s)\mathit{ortho}_n(o), 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle \mathit{scale}_n(s)\mathit{ortho}_n(o), 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 \cdot e_2 : \mathbb{R}\langle\mathit{scale}_n(s)^2, 0 \rangle}
  \end{mathpar}
  \textbf{Transcendental functions}
  \begin{mathpar}
    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, t \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{exp}~e : \mathbb{R}\langle \mathit{exp}~t, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 2\pi z, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{sin}~e : \mathbb{R}\langle 1, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 2\pi z, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{cos}~e : \mathbb{R}\langle 1, 0 \rangle}
  \end{mathpar}
  FIXME: products, division, n-ary products (sum, zip, reduce), constants
  \caption{Type system for smooth functions}
  \label{fig:smooth-terms}
\end{figure*}

\begin{theorem}
  If $\Theta \sepbar \Gamma; \Delta \vdash e : X$ then $\Theta \sepbar
  \Gamma \vdash \elab{e} : C^\infty(\elab{\Delta}, X)$.
\end{theorem}

\section{Examples}

$- \sepbar V : \forall ... . C^\infty(\mathbb{R}^3\langle o, t \rangle, \mathbb{R}\langle 1, 0 \rangle); t : \mathbb{R}\langle 1, t_t \rangle, q : \mathbb{R}^3\langle o, t_q \rangle, \dot{q} : \mathbb{R}^3\langle o, 0 \rangle \vdash \frac{1}{2} * m * (\dot{q} \cdot \dot{q}) - V(q) : \mathbb{R}\langle 1, 0 \rangle$

\begin{enumerate}
\item free particle
\item $2$-body
\item $n$-body (use length-indexed vectors)
\item pendulums
\item oscillators
\end{enumerate}

\section{Basic Field Theory}

\subsection{Principle of Stationary Action for Fields}

\subsection{Noether's Theorem for Fields}

\subsection{The Klein-Gordon Equation}

\section{Conclusions}

\begin{enumerate}
\item Gauge theory
\item Quantum field theory
\item Dependent types
\item Homotopy type theory
\item Distance indexed types and differential privacy.
\item Languages with function spaces with more structure (Alex's
  Enriched Effect Calculus) (Idea for differential privacy)
\end{enumerate}

\newpage

Overall plan:
\begin{enumerate}
\item A type system for writing smooth functions between smooth
  manifolds, subject to invariance with respect to certain Lie groups
\item Use this to write Lagrangians that describe classical mechanics
  situations, in such a way that Noether's theorem can be applied.
\item Need to explain:
  \begin{enumerate}
  \item Lagrangians
  \item Euler-Lagrange equations for $L(t,q,q')$
    \begin{displaymath}
      \frac{\mathrm{d}}{\mathrm{d}t} \frac{\partial L}{\partial q'} - \frac{\partial L}{\partial q} = 0
    \end{displaymath}
  \item Lie groups (manifolds!?), and their infinitesimal generators
    (see the exposition in Olver's book).
  \item Noether's theorem, with (lots of) examples. Explain what was
    in Noether's original paper (first theorem vs. second theorem).
  \end{enumerate}
  Interleave the presentation of the maths/physics and the type
  system?
\item Stick to being relatively uncomplicated with respect to the
  manifold stuff? Or use dependent types to handle the tangent spaces
  ``properly''?
\item Operations we want (in order to do the examples): distance
  between points (arclength/geodesic length), trigonometric functions
  (what continuous invariance properties do these have?),
\item Possibly use a variant of System F$\omega$ as the
  superstructure: System F$\omega$ plus a sub-language for dealing
  with smooth functions (which, in principle, could be handled just by
  using combinators to build something in a arrow-style). System
  F$\omega$ gives us a lot of power, and means that we don't need to
  go anywhere near dependent types...
  \begin{enumerate}
  \item So manifolds are a kind
  \item And so are Lie groups (?), not really, they have a different
    status (due to the identity extension requirement)
  \item Flat spaces are type operators from groups that act on them to
    manifolds
  \item Have a type constructor that builds the space of infinitely
    differentiable functions between two spaces
  \item Elaboration process that desugars a nicer syntax for smooth
    functions...
  \item Possibly a bit of natural numbers and vectors to allow for
    $n$-body problems...
  \item Need to restrict to predicativity in order to get classical
    mathematics? (probably should just forget about type
    quantification, and restrict the set of things we can quantify
    over to certain ``small'' kinds; just the groups?). Mention that
    going for SDG as a basis might offer a way out of this.
  \item Sub-kinding?
  \end{enumerate}
\item Examples:
  \begin{enumerate}
  \item $n$-body problem in $\mathbb{R}^3$, with all the consequent
    conservation laws.
  \item Central force problem
  \item Pendulums
  \item Rigid body problem? Use quotients in the definitions of the
    trigonometric functions? $\sin : \forall n :
    \mathbb{Z}.~\mathbb{R}\langle2\pi n\rangle \to \mathbb{R}\langle 1
    \rangle$. This gives us a way of doing circles in a neat way.
  \end{enumerate}
\end{enumerate}

\begin{enumerate}
\item Be clear that we don't really expect these programs to
  necessarily be (directly) executable.
\item Manifolds?
  \begin{enumerate}
  \item Spheres
  \item Tangent spaces
  \item 'Flat' spaces
  \item (Quotient spaces?) (have a look at Olver's book)
  \end{enumerate}
\end{enumerate}

Groups and their operations:
\begin{enumerate}
\item General linear group
\item Orthogonal group
\item Rotations around certain axes (in 
\end{enumerate}

Types (maybe include dependent types, for $n$-ary sized problems?)
\begin{displaymath}
  \begin{array}{rcl}
    M,M' & ::= & \mathbb{R}^n\langle e \rangle \sepbar \Sigma x : M.~M' \sepbar T_xM \\
    A, B & ::= & S \sepbar C^\infty(S,T) \sepbar A \times B \sepbar \forall G.~A \sepbar A \to B \sepbar A + B
  \end{array}
\end{displaymath}

\begin{mathpar}
  \inferrule*
  {\Gamma \sepbar x : S \vdash M :^\infty T}
  {\Gamma \vdash \lambda^\infty x : S.~M : C^\infty(S,T)}
\end{mathpar}

So a Lagrangian is something of type:
\begin{displaymath}
  \forall G:\mathsf{E}(3).~C^\infty(\mathbb{R} \times \mathbb{R}^3\langle G \rangle \times \mathbb{R}^3\langle G \rangle, \mathbb{R})
\end{displaymath}

\section{Introduction}

\section{Lagrangian Mechanics}

\section{Symmetry and Conservation Laws}

\section{A Type System for Lagrangians}

Syntax and Semantics, and the abstraction theorem...

\section{Examples}

\section{Field Theory}

\end{document}

### 2013-07-10 mechanics/paper.tex
Work on the introduction

@@
 
+\category{CR-number}{subcategory}{third-level}
+
+\terms
+term1, term2
+
+\keywords
+keyword1, keyword2
+
 \section{Introduction}
 
+Reynolds' theory of relational parametricity \cite{reynolds82} tells
+us that parametrically polymorphic programs automatically satisfy
+invariance properties. Such invariance properties are often called
+\emph{Free Theorems}, after Wadler \cite{wadler89theorems}, since they
+follow ``for free'' from the types of programs, rather than through
+detailed study of the program text itself. A classic example,
+presented by Wadler, is the free theorem for programs $f$ with the
+following type:
+\begin{displaymath}
+  f : \forall \alpha.~\mathsf{List}~\alpha \to \mathsf{List}~\alpha
+\end{displaymath}
+Such programs takes lists of $\alpha$s, for any type $\alpha$, to
+lists of $\alpha$s. Using Reynolds' theory of relational
+parametricity, Wadler showed that \emph{any} $f$ with this type
+satisfies the following property:
+\begin{displaymath}
+  \forall \alpha, \beta, g : \alpha \to \beta, l : \mathsf{List}~\alpha.~\mathrm{map}~g~(f~[\alpha]~l) = f~[\beta]~(\mathrm{map}~g~l)
+\end{displaymath}
+Thus, any $f$ with the type given above commutes with mapping some
+arbitrary function $g$ over lists. Thinking in terms of abstract data
+types and change of data representation, this free theorem states that
+$f$ is invariant under change of data representation from an arbitrary
+type $\alpha$ to another arbitrary type $\beta$, via $g$.
+
+Using invariance under change to derive useful properties is a concept
+much older than programming languages. In physics, Noether's theorem
+\cite{noether} provides a general way to derive conserved properties
+of physical systems from
+
+FIXME: example of translation invariance, with the coupled spring example
+\begin{equation}\label{eq:coupled-spring-langrangian}
+  L(t,q_1,q_2,\dot{q_1},\dot{q_2}) = \frac{1}{2}m(\dot{q_1}^2 + \dot{q_2}^2) - \frac{1}{2}k(q_1 - q_2)^2
+\end{equation}
+As a general rule, Lagrangians for classical mechanical systems take
+the form $L = T - V$, where $T$ is the kinetic energy of the system
+and $V$ is the potential energy. From the Lagrangian, using the
+principle of stationary action, we can derive the following two
+equations of motion for this system. (We describe the principle of
+stationary action, and the process for deriving the equations of
+motion fully in Section~\ref{sec:conservation-laws-from-symmetry}.)
+For this system, the equations of motion are a pair of ordinary
+differential equations (ODEs) that describe how the positions and
+velocities of the particles evolve over time:
+\begin{displaymath}
+  FIXME
+\end{displaymath}
+We could now proceed to solve these ODEs to further analyse the
+behaviour of this system. However, Noether's theorem gives us a
+powerful way of gaining insight into properties of these ODEs ``for
+free'', without necessarily having to find solutions to them.
+
+The Lagrangian (\ref{eq:coupled-spring-langrangian}) does not refer to
+any fixed point in space; only the relative distance between the two
+particles, along with their velocities, is relevant. Therefore, the
+Langrangian (\ref{eq:coupled-spring-langrangian}) is invariant under
+translation in space by some arbitrary displacement $x$:
+\begin{equation}\label{eq:coupled-spring-translation-invariance}
+  L(t,q_1,q_2,\dot{q_1},\dot{q_2}) = L(t,q_1+x,q_2+x,\dot{q_1},\dot{q_2})
+\end{equation}
+By Noether's theorem, invariance under spatial translation implies
+that the linear momentum of the whole system is constant for all
+time. For this system, conservation of linear momentum is stated
+mathematically like so:
+\begin{equation}\label{eq:coupled-spring-linear-momentum}
+  FIXME
+\end{equation}
+In general, Noether's theorem gives us a way of deriving conserved
+properties like (\ref{eq:coupled-spring-linear-momentum}) from
+invariance properties like
+(\ref{eq:coupled-spring-translation-invariance}). In this case, we
+have used invariance under translation in space to derive conservation
+of linear momentum. Other common examples include invariance under
+translation in time, yielding conservation of energy, and invariance
+under rotation, yield conservation of angular momentum. We will see
+examples of each of these kinds of invariance, and consequent
+conservation laws, in Section~\ref{sec:examples}.
+
+The invariance property stated in
+Equation~(\ref{eq:coupled-spring-translation-invariance}) is highly
+reminisicent of the
+
+In this paper, we present a way of using a generalised version of
+Reynolds' theory of relational parametricity to prove the geometric
+invariance properties required for the hypotheses of Noether's
+theorem. We build an extension of System F$\omega$ suitable for
+writing invariant Lagrangians. In our system, the Lagrangian
+(\ref{eq:coupled-spring-langrangian}) we gave above for describing a
+system of two particles coupled by a spring will have the type:
+\begin{displaymath}
+  \forall x \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, x \rangle \times \mathbb{R}\langle 1, x \rangle \times \mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
+\end{displaymath}
+where... FIXME
+
+Atkey, Johann and Kennedy \cite{atkey13abstraction} have already
+presented a polymorphic type system for expressing geometric
+invariance properties similar to the
+
+FIXME: area of triangle type
+
+By extending System F$\omega$, we get to
+
+\paragraph{Contributions}
+
+Our major contributions are threefold:
 \begin{enumerate}
-\item Symmetry in programming and physics
-  \begin{enumerate}
-  \item In programming: replacement of one data type by another
-  \item In physics: invariance under change of unobservable
-    properties. FIXME: check this.
-  \item What unifies these, at a conceptual level? Inability to
-    distinguish between situations, given the tools provided. In both
-    cases, this leads to interesting properties of such systems. In
-    programming, we get ``free theorems''. In Physics, via Noether's
-    theorem, we get ``conservation laws''.
-  \item In this paper, we show that we can define a programming
-    language for classical mechanics, based on System F$\omega$, where
-    the abstraction theorem directly fulfils the hypothesis of
-    Noether's theorem, allowing us to derive conservation laws.
-  \end{enumerate}
-\item Parametricity: polymorphic programs have symmetry properties
-  \begin{enumerate}
-  \item Example of a free theorem
-  \item Example of a geometric free theorem (the area of a triangle
-    example)
-  \end{enumerate}
-\item Noether's theorem: symmetry properties of descriptions of
-  systems lead to conservation laws
-\item Let's put the two together
-  \begin{enumerate}
-  \item Need invariance properties of smooth functions
-  \item So we extend System F$\omega$ with 
-  \end{enumerate}
-\item We present the classical mechanics and Noether's theorem first,
-  since it is likely to be more unfamiliar to readers.
+\item FIXME: reformulate AJK in terms of the reflexive graph model of
+  relational parametricity in order to unify that paper with the
+  natural extension of Reynolds' relational parametricity to indexed
+  types.
+\item FIXME: Present an extension of System F$\omega$ suitable for
+  writing invariant Lagrangians. This requires special attention to
+  ensure that all the functions that we can write are actually
+  continuous and differentiable. Hence we include a special type of
+  smooth, invariant functions between cartesian spaces.
+\item FIXME: Present many examples of 
 \end{enumerate}
 
-We restate the results of Atkey, Johann and Kennedy in the more
-general setting of System F$\omega$... which is good because it allows
-for a deeper interpretation of what the unit kinds are. Also, we can
-introduce useful stuff like nat-indexed types without too much fuss.
-
-\paragraph{Summary of the Paper}
+\paragraph{Outline}
 
@@
 
+FIXME: Define one-parameter groups
+
 \paragraph{Invariance of the Action} Let
@@
 \to \mathbb{R}$ that transforms time in some way and a function
-$\sigma : \mathbb{R} \times \mathbb{R}^n \to \mathbb{R}^n$ that
-transforms the vector of generalised coordinates $\vec{q}$ at time $t$
-to another vector of generalised coordinates $\sigma(t,\vec{q})$.
+$\sigma : \mathbb{R}^n \to \mathbb{R}^n$ that transforms the vector of
+generalised coordinates $\vec{q}$ to another vector of generalised
+coordinates $\sigma(\vec{q})$.
 
@@
 $a$, $b$ and $q$ that:
+\begin{equation}\label{eq:invariance}
+    \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
+    = \int_{\tau(a)}^{\tau(b)} L(s, q^*(s), \dot{q^*}(s)) \mathit{ds}
+\end{equation}
+where $q^*(s) = (\sigma \circ q \circ \tau^{-1})(s)$ is the path $q$
+transformed by $\tau$ and $\sigma$.  FIXME: Remark on the use of $t$
+and $s$ in the integrals. By change of variables, the right hand
+integral in the above equation is equal to the following integral:
 \begin{displaymath}
-  \int_a^b L(t,q,\dot{q}) \mathit{dt} = \int_{\tau(a)}^{\tau(b)} L(t', \sigma(\tau^{-1}(t'), q(t')), \mathbf{FIXME}) \mathit{dt'}
+  \int_a^b L(\tau(t), q^*(\tau(t)), \dot{q^*}(\tau(t)))\cdot \dot{\tau}(t) \mathit{dt}
 \end{displaymath}
-FIXME: Remark on the use of $t$ and $t'$ in the integrals. FIXME:
-state that all we are doing is transforming the time and the path by
-the transformations $\tau$ and $\sigma$ respectively. State why it
-will be useful for the
-
-Remark on how to give a simpler version of the invariance principle by
-applying some simple transformations of integrals (see notes).
+Since the endpoints $a$ and $b$ are arbitrary
 
@@
 small reflexive graphs is a key property of our semantics that we will
-maintain as we add additional kinds in Sections XXX below.
+maintain as we add additional kinds in
+Sections~\ref{sec:discrete-kinds} and~\ref{sec:groupoid-kinds},
+below. We state this as Property~\ref{property:semantic-kinds}:
 
@@
 In Section~\ref{sec:refl-graphs-for-fomega}, we only had a single base
-kind: the kind $*$ of proper types.
+kind: the kind $*$ of proper types, with a specific intepretation as
+the reflexive graph of sets and relations. We now describe two
+families of base kinds with interpretations that are particular sorts
+of reflexive graphs. In this section, we look at discrete kinds; kinds
+whose reflexive graph interpretations are such that the reflexive
+edges are the only edges between objects. In the following section
+(Section~\ref{sec:groupoid-kinds}), we look at groupoid kinds, where
+edges are composable and invertible.
 
-Natural numbers as a discrete kind. The type of length-indexed
-vectors.
+Discrete kinds can be seen as the natural way of lifting base types up
+to the kind level.
 
@@
 
+\begin{mathpar}
+  \inferrule*
+  { }
+  {\Theta \vdash \mathit{zero} : \mathsf{nat}}
+
+  \inferrule*
+  {\Theta \vdash A : \mathsf{nat}}
+  {\Theta \vdash \mathit{succ}~A : \mathsf{nat}}
+
+  \inferrule*
+  {\Theta \vdash A : \mathsf{nat} \\
+    \Theta \vdash B : \kappa \\
+    \Theta \vdash C : \kappa \to \kappa}
+  {\Theta \vdash \mathit{natrec}~A~B~C : \kappa}
+\end{mathpar}
+
 Make the point that any set can be used as the interpretation of a
@@
 
-\subsection{Group(oid) Kinds}
+\subsection{Groupoid Kinds}
 \label{sec:groupoid-kinds}
@@
 
-\begin{enumerate}
-\item Each (matrix) group gives a kind, which we can use as parameter
-  to a type.
-\item The collection of cartesian spaces, with diffeomorphisms as the
-  edges.
-\item Do units of measure as an example (include one of the results
-  from that paper, e.g., type iso)
-\item Mention groupoid kinds, and link to Hofmann and Streicher. Do
-  finite sets and bijections as an example.
-\end{enumerate}
+\paragraph{The additive group of integers}
 
-\paragraph{Why not Groupoids Everywhere?} Explain why we aren't using
-groupoids as kinds: don't need to, doesn't directly generalise
-relational parametricity, see Robinson's paper. Distance-indexed
-types.
+\paragraph{Matrix groups}
+
+GL and O, with group operations
+
+$\mathit{ortho}_n : \mathsf{O}(n) \to \mathsf{GL}(n)$
+$\mathit{scale}_n : \mathsf{GL}(1) \to \mathsf{GL}(n)$
+
+\paragraph{Translation groups}
+
+$\mathit{exp} : \mathsf{T}(1) \to \mathsf{GL}(1)$
+$\underline{c}\cdot- : \mathsf{Z} \to \mathsf{T}(1)$
+
+\paragraph{Groupoid of Cartesian Spaces}
+
+\subsection{Why not Groupoids Everywhere?}
+
+Explain why we aren't using groupoids as kinds: don't need to, doesn't
+directly generalise relational parametricity, see Robinson's
+paper. Distance-indexed types.
 
@@
 
-\subsection{Extending System F$\omega$ for Classical Mechanics}
+%\subsection{Extending System F$\omega$ for Classical Mechanics}
 
@@
   is the plan back in Sec 2}.
+
+\subsection{Extending System F$\omega$ with Smooth Functions}
+
+\begin{enumerate}
+\item We have already done the matrix kinds and the CartSp kind back
+  in the previous section. Add $n-ary$ products for CartSp.
+  \begin{displaymath}
+    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+      C^\infty &:& \mathsf{CartSp} \to \mathsf{CartSp} \to * \\
+      \mathbb{R}^n &:& \mathsf{GL}(n) \to \mathsf{T}(n) \to \mathsf{CartSp} \\
+%      (\times) &:& \mathsf{CartSp} \to \mathsf{CartSp} \to \mathsf{CartSp} \\
+      \mathit{vec} &:& \mathsf{nat} \to \mathsf{CartSp} \to \mathsf{CartSp}
+    \end{array}
+\end{displaymath}
+\item Note that the $n$ in $\mathbb{R}^n$ is static
+\item Add the type former for the smooth function space, and all the
+  relevant constants
+\item Give a (very) small example, and complain that it is very
+  difficult to write anything in this point-free style.
+\end{enumerate}
+
+Administrative things: composition, products
+
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    (+) &:&
+    \begin{array}[t]{@{}l}
+      \forall g \mathord: \mathsf{GL}(n), t_1,t_2 \mathord: \mathsf{T}(n). \\
+      \hspace{1.5cm} C^\infty(\mathbb{R}^n\langle g, t_1 \rangle \times \mathbb{R}^n\langle g, t_2 \rangle, \mathbb{R}^n\langle g, t_1 + t_2 \rangle)\\
+    \end{array} \\
+    (-) &:&
+    \begin{array}[t]{@{}l}
+      \forall g \mathord: \mathsf{GL}(n), t_1,t_2 \mathord: \mathsf{T}(n). \\
+      \hspace{1.5cm} C^\infty(\mathbb{R}^n\langle g, t_1 \rangle \times \mathbb{R}^n\langle g, t_2 \rangle, \mathbb{R}^n\langle g, t_1 - t_2 \rangle)\\
+    \end{array} \\
+    \sin &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, 2\pi \cdot z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
+    \cos &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, 2\pi \cdot z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
+    \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, t \rangle, \mathbb{R}\langle \mathit{exp}~t, 0 \rangle)
+  \end{array}
+\end{displaymath}
+
+\subsection{A Surface Syntax for Smooth Functions}
 
@@
   product, vector space operations, cross product(?), trigonometric
-  functions (at least sin and cos)).\footnote{FIXME: fix the notation of $\forall$}
-  \begin{displaymath}
-    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
-      \sin &:& \forall z \mathord: \mathsf{Z}.~\mathbb{R}\langle 2\pi \cdot z, 1 \rangle \smoothrightarrow \mathbb{R} \langle 0, 1 \rangle \\
-      \cos &:& \forall z \mathord: \mathsf{Z}.~\mathbb{R}\langle 2\pi \cdot z, 1 \rangle \smoothrightarrow \mathbb{R} \langle 0, 1 \rangle \\
-    \end{array}
-  \end{displaymath}
+  functions (at least sin and cos)).
 \end{enumerate}
@@
 \begin{displaymath}
-  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
-    \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~\mathbb{R}\langle t, 1 \rangle \smoothrightarrow \mathbb{R}\langle 0, \mathit{exp}~t \rangle
-  \end{array}
-\end{displaymath}
-
-Group homomorphisms:
-\begin{itemize}
-\item $\mathit{exp} : \mathsf{T}(1) \to \mathsf{GL}(1)$
-\item $\mathit{scale}_n : \mathsf{GL}(1) \to \mathsf{GL}(n)$
-\item $\mathit{ortho}_n : \mathsf{O}(n) \to \mathsf{GL}(n)$
-\item $\underline{c}\cdot- : \mathsf{Z} \to \mathsf{T}(1)$
-\end{itemize}
-Group operations on translations (for all $n$), and on
-$\mathsf{GL}(1)$ (no abs, because it is not smooth).
-
-\begin{displaymath}
-  \kappa ::= \cdots \sepbar \mathsf{CartSp} \sepbar \mathsf{T}(n) \sepbar \mathsf{GL}(n) \sepbar \mathsf{O}(n) \sepbar \mathsf{Z} \sepbar XXX
-\end{displaymath}
-
-\begin{displaymath}
-  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
-    (\smoothrightarrow) &:& \mathsf{CartSp} \to \mathsf{CartSp} \to * \\
-    \mathbb{R}^n &:& \mathsf{T}(n) \to \mathsf{GL}(n) \to \mathsf{CartSp} \\
-    (\times) &:& \mathsf{CartSp} \to \mathsf{CartSp} \to \mathsf{CartSp}
-  \end{array}
+  \kappa ::= \cdots \sepbar \mathsf{CartSp} \sepbar \mathsf{T}(n) \sepbar \mathsf{GL}(n) \sepbar \mathsf{O}(n) \sepbar \mathsf{Z}
 \end{displaymath}
@@
       \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle \mathit{scale}_n(s)\mathit{ortho}_n(o), 0 \rangle}
-    {\Theta \sepbar \Gamma; \Delta \vdash e_1 \cdot e_2 : \mathbb{R}\langle\mathit{scale}_n(s)^2, 0 \rangle}
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 \cdot e_2 : \mathbb{R}\langle\mathit{scale}_n(s)^n, 0 \rangle}
   \end{mathpar}
@@
 
-\newpage
-
-Overall plan:
-\begin{enumerate}
-\item A type system for writing smooth functions between smooth
-  manifolds, subject to invariance with respect to certain Lie groups
-\item Use this to write Lagrangians that describe classical mechanics
-  situations, in such a way that Noether's theorem can be applied.
-\item Need to explain:
-  \begin{enumerate}
-  \item Lagrangians
-  \item Euler-Lagrange equations for $L(t,q,q')$
-    \begin{displaymath}
-      \frac{\mathrm{d}}{\mathrm{d}t} \frac{\partial L}{\partial q'} - \frac{\partial L}{\partial q} = 0
-    \end{displaymath}
-  \item Lie groups (manifolds!?), and their infinitesimal generators
-    (see the exposition in Olver's book).
-  \item Noether's theorem, with (lots of) examples. Explain what was
-    in Noether's original paper (first theorem vs. second theorem).
-  \end{enumerate}
-  Interleave the presentation of the maths/physics and the type
-  system?
-\item Stick to being relatively uncomplicated with respect to the
-  manifold stuff? Or use dependent types to handle the tangent spaces
-  ``properly''?
-\item Operations we want (in order to do the examples): distance
-  between points (arclength/geodesic length), trigonometric functions
-  (what continuous invariance properties do these have?),
-\item Possibly use a variant of System F$\omega$ as the
-  superstructure: System F$\omega$ plus a sub-language for dealing
-  with smooth functions (which, in principle, could be handled just by
-  using combinators to build something in a arrow-style). System
-  F$\omega$ gives us a lot of power, and means that we don't need to
-  go anywhere near dependent types...
-  \begin{enumerate}
-  \item So manifolds are a kind
-  \item And so are Lie groups (?), not really, they have a different
-    status (due to the identity extension requirement)
-  \item Flat spaces are type operators from groups that act on them to
-    manifolds
-  \item Have a type constructor that builds the space of infinitely
-    differentiable functions between two spaces
-  \item Elaboration process that desugars a nicer syntax for smooth
-    functions...
-  \item Possibly a bit of natural numbers and vectors to allow for
-    $n$-body problems...
-  \item Need to restrict to predicativity in order to get classical
-    mathematics? (probably should just forget about type
-    quantification, and restrict the set of things we can quantify
-    over to certain ``small'' kinds; just the groups?). Mention that
-    going for SDG as a basis might offer a way out of this.
-  \item Sub-kinding?
-  \end{enumerate}
-\item Examples:
-  \begin{enumerate}
-  \item $n$-body problem in $\mathbb{R}^3$, with all the consequent
-    conservation laws.
-  \item Central force problem
-  \item Pendulums
-  \item Rigid body problem? Use quotients in the definitions of the
-    trigonometric functions? $\sin : \forall n :
-    \mathbb{Z}.~\mathbb{R}\langle2\pi n\rangle \to \mathbb{R}\langle 1
-    \rangle$. This gives us a way of doing circles in a neat way.
-  \end{enumerate}
-\end{enumerate}
-
-\begin{enumerate}
-\item Be clear that we don't really expect these programs to
-  necessarily be (directly) executable.
-\item Manifolds?
-  \begin{enumerate}
-  \item Spheres
-  \item Tangent spaces
-  \item 'Flat' spaces
-  \item (Quotient spaces?) (have a look at Olver's book)
-  \end{enumerate}
-\end{enumerate}
-
-Groups and their operations:
-\begin{enumerate}
-\item General linear group
-\item Orthogonal group
-\item Rotations around certain axes (in 
-\end{enumerate}
-
-Types (maybe include dependent types, for $n$-ary sized problems?)
-\begin{displaymath}
-  \begin{array}{rcl}
-    M,M' & ::= & \mathbb{R}^n\langle e \rangle \sepbar \Sigma x : M.~M' \sepbar T_xM \\
-    A, B & ::= & S \sepbar C^\infty(S,T) \sepbar A \times B \sepbar \forall G.~A \sepbar A \to B \sepbar A + B
-  \end{array}
-\end{displaymath}
-
-\begin{mathpar}
-  \inferrule*
-  {\Gamma \sepbar x : S \vdash M :^\infty T}
-  {\Gamma \vdash \lambda^\infty x : S.~M : C^\infty(S,T)}
-\end{mathpar}
-
-So a Lagrangian is something of type:
-\begin{displaymath}
-  \forall G:\mathsf{E}(3).~C^\infty(\mathbb{R} \times \mathbb{R}^3\langle G \rangle \times \mathbb{R}^3\langle G \rangle, \mathbb{R})
-\end{displaymath}
-
-\section{Introduction}
-
-\section{Lagrangian Mechanics}
-
-\section{Symmetry and Conservation Laws}
-
-\section{A Type System for Lagrangians}
-
-Syntax and Semantics, and the abstraction theorem...
-
-\section{Examples}
-
-\section{Field Theory}
-
 \end{document}

### 2013-07-10 mechanics/paper.tex
Work on the Lagrangian mechanics section

@@
 follow ``for free'' from the types of programs, rather than through
-detailed study of the program text itself. A classic example,
-presented by Wadler, is the free theorem for programs $f$ with the
-following type:
-\begin{displaymath}
-  f : \forall \alpha.~\mathsf{List}~\alpha \to \mathsf{List}~\alpha
+detailed study of the program text itself. An illustrative example is
+the free theorem for programs $f$ with the following type:
+\begin{displaymath}
+  f : \forall \alpha.~\mathsf{List}~\alpha \to \mathsf{Nat}
 \end{displaymath}
 Such programs takes lists of $\alpha$s, for any type $\alpha$, to
-lists of $\alpha$s. Using Reynolds' theory of relational
-parametricity, Wadler showed that \emph{any} $f$ with this type
-satisfies the following property:
-\begin{displaymath}
-  \forall \alpha, \beta, g : \alpha \to \beta, l : \mathsf{List}~\alpha.~\mathrm{map}~g~(f~[\alpha]~l) = f~[\beta]~(\mathrm{map}~g~l)
-\end{displaymath}
-Thus, any $f$ with the type given above commutes with mapping some
-arbitrary function $g$ over lists. Thinking in terms of abstract data
-types and change of data representation, this free theorem states that
-$f$ is invariant under change of data representation from an arbitrary
-type $\alpha$ to another arbitrary type $\beta$, via $g$.
-
-Using invariance under change to derive useful properties is a concept
-much older than programming languages. In physics, Noether's theorem
-\cite{noether} provides a general way to derive conserved properties
-of physical systems from
-
-FIXME: example of translation invariance, with the coupled spring example
-\begin{equation}\label{eq:coupled-spring-langrangian}
-  L(t,q_1,q_2,\dot{q_1},\dot{q_2}) = \frac{1}{2}m(\dot{q_1}^2 + \dot{q_2}^2) - \frac{1}{2}k(q_1 - q_2)^2
+natural numbers. Using Reynolds' theory of relational parametricity,
+Wadler showed that \emph{any} $f$ with this type satisfies the
+following property:
+\begin{displaymath}
+  \forall \alpha, \beta, g : \alpha \to \beta, l : \mathsf{List}~\alpha.~f~[\alpha]~l = f~[\beta]~(\mathrm{map}~g~l)
+\end{displaymath}
+Thus, any $f$ with the type given above is invariant under mapping
+some arbitrary function $g$ over its input. Thinking in terms of
+abstract data types and change of data representation, this free
+theorem states that $f$ is invariant under change of data
+representation from an arbitrary type $\alpha$ to another arbitrary
+type $\beta$, via $g$.
+
+The use of invariance under change to derive useful consequences is a
+technique much older than programming languages. In physics,
+\emph{Noether's theorem} \cite{noether} provides a general way to
+derive conservation laws for physical systems from invariance
+properties. Informally speaking, Noether's theorem states that every
+continuous symmetry of a physical system implies the existence of a
+corresponding conserved quantity. A conserved quantity is a function
+of the state of the system that is constant for all time. We
+illustrate the use of Noether's theorem with a simple example of a
+classical system consisting of two particles of equal mass $m$
+connected by a spring with coefficient $k$, such that the particles
+are constrained to move one-dimensionally.
+
+Noether's theorem applies to systems described in terms of
+\emph{Lagrangians}. For classical mechanics, a Lagrangian is a
+function of time, and the positions and velocities of all the
+particles in the system, to the difference between the kinetic energy
+and the potential energy of the system. For our system with two
+particles at positions $x_1$ and $x_2$ with velocities $\dot{x_1}$ and
+$\dot{x_2}$ the Lagrangian is:
+\begin{equation}\label{eq:coupled-spring-lagrangian}
+  L(t,x_1,x_2,\dot{x_1},\dot{x_2}) = \frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) - \frac{1}{2}k(x_1 - x_2)^2
 \end{equation}
-As a general rule, Lagrangians for classical mechanical systems take
-the form $L = T - V$, where $T$ is the kinetic energy of the system
-and $V$ is the potential energy. From the Lagrangian, using the
-principle of stationary action, we can derive the following two
+The first summand represents the total kinetic energy of the system in
+terms of the mass and velocities, while the second summand represents
+the potential energy contained within the spring due to the distance
+between the particles.
+
+From the Lagrangian (\ref{eq:coupled-spring-lagrangian}), using the
+\emph{principle of stationary action}, we can derive the following two
 equations of motion for this system. (We describe the principle of
@@
 differential equations (ODEs) that describe how the positions and
-velocities of the particles evolve over time:
-\begin{displaymath}
-  FIXME
-\end{displaymath}
+velocities of the particles evolve over time (the second deivative
+with respect to time, $\ddot{x_i}$, denotes the acceleration of the
+$i$th particle):
+\begin{mathpar}
+  m\ddot{x_1} = k(x_1 - x_2)
+
+  m\ddot{x_2} = k(x_1 - x_2)
+\end{mathpar}
 We could now proceed to solve these ODEs to further analyse the
@@
 
-The Lagrangian (\ref{eq:coupled-spring-langrangian}) does not refer to
+The Lagrangian (\ref{eq:coupled-spring-lagrangian}) does not refer to
 any fixed point in space; only the relative distance between the two
 particles, along with their velocities, is relevant. Therefore, the
-Langrangian (\ref{eq:coupled-spring-langrangian}) is invariant under
-translation in space by some arbitrary displacement $x$:
+Lagrangian (\ref{eq:coupled-spring-lagrangian}) is invariant under
+translation in space by some arbitrary displacement $y$:
 \begin{equation}\label{eq:coupled-spring-translation-invariance}
-  L(t,q_1,q_2,\dot{q_1},\dot{q_2}) = L(t,q_1+x,q_2+x,\dot{q_1},\dot{q_2})
+  L(t,x_1,x_2,\dot{x_1},\dot{x_2}) = L(t,x_1+y,x_2+y,\dot{x_1},\dot{x_2})
 \end{equation}
@@
 time. For this system, conservation of linear momentum is stated
-mathematically like so:
+mathematically as a vanishing derivative with respect to time:
 \begin{equation}\label{eq:coupled-spring-linear-momentum}
-  FIXME
+  \frac{d}{\mathit{dt}}~m(\dot{x}_1 + \dot{x_2}) = 0
 \end{equation}
@@
 translation in time, yielding conservation of energy, and invariance
-under rotation, yield conservation of angular momentum. We will see
-examples of each of these kinds of invariance, and consequent
+under rotation, yielding conservation of angular momentum. We will see
+examples of each of these kinds of invariance, and their consequent
 conservation laws, in Section~\ref{sec:examples}.
@@
 Equation~(\ref{eq:coupled-spring-translation-invariance}) is highly
-reminisicent of the
+reminiscent of the free theorem we stated for the polymorphic program
+at the start of this introduction. FIXME
 
@@
 writing invariant Lagrangians. In our system, the Lagrangian
-(\ref{eq:coupled-spring-langrangian}) we gave above for describing a
+(\ref{eq:coupled-spring-lagrangian}) we gave above for describing a
 system of two particles coupled by a spring will have the type:
@@
 
-\section{Conservation Laws from Symmetry}
+\section{Conservation Laws from Invariance}
 \label{sec:conservation-laws-from-symmetry}
@@
 
-\subsection{The Principle of Stationary Action}
+\subsection{Lagrangian Mechanics}
 
@@
 
-Two points: the Langrangian and generalised coordinates. Do
+Two points: the Lagrangian and generalised coordinates. Do
 generalised coordinates as a secondary thing.
@@
 \begin{example}
-  A single particle under a potential field. Set the potential field
-  to be (downward) gravity.
+  We have already seen an example of a Lagrangian, in the
+  introduction, describing a system of two particles coupled by a
+  spring. Another example of a Lagrangian describes a system
+  consisting of a single particle of mass $m$ acted upon by a constant
+  downward gravitational field with gravitational acceleration\footnote{is this the right word?} $g$:
+  \begin{equation}\label{eq:simple-g-lagrangian}
+    L(t,x,y,\dot{x},\dot{y}) = \frac{1}{2}m(\dot{x}^2 + \dot{y}^2) - mgy
+  \end{equation}
+  Again, the left summand represents the kinetic energy of the system
+  in terms of the mass and horizontal and vertical velocities of the
+  particle. The right summand represents the potential energy of the
+  system due to the gravitational field's strength determined by the
+  particle's distance from a fixed baseline at $y = 0$.
 \end{example}
@@
 
-The principle of stationary action
-
-\begin{displaymath}
+The \emph{principle of stationary action} states that the physically
+realisable paths $q$ are the paths that yield stationary points (i.e.,
+either minima or maxima) of the action $\mathcal{S}[q;a;b]$, for all
+$a$ and $b$. Defining what is exactly meant by a stationary point of
+the action is beyond the scope of this paper, and can be found in
+standard references on the calculus of variations, such as Gelfand and
+Fomin \cite{gelfandXXcalculus}. The key point for our purposes is that
+a path $q$ satisfies the principle of stationary action if and only if
+it satisfies a system of ODEs derived from the Lagrangian $L$, called
+the Euler-Lagrange equations:
+\begin{equation}\label{eq:euler-lagrange}
   \frac{d}{dt}\frac{\partial L}{\partial \dot{q_i}} - \frac{\partial L}{\partial q_i} = 0
-\end{displaymath}
-
-Explain the different derivatives going on here
-
-Example: Newton's second law...
-
-\subsection{Noether's Theorem} Noether's theorem provides us with deep
-insights into the properties of the solutions of the Euler-Lagrange
-equations. FIXME: motivate invariance.
-
-FIXME: Define one-parameter groups
+\end{equation}
+The ODEs derived from the Lagrangian for a classical mechanical system
+are called the \emph{equations of motion} for the system.
+
+The notation in the Euler-Lagrange equation, while standard, is not
+exactly clear, especially to a programming language theorist used to a
+more careful treatment of variables. The two partial derivatives
+$\frac{\partial L}{\partial \dot{q_i}}$ and $\frac{\partial
+  L}{\partial q_i}$ take partial derivatives of $L$ with respect to
+$\dot{q_i}$ and $q_i$ \emph{as variables}. The derivative with respect
+to time, $\frac{d}{dt}$ then treats $\dot{q_i}$ as a function of time,
+hence the appearance of the second derivatives $\ddot{q_i}$ in the
+equations of motion. Sussman and XXX \cite{sussman} criticise the
+standard presentation of Lagrangian mechanics for this kind of
+confusing notation, and define a clearer notation based on the Scheme
+programming language. We elect to stick with the standard notation as
+used in physics textbooks just for reasons of familiarity.
+
+\begin{example}
+  Computing the Euler-Lagrange equations for the Lagrangian
+  (\ref{eq:simple-g-lagrangian}), above, yields a pair of ODEs, for
+  the $x$ and $y$ coordinates of the single particle of the system:
+  \begin{mathpar}
+    m\ddot{x} = 0
+
+    m\ddot{y} = mg
+  \end{mathpar}
+  Note that both these ODEs have the form $F = m\ddot{x}$, which is
+  exactly Newton's second law. Lagrangian mechanics allows, from the
+  principle of stationary action, the derivation of Newton's second
+  law, rather than postulating it as an axiom.
+\end{example}
+
+\subsection{Noether's Theorem}
+
+As we described in the introduction, Noether's theorem provides us
+with deep insights into the properties of the solutions of the
+Euler-Lagrange equations. Formally, Noether's theorem requires the
+invariance of the action $\mathcal{S}$, rather than invariance of the
+Lagrangian itself. We therefore first define what is meant by
+invariance of the action, and then state Noether's theorem.
 
@@
 be the action of some physical system described by the Lagrangian
-$L$. Assume a differentiable invertible function $\tau : \mathbb{R}
+$L$. Assume a differentiable invertible function $\Phi : \mathbb{R}
 \to \mathbb{R}$ that transforms time in some way and a function
-$\sigma : \mathbb{R}^n \to \mathbb{R}^n$ that transforms the vector of
+$\Psi : \mathbb{R}^n \to \mathbb{R}^n$ that transforms the vector of
 generalised coordinates $\vec{q}$ to another vector of generalised
-coordinates $\sigma(\vec{q})$.
-
-We say that the action $\mathcal{S}$ is invariant under the
-transformations $\tau$ and $\sigma$ if it is the case that for all
-$a$, $b$ and $q$ that:
+coordinates $\Psi(\vec{q})$.
+
+The action $\mathcal{S}$ is \emph{invariant} between the endpoints $a$
+and $b$, under the transformations $\Phi$ and $\Psi$, if it is the
+case that for all paths $q$:
 \begin{equation}\label{eq:invariance}
     \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
-    = \int_{\tau(a)}^{\tau(b)} L(s, q^*(s), \dot{q^*}(s)) \mathit{ds}
+    = \int_{\Phi(a)}^{\Phi(b)} L(s, q^*(s), \dot{q^*}(s)) \mathit{ds}
 \end{equation}
-where $q^*(s) = (\sigma \circ q \circ \tau^{-1})(s)$ is the path $q$
-transformed by $\tau$ and $\sigma$.  FIXME: Remark on the use of $t$
-and $s$ in the integrals. By change of variables, the right hand
-integral in the above equation is equal to the following integral:
-\begin{displaymath}
-  \int_a^b L(\tau(t), q^*(\tau(t)), \dot{q^*}(\tau(t)))\cdot \dot{\tau}(t) \mathit{dt}
-\end{displaymath}
-Since the endpoints $a$ and $b$ are arbitrary
+where $q^*(s) = (\Psi \circ q \circ \Phi^{-1})(s)$ is the path $q$
+transformed by $\Phi$ and $\Psi$.  By change of variables, the right
+hand integral in the above equation is equal to the following
+integral:
+\begin{displaymath}
+  \int_a^b L(\Phi(t), q^*(\Phi(t)), \dot{q^*}(\Phi(t)))\cdot \dot{\Phi}(t) \mathit{dt}
+\end{displaymath}
+For our application to classical mechanics, the endpoints $a$ and $b$
+are arbitrary, so we can rewrite Equation (\ref{eq:invariance}) to the
+following equation between applications of the Lagrangian:
+\begin{equation}\label{eq:simplified-invariance}
+  L(t,q(t),\dot{q}(t)) = L(\Phi(t), q^*(\Phi(t)), \dot{q^*}(\Phi(t)))\cdot \dot{\Phi}(t)
+\end{equation}
+For most examples, it will be the case that $\Phi$ has the form
+$\Phi(t) = t + t'$ and $\Psi$ is an affine transformation $\Psi(q)
+= Gq + x$, where $G$ is an $n\mathord\times n$ real matrix and $x \in
+\mathbb{R}^n$. In this case, Equation (\ref{eq:simplified-invariance})
+simplifies further, using the definition of $q^*$ and the chain rule
+of differentiation, to:
+\begin{equation}
+  \label{eq:simple-invariance}
+  L(t,q(t),\dot{q}(t)) = L(t + t', Gq(t) + x, G\dot{q}(t))
+\end{equation}
+Equation (\ref{eq:simple-invariance}) is obviously implied by the
+following equation, where we treat $q$ and $\dot{q}$ as universally
+quantified variables, instead of as functions of time:
+\begin{equation}
+  \label{eq:simplest-invariance}
+  L(t,q,\dot{q}) = L(t + t', Gq + x, G\dot{q})
+\end{equation}
+Equation (\ref{eq:simplest-invariance}) is the form of invariance that
+the free theorems we derive from the types of our typed language for
+classical mechanics will usually take. We define our type system for
+classical mechanics in Sections~\ref{sec:symmetry-from-types}
+and~\ref{sec:types-for-classical-mech} below.
+
+\begin{example}\label{ex:invariance-examples}
+  In the introduction, we stated that the Lagrangian
+  (\ref{eq:coupled-spring-lagrangian}) describing the spring-coupled
+  system is invariant under translation in space. Our running example
+  in this section, the Lagrangian (\ref{eq:simple-g-lagrangian}) for
+  the single particle under gravity, is not invariant under all
+  spatial translations, due to the reference in the potential energy
+  term to the distance from the fixed baseline at $y = 0$. It is
+  invariant under translations along the $x$-axis, however. Both
+  Lagrangians are also invariant in translations in time, a special
+  case of Equation~(\ref{eq:simplest-invariance}):
+  \begin{displaymath}
+    L(t,q,\dot{q}) = L(t+t',q,\dot{q})
+  \end{displaymath}
+  This invariance property is simply a consequence of the fact that
+  neither Lagrangian explicitly mentions the time $t$.
+\end{example}
+
+\paragraph{Noether's Theorem} Noether's theorem applies to actions
+that are \emph{continuously} invariant. Being continuously invariant
+means that, instead of having fixed transformations $\Phi$ and $\Psi$,
+we have two families of transformations, $\Phi_\epsilon$ and
+$\Psi_\epsilon$, differentiably indexed by a real parameter
+$\epsilon$, such that $\Phi_0$ and $\Psi_0$ are the identity function.
+
+\begin{theorem}[Noether]
+  If the action
+  \begin{displaymath}
+    \mathcal{S}[q;a;b] = \int_a^b L(t,q(t),\dot{q}(t))~dt
+  \end{displaymath}
+  is invariant under $\epsilon$-indexed families of transformations
+  $\Phi_\epsilon$ and $\Psi_\epsilon$, then the equation
+  \begin{equation}
+    \label{eq:conservation-law}
+    \frac{d}{dt}\left(\sum_{i=1}^n \frac{\partial L}{\partial \dot{q_i}}\psi_i + \left(L - \sum_{i=1}^n \dot{q_i}\frac{\partial L}{\partial \dot{q_i}}\right) \phi \right) = 0
+  \end{equation}
+  holds for all paths $q$ satisfying the Euler-Lagrange equations
+  (\ref{eq:euler-lagrange}), where $\psi = \left. \frac{\partial
+      \Psi}{\partial \epsilon} \right|_{\epsilon = 0}$ and $\phi =
+  \left. \frac{\partial \Phi}{\partial \epsilon} \right|_{\epsilon =
+    0}$.
+\end{theorem}
+
+\begin{proof}
+  E.g., Gelfand and Fomin \cite{gelfandXXcalculus}, Section 20.
+\end{proof}
 
 \begin{example}
-  Continue a simple example from above to show that its action is
-  invariant under (e.g.) time translation.
+  In Example~\ref{ex:invariance-examples}, we noted that the
+  Lagrangians (\ref{eq:coupled-spring-lagrangian}) and
+  (\ref{eq:simple-g-lagrangian}) are both invariant under arbitrary
+  translations in time. In terms of continuous invariance, their
+  corresponding actions are invariant under the families
+  $\Phi_\epsilon(t) = t + \epsilon$ and $\Psi_\epsilon(q) =
+  q$. Plugging these definitions into
+  Equation~(\ref{eq:conservation-law}), we learn that, for these
+  systems:
+  \begin{displaymath}
+    \frac{d}{dt}~\left(L - \sum_{i=1}^2 \dot{q_i}\frac{\partial L}{\partial \dot{q_i}} \right) = 0
+  \end{displaymath}
+  For the spring-coupled particles Lagrangian,
+  (\ref{eq:coupled-spring-lagrangian}), we obtain:
+  \begin{displaymath}
+    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) + \frac{1}{2}k(x_1 - x_2) \right) = 0
+  \end{displaymath}
+  and for the single particle under gravity Lagrangian
+  (\ref{eq:simple-g-lagrangian}), we obtain:
+  \begin{displaymath}
+    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x}^2 + \dot{y}^2) + mgy\right) = 0
+  \end{displaymath}
+  In both cases, we have discovered that, as a consequence of
+  invariance under translation in time, the total energy of the system
+  is conserved (note that the $-$ signs in the original Lagrangians
+  have turned into $+$s). In general, invariance under translation in
+  time implies, via Noether's theorem, conservation of energy.
 \end{example}
 
-\paragraph{Noether's Theorem} Noether's theorem applies to actions
-that are \emph{continuously} invariant. FIXME: not quite right: write
-out the maths first.
-
-(a) parameterise $\tau$ and $\sigma$ by a real value $\epsilon$ such
-that we get a continuous symmetry of the action; (b) state Noether's
-theorem as we will use it.
-
-\begin{theorem}
-  Noether's theorem
-\end{theorem}
-
-\begin{example}
-  Continue the simple example from above to show that the invariance
-  under time translation yields conservation of energy.
-\end{example}
-
-State that we will do more examples later on, after we discuss how to
-derive symmetry from types.
-
-\section{Symmetry from Types}
+This finishes our short introduction to Lagrangian mechanics and
+Noether's theorem. FIXME: State that we will do more examples later
+on, after we discuss how to derive invariance properties from types.
+
+\section{Invariance from Types}
+\label{sec:symmetry-from-types}
 
@@
 
-\section{Examples}
+\section{Examples}\label{sec:examples}
 

### 2013-07-11 mechanics/paper.tex
Introduction, and Section 2

@@
 powerful way of gaining insight into properties of these ODEs ``for
-free'', without necessarily having to find solutions to them.
+free'', without necessarily having to find solutions to them. This is
+accomplished by finding invariance properties of the Lagrangian.
 
@@
 time. For this system, conservation of linear momentum is stated
-mathematically as a vanishing derivative with respect to time:
+mathematically as the vanishing of the derivative with respect to time
+of the total linear momentum of the system:
 \begin{equation}\label{eq:coupled-spring-linear-momentum}
@@
 reminiscent of the free theorem we stated for the polymorphic program
-at the start of this introduction. FIXME
-
-In this paper, we present a way of using a generalised version of
-Reynolds' theory of relational parametricity to prove the geometric
-invariance properties required for the hypotheses of Noether's
-theorem. We build an extension of System F$\omega$ suitable for
-writing invariant Lagrangians. In our system, the Lagrangian
-(\ref{eq:coupled-spring-lagrangian}) we gave above for describing a
-system of two particles coupled by a spring will have the type:
-\begin{displaymath}
-  \forall x \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, x \rangle \times \mathbb{R}\langle 1, x \rangle \times \mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
-\end{displaymath}
-where... FIXME
-
-Atkey, Johann and Kennedy \cite{atkey13abstraction} have already
-presented a polymorphic type system for expressing geometric
-invariance properties similar to the
-
-FIXME: area of triangle type
-
-By extending System F$\omega$, we get to
+at the start of this introduction. A change in representation in the
+input, whether a change in data representation from a type $\alpha$ to
+a type $\beta$, or a change in where time zero is counted from,
+results in no change in the output. Since Reynolds' relational
+parametricity was allowed us to derive the invariance properties for
+the program $f$ just by looking at its type, we ask the following
+question:
+
+\medskip
+
+\noindent
+\textbf{Question:} \emph{Is it possible to use Reynolds' theory of
+  relational parametricity to derive the invariance properties of
+  Lagrangians required for Noether's theorem to derive conservation
+  laws?}
+
+\medskip
+
+In this paper, we answer this question positively. We show that it is
+indeed the case that we can use a generalised version of Reynolds'
+theory of relational parametricity to prove the geometric invariance
+properties required for the hypotheses of Noether's theorem. We
+construct an extension of System F$\omega$ suitable for writing
+invariant Lagrangians, with a relationally parametric model that
+allows us to prove invariance properties as free theorems. In our
+system, the Lagrangian (\ref{eq:coupled-spring-lagrangian}) we gave
+above for describing a system of two particles coupled by a spring
+will have the type:
+\begin{displaymath}
+  \begin{array}{@{}l}
+  \forall y \mathord: \mathsf{T}(1).\\
+  \quad C^\infty(\mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, y \rangle \times \mathbb{R}\langle 1, y \rangle \times \mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
+\end{array}
+\end{displaymath}
+We explain this type by breaking it down from left to right. The
+quantifier $\forall x \mathord: \mathsf{T}(1)$ indicates that we are
+quantifying over all translations $y$ in one-dimensional space, just
+as we did implicity in Equation
+(\ref{eq:coupled-spring-translation-invariance}). The notation
+$C^\infty(-,-)$ denotes the type of \emph{smooth} functions between
+spaces. A smooth function is a function that is infinitely
+differentiable. All the functions that we wish to use as Lagrangians
+must be smooth (at least up to second order) for the theory to
+work. The type $\mathbb{R}\langle g, x \rangle$ denotes real numbers
+that vary with some linear transformation $g$ and translation $x$. In
+this case, all the linear transformations are $1$, indicating no
+transformation. Likewise, when the translation component is $0$, there
+is no translation. Thus a value of type $\mathbb{R}\langle 1, 0
+\rangle$ is invariant, but a value of type $\mathbb{R}\langle 1, y
+\rangle$ varies with the translation $y$. In the smooth function type
+above, in domain space consists of vectors of five real numbers: the
+time, the two particle positions, which vary with the translation $y$,
+and the two particle velocities.
+
+Atkey, Johann and Kennedy \cite{atkey13abstraction}, building on older
+work by Kennedy \cite{kennedy94para}, have already presented a
+polymorphic type system for expressing geometric invariance properties
+similar to the translation invariance property in Equation
+(\ref{eq:coupled-spring-translation-invariance}). In the present work,
+we extend their type system by embedding it within System
+F$\omega$. The benefits of this embedding are twofold. Firstly, by
+using the richer type-level structure of System F$\omega$, we can
+easily add useful indexed types like length indexed vectors and smooth
+function types, both of which would have required special treatment in
+the special purpose type system presented by Atkey \emph{et
+  al.}. Secondly, by using System F$\omega$ as our base, we are able
+to reuse the reflexive graph-based relationally parametric semantics
+presented by Atkey \cite{atkey12relational}, extending it as
+appropriate for our geometric setting.
 
@@
 \begin{enumerate}
-\item FIXME: reformulate AJK in terms of the reflexive graph model of
-  relational parametricity in order to unify that paper with the
-  natural extension of Reynolds' relational parametricity to indexed
-  types.
-\item FIXME: Present an extension of System F$\omega$ suitable for
-  writing invariant Lagrangians. This requires special attention to
-  ensure that all the functions that we can write are actually
-  continuous and differentiable. Hence we include a special type of
-  smooth, invariant functions between cartesian spaces.
-\item FIXME: Present many examples of 
+\item We reformulate the type system for geometric invariance of
+  Atkey, Johann and Kennedy \cite{atkey13abstraction} as an extension
+  of System F$\omega$. This shows that the special status of the
+  group-indexed types in Atkey \emph{et al.}'s work can be
+  incorporated into a standard framework for presenting indexed types.
+\item We present a relationally parametric semantics of our extension
+  of System F$\omega$. In this model, following Atkey's model for
+  basic System F$\omega$ \cite{atkey12relational}, each kind is
+  interpreted as a reflexive graph. Our key technical contribution is
+  to note that the many geometric groups that we wish to incorporate
+  are expressible as groupoids, and hence as reflexive graphs. We
+  therefore show that Reynolds' theory of relational parametricity,
+  suitably generalised to indexed types by considering reflexive
+  graphs, already accomodates geometric invariance properties.
+\item Finally, we connect the free theorems derivable from our
+  relationally parametric semantics with Noether's theorem for
+  deriving conservation laws. We present many examples of invariant
+  Lagrangians expressible in our type system, along with the
+  consequent conservation laws for physical systems. In each case, the
+  necessary invariance property is derived from the type we assign to
+  the Lagrangian.
 \end{enumerate}
@@
 \begin{itemize}
-\item FIXME: Recall the necessary background for Noether's theorem
-  (Section~\ref{sec:conservation-laws-from-symmetry}).
-\item FIXME: Show how to get symmetry properties from types
-  (Section~\ref{sec:symmetry-from-types}).
-\item FIXME: In Section~\ref{sec:types-for-classical-mech}, we
-  construct a type system for classical mechanics,
-\item FIXME: Section~\ref{sec:examples} we present examples of using
-  our extension of System F$\omega$ for writing invariant Lagrangians
-  that describe physical systems.
-\item FIXME: field theory
+\item In the next section,
+  Section~\ref{sec:conservation-laws-from-symmetry}, we describe the
+  necessary background to understand Noether's theorem and the
+  conservation laws it generates. We briefly introduce the Lagrangian
+  formulation of classical mechanics, define precisely what it means
+  for a physical system to be invariant under transformations, and
+  state Noether's theorem.
+\item Having shown how invariance generates conservation laws in
+  Section~\ref{sec:conservation-laws-from-symmetry}, in
+  Section~\ref{sec:symmetry-from-types}, we show how to derive
+  invariance properties from indexed types. We present a relationally
+  parametric semantics for System F$\omega$, and show how geometric
+  invariance properties can be accomodated within this semantics.
+\item In Section~\ref{sec:types-for-classical-mech}, we construct a
+  type system for classical mechanics, by extending System F$\omega$
+  with the necessary constants for smooth function types and
+  combinators for constructing smooth functions. Programming directly
+  with our combinators in raw System F$\omega$ is awkward, due to the
+  point-free style required, so in Section~\ref{sec:surface-syntax} we
+  define a surface syntax for defining smooth functions that is
+  translated into our applied System F$\omega$.
+\item In Section~\ref{sec:examples} we present several examples of
+  using our extension of System F$\omega$ for writing invariant
+  Lagrangians that describe physical systems.
+\item Section~\ref{sec:conclusions} concludes, and offers directions
+  for further work.
 \end{itemize}
@@
 
-Make sure to state up front the necessary background required. To
-understand: basic(ish) calculus. We will be using the Calculus of
-Variations, but only by stating the main results. We go for standard
-physics notation, rather than anything more complicated.
+Noether's theorem applies to systems described using the formalism of
+Lagrangian mechanics, a reformulation of Newtonian classical
+mechanics. In this section we can only offer a very brief introduction
+to the concepts of Lagrangian mechanics. The reader is referred to
+other sources, such as Landau and Lifschitz \cite{landau60mechanics}
+or Arnol'd \cite{arnoldXXmathematical} for the necessary background in
+the mathematical theory of classical mechanics. Lagrangian mechanics,
+and Noether's theorem, make crucial use of the Calculus of
+Variations. Gelfand and Fomin \cite{gelfandXXcalculus} provide a good
+introduction to this topic.
+
+In this section, we will endeavour to keep the presentation relatively
+elementary. However, some knowledge of differential and integral
+calculus will be required. Knowledge of basic Newtonian mechanics will
+also be helpful. We will use standard notation from physics and
+mathematics for integrals and (partial) derivatives. In particular,
+when discussing derivatives with respect to time, we use dot notation:
+$\dot{x}$ for the first derivative of $x$ with respect to time, and
+$\ddot{x}$ for the second derivative.
 
@@
 
-Introduce by stating Newton's second law, and saying that it can be
-derived from the principle of stationary action.
-
-Analytical mechanics, Lagrangian and Hamiltonian mechanics. Basically
-different mathematical reformulations of Newtonian mechanics that
-yield new insights into the underlying structure of the kinds of
-physical systems studied in classical mechanics.
+Lagrangian mechanics is a branch of analytical mechanics, along with
+Hamiltonian mechanics
+\cite{landau60mechanics,arnoldXXmathematical}. Analytical mechanics
+seeks to discover the underlying structure of Newtonian classical
+mechanics by reformulating it in different terms.
 
@@
 
-Lagrangian mechanics reformulates Newtonian mechanics in terms of
-
-Two points: the Lagrangian and generalised coordinates. Do
-generalised coordinates as a secondary thing.
-
+For Lagrangian mechanics, classical mechanical systems are described
+in terms of \emph{Lagrangians} $L = T - V$, where $T$ is the total
+kinetic energy of the system, and $V$ is the total potential
+energy. The Lagrangian is a function of time, the positions of all the
+particles in the system, and their velocities. This is usually written
+like so, with abuse of notation for $\dot{q}$ which is a variable
+here, not a derivative of a function:
+\begin{displaymath}
+  L(t, q, \dot{q}) = T - V
+\end{displaymath}
+In this expression, $q$ and $\dot{q}$ are actually vectors containing
+the components of the position and velocity vectors for each
+particle. A key feature of Lagrangian mechanics is that the coordinate
+system used to describe the system need not necessarily be the usual
+cartesian coordinates. For example, one of the components of $q$ might
+be the angle of swing of a pendulum. This use of generalised
+coordinates is one of the advantages of Lagrangian mechanics over the
+Newtonian presentation in terms of positions and forces.
+
+The Lagrangian for a particular system is used to define the
+\emph{action} of that system. For a particular path $q : \mathbb{R}
+\to \mathbb{R}^n$ and endpoints $a < b$ in time, the action of the
+Lagrangian $L$ over this path is defined by the following integral:
 \begin{displaymath}
@@
 \end{displaymath}
+Below, we will use the principle of stationary action to select the
+physically realisable paths from all the possible paths $q :
+\mathbb{R} \to \mathbb{R}^n$.
 
@@
   \begin{displaymath}
-    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) + \frac{1}{2}k(x_1 - x_2) \right) = 0
+    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) + \frac{1}{2}k(x_1 - x_2)^2 \right) = 0
   \end{displaymath}
@@
 This finishes our short introduction to Lagrangian mechanics and
-Noether's theorem. FIXME: State that we will do more examples later
-on, after we discuss how to derive invariance properties from types.
+Noether's theorem. We have shown how Noether's theorem can be used to
+derive conservation laws from invariance properties. We now go on to
+show how suitable invariance properties can be derived from types. In
+Section~\ref{sec:examples}, we give examples of invariant Lagrangian's
+and their consequent conservation laws.
 
@@
 \subsection{A Surface Syntax for Smooth Functions}
+\label{sec:surface-syntax}
 
@@
 
-\section{Basic Field Theory}
-
-\subsection{Principle of Stationary Action for Fields}
-
-\subsection{Noether's Theorem for Fields}
-
-\subsection{The Klein-Gordon Equation}
+% \section{Basic Field Theory}
+
+% \subsection{Principle of Stationary Action for Fields}
+
+% \subsection{Noether's Theorem for Fields}
+
+% \subsection{The Klein-Gordon Equation}
 
 \section{Conclusions}
+\label{sec:conclusions}
 

### 2013-07-11 mechanics/paper.tex
Nearly there

@@
 
+\newcommand{\comp}{\mathrel{>\kern-0.45em>\kern-0.45em>}}
+
 \newcommand{\smoothrightarrow}{\stackrel{C^\infty}\longrightarrow}
@@
 classical system consisting of two particles of equal mass $m$
-connected by a spring with coefficient $k$, such that the particles
-are constrained to move one-dimensionally.
+connected by a spring with spring constant $k$, such that the
+particles are constrained to move one-dimensionally.
+
+% Picture here?
 
@@
 \begin{mathpar}
-  m\ddot{x_1} = k(x_1 - x_2)
-
-  m\ddot{x_2} = k(x_1 - x_2)
+  m\ddot{x_1} = - k(x_1 - x_2)
+
+  m\ddot{x_2} = - k(x_2 - x_1)
 \end{mathpar}
@@
 $C^\infty(-,-)$ denotes the type of \emph{smooth} functions between
-spaces. A smooth function is a function that is infinitely
-differentiable. All the functions that we wish to use as Lagrangians
+spaces. (A smooth function is a function that is infinitely
+differentiable.) All the functions that we wish to use as Lagrangians
 must be smooth (at least up to second order) for the theory to
@@
 
-Our major contributions are threefold:
+Our core contributions are threefold:
 \begin{enumerate}
@@
   consisting of a single particle of mass $m$ acted upon by a constant
-  downward gravitational field with gravitational acceleration\footnote{is this the right word?} $g$:
+  downward gravitational field with gravitational potential $g$:
   \begin{equation}\label{eq:simple-g-lagrangian}
@@
 hence the appearance of the second derivatives $\ddot{q_i}$ in the
-equations of motion. Sussman and XXX \cite{sussman} criticise the
+equations of motion. Sussman and Wisdom \cite{sussman} criticise the
 standard presentation of Lagrangian mechanics for this kind of
@@
 programming language. We elect to stick with the standard notation as
-used in physics textbooks just for reasons of familiarity.
+used in physics textbooks just for familiarity.
 
@@
   Computing the Euler-Lagrange equations for the Lagrangian
-  (\ref{eq:simple-g-lagrangian}), above, yields a pair of ODEs, for
-  the $x$ and $y$ coordinates of the single particle of the system:
+  (\ref{eq:simple-g-lagrangian}), above, yields a pair of ODEs, one
+  each for the $x$ and $y$ coordinates of the single particle of the
+  system:
   \begin{mathpar}
@@
 
-    m\ddot{y} = mg
+    m\ddot{y} = -mg
   \end{mathpar}
-  Note that both these ODEs have the form $F = m\ddot{x}$, which is
-  exactly Newton's second law. Lagrangian mechanics allows, from the
-  principle of stationary action, the derivation of Newton's second
-  law, rather than postulating it as an axiom.
+  Note that both these ODEs have the form $F = m\ddot{x}$, just as in
+  Newton's second law. Lagrangian mechanics allows, from the principle
+  of stationary action, the derivation of Newton's second law, instead
+  of postulating it as an axiom.
 \end{example}
@@
 derive conservation laws from invariance properties. We now go on to
-show how suitable invariance properties can be derived from types. In
-Section~\ref{sec:examples}, we give examples of invariant Lagrangian's
+showing how suitable invariance properties can be derived from types,
+building towards Section~\ref{sec:types-for-classical-mech}, where we
+construct a type system for writing Lagrangians that expresses
+invariance properties directly in the types. In
+Section~\ref{sec:examples}, we give examples of invariant Lagrangians
 and their consequent conservation laws.
@@
 
-The first part here will be at least partly taken from Atkey
-2012. Work out a way of emphasising that we do not need to do anything
-to extend the basic underlying semantic structures involved: reflexive
-graphs are enough. This presentation differs from Atkey 2012 in that
-we emphasise the reflexive graph structure more, in order to be able
-to better expose the connection with the groupoid kinds in the section
-below.
-
-\subsection{Relational Parametricity for System F$\omega$}
+We derive invariance properties from types by constructing a
+relationally parametric model of System F$\omega$, which we then
+extend with suitable kind-, type-, and term-level constants for
+writing invariant Lagrangians. Our relationally parametric model is
+based on the reflexive graph model presented by
+Atkey~\cite{atkey12relational} (based on previous work on reflexive
+graph models of relational parametricity by Robinson and
+Rosolini~\cite{robinson94} and Hasegawa~\cite{hasegawa94}). In
+reflexive graph models of relational parametricity, the kinds of
+System F$\omega$ are interpreted as reflexive graphs, while types are
+interpreted as morphisms of reflexive graphs. As demonstrated by
+Atkey, uniformly interpreting kinds as reflexive graphs allows for a
+straightforward interpretation of higher kinds like $* \to *$. In the
+present setting, reflexive graphs will allow us to accomodate
+arbitrary groupoids as kinds
+(Section~\ref{sec:groupoid-kinds}). Groupoid kinds are the key
+technical tool that we need to derive the kinds of geometric
+invariance properties we need for Noether's theorem.
+
+\subsection{System F$\omega$: Syntax and Relationally Parametric Model}
 \label{sec:refl-graphs-for-fomega}
 
-Reflexive graphs give us a way of capturing the higher-dimensional
-structure of types in a
-
-Relational parametricity (Reynolds, BFSS), reflexive graphs
-(Robinson/Rosolini, Hasegawa, me).
-
-Explain the semantic setting. Explain the ``property'' thing.
-
-\paragraph{Syntax of System F$\omega$} Present the basic syntax of the
-kinds, types and terms of (predicative) System F$\omega$.
-
-Small kinds. We will extend the kinds, types and terms later to
-encompass the necessary geometrical constructs for classical
-mechanics.
-
+We now present the syntax and relationally parametric semantics of
+System F$\omega$. Since we will incrementally extend the syntax of the
+system throughout this section and the next, we do not state a final
+triple of theorems stating that we have correctly interpreted kinds,
+types and terms. Rather, we state the necessary properties of the
+interpretation as three separate properties
+(Properties~\ref{property:semantic-kinds},
+\ref{property:semantic-types}, and~\ref{property:semantic-terms},
+below), and maintain them as we extend the system.
+
+We deviate from the standard presentation of System F$\omega$ in that
+we distinguish between large and small kinds, and only allow
+quantification over small kinds. In particular, the kind of types $*$
+is \emph{not} small. The reason for this is that we desire a simple
+set-theoretic model. However, requiring that we have a set of sets
+closed under large products, in order to interpret quantification over
+the kind $*$, would force us to use an intuitionistic metatheory
+\cite{pitts}. The basic results of the calculus of variations that we
+are relying on, such as the derivation of the Euler-Lagrange
+equations, use excluded middle in their proofs, so an intuitionistic
+metatheory is not an option. We therefore just assume that we have a
+set of ``small'' sets, which includes the real numbers and is closed
+under subsets, products, function space and comprehension.
+
+\paragraph{Syntax} The kinds of System F$\omega$ include at least the
+base kind of types, $*$, and function and product kinds, as generated
+by the following grammar:
 \begin{displaymath}
@@
 \end{displaymath}
-
+We will extend the collection of kinds in
+Section~\ref{sec:discrete-kinds} with type-level natural numbers, and
+in Section~\ref{sec:groupoid-kinds}
+
+We use the judgement $\kappa~\mathrm{small}$ to denote when the kind
+$\kappa$ will be interpreted by a small reflexive graph (to be defined
+below). The kind $*$ is not small, but the construction of function
+and product kinds preserves smallness:
 \begin{mathpar}
@@
 \end{mathpar}
-
-Type equality... $\Theta \vdash A \equiv B : \kappa$
+The additional kinds we introduce below will all be small.
+
+The well-kinded types, type equalities, and well-typed terms of System
+F$\omega$ are shown in Figures~\ref{fig:types},
+\ref{fig:type-equality} and~\ref{fig:terms}, respectively. Note that
+typing contexts $\Gamma$ only contain well-kinded types of kind
+$*$. Terms also have an equational theory, with $\beta\eta$-laws for
+functions, products, and universal quantification, which we omit. Our
+presentation of System F$\omega$ is entirely standard (see, e.g.,
+Pierce~\cite{pierce01types}), except for the restriction to small
+kinds in the formation of universal kinds $\forall
+\alpha\mathord:\kappa. A$. As is the case for kinds, we will extend
+the types, type equalities and terms in
+Sections~\ref{sec:discrete-kinds},~\ref{sec:groupoid-kinds}
+and~\ref{sec:types-for-classical-mech}, below.
 
@@
   \label{fig:types}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \begin{mathpar}
+    \inferrule* [right=$\beta$]
+    {\Theta, \alpha : \kappa_1 \vdash A : \kappa_2 \\
+      \Theta \vdash B : \kappa_1}
+    {\Theta \vdash (\lambda \alpha : \kappa_1. A)~B \equiv A\{B/\alpha\} : \kappa_2}
+
+    \inferrule* [right=$\eta$]
+    {\Theta \vdash A : \kappa_1 \to \kappa_2}
+    {\Theta \vdash (\lambda \alpha : \kappa_1.A~\alpha) \equiv A : \kappa_1 \to \kappa_2}
+
+    \inferrule* [right=$\beta$]
+    {\Theta \vdash A_1 : \kappa_1 \\ \Theta \vdash A_2 : \kappa_2}
+    {\Theta \vdash \pi_i~\langle A_1, A_2 \rangle \equiv A_i : \kappa_i}
+
+    \inferrule* [right=$\eta$]
+    {\Theta \vdash A : \kappa_1 \times \kappa_2}
+    {\Theta \vdash \langle \pi_1 A, \pi_2 A \rangle \equiv A : \kappa_1 \times \kappa_2}
+  \end{mathpar}
+  {\tiny plus: reflexivity, symmetry, transitivity and congruence}
+  \caption{Type equality}
+  \label{fig:type-equality}
 \end{figure}
@@
 Sections~\ref{sec:discrete-kinds} and~\ref{sec:groupoid-kinds},
-below. We state this as Property~\ref{property:semantic-kinds}:
+below. We state this as Property~\ref{property:semantic-kinds} of our
+semantics:
 
@@
 \end{displaymath}
-The objects of the interpretation of a product kind is the product of
-the underlying collections of objects of the two parts, and the
-relational component is simply the product of the relational
+The collection of objects of the interpretation of a product kind is
+the product of the underlying collections of objects of the two parts,
+and the relational component is simply the product of the relational
 components. This naturally leads to the identity component being
@@
 well-defined function interpreting each well-typed term, with the
-property that it takes related environments to related results:
-
-\begin{property}
+property that it takes related environments to related results, this
+is the fundamental theorem of logical relations for System F$\omega$:
+
+\begin{property}\label{property:semantic-terms}
   For all well-typed terms $\Theta \sepbar \Gamma \vdash e : A$ there
@@
 In Section~\ref{sec:refl-graphs-for-fomega}, we only had a single base
-kind: the kind $*$ of proper types, with a specific intepretation as
+kind: the kind $*$ of proper types, with a specific interpretation as
 the reflexive graph of sets and relations. We now describe two
 families of base kinds with interpretations that are particular sorts
-of reflexive graphs. In this section, we look at discrete kinds; kinds
+of reflexive graph. In this section, we look at discrete kinds; kinds
 whose reflexive graph interpretations are such that the reflexive
@@
 
-Discrete kinds can be seen as the natural way of lifting base types up
-to the kind level.
-
-\begin{displaymath}
+Discrete kinds can be seen as the natural way of lifting types up to
+the kind level. For our purposes, we will only require a kind of
+natural numbers, which we will use for constructing $n$-ary vectors of
+positions and velocities when we use our type system for writing
+Lagrangians in Section~\ref{sec:types-for-classical-mech}. The kind of
+natural numbers is small:
+\begin{mathpar}
   \kappa ::= \cdots \sepbar \mathsf{nat}
-\end{displaymath}
-
-\begin{displaymath}
+
   \inferrule*
@@
   {\mathsf{nat}~\mathrm{small}}
-\end{displaymath}
-
+\end{mathpar}
+The interpretation of this kind as a reflexive graph goes as follows:
 \begin{displaymath}
@@
 \end{displaymath}
-
+The collection of objects of this reflexive graph is simply the set of
+natural numbers. The only edges in the graph are unique edges between
+equal numbers. Since we have assumed that our collection of small sets
+contains the natural numbers, we have clearly maintained
+Property~\ref{property:semantic-kinds} of our semantics.
+
+We add new type-level constructs for the kind $\mathsf{Nat}$ of
+natural numbers, representing zero and successor, and a kind-generic
+recursion operator:
 \begin{mathpar}
@@
     \Theta \vdash C : \kappa \to \kappa}
-  {\Theta \vdash \mathit{natrec}~A~B~C : \kappa}
+  {\Theta \vdash \mathit{natrec}_\kappa~A~B~C : \kappa}
 \end{mathpar}
-
-Make the point that any set can be used as the interpretation of a
-kind. Also, the utility of singleton types.
+These three constructs all have the evident interpretations in terms
+of the inductive structure of the natural numbers, and satisfy the
+following $\beta$-laws, maintaining
+Property~\ref{property:semantic-types} of our semantics:
+\begin{displaymath}
+  \begin{array}{@{}l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    \mathit{natrec}_\kappa~\mathit{zero}~B~C & \equiv & B \\
+    \mathit{natrec}_\kappa~(\mathit{succ}~A)~B~C & \equiv & C~(\mathit{natrec}_\kappa~A~B~C)
+  \end{array}
+\end{displaymath}
 
@@
 
-There is a full and faithful functor from the category of groupoids to
-the category of reflexive graphs, that preserves smallness.
-
-\paragraph{The additive group of integers}
-
-\paragraph{Matrix groups}
-
-GL and O, with group operations
+Discrete kinds are useful for lifting types up to the kind level, but
+they do not really exploit the flexibility of reflexive
+graphs. Moreover, they do not provide us with the invariance
+properties we require to apply Noether's theorem. By considering kinds
+whose interpretations are groupoids, we will be able to derive the
+invariance properties that we need.
+
+Recall that a groupoid is a category in which all morphisms have
+inverses \cite{maclane98categories}. Functors between categories
+always preserve isomorphisms, so morphisms of groupoids are just
+functors. Every group is a groupoid with one object and a morphism for
+every element of the group: composition is the group operation,
+inverses are given by the group inverses, and the identity morphism is
+given by the group unit. Homomorphisms between groups are in
+one-to-one correspondence with functors between the corresponding
+groupoids. An example of a groupoid that is not a group is given by
+the collection of cartesian spaces (i.e., $\mathbb{R}^n$ for some
+$n$), with all diffeomorphisms (smooth invertible functions) as the
+morphisms.
+
+Trivially, every groupoid is a reflexive graph, simply by forgetting
+the composition and inverses. Likewise, every functor between
+groupoids is a morphism of reflexive graphs, by forgetting the
+preservation of composition. Therefore, given any groupoid, we can add
+a new kind to System F$\omega$ that is interpreted by that groupoid,
+and given any functor $f : \sem{\mathsf{G_1}} \to \sem{\mathsf{G_2}}$
+between groupoids interpreting kinds $\mathsf{G_1}$ and
+$\mathsf{G_2}$, we obtain a type constant of kind $\mathsf{G_1} \to
+\mathsf{G_2}$, with interpretation $f$.
+
+\paragraph{Integers} Our first groupoid kind is $\mathsf{Z}$,
+interpreted by additive group of integers. We will use this kind to
+state the types of the trigonometric functions $\sin$ and $\cos$ in
+Section~\ref{sec:types-for-classical-mech-1}, which are periodic with
+period $2\pi$. The kind $\mathsf{Z}$ is small:
+\begin{mathpar}
+  \kappa ::= \cdots \sepbar \mathsf{Z}
+
+  \inferrule*{ }{\mathsf{Z}~\mathrm{small}}
+\end{mathpar}
+The kind $\mathsf{Z}$ has the following interpretation as a reflexive
+graph:
+\begin{displaymath}
+  \sem{\mathsf{Z}} = (\{*\}, \lambda *, *.~\mathbb{Z}, \lambda \mathord*.~0)
+\end{displaymath}
+Note the difference between this interpretation, and the
+interpretation of the kind $\mathsf{Nat}$ in the previous section. In
+the interpretation of $\mathsf{Nat}$, the objects were complex and the
+relations were trivial. The reflexive graph interpreting $\mathsf{Z}$
+has a trivial collection of objects, but a rich structure at the
+relation level, given by the group of integers. This structure means
+that we can add the following type-level constants to our system:
+\begin{mathpar}
+  \inferrule*
+  { }
+  {\Theta \vdash 0 : \mathsf{Z}}
+
+  \inferrule*
+  {\Theta \vdash A : \mathsf{Z} \\ \Theta \vdash B : \mathsf{Z}}
+  {\Theta \vdash A + B : \mathsf{Z}}
+
+  \inferrule*
+  {\Theta \vdash A : \mathsf{Z}}
+  {\Theta \vdash -A : \mathsf{Z}}
+\end{mathpar}
+Each of these constants has a trivial object-level interpretation, due
+to the trivial collection of objects in the reflexive graph
+$\sem{\mathsf{Z}}$. At the relation-level, the interpretations are
+simply given using the group structure of the integers. These
+interpretations satisfy the abelian group axioms, so we add these
+axioms to the type-level equations.
+
+\paragraph{Translation and Matrix Groups} The group of integers is not
+suitable for stating the invariance properties we require to apply
+Noether's theorem. To do so, we require groups of translations and
+invertible linear transformations. We extend our type system with
+small kinds representing $n$-dimensional ($n \geq 0$) translations,
+invertible linear transformations, and orthogonal transformations:
+\begin{mathpar}
+  \kappa ::= \cdots \sepbar \mathsf{T}(n) \sepbar \mathsf{GL}(n) \sepbar \mathsf{O}(n)
+
+  \inferrule*{ }{\mathsf{T}(n), \mathsf{GL}(n), \mathsf{O}(n)~\mathrm{small}}
+\end{mathpar}
+The interpretations of these kinds follow the same pattern as for the
+kind $\mathsf{Z}$ above. Each interpretation is a reflexive graph with
+a trivial collection of objects, and relations taken from the
+appropriate group:
+\begin{displaymath}
+  \begin{array}{@{}l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    \sem{\mathsf{T}(n)} &=& (\{*\}, \lambda *,*.~\mathbb{R}^n, \lambda \mathord*.~\vec{0}) \\
+    \sem{\mathsf{GL}(n)} &=& (\{*\}, \lambda *,*.~\mathrm{GL}(n), \lambda \mathord*. I) \\
+    \sem{\mathsf{O}(n)} &=& (\{*\}, \lambda *,*.~\mathrm{O}(n), \lambda \mathord*. I)
+  \end{array}
+\end{displaymath}
+We add the group operations and group axioms for each of these
+groupoid kinds in the same way as we did for the kind $\mathsf{Z}$
+above, except that we use multiplicative notation for the kinds
+$\mathsf{GL}(n)$ and $\mathsf{O}(n)$. Note that, except for
+$\mathsf{GL}(1)$, the groups $\mathsf{GL}(n)$ and $\mathsf{O}(n)$ are
+not abelian.
+
+FIXME: the following group homomorphisms become type-level constants,
+using the interpretation we defined at the start of this section.
+
+$\mathit{exp} : \mathsf{T}(1) \to \mathsf{GL}(1)$
+
+$\underline{c}\cdot- : \mathsf{Z} \to \mathsf{T}(1)$
 
 $\mathit{ortho}_n : \mathsf{O}(n) \to \mathsf{GL}(n)$
+
 $\mathit{scale}_n : \mathsf{GL}(1) \to \mathsf{GL}(n)$
 
-\paragraph{Translation groups}
-
-$\mathit{exp} : \mathsf{T}(1) \to \mathsf{GL}(1)$
-$\underline{c}\cdot- : \mathsf{Z} \to \mathsf{T}(1)$
-
-\paragraph{Groupoid of Cartesian Spaces}
-
-\subsection{Why not Groupoids Everywhere?}
-
-Explain why we aren't using groupoids as kinds: don't need to, doesn't
-directly generalise relational parametricity, see Robinson's
-paper. Distance-indexed types.
+\paragraph{The Groupoid of Cartesian Spaces} Finally, we introduce the
+groupoid kind of cartesian spaces, which we will use for the
+configuration spaces of the classical mechanical systems we describe
+in our system. As we noted above, the collection of cartesian spaces
+is a groupoid with a non-trivial collection of objects. We extend our
+system with a kind of cartesian spaces:
+\begin{displaymath}
+  \kappa ::= \cdots \sepbar \mathsf{CartSp}
+\end{displaymath}
+with the following interpretation:
+\begin{displaymath}
+  \begin{array}{@{}l}
+    \sem{\mathsf{CartSp}} = \\
+    \quad (\mathbb{N}, \lambda m,n.~\{ f : \mathbb{R}^m \to \mathbb{R}^n \sepbar f\textrm{ a diffeomorphism}\}, \lambda n.~\mathrm{id}_{\mathbb{R}^n})
+  \end{array}
+\end{displaymath}
+We have chosen to represent cartesian spaces of dimension $n$ just as
+the natural number $n$. This gives us a canonical representation of
+$n$-dimensional space for each $n$, avoids tricky problems with
+isomorphic constructions of the same space.
+
+Cartesian spaces are closed under taking their product:
+\begin{displaymath}
+  \inferrule*
+  {\Theta \vdash X : \mathsf{CartSp} \\ \Theta \vdash Y : \mathsf{CartSp}}
+  {\Theta \vdash X \times Y : \mathsf{CartSp}}
+\end{displaymath}
+with the following interpretation, where the cartesian product of
+cartesian spaces adds their dimensions, and composes diffeomorphisms
+``in parallel'':
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    \sem{X \times Y}^f \theta & = & \sem{X}^f \theta + \sem{Y}^f \theta \\
+    \sem{X \times Y}^r \theta \theta' \rho & = & \lambda (\vec{x},\vec{y}).~(\sem{X}^r\theta\theta'\rho \vec{x}, \sem{Y}^r\theta\theta'\rho \vec{y})
+  \end{array}
+\end{displaymath}
+
+Cartesian spaces are generated by the following construct that yields
+the $n$-dimensional cartesian space ($n \geq 0$) that varies with the
+given invertible linear transformation and translation:
+\begin{displaymath}
+  \inferrule*
+  {\Theta \vdash G : \mathsf{GL}(n) \\ \Theta \vdash T : \mathsf{T}(n)}
+  {\Theta \vdash \mathbb{R}^n\langle G, T \rangle : \mathsf{CartSp}}
+\end{displaymath}
+As a notational convenience, we write $\mathbb{R}^1\langle G, T
+\rangle$ as just $\mathbb{R}\langle G, T \rangle$ and
+$\mathbb{R}^0\langle G, T \rangle$ as just $\mathbb{R}^0$.
+
+The type $\mathbb{R}^n\langle G, T \rangle$ has the following
+interpretation as a reflexive graph morphism:
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    \sem{\mathbb{R}^n\langle G, T \rangle}^f \theta & = & n \\
+    \sem{\mathbb{R}^n\langle G, T \rangle}^r \theta \theta' \rho & = & \lambda \vec{x}.~(\sem{G}^r \theta \theta' \rho) \vec{x} + \sem{T}^r \theta \theta' \rho
+  \end{array}
+\end{displaymath}
+Note that $\sem{\mathbb{R}^n\langle G, T \rangle}^r \theta \theta'
+(\semKI{\Theta}\theta)$ is the identity diffeomorphism, so this
+interpretation preserves identity edges, exactly as required in the
+definition of reflexive graph morphisms. Consequently, the
+interpretation of $\mathbb{R}^n\langle G, T \rangle$, and also the
+interpretation for products of cartesian spaces, preserves
+Property~\ref{property:semantic-types}.
+
+Finally, we include an operator that takes any cartesian space to
+its corresponding proper type:
+\begin{displaymath}
+  \inferrule*
+  {\Theta \vdash X : \mathsf{CartSp}}
+  {\Theta \vdash \typeOfCartSp{X} : *}
+\end{displaymath}
+with the following straightforward interpretation:
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    \sem{\typeOfCartSp{X}}^f\theta & = & \mathbb{R}^{\sem{X}^f\theta} \\
+    \sem{\typeOfCartSp{X}}^r\theta\theta'\rho & = & \{ (\vec{x}, \vec{x'}) \sepbar (\sem{X}^r\theta\theta'\rho)\vec{x} = \vec{x'} \}
+  \end{array}
+\end{displaymath}
+
+% \subsection{Why not Groupoids Everywhere?}
+
+% Given that we crucially use groupoid kinds in our model in order
+
+% Explain why we aren't using groupoids as kinds: don't need to, doesn't
+% directly generalise relational parametricity, see Robinson's
+% paper. Distance-indexed types.
 
@@
 
-FIXME: rewrite this para to refer back concretely
-
-We have a way of getting symmetry properties from types, and a way of
-using symmetry properties to derive conservation laws for physical
-systems. We now put the two together by building an applied version
-System F$\omega$ with new kind-, type- and term-level constants that
-will deliver us the exact symmetry properties we require to be able to
-apply Noether's theorem.
-
-%\subsection{Extending System F$\omega$ for Classical Mechanics}
-
-Let us enumerate our requirements for an extension of System F$\omega$
-for writing invariant Lagrangians\footnote{Make sure that I say this
-  is the plan back in Sec 2}.
+In the previous section we constructed a way to derive invariance
+properties from types. In
+Section~\ref{sec:conservation-laws-from-symmetry}, we saw that
+Noether's theorem is a method for deriving conservation laws for
+classical mechanical systems from their invariance properties. In this
+section, we construct a type system for constructing invariant
+Lagrangians. In the following section, Section~\ref{sec:examples}, we
+present several uses of our type system for defining invariant
+Lagrangians that describe many different kinds of classical mechanical
+systems.
 
 \subsection{Extending System F$\omega$ with Smooth Functions}
-
-\begin{enumerate}
-\item We have already done the matrix kinds and the CartSp kind back
-  in the previous section. Add $n-ary$ products for CartSp.
+\label{sec:types-for-classical-mech-1}
+
+\paragraph{The Type of Smooth Functions} Given a pair of cartesian
+spaces $X$ and $Y$ we provide a way to construct the type of smooth
+invariant smooth functions from $X$ to $Y$:
+\begin{displaymath}
+  \inferrule*
+  {\Theta \vdash X : \mathsf{CartSp} \\
+    \Theta \vdash Y : \mathsf{CartSp}}
+  {\Theta \vdash C^\infty(X, Y) : *}
+\end{displaymath}
+with the following interpretation. Since $C^\infty(X,Y)$ is a proper
+type, the relational interpetation in this case is a genuine
+relation. We relate functions that are invariant under the
+diffeomorphism interpretations of $X$ and $Y$ in the current
+relational context $\rho$:
+\begin{displaymath}
+  \begin{array}{@{\hspace{0.05em}}l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    \sem{C^\infty(X,Y)}^f\theta & = & \{ f : \mathbb{R}^{\sem{X}^f\theta} \to \mathbb{R}^{\sem{Y}^f\theta} \sepbar f\textrm{ smooth} \} \\
+    \sem{C^\infty(X,Y)}^f\theta\theta'\rho & = & \{
+    \begin{array}[t]{@{}l}
+      (f,f') \sepbar \forall \vec{x} \in \mathbb{R}^{\sem{X}^f\theta}.\\
+      \quad (\sem{Y}^r\theta\theta'\rho)(f~\vec{x}) = f'~(\sem{X}^r\theta\theta'\rho \vec{x}) \}
+    \end{array}
+  \end{array}
+\end{displaymath}
+
+\paragraph{An Example Free Theorem} We are now in a position to
+formally show how we can use the relationally parametric model we have
+defined to derive invariance properties. Consider a closed term with
+the following type:
+\begin{displaymath}
+  e : \forall o : \mathsf{O}(n).~C^\infty(\mathbb{R}^n\langle \mathit{ortho}_n(o), 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
+\end{displaymath}
+We can deduce, from Property~\ref{property:semantic-terms} and the
+relational interpretations of all the types involved, that the
+denotation of $e$ satisfies the following free theorem:
+\begin{displaymath}
+  \forall O \in \mathrm{O}(n).~\forall \vec{x} \in \mathbb{R}^n.~\sem{e}(O\vec{x}) = \sem{e}(\vec{x})
+\end{displaymath}
+This free theorem is exactly the kind of invariance property we
+require to apply Noether's theorem. Compare the form of the statement
+here, with the mildly more general statement of
+Equation~\ref{eq:simplest-invariance}, back in
+Section~\ref{sec:conservation-laws-from-symmetry}. Moreover, $\sem{e}$
+is a smooth function from $\mathbb{R}^n$ to $\mathbb{R}$, just we
+require to for Lagrangian mechanics.
+
+\paragraph{Combinators for Smooth Functions} Of course, such free
+theorems are pointless without ways to build elements of the smooth
+function spaces. We now extend our type system, for the last time,
+with a set of term-level constants for constructing smooth functions,
+invariant under linear transformations and translations.
+
+FIXME: Administrative things: composition, products
+
 \begin{displaymath}
   \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
-      C^\infty &:& \mathsf{CartSp} \to \mathsf{CartSp} \to * \\
-      \mathbb{R}^n &:& \mathsf{GL}(n) \to \mathsf{T}(n) \to \mathsf{CartSp} \\
-%      (\times) &:& \mathsf{CartSp} \to \mathsf{CartSp} \to \mathsf{CartSp} \\
-      \mathit{vec} &:& \mathsf{nat} \to \mathsf{CartSp} \to \mathsf{CartSp}
-    \end{array}
-\end{displaymath}
-\item Note that the $n$ in $\mathbb{R}^n$ is static
-\item Add the type former for the smooth function space, and all the
-  relevant constants
-\item Give a (very) small example, and complain that it is very
-  difficult to write anything in this point-free style.
-\end{enumerate}
-
-Administrative things: composition, products
+    \mathrm{const} &:& \typeOfCartSp{Y} \to C^\infty(X, Y) \\
+    \mathrm{pair} &:& C^\infty(X,Y) \to C^\infty(X,Z) \to C^\infty(X, Y \times Z) \\
+    \mathrm{proj_1}&:& C^\infty(X \times Y, X) \\
+    \mathrm{proj_2}&:& C^\infty(X \times Y, Y) \\
+    (\comp) &:& C^\infty(X,Y) \to C^\infty(Y,Z) \\
+    \mathrm{id} &:& C^\infty(X,X)
+  \end{array}
+\end{displaymath}
+
+FIXME: vector spacey things
 
@@
   \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    \underline{\vec{x}} &:& \typeOfCartSp{\mathbb{R}^n\langle 1, 0 \rangle} \\
+    0 &:& \forall g \mathord: \mathsf{GL}(n).~\typeOfCartSp{\mathbb{R}^n\langle g, 0 \rangle} \\
     (+) &:&
@@
     \end{array} \\
-    \sin &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, 2\pi \cdot z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
-    \cos &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, 2\pi \cdot z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
-    \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, t \rangle, \mathbb{R}\langle \mathit{exp}~t, 0 \rangle)
-  \end{array}
-\end{displaymath}
+    (*) &:& FIXME \\
+    (\cdot) &:& FIXME \\
+  \end{array}
+\end{displaymath}
+
+FIXME: trigonometic functions, and exponentials.
+
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    \sin &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
+    \cos &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
+    \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, t \rangle, \mathbb{R}\langle \mathit{exp}~t, 0 \rangle) \\
+  \end{array}
+\end{displaymath}
+
+FIXME: division, excusing the partiality
+
+\begin{displaymath}
+  (/) : \forall 
+\end{displaymath}
+
+FIXME: sum and cross
 
@@
 \label{sec:surface-syntax}
-
-\begin{enumerate}
-\item Kind of cartesian spaces, parameterised by a matrix group and a
-  translation group
-\item Smooth functions between spaces: motivate this by looking at the
-  requirements of the Euler-Lagrange equations.
-\item Primitives for dealing with the geometry of the spaces (inner
-  product, vector space operations, cross product(?), trigonometric
-  functions (at least sin and cos)).
-\end{enumerate}
-
-\begin{displaymath}
-  \kappa ::= \cdots \sepbar \mathsf{CartSp} \sepbar \mathsf{T}(n) \sepbar \mathsf{GL}(n) \sepbar \mathsf{O}(n) \sepbar \mathsf{Z}
-\end{displaymath}
-
-\begin{itemize}
-\item $\mathsf{CartSp}$ is the kind of cartesian spaces
-  ($\mathbb{R}^n$, for some $n$) and diffeomorphisms between them,
-  with the standard definition of diffeomorphism on these spaces. The
-  identity edge is just the identity function.
-\item $\mathbb{R}^n~t~g$ is the cartesian space with the
-  diffeomorphism defined by the action of the translation and matrix
-  group.
-\item $X \times Y$ takes the product of two cartesian spaces, and the
-  corresponding product of the diffeomorphisms.
-\item $X \smoothrightarrow Y$ consists of the infinitely
-  differentiable functions from $X$ to $Y$ that preserve the
-  diffeomorphisms.
-\end{itemize}
-
-\begin{enumerate}
-\item Need coercions between the different groups, as required by the
-  examples.
-\item Need operations on vectors, e.g., the usual linear space ones,
-  and inner product. Perhaps also the cross product one.
-\item What operations on the smooth function space do we need?
-  composition, products, identities.
-\end{enumerate}
 
@@
     \inferrule*
-    {x : \typeOfCartSp{X} \in \Gamma}
-    {\Theta \sepbar \Gamma; \Delta \vdash x : X}
+    {z : \typeOfCartSp{X} \in \Gamma}
+    {\Theta \sepbar \Gamma; \Delta \vdash z : X}
 
@@
     \inferrule*
-    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : X \\
-      \Theta \sepbar \Gamma; \Delta, x : X \vdash e_2 : Y}
-    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{let}~x = e_1~\mathrm{in}~e_2 : Y}
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : X_1 \times ... \times X_n \\
+      \Theta \sepbar \Gamma; \Delta, x_1 : X_1, ..., x_n : X_n \vdash e_2 : Y}
+    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{let}~x_1, ..., x_n = e_1~\mathrm{in}~e_2 : Y}
 
@@
     {\Theta \sepbar \Gamma; \Delta \vdash e_1 (e_2) : Y}
+
+    \inferrule*
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : X \\
+      ... \\
+      \Theta \sepbar \Gamma; \Delta \vdash e_n : X_n}
+    {\Theta \sepbar \Gamma; \Delta \vdash (e_1, ..., e_n) : X_1 \times ... \times X_n}
   \end{mathpar}
+
+  \medskip
   \textbf{Vector space operations, and dot product}
@@
   \end{mathpar}
+
+  \medskip
   \textbf{Transcendental functions}
@@
   \end{mathpar}
-  FIXME: products, division, n-ary products (sum, zip, reduce), constants
-  \caption{Type system for smooth functions}
+  FIXME: division, n-ary products (sum, zip, reduce)
+  \caption{A surface syntax for smooth functions}
   \label{fig:smooth-terms}
@@
 
+\begin{figure}
+  \centering
+  \begin{displaymath}
+    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+      \elab{z}_\Delta & = & \mathrm{const}~z \\
+      \elab{x}_\Delta & = & \mathrm{project}(x, \Delta) \\
+    \end{array}
+  \end{displaymath}
+  \caption{Desugaring of the surface syntax}
+  \label{fig:desugar}
+\end{figure}
+
+FIXME: define the translation
+
+Define $\Delta$ and $\elab{\Delta}$.
+
+Define project
+
 \begin{theorem}
-  If $\Theta \sepbar \Gamma; \Delta \vdash e : X$ then $\Theta \sepbar
+  If $\Theta |\Gamma; \Delta \vdash e : X$ then $\Theta \sepbar
   \Gamma \vdash \elab{e} : C^\infty(\elab{\Delta}, X)$.
@@
 
-\section{Examples}\label{sec:examples}
-
-$- \sepbar V : \forall ...
+\section{Examples of Conservation Laws from Types}
+\label{sec:examples}
+
+We now present a number of examples of classical mechanical systems
+expressible in the type system we constructed in the previous section,
+and derive the free theorems and consequent conserved properties for
+each one. These examples are all standard examples demonstrating the
+application of Lagrangian mechanics (see, for example, Landau and
+Lifschitz \cite{landau60mechanics}). Our contribution here is to
+express them in a type system that makes clear their invariance
+properties.
+
+\subsection{Single Particles}
+
+% $- \sepbar V : \forall ...
 .
@@
 
-\begin{enumerate}
-\item free particle
-\item $2$-body
-\item $n$-body (use length-indexed vectors)
-\item pendulums
-\item oscillators
-\end{enumerate}
-
-% \section{Basic Field Theory}
-
-% \subsection{Principle of Stationary Action for Fields}
-
-% \subsection{Noether's Theorem for Fields}
-
-% \subsection{The Klein-Gordon Equation}
+\paragraph{Free Particle}
+
+\paragraph{Particle in a Potential Field}
+
+\subsection{$n$-Body Problems}
+
+\paragraph{Two Bodies}
+
+\paragraph{$n$-Bodies}
+
+\subsection{Pendulums}
+
+\paragraph{A single pendulum}
+
+\paragraph{Coupled pendulum}
+
+\paragraph{Pendulum on a Moving Base}
+
+\paragraph{Pendulum on a Rotating Base}
+
+No time translation invariance.
+
+\subsection{Oscillators}
+
+\paragraph{Coupled Particles} Our first example of a Lagrangian in the
+introduction was of a pair of particles of equal mass coupled by a
+spring.
+
+\paragraph{Fixed Endpoint Oscillator} We now look at the slightly
+simpler case of a single particle connected to a spring attached to a
+fixed point.
+
+\paragraph{Damped Oscillator}
 
@@
 \label{sec:conclusions}
+
+FIXME: sum up what's happened
+
+\paragraph{Related Work} We are not aware of any other work linking
+type systems, programming languages or relational parametricity with
+invariance properties for classical mechanics, or with Noether's
+theorem. We have already mentioned the work of Atkey, Johann and
+Kennedy \cite{atkey13abstraction} that uses
+
+\paragraph{Future Work}
 

### 2013-07-12 mechanics/paper.tex
Finish Sec 3, at last

@@
 
-FIXME: the following group homomorphisms become type-level constants,
-using the interpretation we defined at the start of this section.
-
-$\mathit{exp} : \mathsf{T}(1) \to \mathsf{GL}(1)$
-
-$\underline{c}\cdot- : \mathsf{Z} \to \mathsf{T}(1)$
-
-$\mathit{ortho}_n : \mathsf{O}(n) \to \mathsf{GL}(n)$
-
-$\mathit{scale}_n : \mathsf{GL}(1) \to \mathsf{GL}(n)$
-
-\paragraph{The Groupoid of Cartesian Spaces} Finally, we introduce the
-groupoid kind of cartesian spaces, which we will use for the
-configuration spaces of the classical mechanical systems we describe
-in our system. As we noted above, the collection of cartesian spaces
-is a groupoid with a non-trivial collection of objects. We extend our
-system with a kind of cartesian spaces:
+We also extend our type system with the following type-level
+constants. Each of these is interpreted by the correspondingly named
+homomorphism between the groups involved. For example,
+$\mathit{scale}_n$ is interpreted by the homomorphism $\mathrm{GL}(1)
+\to \mathrm{GL}(n)$ that takes a non-zero real number $s$ to the
+matrix with $s$s along the diagonal and zeros elsewhere.
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    \mathit{exp} &:& \mathsf{T}(1) \to \mathsf{GL}(1) \\
+    \mathit{ortho}_n &:& \mathsf{O}(n) \to \mathsf{GL}(n) \\
+    \mathit{scale}_n &:& \mathsf{GL}(1) \to \mathsf{GL}(n) \\
+    \underline{c}*- &:& \mathsf{Z} \to \mathsf{T}(1) 
+  \end{array}
+\end{displaymath}
+In the last of these $\underline{c}$ stands for an arbitrary real
+constant.  Since each of these is interpreted by a group homomorphism,
+we are justified in adding the group homomorphism laws as axioms to
+our type-level equational theory, maintaining
+Property~\ref{property:semantic-types}.
+
+\paragraph{The Groupoid of Cartesian Spaces} Finally in this section,
+we introduce the groupoid kind of cartesian spaces, which we will use
+for the configuration spaces of the classical mechanical systems we
+describe in our system. As we noted above, the collection of cartesian
+spaces is a groupoid with a non-trivial collection of objects. We
+extend our system with a kind of cartesian spaces:
 \begin{displaymath}

### 2013-07-12 mechanics/paper.tex
Work on Sec 4

@@
 
+Using the type-level natural numbers and their recursion operator, we
+can define a type of $n$-ary products of a cartesian space:
+\begin{displaymath}
+  \mathit{vec} = \lambda n\mathord:\mathsf{Nat}, x\mathord:\mathsf{CartSp}.~\mathit{natrec}~n~\mathbb{R}^0~(\lambda x.~X \times x)
+\end{displaymath}
+$n$-ary products of cartesian spaces will be useful in
+Section~\ref{sec:examples} when we define Lagrangians describing
+systems that are generic in the number of particles involved.
+
 Finally, we include an operator that takes any cartesian space to
@@
 with the following interpretation. Since $C^\infty(X,Y)$ is a proper
-type, the relational interpetation in this case is a genuine
+type, the relational interpretation in this case is a genuine
 relation. We relate functions that are invariant under the
@@
 
-FIXME: Administrative things: composition, products
-
+Our first set of combinators constructs constant and identity smooth
+functions, and composes smooth functions between cartesian spaces. We
+also have combinators for pairing and projection for the product of
+cartesian spaces.
 \begin{displaymath}
@@
     \mathrm{const} &:& \typeOfCartSp{Y} \to C^\infty(X, Y) \\
+    \mathrm{id} &:& C^\infty(X,X) \\
+    (\comp) &:& C^\infty(X,Y) \to C^\infty(Y,Z) \\
     \mathrm{pair} &:& C^\infty(X,Y) \to C^\infty(X,Z) \to C^\infty(X, Y \times Z) \\
@@
     \mathrm{proj_2}&:& C^\infty(X \times Y, Y) \\
-    (\comp) &:& C^\infty(X,Y) \to C^\infty(Y,Z) \\
-    \mathrm{id} &:& C^\infty(X,X)
-  \end{array}
-\end{displaymath}
-
-FIXME: vector spacey things
-
+  \end{array}
+\end{displaymath}
+
+Next, we include combinators for constant, invariant, vectors, the
+zero vector, vector addition, subtraction and scaling. We also include
+the Euclidean dot product of vectors. The type dot product operation
+is where the orthogonal group kind $\mathsf{O}(n)$ is used: the dot
+product of vectors is invariant under orthogonal transformations.
 \begin{displaymath}
   \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
-    \underline{\vec{x}} &:& \typeOfCartSp{\mathbb{R}^n\langle 1, 0 \rangle} \\
+    \underline{\vec{c}} &:& \typeOfCartSp{\mathbb{R}^n\langle 1, 0 \rangle} \\
     0 &:& \forall g \mathord: \mathsf{GL}(n).~\typeOfCartSp{\mathbb{R}^n\langle g, 0 \rangle} \\
@@
     \end{array} \\
-    (*) &:& FIXME \\
-    (\cdot) &:& FIXME \\
-  \end{array}
-\end{displaymath}
-
-FIXME: trigonometic functions, and exponentials.
-
+    (*) &:&
+    \begin{array}[t]{@{}l}
+      \forall g_1 \mathord: \mathsf{GL}(1), g_2 \mathord: \mathsf{GL}(n).\\
+      \quad C^\infty(\mathbb{R}\langle g_1, 0 \rangle \times \mathbb{R}^n\langle g_2, 0\rangle, \mathbb{R}\langle \mathit{scale}_n(g_1)g_2, 0 \rangle)
+    \end{array}
+    \\
+    (\cdot) &:& 
+    \begin{array}[t]{@{}l}
+      \forall g \mathord: \mathsf{GL}(1), o \mathord: \mathsf{O}(n).~\\
+      C^\infty(\mathbb{R}^n\langle (\mathit{scale}_n~g)(\mathit{ortho}_n~o), 0 \rangle \times \\
+      \quad\quad\; \mathbb{R}^n\langle (\mathit{scale}_n~g)(\mathit{ortho}_n~o), 0\rangle, \mathbb{R}\langle (\mathit{scale}_n~g)^n, 0 \rangle)
+    \end{array}
+  \end{array}
+\end{displaymath}
+Each of these combinators has a straightforward interpretation as a
+smooth function, and the proof that the fundamental theorem of logical
+relations is maintained for our system (i.e., that
+Property~\ref{property:semantic-terms} is maintained), follows
+directly from elementary facts about linear algebra.
+
+It will be useful for our examples to have the following trigonometric
+functions, exponential function and division operation. The types of
+each of these functions describes some of their invariance properties,
+for example the periodicity of the $\sin$ and $\cos$ functions. A
+technical problem with our current system is that we do not account
+for the non-definedness of division at $0$. However, division will be
+useful in many of our examples, so we take a pragmatic approach, in
+common with most physics textbooks, and informally treat division as
+total function.
 \begin{displaymath}
   \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
-    \sin &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
-    \cos &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
+    \sin &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} * z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
+    \cos &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} * z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
     \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, t \rangle, \mathbb{R}\langle \mathit{exp}~t, 0 \rangle) \\
-  \end{array}
-\end{displaymath}
-
-FIXME: division, excusing the partiality
-
-\begin{displaymath}
-  (/) : \forall 
-\end{displaymath}
-
-FIXME: sum and cross
+    (/) &:& \forall g_1, g_2 \mathord: \mathsf{GL}(1).~C^\infty(\mathbb{R}\langle g_1, 0 \rangle \times \mathbb{R}\langle g_2, 0\rangle, \mathbb{R}\langle g_1g_2^{-1}, 0 \rangle)
+  \end{array}
+\end{displaymath}
+
+Finally, we include two primitive combinators for dealing with $n$-ary
+products of vectors. Both of these combinators will be useful when
+defining Lagrangians that are generic in the number of particles.
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    \mathrm{sum} &:& \forall n \mathord: \mathsf{Nat}, g \mathord: \mathsf{GL}(1). C^\infty(\mathit{vec}~n~(\mathbb{R}\langle g, 0 \rangle), \mathbb{R}\langle g, 0\rangle) \\
+    \mathrm{cross} &:& 
+    \begin{array}[t]{@{}l}
+      \forall m,n \mathord: \mathsf{Nat}.\\
+      \quad C^\infty(X \times Y, Z) \to \\
+      \quad\quad C^\infty(\mathit{vec}~m~X \times \mathit{vec}~n~Y, \mathit{vec}~(m*n)~Z)
+    \end{array}
+  \end{array}
+\end{displaymath}
+The combinator $\mathrm{sum}$ sums the list of real numbers it is
+given, while $\mathrm{cross}$ takes the cartesian product of two
+vectors of elements of cartesian spaces, and maps the given smooth
+function over them. The multiplication of natural numbers $m*n$ is
+straightforwardly defined in terms of the $\mathit{natrec}$ recursion
+operator. We will use $\mathrm{cross}$ in Section~\ref{sec:examples}
+to define Lagrangians in terms of the interactions between systems of
+particles.
 
@@
     \inferrule*
-    {\vec{x} \in \mathbb{R}^n}
-    {\Theta \sepbar \Gamma; \Delta \vdash \vec{x} : \mathbb{R}^n\langle 1, 0 \rangle}
-
-    \inferrule*
-    {\Theta \vdash g : \mathsf{GL}(n)}
-    {\Theta \sepbar \Gamma; \Delta \vdash 0 : \mathbb{R}^n\langle g, 0 \rangle}
-
-    \inferrule*
-    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle g, t_1 \rangle \\
-      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle g, t_2 \rangle}
-    {\Theta \sepbar \Gamma; \Delta \vdash e_1 + e_2 : \mathbb{R}^n\langle g, t_1 + t_2 \rangle}
-
-    \inferrule*
-    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle g, t_1 \rangle \\
-      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle g, t_2 \rangle}
-    {\Theta \sepbar \Gamma; \Delta \vdash e_1 - e_2 : \mathbb{R}^n\langle g, t_1 - t_2 \rangle}
-
-    \inferrule*
-    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}\langle g_1, 0 \rangle \\
-      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle g_2, 0 \rangle}
-    {\Theta \sepbar \Gamma; \Delta \vdash e_1 * e_2 : \mathbb{R}^n\langle \mathit{scale}_n(g_1)g_2, 0 \rangle}
-
-    \inferrule*
-    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle \mathit{scale}_n(s)\mathit{ortho}_n(o), 0 \rangle \\
-      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle \mathit{scale}_n(s)\mathit{ortho}_n(o), 0 \rangle}
-    {\Theta \sepbar \Gamma; \Delta \vdash e_1 \cdot e_2 : \mathbb{R}\langle\mathit{scale}_n(s)^n, 0 \rangle}
+    {\vec{c} \in \mathbb{R}^n}
+    {\Theta \sepbar \Gamma; \Delta \vdash \vec{c} : \mathbb{R}^n\langle 1, 0 \rangle}
+
+    \inferrule*
+    {\Theta \vdash G : \mathsf{GL}(n)}
+    {\Theta \sepbar \Gamma; \Delta \vdash 0 : \mathbb{R}^n\langle G, 0 \rangle}
+
+    \inferrule*
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle G, T_1 \rangle \\
+      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle G, T_2 \rangle}
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 + e_2 : \mathbb{R}^n\langle G, T_1 + T_2 \rangle}
+
+    \inferrule*
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle G, T_1 \rangle \\
+      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle G, T_2 \rangle}
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 - e_2 : \mathbb{R}^n\langle G, T_1 - T_2 \rangle}
+
+    \inferrule*
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}\langle G_1, 0 \rangle \\
+      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle G_2, 0 \rangle}
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 * e_2 : \mathbb{R}^n\langle \mathit{scale}_n(G_1)G_2, 0 \rangle}
+
+    \inferrule*
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle \mathit{scale}_n(G)\mathit{ortho}_n(O), 0 \rangle \\
+      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle \mathit{scale}_n(G)\mathit{ortho}_n(O), 0 \rangle}
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 \cdot e_2 : \mathbb{R}\langle\mathit{scale}_n(G)^n, 0 \rangle}
   \end{mathpar}
@@
     \inferrule*
-    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, t \rangle}
-    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{exp}~e : \mathbb{R}\langle \mathit{exp}~t, 0 \rangle}
-
-    \inferrule*
-    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 2\pi z, 0 \rangle}
+    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, T \rangle}
+    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{exp}~e : \mathbb{R}\langle \mathit{exp}~T, 0 \rangle}
+
+    \inferrule*
+    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, \underline{2\pi} * Z \rangle}
     {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{sin}~e : \mathbb{R}\langle 1, 0 \rangle}
@@
     \inferrule*
-    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 2\pi z, 0 \rangle}
+    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, \underline{2\pi} * Z \rangle}
     {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{cos}~e : \mathbb{R}\langle 1, 0 \rangle}
@@
       \elab{z}_\Delta & = & \mathrm{const}~z \\
-      \elab{x}_\Delta & = & \mathrm{project}(x, \Delta) \\
+      \elab{x}_\Delta & = & \underline{\mathrm{project}_\Delta}(x) \\
+      \elab{\mathrm{let}~\vec{x} = e_1\textrm{ in }e_2}_\Delta & = & \mathrm{pair}~\mathrm{id}~\elab{e_1}_\Delta \comp \elab{e_2}_{\Delta, \vec{x}} \\
+      \elab{(e_1,...,e_n)}_\Delta &=& \underline{\mathrm{pair}_n}(\elab{e_1}_\Delta, ..., \elab{e_n}_\Delta) \\
+      \elab{e_1(e_2)}_\Delta &=& \elab{e_2}_\Delta \comp e_1 \\
+      \elab{\vec{c}}_\Delta &=& 
     \end{array}

### 2013-07-12 mechanics/paper.tex
Finish Sec 4

@@
     \mathrm{sum} &:& \forall n \mathord: \mathsf{Nat}, g \mathord: \mathsf{GL}(1). C^\infty(\mathit{vec}~n~(\mathbb{R}\langle g, 0 \rangle), \mathbb{R}\langle g, 0\rangle) \\
+    \mathrm{map} &:& \forall n \mathord: \mathsf{Nat}. C^\infty(Z \times X,Y) \to C^\infty(Z \times \mathrm{vec}~n~X, \mathrm{vec}~n~Y) \\
     \mathrm{cross} &:& 
@@
       \forall m,n \mathord: \mathsf{Nat}.\\
-      \quad C^\infty(X \times Y, Z) \to \\
-      \quad\quad C^\infty(\mathit{vec}~m~X \times \mathit{vec}~n~Y, \mathit{vec}~(m*n)~Z)
+      \quad\quad C^\infty(\mathit{vec}~m~X \times \mathit{vec}~n~Y, \mathit{vec}~(m*n)~(X \times Y))
     \end{array}
@@
 given, while $\mathrm{cross}$ takes the cartesian product of two
-vectors of elements of cartesian spaces, and maps the given smooth
-function over them. The multiplication of natural numbers $m*n$ is
-straightforwardly defined in terms of the $\mathit{natrec}$ recursion
-operator. We will use $\mathrm{cross}$ in Section~\ref{sec:examples}
-to define Lagrangians in terms of the interactions between systems of
-particles.
+vectors of elements of cartesian spaces, and $\mathrm{map}$ maps the
+given smooth function over a vector. The multiplication of natural
+numbers $m*n$ is straightforwardly defined in terms of the
+$\mathit{natrec}$ recursion operator. We will use $\mathrm{cross}$ in
+Section~\ref{sec:examples} to define Lagrangians in terms of the
+interactions between systems of particles.
 
@@
       \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle G_2, 0 \rangle}
-    {\Theta \sepbar \Gamma; \Delta \vdash e_1 * e_2 : \mathbb{R}^n\langle \mathit{scale}_n(G_1)G_2, 0 \rangle}
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 e_2 : \mathbb{R}^n\langle \mathit{scale}_n(G_1)G_2, 0 \rangle}
 
@@
   \end{mathpar}
-  FIXME: division, n-ary products (sum, zip, reduce)
+
+  \medskip
+  \textbf{Division, sum, and cross}
+  \begin{mathpar}
+    \inferrule*
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}\langle G_1, 0 \rangle \\
+      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}\langle G_2, 0 \rangle}
+    {\Theta \sepbar \Gamma; \Delta \vdash \frac{e_1}{e_2} : \mathbb{R}\langle G_1G_2^{-1}, 0 \rangle}
+
+    \inferrule*
+    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathit{vec}~N~(\mathbb{R}\langle G, 0 \rangle)}
+    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{sum}~e : \mathbb{R}\langle G, 0 \rangle}
+
+    \inferrule*
+    {\Theta \sepbar \Gamma; \Delta, x : X \vdash e_1 : Y \\
+      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathit{vec}~N~X}
+    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{map}~(x.~e_1)~e_2 : \mathit{vec}~N~Y}
+
+    \inferrule*
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathit{vec}~M~X \\
+      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathit{vec}~N~Y}
+    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{cross}~e_1~e_2 : \mathit{vec}~(M*N)~(X \times Y)}
+  \end{mathpar}
+
   \caption{A surface syntax for smooth functions}
@@
       \elab{e_1(e_2)}_\Delta &=& \elab{e_2}_\Delta \comp e_1 \\
-      \elab{\vec{c}}_\Delta &=& 
+      \elab{\vec{c}}_\Delta &=& \mathrm{const}~\vec{c} \\
+      \elab{0}_\Delta &=& \mathrm{const}~0 \\
+      \elab{e_1 + e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (+) \\
+      \elab{e_1 - e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (-) \\
+      \elab{e_1 e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (*) \\
+      \elab{e_1 \cdot e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (\cdot) \\
+      \elab{\sin~e}_\Delta &=& \elab{e}_\Delta \comp \sin \\
+      \elab{\cos~e}_\Delta &=& \elab{e}_\Delta \comp \cos \\
+      \elab{\mathrm{exp}~e}_\Delta &=& \elab{e}_\Delta \comp \mathrm{exp} \\
+      \elab{e_1 / e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (/) \\
+      \elab{\mathrm{sum}~e}_\Delta &=& \elab{e}_\Delta \comp \mathrm{sum} \\
+      \elab{\mathrm{map}~(x.~e_1)~e_2}_\Delta &=& \mathrm{pair}~\mathrm{id}~\elab{e_2}_\Delta \comp (\mathrm{map}~\elab{e_1}_{\Delta,x}) \\
+      \elab{\mathrm{cross}~e_1~e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp \mathrm{cross}
     \end{array}
@@
 
-FIXME: define the translation
-
-Define $\Delta$ and $\elab{\Delta}$.
-
-Define project
-
+The combinators we have presented allow the construction of smooth
+invariant functions, but doing so is painful due to the point-free
+style that they force. In order to be able to actually define readable
+Lagrangians in our calculus, we define a surface syntax for writing
+smooth functions that is desugared into our extended version of System
+F$\omega$.
+
+We define a typing judgement of smooth terms $\Theta \sepbar \Gamma;
+\Delta \vdash e : X$ by the rules in Figure~\ref{fig:smooth-terms},
+where $\Theta$ and $\Gamma$ are the kinding and typing contexts of our
+extension of F$\omega$, and $\Delta = x_1 : X_1, ..., x_n : X_n$ is a
+context of cartesian spaces. That is, for each $x : X$ in $\Delta$, we
+have $\Theta \vdash X : \mathsf{CartSp}$.
+
+The desugaring of smooth function terms into our extension of System
+F$\omega$ is via the operation $\elab{-}_\Delta$, defined in
+Figure~\ref{fig:desugar}. In this definition, we make use of two
+derived combinators for projection and pairing:
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    \underline{\mathrm{project}_{x : X, \Delta}}(x) & = & \mathrm{proj_1} \\
+    \underline{\mathrm{project}_{y : Y, \Delta}}(x) & = & \mathrm{proj_2} \comp \underline{\mathrm{project}_\Delta}(x) \quad (x \not= y)
+  \end{array}
+\end{displaymath}
+and
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
+    \underline{\mathrm{pair}_1}(e) & = & e \\
+    \underline{\mathrm{pair}_{n+1}}(e, \vec{e}) & = & \mathrm{pair}~e~(\underline{\mathrm{pair}_n}(\vec{e}))
+  \end{array}
+\end{displaymath}
+The following theorem follows easily by induction on the typing
+derivations of smooth terms, where $\elab{\Delta} = \elab{x_1 : X_1,
+  ..., x_n : X_n} = X_1 \times ... \times X_n$:
 \begin{theorem}
-  If $\Theta |\Gamma; \Delta \vdash e : X$ then $\Theta \sepbar
-  \Gamma \vdash \elab{e} : C^\infty(\elab{\Delta}, X)$.
+  If $\Theta |\Gamma; \Delta \vdash e : X$ then $\Theta |
+  \Gamma \vdash \elab{e}_\Delta : C^\infty(\elab{\Delta}, X)$.
 \end{theorem}
@@
 Kennedy \cite{atkey13abstraction} that uses
+
+Quantum programming languages
+
+Information effects
 

### 2013-07-12 mechanics/paper.tex
Final changes

@@
 
-Reynolds' theory of relational parametricity \cite{reynolds82} tells
-us that parametrically polymorphic programs automatically satisfy
-invariance properties. Such invariance properties are often called
-\emph{Free Theorems}, after Wadler \cite{wadler89theorems}, since they
-follow ``for free'' from the types of programs, rather than through
-detailed study of the program text itself. An illustrative example is
-the free theorem for programs $f$ with the following type:
+Reynolds' theory of relational parametricity \cite{reynolds83types}
+tells us that parametrically polymorphic programs automatically
+satisfy invariance properties. Such invariance properties are often
+called \emph{Free Theorems}, after Wadler \cite{wadler89theorems},
+since they follow ``for free'' from the types of programs, rather than
+through detailed study of the program text itself. An illustrative
+example is the free theorem for programs $f$ with the following type:
 \begin{displaymath}
@@
 technique much older than programming languages. In physics,
-\emph{Noether's theorem} \cite{noether} provides a general way to
+\emph{Noether's theorem} \cite{noether71} provides a general way to
 derive conservation laws for physical systems from invariance
@@
 Atkey, Johann and Kennedy \cite{atkey13abstraction}, building on older
-work by Kennedy \cite{kennedy94para}, have already presented a
+work by Kennedy \cite{kennedy97relational}, have already presented a
 polymorphic type system for expressing geometric invariance properties
@@
 other sources, such as Landau and Lifschitz \cite{landau60mechanics}
-or Arnol'd \cite{arnoldXXmathematical} for the necessary background in
+or Arnol'd \cite{arnold89mathematical} for the necessary background in
 the mathematical theory of classical mechanics. Lagrangian mechanics,
 and Noether's theorem, make crucial use of the Calculus of
-Variations. Gelfand and Fomin \cite{gelfandXXcalculus} provide a good
+Variations. Gelfand and Fomin \cite{gelfand00calculus} provide a good
 introduction to this topic.
@@
 Hamiltonian mechanics
-\cite{landau60mechanics,arnoldXXmathematical}. Analytical mechanics
+\cite{landau60mechanics,arnold89mathematical}. Analytical mechanics
 seeks to discover the underlying structure of Newtonian classical
@@
 standard references on the calculus of variations, such as Gelfand and
-Fomin \cite{gelfandXXcalculus}. The key point for our purposes is that
+Fomin \cite{gelfand00calculus}. The key point for our purposes is that
 a path $q$ satisfies the principle of stationary action if and only if
@@
 hence the appearance of the second derivatives $\ddot{q_i}$ in the
-equations of motion. Sussman and Wisdom \cite{sussman} criticise the
-standard presentation of Lagrangian mechanics for this kind of
-confusing notation, and define a clearer notation based on the Scheme
-programming language. We elect to stick with the standard notation as
-used in physics textbooks just for familiarity.
+equations of motion. Sussman and Wisdom \cite{sussman01structure}
+criticise the standard presentation of Lagrangian mechanics for this
+kind of confusing notation, and define a clearer notation based on the
+Scheme programming language. We elect to stick with the standard
+notation as used in physics textbooks just for familiarity.
 
@@
 \begin{proof}
-  E.g., Gelfand and Fomin \cite{gelfandXXcalculus}, Section 20.
+  E.g., Gelfand and Fomin \cite{gelfand00calculus}, Section 20.
 \end{proof}
 
-\begin{example}
+\begin{example}\label{ex:conservation-laws}
   In Example~\ref{ex:invariance-examples}, we noted that the
@@
 graph models of relational parametricity by Robinson and
-Rosolini~\cite{robinson94} and Hasegawa~\cite{hasegawa94}). In
-reflexive graph models of relational parametricity, the kinds of
-System F$\omega$ are interpreted as reflexive graphs, while types are
-interpreted as morphisms of reflexive graphs. As demonstrated by
-Atkey, uniformly interpreting kinds as reflexive graphs allows for a
-straightforward interpretation of higher kinds like $* \to *$. In the
-present setting, reflexive graphs will allow us to accomodate
-arbitrary groupoids as kinds
-(Section~\ref{sec:groupoid-kinds}). Groupoid kinds are the key
+Rosolini~\cite{robinson94reflexive} and
+Hasegawa~\cite{hasegawa94relational}). In reflexive graph models of
+relational parametricity, the kinds of System F$\omega$ are
+interpreted as reflexive graphs, while types are interpreted as
+morphisms of reflexive graphs. As demonstrated by Atkey, uniformly
+interpreting kinds as reflexive graphs allows for a straightforward
+interpretation of higher kinds like $* \to *$. In the present setting,
+reflexive graphs will allow us to accomodate arbitrary groupoids as
+kinds (Section~\ref{sec:groupoid-kinds}). Groupoid kinds are the key
 technical tool that we need to derive the kinds of geometric
@@
 the kind $*$, would force us to use an intuitionistic metatheory
-\cite{pitts}. The basic results of the calculus of variations that we
-are relying on, such as the derivation of the Euler-Lagrange
-equations, use excluded middle in their proofs, so an intuitionistic
-metatheory is not an option. We therefore just assume that we have a
-set of ``small'' sets, which includes the real numbers and is closed
-under subsets, products, function space and comprehension.
+\cite{pitts87polymorphism}. The basic results of the calculus of
+variations that we are relying on, such as the derivation of the
+Euler-Lagrange equations, use excluded middle in their proofs, so an
+intuitionistic metatheory is not an option. We therefore just assume
+that we have a set of ``small'' sets, which includes the real numbers
+and is closed under subsets, products, function space and
+comprehension.
 
@@
 presentation of System F$\omega$ is entirely standard (see, e.g.,
-Pierce~\cite{pierce01types}), except for the restriction to small
+Pierce~\cite{pierce02types}), except for the restriction to small
 kinds in the formation of universal kinds $\forall
@@
 technical problem with our current system is that we do not account
-for the non-definedness of division at $0$. However, division will be
-useful in many of our examples, so we take a pragmatic approach, in
-common with most physics textbooks, and informally treat division as
-total function.
+for the non-definedness of division at $0$, nor do we account for the
+square root of negative numbers. However, division and square root
+will be useful in our examples in Section~\ref{sec:examples}, so we
+take a pragmatic approach, in common with most physics textbooks, and
+informally treat them as total functions.
 \begin{displaymath}
@@
     \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, t \rangle, \mathbb{R}\langle \mathit{exp}~t, 0 \rangle) \\
-    (/) &:& \forall g_1, g_2 \mathord: \mathsf{GL}(1).~C^\infty(\mathbb{R}\langle g_1, 0 \rangle \times \mathbb{R}\langle g_2, 0\rangle, \mathbb{R}\langle g_1g_2^{-1}, 0 \rangle)
-  \end{array}
-\end{displaymath}
-
-Finally, we include two primitive combinators for dealing with $n$-ary
-products of vectors. Both of these combinators will be useful when
+    (/) &:& \forall g_1, g_2 \mathord: \mathsf{GL}(1).~C^\infty(\mathbb{R}\langle g_1, 0 \rangle \times \mathbb{R}\langle g_2, 0\rangle, \mathbb{R}\langle g_1g_2^{-1}, 0 \rangle) \\
+    \mathrm{sqrt} &:& \forall g \mathord: \mathsf{GL}(1).~C^\infty(\mathbb{R}\langle g^2, 0 \rangle, \mathbb{R}\langle g, 0 \rangle)
+  \end{array}
+\end{displaymath}
+
+Finally, we include three primitive combinators for dealing with
+$n$-ary products of vectors. These combinators will be useful when
 defining Lagrangians that are generic in the number of particles.
@@
   \medskip
-  \textbf{Division, sum, and cross}
+  \textbf{Division, square root, sum, and cross}
   \begin{mathpar}
@@
     {\Theta \sepbar \Gamma; \Delta \vdash \frac{e_1}{e_2} : \mathbb{R}\langle G_1G_2^{-1}, 0 \rangle}
+
+    \inferrule*
+    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle G^2, 0 \rangle}
+    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{sqrt}~e : \mathbb{R}\langle G, 0 \rangle}
 
@@
       \elab{e_1 / e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (/) \\
+      \elab{\mathrm{sqrt}~e}_\Delta &=& \elab{e}_\Delta \comp \mathrm{sqrt} \\
       \elab{\mathrm{sum}~e}_\Delta &=& \elab{e}_\Delta \comp \mathrm{sum} \\
@@
 
+Our first example is of a simple Lagrangian for a single free particle
+in $3$-dimensional space, with no external forces. We define the
+Lagrangian using the syntax for smooth terms we defined in the
+previous section, in the following kinding, typing and cartesian space
+contexts:
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}l}
+    \Theta = & t_t : \mathsf{T}(1), t_x : \mathsf{T}(3), o : \mathsf{O}(3) \\
+    \Gamma = & m : \typeOfCartSp{\mathbb{R}\langle 1, 0 \rangle} \\
+    \Delta = & t : \mathbb{R}\langle 1, t_t \rangle, x : \mathbb{R}^3\langle \mathit{ortho}_3(o), t_x \rangle, \dot{x} : \mathbb{R}^3\langle \mathit{ortho}_3(o), 0 \rangle
+  \end{array}
+\end{displaymath}
+where $o$ represents an arbitrary orthogonal transformation of the
+space, $t_t$ represents a translation in time, $t_x$ represents a
+$3$-dimensional translation in space, $m$ is the constant mass of the
+free particle, $t$ is the current time, $x$ is the current position of
+the particle, and $\dot{x}$ is the current velocity.
+
+We write the Lagrangian like so, which in the absence of external
+forces just consists of the kinetic energy term:
+\begin{displaymath}
+  L = \frac{1}{2}m(\dot{x} \cdot \dot{x}) : \mathbb{R}\langle 1, 0 \rangle
+\end{displaymath}
+
+By the fundamental theorem of logical relations for our calculus
+(Property~\ref{property:semantic-terms}), we can derive the following
+free theorems, one each for the three group parameters $t_t$, $t_x$
+and $o$. Quantifying over translations in time, $t_t$, gives us
+invariance under translation in time:
+\begin{displaymath}
+  \forall t_t \in \mathbb{R}.~\sem{L}(t + t_t, \vec{x}, \dot{\vec{x}}) = \sem{L}(t, \vec{x}, \dot{\vec{x}})
+\end{displaymath}
+As we saw in Example~\ref{ex:conservation-laws}, Noether's theorem
+tells us that invariance under translation in time yields conservation
+of energy. Likewise, quantifying over translations in space, $t_x$,
+gives us invariance under translation in space:
+\begin{displaymath}
+  \forall \vec{t_x} \in \mathbb{R}^3.~\sem{L}(t, \vec{x} + \vec{t_x}, \dot{\vec{x}}) = \sem{L}(t, \vec{x}, \dot{\vec{x}})
+\end{displaymath}
+Similar to the example of the spring-coupled particles in the
+introduction, invariance under translations in space yields
+conservation of linear momentum.
+
+Quantification over orthogonal transformations $o : \mathsf{O}(3)$
+yields the following free theorem, which states that the Lagrangian is
+invariant under modification of the positions and velocities by
+arbitrary orthogonal transformations $O$:
+\begin{displaymath}
+  \forall O \in \mathrm{O}(3).~\sem{L}(t, O\vec{x}, O\dot{\vec{x}}) = \sem{L}(t, \vec{x}, \dot{\vec{x}})
+\end{displaymath}
+We are particularly interested in the orthogonal transformations
+derived from rotations, since these will give us the continuous
+transformations we need to apply Noether's theorem. For example,
+rotation by an angle $\epsilon$ around the $x_3$-axis is given by the
+following $\epsilon$-indexed family of orthogonal matricies:
+\begin{displaymath}
+  O_\epsilon =
+  \left(
+    \begin{array}{ccc}
+      \cos \epsilon  & \sin \epsilon & 0 \\
+      -\sin \epsilon & \cos \epsilon & 0 \\
+      0 & 0 & 0
+    \end{array}
+  \right)
+\end{displaymath}
+This leads to the following $\epsilon$-indexed family of
+transformations, suitable for Noether's theorem:
+\begin{displaymath}
+  \Psi_\epsilon\left(\begin{array}{@{}c@{}} x_1 \\ x_2 \\ x_3 \end{array}\right) =
+  O_\epsilon \left(\begin{array}{@{}c@{}} x_1 \\ x_2 \\ x_3 \end{array}\right) =
+  \left(\begin{array}{@{}c@{}} x_1 \cos \epsilon + x_2 \sin \epsilon \\ - x_1 \sin \epsilon + x_2 \cos \epsilon \\ x_3 \end{array}\right)
+\end{displaymath}
+Plugging this into Equation~\ref{eq:conservation-law}, and using the
+definition of $L$ above, we obtain the following conservation law:
+\begin{displaymath}
+  \frac{d}{dt}(m\dot{x}x_2 - m\dot{x}x_1) = 0
+\end{displaymath}
+This is exactly conservation of angular momentum around the $x_3$-axis
+\cite{landau60mechanics}. By considering the families of orthogonal
+matrices for rotation around other axes, we can also derive
+conservation laws for angular momentum around arbitrary axes.
+
 \paragraph{Particle in a Potential Field}
 
-\subsection{$n$-Body Problems}
-
-\paragraph{Two Bodies}
-
-\paragraph{$n$-Bodies}
-
-\subsection{Pendulums}
-
-\paragraph{A single pendulum}
-
-\paragraph{Coupled pendulum}
-
-\paragraph{Pendulum on a Moving Base}
-
-\paragraph{Pendulum on a Rotating Base}
-
-No time translation invariance.
+If we now extend the typing context of our example with a arbitrary
+potential energy term, depending on the position of the particle, we
+can define the Lagrangian for a single particle acted upon by a
+potential field that depends on the position of the particle. The
+invariance properties of the whole system will depend upon the
+invariance properties of the potential field. If we assume the
+following contexts:
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}l}
+    \Theta = & t_t : \mathsf{T}(1), o : \mathsf{O}(3) \\
+    \Gamma = &
+    \begin{array}[t]{@{}l}
+      m : \typeOfCartSp{\mathbb{R}\langle 1, 0 \rangle}, \\
+      V : \forall o \mathord: \mathsf{O}(3).~C^\infty(\mathbb{R}^3\langle \mathit{ortho}_3(o), 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
+    \end{array}\\
+    \Delta = & t : \mathbb{R}\langle 1, t_t \rangle, x : \mathbb{R}^3\langle \mathit{ortho}_3(o), 0 \rangle, \dot{x} : \mathbb{R}^3\langle \mathit{ortho}_3(o), 0 \rangle
+  \end{array}
+\end{displaymath}
+where everything is as in the previous example, except for the
+addition of the polymophic potential energy function $V$. We
+incorporate this into our Lagrangian like so:
+\begin{displaymath}
+  L = \frac{1}{2}m(\dot{x} \cdot \dot{x}) - V(x) : \mathbb{R}\langle 1, 0 \rangle
+\end{displaymath}
+Even though $V$ is left abstract, the types of $t$, $x$, $\dot{x}$ and
+$L$ tell us that this Lagrangian is invariant under translations in
+time and also under all orthogonal transformations. Hence, by
+Noether's theorem, the system this Lagrangian describes has energy and
+angular momentum as conserved quantities.
+
+\subsection{The $n$-Body Problem}
+
+Using the constructs of our calculus for dealing with $n$-ary vectors
+of positions and velocities, we can compactly write Lagrangians that
+describe systems of $n$ particles interacting through Newtonian
+gravitational attraction. We use the following kinding, typing and
+cartesian space contexts, where the type-level parameter $n$
+represents the number of particles that we are considering. For
+simplicity, we assume that all the bodies involved in the system have
+equal mass $m$.
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}l}
+    \Theta = & n : \mathsf{Nat}, t_t : \mathsf{T}(1), t_x : \mathsf{T}(3), o : \mathsf{O}(3) \\
+    \Gamma = & m : \typeOfCartSp{\mathbb{R}\langle 1, 0 \rangle} \\
+    \Delta = &
+    \begin{array}[t]{@{}l}
+      t : \mathbb{R}\langle 1, t_t \rangle, \\
+      x : \mathit{vec}~n~(\mathbb{R}^3\langle \mathit{ortho}_3(o), t_x \rangle), \\
+      \dot{x} : \mathit{vec}~n~(\mathbb{R}^3\langle \mathit{ortho}_3(o), 0 \rangle)
+    \end{array}
+  \end{array}
+\end{displaymath}
+The Lagrangian for this system is defined as follows in our surface
+syntax for smooth terms. The kinetic energy component is the sum of
+the kinetic energies of all the particles in the system. The potential
+energy components sums up all the gravitational potential energies due
+to the interactions between each pair of bodies, where $G$ is the
+gravitational constant.
+\begin{displaymath}
+  L =
+  \begin{array}[t]{@{}l}
+    \frac{1}{2}m(\mathrm{sum}~(\mathrm{map}~(\dot{x_i}.~\dot{x_i}\cdot\dot{x_i}))~\dot{x}) - \\
+    \quad \mathrm{sum}~(\mathrm{map}~((x_i,x_j).~Gm^2/|x_i - x_j|)~(\mathrm{cross}~x~x)) : \mathbb{R}\langle 1, 0 \rangle
+  \end{array}
+\end{displaymath}
+We have used the notation $|e|$ as shorthand for $\mathrm{sqrt}~(e
+\cdot e)$, i.e., the norm of the vector $e$.
+
+Again, just by looking at the types of $t$, $x$, $\dot{x}$, and $L$,
+we can determine via free theorems that this Lagrangian is invariant
+under translation in time and space, and under all orthogonal
+transformations. Hence, by Noether's theorem, this system has energy
+and linear and angular momentum as conserved quantities.
+
+\subsection{Pendulum}
+
+All our examples above have used rectangular coordinate systems. Part
+of the power of the Lagrangian formulation of classical mechanics is
+the ability to use appropriate generalised coordinates to describe
+systems in simple terms. A standard example is that of a pendulum,
+where we take the angle of swing from the vertical, $\theta$ as the
+coordinate.
+
+We use the following kinding, typing and cartesian space contexts:
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}l}
+    \Theta = & t_t : \mathsf{T}(1), z : \mathsf{Z} \\
+    \Gamma = & m : \typeOfCartSp{\mathbb{R}\langle 1, 0 \rangle}, l : \typeOfCartSp{\mathbb{R}\langle 1, 0 \rangle} \\
+    \Delta = &
+    \begin{array}[t]{@{}l}
+      t : \mathbb{R}\langle 1, t_t \rangle, \theta : \mathbb{R}\langle 1, \underline{2\pi}*z \rangle. \dot{\theta} : \mathbb{R}\langle 1, 0 \rangle
+\end{array}
+  \end{array}
+\end{displaymath}
+where $m$ is the mass at the end of the pendulum (we assume the rod of
+the pendulum to be massless), $l$ is the length of the pendulum,
+$\theta$ is the angle of swing and $\dot{\theta}$ is the current rate
+of change of the angle of swing.
+
+Defining the Lagrangian for this system is a matter of simple
+trigonometry and differential calculus:
+\begin{displaymath}
+  L =
+  \begin{array}[t]{@{}l}
+    \textrm{let }y = l\sin \theta \textrm{ in} \\
+    \textrm{let }\dot{x} = l\dot{\theta}\cos \theta \textrm{ in} \\
+    \textrm{let }\dot{y} = - l\dot{\theta}\sin \theta \textrm{ in} \\
+    \quad \frac{1}{2}m(\dot{x}^2 + \dot{y}^2) - mgy : \mathbb{R}\langle 1, 0 \rangle
+  \end{array}
+\end{displaymath}
+From the type of $t$, we can easily determine that this Lagrangian is
+invariant under translation in time, and hence the total energy is a
+conserved quantity of this system. We can also derive another free
+theorem due to the quantification over the group of integers via the
+variable $z$. However, this does not yield a continuous invariance of
+the Lagrangian, and hence no conserved quantity via Noether's theorem.
 
@@
 introduction was of a pair of particles of equal mass coupled by a
-spring.
-
-\paragraph{Fixed Endpoint Oscillator} We now look at the slightly
-simpler case of a single particle connected to a spring attached to a
-fixed point.
-
-\paragraph{Damped Oscillator}
+spring, for which we derived conservation of linear momentum and
+energy. In terms of our type system, we have the following kinding,
+typing and cartesian space contexts, where $m$ is the individual mass
+of the two particles, and $k$ is the spring constant:
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}l}
+    \Theta = & t_t : \mathsf{T}(1), t_x : \mathsf{T}(1) \\
+    \Gamma = & m : \typeOfCartSp{\mathbb{R}\langle 1, 0 \rangle}, k : \typeOfCartSp{\mathbb{R}\langle 1, 0 \rangle} \\
+    \Delta = &
+    \begin{array}[t]{@{}l}
+      t : \mathbb{R}\langle 1, t_t \rangle, x_1 : \mathbb{R}\langle 1, t_x \rangle, x_2 : \mathbb{R}\langle 1, t_x \rangle, \\
+      \dot{x_1} : \mathbb{R}\langle 1, 0 \rangle, \dot{x_2} : \mathbb{R}\langle 1, 0 \rangle
+\end{array}
+  \end{array}
+\end{displaymath}
+In our surface syntax for smooth terms, the Lagrangian is written just
+as it was in the introduction:
+\begin{displaymath}
+  L = \frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) - \frac{1}{2}k^2(x_1 - x_2)^2 : \mathbb{R}\langle 1, 0 \rangle
+\end{displaymath}
+The free theorems for this term state the invariance in time and space
+translation:
+\begin{displaymath}
+  \begin{array}{@{}l}
+    \forall t_t : \mathbb{R}.~\sem{L}(t + t_t, x_1, x_2, \dot{x_1}, \dot{x_2}) = \sem{L}(t, x_1, x_2, \dot{x_1}, \dot{x_2}) \\
+    \forall t_x : \mathbb{R}.~\sem{L}(t, x_1 + t_x, x_2 + t_x, \dot{x_1}, \dot{x_2}) = \sem{L}(t, x_1, x_2, \dot{x_1}, \dot{x_2})
+  \end{array}
+\end{displaymath}
+As we saw in Section~\ref{sec:conservation-laws-from-symmetry}, these
+invariance properties allow us to deduce conservation of energy and
+linear momentum for this system.
+
+% \paragraph{Fixed Endpoint Oscillator} We now look at the slightly
+% simpler case of a single particle connected to a spring attached to a
+% fixed point.
+
+\paragraph{Damped Oscillator} All of the example Lagrangians we have
+looked at so far have had no dependence on time, and so they have all
+described systems with total energy as a conserved quantity. A system
+with a single particle attached to a damped spring (whose other
+endpoint is fixed at position $0$) provides an example of system that
+does not have energy as a conserved quantity. This example is taken
+from Neuenschwander's book \cite{neuenschwander11}.  We use the
+following kinding, typing and cartesian space contexts:
+\begin{displaymath}
+  \begin{array}{l@{\hspace{0.3em}}l}
+    \Theta = & t_t : \mathsf{T}(1) \\
+    \Gamma = & k : \typeOfCartSp{\mathbb{R}\langle 1, 0 \rangle} \\
+    \Delta = &
+    \begin{array}[t]{@{}l}
+      t : \mathbb{R}\langle 1, t_t + t_t \rangle, x : \mathbb{R}\langle \mathit{exp}(-t_t), 0 \rangle, \dot{x} : \mathbb{R}\langle \mathit{exp}(-t_t), 0 \rangle
+    \end{array}
+  \end{array}
+\end{displaymath}
+Here, the position and velocity of the single particle \emph{scale}
+with the exponential of the translation in time. This will account for
+the exponential damping we apply to the system. For simplicity, we
+assume that the mass of the particle and the spring constant are both
+$1$. The Lagrangian for this system looks similar to the coupled pair
+above, except for an additional damping factor:
+\begin{displaymath}
+  L = \left(\frac{1}{2}\dot{x}^2 - \frac{1}{2}x^2\right)\mathrm{exp}(t) : \mathbb{R}\langle 1, 0 \rangle
+\end{displaymath}
+From the types of $t$, $x$ and $\dot{x}$, we learn that this system is
+invariant under the families of transformations $\Phi_\epsilon(t) = t
++ 2\epsilon$ and $\Psi_\epsilon(x) = e^\epsilon x$. This leads to the
+following conservation law for this system, linking energy and linear
+momentum:
+\begin{displaymath}
+  \frac{d}{dt}\left[\left(\frac{1}{2}x\dot{x} + \frac{1}{2}\dot{x}^2 + \frac{1}{2}x^2\right)e^t\right] = 0
+\end{displaymath}
 
@@
 
-FIXME: sum up what's happened
+We have presented a type system for writing invariant Langrangians,
+with a relationally parametric semantics that allows the derivation of
+free theorems that can be used with Noether's theorem to discover
+conservation laws for classical mechanical systems. Our key technical
+contribution has been the observation that relationally parametric
+models of System F$\omega$ admit kinds that are interpreted as
+groupoids, allowing geometric invariance properties to be integrated
+directly into the model.
 
 \paragraph{Related Work} We are not aware of any other work linking
-type systems, programming languages or relational parametricity with
-invariance properties for classical mechanics, or with Noether's
-theorem. We have already mentioned the work of Atkey, Johann and
-Kennedy \cite{atkey13abstraction} that uses
-
-Quantum programming languages
-
-Information effects
-
-\paragraph{Future Work}
-
-\begin{enumerate}
-\item Gauge theory
-\item Quantum field theory
-\item Dependent types
-\item Homotopy type theory
-\item Distance indexed types and differential privacy.
-\item Languages with function spaces with more structure (Alex's
-  Enriched Effect Calculus) (Idea for differential privacy)
-\end{enumerate}
+type systems or relational parametricity with invariance properties
+for classical mechanics, or with Noether's theorem. We have already
+mentioned the work of Atkey, Johann and Kennedy
+\cite{atkey13abstraction} that uses relational parametricity with a
+specialised type system to derive geometric invariance properties,
+albeit in a setting without smooth functions, and without an
+application to Noether's theorem. In this paper, we have placed Atkey
+\emph{et al.}'s work in a more general setting by framing it as an
+extension of System F$\omega$, which allows us to (a) incorporate
+type-level computation, as we used to define the $n$-ary vectors of
+coordinates; and (b) to reuse previous work on relationally parametric
+models of System F$\omega$ by Atkey \cite{atkey12relational}.
+
+Sussman and Wisdom \cite{sussman01structure} reformulate Lagrangian
+and Hamiltonian mechanics in a more programming language style so that
+it can be implemented within the programming language Scheme. In doing
+so, they clear up some of the ambiguous syntax common in the standard
+presentations of classical mechanics. However, they do not attempt to
+incorporate invariance properties into types, as we have done here.
+
+There has been prior work on programming languages inspired by
+theoretical physics. Quantum programming languages, which hope to
+exploit the properties of quantum computation in an understandable way
+have been the subject of study for several years now. We cite Abramsky
+\cite{abramsky04} as an introduction to this field.
+
+\paragraph{Future Work} We have barely scratched the surface of the
+applicability of Noether's theorem to theoretical
+physics. Conservation laws for classical and quantum field theories
+are derivable from generalisations of Noether's
+theorem. Neuenschwander's book \cite{neuenschwander11} describes in an
+easy to read way how Noether's theorem applies to field theories,
+allowing for relativistic and quantum systems to be studied, and how
+it naturally leads to the notion of Gauge invariance, an important
+principle in modern theories of particle physics.
+
+From the type system point of view, it seems clear that greater
+expressiveness will be required in order to accomodate more detailed
+physical theories. Study of relational parametricity for dependent
+types, as done for example by Bernardy \emph{et al.}
+\cite{bernardy12proofs} may be useful here. Also, recent work on
+homotopy type theory and, in particular $\infty$-groupoid models of
+type theory is almost certainly relevant \cite{hottbook2013}.
+
+For each of the examples in Section~\ref{sec:examples}, we computed
+the necessary typing derivations by hand. Type inference for the
+system was have presented here, or some variant of it, is a key item
+of future work. Ideally, the user would enter the description of a
+Lagrangian, and the system would be able to tell them the free
+theorems, and consequent conservation laws that hold. Kennedy
+\cite{kennedy94} and Gundry \cite{gundry11} have studied type
+inference for the related, but simpler, setting of dimension types.
+
+\begin{thebibliography}{12}
+\softraggedright
+
+\bibitem{abramsky04}
+S.~Abramsky. High-Level Methods for Quantum Computation and Information.
+Proceedings, LICS, 2004.
+
+\bibitem{arnold89mathematical} V.~I.~Arnol'd. {\em Mathematical Methods of
+  Classical Mechanics}. Springer, 1989.
+
+\bibitem{atkey12relational}
+R.~Atkey. Relational Parametricity for Higher Kinds.
+Proceedings, CSL, 2012.
+
+\bibitem{atkey13abstraction}
+R.~Atkey, P.~Johann, and A.~J.~Kennedy. Abstraction and Invariance for Algebraically Indexed Types.
+Proceedings, POPL, 2013.
+
+\bibitem{bernardy12proofs} 
+J.-P.~Bernardy, P.~Jansson, R.~Paterson.
+Proofs for Free: Parametricity for Dependent Types.
+{\em Journal of Functional Programming} 22(2), pp. 107-152, 2012.
+
+\bibitem{gelfand00calculus} I.~M.~Gelfand and S.~V.~Fomin,
+  R.~A.~Silverman (ed.). {\em Calculus of Variations}. Dover Publications,
+  2000.
+
+\bibitem{gundry11} A.~Gundry. Type Inference for Units of
+  Measure. Technical Report, University of Strathclyde, 2011.
+
+\bibitem{hasegawa94relational} R. Hasegawa.  Relational Limits in
+  General Polymorphism.  {\em Publications of the Research Institute
+    for Mathematical Sciences} 30, pp. 535--576, 1994.
+
+\bibitem{kennedy94}
+A.~J.~Kennedy. Dimension Types.
+Proceedings, ESOP, 1994.
+
+\bibitem{kennedy97relational} 
+A.~J.~Kennedy. Relational Parametricity and Units of Measure.
+Proceedings, POPL, pp. 442-455, 1997.
+
+\bibitem{landau60mechanics} L.~D.~Landau and
+  E.~M.~Lifschitz. {\em Mechanics}. Pergamon Press. 1967.
+
+\bibitem{neuenschwander11}
+D.~E.~Neuenschwander. {\em Emmy Noether's Wonderful Theorem}.
+The John Hopkins University Press, 2011.
+
+\bibitem{noether71} E.~Noether, M.~Tavel (translator). Invariant
+  Variation Problems.  {\em Transport Theory and Statistical Physics}
+  1(3), pp. 186-207, 1971. Original in {\em Gott. Nachr.},
+  1918:235-257, 1918.
+
+\bibitem{maclane98categories} S.~Mac~Lane. {\em Categories for the Working
+  Mathematician}, 2nd edition. Springer, 1998.
+
+\bibitem{pierce02types} B.~Pierce. {\em Types and Programming
+  Languages}. MIT Press, 2002.
+
+\bibitem{pitts87polymorphism} A.~M.~Pitts.  Polymorphism is Set
+  Theoretic, Constructively.  {\em Proc., Category Theory and Computer
+    Science}, pp. 12--39, 1987.
+
+\bibitem{reynolds83types}
+J.~C.~Reynolds. Types, Abstraction and Parametric Polymorphism.
+{\em Information Processing} 83, pp. 513-523, 1983.
+
+\bibitem{robinson94reflexive} E. Robinson and G. Rosolini.  Reflexive
+Graphs and Parametric Polymorphism.  {\em Proc., Logic in Computer
+  Science}, pp. 364--371, 1994.
+
+\bibitem{sussman01structure} G.~J.Sussman and J.~Wisdom. Structure and
+  Interpretation of Classical Mechanics. MIT Press, 2001
+
+\bibitem{hottbook2013} The Univalent Foundations Program. {\em
+    Homotopy Type Theory}. Institute for Advanced Study, 2013.
+
+\bibitem{wadler89theorems}
+P.~Wadler.  Theorems for Free!.  Proceedings, FPCA,
+pp. 347-359, 1989.
+
+\end{thebibliography}
 

### 2013-11-09 mechanics/paper.tex
Final POPL2014 version

@@
-\documentclass[preprint]{sigplanconf}
+\documentclass{sigplanconf}
+
+\pdfpagewidth=8.5in
+\pdfpageheight=11in
 
@@
 
-\authorinfo{}{}{}
-% \authorinfo{Robert Atkey}
-%            {}
-%            {bob.atkey@gmail.com}
+\authorinfo{Robert Atkey}{}{bob.atkey@gmail.com}
 
@@
 
+\newcommand{\abs}[1]{\lvert #1 \rvert}
+
 \newtheoremstyle{examplestyle}
@@
 
+\exclusivelicense
+\conferenceinfo{POPL~'14}{January 22--24, 2014, San Diego, CA, USA}
+\copyrightyear{2014}
+\copyrightdata{978-1-4503-2544-8/14/01}
+\doi{2535838.2535867}
+
 \maketitle
@@
   deep properties of physical systems. For example, the conservation
-  of energy, which by Noether's theorem is a consequence of a system's
+  of energy is by Noether's theorem a consequence of a system's
   invariance under time-shifting.
@@
 
-\category{CR-number}{subcategory}{third-level}
+\category{D.3.1}{Programming Languages}{Formal Definitions and Theory}
+\category{J.2}{Physical Sciences and Engineering}{Physics}
 
 \terms
-term1, term2
+  Languages, Theory, Types, Physics
 
 \keywords
-keyword1, keyword2
+  classical mechanics; relational parametricity; higher-kinded types; invariance; conservation laws
 
@@
 since they follow ``for free'' from the types of programs, rather than
-through detailed study of the program text itself. An illustrative
-example is the free theorem for programs $f$ with the following type:
+through detailed study of the program text. An illustrative example is
+the free theorem for programs $f$ with the following type:
 \begin{displaymath}
@@
 \end{displaymath}
-Such programs takes lists of $\alpha$s, for any type $\alpha$, to
-natural numbers. Using Reynolds' theory of relational parametricity,
-Wadler showed that \emph{any} $f$ with this type satisfies the
-following property:
+Such programs take lists of $\alpha$s, for any type $\alpha$, and
+return natural numbers. Using Reynolds' theory of relational
+parametricity, Wadler showed that \emph{any} $f$ with this type
+satisfies the following property:
 \begin{displaymath}
@@
 \emph{Lagrangians}. For classical mechanics, a Lagrangian is a
-function of time, and the positions and velocities of all the
+function of time and the positions and velocities of all the
 particles in the system, to the difference between the kinetic energy
@@
 differential equations (ODEs) that describe how the positions and
-velocities of the particles evolve over time (the second deivative
+velocities of the particles evolve over time (the second derivative
 with respect to time, $\ddot{x_i}$, denotes the acceleration of the
@@
 results in no change in the output. Since Reynolds' relational
-parametricity was allowed us to derive the invariance properties for
-the program $f$ just by looking at its type, we ask the following
+parametricity allowed us to derive the invariance properties for the
+program $f$ just by looking at its type, we ask the following
 question:
@@
 We explain this type by breaking it down from left to right. The
-quantifier $\forall x \mathord: \mathsf{T}(1)$ indicates that we are
+quantifier $\forall y \mathord: \mathsf{T}(1)$ indicates that we are
 quantifying over all translations $y$ in one-dimensional space, just
-as we did implicity in Equation
+as we did implicitly in Equation
 (\ref{eq:coupled-spring-translation-invariance}). The notation
@@
 \rangle$ varies with the translation $y$. In the smooth function type
-above, in domain space consists of vectors of five real numbers: the
+above, the domain space consists of vectors of five real numbers: the
 time, the two particle positions, which vary with the translation $y$,
@@
 
-Atkey, Johann and Kennedy \cite{atkey13abstraction}, building on older
+Atkey, Johann and Kennedy \cite{atkey13abstraction}, building on
 work by Kennedy \cite{kennedy97relational}, have already presented a
@@
   group-indexed types in Atkey \emph{et al.}'s work can be
-  incorporated into a standard framework for presenting indexed types.
+  incorporated into a standard framework for indexed types.
 \item We present a relationally parametric semantics of our extension
@@
   suitably generalised to indexed types by considering reflexive
-  graphs, already accomodates geometric invariance properties.
+  graphs, already accommodates geometric invariance properties.
 \item Finally, we connect the free theorems derivable from our
@@
   consequent conservation laws for physical systems. In each case, the
-  necessary invariance property is derived from the type we assign to
-  the Lagrangian.
+  necessary invariance property is derived from the type of the
+  Lagrangian.
 \end{enumerate}
@@
   Section~\ref{sec:conservation-laws-from-symmetry}, in
-  Section~\ref{sec:symmetry-from-types}, we show how to derive
+  Section~\ref{sec:symmetry-from-types} we show how to derive
   invariance properties from indexed types. We present a relationally
@@
   invariance properties can be accomodated within this semantics.
-\item In Section~\ref{sec:types-for-classical-mech}, we construct a
-  type system for classical mechanics, by extending System F$\omega$
+\item In Section~\ref{sec:types-for-classical-mech} we construct a
+  type system for classical mechanics by extending System F$\omega$
   with the necessary constants for smooth function types and
@@
 like so, with abuse of notation for $\dot{q}$ which is a variable
-here, not a derivative of a function:
+here, not the derivative of a function:
 \begin{displaymath}
@@
 $\dot{q_i}$ and $q_i$ \emph{as variables}. The derivative with respect
-to time, $\frac{d}{dt}$ then treats $\dot{q_i}$ as a function of time,
-hence the appearance of the second derivatives $\ddot{q_i}$ in the
-equations of motion. Sussman and Wisdom \cite{sussman01structure}
+to time, $\frac{d}{dt}$, then treats $\dot{q_i}$ as a function of
+time, hence the appearance of the second derivatives $\ddot{q_i}$ in
+the equations of motion. Sussman and Wisdom \cite{sussman01structure}
 criticise the standard presentation of Lagrangian mechanics for this
@@
     \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
-    = \int_{\Phi(a)}^{\Phi(b)} L(s, q^*(s), \dot{q^*}(s)) \mathit{ds}
+    = \int_{\Phi(a)}^{\Phi(b)} L(s, q^*(s), \frac{d q^*}{ds}(s)) \mathit{ds}
 \end{equation}
@@
 \begin{displaymath}
-  \int_a^b L(\Phi(t), q^*(\Phi(t)), \dot{q^*}(\Phi(t)))\cdot \dot{\Phi}(t) \mathit{dt}
+  \int_a^b L(\Phi(t), q^*(\Phi(t)), \frac{d q^*}{ds}(\Phi(t)))\cdot \abs{\dot{\Phi}(t)} \mathit{dt}
 \end{displaymath}
@@
 \begin{equation}\label{eq:simplified-invariance}
-  L(t,q(t),\dot{q}(t)) = L(\Phi(t), q^*(\Phi(t)), \dot{q^*}(\Phi(t)))\cdot \dot{\Phi}(t)
+  L(t,q(t),\dot{q}(t)) = L(\Phi(t), q^*(\Phi(t)), \frac{d q^*}{ds}(\Phi(t)))\cdot \abs{\dot{\Phi}(t)}
 \end{equation}
@@
   invariant under translations along the $x$-axis, however. Both
-  Lagrangians are also invariant in translations in time, a special
+  Lagrangians are also invariant under translations in time, a special
   case of Equation~(\ref{eq:simplest-invariance}):
@@
 
-\begin{theorem}[Noether]
+\begin{theorem}[Noether]\label{thm:noether}
   If the action
@@
 \end{proof}
+
+% Theorem \ref{thm:noether} is a special case of what is known as
+% Noether's ``first theorem''. The general case of Theorem
+% \ref{thm:noether} applies to situations where the action is an
+% integral over regions of $n$-dimensional space, and the
+% transformations $\Psi$ can also depend on time. Noether's second
+% theorem generalises further to the case when there are infinitely many
+% parameters $\epsilon$. This is the case when considering local
+% symmetries that vary at different points in spacetime.
 
@@
 quantification over small kinds. In particular, the kind of types $*$
-is \emph{not} small. The reason for this is that we desire a simple
-set-theoretic model. However, requiring that we have a set of sets
-closed under large products, in order to interpret quantification over
-the kind $*$, would force us to use an intuitionistic metatheory
-\cite{pitts87polymorphism}. The basic results of the calculus of
-variations that we are relying on, such as the derivation of the
+is \emph{not} small. The reason for this choice is that we desire a
+simple set-theoretic model. However, requiring that we have a set of
+sets closed under large products, in order to interpret quantification
+over the kind $*$, would force us to use an intuitionistic metatheory
+\cite{pitts87polymorphism}. The basic results of the Calculus of
+Variations that we are relying on, such as the derivation of the
 Euler-Lagrange equations, use excluded middle in their proofs, so an
@@
 that we have a set of ``small'' sets, which includes the real numbers
-and is closed under subsets, products, function space and
-comprehension.
+and is closed under subsets, products, function spaces and
+set comprehensions.
 
@@
 Section~\ref{sec:discrete-kinds} with type-level natural numbers, and
-in Section~\ref{sec:groupoid-kinds}
+in Section~\ref{sec:groupoid-kinds} with groupoid kinds.
 
@@
 F$\omega$ are shown in Figures~\ref{fig:types},
-\ref{fig:type-equality} and~\ref{fig:terms}, respectively. Note that
-typing contexts $\Gamma$ only contain well-kinded types of kind
-$*$. Terms also have an equational theory, with $\beta\eta$-laws for
-functions, products, and universal quantification, which we omit. Our
-presentation of System F$\omega$ is entirely standard (see, e.g.,
-Pierce~\cite{pierce02types}), except for the restriction to small
-kinds in the formation of universal kinds $\forall
-\alpha\mathord:\kappa. A$. As is the case for kinds, we will extend
-the types, type equalities and terms in
+\ref{fig:type-equality} and~\ref{fig:terms}, respectively. Kinding
+contexts $\alpha_1 : \kappa_1, ..., \alpha_n : \kappa_n$ are denoted
+by $\Theta$ and typing contexts $x_1 : A_1, ..., x_n : A_n$ are
+denoted by $\Gamma$. Note that typing contexts $\Gamma$ only contain
+well-kinded types of kind $*$. Terms also have an equational theory,
+with $\beta\eta$-laws for functions, products, and universal
+quantification, which we omit. Our presentation of System F$\omega$ is
+entirely standard (see, e.g., Pierce~\cite{pierce02types}), except for
+the restriction to small kinds in the formation of universal kinds
+$\forall \alpha\mathord:\kappa. A$. As is the case for kinds, we will
+extend the types, type equalities, and terms in
 Sections~\ref{sec:discrete-kinds},~\ref{sec:groupoid-kinds}
@@
 \paragraph{Reflexive Graphs and the Interpretation of Kinds} We will
-interpret every kind $\kappa$ is as a reflexive graph, which we now
+interpret every kind $\kappa$ as a reflexive graph, which we now
 define. A \emph{reflexive graph} is a triple $(O, R, \mathrm{id})$,
@@
 a small set of directed `edges' to each pair of objects, and
-$\mathrm{id} : \forall o \in O.~R(o,o)$ assigns a distinguished
+$\mathrm{id} : \Pi o \in O.~R(o,o)$ assigns a distinguished
 `identity' edge from every object to itself. We think of the edges of
@@
 Property~\ref{property:semantic-kinds}. At base kind, the collection
-of objects is simply the type of all small sets; edges between $A$ and
-$B$ are binary relations on $A$ and $B$ (i.e., subsets of $A \times
-B$); and the distinguished identity edge is exactly the equality
-relation:
+of objects is simply the (large) set of all small sets; edges between
+$A$ and $B$ are binary relations on $A$ and $B$ (i.e., subsets of $A
+\times B$); and the distinguished identity edge is exactly the
+equality relation:
 \begin{displaymath}
@@
       (& \{ (f,r) \sepbar (f, r) : \sem{\kappa_1} \to \sem{\kappa_2} \},\\
-      &((f,r),(f',r')) \mapsto \forall o,o'. \semKR{\kappa_1}(o,o') \to \semKR{\kappa_2}(f~o, f'~o'), \\
+      &((f,r),(f',r')) \mapsto \Pi o,o'. \semKR{\kappa_1}(o,o') \to \semKR{\kappa_2}(f~o, f'~o'), \\
       &(f,r) \mapsto r\ )
@@
   reflexive graph morphism $\sem{A} : \sem{\Theta} \to \sem{\kappa}$,
-  such that if $\Theta \vdash A \equiv B : \kappa$, then $\sem{A} =
+  such that if $\Theta \vdash A \equiv B : \kappa$ then $\sem{A} =
   \sem{B}$.
@@
 of the basic type constructors for function and product types, and
-universal quantification. Each of these constructs builds an object of
+of universal quantification. Each of these constructs builds an object of
 kind $*$, so the object-level interpretation is a small set in $\Set$,
@@
 well-defined function interpreting each well-typed term, with the
-property that it takes related environments to related results, this
+property that it takes related environments to related results; this
 is the fundamental theorem of logical relations for System F$\omega$:
@@
 always preserve isomorphisms, so morphisms of groupoids are just
-functors. Every group is a groupoid with one object and a morphism for
-every element of the group: composition is the group operation,
-inverses are given by the group inverses, and the identity morphism is
-given by the group unit. Homomorphisms between groups are in
-one-to-one correspondence with functors between the corresponding
-groupoids. An example of a groupoid that is not a group is given by
-the collection of cartesian spaces (i.e., $\mathbb{R}^n$ for some
-$n$), with all diffeomorphisms (smooth invertible functions) as the
-morphisms.
+functors. Every group is a groupoid with one object, and a morphism
+from this object to itself for every element of the group: composition
+is the group operation, inverses are given by the group inverses, and
+the identity morphism is given by the group unit. Homomorphisms
+between groups are in one-to-one correspondence with functors between
+the corresponding groupoids. An example of a groupoid that is not a
+group is given by the collection of cartesian spaces (i.e.,
+$\mathbb{R}^n$ for some $n$), with all diffeomorphisms (smooth
+functions with smooth inverses) as the morphisms.
 
@@
 \begin{displaymath}
-  \sem{\mathsf{Z}} = (\{*\}, \lambda *, *.~\mathbb{Z}, \lambda \mathord*.~0)
-\end{displaymath}
-Note the difference between this interpretation, and the
-interpretation of the kind $\mathsf{Nat}$ in the previous section. In
-the interpretation of $\mathsf{Nat}$, the objects were complex and the
+  \sem{\mathsf{Z}} = (\{*\}, \lambda (*, *).~\mathbb{Z}, \lambda \mathord*.~0)
+\end{displaymath}
+Note the difference between this interpretation and the interpretation
+of the kind $\mathsf{Nat}$ in the previous section. In the
+interpretation of $\mathsf{Nat}$, the objects were complex and the
 relations were trivial. The reflexive graph interpreting $\mathsf{Z}$
@@
 
-\paragraph{Translation and Matrix Groups} The group of integers is not
-suitable for stating the invariance properties we require to apply
-Noether's theorem. To do so, we require groups of translations and
-invertible linear transformations. We extend our type system with
-small kinds representing $n$-dimensional ($n \geq 0$) translations,
-invertible linear transformations, and orthogonal transformations:
+\paragraph{Translation and Linear Transformation Groups} The group of
+integers is not suitable for stating the invariance properties we
+require to apply Noether's theorem. To do so, we require groups of
+translations and invertible linear transformations. We extend our type
+system with small kinds representing $n$-dimensional ($n \geq 0$)
+translations, invertible linear transformations, and orthogonal
+transformations:
 \begin{mathpar}
@@
   \begin{array}{@{}l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
-    \sem{\mathsf{T}(n)} &=& (\{*\}, \lambda *,*.~\mathbb{R}^n, \lambda \mathord*.~\vec{0}) \\
-    \sem{\mathsf{GL}(n)} &=& (\{*\}, \lambda *,*.~\mathrm{GL}(n), \lambda \mathord*. I) \\
-    \sem{\mathsf{O}(n)} &=& (\{*\}, \lambda *,*.~\mathrm{O}(n), \lambda \mathord*. I)
-  \end{array}
-\end{displaymath}
-We add the group operations and group axioms for each of these
-groupoid kinds in the same way as we did for the kind $\mathsf{Z}$
-above, except that we use multiplicative notation for the kinds
-$\mathsf{GL}(n)$ and $\mathsf{O}(n)$. Note that, except for
-$\mathsf{GL}(1)$, the groups $\mathsf{GL}(n)$ and $\mathsf{O}(n)$ are
-not abelian.
+    \sem{\mathsf{T}(n)} &=& (\{*\}, \lambda (*,*).~\mathbb{R}^n, \lambda \mathord*.~\vec{0}) \\
+    \sem{\mathsf{GL}(n)} &=& (\{*\}, \lambda (*,*).~\mathrm{GL}(n), \lambda \mathord*. I) \\
+    \sem{\mathsf{O}(n)} &=& (\{*\}, \lambda (*,*).~\mathrm{O}(n), \lambda \mathord*. I)
+  \end{array}
+\end{displaymath}
+Here, we take $\mathrm{GL}(n)$ to be exactly the group of invertible
+real $n\times n$ matricies, and $\mathrm{O}(n)$ to be the group of
+$n\times n$ orthogonal matricies (recall that an orthogonal matrix is
+one whose transpose is equal to its inverse).  We add the group
+operations and group axioms for each of these groupoid kinds in the
+same way as we did for the kind $\mathsf{Z}$ above, except that we use
+multiplicative notation for the kinds $\mathsf{GL}(n)$ and
+$\mathsf{O}(n)$. Note that, except for $\mathsf{GL}(0)$,
+$\mathsf{GL}(1)$, $\mathsf{O}(0)$ and $\mathsf{O}(1)$, the groups
+$\mathsf{GL}(n)$ and $\mathsf{O}(n)$ are not abelian.
 
@@
 the natural number $n$. This gives us a canonical representation of
-$n$-dimensional space for each $n$, avoids tricky problems with
+$n$-dimensional space for each $n$, avoiding tricky problems with
 isomorphic constructions of the same space.
 
-Cartesian spaces are closed under taking their product:
+Cartesian spaces are closed under products:
 \begin{displaymath}
@@
 \begin{displaymath}
-  \mathit{vec} = \lambda n\mathord:\mathsf{Nat}, x\mathord:\mathsf{CartSp}.~\mathit{natrec}~n~\mathbb{R}^0~(\lambda x.~X \times x)
+  \mathit{vec} = \lambda n\mathord:\mathsf{Nat}, X\mathord:\mathsf{CartSp}.~\mathit{natrec}~n~\mathbb{R}^0~(\lambda x.~X \times x)
 \end{displaymath}
@@
 Section~\ref{sec:conservation-laws-from-symmetry}. Moreover, $\sem{e}$
-is a smooth function from $\mathbb{R}^n$ to $\mathbb{R}$, just we
-require to for Lagrangian mechanics.
+is a smooth function from $\mathbb{R}^n$ to $\mathbb{R}$, just as we
+require for Lagrangian mechanics.
 
@@
     \mathrm{id} &:& C^\infty(X,X) \\
-    (\comp) &:& C^\infty(X,Y) \to C^\infty(Y,Z) \\
+    (\comp) &:& C^\infty(X,Y) \to C^\infty(Y,Z) \to C^\infty(X,Z) \\
     \mathrm{pair} &:& C^\infty(X,Y) \to C^\infty(X,Z) \to C^\infty(X, Y \times Z) \\
@@
       \forall g_1 \mathord: \mathsf{GL}(1), g_2 \mathord: \mathsf{GL}(n).\\
-      \quad C^\infty(\mathbb{R}\langle g_1, 0 \rangle \times \mathbb{R}^n\langle g_2, 0\rangle, \mathbb{R}\langle \mathit{scale}_n(g_1)g_2, 0 \rangle)
+      \quad C^\infty(\mathbb{R}\langle g_1, 0 \rangle \times \mathbb{R}^n\langle g_2, 0\rangle, \mathbb{R}^n\langle \mathit{scale}_n(g_1)g_2, 0 \rangle)
     \end{array}
@@
       C^\infty(\mathbb{R}^n\langle (\mathit{scale}_n~g)(\mathit{ortho}_n~o), 0 \rangle \times \\
-      \quad\quad\; \mathbb{R}^n\langle (\mathit{scale}_n~g)(\mathit{ortho}_n~o), 0\rangle, \mathbb{R}\langle (\mathit{scale}_n~g)^n, 0 \rangle)
+      \quad\quad\; \mathbb{R}^n\langle (\mathit{scale}_n~g)(\mathit{ortho}_n~o), 0\rangle, \mathbb{R}\langle (\mathit{scale}_n~g)^2, 0 \rangle)
     \end{array}
@@
 relations is maintained for our system (i.e., that
-Property~\ref{property:semantic-terms} is maintained), follows
+Property~\ref{property:semantic-terms} is maintained) follows
 directly from elementary facts about linear algebra.
@@
 each of these functions describes some of their invariance properties,
-for example the periodicity of the $\sin$ and $\cos$ functions. A
+such as the periodicity of the $\sin$ and $\cos$ functions. A
 technical problem with our current system is that we do not account
@@
     (/) &:& \forall g_1, g_2 \mathord: \mathsf{GL}(1).~C^\infty(\mathbb{R}\langle g_1, 0 \rangle \times \mathbb{R}\langle g_2, 0\rangle, \mathbb{R}\langle g_1g_2^{-1}, 0 \rangle) \\
-    \mathrm{sqrt} &:& \forall g \mathord: \mathsf{GL}(1).~C^\infty(\mathbb{R}\langle g^2, 0 \rangle, \mathbb{R}\langle g, 0 \rangle)
+    \mathrm{sqrt} &:& \forall g \mathord: \mathsf{GL}(1).~C^\infty(\mathbb{R}\langle g\cdot g, 0 \rangle, \mathbb{R}\langle g, 0 \rangle)
   \end{array}
@@
       \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle \mathit{scale}_n(G)\mathit{ortho}_n(O), 0 \rangle}
-    {\Theta \sepbar \Gamma; \Delta \vdash e_1 \cdot e_2 : \mathbb{R}\langle\mathit{scale}_n(G)^n, 0 \rangle}
+    {\Theta \sepbar \Gamma; \Delta \vdash e_1 \cdot e_2 : \mathbb{R}\langle\mathit{scale}_n(G)^2, 0 \rangle}
   \end{mathpar}
@@
       -\sin \epsilon & \cos \epsilon & 0 \\
-      0 & 0 & 0
+      0 & 0 & 1
     \end{array}
@@
 
-If we now extend the typing context of our example with a arbitrary
-potential energy term, depending on the position of the particle, we
-can define the Lagrangian for a single particle acted upon by a
-potential field that depends on the position of the particle. The
-invariance properties of the whole system will depend upon the
-invariance properties of the potential field. If we assume the
-following contexts:
+If we now extend the typing context of our example with an arbitrary
+potential energy function $V$, depending on the position of the
+particle, we can define the Lagrangian for a single particle acted
+upon by a potential field that depends on the position of the
+particle. The invariance properties of the whole system will depend
+upon the invariance properties of the potential field. If we assume
+the following contexts:
 \begin{displaymath}
@@
 systems in simple terms. A standard example is that of a pendulum,
-where we take the angle of swing from the vertical, $\theta$ as the
+where we take the angle $\theta$ of swing from the vertical as the
 coordinate.
@@
     \begin{array}[t]{@{}l}
-      t : \mathbb{R}\langle 1, t_t \rangle, \theta : \mathbb{R}\langle 1, \underline{2\pi}*z \rangle. \dot{\theta} : \mathbb{R}\langle 1, 0 \rangle
+      t : \mathbb{R}\langle 1, t_t \rangle, \theta : \mathbb{R}\langle 1, \underline{2\pi}*z \rangle, \dot{\theta} : \mathbb{R}\langle 1, 0 \rangle
 \end{array}
@@
 theoretical physics. Quantum programming languages, which hope to
-exploit the properties of quantum computation in an understandable way
-have been the subject of study for several years now. We cite Abramsky
-\cite{abramsky04} as an introduction to this field.
+exploit the properties of quantum computation in an understandable
+way, have been the subject of study for several years now. We cite
+Abramsky \cite{abramsky04} as an introduction to this field.
 
@@
 physics. Conservation laws for classical and quantum field theories
-are derivable from generalisations of Noether's
-theorem. Neuenschwander's book \cite{neuenschwander11} describes in an
-easy to read way how Noether's theorem applies to field theories,
-allowing for relativistic and quantum systems to be studied, and how
-it naturally leads to the notion of Gauge invariance, an important
-principle in modern theories of particle physics.
+are derivable from the much more general theorem that Noether
+originally proved \cite{noether71}. Neuenschwander's book
+\cite{neuenschwander11} describes in an easy to read way how Noether's
+theorem applies to field theories, allowing for relativistic and
+quantum systems to be studied, and how it naturally leads to the
+notion of Gauge invariance, an important principle in modern theories
+of particle physics.
 
 From the type system point of view, it seems clear that greater
-expressiveness will be required in order to accomodate more detailed
-physical theories. Study of relational parametricity for dependent
-types, as done for example by Bernardy \emph{et al.}
-\cite{bernardy12proofs} may be useful here. Also, recent work on
+precision in the types will be required in order to accomodate more
+detailed physical theories. Study of relational parametricity for
+dependent types, as done for example by Bernardy \emph{et al.}
+\cite{bernardy12proofs} and Atkey \emph{et al.}
+\cite{atkey14relational} may be useful here. Also, recent work on
 homotopy type theory and, in particular $\infty$-groupoid models of
@@
 
+\bibitem{atkey14relational} R.~Atkey, N.~Ghani, and P.~Johann. A
+  Relationally Parametric Model of Dependent Type Theory. Proceedings,
+  POPL, 2014.
+
 \bibitem{atkey13abstraction}
@@
 
-\bibitem{sussman01structure} G.~J.Sussman and J.~Wisdom. Structure and
+\bibitem{sussman01structure} G.~J.~Sussman and J.~Wisdom. Structure and
   Interpretation of Classical Mechanics. MIT Press, 2001
